It is useful to go through the puzzles because the puzzle free condition
is reached by untying the knots produced by the puzzles in advance.

What is the end of this inquiry?

[P1] Whether it belongs to one science or more than one to get
a theoretical grasp on causes.

[P2] Whether it only has to look at primary starting points of substance
or is it also concerned with the starting points on which everyone depends
when proving things.

[P3] Concerning substance, whether there is one science concerned with all of them
or more than one. If more than one, whether they are all of the same kind
or some of them sorts of wisdom and others something else.

[P4] Whether only perceptible substances exist or others beyond these such as the
Forms and intermediate objects of Mathematics.

[P5] Whether it is concerned only with substances or also the 
intrinsic coincidentals of substances.

[P6] Whether the starting-points and elements are the genera
or the components present in each thing into which it is divided.

[P7] Whether the genera are the first or ultimate ones said of indivisible things.
For example, whether animal or human is a starting-point
and is to a higher degree beyond what is particular.

[P8] Whether there is something, beyond the matter, which is an instrinsic cause,
whether or not this is separable, whether it is one or more than one in number,
and whether there is something beyond the compound.

[P9] Whether the starting-points are definite in number or sabija,
both those in the accounts and those in the underlying subject.

[P10] Whether the starting-points of what passes away and what does not pass away are the same or distint,
and whether none of them passes away or those of things that pass away do pass away. 

[P11] Whether the one and being are not, as the Pythagoreans and Plato used to say,
another thing, but rather the substance of the beings, or whether this is not so,
but the underlying subject is another thing, which Empedocles says is love, or fire, water ...

[P12] Whether the starting-points are universal or like particular things.

[P13] Whether the starting-points are causes potentially or actively or any other way that regards movement.

[P14] Whether or not numbers, lines, figures, and points are substances of some sort or not,
and if they are substances, are they separate from the perceptible ones or components present in them.
