Prolegomena to any future metaphysics
that will be able to come forward as scienceContents a
Preface
Preamble on the Distinguishing Feature of All Metaphysical
Cognition
On the sources of metaphysics
On the type of cognition, that alone can be
called metaphysical
a. On the distinction between synthetic and analytic
judgments in general
b. The common principle of all analytic judgments is
the principle of contradiction
c. Synthetic judgments require a principle other than
the principle of contradiction
Note on the general division of judgments into analytic
and synthetic
General Question of the Prolegomena:
Is metaphysics possible at all?
General Question:
How is cognition from pure reason possible?
Main Transcendental Question, First Part:
How is pure mathematics possible?
Note I
Note II
Note III
Main Transcendental Question, Second Part:
How is pure natural science possible?
Tables: Logical table of judgments
Transcendental table of concepts of
the understanding
Pure physiological table of universal principles
of natural science
How is nature itself possible?
Appendix to pure natural science: On the system
of categories
a
53
61
61
62
62
62
63
67
69
72
77
82
83
85
89
96
97
97
110
114
This table of contents has been constructed from the section titles. The original editions
did not contain a table of contents.
51Prolegomena to any future metaphysics
Main Transcendental Question, Third Part:
How is metaphysics in general possible?
Preliminary remark on the Dialectic of Pure Reason
I. Psychological ideas
II. Cosmological ideas
III. Theological ideas
General note to the transcendental ideas
Conclusion: On determining the boundary of pure reason
Solution to the General Question of the Prolegomena:
How is metaphysics possible as science?
Appendix, On what Can Be Done in Order to Make
Metaphysics as Science Actual
Specimen of a judgment about the Critique which precedes
the investigation
Proposal for an investigation of the Critique, after which
the judgment can follow
52
119
124
125
129
138
138
140
154
159
160
167Preface a
These prolegomena are not for the use of apprentices, but of future
teachers, and indeed are not to help them to organize the presentation
of an already existing science, but to discover this science itself for the
ﬁrst time.
There are scholars for whom the history of philosophy (ancient as
well as modern) is itself their philosophy; the present prolegomena have
not been written for them. They must wait until those who endeavor to
draw from the wellsprings of reason itself have ﬁnished their business,
and then it will be their turn to bring news of these events to the world.
Otherwise, in their opinion nothing can be said that has not already
been said before; and in fact this opinion can stand for all time as an
infallible prediction, for since the human understanding has wandered
over countless subjects in various ways through many centuries, it can
hardly fail that for anything new something old should be found that has
some similarity to it.
My intention is to convince all of those who ﬁnd it worthwhile to
occupy themselves with metaphysics that it is unavoidably necessary to
suspend their work for the present, to consider all that has happened until
now as if it had not happened, and before all else to pose the question:
“whether such a thing as metaphysics is even possible at all.”
If metaphysics is a science, why is it that it cannot, like other sciences,
attain universal and lasting acclaim? If it is not, how does it happen
that, under the pretense of a science it incessantly shows off, and strings
along the human understanding with hopes that never dim but are never
fulﬁlled? Whether, therefore, we demonstrate our knowledge or our
ignorance, for once we must arrive at something certain concerning the
nature of this self-proclaimed science; for things cannot possibly remain
on their present footing. It seems almost laughable that, while every other
science makes continuous progress, metaphysics, which desires to be
wisdom itself, and which everyone consults as an oracle, perpetually turns
round on the same spot without coming a step further. Furthermore,
it has lost a great many of its adherents, and one does not ﬁnd that
those who feel strong enough to shine in other sciences wish to risk
their reputations in this one, where anyone, usually ignorant in all other
things, lays claim to a decisive opinion, since in this region there are
a
Section heading supplied, with Vorländer, by analogy with Vorrede in A/B.
53
4: 255
4: 256Prolegomena to any future metaphysics
4: 257
in fact still no reliable weights and measures with which to distinguish
profundity from shallow babble.
It is, after all, not completely unheard of, after long cultivation of a
science, that in considering with wonder how much progress has been
made someone should ﬁnally allow the question to arise: whether and
how such a science is possible at all. For human reason is so keen on
building that more than once it has previously erected a tower, but has
afterwards torn it down again in order to see how well constituted its
foundation may have been. It is never too late to grow reasonable and
wise; but if the insight comes late, it is always harder to bring it into
play.
To ask whether a science might in fact be possible assumes a doubt
about its actuality. a Such a doubt, though, offends everyone whose entire
belongings may perhaps consist in this supposed jewel; hence he who
allows this doubt to develop had better prepare for opposition from all
sides. Some, with their metaphysical compendia in hand, will look down
on him with scorn, in proud consciousness of their ancient, and hence
ostensibly legitimate, possession; others, who nowhere see anything that
is not similar to something they have seen somewhere else before, will
not understand him; and for a time everything will remain as if nothing at
all had happened that might yield fear or hope of an impending change.
Nevertheless I venture to predict that the reader of these prolegom-
ena who thinks for himself will not only come to doubt his previous
science, but subsequently will be fully convinced that there can be no
such science unless the requirements expressed here, on which its pos-
sibility rests, are met, and, as this has never yet been done, that there is
as yet no metaphysics at all. Since, however, the demand for it can never
be exhausted, ∗ because the interest of human reason in general is much
too intimately interwoven with it, the reader will admit that a complete
reform or rather a rebirth of metaphysics, according to a plan completely
unknown before now, is inevitably approaching, however much it may
be resisted in the meantime. 1
Since the Essays of Locke 2 and Leibniz, 3 or rather since the rise of
metaphysics as far as the history of it reaches, no event has occurred that
could have been more decisive with respect to the fate of this science
than the attack made upon it by David Hume. 4 He brought no light to
this kind of knowledge, c but he certainly struck a spark from which a
∗ Rusticus exspectat, dum deﬂuat amnis, at ille
Labitur et labetur in omne volubilis aevum. Horace. b
a Wirklichkeit
“A rustic waits for the river to ﬂow away, but it ﬂows on, and will so ﬂow for all eternity.”
Horace, Epistles, I.ii.42–3.
Erkenntniss
b
c
54Prolegomena to any future metaphysics
light could well have been kindled, if it had hit some welcoming tinder
whose glow had then been carefully kept going and made to grow.
Hume started mainly from a single but important concept in meta-
physics, namely, that of the connection a of cause and effect (and of course
also its derivative concepts, of force and action, etc.), and called upon
reason, which pretends to have generated this concept in her womb, to
give him an account of by what right she thinks: that something could
be so constituted that, if it is posited, something else necessarily must
thereby be posited as well; for that is what the concept of cause says.
He undisputably proved that it is wholly impossible for reason to think
such a connection b a priori and from concepts, because this connec-
tion contains necessity; and it is simply not to be seen how it could be,
that because something is, something else necessarily must also be, and
therefore how the concept of such a connection c could be introduced a
priori. From this he concluded that reason completely and fully deceives
herself with this concept, falsely taking it for her own child, when it is
really nothing but a bastard of the imagination, which, impregnated by
experience, and having brought certain representations under the law of
association, passes off the resulting subjective necessity (i.e., habit) for
an objective necessity (from insight). 5 From which he concluded that
reason has no power at all to think such connections, not even merely
in general, because its concepts would then be bare ﬁctions, and all of
its cognitions allegedly established a priori would be nothing but falsely
marked ordinary experiences; which is as much as to say that there is no
metaphysics at all, and cannot be any. ∗
As premature and erroneous as his conclusion was, nevertheless it was
at least founded on inquiry, and this inquiry was of sufﬁcient value, that
the best minds of his time might have come together to solve (more
happily if possible) the problem in the sense in which he presented it,
from which a complete reform of the science must soon have arisen.
∗ All the same, Hume named this destructive philosophy itself metaphysics and
placed great value on it. “Metaphysics and morals,” he said (Essays, 4th pt.,
p. 214, German translation), “are the most important branches of science;
mathematics and natural science are not worth half so much.” 6 The acute man
was, however, looking only to the negative beneﬁt that curbing the excessive
claims of speculative reason would have, in completely abolishing so many
endless and continual conﬂicts that perplex the human species; he meanwhile
lost sight of the positive harm that results if reason is deprived of the most
important vistas, from which alone it can stake out for the will the highest goal
of all the will’s endeavors.
a Verknüpfung
Verbindung
Verknüpfung
b
c
55
4: 258Prolegomena to any future metaphysics
4: 259
But fate, ever ill-disposed toward metaphysics, would have it that
Hume was a understood by no one. One cannot, without feeling a
certain pain, behold how utterly and completely his opponents, Reid,
Oswald, Beattie, and ﬁnally Priestley, 7 missed the point of his problem, and
misjudged his hints for improvement – constantly taking for granted just
what he doubted, and, conversely, proving with vehemence and, more
often than not, with great insolence exactly what it had never entered
his mind to doubt – so that everything remained in its old condition, as
if nothing had happened. The question was not, whether the concept of
cause is right, useful, and, with respect to all cognition of nature, indis-
pensable, for this Hume had never put in doubt; it was rather whether
it is thought through reason a priori, and in this way has an inner truth
independent of all experience, and therefore also a much more widely
extended use which is not limited merely to objects of experience: re-
garding this Hume awaited enlightenment. b The discussion was only
about the origin of this concept, not about its indispensability in use; if
the former were only discovered, the conditions of its use and the sphere
in which it can be valid would already be given.
In order to do justice to the problem, however, the opponents of this
celebrated man would have had to penetrate very deeply into the nature
of reason so far as it is occupied solely with pure thought, something
that did not suit them. They therefore found a more expedient means to
be obstinate without any insight, namely, the appeal to ordinary common
sense. c It is in fact a great gift from heaven to possess right (or, as it
has recently been called, plain) c common sense. But it must be proven
through deeds, by the considered and reasonable things one thinks and
says, and not by appealing to it as an oracle when one knows of nothing
clever to advance in one’s defense. To appeal to ordinary common sense
when insight and science run short, and not before, is one of the subtle
discoveries of recent times, whereby the dullest windbag can conﬁdently
take on the most profound thinker and hold his own with him. So long
as a small residue of insight remains, however, one would do well to
avoid resorting to this emergency help. And seen in the light of day,
this appeal is nothing other than a call to the judgment of the multitude;
applause at which the philosopher blushes, but at which the popular wag
becomes triumphant and deﬁant. I should think, however, that Hume
could lay just as much claim to sound common sense d as Beattie, and on
top of this to something that the latter certainly did not possess, namely,
a critical reason, which keeps ordinary common sense in check, so that
a
b
c
d
Reading wurde for würde, with Ak.
Eröffnung
gemeinen Menschenverstand
gesunden Verstand
56Prolegomena to any future metaphysics
it doesn’t lose itself in speculations, or, if these are the sole topic of
discussion, doesn’t want to decide anything, since it doesn’t understand
the justiﬁcation for its own principles; for only so will it remain sound
common sense. Hammer and chisel are perfectly ﬁne for working raw
lumber, but for copperplate one must use an etching needle. 9 Likewise,
sound common sense and speculative understanding are both useful,
but each in its own way; the one, when it is a matter of judgments that
ﬁnd their immediate application in experience, the other, however, when
judgments are to be made in a universal mode, out of mere concepts, as
in metaphysics, where what calls itself (but often per antiphrasin) a sound
common sense has no judgment whatsoever.
I freely admit that the remembrance of David Hume was the very
thing that many years ago ﬁrst interrupted my dogmatic slumber 10 and
gave a completely different direction to my researches in the ﬁeld of
speculative philosophy. I was very far from listening to him with respect
to his conclusions, which arose solely because he did not completely set
out his problem but only touched on a part of it, which, without the
whole being taken into account, can provide no enlightenment. b If we
begin from a well-grounded though undeveloped thought that another
bequeaths us, then we can well hope, by continued reﬂection, to take it
further than could the sagacious man whom one has to thank for the ﬁrst
spark of this light.
So I tried ﬁrst whether Hume’s objection might not be presented in a
general manner, and I soon found that the concept of the connection of
cause and effect is far from being the only concept through which the
understanding thinks connections of things a priori; rather, metaphysics
consists wholly of such concepts. I sought to ascertain their number, and
once I had successfully attained this in the way I wished, namely from
a single principle, I proceeded to the deduction of these concepts, from
which I henceforth became assured that they were not, as Hume had
feared, derived from experience, but had arisen from the pure under-
standing. This deduction, which appeared impossible to my sagacious
predecessor, and which had never even occurred to anyone but him, even
though everyone conﬁdently made use of these concepts without asking
what their objective validity is based on – this deduction, I say, was the
most difﬁcult thing that could ever be undertaken on behalf of meta-
physics; and the worst thing about it is that metaphysics, as much of it
as might be present anywhere at all, could not give me the slightest help
with this, because this very deduction must ﬁrst settle the possibility of
a metaphysics. As I had now succeeded in the solution of the Humean
problem not only in a single case but with respect to the entire faculty of
a
b
“by way of expression through the opposite”
Auskunft
57
4: 260Prolegomena to any future metaphysics
4: 261
4: 262
pure reason, I could therefore take sure, if still always slow, steps toward
ﬁnally determining, completely and according to universal principles,
the entire extent of pure reason with regard to its boundaries as well as
its content, which was indeed the very thing that metaphysics requires
in order to build its system according to a sure plan.
But I fear that the elaboration of the Humean problem in its greatest
possible ampliﬁcation (namely, the Critique of Pure Reason) may well fare
just as the problem itself fared when it was ﬁrst posed. It will be judged in-
correctly, because it is not understood; it will not be understood, because
people will be inclined just to skim through the book, but not to think
through it; and they will not want to expend this effort on it, because
the work is dry, because it is obscure, because it opposes all familiar con-
cepts and is long-winded as well. Now I admit that I do not expect to hear
complaints from a philosopher regarding lack of popularity, entertain-
ment, and ease, when the matter concerns the existence of highly prized
knowledge that is indispensable to humanity, knowledge that cannot be
constituted except according to the strictest rules of scholarly exactitude,
and to which popularity may indeed come with time but can never be
there at the start. But with regard to a certain obscurity – arising in part
from the expansiveness of the plan, which makes it difﬁcult to survey the
main points upon which the investigation depends – in this respect the
complaint is just; and I will redress it through the present Prolegomena. a
The previous work, which presents the faculty of pure reason in its
entire extent and boundaries, thereby always remains the foundation to
which the Prolegomena refer only as preparatory exercises; for this Critique
must stand forth as science, systematic and complete to its smallest parts,
before one can think of permitting metaphysics to come forward, or even
of forming only a distant hope for metaphysics.
We have long been accustomed to seeing old, threadbare cognitions
newly trimmed by being taken from their previous connections and
ﬁtted out by someone in a systematic garb of his own preferred cut,
but under new titles; and most readers will beforehand expect nothing
else from the Critique. Yet these Prolegomena will bring them to un-
derstand that there exists a completely new science, of which no one
had previously formed so much as the thought, of which even the bare
idea was unknown, and for which nothing from all that has been pro-
vided before now could be used except the hint that Hume’s doubts
had been able to give; Hume also foresaw nothing of any such possi-
ble formal science, but deposited his ship on the beach (of skepticism)
for safekeeping, 11 where it could then lie and rot, whereas it is important
to me to give it a pilot, who, provided with complete sea-charts and a
compass, might safely navigate the ship wherever seems good to him,
a
Emphasis in original.
58Prolegomena to any future metaphysics
following sound principles of the helmsman’s art drawn from a knowl-
edge of the globe.
To approach a new science – one that is entirely isolated and is the
only one of its kind – with the prejudice that it can be judged by means
of one’s putative cognitions already otherwise obtained, even though
it is precisely the reality of those that must ﬁrst be completely called
into question, results only in believing that one sees everywhere some-
thing that was already otherwise known, because the expressions per-
haps sound similar; except that everything must seem to be extremely
deformed, contradictory, and nonsensical, because one does not thereby
make the author’s thoughts fundamental, but always simply one’s own,
made natural through long habit. Yet the copiousness of the work, in-
sofar as it is rooted in the science itself and not in the presentation, and
the inevitable dryness and scholastic exactitude that result, are qualities
that indeed may be extremely advantageous to the subject matter itself,
but must of course be detrimental to the book itself.
It is not given to everyone to write so subtly and yet also so alluringly as
David Hume, or so profoundly and at the same time so elegantly as Moses
Mendelssohn; 12 but I could well have given my presentation popularity
(as I ﬂatter myself) if all I had wanted to do was to sketch a plan and to
commend its execution to others, and had I not taken to heart the well-
being of the science that kept me occupied for so long; for after all it
requires great perseverance and also indeed not a little self-denial to set
aside the enticement of an earlier, favorable reception for the expectation
of an admittedly later, but lasting approval.
To make plans is most often a presumptuous, boastful mental preoccu-
pation, through which one presents the appearance of creative genius,
in that one requires what one cannot provide oneself, censures what one
cannot do better, and proposes what one does not know how to attain
oneself – though merely for a sound plan for a general critique of rea-
son, somewhat more than might be expected would already have been
required if it were not, as is usual, to be merely a recitation of idle wishes.
But pure reason is such an isolated domain, within itself so thoroughly
connected, that no part of it can be encroached upon without disturbing
all the rest, nor adjusted without having previously determined for each
part its place and its inﬂuence on the others; for, since there is nothing
outside of it that could correct our judgment within it, the validity and
use of each part depends on the relation in which it a stands to the others
within reason itself, and, as with the structure of an organized body, the
purpose of any member can be derived only from the complete concept
of the whole. That is why it can be said of such a critique, that it is never
trustworthy unless it is entirely complete down to the least elements of
a
Reading er for es, with Ak.
59
4: 263Prolegomena to any future metaphysics
4: 264
pure reason, and that in the domain of this faculty one must determine
and settle either all or nothing.
But although a mere plan that might precede the Critique of Pure
Reason would be unintelligible, undependable, and useless, it is by con-
trast all the more useful if it comes after. For one will thereby be put in
the position to survey the whole, to test one by one the main points at
issue in this science, and to arrange many things in the exposition better
than could be done in the ﬁrst execution of the work.
Here then is such a plan subsequent to the completed work, which
now can be laid out according to the analytic method, a whereas the work
itself absolutely had to be composed according to the synthetic method, b
so that the science might present all of its articulations, as the structural
organization of a quite peculiar faculty of cognition, in their natural
connection. 13 Whosoever ﬁnds this plan itself, which I send ahead as
prolegomena for any future metaphysics, still obscure, may consider
that it simply is not necessary for everyone to study metaphysics, that
there are some talents that proceed perfectly well in fundamental and
even deep sciences that are closer to intuition, but that will not succeed
in the investigation of purely abstract concepts, and that in such a case
one should apply one’s mental gifts to another object; that, however,
whosoever undertakes to judge or indeed to construct a metaphysics,
must thoroughly satisfy the challenge made here, whether it happens that
they accept my solution, or fundamentally reject it and replace it with
another – for they cannot dismiss it; and ﬁnally, that the much decried
obscurity (a familiar cloaking for one’s own indolence or dimwittedness)
has its use as well, since everybody, who with respect to all other sciences
observes a wary silence, speaks masterfully, and boldly passes judgment
in questions of metaphysics, because here to be sure their ignorance does
not stand out clearly in relation to the science of others, but in relation
to genuine critical principles, which therefore can be praised:
Ignavum, fucos, pecus a praesepibus arcent. Virg. c
a
b
c
analytischer Methode
synthetischer Lehrart
“They protect the hives from the drones, an idle bunch.” Virgil, Georgica, IV.168.
60Preamble
on the
Distinguishing Feature of All
Metaphysical Cognition
4: 265
§ 1
On the sources of metaphysics
If one wishes to present a body of cognition as science, a then one must ﬁrst
be able to determine precisely the differentia it has in common with no
other science, and which is therefore its distinguishing feature; otherwise
the boundaries of all the sciences run together, and none of them can be
dealt with thoroughly according to its own nature.
Whether this distinguishing feature consists in a difference of the
object or the source of cognition, or even of the type of cognition, or several
if not all of these things together, the idea of the possible science and its
territory depends ﬁrst of all upon it.
First, concerning the sources of metaphysical cognition, it already
lies in the concept of metaphysics that they cannot be empirical. The
principles b of such cognition (which include not only its fundamental
propositions, c but also its fundamental concepts) must therefore never
be taken from experience; for the cognition is supposed to be not phys-
ical but metaphysical, i.e., lying beyond experience. Therefore it will
be based upon neither outer experience, which constitutes the source
of physics proper, nor inner, which provides the foundation of empirical
psychology. It is therefore cognition a priori, or from pure understanding
and pure reason.
In this, however, there would be nothing to differentiate it from
pure mathematics; it must therefore be denominated pure philosophical
cognition; but concerning the meaning of this expression I refer to the
Critique of Pure Reason, pp. 712 f., 14 where the distinction between these
two types of use of reason has been presented clearly and sufﬁciently. –
So much on the sources of metaphysical cognition. 15
a
b
c
eine Erkenntniss als Wissenschaft
Principien
Grundsätze
61
4: 266Prolegomena to any future metaphysics
§ 2
On the type of cognition,
that alone can be called metaphysical
(a) On the distinction between synthetic and analytic
judgments in general
4: 267
Metaphysical cognition must contain nothing but judgments a priori,
as required by the distinguishing feature of its sources. But judgments
may have any origin whatsoever, or be constituted in whatever manner
according to their logical form, and yet there is nonetheless a distinction
between them according to their content, by dint of which they are
either merely explicative and add nothing to the content of the cognition,
or ampliative and augment the given cognition; the ﬁrst may be called
analytic judgments, the second synthetic.
Analytic judgments say nothing in the predicate except what was ac-
tually thought already in the concept of the subject, though not so clearly
nor with the same consciousness. If I say: All bodies are extended, then
I have not in the least ampliﬁed my concept of body, but have merely
resolved it, since extension, although not explicitly said of the former
concept prior to the judgment, nevertheless was actually thought of it;
the judgment is therefore analytic. By contrast, the proposition: Some
bodies are heavy, contains something in the predicate that is not actu-
ally thought in the general concept of body; it therefore augments my
cognition, since it adds something to my concept, and must therefore be
called a synthetic judgment. 16
(b) The common principle a of all analytic judgments
is the principle of contradiction
All analytic judgments rest entirely on the principle of contradiction
and are by their nature a priori cognitions, whether the concepts that
serve for their material be empirical or not. For since the predicate
of an afﬁrmative analytic judgment is already thought beforehand in
the concept of the subject, it cannot be denied of that subject without
contradiction; exactly so is its opposite necessarily denied of the subject
in an analytic, but negative, judgment, and indeed also according to the
principle of contradiction. So it stands with the propositions: Every body
is extended, and: No body is unextended (simple).
For that reason all analytic propositions are still a priori judgments
even if their concepts are empirical, as in: Gold is a yellow metal; for
in order to know this, I need no further experience outside my concept
of gold, which includes that this body is yellow and a metal; for this
a
Princip
62Prolegomena to any future metaphysics
constitutes my very concept, and I did not have to do anything except
analyze it, without looking beyond it to something else. 17
(c) Synthetic judgments require a principle other than
the principle of contradiction
There are synthetic judgments a posteriori whose origin is empirical; but
there are also synthetic judgments that are a priori certain and that arise
from pure understanding and reason. Both however agree in this, that
they can by no means arise solely from the principle a of analysis, namely
the principle of contradiction; they demand yet a completely different
principle, b though they always must be derived from some fundamen-
tal proposition, c whichever it may be, in accordance with the principle of
contradiction; for nothing can run counter to this principle, even though
everything cannot be derived from it. I shall ﬁrst classify the synthetic
judgments.
1. Judgments of experience are always synthetic. For it would be ab-
surd to base an analytic judgment on experience, since I do not at all
need to go beyond my concept in order to formulate the judgment and
therefore have no need for any testimony from experience. That a body is
extended, is a proposition that stands certain a priori, and not a judgment
of experience. For before I go to experience, I have all the conditions
for my judgment already in the concept, from which I merely extract
the predicate in accordance with the principle of contradiction, and by
this means can simultaneously become conscious of the necessity of the
judgment, which experience could never teach me.
2. Mathematical judgments are one and all synthetic. This proposition
appears to have completely escaped the observations of analysts of human
reason up to the present, and indeed to be directly opposed to all of their
conjectures, although it is incontrovertibly certain and very important
in its consequences. Because they found that the inferences of the math-
ematicians all proceed in accordance with the principle of contradiction
(which, by nature, is required of any apodictic certainty), they were per-
suaded that even the fundamental propositions were known through the
principle of contradiction, in which they were very mistaken; for a syn-
thetic proposition can of course be discerned in accordance with the
principle of contradiction, but only insofar as another synthetic proposi-
tions is presupposed from which the ﬁrst can be deduced, never however
in itself.
a
b
c
Grundsatze
Princip
Grundsatze
63
4: 268Prolegomena to any future metaphysics
4: 269
First of all it must be observed: that properly mathematical proposi-
tions are always a priori and not empirical judgments, because they carry
necessity with them, which cannot be taken from experience. But if this
will not be granted me, very well, I will restrict my proposition to pure
mathematics, the concept of which already conveys that it contains not
empirical but only pure cognition a priori.
One might well at ﬁrst think: that the proposition 7 + 5 = 12 is a
purely analytical proposition that follows from the concept of a sum
of seven and ﬁve according to the principle of contradiction. However,
upon closer inspection, one ﬁnds that the concept of the sum of 7 and
5 contains nothing further than the uniﬁcation of the two numbers into
one, through which by no means is thought what this single number may
be that combines the two. The concept of twelve is in no way already
thought because I merely think to myself this uniﬁcation of seven and
ﬁve, and I may analyze my concept of such a possible sum for as long as
may be, still I will not meet with twelve therein. One must go beyond
these concepts, in making use of the intuition that corresponds to one of
the two, such as one’s ﬁve ﬁngers, or (like Segner in his arithmetic) 18 ﬁve
points, and in that manner adding the units of the ﬁve given in intuition
step by step to the concept of seven. One therefore truly ampliﬁes one’s
concept through this proposition 7 + 5 = 12 and adds to the ﬁrst concept
a new one that was not thought in it; that is, an arithmetical proposition
is always synthetic, which can be seen all the more plainly in the case
of somewhat larger numbers, for it is then clearly evident that, though
we may turn and twist our concept as we like, we could never ﬁnd the
sum through the mere analysis of our concepts, without making use of
intuition.
Nor is any fundamental proposition of pure geometry analytic. That
the straight line between two points is the shortest is a synthetic propo-
sition. For my concept of the straight contains nothing of magnitude, a
but only a quality. The concept of the shortest is therefore wholly an
addition and cannot be extracted by any analysis from the concept of the
straight line. Intuition must therefore be made use of here, by means of
which alone the synthesis is possible.
Some other fundamental propositions that geometers presuppose
are indeed actually analytic and rest on the principle of contradic-
tion; however, they serve only, like identical propositions, as links in
the chain of method and not as b principles: e.g., a = a, the whole is
equal to itself, or (a + b) > a, i.e., the whole is greater than its part.
And indeed even these, although they are valid from concepts alone,
a
b
Grösse
Reading als for aus, with Ak (and B 17).
64Prolegomena to any future metaphysics
are admitted into mathematics only because they can be exhibited in
intuition. a
It is merely ambiguity of expression which makes us commonly believe
here that the predicate of such apodictic judgments already lies in our
concept and that the judgment is therefore analytic. Namely, we are re-
quired b to add in thought a particular predicate to a given concept, and
this necessity is already attached to the concepts. But the question is not,
what we are required to add in thought to a given concept, but what we
actually think in it, c even if only obscurely, and then it becomes evident
that the predicate attaches to such concepts indeed necessarily, though
not immediately, but rather through an intuition that has to be added. d
The essential feature of pure mathematical cognition, differentiating
it from all other a priori cognition, is that it must throughout proceed not
from concepts, but always and only through the construction of concepts
(Critique, p. 713). 19 Because pure mathematical cognition, in its propo-
sitions, must therefore go beyond the concept to that which is contained
in the intuition corresponding to it, its propositions can and must never
arise through the analysis of concepts, i.e., analytically, and so are one
and all synthetic. 20
I cannot, however, refrain from noting the damage that neglect of
this otherwise seemingly insigniﬁcant and unimportant observation has
brought upon philosophy. Hume, when he felt the call, worthy of a
philosopher, to cast his gaze over the entire ﬁeld of pure a priori cognition,
in which the human understanding claims such vast holdings, inadver-
tently lopped off a whole (and indeed the most considerable) province
of the same, namely pure mathematics, by imagining that the nature and
so to speak the legal constitution of this province rested on completely
different principles, namely solely on the principle of contradiction; and
although he had by no means made a classiﬁcation of propositions so
formally and generally, or with such nomenclature, as I have here, it
was nonetheless just as if he had said: Pure mathematics contains only
analytic propositions, but metaphysics contains synthetic propositions a
priori. Now he erred severely in this, and this error had decisively dam-
aging consequences for his entire conception. For had he not done this,
he would have expanded his question about the origin of our synthetic
judgments far beyond his metaphysical concept of causality and extended
a
b
c
d
Paragraph break added to reﬂect continuity of the new paragraph with the three para-
graphs prior to the preceding two sentences.
sollen
Reading ihm for ihnen, with Ak (and B 17).
The following ﬁve paragraphs are taken from §4 in accordance with the Vaihinger-Sitzler
galley-switching thesis (see Translator’s Introduction).
65
4: 272Prolegomena to any future metaphysics
4: 273
4: 274
it also to the possibility of a priori mathematics; for he would have had
to accept mathematics as synthetic as well. But then he would by no
means have been able to found his metaphysical propositions on mere
experience, for otherwise he would have had to subject the axioms of
pure mathematics to experience as well, which he was much too reason-
able to do. 21 The good company in which metaphysics would then have
come to be situated would have secured it against the danger of scornful
mistreatment; for the blows that were intended for the latter would have
had to strike the former as well, which was not his intention, and could
not have been; and so the acute man would have been drawn into re-
ﬂections which must have been similar to those with which we are now
occupied, but which would have gained inﬁnitely from his inimitably
ﬁne presentation. 22
3. a Properly metaphysical judgments are one and all synthetic. Judg-
ments belonging to metaphysics must be distinguished from properly meta-
physical judgments. Very many among the former are analytic, but they
merely provide the means to metaphysical judgments, toward which the
aim of the science is completely directed, and which are always synthetic.
For if concepts belong to metaphysics, e.g., that of substance, then nec-
essarily the judgments arising from their mere analysis belong to meta-
physics as well, e.g., substance is that which exists only as subject, etc.,
and through several such analytic judgments we try to approach the def-
inition of those concepts. Since, however, the analysis of a pure concept
of the understanding (such as metaphysics contains) does not proceed in
a different manner from the analysis of any other, even empirical, con-
cept which does not belong to metaphysics (e.g., air is an elastic ﬂuid, the
elasticity of which is not lost with any known degree of cold), therefore
the concept may indeed be properly metaphysical, but not the analytic
judgment; for this science possesses something special and proper to it in
the generation of its a priori cognitions, which generation must therefore
be distinguished from what this science has in common with all other
cognitions of the understanding; thus, e.g., the proposition: All that is
substance in things persists, is a synthetic and properly metaphysical
proposition.
If one has previously assembled, according to ﬁxed principles, the a
priori concepts that constitute the material of metaphysics and its tools,
then the analysis of these concepts is of great value; it can even be pre-
sented apart from all the synthetic propositions that constitute meta-
physics itself, as a special part (as it were a philosophia deﬁnitiva) 23 contain-
ing nothing but analytic propositions belonging to metaphysics. For in
fact such analyses do not have much use anywhere except in metaphysics,
a
The numeral three is added in accordance with the Vaihinger–Sitzler thesis.
66Prolegomena to any future metaphysics
i.e., with a view toward the synthetic propositions that are to be generated
from such previously analyzed concepts.
The conclusion of this section is therefore: that metaphysics properly
has to do with synthetic propositions a priori, and these alone constitute
its aim, for which it indeed requires many analyses of its concepts (there-
fore many analytic judgments), in which analyses, though, the procedure
is no different from that in any other type of cognition when one seeks
simply to make its concepts clear through analysis. But the generation
of cognition a priori in accordance with both intuition and concepts,
ultimately of synthetic propositions a priori as well, and speciﬁcally in
philosophical cognition, forms a the essential content of metaphysics.
§ 3
Note on the general division of judgments
into analytic and synthetic
This division is indispensable with regard to the critique of human un-
derstanding, and therefore deserves to be classical in it; other than that I
don’t know that it has much utility anywhere else. And in this I ﬁnd the
reason why dogmatic philosophers (who always sought the sources of
metaphysical judgments only in metaphysics itself, and not outside it in
the pure laws of reason in general) neglected this division, which appears
to come forward of itself, and, like the famous Wolf, or the acute Baum-
garten following in his footsteps, could try to ﬁnd the proof of the prin-
ciple of sufﬁcient reason, which obviously is synthetic, in the principle of
contradiction. 24 By contrast I ﬁnd a hint of this division already in Locke’s
essays on human understanding. For in Book IV, Chapter III, §9 f., after
he had already discussed the various connections of representations b in
judgments and the sources of the connections, of which he located the
one in identity or contradiction (analytic judgments) but the other in the
existence of representations in a subject (synthetic judgments), he then
acknowledges in §10 that our cognition (a priori) of these last is very con-
stricted and almost nothing at all. 25 But there is so little that is deﬁnite
and reduced to rules in what he says about this type of cognition, that it
is no wonder if no one, and in particular not even Hume, was prompted
by it to contemplate propositions of this type. 26 For such general yet
nonetheless deﬁnite principles are not easily learned from others who
have only had them ﬂoating obscurely before them. One must ﬁrst have
a
b
Reading macht for machen, with Vorländer.
Vorstellungen, translated as “representations” here as elsewhere, even though correspond-
ing to Locke’s word “ideas.” German translators of philosophy at this time tended to
avoid the loan word “Idee,” usually rendering the English “idea” as Begriff (on which, see
Poley’s translation of the Essay, his n. 6, on p. 8).
67
4: 270Prolegomena to any future metaphysics
come to them oneself through one’s own reﬂection, after which one also
ﬁnds them elsewhere, where one certainly would not have found them
before, because the authors did not even know themselves that their
own remarks were grounded on such an idea. Those who never think for
themselves in this way nevertheless possess the quick-sightedness to spy
everything, after it has been shown to them, in what has already been
said elsewhere, where no one at all could see it before.
68General Question of the Prolegomena
Is metaphysics possible at all?
4: 271
§ 4
If a metaphysics that could assert itself as science were actual, if one could
say: here is metaphysics, you need only to learn it, and it will convince
you of its truth irresistibly and immutably, then this question would be
unnecessary, and there would remain only that question which would
pertain more to a test of our acuteness than to a proof of the existence of
the subject matter itself, namely: how it is possible, and how reason should
set about attaining it. Now it has not gone so well for human reason in
this case. One can point to no single book, as for instance one presents
a Euclid, and say: this is metaphysics, here you will ﬁnd the highest aim
of this science, knowledge a of a supreme being and a future life, proven
from principles of pure reason. For one can indeed show us many propo-
sitions that are apodictically certain and have never been disputed; but
they are one and all analytic and pertain more to the materials and imple-
ments of metaphysics than to the expansion of knowledge, which after
all ought to be our real aim for it. (§2c) But although you present syn-
thetic propositions as well (e.g., the principle of sufﬁcient reason), which
you have never proven from bare reason and consequently a priori, as
was indeed your obligation, and which are gladly ceded to you all the
same: then if you want to use them toward your main goal, you still fall
into assertions so illicit and precarious that one metaphysics has always
contradicted the other, either in regard to the assertions themselves or
their proofs, and thereby metaphysics has itself destroyed its claim to
lasting approbation. The very attempts to bring such a science into exis-
tence were without doubt the original cause of the skepticism that arose
so early, 27 a mode of thinking in which reason moves against itself with
such violence that it never could have arisen except in complete despair
as regards satisfaction of reason’s most important aims. For long before
we began to question nature methodically, we questioned just our iso-
lated reason, which already was practiced to a certain extent through
common experience: for reason surely is present to us always, but laws
of nature must normally be sought out painstakingly; and so metaphysics
a
Erkenntniss
69
4: 272Prolegomena to any future metaphysics
was ﬂoating at the top like foam, though in such a way that as soon as
what had been drawn off had dissolved, more showed itself on the sur-
face, which some always gathered up eagerly, while others, instead of
seeking the cause of this phenomenon in the depths, thought themselves
wise in mocking the fruitless toilof the former. a
4: 274
4: 275
Weary therefore of dogmatism, which teaches us nothing, and also
of skepticism, which promises us absolutely nothing at all, not even
the tranquility of a permitted ignorance; summoned by the importance
of the knowledge b that we need, and made mistrustful, through long
experience, with respect to any knowledge that we believe we possess or
that offers itself to us under the title of pure reason, there remains left
for us but one critical question, the answer to which can regulate our
future conduct: Is metaphysics possible at all ? But this question must not
be answered by skeptical objections to particular assertions of an actual
metaphysics (for at present we still allow none to be valid), but out of
the still problematic concept of such a science.
In the Critique of Pure Reason c I worked on this question synthetically, d
namely by inquiring within pure reason itself, and seeking to determine
within this source both the elements and the laws of its pure use, accord-
ing to principles. This work is difﬁcult and requires a resolute reader to
think himself little by little into a system that takes no foundation as given
except reason itself, and that therefore tries to develop cognition out of its
original seeds without relying on any fact whatever. Prolegomena e should
by contrast be preparatory exercises; they ought more to indicate what
needs to be done in order to bring a science into existence if possible,
than to present the science itself. They must therefore rely on something
already known to be dependable, from which we can go forward with
conﬁdence and ascend to the sources, which are not yet known, and
whose discovery not only will explain what is known already, but will
also exhibit an area with many cognitions that all arise from these same
sources. The methodological procedure of prolegomena, and especially
of those that are to prepare for a future metaphysics, will therefore be
analytic.
Fortunately, it happens that, even though we cannot assume that meta-
physics as science is actual, we can conﬁdently say that some pure syn-
thetic cognition a priori is actual and given, namely, pure mathematics
a
b
c
d
e
Here followed the ﬁve paragraphs that have been placed in §2 (pp. 65–7), following the
Vaihinger–Sitzler thesis.
Erkenntniss
Emphasis in original.
Emphasis added.
Emphasis in original.
70Prolegomena to any future metaphysics
and pure natural science; for both contain propositions that are fully ac-
knowledged, some as apodictically certain through bare reason, some
from universal agreement with experience (though these are still rec-
ognized as independent of experience). We have therefore some at least
uncontested synthetic cognition a priori, and we do not need to ask whether
it is possible (for it is actual), but only: how it is possible, in order to be able
to derive, from the principle of the possibility of the given cognition, the
possibility of all other synthetic cognition a priori.
71Prolegomena
General Question
How is cognition from pure
reason possible?
§ 5
4: 276
We have seen above the vast difference between analytic and synthetic
judgments. The possibility of analytic propositions could be compre-
hended very easily; for it is founded solely upon the principle of con-
tradiction. The possibility of synthetic propositions a posteriori, i.e., of
such as are drawn from experience, also requires no special explana-
tion; for experience itself is nothing other than a continual conjoin-
ing (synthesis) of perceptions. There remain for us therefore only syn-
thetic propositions a priori, whose possibility must be sought or in-
vestigated, since it must rest on principles other than the principle of
contradiction.
Here, however, we do not need ﬁrst to seek the possibility of such
propositions, i.e., to ask whether they are possible. For there are plenty
of them actually given, and indeed with indisputable certainty, and since
the method we are now following is to be analytic, we will consequently
start from the position: that such synthetic but pure rational cognition is
actual; but we must nonetheless next investigate the ground of this possi-
bility, and ask: how this cognition is possible, so that we put ourselves in
a position to determine, from the principles of its possibility, the condi-
tions of its use and the extent and boundaries of the same. 28 Expressed
with scholastic precision, the exact problem on which everything hinges
is therefore:
How are synthetic propositions a priori possible?
For the sake of popularity I have expressed this problem somewhat dif-
ferently above, namely as a question about cognition from pure reason
which I could well have done on this occasion without disadvantage for
the desired insight; for, since we assuredly have to do here only with
metaphysics and its sources, it will, I hope, always be kept in mind, fol-
lowing the earlier reminders, that when we here speak of cognition from
72Prolegomena to any future metaphysics
pure reason, the discussion is never about analytic cognition, but only
synthetic. ∗
Whether metaphysics is to stand or fall, and hence its existence, now
depends entirely on the solving of this problem. Anyone may present his
contentions on the matter with ever so great a likelihood, piling conclu-
sion on conclusion to the point of suffocation; if he has not been able
beforehand to answer this question satisfactorily then I have the right
to say: it is all empty, baseless philosophy and false wisdom. You speak
through pure reason and pretend as it were to create a priori cognitions,
not only by analyzing given concepts, but by alleging new connections
that are not based on the principle of contradiction and that you nonethe-
less presume to understand completely independently of all experience;
now how do you come to this, and how will you justify such pretenses?
You cannot be allowed to call on the concurrence of general common
sense; b for that is a witness whose standing is based solely on public
rumor. Quodcunque ostendis mihi sic, incredulus odi. c Horat.
As indispensable as it is, however, to answer this question, at the same
time it is just as difﬁcult; and although the principal reason why the an-
swer has not long since been sought rests in the fact that it had occurred
to no one that such a thing could be asked, nonetheless a second reason
is that a satisfactory answer to this one question requires more assidu-
ous, deeper, and more painstaking reﬂection than the most prolix work
of metaphysics ever did, which promised its author immortality on its
ﬁrst appearance. Also, every perceptive reader, if he carefully ponders
what this problem demands, being frightened at ﬁrst by its difﬁculty, is
bound to consider it insoluble and, if such pure synthetic cognitions a
∗ When knowledge a moves forward little by little, it cannot be helped that certain
expressions which have already become classical, having been present from the
very infancy of science, subsequently should be found insufﬁcient and badly
suited, and that a certain newer and more apt usage should fall into danger of
being confused with the old one. The analytic method, insofar as it is opposed
to the synthetic, is something completely different from a collection of analytic
propositions; it signiﬁes only that one proceeds from that which is sought as if
it were given, and ascends to the conditions under which alone it is possible. In
this method one often uses nothing but synthetic propositions, as mathematical
analysis exempliﬁes, and it might better be called the regressive method to
distinguish it from the synthetic or progressive method. Again the name analytic
is also found as a principal division of logic, and there it is the logic of truth and
is opposed to dialectic, without actually looking to see whether the cognitions
belonging to that logic are analytic or synthetic.
a Erkenntniss
allgemeinen Menschenvernunft
“Whatsoever you show me thusly, unbelieving, I hate it.” Horace, Epistles, II.3.188.
b
c
73
4: 277Prolegomena to any future metaphysics
4: 278
priori were not actual, altogether impossible; which is what actually befell
David Hume, although he was far from conceiving the question in such
universality as it is here, and as it must be if the reply is to be decisive for
all metaphysics. For how is it possible, asked the acute man, that when
I am given one concept I can go beyond it and connect another one to
it that is not contained in it, and can indeed do so, as though the latter
necessarily belonged to the former? Only experience can provide us with
such connections (so he concluded from this difﬁculty, which he took
for an impossibility), and all of this supposed necessity – or, what is the
same – this cognition taken for a priori, is nothing but a long-standing
habit of ﬁnding something to be true and consequently of taking sub-
jective necessity to be objective.
If the reader complains about the toil and trouble that I shall give
him with the solution to this problem, he need only make the attempt
to solve it more easily himself. Perhaps he will then feel himself obliged
to the one who has taken on a task of such profound inquiry for him,
and will rather allow himself to express some amazement over the ease
with which the solution could still be given, considering the nature of
the matter; for indeed it cost years of toil to solve this problem in its full
universality a (as this word is understood by the mathematicians, namely,
as sufﬁcient for all cases), and also ultimately to be able to present it in
analytic form, as the reader will ﬁnd it here.
All metaphysicians are therefore solemnly and lawfully suspended
from their occupations until such a time as they shall have satisfactorily
answered the question: How are synthetic cognitions a priori possible? For in
this answer alone consists the credential which they must present if they
have something to advance to us in the name of pure reason; in default
of which, however, they can expect only that reasonable persons, who
have been deceived so often already, will reject their offerings without
any further investigation.
If, on the contrary, they want to put forth their occupation not as
science, but as an art of beneﬁcial persuasions accommodated to general
common sense, then they cannot justly be barred from this trade. They
will then use the modest language of reasonable belief, they will acknowl-
edge that it is not allowed them even once to guess, let alone to know, b
something about that which lies beyond the boundaries of all possible
experience, but only to assume something about it (not for speculative
use, for they must renounce that, but solely for practical use), as is pos-
sible and even indispensable for the guidance of the understanding and
will in life. Only thus will they be able to call themselves useful and wise
men, the more so, the more they renounce the name of metaphysicians;
a
b
Allgemeinheit
wissen
74Prolegomena to any future metaphysics
for metaphysicians want to be speculative philosophers, and since one
cannot aim for vapid probabilities when judgments a priori are at stake
(for what is alleged to be cognized a priori is thereby announced as neces-
sary), it cannot be permitted them to play with guesses, but rather their
assertions must be science or they are nothing at all.
It can be said that the whole of transcendental philosophy, which nec-
essarily precedes all of metaphysics, is itself nothing other than simply
the complete solution of the question presented here, but in systematic
order and detail, and that until now there has therefore been no tran-
scendental philosophy; for what goes under this name is really a part of
metaphysics, but this science is to settle the possibility of metaphysics
in the ﬁrst place, and therefore must precede all metaphysics. Hence
there need be no surprise because a science is needed that is utterly de-
prived of assistance from other sciences and hence is in itself completely
new, in order just to answer a single question adequately, when the so-
lution to it is conjoined with trouble and difﬁculty and even with some
obscurity.
In now setting to work on this solution – and indeed following the
analytic method, in which we presuppose that such cognitions from pure
reason are actual – we can appeal to only two sciences of theoretical knowl-
edge (which alone is being discussed here), namely, pure mathematics and
pure natural science; for only these can present objects to us in intuition,
and consequently, if they happen to contain an a priori cognition, can
show its truth or correspondence with the object in concreto, i.e., its ac-
tuality, from which one could then proceed along the analytic path to
the ground of its possibility. This greatly facilitates the work, in which
general considerations are not only applied to facts, but even start from
them, instead of, as in the synthetic procedure, having to be derived
wholly in abstracto from concepts.
But in order to ascend from these pure a priori cognitions (which are
not only actual but also well-founded) to a possible cognition that we
seek – namely, a metaphysics as science – we need to comprehend un-
der our main question that which gives rise to metaphysics and which
underlies its purely naturally given (though not above suspicion as re-
gards truth) cognition a priori (which cognition, when pursued without
any critical investigation of its possibility, is normally called metaphysics
already) – in a word, the natural disposition to such a science; and so the
main transcendental question, divided into four other questions, will be
answered step by step:
1.
2.
3.
4.
How is pure mathematics possible?
How is pure natural science possible?
How is metaphysics in general possible?
How is metaphysics as science possible?
75
4: 279
4: 280Prolegomena to any future metaphysics
It can be seen that even if the solution to these problems is intended
principally to present the essential content of the Critique, still it also
possesses something distinctive that is worthy of attention in its own
right, namely, the search for the sources of given sciences in reason itself,
in order to investigate and to measure out for reason, by way of the deed
itself, its power to cognize something a priori; whereby these sciences
themselves then beneﬁt, if not with respect to their content, nonetheless
as regards their proper practice, and, while bringing light to a higher
question regarding their common origin, they simultaneously provide
occasion for a better explanation of their own nature.
76Main Transcendental Question,
First Part
How is pure mathematics possible?
§ 6
Here now is a great and proven body of cognition, a which is already of
admirable extent and promises unlimited expansion in the future, which
carries with it thoroughly apodictic certainty (i.e., absolute necessity),
hence rests on no grounds of experience, and so is a pure product of rea-
son, but beyond this is thoroughly synthetic. “How is it possible then for
human reason to achieve such cognition wholly a priori?” Does not this
capacity, since it is not, and cannot be, based on experience, presuppose
some a priori basis for cognition, which lies deeply hidden, but which
might reveal itself through these its effects, if their ﬁrst beginnings were
but diligently tracked down?
§ 7
We ﬁnd, however, that all mathematical cognition has this distinguishing
feature, that it must present its concept beforehand in intuition and in-
deed a priori, consequently in an intuition that is not empirical but pure,
without which means it cannot take a single step; therefore its judg-
ments are always intuitive, b in the place of which philosophy can content
itself with discursive judgments from mere concepts, and can indeed exem-
plify its apodictic teachings through intuition c but can never derive them
from it. This observation with respect to the nature of mathematics al-
ready guides us toward the ﬁrst and highest condition of its possibility;
namely, it must be grounded in some pure intuition or other, in which it
can present, or, as one calls it, construct all of its concepts in concreto yet
a priori. ∗ If we could discover this pure intuition and its possibility, then
from there it could easily be explained how synthetic a priori propositions
are possible in pure mathematics, and consequently also how this science
∗ See Critique p. 713. 29
a eine grosse und bewährte Erkenntniss
intuitiv
Anschauung
b
c
77
4: 281Prolegomena to any future metaphysics
itself is possible; for just as empirical intuition makes it possible for us,
without difﬁculty, to amplify (synthetically in experience) the concept we
form of an object of intuition through new predicates that are presented
by intuition itself, so too will pure intuition do the same only with this
difference: that in the latter case the synthetic judgment will be a priori
certain and apodictic, but in the former only a posteriori and empirically
certain, because the former contains only what is met with in contingent
empirical intuition, while the latter contains what necessarily must be
met with in pure intuition, since it is, as intuition a priori, inseparably
bound with the concept before all experience or individual perception.
§ 8
4: 282
But with this step the difﬁculty seems to grow rather than to diminish.
For now the question runs: How is it possible to intuit something a priori? An
intuition is a representation of the sort which would depend immediately
on the presence of an object. It therefore seems impossible originally to
intuit a priori, since then the intuition would have to occur without an
object being present, either previously or now, to which it could refer,
and so it could not be an intuition. Concepts are indeed of the kind that
we can quite well form some of them for ourselves a priori (namely, those
that contain only the thinking of an object in general) without our being
in an immediate relation to an object, e.g., the concept of quantity, of
cause, etc.; but even these still require, in order to provide them with
signiﬁcation and sense, a certain use in concreto, i.e., application to some
intuition or other, by which an object for them is given to us. But how
can the intuition of an object precede the object itself?
§ 9
If our intuition had to be of the kind that represented things as they are
in themselves, then absolutely no intuition a priori would take place, but it
would always be empirical. For I can only know what may be contained in
the object in itself if the object is present and given to me. Of course, even
then it is incomprehensible how the intuition of a thing that is present
should allow me to cognize it the way it is in itself, since its properties
cannot migrate over into my power of representation; but even granting
such a possibility, the intuition still would not take place a priori, i.e.,
before the object were presented to me, for without that no basis for
the relation of my representation to the object can be conceived; so it
would have to be based on inspiration. There is therefore only one way
possible for my intuition to precede the actuality of the object and occur
as an a priori cognition, namely if it contains nothing else except the form of
sensibility, which in me as subject precedes all actual impressions through which
78Prolegomena to any future metaphysics
I am affected by objects. For I can know a priori that the objects of the
senses can be intuited only in accordance with this form of sensibility.
From this it follows: that propositions which relate merely to this form of
sensory intuition will be possible and valid for objects of the senses; also,
conversely, that intuitions which are possible a priori can never relate to
things other than objects of our senses.
§ 10
Therefore it is only by means of the form of sensory intuition that we
can intuit things a priori, though by this means we can cognize objects
only as they appear to us (to our senses), not as they may be in themselves;
and this supposition is utterly necessary, if synthetic propositions a priori
are to be granted as possible, or, in case they are actually encountered,
if their possibility is to be conceived and determined in advance.
Now space and time are the intuitions upon which pure mathemat-
ics bases all its cognitions and judgments, which come forward as at
once apodictic and necessary; for mathematics must ﬁrst exhibit all of
its concepts in intuition – and pure mathematics in pure intuition – i.e.,
it must ﬁrst construct them, failing which (since mathematics cannot
proceed analytically, namely, through the analysis of concepts, but only a
synthetically) it is impossible for it to advance a step, that is, as long as
it lacks pure intuition, in which alone the material b for synthetic judg-
ments a priori can be given. Geometry bases itself on the pure intuition
of space. Even arithmetic forms its concepts of numbers through succes-
sive addition of units in time, but above all pure mechanics can form its
concepts of motion only by means of the representation of time. 30 Both
representations are, however, merely intuitions; for, if one eliminates
from the empirical intuitions of bodies and their alterations (motion)
everything empirical, that is, that which belongs to sensation, then space
and time still remain, which are therefore pure intuitions that underlie
a priori the empirical intuitions, and for that reason can never themselves
be eliminated; but, by the very fact that they are pure intuitions a priori,
they prove that they are mere forms of our sensibility that must precede
all empirical intuition (i.e., the perception of actual objects), and in ac-
cordance with which objects can be cognized a priori, though of course
only as they appear to us.
§ 11
The problem of the present section is therefore solved. Pure mathemat-
ics, as synthetic cognition a priori, is possible only because it refers to
a
b
Adding nur, with Vorländer.
Stoff
79
4: 283Prolegomena to any future metaphysics
4: 284
no other objects than mere objects of the senses, the empirical intuition
of which is based on a pure and indeed a priori intuition (of space and
time), and can be so based because this pure intuition is nothing but the
mere form of sensibility, which precedes the actual appearance of ob-
jects, since it in fact ﬁrst makes this appearance possible. This faculty of
intuiting a priori does not, however, concern the matter of appearance –
i.e., that which is sensation in the appearance, for that constitutes the
empirical – but only the form of appearance, space and time. If anyone
wishes to doubt in the slightest that the two are a not determinations in-
hering in things in themselves but only mere determinations inhering
in the relation of those things to sensibility, I would very much like to
know how he can ﬁnd it possible to know, a priori and therefore before
all acquaintance with things, how their intuition must be constituted –
which certainly is the case here with space and time. But this is com-
pletely comprehensible as soon as the two are taken for nothing more
than formal conditions of our sensibility, and objects are taken merely
for appearances; for then the form of appearance, i.e., the pure intuition,
certainly can be represented from ourselves, i.e., a priori.
§ 12
4: 285
In order to add something by way of illustration and conﬁrmation, we
need only to consider the usual and unavoidably necessary procedure of
the geometers. All proofs of the thoroughgoing equality of two given
ﬁgures (that one can in all parts be put in the place of the other) ulti-
mately come down to this: that they are congruent with one another;
which plainly is nothing other than a synthetic proposition based upon
immediate intuition; and this intuition must be given pure and a priori,
for otherwise that proposition could not be granted as apodictically cer-
tain but would have only empirical certainty. It would only mean: we
observe it always to be so and the proposition holds only as far as our
perception has reached until now. That full-standing space (a space that
is itself not the boundary of another space) 31 has three dimensions, and
that space in general cannot have more, is built upon the proposition
that not more than three lines can cut each other at right angles in one
point; this proposition can, however, by no means be proven from con-
cepts, but rests immediately upon intuition, and indeed on pure a priori
intuition, because it is apodictically certain; indeed, that we can require
that a line should be drawn to inﬁnity (in indeﬁnitum), or that a series of
changes (e.g., spaces traversed through motion) should be continued to
inﬁnity, presupposes a representation of space and of time that can only
inhere in intuition, that is, insofar as the latter is not in itself bounded
a
Reading sind for seyn, with Ak.
80Prolegomena to any future metaphysics
by anything; 32 for this could never be concluded from concepts. There-
fore pure intuitions a priori indeed actually do underlie mathematics, and
make possible its synthetic and apodictically valid propositions; and con-
sequently our transcendental deduction of the concepts of a space and
time 33 at the same time explains the possibility of a pure mathematics, a
possibility which, without such a deduction, and without our assuming
that “everything which our senses may be given (the outer in space, the
inner in time) is only intuited by us as it appears to us, not as it is in itself,”
could indeed be granted, but into which we could have no insight at all.
§ 13
All those who cannot yet get free of the conception, as if space and time
were actual qualities attaching to things in themselves, can exercise their
acuity on the following paradox, and, if they have sought its solution in
vain, can then, free of prejudice at least for a few moments, suppose that
perhaps the demotion of space and of time to mere forms of our sensory
intuition may indeed have foundation.
If two things are fully the same (in all determinations belonging to
magnitude and quality) in all the parts of each that can always be cog-
nized by itself alone, it should indeed then follow that one, in all cases
and respects, can be put in the place of the other, without this exchange
causing the least recognizable difference. In fact this is how things stand
with plane ﬁgures in geometry; yet various spherical ﬁgures, 34 notwith-
standing this sort of complete inner agreement, nonetheless reveal such
a difference b in outer relation that one cannot in any case be put in the
place of the other; e.g., two spherical triangles from each of the hemi-
spheres, which have an arc of the equator for a common base, can be fully
equal with respect to their sides as well as their angles, so that nothing
will be found in either, when it is fully described by itself, that is not also
in the description of the other, and still one cannot be put in the place
of the other (that is, in the opposite hemisphere); and here is then after
all an inner difference between the triangles that no understanding can
specify as inner, and that reveals itself only through the outer relation
in space. But I will cite more familiar instances that can be taken from
ordinary life.
What indeed can be more similar to, and in all parts more equal to, my
hand or my ear than its image in the mirror? And yet I cannot put such a
hand as is seen in the mirror in the place of its original; for if the one was a
right hand, then the other in the mirror is a left, and the image of the right
ear is a left one, which can never take the place of the former. Now there
a
b
Reading von for im, with Vorländer.
Adding Verschiedenheit, with Vorländer.
81
4: 286Prolegomena to any future metaphysics
are no inner differences here that any understanding could merely think;
and yet the differences are inner as far as the senses teach, for the left hand
cannot, after all, be enclosed within the same boundaries as the right (they
cannot be made congruent), despite all reciprocal equality and similarity;
one hand’s glove cannot be used on the other. What then is the solution?
These objects are surely not representations of things as they are in
themselves, and as the pure understanding would cognize them, rather,
they are sensory intuitions, i.e., appearances, whose possibility rests on
the relation of certain things, unknown in themselves, to something
else, namely our sensibility. Now, space is the form of outer intuition
of this sensibility, and the inner determination of any space is possible
only through the determination of the outer relation to the whole space
of which the space is a part (the relation to outer sense); that is, the
part is possible only through the whole, which never occurs with things
in themselves as objects of the understanding alone, but well occurs
with mere appearances. We can therefore make the difference between
similar and equal but nonetheless incongruent things (e.g., oppositely
spiralled snails) intelligible through no concept alone, but only through
the relation to right-hand and left-hand, which refers immediately to
intuition.
Note I
4: 287
Pure mathematics, and especially pure geometry, can have objective re-
ality only under the single condition that it refers merely to objects of
the senses, with regard to which objects, however, the principle remains
ﬁxed, that our sensory representation is by no means a representation
of things in themselves, but only of the way in which they appear to us.
From this it follows, not at all that the propositions of geometry are a
determinations of a mere ﬁgment of our poetic phantasy, 35 and there-
fore could not with certainty be referred to actual objects, but rather,
that they are valid necessarily for space and consequently for everything
that may be found in space, because space is nothing other than the form
of all outer appearances, under which alone objects of the senses can be
given to us. Sensibility, whose form lies at the foundation of geometry, is
that upon which the possibility of outer appearances rests; these, there-
fore, can never contain anything other than what geometry prescribes
to them. It would be completely different if the senses had to represent
objects as they are in themselves. For then it absolutely would not follow
from the representation of space, a representation that serves a priori,
with all the various properties of space, as foundation for the geometer,
that all of this, together with what is deduced from it, must be exactly so
a
Adding sind, with Ak.
82Prolegomena to any future metaphysics
in nature. The space of the geometer would be taken for mere fabrication
and would be credited with no objective validity, because it is simply not
to be seen how things would have to agree necessarily with the image that
we form of them by ourselves and in advance. If, however, this image –
or, better, this formal intuition – is the essential property of our sensibil-
ity by means of which alone objects are given to us, and if this sensibility
represents not things in themselves but only their appearances, then it
is very easy to comprehend, and at the same time to prove incontrovert-
ibly: that all outer objects of our sensible world must necessarily agree,
in complete exactitude, with the propositions of geometry, because sen-
sibility itself, through its form of outer intuition (space), with which the
geometer deals, ﬁrst makes those objects possible, as mere appearances.
It will forever remain a remarkable phenomenon in the history of phi-
losophy that there was a time when even mathematicians who were at
the same time philosophers began to doubt, not, indeed, the correctness
of their geometrical propositions insofar as they related merely to space,
but the objective validity and application to nature of this concept itself
and all its geometrical determinations, since they were concerned that
a line in nature might indeed be composed of physical points, conse-
quently that true space in objects might be composed of simple parts,
notwithstanding that the space which the geometer holds in thought can
by no means be composed of such things. 36 They did not realize that
this space in thought itself makes possible physical space, i.e., the ex-
tension of matter; that this space is by no means a property of things in
themselves, but only a form of our power of sensory representation; that
all objects in space are mere appearances, i.e., not things in themselves
but representations of our sensory intuition; and that, since space as the
geometer thinks it is precisely the form of sensory intuition which we
ﬁnd in ourselves a priori and which contains the ground of the possibility
of all outer appearances (with respect to their form), these appearances
must of necessity and with the greatest precision harmonize with the
propositions of the geometer, which he extracts not from any fabricated
concept, but from the subjective foundation of all outer appearances,
namely sensibility itself. In this and no other way can the geometer be
secured, regarding the indubitable objective reality of his propositions,
against all the chicaneries of a shallow metaphysics, however strange this
way must seem to such a metaphysics because it does not go back to the
sources of its concepts.
Note II
Everything that is to be given to us as object must be given to us in intu-
ition. But all our intuition happens only by means of the senses; the un-
derstanding intuits nothing, but only reﬂects. Now since, in accordance
83
4: 288Prolegomena to any future metaphysics
4: 289
with what has just been proven, the senses never and in no single instance
enable us to cognize things in themselves, but only their appearances,
and as these are mere representations of sensibility, “consequently all
bodies together with the space in which they are found must be taken
for nothing but mere representations in us, and exist nowhere else than
merely in our thoughts.” Now is this not manifest idealism? 37
Idealism consists in the claim that there are none other than thinking
beings; the other things that we believe we perceive in intuition are only
representations in thinking beings, to which in fact no object existing
outside these beings corresponds. I say in opposition: There are things
given to us as objects of our senses existing outside us, yet we know a
nothing of them as they may be in themselves, but are acquainted b only
with their appearances, i.e., with the representations that they produce
in us because they affect our senses. Accordingly, I by all means avow
that there are bodies outside us, i.e., things which, though completely
unknown c to us as to what they may be in themselves, we know d through
the representations which their inﬂuence on our sensibility provides for
us, and to which we give the name of a body – which word therefore
merely signiﬁes the appearance of this object that is unknown to us but
is nonetheless real. Can this be called idealism? It is the very opposite of it.
That one could, without detracting from the actual existence of outer
things, say of a great many of their predicates: they belong not to these
things in themselves, but only to their appearances and have no existence
of their own outside our representation, is something that was generally
accepted and acknowledged long before Locke’s time, though more com-
monly thereafter. To these predicates belong warmth, color, taste, etc.
That I, however, even beyond these, include (for weighty reasons) also
among mere appearances the remaining qualities of bodies, which are
called primarias: extension, place, and more generally space along with
everything that depends on it (impenetrability or materiality, shape, etc.),
is something against which not the least ground for uncertainty can be
raised; and as little as someone can be called an idealist because he wants
to admit colors as properties that attach not to the object in itself, but
only to the sense of vision as modiﬁcations, just as little can my system
be called idealist simply because I ﬁnd that even more of, nay, all of the
properties that make up the intuition of a body belong merely to its appear-
ance: for the existence of the thing that appears is not thereby nulliﬁed,
as with real idealism, but it is only shown that through the senses we
cannot cognize it at all as it is in itself.
a
b
c
d
wissen
kennen
unbekannt
kennen
84Prolegomena to any future metaphysics
I would very much like to know how then my claims must be framed
so as not to contain any idealism. Without doubt I would have to say: that
the representation of space not only is perfectly in accordance with the
relation that our sensibility has to objects, for I have said that, but that it
is even fully similar to the object; an assertion to which I can attach no
sense, any more than to the assertion that the sensation of red is similar
to the property of cinnabar that excites this sensation in me.
4: 290
Note III
From this an easily foreseen but empty objection can now be quite eas-
ily rejected: “namely that through the ideality of space and time the
whole sensible world would be transformed into sheer illusion.” 38 After
all philosophical insight into the nature of sensory cognition had previ-
ously been perverted by making sensibility into merely a confused kind
of representation, through which we might still cognize things as they
are but without having the ability to bring everything in this represen-
tation of ours to clear consciousness, we showed on the contrary that
sensibility consists not in this logical difference of clarity or obscurity,
but in the genetic difference of the origin of the cognition itself, since
sensory cognition does not at all represent things as they are but only in
the way in which they affect our senses, and therefore that through the
senses mere appearances, not the things themselves, are given to the un-
derstanding for reﬂection; 39 from this necessary correction an objection
arises, springing from an inexcusable and almost deliberate misinterpre-
tation, as if my system transformed all the things of the sensible world
into sheer illusion.
If an appearance is given to us, we are still completely free as to how
we want to judge things from it. The former, namely the appearance,
was based on the senses, but the judgment on the understanding, and
the only question is whether there is truth in the determination of the
object or not. The difference between truth and dream, however, is not
decided through the quality of the representations that are referred to
objects, for they are the same in both, but through their connection ac-
cording to the rules that determine the combination of representations
in the concept of an object, and how far they can or cannot stand to-
gether in one experience. And then it is not the fault of the appearances
at all, if our cognition takes illusion for truth, i.e., if intuition, through
which an object is given to us, is taken for the concept of the object,
or even for its existence, which only the understanding can think. The
course of the planets is represented to us by the senses as now pro-
gressive, now retrogressive, and herein is neither falsehood nor truth,
because as long as one grants that this is as yet only appearance, one
still does not judge at all the objective quality of their motion. Since,
85
4: 291Prolegomena to any future metaphysics
4: 292
however, if the understanding has not taken good care to prevent this
subjective mode of representation from being taken for objective, a false
judgment can easily arise, one therefore says: they appear to go back-
wards; but the illusion is not ascribed to the senses, but to the under-
standing, whose lot alone it is to render an objective judgment from the
appearance.
In this manner, if we do not reﬂect at all on the origin of our represen-
tations, and we connect our intuitions of the senses, whatever they may
contain, in space and time according to rules for the combination of all
cognition in one experience, then either deceptive illusion or truth can
arise, according to whether we are heedless or careful; that concerns only
the use of sensory representations in the understanding, and not their
origin. In the same way, if I take all the representations of the senses
together with their form, namely space and time, for nothing but ap-
pearances, and these last two for a mere form of sensibility that is by no
means to be found outside it in the objects, and I make use of these same
representations only in relation to possible experience: then in the fact
that I take a them for mere appearances is contained not the least illusion
or temptation toward error; for they nonetheless can be connected to-
gether correctly in experience according to rules of truth. In this manner
all the propositions of geometry hold good for space as well as for all
objects of the senses, and hence for all possible experience, whether I
regard space as a mere form of sensibility or as something inhering in
things themselves; though only in the ﬁrst case can I comprehend how
it may be possible to know those propositions a priori for all objects of
outer intuition; otherwise, with respect to all merely possible experience,
everything remains just as if I had never undertaken this departure from
the common opinion.
But if I venture to go beyond all possible experience with my concepts
of space and time – which is inevitable if I pass them off for qualities that
attach to things in themselves (for what should then prevent me from
still permitting them to hold good for the very same things, even if my
senses might now be differently framed and either suited to them or
not?) – then an important error can spring up which rests on an illusion,
since I passed off as universally valid that which was a condition for the
intuition of things (attaching merely to my subject, and surely valid for all
objects of the senses, hence for all merely possible experience), because I
referred it to the things in themselves and did not restrict it to conditions
of experience.
Therefore, it is so greatly mistaken that my doctrine of the ideality of
space and time makes the whole sensible world a mere illusion, that, on
the contrary, my doctrine is the only means for securing the application
a
Reading halte for enthalte, with Ak.
86Prolegomena to any future metaphysics
to actual objects of one of the most important bodies of cognition –
namely, that which mathematics expounds a priori – and for prevent-
ing it from being taken for nothing but mere illusion, since without
this observation it would be quite impossible to make out whether the
intuitions of space and time, which we do not derive from experience
but which nevertheless lie a priori in our representations, were not mere
self-produced fantasies, to which no object at all corresponds, at least
not adequately, and therefore geometry itself a mere illusion, whereas
we have been able to demonstrate the incontestable validity of geometry
with respect to all objects of the sensible world for the very reason that
the latter are mere appearances.
Second, it is so greatly mistaken that these principles of mine, be-
cause they make sensory representations into appearances, are supposed,
in place of the truth of experience, to transform sensory representations
into mere illusion, that, on the contrary, my principles are the only means
of avoiding the transcendental illusion by which metaphysics has always
been deceived and thereby tempted into the childish endeavor of chasing
after soap bubbles, because appearances, which after all are mere rep-
resentations, were taken for things in themselves; from which followed
all those remarkable enactments of the antinomy of reason, which I will
mention later on, and which is removed through this single observation:
that appearance, as long as it is used in experience, brings forth truth, but
as soon as it passes beyond the boundaries of experience and becomes
transcendent, brings forth nothing but sheer illusion.
Since I therefore grant their reality to the things that we represent
to ourselves through the senses, and limit our sensory intuition of these
things only to the extent that in no instance whatsoever, not even in the
pure intuitions of space and time, does it represent a anything more than
mere appearances of these things, and never their quality in themselves,
this is therefore no thoroughgoing illusion ascribed by me to nature,
and my protestation against all imputation of idealism is so conclusive
and clear that it would even seem superﬂuous if there were not unautho-
rized judges who, being glad to have an ancient name for every deviation
from their false though common opinion, and never judging the spirit
of philosophical nomenclatures but merely clinging to the letter, were
ready to put their own folly in the place of well-determined concepts,
and thereby to twist and deform them. For the fact that I have myself
given to this theory of mine the name of transcendental idealism cannot
justify anyone in confusing it with the empirical idealism of Descartes
(although this idealism was only a problem, whose insolubility left ev-
eryone free, in Descartes’ opinion, to deny the existence of the corporeal
world, since the problem could never be answered satisfactorily) or with
a
Reading vorstelle for vorstellen, with Ak.
87
4: 293Prolegomena to any future metaphysics
4: 294
the mysticaland visionary a, 40 idealism of Berkeley (against which, along
with other similar fantasies, our Critique, on the contrary, contains the
proper antidote). 41 For what I called idealism did not concern the ex-
istence of things (the doubting of which, however, properly constitutes
idealism according to the received meaning), for it never came into my
mind to doubt that, but only the sensory representation of things, to
which space and time above all belong; and about these last, hence in
general about all appearances, I have only shown: that they are not things
(but mere modes of representation), nor are they determinations that be-
long to things in themselves. The word transcendental, however, which
with me never signiﬁes a relation of our cognition to things, but only
to the faculty of cognition, was intended to prevent this misinterpretation.
But before it prompts still more of the same, b I gladly withdraw this
name, and I will have it called critical idealism. But if it is an in fact
reprehensible idealism to transform actual things (not appearances) into
mere representations, 42 with what name shall we christen that idealism
which, conversely, makes mere representations into things? I think it
could be named dreaming idealism, to distinguish it from the preced-
ing, which may be called visionary idealism, both of which were to have
been held off by my formerly so-called transcendental, or better, critical
idealism.
a
b
schwaermerischen
Reading dieselbe for denselben, with Ak.
88Main Transcendental Question,
Second Part
How is pure natural science possible?
§ 14
Nature is the existence of things, insofar as that existence is determined
according to universal laws. If nature meant the existence of things in
themselves, we would never be able to cognize it, either a priori or a pos-
teriori. Not a priori, for how are we to know what pertains to things in
themselves, inasmuch as this can never come about through the analysis
of our concepts (analytical propositions), since I do not want to know
what may be contained in my concept of a thing (for that belongs to
its logical essence), but what would be added to this concept in the ac-
tuality of a thing, and what the thing itself would be determined by in
its existence apart from my concept. My understanding, and the con-
ditions under which alone it can connect the determinations of things
in their existence, prescribes no rule to the things themselves; these do
not conform to my understanding, but my understanding would have to
conform to them; they would therefore have to be given to me in advance
so that these determinations could be drawn from them, but then they
would not be cognized a priori.
Such cognition of the nature of things in themselves would also be
impossible a posteriori. For if experience were supposed to teach me
laws to which the existence of things is subject, then these laws, inso-
far as they relate to things in themselves, would have to apply to them
necessarily even apart from my experience. Now experience teaches me
what there is and how it is, but never that it necessarily must be so and
not otherwise. Therefore it can never teach me the nature of things in
themselves.
§ 15
Now we are nevertheless actually in possession of a pure natural sci-
ence, which, a priori and with all of the necessity required for apodictic
propositions, propounds laws to which nature is subject. Here I need call
to witness only that propaedeutic to the theory of nature which, under
89
4: 295Prolegomena to any future metaphysics
the title of universal natural science, precedes all of physics (which is
founded on empirical principles). a Therein we ﬁnd mathematics applied
to appearances, and also merely discursive principles b (from concepts),
which make up the philosophical part of pure cognition of nature. 43
But indeed there is also much in it that is not completely pure and in-
dependent of sources in experience, such as the concept of motion, of
impenetrability (on which the empirical concept of matter is based), of
inertia, among others, so that it cannot be called completely pure natu-
ral science; furthermore it refers only to the objects of the outer senses,
and therefore does not provide an example of a universal natural sci-
ence in the strict sense; for that would have to bring nature in general –
whether pertaining to an object of the outer senses or of the inner sense
(the object of physics as well as psychology) – under universal laws. But
among the principles of this universal physics 44 a few are found that
actually have the universality we require, such as the proposition: that
substance remains and persists, that everything that happens always pre-
viously is determined by a cause according to constant laws, and so on.
These are truly universal laws of nature, that exist fully a priori. There
is then in fact a pure natural science, and now the question is: How is it
possible?
§ 16
4: 296
The word nature assumes yet another meaning, namely one that de-
termines the object, whereas in the above meaning it only signiﬁed the
conformity to law of the determinations of the existence of things in gen-
eral. Nature considered materialiter 45 is the sum total of all objects of expe-
rience. We are concerned here only with this, since otherwise things
that could never become objects of an experience if they had to be
cognized according to their nature would force us to concepts whose
signiﬁcance could never be given in concreto (in any example of a pos-
sible experience), and we would therefore have to make for ourselves
mere concepts of the nature of those things, c the reality of which con-
cepts, i.e., whether they actually relate to objects or are mere beings
of thought, could not be decided at all. Cognition of that which can-
not be an object of experience would be hyperphysical, and here we
are not concerned with such things at all, but rather with that cogni-
tion of nature, the reality of which can be conﬁrmed through experi-
ence, even though such cognition is possible a priori and precedes all
experience.
a
b
c
Principien
Grundsätze
Reading deren for dessen, with Ak.
90Prolegomena to any future metaphysics
§ 17
The formal in nature in this narrower meaning is therefore the confor-
mity to law of all objects of experience, and, insofar as this conformity
is cognized a priori, the necessary conformity to law of those objects. But
it has just been shown: that the laws of nature can never be cognized a
priori in objects insofar as these objects are considered, not in relation to
possible experience, but as things in themselves. We are here, however,
concerned not with things in themselves (the properties of which we
leave undetermined), but only with things as objects of a possible expe-
rience, and the sum total of such objects is properly what we here call
nature. And now I ask whether, if the discussion is of the possibility of a
cognition of nature a priori, it would be better to frame the problem in
this way: How is it possible to cognize a priori the necessary conformity
to law of things as objects of experience, or: How is it possible in general
to cognize a priori the necessary conformity to law of experience itself with
regard to all of its objects?
On closer examination, whether the question is posed one way or the
other, its solution will come out absolutely the same with regard to the
pure cognition of nature (which is actually the point of the question).
For the subjective laws under which alone a cognition of things through
experience a is possible also hold good for those things as objects of a
possible experience (but obviously not for them as things in themselves,
which, however, are not at all being considered here). It is completely the
same, whether I say: A judgment of perception can never be considered as
valid for experience without the law, that if an event is perceived then it is
always referred to something preceding from which it follows according
to a universal rule; or if I express myself in this way: Everything of which
experience shows that it happens must have a cause.
It is nonetheless more appropriate to choose the ﬁrst formulation.
For since we can indeed, a priori and previous to any objects being given,
have a cognition of those conditions under which alone an experience
regarding objects is possible, but never of the laws to which objects may
be subject in themselves without relation to possible experience, we will
therefore be able to study a priori the nature of things in no other way
than by investigating the conditions, and the universal (though subjec-
tive) laws, under which alone such a cognition is possible as experience (as
regards mere form), and determining the possibility of things as objects
of experience accordingly; for were I to choose the second mode of ex-
pression and to seek the a priori conditions under which nature is possible
as an object of experience, I might then easily fall into misunderstanding
a
Erfahrungserkenntniss; not translated as “empirical cognition,” which translates Kant’s
empirische Erkenntniss, which he distinguished from the former (§18).
91
4: 297Prolegomena to any future metaphysics
and fancy that I had to speak about nature as a thing in itself, and in that
case I would be wandering about fruitlessly in endless endeavors to ﬁnd
laws for things about which nothing is given to me.
We will therefore be concerned here only with experience and with
the universal conditions of its possibility which are given a priori, and
from there we will determine nature as the whole object of all possible
experience. I think I will be understood: that here I do not mean the rules
for the observation of a nature that is already given, which presuppose
experience already; and so do not mean, how we can learn the laws from
nature (through experience), for these would then not be laws a priori
and would provide no pure natural science; but rather, how the a priori
conditions of the possibility of experience are at the same time the sources
out of which all universal laws of nature must be derived.
§ 18
4: 298
We must therefore ﬁrst of all note: that, although all judgments of ex-
perience are empirical, i.e., have their basis in the immediate perception
of the senses, nonetheless the reverse is not the case, that all empirical
judgments are therefore judgments of experience; rather, beyond the
empirical and in general beyond what is given in sensory intuition, spe-
cial concepts must yet be added, which have their origin completely a
priori in the pure understanding, and under which every perception can
ﬁrst be subsumed and then, by means of the same concepts, transformed
into experience.
Empirical judgments, insofar as they have objective validity, are judgments
of experience; those, however, that are only subjectively valid I call mere
judgments of perception. The latter do not require a pure concept
of the understanding, but only the logical connection of perceptions in
a thinking subject. But the former always demand, in addition to the
representations of sensory intuition, special concepts originally generated
in the understanding, which are precisely what make the judgment of
experience objectively valid.
All of our judgments are at ﬁrst mere judgments of perception; they
hold only for us, i.e., for our subject, and only afterwards do we give
them a new relation, namely to an object, and intend that the judgment
should also be valid at all times for us and for everyone else; for if a
judgment agrees with an object, then all judgments of the same object
must also agree with one another, and hence the objective validity of a
judgment of experience signiﬁes nothing other than its necessary uni-
versal validity. But also conversely, if we ﬁnd cause to deem a judgment
necessarily, universally valid (which is never based on the perception, but
on the pure concept of the understanding under which the perception
is subsumed), we must then also deem it objective, i.e., as expressing
92Prolegomena to any future metaphysics
not merely a relation of a perception to a subject, but a property of an
object; for there would be no reason why other judgments necessarily
would have to agree with mine, if there were not the unity of the object –
an object to which they all refer, with which they all agree, and, for that
reason, also must all harmonize among themselves.
§ 19
Objective validity and necessary universal validity (for everyone) are
therefore interchangeable concepts, and although we do not know the
object in itself, nonetheless, if we regard a judgment as universally valid
and hence necessary, objective validity is understood to be included.
Through this judgment we cognize the object (even if it otherwise re-
mains unknown as it may be in itself) by means of the universally valid
and necessary connection of the given perceptions; and since this is the
case for all objects of the senses, judgments of experience will not de-
rive their objective validity from the immediate cognition of the object
(for this is impossible), but merely from the condition for the universal
validity of empirical judgments, which, as has been said, never rests on
empirical, or indeed sensory conditions at all, but on a pure concept
of the understanding. The object always remains unknown in itself; if,
however, through the concept of the understanding the connection of
the representations which it provides to our sensibility is determined as
universally valid, then the object is determined through this relation, and
the judgment is objective.
Let us provide examples: that the room is warm, the sugar sweet,
the wormwood 46 repugnant, ∗ are merely subjectively valid judgments.
I do not at all require that I should ﬁnd it so at every time, or that
everyone else should ﬁnd it just as I do; they express only a relation of
two sensations to the same subject, namely myself, and this only in my
present state of perception, and are therefore not expected to be valid for
the object: these I call judgments of perception. The case is completely
different with judgments of experience. What experience teaches me
∗
I gladly admit that these examples do not present judgments of perception such
as could ever become judgments of experience if a concept of the understand-
ing were also added, because they refer merely to feeling – which everyone
acknowledges to be merely subjective and which must therefore never be at-
tributed to the object – and therefore can never become objective; I only wanted
to give for now an example of a judgment that is merely subjectively valid and
that contains in itself no basis for necessary universal validity and, thereby, for a
relation to an object. An example of judgments of perception that become judg-
ments of experience through the addition of a concept of the understanding
follows in the next note.
93
4: 299Prolegomena to any future metaphysics
under certain circumstances, it must teach me at every time and teach
everyone else as well, and its validity is not limited to the subject or its
state at that time. Therefore I express all such judgments as objectively
valid; as, e.g., if I say: the air is elastic, then this judgment is to begin
with only a judgment of perception; I relate two sensations in my senses
only to one another. If I want it to be called a judgment of experience, I
then require that this connection be subject to a condition that makes it
universally valid. I want therefore that I, at every time, and also everyone
else, would necessarily have to connect the same perceptions a under the
same circumstances.
§ 20
4: 300
We will therefore have to analyze experience in general, in order to see
what is contained in this product of the senses and the understanding,
and how the judgment of experience is itself possible. At bottom lies the
intuition of which I am conscious, i.e., perception ( perceptio), which be-
longs solely to the senses. But, secondly, judging (which pertains solely
to the understanding) also belongs here. Now this judging can be of two
types: ﬁrst, when I merely compare the perceptions and connect them
in a consciousness of my state, or, second, when I connect them in a
consciousness in general. The ﬁrst judgment is merely a judgment of
perception and has thus far only subjective validity; it is merely a con-
nection of perceptions within my mental state, without reference to the
object. Hence it is not, as is commonly imagined, sufﬁcient for experi-
ence to compare perceptions and to connect them in one consciousness
by means of judging; from that there arises no universal validity and ne-
cessity of the judgment, on account of which alone it can be objectively
valid and so can be experience.
A completely different judgment therefore occurs before experience
can arise from perception. The given intuition must be subsumed under
a concept that determines the form of judging in general with respect
to the intuition, connects the empirical consciousness of the latter in
a consciousness in general, and thereby furnishes empirical judgments
with universal validity; a concept of this kind is a pure a priori concept
of the understanding, which does nothing but simply determine for an
intuition the mode in general in which it can serve for judging. The
concept of cause being such a concept, it therefore determines the intu-
ition which is subsumed under it, e.g., that of air, with respect to judging
in general – namely, so that the concept of air serves, with respect to
expansion, in the relation of the antecedent to the consequent in a hy-
pothetical judgment. The concept of cause is therefore a pure concept
a
Reading Wahrnehmungen for Wahrnehmung, as suggested at Ak 4:617.
94Prolegomena to any future metaphysics
of the understanding, which is completely distinct from all possible per-
ception, and serves only, with respect to judging in general, to determine
that representation which is contained under it and so to make possible
a universally valid judgment.
Now before a judgment of experience can arise from a judgment of
perception, it is ﬁrst required: that the perception be subsumed under
a concept of the understanding of this kind; e.g., the air belongs under
the concept of cause, a which determines the judgment about the air as
hypothetical with respect to expansion. ∗ This expansion is thereby rep-
resented not as belonging merely to my perception of the air in my state
of perception or in several of my states or in the state of others, but as
necessarily belonging to it, and the judgment: the air is elastic, becomes
universally valid and thereby for the ﬁrst time a judgment of experi-
ence, because certain judgments occur beforehand, which subsume the
intuition of the air under the concept of cause and effect, and thereby
determine the perceptions not merely with respect to each other in my
subject, but with respect to the form of judging in general (here, the hypo-
thetical), and in this way make the empirical judgment universally valid.
If one analyzes all of one’s synthetic judgments insofar as they are ob-
jectively valid, one ﬁnds that they never consist in mere intuitions that
have, as is commonly thought, merely been connected in a b judgment
through comparison, 47 but rather that they would not be possible if,
over and above the concepts drawn from intuition, a pure concept of
the understanding had not been added under which these concepts had
been subsumed and in this way ﬁrst combined into an objectively valid
judgment. Even the judgments of pure mathematics in its simplest ax-
ioms are not exempt from this condition. The principle: a straight line is
the shortest line between two points, presupposes that the line has been
subsumed under the concept of magnitude, which certainly is no mere
intuition, but has its seat solely in the understanding and serves to deter-
mine the intuition (of the line) with respect to such judgments as may be
passed on it as regards the quantity of these judgments, namely plurality
∗ To have a more easily understood example, consider the following: If the sun
shines on the stone, it becomes warm. This judgment is a mere judgment of
perception and contains no necessity, however often I and others also have
perceived this; the perceptions are only usually found so conjoined. But if I
say: the sun warms the stone, then beyond the perception is added the under-
standing’s concept of cause, which connects necessarily the concept of sunshine
with that of heat, and the synthetic judgment becomes necessarily universally
valid, hence objective, and changes from a perception into experience.
a Reading Ursache for Ursachen, with Ak.
Reading einem for ein, with Ak.
b
95
4: 301Prolegomena to any future metaphysics
4: 302
(as judicia plurativa ∗ ), since through such judgments it is understood that
in a given intuition a homogeneous plurality is contained.
§ 21
In order therefore to explain the possibility of experience insofar as it
rests on pure a priori concepts of the understanding, we must ﬁrst present
that which belongs to a judgments in general, and the various moments of
the understanding therein, in a complete table; for the pure concepts of
the understanding – which are nothing more than concepts of intuitions
in general insofar as these intuitions are, with respect to one or another
of these moments, in themselves determined to judgments and therefore
determined necessarily and with universal validity – will come out exactly
parallel to them. By this means the a priori principles of the possibility
of all experience as objectively valid empirical cognition will also be
determined quite exactly. For they are nothing other than propositions
that subsume all perception (according to certain universal conditions of
intuition) under those pure concepts of the understanding.
logical table
of Judgments
1.
According to Quantity
Universal
Particular
Singular
2.
According to Quality 3.
According to Relation
Afﬁrmative
Negative
Inﬁnite Categorical
Hypothetical
Disjunctive
4: 303
4.
According to Modality
Problematic
Assertoric
Apodictic
∗ So I would prefer those judgments to be called, which are called particularia in
logic. For the latter expression already contains the thought that they are not
universal. If, however, I commence from unity (in singular judgments) and then
continue on to the totality, I still cannot mix in any reference to the totality; I
think only a plurality without totality, not the exception to the latter. 48 This is
necessary, if the logical moments are to be placed under the pure concepts of
the understanding; in logical usage things can remain as they were.
a Reading zu for zum, with Vorländer.
96Prolegomena to any future metaphysics
transcendental table
of Concepts of the Understanding
1.
According to Quantity
Unity (measure)
Plurality (magnitude)
Totality (the whole)
2.
According to Quality 3.
According to Relation
Reality
Negation
Limitation Substance
Cause
Community
4.
According to Modality
Possibility
Existence
Necessity
pure physiological table
of Universal Principles of Natural Science
1.
Axioms
of intuition
2.
Anticipations
of perception
3.
Analogies
of experience
4.
Postulates
of empirical thinking in general
§ 21[a] a
In order to comprise all the preceding in one notion, it is ﬁrst of all
necessary to remind the reader that the discussion here is not about
the genesis of experience, but about that which lies in experience. The
former belongs to empirical psychology and could never be properly
developed even there without the latter, which belongs to the critique of
cognition and especially of the understanding.
a
Adding the letter “a,” with Ak and Vorländer, to distinguish this section from the pre-
ceding one, both of which are shown as “§21” in the original edition.
97
4: 304Prolegomena to any future metaphysics
Experience consists of intuitions, which belong to sensibility, and of
judgments, which are solely the understanding’s business. Those judg-
ments that the understanding forms solely from sensory intuitions are,
however, still not judgments of experience by a long way. For in the one a
case the judgment would only connect perceptions as they are given in
sensory intuition; but in the latter case the judgments are supposed to
say what experience in general contains, therefore not what mere per-
ception – whose validity is merely subjective – contains. The judgment
of experience must still therefore, beyond the sensory intuition and its
logical connection (in accordance with which the intuition has been ren-
dered universal through comparison in a judgment), add something that
determines the synthetic judgment as necessary, and thereby as univer-
sally valid; and this can be nothing but that concept which represents the
intuition as in itself determined with respect to one form of judgment
rather than the others, b i.e., c a concept of that synthetic unity of intu-
itions which can be represented only through a given logical function of
judgments.
§ 22
4: 305
To sum this up: the business of the senses is to intuit; that of the
understanding, to think. To think, however, is to unite representations
in a consciousness. This uniﬁcation either arises merely relative to the
subject and is contingent and subjective, or it occurs without condition
and is necessary or objective. The uniﬁcation of representations in a
consciousness is judgment. Therefore, thinking is the same as judging
or as relating representations to judgments in general. Judgments are
therefore either merely subjective, if representations are related to
one consciousness in one subject alone and are united in it, or they
are objective, if they are united in a consciousness in general, i.e., are
united necessarily therein. The logical moments of all judgments are
so many possible ways of uniting representations in a consciousness. If,
however, the very same moments serve as concepts, they are concepts
of the necessary uniﬁcation of these representations in a consciousness,
and so are principles d of objectively valid judgments. This uniﬁcation
in a consciousness is either analytic, through identity, or synthetic,
through combination and addition of various representations with one
another. Experience consists in the synthetic connection of appearances
(perceptions) in a consciousness, insofar as this connection is necessary.
a
b
c
d
Reading dem einen for einem, as suggested by Vorländer.
Reading anderen for andere, with Ak.
Reading d. i. for die, with Ak.
Principien
98Prolegomena to any future metaphysics
Therefore pure concepts of the understanding are those under which all
perceptions must ﬁrst be subsumed before they can serve in judgments
of experience, in which the synthetic unity of perceptions is represented
as necessary and universally valid. ∗
§ 23
Judgments, insofar as they are regarded merely as the condition for the
uniﬁcation of given representations in a consciousness, are rules. These
rules, insofar as they represent the uniﬁcation as necessary, are a priori
rules, and provided that there are none above them from which they can
be derived, are principles. Now since, with respect to the possibility of all
experience, if merely the form of thinking is considered in the experience,
no conditions on judgments of experience are above those that bring
the appearance (according to the varying form of their intuition) under
pure concepts of the understanding (which make the empirical judgment
objectively valid), these conditions are therefore the a priori principles
of possible experience.
Now the principles of possible experience are, at the same time, uni-
versal laws of nature that can be cognized a priori. And so the problem
that lies in our second question, presently before us: how is pure natural
science a possible? is solved. For the systematization that is required for the
form of a science is here found to perfection, since beyond the aforemen-
tioned formal conditions of all judgments in general, hence of all rules
whatsoever furnished by logic, no others are possible, and these form
a logical system; but the concepts based thereon, which contain b the a
priori conditions for all synthetic and necessary judgments, for that very
reason form a transcendental system; ﬁnally, the principles by means of
∗ But how does this proposition: that judgments of experience are supposed to
contain necessity in the synthesis of perceptions, square with my proposition,
urged many times above: that experience, as a posteriori cognition, can provide
merely contingent judgments? If I say: Experience teaches me something, I
always mean only the perception that is in it – e.g., that upon illumination of
the stone by the sun, warmth always follows – and hence the proposition from
experience is, so far, always contingent. That this warming follows necessarily
from illumination by the sun is indeed contained in the judgment of experience
(in virtue of the concept of cause), but I do not learn it from experience; rather,
conversely, experience is ﬁrst generated through this addition of a concept of
the understanding (of cause) to the perception. Concerning how the perception
may come by this addition, the Critique must be consulted, in the section on
transcendental judgment, pp. 137ff. 49
a Reading Naturwissenschaft for Vernunftwissenschaft, with Ak.
Reading enthalten for erhalten, with Ak.
b
99
4: 306Prolegomena to any future metaphysics
which all appearances are subsumed under these concepts form a phys-
iological system, i.e., a system of nature, a which precedes all empirical
cognition of nature and ﬁrst makes it possible, and can therefore be called
the true universal and pure natural science.
§ 24
4: 307
The b ﬁrst ∗ of the physiological principles subsumes all appearances, as
intuitions in space and time, under the concept of magnitude and is to
that extent a principle for the application of mathematics to experience.
The c second does not subsume the properly empirical – namely sensa-
tion, which signiﬁes the real d in intuitions – directly under the concept
of magnitude, since sensation is no intuition containing space or time, al-
though it does place the object corresponding to it in both; but there
nonetheless is, between reality (sensory representation) and nothing,
i.e., the complete emptiness of intuition in time, a difference that has
a magnitude, for indeed between every given degree of light and dark-
ness, every degree of warmth and the completely cold, every degree of
heaviness and absolute lightness, every degree of the ﬁlling of space and
completely empty space, ever smaller degrees can be thought, just as be-
tween consciousness and total unconsciousness (psychological darkness)
ever smaller degrees occur; therefore no perception is possible that would
show a complete absence, e.g., no psychological darkness is possible that
could not be regarded as a consciousness that is merely outweighed by
another, stronger one, and thus it is in all cases of sensation; as a result
of which the understanding can anticipate even sensations, which form
the proper quality of empirical representations (appearances), by means
of the principle that they all without exception, hence the real in all ap-
pearance, have degrees – which is the second application of mathematics
(mathesis intensorum) to natural science. 51
§ 25
With respect to the relation of appearances, and indeed exclusively
with regard to their existence, the determination of this relation is
∗ The three subsequent sections could be difﬁcult to understand properly, if one
does not have at hand what the Critique says about principles as well; but they
might have the advantage of making it easier to survey the general features of
such principles and to attend to the main points. 50
a Natursystem
Reading der for das, with Ak.
Reading der for das, with Ak.
Reale
b
c
d
100Prolegomena to any future metaphysics
not mathematical but dynamical, and it can a never be objectively valid,
hence ﬁt for experience, if it is not subject to a priori principles, which
ﬁrst make cognition through experience possible with respect to that
determination. 52 Therefore appearances must be subsumed under the
concept of substance, which, as a concept of the thing itself, underlies all
determination of existence; or second, insofar as a temporal sequence,
i.e., an event, is met with among the appearances, they must be sub-
sumed under the concept of an effect in relation to a b cause; or, insofar
as coexistence is to be cognized objectively, i.e., through a judgment of
experience, they must be subsumed under the concept of community
(interaction): and so a priori principles underlie objectively valid, though
empirical, judgments, i.e., they underlie the possibility of experience
insofar as it is supposed to connect objects in nature according to exis-
tence. These principles are the actual laws of nature, which can be called
dynamical.
Finally, there also belongs to judgments of experience the cognition
of agreement and connection: not so much of the appearances among
themselves in experience, but of their relation to experience in general, a
relation that contains either their agreement with the formal conditions
that the understanding cognizes, or their connection with the material
of the senses and perception, or both c united in one concept, and thus
possibility, existence, and necessity according to universal laws of nature;
all of which would constitute the physiological theory of method (the
distinction of truth and hypotheses, and the boundaries of the reliability
of the latter).
§ 26
Although the third table of principles, which is drawn from the nature of
the understanding itself according to the critical method, in itself exhibits
a perfection through which it raises itself far above every other that
has (albeit vainly) ever been attempted or may yet be attempted in the
future from the things themselves through the dogmatic method: namely,
that in it d all of the synthetic principles a priori are exhibited completely
and according to a principle, e namely that of the faculty for judging
in general (which constitutes the essence of experience with respect to
the understanding), so that one can be certain there are no more such
principles (a satisfaction that the dogmatic method can never provide) –
nevertheless this is still far from being its greatest merit.
a
b
c
d
e
Adding kann, with Ak.
Adding eine, with Vorländer.
Reading beides for beiden, as suggested at Ak 4:617.
Reading in ihr for sie, with Vorländer.
Princip
101
4: 308Prolegomena to any future metaphysics
4: 309
Notice must be taken of the ground of proof that reveals the pos-
sibility of this a priori cognition and at the same time limits all such
principles to a condition that must never be neglected if they are not to
be a misunderstood and extended in use further than the original sense
which the understanding places in them will allow: namely, that they
contain only the conditions of possible experience in general, insofar as
it is subject to a priori laws. Hence I do not say: that things in themselves
contain b a magnitude, their reality a degree, their existence a connection
of accidents in a substance, and so on; for that no one can prove, because
such a synthetic connection out of mere concepts, in which all relation
to sensory intuition on the one hand and all connection of such intuition
in a possible experience on the other is lacking, is utterly impossible.
Therefore the essential limitation on the concepts in these principles is:
that only as objects of experience are all things necessarily subject a priori
to the aforementioned conditions.
From this there follows then secondly a speciﬁcally characteristic
mode of proving the same thing: that the above-mentioned principles
are not referred directly to appearances and their relation, but to the pos-
sibility of experience, for which appearances constitute only the matter
but not the form, i.e., they are referred to the objectively and univer-
sally valid synthetic propositions through which judgments of experience
are distinguished from mere judgments of perception. This happens be-
cause the appearances, as mere intuitions that ﬁll a part of space and time,
are subject to the concept of magnitude, which synthetically uniﬁes the
multiplicity of intuitions a priori according to rules; and because the real
in the appearances must have a degree, insofar as perception contains,
beyond intuition, sensation as well, between which and nothing, i.e.,
the complete disappearance of sensation, a transition always occurs by
diminution, insofar, that is, as sensation itself ﬁlls no part of space and time, ∗
∗ Warmth, light, etc. are just as great (according to degree) in a small space as in a
large one; just as the inner representations (pain, consciousness in general) are
not smaller according to degree whether they last a short or a long time. Hence
the magnitude here is just as great in a point and in an instant as in every space
and time however large. Degrees are therefore magnitudes, c not, however, in
intuition, but in accordance with mere sensation, or indeed with the magnitude
of the ground of an intuition, and can be assessed as magnitudes only through
the relation of 1 to 0, i.e., in that every sensation can proceed in a certain
time to vanish through inﬁnite intermediate degrees, or to grow from nothing
to a determinate sensation through inﬁnite moments of accretion. (Quantitas
qualitatis est gradus.) d
a Reading sollen for soll, with Vorländer.
Reading enthalten for enthalte, as suggested at Ak 4:617.
Reading Grössen for grösser, with Vorländer.
“The magnitude of quality is degree.”
b
c
d
102Prolegomena to any future metaphysics
but yet the transition to sensation from empty time or space is possi-
ble only in time, with the consequence that although sensation, as the
quality of empirical intuition with respect to that by which a sensation
differs speciﬁcally from other sensations, can never be cognized a priori,
it nonetheless can, in a possible experience in general, as the magnitude
of perception, be distinguished intensively from every other sensation
of the same kind; from which, then, the application of mathematics to
nature, with respect to the sensory intuition whereby nature is given to
us, is ﬁrst made possible and determined.
Mostly, however, the reader must attend to the mode of proving the
principles that appear under the name of the Analogies of Experience.
For since these do not concern the generation of intuitions, as do the
principles for applying mathematics to natural science in general, but the
connection of their existence in one experience, and since this connec-
tion can be nothing a other than the determination of existence in time
according to necessary laws, under which alone the connection is objec-
tively valid and therefore is experience: it follows that the proof does not
refer to synthetic unity in the connection of things in themselves, but of
perceptions, and of these indeed not with respect to their content, but to
the determination of time and to the relation of existence in time in ac-
cordance with universal laws. These universal laws contain therefore the
necessity of the determination of existence in time in general (hence a
priori according to a rule of the understanding), if the empirical determi-
nation in relative time is to be objectively valid, and therefore to be expe-
rience. For the reader who is stuck in the long habit of taking experience
to be a mere empirical combining of perceptions – and who therefore has
never even considered that it extends much further than these reach, that
is, that it gives to empirical judgments universal validity and to do so re-
quires a pure unity of the understanding that precedes a priori – I cannot
adduce more here, these being prolegomena, except only to recommend:
to heed well this distinction of experience from a mere aggregate of per-
ceptions, and to judge the mode of proof from this standpoint.
§ 27
Here is now the place to dispose thoroughly of the Humean doubt. He
rightly afﬁrmed: that we in no way have insight through reason into
the possibility of causality, i.e., the possibility of relating the existence
of one thing to the existence of some other thing that would necessar-
ily be posited through the ﬁrst one. I add to this that we have just as
little insight into the concept of subsistence, i.e., of the necessity that
a subject, which itself cannot be a predicate of any other thing, should
a
Reading nichts for nicht, with Ak.
103
4: 310Prolegomena to any future metaphysics
4: 311
underlie the existence of things – nay, that we cannot frame any concept
of the possibility of any such thing (although we can point out examples
of its use in experience); and I also add that this very incomprehensi-
bility affects the community of things as well, since we have no insight
whatsoever into how, from the state of one thing, a consequence could
be drawn about the state of completely different things outside it (and
vice versa), and into how substances, each of which has its own sepa-
rate existence, should depend on one another and should indeed do so
necessarily. Nonetheless, I am very far from taking these concepts to be
merely borrowed from experience, and from taking the necessity repre-
sented in them to be falsely imputed and a mere illusion through which
long habit deludes us; rather, I have sufﬁciently shown that they and the
principles taken from them stand ﬁrm a priori prior to all experience,
and have their undoubted objective correctness, though of course only
with respect to experience.
§ 28
Although I therefore do not have the least concept of such a connection
of things in themselves, how they can exist as substances or act as causes
or stand in community with others (as parts of a real whole), and though
I can still less think such properties of appearances as appearances (for
these concepts do not contain what lies in appearances, but what the
understanding alone must think), we nonetheless do have a concept of
such a connection of representations in our understanding, and indeed
in judging in general, namely: that representations belong in one kind
of judgments as subject in relation to predicate, in another as ground in
relation to consequence, and in a third as parts that together make up
a whole possible experience. Further, we cognize a priori: that, without
regarding the representation of an object as determined with respect to
one or another a of these moments, we could not have any cognition at
all that was valid for the object; and if we were to concern ourselves with
the object in itself, then no unique characteristic would be possible by
which I could cognize that it b had been determined with respect to one
or another of the above-mentioned moments, i.e., that it belonged under
the concept of substance, or of cause, or (in relation to other substances)
under the concept of community; for I have no concept of the possibility
of such a connection of existence. The question is not, however, how
things in themselves, but how the cognition of things in experience is
determined with respect to said moments of judgments in general, i.e.,
how things as objects of experience can and should be subsumed under
a
b
Reading eines oder des for einer oder der, with Ak.
Reading er for es, with Ak.
104Prolegomena to any future metaphysics
those concepts of the understanding. And then it is clear that I have
complete insight into not only the possibility but also the necessity of
subsuming all appearances under these concepts, i.e., of using them as
principles of the possibility of experience.
§ 29
For having a try at Hume’s problematic concept (this, his crux
metaphysicorum), a namely the concept of cause, there is ﬁrst given to
me a priori, by means of logic: the form of a conditioned judgment in
general, that is, the use of a given cognition as ground and another as
consequent. It is, however, possible that in perception a rule of relation
will be found, which says this: that a certain appearance is constantly
followed by another (though not the reverse); and this is a case for me to
use hypothetical judgment and, e.g., to say: If a body is illuminated by
the sun for long enough, then it becomes warm. Here there is of course
not yet a necessity of connection, hence not yet the concept of cause.
But I continue on, and say: if the above proposition, which is merely a
subjective connection of perceptions, is to be a proposition of experi-
ence, then it must be regarded as necessarily and universally valid. But a
proposition of this sort would be: The sun through its light is the cause
of the warmth. The foregoing empirical rule is now regarded as a law,
and indeed as valid not merely of appearances, but of them on behalf of a
possible experience, which requires universally and therefore necessarily
valid rules. I therefore have quite good insight into the concept of cause,
as a concept that necessarily belongs to the mere form of experience,
and into its possibility as a synthetic uniﬁcation of perceptions in a con-
sciousness in general; but I have no insight at all into the possibility of
a thing in general as a cause, and that indeed because the concept of
cause indicates a condition that in no way attaches to things, but only to
experience, namely, that experience can be an objectively valid cognition
of appearances and their sequence in time only insofar as the antecedent
appearance can be connected with the subsequent one according to the
rule of hypothetical judgments.
§ 30
Consequently, even the pure concepts of the understanding have no
signiﬁcance at all if they depart from objects of experience and want to
be referred to things in themselves (noumena). b They serve as it were
a
b
“cross of metaphysics”
Noumena is a latinized Greek word (singular: noumenon) meaning literally “that which
is thought” or “that which is conceived,” but used by Kant in connection with the
105
4: 312Prolegomena to any future metaphysics
4: 313
only to spell out appearances, so that they can be read as experience;
the principles that arise from their relation to the sensible world serve
our understanding for use in experience only; beyond this there are
arbitrary connections without objective reality whose possibility cannot
be cognized a priori and whose relation to objects cannot, through any
example, be conﬁrmed or even made intelligible, since all examples can
be taken only from some possible experience or other and hence the
objects of these concepts can be met with nowhere else but in a possible
experience.
This complete solution of the Humean problem, though coming out
contrary to the surmise of the originator, thus restores to the pure con-
cepts of the understanding their a priori origin, and to the universal laws
of nature their validity as laws of the understanding, but in such a way
that it restricts their use to experience only, because their possibility is
founded solely in the relation of the understanding to experience: not,
however, in such a way that they are derived from experience, but that ex-
perience is derived from them, a completely reversed type of connection
that never occurred to Hume.
From this now ﬂows the following result of all the foregoing investiga-
tions: “All synthetic a priori principles a are nothing more than principles b
of possible experience,” and can never be related to things in themselves,
but only to appearances as objects of experience. Therefore both pure
mathematics and pure natural science can never refer to anything more
than mere appearances, and they can only represent either that which
makes experience in general possible, or that which, being derived from
these principles, c must always be able to be represented in some possible
experience or other.
§ 31
And so for once one has something determinate, and to which one can
adhere in all metaphysical undertakings, which have up to now boldly
enough, but always blindly, run over everything without distinction. It
never occurred to dogmatic thinkers that the goal of their efforts might
have been set up so close, nor even to those who, obstinate in their so-
called sound common sense, d sallied forth to insights with concepts and
principles of the pure understanding that were indeed legitimate and
a
b
c
d
philosophical meaning of nous as “intellect” to mean “intelligible objects,” or “intelligible
beings” or “beings of the understanding.” In §32 he contrasts noumena with phaenomena,
which he speaks of as “sensible beings” or “appearances.”
Grundsätze
Principien
Principien
gesunde Vernunft
106Prolegomena to any future metaphysics
natural, but were intended for use merely in experience, and for which
they neither recognized nor could recognize any determinate bound-
aries, because they neither had reﬂected on nor were able to reﬂect on
the nature or even the possibility of such a pure understanding.
Many a naturalist of pure reason (by which I mean he who trusts
himself, without any science, to decide in matters of metaphysics) would
like to pretend that already long ago, through the prophetic spirit of
his sound common sense, he had not merely suspected, but had known
and understood, that which is here presented with so much prepara-
tion, or, if he prefers, with such long-winded pedantic pomp: “namely
that with all our reason we can never get beyond the ﬁeld of experi-
ences.” But since, if someone gradually questions him on his rational
principles, a he must indeed admit that among them there are many that
he has not drawn from experience, which are therefore independent of
it and valid a priori – how and on what grounds will he then hold within
limits the dogmatist (and himself), who makes use of b these concepts
and principles beyond all possible experience for the very reason that
they are cognized independently of experience. And even he, this adept
of sound common sense, is not so steadfast that, despite all of his pre-
sumed and cheaply gained wisdom, he will not stumble unawares out
beyond the objects of experience into the ﬁeld of chimeras. Ordinarily,
he is indeed deeply enough entangled therein, although he cloaks his
ill-founded claims in a popular style, since he gives everything out as
mere probability, reasonable conjecture, or analogy.
4: 314
§ 32
Already from the earliest days of philosophy, apart from the sensible
beings c or appearances (phaenomena) that constitute the sensible world,
investigators of pure reason have thought of special intelligible beings d
(noumena), which were supposed to form an intelligible world; e and they
have granted reality to the intelligible beings alone, because they took
appearance and illusion to be one and the same thing (which may well
be excused in an as yet uncultivated age). 53
In fact, if we view the objects of the senses as mere appearances, as is
ﬁtting, then we thereby admit at the very same time that a thing in itself
underlies them, although we are not acquainted with this thing as it may
be constituted in itself, but only with its appearance, i.e., with the way in
which our senses are affected by this unknown something. Therefore the
a
b
c
d
e
Vernunftprincipien
Adding sich, with Ak.
Sinnenwesen
Verstandeswesen
Verstandeswelt
107
4: 315Prolegomena to any future metaphysics
understanding, just by the fact that it accepts appearances, also admits to
the existence of things in themselves, and to that extent we can say that
the representation of such beings as underlie the appearances, hence of
mere intelligible beings, is not merely permitted but also unavoidable.
Our critical deduction in no way excludes things of such kind
(noumena), but rather restricts the principles of aesthetic 54 in such a way
that they are not supposed to extend to all things, whereby everything
would be transformed into mere appearance, but are to be valid only for
objects of a possible experience. Hence intelligible beings are thereby al-
lowed only with the enforcement of this rule, which brooks no exception
whatsoever: that we do not know and cannot know anything determinate
about these intelligible beings at all, because our pure concepts of the
understanding as well as our pure intuitions refer to nothing but objects
of possible experience, hence to mere beings of sense, and that as soon as
one departs from the latter, not the least signiﬁcance remains for those
concepts.
§ 33
4: 316
There is in fact something insidious in our pure concepts of the under-
standing, as regards enticement toward a transcendent use; for so I call
that use which goes out beyond all possible experience. It is not only
that our concepts of substance, of force, of action, of reality, etc., are
wholly independent of experience, likewise contain no sensory appear-
ance whatsoever, and so in fact seem to refer to things in themselves
(noumena); but also, which strengthens this supposition yet further, that
they contain in themselves a necessity of determination which experience
never equals. The concept of cause contains a rule, according to which
from one state of affairs another follows with necessity; but experience
can only show us that from one state of things another state often, or,
at best, commonly, follows, and it can therefore furnish neither strict
universality nor necessity (and so forth).
Consequently, the concepts of the understanding appear to have much
more signiﬁcance and content than they would if their entire vocation
were exhausted by mere use in experience, and so the understanding
unheededly builds onto the house of experience a much roomier wing,
which it crowds with mere beings of thought, without once noticing that
it has taken its otherwise legitimate concepts far beyond the boundaries
of their use.
§ 34
Two important, nay completely indispensable, though utterly dry inves-
tigations were therefore needed, which were carried out in the Critique,
108Prolegomena to any future metaphysics
pp.137ff. and 235ff. 55 Through the ﬁrst of these it was shown that the
senses do not supply pure concepts of the understanding in concreto, but
only the schema for their use, and that the object appropriate to this
schema is found only in experience (as the product of the understand-
ing from materials of sensibility). In the second investigation (Critique,
p. 235) it is shown: that notwithstanding the independence from expe-
rience of our pure concepts of the understanding and principles, and
even their apparently larger sphere of use, nonetheless, outside the ﬁeld
of experience nothing at all can be thought by means of them, because
they can do nothing but merely determine the logical form of judgment
with respect to given intuitions; but since beyond the ﬁeld of sensibility
there is no intuition at all, these pure concepts lack completely all signif-
icance, in that there are no means through which they can be exhibited
in concreto, and so all such noumena, together with their aggregate – an
intelligible ∗ world – are nothing but representations of a problem, whose
object is in itself perfectly possible, but whose solution, given the nature
of our understanding, is completely impossible, since our understanding
is no faculty of intuition but only of the connection of given intuitions
in an experience; and experience therefore has to contain all the ob-
jects for our concepts, whereas apart from it all concepts will be without
signiﬁcance, since no intuition can be put under them.
§ 35
The imagination may perhaps be excused if it daydreams c every now
and then, i.e., if it does not cautiously hold itself inside the limits of
experience; for it will at least be enlivened and strengthened through
such free ﬂight, and it will always be easier to moderate its boldness
than to remedy its languor. That the understanding, however, which is
supposed to think, should, instead of that, daydream – for this it can never
∗ Not (as is commonly said) an intellectual a world. For the cognitions through the
understanding are intellectual, and the same sort of cognitions also refer to our
sensible world; but intelligible b means objects insofar as they can be represented
only through the understanding, and none of our sensory intuitions can refer to
them. Since, however, to each object there must nonetheless correspond some
possible intuition or other, we would therefore have to think of an understand-
ing that intuits things immediately; of this sort of understanding, however, we
have not the least concept, hence also not of the intelligible beings to which it is
supposed to refer.
a intellectuellen
intelligibel
schwärmt
b
c
109
4: 317Prolegomena to any future metaphysics
be forgiven; for all assistance in setting bounds, where needed, to the
revelry a of the imagination depends on it alone.
The understanding begins all this very innocently and chastely. First,
it puts in order the elementary cognitions that dwell in it prior to all expe-
rience but must nonetheless always have their application in experience.
Gradually, it removes these constraints, and what is to hinder it from
doing so, since the understanding has quite freely taken its principles
from within itself? And now reference is made ﬁrst to newly invented
forces in nature, soon thereafter to beings outside nature, in a word, to a
world for the furnishing of which building materials cannot fail us, since
they are abundantly supplied through fertile invention, and though not
indeed conﬁrmed by experience, are also never refuted by it. That is
also the reason why young thinkers so love metaphysics of the truly dog-
matic sort, and often sacriﬁce their time and their otherwise useful talent
to it.
It can, however, help nothing at all to want to curb these fruitless
endeavors of pure reason by all sorts of admonitions about the difﬁculty
of resolving such deeply obscure questions, by complaints over the limits
of our reason, and by reducing assertions to mere conjectures. For if the
impossibility of these endeavors has not been clearly demonstrated, and if
reason’s knowledge of itself b does not become true science, in which the
sphere of its legitimate use is distinguished with geometrical certainty
(so to speak) from that of its empty and fruitless use, then these futile
efforts will never be fully abandoned.
§ 36
How is nature itself possible?
4: 318
This question, which is the highest point that transcendental philosophy
can ever reach, and up to which, as its boundary and completion, it must
be taken, actually contains two questions.
ﬁrst: How is nature possible in general in the material sense, namely,
according to intuition, as the sum total of appearances; how are space,
time, and that which ﬁlls them both, the object of sensation, possible in
general? The answer is: by means of the constitution of our sensibility,
in accordance with which our sensibility is affected in its characteristic
way by objects that are in themselves unknown to it and that are wholly
distinct from said appearances. This answer is, in the book itself, given
in the Transcendental Aesthetic, but here in the Prolegomena through the
solution of the ﬁrst main question.
a
b
Schwärmerei
Selbsterkenntniss
110Prolegomena to any future metaphysics
second: How is nature possible inthe formal a sense, as the sum total
of the rules to which all appearances must be subject if they are to be
thought as connected in one experience? The answer cannot come out
otherwise than: it is possible only by means of the constitution of our
understanding, in accordance with which all these representations of
sensibility are necessarily referred to one consciousness, and through
which, ﬁrst, the characteristic mode of our thinking, namely by means
of rules, is possible, and then, by means of these rules, experience is
possible – which is to be wholly distinguished from insight into objects in
themselves. This answer is, in the book itself, given in the Transcendental
Logic, 56 but here in the Prolegomena, in the course of solving the second
main question.
But how this characteristic property of our sensibility itself may be
possible, or that of our understanding and of the necessary apperception
that underlies it and all thinking, cannot be further solved and answered,
because we always have need of them in turn for all answering and for
all thinking of objects.
There are many laws of nature that we can know only through ex-
perience, but lawfulness in the connection of appearances, i.e., nature
in general, we cannot come to know through any experience, because
experience itself has need of such laws, which lie a priori at the basis of
its possibility.
The possibility of experience in general is thus at the same time the
universal law of nature, and the principles of the former are themselves
the laws of the latter. For we are not acquainted with nature except as
the sum total of appearances, i.e., of the representations in us, and so we
cannot get the laws of their connection from anywhere else except the
principles of their connection in us, i.e., from the conditions of neces-
sary uniﬁcation in one consciousness, which uniﬁcation constitutes the
possibility of experience.
Even the main proposition that has been elaborated throughout this
entire part, that universal laws of nature can be cognized a priori, already
leads by itself to the proposition: that the highest legislation for nature
must lie in ourselves, i.e., in our understanding, and that we must not seek
the universal laws of nature from nature by means of experience, but, con-
versely, must seek nature, as regards its universal conformity to law, solely
in the conditions of the possibility of experience that lie in our sensibility
and understanding; for how would it otherwise be possible to become ac-
quainted with these laws a priori, since they are surely not rules of analytic
cognition, but are genuine synthetic ampliﬁcations of cognition? Such
agreement, and indeed necessary agreement, between the principles b of
a
b
Emphasis added, with Vorländer.
Principien
111
4: 319Prolegomena to any future metaphysics
4: 320
possible experience and the laws of the possibility of nature, can come
about only from one of two causes: either these laws are taken from na-
ture by means of experience, or, conversely, nature is derived from the
laws of the possibility of experience in general and is fully identical with
the mere universal lawfulness of experience. The ﬁrst one contradicts
itself, for the universal laws of nature can and must be cognized a priori
(i.e., independently of all experience) and set at the foundation of all
empirical use of the understanding; so only the second remains. ∗
We must, however, distinguish empirical laws of nature, which al-
ways presuppose particular perceptions, from the pure or universal laws
of nature, which, without having particular perceptions underlying them,
contain merely the conditions for the necessary uniﬁcation of such per-
ceptions in one experience; with respect to the latter laws, nature and
possible experience are one and the same, and since in possible experience
the lawfulness rests on the necessary connection of appearances in one
experience (without which we would not be able to cognize any object of
the sensible world at all), and so on the original laws of the understand-
ing, then, even though it sounds strange at ﬁrst, it is nonetheless certain,
if I say with respect to the universal laws of nature: the understanding does
not draw its (a priori) laws from nature, but prescribes them to it.
§ 37
We will elucidate this seemingly daring proposition through an exam-
ple, which is supposed to show: that laws which we discover in objects
of sensory intuition, especially if these laws have been cognized as nec-
essary, are already held by us to be such as have been put there by the
understanding, although they are otherwise in all respects like the laws
of nature that we attribute to experience.
§ 38
If one considers the properties of the circle by which this ﬁgure uniﬁes
in a universal rule at once so many arbitrary determinations of the space
within it, one cannot refrain from ascribing a nature to this geometrical
thing. Thus, in particular, two lines that intersect each other and also the
∗
Crusius 57 alone knew of a middle way: namely that a spirit who can neither
err nor deceive originally implanted these natural laws in us. But, since false
principles are often mixed in as well – of which this man’s system itself provides
not a few examples – then, with the lack of sure criteria for distinguishing an
authentic origin from a spurious one, the use of such a principle looks very
precarious, since one can never know for sure what the spirit of truth or the
father of lies may have put into us.
112Prolegomena to any future metaphysics
circle, 58 however they happen to be drawn, nonetheless always partition
each other in a regular manner such that the rectangle from the parts of
one line is equal to that from the other. Now I ask: “Does this law lie
in the circle, or does it lie in the understanding?” i.e., does this ﬁgure,
independent of the understanding, contain the basis for this law in itself,
or does the understanding, since it has itself constructed the ﬁgure in ac-
cordance with its concepts (namely, the equality of the radii), at the same
time insert into it the law that chords cut one another in geometrical
proportion? If one traces the proofs of this law, one soon sees that it can
be derived only from the condition on which the understanding based
the construction of this ﬁgure, namely, the equality of the radii. If we
now expand upon this concept so as to follow up still further the unity
of the manifold properties of geometrical ﬁgures under common laws,
and we consider the circle as a conic section, which is therefore sub-
ject to the very same fundamental conditions of construction as other
conic sections, we then ﬁnd that all chords that intersect within these
latter (within the ellipse, the parabola, and the hyperbola) always do so
in such a way that the rectangles from their parts are a not indeed equal,
but always stand to one another in equal proportions. If from there we
go still further, namely to the fundamental doctrines of physical astron-
omy, there appears a physical law of reciprocal attraction, extending to
all material nature, the rule of which is that these attractions b decrease
inversely with the square of the distance from each point of attraction,
exactly as the spherical surfaces into which this force spreads itself in-
crease, something that seems to reside as necessary in the nature of the
things themselves and which therefore is customarily presented as cog-
nizable a priori. As simple as are the sources of this law – in that they
rest merely on the relation of spherical surfaces c with different radii –
the consequence therefrom is nonetheless so excellent with respect to
the variety and regularity of its agreement that not only does it follow
that all possible orbits of the celestial bodies are conic sections, but
also that their mutual relations are such that no other law of attrac-
tion save that of the inverse square of the distances can be conceived as
suitable for a system of the world.
Here then is nature that rests on laws that the understanding cognizes
a priori, and indeed chieﬂy from universal principles d of the determina-
tion of space. Now I ask: do these laws of nature lie in space, and does
the understanding learn them in that it merely seeks to investigate the
wealth of meaning that lies in space, or do they lie in the understanding
a
b
c
d
Adding sind, with Ak.
Reading sie as plural, with the singular antecedent Attraction, with Vorländer.
Reading Kugelﬂächen for Kugelﬂäche, with Vorländer.
allgemeinen Principien
113
4: 321Prolegomena to any future metaphysics
4: 322
and in the way in which it determines space in accordance with the con-
ditions of the synthetic unity toward which its concepts are one and all
directed? Space is something so uniform, and so indeterminate with re-
spect to all speciﬁc properties, that certainly no one will look for a stock
of natural laws within it. By contrast, that which determines space into
the ﬁgure of a circle, a cone, or a sphere is the understanding, insofar
as it contains the basis for the unity of the construction of these ﬁgures.
The mere universal form of intuition called space is therefore certainly
the substratum of all intuitions determinable upon particular objects,
and, admittedly, the condition for the possibility and variety of those
intuitions lies in this space; but the unity of the objects is determined
solely through the understanding, and indeed according to conditions
that reside in its own nature; and so the understanding is the origin of
the universal order of nature; in that it comprehends all appearances un-
der its own laws and thereby ﬁrst brings about experience a priori (with
respect to its form), in virtue of which everything that is to be cognized
only through experience is necessarily subject to its laws. For we are
not concerned with the nature of the things in themselves, which is in-
dependent of the conditions of both our senses and understanding, but
with nature as an object of possible experience, and here the understand-
ing, since it makes experience possible, at the same time makes it that
the sensible world is either not an object of experience at all, or else is
nature.
§ 39
Appendix to pure natural science
On the system of categories
Nothing can be more desirable to a philosopher than to be able to de-
rive, a priori from one principle, a the multiplicity of concepts or basic
principles b that previously had exhibited themselves to him piecemeal
in the use he had made of them in concreto, and in this way to be able to
unite them all in one cognition. Previously, he believed simply that what
was left to him after a certain abstraction, and that appeared, through
mutual comparison, to form a distinct kind of cognitions, had been com-
pletely assembled: but this was only an aggregate; now he knows that
only precisely so many, not more, not fewer, can constitute this c kind of
cognition, and he has understood the necessity of his division: this is a
comprehending, d and only now does he have a system.
a
b
c
d
Princip
Grundsätze
Reading diese for die, as suggested in Vorländer.
ein Begreifen; contrasted with the “comparison” mentioned earlier.
114Prolegomena to any future metaphysics
To pick out from ordinary cognition the concepts that are not based
on any particular experience and yet are present in all cognition from
experience (for which they constitute as it were the mere form of con-
nection) required no greater reﬂection or more insight than to cull from
a language rules for the actual use of words in general, and so to com-
pile the elements for a grammar (and in fact both investigations are very
closely related to one another) without, for all that, even being able to
give a reason why any given language should have precisely this and no
other formal constitution, and still less why precisely so many, neither
more nor fewer, of such formal determinations of the language can be
found at all.
Aristotle had compiled ten such pure elementary concepts under the
name of categories. ∗ To these, which were also called predicaments, he
later felt compelled to append ﬁve post-predicaments, ∗∗ some of which
(like prius, simul, motus) are indeed already found in the former; but this
rhapsody 59 could better pass for, and be deserving of praise as, a hint for
future inquirers than as an idea worked out according to rules, and so
with the greater enlightenment of philosophy it too has been rejected as
completely useless.
During an investigation of the pure elements of human cognition
(containing nothing empirical), I was ﬁrst of all able after long reﬂection
to distinguish and separate with reliability the pure elementary concepts
of sensibility (space and time) from those of the understanding. By this
means the seventh, eighth, and ninth categories were now excluded from
the above list. The others could be of no use to me, because no principle c
was available whereby the understanding could be fully surveyed and all
of its functions, from which its pure concepts arise, determined exhaus-
tively and with precision.
In order, however, to discover such a principle, d I cast about for an
act of the understanding that contains all the rest and that differentiates
itself only through various modiﬁcations or moments in order to bring
the multiplicity of representation under the unity of thinking in general;
and there I found that this act of the understanding consists in judg-
ing. Here lay before me now, already ﬁnished though not yet wholly
free of defects, the work of the logicians, through which I was put in
∗
1. Substantia. 2. Qualitas. 3. Quantitas. 4. Relatio. 5. Actio. 6. Passio. 7. Quando.
8. Ubi. 9. Situs. 10. Habitus. a
∗ ∗ Oppositum, Prius, Simul, Motus, Habere. b
a
b
c
d
Substance, quality, quantity, relation, action, affection, time, place, position, state.
Opposition, priority, simultaneity, motion, possession.
Princip
Princip
115
4: 323Prolegomena to any future metaphysics
4: 324
4: 325
the position to present a complete table of pure functions of the un-
derstanding, which were however undetermined with respect to every
object. Finally, I related these functions of judging to objects in gen-
eral, or rather to the condition for determining judgments as objectively
valid, and there arose pure concepts of the understanding, about which I
could have no doubt that precisely these only, and of them only so many,
neither more nor fewer, can make up our entire cognition of things out
of the bare understanding. As was proper, I called them categories, af-
ter their ancient name, whereby I reserved for myself to append in full,
under the name of predicables, all the concepts derivable from them –
whether by connecting them with one another, or with the pure form of
appearance (space and time) or its matter, provided the latter is not yet
determined empirically (the object of sensation in general) – just as soon
as a system of transcendental philosophy should be achieved, on behalf of
which I had, at the time, been concerned only with the critique of reason
itself.
The essential thing, however, in this system of categories, by which
it is distinguished from that ancient rhapsody (which proceeded without
any principle), a and in virtue of which it alone deserves to be counted
as philosophy, consists in this: that through it b the true signiﬁcation of
the pure concepts of the understanding and the condition of their use
could be exactly determined. For here it became apparent that the pure
concepts of the understanding are, of themselves, nothing but logical
functions, but that as such they do not constitute the least concept of an
object in itself but rather need sensory intuition as a basis, and even then
they serve only to determine empirical judgments, which are otherwise
undetermined and indifferent with respect to all the functions of judging,
with respect to those functions, so as to procure universal validity for
them, and thereby to make judgments of experience possible in general.
This sort of insight into the nature of the categories, which would
at the same time restrict their use merely to experience, never occurred
to their ﬁrst originator, or to anyone after him; but without this insight
(which depends precisely on their derivation or deduction), they are
completely useless and are a paltry list of names, without explanation
or rule for their use. Had anything like it ever occurred to the ancients,
then without doubt the entire study of cognition through pure reason,
which under the name of metaphysics has ruined so many good minds
over the centuries, would have come down to us in a completely different
form and would have enlightened the human understanding, instead of,
as has actually happened, exhausting it in murky and vain ruminations
and making it unserviceable for true science.
a
b
Princip
Reading desselben for derselben, with Ak.
116Prolegomena to any future metaphysics
This system of categories now makes all treatment of any object of
pure reason itself systematic in turn, and it yields an undoubted instruc-
tion or guiding thread as to how and through what points of inquiry any
metaphysical contemplation must be directed if it is to be complete; for
it exhausts all moments of the understanding, under which every other
concept must be brought. Thus too has arisen the table of principles,
of whose completeness we can be assured only through the system of
categories; and even in the division of concepts that are supposed to go
beyond the physiological use of the understanding (Critique, p. 344, also
p. 415), 60 there is always the same guiding thread, which, since it always
must be taken through the same ﬁxed points determined a priori in the
human understanding, forms a closed circle every time, leaving no room
for doubt that the object of a pure concept of the understanding or rea-
son, insofar as it is to be examined philosophically and according to a
priori principles, can be cognized completely in this way. I have not even
been able to refrain from making use of this guide with respect to one
of the most abstract of ontological classiﬁcations, namely the manifold
differentiation of the concepts of something and nothing, and accordingly
from achieving a rule-governed and necessary table (Critique, p. 292). ∗61
This very system, like every true system founded on a universal
principle, c also exhibits its inestimable usefulness in that it expels all the
∗ All sorts of nice notes can be made on a laid-out table of categories, such as:
1. that the third arises from the ﬁrst and second, conjoined into one concept,
2. that in those for quantity and quality there is merely a progression a from
Unity to Totality, or from something to nothing (for this purpose the categories
of quality must stand thus: Reality, Limitation, full Negation), without correlata
or opposita, while those of relation and modality carry the latter with them,
3. that, just as in the logical table, categorical judgments underlie all the others,
so the category of substance underlines all concepts of real things, 4. that,
just as modality in a judgment is not a separate predicate, so too the modal
concepts b do not add a determination to things, and so on. Considerations
such as these all have their great utility. If beyond this all the predicables are
enumerated – they can be extracted fairly completely from any good ontology
(e.g., Baumgarten’s) 62 – and if they are ordered in classes under the categories
(in which one must not neglect to add as complete an analysis as possible of all
these concepts), then a solely analytical part of metaphysics will arise, which
as yet contains no synthetic proposition whatsoever and could precede the
second (synthetic) part, and, through its determinateness and completeness,
might have not only utility, but beyond that, in virtue of its systematicity, a
certain beauty. 63
a Reading fortgehe for forgehen, with Vorländer.
Reading Modalbegriffe for Modelbegriffe, with Ak.
Princip
b
c
117
4: 326Prolegomena to any future metaphysics
extraneous concepts that might otherwise creep in among these pure
concepts of the understanding, and it assigns each cognition its place.
Those concepts that, under the name of concepts of reﬂection, I had also
put into a table under the guidance of the categories mingle in ontology
with the pure concepts of the understanding without privilege and legit-
imate claims, although the latter are concepts of connection and thereby
of the object itself, whereas the former are only concepts of the mere
comparison of already given concepts, and therefore have an entirely
different nature and use; through my law-governed division (Critique,
p. 260) 64 they are extricated from this amalgam. But the usefulness of
this separated table of categories shines forth yet more brightly if, as
will soon be done, we separate from the categories the table of transcen-
dental concepts of reason, which have a completely different nature and
origin than the concepts of the understanding (so that the table must
also have a different form), a separation that, necessary as it is, has never
occurred in any system of metaphysics, as a result of which a these ideas
of reason and concepts of the understanding run confusedly together as
if they belonged to one family, like siblings, an intermingling that also
could never have been avoided in the absence of a separate system of
categories.
a
Adding wo daher, with Ak.
118Main Transcendental Question,
Third Part
How is metaphysics in general possible?
§ 40
Pure mathematics and pure natural science would not have needed, for
the purpose of their own security and certainty, a deduction of the sort
that we have hitherto accomplished for them both; for the ﬁrst is sup-
ported by its own evidence, whereas the second, though arising from
pure sources of the understanding, is nonetheless supported from ex-
perience and thoroughgoing conﬁrmation by it – experience being a
witness that natural science cannot fully renounce and dispense with,
because, as philosophy, 65 despite all its certainty it can never rival math-
ematics. Neither science had need of the aforementioned investigation
for itself, but for another science, namely metaphysics.
Apart from concepts of nature, which always ﬁnd their application
in experience, metaphysics is further concerned with pure concepts
of reason that are never given in any possible experience whatsoever,
hence with concepts whose objective reality (that they are not mere
fantasies) and with assertions whose truth or falsity cannot be conﬁrmed
or exposed by any experience; and this part of metaphysics is moreover
precisely that which forms its essential end, toward which all the rest
is only a means – and so this science needs such a deduction for its own
sake. The third question, now put before us, therefore concerns as it
were the core and the characteristic feature of metaphysics, namely,
the preoccupation of reason simply with itself, and that acquaintance a
with objects which is presumed to arise immediately from reason’s
brooding over its own concepts without its either needing mediation
from experience for such an acquaintance, or being able to achieve such
an acquaintance through experience at all. ∗
∗ If it can be said that a science is actual at least in the thought b of all humankind
from the moment it has been determined that the problems which lead to it are
set before everyone by the nature of human reason, and therefore that many (if
faulty) attempts at those problems are always inevitable, it will also have to be
said: Metaphysics is subjectively actual (and necessarily so); and then we will
rightly ask: How is it (objectively) possible?
a Bekanntschaft
Idee
b
119
4: 327Prolegomena to any future metaphysics
4: 328
Without a solution to this question, reason will never be satisﬁed
with itself. The use in experience to which reason limits the pure under-
standing does not entirely fulﬁll reason’s own vocation. Each individual
experience is only a part of the whole a sphere of the domain of expe-
rience, but the absolute totality b of all possible experience is not itself an
experience, and yet is still a necessary problem for reason, for the mere
representation of which reason needs concepts entirely different from
the pure concepts of the understanding, whose use is only immanent, i.e.,
refers to experience insofar as such experience can be given, whereas the
concepts of reason extend to the completeness, i.e., the collective unity
of the whole of possible experience, and in that way exceed any given
experience and become transcendent.
Hence, just as the understanding needed the categories for experience,
reason contains in itself the basis for ideas, by which I mean necessary
concepts whose object nevertheless cannot be given in any experience.
The latter are just as intrinsic to the nature of reason as are the former
to that of the understanding; and if the ideas carry with them an illusion
that can easily mislead, this illusion is unavoidable, although it can very
well be prevented “from leading us astray.”
Since all illusion consists in taking the subjective basis for a judgment
to be objective, pure reason’s knowledge of itself in its transcendent
(overreaching) c use will be the only prevention against the errors into
which reason falls if it misconstrues its vocation and, in transcendent
fashion, refers to the object in itself that which concerns only its own
subject and the guidance of that subject in every use that is immanent.
§ 41
4: 329
The distinction of ideas, i.e., of pure concepts of reason, from categories,
or pure concepts of the understanding, as cognitions of completely dif-
ferent type, origin, and use, is so important a piece of the foundation of
a science which is to contain a system of all these cognitions a priori that,
without such a division, metaphysics is utterly impossible, or at best is
a disorderly and bungling endeavor to patch together a house of cards,
without knowledge of the materials with which one is preoccupied and
of their suitability for one or another end. If the Critique of Pure Reason
had done nothing but ﬁrst point out this distinction, it would thereby
have already contributed more to elucidating d our conception of, and to
guiding inquiry in, the ﬁeld of metaphysics, than have all the fruitless
a
b
c
d
ganzen
absolute Ganze
überschwenglichen
Aufklärung
120Prolegomena to any future metaphysics
efforts undertaken previously to satisfy the transcendent problems of
pure reason, without it ever being imagined that one may have been sit-
uated in a completely different ﬁeld from that of the understanding, and
as a result was listing the concepts of the understanding together with
those of reason as if they were of the same kind.
§ 42
All the pure cognitions of the understanding are such that their concepts
can be given in experience and their principles conﬁrmed through expe-
rience; by contrast, the transcendent cognitions of reason neither allow
what relates to their ideas to be given in experience, nor their theses a ever
to be conﬁrmed or refuted through experience; hence, only pure reason
itself can detect the error that perhaps creeps into them, though this is
very hard to do, because this selfsame reason by nature becomes dialec-
tical through its ideas, and this unavoidable illusion cannot be kept in
check through any objective and dogmatic investigation b of things, but
only through a subjective investigation of reason itself, as a source c of
ideas.
§ 43
In the Critique I always gave my greatest attention not only to how I
could distinguish carefully the types of cognition, but also to how
I could derive all d the concepts belonging to each type from their
common source, so that I might not only, by learning their origin, be
able to determine their use with certainty, but also have the inestimable
advantage (never yet imagined) of cognizing a priori, hence according
to principles, e the completeness of the enumeration, classiﬁcation, and
speciﬁcation of the concepts. Failing this, everything in metaphysics
is nothing but rhapsody, in which one never knows whether what one
has is enough, or whether and where something may still be lacking.
Such an advantage is, of course, available only in pure philosophy, but
it constitutes the essence of that philosophy.
Since I had found the origin of the categories in the four logical
functions of all judgments of the understanding, it was completely
natural to look for the origin of the ideas in the three functions of
syllogisms; f for once such pure concepts of reason (transc. ideas) have
a
b
c
d
e
f
Sätze
Reading Untersuchung for Untersuchungen, as suggested in Vorländer.
Reading eines Quells for einem Quell, with Ak.
Reading alle for allein, with Ak.
Principien
Vernunftschlüsse; literally, “inferences of reason.”
121
4: 330Prolegomena to any future metaphysics
been granted; then, if they are not to be taken for innate, they could
indeed be found nowhere else except in this very act of reason which,
insofar as it relates merely to form, constitutes the logical in syllogisms,
but, insofar as it represents the judgments of the understanding as
determined with respect to one or another a priori form, constitutes the
transcendental concepts of pure reason.
The formal distinction of syllogisms necessitates their division into
categorical, hypothetical, and disjunctive. Therefore the concepts of rea-
son based thereupon contain ﬁrst, the idea of the complete subject (the
substantial), second, the idea of the complete series of conditions, and
third, the determination of all concepts in the idea of a complete sum
total of the possible. ∗ The ﬁrst idea was psychological, c the second cos-
mological, the third theological; and since all three give rise to a dialectic,
but each in its own way, all this provided the basis for dividing the entire
dialectic of pure reason into the paralogism, the antinomy, and ﬁnally
the ideal of pure reason – through which derivation it is rendered com-
pletely certain that all claims of pure reason are represented here in full,
and not one can be missing, since the faculty of reason itself, whence
they all originate, is thereby fully surveyed.
§ 44
4: 331
In this examination it is in general further noteworthy: that the ideas
of reason d are not, like the categories, helpful to us in some way in us-
ing the understanding with respect to experience, but are completely
dispensable with respect to such use, nay, are contrary to and obstruc-
tive of the maxims for the cognition of nature through reason, although
they are still quite necessary in another respect, yet to be determined. 66
∗ In disjunctive judgments we consider all possibility as divided with respect to
a certain concept. The ontological principle a of the thoroughgoing determi-
nation of a thing in general (out of all possible opposing predicates, one is
attributed to each thing), which is at the same time the principle of all disjunc-
tive judgments, founds itself upon the sum total of all possibility, in which the
possibility of each thing in general is taken to be determinable. b The follow-
ing helps provide a small elucidation of the above proposition: That the act of
reason in disjunctive syllogisms is the same in form with that by which reason
achieves the idea of a sum total of all reality, which contains in itself the positive
members of all opposing predicates.
a Princip
Reading bestimmbar for bestimmter, with Ak.
Reading psychologisch for physiologisch, with Ak.
Reading Vernunftideen for Vernunftidee, with Ak.
b
c
d
122Prolegomena to any future metaphysics
In explaining the appearances of the soul, we can be completely indif-
ferent as to whether it is a simple substance or not; for we are unable
through any possible experience to make the concept of a simple being
sensorily intelligible, hence intelligible in concreto; and this concept is
therefore completely empty with respect to all hoped-for insight into
the cause of the appearances, and cannot serve as a principle a of expla-
nation of that which supplies inner or outer experience. Just as little can
the cosmological ideas of the beginning of the world or the eternity of
the world (a parte ante) b help us to explain any event in the world itself.
Finally, in accordance with a correct maxim of natural philosophy, we
must refrain from all explanations of the organization of nature drawn
from the will of a supreme being, c because this is no longer natural phi-
losophy but an admission that we have come to the end of it. These
ideas therefore have a completely different determination of their use
from that of the categories, through which (and through the principles
built upon them) experience itself ﬁrst became possible. Nevertheless
our laborious analytic of the understanding 67 would have been entirely
superﬂuous, if our aim had been directed toward nothing other than
mere cognition of nature insofar as such cognition can be given in ex-
perience; for reason conducts its affairs in both mathematics and natural
science quite safely and quite well, even without any such subtle de-
duction; hence our critique of the understanding joins with the ideas of
pure reason for a purpose that lies beyond the use of the understanding
in experience, though we have said above that the use of the under-
standing in this regard is wholly impossible and without object or sig-
niﬁcance. There must nonetheless be agreement between what belongs
to the nature of reason and of the understanding, and the former must
contribute to the perfection of the latter and cannot possibly confuse
it. 68
The solution to this question is as follows: Pure reason does not,
among its ideas, have in view particular objects that might lie beyond
the ﬁeld of experience, but it merely demands completeness in the use of
the understanding in the connection of experience. This completeness
can, however, only be a completeness of principles, d but not of intuitions
and objects. Nonetheless, in order to represent these principles deter-
minately, reason conceives of them as the cognition of an object, cog-
nition of which is completely determined with respect to these rules –
though the object is only an idea – so as to bring cognition through
a
b
c
d
Princip
“up until now,” literally, “on the side of the previous.”
höchsten Wesens
Principien
123
4: 332Prolegomena to any future metaphysics
the understanding as close as possible to the completeness that this idea
signiﬁes.
§ 45
Preliminary Remark
On the Dialectic of Pure Reason
4: 333
We have shown above (§§33, 34): that the purity of the categories from
all admixture with sensory determinations can mislead reason into ex-
tending their use entirely beyond all experience to things in themselves;
and yet, because the categories are themselves unable to ﬁnd any in-
tuition that could provide them with signiﬁcance and sense in concreto,
they cannot in and of themselves provide any determinate concept of
anything at all, though they can indeed, as mere logical functions, rep-
resent a thing in general. Now hyperbolical objects of this kind are what
are called noumena or pure beings of the understanding (better: beings
of thought) a – such as, e.g., substance, but which is thought without persis-
tence in time, or a cause, which would however not act in time, and so on –
because such predicates are attributed to these objects as serve only to
make the lawfulness of experience possible, and yet they are nonetheless
deprived of all the conditions of intuition under which alone experi-
ence is possible, as a result of which the above concepts again lose all
signiﬁcance.
There is, however, no danger that the understanding will of itself
wantonly stray beyond its boundaries into the ﬁeld of mere beings of
thought, without being urged by alien laws. But if reason, which can
never be fully satisﬁed with any use of the rules of the understanding in
experience because such use is always still conditioned, requires com-
pletion of this chain of conditions, then the understanding is driven out
of its circle, in order partly to represent the objects of experience in a
series stretching so far that no experience can comprise the likes of it,
partly (in order to complete the series) even to look for noumena entirely
outside said experience to which reason can attach the chain and in that
way, independent at last of the conditions of experience, nonetheless can
make its hold complete. These then are the transcendental ideas, which,
although in accordance with the true but hidden end of the natural deter-
mination of our reason they may be aimed not at overreaching concepts
but merely at the unbounded expansion of the use of concepts in experi-
ence, may nonetheless, through an unavoidable illusion, elicit from the
understanding a transcendent use, which, though deceitful, nonetheless
cannot be curbed by any resolve to stay within the bounds of experience,
but only through scientiﬁc instruction and hard work.
a
Gedankenwesen, contrasted with the just previous Verstandeswesen.
124Prolegomena to any future metaphysics
§ 46
I. Psychological ideas (Critique, pp. 341ff.) 69
It has long been observed that in all substances the true subject – namely
that which remains after all accidents (as predicates) have been removed –
and hence the substantial itself, is unknown to us; and various complaints
have been made about these limits to our insight. But it needs to be said
that human understanding is not to be blamed because it does not know
the substantial in things, i.e., cannot determine it by itself, but rather
because it wants to cognize determinately, like an object that is given,
what is only an idea. Pure reason demands that for each predicate of a
thing we should seek its appropriate subject, but that for this subject,
which is in turn necessarily only a predicate, we should seek its subject
again, and so forth to inﬁnity (or as far as we get). But from this it
follows that we should take nothing that we can attain for a ﬁnal subject,
and that the substantial itself could never be thought by our ever-so-
deeply penetrating understanding, even if the whole of nature were laid
bare before it; for the speciﬁc nature of our understanding consists in
thinking everything discursively, i.e., through concepts, hence through
mere predicates, among which the absolute subject must therefore always
be absent. Consequently, all real properties by which we cognize bodies
are mere accidents for which we lack a subject – even impenetrability,
which must always be conceived only as the effect of a force.
Now it does appear as if we have something substantial in the con-
sciousness of ourselves (i.e., in the thinking subject), and indeed have it
in immediate intuition; for all the predicates of inner sense are referred
to the I as subject, and this I cannot again be thought as the predicate of
some other subject. It therefore appears that in this case completeness in
referring the given concepts to a subject as predicates is not a mere idea,
but that the object, namely the absolute subject itself, is given in experi-
ence. But this expectation is disappointed. For the I is not a concept ∗ at
all, but only a designation of the object of inner sense insofar as we do
not further cognize it through any predicate; hence although it cannot
itself be the predicate of any other thing, just as little can it be a deter-
minate concept of an absolute subject, but as in all the other cases it can
only be the referring of inner appearances to their unknown subject.
∗ If the representation of apperception, the I, were a concept through which
anything might be thought, it could then be used as a predicate for other
things, or contain such predicates in itself. But it is nothing more than a feeling
of an existence without the least concept, and is only a representation of that
to which all thinking stands in relation (relatione accidentis). a
a “relation of accident”
125
4: 334Prolegomena to any future metaphysics
Nevertheless, through a wholly natural misunderstanding, this idea
(which, as a regulative principle, serves perfectly well to destroy com-
pletely all materialistic explanations of the inner appearances of our
soul) a gives rise to a seemingly plausible argument for inferring the
nature of our thinking being from this presumed cognition of the sub-
stantial in it, inasmuch as knowledge of its nature falls completely outside
the sum total of experience.
§ 47
4: 335
This thinking self (the soul), as the ultimate subject of thinking, which
cannot itself be represented as the predicate of another thing, may now
indeed be called substance: but this concept nonetheless remains com-
pletely empty and without any consequences, if persistence (as that which
renders the concept of substances fertile within experience) cannot be
proven of it.
Persistence, however, can never be proven from the concept of a
substance as a thing in itself, but only for the purposes of experience.
This has been sufﬁciently established in the ﬁrst Analogy of Experience
(Critique, p. 182); 70 and anyone who will not grant this proof can test for
themselves whether they succeed in proving, from the concept of a sub-
ject that does not exist as the predicate of another thing, that the existence
of that subject is persistent throughout, and that it can neither come into
being nor pass away, either in itself or through any natural cause. Syn-
thetic a priori propositions of this type can never be proven in themselves,
but only in relation to things as objects of a possible experience.
§ 48
If, therefore, we want to infer the persistence of the soul from the concept
of the soul as substance, this can be valid of the soul only for the purpose of
possible experience, and not of the soul as a thing in itself and beyond all
possible experience. But life is the subjective condition of all our possible
experience: consequently, only the persistence of the soul during life can
be inferred, for the death of a human being is the end of all experience
as far as the soul as an object of experience is concerned (provided that
the opposite has not been proven, which is the very matter in question).
Therefore the persistence of the soul can be proven only during the life
of a human being (which proof will doubtless be granted us), but not
after death (which is actually our concern) – and indeed then only from
the universal ground that the concept of substance, insofar as it is to
be considered as connected necessarily with the concept of persistence,
a
The original has an asterisk here, with no corresponding note.
126Prolegomena to any future metaphysics
can be so connected only in accordance with a principle of possible
experience, and hence only for the purpose of the latter. ∗
§ 49
That our outer perceptions not only do correspond to something real b
outside us, but must so correspond, also can never be proven as a connec-
tion of things in themselves, but can well be proven for the purpose of
experience. This is as much as to say: it can very well be proven that there
is something outside us of an empirical kind, and hence as appearance in
space; for we are not concerned with objects other than those which be-
long to a possible experience, just because such objects cannot be given to
us in any experience and therefore are nothing for us. Outside me empir-
ically is that which is intuited in space; and because this space, together
with all the appearances it contains, belongs to those representations
whose connection according to laws of experience proves their objective
truth, just as the connection of the appearances of the inner sense proves
the reality c of my soul (as an object of inner sense), it follows that I am,
by means of outer appearances, just as conscious of the reality of bodies
∗ It is in fact quite remarkable that metaphysicians have always slid so blithely
over the principle of the persistence of substances, without ever attempting to
prove it; doubtless because they found themselves completely forsaken by all
grounds of proof as soon as they commenced with the concept of substance.
Common sense, being well aware that without this assumption no uniﬁcation
of perceptions in an experience would be possible, made up for this defect
with a postulate; for it could never extract this principle from experience itself,
partly because experience cannot follow the materials (substances) through all
their alterations and dissolutions far enough to be able to ﬁnd matter always
undiminished, partly because the principle contains necessity, which is always
the sign of an a priori principle. a But the metaphysicians applied this principle
conﬁdently to the concept of the soul as a substance and inferred its neces-
sary continuation after the death of a human being (principally because the
simplicity of this substance, which had been inferred from the indivisibility
of consciousness, saved it from destruction through dissolution). Had they
found the true source of this principle, which however would have required
far deeper investigations than they ever wanted to start, then they would have
seen: that this law of the persistence of substances is granted only for the pur-
pose of experience and therefore can hold good only for things insofar as they
are to be cognized in experience and connected with other things, but never
for things irrespective of all possible experience, hence not for the soul after
death.
a Princips
Wirkliches
Wirklichkeit
b
c
127
4: 336Prolegomena to any future metaphysics
4: 337
as outer appearances in space, as I am, by means of inner experience,
conscious of the existence of my soul in time – which soul I cognize a
only as an object of inner sense through the appearances constituting an
inner state, and whose being as it is in itself, b which underlies these ap-
pearances, is unknown to me. Cartesian idealism therefore distinguishes
only outer experience from dream, and lawfulness as a criterion of the
truth of the former from the disorder and false illusion of the latter. c In
both cases it presupposes space and time as conditions for the existence
of objects and merely asks whether the objects of the outer senses are
actually d to be found in the space in which we put them while awake,
in the way that the object of inner sense, the soul, actually is in time,
i.e., whether experience carries with itself sure criteria to distinguish it
from imagination. Here the doubt can easily be removed, and we always
remove it in ordinary life by investigating the connection of appearances
in both space and time according to universal laws of experience, and
if the representation of outer things consistently agrees therewith, we
cannot doubt that those things should not constitute truthful experience.
Because appearances are considered as appearances only in accordance
with their connection within experience, material idealism can therefore
very easily be removed; and it is just as secure an experience that bodies
exist outside us (in space) as that I myself exist in accordance with the
representation of inner sense (in time) – for the concept: outside us, sig-
niﬁes only existence in space. Since, however, the I in the proposition
I am does not signify merely the object of inner intuition (in time) but
also the subject of consciousness, just as body does not signify merely
outer intuition (in space) but also the thing in itself that underlies this ap-
pearance, accordingly the question of whether bodies (as appearances of
outer sense) exist outside my thought as bodies in nature e can without hes-
itation be answered negatively; but here matters do not stand otherwise
for the question of whether I myself as an appearance of inner sense (the
soul according to empirical psychology) exist in time outside my power
of representation, for this question must also be answered negatively. In
this way everything is, when reduced to its true signiﬁcation, conclusive
and certain. Formal idealism (elsewhere called transcendental idealism
by me) actually destroys f material or Cartesian idealism. For if space is
nothing but a form of my sensibility, then it is, as a representation in me,
a
b
c
d
e
f
Reading erkenne for erkennen, with Vorländer.
Wesen an sich selbst
der letzteren, plural for dreams; rejecting the emendation to des letztern in Ak.
wirklich
Placing in der Natur after ausser meinen Gedanken, with Vorländer, as opposed to its
original placement after ohne Bedenken.
aufhebt
128Prolegomena to any future metaphysics
just as real a as I am myself, and the only question remaining concerns
the empirical truth of the appearances in this space. If this is not the case,
but rather space and the appearances in it are something existing outside
us, then all the criteria of experience can never, outside our perception,
prove the reality of these objects outside us.
§ 50
II. Cosmological ideas (Critique, pp. 405 ff.) 71
This product of pure reason in its transcendent use is its most remarkable
phenomenon, and it works the most strongly of all to awaken philosophy
from its dogmatic slumber, and to prompt it toward the difﬁcult business
of the critique of reason itself.
I call this idea cosmological because it always ﬁnds its object only in
the sensible world and needs no other world than that whose object b is
an object c for the senses, and so, thus far, is immanent d and not tran-
scendent, and therefore, up to this point, is not yet an idea; by contrast,
to think of the soul as a simple substance already amounts to thinking of
it as an object (the simple) the likes of which cannot be represented at all
to the senses. Notwithstanding all that, the cosmological idea expands
the connection of the conditioned with its condition (be it mathematical
or dynamic) so greatly that experience can never match it, and therefore
it is, with respect to this point, always an idea whose object can never be
adequately given in any experience whatever.
§ 51
In the ﬁrst place, the usefulness of a system of categories is here revealed
so clearly and unmistakably that even if there were no further grounds
of proof of that system, this alone would sufﬁciently establish their in-
dispensability in the system of pure reason. There are no more than four
such transcendental e ideas, as many as there are classes of categories; in
each of them, however, they refer only to the absolute completeness of
the series of conditions for a given conditioned. In accordance with these
cosmological ideas there are also only four kinds of dialectical assertions
of pure reason, which show themselves to be dialectical because for each
such assertion a contradictory one stands in opposition in accordance
a
b
c
d
e
wirklich
Gegenstand
Object
einheimisch
Reading transscendentalen for transscendenten, as suggested by Vorländer, and in accordance
with §§45, 55–7, and 60.
129
4: 338Prolegomena to any future metaphysics
4: 339
with equally plausible principles of pure reason, a conﬂict that cannot
be avoided by any metaphysical art of the most subtle distinctions, but
that requires the philosopher to return to the ﬁrst sources of pure reason
itself. This antinomy, by no means arbitrarily contrived, but grounded
in the nature of human reason and so unavoidable and neverending,
contains the following four theses together with their antitheses. a
1. b
Thesis
The world has, as to time and space,
a beginning (a boundary).
Antithesis
The world is, as to time and space,
inﬁnite.
2.
Thesis
Everything in the world
is constituted out of the
simple. 3.
Thesis
There exist in the world
causes through
freedom.
Antithesis
There is nothing simple,
but everything is
composite. Antithesis
There is no freedom,
but everything is
nature.
4.
Thesis
In the series of causes in the world there is a
necessary being.
Antithesis
There is nothing necessary in this series, but in it
everything is contingent.
§ 52a
4: 340
Here is now the strangest phenomenon of human reason, no other ex-
ample of which can be pointed to in any of its other uses. If (as normally
happens) we think of the appearances of the sensible world as things in
themselves, if we take the principles of their connection to be princi-
ples that are universally valid for things in themselves and not merely
for experience (as is just as common, nay, is unavoidable without our
a
b
Sätze and Gegensätzen; in the Kritik, assertion and counterassertion are labelled These and
Antithese, terms that are used below, §§52b, 52c.
Stress added to the last line of the thesis and antithesis, to parallel the typography in 2,
3, and 4.
130Prolegomena to any future metaphysics
Critique): then an unexpected conﬂict comes to light, which can never
be settled in the usual dogmatic manner, since both thesis and antithe-
sis can be established through equally evident, clear, and incontestable
proofs – for I will vouch for the correctness of all these proofs – and
therefore reason is seen to be divided against itself, a situation that makes
the skeptic rejoice, but must make the critical philosopher pensive and
uneasy.
§ 52b
One can tinker around with metaphysics in sundry ways without even
suspecting that one might be venturing into untruth. For if only we do
not contradict ourselves – something that is indeed entirely possible with
synthetic, though completely fanciful, propositions – then we can never
be refuted by experience in all such cases where the concepts we connect
are mere ideas, which can by no means be given (in their entire content) in
experience. For how would we decide through experience: Whether the
world has existed from eternity, or has a beginning? Whether matter is
inﬁnitely divisible, or is constituted out of simple parts? Concepts such as
these cannot be given in any experience (even the greatest possible), and
so the falsity of the afﬁrmative or negative thesis cannot be discovered
through that touchstone.
The single possible case in which reason would reveal (against its
will) its secret dialectic (which it falsely passes off as dogmatics) would
be that in which it based an assertion on a universally a acknowledged
principle, and, with the greatest propriety in the mode of inference,
derived the direct opposite from another equally accredited principle.
Now this case is here actual, and indeed is so with respect to four natural
ideas of reason, from which there arise – each with proper consistency
and from universally acknowledged principles – four assertions on one
side and just as many counterassertions on the other, thereby revealing
the dialectical illusion of pure reason in the use of these principles, which
otherwise would have had to remain forever hidden.
Here is, therefore, a decisive test, which must necessarily disclose to us
a fault that lies hidden in the presuppositions of reason. ∗ Of two mutually
∗
I therefore desire that the critical reader concern himself mainly with this
antinomy, because nature itself seems to have set it up to make reason suspicious
in its bold claims and to force a self-examination. I promise to answer for each
proof I have given of both thesis b and antithesis, c and thereby to establish the
certainty of the inevitable antinomy of reason. If the reader is induced, through
this strange phenomenon, to reexamine the presupposition that underlies it,
he will then feel constrained to investigate more deeply with me the primary
foundation of all cognition through pure reason.
131
4: 341Prolegomena to any future metaphysics
contradictory propositions, both cannot be false save when the concept
underlying them both is itself contradictory; e.g., the two propositions:
a square circle is round, and: a square circle is not round, are both false.
For, as regards the ﬁrst, it is false that the aforementioned circle is round,
since it is square; but it is also false that it is not round, i.e., has corners,
since it is a circle. The logical mark of the impossibility of a concept
consists, then, in this: that under the presupposition of this concept,
two contradictory propositions would be false simultaneously; and since
between these two no third proposition can be thought, through this
concept nothing at all is thought.
§ 52c
4: 342
Now underlying the ﬁrst two antinomies, which I call mathematical
because they concern adding together or dividing up the homogeneous,
is a contradictory concept of this type; and by this means I explain how
it comes about that thesis and antithesis are false in both.
If I speak of objects in time and space, I am not speaking of things
in themselves (since I know nothing of them), but only of things in
appearance, i.e., of experience as a distinct mode of cognition of objects
that is granted to human beings alone. I must not say of that which I
think in space or time: that it is in itself in space and time, independent
of this thought of mine; for then I would contradict myself, since space
and time, together with the appearances in them, are nothing existing
in themselves and outside my representations, but are themselves only
modes of representation, and it is patently contradictory to say of a mere
mode of representation that it also exists outside our representation. The
objects of the senses therefore exist only in experience; by contrast, to
grant them a self-subsistent existence of their own, without experience
or prior to it, is as much as to imagine that experience is also real without
experience or prior to it.
Now if I ask about the magnitude of the world with respect to space
and time, for all of my concepts it is just as impossible to assert that it is
inﬁnite as that it is ﬁnite. For neither of these can be contained in expe-
rience, because it is not possible to have experience either of an inﬁnite
space or inﬁnitely ﬂowing time, or d of a bounding of the world by an
empty space or by an earlier, empty time; these are only ideas. There-
fore the magnitude of the world, determined one way or the other, must
a
b
c
d
Reading allgemein for allgemeinen, with Ak.
Thesis
Antithesis
Reading noch for nach, with Ak.
132Prolegomena to any future metaphysics
lie in itself, apart from all experience. But this contradicts the concept
of a sensible world, which is merely a sum total of appearance, whose
existence and connection takes place only in representation, namely in
experience, since it is not a thing in itself, a but is itself nothing but a
kind of representation. From this it follows that, since the concept of a
sensible world existing for itself is self-contradictory, any solution to this
problem as to its magnitude will always be false, whether the attempted
solution be afﬁrmative or negative.
The same holds for the second antinomy, which concerns dividing
up the appearances. For these appearances are mere representations,
and the parts exist only in the representation of them, hence in the
dividing, i.e., in a possible experience in which they are given, and the
dividing therefore proceeds only as far as possible experience reaches. To
assume that an appearance, e.g., of a body, contains within itself, before
all experience, all of the parts to which possible experience can ever attain,
means: to give to a mere appearance, which can exist only in experience,
at the same time an existence of its own previous to experience, which is to
say: that mere representations are present before they are encountered
in the representational power, which contradicts itself and hence also
contradicts every solution to this misunderstood problem, whether that
solution asserts that bodies in themselves consist of inﬁnitely many parts
or of a ﬁnite number of simple parts.
§ 53
In the ﬁrst (mathematical) class of antinomy, the falsity of the presup-
position consisted in the following: that something self-contradictory
(namely, appearance as a thing in itself) b would be represented as being
uniﬁable in a concept. But regarding the second, namely the dynamical,
class of antinomy, the falsity of the presupposition consists in this: that
something that is uniﬁable is represented as contradictory; consequently,
while in the ﬁrst case both of the mutually opposing assertions were false,
here on the contrary the assertions, which are set in opposition to one
another through mere misunderstanding, can both be true.
Speciﬁcally, mathematical combination necessarily presupposes the
homogeneity of the things combined (in the concept of magnitude), but
dynamical connection does not require this at all. If it is a question of
the magnitude of something extended, all parts must be homogeneous
among themselves and with the whole; by contrast, in the connection of
cause and effect homogeneity can indeed be found, but is not necessary;
a
b
Sache an sich
Sache an sich selbst
133
4: 343Prolegomena to any future metaphysics
4: 344
for the concept of causality (whereby through one thing, something
completely different from it is posited) at least does not require it.
If the objects of the sensible world were taken for things in themselves,
and the previously stated natural laws for laws of things in themselves,
contradiction would be unavoidable. In the same way, if the subject of
freedom were represented, like the other objects, as a mere appearance,
contradiction could again not be avoided, for the same thing would be
simultaneously afﬁrmed and denied of the same object in the same sense.
But if natural necessity is referred only to appearances and freedom only
to things in themselves, then no contradiction arises if both kinds of
causality are assumed or conceded equally, however difﬁcult or impos-
sible it may be to make causality of the latter kind conceivable.
Within appearance, every effect is an event, or something that hap-
pens in time; the effect must, in accordance with the universal law of
nature, be preceded by a determination of the causality of its cause (a
state of the cause), from which the effect follows in accordance with a
constant law. But this determination of the cause to causality must also
be something that occurs or takes place; the cause must have begun to act,
for otherwise no sequence in time could be thought between it and the
effect. Both the effect and the causality of the cause would have always ex-
isted. Therefore the determination of the cause to act must also have arisen
among the appearances, and so must, like its effect, be an event, which
again must have its cause, and so on, and hence natural necessity must be
the condition in accordance with which efﬁcient causes are determined.
Should, by contrast, freedom be a property of certain causes of appear-
ances, then that freedom must, in relation to the appearances as events,
be a faculty of starting those events from itself (sponte), a i.e., without the
causality of the cause itself having to begin, and hence without need for
any other ground to determine its beginning. But then the cause, as to its
causality, would not have to be subject to temporal determinations of its
state, i.e., would not have to be appearance at all, i.e., would have to be
taken for a thing in itself, and only the effects would have to be taken for
appearances. ∗ If this sort of inﬂuence of intelligible beings on appearances
∗
The idea of freedom has its place solely in the relation of the intellectual, b as
cause, to the appearance, as effect. Therefore we cannot bestow freedom upon
matter in consideration of the unceasing activity by which it ﬁlls its space,
even though this activity occurs through an inner principle. We can just as
little ﬁnd any concept of freedom to ﬁt a purely intelligible being, e.g., God,
insofar as his action is immanent. For his action, although independent of
causes determining it from outside, nevertheless is determined in his eternal
reason, hence in the divine nature. Only if something should begin through an
action, hence the effect be found in the time series, and so in the sensible
world (e.g., the beginning of the world), does the question arise of whether the
causality of the cause must itself also have a beginning, or whether the cause
134Prolegomena to any future metaphysics
can be thought without contradiction, then natural necessity will indeed
attach to every connection of cause and effect in the sensible world,
and yet that cause which is itself not an appearance (though it under-
lies appearance) will still be entitled to freedom, and therefore nature
and freedom will be attributable without contradiction to the very same
thing, but in different respects, in the one case as appearance, in the
other as a thing in itself.
We have in us a faculty that not only stands in connection with its
subjectively determining grounds, which are the natural causes of its
actions – and thus far is the faculty of a being which itself belongs to
appearances – but that also is related to objective grounds that are mere
ideas, insofar as these ideas can determine this faculty, a connection that
is expressed by ought. c This faculty is called reason, and insofar as we are
considering a being (the human being) solely as regards this objectively
determinable reason, this being cannot be considered as a being of the
senses; rather, the aforesaid property is the property of a thing in itself,
and the possibility of that property – namely, how the ought, which has
never yet happened, can determine the activity of this being and can be
the cause of actions whose effect is an appearance in the sensible world –
we cannot comprehend at all. Yet the casuality of reason with respect to
effects in the sensible world would nonetheless be freedom, insofar as
objective grounds, which are themselves ideas, are taken to be determining
with respect to that causality. For the action of that causality would in
that case not depend on any subjective, hence also not on any temporal
conditions, and would therefore also not depend on the natural law that
serves to determine those conditions, because grounds of reason provide
the rule for actions universally, from principles, without inﬂuence from
the circumstances of time or place.
What I adduce here counts only as an example, for intelligibility, and
does not belong necessarily to our question, which must be decided from
mere concepts independently of properties that we ﬁnd in the actual
world.
I can now say without contradiction: all actions of rational beings,
insofar as they are appearances (are encountered in some experience or
can originate an effect without its causality itself having a beginning. In the ﬁrst
case the concept of this causality is a concept of natural necessity, in the second
of freedom. From this the reader will see that, since I have explained freedom
as the faculty to begin an event by oneself, I have exactly hit that concept which
is the problem of metaphysics.
a
b
c
“spontaneously”
des Intellektuellen
Sollen
135Prolegomena to any future metaphysics
4: 346
other), are subject to natural necessity; but the very same actions, with
respect only to the rational subject and its faculty of acting in accordance
with bare reason, are free. What, then, is required for natural necessity?
Nothing more than the determinability of every event in the sensible
world according to constant laws, and therefore a relation to a cause
within appearance; whereby the underlying thing in itself and its causality
remain unknown. But I say: the law of nature remains, whether the rational
being be a cause of effects in the sensible world through reason and hence
through freedom, or whether that being does not determine such effects
through rational grounds. For if the ﬁrst is the case, the action takes place
according to maxims whose effect within appearance will always conform
to constant laws; if the second is the case, and the action does not take
place according to principles of reason, then it is subject to the empirical
laws of sensibility, and in both cases the effects are connected according
to constant laws; but we require nothing more for natural necessity, and
indeed know nothing more of it. In the ﬁrst case, however, reason is the
cause of these natural laws and is therefore free, in the second case the
effects ﬂow according to mere natural laws of sensibility, because reason
exercises no inﬂuence on them; but, because of this, reason is not itself
determined by sensibility (which is impossible), and it is therefore also
free in this case. Therefore freedom does not impede the natural law of
appearances, any more than this law interferes with the freedom of the
practical use of reason, a use that stands in connection with things in
themselves as determining grounds.
In this way practical freedom – namely, that freedom in which rea-
son has causality in accordance with objective determining grounds – is
rescued, without natural necessity suffering the least harm with respect
to the very same effects, as appearances. This can also help elucidate
what we have had to say about transcendental freedom and its uniﬁca-
tion with natural necessity (in the same subject, but not taken in one and
the same respect). For, as regards transcendental freedom, any beginning
of an action of a being out of objective causes is always, with respect to
these determining grounds, a ﬁrst beginning, although the same action
is, in the series of appearances, only a subalternate beginning, which has
to be preceded by a state of the cause which determines that cause, and
which is itself determined in the same way by an immediately preceding
cause: so that in rational beings (or in general in any beings, provided
that their causality is determined in them as things in themselves) one
can conceive of a faculty for beginning a series of states spontaneously,
without falling into contradiction with the laws of nature. For the re-
lation of an action to the objective grounds of reason is not a tempo-
ral relation; here, that which determines the causality does not precede
the action as regards time, because such determining grounds do not
represent the relation of objects to the senses (and so to causes within
136Prolegomena to any future metaphysics
appearance), but rather they represent determining causes as things in
themselves, which are not subject to temporal conditions. Hence the
action can be regarded as a ﬁrst beginning with respect to the causal-
ity of reason, but can nonetheless at the same time be seen as a mere
subordinated beginning with respect to the series of appearances, and
can without contradiction be considered in the former respect as free,
in the latter (since the action is mere appearance) as subject to natural
necessity.
As regards the fourth antinomy, it is removed in a similar a manner
as was the conﬂict of reason with itself in the third. For if only the
cause in the appearances is distinguished from the cause of the appearances
insofar as the latter cause can be thought as a thing in itself, then these
two propositions can very well exist side by side, as follows: that there
occurs no cause of the sensible world (in accordance with similar laws
of causality) whose existence is absolutely necessary, as also on the other
side: that this world is nonetheless connected with a necessary being
as its cause (but of another kind and according to another law) – the
inconsistency of these two propositions resting solely on the mistake of
extending what holds merely for appearances to things in themselves,
and in general of mixing the two of these up into one concept.
§ 54
This then is the statement and solution of the whole antinomy in which
reason ﬁnds itself entangled in the application of its principles b to the
sensible world, and of which the former (the mere statement) even by
itself would already be of considerable beneﬁt toward a knowledge c of
human reason, even if the solution of this conﬂict should not yet fully
satisfy d the reader, who has here to combat a natural illusion that has
only recently been presented to him as such, after he had hitherto always
taken that illusion for the truth. One consequence of all this is, indeed,
unavoidable; namely, that since it is completely impossible to escape from
this conﬂict of reason with itself as long as the objects of the sensible
world are taken for things in themselves e – and not for what they in fact
are, that is, for mere appearances – the reader is obliged, for that reason,
to take up once more the deduction of all our cognition a priori (and
the examination of that deduction which I have provided), in order to
come to a decision about it. For the present I do not require more; for
a
b
c
d
e
Reading ähnliche for die ähnliche, with Vorländer.
Principien
Kenntniss
Reading noch nicht völlig befriedigen for hiedurch noch nicht völlig befriedigt werden, with
Vorländer.
Sachen an sich selbst
137
4: 347Prolegomena to any future metaphysics
4: 348
if, through this pursuit, he has ﬁrst thought himself deeply enough into
the nature of pure reason, then the concepts by means of which alone
the solution to this conﬂict of reason is possible will already be familiar
to him, a circumstance without which I cannot expect full approbation
from even the most attentive reader.
§ 55
III. Theological idea (Critique, pp. 571ff.) 72
The third transcendental idea, which provides material for the most im-
portant among all the uses of reason – but one that, if pursued merely
speculatively, is overreaching (transcendent) and thereby dialectical – is
the ideal of pure reason. Here reason does not, as with the psychological
and the cosmological idea, start from experience and become seduced
by the ascending sequence of grounds into aspiring, if possible, to ab-
solute completeness in their series, but instead breaks off entirely from
experience and descends from bare concepts of what would constitute
the absolute completeness of a thing in general – and so by means of
the idea of a supremely perfect ﬁrst being a – to determination of the
possibility, hence the reality, of all other things; in consequence, here
the bare presupposition of a being that, although not in the series of
experiences, is nonetheless thought on behalf of experience, for the sake
of comprehensibility in the connection, ordering, and unity of that ex-
perience – i.e., the idea – is easier to distinguish from the concept of the
understanding than in the previous cases. Here therefore the dialectical
illusion, which arises from our taking the subjective conditions of our
thinking for objective conditions of things themselves b and our taking a
hypothesis that is necessary for the satisfaction of our reason for a dogma,
is easily exposed, and I therefore need mention nothing more about the
presumptions of transcendental theology, since what the Critique says
about them is clear, evident, and decisive.
§ 56
General Note
to
the Transcendental Ideas
4: 349
The objects that are given to us through experience are incomprehen-
sible to us in many respects, and there are many questions to which
natural law carries us, which, if pursued to a certain height (yet al-
ways in conformity with those laws) cannot be solved at all; e.g., how
a
b
höchst vollkommenen Urwesens
der Sachen selbst
138Prolegomena to any future metaphysics
pieces of matter attract one another. But if we completely abandon na-
ture, or transcend a all possible experience in advancing the connection
of nature and so lose ourselves in mere ideas, then we are unable to say
that the object b is incomprehensible to us and that the nature of things
presents us with unsolvable problems; for then we are not concerned with
nature or in general with objects c that are given, but merely with con-
cepts that have their origin solely in our reason, and with mere beings of
thought, with respect to which all problems, which must originate from
the concepts of those very beings, can be solved, since reason certainly
can and must be held fully accountable for its own proceedings. ∗ Be-
cause the psychological, d cosmological, and theological ideas are nothing
but pure concepts of reason, which cannot be given in any experience,
the questions that reason puts before us with respect to them are not
set for us through objects, but rather through mere maxims of reason
for the sake of its self-satisfaction, and these questions must one and
all be capable of sufﬁcient answer – which occurs by its being shown
that they are principles for bringing the use of our understanding into
thoroughgoing harmony, completeness, and synthetic unity, and to that
extent are valid only for experience, though in the totality of that expe-
rience. But although an absolute totality of experience is not possible,
nonetheless the idea of a totality of cognition according to principles
in general is what alone can provide it with a special kind of unity,
namely that of a system, without which unity our cognition is noth-
ing but piecework and cannot be used for the highest end (which is
nothing other than the system of all ends); and here I mean not only
the practical use of reason, but also the highest end of its speculative
use.
∗ Herr Platner in his Aphorisms therefore says with astuteness ( §§ 728–729): “If
reason is a criterion, then there cannot possibly be a concept that is incompre-
hensible to human reason. – Only in the actual does incomprehensibility have
a place. Here the incomprehensibility arises from the inadequacy of acquired
ideas.” 73 – It therefore only sounds paradoxical, and is otherwise not strange
to say: that in nature much is incomprehensible to us (e.g., the procreative fac-
ulty), but if we rise still higher and even go out beyond nature, then once again
all will be comprehensible to us; for then we entirely leave behind the objects
that can be given to us, and concern ourselves merely with ideas, with respect
to which we can very well comprehend the law that reason thereby prescribes
to the understanding for its use in experience, since that law is reason’s own
product.
a übersteigen
Gegenstand
Objecten
Reading psychologischen for physiologische, with Ak.
b
c
d
139
4: 350Prolegomena to any future metaphysics
Therefore the transcendental ideas express the peculiar vocation of
reason, namely to be a principle a of the systematic unity of the use of
the understanding. But if one looks upon this unity of mode of cognition
as if it were inhering in the object of cognition, if one takes that which
really is only regulative to be constitutive, and becomes convinced that by
means of these ideas one’s knowledge b can be expanded far beyond all
possible experience, hence can be expanded transcendently, even though
this unity serves only to bring experience in itself as near as possible to
completeness (i.e., to have its advance constrained by nothing that cannot
belong to experience), then this is a mere misunderstanding in judging
the true vocation of our reason and its principles, and it is a dialectic,
which partly confounds the use of reason in experience, and partly divides
reason against itself.
Conclusion
on
Determining the Boundary of Pure Reason
§ 57
4: 351
After the extremely clear proofs we have given above, it would be an
absurdity for us, with respect to any object, to hope to cognize more
than belongs to a possible experience of it, or for us, with respect to any
thing that we assume not to be an object of possible experience, to claim
even the least cognition for determining it according to its nature as it is
in itself; for by what means will we reach this determination, since time,
space, and all the concepts of the understanding, and especially the con-
cepts drawn from empirical intuition or perception in the sensible world,
do not and cannot have any other use than merely to make experience
possible, and if we relax this condition even for the pure concepts of the
understanding, they then determine no object whatsoever, and have no
signiﬁcance anywhere.
But, on the other hand, it would be an even greater absurdity for us
not to allow any things in themselves at all, or for us to want c to pass
off our experience for the only possible mode of cognition of things –
hence our intuition in space and time for the only possible intuition
and our discursive understanding for the archetype of every possible
understanding – and so to want to take principles d of the possibility of
experience for universal conditions on things in themselves.
a
b
c
d
Princips
Kenntniss
Reading wollten for wollte, with Ak.
Principien
140Prolegomena to any future metaphysics
Our principles, a which limit the use of reason to possible experi-
ence alone, could accordingly themselves become transcendent and could
pass off the limits of our reason for limits on the possibility of things
themselves (for which Hume’s Dialogues 74 can serve as an example), if
a painstaking critique did not both guard the boundaries of our reason
even with respect to its empirical use, and set a limit to its pretensions.
Skepticism originally arose from metaphysics and its unpoliced dialectic.
At ﬁrst this skepticism wanted, solely for the beneﬁt of the use of reason
in experience, to portray everything that surpasses this use as empty and
deceitful; but gradually, as it came to be noticed that it was the very same a
priori principles b which are employed in experience that, unnoticed, had
led still further than experience reaches – and had done so, as it seemed,
with the very same right – then even the principles of experience began
to be doubted. There was no real trouble with this, for sound common
sense c will always assert its rights in this domain; but there did arise a
particular confusion in science, which cannot determine how far (and
why only that far and not further) reason is to be trusted, and this con-
fusion can be remedied and all future relapses prevented only through
a formal determination, derived from principles, of the boundaries for
the use of our reason.
It is true: we cannot provide, beyond all possible experience, any
determinate concept of what things in themselves may be. But we are
nevertheless not free to hold back entirely in the face of inquiries about
those things; for experience never fully satisﬁes reason; it directs us ever
further back in answering questions and leaves us unsatisﬁed as regards
their full elucidation, as everyone can sufﬁciently observe in the dialectic
of pure reason, which for this very reason has its good subjective ground.
Who can bear being brought, as regards the nature of our soul, both to
the point of a clear consciousness of the subject and to the conviction
that the appearances of that subject cannot be explained materialistically,
without asking what then the soul really is, and, if no concept of expe-
rience sufﬁces thereto, without perchance adopting a concept of reason
(that of a simple immaterial d being) just for this purpose, although we can
by no means prove the objective reality of that concept? Who can satisfy
himself with mere cognition through experience in all the cosmological
questions, of the duration and size of the world, of freedom or natural
necessity, since, wherever we may begin, any answer given according
to principles of experience e always begets a new question which also
a
b
c
d
e
Principien
Grundsätze
gesunde Verstand
Reading immateriellen for materiellen, with Ak.
Reading Erfahrungsgrundsätzen for Erfahrungsgrundgesetzen, with Ak.
141
4: 352Prolegomena to any future metaphysics
4: 353
requires an answer, and for that reason clearly proves the insufﬁciency
of all physical modes of explanation for the satisfaction of reason? Finally,
who cannot see, from the throughgoing contingency and dependency of
everything that he might think or assume according to principles of ex-
perience, the impossibility of stopping with these, and who does not feel
compelled, regardless of all prohibition against losing himself in tran-
scendent ideas, nevertheless to look for peace and satisfaction beyond
all concepts that he can justify through experience, in the concept of a
being the idea of which indeed cannot in itself be understood as regards
possibility – though it cannot be refuted either, because it pertains to
a mere being of the understanding – an idea without which, however,
reason would always have to remain unsatisﬁed? a
Boundaries b (in extended things) always presuppose a space that is
found outside a certain ﬁxed location, and that encloses that location;
limits c require nothing of the kind, but are mere negations that affect
a magnitude insofar as it does not possess absolute completeness. Our
reason, however, sees around itself as it were a space for the cognition of
things in themselves, although it can never have determinate concepts
of those things and is limited to appearances alone.
As long as reason’s cognition is homogeneous, no determinate bound-
aries can be thought for it. In mathematics and natural science, human
reason recognizes limits but not boundaries, i.e., it indeed recognizes
that something lies beyond it to which it can never reach, but not that
it would itself at any point ever complete its inner progression. The
expansion of insight in mathematics, and the possibility of ever-new in-
ventions, goes to inﬁnity; so too does the discovery of new properties
in nature (new forces and laws) through continued experience and the
uniﬁcation of that experience by reason. But limits here are nonethe-
less unmistakable, for mathematics refers only to appearances, and that
which cannot be an object of sensory intuition, like the concepts of meta-
physics and morals, lies entirely outside its sphere, and it can never lead
there; but it also has no need whatsoever for such concepts. There is
therefore no continuous progress and advancement toward those sci-
ences, or any point or line of contact, as it were. Natural science will
never reveal to us the inside of things, i.e., that which is not appear-
ance but can nonetheless serve as the highest ground of explanation
for the appearances; but it does not need this for its physical expla-
nations; nay, if such were offered to it from elsewhere (e.g., the inﬂu-
ence of immaterial beings), natural science should indeed reject it and
ought by no means bring it into the progression of its explanations,
a
b
c
Final question mark added, with Vorländer.
Grenzen
Schranken
142Prolegomena to any future metaphysics
but should always base its explanations only on that which can be-
long to experience as an object of the senses and which can be brought
into connection with our actual perceptions in accordance with laws of
experience.
But metaphysics, in the dialectical endeavors of pure reason (which
are not initiated arbitrarily or wantonly, but toward which the nature
of reason itself drives), does lead us to the boundaries; and the tran-
scendental ideas, just because they cannot be avoided and yet will never
be realized, serve not only actually to show us the boundaries of rea-
son’s pure use, but also to show us the way to determine such bound-
aries; and that too is the end and use of this natural predisposition of
our reason, which bore metaphysics as its favorite child, whose procre-
ation (as with any other in the world) is to be ascribed not to chance
accident but to an original seed that is wisely organized toward great
ends. For metaphysics, perhaps more than any other science, is, as re-
gards its fundamentals, placed in us by nature itself, and cannot at all be
seen as the product of an arbitrary choice, or as an accidental extension
from the progression of experiences (it wholly separates itself from those
experiences).
Reason, through all of its concepts and laws of the understanding,
which it ﬁnds to be adequate for empirical use, and so adequate within the
sensible world, nonetheless does not thereby ﬁnd satisfaction for itself;
for, as a result of questions that keep recurring to inﬁnity, it is denied
all hope of completely answering those questions. The transcendental
ideas, which have such completion as their aim, are such problems for
reason. Now reason clearly sees: that the sensible world could not contain
this completion, any more than could therefore all of the concepts that
serve solely for understanding that world: space and time, and everything
that we have put forward under the name of the pure concepts of the
understanding. The sensible world is nothing but a chain of appearances
connected in accordance with universal laws, which therefore has no
existence for itself; it truly is not the thing in itself, and therefore it
necessarily refers to that which contains the ground of those appearances,
to beings that can be cognized not merely as appearances, a but as things
in themselves. Only in the cognition of the latter can reason hope to see
its desire for completeness in the progression from the conditioned to
its conditions satisﬁed for once.
Above (§§33, 34) we noted limits of reason with respect to all cog-
nition of mere beings of thought; now, since the transcendental ideas
nevertheless make the progression up to these limits necessary for us,
and have therefore led us, b as it were, up to the contiguity of the ﬁlled
a
b
Reading Erscheinungen for Erscheinung (here and just previously), with Vorländer.
Reading uns for nur, with Vorländer.
143
4: 354Prolegomena to any future metaphysics
4: 355
space (of experience) with empty space (of which we can know nothing –
the noumena), we can also determine the boundaries of pure reason; for
in all boundaries there is something positive (e.g., a surface is the bound-
ary of corporeal space, yet is nonetheless itself a space; a line is a space,
which is the boundary of a surface; a point is the boundary of a line,
yet is nonetheless a locus in space), whereas limits contain mere nega-
tions. The limits announced in the cited sections are still not enough
after we have found that something lies beyond them (although we will
never cognize what that something may be in itself). For the question
now arises: How does our reason cope with this connection of that with
which we are acquainted to that with which we are not acquainted, and
never will be? Here is a real connection of the known to a wholly un-
known (which will always remain so), and even if the unknown should
not become the least bit better known – as is not in fact to be hoped –
the concept of this connection must still be capable of being determined
and brought to clarity.
We should, then, think for ourselves an immaterial being, an intelli-
gible world, and a highest of all beings (all noumena), because only in
these things, as things in themselves, does reason ﬁnd completion and
satisfaction, which it can never hope to ﬁnd in the derivation of the ap-
pearances from the homogeneous grounds of those appearances; and we
should think such things for ourselves because the appearances actually
do relate to something distinct from them (and so entirely heteroge-
neous), in that appearances always presuppose a thing in itself, a and so
provide notice of such a thing, whether or not it can be cognized more
closely.
Now since we can, however, never cognize these intelligible beings
according to what they may be in themselves, i.e., determinately – though
we must nonetheless assume such beings in relation to the sensible world,
and connect them with it through reason – we can still at least think this
connection by means of such concepts as express the relation of those
beings to the sensible world. For, if we think an intelligible being through
nothing but pure concepts of the understanding, we really think nothing
determinate thereby, and so our concept is without signiﬁcance; if we
think it through properties borrowed from the sensible world, it is no
longer an intelligible being: it is thought as one of the phenomena and
belongs to the sensible world. We will take an example from the concept
of the supreme being.
The deistic concept is a wholly pure concept of reason, which however
represents merely a thing that contains every reality, without being able
to determine a single one of them, since for that an example would have
a
Sache an sich selbst
144Prolegomena to any future metaphysics
to be borrowed from the sensible world, in which case I would always
have to do only with an object of the senses, and not with something
completely heterogeneous which cannot be an object of the senses at
all. For I would, for instance, attribute understanding to it; but I have
no concept whatsoever of any understanding save one like my own, that
is, one such that intuitions must be given to it through the senses, and
that busies itself with bringing them under rules for the unity of con-
sciousness. But then the elements of my concept would still lie within
appearance; I was, however, forced by the inadequacy of the appearances
to go beyond them, to the concept of a being that is in no way dependent
on appearances nor bound up with them as conditions for its determi-
nation. If, however, I separate understanding from sensibility, in order
to have a pure understanding, then nothing but the mere form of think-
ing, without intuition, is left; through which, by itself, I cannot cognize
anything determinate, hence cannot cognize any object. To that end I
would have to think to myself a different understanding, which intu-
its objects, 75 of which, however, I do not have the least concept, since
the human understanding is discursive and can cognize only by means
of universal concepts. The same thing happens to me if I attribute a
will to the supreme being: For I possess this concept only by drawing
it from my inner experience, where, however, my a dependence on sat-
isfaction through objects whose existence we need, and so sensibility, is
the basis – which completely contradicts the pure concept of a supreme
being.
Hume’s objections to deism are weak and always concern the grounds
of proof but never the thesis of the deistic assertion itself. But with
respect to theism, which is supposed to arise through a closer deter-
mination of our (in deism, merely transcendent) concept of a supreme
being, they are very strong, and, depending on how this concept has
been framed, they are in certain cases (in fact, in all the usual ones) ir-
refutable. Hume always holds to this: that through the mere concept of
a ﬁrst being to which we attribute none but ontological predicates (eter-
nity, omnipresence, omnipotence), we actually do not think anything
determinate at all; rather, properties would have to be added that can
yield a concept in concreto: it is not enough to say: this being is a cause,
rather we need to say how its causality is constituted, e.g., by under-
standing and willing – and here begin Hume’s attacks on the matter in
question, namely on theism, whereas he had previously assaulted only
the grounds of proof for deism, an assault that carries no special danger
with it. His dangerous arguments relate wholly to anthropomorphism,
of which he holds that it is inseparable from theism and makes theism
a
Rejecting various emendations recorded by Vorländer.
145
4: 356Prolegomena to any future metaphysics
4: 357
self-contradictory, but that if it is eliminated, theism falls with it and
nothing but deism remains – from which nothing can be made, which
can be of no use to us, and can in no way serve as a foundation for
religion and morals. If this unavoidability of anthropomorphism were
certain, then the proofs for the existence of a supreme being might be
what they will, and might all be granted, and still the concept of this be-
ing could never be determined by us without our becoming entangled in
contradictions.
If we combine the injunction to avoid all transcendent judgments
of pure reason with the apparently conﬂicting command to proceed
to concepts that lie beyond the ﬁeld of immanent (empirical) use, we
become aware that both can subsist together, but only directly on the
boundary of all permitted use of reason – for this boundary belongs just as
much to the ﬁeld of experience as to that of beings of thought – and we
are thereby at the same time taught how those remarkable ideas serve
solely for determining the boundary of human reason: that is, we are
taught, on the one hand, not to extend cognition from experience without
bound, so that nothing at all remains for us to cognize except merely the
world, and, on the other, nevertheless not to go beyond the boundary
of experience and to want to judge of things outside that boundary as
things in themselves.
But we hold ourselves to this boundary if we limit our judgment
merely to the relation that the world may have to a being whose concept
itself lies outside all cognition that we can attain within the world. For
we then do not attribute to the supreme being any of the properties in
themselves by which we think the objects of experience, and we thereby
avoid dogmatic anthropomorphism; but we attribute those properties,
nonetheless, to the relation of this being to the world, and allow ourselves
a symbolic anthropomorphism, which in fact concerns only language and
not the object itself.
If I say that we are compelled to look upon the world as if it were the
work of a supreme understanding and will, I actually say nothing more
than: in the way that a watch, a ship, and a regiment are related to an
artisan, a builder, and a commander, the sensible world (or everything
that makes up the basis of this sum total of appearances) is related to the
unknown – which I do not thereby cognize according to what it is in
itself, but only according to what it is for me, that is, with respect to the
world of which I am a part.
§ 58
This type of cognition is cognition according to analogy, which surely does
not signify, as the word is usually taken, an imperfect similarity between
146Prolegomena to any future metaphysics
two things, but rather a perfect similarity between two relations in wholly
dissimilar things. ∗ By means of this analogy there still remains a concept
of the supreme being sufﬁciently determinate for us, though we have
omitted everything that could have determined this concept uncondition-
ally and in itself; for we determine the concept only with respect to the
world and hence with respect to us, and we have no need of more. The
attacks that Hume makes against those who want to determine this con-
cept absolutely – since they borrow the materials for this determination
from themselves and from the world – do not touch us; he also cannot
reproach us that nothing whatsoever would remain for us if objective
anthropomorphism were subtracted from the concept of the supreme
being.
For if one only grants us, at the outset, the deistic concept of a ﬁrst
being as a necessary hypothesis (as does Hume in his Dialogues in the
person of Philo as opposed to Cleanthes), which is a concept in which
one thinks the ﬁrst being by means of ontological predicates alone, of
substance, cause, etc. (something that one must do, since reason, being
driven in the sensible world solely by conditions that are always again
conditioned, cannot have any satisfaction at all without this being done,
and something that one very well can do without falling into that anthro-
pomorphism which transfers predicates from the sensible world onto a
being wholly distinct from the world, since the predicates listed here are
mere categories, which cannot indeed provide any determinate concept
of that being, but which, for that very reason, do not provide a concept
of it that is limited to the conditions of sensibility) – then nothing can
keep us from predicating of this being a causality through reason with re-
spect to the world, and thus from crossing over to theism, but without
our being compelled to attribute this reason to that being in itself, as a
∗
Such is an analogy between the legal relation of human actions and the me-
chanical relation of moving forces: I can never do anything to another without
giving him a right to do the same to me under the same conditions; just as a
body cannot act on another body with its motive force without thereby caus-
ing the other body to react just as much on it. Right and motive force are here
completely dissimilar things, but in their relation there is nonetheless complete
similarity. By means of such an analogy I can therefore provide a concept of a
relation to things that are absolutely unknown to me. E.g., the promotion of
the happiness of the children = a is to the love of the parents = b as the welfare
of humankind = c is to the unknown in God = x, which we call love: not as if
this unknown had the least similarity with any human inclination, but because
we can posit the relation between God’s love and the world to be similar to
that which things in the world have to one another. But here the concept of the
relation is a mere category, namely the concept of cause, which has nothing to
do with sensibility.
147
4: 358Prolegomena to any future metaphysics
4: 359
4: 360
property inhering in it. For, concerning the ﬁrst point, a the only possible
way to compel the use of reason in the sensible world (with respect to all
possible experience) into the most thoroughgoing harmony with itself is
to assume, in turn, a supreme reason as a cause of all connections in the
world; such a principle must be thoroughly advantageous to reason and
can nowhere harm it in its use in nature. Regarding the second point, b how-
ever, reason is not thereby transposed as a property onto the ﬁrst being
in itself, but only onto the relation of that being to the sensible world, and
therefore anthropomorphism is completely avoided. For here only the
cause of the rational form found everywhere in the world is considered,
and the supreme being, insofar as it contains the basis of this rational
form of the world, is indeed ascribed reason, but only by analogy, i.e.,
insofar as this expression signiﬁes only the relation that the highest cause
(which is unknown to us) has to the world, in order to determine every-
thing in it with the highest degree of conformity to reason. We thereby
avoid using the property of reason in order to think God, but instead
think the world through it in the manner necessary to have the greatest
possible use of reason with respect to the world in accordance with a
principle. We thereby admit that the supreme being, as to what it may
be in itself, is for us wholly inscrutable and is even unthinkable by us in a
determinate manner; and we are thereby prevented from making any tran-
scendent use of c the concepts that we have of reason as an efﬁcient cause
(through willing) in order to determine the divine nature through prop-
erties that are in any case always borrowed only from human nature, and
so from losing ourselves in crude or fanatical concepts, and, on the other
hand, we are also prevented from swamping the contemplation of the
world with hyperphysical modes of explanation according to concepts
of human reason we have transposed onto God, and so from diverting
this contemplation from its true vocation, according to which it is sup-
posed to be a study of mere nature through reason, and not an audacious
derivation of the appearances of nature from a supreme reason. The ex-
pression suitable to our weak concepts will be: that we think the world
as if it derived from a supreme reason as regards its existence and inner
determination; whereby we in part cognize the constitution belonging
to it (the world) itself, without presuming to want to determine that of
its cause in itself, and, on the other hand, we in part posit the basis of this
constitution (the rational form of the world) in the relation of the highest
cause to the world, not ﬁnding the world by itself sufﬁcient thereto. ∗
∗
I will say: the causality of the highest cause is that, with respect to the world,
which human reason is with respect to its works of art. Thereby the nature of
the highest cause itself remains unknown to me: I compare only its effect (the
order of the world), which is known to me, and the conformity with reason
of this effect, with the effects of human reason that are known to me, and in
148Prolegomena to any future metaphysics
In this way the difﬁculties that appear to oppose theism disappear,
in that to Hume’s principle, not to drive the use of reason dogmatically
beyond the ﬁeld of all possible experience, we conjoin another principle
that Hume completely overlooked, namely: not to look upon the ﬁeld
of possible experience as something that bounds itself in the eyes of our
reason. A critique of reason indicates the true middle way between the
dogmatism that Hume fought and the skepticism he wanted to introduce
instead – a middle way that, unlike other middle ways, which we are
advised to determine for ourselves as it were mechanically (something
from one side, and something from the other), and by which no one
is taught any better, is d one, rather, that can be determined precisely,
according to principles. 76
§ 59
At the beginning of this note I made use of the metaphor of a boundary
in order to ﬁx the limits of reason with respect to its own appropriate
use. The sensible world contains only appearances, which are still not
things in themselves, which latter things (noumena) the understanding
must therefore assume for the very reason that it cognizes the objects of
experience as mere appearances. Both are considered together in our
reason, and the question arises: how does reason proceed in setting
boundaries for the understanding with respect to both ﬁelds? Expe-
rience, which contains everything that belongs to the sensible world,
does not set a boundary for itself: From every conditioned it always
arrives merely at another conditioned. That which is to set its bound-
ary must lie completely outside it, and this is the ﬁeld of pure intel-
ligible beings. For us, however, as far as concerns the determination of
the nature of these intelligible beings, this is an empty space, and to
that extent, if dogmatically determined concepts are intended, we can-
not go beyond the ﬁeld of possible experience. But since a boundary
is itself something positive, which belongs as much to what is within it
as to the space lying outside a given totality, reason therefore, merely
by expanding up to this boundary, partakes of a real, positive cogni-
tion, provided that it does not try to go out beyond the boundary, since
there it ﬁnds an empty space before it, in which it can indeed think the
consequence I call the highest cause a reason, without thereby ascribing to it
as its property the same thing I understand by this expression in humans, or in
anything else known to me.
a
b
c
d
“something that one must do . . .”
“something that one very well can do . . .”
Reading von for nach, with Vorländer.
Adding ist after Mittelwege, as suggested by Vorländer.
149
4: 361Prolegomena to any future metaphysics
4: 362
forms to things, but no things themselves. But setting the boundary to
the ﬁeld of experience through something that is otherwise unknown to
it is indeed a cognition that is still left to reason from this standpoint,
whereby reason is neither locked inside the sensible world nor adrift
outside it, but, as beﬁts knowledge of a boundary, restricts itself solely
to the relation of what lies outside the boundary to what is contained
within.
Natural theology is a concept of this kind, on the boundary of human
reason, since reason ﬁnds itself compelled to look out toward the idea of
a supreme being (and also, in relation to the practical, to the idea of an
intelligible world), not in order to determine something with respect to
this mere intelligible being (and hence outside the sensible world), but
only in order to guide its own use within the sensible world in accor-
dance with principles of the greatest possible unity (theoretical as well
as practical), and to make use (for this purpose) of the relation of that
world to a free-standing reason as the cause of all of these connections –
not, however, in order thereby merely to fabricate a being, but, since
beyond the sensible world there must necessarily be found something
that is thought only by the pure understanding, in order, in this way, to
determine this being, a though of course merely through analogy.
In this manner our previous proposition, which is the result of the
entire Critique, remains: “that reason, through all its a priori principles,
never teaches us about anything more than objects of possible experi-
ence alone, and of these, nothing more than what can be cognized in
experience”; but this limitation does not prevent reason from carrying
us up to the objective boundary of experience – namely, to the relation to
something that cannot itself be an object of experience, but which must
nonetheless be the highest ground of all experience – without, however,
teaching us anything about this ground in itself, but only in relation to
reason’s own complete use in the ﬁeld of possible experience, as directed
to the highest ends. This is, however, all of the beneﬁt that can reasonably
even be wished for here, and there is cause to be satisﬁed with it.
§ 60
We have thus fully exhibited metaphysics in accordance with its subjec-
tive possibility, as metaphysics is actually given in the natural predisposition
of human reason, and with respect to that which forms the essential goal
of its cultivation. But because we found that, if reason is not reined in and
given limits by a discipline of reason, which is only possible through a sci-
entiﬁc critique, this wholly natural use of this sort of predisposition of our
reason entangles it in transcendent dialectical inferences, which are partly
a
Omitting nur after dieses, with Vorländer.
150Prolegomena to any future metaphysics
specious, partly even in conﬂict among themselves; and, moreover, be-
cause we found that this sophistical metaphysics is superﬂuous, nay, even
detrimental to the advancement of the cognition of nature, it therefore
still remains a problem worthy of investigation, to discover the natural
purposes toward which this predisposition of our a reason b to transcendent
concepts may be aimed, since everything found in nature must originally
be aimed at some beneﬁcial purpose or other.
Such an investigation is in fact uncertain; I also admit that it is merely
conjectural (as is everything I know to say concerning the original pur-
poses of nature), something I may be permitted in this case only, since
the question does not concern the objective validity of metaphysical
judgments, but rather the natural predisposition to such judgments, and
therefore lies outside the system of metaphysics, in anthropology. 77
If I consider c all the transcendental ideas, which together constitute
the real problem for natural pure reason – a problem that compels reason
to forsake the mere contemplation of nature and go beyond all possible
experience, and, in this endeavor, to bring into existence the thing called
metaphysics (be it knowledge or sophistry) – then I believe I perceive
that this natural predisposition is aimed at making our concept sufﬁ-
ciently free from the fetters of experience and the limits of the mere
contemplation of nature that it at the least sees a ﬁeld opening before it
that contains only objects for the pure understanding which no sensibil-
ity can reach: not with the aim that we concern ourselves speculatively
with these objects (for we ﬁnd no ground on which we can gain footing),
but rather with practical principles, d which, without ﬁnding such a space
before them for their necessary expectations and hopes, could not ex-
tend themselves to the universality that reason ineluctably requires with
respect to morals.
Here I now ﬁnd that the psychological idea, however little insight I may
gain through it into the pure nature of the human soul elevated beyond
all concepts of experience, at least reveals clearly enough the inadequacy
of those concepts of experience, and thereby leads me away from mate-
rialism, as a psychological concept unsuited to any explanation of nature
and one that, moreover, constricts reason with respect to the practical.
Similarly, the cosmological ideas, through the manifest inadequacy of all
possible cognition of nature to satisfy reason in its rightful demands,
a
b
c
d
Reading unserer for unsere, with Ak.
Reading Vernunft, with the original edition; Ak has Natur (a typographical error).
Supplying betrachte as the verb, with Vorländer; Ak supplies zusammennehme.
There is a generally acknowledged ellipsis at this point. I avoid adopting any of the
lengthy interpolations recorded in Vorländer by extending the meaning of the previous
verb, zu beschäftigen, beyond sondern, so as to provide a context for practische Principien (a
solution that also would require some emendation of the German).
151
4: 363Prolegomena to any future metaphysics
4: 364
serve to deter us from naturalism, which would have it that nature is
sufﬁcient unto itself. Finally, since all natural necessity in the sensible
world is always conditioned, in that it always presupposes the depen-
dence of one thing on another, and since unconditioned necessity must
be sought only in the unity of a cause distinct from the sensible world,
although the causality of that cause, in turn, if it were merely nature,
could never make comprehensible the existence of the contingent as its
consequence, reason, therefore, by means of the theological idea, frees
itself from fatalism – from blind natural necessity both in the connec-
tion of nature itself, without a ﬁrst principle, and in the causality of this
principle itself – and leads the way to the concept of a cause through
freedom, and so to that of a highest intelligence. The transcendental
ideas therefore serve, if not to instruct us positively, at least to negate
the impudent assertions of materialism, naturalism, and fatalism which
constrict the ﬁeld of reason, and in this way they serve to provide moral
ideas with space outside the ﬁeld of speculation; and this would, I should
think, to some extent explain the aforementioned natural predisposition.
The practical beneﬁt that a purely speculative science may have lies
outside the boundaries of this science; such beneﬁt can therefore be seen
simply as a scholium, and like all scholia does not form part of the sci-
ence itself. Nonetheless, this relation at least lies within the boundaries
of philosophy, and especially of that philosophy which draws from the
wellsprings of pure reason, where the speculative use of reason in meta-
physics must necessarily have unity with its practical use in morals. Hence
the unavoidable dialectic of pure reason deserves, in a metaphysics con-
sidered as natural predisposition, not only to be explained as an illusion
that needs to be resolved, but also (if one can) as a natural institution in
accordance with its purpose – although this endeavor, as supererogatory,
cannot rightly be required of metaphysics proper.
The solution to the questions that proceed in the Critique from pages
647 to 668 would have to be taken for a second scholium, more closely
related to the content of metaphysics. 78 For there certain principles of
reason are put forward that determine the order of nature a priori, or
rather determine the understanding a priori, which is supposed to search
for the laws of this order by means of experience. These principles seem
to be constitutive and law-giving with respect to experience, though
they spring from mere reason, which cannot, like the understanding,
be regarded as a principle of possible experience. Now whether this
agreement rests on the fact that, just as nature does not in itself inhere
in the appearances or in their source, sensibility, but is found only in
the relation of sensibility to the understanding, so too, a thoroughgoing
unity in the use of this understanding, for the sake of a uniﬁed possible
experience (in a system), can belong to the understanding only in relation
to reason, hence experience, too, be indirectly subject to the legislation
152Prolegomena to any future metaphysics
of reason – this may be further pondered by those who want to track the
nature of reason even beyond its use in metaphysics, into the universal
principles for making natural history generally systematic; for in the
book itself I have indeed presented this problem as important, but have
not attempted its solution. ∗
And thus I conclude the analytic 79 solution of the main question I
myself have posed: How is metaphysics in general possible?, since I have
ascended from the place where its use is actually given, at least in the
consequences, to the grounds of its possibility.
∗
It was my unremitting intention throughout the Critique not to neglect any-
thing that could bring to completion the investigation of the nature of pure
reason, however deeply hidden it might lie. Afterwards it is in each person’s
discretion, how far he will take his investigation, if he only has been apprised
of what may still need to be done; for it can properly be expected, from one
who has made it his business to survey this entire ﬁeld, that afterward he leave
future additions and optional divisions to others. Hereto belong both of the
scholia, which, on account of their dryness, could hardly be recommended to
amateurs, and have therefore been set out only for experts.
153
4: 365Solution
to the General Question
of the Prolegomena
How is metaphysics possible as science?
4: 366
Metaphysics, as a natural predisposition of reason, is actual, but it is also
of itself (as the analytical solution to the third main question proved)
dialectical and deceitful. The desire to derive principles from it, and to
follow the natural but nonetheless false illusion in their use, can therefore
never bring forth science, but only vain dialectical art, in which one
school can outdo another but none can ever gain legitimate and lasting
approbation.
In order that metaphysics might, as science, be able to lay claim, not
merely to deceitful persuasion, but to insight and conviction, a critique
of reason itself must set forth the entire stock of a priori concepts, their
division according to the different sources (sensibility, understanding,
and reason), further, a complete table of those concepts, and the analysis
of all of them along with everything that can be derived from that analysis;
and then, especially, such a critique must set forth the possibility of
synthetic cognition a priori through a deduction of these concepts, it
must set forth the principles of their use, and ﬁnally also the boundaries
of that use; and all of this in a complete system. Therefore a critique, and
that alone, contains within itself the whole well-tested and veriﬁed plan
by which metaphysics as science can be achieved, and even all the means
for carrying it out; by any other ways or means it is impossible. Therefore
the question that arises here is not so much how this enterprise is possible,
but only how it is to be set in motion, and good minds stirred from the
hitherto ill-directed and fruitless endeavor to one that will not deceive,
and how such an alliance might best be turned toward the common end.
This much is certain: whosoever has once tasted of critique forever
loathes all the dogmatic chatter which he previously had to put up with
out of necessity, since his reason was in need of something and could not
ﬁnd anything better for its sustenance. Critique stands to the ordinary
school metaphysics precisely as chemistry stands to alchemy, or astronomy
to the fortune-teller’s astrology. I’ll guarantee that no one who has thought
through and comprehended the principles of critique, even if only in
154Prolegomena to any future metaphysics
these prolegomena, will ever again return to that old and sophistical
pseudoscience; he will on the contrary look out with a certain delight
upon a metaphysics that is now fully in his power, that needs no more
preliminary discoveries, and that can for the ﬁrst time provide reason
with lasting satisfaction. For this is an advantage upon which metaphysics
alone, among all the possible sciences, can rely with conﬁdence, namely,
that it can be completed and brought into a permanent state, since it
cannot be further changed and is not susceptible to any augmentation
through new discoveries – because here reason has the sources of its
cognition not in objects and their intuition (through which reason cannot
be taught one thing more), but in itself, and, if reason has presented
the fundamental laws of its faculty fully and determinately (against all
misinterpretation), nothing else remains that pure reason could cognize
a priori, or even about which it could have cause to ask. The sure prospect
of a knowledge so determinate and ﬁnal has a certain attraction to it, even
if all usefulness (of which I will say more hereafter) is set aside.
All false art, all empty wisdom lasts for its time; for it ﬁnally destroys
itself, and the height of its cultivation is simultaneously the moment of its
decline. That this time has now come as regards metaphysics is proven by
the condition into which it has fallen among all learned peoples, amidst
all the zeal with which sciences of all kinds are otherwise being developed.
The old organization of university studies still preserves the shadow of
metaphysics, a lone academy of sciences now and then, by offering prizes,
moves someone or other to make an effort in it, but metaphysics is no
longer reckoned among serious sciences, and each may judge for himself
how a clever man, whom one wished to call a great metaphysician, would
perhaps receive this encomium, which might be well meant but would
hardly be envied by anyone.
But although the time for the collapse of all dogmatic metaphysics
is undoubtedly here, much is still lacking in order to be able to say
that, on the contrary, the time for its rebirth, through a thorough and
completed critique of reason, has already appeared. All transitions from
one inclination to its opposite pass through a state of indifference, and
this moment is the most dangerous for an author, but nonetheless, it
seems to me, the most favorable for the science. For if the partisan spirit
has been extinguished through the complete severance of former ties,
then minds are best disposed to a hear out, bit by bit, proposals for an
alliance according to another plan.
If I say that I hope these Prolegomena will perhaps excite investigation
in the ﬁeld of critique, and provide the universal spirit of philosophy,
which seems to want nourishment in its speculative part, with a new and
a
Reading um for nur, with Ak.
155
4: 367Prolegomena to any future metaphysics
4: 368
quite promising object of sustenance, I can already imagine beforehand
that everyone who a has been made weary and unwilling by the thorny
paths on which I have led him in the Critique will ask me: On what do I
base this hope? I answer: On the irresistible law of necessity.
That the human mind would someday entirely give up metaphysical
investigations is just as little to be expected, as that we would someday
gladly stop all breathing so as never to take in impure air. There will
therefore be metaphysics in the world at every time, and what is more,
in every human being, and especially the reﬂective ones; metaphysics
that each, in the absence of a public standard of measure, will carve
out for himself in his own manner. Now what has hitherto been called
metaphysics can satisfy no inquiring mind, and yet it is also impossible to
give up metaphysics completely; therefore, a critique of pure reason itself
must ﬁnally be attempted, or, if one exists, it must be examined and put
to a general test, since there are no other means to relieve this pressing
need, which is something more than a mere thirst for knowledge.
Ever since I have known critique, I have been unable to keep myself
from asking, upon ﬁnishing reading through a book with metaphysical
content, which has entertained as well as cultivated me by the determi-
nation of its concepts and by variety and organization and by an easy
presentation: has this author advanced metaphysics even one step? I ask for-
giveness of the learned men whose writings have in other respects been
useful to me and have always contributed to a cultivation of mental pow-
ers, because I confess that I have not been able to ﬁnd, either in their
attempts or in my own inferior ones (with self-love speaking in their
favor), that the science has thereby been advanced in the least, and this
for the wholly natural reason that the science did not yet exist, and also
that it cannot be assembled bit by bit but rather its seed must be fully
preformed beforehand in the critique. However, in order to avoid all
misunderstanding, it must be recalled from the preceding that although
the understanding certainly beneﬁts very much from the analytical treat-
ment of our concepts, the science (of metaphysics) is not advanced the
least bit thereby, since these analyses of concepts are only materials, out
of which the science must ﬁrst be constructed. The concept of substance
and accident may be analyzed and determined ever so nicely; that is
quite good as preparation for some future use. But if I simply cannot
prove that in all that exists the substance persists and only the accidents
change, then through all this analysis the science has not been advanced
in the least. Now metaphysics has not as yet been able to prove, as a pri-
ori valid, either this proposition or the principle of sufﬁcient reason, still
less any more composite proposition, such as, for instance, one belonging
a
Reading den for der, with Ak.
156Prolegomena to any future metaphysics
to psychology or cosmology, nor, in general, any synthetic proposition
whatsoever; hence, through all this analysis nothing has been achieved,
nothing created and advanced, and, after so much bustle and clatter,
the science is still right where it was in Aristotle’s time, although the
preparations for it incontestably have been much better laid than be-
fore, if only the guiding thread to synthetic cognition had ﬁrst been
found.
If anyone believes himself wronged in this, he can easily remove the
above indictment if he will cite only a single synthetic proposition be-
longing to metaphysics that he offers to prove a priori in the dogmatic
manner; for only when he accomplishes this will I grant to him that
he has actually advanced the science (even if the proposition may oth-
erwise have been sufﬁciently established through common experience).
No challenge can be more moderate and more equitable, and in the (in-
fallibly certain) event of nonfulﬁllment, no verdict more just, than this:
that up to now metaphysics as science has never existed at all.
In case the challenge is accepted, I must forbid only two things: ﬁrst,
the plaything of probability and conjecture, which suits metaphysics just
as poorly as it does geometry; second, decision by means of the divining
rod of so-called sound common sense, a which does not bend for everyone,
but is guided by personal qualities.
For, as regards the ﬁrst, there can be nothing more absurd than to want
to base one’s judgments in metaphysics, a philosophy from pure reason,
on probability and conjecture. Everything that is to be cognized a priori is
for that very reason given out as apodictically certain and must therefore
also be proven as such. One might just as well want to base a geometry
or an arithmetic on conjectures; for as concerns the calculus probabilium b
of arithmetic, it contains not probable but completely certain judgments
about the degree of possibility of certain cases under given homogeneous
conditions, judgments which, in the sum total of all possible cases, must
be found to conform to the rule with complete infallibility, even though
this rule is not sufﬁciently determinate with respect to any single case.
Only in empirical natural science can conjectures (by means of induction
and analogy) be tolerated, and even then, the possibility at least of what
I am assuming must be fully certain.
Matters are, if possible, even worse with the appeal to sound common
sense, if the discussion concerns c concepts and principles, not insofar as
they are supposed to be valid with respect to experience, but rather inso-
far as they are to be taken as valid beyond the conditions of experience.
a
b
c
gesunden Menschenverstandes
“calculus of probability”
Adding die Rede ist, with Ak.
157
4: 369Prolegomena to any future metaphysics
4: 370
For what is sound common sense? a It is the ordinary understanding, b insofar
as it judges correctly. And what now is the ordinary understanding? It is
the faculty of cognition and of the use of rules in concreto, as distinguished
from the speculative understanding, which is a faculty of the cognition of
rules in abstracto. The ordinary understanding will, then, hardly be able
to understand the rule: that everything which happens is determined by
its cause, and it will never be able to have insight into it in such a gen-
eral way. It therefore demands an example from experience, and when it
hears that this rule means nothing other than what it had always thought
when a windowpane was broken or a household article had disappeared,
it then understands the principle and grants it. Ordinary understanding,
therefore, has a use no further than the extent to which it can see its
rules conﬁrmed in experience (although these rules are actually present
in it a priori); consequently, to have insight into these rules a priori and
independently of experience falls to the speculative understanding, and
lies completely beyond the horizon of the ordinary understanding. But
metaphysics is concerned indeed solely with this latter type of cogni-
tion, and it is certainly a poor sign of sound common sense to appeal to
this guarantor, who has no judgment here, and whom we otherwise look
down upon, except if we ﬁnd ourselves in trouble, and without either
advice or help in our speculation.
It is a common excuse, habitually employed by these false friends of or-
dinary common sense (which they extol on occasion, but usually despise),
to say: There must in the end be some propositions that are c immedi-
ately certain, and for which not only no proof, but indeed no account at
all need be given, since otherwise there would never come an end to the
grounds for one’s judgments; but in proof of this right they can never cite
anything else (other than the principle of contradiction, which is how-
ever inadequate for establishing the truth of synthetic judgments) that is
undoubted and can be ascribed directly to ordinary common sense, ex-
cept for mathematical propositions: e.g., that two times two makes four,
that between two points there is only one straight line, and still others.
These judgments are, however, worlds apart from those of metaphysics.
For in mathematics, everything that I conceive through a concept as
possible I can make for myself (construct) by means of my thought; to
one two I successively add the other two, and myself make the number
four, or I draw in thought all kinds of lines from one point to the other,
and can draw only one that is self-similar in all its parts (equal as well as
unequal). 80 But from the concept of a thing I cannot, with all my powers
of thought, draw forth the concept of something else whose existence is
a
b
c
gesunde Verstand
gemeine Verstand
Reading seien for seyn, with Vorländer.
158Prolegomena to any future metaphysics
necessarily connected with the ﬁrst thing, but must consult experience;
and, although my understanding provides me a priori (though always only
in relation to possible experience) with the concept of a connection of
this sort (causality), I nevertheless cannot exhibit this concept in intuition
a priori, like the concepts of mathematics, and thus exhibit its possibility a
priori; rather, this concept (together with principles of its application), if
it is to be valid a priori – as is indeed required in metaphysics – always has
need of a justiﬁcation and deduction of its possibility, for otherwise one
does not know the extent of its validity and whether it can be used only in
experience or also outside it. Therefore in metaphysics, as a speculative
science of pure reason, one can never appeal to ordinary common sense,
but one can very well do so if one is forced to abandon metaphysics
and to renounce all pure speculative cognition, which must always be
knowledge, a hence to renounce metaphysics itself and its teaching (on
certain matters), and if a reasonable belief b is alone deemed possible for
us, as well as sufﬁcient for our needs (perhaps more wholesome indeed
than knowledge itself ). For then the shape of things is completely al-
tered. Metaphysics must be science, not only as a whole but also in all its
parts; otherwise it is nothing at all, since, as speculation of pure reason,
it has a hold on nothing else save universal insights. But outside meta-
physics, probability and sound common sense can very well have their
beneﬁcial and legitimate use, though following principles entirely their
own, whose importance always depends on a relation to the practical.
That is what I consider myself entitled to require for the possibility
of a metaphysics as science.
4: 371
Appendix
On what can be done in order to make
metaphysics as science actual
Since all paths hitherto taken have not attained this end, and it may never
be reached without a preceding critique of pure reason, the demand that
the attempt at such a critique which is now before the public be subjected
to an exact and careful examination does not seem unreasonable – unless
it is considered more advisable still to give up all claims to metaphysics
entirely, in which case, if one only remains true to one’s intention, there
is nothing to be said against it. If the course of events is taken as it actually
runs and not as it should run, then there are two kinds of judgments: a
judgment that precedes the investigation, and in our case this is one in which
the reader, from his own metaphysics, passes judgment on the Critique of
Pure Reason (which is supposed ﬁrst of all to investigate the possibility of
a
b
ein Wissen
vernünftiger Glaube
159
4: 372Prolegomena to any future metaphysics
that metaphysics); and then a different judgment that comes after the investi-
gation, in which the reader is able to set aside for a while the consequences
of the critical investigation, which might tell pretty strongly against the
metaphysics he otherwise accepts, and ﬁrst tests the grounds from which
these consequences may have been derived. If what ordinary metaphysics
presents were undeniably certain (like geometry, for instance), the ﬁrst
way of judging would be valid; for if the consequences of certain princi-
ples conﬂict with undeniable truths, then those principles are false and
are to be rejected without any further investigation. But if it is not the
case that metaphysics has a supply of incontestably certain (synthetic)
propositions, and perhaps is the case that a good number of them, which
are as plausible as the best among them, nevertheless are, a in their con-
sequences, in conﬂict even among themselves, while there is not to be
found overall in metaphysics any secure criterion whatsoever of the truth
of properly metaphysical (synthetic) propositions: then the ﬁrst way of
judging cannot be allowed, but rather the investigation of the principles
of the Critique must precede all judgment of its worth or unworth.
Specimen of a judgment about the Critique
which precedes the investigation
4: 373
This sort of judgment is to be found in the Göttingische gelehrte Anzeigen,
the third part of the supplement, from January 19, 1782, pages 40 ff. 81
If an author who is well acquainted with the object of his work, who
has been assiduous throughout in putting reﬂection into its composition
that is completely his own, falls into the hands of a reviewer who for
his part is sufﬁciently clear-sighted to espy the moments upon which b
the worth or unworth of the piece actually rests, who does not hang on
words but follows the subject matter, and who examines and tests only c
the principles from which the author has proceeded, then although the
severity of the judgment may certainly displease the author, the pub-
lic is, by contrast, indifferent to it, for it proﬁts thereby; and the au-
thor himself can be content that he gets the opportunity to correct or
to elucidate his essays, which have been examined early on by an ex-
pert, and, if he believes he is basically right, in this way to remove in
good time a stumbling block that could eventually be detrimental to his
work.
I ﬁnd myself in a completely different situation with my reviewer.
He appears not at all to see what really mattered in the investigation
with which I have (fortunately or unfortunately) occupied myself, and,
a
b
c
Reading sind for seyn, with Ak.
Reading denen for die, with Vorländer.
Deleting nicht prior to bloss, with Vorländer.
160Prolegomena to any future metaphysics
whether it was impatience with thinking through a lengthy work, or
ill-temper over the threatened reform of a science in which he believed
he had long since put everything in order, or whether, as I reluctantly
surmise, it was the fault of a truly limited conception, through which he
could never think himself beyond his school metaphysics – in short, he
impetuously runs through a long series of propositions, with which one
can think nothing at all without knowing their premises, he disperses his
rebukes to and fro, for which the reader no more sees any basis than he
understands the propositions toward which they are supposedly directed,
and therefore the reviewer can neither help to inform the public nor do
me the least bit of harm in the judgment of experts; consequently, I would
have passed over this review completely, if it did not provide me occasion
for a few elucidations that in some cases might save the reader of these
Prolegomena from misconception.
In order, however, that the reviewer might adopt a viewpoint from
which he could, without having to trouble himself with any special inves-
tigation, most easily present the entire work in a manner disadvantageous
to the author, he begins and also ends by saying: “this work is a system
of transcendental a (or, as he construes it, higher) ∗ idealism.” b
At the sight of this line I quickly perceived what sort of review would
issue thence – just about as if someone who had never seen or heard
anything of geometry were to ﬁnd a Euclid, and, being asked to pass
judgment on it, were perhaps to say, after stumbling onto a good many
ﬁgures by turning the pages: “the book is a systematic guide to drawing;
the author makes use of a special language in order to provide obscure,
unintelligible instructions, which in the end can achieve nothing more
than what anyone can accomplish with a good natural eye, and so on.”
∗ On no account higher. High towers and the metaphysically-great men who
resemble them, around both of which there is usually much wind, are not for
me. My place is the fertile bathos of experience, and the word: transcendental –
whose signiﬁcation, which I indicated so many times, was not caught once by the
reviewer (so hastily had he looked at everything) – does not signify something
that surpasses all experience, but something that indeed precedes experience (a
priori), but that, all the same, is destined to nothing more than solely to make
cognition from experience possible. If these concepts cross beyond experience,
their use is then called transcendent, which is distinguished from the immanent
use (i.e., use limited to experience). All misinterpretations of this kind have been
sufﬁciently forestalled in the work itself; but the reviewer found his advantage
in misinterpretations.
a Reading transcendentalen for transscendenten, in accordance with Kant’s wording in his
footnote; the Garve–Feder review itself has the word transscendentellen here (p. 40), a
spelling that Kant did not use.
The parenthetical aside is Kant’s insertion.
b
161
4: 374Prolegomena to any future metaphysics
4: 375
Let us, however, look at what sort of idealism it is that runs through
my entire work, although it does not by far constitute the soul of the
system.
The thesis of all genuine idealists, from the Eleatic School up to
Bishop Berkeley, 82 is contained in this formula: “All cognition through
the senses and experience is nothing but sheer illusion, and there is truth
only in the ideas of pure understanding and reason.”
The principle that governs and determines my idealism throughout is,
on the contrary: “All cognition of things out of mere pure understanding
or pure reason is nothing but sheer illusion, and there is truth only in
experience.”
But this is, of course, the direct opposite of the previous, genuine
idealism; how then did I come to use this expression with a completely
opposite intention, and how did the reviewer come to see genuine ide-
alism everywhere?
The solution to this difﬁculty rests upon something that could have
been seen very easily from the context of the work, if one had wanted
to. Space and time, together with everything contained in them, are
not things (or properties of things) in themselves, but belong instead
merely to the appearances of such things; thus far I am of one creed
with the previous idealists. But these idealists, and among them espe-
cially Berkeley, viewed space as a merely empirical representation, a rep-
resentation which, just like the appearances in space together with all
of the determinations of space, would be known to us only by means
of experience or perception; I show, on the contrary, ﬁrst: that space
(and time as well, to which Berkeley gave no attention), together with
all its determinations, can be cognized by us a priori, since space (as
well as time) inheres in us before all perception or experience as a pure
form of our sensibility and makes possible all intuition from sensibility,
and hence all appearances. From this it follows: that, since truth rests
upon universal and necessary laws as its criteria, for Berkeley experience
could have no criteria of truth, because its appearances (according to
him) had nothing underlying them a priori; from which it then followed
that experience is nothing but sheer illusion, whereas for us space and
time (in combination with the pure concepts of the understanding) pre-
scribe a priori their law to all possible experience, which law at the same
time provides the sure criterion for distinguishing truth from illusion in
experience. ∗
∗
Genuine idealism always has a visionary purpose and can have no other; but
my idealism is solely for grasping the possibility of our a priori cognition of
the objects of experience, which is a problem that has not been solved before
now, nay, has not even once been posed. By that means all visionary idealism
162Prolegomena to any future metaphysics
My so-called (properly, critical) idealism is therefore of a wholly pecu-
liar kind, namely such that it overturns ordinary idealism, and such that
by means of it all cognition a priori, even that of geometry, ﬁrst acquires
objective reality, which, without my proven ideality of space and time,
could not have been asserted by even the most zealous of realists. With
matters standing so, I have wished that I could name this concept of
mine something else, in order to b prevent all misunderstanding; but this
concept cannot be completely changed. I may therefore be permitted in
the future, as has already been stated above, to call it formal, or better,
critical idealism, in order to distinguish it from the dogmatic idealism of
Berkeley and the skeptical idealism of Descartes.
I ﬁnd nothing else worthy of note in the review of this book. Its au-
thor judges en gros c throughout, a mode that is cleverly chosen, since
it does not betray one’s own knowledge or ignorance; a single com-
prehensive judgment en détail, d if, as is proper, it had considered the
main question, would have perhaps exposed my error, perhaps also the
degree of the reviewer’s insight into investigations of this kind. It was
no ill-considered trick, for removing early on the desire to read the
book itself from readers who are used to forming a conception of books
from newspaper articles only, to recite one after another a great many
propositions, which, torn from the context of their arguments and ex-
plications (especially as antipodean as these propositions are in rela-
tion to all school metaphysics), must of necessity sound nonsensical;
to assault the reader’s patience to the point of disgust; and then, after
having introduced me to the witty proposition that constant illusion
is truth, to conclude with the harsh, though paternal, reprimand: To
what end, then, the conﬂict with accepted e language, to what end, and
whence, the idealistic distinction? 83 A judgment that ultimately renders
everything peculiar to my book into merely verbal innovation (though
previously the book was supposed to be metaphysically heretical), and
that clearly proves that my would-be judge has not correctly understood
collapses, which (as was already to be seen with Plato) always inferred, from
our cognitions a priori (even those a of geometry), to another sort of intuition
(namely, intellectual) than that of the senses, since it did not occur to anyone
that the senses might also intuit a priori.
a
b
c
d
e
Reading denen for derer, with Ak.
Reading um for nun, with Ak.
“in the large”
“in detail”
Omitting the antecedent addition of gemein or gemeine by Ak and Vorländer (respectively)
as not required; Vorländer’s emendation was based on the draft ms. (Ak 23:55).
163
4: 376Prolegomena to any future metaphysics
4: 377
the least bit of it, and, what’s more, has not correctly understood
himself. ∗
The reviewer, however, talks like a man who must be aware of im-
portant and exquisite insights, which, however, he still keeps secret; for
nothing has become known to me of late regarding metaphysics that
could justify such a tone. But he is doing a great wrong in withholding
his discoveries from the world; for there are doubtless many others like
me who, with all the ﬁne things that have been written in this ﬁeld for
some time now, have still been unable to ﬁnd that the science has thereby
been advanced a ﬁnger’s breadth. In other respects, we do indeed ﬁnd
deﬁnitions being sharpened, lame proofs provided with new crutches,
the patchwork garment of metaphysics given new pieces, or an altered
cut – but that is not what the world demands. The world is tired of meta-
physical assertions; what’s wanted b are the possibility of this science, the
sources from which certainty could be derived in it, and sure criteria for
distinguishing truth from the dialectical illusion of pure reason. The re-
viewer must possess the key to all this, otherwise he surely would never
have spoken in so high a tone.
But I come to suspect that this sort of need of the science perhaps may
never have come into his head; for otherwise he would have directed his
review toward this point, and in such an important matter even a failed
attempt would have gained his respect. If that is so, then we are good
friends again. He may think himself as deeply into his metaphysics as
seems good to him, no one will stop him; only he is not permitted to
judge of something that lies outside metaphysics, i.e., its source located
in reason. But that my suspicion is not unfounded, I prove by the fact
that he did not say a word about the possibility c of synthetic cognition
a priori, which was the real problem, on the solution of which the fate
∗ The reviewer mostly ﬁghts his own shadow. When I oppose the truth of experi-
ence to dream, it never enters his head that the point of discussion is merely the
notorious somnio objective sumto a of the Wolfﬁan philosophy, which is merely
formal, and whereby no regard at all is given to the difference between sleeping
and waking, which also cannot be seen in transcendental philosophy. 84 More-
over, he calls my deduction of the categories and the table of principles of the
understanding, “commonly known principles of logic and ontology, expressed
in the manner of idealism.” 85 The reader need only examine these Prolegom-
ena on this subject to be convinced that a more deplorable, and even a more
historically incorrect judgment could not be given.
a “dreams taken objectively”
Not adopting Vorländer’s additions, loosely based on the the draft ms. (Ak 23:58, 62), of
untersucht wissen after werden könne, and haben after Criterien.
Reading Möglichkeit for Metaphysik, with Ak, as conﬁrmed by the draft ms. (Ak 23:62).
b
c
164Prolegomena to any future metaphysics
of metaphysics wholly rests, and to which my Critique (just as here my
Prolegomena) was entirely directed. The idealism upon which he chanced,
and to which he then held fast, was taken up into the system only as the
sole means for solving this problem (although it then also received its a
conﬁrmation on yet other grounds); and so he would have had to show
either that this problem does not have the importance that I attribute
to it (as also now in the Prolegomena), or that it could not be solved at
all by my concept of appearances, or could better be solved in another
way; but I ﬁnd not a word of this in the review. The reviewer therefore
understood nothing of my work and perhaps also nothing of the spirit
and nature of metaphysics itself, unless on the contrary, which I prefer to
assume, a reviewer’s haste, indignant at the difﬁculty of plowing his way
through so many obstacles, cast an unfavorable shadow over the work
lying before him and made it unrecognizable to him in its fundamentals.
There is still a great deal needed for a learned gazette, however well-
chosen and carefully selected its contributors may be, to be able to uphold
its otherwise well-deserved reputation in the ﬁeld of metaphysics (just as
elsewhere). Other sciences and areas of learning b have their standards.
Mathematics has its standard within itself, history and theology in sec-
ular or sacred books, natural science and medicine in mathematics and
experience, jurisprudence in law books, and even matters of taste in an-
cient paradigms. But in order to assess the thing called metaphysics, the
standard must ﬁrst be found (I have made an attempt to determine this
standard as well as its use). Until it is ascertained, what is to be done
when works of this kind must be judged? If they are of the dogmatic
kind, one may do as one likes; no one will for long play the master over
others in this without ﬁnding someone who repays him in kind. But if
they are of the critical kind, and not indeed with regard to other writings
but to reason itself, so that the standard of appraisal cannot be already
assumed but must ﬁrst be sought: then objection and censure are not
to be forbidden, but they must be rooted in tolerance, since the need is
common to us all, and the lack of the required insight makes an air of
judicial decisiveness unsuitable.
But in order at the same time to tie this my defense to the interest
of the philosophizing community, I propose a test, which is decisive as
to the way in which all metaphysical investigations must be directed
toward their common end. This is nothing else than what mathemati-
cians have done before, in order to decide the merits of their methods
in a contest – that is, a challenge to my reviewer to prove in his own
way any single truly metaphysical (i.e., synthetic, and cognized a priori
a
b
Reading seine for ihre, with Ak.
Kenntnisse
165
4: 378Prolegomena to any future metaphysics
4: 379
4: 380
from concepts) proposition a he holds, and at best one of the most indis-
pensable, such as the principle of the persistence of substance or of the
necessary determination of the events in the world through their cause
– but, as is ﬁtting, to prove it on a priori grounds. If he can’t do this (and
silence is confession), then he must admit: that, since metaphysics is ab-
solutely nothing without the apodictic certainty of propositions of this
sort, their possibility or impossibility would ﬁrst, before all else, have to
be settled in a critique of pure reason, and hence he is obliged either to
acknowledge that my principles of critique are correct or to prove their
invalidity. Since, however, I already foresee that, as heedlessly as he has
hitherto been relying on the certainty of his principles, still, now that it
comes down to a rigorous test, he will not ﬁnd a single principle in the
whole compass of metaphysics with which he can dare come forward, I
will therefore grant him the most favorable terms that can ever be ex-
pected in a competition; namely, I will take the onus probandi b from him
and will have it put on me.
In particular, in these Prolegomena and in my Critique, pp. 426–61, 86 he
will ﬁnd eight propositions which are, pair by pair, always in conﬂict with
one another, but each of which belongs necessarily to metaphysics, which
must either accept it or refute it (although there is not a single one of
them that has not in its day been accepted by some philosopher or other).
He now has the freedom to pick any one of these eight propositions he
likes, and to assume it without proof (which I concede to him), but only
one (for wasting time will be no more useful to him than to me), and
then he is to attack my proof of the antithesis. But if I can rescue it,
and in this way show that the opposite of the proposition he adopted
can be proven exactly as clearly, in accordance with principles that every
dogmatic metaphysics must of necessity acknowledge, then by this means
it is settled that there is an hereditary defect in metaphysics that cannot
be explained, much less removed, without ascending to its birthplace,
pure reason itself, and so my Critique must either be accepted or a better
one put in its place, and therefore it must at least be studied; which is
the only thing I ask for now. If, on the contrary, I cannot rescue my
proof, then a synthetic a priori proposition is established from dogmatic
principles on my opponent’s side, my indictment of ordinary metaphysics
was therefore unjust, and I offer to recognize his censure of my Critique as
legitimate (although this is far from being the likely outcome). But hereto
it would be necessary, I should think, to emerge from being incognito, since
I do not otherwise see how to prevent my being honored or assailed with
many problems from unknown and indeed unbidden opponents, instead
of just one. 87
a
b
Adding Satz, with Vorländer.
“burden of proof”
166Prolegomena to any future metaphysics
Proposal for an investigation of the Critique,
after which the judgment can follow
I am obliged to the learned public for the silence with which it has
honored my Critique for so long a time; for this after all demonstrates
a suspension of judgment, and thus some suspicion that, in a work that
abandons all the usual paths and pursues a new one in which one cannot
immediately ﬁnd one’s way, something might nonetheless perhaps be
found through which an important but now moribund branch of human
knowledge could receive new life and fertility, and so demonstrates a
cautiousness, not to break off and destroy the still fresh graft through an
overly hasty judgment. A specimen of a judgment that was delayed for
such reasons has only just now come before me in the Gothaische gelehrte
Zeitung, 88 a judgment whose well-foundedness every reader will perceive
for himself (without taking into account my own suspect praise) from
the clear and candid presentation of a portion of the ﬁrst principles of
my work.
And now I propose, since a large ediﬁce cannot possibly be instantly
judged as a whole through a quick once-over, that it be examined piece by
piece from its foundation, and that in this the present Prolegomena be used
as a general synopsis, with which the work itself could then be compared
on occasion. This suggestion, if it were based on nothing more than the
imagined importance that vanity customarily imparts to all one’s own
products, would be immodest and would deserve to be dismissed with
indignation. But the endeavors of all speculative philosophy now stand
at the point of total dissolution, although human reason clings to them
with undying affection, an affection that now seeks, though vainly, to turn
itself into indifference, only because it has been constantly betrayed.
In our thinking age it is not to be expected but that many merito-
rious men would use every good opportunity to work together toward
the common interest of an ever more enlightened reason, if only there
appears some hope of thereby attaining the goal. Mathematics, natural
science, law, the arts, even morals (and so on) do not completely ﬁll up
the soul; there still remains a space in it that is marked off for mere
pure and speculative reason, and its emptiness drives us to seek out, in
grotesques and trivialities, or else in delusions, what seems to be occu-
pation and amusement, but is at bottom only distraction to drown out
the troublesome call of reason, which, as beﬁts its vocation, demands
something that satisﬁes it for itself and does not merely stir it to activity
on behalf of other purposes or in the service of inclinations. Therefore,
for everyone who has even tried to enlarge his conception in this way,
contemplation that occupies itself only with this sphere of reason exist-
ing for itself has a great attraction, because exactly in this sphere all other
areas of learning and even ends must, as I have reason to suppose, join
167
4: 381Prolegomena to any future metaphysics
4: 382
together and unite in a whole – and, I dare say, it has a greater attraction
than any other theoretical knowledge, for which one would not readily
exchange it.
But I propose these Prolegomena as the plan and guide for the inves-
tigation, and not the work a itself, because, with respect to the latter,
though I am even now quite satisﬁed as regards the content, order, and
method, and the care that was taken to weigh and test each proposi-
tion accurately before setting it down (for it took years for me to be
fully satisﬁed not only with the whole, but sometimes also with only a
single proposition, as regards its sources), I am not fully satisﬁed with
my presentation in some chapters of the Doctrine of Elements, e.g., the
Deduction of the Concepts of the Understanding or the chapter on the
Paralogisms of Pure Reason, 89 since in them a certain prolixity obstructs
the clarity, and in their stead the examination can be based on what the
Prolegomena here say with respect to these chapters.
The Germans are praised for being able to advance things further than
other peoples in matters where persistence and unremitting industry are
called for. If this opinion is well-founded, then an opportunity presents
itself here to bring to completion an endeavor whose happy outcome
is hardly to be doubted and in which all thinking persons share equal
interest, but which has not succeeded before now – and to conﬁrm that
favorable opinion; especially since the science concerned is of such a
peculiar kind that it can be brought all at once to its full completion, and
into a permanent state such that it cannot be advanced the least bit further
and can be neither augmented nor altered by later discovery (herein I
do not include embellishment through enhanced clarity here and there,
or through added utility in all sorts of respects): an advantage that no
other science has or can have, since none is concerned with a cognitive
faculty that is so fully isolated from, independent of, and unmingled
with other faculties. The present moment does not seem unfavorable to
this expectation of mine, since in Germany nowadays one hardly knows
how he could keep himself otherwise still occupied outside the so-called
useful sciences and have it be, not mere sport, but at the same time an
endeavor through which an enduring goal is reached.
I must leave it to others to contrive the means by which the efforts
of the learned could be united toward such an end. In the meantime
it is not my intention to expect of anyone a simple adherence to my
theses, nor even to ﬂatter myself with hope of that; rather, whether it
should, as it happens, be attacks, revisions, and qualiﬁcations that bring
it about, or conﬁrmation, completion, and extension, if only the matter
is investigated from the ground up, then it now can no longer fail that
a
Reading das Werk for des Werks, with Ak.
168Prolegomena to any future metaphysics
a system would thereby come into being (even if it were not mine) that
could become a legacy to posterity for which it would have reason to be
thankful.
It would be too much to show here what sort of metaphysics could be
expected to follow if one were ﬁrst right about the principles of a critique,
and how it would by no means have to appear paltry and cut down to just a
small ﬁgure because its false feathers had been plucked, but could in other
respects appear richly and respectably outﬁtted; but other large beneﬁts
that such a reform would bring with it are apparent at once. The ordinary
metaphysics has indeed already produced beneﬁts, because it searched
for the elementary concepts of the pure understanding in order to render
them clear through analysis and determinate through explication. It was
thereby a cultivation of reason, wherever reason might subsequently
think ﬁt to direct itself. But that was all the good that it did. For it
undid this merit again by promoting self-conceit through rash assertions,
sophistry through subtle evasions and glosses, and shallowness through
the facility with which it overcame the most difﬁcult problems with a little
school wisdom – a shallowness that is all the more enticing the more it has
the option of, on the one hand, taking on something from the language
of science, and, on the other, from popularity, and thereby is everything
to everyone, but in fact is nothing at all. By contrast, through critique our
judgment is afforded a standard by which knowledge can be distinguished
with certainty from pseudo-knowledge; and, as a result of being brought
fully into play in metaphysics, critique establishes a mode of thinking
that subsequently extends its wholesome inﬂuence to every other use
of reason, and for the ﬁrst time excites the true philosophical spirit.
Moreover, the service it renders to theology, by making it independent of
the judgment of dogmatic speculation and in that way securing it against
all attacks from such opponents, is certainly not to be underrated. For the
ordinary metaphysics, although promising to assist theology greatly, was
subsequently unable to fulﬁll this promise, and beyond this, in calling
speculative dogmatism to its aid, had done nothing other than to arm
enemies against itself. Fanaticism, which cannot make headway in an
enlightened age except by hiding behind a school metaphysics, under
the protection of which it can venture, as it were, to rave rationally,
will be driven by critical philosophy from this its ﬁnal hiding place; and
beyond all this it cannot fail to be important to a teacher of metaphysics
to be able, for once with universal assent, to say that what he propounds
is now at last science, and that through it genuine beneﬁt is rendered to
the commonweal.
169
4: 383Metaphysical foundations of natural scienceTranslator’s introduction
The Metaphysische Anfangsgründe der Naturwissenschaft (hereinafter:
Metaphysical Foundations) ﬁrst appeared in 1786 (with second and third
printings in 1787 and 1800 respectively). This work thus belongs to the
most creative decade of Kant’s so-called critical period: the decade of the
ﬁrst edition of the Critique of Pure Reason (1781), the Prolegomena (1783),
the Groundwork of the Metaphysics of Morals (1785), the second edition of
the Critique of Pure Reason (1787), the Critique of Practical Reason (1788),
and ﬁnally the Critique of Judgement (1790). Until very recently, however,
the Metaphysical Foundations has had by far the least impact of any of these
works, and has accordingly attracted the least amount of scholarly atten-
tion. Both the content and the form of the work have contributed to this
situation. For, on the one hand, the Metaphysical Foundations is concerned
with relatively specialized questions belonging to natural philosophy and
even to physics: questions about the character and behavior of attractive
and repulsive forces, for example, or about impact and the communica-
tion of motion. And, on the other hand, it is written in an inhospitable
and forbidding style – organized in quasi-mathematical fashion into def-
initions (“explications”), propositions, proofs, remarks, and so on. In
both of these respects the Metaphysical Foundations is more akin to some
of Kant’s precritical writings on natural philosophy – especially the Phys-
ical Monadology (1756) – than to the great works of the critical period.
It is tempting, then, to dismiss the Metaphysical Foundations as a more or
less unaccountable regression into the outmoded preoccupations of the
precritical period – thereby both explaining and justifying its traditional
neglect. This, however, would be a serious mistake.
First of all, the close kinship between the Metaphysical Foundations and
Kant’s precritical writings on natural philosophy is of the greatest inter-
est. For we thereby learn more than we can from any other work of the
critical period about both the continuities and the discontinuities linking
the concerns of the later period with those of the earlier one. In both
the Physical Monadology and the Metaphysical Foundations, for example,
Kant formulates a so-called “dynamical theory of matter,” according to
which attractive and repulsive forces constitute the very nature or essence
of matter. In particular, the impenetrability of matter is grounded in a
force of repulsion rather than being taken as a primitive and irreducible
nondynamical property of hardness or solidity. Yet there is, nonethe-
less, a striking and fundamental difference between the two works in
173Metaphysical foundations of natural science
this regard. In the precritical Physical Monadology, Kant is still operating
within the context of the Leibniz–Wolfﬁan monadology. Matter consists
of ultimate simple substances which determine, through the “sphere of
activity” of their repulsive forces, elementary small volumes of space oth-
erwise empty of intervening material. Matter is thus only ﬁnitely divisible
into such elementary volumes. In the Metaphysical Foundations, by con-
trast, matter is explicitly taken to be continuous or inﬁnitely divisible,
and material substance, in particular, is now characterized precisely by
the impossibility of elementary monadic simple elements. Accordingly,
the problem posed by the inﬁnite divisibility of space that the Mon-
adology had solved by invoking ﬁnite “spheres of activity” is now solved
by invoking the transcendental idealism articulated in the Antinomy of
Pure Reason of the ﬁrst Critique – and, speciﬁcally, the argument of the
Second Antinomy. Matter is inﬁnitely divisible but never, in experience,
ever inﬁnitely divided; hence, since matter is a mere appearance and is
thus given only in the “progress of experience,” it consists neither in
ultimate simple elements nor in an actual or completed inﬁnity of ever
smaller spatial parts. Here, therefore, the deeply critical character of the
Metaphysical Foundations becomes evident – and in a way, moreover, that
illuminates both the relationship of the critical to the precritical period
and, via application to a concrete example, such characteristically critical
doctrines as the transcendental idealism of the Antinomies.
Indeed, there is an even more fundamental aspect of transcenden-
tal idealism that is centrally implicated here – the doctrine, namely, that
space is not a property or relation of things in themselves but a pure form
of our sensible intuition. In the precritical period of the Physical Mon-
adology, space is viewed as a derivative “phenomenon” of the relations (of
interaction) between the ultimate simple substances – substances that can
be conceived purely intellectually independently of all spatial properties.
The ultimate simple substances are not actually in space; rather, space
itself is ﬁrst constituted by their (interactive) relations. On the precritical
view, therefore, space, on broadly Leibnizean grounds, just is a relation
between “noumena” or things in themselves. In the critical period, how-
ever, Kant’s transcendental idealism about space is explicitly intended as a
middle ground between Leibnizean “relationalism” and the Newtonian
doctrine of absolute space (cf. A 39–41/B 56–58), and the Metaphysical
Foundations greatly illuminates the precise sense in which the critical posi-
tion secures this middle ground. In particular, Kant here attempts to pre-
serve the distinction between “true” and “apparent” motion required by
Newtonian mathematical physics without being thereby committed to a
metaphysically objectionable postulation of absolute space. Thus, by fo-
cusing from the outset on the problem of motion (on matter as “the mov-
able in space”), the Metaphysical Foundations shows how transcendental
idealism is intertwined with the profound conceptual problems about
174Translator’s introduction
the proper dynamical characterization of space, time, and motion that
divided the natural philosophies of Newton and Leibniz.
More generally, a continuous struggle aimed at integrating the
Leibnizean and Newtonian natural philosophies is clearly visible
throughout the precritical period: beginning with Kant’s ﬁrst published
work, Thoughts on the True Estimation of Living Forces (1747), continuing
through the Universal Natural History and Theory of the Heavens (1755)
and Physical Monadology, and on into the Attempt to Introduce the Concept of
Negative Magnitude into Philosophy (1763) and Enquiry Concerning the Clar-
ity of the Principles of Natural Theology and Ethics (1764). In the latter two
works the opposition between Leibniz–Wolfﬁanism and Newtonianism
is considered especially in its methodological aspect. For the Wolff-
ian tradition had promulgated a deductive, quasimathematical method
for philosophy, while simultaneously claiming that the actual results of
mathematical natural science (i.e., Newtonian natural science) must yield
pride of place to the more certain doctrines of metaphysics (i.e., Leibniz–
Wolfﬁan metaphysics). For the Newtonians, by contrast, metaphysics or
philosophy – if it is now even possible at all – must rather receive its ori-
entation from the more certain results of mathematical natural science:
metaphysics is now possible at all, that is, only by following an inductive
rather than a deductive method. Kant, in the methodological works of the
early 1760s, is in explicit agreement with the Newtonians in this respect:
metaphysics is possible only in the context of an “analytic” as opposed to
a “synthetic” method – whereby, in particular, we inductively follow the
secure results of mathematical natural science rather than attempting de-
ductively to legislate for such science. And this “Newtonian” conception
of metaphysical method ﬁnds clear echoes in the critical period as well –
especially in the Prolegomena, where the question, “How is metaphysics
in general possible?” is approached by way of the questions, “How is pure
mathematics possible?” and “How is pure natural science possible?”
We have now arrived at the second point I wish to emphasize: namely,
the central place of the Metaphysical Foundations within the critical period.
For the answer to the question concerning pure natural science posed by
the Prolegomena (which question, we should remember, corresponds to
the Transcendental Analytic of the ﬁrst Critique) is given its most explicit
and developed answer in the Metaphysical Foundations. Indeed, the task
of this work is to articulate in detail what Kant calls the “metaphysical
part” of pure natural science – the “metaphysical foundations of the doc-
trine of body.” Moreover, Kant’s preoccupation, at precisely this time,
with the metaphysical foundations of the doctrine of body is intimately
bound up with his further articulation of the critical philosophy. As is
well known, the ﬁrst (1781) edition of the Critique received a hostile
review from Garve and Feder in 1782, in which Kant was accused of
presenting a system of subjective idealism of the Berkeleyan variety. The
175Metaphysical foundations of natural science
Prolegomena of 1783 then explicitly replies to this review by emphasiz-
ing that Kantian “appearances” are bodies external to us in space and
that the doctrine of space as an a priori form of intuition is quite in-
compatible with Berkeley’s empiricist account of space. In this context,
the metaphysical foundations of the doctrine of body serve precisely to
emphasize that the objects whose possibility the critical philosophy is
to explain – i.e., “appearances” – are spatiotemporal objects of natural
scientiﬁc knowledge and, accordingly, that the point of the Critique is to
secure the a priori foundations of this type of knowledge rather than to
reduce the external physical world to a play of subjective mental states.
Indeed, the central importance of the Metaphysical Foundations in this
regard is clearly visible in the second (1787) edition of the Critique. This
is most evident in the General Remark to the System of the Principles
that Kant added there (B 288–294), which claims, in particular, that “in
order to understand the possibility of things in conformity with the cat-
egories, and thus to verify the objective reality of the latter, we require not
merely intuitions, but always even outer intuitions,” so that “in order to
give something permanent in intuition corresponding to the concept of
substance (and thereby to verify the objective reality of this concept) we
require an intuition in space (of matter)” (B 291). This passage is closely
related, in turn, to the Refutation of Idealism that Kant also added to the
second edition (B 274–279), which claims, in particular, that “we have
nothing permanent on which we could base the concept of substance, as
intuition, except only matter” (B 278). Both passages make it as explicit
as possible that the objects of cognition with which the transcendental
idealism of the ﬁrst Critique is concerned are, ﬁrst and foremost, phys-
ical objects or bodies – the objects of natural science – and thus, along
with other related additions and changes in the second edition, they
thereby function to distance Kant’s doctrine from the subjective ideal-
ism of Berkeley. And it is perhaps for this reason, above all, that Kant also
emphasizes, in a passage from the Preface to the Metaphysical Foundations
closely related to the above passage from the General Remark to the
System of Principles, that “general metaphysics, in all instances where it
requires examples (intuitions) in order to provide meaning for its pure
concepts of the understanding, must always take them from the gen-
eral doctrine of body,” which thereby “does excellent and indispensable
service for general metaphysics, in that the former furnishes examples
(instances in concreto) in which to realize the concepts and propositions
of the latter (properly speaking, transcendental philosophy), that is, to
give a mere form of thought sense and meaning.”
The third point I wish to emphasize is that Kant’s concern with the
foundations of natural science extends beyond the critical period as well.
From approximately 1796 until his death in 1804, Kant was struggling
with a work which he thought, at that time, was necessary to complete
176Translator’s introduction
the system of the critical philosophy. This work was never successfully
brought into publishable form, and has instead come down to us in a
collection of drafts, fragments, and sketches bound together as the Opus
postumum. Despite its fragmentary and unﬁnished character, however,
the Opus postumum is of the greatest interest in indicating how Kant en-
visioned a radical transformation of the critical philosophy toward the
end of his creative life. But what is of most importance, from our present
point of view, is the title of the proposed work that Kant was attempting
to draft: Transition from the Metaphysical Foundations of Natural Science to
Physics. As this title suggests, the transformation of the critical philos-
ophy with which Kant was struggling here took its impetus precisely
from a fundamental reconsideration of the philosophy of natural science
articulated in the Metaphysical Foundations of 1786. Thus, the Metaphys-
ical Foundations is of central importance, not only, as emphasized above,
for understanding both the relationship of the critical to the precritical
period and the internal development of the critical period itself, but also
for understanding the later trajectory and ultimate transformation of the
critical philosophy in what we might call Kant’s postcritical period. In-
deed, from this point of view, the Metaphysical Foundations appears as the
central work linking all three periods together. It is no wonder, then, that
serious scholarly attention to this work – and to Kant’s evolving philos-
ophy of natural science more generally – has been markedly increasing
in recent years (see the following bibliography).
There is one ﬁnal issue worth considering here: namely, the forbid-
ding quasi-mathematical form of the Metaphysical Foundations – its or-
ganization into deﬁnitions (“explications”), propositions, proofs, obser-
vations, and so on. This form is not merely an annoying hindrance to
the reader, it also poses a fundamental philosophical problem. For one
of the key doctrines of the Critique of Pure Reason is that there is an es-
sential distinction between mathematical and philosophical reasoning –
so that, in particular, it is entirely inappropriate for philosophy to
attempt to “imitate” mathematics (see the Discipline of Pure Reason
in its Dogmatic Employment: A 712–738/B 740–766). Indeed, as noted
above, we ﬁnd this doctrine already in the precritical methodological
works of the early 1760s, where Kant advocates a Newtonian inductive or
“analytic” style of metaphysics that is explicitly opposed to the deductive,
quasi-mathematical or “synthetic” style of metaphysics of the Wolfﬁan
school. So why does Kant, at the height of the critical period, employ
precisely such a quasi-mathematical style in the Metaphysical Foundations?
I cannot attempt adequately to answer this question here, but I can offer
some preliminary suggestions.
First, we should take this quasi-mathematical style, as stated above,
as presenting us with a philosophical problem – a problem that is con-
nected, in turn, with the content of the Metaphysical Foundations. In the
177Metaphysical foundations of natural science
Preface, for example, Kant explains that pure natural science has two
interdependent parts: a metaphysical part (the metaphysical foundations
of the doctrine of body), but also a mathematical part. For, as perhaps the
best-known sentence of the Metaphysical Foundations puts it, “in any spe-
cial doctrine of nature there can be only as much proper science as there
is mathematics therein.” For this reason, in fact, even the metaphysical
part of pure natural science must contain “principles for the construction
of the concepts that belong to the possibility of matter in general.” In-
deed, at one point Kant even appears to say that pure natural science
contains both “metaphysical and mathematical constructions” – thereby
seeming explicitly to contradict the doctrine of the ﬁrst Critique, noted
above, that philosophy, as opposed to mathematics, cannot proceed by
“construction.”
Second, although a quasi-mathematical organization into deﬁnitions,
propositions, proofs, and observations is quite rare in the critical period,
it is not entirely absent from other critical works. It is used, for example,
throughout §§1–8 of the Critique of Practical Reason (where we also ﬁnd
“corollaries” and “problems”) and, which is of more relevance to our
present concerns, in the Refutation of Idealism of the Critique of Pure
Reason. Due to the close connection, noted above, between the content
of the Refutation of Idealism and that of the Metaphysical Foundations,
this correspondence in form as well is of particular interest.
Third, although the quasi-mathematical style of the Metaphysical
Foundations exhibits the deductive form rejected in the methodologi-
cal works of the 1760s, this is by no means in the service of a Wolfﬁan
devaluation of mathematical-physical content. Indeed, as Kant himself
explains in the Preface, he uses a quasi-mathematical style here precisely
to encourage a closer connection between metaphysics of nature and
mathematical natural science:
In this treatise, although I have not followed the mathematical method with
thoroughgoing rigor (which would have required more time than I had to spend
thereon), I have nonetheless imitated that method – not in order to obtain a
better reception for the treatise through an ostentatious display of exactitude,
but rather because I believe that such a system is certainly capable of this rigor
and also that such perfection can certainly be reached in time by a more adept
hand if, stimulated by this sketch, mathematical natural scientists should ﬁnd it
not unimportant to treat the metaphysical part, which they cannot leave out in
any case, as a special fundamental part in their general physics and to bring it
into union with the mathematical doctrine of motion.
Accordingly, Kant’s own attitude towards the relative claims of meta-
physics and mathematical natural science is one of striking modesty:
Newton, in the preface to his Mathematical First Principles of Natural Science,
says (after he had remarked that geometry requires only two of the mechanical
178Translator’s introduction
operations that it postulates: namely, to describe a straight line and a circle):
Geometry is proud of the fact that with so little derived from without it is able to produce
so much. By contrast, one can say of metaphysics: it is dismayed that with so much
offered to it by pure mathematics it can still accomplish so little. Nevertheless, this
small amount is still something that even mathematics unavoidably requires in
its application to natural science, and thus, since it must here necessarily borrow
from metaphysics, need also not be ashamed to let itself be seen in community
with the latter.
Kant’s view of the relationship between metaphysics and mathematical
natural science – and, in particular, the mathematical natural science
of Newton’s Principia – could not be more different from that of the
Wolfﬁan school.
In carrying out the translation I have consulted the two previous English
renderings: those of E. Belfort Bax (London, 1883) and James Ellington
(Indianapolis and New York, 1970). Although he revises the earlier trans-
lation in several signiﬁcant respects, Ellington still follows Bax rather
closely. I have chosen, however, to deviate quite fundamentally from
both Ellington and Bax, and to begin again from scratch. First, I have
attempted to be as scrupulous as possible about respecting the nuances
of Kant’s technical terminology and phraseology – so that, in particu-
lar, Kant’s conceptual distinctions are reﬂected as faithfully as possible.
Second, I have also attempted, as far as possible, to render Kant’s sen-
tences into readable English – so as not to create additional problems
for the reader in grappling with an already forbidding text.
The translation is based on the text in volume 4 of the Akademie
Edition of Kant’s gesammelte Schriften, the page numbers of which appear
in the margins. Where I have deviated from that text in favor of the
original edition I have appended an explanatory note.
For help and advice on the translation I am indebted to Frederick
Beiser, Robert Butts, Eckart Förster, Hannah Ginsborg, Paul Guyer,
Peter McLaughlin, and Daniel Warren. For carefully reading and cor-
recting the entire text I am indebted to Martin Carrier, Martin Schönfeld
(who also helped with the notes and glossary), Henry Allison, and, espe-
cially, Peter Heath. I am of course solely responsible for any errors that
remain.
I am grateful for a Fellowship from the National Endowment for the
Humanities in support of this project during the academic year 1992–93.
bibliography
adickes, e. (1924). Kant als Naturforscher, Berlin.
beck, l. , ed. (1974). Kant’s Theory of Knowledge, Dordrecht.
brittan, g. (1978). Kant’s Philosophy of Science, Princeton.
179Metaphysical foundations of natural science
buchdahl, g. (1969). Metaphysics and the Philosophy of Science, Oxford.
busse, f. (1828). Metaphysische Anfangsgründe der Naturwissenschaft, Dresden
und Leipzig.
butts, r. , ed. (1986). Kant’s Philosophy of Physical Science, Dordrecht.
cramer, k. (1985). Nicht-reine synthetische Urteile a priori, Heidelberg.
drews, a. (1894). Kants Naturphilosophie als Grundlage seines Systems.
friedman, m. (1992). Kant and the Exact Sciences, Cambridge, Mass.
fries, j. (1882). Die mathematische Naturphilosophie nach philosophischer Methode
bearbeitet, Heidelberg.
gloy, k. (1976). Die Kantische Theorie der Naturwissenschaft, Berlin.
harman, p. (1982). Metaphysics and Natural Philosophy, Brighton.
hoppe, h. (1969). Kants Theorie der Physik, Frankfurt.
martin, g. (1951). Immanuel Kant. Ontologie und Wissenschaftslehre, Köln; trans-
lated as Kant’s Metaphysics and Theory of Science (Manchester, 1955).
plaass, p. (1965). Kants Theorie der Naturwissenschaft, Göttingen; translated as
Kant’s Theory of Natural Science (Dordrecht, 1994).
sch äfer, l. (1966). Kants Metaphysik der Natur, Berlin.
stadler, a. (1883). Kants Theorie der Materie, Leipzig.
st öhr, a. (1884). Analyse der reinen Naturwissenschaft Kants, Wien.
tuschling, b. (1971). Metaphysische und transzendentale Dynamik in Kants Opus
postumum, Berlin.
vuillemin, j. (1955). Physique et Métaphysique Kantiennes, Paris.
waidhaus, d. (1985). Kants System der Natur, Frankfurt.
wood, a. , ed. (1984). Self and Nature in Kant’s Philosophy, Ithaca.
180Metaphysical foundations of natural scienceContents
Preface
First Chapter: Metaphysical Foundations of Phoronomy
Second Chapter: Metaphysical Foundations of Dynamics
Third Chapter: Metaphysical Foundations of Mechanics
Fourth Chapter: Metaphysical Foundations of Phenomenology
183
194
209
245
260Preface
If the word nature is taken simply in its formal meaning, where it means
the ﬁrst inner principle of all that belongs to the existence of a thing, ∗
then there can be as many different natural sciences as there are speciﬁ-
cally different things, each of which must contain its own peculiar inner
principle of the determinations belonging to its existence. But nature is
also taken otherwise in its material meaning, not as a constitution, a but
as the sum total of all things, insofar as they can be objects of our senses, and
thus also of experience. Nature, in this meaning, is therefore understood
as the whole of all appearances, that is, the sensible world, excluding
all nonsensible objects. Now nature, taken in this meaning of the word,
has two principal parts, in accordance with the principal division of our
senses, where the one contains the objects of the outer senses, the other
the object of inner sense. In this meaning, therefore, a twofold doctrine
of nature is possible, the doctrine of body and the doctrine of the soul, where
the ﬁrst considers extended nature, the second thinking nature.
Every doctrine that is supposed to be a system, that is, a whole of
cognition ordered according to principles, is called a science. And, since
such principles may be either principles of empirical or of rational connec-
tion of cognitions into a whole, then natural science, be it the doctrine of
body or the doctrine of the soul, would have to be divided into historical
or rational natural science, were it not that the word nature (since this
signiﬁes a derivation of the manifold belonging to the existence of things
from their inner principle) makes necessary a cognition through reason
of the interconnection of natural things, insofar as this cognition is to
deserve the name of a science. Therefore, the doctrine of nature can be
better divided into historical doctrine of nature, which contains nothing
but systematically ordered facts about natural things (and would in turn
consist of natural description, as a system of classiﬁcation for natural things
in accordance with their similarities, and natural history, as a systematic
presentation of natural things at various times and places), and natural
∗ Essence is the ﬁrst inner principle of all that belongs to the possibility of a
thing. Therefore, one can attribute only an essence to geometrical ﬁgures, but
not a nature (since in their concept nothing is thought that would express an
existence).
a Beschaffenheit
183
4: 467
4: 468Metaphysical foundations of natural science
4: 469
science. Natural science would now be either properly or improperly so-
called natural science, where the ﬁrst treats its object wholly according
to a priori principles, the second according to laws of experience.
What can be called proper science is only that whose certainty is
apodictic; cognition that can contain mere empirical certainty is only
knowledge a improperly so-called. Any whole of cognition that is system-
atic can, for this reason, already be called science, and, if the connection
of cognition in this system is an interconnection of grounds and con-
sequences, even rational science. If, however, the grounds or principles
themselves are still in the end merely empirical, as in chemistry, for ex-
ample, and the laws from which the given facts are explained through
reason are mere laws of experience, then they carry with them no con-
sciousness of their necessity (they are not apodictally certain), and thus
the whole of cognition does not deserve the name of a science in the
strict sense; chemistry should therefore be called a systematic art rather
than a science.
A rational doctrine of nature thus deserves the name of a natural
science, only in case the fundamental natural laws therein are cognized
a priori, and are not mere laws of experience. One calls a cognition
of nature of the ﬁrst kind pure, but that of the second kind is called
applied rational cognition. Since the word nature already carries with
it the concept of laws, and the latter carries with it the concept of the
necessity of all determinations of a thing belonging to its existence, one
easily sees why natural science must derive the legitimacy of this title only
from its pure part – namely, that which contains the a priori principles
of all other natural explanations – and why only in virtue of this pure
part is natural science to be proper science. Likewise, [one sees] that,
in accordance with demands of reason, every doctrine of nature must
ﬁnally lead to natural science and conclude there, because this necessity
of laws is inseparably attached to the concept of nature, and therefore
makes claim to be thoroughly comprehended. Hence, the most complete
explanation of given appearances from chemical principles still always
leaves behind a certain dissatisfaction, because one can adduce no a priori
grounds for such principles, which, as contingent laws, have been learned
merely from experience.
All proper natural science therefore requires a pure part, on which the
apodictic certainty that reason seeks therein can be based. And because
this pure part is wholly different, in regard to its principles, from those
that are merely empirical, it is also of the greatest utility to expound
this part as far as possible in its entirety, separated and wholly unmixed
with the other part; indeed, in accordance with the nature of the case
it is an unavoidable duty with respect to method. This is necessary in
a
Wissen. Cf. “science [Wissenschaft]” in the previous sentence.
184Metaphysical foundations of natural science
order that one may precisely determine what reason can accomplish for
itself, and where its power begins to require the assistance of principles
of experience. Pure rational cognition from mere concepts is called pure
philosophy or metaphysics; by contrast, that which grounds its cognition
only on the construction of concepts, by means of the presentation of the
object in an a priori intuition, is called mathematics.
Properly so-called natural science presupposes, in the ﬁrst place, meta-
physics of nature. For laws, that is, principles of the necessity of that
which belongs to the existence of a thing, are concerned with a concept
that cannot be constructed, since existence cannot be presented a priori
in any intuition. Thus proper natural science presupposes metaphysics of
nature. Now this latter must always contain solely principles that are not
empirical (for precisely this reason it bears the name of a metaphysics),
but it can still either: ﬁrst, treat the laws that make possible the concept
of a nature in general, even without relation to any determinate object of
experience, and thus undetermined with respect to the nature of this or
that thing in the sensible world, in which case it is the transcendental part
of the metaphysics of nature; or second, concern itself with a particular na-
ture of this or that kind of things, for which an empirical concept is given,
but still in such a manner that, outside of what lies in this concept, no
other empirical principle is used for its cognition (for example, it takes the
empirical concept of matter or of a thinking being as its basis, and it seeks
that sphere of cognition of which reason is capable a priori concerning
these objects), and here such a science must still always be called a meta-
physics of nature, namely, of corporeal or of thinking nature. However,
[in this second case] it is then not a general, but a special metaphysical nat-
ural science (physics or psychology), in which the above transcendental
principles are applied to the two species of objects of our senses. 1
I assert, however, that in any special doctrine of nature there can be
only as much proper science as there is mathematics therein. For, according
to the preceding, proper science, and above all proper natural science,
requires a pure part lying at the basis of the empirical part, and resting on
a priori cognition of natural things. Now to cognize something a priori
means to cognize it from its mere possibility. But the possibility of de-
terminate natural things cannot be cognized from their mere concepts;
for from these the possibility of the thought (that it does not contradict
itself ) can certainly be cognized, but not the possibility of the object, as a
natural thing that can be given outside the thought (as existing). Hence,
in order to cognize the possibility of determinate natural things, and thus
to cognize them a priori, it is still required that the intuition corresponding
to the concept be given a priori, that is, that the concept be constructed.
Now rational cognition through construction of concepts is mathemat-
ical. Hence, although a pure philosophy of nature in general, that is,
that which investigates only what constitutes the concept of a nature in
185
4: 470Metaphysical foundations of natural science
4: 471
general, may indeed be possible even without mathematics, a pure doc-
trine of nature concerning determinate natural things (doctrine of body
or doctrine of soul) is only possible by means of mathematics. And, since
in any doctrine of nature there is only as much proper science as there is
a priori knowledge therein, a doctrine of nature will contain only as much
proper science as there is mathematics capable of application there.
So long, therefore, as there is still for chemical actions of matters on
one another no concept to be discovered that can be constructed, that is,
no law of the approach or withdrawal of the parts of matter can be speci-
ﬁed according to which, perhaps in proportion to their density or the like,
their motions and all the consequences thereof can be made intuitive and
presented a priori in space (a demand that will only with great difﬁculty
ever be fulﬁlled), then chemistry can be nothing more than a system-
atic art or experimental doctrine, but never a proper science, because its
principles are merely empirical, and allow of no a priori presentation in
intuition. Consequently, they do not in the least make the principles of
chemical appearances conceivable with respect to their possibility, for
they are not receptive to the application of mathematics.
Yet the empirical doctrine of the soul must remain even further from
the rank of a properly so-called natural science than chemistry. In the
ﬁrst place, because mathematics is not applicable to the phenomena of
inner sense and their laws, the only option one would have would be
to take the law of continuity in the ﬂux of inner changes into account –
which, however, would be an extension of cognition standing to that
which mathematics provides for the doctrine of body approximately as
the doctrine of the properties of the straight line stands to the whole
of geometry. For the pure inner intuition in which the appearances of
the soul are supposed to be constructed is time, which has only one
dimension. [In the second place,] however, the empirical doctrine of the
soul can also never approach chemistry even as a systematic art of analysis
or experimental doctrine, for in it the manifold of inner observation can
be separated only by mere division in thought, and cannot then be held
separate and recombined at will (but still less does another thinking
subject suffer himself to be experimented upon to suit our purpose), and
even observation by itself already changes and displaces the state of the
observed object. Therefore, the empirical doctrine of the soul can never
become anything more than an historical doctrine of nature, and, as
such, a natural doctrine of inner sense which is as systematic as possible,
that is, a natural description of the soul, but never a science of the soul,
nor even, indeed, an experimental psychological doctrine. This is also
the reason for our having used, in accordance with common custom, the
general title of natural science for this work, which actually contains the
principles of the doctrine of body, for only to it does this title belong in
the proper sense, and so no ambiguity is thereby produced. 2
186Metaphysical foundations of natural science
But in order to make possible the application of mathematics to the
doctrine of body, which only through this can become natural science,
principles for the construction of the concepts that belong to the possibility
of matter in general must ﬁrst be introduced. Therefore, a complete
analysis of the concept of a matter in general will have to be taken as
the basis, and this is a task for pure philosophy – which, for this purpose,
makes use of no particular experiences, but only that which it ﬁnds in the
isolated (although intrinsically empirical) concept itself, in relation to
the pure intuitions in space and time, and in accordance with laws that
already essentially attach to the concept of nature in general, and is
therefore a genuine metaphysics of corporeal nature.
Hence all natural philosophers who have wished to proceed mathe-
matically in their occupation have always, and must have always, made
use of metaphysical principles (albeit unconsciously), even if they them-
selves solemnly guarded against all claims of metaphysics upon their
science. Undoubtedly they have understood by the latter the folly of
contriving possibilities at will and playing with concepts, which can per-
haps not be presented in intuition at all, and have no other certiﬁcation
of their objective reality than that they merely do not contradict them-
selves. All true metaphysics is drawn from the essence of the faculty of
thinking itself, and is in no way ﬁctitiously invented a on account of not
being borrowed from experience. Rather, it contains the pure actions of
thought, and thus a priori concepts and principles, which ﬁrst bring the
manifold of empirical representations into the law-governed connection
through which it can become empirical cognition, that is, experience.
Thus these mathematical physicists could in no way avoid metaphysical
principles, and, among them, also not those that make the concept of
their proper object, namely, matter, a priori suitable for application to
outer experience, such as the concept of motion, the ﬁlling of space,
inertia, and so on. 3 But they rightly held that to let merely empirical
principles govern these concepts would in no way be appropriate to the
apodictic certainty they wished their laws of nature to possess, so they
preferred to postulate such [principles], without investigating them with
regard to their a priori sources.
Yet it is of the greatest importance to separate heterogeneous prin-
ciples from one another, for the advantage of the sciences, and to place
each in a special system so that it constitutes a science of its own kind,
in order to guard against the uncertainty arising from mixing things to-
gether, where one ﬁnds it difﬁcult to distinguish to which of the two
the limitations, and even mistakes, that might occur in their use may be
assigned. For this purpose I have considered it necessary [to isolate] the
former from the pure part of natural science ( physica generalis), where
a
erdichtet
187
4: 472
4: 473Metaphysical foundations of natural science
4: 474
4: 475
metaphysical and mathematical constructions customarily run together,
and to present them, together with principles of the construction of these
concepts (and thus principles of the possibility of a mathematical doc-
trine of nature itself ), in a system. 4 Aside from the already mentioned
advantage that it provides, this isolation has also a special charm arising
from the unity of cognition, when one takes care that the boundaries
of the sciences do not run together, but rather each takes in its own
separated ﬁeld.
The following can serve as still another ground for commending this
procedure. In everything that is called metaphysics one can hope for the
absolute completeness of the sciences, of such a kind one may expect in no
other type of cognition. Therefore, just as in the metaphysics of nature
in general, here also the completeness of the metaphysics of corporeal
nature can conﬁdently be expected. The reason is that in metaphysics the
object is only considered in accordance with the general laws of thought,
whereas in other sciences it must be represented in accordance with data
of intuition (pure as well as empirical), where the former, because here the
object has to be compared always with all the necessary laws of thought,
must yield a determinate number of cognitions that may be completely
exhausted, but the latter, because they offer an inﬁnite manifold of
intuitions (pure or empirical), and thus an inﬁnite manifold of objects of
thought, never attain absolute completeness, but can always be extended
to inﬁnity, as in pure mathematics and empirical doctrine of nature. I also
take myself to have completely exhausted this metaphysical doctrine of
body, so far as it may extend, but not to have thereby accomplished any
great [piece of ] work.
But the schema for completeness of a metaphysical system, whether it
be of nature in general, or of corporeal nature in particular, is the table of
categories. ∗ For there are no more pure concepts of the understanding
which can be concerned with the nature of things. All determinations of
the general concept of a matter in general must be able to be brought
∗
In the Allgemeine Literatur Zeitung, No. 295, in the review of Institutiones Logi-
cae et Metaphysicae by Prof. Ulrich, I ﬁnd doubts, which are not directed against
this table of pure concepts of the understanding, but rather against the infer-
ences drawn therefrom to the determination of the limits of the entire faculty
of pure reason, and thus all metaphysics, [doubts] with respect to which the
deeply delving reviewer declares himself to be in agreement with the no less
penetrating author. And, in fact, since these doubts are supposed to concern
precisely the principal basis of my system articulated in the Critique, they would
be grounds for thinking that this system, with respect to its principal aim, does
not come close to carrying that apodictic conviction that is required for elic-
iting an unqualiﬁed acceptance. This principal basis is said to be the deduction
of the pure concepts of the understanding, which is expounded partly in the
Critique and partly in the Prolegomena, and which, however, in the part of the
188Metaphysical foundations of natural science
under the four classes of [pure concepts of the understanding], those of
quantity, of quality, of relation, and ﬁnally of modality – and so, too, [must]
all that may be either thought a priori in this concept, or presented in
Critique that ought to be precisely the most clear, is rather the most obscure,
or even revolves in a circle, etc. I direct my reply to these objections only to
their principal point, namely, the claim that without an entirely clear and sufﬁcient
deduction of the categories the system of the Critique of Pure Reason totters on its
foundation. I assert, on the contrary, that the system of the Critique must carry
apodictic certainty for whoever subscribes (as the reviewer does) to my propo-
sitions concerning the sensible character of all our intuition, and the adequacy
of the table of categories, as determinations of our consciousness derived from
the logical functions in judgments in general, because it is erected upon the
proposition that the entire speculative use of our reason never reaches further than to
objects of possible experience. For if we can prove that the categories which reason
must use in all its cognition can have no other use at all, except solely in rela-
tion to objects of possible experience (insofar as they simply make possible the
form of thought in such experience), then, although the answer to the question
how the categories make such experience possible is important enough for
completing the deduction where possible, with respect to the principal end of
the system, namely, the determination of the limits of pure reason, it is in no
way compulsory, but merely meritorious. For the deduction is already carried far
enough for this purpose if it shows that categories of thought are nothing but
mere forms of judgments insofar as they are applied to intuitions (which for us
are always sensible), and that they thereby ﬁrst of all obtain objects and become
cognitions; because this already sufﬁces to ground with complete certainty the
entire system of Critique properly speaking. Thus Newton’s system of universal
gravitation stands ﬁrm, even though it involves the difﬁculty that one cannot
explain how attraction at a distance is possible; but difﬁculties are not doubts. That
the above fundamental basis stands ﬁrm, even without a complete deduction
of the categories, I now prove from the following granted propositions:
1. Granted: that the table of categories contains all pure concepts of the under-
standing, just as it contains all formal actions of the understanding in judging,
from which the concepts of the understanding are derived, and from which
they differ only in that, through the concepts of the understanding, an object
is thought as determined with respect to one or another function of judgment.
(Thus, for example, in the categorical judgement the stone is hard, the stone is
used as subject, and hard as predicate, in such a way that the understanding
is still free to exchange the logical function of these concepts, and to say that
something hard is a stone. By contrast, if I represent it to myself as determined
in the object that the stone must be thought only as subject, but hardness only as
predicate, in any possible determination of an object (not of the mere concept),
then the very same logical functions now become pure concepts of the understand-
ing of objects, namely, as substance and accident.)
2. Granted: that the understanding by its nature contains synthetic a priori
principles, through which it subjects all objects that may be given to it to
these categories, and, therefore, there must also be intuitions given a priori
189
4: 476
4: 475Metaphysical foundations of natural science
mathematical construction, or given as a determinate object of experi-
ence. There is no more to be done, or to be discovered, or to be added
4: 476
that contain the conditions required for the application of these pure concepts
of the understanding, because without intuition there can be no object, with re-
spect to which the logical function could be determined as category, and thus
no cognition of any object whatsoever, and hence without pure intuition no
principle that determines it a priori for this purpose.
3. Granted: that these pure intuitions can never be anything other than mere
forms of the appearances of outer or of inner sense (space and time), and there-
fore of the objects of possible experience alone.
It then follows: that all use of pure reason can never extend to anything other
than objects of experience, and, since nothing empirical can be the condition
of a priori principles, the latter can be nothing more than principles of the pos-
sibility of experience in general. This alone is the true and sufﬁcient basis for the
determination of the limits of pure reason, but not the solution to the problem
how experience is now possible by means of these categories, and only through
these categories alone. The latter problem, although without it the structure
still stands ﬁrm, has great importance nonetheless, and, as I now understand it,
[it can be solved with] just as much ease, since it can almost be accomplished
through a single inference from the precisely determined deﬁnition of a judg-
ment in general (an action through which given representations ﬁrst become
cognitions of an object). The obscurity that attaches to my earlier discussions
in this part of the deduction (and which I do not deny), is to be attributed to
the common fortunes of the understanding in its investigations, in which the
shortest way is commonly not the ﬁrst way that it becomes aware of. There-
fore, I shall take up the next opportunity to make up for this deﬁciency (which
concerns only the manner of presentation, and not the ground of explanation,
which is already stated correctly there), so that the perceptive reviewer may
not be left with the necessity, certainly unwelcome even to himself, of tak-
ing refuge in a preestablished harmony to explain the surprising agreement of
appearances with the laws of the understanding, despite their having entirely
different sources from the former. This remedy would be much worse than the
evil it is supposed to cure, and, on the contrary, actually cannot help at all. For
the objective necessity that characterizes the pure concepts of the understanding
(and the principles of their application to appearances), in the concept of cause
in connection with the effect, for example, is still not forthcoming. Rather, it
all remains only subjectively necessary, but objectively merely contingent, placing
together, precisely as Hume has it when he calls this mere illusion from custom.
No system in the world can derive this necessity from anywhere else than the
principles lying a priori at the basis of the possibility of thinking itself, through
which alone the cognition of objects whose appearance is given to us, that is, ex-
perience, becomes possible. Even if we suppose, therefore, that the explanation
of how experience thereby becomes possible in the ﬁrst place could never be
sufﬁciently carried out, it still remains incontrovertibly certain that it is possible
solely through these concepts, and, conversely, that these concepts are capable
of meaning and use in no other relation than to objects of experience.
190Metaphysical foundations of natural science
here, except, if need be, to improve it where it may lack in clarity or
exactitude. a
The concept of matter had therefore to be carried through all four
of the indicated functions of the concepts of the understanding (in four
chapters), where in each a new determination of this concept was added.
The basic determination of something that is to be an object of the
outer senses had to be motion, because only thereby can these senses be
affected. The understanding traces back all other predicates of matter
belonging to its nature to this, and so natural science, therefore, is either a
pure or applied doctrine of motion. The metaphysical foundations of natural
science are therefore to be brought under four chapters. The ﬁrst consid-
ers motion as a pure quantum in accordance with its composition, without
any quality of the movable, and may be called phoronomy. The second
takes into consideration motion as belonging to the quality of matter, un-
der the name of an original moving force, and is therefore called dynam-
ics. The third considers matter with this quality as in relation to another
through its own inherent motion, and therefore appears under the name
of mechanics. The fourth chapter, however, determines matter’s motion
or rest merely in relation to the mode of representation or modality, and
thus as appearance of the outer senses, and is called phenomenology.
Yet aside from the inner necessity to isolate the metaphysical foun-
dations of the doctrine of body, not only from physics, which needs
empirical principles, but even from the rational premises of physics that
concern the use of mathematics therein, there is still an external, cer-
tainly only accidental, but nonetheless important reason for detaching
its detailed treatment from the general system of metaphysics, and pre-
senting it systematically as a special whole. For if it is permissible to draw
the boundaries of a science, not simply according to the constitution of
the object and its speciﬁc mode of cognition, but also according to the
end that one has in mind for this science itself in uses elsewhere; and
if one ﬁnds that metaphysics has busied so many heads until now, and
will continue to do so, not in order thereby to extend natural knowledge
(which takes place much more easily and surely through observation,
experiment, and the application of mathematics to outer appearances),
but rather so as to attain cognition of that which lies wholly beyond
all boundaries of experience, of God, Freedom, and Immortality; then
one gains in the advancement of this goal if one frees it b from an off-
shoot that certainly springs from its root, but nonetheless only hinders
its regular growth, and one plants this offshoot specially, yet without
failing to appreciate the origin of [this offshoot] from it, c and without
a
b
c
Gründlichkeit
Sie. The reference is most likely to “metaphysics” – or possibly to “this science.”
Jener. See the previous note.
191
4: 477Metaphysical foundations of natural science
4: 478
4: 479
omitting the mature plant from the system of general metaphysics. This
does not impair the completeness of general metaphysics, and in fact
facilitates the uniform progress of this science toward its end, if, in all
instances where one requires the general doctrine of body, one may call
only upon the isolated system, without swelling this greater system with
the latter. It is also indeed very remarkable (but cannot be expounded in
detail here) 5 that general metaphysics, in all instances where it requires
examples (intuitions) in order to provide meaning for its pure concepts
of the understanding, must always take them from the general doctrine
of body, and thus from the form and principles of outer intuition; and, if
these are not exhibited completely, it gropes uncertainly and unsteadily
among mere meaningless concepts. This is the source of the well-known
disputes, or at least obscurity, in the questions concerning the possibility
of a conﬂict of realities, of intensive magnitude, and so on, in which the
understanding is taught only by examples from corporeal nature what
the conditions are under which such concepts can alone have objec-
tive reality, that is, meaning and truth. And so a separated metaphysics
of corporeal nature does excellent and indispensable service for general
metaphysics, in that the former furnishes examples (instances in concreto)
in which to realize the concepts and propositions of the latter (prop-
erly speaking, transcendental philosophy), that is, to give a mere form
of thought sense and meaning.
In this treatise, although I have not followed the mathematical method
with thoroughgoing rigor (which would have required more time than I
had to spend thereon), I have nonetheless imitated that method 6 – not in
order to obtain a better reception for the treatise, through an ostentatious
display of exactitude, a but rather because I believe that such a system is
certainly capable of this rigor, and also that such perfection can certainly
be reached in time by a more adept hand, if, stimulated by this sketch,
mathematical natural scientists should ﬁnd it not unimportant to treat
the metaphysical part, which they cannot leave out in any case, as a special
fundamental part in their general physics, and to bring it into union with
the mathematical doctrine of motion.
Newton, in the preface to his Mathematical First Principles of Natu-
ral Science, says (after he had remarked that geometry requires only two
of the mechanical operations that it postulates, namely, to describe a
straight line and a circle): Geometry is proud of the fact that with so little
7
derived from without it is able to produce so much. ∗, By contrast, one can
∗ “It is the glory of geometry that from those few principles, brought from
without, it is able to produce so many things.” Newton Princ. Phil. Nat. Math.
praefat.
a Gründlichkeit
192Metaphysical foundations of natural science
say of metaphysics: it is dismayed that with so much offered to it by pure
mathematics it can still accomplish so little. Nevertheless, this small amount
is still something that even mathematics unavoidably requires in its ap-
plication to natural science; and thus, since it must here necessarily bor-
row from metaphysics, need also not be ashamed to let itself be seen in
community with the latter.
193First Chapter
Metaphysical foundations of phoronomy
4: 480
EXPLICATION 8 1
Matter is the movable in space. That space which is itself movable is
called material, or also relative space; that in which all motion must ﬁnally
be thought (and which is therefore itself absolutely a immovable) is called
pure, or also absolute space.
Remark 1
Since in phoronomy nothing is to be at issue except motion, no other
property is here ascribed to the subject of motion, namely, matter, aside
from movability. It can itself so far, therefore, also be considered as a
point, and one abstracts in phoronomy from all inner constitution, and
therefore also from the quantity of the movable, and concerns oneself
only with motion and what can be considered as quantity in motion
(speed and direction). 9 – If the expression “body” should nevertheless
sometimes be used here, this is only to anticipate to some extent the
application of the principles of phoronomy to the more determinate
concepts of matter that are still to follow, so that the exposition may be
less abstract and more comprehensible.
4: 481
Remark 2
If I am to explicate the concept of matter, not through a predicate that
belongs to it itself as object, but only by relation to that cognitive faculty
in which the representation can ﬁrst of all be given to me, then every object
of the outer senses is matter, and this would be the merely metaphysical
explication thereof. Space, however, would be merely the form of all
outer sensible intuition (we here leave completely aside the question
whether just this form also belongs in itself to the outer object we call
matter, or remains only in the constitution of our sense). Matter, as
opposed to form, would be that in the outer intuition which is an object of
sensation, and thus the properly empirical element of sensible and outer
a
schlechterdings
194Metaphysical foundations of natural science
intuition, because it can in no way be given a priori. In all experience
something must be sensed, and that is the real of sensible intuition, and
therefore the space, in which we are to arrange our experience of motion,
must also be sensible – that is, it must be designated a through what can
be sensed – and this, as the totality of all objects of experience, and itself
an object of experience, is called empirical space. But this, as material, is
itself movable. But a movable space, if its motion is to be capable of being
perceived, presupposes in turn an enlarged material space, in which it is
movable; this latter presupposes in precisely the same way yet another;
and so on to inﬁnity.
Thus all motion that is an object of experience is merely relative; and
the space in which it is perceived is a relative space, which itself moves
in turn in an enlarged space, perhaps in the opposite direction, so that
matter moved with respect to the ﬁrst can be called at rest in relation to
the second space, and these variations in the concept of motions progress
to inﬁnity along with the change of relative space. To assume an absolute
space, that is, one such that, because it is not material, it can also not be an
object of experience, as given in itself, is to assume something, which can
be perceived neither in itself nor in its consequences (motion in absolute
space), for the sake of the possibility of experience – which, however, must
always be arranged without it. Absolute space is thus in itself nothing, and
no object at all, but rather signiﬁes b only any other relative space, which
I can always think beyond the given space, and which I can only defer to
inﬁnity beyond any given space, so as to include it and suppose it to be
moved. Since I have the enlarged, although still always material, space
only in thought, and since nothing is known to me of the matter that
designates it, I abstract from the latter, and it is therefore represented
as a pure, nonempirical, and absolute space, with which I compare any
empirical space, and in which I can represent the latter as movable (so
that the enlarged space always counts as immovable). To make this into
an actual thing is to transform the logical universality of any space with
which I can compare any empirical space, as included therein, into a
physical universality of actual extent, and to misunderstand reason in its
idea. 10
Finally, I further remark that, since the movability of an object in space
cannot be cognized a priori, and without instruction through experience,
I could not, for precisely this reason, enumerate it under the pure con-
cepts of the understanding in the Critique of Pure Reason; 11 and that this
concept, as empirical, could ﬁnd a place only in a natural science, as ap-
plied metaphysics, which concerns itself with a concept given through
experience, although in accordance with a priori principles.
a
b
bezeichnet
bedeutet
195
4: 482Metaphysical foundations of natural science
EXPLICATION 2
Motion of a thing is the change of its outer relations to a given space.
Remark 1
4: 483
I have so far placed the concept of motion at the basis of the concept of
matter. For, since I wanted to determine this concept independently of
the concept of extension, and could therefore consider matter also in a
point, I could allow the common explication of motion as change of place to
be used. Now, since the concept of a matter is to be explicated generally,
and therefore as beﬁtting also moving bodies, this deﬁnition is no longer
sufﬁcient. For the place of any body is a point. If one wants to determine
the distance of the moon from the earth, one wants to know the distance
between their places, and for this purpose one does not measure from
an arbitrary point of the surface or interior of the earth to any arbitrary
point of the moon, but chooses the shortest line from the central point of
the one to the central point of the other, so that for each of these bodies
there is only one point constituting its place. Now a body can move
without changing its place, as in the case of the earth rotating around
its axis. But its relation to external space still changes thereby; since it
turns, for example, its different sides toward the moon in 24 hours, from
which all kinds of varying effects then follow on the earth. Only of a
movable, that is, physical, point can one say that motion is always change
of place. One could object to this explication by pointing out that inner
motion, fermentation, for example, is not included; but the thing one
calls moving must to that extent be considered as a unity. For example,
that matter, as a cask of beer, is moved, means something different from
the beer in the cask being in motion. The motion of a thing is not the same
as motion in this thing, but here we are concerned only with the former
case. But the application of this concept to the second case is then easy.
Remark 2
Motions can be either rotating (without change of place) or progressive,
and the latter can be either motions that enlarge the space, or are limited
to a given space. Of the ﬁrst kind are the rectilinear motions, and also the
curvilinear motions that do not return on themselves. Of the second [kind]
are the motions that return on themselves. The latter in turn are either
circulating or oscillating, that is, either circular or oscillatory motions. The
former always traverse precisely the same space in the same direction,
the latter always alternately back in the opposite direction, as in the
case of oscillating pendulums. Belonging to both is still tremor (motus
tremulus), which is not the progressive motion of a body, but nonetheless
196Metaphysical foundations of natural science
a reciprocating motion of a matter, which does not thereby change its
place as a whole, as in the vibrations of a struck bell, or the tremors of
the air set in motion by the sound. I only make mention of these various
kinds of motion in phoronomy, because one commonly uses the word
speed, in the case of all nonprogressive [motions], in another meaning
than in the case of the progressive ones, as the following remark shows.
Remark 3
In every motion direction and speed are the two moments for consider-
ing motion, if one abstracts from all other properties of the movable. I
here presuppose the usual deﬁnitions of both, but that of direction still
requires various qualiﬁcations. A body moving in a circle changes its di-
rection continuously, in such a way that it follows all possible directions
in a surface along its way back to the point from which it set off, and yet
one says that it moves always in the same direction, for example, a planet
from west to east. a
But what is here the side toward which the motion is directed? This
has a kinship with the following question: On what rests the inner dif-
ference of snails, which are otherwise similar and even equal, but among
which one species is wound rightward, the other leftward; or the winding
of the kidney bean and the hop, where the ﬁrst runs around its pole like
a corkscrew, or, as sailors would express it, against the sun, whereas the
second runs with the sun? This is a concept which can certainly be con-
structed, but, as a concept, can in no way be made clear in itself by means
of universal characteristics and in the discursive mode of cognition, and
can yield no thinkable difference in the inner consequences in the things
themselves (for example, in the case of those unusual people where all
parts on dissection were found in agreement physiologically with other
humans, except that all organs were transposed leftward or rightward
contrary to the usual order), but is nevertheless a genuine mathemati-
cal, and indeed inner difference, which is connected with, although not
identical to, the difference between two circular motions that are other-
wise equal in all parts, but differ in direction. I have shown elsewhere 12
that, since this difference can certainly be given in intuition, but can
in no way be captured in clear concepts, and thus cannot be rationally
explicated (dari, non intelligi), it supplies a good conﬁrming ground of
proof for the proposition that space in general does not belong to the
properties or relations of things in themselves, which would necessarily
have to be reducible to objective concepts, b but rather belongs merely to
the subjective form of our sensible intuition of things or relations, which
a
b
von Abend gegen Morgen
die sich notwendig auf objective Begriffe mü ß ten bringen lassen
197
4: 484Metaphysical foundations of natural science
must remain completely unknown to us as to what they may be in them-
selves. Yet this is a digression from our present business, in which we
must necessarily treat space as a property of the things under considera-
tion, namely, corporeal beings, because these things are themselves only
appearances of the outer senses, and only require to be explicated as such
here. As far as the concept of speed is concerned, this expression some-
times acquires in use a deviant meaning. We say that the earth rotates
faster around its axis than the sun, because it does this in a shorter time,
although the motion of the latter is much faster. The circulation of the
blood in a small bird is much faster than that in a human being, although
its ﬂowing motion in the ﬁrst case has without any doubt less speed, and
so also in the case of vibrations in elastic materials. The brevity of the
time of return, whether it be circulating or oscillating motion, constitutes
the basis for this usage, which, so long as one avoids misunderstanding,
is also not incorrect. For this mere increase in the rapidity of the return,
without increase in spatial speed, has its own very important effects in
nature, concerning which, in the circulation of ﬂuids in animals, perhaps
not enough notice has yet been taken. In phoronomy we use the word
“speed” purely in a spatial meaning C = S/T. 13
EXPLICATION 3
4: 485
Rest is perduring presence a ( praesentia perdurabilis) at the same place;
what is perduring is that which exists throughout a time, that is, endures.
Remark
A body in motion is at every point of the line that it traverses for a
moment. The question is now whether it rests there or moves. Without
a doubt one will say the latter; for it is present at this point only insofar
as it moves. Assume, however, that the motion of the body is such:
that the body travels along the line AB with uniform speed forwards and
backwards from B to A, and that, since the moment when it is at B is
common to both motions, the motion from A to B is traversed in 1/2 sec.,
that from B to A also in 1/2 sec., and both together in one whole second,
so that not even the smallest part of the time pertains to the presence of
the body at B; then, without the least increase of these motions, the latter,
having taken place in the direction BA, can be transformed into that in
a
beharrliche Gegenwart
198Metaphysical foundations of natural science
the direction Ba, lying in a straight line with AB, in which case the body,
when it is at B, must be viewed as not at rest there, but as moved. It would
therefore have to be also viewed as moved at the point B in the ﬁrst case
of motion returning back on itself – which, however, is impossible; since
according to what was assumed, this point comprises only a moment –
belonging to the motion AB and simultaneously to the motion BA, which
is opposite to AB and joined to AB in one and the same moment – of
complete lack of motion. Therefore, if this constituted the concept of
rest, then rest of the body would also have to be manifested in the uniform
motion Aa at every point, for example, at B, which contradicts the above
assertion. Suppose, however, that one imagines the line AB as erected
above the point A, so that a body rising from A to B falls back again from
B to A after it has lost its motion at B through gravity; I then ask whether
the body at B can be viewed as moved or as at rest. Without a doubt one
will say at rest; for all previous motion has been taken from it once it has
reached this point, and after this an equivalent motion back is about to
follow, and thus is not yet there; but the lack of motion, one will add,
is rest. Yet in the ﬁrst case of an assumed uniform motion, the motion
BA could also not take place except through the fact that the motion AB
had previously ceased, and that from B to A was not yet there, so that a
lack of all motion at B had to be assumed, and, according to the usual
explication, also rest – but one may not assume it, because no body at
a point of its uniform motion at a given speed can be thought of as at
rest. So on what is based the appropriateness of the concept of rest in the
second case, where the rising and falling are likewise separated from one
another only by a moment? The reason for this lies in the fact that the
latter motion is not thought of as uniform at a given speed, but rather
ﬁrst as uniformly decelerated and thereafter as uniformly accelerated, 14
so that the speed at point B is not completely diminished, but only to a
degree that is smaller than any given speed. With this [speed], therefore,
the body, if it were to be viewed always as still rising, so that instead of
falling back the line of its fall BA were to be erected in the direction Ba,
would uniformly traverse, with a mere moment of speed (the resistance
of gravity here being set aside), a space smaller than any given space in
any given time, however large, and thus would in no way change its place
(for any possible experience) in all eternity. It is therefore put into the
state of an enduring presence at the same place, that is, a state of rest,
even though, because of the continual inﬂuence of gravity, that is, the
change of this state, it is immediately destroyed. To be in a perduring state
and to perdure in this state (if nothing else displaces it) are two different,
although not incompatible, concepts. Thus rest cannot be explicated as
lack of motion, which, as = 0, can in no way be constructed, but must
rather be explicated as perduring presence at the same place, since this
concept can also be constructed, through the representation of a motion
199
4: 486Metaphysical foundations of natural science
with inﬁnitely small speed throughout a ﬁnite time, and can therefore
be used for the ensuing application of mathematics to natural science.
EXPLICATION 4
To construct the concept of a composite motion means to present a motion
a priori in intuition, insofar as it arises from two or more given motions
united in one movable.
Remark
4: 487
It is required for the construction of concepts that the conditions of their
presentation not be borrowed from experience, and thus not presuppose
certain forces whose existence can only be derived from experience; or,
in general, that the condition of the construction must not itself be a
concept that can by no means be given a priori in intuition, such as, for
example, the concept of cause and effect, action and resistance, etc. Now
here it is above all to be noted that phoronomy has ﬁrst to determine the
construction of motions in general as quantities, and, since it has matter
merely as something movable as its object, in which no attention at all
is therefore paid to its quantity, [it has to determine] these motions a
priori solely as quantities, with respect to both their speed and direction,
and, indeed, with respect to their composition. For so much must be
constituted wholly a priori, and indeed intuitively, on behalf of applied
mathematics. For the rules for the connection of motions by means of
physical causes, that is, forces, can never be rigorously expounded, until
the principles of their composition in general have been previously laid
down, purely mathematically, as basis.
PRINCIPLE
Every motion, as object of a possible experience, can be viewed arbitrarily
as motion of the body in a space at rest, or else as rest of the body, and,
instead, as motion of the space in the opposite direction with the same
speed.
Remark
To make the motion of a body into an experience, it is required that not
only the body, but also the space in which it moves, be objects of outer
experience, and thus material. Hence an absolute motion, that is, [one]
in relation to a nonmaterial space, cannot be experienced at all, and thus
is nothing for us (even if one wanted to grant that absolute space were
something in itself ). But in all relative motion the space itself, since it is
assumed to be material, can in turn be represented as either at rest or as
200Metaphysical foundations of natural science
moved. The ﬁrst occurs when, beyond the space in which I view the body
as moved, no further enlarged space is given to me that includes it (as
when I see a ball moving on the table in the cabin of a ship); the second
occurs when, beyond the given space, another space that includes it is
given (in the example mentioned, the bank of the river), since I can then,
in relation to the latter, view the nearest space (the cabin) as moved, and
the body itself as possibly at rest. Now because it is completely impossible
to determine for an empirically given space, no matter how enlarged it
may be, whether it may or may not be moved in turn, in relation to
an inclusive space of still greater extent, it must then be completely the
same for all experience, and every consequence of experience, whether
I wish to view a body as moved, or as at rest, but the space as moved
in the opposite direction with the same speed. Further, since absolute
space is nothing for all possible experience, the concepts are also the
same whether I say that a body moves in relation to this given space,
in such and such direction with such and such speed, or I wish to think
the body as at rest, and to ascribe all this, but in the opposite direction,
to the space. For any concept is entirely the same as a concept whose
differences from it have no possible example at all, being only different
with respect to the connection we wish to give it in the understanding.
We are also incapable, in any experience at all, of assigning a ﬁxed
point in relation to which it would be determined what motion and rest
are to be absolutely; for everything given to us in this way is material, and
thus movable, and (since we are acquainted with no outermost limit of
possible experience in space) is perhaps also actually moved, without our
being able to perceive this motion. – Of this motion of a body in empirical
space, I can give a part of the given speed to the body, and the other to the
space, but in the opposite direction, and the whole possible experience,
with respect to the consequences of these two combined motions, is
entirely the same as that experience in which I think the body as alone
moved with the whole speed, or the body as at rest and the space as
moved with the same speed in the opposite direction. But here I assume
all motions to be rectilinear. For in regard to curvilinear motions, it is not in
all respects the same whether I am authorized to view the body (the earth
in its daily rotation, for example) as moved and the surrounding space
(the starry heavens) as at rest, or the latter as moved and the former as at
rest, which will be specially treated in what follows. Thus in phoronomy,
where I consider the motion of a body only in relation to the space (which
has no inﬂuence at all on the rest or motion of the body), it is completely
undetermined and arbitrary how much speed, if any, I wish to ascribe
to the one or to the other. Later, in mechanics, where a moving body
is to be considered in active relation to other bodies in the space of its
motion, this will no longer be entirely the same, as will be shown in the
proper place. 15
201
4: 488Metaphysical foundations of natural science
4: 489
EXPLICATION 5
The composition of motion is the representation of the motion of a point
as the same as two or more motions of [this point] combined together.
Remark
4: 490
In phoronomy, since I am acquainted with matter through no other prop-
erty but its movability, and may thus consider it only as a point, motion
can only be considered as the describing of a space – in such a way, however,
that I attend not solely, as in geometry, to the space described, but also
to the time in which, and thus to the speed with which, a point describes
the space. Phoronomy is thus the pure theory of quantity (mathesis) of
motions. The determinate concept of a quantity is the concept of the
generation of the representation of an object through the composition
of the homogeneous. 16 Now since nothing is homogeneous with motion
except motion in turn, phoronomy is a doctrine of the composition of
the motions of one and the same point in accordance with its speed and
direction, that is, the representation of a single motion as one that con-
tains two or more motions at the same time, or two motions of precisely
the same point at the same time, insofar as they constitute one motion
combined – that is, [they] are the same as the latter, and do not, for exam-
ple, produce it, as causes produce their effect. In order to ﬁnd the motion
arising from the composition of several motions, as many as one wishes,
one need only, as in all generation of quantity, ﬁrst seek for that motion
which, under the given conditions, is composed from two motions; this
is then combined with a third; and so on. Therefore, the doctrine of
the composition of all motions can be reduced to that of two. But two
motions of one and the same point, which are found there at the same
time, can be distinguished in two ways, and, as such, can be combined
there in three ways. First, they occur either in one and the same line, or
in different lines at the same time; the latter are motions comprising an
angle. Those occurring in one and the same line are now, with respect to
direction, either opposite to one another or have the same direction. Be-
cause all these motions are considered as occurring at the same time, the
ratio of speed results immediately from the ratio of the lines, that is, from
the described spaces of the motion in the same time. There are therefore
three cases: (1) The two motions (they may have equal or unequal speeds)
are to constitute a composite motion combined in one body in the same
direction. (2) The two motions of the same point (of equal or unequal
speed), combined in opposite directions, are to constitute, through their
composition, a third motion in the same line. (3) The two motions of a
point are considered as composed with equal or unequal speeds, but in
different lines comprising an angle.
202Metaphysical foundations of natural science
PROPOSITION
The composition of two motions of one and the same point can only be
thought in such a way that one of them is represented in absolute space,
and, instead of the other, a motion of the relative space with the same
speed occurring in the opposite direction is represented as the same as
the latter.
Proof
First case. Two motions in one and the same line and direction belong to one
and the same point.
Two speeds AB and ab are to be represented as contained in one speed
of motion. If one assumes these speeds to be equal for the moment, so
that AB = ab, then I say that they cannot be represented at the same
time in one and the same space (whether absolute or relative) in one and
the same point. For, since the lines AB and ab designating the speeds
are, properly speaking, the spaces they traverse in equal times, then the
composition of these spaces AB and ab = BC, and hence the line AC as
the sum of the spaces, would have to express the sum of the two speeds.
But neither the part AB nor the part BC represents the speed = ab, for
they are not traversed in the same time as ab. Therefore, the doubled
line AC, traversed in the same time as the line ab, does not represent the
twofold speed of the latter, which, however, was required. Therefore,
the composition of two speeds in one direction cannot be represented
intuitively in the same space.
By contrast, if the body A with speed AB is represented as moved in
absolute space, and, moreover, I give to the relative space a speed ab =
AB in the opposite direction ba = CB, then this is precisely the same
as if I had imparted the latter speed to the body in the direction AB
(Principle). The body then moves through the sum of the lines AB and
BC = 2ab in the same time in which it would have traversed the line
ab = AB alone, and its speed is thus represented as the sum of the two
equal speeds AB and ab, which is what was required.
Second case. Two motions are to be combined in precisely opposite direc-
tions in one and the same point.
203
4: 491Metaphysical foundations of natural science
4: 492
Let AB be one of these motions and AC the other in the opposite
direction, whose speed we will assume here as equal to the ﬁrst. Then
even the thought of representing two such motions at the same time in
exactly the same point within one and the same space would be impossi-
ble, and thus so would the case of such a composition of motions itself,
which is contrary to the presupposition.
By contrast, think instead the motion AB in absolute space, but, in-
stead of the motion AC in the same absolute space, the opposite motion
CA of the relative space with exactly the same speed, which (according
to the Principle) counts as entirely the same as the motion AC, and can
therefore be posited wholly in place of the latter. Then two precisely
opposite and equal motions of the same point at the same time can per-
fectly well be represented. Because the relative space now moves with the
same speed CA = AB in the same direction as the point A, this point, or
the body found there, does not change its place in relation to the relative
space. That is, a body moved in two exactly opposite directions with the
same speed is at rest, or, expressed in general: its motion is equal to the
difference of the speeds in the direction of the greater (which can easily
be concluded from what has been proved).
Third case. Two motions of one and the same point are to be repre-
sented as combined in directions comprising an angle.
The two given motions are AB and AC, whose speeds and direc-
tions are expressed by these lines, but the angle comprised by these lines
204Metaphysical foundations of natural science
[is expressed] by BAC (it may, as here, be a right angle, but also any
arbitrary oblique angle). Now if these two motions were to occur at the
same time in the directions AB and AC, and in one and the same space,
then they would still not be able to occur at the same time in these
two lines AB and AC, but only in lines running parallel to these. It would
therefore have to be assumed that one of these motions effected a change
in the other (namely, directing it from the given path), if both directions
were to remain the same. But this is contrary to the presupposition of
the Proposition, which indicates by the word “composition” that the two
given motions are to be contained in a third, and therefore are to be the
same as the latter, and are not to produce a third, in that one changes the
other.
By contrast, assume the motion AC as proceeding in absolute space,
but, instead of the motion AB, the motion of the relative space in the
opposite direction. Let the line AC be divided into three equal parts AE,
EF, FC. While the body A traverses the line AE in absolute space, the
relative space, together with the point E, traverses the space Ee = MA.
While the body traverses the two parts together = AF, the relative space,
together with the point F, describes the line Ff = NA. Finally, while the
body traverses the whole line AC, the relative space, together with the
point C, describes the line Cc = BA. But all of this is precisely the same
as if the body A had traversed the lines Em, Fn, and CD = AM, AN,
AB in these three parts of the time, and in the whole time, in which it
traverses AC, the line CD = AB. It is thus in the last moment at the point
D, and in this whole time successively at all points of the diagonal AD,
which therefore expresses both the direction and speed of the composite
motion.
Remark 1
Geometrical construction requires that one quantity be the same as another
or that two quantities in composition be the same as a third, not that they
produce the third as causes, which would be mechanical construction.
Complete similarity and equality, insofar as it can be cognized only in
intuition, is congruence. All geometrical construction of complete identity
rests on congruence. Now this congruence of two combined motions
with a third (as with the motus compositus itself ) can never take place if
these two combined motions are represented in one and the same space,
for example, in relative space. Therefore, all attempts to prove the above
Proposition in its three cases were always only mechanical analyses –
namely, where one allows moving causes to produce a third motion by
combining one given motion with another 17 – but not proofs that the
two motions are the same as the third, and can be represented as such a
priori in pure intuition.
205
4: 493Metaphysical foundations of natural science
Remark 2
4: 494
If, for example, a speed AC is called doubled, nothing else can be un-
derstood by this except that it consists of two simple and equal speeds
AB and BC (see Figure 1). If, however, one explicates a doubled speed
by saying that it is a motion through which a doubled space is traversed
in the same time, then something is assumed here that is not obvious in
itself – namely, that two equal speeds can be combined in precisely the
same way as two equal spaces – and it is not clear in itself that a given
speed consists of smaller speeds, and a rapidity of slownesses, in precisely
the same way that a space consists of smaller spaces. For the parts of the
speed are not external to one another like the parts of the space, and
if the former is to be considered as a quantity, then the concept of its
quantity, since this is intensive, must be constructed in a different way
from that of the extensive quantity of space. 18 But this construction is
possible in no other way than through the mediate composition of two
equal motions, such that one is the motion of the body, and the other the
motion of the relative space in the opposite direction, which, however,
for precisely this reason, is entirely the same as a motion of the body in
the original direction that is equal to it. For two equal speeds cannot be
combined in the same body in the same direction, except through external
moving causes, for example, a ship, which carries the body with one of
these speeds, while another moving force combined immovably with the
ship impresses on the body the second speed that is equal to the ﬁrst.
But here it must always be presupposed that the body conserves itself
in free motion with the ﬁrst speed, while the second is added – which,
however, is a law of nature of moving forces that can in no way be at
issue here, where the question is solely how the concept of speed as a
quantity is to be constructed. So much, then, for the addition of speeds
to one another. If, however, the subtraction of one from the other is
at issue, then this can indeed easily be thought, as soon as the possibil-
ity of speed as a quantity through addition is granted, but this concept
cannot so easily be constructed. For, to this end two opposite motions
must be combined in one body; and how is this supposed to happen? It
is impossible to think two equal motions in the same body in opposite
directions immediately, that is, in relation to precisely the same space at
rest. But the representation of the impossibility of these two motions in
one body is not the concept of its rest, but rather of the impossibility of
constructing this composition of opposite motions, which is nonetheless
assumed as possible in the Proposition. This construction is possible in
no other way, however, except through the combination of the motion
of the body with the motion of the space, as was shown. Finally, with re-
spect to the composition of two motions with directions comprising an
angle, this cannot be thought in the body in reference to one and the
206Metaphysical foundations of natural science
same space either, unless we assume that one of them is effected through
an external continually inﬂuencing force (for example, a vehicle carrying
the body forward), while the other is conserved unchanged – or, in gen-
eral, one must take as basis moving forces, and the generation of a third
motion from two united forces, which is indeed the mechanical execution
of what is contained in a concept, but not its mathematical construction,
which should only make intuitive what the object (as quantum) is to be,
not how it may be produced by nature or art by means of certain instru-
ments and forces. – The composition of motions, in order to determine
their ratio to others as quantity, must take place in accordance with the
rules of congruence, which is only possible in all three cases by means
of the motion of the space congruent to one of the two given motions,
so that the two [together] are congruent to the composite [motion].
Remark 3
Phoronomy, not as pure doctrine of motion, but merely as pure doctrine
of the quantity of motion, in which matter is thought with respect to
no other property than its mere movability, therefore contains no more
than this single Proposition, carried out through the above three cases, of
the composition of motion – and, indeed, of the possibility of rectilinear
motions only, not curvilinear [ones]. For since in these latter the motion is
continually changed (in direction), a cause of this change must be brought
forward, which cannot now be the mere space. But that one normally
understood, by the term composite motion, only the single case where the
directions comprise an angle, did no harm to physics, but rather to the
principle of classiﬁcation of a pure philosophical science in general. For,
with respect to the former, all three cases treated in the above Proposition
can be sufﬁciently presented in the third alone. For if the angle comprised
by the two given motions is thought as inﬁnitely small, then it contains
the ﬁrst case; but if it is represented as different from a single straight
line only by an inﬁnitely small amount, then it contains the second case;
so that all three cases named by us can certainly be given in the well-
known proposition of composite motion in a general formula. But one
could not, in this way, learn to comprehend the doctrine of the quantity
of motion a priori with respect to its parts, which also has its uses for
several purposes.
If anyone is interested in connecting the above three parts of the
general phoronomic Proposition with the schema of classiﬁcation of all
pure concepts of the understanding – namely, here that of the concept
of quantity – then he will note that, since the concept of quantity always
contains that of the composition of the homogeneous, the doctrine of
the composition of motion is, at the same time, the pure doctrine of the
quantity of motion, and, indeed, in accordance with all three moments
207
4: 495Metaphysical foundations of natural science
suggested by [the structure of ] space: unity of line and direction, plurality
of directions in one and the same line, and the totality of directions, as
well as lines, in accordance with which the motion may occur, which
contains the determination of all possible motion as a quantum, even
though the quantity of motion (in a movable point) consists merely in
the speed. This remark has its uses only in transcendental philosophy.
208Second Chapter
Metaphysical foundations of dynamics
4: 496
EXPLICATION 1
Matter is the movable insofar as it ﬁlls a space. To ﬁll a space is to resist
every movable that strives through its motion to penetrate into a certain
space. A space that is not ﬁlled is an empty space.
Remark
This is now the dynamical explication of the concept of matter. It pre-
supposes the phoronomical [explication], but adds a property relating as
cause to an effect, namely, the power to resist a motion within a certain
space; there could be no mention of this in the preceding science, not
even when dealing with motions of one and the same point in opposite
directions. This ﬁlling of space keeps a certain space free from the pen-
etration of any other movable, when its motion is directed toward any
place in this space. Now the basis for the resistance of matter exerted
in all directions, a and what this resistance is, must still be investigated.
But one already sees this much from the above explication: Matter is
not here considered as it resists, when it is driven out of its place, and thus
moved itself (this case will be considered later, as mechanical resistance),
but rather when merely the space of its own extension is to be dimin-
ished. One uses the expression to occupy a space, that is, to be immediately
present in all points of this space, in order to designate the extension of a
thing in space. However, it is not determined in this concept what effect
arises from this presence, or even whether there is any effect at all –
whether to resist others that are striving to penetrate within; or whether
it means merely a space without matter, in so far as it is a complex of
several spaces, as one can say of any geometrical ﬁgure that it occupies
a space by being extended; or even whether there is something in the
space that compels another movable to penetrate deeper into it (by at-
tracting others) – because, I say, all this is undetermined by the concept
of occupying a space, ﬁlling a space is a more speciﬁc determination of
the concept of occupying a space.
a
nach allen Seiten gerichtete
209
4: 497Metaphysical foundations of natural science
PROPOSITION 1
Matter ﬁlls a space, not through its mere existence, but through a particular
moving force.
Proof
Penetration into a space (in the initial moment this is called a striving to
penetrate) is a motion. Resistance to motion is the cause of its diminu-
tion, or even of the change of this motion into rest. Now nothing can
be combined with a motion, which diminishes it or destroys it, except
another motion of precisely the same movable in the opposite direction
(Phoron. Prop.). Therefore, the resistance that a matter offers in the
space that it ﬁlls to every penetration by other matters is a cause of the
motion of the latter in the opposite direction. But the cause of a motion
is called a moving force. Thus matter ﬁlls its space through a moving
force, and not through its mere existence.
Remark
4: 498
Lambert and others called the property of matter by which it ﬁlls a space
solidity (a rather ambiguous expression), and claim that one must assume
this in every thing that exists (substance), at least in the outer sensible
world. According to their ideas a the presence of something real in space 19
must already, through its concept, and thus in accordance with the prin-
ciple of noncontradiction, imply this resistance, and bring it about that
nothing else can be simultaneously in the space where such a thing is
present. But the principle of noncontradiction does not repel a matter
advancing to penetrate into a space where another is found. Only when I
ascribe to that which occupies a space a force to repel every external mov-
able that approaches, do I understand how it contains a contradiction for
yet another thing of the same kind to penetrate into the space occupied
by a thing. Here the mathematician has assumed something, as a ﬁrst
datum for constructing the concept of a matter, which is itself incapable
of further construction. Now he can indeed begin his construction of a
concept from any chosen datum, without engaging in the explication of
this datum in turn. But he is not therefore permitted to declare this to
be something entirely incapable of any mathematical construction, b so
as thereby to obstruct us from going back to ﬁrst principles in natural
science.
a
b
Begriffen
jenes für etwas aller mathematischen Construction ganz Unfähiges zu erklären
210Metaphysical foundations of natural science
EXPLICATION 2
Attractive force is that moving force by which a matter can be the cause
of the approach of others to it (or, what is the same, by which it resists
the removal of others from it).
Repulsive force is that by which a matter can be the cause of others
removing themselves from it (or, what is the same, by which it resists the
approach of others to it). The latter force will also sometimes be called
driving force, the former drawing force.
Note
Only these two moving forces of matter can be thought. For all motion
that one matter can impress on another, since in this regard each of them
is considered only as a point, must always be viewed as imparted in the
straight line between the two points. But in this straight line there are
only two possible motions: the one through which the two points remove
themselves from one another, the second through which they approach
one another. But the force causing the ﬁrst motion is called repulsive force,
whereas the second is called attractive force. Therefore, only these two
kinds of forces can be thought, as forces to which all moving forces in
material nature must be reduced.
PROPOSITION 2
Matter ﬁlls its space through the repulsive forces of all of its parts, that
is, through an expansive force of its own, having a determinate degree,
such that smaller or larger degrees can be thought to inﬁnity.
Proof
Matter ﬁlls a space only through moving force (Prop. 1), a force resisting
the penetration (that is, the approach) of others. Now this is a repulsive
force (Explication 2). Therefore, matter ﬁlls its space only through re-
pulsive forces, and, indeed, through repulsive forces of all of its parts.
For otherwise a part of its space (contrary to the presupposition) would
not be ﬁlled, but only enclosed. But the force of something extended in virtue
of the repulsion of all of its parts is an expansive force. So matter ﬁlls its space
only through an expansive force of its own, which was the ﬁrst [thing to be
shown]. Now, beyond any given force a greater force must be thinkable,
for that force beyond which no greater is possible would be one whereby
an inﬁnite space would be traversed in a ﬁnite time (which is impossible).
Further, below any given force a smaller force must be thinkable (for
the smallest force would be one whereby its inﬁnite addition to itself
211
4: 499Metaphysical foundations of natural science
throughout a given time could generate no ﬁnite speed, which, however,
means the absence of all moving force). Thus, below any given degree
of a moving force a smaller must always be capable of being given, which
is the second [thing to be shown]. Consequently, the expansive force by
which every matter ﬁlls its space has a degree, which is never the great-
est or the smallest, but is such that beyond it both greater and smaller
degrees can be found to inﬁnity.
4: 500
Note 1
The expansive force of a matter is also called elasticity. Now, since it
is the basis on which the ﬁlling of space rests, as an essential property
of all matter, this elasticity must therefore be called original, because it
can be derived from no other property of matter. All matter is therefore
originally elastic.
Note 2
Beyond every expanding force a greater moving force can be found. But
the latter can also act contrary to the former, whereby it would then
decrease the space that the former strives to enlarge, in which case the
latter would be called compressing force. Therefore, for every matter a
compressing force must also be discoverable, which can drive it from the
space it ﬁlls into a decreased space.
EXPLICATION 3
A matter penetrates another in its motion, when it completely destroys
the space of the latter’s extension through compression.
Remark
When, in the barrel of an air pump ﬁlled with air, the piston is driven
closer and closer to the bottom, the air-matter is compressed. If this com-
pression could now be driven so far that the piston completely touched
the bottom (without the least amount of air escaping), then the air-matter
would be penetrated. For the matters enclosing the air would leave no
remaining space for it, and it would thus be found between the piston
and the bottom without occupying a space. This penetrability of matter
through external compressing forces, if someone wished to assume or
even to think such a thing, could be called mechanical penetration. I have
reason thus to distinguish this penetrability of matter from another kind,
whose concept is perhaps just as impossible as the ﬁrst, but of which I
may yet have occasion to say something later on. 20
212Metaphysical foundations of natural science
PROPOSITION 3
4: 501
Matter can be compressed to inﬁnity, but can never be penetrated by a
matter, no matter how great the compressing force of the latter may be.
Proof
An original force, with which a matter strives to extend itself on all sides
beyond a given space that it occupies, must be greater when enclosed
in a smaller space, and inﬁnite when compressed into an inﬁnitely small
space. Now, for a given expanding force of matter, a greater compressing
force can be found, which forces the former into a smaller space, and
so on to inﬁnity, which was the ﬁrst [thing to be shown]. But a com-
pression of matter into an inﬁnitely small space, and thus an inﬁnite
compressing force, would be required for its penetration, and this is im-
possible. Therefore, a matter cannot be penetrated by any other through
compression, which is the second [thing to be shown].
Remark
In this proof I have assumed from the very beginning that an expanding
force must counteract all the more strongly, the more it is driven into a
smaller space. But this would not in fact hold for every kind of merely
derivative elastic forces. However, it can be postulated in matter, insofar
as essential elasticity belongs to it, as matter in general ﬁlling a space. For
expansive force, exerted from every point, and in every direction, actually
constitutes this concept. But the same quantum of extending forces, when
brought into a smaller space, must repel all the more strongly at every
point, the smaller the space in which this quantum diffuses its activity.
EXPLICATION 4
I call the impenetrability of matter that rests on resistance increasing in
proportion to the degree of compression relative impenetrability. But
that resting on the presupposition that matter as such is capable of no
compression at all is called absolute impenetrability. The ﬁlling of space
with absolute impenetrability can be called mathematical ﬁlling of space,
whereas that with mere relative impenetrability can be called dynamical
ﬁlling of space. 21
Remark 1
According to the purely mathematical concept of impenetrability (which
presupposes no moving force as originally belonging to matter), matter
213
4: 502Metaphysical foundations of natural science
is not capable of compression except insofar as it contains empty spaces
within itself. Hence matter as matter resists all penetration utterly a and
with absolute necessity. However, according to our discussion of this
property, impenetrability rests on a physical basis. For expanding force
ﬁrst makes matter itself possible, as an extended thing ﬁlling its space.
But this force has a degree that can be overpowered, and thus the space
of its extension can be diminished, that is, penetrated up to a certain
amount by a given compressing force, but only in such a way that com-
plete penetration is impossible, because this would require an inﬁnite
compressing force; therefore the ﬁlling of space must be viewed only as rela-
tive impenetrability.
Remark 2
Absolute impenetrability is in fact nothing more nor less than an occult
quality. b For one asks what the cause is for the inability of matters to
penetrate one another in their motion, and one receives the answer: be-
cause they are impenetrable. The appeal to repulsive force is not subject
to this reproach. For, although this force cannot be further explicated in
regard to its possibility, and therefore must count as a fundamental force,
it does yield a concept of an acting cause, together with its laws, whereby
the action, namely, the resistance in the ﬁlled space, can be estimated in
regard to its degrees.
EXPLICATION 5
4: 503
Material substance is that in space which is movable in itself, that is, in
isolation from everything else existing external to it in space. The motion
of a part of matter, whereby it ceases to be a part, is separation. The
separation of the parts of a matter is physical division.
Remark
The concept of a substance means the ultimate subject of existence, that
is, that which does not itself belong in turn to the existence of another
merely as a predicate. Now matter is the subject of everything that may be
counted in space as belonging to the existence of things. For, aside from
matter, no other subject would be thinkable except space itself, which,
however, is a concept that contains nothing existent at all, but merely
the necessary conditions for the external relations of possible objects of
the outer senses. Thus matter, as the movable in space, is the substance
a
b
schlechterdings
qualitas occulta
214Metaphysical foundations of natural science
therein. But all parts of matter must likewise be called substances, and
thus themselves matter in turn, insofar as one can say of them that they
are themselves subjects, and not merely predicates of other matters. They
are themselves subjects, however, if they are movable in themselves, and
thus exist in space outside their connection with other neighboring parts.
Therefore, the movability belonging to matter, or any part of it, is at the
same time a proof that this movable thing, and any movable part thereof,
is substance.
PROPOSITION 4
Matter is divisible to inﬁnity, and, in fact, into parts such that each is matter
in turn.
Proof
Matter is impenetrable, through its original expansive force (Prop. 3). 22
But this is only a consequence of the repulsive forces of each point in a
space ﬁlled with matter. Now the space ﬁlled by matter is mathematically
divisible to inﬁnity, that is, its parts can be distinguished to inﬁnity,
although they cannot be moved, and thus cannot be divided (according
to geometrical proofs). But in a space ﬁlled with matter, every part of it
contains repulsive force, so as to counteract all the rest in all directions,
and thus to repel them and to be repelled by them, that is, to be moved
a distance from them. Hence, every part of a space ﬁlled with matter is
in itself movable, and thus separable from the rest as material substance
through physical division. Therefore, the possible physical division of the
substance that ﬁlls space extends as far as the mathematical divisibility
of the space ﬁlled by matter. But this mathematical divisibility extends
to inﬁnity, and thus so does the physical [divisibility] as well. That is,
all matter is divisible to inﬁnity, and, in fact, into parts such that each is
itself material substance in turn.
Remark 1
The proof of the inﬁnite divisibility of space has not yet come close to
proving the inﬁnite divisibility of matter, if it has not previously been
shown that there is material substance in every part of space, that is, that
parts movable in themselves are to be found there. For suppose that a
monadist wished to assume that matter consisted of physical points, each
of which (for precisely this reason) had no movable parts, but nonetheless
ﬁlled a space through mere repulsive force. Then he could grant that
space would be divided, but not the substance that acts in space – that
the sphere of activity of this substance [would be divided] by the division
215
4: 504Metaphysical foundations of natural science
of space, but not the acting movable subject itself. 23 Thus he would
assemble matter out of physically indivisible parts, and yet allow them
to occupy a space in a dynamical fashion.
But this way out is completely taken away from the monadist by the
above proof. For it is thereby clear that there can be no point in a ﬁlled
space that does not exert repulsion in all directions, and is itself repelled,
and thus would be movable in itself, as a reacting subject external to
every other repelling point. Hence the hypothesis of a point that would
ﬁll a space through mere driving force, and not by means of other equally
repelling forces, is completely impossible. In order to make this intuitive,
and hence also the proof of the preceding Proposition, let us assume that
A is the place of a monad in space, and ab is the diameter of the sphere
of its repulsive force, so that aA is the radius of this sphere:
4: 505
Then between a, where the penetration of an external monad into the
space occupied by this sphere is resisted, and the center A, it is possible
to specify a point c (according to the inﬁnite divisibility of space). But
if A resists that which strives to penetrate into a, then c must also resist
the two points A and a. For, if this were not so, they would approach
one another without hindrance, and thus A and a would meet at the
point c, that is, the space would be penetrated. Therefore, there must
be something at c that resists the penetration of A and a, and thus repels
the monad A, the same as it is also repelled by A. But since repelling is
a [kind of] moving, c is something movable in space, and thus matter,
and the space between A and a could not be ﬁlled through the sphere of
activity of a single monad, nor could the space between c and A, and so
on to inﬁnity.
When mathematicians represent the repulsive forces of the parts of
elastic matters as increasing or decreasing, in accordance with a certain
proportion of their distances from one another, at greater or lesser com-
pression of these parts (for example, that the smallest parts of the air
repel one another in inverse ratio to their distances from one another,
because the elasticity of these parts stands in inverse ratio to the spaces in
which they are compressed), then one completely misses their meaning,
and misinterprets their language, if one ascribes that which necessarily
belongs to the procedure of constructing a concept to the concept in
the object itself. For by the latter [procedure], any contact can be rep-
resented as an inﬁnitely small distance – which must also necessarily be
so in those cases where a greater or smaller space is to be represented
216Metaphysical foundations of natural science
as completely ﬁlled by one and the same quantity of matter, that is, one
and the same quantum of repulsive forces. So even in the case of some-
thing divisible to inﬁnity, no actual distance of the parts may therefore be
assumed – they always constitute a continuum, no matter how enlarged
is the space, even though the possibility of such an enlargement can only
be made intuitive under the idea of an inﬁnitely small distance.
Remark 2
To be sure, mathematics in its internal use can be entirely indifferent
with regard to the chicanery of a misguided metaphysics, and can persist
in the secure possession of its evident claims as to the inﬁnite divisibility
of space, whatever objections may be put in its way by a sophistry splitting
hairs on mere concepts. However, in the application of its propositions
governing space to the substance that ﬁlls it, mathematics must nonethe-
less accede to an examination in accordance with mere concepts, and thus
to metaphysics. The above Proposition is already a proof of this. For it
does not necessarily follow that matter is physically divisible to inﬁnity,
even if it is so from a mathematical point of view, even if every part of
space is a space in turn, and thus always contains [more] parts external
to one another. For so far it cannot be proved that in each of the pos-
sible parts of this ﬁlled space there is also substance, which therefore also
exists in separation from all else as movable in itself. Thus something
without which this proof could not ﬁnd secure application to natural
science was until now still missing in the mathematical proof, and this
deﬁciency is remedied in the above Proposition. Now, however, when
it comes to the remaining metaphysical attacks on what will henceforth
be the physical Proposition of the inﬁnite divisibility of matter, the math-
ematician must leave them entirely to the philosopher, who in any case
ventures, by means of these objections, into a labyrinth, from which it
becomes difﬁcult for him to extricate himself, even in those questions
immediately pertaining to him. He therefore has quite enough to do for
himself, without the mathematician being permitted to involve himself
in this business. For if matter is divisible to inﬁnity then (concludes the
dogmatic metaphysician) it consists of an inﬁnite aggregate of parts; for a
whole must already contain in advance all of the parts in their entirety,
into which it can be divided. And this last proposition is undoubtedly
certain for every whole as thing in itself. But one cannot admit that matter,
or even space, consists of inﬁnitely many parts (because it is a contradiction
to think an inﬁnite aggregate, whose concept already implies that it can
never be represented as completed, as entirely completed). One would
therefore have to conclude either, in spite of the geometer, that space is
not divisible to inﬁnity, or, to the annoyance of the metaphysician, that
217
4: 506Metaphysical foundations of natural science
4: 507
space is not a property of a thing in itself, and thus that matter is not a thing
in itself, but merely an appearance of our outer senses in general, just as
space is the essential form thereof.
But here the philosopher is caught between the horns of a dangerous
dilemma. To deny the ﬁrst proposition, that space is divisible to inﬁnity,
is an empty undertaking; for nothing can be argued away from mathe-
matics by sophistical hairsplitting. a But viewing matter as a thing in itself,
and thus space as a property of the thing in itself, amounts to the de-
nial of this proposition. The philosopher therefore ﬁnds himself forced
to deviate from this last proposition, 24 however common and congenial
to the common understanding it may be. But he does this, of course,
only provided that, after making matter and space into mere appear-
ances (and thus the latter into the form of our outer sensible intuition,
so that both [are made] not into things in themselves, but only into sub-
jective modes of representation of objects unknown to us in themselves),
he is thereby helped out of that difﬁculty due to the inﬁnite divisibility
of matter, whereby it still does not consist of inﬁnitely many parts. Now
this latter can perfectly well be thought through reason, even though
it cannot be made intuitive and constructed. For what is only actual by
being given in the representation also has no more given of it than what
is met with in the representation – no more, that is, than the progress
of representations reaches. Therefore, one can only say of appearances,
whose division proceeds to inﬁnity, that there are just so many parts in
the appearance as we may provide, that is, so far as we may divide. For
the parts, as belonging to the existence of an appearance, exist only in
thought, namely, in the division itself. Now, the division does of course
proceed to inﬁnity, but it is still never given as inﬁnite. Thus it does
not follow, from the fact that its division proceeds to inﬁnity, that the
divisible contains an inﬁnite aggregate of parts in itself, and outside of
our representation. For it is not the thing, but only this representation
of it, whose division, although it can indeed be continued to inﬁnity, and
there is also a ground for this in the object (which is unknown in itself ),
can nonetheless never be completed, and thus be completely given; and
this also proves no actual inﬁnite aggregate in the object (which would be
an explicit contradiction). 25 A great man, 26 who has contributed perhaps
more than anyone else to preserving the reputation of mathematics in
Germany, has frequently rejected the presumptuous metaphysical claims
to overturn the theorems of geometry concerning the inﬁnite divisibility
of space by the well-founded reminder that space belongs only to the appear-
ance of outer things. But he has not been understood. This proposition was
taken to be asserting that space appears to us, though it is otherwise a
thing, or relation of things, in itself, but that the mathematician considers
a
dem Mathematik lä ß t sich nichts wegvernünfteln
218Metaphysical foundations of natural science
it only as it appears. Instead, it should have been understood as saying
that space is in no way a property that attaches in itself to any thing at all
outside our senses. It is, rather, only the subjective form of our sensibility,
under which objects of the outer senses, with whose constitution in itself
we are not acquainted, appear to us, and we then call this appearance
matter. Through this misunderstanding one went on thinking of space
as a property also attaching to things outside our faculty of representa-
tion, but such that the mathematician thinks it only in accordance with
common concepts, a that is, confusedly (for it is thus that one commonly
explicates appearance). And one thus attributed the mathematical the-
orem of the inﬁnite divisibility of matter, a proposition presupposing
the highest [degree of] clarity in the concept of space, to a confused
representation of space taken as basis by the geometer – whereby the
metaphysician was then free to compose space out of points, and matter
out of simple parts, and thus (in his opinion) to bring clarity into this con-
cept. The ground for this aberration lies in a poorly understood monad-
ology, [a theory] which has nothing at all to do with the explanation of
natural appearances, but is rather an intrinsically correct platonic concept
of the world devised by Leibniz, insofar as it is considered, not at all as
object of the senses, but as thing in itself, and is merely an object of the
understanding, which, however, does indeed underlie the appearances of
the senses. Now the composite of things in themselves must certainly consist
of the simple, for the parts must here be given prior to all composition.
But the composite in the appearance does not consist of the simple, because
in the appearance, which can never be given otherwise than as com-
posed (extended), the parts can only be given through division, and thus
not prior to the composite, but only in it. Therefore, Leibniz’s idea, b
so far as I comprehend it, was not to explicate space through the order
of simple beings next to one another, but was rather to set this order
alongside space as corresponding to it, but as belonging to a merely in-
telligible world (unknown to us). Thus he asserts nothing but what has
been shown elsewhere: namely, that space, together with the matter of
which it is the form, does not contain the world of things in themselves,
but only their appearance, and is itself only the form of our outer sensible
intuition.
PROPOSITION 5
The possibility of matter requires an attractive force as the second essential
fundamental force of matter.
a
b
gemeinen Begriffen
Meinung
219
4: 508Metaphysical foundations of natural science
Proof
4: 509
Impenetrability, as the fundamental property of matter, whereby it ﬁrst
manifests itself to our outer senses, as something real in space, is nothing
but the expansive power of matter (Proposition 2). Now an essential
moving force, whereby the parts of matter ﬂee from one another, cannot,
in the ﬁrst place, be limited by itself, for matter is thereby striving instead
continuously to enlarge the space that it ﬁlls; in the second place, [such a
force] can also not be determined by space alone to a certain limit of
extension, for the latter, although it can certainly contain the ground
for the expansive force becoming weaker in inverse proportion to the
increase of volume of an expanding matter, can never contain the ground
for this force ceasing anywhere, because smaller degrees are possible
to inﬁnity for any moving force. Hence matter, by its repulsive force
(containing the ground of impenetrability), would, [through itself] alone
and if no other moving force counteracted it, be conﬁned within no limit
of extension; that is, it would disperse itself to inﬁnity, and no speciﬁed
quantity of matter would be found in any speciﬁed space. Therefore, with
merely repulsive forces of matter, all spaces would be empty, and thus,
properly speaking, no matter would exist at all. So all matter requires
for its existence forces that are opposed to the expansive forces, that is,
compressing forces. But these, in turn, cannot originally be sought in
the contrary striving a of another matter, for this latter itself requires a
compressive force in order to be matter. Hence there must somewhere
be assumed an original force of matter acting in the opposite direction to
the repulsive force, and thus to produce approach, that is, an attractive
force. Yet since this attractive force belongs to the possibility of a matter
as matter in general, and thus precedes all differences of matter, it may
not be ascribed merely to a particular species of matter, but must rather
be ascribed to all matter originally and as such. Therefore, an original
attraction is attributed to all matter, as a fundamental force belonging to
its essence.
Remark
In this transition from one property of matter to another, speciﬁcally
different from it, and belonging equally to the concept of matter, even
though not contained in it, the procedure of our understanding must be
considered more closely. If attractive force is originally required even
for the possibility of matter, why do we not use it, just as much as impen-
etrability, as the ﬁrst distinguishing mark of a matter? Why is the latter
immediately given with the concept of a matter, whereas the former is
a
Entgegenstrebung
220Metaphysical foundations of natural science
not thought in the concept, but only adjoined to it through inferences?
That our senses do not allow us to perceive this attraction so immediately
as the repulsion and resistance of impenetrability cannot yet provide a
sufﬁcient answer to the difﬁculty. For even if we had such a capacity, it
is still easy to see that our understanding would nonetheless choose the
ﬁlling of space in order to designate substance in space, that is, matter,
and how precisely this ﬁlling, or, as one otherwise calls it, solidity, is then
posited to be characteristic of matter, as a thing different from space.
Attraction, even if we sensed it equally well, would still never disclose to
us a matter of determinate volume and ﬁgure, but only the striving of our
organ to approach a point outside us (the center of the attracting body).
For the attractive force of all parts of the earth can affect us no more, and
in no other way, than as if it were wholly united in the earth’s center, and
this alone inﬂuenced our sense, and the same holds for the attraction of
a mountain, or any stone, etc. 27 But we thereby obtain no determinate
concept of any object in space, since neither ﬁgure, nor quantity, nor
even the place where it would be found can strike our senses. (The mere
direction of attraction would be perceivable, as in the case of weight: the
attracting point would be unknown, and I do not even see how it could be
ascertained through inferences, without perception of matter insofar as
it ﬁlls space). It is therefore clear that the ﬁrst application of our concepts
of quantity to matter, through which it ﬁrst becomes possible for us to
transform our outer perceptions into the empirical concept of a matter,
as object in general, is grounded only on that property whereby it ﬁlls
a space – which, by means of the sense of feeling, a provides us with the
quantity and ﬁgure of something extended, and thus with the concept of
a determinate object in space, which forms the basis of everything else
one can say about this thing. Precisely this circumstance is undoubtedly
the reason, despite the clearest proofs from elsewhere that attraction
must belong to the fundamental forces of matter, just as much as
repulsion, that one nevertheless struggles so much against the former,
and will admit no other moving forces at all except those through impact
and pressure (both mediated by impenetrability). For that whereby space
is ﬁlled is substance, one says, and this is also perfectly correct. But this
substance discloses its existence to us in no other way than through that
sense whereby we perceive its impenetrability, namely, feeling, and thus
only in relation to contact, whose onset (in the approach of one matter
to another) is called impact, and whose persistence is called pressure. It
therefore seems as if every immediate action of one matter on the other
could never be anything but pressure or impact, the only two inﬂuences
we can sense immediately. Attraction, on the other hand, can give us
in itself either no sensation at all, or at least no determinate object of
a
Gefühl
221
4: 510Metaphysical foundations of natural science
sensation, and is therefore difﬁcult for us to understand as a fundamental
force.
PROPOSITION 6
No matter is possible through mere attractive force without repulsion.
Proof
4: 511
Attractive force is that moving force of matter whereby it impels another
to approach it; consequently, if it is found between all parts of matter,
matter thereby strives to diminish the distance of its parts from one an-
other, and thus the space that they occupy together. But nothing can
hinder the action of a moving force except another moving force op-
posed to it, and that which opposes attraction is repulsive force. Hence,
without repulsive forces, through mere convergence, a all parts of matter
would approach one another unhindered, and would diminish the space
that they occupy. But since, in the case assumed, there is no distance of
the parts at which a greater approach due to attraction would be made
impossible by a repulsive force, they would move toward one another so
far, until no distance at all would be found between them; that is, they
would coalesce into a mathematical point, and space would be empty,
and thus without any matter. Therefore, matter is impossible through
mere attractive forces without repulsive forces.
Note
A property on which the inner possibility of a thing rests, as a condition,
is an essential element thereof. Hence repulsive force belongs to the
essence of matter just as much as attractive force, and neither can be
separated from the other in the concept of matter.
Remark
Since only two moving forces can be thought everywhere in space, b
repulsion and attraction, it was previously necessary, in order to prove a
priori the uniting of the two in the concept of a matter in general, that
each be considered on its own, so as to see what either in isolation could
achieve for the presentation of a matter. It is now manifest that, whether
one takes neither as basis, or assumes merely one of them, space would
always remain empty, and no matter would be found therein.
a
b
Annährung
Weil überall nur zwei bewegende Kräfte im Raum gedacht werden kônnen
222Metaphysical foundations of natural science
EXPLICATION 6
Contact in the physical sense is the immediate action and reaction of
impenetrability. The action of one matter on another in the absence of
contact is action at a distance (actio in distans). This action at a distance,
which is possible even without the mediation of matter lying in between,
is called immediate action at a distance, or the action of matters on one
another through empty space.
Remark
Contact in the mathematical sense is the common boundary of two
spaces, which is therefore within neither the one nor the other space.
Thus two straight lines cannot be in contact with one another; rather, if
they have a point in common, it belongs as much to one of these lines
as to the other when they are produced, a that is, they intersect. But a
circle and a straight line, or two circles, are in contact at a point, surfaces
at a line, and bodies at surfaces. Mathematical contact is the basis for
physical contact, but does not yet constitute the latter by itself, since for
the one to arise from the other a dynamical relation must also be added
in thought – and, indeed, not of attractive, but of repulsive forces, that is,
of impenetrability. Physical contact is the interaction of repulsive forces
at the common boundary of two matters.
PROPOSITION 7
The attraction essential to all matter is an immediate action of matter on
other matter through empty space.
Proof
The original attractive force contains the very ground of the possibility
of matter, as that thing which ﬁlls a space to a determinate degree, and
so contains even [the ground] of the possibility of a physical contact
thereof. It must therefore precede the latter, and its action must thus
be independent of the condition of contact. But the action of a moving
force that is independent of all contact is also independent of the ﬁlling
of space between the moving and the moved [matters]; that is, it must
also take place without the space between the two being ﬁlled, and thus
as action through empty space. Hence the original attraction essential
to all matter is an immediate action of matter on other matter through
empty space.
a
fortgezogen
223
4: 512Metaphysical foundations of natural science
4: 513 Remark 1
4: 514 That the possibility of the fundamental forces should be made conceiv-
able is a completely impossible demand; for they are called fundamental
forces precisely because they cannot be derived from any other, that is,
they can in no way be conceived. But the original attractive force is in
no way more inconceivable than the original repulsion. It simply does not
present itself so immediately to the senses as impenetrability, so as to
furnish us with concepts of determinate objects in space. Thus, because
it is not felt, but is only to be inferred, a it has so far the appearance of
a derived force, exactly as if it were only a hidden play of moving forces
through repulsion. On closer consideration we see that it can in no way
be further derived from anywhere else, least of all from the moving force
of matters through their impenetrability, since its action is precisely the
reverse of the latter. The most common objection to immediate action
at a distance is that a matter cannot act immediately where it is not. When
the earth immediately impels the moon to approach it, the earth acts on a
thing that is many thousands of miles away from it, and yet immediately;
the space between it and the moon may well be viewed as completely
empty. For even though matter may lie between the two bodies, it still
contributes nothing to this attraction. It therefore acts immediately at a
place where it is not, which is apparently contradictory. In truth, how-
ever, it is so far from being contradictory that one may rather say that
every thing in space acts on another only at a place where the acting
thing is not. For if it should act at the same place where it itself is, then
the thing on which it acts would not be outside it at all; for this outsideness
means presence at a place where the other is not. If earth and moon were
to be in contact with one another, the point of contact would still be a
place where neither the earth nor the moon is, for the two are distanced
from one another by the sum of their radii. Moreover, no part of either
the earth or the moon would be found at the point of contact, for this
point lies at the boundary of the two ﬁlled spaces, which constitutes no
part of either the one or the other. Hence to say that matters cannot act
immediately on one another at a distance, would amount to saying that
they cannot act immediately on one another except through the forces of
impenetrability. But this would be as much as to say that repulsive forces
are the only ones whereby matters can be active, or that they are at
least the necessary conditions under which alone matters can act on one
another, which would declare attractive force to be either completely im-
possible or always dependent on the action of repulsive forces. But these
are both groundless assertions. The confusion of mathematical contact
of spaces and physical contact through repulsive forces constitutes the
a
Weil sie also nicht gefühlt, sondern nur geschlossen werden will
224Metaphysical foundations of natural science
ground of misunderstanding here. To attract one another immediately
in the absence of contact means to approach one another in accordance
with an invariable law, without a force of repulsion containing the condi-
tion for this. And this must be just as thinkable as an immediate repulsion
of one another, that is, to ﬂee from one another in accordance with an
invariable law, without the force of attraction having any part therein.
For the two moving forces are of completely different kinds, and there
is not the slightest ground for making one of them dependent on the
other, and contesting its possibility unmediated by the other.
Remark 2
No motion at all can arise from attraction in contact; for contact is inter-
action of impenetrability, which therefore prevents all motion. Hence,
some sort of immediate attraction must be found in the absence of con-
tact, and thus at a distance. For otherwise even the forces of pressure and
impact, which are supposed to bring about the striving to approach by
acting in the opposite direction to that of the repulsive force of matter,
would have no cause, or at least none lying originally in the nature of
matter. We may call that attraction which takes place without mediation
of the repulsive forces true attraction, whereas that which takes place
merely in that way is apparent attraction. For, properly speaking, the
body which another is striving to approach, merely because the latter
has been driven toward it from elsewhere by impact, exerts no attractive
force at all on this body. But even these apparent attractions must in
the end have a true one as their ground. For matter whose pressure or
impact is supposed to serve instead of attraction would not even be mat-
ter without attractive forces (Prop. 5), and so the mode of explaining all
phenomena of approach by merely apparent attraction revolves in a circle.
It is commonly supposed that Newton did not at all ﬁnd it necessary for
his system to assume an immediate attraction of matter, but, with the
most rigorous abstinence of pure mathematics, allowed the physicists
full freedom to explain the possibility of attraction as they might see ﬁt,
without mixing his propositions with their play of hypotheses. But how
could he ground the proposition that the universal attraction of bodies,
which they exert at equal distances around them, is proportional to the
quantity of their matter, if he did not assume that all matter, merely as
matter, therefore, and through its essential property, exerts this moving
force? For although between two bodies, when one attracts the other,
whether their matter be similar or not, the mutual approach (in accor-
dance with the law of equality of interaction) must always occur in inverse
ratio to the quantity of matter, this law still constitutes only a principle of
mechanics, but not of dynamics. That is, it is a law of the motions that fol-
low from attracting forces, not of the proportion of the attractive forces
225
4: 515Metaphysical foundations of natural science
themselves, and it holds for all moving forces in general. 28 Thus, if a
magnet is at one time attracted by another equal magnet, and at another
by the same magnet enclosed in a wooden box of double the weight, the
latter will impart more relative motion to the former in the second case
than in the ﬁrst, even though the wood, which increases the quantity of
matter of this second magnet, adds nothing at all to its attractive force,
and manifests a no magnetic attraction of the box. Newton says (Cor. 2,
Prop. 6, Book III, Principia): “if the aether or any other body were with-
out weight, it could, since it differs from every other matter only in its
form, be transformed successively by gradual change of this form into a
matter of the same kind as those which on earth have the most weight;
and so the latter, conversely, by gradual change of their form, could lose
all their weight, which is contrary to experience, etc.” 29 Thus he did
not himself exclude the aether (much less other matters) from the law
of attraction. So what other kind of matter could he then have left, by
whose impact the approach of bodies to one another might be viewed as
mere apparent attraction? Thus, one cannot adduce this great founder
of the theory of attraction as one’s predecessor, if one takes the liberty
of substituting an apparent attraction for the true attraction he did as-
sert, and assumes the necessity of an impulsion through impact to explain
the phenomenon of approach. He rightly abstracted from all hypothe-
ses purporting to answer the question as to the cause of the universal
attraction of matter, for this question is physical or metaphysical, but
not mathematical. And, even though he says in the advertisement to the
second edition of his Optics, “to show that I do not take gravity for an
essential property of bodies, I have added one question concerning its
cause,” 30 it is clear that the offense taken by his contemporaries, and
perhaps even by Newton himself, at the concept of an original attraction
set him at variance with himself. For he could by no means say that the
attractive forces of two planets, those of Jupiter and Saturn for exam-
ple, manifested at equal distances from their satellites (whose mass is
unknown), are proportional to the quantity of matter of these heavenly
bodies, 31 if he did not assume that they attracted other matter merely as
matter, and thus according to a universal property of matter.
EXPLICATION 7
4: 516
I call a moving force whereby matters can act immediately on one another
only at the common surface of contact, a surface force. But that whereby
a matter can act immediately on the parts of others, even beyond the
surface of contact, I call a penetrating force.
a
beweiset
226Metaphysical foundations of natural science
Note
The repulsive force whereby matter ﬁlls a space is a mere surface force,
for the parts in contact mutually limit their spaces of action. Repulsive
force cannot move a part at a distance without the mediation of those
lying in between, and an immediate action, passing straight through the
latter, of one matter on another by expansive forces, is impossible. By
contrast, no intervening matter sets limits to the action of an attractive
force, whereby matter occupies a space without ﬁlling it, so that it thereby
acts on other distant matter through empty space. Now the original attrac-
tion, which makes matter itself possible, must be thought in this way,
and it is therefore a penetrating force, and for this reason alone is always
proportional to the quantity of matter.
PROPOSITION 8
The original attractive force, on which the very possibility of matter as
such rests, extends immediately to inﬁnity throughout the universe, from
every part of matter to every other part.
Proof
Because the original attractive force belongs to the essence of matter, it
also pertains to every part of matter to act immediately at a distance as
well. But suppose there were a distance beyond which it did not extend.
Then this limiting of the sphere of its activity would rest either on the
matter lying within this sphere, or merely on the magnitude of the space
in which it diffuses this inﬂuence. The ﬁrst [case] does not hold; for
this attraction is a penetrating force and acts immediately at a distance
through that space, as an empty space, regardless of any matter lying
in between. The second [case] likewise does not hold; for, since every
attraction is a moving force having a degree, below which ever smaller
degrees can always be thought to inﬁnity, a greater distance would indeed
be a reason for the degree of attraction to diminish in inverse ratio, in
accordance with the measure of the diffusion of this force, but never for
it to cease altogether. Thus, since there is nothing that has anywhere
limited the sphere of activity of the original attraction of every part of
matter, it extends beyond all speciﬁed limits to every other matter, and
thus throughout the universe to inﬁnity.
Note 1
From this original attractive force, as a penetrating force exerted by all
matter, and hence in proportion to its quantity, and extending its action to
227
4: 517Metaphysical foundations of natural science
4: 518
all matter at all possible distances, it should now be possible, in combina-
tion with the force counteracting it, namely, repulsive force, to derive the
limitation of the latter, and thus the possibility of a space ﬁlled to a deter-
minate degree. And thus the dynamical concept of matter, as that of the
movable ﬁlling its space (to a determinate degree), would be constructed.
But for this one needs a law of the ratio of both original attraction and
repulsion at various distances of matter and its parts from one another,
which, since it now rests simply on the difference in direction of these
two forces (where a point is driven either to approach others or to move
away from them), and on the magnitude of the space into which each of
these forces diffuses at various distances, is a purely mathematical task,
which no longer belongs to metaphysics – nor is metaphysics responsi-
ble if the attempt to construct the concept of matter in this way should
perhaps not succeed. For it is responsible only for the correctness of the
elements of the construction granted to our rational cognition, not for
the insufﬁciency and limits of our reason in carrying it out.
Note 2
Since every given matter must ﬁll its space with a determinate degree of
repulsive force, in order to constitute a determinate material thing, only
an original attraction in conﬂict with the original repulsion can make
possible a determinate degree of the ﬁlling of space, and thus matter.
Now it may be that the former ﬂows from the individual attraction of
the parts of the compressed matter among one another, or from the
uniting of this attraction with that of all matter in the universe. a
The original attraction is proportional to the quantity of matter and
extends to inﬁnity. Therefore, the determinate ﬁlling, in accordance with
its measure, of a space by matter, b can in the end be effected only by the
attraction of matter extending to inﬁnity, and imparted to each matter
in accordance with the measure of its repulsive force.
The action of the universal attraction immediately exerted by each
matter on all matters, and at all distances, is called gravitation; the ten-
dency to move in the direction of greater gravitation is weight. The
action of the general c repulsive force of the parts of every given matter
is called its original elasticity. Hence this property and weight constitute
the sole universal characteristics of matter, which are comprehensible
a priori, the former internally, and the latter in external relations. For
the possibility of matter itself rests on these two properties. Cohesion, if
this is explicated as the mutual attraction of matter limited solely to the
a
b
c
aller Weltmaterie
die dem Ma ß e nach bestimmte Erfüllung eines Raumes durch Materie
durchgängig
228Metaphysical foundations of natural science
condition of contact, does not belong to the possibility of matter in gen-
eral, and cannot therefore be cognized a priori as bound up with this.
This property would therefore not be metaphysical but rather physical,
and so would not belong to our present considerations.
Remark 1
Yet I cannot forbear adding a small preliminary suggestion on behalf of
the attempt at such a perhaps possible construction.
(1) Of any force that acts immediately at various distances, and is
limited, as to the degree with which it exerts moving force on any given
point at a certain distance, only by the magnitude of the space into which
it must diffuse so as to act on this point, one can say that in all the spaces,
large or small, into which it diffuses, it always constitutes an equal quan-
tum, but [also] that the degree of its action on that point in this space is
always in inverse ratio to the space, into which it has had to diffuse, so
that it could act on this point. Thus light, for example, diffuses from an il-
luminating point in all directions on spherical surfaces, which constantly
increase with the squares of the distance, and the quantum of illumina-
tion on all of these spherical surfaces, which become greater to inﬁnity,
is always the same in total. But it follows from this that a given equal
part of one of these spherical surfaces must become ever less illuminated
with respect to its degree, as the surface of diffusion of precisely the same
light quantum becomes greater. And so, too, with all other forces, and
the laws whereby they must diffuse, either on surfaces or in volumes, a so
as to act on distant objects in accordance with their nature. It is better to
represent the diffusion of a moving force at all distances from a point in
this way, rather than as is customary, for example, in optics, by means of
rays diverging from one another radially from a central point. For since,
as an unavoidable consequence of their divergence, lines drawn in this
way can never ﬁll the space through which they spread, nor the surfaces
on which they fall, no matter how many are drawn or plotted, they give
rise only to troublesome inferences, and these in turn to hypotheses,
which might well be avoided by merely taking into consideration the
magnitude of the whole spherical surface – which is to be uniformly illu-
minated by the same quantity of light; and the degree of its illumination
at every place is then naturally taken in inverse ratio of its magnitude to
the whole, and similarly for any other diffusion of a force through spaces
of different magnitudes.
(2) If the force is an immediate attraction at a distance, then it is even
more necessary to represent the directed lines of attraction, not as if they
a
körperlichen Raum
229
4: 519Metaphysical foundations of natural science
4: 520
4: 520
diverged like rays from the attracting point, but rather as converging
from every point of the surrounding spherical surface (whose radius is
the given distance) toward it. For the very directed line of motion toward
the point, which is the cause and goal of this motion, already yields the
terminus a quo from which the lines must begin, namely, from every
point of the surface from which they take their direction toward the
attracting central point, and not conversely. For this magnitude of the
surface alone determines the aggregate of lines; the central point leaves
this undetermined. ∗
(3) If the force is an immediate repulsion, by which a point (in the
merely mathematical presentation) ﬁlls a space dynamically, and the ques-
tion is by what law of inﬁnitely small distances (which here count as
equivalent to contacts) an original repulsive force (whose limitation thus
rests simply on the space in which it is diffused) acts at various distances,
then it is even less possible to make this force representable by diverging
rays of repulsion from the assumed repelling point, even though the di-
rection of motion has this point as its terminus a quo. For the space into
∗
It is impossible by lines radiating from a point to represent surfaces at given
distances as completely ﬁlled with their action, whether of illumination or
attraction. Thus, in the case of such diverging light rays, the lesser illumination
of a distant surface would rest merely on the circumstance that between the
illuminated places remain unilluminated ones, and the more distant the surface
the larger they are. Euler’s hypothesis avoids this impropriety, but has all the
more difﬁculty in making the rectilinear motion of light conceivable. Yet this
difﬁculty ﬂows from an easily avoidable mathematical representation of light
matter as an agglomeration of little spheres, which would certainly yield a lateral
motion of light in accordance with their varying obliquity to the direction of
impact. Instead of this, however, there is no obstacle to thinking the matter in
question as an original ﬂuid, and, indeed, as ﬂuid throughout, without being
divided into rigid particles. If the mathematician wants to make intuitive the
decrease of light at increasing distances, he uses rays diverging radially to
represent the magnitude of the space on the spherical surface of its diffusion,
wherein the same quantity of light is supposed to be uniformly diffused between
these rays, and thus to represent the decrease of the degree of illumination.
But he does not want one to view these rays as the only sources of illumination,
as if places empty of light, which would be greater at greater distances, were
always to be found between them. If one wants to imagine each such surface
as illuminated throughout, then the same quantity of illumination as covers
the smaller surfaces must be thought as uniformly [spread] over the larger
surfaces; so, in order to indicate the rectilinear direction, straight lines from the
surface, and all of its points, must be drawn toward the illuminating point. The
action and its quantity must be thought of beforehand, and the cause thereupon
speciﬁed. Precisely the same holds for rays of attraction, if one wants to call
them that, and indeed, for all directions of forces that are supposed to ﬁll a
space, and even a volume, proceeding from a point.
230Metaphysical foundations of natural science
which the force must be diffused in order to act at a distance is a volume,
which is supposed to be thought as ﬁlled. (The manner in which a point
could do this by moving force, that is, dynamically ﬁll a volume, is cer-
tainly not capable of further mathematical presentation.) And diverging
rays from a point cannot possibly make representable the repulsive force
of a ﬁlled volume. Rather, one would simply estimate the repulsion, at
various inﬁnitely small distances of these mutually impelling points, as
merely in inverse ratio to the volumes that each of them ﬁlls dynami-
cally, and thus to the cube of their distances, without being able to con-
struct it.
(4) Thus the original attraction of matter would act in inverse ratio
to the squares of the distance at all distances, the original repulsion in
inverse ratio to the cubes of the inﬁnitely small distances, and, through
such an action and reaction of the two fundamental forces, matter ﬁlling
its space to a determinate degree would be possible. For since repul-
sion increases with the approach of the parts to a greater extent than
attraction, the limit of approach, beyond which no greater is possible
by the given attraction, is thereby determined, and so is that degree of
compression which constitutes the measure of the intensive ﬁlling of
space.
Remark 2
I am well aware of the difﬁculty in this mode of explaining the possi-
bility of a matter in general. It consists in this, that if a point cannot
immediately propel another by repulsive force, without at the same time
ﬁlling the entire volume up to the given distance with its force, then
it appears to follow that this volume would have to contain several im-
pelling points, which contradicts the presupposition, and was refuted
above (Proposition 4) under the name of a sphere of repulsion of the
simple [elements] in space. But there is a difference between the concept
of an actual space, which can be given, and the mere idea of a space, which
is thought simply for determining the ratio of given spaces, but is not in
fact a space. In the case put forward, of a supposed physical monadology,
there were supposed to be actual spaces ﬁlled dynamically by a point,
namely, through repulsion; 32 for they would exist as points prior to any
possible generation of matter therefrom, and would determine, through
their own spheres of activity, that part of the space to be ﬁlled which
could belong to them. So on this hypothesis, matter cannot be viewed as
divisible to inﬁnity, and as quantum continuum. For the parts that imme-
diately repel one another have a determinate distance from one another
(the sum of the radii of the spheres of their repulsion). By contrast, if,
as is actually the case, we think matter as a continuous quantity, there
is then no distance at all between the points immediately repelling one
231
4: 521Metaphysical foundations of natural science
4: 522
another, and thus no increasing or decreasing sphere of their activity.
But matters can expand or be compressed (like air), and here one does
represent to oneself a distance of their adjacent parts, which can increase
and decrease. Yet since the adjacent parts of a continuous matter are in
contact with one another, whether it is further expanded or compressed,
one then thinks these distances as inﬁnitely small, and this inﬁnitely small
space as ﬁlled by its repulsive force to a greater or lesser degree. But
the inﬁnitely small intervening space is not at all different from contact.
Hence it is only the idea of a space, which serves to make intuitive the
enlargement of a matter as a continuous quantity, although it cannot, in
fact, be actually conceived in this way. If it is said, therefore, that the re-
pulsive forces of the parts of matter that immediately impel one another
stand in inverse ratio to the cubes of their distances, this means only that
they stand in inverse ratio to the volumes one imagines between parts
that are nevertheless in immediate contact, and whose distance must for
precisely this reason be called inﬁnitely small, so as to be distinguished
from every actual distance. Hence one must not object to a concept itself
because of difﬁculties in constructing it, or, rather, because of a misun-
derstanding of this construction. For otherwise it would apply to the
mathematical presentation of the proportion in accordance with which
attraction takes place at various distances, no less than to that whereby
every point in an expanding or contracting whole of matter immediately
repels the others. The universal law of dynamics would in both cases be
this: the action of the moving force, exerted by a point on every other
point external to it, stands in inverse ratio to the space into which the
same quantum of moving force would need to have diffused, in order to
act immediately on this point at the determinate distance.
From the law of the parts of matter repelling one another originally in
inverse cubic ratio to their inﬁnitely small distances, a law of expansion
and contraction of matter completely different from Mariotte’s law for
the air must therefore necessarily follow; for the latter proves ﬂeeing
forces of its adjacent parts standing in inverse ratio to their distances,
as Newton demonstrates (Principia, Book II, Prop. 23, Schol.). 33 But we
may also view the expansive force of air, not as the action of originally
repelling forces, but as resting rather on heat, which compels the proper
parts of air (to which, moreover, actual distances from one another are
attributable) to ﬂee one another, not merely as a matter penetrating it, but
rather, to all appearances, through its vibrations. But that these tremors
must impart a ﬂeeing force to the adjacent parts, standing in inverse ratio
to their distances, can doubtless be made conceivable in accordance with
the laws of communication of motion through the oscillation of elastic
matters.
I declare, furthermore, that I do not want the present exposition of
the law of an original repulsion to be viewed as necessarily belonging to
232Metaphysical foundations of natural science
the goals of my metaphysical treatment of matter. Nor do I want this
latter (for which it is enough to have presented the ﬁlling of space as
a dynamical property of matter) to be mixed up with the conﬂicts and
doubts that could afﬂict the former.
4: 523
GENERAL NOTE TO DYNAMICS
If we look back over all our discussions of the subject, we will notice
that we have therein considered the following: ﬁrst, the real in space
(otherwise called the solid), in the ﬁlling of space through repulsive force;
second, that which in relation to the ﬁrst, as the proper object of our outer
perception, is negative, namely, attractive force, whereby, for its own part,
all space would be penetrated, and thus the solid would be completely
destroyed; third, the limitation of the ﬁrst force by the second, and the de-
termination of the degree of ﬁlling of a space that rests on this. Hence, the
quality of matter, under the headings of reality, negation, and limitation, has
been treated completely, so far as pertains to a metaphysical dynamics.
GENERAL REMARK TO DYNAMICS
The general principle of the dynamics of material nature is that every-
thing real in the objects of the outer senses, which is not merely a
determination of space (place, extension, and ﬁgure), must be viewed
as moving force. So by this principle the so-called solid or absolute
impenetrability is banished from natural science, as an empty concept,
and repulsive force is posited in its stead. But the true and immediate
attraction, by contrast, is thereby defended against all sophistries of a
metaphysics that misunderstands itself, and, as a fundamental force, is
declared necessary for the very possibility of the concept of matter. Now
from this it follows that space, if it should be necessary, can be assumed
to be completely ﬁlled, and in different degrees, even without dispersing
empty interstices within matter. 34 For, in accordance with the originally
different degree of the repulsive forces, on which rests the ﬁrst property
of matter, namely, that of ﬁlling a space, their relation to the original
attraction (whether of any [piece of] matter separately, or to the united
attraction of all matter in the universe) can be thought of as inﬁnitely
various. This is because attraction rests on the aggregate of matter in a
given space, whereas its expansive force, by contrast, rests on the degree
of ﬁlling of this space, which can be very different speciﬁcally (as the
same quantity of air, say, in the same volume, manifests more or less
elasticity in accordance with its greater or lesser heating). The general
ground for this is that through true attraction all parts of a matter act
immediately on every part of another, whereas through expansive force
only those at the surface of contact act, so that it is all the same whether
233
4: 524Metaphysical foundations of natural science
4: 525
much or little of this matter is found behind that surface. Now a great
advantage for natural science already arises here, since it is thereby re-
lieved of the burden of fabricating a world from the full and the empty in
accordance with mere fantasy. On the contrary, all spaces can be thought
of as full, and yet as ﬁlled in different measures, whereby empty space at
least loses its necessity, and is demoted to the value of an hypothesis. For
it could otherwise usurp the title of a principle, under the pretense of
being a necessary condition for explaining the different degrees of the
ﬁlling of space.
In all this the advantage of a metaphysics that is here used methodi-
cally, to get rid of principles that are equally metaphysical, but have not
been brought to the test of criticism, is apparently only negative. Nev-
ertheless, the ﬁeld of the natural scientist is thereby indirectly enlarged.
For the conditions by which he formerly limited himself, and through
which all original moving forces were philosophized away, now lose their
validity. But one should guard against going beyond that which makes
possible the general concept of a matter as such, and wishing to explain
a priori its particular, or even speciﬁc, determination and variety. The
concept of matter is reduced to nothing but moving forces, and one could
not expect anything else, since no activity or change can be thought in
space except mere motion. But who pretends to comprehend the pos-
sibility of the fundamental forces? They can be assumed only if they
unavoidably belong to a concept that is demonstrably fundamental and
not further derivable from any other (like that of the ﬁlling of space), and
these, in general, are repulsive forces and the attractive forces that coun-
teract them. We can indeed certainly judge a priori about the connection
and consequences of these forces, whatever relations among them one
can think without contradiction, but cannot yet presume to suppose one
of them as actual. For to be authorized in erecting an hypothesis, it is un-
avoidably required that the possibility of what we suppose be completely
certain, but with fundamental forces their possibility can never be com-
prehended. And here the mathematical-mechanical mode of explanation
has an advantage over the metaphysical-dynamical [mode], which cannot
be wrested from it, namely, that of generating from a thoroughly homo-
geneous material a great speciﬁc variety of matters, which vary in both
density and (if foreign forces are added) mode of action, through the
varying shape of the parts and the empty interstices interspersed among
them. For the possibility of both the shapes and the empty interstices can
be veriﬁed with mathematical evidence. By contrast, if the material itself
is transformed into fundamental forces (whose laws we cannot determine
a priori, and are even less capable of enumerating reliably a manifold of
such forces sufﬁcient for explaining the speciﬁc variety of matter), we
lack all means for constructing this concept of matter, and presenting what
we thought universally as possible in intuition. Conversely, however, a
234Metaphysical foundations of natural science
merely mathematical physics pays doubly for this advantage on the other
side. First, it must take an empty concept (of absolute impenetrability)
as basis; and second, it must give up all forces inherent in matter; and
beyond this, further, with its original conﬁgurations of the fundamental
material and its interspersing of empty spaces, as the need for explana-
tion requires them, such a physics must allow more freedom, and indeed
rightful claims, to the imagination in the ﬁeld of philosophy than is truly
consistent with the caution of the latter.
Instead of a sufﬁcient explanation for the possibility of matter and its
speciﬁc variety from these fundamental forces, which I cannot provide,
I will present completely, so I hope, the moments to which its speciﬁc
variety must collectively be reducible (albeit not conceivable in regard
to its possibility). The remarks inserted between the deﬁnitions a will
explain their application.
1. A body, in the physical sense, is a matter between determinate bound-
aries (which therefore has a ﬁgure). The space between these boundaries,
considered in accordance with its magnitude, is the volume [of the body]. b
The degree of the ﬁlling of a space with determinate content c is called
density. (Otherwise the term dense is also used absolutely for what is not
hollow, that is, vesicular or porous.) In this sense, there is an absolute
density in the system of absolute impenetrability, that is, when a matter
contains no empty interstices at all. In accordance with this concept of
the ﬁlling of space we make comparisons, and call one matter denser
than another when it contains less emptiness, until ﬁnally that in which
no part of the space is empty is called perfectly dense. One can only make
use of the latter expression in connection with the merely mathematical
concept of matter, but in the dynamical system of a merely relative im-
penetrability there is no maximum or minimum of density, and yet every
matter, however rareﬁed, can still be called completely dense, if it ﬁlls its
space entirely without containing empty interstices, and is thus a contin-
uum, not an interruptum. In comparison with another matter, however,
it is less dense, in the dynamical sense, if it ﬁlls its space entirely, but not
to the same degree. But in this system, too, it is inappropriate to think
of matters as related with respect to their density, if we do not imagine
them as speciﬁcally of the same kind, so that one can be generated from
the other by mere compression. Now since the latter [condition] by no
means appears to be necessary to the nature of all matter in itself, no
comparison with regard to their density can properly take place between
matters of different kinds, between water and mercury, for example, even
though it is customary.
a
b
c
Deﬁnitionen
Raumesinhalt (volumen)
Der Grad der Erfüllung eines Raumes von bestimmtem Inhalt
235
4: 526Metaphysical foundations of natural science
4: 527
2. Attraction, insofar as it is thought merely as active in contact, is called
cohesion. (To be sure, it is conﬁrmed by very good experiments that the
same force, which in contact is called cohesion, is also found to be active at
a very small distance. But attraction is still called cohesion, only insofar as I
think it merely in contact, in accordance with common experience, where
it is hardly ever observed at small distances. Cohesion is commonly taken
for an entirely general property of matter, not because one is already
led to it by the concept of a matter, but because experience shows it
everywhere. But this generality must not be understood collectively, as if
every matter acted, through this kind of attraction, on every other matter
in the universe at once, like gravitation; it must rather be understood
merely disjunctively, as acting, that is, on one matter or another with
which it comes into contact, of whatever kind it may be. For this reason,
and since such attraction, as various grounds of proof can show, is not
a penetrating, but only a surface force; since it is not even determined
everywhere in accordance with density; since for full strength of cohesion
a prior state of ﬂuidity of the matters and their subsequent rigidiﬁcation
is required, whereby the closest possible contact of broken solid matters
at precisely the same surfaces where they previously cohered so strongly,
in a cracked mirror, for example, is still very far from permitting any
longer that degree of attraction it had gained on rigidiﬁcation from a
ﬂuid state; I therefore take this attraction in contact to be no fundamental
force of matter, but only a derivative one; of which more below.) A
matter whose parts, however strong their mutual cohesion, can nonetheless be
mutually displaced by every moving force, however small, is ﬂuid. But parts of a
matter are so displaced, when they are merely compelled, without reducing the
quantum of contact, to interchange such contact. Parts, and thus also matters, are
separated, when the contact is not merely exchanged with others, but destroyed
or reduced in quantity. A solid – or better a rigid – body (corpus rigidum)
is one whose parts cannot be so displaced by every force – and therefore resist
displacement with a certain degree of force. – The resistance to such mutual
displacement of matters is friction. The resistance to separation of matters
in contact is cohesion. Fluid matters therefore undergo no friction when
divided; where it occurs, the matters, at least in their smallest parts, are
taken to be rigid, in greater or lesser degree, where the latter is called
viscosity. a A rigid body is brittle, when its parts cannot be mutually displaced
without breaking apart, and thus when their cohesion cannot be changed
without at the same time being destroyed. (It is quite wrong to locate
the difference between ﬂuid and solid matters in the different degree of
cohesion of their parts. For to call a matter ﬂuid does not depend on
the degree of resistance it opposes to the breaking up of its parts, but
only on its opposition to their mutual displacement. The former can be
a
Klebrigkeit (viscositas)
236Metaphysical foundations of natural science
as large as one wishes, but in a ﬂuid matter the latter is still always = 0.
Consider a drop of water. If a particle within it is drawn to one side by
an attraction, however great, of the neighboring parts that are in contact
with it, it is still drawn just as much to the opposite side as well; and
since the attractions mutually cancel their effects, the particle is just as
easily movable as if it were in empty space. That is, the force that is
to move it has no cohesion to overcome, but only the so-called inertia,
which it would have to overcome in all matter, even if the latter did not
cohere with anything. Thus a small microscopic organism will move just
as easily within the drop as if there were no cohesion at all to separate.
For it actually has no cohesion of the water to destroy, nor any internal
contact thereof to diminish – it needs only to change this contact. But
if you imagine that this same small organism wants to work its way out
through the external surface of the drop, then it should ﬁrst be noticed
that the mutual attraction of the parts of this water droplet causes them
to keep moving until they have attained the greatest contact with one
another, and thus the smallest contact with empty space, that is, until they
have formed a spherical shape. Now if the insect in question is striving
to work its way out beyond the surface of the drop, it must change the
spherical shape, and thus create more contact between the water and
empty space, and hence less contact of its parts with one another, that
is, [it must] diminish the cohesion of these parts. And here the water
resists it primarily through its cohesion, but not within the drop, where
the mutual contact of the parts is not diminished at all, but only changed
into contact with other parts, so that they are not in the least separated,
but only displaced. One can also apply to the microscopic organism, and
indeed on similar grounds, what Newton says of the light ray: that it
is repulsed, not by dense matter, but only by empty space. 35 It is clear,
therefore, that the increase in cohesion of the parts of a matter does
not impair its ﬂuidity in the least. Water coheres far more strongly in
its parts than is commonly believed, when one relies on the experiment
of a metal plate pulled off the surface of the water; this settles nothing,
because here the water is not torn loose over the whole surface of ﬁrst
contact, but on a much smaller one, which, in fact, it has ﬁnally arrived
at through the displacement of its parts, as a stick of soft wax, say, can be
ﬁrst drawn out thinner by a hanging weight, and must then rupture at a
much smaller surface than was originally assumed. But what is entirely
decisive in regard to our concept of ﬂuidity is this: that ﬂuid matters
can also be deﬁned a as those, in which every point endeavors to move in all
directions with precisely the same force with which it is pressed toward any one of
them, a property on which rests the ﬁrst law of hydrodynamics, although
it can never be attributed to an agglomeration of smooth and yet solid
a
erklärt
237
4: 528Metaphysical foundations of natural science
4: 529
corpuscles, as can be shown by a very easy calculation of its pressure
in accordance with the laws of composite motion, thereby proving the
original character of the property of ﬂuidity. If the ﬂuid matter were to
suffer the least resistance to displacement, and thus even the smallest
amount of friction, then the latter would increase with the strength of
the pressure by which its parts are pressed against one another, and a
pressure would ﬁnally obtain at which its parts could not be displaced
along one another by any small force. Consider, for example, a bent
tube with two arms, one of which may be arbitrarily wide, and the other
arbitrarily narrow, so long as it is not a capillary tube; 36 if one imagines
both arms several hundred feet high, then, according to the laws of
hydrostatics, the ﬂuid matter in the narrow arm would stand precisely
as high as in the wide one. But since the pressure on the bases of the
tubes, and hence also on the part that joins them in common, can be
thought as increasing to inﬁnity in proportion to the heights, it follows
that if the least amount of friction occurred between the parts of the
ﬂuid, a height for the tubes could be found, at which a small quantity of
water, poured into the narrower tube, did not disturb that in the wider
one from its place. So the water column in the former would come to
stand higher than that in the latter, because the lower parts, at such great
pressure against one another, could no longer be displaced by so small a
moving force as that of the added weight of water. But this is contrary to
experience, and even to the concept of a ﬂuid. The same holds if, instead
of pressure by weight, one posits cohesion of the parts, however great
one cares to make it. The cited second deﬁnition of ﬂuidity, on which
rests the fundamental law of hydrostatics – namely, that it is that property
of a matter whereby any part of it strives to move in all directions with
precisely the same force by which it is pushed in any given direction –
follows from the ﬁrst deﬁnition, if one combines it with the principle
of general dynamics that all matter is originally elastic. For this matter
must then be striving to expand in all directions of the space in which
it is compressed, with the same force by which the pressure occurs in
any direction, whatever it may be, that is, if the parts of a matter can
be displaced along one another by any force, without resistance, as is
actually the case with ﬂuids, it must be striving to move in all directions.
Hence friction, properly speaking, is attributable only to rigid matters
(whose possibility requires yet another ground of explanation besides the
cohesion of the parts), and friction already presupposes the property of
rigidity. But why certain matters, even though they may have no greater,
and perhaps even a lesser force of cohesion than other matters that are
ﬂuid, nevertheless resist the displacement of their parts so strongly, and
hence can be separated in no other way than by destroying the cohesion
of all parts in a given surface at once, which then yields the semblance
of a superior cohesion – how, that is, rigid bodies are possible – is still an
238Metaphysical foundations of natural science
unsolved problem, no matter how easily the common doctrine of nature
presumes to have settled it.
3. Elasticity (spring-force) is the capacity of a matter, when its mag-
nitude or ﬁgure are changed by another moving force, to reassume them again
when this latter is diminished. It is either expansive or attractive elasticity:
one to regain a previously greater volume after compression, the other
a previously smaller volume after expansion. (Attractive elasticity is ob-
viously derivative, as the term already shows. An iron wire, stretched by
a hanging weight, springs back into its volume when the band is cut.
In virtue of the same attraction that is the cause of its cohesion, or, in
the case of ﬂuid matters, if heat were suddenly to be extracted from
mercury, the matter would quickly reassume the previously smaller vol-
ume. The elasticity that consists only in regeneration of the previous
ﬁgure is always attractive, as in the case of a bent sword-blade, where
the parts, stretched away from one another on the convex surface, strive
to reassume their previous proximity, and so a small drop of mercury
can likewise be called elastic. Expansive elasticity, however, can be either
original or derivative. Thus air has a derivative elasticity in virtue of the
matter of heat, which is most intimately united with it, and whose own
elasticity is perhaps original. By contrast, the fundamental material of
the ﬂuid we call air must nonetheless, as matter in general, already have
original elasticity in itself. It is not possible to decide with certainty to
which type an observed elasticity belongs in any given case.)
4. The action of moved bodies on one another by communication of their mo-
tion is called mechanical; but the action of matters is called chemical, insofar
as they mutually change, even at rest, the combination of their parts through
their inherent forces. This chemical inﬂuence is called dissolution, inso-
far as it has the separation of the parts of a matter as its effect. (Mechanical
separation, by means of a wedge driven between the parts of a matter,
for example, is therefore entirely different from chemical separation,
because the wedge does not act by means of inherent force.) But that
chemical inﬂuence whose effect is to isolate two matters dissolved in one
another is decomposition. A dissolution of speciﬁcally different matters
by one another, in which no part of the one is found that would not be
united with a speciﬁcally different part of the other, in the same propor-
tion as the whole, is absolute dissolution, which can also be called chemical
penetration. Whether the dissolving forces that are actually to be found
in nature are capable of effecting a complete dissolution may remain un-
decided. Here it is only a question of whether such a dissolution can be
thought. Now it is obvious that, so long as the parts of a dissolved matter
remain small clots (moleculae), a dissolution of them is no less possible
than that of the larger parts. Indeed, if dissolving force remains, such a
dissolution must actually proceed until there is no longer any part that
is not made up of the solvent and the solute, in the same proportion in
239
4: 530Metaphysical foundations of natural science
4: 531
which the two are found in the whole. Thus, because in such a case there
can be no part of the volume of the solution that would not contain a
part of the solvent, the latter must ﬁll this volume completely as a con-
tinuum. In precisely the same way, because there can be no part of this
same volume of the solution that would not contain a proportional part
of the solute, the latter must also ﬁll the whole space constituting the
volume of the mixture, as a continuum. But if two matters ﬁll one and
the same space, and each of them does this completely, they penetrate one
another. Hence a complete chemical dissolution would be a penetration
of matters, which would nonetheless be entirely different from mechan-
ical penetration. For in the latter case it is thought that, as the moved
matters approach one another more closely, the repulsive force of the
one can completely surpass that of the other, so that one or both can
have their extension shrink to nothing. Here, by contrast, the extension
remains, and it is only that the matters together occupy a space, which
accords with the sum of their densities, not outside, but inside one an-
other, that is, through intussusception 37 (as it is customarily called). It is
not easy to make any objection to the possibility of this complete dissolu-
tion, and thus chemical penetration, even though it contains a completed
division to inﬁnity, which, in this case, still involves no contradiction,
because the dissolution takes place continuously throughout a time, and
thus equally through an inﬁnite series of moments with acceleration. By
the division, moreover, the sum of the surfaces of the matters yet to be
divided increases, and, since the dissolving force acts continuously, the
entire dissolution can be completed in a speciﬁable time. The inconceiv-
ability of such a chemical penetration of two matters is to be attributed to
the inconceivability of dividing any such continuum in general to inﬁn-
ity. If one recoils from this complete dissolution, then one must assume
that it proceeds only up to certain small clots of the solute, which swim
in the solvent at given distances from one another, without being able
to offer the slightest reason why these clots are not equally dissolved,
since they are still always divisible matter. It may always be true in na-
ture, so far as experience reaches, that the solvent acts no further; but
all that is at issue here is the possibility of a dissolving force that also
dissolves this clot, as well as anything left over from that, until solution
is completed. The volume occupied by the solution may be equal to,
smaller than, or even greater than the sum of the spaces occupied by
the mutually dissolving matters before mixing, depending on the ratio
of the attracting forces to the repulsions. In the solution, each matter
by itself, and both united, constitute an elastic medium, and this, on its
own, can supply a sufﬁcient reason why the solute does not again sepa-
rate from the solvent through its weight. For the attraction of the latter,
since it takes place equally strongly in all directions, itself destroys the
resistance of the solute; and to assume a certain viscosity in the ﬂuid by
240Metaphysical foundations of natural science
no means accords with the great force that such dissolved matters, acids
diluted with water, for example, exert on metallic bodies. They do not
merely lie on them, as would have to happen if they were merely aﬂoat
in their medium; rather, they split them up with great attractive force,
and spread throughout the entire space of the vehicle. Moreover, even
if we supposed that art had no such dissolving forces at its disposal as
to effect a complete dissolution, nature, in its vegetable or animal oper-
ations, could still perhaps manifest them, and thereby generate matters
that, although certainly mixed, can be separated again by no art. 38 This
chemical penetration could also be found even where one of two matters
is not in fact separated by and literally dissolved in the other, as caloric,
for example, penetrates bodies; for if it merely dispersed into their empty
interstices, the solid substance would itself remain cold, since it could
not absorb anything from it. We might even imagine, in this way, an
apparently free passage of certain matters through others, for example,
of magnetic matter, without preparing, for this purpose, open passages
and empty interstices in all matters, even the most dense. Yet here is not
the place to uncover hypotheses for particular phenomena, but only the
principle in accordance with which they are all to be judged. Everything
that relieves us of the need to resort to empty spaces is a real gain for
natural science, for they give the imagination far too much freedom to
make up by fabrication a for the lack of any inner knowledge of nature.
In the doctrine of nature, the absolutely empty and the absolutely dense
are approximately what blind accident and blind fate are in metaphysical
science, namely, an obstacle to the governance of reason, whereby it is
either supplanted by fabrication or lulled to rest on the pillow of occult
qualities.
But now as to the procedure of natural science with respect to the
most important of all its tasks – namely, that of explaining a potentially
inﬁnite speciﬁc variety of matters – one can take only two paths in this
connection: the mechanical, by combination of the absolutely full with
the absolutely empty, and an opposing dynamical path, by mere variety
in combining the original forces of repulsion and attraction to explain
all differences of matters. The ﬁrst has as materials for its derivation
atoms and the void. An atom is a small part of matter that is physically
indivisible. A matter is physically indivisible when its parts cohere with
a force that cannot be overpowered by any moving force in nature. An
atom, insofar as it is speciﬁcally distinguished from others by its ﬁg-
ure, is called a primary particle. A body (or particle) whose moving force
depends on its ﬁgure is called a machine. The mode of explaining the
speciﬁc variety of matters by the constitution and composition of their
smallest parts, as machines, is the mechanical natural philosophy. But that
a
Erdichtung
241
4: 532Metaphysical foundations of natural science
4: 533
which derives this speciﬁc variety from matters, not as machines, that is,
mere instruments of external moving forces, but from the moving forces
of attraction and repulsion originally inherent in them, can be called the
dynamical natural philosophy. The mechanical mode of explanation, since
it is the most tractable for mathematics, has, under the name of atomism
or the corpuscular philosophy, always retained its authority and inﬂuence
on the principles of natural science, with few changes from Democritus
of old, up to Descartes, and even to our time. What is essential therein
is the presupposition of the absolute impenetrability of the primitive mat-
ter, the absolute homogeneity of this material, leaving only differences in
the shape, and the absolute insurmountability of the cohesion of matter
in these fundamental particles themselves. These were the materials for
generating speciﬁcally different matters, so as not only to have at hand
an invariable, and, at the same time, variously shaped fundamental ma-
terial for explaining the invariability of species and kinds, but also to
explain mechanically, from the shapes of these primary parts, as machines
(where nothing further is lacking but an external impressed force), the
manifold workings of nature. But the ﬁrst and foremost authentication
for this system rests on the apparently unavoidable necessity for using
empty spaces on behalf of the speciﬁc difference in the density of matters. These
spaces were taken to be distributed within the matters, and between
these particles, in any proportion found necessary, so great, indeed, for
the sake of some phenomena, that the ﬁlled part of the volume of even
the densest matter is virtually negligible relative to the empty part. –
In order now to introduce a dynamical mode of explanation (which is
much more appropriate and conducive to experimental philosophy, in
that it leads us directly to the discovery of matter’s inherent moving
forces and their laws, while restricting our freedom to assume empty
interstices and fundamental particles of determinate shapes, neither of
which are determinable or discoverable by any experiment), it is not at
all necessary to frame new hypotheses. It is only necessary to refute the
postulate of the merely mechanical mode of explanation – namely, that
it is impossible to think a speciﬁc difference in the density of matters without
interposition of empty spaces – by simply advancing a mode of explana-
tion in which this can be thought without contradiction. For once the
postulate in question, on which the merely mechanical mode of expla-
nation rests, is shown to be invalid as a principle, then it obviously does
not have to be adopted as an hypothesis in natural science, so long as a
possibility remains for thinking the speciﬁc difference in densities even
without any empty interstices. But this necessity 39 rests on the circum-
stance that matter does not ﬁll its space (as merely mechanical natural
scientists assume) by absolute impenetrability, but rather by repulsive
force, which has a degree that can be different in different matters; and,
since in itself it has nothing in common with the attractive force, which
242Metaphysical foundations of natural science
depends on the quantity of matter, it may be originally different in degree
in different matters whose attractive force is the same. Thus the degree
of expansion of these matters, when the quantity is the same, and, con-
versely, the quantity of matter at the same volume, that is, its density,
originally admit of very large speciﬁc differences. In this way, one would
not ﬁnd it impossible to think a matter (as one imagines the aether, for
example) that completely ﬁlled its space without any emptiness, and yet
with an incomparably smaller quantity of matter, at the same volume,
than any bodies we can subject to our experiments. In the aether, the
repulsive force must be thought as incomparably larger in proportion to
its inherent attractive force than in any other matters known to us. And
this, then, is the one and only assumption that we make, simply because
it can be thought, but only to controvert an hypothesis (of empty spaces),
which rests solely on the pretension that such a thing cannot be thought
without empty spaces. For, aside from this, no law of either attractive or
repulsive force may be risked on a priori conjectures. Rather, everything,
even universal attraction as the cause of weight, must be inferred, to-
gether with its laws, from data of experience. Still less may such laws be
attempted for chemical afﬁnities otherwise than by way of experiments.
For it lies altogether beyond the horizon of our reason to comprehend
original forces a priori with respect to their possibility; all natural phi-
losophy consists, rather, in the reduction of given, apparently different
forces to a smaller number of forces and powers that explain the actions
of the former, although this reduction proceeds only up to fundamental
forces, beyond which our reason cannot go. And so metaphysical inves-
tigation behind that which lies at the basis of the empirical concept of
matter is useful only for the purpose of guiding natural philosophy, so
far as this is ever possible, to explore dynamical grounds of explanation.
For these alone permit the hope of determinate laws, and thus a true
rational coherence of explanations.
This is now all that metaphysics can ever achieve toward the con-
struction of the concept of matter, and thus to promote the application of
mathematics to natural science, with respect to those properties whereby
matter ﬁlls a space in a determinate measure – namely, to view these prop-
erties as dynamical, and not as unconditioned original positings, a as a
merely mathematical treatment might postulate them.
The well-known question as to the admissibility of empty spaces in
the world may serve as our conclusion. The possibility of such spaces can-
not be disputed. For space is required for all forces of matter, and, since
it also contains the conditions of the laws of diffusion of these forces,
it is necessarily presupposed prior to all matter. Thus attractive force is
attributed to matter insofar as it occupies a space around itself, through
a
Positionen
243
4: 534
4: 535Metaphysical foundations of natural science
attraction, without at the same time ﬁlling this space. Thus this space
can be thought as empty, even where matter is active, because matter is
not active there by repulsive forces, and hence does not ﬁll this space.
But no experience, or inference therefrom, or necessary hypothesis for
their explanation, can justify us in assuming empty spaces as actual. For
all experience yields only comparatively empty spaces for our cognition,
which can be completely explained, to any arbitrary degree, by the mat-
ter’s property of ﬁlling its space with greater or inﬁnitely diminishing
expansive force, without requiring empty spaces.
244Third Chapter
Metaphysical foundations of mechanics
4: 536
EXPLICATION 1
Matter is the movable insofar as it, as such a thing, has moving force.
Remark
This is now the third deﬁnition of matter. The merely dynamical concept
could consider matter also as at rest; for the moving force there dealt with
had merely to do with the ﬁlling of a certain space, without the matter
ﬁlling it needing to be seen as itself moved. a Repulsion was therefore an
originally-moving force for imparting motion. In mechanics, by contrast,
the force of a matter set in motion is considered as communicating this
motion to another. It is clear, however, that the movable would have
no moving force by means of its motion, if it did not possess originally-
moving forces, by which it is active in every place where it is found,
prior to any inherent motion of its own. b No matter would impress
proportionate motion on another matter lying straight ahead and in the
way of its motion, if both did not possess original laws of repulsion; nor
could a matter, by its motion, compel another to follow straight behind
it (to drag it along behind), if both did not possess attractive forces.
Thus all mechanical laws presuppose dynamical laws, and a matter, as
moved, can have no moving force except by means of its repulsion or
attraction, on which, and with which, it acts immediately in its motion,
and thereby communicates its own inherent motion to another. I will be
forgiven if I do not here further discuss the communication of motion
by attraction (for example, if a comet, perhaps, with stronger attractive
power than the earth, were to drag the latter in its wake in passing ahead
of it), but only that by means of repulsive forces, and thus by pressure
(as by means of tensed springs), or through impact. For, in any event,
the application of the laws of the one case to those of the other differs
only in regard to the line of direction, but is otherwise the same in both
cases.
a
b
ohne daβ die Materie, die ihn erfüllte, selbst als bewegt angesehen werden durfte
vor aller eigener Bewegung
245
4: 537Metaphysical foundations of natural science
EXPLICATION 2
The quantity of matter is the aggregate of the movable in a determinate
space. Insofar as all its parts are considered as acting (moving) together in
their motion, it is called mass, and one says that a matter acts in mass, when
all its parts, moved in the same direction, together exert their moving force
externally. A mass of determinate shape is called a body (in the mechanical
meaning). The quantity of motion (estimated mechanically) is that which
is estimated by the quantity of the moved matter and its speed together;
phoronomically it consists merely in the degree of speed.
PROPOSITION 1
The quantity of matter, in comparison with every other matter, can be
estimated only by the quantity of motion at a given speed.
Proof
4: 538
Matter is inﬁnitely divisible. So its quantity cannot be immediately deter-
mined by an aggregate of its parts. For even if this occurs in comparing the
given matter with another of the same kind, in which case the quantity of
matter is proportional to the size of the volume, it is still contrary to the
requirement of the proposition, that it is to be estimated in comparison
with every other (including the speciﬁcally different). Hence matter can-
not be validly estimated, either immediately or mediately, in comparison
with every other, so long as we abstract from its own inherent motion;
no other generally valid measure remains, therefore, except the quantity
of its motion. But here the difference of motion, resting on the differing
quantity of matters, can be given only when the speed of the compared
matters is assumed to be the same; hence, etc.
Note
The quantity of motion of bodies is in compound ratio to that of the
quantity of their matter and their speed, that is, it is one and the same
whether I make the quantity of matter in a body twice as large, and
retain the same speed, or double the speed, and retain precisely this
mass. For the determinate concept of a quantity is possible only through
the construction of the quantum. But in regard to the concept of quan-
tity, this is nothing but the composition of the equivalent; so construction
of the quantity of a motion is the composition of many motions equiva-
lent to one another. Now according to the phoronomical propositions,
it is one and the same whether I impart to a single movable a certain de-
gree of speed, or to each of many movables all smaller degrees of speed,
246Metaphysical foundations of natural science
resulting from the given speed divided by the aggregate of movables.
From this ﬁrst arises a seemingly phoronomical concept of the quantity
of a motion, as composed of many motions of movable points, external
to one another yet united in a whole. If these points are now thought
as something that has moving force through its motion, then there arises
from this the mechanical concept of the quantity of motion. In phoron-
omy, however, it is not appropriate to represent a motion as composed
of many motions external to one another, since the movable, as it is here
represented as devoid of moving force, yields no other difference in the
quantity of motion, in any composition with several of its kind, than
that which consists merely in speed. As the quantity of motion in a body
relates to that of another, so also does the magnitude of their action,
but this is to be understood as the entire action. Those who merely took
the quantity of a space ﬁlled with resistance as the measure of the entire
action (for example, the height to which a body with a certain speed
can rise against gravity, or the depth to which it can penetrate into soft
matters) came out with another law of moving forces for actual motions –
namely, that of the compound ratio of the quantity of matters and the
squares of their speeds. But they overlooked the magnitude of action in
the given time, during which the body traverses its space at a lower speed;
and yet this alone can be the measure of a motion that is exhausted by
a given uniform resistance. Hence there can be no difference, either,
between living and dead forces, if the moving forces are considered me-
chanically, that is, as those which bodies have insofar as they themselves
are moved, whether the speed of their motion be ﬁnite or inﬁnitely small
(mere striving toward motion). Rather, it would be much more appro-
priate to call dead forces those, such as the original moving forces of
dynamics, whereby matter acts on another, even when we abstract com-
pletely from its own inherent motion, and also even from its striving to
move; by contrast, one could call living forces all mechanical moving
forces, that is, those moving by inherent motion, without attending to
the difference of speed, whose degree may even be inﬁnitely small – if in
fact these terms for dead and living forces still deserve to be retained. 40
Remark
In order to avoid prolixity we will merge the explanation of the previous
three statements into one remark.
That the quantity of matter can only be thought as the aggregate of
movables (external to one another), as the deﬁnition expresses it, is a
remarkable and fundamental proposition of general mechanics. For it
is thereby indicated that matter has no other magnitude than that con-
sisting in the aggregate of manifold [elements] external to one another,
and hence has no degree of moving force at a given speed that would be
247
4: 539Metaphysical foundations of natural science
4: 540
independent of this aggregate, and could be considered merely as inten-
sive magnitude – which would be the case, however, if matter consisted
of monads, whose reality in every relation must have a degree that can be
larger or smaller, without depending on an aggregate of parts external
to one another. As to the concept of mass in this same explication, one
cannot take it in the customary way to be the same as that of quantity
[of matter]. Fluid matters can act by their own inherent motion in a
mass, a but they can also act as a ﬂuid. b In the so-called water hammer,
the impulsive water acts in a mass, that is, with all its parts together. 41
The same thing happens when water enclosed in a vessel presses down
with its weight on the scale on which it stands. By contrast, the water
of a millstream does not act on the paddle of an undershot waterwheel
in a mass, that is, with all its parts impinging on this paddle together,
but only one after the other. 42 Thus, if the quantity of matter, which is
moved with a certain speed, and has moving force, is to be determined
here, one must ﬁrst look for the water body, that is, that quantity of mat-
ter which, if it acts in a mass with a certain speed (with its weight), can
bring about the same effect. So we also customarily understand by the
word mass the quantity of matter in a solid body (the vessel in which a
ﬂuid is contained can also stand proxy for its solidity). Finally, there is
something peculiar in the Proposition together with its appended Note.
According to the former the quantity of matter must be estimated by the
quantity of motion at a given speed, but according to the latter the quan-
tity of motion (of a body, for that of a point consists merely in the degree
of speed) must, at the same speed, in turn be estimated by the quantity of
the matter moved. And this seems to revolve in a circle, and to promise
no determinate concept from either the one or the other. This alleged
circle would be an actual one, if it were a reciprocal derivation of two
identical concepts from one another. But it contains only the explication
of a concept, on the one hand, and that of its application to experience,
on the other. The quantity of the movable in space is the quantity of
matter; but this quantity of matter (the aggregate of the movable) man-
ifests itself c in experience only by the quantity of motion at equal speed
(for example, by equilibrium). d
It is to be noted, further, that the quantity of matter is the quantity of
substance in the movable, and thus not the magnitude of a certain quality
of the movable (the repulsion or attraction that are cited in dynamics),
and that the quantum of substance here means nothing else but the mere
aggregate of the movable that constitutes matter. For only this aggregate
a
b
c
d
in Masse
im Flusse
beweiset sich
Gleichgewicht
248Metaphysical foundations of natural science
of the moved can yield, at the same speed, a difference in the quantity
of motion. But that the moving force a matter has in its own inherent
motion alone manifests a the quantity of substance, rests on the concept of
the latter as the ultimate subject in space (which is in turn no predicate of
another) – which, for precisely this reason, can have no other magnitude
than that of the aggregate of homogeneous [elements] external to one
another. Now since the inherent motion of matter is a predicate that
determines its subject (the movable), and indicates in a matter, as an ag-
gregate of movables, a plurality of the subjects moved (at the same speed
and in the same way), which is not the case for dynamical properties,
whose magnitude can also be that of the action of a single subject (where
an air particle, for example, can have more or less elasticity); it therefore
becomes clear how the quantity of substance in a matter has to be
estimated mechanically only, that is, by the quantity of its own inherent
motion, and not dynamically, by that of the original moving forces. 43
Nevertheless, original attraction, as the cause of universal gravitation,
can still yield a measure of the quantity of matter, and of its substance
(as actually happens in the comparison of matters by weighing), even
though a dynamical measure – namely, attractive force – seems here to
be the basis, rather than the attracting matter’s own inherent motion. 44
But since, in the case of this force, the action of a matter with all its
parts is exerted immediately on all parts of another, and hence (at equal
distances) is obviously proportional to the aggregate of the parts, the
attracting body also thereby imparts to itself a speed of its own inherent
motion (by the resistance of the attracted body), which, in like external
circumstances, is exactly proportional to the aggregate of its parts; so the
estimation here is still in fact mechanical, although only indirectly so.
4: 541
PROPOSITION 2
First Law of Mechanics. In all changes of corporeal nature the total quan-
tity of matter remains the same, neither increased nor diminished.
Proof
(From general metaphysics we take as basis the proposition that in all
changes of nature no substance either arises or perishes, and here it is only
shown what substance shall be in matter.) 45 In every matter the movable
in space is the ultimate subject of all accidents inhering in matter, and the
aggregate of these movables, external to one another, is the quantity of
substance. Hence the quantity of matter, with respect to its substance, is
a
beweise
249
4: 542Metaphysical foundations of natural science
nothing else but the aggregate of substances of which it consists. There-
fore, the quantity of matter cannot be increased or diminished except in
such a way that new substance thereof arises or perishes. Now substance
never arises or perishes in any change of matter; so the quantity of matter
is also neither increased nor diminished thereby, but remains always the
same, and, indeed, as a whole – in such a way, that is, that somewhere
in the world it persists in the same quantity, although this or that mat-
ter can be increased or diminished, through addition or separation of
parts.
Remark
4: 543
What is essential in this proof to the characterization of the substance
that is possible only in space, and in accordance with its condition, and
thus possible only as object of the outer senses, is that its quantity cannot
be increased or diminished without substance arising or perishing. For,
since all quantity of an object possible merely in space must consist of
parts external to one another, these, if they are real (something movable),
must therefore necessarily be substances. By contrast, that which is con-
sidered as object of inner sense can have a magnitude, as substance, which
does not consist of parts external to one another; and its parts, therefore, are
not substances; and hence their arising or perishing need not be the
arising or perishing of a substance; and their augmentation or diminu-
tion, then, is possible without violating the principle of the persistence
of substance. So consciousness, and thus the clarity of representations in
my soul, and therefore the faculty of consciousness, apperception, and
even, along with this, the very substance of the soul, have a degree, which
can be greater or smaller, without any substance at all needing to arise
or perish for this purpose. But since, from its gradual diminution, the
complete disappearance of the faculty of apperception would ﬁnally have
to result, the very substance of the soul would still be subject to a gradual
perishing, even if it were of a simple nature; for this disappearance of its
fundamental force could result, not by division (separation of substance
from a composite), but rather, as it were, by expiration a – and this, too,
not in a moment, but by a gradual waning of its degree, whatever the
cause of this might be. 46 The I, the general correlate of apperception,
and itself merely a thought, designates, as a mere preﬁx, a thing of unde-
termined meaning – namely, the subject of all predicates – without any
condition at all that would distinguish this representation of the subject
from that of a something in general: a substance, therefore, of which,
by this term, one has no concept of what it may be. By contrast, the
concept of a matter as substance is the concept of the movable in space.
a
Erlöschen
250Metaphysical foundations of natural science
It is therefore no wonder if the persistence of substance can be proved
of the latter, but not of the former, since, in the case of matter, it already
results from its concept – namely, that it is the movable, which is possible
only in space – that what has quantity therein contains a plurality of the
real external to one another, and thus a plurality of substances; and hence
the quantity of matter can be diminished only by division, which is not
disappearance – and the latter would also be impossible in matter ac-
cording to the law of constancy. a The thought I, by contrast, is no concept
at all, but only inner perception, and so nothing at all can be inferred
from it (except for the total distinctness of an object of inner sense from
that which is thought merely as object of the outer senses) – including,
in particular, the persistence of the soul as substance.
PROPOSITION 3
Second Law of Mechanics. Every change in matter has an external cause.
(Every body persists in its state of rest or motion, in the same direction,
and with the same speed, if it is not compelled by an external cause to
leave this state.)
Proof
(From general metaphysics we take as basis the proposition that every
change has a cause, and here it is only to be proved of matter that its
change must always have an external cause.) 47 Matter, as mere object of
the outer senses, has no other determinations except those of external
relations in space, and therefore undergoes no change except by motion.
With respect to the latter, as change of one motion into another, or of a
motion into rest, or conversely, a cause must be found (by the principle
of metaphysics). But this cause cannot be internal, for matter has no
essentially internal determinations or grounds of determination. Hence
every change in a matter is based on external causes (that is, a body
persists, etc.).
Remark
This mechanical law must alone be called the law of inertia (lex inertiae);
the law of an equal and opposite reaction for every action cannot bear
this name. For the latter says what matter does, but the former only
what it does not do, which is more appropriate to the term inertia. The
inertia of matter is, and means, nothing else than its lifelessness, as matter
in itself. Life is the faculty of a substance to determine itself to act from an
a
nach dem Gesetze der Stetigkeit
251
4: 544Metaphysical foundations of natural science
internal principle, of a ﬁnite substance to change, and of a material substance
[to determine itself] to motion or rest, as change of its state. Now we
know no other internal principle in a substance for changing its state ex-
cept desiring, and no other internal activity at all except thinking, together
with that which depends on it, the feeling of pleasure or displeasure, and
desire or willing. But these actions and grounds of determination in no
way belong to representations of the outer senses, and so neither [do they
belong] to the determinations of matter as matter. Hence all matter, as
such, is lifeless. The principle of inertia says this, and nothing more. If
we seek the cause of any change of matter in life, we will have to seek
it forthwith in another substance, different from matter, yet combined
with it. For in natural knowledge we ﬁrst have to be acquainted with the
laws of matter, as such a thing, and to purge them from the admixture of
all other active causes, before we connect them with these latter, in order
properly to distinguish how, and in what manner, each of them acts in
itself alone. The possibility of a proper natural science rests entirely and
completely on the law of inertia (along with that of the persistence of
substance). The opposite of this, and thus also the death of all natural
philosophy, would be hylozoism. From this very same concept of inertia,
as mere lifelessness, it follows at once that it does not mean a positive striv-
ing to conserve its state. Only living beings are called inert in this latter
sense, because they have a representation of another state, which they
abhor, and against which they exert their power.
PROPOSITION 4
Third mechanical law. In all communication of motion, action and reaction
are always equal to one another.
Proof
4: 545
(From general metaphysics we must borrow the proposition that all
external action in the world is interaction. Here, in order to stay within
the bounds of mechanics, it is only to be shown that this interaction (ac-
tio mutua) is at the same time reaction (reactio); but here I cannot wholly
leave aside this metaphysical law of community, without detracting from
the completeness of the insight.) 48 All active relations of matters in space,
and all changes of these relations, insofar as they may be causes of certain
actions or effects, a must always be represented as mutual; that is, because
all change of matter is motion, we cannot think any motion of a body
in relation to another absolutely at rest that is thereby also to be set in
motion. Rather, the latter must be represented as only relatively at rest
a
Wirkungen [actions or effects]
252Metaphysical foundations of natural science
with respect to the space that we relate it to, but as moved, together with
this space, in the opposite direction, with precisely the same quantity of
motion in absolute space as the moved body there has toward it. For the
change of relation (and thus the motion) between the two is completely
mutual; as much as the one body approaches every part of the other, by
so much does the other approach every part of the ﬁrst. And, since it is
here a question, not of the empirical space surrounding the two bodies,
but only of the line lying between them (in that they are considered sim-
ply in relation to one another, in accordance with the inﬂuence that the
motion of the one can have on the change of state of the other, abstract-
ing from all relation to the empirical space), their motion is therefore
considered as determinable merely in absolute space, in which each of
the two must have an equal share in the motion that is ascribed to one
of them in relative space, in that there is no reason to ascribe more of
this motion to one than to the other. On this basis, a the motion of a
body A with respect to another body B at rest, in regard to which it can
thereby be moving, is reduced to absolute space; that is, as a relation of
acting causes merely related to one another, this motion is so considered
that both have an equal share in the motion which, in the appearance,
is ascribed to body A alone. And the only way this can happen is that
the speed ascribed in relative space to body A alone is apportioned be-
tween A and B in inverse ratio to their masses – to A alone its speed in
absolute space, and to B, together with the relative space in which it is at
rest, its speed in the opposite direction. The same appearance of motion
is thereby perfectly maintained, but the action in the community of the
two bodies is constructed as follows.
Let a body A be approaching the body B, with a speed = AB with respect
to the relative space in which B is at rest. One divides the speed AB into
two parts, Ac and Bc, which relate to one another inversely as the masses
B and A, and imagines that A is moved with speed Ac in absolute space,
while B is moved with speed Bc in the opposite direction, together with
the relative space; thus the two motions are equal and opposite to one
another, and, since they mutually cancel one another, the two bodies
place themselves relative to one another, that is, in absolute space, at
rest. 49 But now B was in motion, together with the relative space, with
a
Auf diesem Fuβ
253
4: 546Metaphysical foundations of natural science
4: 547
4: 547
speed Bc, in the direction BA exactly opposite to that of body A, namely,
AB. Hence, if the motion of body B is canceled through the impact, the
motion of the relative space is not thereby canceled as well. Therefore,
the relative space moves after the impact, with respect to the two bodies
A and B (now at rest in absolute space), in the direction BA with speed
Bc, or, equivalently, both bodies move after the impact, with equal speed
Bd = Bc, in the direction of the impacting body AB. But now, according
to the preceding, the quantity of motion of body B in the direction and
with the speed Bc, and hence also that in the direction Bd with the same
speed, is equal to the quantity of motion of body A, with the speed and
in the direction Ac. Therefore, the action or effect, that is, the motion
Bd in relative space that body B receives through the impact, and thus
also the action of body A with speed Ac, is always equal to the reaction
Bc. Precisely the same law (as mathematical mechanics teaches) holds
without modiﬁcation, if, instead of an impact on a body at rest, one
assumes an impact of the same body on one similarly moved. Moreover,
the communication of motion through impact differs from that through
traction only in the direction in which the matters resist one another
in their motions. It follows, then, that in all communication of motion
action and reaction are always equal to one another (that every impact
can communicate the motion of one body to another only by means of
an equal counter-impact, every pressure by means of an equal counter-
pressure, and every traction only through an equal counter-traction). ∗
∗
In phoronomy, where the motion of a body was considered merely with respect
to space, as change of relation in space, it was all the same whether I wanted
to grant motion to the body in space, or an equal but opposite motion to
the relative space instead; the two yielded entirely the same appearance. The
quantity of motion of the space was merely the speed, and hence that of the
body was likewise nothing but its speed (for which reason it could be considered
as a mere movable point). But in mechanics, where a body is considered in
motion relative to another, with regard to which, through its motion, it has
a causal relation – namely, that of moving the body itself – in that it enters
into community with [the body] either in its approach through the force of
impenetrability or in its withdrawal through that of attraction, it is no longer
the same whether I wish to ascribe a motion to one of these bodies, or an
opposite motion to the space. For another concept of the quantity of motion
now comes into play, namely, not that which is thought merely with respect to
space, and consists only in the speed, but rather that whereby the quantity of
substance (as moving cause) must be brought into the calculation at the same
time; and here it is no longer arbitrary, but rather necessary, to assume each of
the two bodies as moved, and, indeed, with equal quantity of motion in the
opposite direction – but if one is relatively at rest with respect to the space, to
ascribe the required motion to it, together with the space. For one body cannot
act on the other through its own inherent motion, except either in approach
by means of repulsive force, or in withdrawal by means of attraction. Now
254Metaphysical foundations of natural science
Note 1
From this there follows a law of nature that is not unimportant for
general mechanics: any body, however great its mass, must be movable
by the impact of any other body, however small its mass or speed. For
to the motion of A in the direction AB there necessarily corresponds an
opposite and equal motion of B in the direction BA. The two motions
cancel one another through their impact in absolute space. But the two
bodies thereby acquire a speed Bd = Bc in the direction of the impacting
body; hence body B is movable by any force of impact, however small.
Note 2
This, then, is the mechanical law of the equality of action and reaction,
which rests on the fact that no communication of motion takes place, ex-
cept insofar as we presuppose a community of these motions, and thus on
the fact that no body impacts another that is at rest relative to it; rather, the
second body is at rest relative to space, only insofar as it moves, together
with this space, in the same amount, but in the opposite direction, with
that motion which then falls to the ﬁrst as its relative share, and together
would originally yield the quantity of motion that we would ascribe to
the ﬁrst in absolute space. For no motion that is to be moving with respect
to another body, can be absolute; but if it is relative with respect to the
latter, then there is no relation in space that would not be mutual and
equal. – There is, however, another law of the equality of action and
reaction among matters – namely, a dynamical law – not insofar as one
matter communicates its motion to another, but rather as it imparts this
motion originally to it, and, at the same time, produces the same in itself
through the latter’s resistance. This can easily be shown in a similar way.
For, if matter A exerts traction on a matter B, then it compels the latter to
approach it, or, equivalently, it resists the force with which the latter might
since both forces always act mutually and equally in opposite directions, no
body can act by means of them on another body through its motion, without
just as much reaction from the other with the same quantity of motion. Hence
no body can impart motion to an absolutely resting body through its motion;
rather, the latter must be moved precisely with the same quantity of motion
(together with the space) in the opposite direction as that which it is supposed
to receive through the motion, and in the direction, of the ﬁrst. – The reader
will easily become aware that, despite the somewhat unaccustomed character
of this mode of representing the communication of motion, it can nonetheless
be set in the clearest light, if one does not shy away from the prolixity of the
explanation.
a
zieht
255
4: 548Metaphysical foundations of natural science
4: 549
strive to remove itself. But since it is all the same whether B removes itself
from A, or A from B, this resistance is, at the same time, a resistance ex-
erted by body B against A, insofar as the latter may be striving to remove
itself from the former; and so traction and counter-traction are equal to
one another. In just the same way, if A repels matter B, then A resists the
approach of B. But, since it is all the same whether B approaches A, or A ap-
proaches B, B also resists the approach of A to precisely the same extent;
so pressure and counter-pressure are also always equal to one another.
Remark 1
Such, then, is the construction of the communication of motion, which
at the same time brings with it, as its necessary condition, the law of
the equality of action and reaction. Newton by no means dared to prove
this law a priori, and therefore appealed rather to experience. Others, for
its sake, introduced into natural science a special force of matter, under
the name, ﬁrst introduced by Kepler, of a force of inertia (vis inertiae), 50
and thus they, too, derived it in principle from experience. Finally, still
others posited in the concept a mere communication of motion, which
they viewed as a gradual transfer of the motion of the one body into
the other, whereby the mover would have to lose precisely as much
motion as it imparts to the moved, until it impresses no more on the
latter (that is, until it has already arrived at equality of speed with the
latter in the same direction). ∗ In this way, they eliminated in principle
all reaction, that is, all actual reacting force of the impacted body on
∗
4: 549
The equality of action with the (in this case falsely so-called) reaction comes out
equally well, if, on the hypothesis of transfusion of motion from one body into
another, one allows the moved body A to deliver its entire motion in an instant
to the body at rest, so that it is itself at rest after the impact – this case was
inevitable as soon as one thought both bodies as absolutely hard (which property
must be distinguished from elasticity). But since this law of motion would
agree neither with experience, nor with itself in application, no other remedy
was known but to deny the existence of absolutely hard bodies, which amounted
to admitting the contingency of this law, in that it was supposed to rest on a
particular quality of matters that move one another. In our presentation of this
law, by contrast, it is all the same whether one wishes to think the colliding
bodies as absolutely hard or not. It is completely inconceivable to me, however,
how the transfusionists of motion would explain, in their fashion, the motion
of elastic bodies through impact. For here it is clear that the resting body does
not, merely as resting, acquire motion lost by the impacting body, but that, in
the collision, it exerts actual force on the latter in the opposite direction, so as
to compress, as it were, a spring between the two, which requires just as much
actual motion on its part (but in the opposite direction) as the moving body
itself has need of for this purpose.
256Metaphysical foundations of natural science
the impacting one (which would be capable, for example, of tensing
a spring). Moreover, they do not demonstrate what is properly meant
in the law in question, and did not at all explain the communication of
motion itself with regard to its possibility. For the term transfer of motion
from one body to another explains nothing, and, if it is not meant to be
taken literally (in violation of the principle that accidents do not wander
from one substance to another), 51 as if motion were poured from one
body into another like water from one glass into another, then we here
have precisely the problem of how to make this possibility conceivable –
where the explanation thereof in fact rests on precisely the same ground
as that from which the law of the equality of action and reaction is
derived. One cannot think at all how the motion of a body A must be
necessarily combined with the motion of another body B, except by
thinking forces in both that pertain to them (dynamically) prior to all
motion (for example, repulsion), and now being able to demonstrate that
the motion of body A, in its approach toward B, is necessarily combined
with the approach of B toward A (and, if B is viewed as at rest, with its
motion, together with its space, toward A), insofar as the bodies, with their
(original) moving forces, are considered to be in motion merely relative
to one another. This latter can be comprehended completely a priori,
once it is seen that, whether body B is at rest or moved relative to the
empirically knowable space, it still must be viewed, relative to body A, as
necessarily moved, and, indeed, in the opposite direction. For otherwise
no inﬂuence of B on the repulsive force of the two bodies would occur,
and without this no mechanical action of the matters on one another –
that is, no communication of motion by impact – is possible at all.
4: 550
Remark 2
Regardless of the famous name of its creator, the terminology of inertial
force (vis inertiae) must therefore be entirely banished from natural sci-
ence, not only because it carries with it a contradiction in terms, nor even
because the law of inertia (lifelessness) might thereby easily be confused
with the law of reaction in every communicated motion, but primarily
because the mistaken idea of those who are not properly acquainted with
the mechanical laws is thereby maintained and even strengthened –
according to which the reaction of bodies discussed under the name of
inertial force would amount to a draining off, diminution, or eradication
of the motion in the world; but the mere communication of motion
would not be effected thereby, because the moving body would have to
apply a part of its motion solely in overcoming the inertia of the one at
rest (which would then be a pure loss), and could set the latter in motion
only with the remaining part; but if none were left over, it would com-
pletely fail to move the latter by its impact, because of its great mass. But
257
4: 551Metaphysical foundations of natural science
nothing can resist a motion except the opposite motion of another body,
and certainly not its state of rest. Thus here the inertia of matter, that is,
the mere incapacity for self-movement, is not the cause of a resistance.
A special, entirely peculiar force merely to resist, without being able to
move a body, under the name of an inertial force, would be a word without
any meaning. The three laws of general mechanics could therefore more
appropriately be named the laws of self subsistence, inertia, and reaction of
matters (lex subsistentiae, inertiae, et antagonismi) in all of their changes. That
these laws, and thus all Propositions of the present science, precisely
answer to the categories of substance, causality, and community, insofar as
these concepts are applied to matter, needs no further discussion.
GENERAL REMARK TO MECHANICS
4: 552
The communication of motion occurs only by means of such moving
forces as also inhere in a matter at rest (impenetrability and attraction).
The action of a moving force on a body in an instant is its solicitation;
the speed effected in the latter through solicitation, insofar as it can
increase in equal proportion to the time, is the moment of acceleration. a
(The moment of acceleration must therefore contain only an inﬁnitely
small speed, because otherwise the body would thereby attain an inﬁnite
speed in a given time, which is impossible; moreover, the possibility of
acceleration b in general, by means of a continued moment thereof, rests
on the law of inertia.) The solicitation of matter by expansive force
(of compressed air bearing a weight, for example) occurs always with
a ﬁnite speed, but the speed thereby impressed on (or extracted from)
another body can only be inﬁnitely small; for expansion is only a surface
force, or, what is the same, the motion of an inﬁnitely small quantum
of matter, which therefore must occur with ﬁnite speed, in order to
be equal to the motion of a body with ﬁnite mass and inﬁnitely small
speed (a weight). By contrast, attraction is a penetrating force, and with
such a force a ﬁnite quantum of matter exerts moving force on another
similarly ﬁnite quantum. The solicitation of attraction must therefore be
inﬁnitely small, because it is equal to the moment of acceleration (which
must always be inﬁnitely small) – which is not the case with repulsion,
since an inﬁnitely small part of matter is to impress a moment on a ﬁnite
one. No attraction can be thought with a ﬁnite speed, without the matter
having to penetrate itself by its own force of attraction. For the attraction
that a ﬁnite quantity of matter exerts on a ﬁnite one, with a ﬁnite speed,
must at all points of the compression be greater than any ﬁnite speed
whereby the matter reacts through its impenetrability, but with only an
a
b
Moment der Acceleration
Beschleunigung
258Metaphysical foundations of natural science
inﬁnitely small part of the quantity of its matter. If the attraction is only a
surface force, as we think cohesion to be, then the opposite of this would
result. But it is impossible to think cohesion in this way if it is to be true
attraction (and not mere external compression).
An absolutely hard body would be one whose parts attracted a one
another so strongly that they could neither be separated, nor changed
in their situation relative to one another, by any weight. Now since the
parts of the matter of such a body would have to attract b one another
with a moment of acceleration that would be inﬁnite with respect to
that of gravity, but ﬁnite with respect to that of the mass that is to be
driven thereby, the resistance by means of impenetrability, as expansive
force, since it always occurs with an inﬁnitely small quantity of matter,
would then have to take place with a more than ﬁnite speed of solic-
itation, that is, the matter would strive to expand with inﬁnite speed,
which is impossible. Hence an absolutely hard body, that is, one that
would, on impact, instantaneously oppose a body moved at ﬁnite speed,
with a resistance equal to the total force of that body, is impossible. 52
Consequently, by means of its impenetrability or cohesion, a matter at-
tains instantaneously only an inﬁnitely small resistance to the force of a
body in ﬁnite motion. And from this there now follows the mechanical
law of continuity (lex continui mechanica): namely, that in no body is the
state of rest or motion, or the speed or direction of the latter, changed
by impact instantaneously, but only in a certain time, through an inﬁnite
series of intermediate states, whose difference from one another is less
than that between the ﬁrst state and the last. A moved body that impacts
on a matter is thus not brought into a state of rest by the latter’s resis-
tance all at once, but only through a continuous retardation; and one that
was at rest is put into motion only through a continuous acceleration; c
and it is changed from one degree of speed to another only in accor-
dance with the same rule. In the same way, the direction of its motion
is changed into one that makes an angle with it no otherwise than by
means of all possible intermediate directions, that is, by means of mo-
tion in a curved line. (And, on similar grounds, this law can be extended
also to the change of state of a body by attraction). This lex continui is
based on the law of the inertia of matter, whereas the metaphysical law
of continuity would have to be extended to all changes in general (inner
as well as outer), and thus would have to be based on the mere concept
of a change in general, as quantity, and on the generation thereof (which
would necessarily proceed continuously in a certain time, as does time
itself ). This metaphysical law can therefore ﬁnd no place here. 53
a
b
c
zögen
ziehen
Acceleration
259
4: 553Fourth Chapter
Metaphysical foundations of phenomenology
4: 554
EXPLICATION
Matter is the movable insofar as it, as such a thing, can be an object of
experience.
Remark
4: 555
Motion, like everything that is represented through the senses, is given
only as appearance. For its representation to become experience, we
require, too, that something be thought through the understanding –
namely, besides the mode in which the representation inheres in the sub-
ject, also the determination of an object thereby. Hence the movable, as
such a thing, becomes an object of experience, when a certain object (here
a material thing) is thought as determined with respect to the predicate of
motion. But motion is change of relation in space. There are thus always
two correlates here, such that either, ﬁrst, the change can be attributed
in the appearance to one just as well as to the other, and either the one or
the other can be said to be moved, because the two cases are equivalent; a
or, second, one must be thought in experience as moved to the exclusion
of the other; or, third, both must be necessarily represented through
reason as equally moved. In the appearance, which contains nothing but
the relation in the motion (with respect to its change), none of these
determinations are contained. But if the movable, as such a thing, namely,
with respect to its motion, is to be thought of as determined for the sake
of a possible experience, it is necessary to indicate the conditions under
which the object (matter) must be determined in one way or another
by the predicate of motion. At issue here is not the transformation of
semblance b into truth, but of appearance into experience; for, in the case
of semblance, the understanding with its object-determining judgments
is always in play, although it is in danger of taking the subjective for
objective; in the appearance, however, no judgment of the understanding
is to be met with at all 54 – which needs to be noted, not merely here, but
in the whole of philosophy, because otherwise, when appearances are in
a
b
weil beides gleichgültig ist
Schein
260Metaphysical foundations of natural science
question, and this term is taken to have the same meaning as semblance,
one is always poorly understood.
PROPOSITION 1
The rectilinear motion of a matter with respect to an empirical space, as
distinct from the opposite motion of the space, is a merely possible pred-
icate. The same when thought in no relation at all to a matter external
to it, that is, as absolute motion, is impossible.
Proof
Whether a body is said to be moved in a relative space, and the latter
at rest, or whether, conversely, the latter shall be said to be moved, with
the same speed in the opposite direction, with the former at rest, is not a
dispute about what pertains to the object, but only about its relation to the
subject, and belongs therefore to appearance and not to experience. For
if the observer locates himself in that space as at rest, the body counts as
moved for him; if he locates himself (at least in thought) in another space
comprehending the ﬁrst, relative to which the body is likewise at rest,
then that relative space counts as moved. Thus in experience (a cognition
that determines the object validly for all appearances) there is no differ-
ence at all between the motion of the body in the relative space, and the
body being at rest in absolute space, together with an equal and oppo-
site motion of the relative space. Now the representation of an object
through one of two predicates, which are equally valid with respect to
the object, and differ from one another only in regard to the subject
and its mode of representation, is not a determination in accordance
with a disjunctive judgement, but merely a choice in accordance with an
alternative judgement. (In the former, of two objectively opposed predi-
cates, one is assumed to the exclusion of the other for the determina-
tion of the object; in the latter, of two judgments objectively equiva-
lent, yet subjectively opposed to one another, one is assumed for the
determination of the object without excluding its opposite – and thus
by mere choice.) ∗ This means that through the concept of motion, as
object of experience, it is in itself undetermined, and therefore equiv-
alent, whether a body be represented as moved in the relative space,
or the latter with respect to the former. Now that which is in itself
undetermined with respect to two opposed predicates is to that extent
merely possible. Hence the rectilinear motion of a matter in empirical
space, as distinct from the equal opposite motion of the space, is a
∗
More will be said about this distinction between disjunctive and alternative
opposition in the General Remark to this Chapter.
261
4: 556Metaphysical foundations of natural science
merely possible predicate in experience – which was the ﬁrst [thing to
be proved].
Since, moreover, a relation, and thus also a change thereof, that is,
motion, can be an object of experience only insofar as both correlates are
objects of experience, whereas the pure space that is also called absolute
space, in contrast to relative (empirical) space, is no object of experience,
and in general a is nothing, a rectilinear motion without reference to
anything empirical, that is, absolute motion, is completely impossible –
which was the second [thing to be proved].
Remark
This Proposition determines the modality of motion with respect to
phoronomy.
PROPOSITION 2
4: 557
The circular motion of a matter, as distinct from the opposite motion
of the space, is an actual predicate of this matter; by contrast, the op-
posite motion of a relative space, assumed instead of the motion of the
body, is no actual motion of the latter, but, if taken to be such, is mere
semblance.
Proof
Circular motion (like all curvilinear motion) is a continuous change of
rectilinear motion, and, since the latter is itself a continuous change of
relation with respect to the external space, circular motion is a change
of a change in these external relations in space, and is thus a continuous
arising of new motions. Now since, according to the law of inertia, a
motion, insofar as it arises, must have an external cause, while the body,
at every point on this circle (according to precisely the same law), is
striving, for its own part, to proceed in the straight line tangent to the
circle, which motion acts in opposition to this external cause, it follows
that every body in circular motion manifests, b by its motion, a moving
force. But the motion of the space, as distinct from that of the body, is
merely phoronomic, and has no moving force. Thus the judgment that
here either the body is moved, or the space is moved in the opposite
direction, is a disjunctive judgment, whereby, if one of the terms (namely,
the motion of the body) is posited, the other, (namely, that of the space)
is excluded. Thus the circular motion of a body, as distinct from that of
a
b
überall
beweiset
262Metaphysical foundations of natural science
the space, is an actual motion, so that the latter, even though it agrees
with the former according to the appearance, nevertheless contradicts it
in the context of all appearances, that is, of a possible experience, and so
is nothing but mere semblance.
Remark
This Proposition determines the modality of motion with respect to
dynamics; for a motion that cannot take place without the inﬂuence of
a continuously acting external moving force manifests, a directly or in-
directly, originally moving forces of matter, whether of attraction or
repulsion. – Moreover, Newton’s Scholium to the Deﬁnitions he has pre-
ﬁxed to his Principia may be consulted on this subject, toward the end,
where it becomes clear that the circular motion of two bodies around a
common central point (and thus also the axial rotation of the earth) can
still be known by experience even in empty space, and thus without any
empirically possible comparison with an external space; 55 so that a mo-
tion, therefore, which is a change of external relations in space, can be
empirically given, even though this space is not itself empirically given,
and is no object of experience. This is a paradox that deserves to be
solved.
PROPOSITION 3
In every motion of a body, whereby it is moving relative to another, an
opposite and equal motion of the latter is necessary.
Proof
According to the Third Law of Mechanics (Proposition 4), the com-
munication of motion of bodies is possible only by the community of
their original moving forces, and the latter only by mutually opposite
and equal motion. The motion of both is therefore actual. But since the
actuality of this motion does not rest (as in the second Proposition) on
the inﬂuence of external forces, but follows immediately and unavoid-
ably from the concept of the relation of the moved in space to anything
else movable thereby, the motion of the latter is necessary.
Remark
This Proposition determines the modality of motion with respect to
mechanics. – Moreover, it is obvious that these three Propositions
a
beweiset
263
4: 558Metaphysical foundations of natural science
determine the motion of matter with respect to its possibility, actuality,
and necessity, and thus with respect to all three categories of modality.
GENERAL REMARK TO PHENOMENOLOGY
4: 559
Thus here appear three concepts, whose use in general natural science
is unavoidable, and whose precise determination is therefore necessary,
although not that easy or comprehensible – namely, [ﬁrst,] the concept
of motion in relative (movable) space, second, that of motion in absolute
(immovable) space, and third, that of relative motion in general, as distinct
from absolute motion. The concept of absolute space is the basis for all of
them. But how do we arrive at this peculiar concept, and what underlies
the necessity of its use?
It cannot be an object of experience, for space without matter is no
object of perception, and yet it is a necessary concept of reason, and
thus nothing more than a mere idea. For in order that motion may be
given, even merely as appearance, an empirical representation of space
is required, with respect to which the movable is to change its relation;
but the space that is to be perceived must be material, and thus itself
movable, in accordance with the concept of a matter in general. Now,
to think of it as moved, one may think it only as contained in a space
of greater extent, and take the latter to be at rest. But the same can be
done with the latter, with respect to a still further extended space, and
so on to inﬁnity, without ever arriving by experience at an immovable
(immaterial) space, with respect to which either motion or rest might
absolutely a be attributed to any matter. Rather, the concept of these
relational determinations will have to be continually revised, according
to the way that we will consider the movable in relation to one or another
of these spaces. Now since the condition for regarding something as at
rest or moved is always conditioned in turn, ad inﬁnitum, in relative space,
it becomes clear, ﬁrst, that all motion or rest can be relative only and never
absolute, that is, that matter can be thought as moved or at rest solely in
relation to matter, and never with respect to mere space without matter,
so that absolute motion, thought without any relation of one matter to
another, is completely impossible; and second, for precisely this reason,
that no concept of motion or rest valid for all appearance is possible in
relative space. Rather, one must think a space in which the latter can
itself be thought as moved, but which depends for its determination on
no further empirical space, and thus is not conditioned in turn – that
is, an absolute space to which all relative motions can be referred, in
which everything empirical is movable, precisely so that in it all motion
a
schlechthin
264Metaphysical foundations of natural science
of material things a may count as merely relative with respect to one
another, as alternatively-mutual, ∗ but none as absolute motion or rest
(where, while one is said to be moved, the other, in relation to which it is
moved, is nonetheless represented as absolutely b at rest). Absolute space
is therefore necessary, not as a concept of an actual object, but rather
as an idea, which is to serve as a rule for considering all motion therein
merely as relative; and all motion and rest must be reduced to absolute
space, if the appearance thereof is to be transformed into a determinate
concept of experience (which unites all appearances).
Thus the rectilinear motion of a body in relative space is reduced
to absolute space, when I think the body as in itself at rest, but this
space as moved in the opposite direction in absolute space (which is
not apprehended by the senses), and when I think this representation
as that which yields precisely the same appearance, whereby all possible
appearances of rectilinear motions that a body may have at the same time
are reduced to the concept of experience which unites them all, namely,
that of merely relative motion and rest.
Because circular motion, according to the second Proposition, can be
given as actual motion in experience, even without reference to the exter-
nal empirically given space, it indeed seems to be absolute motion. For
relative motion with respect to the external space (for example, the axial
rotation of the earth relative to the stars of the heavens) is an appearance,
∗
a
b
In logic the either-or always signiﬁes a disjunctive judgment, where, if the one is
true, the other must be false. For example, a body is either moved or not moved,
that is, at rest. For here [in logic] one speaks solely of the relation of the cog-
nition to the object. In the doctrine of appearance, where it is a matter of the
relation to the subject, so as to determine the relation to the object therefrom,
the situation is different. For here the proposition that the body is either moved
and the space at rest, or conversely, is not a disjunctive proposition in an ob-
jective relation, but only in a subjective one, and the two judgments contained
therein are valid alternatively. In precisely the same phenomenology, where the
motion is considered, not merely phoronomically, but rather dynamically, the
disjunctive proposition is instead to be taken in an objective meaning; that is,
I cannot assume, in place of the rotation of a body, a state of rest of the lat-
ter and the opposite motion of the space instead. But wherever the motion is
considered mechanically (as when a body approaches another seemingly at rest),
then the formally disjunctive judgment must be used distributively with respect
to the object, so that the motion must not be attributed either to one or to the
other, but rather an equal share of it to each. This distinction among alter-
native, disjunctive, and distributive determinations of a concept with respect to
opposing predicates has its importance, but cannot be further discussed here.
Materiellen
schlechthin
265
4: 560
4: 559
4: 560Metaphysical foundations of natural science
4: 561
in place of which the opposite motion of this space (of the heavens) in
the same time can be supposed as completely equivalent to the former;
but, according to the Proposition in question, it certainly may not be so
substituted in experience. Hence, that rotation is not to be represented
as externally relative, which sounds as if this kind of motion is to be taken
as absolute.
But it should be noted that it is here a question of true (actual) motion,
which does not, however, appear as such, so that, if one wished to evaluate
it merely in accordance with empirical relations to space, it could be taken
for rest; it is a question, that is, of true motion as distinct from semblance,
but not of absolute motion in contrast to relative. Thus, circular motion,
although it in fact exhibits no change of place in the appearance, that is,
no phoronomic change in the relations of the moved body to (empirical)
space, exhibits nonetheless a continuous dynamical change, demonstrable
through experience, in the relations of matter within its space, for example,
a continual diminution of attraction in virtue of a striving to escape,
as an action or effect a of the circular motion, and thereby assuredly
indicates its difference from semblance. For example, one may represent
to oneself the earth as rotating on its axis in inﬁnite empty space, and also
verify this motion by experience, even though neither the relation of the
earth’s parts among one another, nor to the space outside it, is changed
phoronomically, that is, in the appearance. For, with respect to the ﬁrst,
as empirical space, nothing changes its position in or on the earth; and,
as regards the second, which is completely empty, no externally changed
relation, and thus no appearance of a motion, can take place anywhere.
But if I represent to myself a deep hole descending to the center of
the earth, and I let a stone fall into it, I ﬁnd, however, that the falling
stone deviates from its perpendicular direction continuously, and, in fact,
from west to east, even though gravity, at all distances from the center
of the earth, is always directed toward it, and I conclude, therefore, that
the earth is rotating on its axis from west to east. b Or, if I also remove the
stone further out from the surface of the earth, and it does not remain
over the same point of the surface, but moves away from it from east
to west, 56 then I will infer to the very same previously mentioned axial
rotation of the earth, and both observations will be sufﬁcient to prove
the actuality of this motion. The change of relation to the external space
(the starry heavens) does not sufﬁce for this, since it is mere appearance,
which may proceed from two in fact opposing grounds, and is not a
cognition derived from the explanatory ground of all appearances of
this change, that is, experience. But that this motion, even though it is
no change of relation to the empirical space, is nevertheless not absolute
a
b
Wirkung [action or effect]
von Abend gegen Morgan
266Metaphysical foundations of natural science
motion, but rather a continuous change in the relations of matters to one
another, which, although represented in absolute space, is thus actually
only relative, and, for just that reason, is true motion – this rests on
the representation of the mutual and continuous withdrawal of any part
of the earth (outside the axis) from any other part lying diametrically
opposite to it at the same distance from the center. For this motion is
actual in absolute space, in that the reduction of the distance in question,
which gravity by itself would induce in the body, is thereby continuously
made up, and, in fact, without any dynamical repulsive cause (as may be
seen from the example chosen by Newton in the Principia, page 10 of
the 1714 edition); ∗ hence it is made up through actual motion, which
relates, however, to the space inside of the moved matter (namely, its
center), and not to that outside it. 57
In the case of the third Proposition, to show the truth of the mutually
opposed and equal motions of the two bodies, even without reference to
the empirical space, we do not even need the active dynamical inﬂuences,
given through experience, that are required in the second case (gravity
or a tensed cord). Rather, the mere dynamical possibility of such an
inﬂuence, as a property of matter (repulsion or attraction), leads by itself,
and from mere concepts of a relative motion, from the motion of one
body to the simultaneous equal and opposite motion of the other, when
the latter is considered in absolute space, that is, in accordance with
truth. Hence like everything sufﬁciently provable from mere concepts,
this is a law of an absolutely necessary counter-motion.
There is thus no absolute motion, even when a body in empty space
is thought as moved with respect to another; their motion here is not
considered relative to the space surrounding them, but only to the space
between them, which, considered as absolute space, alone determines
their external relations to one another, and is in turn only relative. Abso-
lute motion would thus be only that which pertained to a body without
relation to any other matter. Only the rectilinear motion of the cosmos, a
that is, the system of all matter, would be such a motion. For if, outside
∗ There he says: “It is indeed a matter of great difﬁculty to discover, and effec-
tually to distinguish, the true motions of particular bodies from the apparent;
because the parts of that immovable space, in which these motions are per-
formed, do by no means come under the observation of our senses. Yet the
thing is not altogether desperate.” He then lets two spheres connected by a
cord revolve around their common center of gravity in empty space, and shows
how the actuality of their motion, together with its direction, can nonetheless
be discovered by means of experience. I have attempted to show this also in the
case of the earth moved around its axis, in somewhat altered circumstances.
a Weltganzen
267
4: 562Metaphysical foundations of natural science
4: 563
a matter, there were any other at all, even separated from it by empty
space, then the motion would already be relative. For this reason, any
proof of a law of motion, which amounts to showing that its opposite
would have to result in a rectilinear motion of the entire cosmic system, a
is an apodictic proof of its truth, simply because absolute motion would
then follow, which is utterly impossible. Such is the law of antagonism
in all community of matter through motion. For any deviation from it
would shift the common center of gravity of all matter, and thus the en-
tire cosmic system, from its place – which would not happen, however,
if one wanted to imagine this system as rotating on its axis. Hence it is
always possible to think such a motion, although to suppose it would, so
far as one can see, be entirely without any conceivable use.
To the various concepts of motion and moving forces there also corre-
spond the various concepts of empty space. Empty space in the phoronom-
ical sense, which is also called absolute space, should not properly be
called an empty space; for it is only the idea of a space, in which I abstract
from all particular matter that makes it an object of experience, in order
to think therein the material space, or any other empirical space, as
movable, and thereby to think of motion, not merely in a one-sided
fashion as absolute, but always mutually, as a merely relative predicate.
It is therefore nothing at all that belongs to the existence of things, but
merely to the determination of concepts, and to this extent no empty
space exists. Empty space in the dynamical sense is that which is not ﬁlled,
that is, in which no other movable resists the penetration of a movable,
and thus no repulsive force acts; it can either be empty space within the
world (vacuum mundanum) or, if the latter is represented as bounded,
empty space outside the world (vacuum extramundanum). The former,
too, can be represented either as dispersed (vacuum disseminatum, which
constitutes only a part of the volume of matter), or as accumulated
empty space (vacuum coacervatum, which separates bodies, for example,
the heavenly bodies, from one another). This latter distinction is
certainly not an essential one, since it rests only on a difference in the
locations assigned to empty space within the world, but is still employed
for various purposes: the ﬁrst, to derive speciﬁc differences in density,
and the second, to derive the possibility of a motion in the universe free
from all external resistance. That it is not necessary to assume empty
space for the ﬁrst purpose has already been shown in the General Remark
to Dynamics; but that it is impossible can in no way be proved from its
concept alone, in accordance with the principle of noncontradiction.
Nevertheless, even if no merely logical reason for rejecting this kind of
empty space were to be found here, there could still be a more general
a
Weltgebäude
268Metaphysical foundations of natural science
physical reason for expelling it from the doctrine of nature – that of
the possibility of the composition of a matter in general, if only this
were better understood. For if the attraction assumed in order to explain
the cohesion of matter should be only apparent, not true attraction,
and were merely the effect, say, of a compression by external matter (the
aether) distributed everywhere in the universe, which is itself brought
to this pressure only through a universal and original attraction, namely,
gravitation (a view that is supported by several reasons), then empty
space within matter, although not logically impossible, would still be
so dynamically, and thus physically, since any matter would expand of
itself into the empty spaces assumed within it (since nothing resists its
expansive force here), and would always keep them ﬁlled. An empty
space outside the world, understood as the totality of preeminently
attractive matters (the large heavenly bodies), would be impossible for
precisely the same reasons, since in accordance with their mass, as the
distance from them increases, the attractive force on the aether (which
encloses all these bodies, and, driven by that force, conserves them in
their density by compression) decreases in inverse proportion, so that
the latter would itself only decrease indeﬁnitely in density, but nowhere
leave space completely empty. It should not surprise anyone, however,
that this refutation of empty space proceeds entirely hypothetically,
for the assertion of empty space fares no better. Those who venture to
settle this disputed question dogmatically, whether for or against, rely
in the end on plainly metaphysical presuppositions, as can be seen from
the Dynamics – and here it was necessary at least to show that these can
do nothing at all to resolve the problem. As for empty space in the third,
or mechanical sense, it is the emptiness accumulated within the cosmos
to provide the heavenly bodies with free motion. It is easy to see that
the possibility or impossibility of this does not rest on metaphysical
grounds, but on the mystery of nature, difﬁcult to unravel, as to how
matter sets limits to its own expansive force. Nevertheless, if one
grants what was said in the General Remark to Dynamics concerning
the possibility of an ever-increasing expansion of speciﬁcally different
materials, at the same quantity of matter (in accordance with their
weight), it may well be unnecessary to suppose an empty space for the
free and enduring motion of the heavenly bodies; since even in spaces
completely ﬁlled, the resistance can still be thought as small as one
likes.
And so ends the metaphysical doctrine of body with the empty, and there-
fore the inconceivable, wherein it shares the same fate as all other at-
tempts of reason, when it strives after the ﬁrst grounds of things in a
retreat to principles – where, since its very nature entails that it can
269
4: 564Metaphysical foundations of natural science
4: 565
never conceive anything, except insofar as it is determined under given
conditions, and since it can therefore neither come to a halt at the con-
ditioned, nor make the unconditioned comprehensible, nothing is left
to it, when thirst for knowledge invites it to comprehend the absolute
totality of all conditions, but to turn away from the objects to itself, so
as to explore and determine, not the ultimate limits of things, but rather
the ultimate limits of its own unaided powers.
