Logical forms of judgments: Quantity, quality, relation, and modality

The distinctions among judgments in respect of their form may be traced back to
the four principal moments of quantify, quality, relation, and modality,
in regard to which just as many different kinds of judgments are determined.

Quantity of judgments: Universal, particular, singular

As to quantity, judgments are either universal or particular or singular,
accordingly as the subject is either wholly included in or excluded from the
notion of the predicate or is only in part excluded in or excluded from it.
In the universal judgment, the sphere of one concept is
wholly enclosed within the sphere of another;
in the particular, a part of the former is enclosed under the sphere of the other;
and in the singular judgment, finally, a concept that has no sphere at all is
enclosed, merely as part then, under the sphere of another.

    As to logical form, singular judgments are to be assessed as like universal
    ones in use, for in both the predicate holds of the subject without excep-
    tion. In the singular proposition, Caius is mortal, for example, there can
    just as little be an exception as in the universal one, All men are mortal.
    For there is only one Caius.

    In respect of the universality of a cognition, there exists a real distinction
    between general and universal propositions, which of course does not
    have to do with logic, however. General propositions, namely, are those
    that contain merely something of the universal regarding certain objects,
    and consequently do not contain sufficient conditions of subsumption,
    e.g., the proposition that one must make proofs thorough. Universal"
    propositions are those that maintain something universally 1 ' of an object.
    j. Universal rules are either analytically or synthetically universal. The former
    abstract from differences, the latter attend to distinctions and conse-
    quently determine in regard to them too. The more simply an object is
    thought, the more possible is analytic universality in consequence of a
    concept.

    If we cannot have insight into universal propositions in their universality
    without cognizing 01 them in concreto, then they cannot serve as a standard
    and hence cannot hold heuristically in application, but are only assignments
    to investigate the universal grounds for that with which we first became
    acquainted in particular cases. For example, the proposition, He who has no
    interest in lying and who knows the truth will speak the truth; we cannot have
    insight into this proposition in its universality, because we are acquainted
    with the restriction to the condition of the disinterested - namely, that
    men can lie out of interest, which comes because they do not hold fast to
    morality - only through experience. An observation that teaches us the
    weakness of human nature.

    Of particular judgments it is to be noted that if it is to be possible to have
    insight into them through reason, and hence for them to have a rational,
    not merely intellectual (abstracted) form, then the subject must be a
    broader concept (conceptus latior) than the predicate. Let the predicate
    always = O> the subject Q, then
    is a particular judgment, for some of what belongs under a is b, some not
    b - that follows from reason. But let it be
    then at least all a can be contained under b, if it is smaller, but not if it is
    greater, hence it is particular only by accident.

Quality of judgments: Affirmative, negative, infinite

As to quality, judgments are either affirmative or negative or infinite.
In the affirmative judgment the subject is thought under the sphere of a predicate,
in the negative it is posited outside the sphere of the latter, and
in the infinite it is posited in the sphere of a concept that lies outside
the sphere of another.

The infinite judgment indicates not merely that a subject is not contained
under the sphere of a predicate, but that it lies somewhere in the infinite
sphere outside its sphere; consequently this judgment represents the
sphere of the predicate as restricted.
Everything possible is either^ or nonA. If I say, then, something is non
A, e.g., the human Soul is non-mortal, some men are non-learned, etc.,
then this is an infinite judgment. For it is not thereby determined, con-
cerning the finite sphere A, under which concept the object belongs, but
merely that it belongs in the sphere outside A, which is really no sphere at
all but only a sphere's sharing of a limit with the infinite* or the limiting
itself. y Now although exclusion is a negation, the restriction 2 of a concept
is still a positive act. Therefore limits" are positive concepts of restricted*
objects.

According to the principle of the excluded middle (exclusi tertii),
the sphere of one concept relative to another is either exclusive or inclusive.
Now since logic has to do merely with the form of judgment,
not with concepts as to their content,
the distinction of infinite from negative judgments is
not proper to this science.
In negative judgments the negation always affects the copula;
in infinite ones it is not the copula but rather the predicate that is affected,
which may best be expressed in Latin.

Relation of judgments: Categorical, hypothetical, disjunctive

As to relation, judgments are either categorical or hypothetical or disjunctive.
The given representations in judgment are subordinated one to another
for the unity of consciousness, namely, either
as predicate to subject, or
as consequence to ground, or
as member of the division to the divided concept.
Through the first relation categorical judgments are determined,
through the second hypothetical,
and through the third disjunctive.

Categorical judgments

In categorical judgments, subject and predicate constitute their matter;
the form, through which the relation (of agreement or of opposition)
between subject and predicate is determined and expressed,
is called the copula.

    Categorical judgments constitute the matter of the remaining judgments,
    to be sure, but one must not on this account believe,
    as several logicians do, that both hypothetical and disjunctive judgments are
    nothing more than various clothings of categoricals and
    hence may be wholly traced back to these latter.
    All three kinds of judgments rest on essentially
    different logical functions of the understanding and
    must therefore be considered according to their specific difference.

Hypothetical judgments

The matter of hypothetical judgments consists of
two judgments that are connected with one another
as ground and consequence.
One of these judgments, which contains the ground, is
the antecedent (antecedens, prius),
the other, which is related to it as consequence, is
the consequent (consequens, posterius),
and the representation of this kind of connection of
two judgments to one another for the unity of consciousness is called the consequentia,
which constitutes the form of hypothetical judgments.

    What the copula is for categorical judgments, then,
    the consequentia is for hypotheticals: their form.

    Some believe it is easy to transform a
    hypothetical proposition into a categorical.
    But this will not do, because the two are wholly different from
    one another as to their nature.
    In categorical judgments nothing is problematic,
    rather, everything is assertoric;
    but in hypotheticals only the consequentia is assertoric.
    In the latter I can thus connect two false judgments with one another,
    for here it is only a matter of the correctness of the connection:
    the form of the consequentia, on which the logical truth of these judgments rests.
    There is an essential difference between the two propositions,
    All bodies are divisible, and,
    If all bodies are composite, then they are divisible.
    In the former proposition I maintain the thing directly,
    in the latter only under a condition expressed problematically.

Modes of connection in hypothetical judgments: modus ponens and modus tollens

The form of the connection in hypothetical judgments is of two kinds:
the positing (modus ponens) and the denying (modus tollens).
1. If the ground (antecedens) is true, then
the consequence (consequens) determined by it is true too;
called modus ponens.
2. If the consequence (consequens) is false, then
the ground (antecedens) is false too;
modus tollens.

Disjunctive judgments

A judgment is disjunctive if the parts of the sphere of a given concept
determine one another in the whole or toward a whole as complements
(complementa).

Matter and form of disjunctive judgments

The several given judgments of which the disjunctive judgment is composed
constitute its matter and are called the members of the disjunction or opposition.
The form of these judgments consists in the disjunction itself,
in the determination of the relation of the various judgments as
members of the whole sphere of the divided cognition which mutually
exclude one another and complement one another.

    Thus all disjunctive judgments represent various judgments as
    in the community of a sphere and produce each judgment only through
    the restriction of the others in regard to the whole sphere;
    they determine each judgment's relation to the whole sphere, then, and
    thereby at the same time the relation that these various members of
    the division (membra disjuncta) have among themselves.
    Thus one member determines every other here only
    insofar as they stand together in community as parts of a whole sphere of
    cognition, outside of which, in a certain relation, nothing may be thought.

Peculiar character of disjunctive judgments

The peculiar character of all disjunctive judgments, whereby their specific
difference from others, in particular from categorical judgments, is deter-
mined as to the moment of relation, consists in this: that the members of
the disjunction are all problematic judgments, of which nothing else is
thought except that, taken together as parts of the sphere of a cognition,
each the complement of the other toward the whole (complementum ad
totum), they are equal to the sphere of the first. And from this it follows
that in one of these problematic judgments the truth must be contained
or - what is the same - that one of them must hold assertorically, because
outside of them the sphere of the cognition includes nothing more under
the given conditions, and one is opposed to the other, consequently nei-
ther something outside them nor more than one among them can be true.
Note.
In a categorical judgment the thing whose representation is considered as a
part of the sphere of another, subordinated representation is considered as
contained under this, its higher concept; thus here, in the subordination of
the spheres, the part of the part is compared with the whole. But in
disjunctive judgments I go from the whole to all the parts taken together.
What is contained under the sphere of a concept is also contained under
one of the parts of this sphere. Accordingly, the sphere must first be
divided. If, for example, I make the disjunctive judgment, A learned man is
learned either historically or in matters of reason, I thereby determine that
these concepts are, as to the sphere, parts of the sphere of the learned man,
but not in any way parts of one another, and that taken together they are all
complete.
The following schema of comparison between categorical and disjunc-
live judgments may make it more intuitive that in disjunctive judgments the
sphere of the divided concept is not considered as contained in the sphere
of the divisions, but rather that which is contained under the divided
concept is considered as contained under one of the members of the
division.
In categorical judgments x, which is contained under b, is also under a:
In disjunctive ones x, which is contained under a, is contained either under
b or c, etc.:
Thus the division in disjunctive judgments indicates the coordination not
of the parts of the whole concept, but rather all the parts of its sphere.
Here I think many things through one concept, there one thing through' many
concepts, e.g., the definitum through all the marks of coordination.

Modality of judgments: Problematic, assertoric, apodeictic

As to modality, through which moment the relation of
the whole judgment to the faculty of cognition is determined,
judgments are either problematic or assertoric or apodeictic.
The problematic ones are accompanied with the consciousness of
the mere possibility of the judging,
the assertoric ones with the consciousness of its actuality,
the apodeictic ones, finally, with the consciousness of its necessity.

    This moment of modality indicates, then, only
    the way in which something is maintained or denied in judgment:
    anything concerning the truth or untruth of a judgment,
    as in the problematic judgment, The soul of man may be immortal;
    or whether one determines something concerning it,
    as in the assertoric judgment, The human soul is immortal;
    or, finally, whether one even expresses the truth of a judgment
    with the dignity of necessity, as in the apodeictic judgment,

The soul of man must be immortal. This determination of merely possi-
ble or actual or necessary truth concerns only the judgment itself, then, not
in any way the thing about which we judge.
2. In problematic judgments, which one can also explain as ones whose
material is given with the possible relation between predicate and subject,
the subject must always have a smaller sphere than the predicate.
j. On the distinction between problematic and assertoric judgments rests
the true distinction between judgments and propositions, which is custom-
arily placed, wrongly, in the mere expression through words, without
which one simply could not judge at all. In judgment the relation of
various representations to the unity of consciousness is thought merely as
problematic, but in a proposition as assertoric. A problematic proposition
is a contradictio in adjecto. Before I have a proposition I must first judge;
and I judge about much that I cannot decide, which I must do, however,
as soon as I determine a judgment as a proposition. It is good, by the way,
first to judge problematically, before one accepts the judgment as asser-
toric, in order to examine it in this way. Also, it is not always necessary to
our purpose to have assertoric judgments.

Exponible judgments

Judgments in which an affirmation and a negation are contained simultaneously,
but in a covert way, so that the affirmation occurs distinctly but
the negation covertly, are exponible propositions.

    In the exponible judgment, Few men are learned, for example lies (i.), but
    in a covert way, the negative judgment, Many men are not learned, and (2.)
    the affirmative one, Some men are learned. Since the nature of exponible
    propositions depends merely on conditions of language, in accordance
    with which one can express two judgments briefly at once, the observation
    that in our language there can be judgments that must be expounded
    belongs not to logic but to grammar.

Theoretical and practical propositions

Those propositions that relate to the object and determine what belongs
or does not belong to it are called theoretical; practical propositions, on the
other hand, are those diat state the action whereby, as its necessary condi-
tion, an object becomes possible.

    Logic has to deal only with propositions that are practical as to form, which
    are to this extent opposed to theoretical ones. Propositions that are practical
    as to content, and to this extent are distinct from speculative ones, belong to
    morals.

Indemonstrable and demonstrable propositions

Demonstrable propositions are those that are capable of a proof; those not
capable of a proof are called indemonstrable.
Immediately certain judgments are indemonstrable and thus are to be
regarded as elementary propositions.

Principles

Immediately certain judgments a priori can be called principles/ insofar as
other judgments are proved from them, but they themselves cannot be
subordinated to any other. On this account they are also called principles
(beginnings).

Intuitive and discursive principles: Axioms and acroamata

Principles are either intuitive or discursive. The former can be exhibited in
intuition and are called axioms (axiomata), the latter may be expressed only
through concepts and can be called acroamata.

Analytic and synthetic propositions

Propositions whose certainty rests on identity of concepts (of the predicate
with the notion of the subject) are called analytic propositions." 1 Proposi-
tions whose truth is not grounded on identity of concepts must be called
synthetic.

    An example of an analytic proposition is, To everything x, to which the
    concept of body (a + b) belongs, belongs also extension (b)
    An example of a synthetic proposition is, To everything x, to which the
    concept of body (a+b) belongs, belongs also attraction (c). Synthetic
    propositions increase cognition materialiter, analytic ones merely for-
    maliter. The former contain determinations (determinationes), the latter
    only logical" predicates.
    Analytic principles are not axioms, for they are discursive. And even
    synthetic principles are axioms only if they are intuitive.

Tautological propositions

The identity of the concepts in analytic judgments can be either explicit
(explicita) or non-explicit (implicita).
In the first case the analytic propositions are tautological.

    Tautological propositions are empty virtualiter, or empty of consequences,
    for they are without value or use. The tautological proposition, Man is
    man, is of this sort, for example. For if I do not know anything more to say
    of man except that he is a man, then I know nothing more of him at all.
    Propositions that are identical implicite, on the other hand, are not
    empty of consequences or fruitless, for they make clear the predicate that
    lay undeveloped (implicite) in the concept of the subject through develop-
    ment 11 (explicatio).
    Propositions that are empty of consequences must be distinguished from
    ones that are empty of sense, 9 which are empty in meaning' because they
    concern the determination of so-called hidden properties (qualitates oc-
    cultae).

Postulate and problem

A postulate is a practical, immediately certain proposition, or a principle
that determines a possible action, in the case of which it is presupposed
that the way of executing it is immediately certain.

Problems (problemata) are demonstrable propositions that require a directive,
or ones that express an action, the manner of whose execution is not immediately certain.

    There can also be theoretical postulates on behalf of practical reason.
    These are theoretical hypotheses that are necessary for a practical purpose of reason,
    like those of the existence of God, of freedom, and of another world.

    A problem involves

    (1.) the question, which contains what is to be accomplished,
    (2.) the resolution, which contains the way in which what is
    to be accomplished can be executed, and
    (3.) the demonstration that when I have proceeded thus, what is required will occur.

Theorems, corollaries, lemmas, and scholia

Theorems are theoretical propositions that are capable of and require proof.
Corollaries are immediate consequences from a preceding proposition.
Propositions that are not indigenous to the science in which they are
presupposed as proved, but rather are borrowed from other sciences, are
called lemmas (lemmata).
Scholia, finally, are mere elucidative propositions,
which thus do not belong to the whole of the system as members.

    Essential and universal moments of every theorem are
    the thesis and the demonstration.
    Furthermore, one can place the distinction between
    theorems and corollaries in the fact that
    the latter are inferred immediately,
    but the former are drawn through a series of
    consequences from immediately certain propositions.

Judgments of perception and of experience

A judgment of perception is merely subjective,
an objective judgment from perceptions is a judgment of experience.

    A judgment from mere perceptions is really not possible,
    except through the fact that I express my representation as perception:
    I, who perceive a tower, perceive in it the red color.
    But I cannot say: It is red.
    For this would not be merely an empirical judgment, but a judgment of experience, i.e.,
    an empirical judgment through which I get a concept of the object. E.g.,
    In touching the stone I sense warmth, is a judgment of perception:
    but on the other hand, The stone is warm, is a judgment of experience.
    It pertains to the latter that I do not reckon to the object
    what is merely in my subject, for a judgment of experience is perception
    from which a concept of the object arises;
    e.g., whether points of light move on the moon or in the air or in my eye.
