Content and extension of concepts

Every concept, as partial concept, is contained in the representation of things;
as ground of cognition, i.e., as mark, these things are contained under it.
In the former respect every concept has a content, in the other an extension.
The content and extension of a concept stand in inverse relation to one another.
The more a concept contains under itself, namely,
the less it contains in itself, and conversely.

    The universality or universal validity of a concept does not rest on
    the fact that the concept is a partial concept, but rather on
    the fact that it is a ground of cognition.

Quantity of the extension of concepts

The more the things that stand under a concept and can be thought through it,
the greater is its extension or sphere.

    As one says of a ground in general that it contains the consequence under itself,
    so can one also say of the concept that as ground of cognition it contains
    all those things under itself from which it has been abstracted,
    e.g., the concept of metal contains under itself gold, silver, copper, etc.
    For since every concept, as a universally valid representation, contains
    that which is common to several representations of various things,
    all these things, which are to this extent contained under it,
    can be represented through it.
    And it is just this that constitutes the usefulness of a concept.
    The more the things that can be represented through a concept,
    the greater is its sphere.
    Thus the concept body, for example, has a greater extension
    than the concept metal.

Higher and lower concepts

Concepts are called higher (conceptus superiores) insofar as
they have other concepts under themselves,
which, in relation to them, are called lower concepts.
A mark of a mark, a remote mark, is a higher concept,
the concept in relation to a remote mark is a lower one.

    Since higher and lower concepts are so called only relatively,
    one and the same concept can, in various relations, be simultaneously
    a higher one and a lower one. Thus the concept man is a higher one in
    relation to the concept Negro, but a lower one in relation to the concept
    animal.

Genus and species

The higher concept, in respect to its lower one, is called genus,
the lower concept in regard to its higher one species.
Like higher and lower concepts, genus and species concepts are distinguished
not as to their nature, then, but only in regard to their relation to
one another (termini a quo or ad quod) in logical subordination.

Highest genus and lowest species

The highest genus is that which is not a species
(genus summum non est species),
just as the lowest species is that which is not a genus
(species, quae non est genus, est infima).
In consequence of the law of continuity, however,
there cannot be either a lowest or a next species.

    If we think of a series of several concepts subordinated to one another,
    e.g., iron, metal, body, substance, thing, then here we can attain ever
    higher genera - for every species is always to be considered at the same
    time as genus in regard to its lower concept, e.g., the concept learned man
    in regard to the concept philosopher - until we finally come to a genus that
    cannot in turn be a species. And we must finally be able to attain such a
    one, because in the end there must be a highest concept (conceptus
    summus), from which, as such, nothing further may be abstracted without
    the whole concept disappearing. - But in the series of species and gen-
    era there is no lowest concept (conceptus infimus) or lowest species, under
    which no other would be contained, because such a one cannot possibly
    be determined. For even if we have a concept that we apply immediately
    to individuals, there can still be specific differences in regard to it, which
    we either do not note, or which we disregard. Only comparatively for use
    are there lowest concepts, which have attained this significance," 1 as it
    were, through convention, insofar as one has agreed not to go deeper
    here.
    In respect to the determination of species and genus concepts, then, the
    following universal law holds: There is a genus that cannot in turn be a species,
    but there is no species that should not be able in turn to be a genus.

Broader and narrower concept - Convertible concepts

The higher concept is also called a broader concept,
the lower concept a narrower one.

Concepts that have one and the same sphere are called convertible concepts
(conceptus reciproci).

Relation of the lower concept to the higher, of the broader to the narrower

The lower concept is not contained in the higher,
for it contains more in itself than does the higher one;
it is contained under it, however, because the higher contains
the ground of cognition of the lower.

Furthermore, one concept is not broader than another because it contains
more under itself - for one cannot know that - but rather insofar as it
contains under itself the other concept and besides this still more.

Universal rules in respect of the subordination of concepts

In regard to the logical extension of concepts,
the following universal rules hold:
1. What belongs to or contradicts higher concepts also belongs to or contradicts
all lower concepts that are contained under those higher ones; and
2. conversely: What belongs to or contradicts all lower concepts also belongs to
or contradicts their higher concept.

    Because that in which things agree flows from their universal properties,
    and that in which they are different from one another flows from their
    particular properties, one cannot infer that what belongs to or contradicts
    one lower concept also belongs to or contradicts other lower concepts,
    which belong with it to one higher concept. Thus one cannot infer, e.g.,
    that what does not belong to man does not belong to angels either.

Conditions for higher and lower concepts to arise:
Logical abstraction and logical determination

Through continued logical abstraction higher and higher concepts arise,
just as through continued logical determination, on the other hand,
lower and lower concepts arise.
The greatest possible abstraction yields the highest or most abstract concept,
that from which no determination can be further thought away.
The highest, completed determination would yield
a thoroughly determinate concept (conceptus omnimode determinate),
i.e., one to which no further determination might be added in thought.

    Since only individual things, or individuals, are thoroughly determinate,
    there can be thoroughly determinate cognitions only as intuitions,
    but not as concepts; in regard to the latter, logical determination
    can never be regarded as completed.

Use of concepts in abstracto and in concreto

Every concept can be used universally or particularly
(in abstracto or in concreto).
The lower concept is used in abstracto in regard to its higher one,
the higher concept in concreto in regard to its lower one.

    Thus the expressions abstract and concrete relate not to concepts in
    themselves - for every concept is an abstract concept - but rather only to
    their use. And this use can in turn have various degrees, accordingly as
    one treats a concept more or less abstractly or concretely, i.e., as one
    either leaves aside or adds more or fewer determinations.
    Through abstract use a concept comes closer to the highest genus,
    through concrete use, on the other hand, to the individual.

    Which use of concepts, the abstract or die concrete, has an advantage
    over the odier? Nodiing can be decided about this. The worth of the one
    is not to be valued less than the worth of the other. Through very abstract
    concepts we cognize little in many things, through very concrete concepts
    we cognize much m few tilings; what we win on the one side, then, we lose
    again on the other. A concept that has a large sphere is very useful insofar
    as one can apply it to many things; but in return for that, there is that
    much less contained in it. In the concept substance, for example, I do not
    think as much as in the concept chalk.

The art of popularity consists in finding the relation between
representation in abstracto and in concreto in the same cognition,
hence between concepts and their exhibition,
through which the maximum of cognition is achieved,
both as to extension and as to content.
