YS II.44

    svadhyayad ista-devata-samprayoga

YS II.45

    samadhi-siddhi îsvara-pranidhana

SK 22.

Prakriti mahat tata ahamkara tasmat ca sodasaka gana
sodasakat tasmad api pancabhyah panca bhutani

From nature evolves the Great Principle;
from this evolves the I Principle;
from this evolves the set of sixteen;
from the five of this set of sixteen evolves the five elements.

SK 23.

buddhi adhyavasaya dharma jnana viraga aisvarya
etad rupam sattvikam tasmat viparyastam tamasam

Buddhi is self-certainty.
Virtue, knowledge, dispassion, and power are its manifestations
when the sattva attribute abounds.
And the reverse of these, when the tamasa attribute abounds.

SK 24.

ahamkara abhimana tasmat pravartate dvividha sarga eva
ekadasaka gana ca tanmatra pancaka

Ahamkara is self-assertion;
from that proceeds a two-fold evolution
the set of eleven and the five-fold primary elements.

YS II.46

    sthira-sukham asanam

YS II.47

    prayatna-saithilya-ananta-samapattibhyam

YS II.48

    tato dvandva-anabhighata

SK 25.

vaikrita ahamkara ekadasaka sattvika pravartate
tanmatra bhutade sa tamasa taijasa ubhayam

The set of eleven sattvic indriyas proceeds from the Vaikriti form of I-Principle;
the set of five tamasic tanmatras proceed from the Bhutadi form of I-Principle.
From the Taijasa form of I-Principle proceed both of them.

SK 26.

buddhi indriyani akhyani caksu srotram ghrana rasana tvak
karma indriyani ahu vak pani pada payu upastha

Organs of knowledge are called the Eye, the Ear, the Nose, the Tongue, the Skin.
Organs of action are called the Speech, the Hand, the Feet, the excretory organ and the organ of generation.

YS II.49

    tasmin sati svasa-prasvasayo gati-viccheda pranayama

YS II.50

    bahya-abhyantara-stambha-vritti desa-kala-sankhyabhi-paridrsta dirgha-suksma

YS II.51

    bahya-abhyantara-visaya-aksepi caturtha

SK 27.

atra mana ubhayatmakam sankalpakam ca sadharmaya indriyam
nanatvam bahya bheda ca guna-parinama-visesa

Of these (sense organs), the Mind possesses the nature of both.
It is the deliberating principle, and is also called a sense organ
since it posseses properties common to the sense organs.
Its multifariousness and also its external diversities are owing
to special modifications of the Attributes.

SK 28.

pancanam rupadisu alocanamatram isyate
pancanam vritti vacana adana viharana utsarga ca ananda

The function of the five in respect to form and the rest,
is considered to be mere observation.
Speech, manipulation, locomotion, excretion and gratification
are the functions of the other five.

YS II.52

    tata ksiyate prakasa-avaranam

YS II.53

    dharanasu ca yogyata manasa

YS II.54

    sva-visaya-asamprayoge cittasya svarupa-anukara ive indriyanam pratyahara

YS II.55

    tata parama vasyata indriyanam

YS III.1

    desa-bandha cittasya dharana

YS III.2

    tatra pratyaya-eka-tanata dhyanam

YS III.3

    tad evartha-matra-nirbhasa svarupa-sunyam iva samadhi

YS III.4

    trayam ekatra samyama

YS III.5

    taj-jayat prajna-aloka

YS III.6

    tasya bhumisu viniyoga

YS III.7

    trayam antar-angam purvebhya

YS III.8

    tad api bahir-angam nirbijasya

SK 29.

trayasya svalaksanyam vritti sa esa asamanya bhavati
samanya karana-vritti pranadhyah-vayavah-panca

Of the three internal instruments, their own characteristics are their functions;
this is peculiar to each.
The common modification of the instruments is the five airs
such as prana and the rest.

YS III.9

    vyutthana-nirodha-samskarayor abhibhava-pradur-bhavau nirodha-kshana-cittanvayo nirodha-parinama

YS III.10

    tasya prasanta-vahita samskarat

    Manas, the understanding, first the faculty for the cognition of the general (of rules)

    a. The Concept as such

    The concept in general and its distinction from intuition

        All cognitions, that is, all representations related with consciousness to an object,
        are either intuitions or concepts.
        An intuition is a singular representation (repraesentatio singularis),
        a concept a universal (repraesentatio per notas communes) or
        reflected representation (repraesentatio discursiva).
        Cognition through concepts is called thought (cognitio discursiva).

    Matter and form of concepts

    With every concept we are to distinguish matter and form.
    The matter of concepts is the object, their form universality.

    Empirical and pure concept

    A concept is either an empirical or a pure concept
    (vel empiricus vel intellectualis).
    A pure concept is one that is not abstracted from experience but
    arises rather from the understanding even as to content.
    An idea is a concept of reason whose object simply cannot be met with in experience.

    Concepts that are given (a priori or a posteriori) and concepts that are made

    All concepts, as to matter, are either
    given (conceptus dati) or made (conceptus factitii).
    The former are given either a priori or a posteriori.
    All concepts that are given empirically or a posteriori
    are called concepts of experience,
    all that are given a priori are called notions.

    Logical origin of concepts

    The origin of concepts as to mere form rests on reflection and on abstraction from
    the difference among things that are signified by a certain representation.
    And thus arises here the question: Which acts of the understanding constitute a concept? or
    what is the same, Which are involved in the generation of a concept out of given representations?

    Logical Actus of comparison, reflection, and abstraction

    The logical actus of the understanding,
    through which concepts are generated as to their form, are:
    1. comparison of representations among one another in relation to the unity of consciousness;
    2. reflection as to how various representations can be conceived in one consciousness; and finally
    3. abstraction of everything else in which the given representations differ.

SK 30.

drste catustayasya tu yugapat vrttih tasya kramasasca nirdista
tatha api adrste trayasya vrttih tat purvika

Of all the four, the functions are said to be simultaneous and also successive
with regard to the seen objects;
with regard to the unseen objects
the functions of the three are preceeded by that.

YS III.11

    sarva-arthata-ekagratayo kshayodayau cittasya samadhi-parinama

    The logical form of all judgments consists of the objective unity of
    the apperception of the concepts contained therein

    Ahamkara, the determinative power of judgment,
    second the faculty for the subsumption of the particular under the general

    b. Judgment

    Definition of a judgment in general

        A judgment is the representation of the unity of the consciousness of various representations,
        or the representation of their relation insofar as they constitute a concept.

    Matter and form of judgments

        Matter and form belong to every judgment as essential constituents of it.
        The matter of the judgment consists in the given representations that are combined in
        the unity of consciousness in the judgment,
        the form in the determination of the way that the various representations belong, as such,
        to one consciousness.

    Object of logical reflection the mere form of judgments

        Since logic abstracts from all real or objective difference of cognition,
        it can occupy itself as little with the matter of judgments
        as with the content of concepts.
        Thus it has only the difference among judgments in regard to their mere form
        to take into consideration.

SK 31.

svam svam vrittim pratipadyante paraspara akuta hetukam
purushartha eva hetu na kenacit karanam karyate

Instruments enter into their respective functions being incited by mutual impulse.
The meaning of the Spirit is the sole motive (for the activities of the instruments).
By none whatsoever is an instrument made to act.

YS III.12

    tata punashantoditau tulya-pratyayau cittasya-ekagrata-parinama

    Buddhi, reason, third the faculty for the determination of
    the particular from the general (for the derivation from principles)

    c. Syllogism

    Inference in general

        By inferring is to be understood that function of thought whereby one judgment is derived from another.
        An inference is thus in general the derivation of one judgment from the other.

    Immediate and mediate inferences

        All inferences are either immediate or mediate.
        An immediate inference (consequentia immediata) is the derivation (deductio) of
        one judgment from the other without a mediating judgment (judicium intermedium).
        An inference is mediate if, besides the concept that a judgment contains in itself,
        one needs still others in order to derive a cognition therefrom.

    Inferences of the understanding, inferences of reason,
    and inferences of the power of judgment

        Immediate inferences are also called inferences of the understanding;
        all mediate inferences, on the other hand, are either
        inferences of reason or
        inferences of the power of judgment.

SK 32.

karanam trayodasavidham tad aharana dharana prakasakaram
tasya karyam ca dasadha aharyam dharyam prakasyam ca

Instruments are of thirteen kinds performing the functions of
seizing, sustaining and illuminating.
Its objects are of ten kinds, vis
the seized, the sustained, and the illumined.

YS III.13

    etena bhutendriyesu dharma-laksana-vastha-parinama vyakhyata

    Mechanism

SK 33.

antahkaranam trividham bahyam dasadha trayasya visayakhyam
bahyam sampratakalam abhyantaram karanam trikalam

The internal instruments are three-fold.
The external are ten-fold; they are called the objects of the three (internal instruments).
The external instruments function at the present time and
the internal instruments function at all the three times.

YS III.14

    shantoditavyapadesya-dharmanupati dharmi

    Chemism

SK 34.

Tesam panca buddhi visesa-avisesa-visaya
Vak sabda-visaya-bhavati sesani tu pancavisayani

Of these, the five organs of knowledge have as their objects
both the specific and the general.
Speech has speech as its object;
the rest have all the five as its object.

YS III.15

    kramanyatvam parinamanyatve hetu

    Teleology

SK 35.

yasmat buddhi santahkarana sarvam visayam avagahate
tasmat trividham karanam dvari sesani dvarani

Since buddhi along with the other internal instruments
comprehends all of the objects these three instruments are
like the warders while the rest are like the gates.

UNIVERSAL LOGICAL PERFECTIONS OF COGNITION

YS III.16

    parinama-traya-samyamad atitanagata-jnanam

    Manner and method

    All cognition, and a whole of cognition, must be in conformity with a rule.
    (Absence of rules is at the same time unreason.)
    But this rule is either that of manner (free) or that of method (compulsion).

    Form of science — Method

    Cognition, as science, must be arranged in accordance with a method.
    For science is a whole of cognition as a system, and not merely as an aggregate.
    It therefore requires a systematic cognition, hence one composed
    in accordance with rules on which we have reflected.

    Doctrine of method — Its object and end

    As the doctrine of elements in logic has for its content
    the elements and conditions of the perfection of a cognition,
    so the universal doctrine of method, as the other part of logic,
    has to deal with the form of a science in general,
    or with the ways of acting so as to connect
    the manifold of cognition in a science.

    Means for furthering the logical perfection of cognition

    The doctrine of method is supposed to expound the way
    for us to attain the perfection of cognition.
    Now one of the most essential logical perfections of cognition consists in its
    distinctness, thoroughness, and systematic ordering into the whole of a science.
    Accordingly, the doctrine of method will have principally to provide
    the means through which these perfections of cognition are furthered.

    Conditions of the distinctness of cognition

    The distinctness of cognitions and their combination
    in a systematic whole depends on the distinctness of concepts
    both in regard to what is contained in them
    and in respect of what is contained under them.
    The distinct consciousness of the content of concepts is furthered
    by exposition and definition of them,
    while the distinct consciousness of their extension, on the other hand,
    is furthered through logical division of them.

    First of all, then, of the means for furthering the distinctness of concepts
    in regard to their content.

    I. FURTHERING LOGICAL PERFECTION OF COGNITION THROUGH
    DEFINITION, EXPOSITION, AND DESCRIPTION OF CONCEPTS

    Definition

    A definition is a sufficiently distinct and precise concept
    (conceptus rei adaequatus in minimis terminis, complete determinatus).

    Analytic and synthetic definition

    All definitions are either analytic or synthetic. The former are definitions
    of a concept that is given, the latter of one that is made.
    A concept adequate to the thing, in minimal terms, completely determined.

    Concepts that are given and made a priori and a posteriori

    The given concepts of an analytic definition are given either a priori or a
    posteriori, just as the concepts of a synthetic definition, which are made,
    are made either a priori or a posteriori.

    Synthetic definitions through exposition or construction

    The synthesis of concepts that are made,
    out of which synthetic definitions arise,
    is either that of expositions (of appearances) or that of construction.
    The latter is the synthesis of concepts that are made arbitrarily,
    the former the synthesis of concepts that are made empirically,
    from given appearances as their matter
    (conceptus factitii vel a priori vel per synthesin empiricam).
    Concepts that are made arbitrarily are the mathematical ones.

    Impossibility of empirically synthetic definitions

    Since the synthesis of empirical concepts is not arbitrary but
    rather is empirical and as such can never be complete
    (because one can always discover in experience more marks of the concept),
    empirical concepts cannot be defined, either.

    Analytic definitions through analysis of concepts given a priori or a posteriori

    All given concepts, be they given a priori or a posteriori,
    can be defined only through analysis.
    For one can make given concepts distinct only insofar as
    one successively makes their marks clear.
    If all the marks of a given
    concept are made clear, then the concept becomes completely distinct; if it
    does not contain too many marks, then it is at the same time precise, and
    from this there arises a definition of the concept.

    Expositions and descriptions

    Not all concepts can be defined, and not all need to be.
    There are approximations to the definition of certain concepts;
    these are partly expositions (expositiones), partly descriptions (descriptions).
    The expounding of a concept consists in the connected (successive)
    representation of its marks, insofar as these are found through analysis.
    Description is the exposition of a concept, insofar as it is not precise.
    Description can occur only with empirically given concepts.
    It has no determinate rules and contains only the materials for definition.

    Nominal and real definitions

    By mere definitions of names, or nominal definitions, are to be understood
    those that contain the meaning that one wanted arbitrarily to give to a
    certain name, and which therefore signify only the logical essence of their
    object, or which serve merely for distinguishing it from other objects.
    Definitions of things, or real definitions, on the other hand, are ones that
    suffice for cognition of the object according to its inner determinations,
    since they present the possibility of the object from inner marks.

    Principal requirements of definition

    The essential and universal requirements that pertain to the completeness
    of a definition in general may be considered under the four principal
    moments of quantity, quality, relation, and modality:
    as to quantity - what concerns the sphere of the definition - the definition
    and the definitum must be convertible concepts" (conceptus redprocf), and hence
    the definition must be neither broader nor narrower than its definitum;
    as to quality, the definition must be
    a detailed and at the same time precise concept,
    as to relation, it must not be tautological,
    the marks of the definitum must, as grounds of its cognition,
    be different from it itself, and finally
    as to modality, the marks must be necessary, and
    hence not such as are added through experience.

    Rules for testing definitions

    In the testing of definitions four acts are to be performed;
    it is to be investigated, namely,
    1. whether the definition considered as a proposition is true,
    2. whether as a concept it is distinct,
    3. whether as a distinct concept it is also detailed, and finally
    4. whether as a detailed concept it is at the same time determinate,
       adequate to the thing itself.

    Rules for preparation of definitions

    The very same acts that belong to the testing of definition are
    also to be performed in the preparation of them.
    Toward this end, then, seek
    (1.) true propositions,
    (2.) whose predicate does not presuppose the concept of the thing;
    (3.) collect several of them and compare them with the concept of
    the thing itself to see if they are adequate; and finally
    (4.) see whether one mark does not lie in another or is not subordinated to it.

    II. FURTHERING THE PERFECTION OF COGNITION
    THROUGH LOGICAL DIVISION OF CONCEPTS

    Concept of logical division

    Every concept contains a manifold under itself insofar as the manifold agrees,
    but also insofar as it is different.
    The determination of a concept in regard to everything possible
    that is contained under it, insofar as things are opposed to one another,
    are distinct from one another, is called the logical division of the concept.
    The higher concept is called the divided concept (divisus),
    the lower concepts the members of the division (membra dividentia).

    Universal rules of logical division

    In every division of a concept we must see to it:
    1. that the members of the division exclude or are opposed to one another,
    that furthermore they belong under one higher concept (conceptus communis),
    and finally that taken together they constitute the sphere of
    the divided concept or are equal to it.

    Codivision and subdivision

    Various divisions of a concept, which are made in various respects, are
    called codivisions, and division of the members of division is called a
    subdivision (mbdivisio).

    Dichotomy and polytomy

    A division into two members is called dichotomy; but if it has more than two
    members, it is called polytomy.

    Various divisions of method

    Now as for what concerns in particular method itself in working up and
    treating scientific cognitions, there are various principal kinds of it,
    which we can present in accordance with the following division.

    1. Scientific or popular method

    Scientific or scholastic method differs from popular method through the
    fact that the former proceeds from basic and elementary propositions,
    but the latter from the customary and the interesting.
    The former aims for thoroughness and thus removes everything foreign,
    the latter aims at entertainment.

    2. Systematic or fragmentary method

    Systematic method is opposed to fragmentary or rhapsodic method.
    If one has thought in accordance with a method and then also expressed this
    method in the exposition, and if the transition from one proposition to
    another is distinctly presented, then one has treated a cognition systematically.
    If, on the other hand, one has thought according to a method but has
    not arranged the exposition methodically, such a method is to be called rhapsodic.

    3. Analytic or synthetic method

    Analytic is opposed to synthetic method.
    The former begins with the conditioned and grounded and
    proceeds to principles (a prindpiatis ad prindpia),
    while the latter goes from principles to consequences or
    from the simple to the composite.
    The former could also be called regressive,
    as the latter could progressive.

    4. Syllogistic [or] tabular method

    Syllogistic method is that according to which
    a science is expounded in a chain of inferences.
    That method in accordance with which a finished system is exhibited in
    its complete connection is called tabular.
    Method is acroamatic insofar as someone only teaches,
    erotematic insofar as he asks as well.
    The latter method can be divided in turn into dialogic or Socratic method and
    catechistic method, accordingly as the questions are
    directed either to the understanding or merely to memory.

    Meditation

    By meditation is to be understood reflection, or methodical thought.
    Meditation must accompany all reading and learning, and for this
    it is requisite that one first undertake provisional investigations and then
    put his thoughts in order, or connect them in accordance with a method.

SK 36.

ete manas ahamkara pradipa-kalpa paraspara vilaksana
guna-visesa krstnam prakasya purushasya-artham buddhau prayacchati

These (external instruments along with manas and ahamkara)
which are characteristic-wise different from one another
and are different modifications of the attributes
and which resemble a lamp (in action) illuminating all (their respective objects)
present them to the Buddhi for the purpose of the Spirit,

Cognition in general
Intuitive and discursive cognition;
intuition and concept and in particular their difference
Logical and aesthetic perfection of cognition

All our cognition has a twofold relation,
first, a relation to the object,
second a relation to the subject.
In the former respect it is related to representation,
in the latter to consciousness,
the universal condition of all cognition in general.
(Consciousness is really a representation that
another representation is in me.)

In every cognition we must distinguish
matter as the object, and
form as the way in which we cognize the object.

    If a savage sees a house from a distance, for example,
    with whose use he is not acquainted, he admittedly
    has before him in his representation the very same object as
    someone else who is acquainted with it determinately
    as a dwelling established for men.
    But as to form, this cognition of one and the same object is different in the two.
    With the one it is mere intuition,
    with the other it is intuition and concept at the same time.

The difference in the form of the cognition rests on
a condition that accompanies all cognition, on consciousness.
If I am conscious of the representation, it is clear,
if I am not conscious of it, obscure.

    Since consciousness is the essential condition of all logical form of cognitions,
    logic can and may occupy itself only with clear but not with obscure representations.
    In logic we do not see how representations arise,
    but merely how they agree with logical form.
    In general logic cannot deal at all with mere representations and
    their possibility either. This it leaves to metaphysics.
    Logic itself is occupied merely with
    the rules of thought in concepts, judgments, and inferences,
    as that through which all thought takes place.
    Something precedes, of course, before a representation becomes a concept.
    We will indicate that in its place, too.
    But we will not investigate how representations arise.
    Logic deals with cognition too, to be
    sure, because in cognition there is already thought.
    But representation is not yet cognition, rather,
    cognition always presupposes representation.
    And this latter cannot be explained at all.
    For we would always have to explain what representation is
    by means of yet another representation.

All clear representations, to which alone logical rules can be applied,
can now be distinguished in regard to distinctness and indistinctness.
If we are conscious of the whole representation,
but not of the manifold that is contained in it,
then the representation is indistinct.
First, to elucidate this, an example in intuition.

    We glimpse a country house in the distance.
    If we are conscious that the intuited object is a house,
    then we must necessarily have a representation of
    the various parts of this house, the windows, doors, etc.
    For if we did not see the parts, we would not see the house itself either.
    But we are not conscious of this representation of the manifold of its parts,
    and our representation of the object indicated is thus itself an indistinct representation.

If we want an example of indistinctness in concepts, furthermore, then
the concept of beauty may serve. Everyone has a clear concept of beauty.
But in this concept many different marks occur, among others that the
beautiful must be something that (i.) strikes the senses and (2.) pleases
universally. Now if we cannot explicate the manifold of these and other
marks of the beautiful, then our concept of it is still indistinct.

    Wolff's disciples call the indistinct representation a confused one. But
    this expression is not fitting, because the opposite of confusion is not
    distinctness but order. Distinctness is an effect of order, to be sure, and
    indistinctness an effect of confusion; and every confused cognition is thus
    also an indistinct one. But the proposition does not hold conversely; not
    every indistinct cognition is a confused one. For in the case of cognitions
    in which there is no manifold at hand, there is no order, but also no
    confusion.

This is the situation with all simple representations, which never become distinct,
not because there is confusion in them,
but rather because there is no manifold to be found in them.
One must call them indistinct, therefore, but not confused.

And even with compound representations, too, in which
a manifold of marks can be distinguished,
indistinctness often derives not from confusion
but from weakness of consciousness.
Thus something can be distinct as to form,
I can be conscious of the manifold in the representation,
but the distinctness can diminish as to matter if
the degree of consciousness becomes smaller,
although all the order is there.
This is the case with abstract representations.

Distinctness itself can be of two sorts:

First, sensible.
This consists in the consciousness of the manifold in intuition.

I see the Milky Way as a whitish streak, for example;
the light rays from the individual stars located in it
must necessarily have entered my eye.
But the representation of this was merely clear,
and it becomes distinct only through the telescope,
because then I glimpse the individual stars contained in the Milky Way.

Secondly, intellectual;
distinctness in concepts or distinctness of the understanding.
This rests on the analysis of the concept in regard to
the manifold that lies contained within it.

Thus in the concept of virtue, for example, are contained as marks
(1.) the concept of freedom,
(2.) the concept of adherence to rules (to duty),
(3.) the concept of overpowering the force of the inclinations,
in case they oppose those rules.
Now if we break up the concept of virtue into its individual constituent parts,
we make it distinct for ourselves through this analysis.
By thus making it distinct, however, we add nothing to a concept; we only explain it.
With distinctness, therefore, concepts are improved not as to matter but only as to form.

If we reflect on our cognitions in regard to the two essentially different
basic faculties, sensibility and the understanding, from which they arise,
then here we come upon the distinction between intuitions and concepts.
Considered in this respect, all our cognitions are, namely, either intuitions
or concepts. The former have their source in sensibility, the faculty of
intuitions, the latter in the understanding, the faculty of concepts. This is
the logical distinction between understanding and sensibility, according to
which the latter provides nothing but intuitions, the former on the other
hand nothing but concepts. The two basic faculties may of course be
considered from another side and defined in another way: sensibility,
namely, as a faculty of receptivity, the understanding as a faculty of spontane-
ity. But this mode of explanation is not logical but rather metaphysical. It is
also customary to call sensibility the lower faculty, the understanding on
the other hand the higher faculty, on the ground that sensibility gives the
mere material for thought, but the understanding rules over this material
and brings it under rules or concepts.
The difference between aesthetic and logical perfection of cognition is
grounded on the distinction stated here between intuitive and discursive
cognitions, or between intuitions and concepts.

A cognition can be perfect either
according to laws of sensibility or
according to laws of the understanding;
in the former case it is aesthetically perfect,
in the other logically perfect.
The two, aesthetic and logical perfection, are thus of different kinds;
the former relates to sensibility, the latter to the understanding.
The logical perfection of cognition rests on its agreement with the object,
hence on universally valid laws, and hence we
can pass judgment on it according to norms a priori.
Aesthetic perfection
consists in the agreement of cognition with the subject and is grounded on
the particular sensibility of man.
In the case of aesthetic perfection, therefore,
there are no objectively and universally valid laws, in relation to
which we can pass judgment on it a priori in a way that is universally valid
for all thinking beings in general.
Insofar as there are nonetheless universal laws of sensibility,
which have validity subjectively for the whole of humanity
although not objectively and for all thinking beings in general,
we can think of an aesthetic perfection that contains the ground of a
subjectively universal pleasure.
This is beauty, that which pleases the senses in intuition and
can be the object of a universal pleasure just because
the laws of intuition are universal laws of sensibility.
Through this agreement with the universal laws of sensibility
the really, independently beautiful,
whose essence consists in mere form,
is distinguished in kind from the pleasant,
which pleases merely in sensation
through stimulation or excitement,
and which on this account can only be
the ground of a merely private pleasure.
It is this essential aesthetic perfection, too,
which, among all [perfections], is compatible with
logical perfection and may best be combined with it.

Considered from this side, aesthetic perfection in regard to the essentially
beautiful can thus be advantageous to logical perfection. In another
respect it is also disadvantageous, however, insofar as we look, in the case
of aesthetic perfection, only to the non-essentially beautiful, the stimulating
or the exciting, which pleases the senses in mere sensation and does not
relate to mere form but rather to the matter of sensibility. For stimulation
and excitement, most of all, can spoil the logical perfection in our
cognitions and judgments.

In general, however, there always remains a kind of conflict between the
aesthetic and the logical perfection of our cognition, which cannot be fully
removed. The understanding wants to be instructed, sensibility enlivened;
the first desires insight, the second comprehensibility. If cognitions are to
instruct then they must to that extent be thorough; if they are to entertain
at the same time, then they have to be beautiful as well. If an exposition is
beautiful but shallow, then it can only please sensibility but not the understanding,
but if it is thorough yet dry, only the understanding but not
sensibility as well.

Since the needs of human nature and the end of popularity in cognition
demand, however, that we seek to unite the two perfections with one
another, we must make it our task to provide aesthetic perfection for those
cognitions that are in general capable of it, and to make a scholastically
correct, logically perfect cognition popular through its aesthetic form.

But in this effort to combine aesthetic with logical perfection
in our cognitions we must not fail to attend to the following rules, namely:
(1.) that logical perfection is the basis of all other perfections and
hence cannot be wholly subordinated or sacrificed to any other;
(2.) that one should look principally to formal aesthetic perfection,
the agreement of a cognition with the laws of intuition,
because it is just in this that the essentially beautiful,
which may best be combined with logical perfection, consists;
(3.) that one must be very cautious with stimulation and excitement,
whereby a cognition affects sensation and acquires an interest for it,
because attention can thereby so easily be drawn from the object to the subject,
whence a very disadvantageous influence on
the logical perfection of cognition must evidently arise.

To acquaint us better with the essential differences that exist between
the logical and the aesthetic perfection of cognition,
not merely in the universal but from several particular sides,
we want to compare the two with one another in respect to
the four chief moments of quantity, quality, relation, and modality,
on which the passing of judgment as to the perfection of cognition depends.

A cognition is perfect
(1.) as to quantity if it is universal;
(2.) as to quality if it is distinct;
(3.) as to relation if it is true; and finally
(4.) as to modality if it is certain.

    Considered from the viewpoints indicated,
    a cognition will thus be logically perfect
    as to quantity  if it has objective universality
    (universality of the concept or of the rule),
    as to quality if it has objective distinctness
    (distinctness in the concept),
    as to relation if it has objective truth, and finally
    as to modality if it has objective certainty.

To these logical perfections correspond now the following aesthetic
perfections in relation to those four principal moments, namely

aesthetic universality.
    This consists in the applicability of a cognition to a
    multitude of objects that serve as examples,
    to which application of it can be made,
    and whereby it becomes useful at the same time for the end of popularity;
aesthetic distinctness.
    This is distinctness in intuition, in which a concept
    thought abstractly is exhibited or elucidated in concrete through examples;
aesthetic truth.
    A merely subjective truth, which consists only in the agreement
    of cognition with the subject and the laws of sensory illusion,
    and which is consequently nothing more than a universal semblance.
aesthetic certainty.
    This rests on what is necessary in consequence of
    the testimony of the senses;
    what is confirmed through sensation and experience.

With the perfections just mentioned two things are always to be found,
which in their harmonious union make up perfection in general,
namely, manifoldness and unity.
Unity in the concept lies with the understanding,
unity of intuition with the senses.
Mere manifoldness without unity cannot satisfy us.
And thus truth is the principal perfection among them all,
because it is the ground of unity through
the relation of our cognition to the object.
Even in the case of aesthetic perfection,
truth always remains the conditio sine qua non,
the foremost negative condition,
apart from which something cannot please taste universally.
Hence no one may hope to make progress in the belles lettres
if he has not made logical perfection the ground of his cognition.
It is in the greatest possible unification of
logical with aesthetic perfection in general,
in respect to those cognitions that are both
to instruct and to entertain,
that the character and the art of
the genius actually shows itself.

PARTICULAR LOGICAL PERFECTIONS OF COGNITION

YS III.17

    sabda-artha-pratyayanam itaretara-adhyasat sankaras tat-pravibhaga-samyamat sarva-bhuta-ruta-jnanam

YS III.18

    samskara-saksat-karanat purva-jati-jnanam

YS III.19

    pratyayasya para-citta-jnanam

YS III.20

    na ca tat salambanam tasya-avisayi-bhutatvat

SK 37.

yasmat buddhi purusasya upabho-gam sarvam pratisadhyati
sa eva ca suksma pradhana-purusantaram visinasti

Because it is the Buddhi that accomplishes the experiences
with regard to all objects to the Purusha.
It is that again that discriminates the subtle difference
between the Pradhana and the Purusha.

YS III.21

    kaya-rupa-samyamat tad-grahya-sakti-stambhe caksu-prakasa-asamprayoge ‘ntardhanam

YS III.22

    etena sabda-adi-antardhanam uktam
