YS II.44

    svadhyayad ista-devata-samprayoga

YS II.45

    samadhi-siddhi îsvara-pranidhana

SK 22.

Prakriti mahat tata ahamkara tasmat ca sodasaka gana
sodasakat tasmad api pancabhyah panca bhutani

From nature evolves the Great Principle;
from this evolves the I Principle;
from this evolves the set of sixteen;
from the five of this set of sixteen evolves the five elements.

SK 23.

buddhi adhyavasaya dharma jnana viraga aisvarya
etad rupam sattvikam tasmat viparyastam tamasam

Buddhi is self-certainty.
Virtue, knowledge, dispassion, and power are its manifestations
when the sattva attribute abounds.
And the reverse of these, when the tamasa attribute abounds.

SK 24.

ahamkara abhimana tasmat pravartate dvividha sarga eva
ekadasaka gana ca tanmatra pancaka

Ahamkara is self-assertion;
from that proceeds a two-fold evolution
the set of eleven and the five-fold primary elements.

YS II.46

    sthira-sukham asanam

YS II.47

    prayatna-saithilya-ananta-samapattibhyam

YS II.48

    tato dvandva-anabhighata

SK 25.

vaikrita ahamkara ekadasaka sattvika pravartate
tanmatra bhutade sa tamasa taijasa ubhayam

The set of eleven sattvic indriyas proceeds from the Vaikriti form of I-Principle;
the set of five tamasic tanmatras proceed from the Bhutadi form of I-Principle.
From the Taijasa form of I-Principle proceed both of them.

SK 26.

buddhi indriyani akhyani caksu srotram ghrana rasana tvak
karma indriyani ahu vak pani pada payu upastha

Organs of knowledge are called the Eye, the Ear, the Nose, the Tongue, the Skin.
Organs of action are called the Speech, the Hand, the Feet, the excretory organ and the organ of generation.

YS II.49

    tasmin sati svasa-prasvasayo gati-viccheda pranayama

YS II.50

    bahya-abhyantara-stambha-vritti desa-kala-sankhyabhi-paridrsta dirgha-suksma

YS II.51

    bahya-abhyantara-visaya-aksepi caturtha

SK 27.

atra mana ubhayatmakam sankalpakam ca sadharmaya indriyam
nanatvam bahya bheda ca guna-parinama-visesa

Of these (sense organs), the Mind possesses the nature of both.
It is the deliberating principle, and is also called a sense organ
since it posseses properties common to the sense organs.
Its multifariousness and also its external diversities are owing
to special modifications of the Qualities.

SK 28.

pancanam rupadisu alocanamatram isyate
pancanam vritti vacana adana viharana utsarga ca ananda

The function of the five in respect to form and the rest,
is considered to be mere observation.
Speech, manipulation, locomotion, excretion and gratification
are the functions of the other five.

YS II.52

    tata ksiyate prakasa-avaranam

YS II.53

    dharanasu ca yogyata manasa

YS II.54

    sva-visaya-asamprayoge cittasya svarupa-anukara ive indriyanam pratyahara

YS II.55

    tata parama vasyata indriyanam

YS III.1

    desa-bandha cittasya dharana

YS III.2

    tatra pratyaya-eka-tanata dhyanam

YS III.3

    tad evartha-matra-nirbhasa svarupa-sunyam iva samadhi

YS III.4

    trayam ekatra samyama

YS III.5

    taj-jayat prajna-aloka

YS III.6

    tasya bhumisu viniyoga

YS III.7

    trayam antar-angam purvebhya

YS III.8

    tad api bahir-angam nirbijasya

SK 29.

trayasya svalaksanyam vritti sa esa asamanya bhavati
samanya karana-vritti pranadhyah-vayavah-panca

Of the three internal instruments, their own characteristics are their functions;
this is peculiar to each.
The common modification of the instruments is the five airs
such as prana and the rest.

YS III.9

    vyutthana-nirodha-samskarayor abhibhava-pradur-bhavau nirodha-kshana-cittanvayo nirodha-parinama

YS III.10

    tasya prasanta-vahita samskarat

    Manas, the understanding, first the faculty for the cognition of the general (of rules)

    a. The Concept as such

    An idea contains the archetype for the use of the understanding, e.g.,
    the idea of the world whole, which idea must necessarily be,
    not as constitutive principle for the empirical use of the understanding,
    but as regulative principle for the sake of the thoroughgoing connection of
    our empirical use of the understanding.
    Thus it is to be regarded as a necessary basic concept,
    either for objectively completing the understanding's actions of
    subordination or for regarding them as unlimited.
    The idea cannot be attained by composition, either,
    for the whole is prior to the part.
    There are ideas, however, to which an approximation occurs.
    This is the case with mathematical ideas,
    or ideas of the mathematical production of a whole,
    which differ essentially from dynamical ideas,
    which are completely heterogeneous from all concrete concepts,
    because the whole is different from concrete concepts
    not as to quantity (as with the mathematical ideas)
    but rather as to kind.

    One cannot provide objective reality for any theoretical idea, or prove it,
    except for the idea of freedom, because this is the condition of the moral law,
    whose reality is an axiom.
    The reality of the idea of God can only be proved by means of this idea, and
    hence only with a practical purpose;
    to act as if there is a God, and hence only for this purpose.
    In all sciences, above all in those of reason, the idea of the science is
    its universal abstract or outline, hence the extension of all the cognitions
    that belong to it.
    Such an idea of the whole, the first thing one has to look to in a science, and
    which one has to seek is architectonic, as, e.g., the idea of jurisprudence.
    Most men lack the idea of humanity, the idea of a perfect republic,
    of a happy life, etc.
    Many men have no idea of what they want,
    hence they proceed according to instinct and authority.

SK 30.

drste catustayasya tu yugapat vrttih tasya kramasasca nirdista
tatha api adrste trayasya vrttih tat purvika

Of all the four, the functions are said to be simultaneous and also successive
with regard to the seen objects;
with regard to the unseen objects
the functions of the three are preceeded by that.

YS III.11

    sarva-arthata-ekagratayo kshayodayau cittasya samadhi-parinama

    The logical form of all judgments consists of the objective unity of the apperception of the concepts contained therein

    Ahamkara, the determinative power of judgment, second the faculty for the subsumption of the particular under the general

    b. Judgment

SK 31.

svam svam vrittim pratipadyante paraspara akuta hetukam
purushartha eva hetu na kenacit karanam karyate

Instruments enter into their respective functions being incited by mutual impulse.
The purpose of the Spirit is the sole motive (for the activities of the instruments).
By none whatsoever is an instrument made to act.

YS III.12

    tata punashantoditau tulya-pratyayau cittasya-ekagrata-parinama

    c. Syllogism

    Buddhi, reason, third the faculty for the determination of the particular from the general (for the derivation from principles)

SK 32.

karanam trayodasavidham tad aharana dharana prakasakaram
tasya karyam ca dasadha aharyam dharyam prakasyam ca

Instruments are of thirteen kinds performing the functions of
seizing, sustaining and illuminating.
Its objects are of ten kinds, vis
the seized, the sustained, and the illumined.

YS III.13

    etena bhutendriyesu dharma-laksana-vastha-parinama vyakhyata

    a. Mechanism

SK 33.

antahkaranam trividham bahyam dasadha trayasya visayakhyam
bahyam sampratakalam abhyantaram karanam trikalam

The internal instruments are three-fold.
The external are ten-fold; they are called the objects of the three (internal instruments).
The external instruments function at the present time and
the internal instruments function at all the three times.

YS III.14

    shantoditavyapadesya-dharmanupati dharmi

    b. Chemism

SK 34.

Tesam panca buddhi visesa-avisesa-visaya
Vak sabda-visaya-bhavati sesani tu pancavisayani

Of these, the five organs of knowledge have as their objects
both the specific and the general.
Speech has speech as its object;
the rest have all the five as its object.

YS III.15

    kramanyatvam parinamanyatve hetu

    c. Teleology

SK 35.

yasmat buddhi santahkarana sarvam visayam avagahate
tasmat trividham karanam dvari sesani dvarani

Since buddhi along with the other internal instruments
comprehends all of the objects these three instruments are
like the warders while the rest are like the gates.

YS III.16

    parinama-traya-samyamad atitanagata-jnanam

SK 36.

ete manas ahamkara pradipa-kalpa paraspara vilaksana
guna-visesa krstnam prakasya purushasya-artham buddhau prayacchati

These (external instruments along with manas and ahamkara)
which are characteristic-wise different from one another
and are different modifications of the attributes
and which resemble a lamp (in action) illuminating all (their respective objects)
present them to the Buddhi for the purpose of the Spirit,

PARTICULAR LOGICAL PERFECTIONS OF COGNITION

YS III.17

    sabda-artha-pratyayanam itaretara-adhyasat sankaras tat-pravibhaga-samyamat sarva-bhuta-ruta-jnanam

YS III.18

    samskara-saksat-karanat purva-jati-jnanam

YS III.19

    pratyayasya para-citta-jnanam

YS III.20

    na ca tat salambanam tasya-avisayi-bhutatvat

A) Logical perfection of cognition as to quantity

Quantity

    Extensive and intensive quantity
    Extensiveness and thoroughness or importance and fruitfulness of cognition
    Determination of the horizon of our cognition

The quantity of cognition can be understood in two senses,
either as extensive or as intensive quantity.
The former relates to the extension of cognition and
thus consists in its multitude and manifoldness;
the latter relates to its content, which concerns the richness
or the logical importance and fruitfulness of a cognition,
insofar as it is considered as ground of many and great consequences (multa sed multum.)
In expanding our cognitions or in perfecting them as to their extensive quantity
it is good to make an estimate as to how far a cognition agrees with our ends and capabilities.
This reflection concerns the determination of the horizon of our cognitions,
by which is to be understood the congruence of the quantity of all cognitions
with the capabilities and ends of the subject.

The horizon can be determined

1. logically, in accordance with the faculty or the powers of cognition
in relation to the interest of the understanding.

Here we have to pass judgment on how far we can go in our cognitions,
how far we must go, and to what extent certain cognitions serve,
in a logical respect, as means to various principal cognitions as our ends;

2. aesthetically, in accordance with taste in relation to the interest of feeling.
He who determines his horizon aesthetically seeks to arrange science according
to the taste of the public, i.e., to make it popular, or in general to attain only
such cognitions as may be universally communicated, and in which the class
of the unlearned, too, find pleasure and interest;

3. practically, in accordance with use in relation to the interest of the will.
The practical horizon, insofar as it is determined according to the influence
which a cognition has on our morality is pragmatic and is of the greatest importance.

Thus the horizon concerns passing judgment on, and determining,
what man can know, what he is permitted to know, and what he ought to know.
Now as for what concerns the theoretically or logically determined horizon
in particular - and it is of this alone that we can speak here - we can
consider it either from the objective or from the subjective viewpoint.

In regard to objects, the horizon is either historical or rational.
The former is much broader than the other, indeed, it is immeasurably great,
for our historical cognition has no limits.
The rational horizon, on the other hand, may be fixed,
e.g., it may be determined to what kind of objects
mathematical cognition cannot be extended.
So too in respect of philosophical cognition of reason,
as to how far reason can go here a priori without any experience.

In relation to the subject the horizon is either the universal and absolute,
or a particular and conditioned one (a private horizon).

By the absolute and universal horizon is to be understood
the congruence of the limits of human cognitions with
the limits of the whole of human perfection in general.

And here, then, the question is:

In general, what can man, as man, know?

The determination of the private horizon depends upon various
empirical conditions and special considerations,
e.g., age, sex, station, mode of life, etc.
Every particular class of men has its particular horizon in relation
to its special powers of cognition, ends, and standpoints, every mind its
own horizon according to the standard of the individuality of its powers
and its standpoint. Finally, we can also think a horizon of healthy reason
and a horizon of science, which latter still requires principles, in accordance
with which to determine what we can and cannot know.
What we cannot know is beyond f our horizon, what we do not need to
know is outside our horizon. This latter can hold only relatively, however,
in relation to various particular private ends, to whose accomplishment
certain cognitions not only do not contribute anything but could even be
an obstacle. For no cognition is, absolutely and for every purpose, useless
and unusable, although we may not always be able to have insight into its
use. Hence it is an objection as unwise as it is unjust that is made to great
men who labor in the sciences with painstaking industry when shallow
minds ask, What is the use ofthat?' We must simply never raise this ques-
tion if we want to occupy ourselves with the sciences. Even granted that a
science could give results only concerning some possible object, it would
still for that reason alone be useful enough. Every logically perfect cogni-
tion always has some possible use, which, although we are as yet unac-
quainted with it, will perhaps be found by posterity. If in the cultivation of
the sciences one had always looked only to material gain, their use, then
we would have no arithmetic or geometry. Besides, our understanding is
so arranged that it finds satisfaction in mere insight, even more than in the
use that arises therefrom. Plato noted this. Man feels in this his own
excellence, he senses what it means to have understanding. Men who do
not sense this must envy the animals. The inner worth that cognitions have
through logical perfection is not to be compared with the outer, their worth
in application.

Like that which lies outside our horizon, insofar as we, in accordance
with our purposes, do not need to know it, as dispensable for us, that which
lies beneath' our horizon, insofar as we ought not to know it as harmful to
us, is to be understood in a relative sense but never in an absolute one.

With respect to the extension and the demarcation of our cognition,
the following rules are to be recommended:

1. One must determine his horizon early,
but of course only when one can determine it oneself,
which usually does not occur before the 2oth year;

2. not alter it lightly or often
(not turn from one thing to another);

3. not measure the horizon of others by one's own,
and not consider as useless what is of no use to us;
it would be presumptuous to want to determine others' horizons,
because one is not sufficiently acquainted,
in part with their capabilities,
in part with their purposes;

4. neither extend it too far nor restrict it too much.
For he who wants to know too much ends by knowing nothing,
and conversely, he who believes of some things
that they do not concern him, often deceives himself;
as when, e.g., the philosopher believes of history
that it is dispensable for him.

One should also seek

5. to determine in advance the absolute horizon of
the whole human race (as to past and to future time),
as well as also

6. to determine, in particular, the position that
our science occupies in the whole of cognition.
The Universal Encyclopedia serves for this as
a universal map (mappe-monde) of the sciences.

7. In determining his own particular horizon one should carefully consider for
which part of cognition one has the greatest capability and pleasure, what is
more or less necessary in regard to certain duties, what cannot coexist with
the necessary duties; and finally

8. one should of course always seek
to expand his horizon rather than to narrow it.

As for the extension of cognition, there need be no concern in general
about what concerned d'Alembert.
For the burden does not press us down, but rather the volume of space
for our cognitions constrains us.
Critique of reason, of history and historical writings,
a universal spirit that deals with human cognition en gros and
not merely in detail, will always make the extension smaller,
without diminishing anything in the content.
The metal merely separates from the slag, or the inferior vehicle,
the husk, which was necessary for so long.
With the extension of natural history, of mathematics, etc.,
new methods will be invented which will shorten the old and
make the multitude of books dispensable.
It will be because of the invention of such new methods and principles
that we will be able, with their help, to find anything we desire
without burdening memory.
Thus he who brings history under ideas that can always remain
renders it service as a genius.

Opposed to the logical perfection of cognition
in regard to its extension stands ignorance.

A negative imperfection, or imperfection of lack, which,
on account of the restrictions of the understanding,
is inseparable from our cognition.

We can consider ignorance from an objective or from a subjective viewpoint.

1. Taken objectively, ignorance is either material or formal.
The former consists in a lack of historical cognitions,
the other in a lack of rational cognitions.
One does not have to be completely ignorant in any field,
but one can well restrict historical knowledge in order to
devote oneself more to rational knowledge, or conversely.

2. In the subjective sense, ignorance is either learned, scientific, or is common.
He who has distinct insight into the restrictions of cognition,
hence into the field of ignorance from where it begins,
e.g., the philosopher who sees and proves how little one can know of
gold in regard to its structure due to a lack of the requisite data,
is ignorant artfully or in a learned way.
He who is ignorant, on the other hand, without
having insight into the grounds of the limits of knowledge,
and without concerning himself with this, is so in a common, not a scientific way.
Such a one does not even know that he knows nothing.
For one can never represent his ignorance except through science,
as a blind man cannot represent darkness until he has become sighted.

Cognition of one's ignorance presupposes science, then, and
makes one at the same time modest, while imagined knowledge puffs one up.
Hence Socrates' non-knowledge was a laudable ignorance,
a knowledge of non-knowledge, according to his own admission.
It is precisely those who possess very many cognitions then, and
who for all that are astounded at the multitude of what they do not know,
who cannot be reproached with their ignorance.
Ignorance in things whose cognition lies beyond our horizon is in
general irreproachable (inculpabilis), and in regard to the speculative use of
our faculty of cognition it can be allowed (although only in the relative
sense), insofar as the objects here lie not beyond our horizon but yet outside
it. It is shameful, however, in things that it is quite necessary and also easy
to know.

There is a distinction between not knowing something and ignoring
something, i.e., taking no notice of it. It is good to ignore much that it is not
good for us to know. Abstracting is distinct from both of these. One ab-
stracts from a cognition when one ignores its application, whereby one
gets it in abstracto and can better consider it in the universal as a principle.
Such abstraction from what does not belong to our purpose in the cogni-
tion of a thing is useful and praiseworthy.

Scholars in matters of reason are commonly ignorant historically.
Historical knowledge without determinate limits is poly history; this puffs
one up. Polymathy has to do with cognition of reason. Both historical and
rational knowledge, when extended without determinate limits, can be
called pansophy. Historical knowledge includes the science of die tools of
learnedness - philology, which comprises a critical acquaintance with
books and languages (literature and linguistics).
Mere polyhistory is cyclopic learnedness, which lacks one eye, the eye of
philosophy, and a cyclops among mathematicians, historians, natural histo-
rians, philologists, and linguists is a learned man who is great in all these
matters, but who for all that holds all philosophy to be dispensable.
One part of philology is constituted by the humaniora, by which is
understood acquaintance with the ancients, which furthers die unification
of science with taste, which rubs off coarseness and furthers die communica-
bility and urbanity in which humanity consists.
The humaniora, then, concern instruction in what serves the cultivation
of taste, in conformity with the models of the ancients. This includes, e.g.,
eloquence, poetry, wide reading in the classical authors, etc. All these
humanistic cognitions' can be reckoned in the practical part of philology,
which aims in the first instance at the cultivation of taste. If we separate
the mere philologist from the humanist, however, die two would differ
from one anotiier in dial die former seeks die tools of learnedness among
the ancients, the latter die tools for the cultivation of taste.
The belletrist, or bei esprit, 1 is a humanist according to contemporary
models in living languages. He is not learned, then, for only dead lan-
guages are now learned languages, but is radier a mere dilettante in
cognitions of taste' in accordance with fashion, with no need for the
ancients. We could call him one who apes the humanist. The polyhistor
must, as philologist, be a linguist and a literator, and as humanist a
classicist and expositor of the classics. As philologist he is cultivated, as
humanist civilized.

In regard to the sciences, there are
two degenerate forms of prevailing taste: pedantry and gallantry.
The one pursues the sciences only for the school and
thereby restricts them in respect of their use,
the other pursues them merely for intercourse or for the world and
in this way restricts them in respect of their content.

The pedant is either opposed, as learned man, to the man of the world,
and is to this extent the puffed up man of learning, unacquainted with the
world, i.e., with the ways of bringing his science to bear on men; or he is to
be considered, in general, as a man of skill, but only informalities, not as to
essence and as to his end. In the latter sense he is z fanatic for formalities?
restricted in regard to the core of things, he looks only to the clothing and
the shell. He is the unfortunate imitation or caricature of the methodical
mind. Thus pedantry can also be called cavilling fussiness 1 " and useless
exactitude (micrology) in formalities. And such formality of scholastic
method outside the schools is to be found not merely among learned
people and in learned things, but also in other classes and in other things.
What is the ceremonial at court and in intercourse but a pursuit of formalities
and hair-splitting. In the military this is not completely so, although it
seems so. But in conversation, in clothing, in diet, in religion, a good deal
of pedantry often prevails.
Thoroughness (scholarly, scholastic perfection 2 ) is a purposeful exacti-
tude in formalities. Pedantry is thus an affected thoroughness, just as
gallantry, as a mere courtesan seeking the approval of taste, is nothing but
an affected popularity. For gallantry only strives to gain the reader's affec-
tion and thus never to insult him with a hard word.
To avoid pedantry requires extensive cognitions" not only in the sci-
ences themselves but also in regard to their use. Only the true man of
learning, then, can free himself from pedantry, which is always the prop-
erty of a restricted mind.
In striving to procure for our cognition the perfection of scholastic
thoroughness and at the same time of popularity, without falling into the
indicated mistakes of affected thoroughness or of affected popularity, we
must look above all to the scholastic perfection of our cognition, the
scholastically correct form of thoroughness; and only then may we con-
cern ourselves about how we are to make our cognition, methodically
learned in school, truly popular, i.e., easily and universally communicable
to others, in such a way that thoroughness is not displaced by popularity.

For scholastic perfection, without which all science is nothing but tricks
and trifling, must not be sacrificed for the sake of popular perfection, to
please the people.
To learn true popularity, however, one must read the ancients, e.g.,
Cicero's philosophical writings, the poets Horace, Virgil, etc., and among
the moderns Hume, Shaftesbury, et. al. Men who have all had a good deal
of intercourse with the refined world, without which one cannot be popu-
lar. For true popularity demands a good deal of practical acquaintance
with the world and with men, acquaintance with men's concepts, taste,
and inclinations, to which constant regard must be given in presentation
and even in the choice of expressions that are fitting and adequate to
popularity. This ability to descend' to the public's power of comprehen-
sion and to the customary expressions, in which scholastic perfection is
not slighted, but in which the clothing of thoughts is merely so arranged
that the framework, the scholastically correct and technical in that perfection,
may not be seen (just as one draws lines with a pencil, writes on them, and
subsequently erases them) - this truly popular perfection of cognition is
in fact a great and rare perfection, which shows much insight into the
science. It has this merit, too, in addition to many others, that it can
provide a proof of complete insight into a thing. For the merely scholastic
examination of a cognition leaves doubt as to whether that examination is
not one-sided and whether the cognition itself has a worth admitted by all
men. The school has its prejudices, just as does the common understand-
ing. One improves the other here. It is therefore important that a cogni-
tion be examined by men whose understanding does not depend on any
school.

This perfection of cognition, whereby it qualifies for easy and universal
communication, could also be called external extension or the extensive
quantity of a cognition, insofar as it is widespread externally among men.
Since cognitions are so many and manifold, one will do well to make
himself a plan, in accordance with which he orders the sciences in the way
that best agrees with, and contributes to the furtherance of, his ends. All
cognitions stand in a certain natural connection with one another. Now if,
in striving to expand his cognitions, one does not look to their connection,
then extensive knowledge amounts to nothing more than a mere rhapsody.
If one makes one principal science his end, however, and
considers all other cognitions only as means for achieving it,
then he brings a certain systematic character into his knowledge.
And in order to go to work on extending his cognitions
according to such a well ordered and purposive plan,
one must seek, therefore, to become acquainted with
this connection of cognitions among themselves.
For this, the sciences get guidance from architectonic,
which is a system in accordance with ideas,
in which the sciences are considered in regard to their kinship and
systematic connection in a whole of cognition that interests humanity.

Now as for what concerns the intensive quantity of cognition;
its content, or its richness and importance,
which is essentially distinct from its extensive quantity,
its mere extensiveness, as we noted above,
we want here to add only the following few remarks:

1. A cognition that is concerned with what is great,
with the whole in the use of the understanding,
is to be distinguished from subtlety in what is small (micrology).

2. Every cognition that furthers logical perfection
as to form is to be called logically important;
e.g., every mathematical proposition, every law of nature
into which we have distinct insight, every correct philosophical explanation.
Practical importance cannot be foreseen, one must simply wait and watch for it.

3. Importance must not be confused with difficulty.
A cognition can be difficult without being important, and conversely.
Difficulty, then, does not decide either for or against
the worth or the importance of a cognition.
This rests on the quantity or multiplicity of its consequences.
A cognition is the more important accordingly as
it has more or greater consequences, as the use that
may be made of it is more.
Cognition without important consequences is called cavilling;
scholastic philosophy, e.g., was of this sort.

B) Logical perfection of cognition as to relation

Truth

    Material and formal, or logical, truth;
    Criteria of logical truth
    Falsehood and error
    Illusion, as source of error;
    Means for avoiding errors

A principal perfection of cognition, indeed,
the essential and inseparable condition of all its perfection, is truth.
Truth, it is said, consists in the agreement of cognition with its object.
In consequence of this mere nominal explanation, my cognition,
to count as true, is supposed to agree with its object.
Now I can compare the object with my cognition, however,
only by cognizing it.
Hence my cognition is supposed to confirm itself,
which is far short of being sufficient for truth.
For since the object is outside me, the cognition in me,
all I can ever pass judgment on is whether
my cognition of the object agrees with my cognition of the object.
The ancients called such a circle in explanation a diallelon.
And actually the logicians were always reproached with this mistake by the skeptics, who
observed that with this explanation of truth it is just as when someone
makes a statement before a court and in doing so appeals to a witness with
whom no one is acquainted, but who wants to establish his credibility by
maintaining that the one who called him as witness is an honest man.
The accusation was grounded, too.
Only the solution of the indicated problem
is impossible without qualification and for every man.
The question here is, namely, whether and to what extent there is a
criterion of truth that is certain, universal, and useful in application.
For this is what the question, What is truth?, ought to mean.
To be able to decide this important question we must distinguish that
which belongs to the matter in our cognition and is related to the object
from that which concerns its mere form, as that condition without which a
cognition would in general never be a cognition. With respect to this
distinction between the objective, material relation in our cognition and the
subjective, formal relation, the question above thus breaks down into these
two particular ones:
Is there a universal material, and
Is there a universal formal criterion of truth?
A universal material criterion of truth is not possible;
it is even self-contradictory.
For as a universal criterion, valid for all objects in general, it
would have to abstract fully from all difference among objects, and yet at
the same time, as a material criterion, it would have to deal with just this
difference, in order to be able to determine whether a cognition agrees
with just that object to which it is related and not just with any object in
general, in which case nothing would really be said.
Material truth must consist in this agreement of a cognition with just that determinate object
to which it is related, however.
For a cognition that is true in regard to one
object can be false in relation to other objects.
Hence it is absurd to demand a universal material criterion of truth,
which should abstract and at the same time not abstract from all difference among objects.
If the question is about universal formal criteria of truth, however, then
here it is easy to decide that of course there can be such a thing.
For formal truth consists merely in the agreement of cognition with itself,
in complete abstraction from all objects whatsoever and from all difference among them.
And the universal formal criteria of truth are accordingly nothing
other than universal logical marks of the agreement of cognition with itself
or - what is one and the same - with the universal laws of the understanding and of reason.
These formal, universal criteria are of course not sufficient for objective truth,
but they are nonetheless to be regarded as its conditio sine qua non.
For the question of whether cognition agrees with its objects
must be preceded by the question of whether it agrees with itself (as to form).
And this is a matter for logic.

The formal criteria of truth in logic are

1. the principle of contradiction,
2. the principle of sufficient reason.

Through the former the logical possibility of a cognition is determined,
through the latter its logical actuality.
To the logical actuality of a cognition it pertains, namely:
First: that it be logically possible, not contradict itself.
This characteristic of internal logical truth is only negative, however;
for a cognition that contradicts itself is of course false,
but if it does not contradict itself it is not always true.
Second: that it be logically grounded,
that it (a) have grounds and (b) not have false consequences.
This second criterion of external logical truth or of accessibility to reason,
which concerns the logical connection of a cognition with grounds and
consequences, is positive.
And here the following rules are valid:
1. From the truth of the consequence we may infer the truth of the cognition as
ground, but only negatively: if one false consequence flows from a cognition,
then the cognition itself is false. For if the ground were true, then the conse-
quence would also have to be true, because the consequence is determined by
the ground.
But one cannot infer conversely that if no false consequence flows from
a cognition, then it is true; for one can infer true consequences from a
false ground.
2. If all the consequences of a cognition are true, then the cognition is true too. For if
there were something false in the cognition, then there would have to be a
false consequence too.
From the consequence, then, we may infer to a ground, but without
being able to determine this ground. Only from the complex of all conse-
quences can one infer to a determinate ground, infer that it is the true
ground.

The former mode of inference, according to which the consequence
can only be a negatively and indirectly sufficient criterion of the truth of a
cognition, is called in logic the apagogic mode (modus tollens).
This procedure, of which frequent use is made in geometry, has the
advantage that I may derive just one false consequence from a cognition in
order to prove its falsehood. To show, e.g., that the earth is not flat, I may
just infer apagogically and indirectly, without bringing forth positive and
direct grounds: If the earth were flat, then the pole star would always have
to be at the same height; but this is not die case, consequently it is not flat.
With the other, the positive and direct mode of inference (modus ponens)
the difficulty enters that the totality of the consequences cannot be
cognized apodeictically, and that one is therefore led by the indicated
mode of inference only to a probable and hypothetically true cognition
(hypotheses), in accordance witii the presupposition that where many
consequences are true, all the remaining ones' may be true too.
Thus we will be able to advance three principles here as universal,
merely formal or logical criteria of truth; these are
the principle of contradiction and of identity (principium contradictionis and identita-
tis), through which the internal possibility of a cognition is determined for
problematic judgments;
the principle of sufficient reason (principium rationis suffidentis), on which rests
the (logical) actuality of a cognition, the fact that it is grounded, as material for
assertoric judgments;
the principle of the excluded middle (principium exclusi medii inter duo contradic-
toria m ), on which the (logical) necessity of a cognition is grounded - that we
must necessarily judge thus and not otherwise, i.e., that the opposite is false -
for apodeictic judgments.

The opposite of truth is falsehood, which,
insofar as it is taken for truth, is called error.
An erroneous judgment, for there is error as well as truth only in judgment,
is thus one that confuses the illusion of truth with truth itself.
It is easy to have insight into how truth is possible,
since here the understanding acts in accordance with its essential laws.
But it is hard to comprehend how error in the formal sense of the word,
how the form of thought contrary to the understanding is possible,
just as we cannot in general comprehend how any power
should deviate from its own essential laws.
We cannot seek the ground of errors in
the understanding itself and its essential laws, then,
just as little as we can in the restrictions of the understanding,
in which lies the cause of ignorance, to be sure,
but not in any way the cause of error.
Now if we had no other power of cognition
but the understanding, we would never err.
But besides the understanding, there lies in us
another indispensable source of cognition.
That is sensibility, which gives us the material for thought, and
in doing this works according to other laws than those the understanding does.
Error cannot arise from sensibility in and by itself, however,
because the senses simply do not judge.

The ground for the origin of all error will therefore have to be sought
simply and solely in the unnoticed influence of sensibility upon
the understanding, or to speak more exactly, upon judgment.
This influence, namely, brings it about that in judgment
we take merely subjective grounds to be objective, and
consequently confuse the mere illusion of truth with truth itself.
For it is just in this that the essence of illusion consists,
which on this account is to be regarded as a ground
for holding a false cognition to be true.

What makes error possible, then, is illusion,
in accordance with which the merely subjective is
confused in judgment with the objective.
In a certain sense, however, one can make the understanding the author
of errors, namely,  insofar as it allows itself,
due to a lack of requisite attention to that influence of sensibility,
to be misled by the illusion arising therefrom into holding
merely subjective determining grounds of judgment to be objective ones,
or into letting that which is true only according to the laws of sensibility hold
as true in accordance with its own laws.
In the restrictions of the understanding, then, lies only the responsibility for ignorance;
the responsibility for error we have to assign to ourselves.
Nature has denied us many cognitions, to be sure, it leaves us in
unavoidable ignorance concerning so much, but still it does not cause
error.
We are misled into this by our own inclination to judge and to
decide even where, on account of our limitedness, we are not able to
judge and to decide.
Every error into which the human understanding can fall is only partial,
however, and in every erroneous judgment there must always lie some-
thing true. For a total error would be a complete opposition to the laws of
the understanding and of reason. But how could that, as such, in any way
come from the understanding and, insofar as it is still a judgment, be held
to be a product of the understanding.
In respect to the true and the erroneous in our cognition,
we distinguish an exact cognition from a rough one.
Cognition is exact when it is adequate to its object, or
when there is not the slightest error in regard to its object,
and it is rough when there can be errors in it yet without being a hindrance to its purpose.
This distinction concerns the broader or narrower determinateness of our
cognition (cognitio late vel stride determinata).
Initially it is sometimes necessary to determine a cognition in a broader extension (late determinare),
particularly in historical things.
In cognitions of reason everything must be
determined exactly (striae), however.
In the case of broad determination
one says that a cognition is determined praeter propter. Whether a cogni-
tion ought to be determined roughly or exactly always depends on its
purpose. Broad determination leaves a certain play for error, which still
can have its determinate limits, however. Error occurs particularly where a
broad determination is taken for a strict one, e.g., in matters of morality,"
where everything must be determined striae. Those who do not do this are
called by die English latitudinarians.
One can distinguish subtlety, as a subjective perfection of cognition, from
exactness, as an objective perfection - since here cognition is fully congru-
ent with its object.
A cognition is subtle when one discovers in it that which usually escapes
the attention of others. It requires a higher degree of attention, then, and
a greater application of power of the understanding.
Many reprove all subtlety because they cannot attain it. But in itself it
always brings honor to the understanding and is even laudable and neces-
sary, insofar as it is applied to an object worthy of observation. When one
could have attained the same end with less attention and effort of the
understanding, however, and yet one uses more, then one makes a useless
expense and falls into subtleties, which are difficult, to be sure, but do not
have any use (nugae difficiles).
As the rough is opposed to the exact, so is the crude to the subtle.

From the nature of error - whose concept, as we noted, contains as an
essential mark, besides falsehood, also the illusion of truth - we get the
following important rule for the truth of our cognition:

To avoid errors - and no error is unavoidable, at least not absolutely or
without qualification, although it can be unavoidable relatively, for the
cases where it is unavoidable for us to judge, even with the danger of
error -
to avoid errors, then, one must seek to disclose and to explain
their source, illusion. Very few philosophers have done that, however.
They have only sought to refute the errors themselves, without indicating
the illusion from which they arise. This disclosure and breaking up of
illusion is a far greater service to truth, however, than the direct refutation
of errors, whereby one does not block their source and cannot guard
against the same illusion misleading one into errors again in other cases
because one is not acquainted with it. For even if we are convinced that we
have erred, then in case the illusion that grounds our error has not been
removed we still have scruples, however little we can bring forth
in justification of them.

Through the explanation of illusion, furthermore, one grants to the one
who erred a kind of fairness. For no one will admit that he erred without
any illusion of truth, which might even have deceived someone more
acute, because here it is a matter of subjective grounds.

Where the illusion is evident even to the common understanding (sensus
communis), an error is called a stupidity or an absurdity. The charge of
absurdity is always a personal reproof, which one must avoid, particularly
in the refutation of errors.

For to him who maintains an absurdity, the very illusion that lies at the
ground of the evident falsehood is not evident. One must first make this
illusion evident to him. Then if he still persists, he is admittedly stupid;
but then nothing more can be undertaken with him either. He has thereby
made himself just as incapable of further correction and refutation as he is
unworthy of it. For one cannot really prove to anyone that he is absurd;
here all ratiocination would be vain. If one proves absurdity, then one is no
longer speaking with him who erred but with him who is rational. But then
the disclosure of the absurdity (deductio ad absurdum) is not necessary.
One can also call a stupid error one that nothing, not even illusion, serves
to excuse; just as a crude error is an error that proves ignorance in common
cognition or a slip in common attentiveness.

Error in principles is greater than in their application.

An external mark or an external touchstone of truth is the comparison of
our own judgments with those of others, because the subjective will not be
present in all others in the same way, so that illusion can thereby be
cleared up. The incompatibility of the judgments of others with our own is
thus an external mark of error and is to be regarded as a cue to investigate
our procedure in judgment, but not for that reason to reject it at once. For
one can perhaps be right about the thingbut not right in manner, i.e., in the
exposition.

The common human understanding (sensus communis) is also in itself a
touchstone for discovering the mistakes of the artificial use of the under-
standing. This is what it means to orient oneself in thought or in the
speculative use of reason by means of the common understanding, when
one uses the common understanding as a test for passing judgment on the
correctness of the speculative use.

Universal rules and conditions for avoiding error in general are:

1) to think for oneself,
2) to think oneself in the position of someone else, and
3) always to think in agreement with oneself.

The maxim of thinking for oneself can be called
the enlightened mode of thought;
the maxim of putting oneself in the viewpoint of others in thought,
the extended mode of thought;
and the maxim of always thinking in agreement with one self,
the consequent or coherent mode of thought.

C) Logical perfection of cognition as to quality

Clarity

    Concept of a mark in general
    Various kinds of marks
    Determination of the logical essence of a thing
    Its distinction from the real essence
    Distinctness, a higher degree of clarity
    Aesthetic and logical distintness
    Distinction between analytic and synthetic distinctness

From the side of the understanding, human cognition is discursive;
it takes place through representations which take as the ground of cognition
that which is common to many things, hence through marks' as such.
Thus we cognize things through marks and that is called cognizing,
[the German word for which] comes from [the German word for] being acquainted.

A mark is that in a thing which constitutes
a part of the cognition of it, or
what is the same - a partial representation,
insofar as it is considered as
ground of cognition of the whole representation.
All our concepts are marks, accordingly, and
all thought is nothing other than a representing through marks.
Every mark may be considered from two sides:
First, as a representation in itself;
Second, as belonging, as a partial concept, to
the whole representation of a thing, and thereby
as ground of cognition of this thing itself.
All marks, considered as grounds of cognition, have two uses,
internal or an external use.
The internal use consists in derivation,
in order to cognize the thing itself
through marks as its grounds of cognition.
The external use consists in comparison,
insofar as we can compare one thing with others
through marks in accordance with the rules of
identity or diversity.

There are many specific differences among marks,
on which the following classification of them is grounded.

1. Analytic or synthetic marks.

The former are partial concepts of my actual concept
(marks that I already think therein),
while the latter are partial concepts of the merely possible complete concept
(which is supposed to come to be through a synthesis of several parts).

The former are all concepts of reason,
the latter can be concepts of experience.

2. Coordinate or subordinate.

This division of marks concerns their connection after or under one another.

Marks are coordinate insofar as each of them is
represented as an immediate mark of the thing
and are subordinate insofar as one mark is
represented in the thing only by means of the other.
The combination of coordinate marks to form
the whole of a concept is called an aggregate,
the combination of subordinate concepts a series.
The former, the aggregation of coordinate marks,
constitutes the totality of the concept,
regard to synthetic empirical concepts, can never be completed,
but rather resembles a straight line without limits.
The series of subordinate marks terminates a pane ante,
or on the side of the grounds, in concepts which cannot be broken up,
which cannot be further analyzed on account of their simplicity;
a pane post, or in regard to the consequences, it is infinite,
because we have a highest genus but no lowest species.
With the synthesis of every new concept in the aggregation of coordinate marks,
the extensive or extended distinctness grows,
as intensive or deep distinctness grows with the further
analysis of the concept in the series of subordinate marks.
This latter kind of distinctness, since it necessarily
contributes to thoroughness and coherence of the cognition,
is thus principally a matter of philosophy and is pursued
to the highest degree in metaphysical investigations in particular.

3. Affirmative or negative marks.

Through the former we cognize what the thing is,
through the latter what it is not.

Negative marks serve to keep us from errors.
Hence they are unnecessary where it is impossible to err,
and are necessary and of importance only in
those cases where they keep us from an important error
into which we can easily fall.
Thus in regard to the concept, e.g., of a being like God,
negative marks are quite necessary and important.
Through affirmative marks we seek to understand something,
through negative marks, into which all marks can be transformed,
we only seek not to misunderstand or not to err,
even if we should not thereby become acquainted with anything.

4. Important and fruitful, or empty and unimportant, marks.

A mark is important and fruitful if it is a ground of cognition for great
and numerous consequences, partly in regard to its internal use, its use in
derivation, insofar as it is sufficient for cognizing thereby a great deal in
the thing itself, partly in respect to its external use, its use in comparison,
insofar as it thereby contributes to cognizing both the similarity of a thing
to many others and its difference from many others.
We have to distinguish logical importance and fruitfulness from practical,
from usefulness and utility, by the way.

5. Sufficient and necessary or insufficient and accidental marks.

A mark is sufficient insofar as it suffices always
to distinguish the thing from all others;
otherwise it is insufficient, as the mark of barking is,
for example, for dogs.
The sufficiency of marks, as well as their importance,
is to be determined only in a relative sense,
in relation to ends that are intended through a cognition.
Necessary marks, finally, are those that must always be there to be found
in the thing represented.
Marks of this sort are also called essential and are
opposed to extra-essential and accidental marks,
which can be separated from the concept of the thing.
Among necessary marks there is another distinction, however.
Some of them belong to the thing as grounds of other marks of one and
the same thing, while others belong only as consequences of other marks.
The former are primitive and constitutive marks
(constitutiva, essentialia in sensu strictissimo),
the others are called attributes (consectaria, rationata)
and belong admittedly to the essence of the thing,
but only insofar as they must first be derived from its essential points,
as the three angles follow from the three sides
in the concept of the triangle, for example.
Extra-essential marks are again of two kinds;
they concern either internal determinations of a thing (moat) or
its external relations (relationes).
Thus the mark of learnedness signifies an inner determination of a man,
but being a master or a servant only an external relation.

The complex of all the essential parts of a thing, or the sufficiency of its
marks as to coordination or subordination, is the essence
(complexus notarum primitivarum, interne conceptui dato sufficientium;
s. complexus notarum, conceptum aliquem primitive constituentium).

In this explanation, however, we must not think at all of
the real or natural essence of things,
into which we are never able to have insight.
For since logic abstracts from all content of cognition, and
consequently also from the thing itself, in this science
the talk can only be of the logical essence of things.
And into this we can easily have insight.
For it includes nothing further than the cognition of
all the predicates in regard to which
an object is determined through its concept;
whereas for the real essence of the thing (esse ret)
we require cognition of those predicates on which,
as grounds of cognition, everything that belongs
to the existence of the thing depends.
If we wish to determine, e.g., the logical essence of body, then
we do not necessarily have to seek for the data for this in nature;
we may direct our reflection to the marks which, as essential points
(constitutiva, rationes) originally constitute the basic concept of the thing.
For the logical essence is nothing but the first basic concept of
all the necessary marks of a thing (esse conceptus).

The first stage of the perfection of our cognition
as to quality is thus its clarity.
A second stage, or a higher degree of clarity, is distinctness.
This consists in clarity of marks.

First of all we must here distinguish
logical distinctness in general from
aesthetic distinctness.
Logical distinctness rests on objective clarity of marks,
aesthetic distinctness on subjective clarity.
The former is a clarity through concepts,
the latter a clarity through intuition.
The latter kind of distinctness consists, then,
in a mere liveliness and understandability,
in a mere clarity through examples in concreto
(for much that is not distinct can still be understandable, and
conversely, much that is hard to understand can still be distinct,
because it goes back to remote marks, whose connection with intuition
is possible only through a long series).
Objective distinctness frequently causes
subjective obscurity, and conversely.
Hence logical distinctness is often possible only to the detriment of
aesthetic distinctness, and conversely aesthetic distinctness
through examples and similarities which do not fit exactly but are only
taken according to an analogy often becomes harmful to logical distinctness.
Besides, examples are simply not marks and do not belong to the
concept as parts but, as intuitions, to the use of the concept.
Distinctness through examples, mere understandability, is hence
of a completely different kind than distinctness through concepts as marks.
Lucidity consists in the combination of both,
of aesthetic or popular distinctness and
of scholastic or logical distinctness.
For one thinks of a lucid mind as the talent for a
luminous presentation of abstract and thorough cognitions
that is congruent with the common understanding's power of comprehension.
Next, as for what concerns logical distinctness in particular,
it is to be called complete distinctness insofar as all the marks
which, taken together, make up the whole concept have come to clarity.
A completely distinct concept can be so, again,
either in regard to the totality of its coordinate marks or
in respect to the totality of its subordinate marks.
Extensively complete or sufficient distinctness of a concept
consists in the total clarity of its coordinate marks,
which is also called exhaustiveness.
Total clarity of subordinate marks constitutes
intensively complete distinctness, profundity.

The former kind of logical distinctness can also be called the external
completeness (completudo externd) of the clarity of marks, the other the inter-
nal completeness (completudo internet). The latter can be attained only with
pure concepts of reason and with arbitrary concepts, but not with empiri-
cal concepts.

The extensive quantity of distinctness, insofar as it is not superfluous is
called precision. Exhaustiveness (completudo) and precision (praecisio) to-
gether constitute adequacy (cognitio, quae rem adaequat); and the completed
perfection of a cognition (consummata cognitionis perfectio) consists (as to qual-
ity) in intensively adequate cognition, profundity, combined with extensively
adequate cognition, exhaustiveness and precision.

Since, as we have noted, it is the business of logic to make clear concepts
distinct, the question now is in what way it makes them distinct.
Logicians of the Wolffian school place the act of making cognitions
distinct' entirely in mere analysis of them. But not all distinctness rests on
analysis of a given concept. It arises thereby only in regard to those marks
that we already thought in the concept, but not in respect to those marks
that are first added to the concept as parts of the whole possible concept.
The kind of distinctness that arises not through analysis but through
synthesis of marks is synthetic distinctness. And thus there is an essential
difference between the two propositions: to make a distinct concept and to
make a concept distinct.
For when I make a distinct concept, I begin with the parts and proceed
from these toward the whole. Here there are no marks as yet at hand; I
acquire them only through synthesis. From this synthetic procedure
emerges synthetic distinctness, then, which actually extends my concept
as to content through what is added as a mark beyond' the concept in (pure
or empirical) intuition. The mathematician and the natural philosopher
make use of this synthetic procedure in making distinctness in concepts.*
For all distinctness of properly mathematical cognition, as of all cognition
based on experience, rests on such an expansion of it through the synthe-
sis of marks.
When I make a concept distinct, however, my cognition does not grow
at all as to content through this mere analysis. The content remains the
same, only the form is altered, in that I learn to distinguish better, or to
cognize with clearer consciousness, what lay in the given concept already.
As nothing is added to a map through the mere illumination' of it, so a
given concept is not in the least increased through its mere illumination
by means of the analysis of its marks.
To synthesis pertains the making distinct of objects, to analysis the
making distinct of concepts. In the latter case the whole precedes the parts, in
the former the parts precede the whole. The philosopher only makes given
concepts distinct.
Sometimes one proceeds synthetically even when the
concept that one wants to make distinct in this way is already given.
 This is often the case with propositions based on experience, in case one is not
yet satisfied with the marks already thought in a given concept.
The analytic procedure for creating distinctness, with which alone logic
can occupy itself, is the first and principal requirement in making our
cognition distinct.
For the more distinct our cognition of a thing is, the
stronger and more effective it can be too.
 But analysis must not go so far
that in the end the object itself disappears.
If we were conscious of all that we know, we would have to be aston-
ished at the great multitude of our cognitions.

In regard to the objective content of our cognition in general,
we may think the following degrees, in accordance with which
cognition can, in this respect, be graded:

The first degree of cognition is: to represent something;

The second: to represent something with consciousness,
or to perceive (percipere);

The third: to be acquainted with something (noscere), or
to represent something in comparison with other things,
both as to sameness and as to difference;

The fourth: to be acquainted with something with consciousness,
to cognize it (cognoscere).

Animals are acquainted with objects too, but they do not cognize them.

The fifth: to understand something (intelligere),
to cognize something through the understanding by means of concepts,
or to conceive.

One can conceive much, although one cannot comprehend it;
e.g., a perpetuum mobile, whose impossibility is shown in mechanics.

The sixth: to cognize something through reason, or
to have insight into it (perspicere).

With few things do we get this far,
and our cognitions become fewer and fewer in number
the more that we seek to perfect them as to content.

The seventh, finally: to comprehend something (comprehendere),
to cognize something through reason or a priori
to the degree that is sufficient for our purpose.

For all our comprehension is only relative,
sufficient for a certain purpose;
we do not comprehend anything without qualification.
Nothing can be comprehended more than what the mathematician demonstrates,
that all lines in the circle are proportional.
And yet he does not comprehend how it happens that such a simple
figure has these properties.
The field of understanding or of the understanding is thus in general
much greater than the field of comprehension or of reason.

D) Logical perfection of cognition as to modality

Certainty

    Concept of holding-to-be-true in general
    Modi of holding-to-be-true: opining, believing and knowing
    Conviction and persuasion
    Reservation and deferral of a judgment
    Provisional judgments
    Prejudices, their sources and principal kinds

Truth is an objective property of cognition;
the judgment through which something is represented as true,
the relation to an understanding, and thus to a particular subject, is,
subjectively, holding-to-be-true.

Holding-to-be-true is in general of two kinds, certain or uncertain.
Certain holding-to-be-true, or certainty, is combined with
consciousness of necessity,
while uncertain holding-to-be-true, or uncertainty, is combined
with consciousness of the contingency or the possibility of the opposite.
The latter is again either subjectively as well as objectively insufficient, or
objectively insufficient but subjectively sufficient.
The former is called opinion, the latter must be called belief.

Accordingly, there are three kinds or modi of holding-to-be-true:
opining, believing, and knowing.
Opining is problematic judging,
believing is assertoric judging,
and knowing is apodeictic judging.

For what I merely opine I hold in judging, with consciousness, only to be problematic;
what I believe I hold to be assertoric, but not as objectively necessary, only as
subjectively so (holding only for me); what I know, finally, I hold to be
apodeiäically certain, i.e., to be universally and objectively necessary (hold-
ing for all), even granted that the object to which this certain holding-to-
be-true relates should be a merely empirical truth. For this distinction in
holding-to-be-true according to the three modi just named concerns only
the power of judgment in regard to the subjective criteria for subsumption of
a judgment under objective rules.

Thus, for example, our holding-to-be-true of immortality would be
merely problematic in case we only act as if we were immortal,
but it would be assertoric in case we believe that we are immortal,
and it would be apodeictic, finally, in case we all knew that
there is another life after this one.

There is an essential difference, then, between opining, believing, and knowing,
which we wish to expound more exactly and in more detail here.

1. Opining.

Opining, or holding-to-be-true based on a ground of cognition
that is neither subjectively nor objectively sufficient, can be regarded
as provisional judging (sub conditione suspensiva ad interim) that
one cannot easily dispense with.
One must first opine before one accepts and maintains,
but in doing so must guard oneself against holding an opinion to be
something more than mere opinion.
For the most part, we begin with
opining in all our cognizing. Sometimes we have an obscure premonition
of truth, a thing seems to us to contain marks of truth; we suspect its truth
even before we cognize it with determinate certainty.
But now where does mere opining really occur?
Not in any sciences that contain cognitions a priori,
hence neither in mathematics nor in metaphysics nor in morals,
but merely in empirical cognitions: in physics, psychology, etc.
For it is absurd to opine a priori.
In fact, too, nothing could be more ridiculous than, e.g.,
only to opine in mathematics.
Here, as in metaphysics and in morals,
the rule is either to know or not to know.
Thus matters of opinion can only be objects of a cognition by experience,
a cognition which is possible in itself but impossible for us in accordance
with the restrictions and conditions of our faculty of experience and the
attendant degree of this faculty that we possess.
Thus, for example, the ether of modern physicists is a mere matter of opinion.
For with this as with every opinion in general, whatever it may be,
I see that the opposite could perhaps yet be proved.
Thus my holding-to-be-true is here both objectively and subjectively insufficient,
although it can become complete, considered in itself.

2. Believing.

Believing, or holding-to-be-true based on a ground that is
objectively insufficient but subjectively sufficient, relates to objects in
regard to which we not only cannot know anything but also cannot opine
anything, indeed, cannot even pretend there is probability, but can only be
certain that it is not contradictory to think of such objects as one does
think of them. What remains here is a free holding-to-be-true, which is
necessary only in a practical respect given a priori, hence a holding-to-be-
true of what I accept on moral grounds, and in such a way that I am certain
that the opposite can never be proved.
Believing is not a special source of cognition. It is a kind of incomplete holding-to-be-true
with consciousness, and if considered as restricted to a particular kind of object (which
pertains only to believing), it is distinguished from opining not by its degree but rather by the
relation that it has as cognition to action. Thus the businessman, for example, to strike a
deal, needs not just to opine that there will be something to be gained thereby, but to believe
it, i.e., to have his opinion be sufficient for an undertaking into the uncertain. Now we have
theoretical cognitions (of the sensible) in which we can come to certainty, and in regard to
everything that we can call human cognition this latter must be possible. We have just such
certain cognitions, and in fact completely a priori, in practical laws, but these are grounded
on a supersensible principle (of freedom) and in fact in us ourselves, as a principle of practical
reason. But this practical reason is a causality in regard to a likewise supersensible object, the
highest good, which is not possible through our faculty in the sensible world. Nature as object
of our theoretical reason must nonetheless agree with this, for the consequence or effect of this
idea is supposed to be met with in the world of the senses. Thus we ought to act so as to
make this end actual.
Now in the world of the senses we also find traces of an artistic misdom, h and we believe
that the cause of the world also works with moral wisdom toward the highest good. This is a
holding-to-be-true that is enough for action, i.e., a belief. Now we do not need this for action
in accordance with moral laws, for these are given through practical reason alone, but we
need to accept a highest wisdom as the object of our moral will, an object beyond the mere
legitimacy of our actions, toward which we cannot avoid directing our ends. Although
objectively this would not be a necessary relation of our faculty of choice, subjectively the
highest good is still necessarily the object of a good (even of a human) will, and hence belief
in its attainability is necessarily presupposed.
There is no mean between the acquisition of a cognition through experience (a posteriori)
and through reason (apriori). But there is a mean between the cognition of an object and the
mere presupposition of its possibility, namely, an empirical ground or a ground of reason for
Matters of belief are thus I) not objects of empirical cognition. Hence so-
called historical belief cannot really be called belief, either, and cannot be
opposed as such to knowledge, since it can itself be knowledge. Holding-
to-be-true based on testimony is not distinguished from holding-to-be-
true through one's own experience either as to degree or as to kind.
II) [N]or [are they] objects of cognition by reason (cognition a priori),
whether theoretical, e.g., in mathematics and metaphysics, or practical, in
morals.
One can believe mathematical truths of reason on testimony, to be sure,
partly because error here is not easily possible, partly, too, because it can
easily be discovered, but one cannot know them in this way, of course. But
accepting this possibility in relation to a necessary extension of the field of possible objects
beyond those whose cognition is possible for us. This necessity occurs only in regard to that
in which the object is cognized as practical and, through reason, as practically necessary, for
to accept something on behalf of the mere extension of theoretical cognition is always
contingent. This practically necessary presupposition of an object is the presupposition of the
possibility of the highest good as object of choice, hence also of the condition of this
possibility (God, freedom, and immortality). This is a subjective necessity to accept the
reality of the object for the sake of the necessary determination of the will. This is the casus
extravrdinarius, without which practical reason cannot maintain itself in regard to its neces-
sary end, and here a favor necessitate proves useful to it in its own judgment. It cannot acquire
an object logically, but can only oppose what hinders it in the use of this idea, which belongs
to it practically.
This belief is the necessity to accept the objective reality of a concept (of the highest
good), i.e., the possibility of its object, asapriori necessary object of choice. If we look merely
to actions, we do not need this belief. But if we wish to extend ourselves through actions to
possession of the end that is thereby possible, then we must accept that this end is com-
pletely possible. Hence I can only say that / see myself necessitated through my end, in
accordance with laws of freedom, to accept as possible a highest good in the world, but I
cannot necessitate anyone else through grounds (the belief is free).
A belief of reason can never aim at theoretical cognition, then, for there objectively
insufficient holding-to-be-true is merely opinion. It is merely a presupposition of reason for
a subjective though absolutely necessary practical purpose. The sentiment toward moral
laws leads to an object of choice, which [choice] is determinable through pure reason. The
acceptance of the feasibility of this object, and hence of the reality of its cause, is a moral
belief, or a free holding-to-be-true that is necessary for moral purposes for completion of
one's ends.
Fides is really good faith 1 in thepactum, or subjective trust-' in one another, that one will keep
his promise to the other, with full faith and credit.* The first when the pactum is made, the
second when it is to be concluded.
In accordance with the analogy, practical reason is, as it were, the promisor,' man the
promissee," the good expected from the deed the promised."
philosophical truths of reason may not even be believed, they must simply
be known; for philosophy does not allow mere persuasion. And as for
what concerns in particular the objects of practical cognition by reason in
morals, rights and duties, there can just as little be mere belief in regard to
them. One must be fully certain whether something is right or wrong, in
accordance with duty or contrary to duty, allowed or not allowed. In moral
things one cannot risk anything on the uncertain, one cannot decide any-
thing on the danger of trespass against the law. Thus it is not enough for the
judge, for example, that he merely believe that someone accused of a crime
actually committed this crime. He must know it (juridically), or he acts
unconscientiously.
 The only objects that are matters of belief are those in which
holding-to-be-true is necessarily free, i.e., is not determined through
objective grounds of truth that are independent of the nature and the
interest of the subject.
Thus also on account of its merely subjective grounds, believing yields
no conviction that can be communicated and that commands universal
agreement, like the conviction that comes from knowledge. Only / myself
can be certain of the validity and unalterability of my practical belief, and
my belief in the truth of a proposition or the actuality of a thing is what
takes the place of a cognition only in relation to me without itself being a
cognition.
He who does not accept what it is impossible to know but morally neces-
sary to presuppose is morally unbelieving. At the basis of this kind of
unbelief lies always a lack of moral interest. The greater a man's moral
sentiment," the firmer and more lively will be his belief in all that he feels
himself necessitated to accept and to presuppose out of moral interest, for
practically necessary purposes.

3. Knowing.

Holding-to-be-true based on a ground of cognition that is
objectively as well as subjectively sufficient, or certainty, is
either empirical or rational, accordingly as it is grounded either
on experience, one's own as well as that communicated by others, or on reason.
This distinction relates, then, to the two sources from which
the whole of our cognition is drawn: experience and reason.

Rational certainty, again, is either mathematical or philosophical certainty.
The former is intuitive, the latter discursive.
Mathematical certainty is also called evidence, because an intuitive cog-
nition is clearer than a discursive one. Although the two, mathematical
and philosophical cognition of reason, are in themselves equally certain,
the certainty is different in kind in them.
Empirical certainty is original (originarie empirica) insofar as I become
certain of something from my own experience, and derived (derivative em-
pirica) insofar as I become certain through someone else's experience. The
latter is also usually called historical certainty.
Rational certainty is distinguished from empirical certainty by the con-
sciousness of necessity that is combined with it; hence it is apodeictic cer-
tainty, while empirical certainty is only assertoric. We are rationally certain
of that into which we would have had insight a priori even without any
experience. Hence our cognitions can concern objects of experience and
the certainty concerning them can still be both empirical and rational at
the same time, namely, insofar as we cognize an empirically certain propo-
sition from principles a priori.
We cannot have rational certainty of everything, but where we can have
it, we must put it before empirical certainty.
All certainty is either unmediated or mediated, i.e., it either requires a
proof, or it is not capable of and does not require any proof. Even if so
much in our cognition is certain only mediately, i.e., through a proof,
there must still be something indemonstrable or immediately certain, and the
whole of our cognition must proceed from immediately certain proposi-
tions.

The proofs on which any mediated or mediate certainty of a cognition
rests are either direct proofs or indirect, i.e., apagogical ones.
When I prove a truth from its grounds I provide a direct proof for it,
and when I infer the truth of a proposition from the falsehood of its
opposite I provide an indirect one.
If this latter is to have validity, however,
the propositions must be opposed contradictorily or diametraliter.
For two propositions opposed only as contraries (contrarie opposita) can both be false.
A proof that is the ground of mathematical certainty is called a demonstration,
and that which is the ground of philosophical certainty is called an acroamatic proof.
The essential parts of any proof in general are its matter and its form,
or the ground of proof and the consequentia.

From [the German word for] knowing comes [the German word for] science,
by which is to be understood the complex of a cognition as a system.
It is opposed to common cognition;
to the complex of a cognition as mere aggregate.
A system rests on an idea of the whole, which precedes the parts,
while with common cognition on the other hand,
or a mere aggregate of cognitions, the parts precede the whole.
There are historical sciences and sciences of reason.
In a science we often know only the cognitions but
not the things represented through them;
hence there can be a science of that of which
our cognition is not knowledge.

From the foregoing observations concerning
the nature and the kinds of holding-to-be-true
we can now draw the universal result that
all our conviction is thus either logical or practical.
When we know, namely, that we are free of all subjective grounds and
yet the holding-to-be-true is sufficient,
then we are convinced, and in fact logically convinced, or
convinced on objective grounds (the object is certain).
Complete holding-to-be-true on subjective grounds,
which in a practical relation hold just as much as objective grounds,
is also conviction, though not logical but rather pratical conviction (I am certain).
And this practical conviction, or this moral belief of reason,
is often firmer than all knowledge.
With knowledge one still listens to opposed grounds,
but not with belief, because here it does not depend
on objective grounds but on the moral interest of the subject.

Opposed to conviction stands persuasion, a holding-to-be-true on insuffi-
cient grounds, of which one does not know whether they are merely
subjective or also objective.
Persuasion often precedes conviction. We are conscious of many cogni-
tions only in such a way that we cannot judge whether the grounds of our
holding-to-be-true are objective or subjective. To be able to pass from
mere persuasion to conviction, then, we must first of all reflect, i.e., see to
which power of cognition a cognition belongs, and then investigate, i.e.,
test whether the grounds are sufficient or insufficient in regard to the
object. Many remain with persuasion. Some come to reflection, few to
investigation. He who knows what pertains to certainty will not easily mix
up persuasion and conviction, and hence will not let himself be easily
* This practical conviction is thus moral belief of reason," which alone can be called a belief in
the proper sense and be opposed as such to knowledge and to all theoretical or logical
conviction in general, because it can never elevate itself to knowledge. So-called historical
belief, on the other hand, as already observed, may not be distinguished from knowledge,
since as a kind of theoretical or logical holding-to-be-true it can itself be knowledge. We can
accept an empirical truth on the testimony of others with the same certainty as if we had
attained it through facta of our own experience. In the former kind of empirical knowledge
there is something deceptive, but also with the latter kind.
Historical or mediate empirical knowledge rests on die reliability of testimony. The
requirements of an irrefutable witness include authenticity* (competence") and integrity.
persuaded, either. There is a ground of determination to approval, which
is composed of objective and subjective grounds, and most men do not
analyze this mixed effect.
Although all persuasion is false as to form (formaliter), namely, insofar
as an uncertain cognition appears here to be certain, it can nonetheless be
true as to matter (materialiter). And thus it is distinct from opinion, too,
which is an uncertain cognition, insofar as it is held to be uncertain.
The sufficiency of holding-to-be-true (in belief) can be put to the test
by betting and by taking oaths. For the first comparative sufficiency of objec-
tive grounds is necessary, for the second absolute sufficiency, instead of
which, if this is not available, a merely subjectively sufficient holding-to-
be-true nevertheless holds.
It is customary to use the expressions, to agree with someone's judgment, to
reserve, to defer, or give up one's judgment. These and similar expressions
seem to indicate that there is something arbitrary 2 in our judging, in that
we hold something to be true because we want to hold it to be true. The
question arises, accordingly, whether willing has an influence on our judg-
ments.
The will does not have any influence immediately on holding-to-be-
true; this would be quite absurd. When it is said that we gladly believe what
we wish, this means only our benign wishes, e.g., those of a father for his
children. If the will had an immediate influence on our conviction concern-
ing what we wish, we would constantly form for ourselves chimeras of a
happy condition, and always hold them to be true, too. But the will cannot
struggle against convincing proofs of truths that are contrary to its wishes
and inclinations.
Insofar as the will either impels the understanding toward inquiry into a
truth or holds it back therefrom, however, one must grant it an influence
on the use of the understanding, and hence mediately on conviction itself,
since this depends so much upon the use of the understanding.
As for what concerns in particular the deferral or reservation of our
judgment, however, this consists in the resolution not to let a merely
provisional judgment become determining. A provisional judgment is one in
which I represent that while there are more grounds for the truth of a
thing than against it, these grounds still do not suffice for a determining or
definitive judgment, through which I simply decide for the truth. Provi-
sional judging is thus merely problematic judging with consciousness.
Reservation of judgment can happen for two purposes: either in order to
seek for the grounds of the determining judgment, or in order never to
judge. In the former case the deferral of judgment is called critical
(suspensio judicii indagatoria), in the hiter skeptical (suspensiojudiciisceptica).
For the skeptic refrains from all judgment, while the true philosopher
merely suspends his judgment in case he does not yet have sufficient
grounds for holding something to be true.
To suspend one's judgment in accordance with maxims requires a prac-
ticed faculty of judgment, which is found only in advancing age. In gen-
eral, reservation of our approval is a very hard thing, partly because our
understanding is so desirous of expanding itself and enriching itself with
cognitions" by judging, partly because our inclination is always directed
more toward certain things than toward others. He who has often had to
retract his approval, however, and who has thereby become smart and
cautious, will not give it so quickly, out of fear of having subsequently to
retract his judgment again. This revocation is always mortifying and causes
one to mistrust all other cognitions.*
We observe here further that leaving one's judgment in dubio is some-
thing different from leaving it in suspense. In the latter case I always have
an interest in the thing, in the former it is not always in conformity with
my end and interest to decide whether the thing is true or not.
Provisional judgments are quite necessary, indeed, indispensable, for
the use of the understanding in all meditation and investigation. For they
serve to guide the understanding in its inquiries and to provide it with
various means thereto.
When we meditate concerning an object, we must always judge provi-
sionally and, as it were, get the scent of the cognition that is partly to come
to us through the meditation. And when we go after inventions or discover-
ies, we must always make a provisional plan, otherwise our thoughts go on
at random. We can think of provisional judgments, therefore, as maxims
for the investigation of a thing. We could also call them anticipations,
because we anticipate our judgment of a thing even before we have the
determining judgment. Judgments of this sort have their good use, then,
and rules can even be given for how we ought to judge provisionally
concerning an object.
Prejudices must be distinguished from provisional judgments.
Prejudices are provisional judgments insofar as they are accepted as princi-
ples. Every prejudice is to be regarded as a principle of erroneous judg-
ments, and from prejudices arise not prejudices, but rather erroneous
judgments. Hence one must distinguish the false cognition that arises
from prejudice from its source, the prejudice itself. Thus the interpreta-
tion of dreams, for example, is not in itself a prejudice, but rather an error,
which arises from the assumed universal rule that what happens a few
times happens always, or is always to be held to be true. And this princi-
ple, under which the interpretation of dreams belongs, is a prejudice.
Sometimes prejudices are true provisional judgments; what is wrong is
only that they hold for us as principles or as determining judgments. The
cause of this deception is to be sought in the fact that subjective grounds
are falsely held to be objective, due to a lack of reflection, which must
precede all judging. For even if we can accept some cognitions, e.g.,
immediately certain propositions, without investigating them, i.e., without
examining the conditions of their truth, we still cannot and may not judge
concerning anything without reflecting, i.e., without comparing a cognition
with the power of cognition from which it is supposed to arise (sensibility
or the understanding). If we accept judgments without this reflection,
which is necessary even where no investigation occurs, then from this
prejudices arise, or principles for judging based on subjective causes that
are falsely held to be objective grounds.
The principal sources of prejudices are: imitation, custom, and inclination.
Imitation has a universal influence on our judgments, for there is a
strong ground for holding to be true what others have put forth as true.
Hence the prejudice that what the whole world does is right. As for what
concerns prejudices that have arisen from custom, they can only be rooted
out in the course of time, as the understanding, having little by little been
held up and slowed down in judging by opposing grounds, is thereby
gradually brought to an opposite mode of thought. If a prejudice of cus-
tom has arisen at the same time from imitation, however, then the man
who possesses it is very hard to cure. The inclination toward passive use of
reason, or toward the mechanism of reason rather than toward its spontaneity
under laws, can also be called a prejudice of imitation.
Reason is an active principle, to be sure, which ought not to derive
anything from the mere authority of others, nor even, when its pure use is
concerned, from experience. But the indolence of many men is such that
they prefer to follow in the footsteps of others rather than strain their own
powers of understanding. Men of this sort can only be copies of others,
and if everyone were of this kind, the world would remain eternally in one
and the same place. Hence it is most necessary and important not to
confine youths to mere imitation, as customarily happens.
There are so many things that contribute to accustoming us to the
maxim of imitation, and thereby to making reason a fruitful ground of
prejudices. Such aids to imitation include:
i. Formulas. 0 These are rules whose expression serves as a model for imitation.
They are uncommonly useful, by the way, for making complicated proposi-
tions easier, and the most enlightened mind therefore seeks to discover such
things.
Sayings, whose expression has the great precision of pregnant meaning, so
that it seems one could not capture the sense with fewer words. Pronounce-
ments of this sort (dicta\), which must always be borrowed from others whom
one trusts to have a certain infallibility, serve, on account of this authority, as
rules and as laws. The pronouncements of the bible are called sayings xat'
Sentences/ i.e., propositions which recommend themselves and which,
through the force of the thoughts lying within them, often retain their prestige
through centuries as products of a mature power of judgment.
Canones. These are universal rules 1 that serve as foundations for die sciences
and indicate something sublime and thought through. One can express them
in a sententious way, too, so that they are the more pleasing.
Proverbs (proverbia). These are popular rules of the common understanding,
or expressions for signifying its popular judgments. Since provincial proposi-
tions of this sort serve only the common crowd as sentences and canons, tiiey
are not to be found among people of finer upbringing.

From the three universal sources of prejudices stated above, and from
imitation in particular, many particular prejudices arise, among which we
wish to touch here upon the following as the most common,
i. Prejudices of prestige. Among these are to be reckoned:
a) The prejudice of the prestige of the person. If, in things that rest on
experience and on testimony, we build our cognition on the prestige of
other persons, we are not thereby guilty of any prejudice; for in matters of
this kind, since we cannot experience everything ourselves and compre-
hend it with our own understanding, the prestige of the person must be
the foundation of our judgments. When we make the prestige of others
the ground of our holding-to-be-true in respect of cognitions of reason,
however, we accept these cognitions merely on the basis of prejudice. For
truths of reason hold anonymously; the question here is not, Who said it?
but rather, What did he say? It does not matter at all whether a cognition is
of noble descent; but the inclination toward the prestige of great men is
nonetheless very common, partly because of the restrictedness of our own
insight, partly due to a desire to imitate what is described to us as great.
Added to this is the fact that the prestige of the person serves to flatter our
vanity in an indirect way. Just as the subjects of a powerful despot are
proud of the fact that they are all just treated equally by him, since to this
extent the least can fancy himself the equal of the foremost, as they are
both nothing over against the unrestricted power of their ruler, so too do
the admirers of a great man judge themselves to be equal, insofar as the
superiorities that they may have compared to one another are to be re-
garded as insignificant when considered against the merit of the great
man. For more than one reason, therefore, highly prized great men con-
tribute not a little to the inclination toward the prejudice of the prestige of
the person.
b) The prejudice of the prestige of the multitude. It is principally the crowd
who are inclined to this prejudice. For since they are unable to pass
judgment on the merits, the capabilities, and the cognitions' of the person,
they hold rather to the judgment of the multitude, under the presupposi-
tion that what everyone says must surely be true. This prejudice among
the vulgar relates only to historical matters, however; in matters of reli-
gion, where they are themselves interested, they rely on the judgment of
the learned.
It is in general noteworthy that the ignorant man has a prejudice for
learnedness, while the learned man, on the other hand, has a prejudice for
the common understanding.
If, after a learned man has nearly run the course of the sciences, he
does not procure appropriate satisfaction from all his efforts, he finally
acquires a mistrust of learnedness, especially in regard to those specula-
tions where the concepts cannot be made sensible, and whose foundations
are unsettled, as, e.g., in metaphysics. Since he still believes, however, that
it must be possible to find the key to certainty concerning certain objects
somewhere, he seeks it now in the common understanding, after he had
sought it so long in vain on the path of scientific inquiry.
But this hope is quite deceptive, for if the cultivated faculty of reason
can accomplish nothing in respect to the cognition of certain things, the
uncultivated faculty will certainly do so just as little. In metaphysics the
appeal to pronouncements of the common understanding is completely
inadmissible, because here no case can be exhibited in concreto. With
morals, however, the situation is admittedly different. Not only can all
rules in morals be given in concreto, but practical reason even manifests
itself in general more clearly and more correctly through the organ of the
common use of the understanding than through that of its speculative use.
Hence the common understanding often judges more correctly concern-
ing matters of morality and duty than does the speculative.
c) The prejudice of the prestige of the age. Here the prejudice of antiquity is
one of the most significant. We do have reason to judge kindly of antiquity,
to be sure, but that is only a ground for moderate respect, whose limits we
all too often overstep by treating the ancients as treasurers of cognitions
and of sciences, elevating the relative worth of their writings to an absolute
one and trusting blindly to their guidance. To esteem the ancients so
excessively is to lead the understanding back into its childhood and to
neglect the use of one's own talent. We would also err greatly if we
believed that everyone in antiquity had written as classically as those
whose writings have come down to us. For since time sifts everything and
preserves only what has an inner worth, we may assume, not without
reason, that we only possess the best writings of the ancients.
There are several causes by which the prejudice of antiquity is created
and sustained.
If something exceeds expectation as a universal rule, one initially won-
ders at this, and this wonder then often turns to admiration. This is the
case with the ancients when one finds something in them that, in respect
of the circumstances of time in which they lived, one did not seek. An-
other cause lies in the circumstance that acquaintance with the ancients
and with antiquity proves learnedness and wide reading, which always
brings respect, however common and insignificant in themselves the
things may be that one has drawn from the study of the ancients. A third
cause is the gratitude we owe to the ancients for the fact that they blazed
the path toward many cognitions.-' It seems fair to show them special
esteem, whose measure we often overstep, however. A fourth cause, fi-
nally, is to be sought in a certain envy toward our contemporaries. He who
cannot contend with the moderns extols the ancients at their expense, so
that the moderns cannot raise themselves above him.
The opposite of this is the prejudice of modernity. Sometimes the pres-
tige of antiquity and the prejudice in its favor declined, particularly at the
beginning of this century, when the famous Fontenelle 1 ? took the side of
the moderns. In the case of cognitions that are capable of extension, it is
quite natural that we place more trust in the moderns than in the ancients.
But this judgment has ground only as a mere provisional judgment. If we
make it a determining one, it becomes a prejudice.
2. Prejudices based on self-love or logical egoism, in accordance with which
one holds the agreement of one's own judgment with the judgments of
others to be a dispensable criterion of truth. They are opposed to the
prejudices of prestige, since they express themselves in a certain prefer-
ence for that which is the product of one's own understanding, e.g., one's
own system.
Is it good and advisable to let prejudices stand or even to encourage them? It
is astonishing that in our age such questions can still be advanced, espe-
cially that concerning the encouragement of prejudices. Encouraging some-
one's prejudices amounts to deceiving someone with good intent. It would
be permissible to leave prejudices untouched, for who can occupy himself
with exposing and getting rid of every prejudice? But it is another question
whether it would not be advisable to work toward rooting them out with all
one's powers. Old and rooted prejudices are admittedly hard to battle,
because they justify themselves and are, as it were, their own judges. People
also seek to excuse letting prejudices stand on the ground that disadvan-
tages would arise from rooting them out. But let us always accept these
disadvantages; they will subsequently bring all the more good.

SK 37.

yasmat buddhi purusasya upabho-gam sarvam pratisadhyati
sa eva ca suksma pradhana-purusantaram visinasti

Because it is the Buddhi that accomplishes the experiences
with regard to all objects to the Purusha.
It is that again that discriminates the subtle difference
between the Pradhana and the Purusha.

YS III.21

    kaya-rupa-samyamat tad-grahya-sakti-stambhe caksu-prakasa-asamprayoge ‘ntardhanam

YS III.22

    etena sabda-adi-antardhanam uktam
