I. INFERENCES OF THE UNDERSTANDING

Peculiar nature of inferences of the understanding

The essential character of all immediate inferences and
the principle of their possibility consists simply in
an alteration of the mere form of judgments,
while the matter of the judgments,
the subject and predicate, remains unaltered, the same.

    By virtue of the fact that in immediate inferences
    only the form of judgments is altered and not in any way the matter,
    these inferences differ essentially from all mediate inferences,
    in which the judgments are distinct as to matter too,
    since here a new concept must be added as mediating judgment or
    as middle concept (terminus medius) in order to deduce the
    one judgment from the other. If I infer, e.g.,
    All men are mortal, hence Caius is mortal too,
    this is not an immediate inference.
    For here I need for the deduction the mediating judgment, Caius is a man;
    through this new concept, however, the matter of the judgments is altered.
    With inferences of the understanding judicium intermedium may also be made,
    to be sure, but then this mediating judgment is merely tautological.
    As, for example, in the immediate inference,
    All men are mortal, some men are men, hence some men are mortal,
    the middle concept is a tautological proposition.

Modi of inferences of the understanding

Inferences of the understanding run through all the classes of
the logical functions of judgment and consequently are determined
in their principal kinds through the moments of
quantity, quality, relation, and modality.
On this rests the following division of these inferences.

1. Inferences of the understanding
(in relation to the quantity of judgments)
per judicia subalternata

In inferences of the understanding per judicia subalternata
the two judgments are distinct as to quantity, and here
the particular judgment is derived from the universal
in consequence of the principle:
The inference from the universal to the particular is valid
(ab universali ad particulare valet consequentia).

    A judicium is called subalternatum insofar as it is contained under the other,
    as, e.g., particular judgments are under universal ones.

2. Inferences of the understanding
(in relation to the quality of judgments)
per judicia opposita

In inferences of the understanding of this kind,
the alteration concerns the quality of the judgments,
and this considered in relation to opposition.
Now since this opposition can be of three kinds,
the following particular division of immediate inference results:
through contradictorily opposed, through contrary, and subcontrary judgments.

    Inferences of the understanding through equivalent judgments
    (judicia aequipollentia) cannot really be called inferences,
    for there is no consequence here;
    they are rather to be regarded as a mere substitution of words
    that signify one and the same concept, where the judgments themselves
    also remain unaltered as to form.
    E.g., Not all men are virtuous, and, Some men are not virtuous.
    The two judgments say one and the same thing.

a. Inferences of the understanding per judicia contradictorie opposita

In inferences of the understanding through judgments which are opposed
to one another contradictorily and which, as such, constitute genuine,
pure opposition, the truth of one of the contradictorily opposed judgments
is deduced from the falsehood of the other, and conversely.
For genuine opposition, which occurs here, contains no more and no less than what
belongs to opposition.
In consequence of the principle of the excluded middle,
the two contradicting judgments cannot both be true,
and just as little can they both be false.
If the one is true, then the other is false, and conversely.

b. Inferences of the understanding per judicia contrarie opposita

Contrary or conflicting judgments (judicia contrarie opposita) are judgments of which
the one is universally affirmative, the other universally negative.
Now since one of them says more than the other, and since in the excess,
that it says more than the mere negation of the other,
there can lie falsehood, they cannot both be true, of course,
but they can both be false.
In regard to these judgments, therefore, only the inference
from the truth of the one to the falsehood of the other holds, hut not conversely.

c. Inferences of the understanding per judicia subcontrarie opposita

Subcontrary judgments are ones of which the one affirms or denies
in particular (particulariter) what the other denies or affirms in particular.
Since they can both be true but cannot both be false,
only the following inference holds in regard to them:
If one of these propositions is false, the other is true, but not conversely.

    In the case of subcontrary judgments there is no pure, strict opposition,
    for what is affirmed or denied in the one is not denied or affirmed of the same
    objects in the other.
    In the inference, e.g., Some men are learned, hence some men are not learned,
    what is denied in the second judgment is not affirmed of the same men in the first.

3. Inferences of the understanding
(with respect to the relation of judgments)
per judicia conversa sive per conversionem

Immediate inferences through conversion concern the relation of judgments and
consist in the transposition of subject and predicate in the two judgments,
so that the subject of the one judgment is made
the predicate of the other judgment, and conversely.

Pure and altered conversion

In conversion the quantity of the judgments is either altered or it remains
unaltered. In the first case the converted (conversum) is distinct from what
converts (convertens) as to quantity, and the conversion is said to be altered 3 '
(conversio per accidens); in the latter case the conversion is called pure
(conversio simpliciter talis).

Universal rules of conversion

With respect to inferences of the understanding through conversion,
the following rules hold:
Universal affirmative judgments may be converted only per accidens,
for the predicate in these judgments is a broader concept and thus
only some of it is contained in the concept of the subject.
But all universal negative judgments may be converted simpliciter,
for here the subject is removed from the sphere of the predicate.
Thus too, finally,
All particular affirmative propositions may be converted simpliciter;
for in these judgments
a part of the sphere of the subject has been subsumed under the predicate,
hence a part of the sphere of the predicate may be subsumed under the subject.

    In universal affirmative judgments the subject is considered as a con-
    tentum of the predicate, since it is contained under its sphere. Therefore I
    may only infer, e.g., All men are mortal, hence some of those who are
    contained under the concept mortal are men. The cause of the fact that
    universal negative judgments may be converted simpliciter is that two
    concepts that contradict one another universally contradict each other in
    the same extension.
    Some universal affirmative judgments may be converted simpliciter, to be
    sure. But the ground of this lies not in their form but in the particular
    character of their matter; as, e.g., the two judgments, Everything unalter-
    able is necessary, and everything necessary is unalterable.

4. Inferences of the understanding
(in relation to the modality of judgments)
per judicia contraposita

The immediate mode of inference through contraposition consists in that
transposition (metathesis of judgments in which
merely the quantity remains the same while the quality is altered.
They concern only the modality of judgments,
since they transform an assertoric into an apodeictic judgment.

Universal rule of contraposition

With respect to contraposition, the universal rule holds:
All universal affirmative judgments may be contraposed simpliciter.
For if the predicate, as that which contains the subject under itself, is denied,
and hence the whole sphere, then a part of it must also be denied, i.e., the
subject.

    The metathesis of judgments through conversion and that
    through contraposition are thus opposed to one another to the extent that
    the former alters merely quantity, the latter merely quality.

    The immediate modes of inference mentioned relate merely to
    categorical judgments.

II. INFERENCES OF REASON

Inference of reason in general

An inference of reason is the cognition of the necessity of a proposition
through the subsumption of its condition under a given universal rule.

Universal principle of all inferences of reason

The universal principle on which the validity of
all inference through reason rests may be
determinately expressed in the following formula:
What stands under the condition of a rule
also stands under the rule itself.

    The inference of reason premises a universal rule and
    a subsumption under its condition.
    Through this one cognizes the conclusion a priori,
    not in the individual, but as contained in the universal and
    as necessary under a certain condition.
    And this, that everything stands under the universal and
    is determinable in universal rules, is just
    the principle of rationality or of necessity
    (principium rationalitatis sive necessitatis).

Essential components of the inference of reason

To every inference of reason belong the following essential three parts:
1. a universal rule, which is called the major proposition (propositio major),
2. the proposition which subsumes a cognition under the condition of
the universal rule, and which is called the minor proposition (propositio minor), and finally
3. the proposition which affirms or denies the rule's predicate of the subsumed cognition:
the conclusion (conclusio).

The first two propositions, in their combination with one another, are called premises.

    A rule is an assertion under a universal condition.
    The relation of the condition to the assertion, namely,
    how the latter stands under the former, is the exponent of the rule.
    The cognition that the condition (somewhere) exists is the subsumption.
    The combination of that which is subsumed under the condition with
    the assertion of the rule is the inference.

Matter and form of inferences of reason

The matter of inferences of reason consists in
the antecedent propositions or premises,
the form in the conclusion insofar as it contains the consequentia.

    In every inference of reason, then, the truth of the premises is to be
    examined first, and then the correctness of the consequentia.
    In rejecting an inference of reason, one must never reject the conclusion first, but
    rather one must first reject either the premises or the consequentia.
    In every inference of reason the conclusion is given as soon as the
    premises and the consequentia are given.

Division of inferences of reason (as to relation)
into categorical, hypothetical, and disjunctive

All rules (judgments) contain objective unity of consciousness of
the manifold of cognition, hence a condition under which one cognition
belongs with another to one consciousness.
Now only three conditions of this unity may be thought, however, namely:
as subject of the inherence of marks, or
as ground of the dependence of one cognition on another, or, finally,
as combination of parts in a whole (logical division).
Consequently there can only be just as many kinds of universal rules
(propositiones majores), through which the consequentia of one judgment
from another is mediated.
And on this is grounded the division of all inferences of reason into
categorical, hypothetical, and disjunctive.

    Inferences of reason can be divided neither as to quantity,
    for every major is a rule, hence something universal;
    nor in regard to quality, for it is equivalent whether
    the conclusion is affirmative or negative;
    nor, finally, in respect of modality,
    for the conclusion is always accompanied with the consciousness of
    necessity and consequently has the dignity of an apodeictic proposition.
    Thus only relation remains as the sole possible
    ground of division of inferences of reason.
    Many logicians hold only categorical inferences of reason to be ordinary,
    the others to be extraordinary.
    But this is groundless and false.
    For all three of these kinds are products of equally correct,
    though equally essentially different functions of reason.

Peculiar difference between categorical, hypothetical and disjunctive
inferences of reason

The distinguishing feature among the three mentioned kinds of inferences of reason
lies in the major premise.
In categorical inferences of reason the major is a categorical proposition,
in hypothetical ones it is a hypothetical or problematic proposition,
and in disjunctive ones it is a disjunctive proposition.

1. Categorical inferences of reason

In every categorical inference of reason there are three principal concepts 1
(termini), namely:
1.
2.
3.
the predicate in the conclusion, which concept is called the major concept 11
(terminus major), because it has a larger sphere than the subject,
the subject (in the conclusion), whose concept is called the minor concept'
(terminus minor), and
a mediating mark (nota intermedia), which is called the middle concept' (terminus
medius), because through it a cognition is subsumed under the condition of
the rule.
Note.
This distinction among the mentioned termini exists only in categorical
inferences of reason, because these alone infer through a terminus medius;
the others, on the other hand, infer only through the subsumption of a
proposition represented problematically in the m ajo r L and assertorically in
the minor L .
§6 3
Principle of categorical inferences of reason
The principle on which the possibility and validity of all categorical infer-
ences of reason rests is this:
What belongs to the mark of a thing belongs also to the thing itself; and what
617
contradicts the mark of a thing contradicts also the thing itself (nota notae est nota
rei ipsius; repugnans notae, repugnat rei ipsi).
Note.
From the principle just set forth the so-called dictum de omni et nullo may
easily be deduced, and hence it can hold as the first principle neither for
inferences of reason in general nor for the categorical in particular.
Genus and species concepts are universal marks of all things that stand
under these concepts. Accordingly, the rule holds here: What belongs to or
contradicts the genus or species belongs to or contradicts all the objects that are
contained under that genus or species. And this rule is called just the dictum
de omni et nullo.
§64
Rules for categorical inferences of reason
From the nature and the principle of categorical inferences of reason flow
the following rules for them:
124
i.
2.
3.
4.
5.
6.
7.
In no categorical inference of reason can either more or fewer than three
principal concepts (termini) be contained; for here I am supposed to combine
two concepts (subject and predicate) through a mediating mark.
The antecedent propositions or premises may not be wholly negative (expuris
negatives nihil sequitur*); for the subsumption in the minor premise must be
affirmative, as that which states that a cognition stands under the condition of
the rule.
The premises may also not be wholly particular 1 propositions (ex puris particu-
laribus nihil sequitur™); for then there would be no rule, i.e., no universal
proposition, from which a particular cognition could be deduced.
The conclusion always follows the weaker pan of the inference, i.e., the negative and
the particular proposition in the premises, as that which is called the weaker
part of the categorical inference of reason (conclusio sequitur partem debilio-
rem"). If, therefore,
one of the premises is a negative proposition, then the conclusion must be
negative too, and
if one premise is a particular proposition, then the conclusion must be particu-
lar too.
In all categorical inferences of reason the majo r L must be a universal proposi-
tion (universalis), the minor L an affirmative one (affirmans), and from this it
follows, finally,
From purely negative [premises] nothing follows.
"besondere (particulare)."
From purely particular [premises] nothing follows.
The conclusion follows the weaker part.
that the conclusion must follow the major premise in regard to quality, but the
minor premise as to quantity.
Note.
It is easy to see that the conclusion must in every case follow the negative
and the particular proposition in the premises.
If I make the minor premise only particular, and say, Some is contained
under the rule, then I can only say in the conclusion that the predicate of
the rule belongs to some, because I have not subsumed more than this under
the rule. And if I have a negative proposition as the rule (major premise),
then I must also make the conclusion negative. For if the major premise
says that this or that predicate must be denied of everything that stands
under the condition of the rule, then the conclusion must also deny the
predicate of that (subject) which has been subsumed under the condition
of the rule.
§65
Pure and mixed categorical inferences of reason
A categorical inference of reason is pure 0 (punts) if no immediate inference
is mixed in with it, nor is the legitimate order of the premises altered; in
the contrary case it is called impure or mixed (ratiocinium impurum or
hybridum).
§66
Mixed inferences of reason through conversion of
propositions - Figures
Those inferences which arise through the conversion of propositions and
in which the position of these propositions is thus not the legitimate one
are to be counted as mixed inferences. This case occurs in the three latter
so-called figures of the categorical inference of reason.
§6?
Four figures of inferences
By figures are to be understood those four modes of inferring whose
difference is determined through the particular position of the premises
and their concepts.
§68
Ground of the determination of their distinction through the
position of the middle concept
126
different
The middle concept, namely, on whose position things really depend
here, can occupy either (i.) the place of the subject in the major premise
and the place of the predicate in the minor premise, or (2.) the place of the
predicate in both premises, or (3.) the place of the subject in both, or
finally (4.) the place of the predicate in the major premise and the place of
the subject in the minor premise. Through these four cases, the distinc-
tion among the four figures is determined. If we let 5 signify the subject of
the conclusion, P its predicate, and M the terminus medius, then the
schema for the four mentioned figures may be exhibited in the following
table:
M
S
S
P
M
P
P M
S M
S P
M P
M S
S P
P M
M S
S P

Rules for the first figure, as the only legitimate one

The rule of first figure is that
the major be a universal proposition,
the minor an affirmative.
And since this must be the universal rule of
all categorical inferences of reason in general,
it results from this that the first figure is the only legitimate one,
which lies at the basis of all the others, and to which all the others
must be traced back through conversion of premises (metathesis praemissorum),
insofar as they are to have validity.

    The first figure can have a conclusion of any quantity and quality. In the
    other figures there are conclusions only of a certain kind; some modi of
    them are excluded here. This already indicates that these figures are not
    perfect, that there are certain restrictions in them which prevent the conclu-
    sion from occurring in all modi, as in the first figure.

Condition of the reduction of the three latter figures to the first

The condition of the validity of the three latter figures, under which a
correct modus of inference is possible in each of them, amounts to this:
that the medius terminus occupies a place such that from it, through
immediate inferences (consequentias immediatas), their position in accor-
dance with the rules of the first figure can arise. -
From this the following rules for the three latter figures emerge.

Rules of the second figure

In the second figure the minor L stands rightly, hence the major must be
converted, and in such a way that it remains universal (universalis).
This is possible only if it is universally negative;
but if it is affirmative, then it must be contraposed.
In both cases the conclusion becomes negative
(sequitur partem debiliorem).

    The rule of the second figure is, What is contradicted by a mark of a thing
    contradicts the thing itself. Now here I must first convert and say, What is
    contradicted by a mark contradicts this mark; or I must convert the conclusion,
    What is contradicted by a mark of a thing is contradicted by the thing itself,
    consequently it contradicts the thing.

Rule of the third figure

In the third figure the major stands rightly, hence the minor must be converted,
yet in such a way that an affirmative proposition arises therefrom.
This is only possible, however, when the affirmative proposition is particular,
consequently the conclusion is particular.

    The rule of the third figure is, What belongs to or contradicts a mark also
    belongs to or contradicts some things under which this mark is contained.
    Here I must first say, It belongs to or contradicts everything that is con-
    tained under this mark.

Rule of the fourth figure

If in the fourth figure the major is universal negative, then it may be
converted purely (simplidter), just as the minor L may be as particular;
hence the conclusion is negative. If, on the other hand, the major L is
universal affirmative, then it may either be converted only per accidens or it
may be contraposed; hence the conclusion is either particular or negative.
If the conclusion is not to be converted (P S transformed into 5 P), a
transposition of the premises (metathesis praemissorum) or a conversion
(conversio) must occur.

    In the fourth figure it is inferred that the predicate depends on the medius
    terminus, the medius terminus on the subject (of the conclusion), conse-
    quently the subject on the predicate; this simply does not follow, however, but
    at most its converse. To make this possible, the major L must be made the
    minor L and vice versa, and the conclusion must be converted, because in the
    first alteration the terminus minor is transformed into the major L .

Universal results concerning the three latter figures

From the rules stated for the three latter figures it is clear
1. that in none of them is there a universal affirmative conclusion, but rather the
conclusion is always either negative or particular;
2. that in every one an immediate inference (consequentia immediatd) is mixed in,
which is not expressly signified, to be sure, but still must be silently included;
3. that on this account, too, these three latter modi of inference must all be called
impure inferences (ratiocinia hybrida, impurd), not pure ones,
since no pure inference can have more than three principal propositions (termini).

2. Hypothetical inferences of reason

A hypothetical inference is one that has a hypothetical proposition as major.
Thus it consists of two propositions,
(1.) an antecedent proposition (antecedent) and
(2.) a consequent proposition (consequens), and
here the deduction is either according to modus ponens or to modus tollens.

    Hypothetical inferences of reason have no medius terminus, then, but instead
    the consequentia of one proposition from another is only indicated in them.
    In their major, namely, the consequentia of two propositions from one another
    is expressed, of which the first is the premise, the second a conclusion.
    The minor is a transformation of the problematic condition into a categorical proposition.

    From the fact that the hypothetical inference consists only of two propositions,
    without having a middle concept, it may be seen that it is really
    not an inference of reason, but rather only an immediate inference,
    to be proved from an antecedent proposition and a consequent proposition,
    as to matter or form (consequentia immediata demonstrabilis [ex antecedente et
    consequente] vel quoad materiam vel quoad formam).
    Every inference of reason is supposed to be a proof.
    But the hypothetical carries with it only the ground of proof.
    It is clear from this, consequently, that it cannot be an inference of reason.

Principle of hypothetical inferences

The principle of hypothetical inferences is
the principle of the ground,

a ratione ad rationatum; a negatione rationati ad negationem rationis valet consequential

3. Disjunctive inferences of reason

In disjunctive inferences, the major is a disjunctive proposition and as such
must therefore have members of division or disjunction. Here we infer either
(1.) from the truth of one member of the
disjunction to the falsehood of the others, or
(2.) from the falsehood of all members but one to the truth of this one.
The former occurs through the modus ponens (or ponendo tollens),
the latter through the modus tollens (tollendo ponens).
The consequentia from the ground to the grounded,
and from the negation of the grounded to negation of the ground, is valid.

    All the members of the disjunction but one, taken together, constitute the
    contradictory opposite of this one.
    Here there is a dichotomy, then, according to which,
    if one of the two is true, the other must be false, and conversely.
    2. All disjunctive inferences of reason of more than two members of
    disjunction are thus really polysyllogistic.
    For every true disjunction can only be bimembris,
    and logical division is also bimembris,
    but for brevity's sake the membra subdividentia are
    posited among the membra dividentia.

Principle of disjunctive inferences of reason

The principle of disjunctive inferences is
the principle of the excluded middle:

A contradictorie oppositorum negatione unius ad affirmationem
positione unius ad negationem alterius valet consequential alterius

Dilemma

A dilemma is a hypothetical-disjunctive inference of reason,
or a hypothetical inference, whose consequens is a disjunctive judgment.
The hypothetical proposition whose consequens is disjunctive is the major proposition;
the minor proposition affirms that the consequens (per omnia membra) is false, and
the conclusion affirms that the antecedent is false.
(A remotione consequents ad negationem antecedents valet consequential)

    The ancients made a great deal of the dilemma and
    named this inference comutus.
    They knew how to drive an opponent into a corner by rehearsing
    everything to which he could turn and then contradicting everything, too.
    They showed him many difficulties with any opinion he accepted. But it is
    a sophistical trick, not to refute propositions directly but rather only to
    show difficulties, which is feasible with many, indeed, most things.
    Now if we wish to declare as false everything with which we find difficulties,
    then it is an easy game to reject everything. It is good, to be sure, to
    show the impossibility of the opposite, only there still lies something decep-
    tive in this in case one takes the incomprehensibility of the opposite for its
    impossibility. Hence dilemmata have much that is captious in them, even
    though they infer correctly. They can be used to defend true propositions,
    but also to attack true propositions by means of difficulties that one throws
    up against them.

Formal and coven inferences of reason (ratiocinia formalia and cryptica)

A formal inference of reason is one that not only contains everything required
as to matter but also is expressed correctly and completely as to form.
Opposed to formal inferences of reason are coven ones (cryptica),
which all those can be reckoned, in which
either the premises are transposed,
or one of the premises is left out, or, finally,
the middle concept alone is combined with the conclusion.
A covert inference of reason of the second kind,
in which one premise is not expressed but only thought, is called
a truncated one or an enthymeme.
Those of the third kind are called contracted inferences.

III. INFERENCES OF THE POWER OF JUDGMENT

Determinative and reflective power of judgment

The power of judgment is of two kinds:
the determinative or the reflective power of judgment.
The former goes from the universal to the particular,
the second from the particular to the universal.
The latter has only subjective validity,
for the universal to which it proceeds from the particular is
only empirical universality, a mere analogue of the logical.

Inferences of the (reflective) power of judgment

Inferences of the power of judgment are certain modes of inference for
coming from particular concepts to universal ones.
They are not functions of the determinative power of judgment,
then, but rather of the reflective;
hence they also do not determine the object,
but only the mode of reflection concerning it,
in order to attain its cognition.

Principle of these inferences

The principle that lies at the basis of these inferences of the
power of judgment is this:

that the many will not agree in one without a common ground,
but rather that which belongs to the many in this way will be
necessary due to a common ground.

    Since such a principle lies at the basis of
    the inferences of the power of judgment,
    they cannot, on that account, be held to be immediate inferences.

Induction and analogy — The two modes of inference of the power of judgment

The power of judgment, by proceeding from the particular to the univer-
sal in order to draw from experience (empirically) universal - hence not a
priori — judgments, infers either from many to all things of a kind, or from
many determinations and properties, in which things of one kind agree, to
the remaining ones, insofar as they belong to the same principle. The former
mode of inference is called inference through induction, the other infer-
ence according to analogy.

    Induction infers, then, from the particular to the universal (aparticular? ad
    universale) according to the principle of universalization:* What belongs to
    many things of a genus belongs to the remaining ones too. Analogy infers from
    particular to total similarity of two things, according to the principle of
    specification: Things of one genus, which we know to agree in much/ also
    agree in what remains, with which we are familiar in some things of this
    genus but which we do not perceive in others. Induction extends the
    empirically given from the particular to the universal in regard to many
    objects, while analogy extends the given properties of one thing to several
    [other properties] of the very same thing[.] - One in many, hence in all:
    Induction; many in one (which are also in others), hence also what remains
    in the same thing: Analogy. Thus the ground of proof for immortality
    from the complete development of natural dispositions of each creature
    is, for example, an inference according to analogy.

    In the inference according to analogy, however, identity of the ground
    (par ratio) is not required. In accordance with analogy we infer only
    rational inhabitants of the moon, not men. Also, one cannot infer accord-
    ing to analogy beyond the tertium comparationis.
    Every inference of reason must yield necessity. Induction and analogy are
    therefore not inferences of reason, but only logical presumptions, or even
    empirical inferences; and through induction one does get general proposi-
    tions, but not universal ones.
    The mentioned inferences of the power of judgment are useful and
    indispensable for the sake of the extending of our cognition by experi-
    ence/ But since they give only empirical certainty, we must make use of
    them with caution and care.

Simple and composite inferences of reason

An inference of reason is called simple if it consists
of one inference of reason, composite if of several.

Ratiocinatio polysyllogistica

A composite inference, in which the several inferences of reason are
combined with one another not through mere coordination but through
subordination, i.e., as grounds and consequences, is called
a chain of inferences of reason (ratiocinatio polysyllogistica).

Prosyllogisms and episyllogisms

In the series of composite inferences one can infer in two ways,
either from the grounds down to the consequences, or
from the consequences up to the grounds.
The first occurs through episyllogisms,
the other through prosyllogisms.
An episyllogism is that inference, namely, in the series of inferences,
whose premise becomes the conclusion of a prosyllogism,
inference that has the premises of the former as conclusion.

Sorites or chain inference

An inference consisting of several inferences that are shortened and com-
bined with one another for one conclusion is called a sorites or a chain
inference, which can be either progressive or regressive, accordingly as one
climbs from the nearer grounds up to the more distant ones, or from the
more distant grounds down to the nearer ones.

Categorical and hypothetical sorites

Both progressive and regressive chain inferences can in turn be either
categorical or hypothetical. The former consists of categorical propositions, as a
series of predicates, the latter of hypothetical ones, as a series of conse-
quences.

Fallacy — Paralogism - Sophism

An inference of reason that is wrong as to form, although it has for itself
the illusion of a correct inference, is called a fallacy {fallacia). Such an
inference is a paralogism insofar as one deceives oneself through it, a
sophism insofar as one intentionally seeks to deceive others through it.

    The ancients occupied themselves very much with the art of making such
    sophisms. Therefore many of this kind have emerged, e.g., the sophisma
    figurae dictionis, in which the medius terminus is taken in different
    meanings - fallacia a dicta secundum quid ad dictum simpliciter, sophisma
    heterozeteseos, elenchi d ignorationis,' etc.

Leap in inference

A leap (saltus) in inference or proof is the combination of
a premise with the conclusion so that the other premise is left out.
Such a leap is legitimate Reading "elenchi" for "elenchi,".
' the fallacy [of inferring] from what is said with qualification to what is said simpliciter, the
sophism of misdirection, the ignoratio elenchis, etc.
(legitimus) if everyone can easily add the missing premise in thought, but
illegitimate (illegitimus) if the subsumption is not clear.
Here a distant mark is connected with a thing
without an intermediate mark (nota intermedia).

Petitio principii - Circulus in probando

By a petitio principii is understood the acceptance of a proposition as
ground of proof as an immediately certain proposition, although it still
requires a proof. And one commits a circle in proof if one lays at the basis of
its own proof the very proposition that one wanted to prove.

    The circle in proof is often hard to discover, and this mistake is usually
    most frequently committed where the proofs are hard.

Probatio plus and minus probans

A proof can prove too much, but also too little.
In the latter case it proves only a part of what is to be proved,
in the former it also goes on to that which is false.

    A proof that proves too little can be true and thus is not to be rejected. But
    if it proves too much, then it proves more than is true, and that is then
    false. Thus, for example, the proof against suicide, that he who has not
    given life cannot take it either, proves too much; for on this ground we
    would not be permitted to kill animals. Thus it is false.
