
Something is said to BE coincidentally or intrinsically:

[1]     coincidentally

        We say that the just person is musical and the human is musical,
        and the musical is human, and this is quite similar to saying that
        the musical builds because the builder happens to be musical
        (for here "this is this" signifies "this coincides with this".)

[1a]    Both attributes coincide in the same thing, and and in the first because
[1b]    the attribute coincides with the thing that is a being, whereas
[1c]    the musical coincides with a human.

        Things said to be coincidentally because:

[1a]    Both belong to the same thing that is a being
[1b]    or that to which one belongs is a being
[1c]    the thing itself, to which belongs that thing of which it itself is predicated, is a being.

[2]     intrinsically

[2a]    signified by figures of predication.

        For the ways in which things are said to be are the ways in which
        "being" signifies.

[2b]    the ways in which "being" signifies.
        
        When predicates signify what it is, a quality of it,
        a quantity, a relation, a doing, an affection, a where, a when of it,
        "being" signifies the same thing as each of these.

[3]     true being

        Further, "being" or "is" signify that something is true, and "not to be" that it is not true
        but rather a falsehood -- similarly in the case of affirmation and denial.

        For example, "it is" that Socrates is musical signifies that this is true,
        "it is that Socrates is not pale, that this is true; whereas,
        "it is not" that the diagonal is commensurable signifies that this is false.

[4]      potential and actual being

        Further, "is" and "being" signify the things we mentioned
        on the one hand potentially on the other actually.

What potentially sees and what actually sees is a "seeing thing."
What is capable of using scientific knowledge and
what is using it is "a scientifically knowing one."

But when something is capable and when it is not yet capable is determined elsewhere.
