
Something is said to be a CAPACITY [POTENTIALITY, POWER] when:

[1]     when the starting-point of change or movement is in another thing or
        in a thing in so far as it is other.

[2]     when it is caused by another thing or
        by a thing insofar as it other.

[3]     when there is the capacity to accomplish the thing in question well
        or in accord with deliberate choice.

[4]     when similar to the case of being affected.

[5]     when the states in virtue of which something is wholly
        unaffectable or unchangeable or not easily moved for the worse.

What is said to be CAPABLE:

[1]     when the starting-point of change or movement is in another thing or
        in a thing in so far as it is another.

[2]     when something else has a capacity of this sort with regard to it. 

[3]     when it has the capacity to change into something,
        whether for the worse or the better.

[4]     when neither another thing or the thing itself insofar as it is other,
        has a capacity or a starting-point that is destructive of it.

[5]     when either because the thing might happen to come about or
        not come about or because it might do so well.

INCAPACITY is lack of capability, that is, lack of a starting-point
either generally or by what naturally should have it or even by
what at the time should naturally have already have it.

Some things are also said to be INCAPABLE in virtue of this sort of
incapacity, where others are in another way both capable and incapable, namely
POSSIBLE and IMPOSSIBLE, where an impossible thing is that of which
the contrary is of necessity true.

What in geometry is said to be a POWER is so by metaphorical transference.

Those things are capable have a POWER or or POSSIBLE.

And so the strict definition of the primary sort of capacity will be:
what is a starting-point of change in another thing or
in a thing is so far as it is another.
