
Something is said to be a CAUSE if it is:

[1]     the component from which a thing comes to be

        The bronze of a statue, or the silver of a bowl, and also
        the kinds of these.

[2]     the form or paradigm or the account of the essence, and
        the kinds of these.

        The octave, the ratio 2:1 and number in general,
        and the parts included in the account.

[3]     that from which the change or rest from change first starts

        The person who has deliberated is the cause of the action.
        and the father of his child, and in general the producer is
        cause of the thing produced and the change-maker of the change.

[4]     the end, and this is the for-the-sake-of-which

        Of taking walks, health is the end. For why does he take walks?
        In order that he may be healthy, we say. And in speaking that way,
        we think we have presented the cause. Also anything that comes to be
        as an intermediate means to an end when something else has started
        the movement. In the case of health, making thin, purging, drugs,
        or instruments, since these are all for the sake of the end,
        although they differ from each other in that some are instruments
        and others works.

These then are pretty much how things are said to be a cause,
and it follows both that:

[a]     there are many causes of the same things, and not coincidentally

[b]     things can be causes of each other

Both things, presence and lack, are causes in the way moving things are.



These then are the causes and this is the number of their kind.
But the ways of being causes also falls under comparatively few headings.
Things are said to be causes in many ways, even within the same genus.

Further there are coincidental causes and their kinds.

        As a cause of the statue, in one way Polyclitus, in another a sculptor,
        because the sculptor coincides with Polyclitus. And a human is a cause
        of the statue, or in general an animal, because Polyclitus is a human
        and a human is an animal. And among coincidental causes, some are more
        remote than others, some less. For example, if the pale and the musical
        were said to be causes of the statue, but not Polyclitus or a human only.

Some causes are said to be so potentially other actively.

        As a cause of the house's being built, a builder
        on the one hand and a builder building on the other.

Similarly, the things that cause actions are causes of will.

    This statue, a statue, or in general a likeness, or this bronze, bronze,
    or in general, matter; and similarly in the case of coincidental effects.

Further, both the former and the latter may be said to be
causes in combination.

    Not Polyclitus or sculptor, but Polyclitus the sculptor.

All of these things are just six [i - vi] in number,
each said to be in two [a-b] ways:

[a-i]   the particular
[a-ii]  the kind
[a-iii] the coincidental
[a-iv]  the kind of the coincidental
[a-v]   these as combined
[a-vi]  these as simple

[b]     all actually or potentially, but they differ to the extent
        that what is actual or what is particular exists,
        or does not exist, at the same time as the thing it causes

        This person curing at the same time as this one recovering his health,
        this person building with this building being built. But this is not
        always the case with what are potentially causes, since the house does
        not pass away at the same time as the builder.

