What is said to be NATURE is:

[1]     the coming to be of things that grow
[2]     the first component from which a growing thing grows
[3]     that from which the first movement in each of the beings
        that are by nature is present in it insofar as it is itself

        Those things are said to grow that get their increase in size
        through another thing by making contact and growing together
        or by adhesion. But growing together differs from contact,
        since in the case of contact it is not necessary that 
        any other thing exist besides the contact whereas in the case of
        things that grow together there is some one thing, the same in both,
        which makes them grow together instead of merely making contact,
        and makes them one as regards continuity and quantity,
        although not as regards quality.

[4]     that from which some one of the beings that are by nature
        first is or comes to be first, and which is unshaped and cannot
        be changed from the capacity that belongs to it is said to
        be its nature

        For something is from these when the primary matter is preserved
        throughout. For it is in this way that the elements of the beings
        that are by nature are also said to be their nature,
        some saying fire, some earth, sine air, some water, and others
        some or all of these.

[5]     the substance of the beings that are by nature is said to be their nature

[6]     By metaphorical transference from this case every substance in general is
        said to be a nature, because the nature of the thing is a sort of substance.

Nature in the primary and full way is the substance of things that
have a starting-point of movement within themselves, insofar as
they are themselves. For the matter is said to be nature
because it is receptive of this, and comings to be and growing because
they are movements arising from it. The starting-point of change for the beings
that are by nature, which is in some way a component of them, either potentially
or actually is also nature.
