Something is said to be NECESSARY if:

[1a]    that without which, as a contributing cause, living is impossible

        Breathing and nourishment are necessary for an animal.

[1b]    that without which it is not possible for the good to be or come to be

        Drinking the medicine is necessary in order to not get sick,
        as is sailing to Aegina in order to get the money.

[2]     that which is forced or is force

        What is contrary to the impulse or the deliberate choice
        impedes or tends to hinder. For what is forced is said to be
        necessary and is something that cannot be persuaded,
        and correctly so, since it is contrary to the movement
        that is in accord with deliberate choice
        and in accord with rational calculation.

[3]     that which does not admit of being otherwise to be the way it is

        It is in accord with this sort of necessity that all of the others
        are said in some way to be necessary. For what is forced is said to be necessary
        either to do or to suffer implying that necessity is that thing because of which
        it is not possible for a thing to be otherwise. And similarly in the contributing
        causes of living and the good, since the one case, the good, and the other case,
        living and being, are not possible without certain things
        and that cause is a sort of necessity.

        Demonstration is among the things that are necessary
        because it is not possible for something to be otherwise
        if it has been demonstrated unconditionally.
        The cause of this is the primary things,
        if the thing from which a demonstration proceeds cannot be otherwise.

Of some things, then, another thing is their
cause of their being necessary,
of others, nothing is such a cause, rather,
because of them other things are necessary.
The necessary in the primary and full way is the simple.
For it is not possible for this to be in more than one state,
or even to be both in one state and another,
since it would thereby be in more than one state.
Hence if there are certain things that are eternal and immovable,
there is nothing forced or contrary to nature in them.
