
Things are said to be RELATIVE:

[1a]    as double relative to half and triple relative to one third
[1b]    as multiple relative to submultiple
[1c]    as what exceeds relative to what is exceeded

[2a]    what is capable of heating relative to being heated
[2b]    what is capable of cutting relative to being cut
[2c]    what is capable of acting relative to what is capable of being acted on

[3a]    what is measurable relative to the measure
[3b]    what is scientifically knowable to scientific knowledge
[3c]    what is perceptible to perception

[1]     The first lot are said to be relative numerically,
        whether unconditionally or in a definite way,
        and relative to themselves or to one.

        All these things are said to be relative with reference to number
        and are the attributes of number. As are the equal, the similar, and the same
        since they are said with reference to one.
        For things are the same when their substance is one;
        similar when their quality is one;
        and equal when their quantity is one.
        And one is the starting point and measure of number,
        so that all these are said of things with reference to number,
        but not in the same way.

[2]     Things that can act or be acted on are said to be relative
        with reference to a capacity to act or be acted on and
        to the activations of the capacity. Of numerical relations,
        on the other hand, there are no activations, except in the way
        that has been stated elsewhere but activations with reference
        to movement do not belong to them.

[3]     Things that are said to be relative with reference to a number or a capacity
        are all relative because of being said to be just what they are of another thing,
        not because of the other thing's being relative to them.
        But what is measurable or scientifically knowable or thinkable is
        is said to be relative because of another thing's being said to be
        what it is relative to them. For what is thinkable signifies that
        there can be a thought of it, but the thought is not relative to the
        thing it is the thought of, for then we would have said the same thing twice.
        Similarly sight is the sight of something not of what it is the sight of.
        Instead it is relative to color.

Things that are said to be intrinsically relatives are said to be such
sometimes in these ways, sometimes if their genera are of the relevant sort.
Further those things are said to be relative with reference to which
their possessors are said to be relatives.

Other things are said to be relative coincidentally.
For example, human is a relative because double coincides with it, and 
double is a relative, or pale is, if double and pale coincide in the same thing.
