SECTION III

Actuality

Actuality is the unity of essence and concrete existence;
in it, shapeless essence and unstable appearance
(subsistence without determination and manifoldness without permanence)
have their truth.
Although concrete existence is the immediacy that has proceeded from ground,
it still does not have form explicitly posited in it;
inasmuch as it determines and informs itself, it is appearance;
and in developing this subsistence that otherwise
only is a reflection-into-other into an immanent reflection,
it becomes two worlds, two totalities of content,
one determined as reflected into itself and
the other as reflected into other.
But the essential relation exposes the formality of
their connection, and the consummation of the latter is the relation of
the inner and the outer in which the content of both is equally only one
identical substrate and only one identity of form.
this identity has come about also in regard to form,
the form determination of their difference is sublated, and
that they are one absolute totality is posited.
This unity of the inner and outer is absolute actuality.
But this actuality is, first, the absolute as such
(in so far as it is posited as a unity in which the
form has sublated itself)
making itself into the empty or external distinction of an outer and inner.
Reflection relates to this absolute as external to it;
it only contemplates it rather than being its own movement.
But it is essentially this movement and is, therefore,
as the absolute’s negative turning back into itself.
Second, it is actuality proper.
Actuality, possibility, and necessity constitute
the formal moments of the absolute, or its reflection.
Third, the unity of the absolute and its reflection is the absolute relation,
or rather the absolute as relation to itself, substance.

CHAPTER 1

The absolute

The simple solid identity of the absolute is indeterminate, or
rather, every determinateness of essence and concrete existence,
or of being in general as well as of reflection,
has dissolved itself into it.
Accordingly, the determining of what is the absolute appears to be a negating,
and the absolute itself appears only as the negation of all predicates, as the void.
But since it must equally be spoken of as the position of all predicates,
it appears as the most formal of contradictions.
In so far as that negating and this positing belong to external reflection,
what we have is a formal, unsystematic dialectic that has an easy time
picking up a variety of determinations here and there, and is
just as at ease demonstrating, on the one hand, their finitude and relativity, as
declaring, on the other, that the absolute, which it vaguely envisages as totality,
is the dwelling place of all determinations,
yet is incapable of raising either the positions or the negations to a true unity.
The task is indeed to demonstrate what the absolute is.
But this demonstration cannot be either
a determining or an external reflection by virtue of which determinations
of the absolute would result, but is rather the exposition of the absolute,
more precisely the absolute’s own exposition, and only a displaying of what it is.

A. THE EXPOSITION OF THE ABSOLUTE

The absolute is not just being, nor even essence. The former is the first
unreflected immediacy; the latter, the reflected immediacy; further, each
is explicitly a totality, but a determinate totality. Being emerges in essence
as concrete existence, and the connection of being and essence develops
into the relation of inner and outer. The inner is essence, but as a totality
whose essential determination is to be referred to being and to be being
immediately. The outer is being, but with the essential determination of
being immediately connected with reflection and, equally, in a relationless
identity with essence. The absolute itself is the absolute unity of the two;
it is that which constitutes in general the ground of the essential relation
which, as only relation, has yet to return into this its identity and whose
ground is not yet posited.
It follows that the determination of the absolute is to be absolute form,
but at the same time not as an identity whose moments only are simple
determinacies, but, on the contrary, as an identity whose moments are each
explicitly the totality and hence, indifferent with respect to the form, the
complete content of the whole. But, conversely, the absolute is absolute
content in such a way that this content, which is as such indifferent
plurality, explicitly has the negative connection of form by virtue of which
its manifold is only one substantial identity.
Thus the identity of the absolute is for this reason absolute identity,
because each of its parts is itself the whole or each determinateness is the
totality, that is, because determinateness has become as such a thoroughly
transparent reflective shine, a difference that has disappeared in its posited-
ness. Essence, concrete existence, the world existing in itself, whole, parts, force –
these reflected determinations appear to representation as true being valid
in and for itself; but against them the absolute is the ground into which
they have foundered. – Because in the absolute the form is now only simple
self-identity, the absolute does not determine itself, for the determination is
a difference of form which is valid as such from the start. But because the
absolute at the same time contains every difference and form determination
in general, or because it is itself absolute form and reflection, the difference
of content must also come into it. But the absolute itself is the absolute
identity; to be this identity is its determination, for the manifoldness of the
world-in-itself and of the phenomenal world has all been sublated in it. –
In the absolute itself there is no becoming, since the absolute is not being;
nor does the absolute determine itself reflectively, for it is not the essence
which determines itself only inwardly; and it also does not externalize itself,
for it is the identity of inner and outer. – But in this way the movement of
reflection stands over against its absolute identity. The movement is sub-
lated in this identity and is thus only its inner; but consequently its outer.
– At first, therefore, the movement consists only in sublating its act in the
absolute. It is the beyond of the manifold differences and determinations
and of their movement, a beyond that lies at the back of the absolute.
It is thus the negative exposition of the absolute earlier alluded to. 1 – In
its true presentation, this exposition is the preceding whole of the logical
movement of the spheres of being and essence, the content of which has
not been gathered in from outside as something given and contingent; nor
has it been sunk into the abyss of the absolute by a reflection external to
it; on the contrary, it has determined itself within it by virtue of its inner
necessity, and, as being’s own becoming and as the reflection of essence, has
returned into the absolute as into its ground.
But this exposition has itself also a positive side, for in foundering to
the ground the finite demonstrates that its nature is to be referred to the
absolute, or to contain the absolute within. However, this side is not as
much the positive exposition of the absolute as it is rather the exposition
of the determinations, namely that these have the absolute for their abyss,
but also for their ground, 2 or that that which imparts subsistence to them,
to their reflective shine, is the absolute itself. – Being as shine is not nothing
but reflection, reference to the absolute; or it is a shine inasmuch as that
which shines in it is the absolute. This positive exposition thus halts the
finite just before its disappearing: it considers it an expression and a copy
of the absolute. But this transparency of the finite that lets only the absolute
transpire through it ends up in complete disappearance, for there is nothing
in the finite which would retain for it a difference over against the absolute;
as a medium, it is absorbed by that through which it shines.
This positive exposition of the absolute is therefore itself only a reflec-
tive shine, for the true positive, that which contains the exposition and the
expounded content, is the absolute itself. Whatever the further determina-
tions that may occur, the form in which the absolute reflectively shines is a
nullity which the exposition gathers up from outside and in which it gains
for itself a starting point for its activity. Any such determination has in the
absolute, not its beginning but its end. This expository process, therefore,
though it is an absolute act because of its reference to the absolute into
which it returns, is not so at its starting point which is a determination
external to the absolute.
But in actual fact the exposition of the absolute is the absolute’s own
doing, an act that begins from itself and arrives at itself. The absolute, only
as absolute identity, is absolute in a determined guise, that is, as identical
absolute; it is posited as such by reflection over against opposition and
manifoldness; or it is only the negative of reflection and determination in
general. – It is not just the exposition of the absolute which is therefore
something incomplete, but this absolute itself which is only arrived at. Or
again, the absolute which is only as absolute identity is only the absolute of
an external reflection. It is, therefore, not the absolutely absolute but the
absolute in a determination, or it is attribute.
But the absolute is not attribute just because it is the subject matter of
an external reflection and is consequently something determined by it. –
Or, reflection is not only external to it; but, precisely because it is external
to it, it is immediately internal to it. The absolute is absolute only because
it is not abstract identity but is the identity of being and essence, or the
identity of the inner and the outer. It is therefore itself the absolute form
that makes it reflectively shine within itself and determines it as attribute.

B. THE ABSOLUTE ATTRIBUTE

The expression which we have used, “the absolute absolute,” denotes the
absolute which in its form has returned back into itself or whose form is
equal to its content. The attribute is just the relative absolute, a combination
which only signifies the absolute in a form determination. For at first, before
its complete exposition, the form is only internally or, which is the same,
only externally; it is at first determinate form in general or negation in
general. But because form is at the same time as the form of the absolute,
the attribute is the whole content of the absolute; it is the totality which
earlier appeared as a world, or as one of the sides of the essential relation,
each of which is itself the whole. But both worlds, the phenomenal world
and the world that exists in and for itself, were supposed to be opposed to
each other in their essence. Each side of the essential relation was indeed
equal to the other – the whole as much as the parts, the expression of force
the same content as force itself, and the outer everywhere the same as the
inner. But these sides were at the same time supposed each to have still
an immediate subsistence of its own, the one side as existent immediacy
and the other as reflected immediacy. In the absolute, on the contrary,
these different immediacies have been reduced to a reflective shine, and the
totality that the attribute is is posited as its true and single subsistence, while
the determination in which it is is posited as unessential subsistence.
The absolute is attribute because, as simple absolute identity, it is in
the determination of identity; now to the determination as such other
determinations can be attached, for instance, also that there are several
attributes. But because absolute identity has only this meaning, that not
only all determinations have been sublated but that reflection itself has
also sublated itself, all determinations are thus posited in it as sublated.
Or the totality is posited as absolute totality. Or again, the attribute has
the absolute for its content and subsistence and, consequently, its form
determination by which it is attribute is also posited, posited immediately
as mere reflective shine; the negative is posited as negative. The positive
reflective shine that the exposition gives itself through the attribute – in
that it does not take the finite in its limitation as something that exists in
and for itself but dissolves its subsistence into the absolute and expands
it into attribute – sublates precisely this, that the attribute is attribute; it
sinks it and its differentiating act into the simple absolute.
But since reflection thus reverts from its differentiating act only to the
identity of the absolute, it has not at the same time left its externality behind
and has not arrived at the true absolute. It has only reached the indetermi-
nate, abstract identity, which is to say, the identity in the determinateness
of identity. – Or, since reflection determines the absolute into attribute as
inner form, this determining is something still distinct from externality;
the inner determination does not penetrate the absolute; the attribute’s
expression, as something merely posited, is to disappear into the absolute.
The form by virtue of which the absolute would be attribute, whether
it is taken as outer or inner, is therefore posited as something null in itself,
an external reflective shine, or a mere way and manner.

C. THE MODE OF THE ABSOLUTE

The attribute is first the absolute in simple self-identity. Second, it is nega-
tion, a negation which is as such formal immanent reflection. These two
sides constitute at first the two extremes of the attribute, the middle term
of which is the attribute itself, since it is both the absolute and the deter-
minateness. – The second of these extremes is the negative as negative, the
reflection external to the absolute. – Or inasmuch as the negative is taken
as the inner of the absolute and its own determination is to posit itself
as mode, it is then the self-externality of the absolute, the loss of itself in
the changeability and contingency of being, its having passed over into its
opposite without turning back into itself, the manifoldness of form and
content determinations that lacks totality. –
But the mode, the externality of the absolute, is not just this. It is
rather externality posited as externality, a mere way and manner, hence the
reflective shine as reflective shine, or the reflection of form into itself; hence,
the self-identity which is the absolute. In actual fact, therefore, the absolute
is first posited as absolute identity only in the mode; it is what it is, namely
self-identity, only as self-referring negativity, as reflective shining which is
posited as reflective shining.
Hence, in so far as the exposition of the absolute begins from its absolute
identity and passes over to the attribute and from there to the mode, it
has therein exhaustively run through its moments. But first, in this course
it does not just behave negatively towards these determinations; its act is
rather the reflective movement itself, and it is only as such a movement
that the absolute truly is absolute identity. – Second, the exposition does
not thereby deal with mere externality, and the mode is not only the most
external externality. Rather, since the mode is reflective shine as shine, it is
an immanent turning back, the self-dissolving reflection, and it is in being
this reflection that the absolute is absolute being. – Third, the reflective
act of exposition seems to begin from its own determinations and from
something external, to take up the modes or even the determinations of
the attribute as if they were found outside the absolute and its contri-
bution were only to reduce them to undifferentiated identity. But it has
in fact found the determinateness from which it begins in the absolute
itself. For as first undifferentiated identity, the absolute is itself only the
determinate absolute, or attribute, because it is the unmoved, still unre-
flected absolute. This determinateness, since it is determinateness, belongs
to the reflective movement, and it is through this movement alone that
the absolute is determined as the first identity; through it alone that it
has absolute form and does not just exist as self-equal but posits itself as
self-equal.
Accordingly the true meaning of mode is that it is the absolute’s own
reflective movement; it is a determining by virtue of which the absolute
would become, not an other, but what it already is; a transparent externality
which is a pointing to itself; a movement out of itself, but in such a way
that being outwardly is just as much inwardness, and consequently equally
a positing which is not mere positedness but absolute being.
When therefore one asks for a content of the exposition, for what the
absolute manifests, the reply is that the distinction of form and content
in the absolute has been dissolved; or that just this is the content of the
absolute, that it manifests itself. The absolute is the absolute form which
in its diremption of itself is utterly identical with itself, is the negative
as negative or the negative that rejoins itself and in this way alone is the
absolute self-identity which equally is indifferent towards its distinctions or
is absolute content. The content is therefore only this exposition itself.
As this self-bearing movement of exposition, as a way and manner which
is its absolute identity with itself, the absolute is expression, not of an inner,
nor over against an other, but simply as absolute manifestation of itself for
itself. Thus it is actuality.

CHAPTER 2

Actuality

The absolute is the unity of inner and outer as a first implicitly existent unit.
The exposition appeared as an external reflection which, for its part, has the
immediate as something it has found, but it equally is its movement and
the reference connecting it to the absolute and, as such, it leads it back
to the latter, determining it as a mere “way and manner.” But this “way
and manner” is the determination of the absolute itself, namely its first
identity or its mere implicitly existent unity. And through this reflection, not
only is that first in-itself posited as essenceless determination, but, since the
reflection is negative self-reference, it is through it that the in-itself becomes
a mode in the first place. It is this reflection that, in sublating itself in its
determinations and as a movement which as such turns back upon itself,
is first truly absolute identity and, at the same time, the determining of
the absolute or its modality. The mode, therefore, is the externality of the
absolute, but equally so only its reflection into itself; or again, it is the
absolute’s own manifestation, so that this externalization is its immanent
reflection and therefore its being in-and-for-itself.
So, as the manifestation that it is nothing, that it has no content, save
to be the manifestation of itself, the absolute is absolute form. Actuality is
to be taken as this reflected absoluteness. Being is not yet actual; it is the
first immediacy; its reflection is therefore becoming and transition into an
other; or its immediacy is not being-in-and-for-itself. Actuality also stands
higher than concrete existence. It is true that the latter is the immediacy
that has proceeded from ground and conditions, or from essence and
its reflection. In itself or implicitly, it is therefore what actuality is, real
reflection; but it is still not the posited unity of reflection and immediacy.
Hence concrete existence passes over into appearance as it develops the
reflection contained within it. It is the ground that has foundered to the
ground; its determination, its vocation, 14 is to restore this ground, and
therefore it becomes essential relation, and its final reflection is that its
immediacy be posited as immanent reflection and conversely. This unity,
in which concrete existence or immediacy and the in-itself, the ground or
the reflected, are simply moments, is now actuality. The actual is therefore
manifestation. It is not drawn into the sphere of alteration by its externality,
nor is it the reflective shining of itself in an other. It just manifests itself, and
this means that in its externality, and only in it, it is itself, that is to say,
only as a self-differentiating and self-determining movement.
Now in actuality as this absolute form, the moments only are as sublated
or formal, not yet realized; their differentiation thus belongs at first to
external reflection and is not determined as content.
Actuality, as itself immediate form-unity of inner and outer, is thus in
the determination of immediacy as against the determination of immanent
reflection; or it is an actuality as against a possibility. The connection of the
two to each other is the third, the actual determined both as being reflected
into itself and as this being immediately existing. This third is necessity.
But first, since the actual and the possible are formal distinctions, their
connection is likewise only formal, and consists only in this, that the one
just like the other is a positedness, or in contingency.
Second, because in contingency the actual as well as the possible are a
positedness, because they have retained their determination, real actuality
now arises, and with it also real possibility and relative necessity.
Third, the reflection of relative necessity into itself yields absolute neces-
sity, which is absolute possibility and actuality.

A. CONTINGENCY OR FORMAL ACTUALITY, POSSIBILITY, AND NECESSITY

1. Actuality is formal inasmuch as, as a first actuality, it is only immediate,
unreflected actuality, and hence is only in this form determination but not
as the totality of form. And so it is nothing more than a being, or concrete
existence in general. But because by essence it is not mere concrete existence
but is the form-unity of the in-itselfness or inwardness and externality, it
immediately contains in-itselfness or possibility. What is actual is possible.
2. This possibility is actuality reflected into itself. But this reflected-
ness, itself a first, is equally something formal and consequently only the
determination of self-identity or of the in-itself in general.
But because the determination is here totality of form, this in-itself is
determined as sublated or essentially only with reference to actuality; as the
negative of actuality, it is posited as negative. Possibility entails, therefore,
two moments. It has first the positive moment of being a being-reflected-
into-itself. But this being-reflected-into-itself, since in the absolute form it
is reduced to a moment, no longer has the value of essence but has rather
the negative meaning that possibility is (in a second moment) something
deficient, that it points to an other, to actuality, and is completed in this
other.
According to the first, merely positive side, possibility is therefore the
mere form determination of self-identity, or the form of essentiality. As such
it is the relationless, indeterminate receptacle of everything in general. –
In this formal sense of possibility, everything is possible that does not contract
itself; the realm of possibility is therefore limitless manifoldness. But every
manifold is determined in itself and as against an other: it possesses negation
within. Indifferent diversity passes over as such into opposition; but opposi-
tion is contradiction. Therefore, all things are just as much contradictory
and hence impossible.
– When we therefore say of something that “it is possible,” this purely
formal assertion is just as superficial and empty as the principle of contra-
diction, and any content that we put into it, “A is possible,” says no more
than “A is A.” Left undeveloped, this content has the form of simplicity;
only after being resolved into its determinations, does difference emerge
within it. To the extent that we stop at that simple form, the content
remains something self-identical and hence a possible. But we do not say
anything by it, just as we do not with the principle of identity.
Yet the possible amounts to more than just the principle of identity.
The possible is reflected immanent reflectedness; or the identical simply as a
moment of the totality, hence also as determined not to be in itself; it therefore
has the second determination of being only a possible and the ought-to-be of
the totality of form. Without this ought-to-be, possibility is essentiality as
such; but the absolute form entails this, that essence itself is only a moment
and that it has no truth without being. Possibility is this mere essentiality,
but so posited as to be only a moment, to be disproportionate with respect
to the absolute form. It is the in-itself, determined as only a posited or,
equally, as not to be in itself. –
Internally, therefore, possibility is contradiction, or it is impossibility.
This finds expression at first in this way, that possibility as form deter-
mination posited as sublated possesses a content in general. As possible, this
content is an in-itself which is at the same time something sublated or an
otherness. But because this content is only a possible, an other opposite to it
is equally possible. “A is A”; then, too, “– A is – A.” These two statements
each express the possibility of its content determination. But, as identical
statements, they are indifferent to each other; that the other is also added,
is not posited in either. Possibility is the connection comparing the two; as
a reflection of the totality, it implies that the opposite also is possible. It is
therefore the ground for drawing the connection that, because A equals A, –
A also equals – A; entailed in the possible A there is also the possible not-A,
and it is this reference itself connecting them which determines both as
possible.
But this connection, in which the one possible also contains its other, is
as such a contradiction that sublates itself. Now, since it is determined to
be reflective and, as we have just seen, reflectively self-sublating, it is also
therefore an immediate and it consequently becomes actuality.
3. This actuality is not the first actuality but reflected actuality, posited
as unity of itself and possibility. What is actual is as such possible; it is in
immediate positive identity with possibility; but the latter has determined
itself as only possibility; consequently the actual is also determined as only a
possible. And because possibility is immediately contained in actuality, it is
immediately in it as sublated, as only possibility. Conversely, actuality which
is in unity with possibility is only sublated immediacy; or again, because
formal actuality is only immediate first actuality, it is only a moment, only
sublated actuality, or only possibility.
With this we also have a more precise expression of the extent to which
possibility is actuality. Possibility is not yet all actuality; there has been no
talk yet of real and absolute actuality. It is still only the possibility as it first
presented itself, namely the formal possibility that has determined itself
as being only possibility and hence the formless actuality which is only
being or concrete existence in general. Everything possible has therefore in
general a being or a concrete existence.
This unity of possibility and actuality is contingency. – The contingent
is an actual which is at the same time determined as only possible, an
actual whose other or opposite equally is. This actuality is, therefore, mere
being or concrete existence, but posited in its truth as having the value of
a positedness or a possibility. Conversely, possibility is immanent reflection
or the in-itself posited as positedness; what is possible is an actual in this
sense of actuality, that it has only as much value as contingent actuality; it
is itself something contingent.
The contingent thus presents these two sides. First, in so far as it has
possibility immediately in it, or, what is the same, in so far as this possibility
is sublated in it, it is not positedness, nor is it mediated, but is immediate
actuality; it has no ground. – Because this immediate actuality pertains also
to the possible, the latter is determined no less than the actual as contingent
and is likewise groundless.
But, second, the contingent is the actual as what is only possible, or as
a positedness; thus the possible also, as formal in-itself, is only positedness.
Consequently, the two are both not in and for themselves but have their
immanent reflection in an other, or they do have a ground.
The contingent thus has no ground because it is contingent; and for
that same reason it has a ground, because it is contingent.
It is the posited, immediate conversion of inner and outer, or of
immanently-reflected-being and being, each into the other – posited,
because possibility and actuality both have this determination in them
by being moments of the absolute form. – So actuality, in its immedi-
ate unity with possibility, is only concrete existence and is determined as
groundless, something only posited or only possible; or, as reflected and deter-
mined over against possibility, it is separated from possibility, from imma-
nent reflectedness, and then, too, is no less immediately only a possible. –
Likewise possibility, as simple in-itself, is something immediate, only an
existent in general; or, opposed to actuality, it equally is an in-itself without
actuality, only a possible, but, for that very reason, again only a concrete,
not immanently reflected, existence in general.
This absolute restlessness of the becoming of these two determinations is
contingency. But for this reason, because each determination immediately
turns into the opposite, in this opposite each equally rejoins itself, and this
identity of the two, of each in the other, is necessity.
The necessary is an actual; as such it is immediate, groundless; but it
equally has its actuality through an other or in its ground and is at the
same time the positedness of this ground and its reflection into itself; the
possibility of the necessary is a sublated one. The contingent is therefore
necessary because the actual is determined as a possible; its immediacy
is consequently sublated and is repelled into the ground or the in-itself,
and into the grounded, equally because its possibility, this ground-grounded-
connection, is simply sublated and posited as being. What is necessary is,
and this existent is itself the necessary. At the same time it is in itself; this
immanent reflection is an other than that immediacy of being, and the
necessity of the existent is an other. Thus the existent is not the necessary;
but this in-itself is itself only positedness; it is sublated and itself immediate.
And so actuality, in that from which it is distinguished, in possibility, is
identical with itself. As this identity, it is necessity.

B. RELATIVE NECESSITY OR REAL ACTUALITY, POSSIBILITY, AND NECESSITY

1. The necessity which has resulted is formal because its moments are formal,
that is, simple determinations which are a totality only as an immediate
unity, or as an immediate conversion of the one into the other, and thus
lack the shape of self-subsistence. – The unity in this formal necessity is
therefore simple at first, and indifferent to its differences. As the immediate
unity of the form determinations, this necessity is actuality, but an actuality
which, since its unity is now determined as indifferent to the difference of
the form determinations, has a content. This content as an indifferent
identity contains the form also as indifferent, that is, as a mere variety of
determinations, and is a manifold content in general. This actuality is real
actuality.
Real actuality is as such at first the thing of many properties, the con-
cretely existing world; but it is not the concrete existence that dissolves into
appearance but, as actuality, it is at the same time an in-itself and immanent
reflection; it preserves itself in the manifoldness of mere concrete existence;
its externality is an inner relating only to itself. What is actual can act; some-
thing announces its actuality by what it produces. Its relating to an other
is the manifestation of itself, and this manifestation is neither a transition
(the immediate something refers to the other in this way) nor an appearing
(in this way the thing only is in relation to an other); it is a self-subsistent
which has its immanent reflection, its determinate essentiality, in another
self-subsistent.
Now real actuality likewise has possibility immediately present in it. It
contains the moment of the in-itself; but, since it is in the first instance
only immediate unity, it is in one of the determinations of form and hence
distinguished, as immediate existent, from the in-itself or possibility.
2. This possibility, as the in-itself of real actuality, is itself real possibility, at
first the in-itself full of content. – Formal possibility is immanent reflection
only as abstract identity, the absence of contradiction in a something. But
when we delve into the determinations, the circumstances, the conditions
of a fact in order to discover its possibility, we do not stop at this formal
possibility but consider its real possibility.
This real possibility is itself immediate concrete existence, but no longer
because possibility as such, as a formal moment, is immediately its opposite,
a non-reflected actuality, but because this determination pertains to it by
the very fact of being real possibility. The real possibility of a fact is therefore
the immediately existent manifoldness of circumstances that refer to it.
This manifoldness of existence is therefore indeed both possibility and
actuality, but their identity is at first only the content which is indifferent to
these form determinations; they therefore constitute the form, determined
as against their identity. – Or the immediate real actuality, because it is
immediate, is determined as against its possibility; as this determinate and
hence reflected actuality, it is real possibility. This real possibility is now
indeed the posited whole of the form, but of the form in the determinateness
of actuality as formal or immediate and equally of possibility as the abstract
in-itself. This actuality, therefore, which constitutes the possibility of a fact,
is not its own possibility but the in-itself of an other actual; itself, it is the
actuality that ought to be sublated, the possibility as only possibility. – Real
possibility thus constitutes the totality of conditions, a dispersed actuality
which is not reflected into itself but is determined to be the in-itself of an
other and intended in this determination to return to itself.
What is really possible is, therefore, something formally identical accord-
ing to its in-itself, free of contradiction because of its simple content deter-
mination; but, as self-identical, this something must also not contradict
itself according to its developed and differentiated circumstances and all
else connected with it. But, secondly, because it is manifold in itself and in
manifold connection with others, and variety inherently passes over into
opposition, it is contradictory. Whenever a possibility is in question, and
the issue is to demonstrate its contradiction, one need only fasten on to
the multiplicity that it contains as content or as its conditioned concrete
existence, and from this the contradiction will easily be discovered. – And
this contradiction is not just a function of comparing; on the contrary, the
manifold of concrete existence is in itself this, to sublate itself and to founder
to the ground: in this it explicitly has the determination of being only a
possibility. – Whenever all the conditions of a fact are completely present,
the fact is actually there; the completeness of the conditions is the totality
as in the content, and the fact is itself this content determined as being
equally actual as possible. In the sphere of the conditioned ground, the
conditions have the form (that is, the ground or the reflection that stands
on its own) outside them, and it is this form that makes them moments
of the fact and elicits concrete existence in them. 16 Here, on the contrary,
the immediate actuality is not determined to be condition by virtue of a
presupposing reflection, but the supposition is rather that the immediate
actuality is itself the possibility.
In self-sublating real possibility, it is a twofold that is now sub-
lated; for this possibility is itself the twofold of actuality and possibility.
(1) The actuality is formal, or is a concrete existence which appeared to
subsist immediately, and through its sublating becomes reflected being, the
moment of an other, and thus comes in possession of the in-itself. (2) That
concrete existence was also determined as possibility or as the in-itself, but
of an other. As it sublates itself, this in-itself of the other is also sublated
and passes over into actuality. – This movement of self-sublating real pos-
sibility thus produces the same moments that are already present, but each as
it comes to be out of the other; in this negation, therefore, the possibility
is also not a transition but a self-rejoining. – In formal possibility, if some-
thing was possible, then an other than it, not itself, was also possible. Real
possibility no longer has such an other over against it, for it is real in so
far as it is itself also actuality. Therefore, as its immediate concrete existence,
the circle of conditions, sublates itself, it makes itself into the in-itselfness
which it already is, namely the in-itself of an other. And conversely, since
its moment of in-itselfness thereby sublates itself at the same time, it
becomes actuality, hence the moment which it likewise already is. – What
disappears is consequently this, that actuality was determined as the pos-
sibility or the in-itself of an other, and, conversely, the possibility as an
actuality which is not that of which it is the possibility.
3. The negation of real possibility is thus its self-identity; inasmuch as in
its sublating it is thus within itself the recoiling of this sublating, it is real
necessity.
What is necessary cannot be otherwise; but what is only possible can be,
for possibility is the in-itself which is only positedness and hence essentially
otherness. Formal possibility is this identity as transition into the other as
such; but real possibility, since it has the other moment of actuality within
it, is already itself necessity. Hence what is really possible can no longer
be otherwise; under the given conditions and circumstances, nothing else
can follow. Real possibility and necessity are, therefore, only apparently
distinguished; theirs is an identity that does not first come to be but is
already presupposed at their base. Real possibility is therefore a connection
full of content, for the content is that identity, existing in itself, which is
indifferent to form.
But this necessity is at the same time relative. – For it has a presupposition
from which it begins; it takes its start from the contingent. For the real
actual is as such the determinate actual, and first has its determinateness as
immediate being in that it is a multiplicity of concretely existing circum-
stances; but this immediate being as determinateness is also the negative of
itself, is an in-itself or possibility and so real possibility. As this unity of the
two moments, it is the totality of form, but a totality which is still external
to itself; it is the unity of possibility and actuality in such a way that (1) the
manifold concrete existence is possibility immediately or positively: it is a
possible, something self-identical as such, because it is an actual; (2) inas-
much as this possibility of concrete existence is posited, it is determined as
only possibility, as the immediate conversion of actuality into its opposite –
or as contingency. Hence this possibility which immediate actuality has
within in so far as it is condition, is only the in-itself or the possibility of an
other. Because this in-itself, as shown, 17 sublates itself and this positedness
is itself posited, real possibility becomes indeed necessity; but this necessity
thus begins from that unity of the possible and the actual which is not yet
reflected into itself – this presupposing and the movement which turns back
unto itself are still separate – or necessity has not yet determined itself out of
itself into contingency.
The relativity of real possibility is manifested in the content by the fact
that the latter is at first only the identity indifferent to form, is therefore
distinct from it and a determinate content in general. A necessary reality is
for this reason any limited actuality which, because of its limitation, is in
some other respect also only something contingent.
In actual fact, therefore, real necessity is in itself also contingency. – This
first becomes apparent because real necessity, although something necessary
according to form, is still something limited according to content, and
derives its contingency through the latter. But this contingency is to be
found also in the form of real necessity because, as shown, 18 real possibility
is the necessary only in itself, but as posited it is the mutual otherness of
actuality and possibility. Real necessity thus contains contingency; it is the
turning back into itself from the restless being-the-other-of-each-other of
actuality and possibility, but not the turning back from itself to itself.
In itself, therefore, we have here the unity of necessity and contingency;
this unity is to be called absolute actuality.

C. ABSOLUTE NECESSITY

Real necessity is determinate necessity; formal necessity does not yet have
any content and determinateness in it. The determinateness of necessity
consists in its having its negation, contingency, within it. This is how it
has shown itself to be.
But in its first simplicity this determinateness is actuality; determinate
necessity is therefore immediate actual necessity. This actuality which is
itself as such necessary, since it contains necessity as its in-itself, is absolute
actuality – an actuality which can no longer be otherwise, for its in-itself is
not possibility but necessity itself.
But because this actuality is posited to be absolute, that is to say, to
be itself the unity of itself and possibility, it is consequently only an empty
determination, or it is contingency. – This emptiness of its determination
makes it into a mere possibility, one which can just as well be an other and
is determined as possibility. But this possibility is itself absolute possibility,
for it is precisely the possibility of being equally determined as possibility
and actuality. For this reason, because it is this indifference towards itself,
it is posited as empty, contingent determination.
Thus real necessity not only contains contingency implicitly, but the
latter also becomes in it; but this becoming, as externality, is itself only the
in-itself of the necessity, because it is only an immediate determinateness. But
it is not only this but the necessity’s own becoming – or the presupposition
which it had is its own positing. For as real necessity, it is the sublatedness
of actuality into possibility and of possibility into actuality; because it is
this simple conversion of one of these moments into the other, it is also
their positive unity, for in the other each rejoins itself. And so it is actuality,
yet an actuality which is nothing but this rejoining of form with itself. Its
negative positing of these moments is thereby itself the presupposing or the
positing of itself as sublated, or the positing of immediacy.
But it is precisely in this positing that this actuality is determined as the
negative; it rejoins itself from the actuality which was real possibility; this
new actuality thus comes to be only out of its in-itself, out of the negation
of itself. – Consequently, it is at the same time immediately determined
as possibility, as mediated by virtue of its negation. But accordingly, this
possibility is immediately nothing but this mediating in which the in-itself,
namely the possibility itself and the mediating, both in the same manner,
are positedness. – Thus it is necessity which is equally the sublating of this
positedness, or the positing of immediacy and of the in-itself, just as in this
very sublating it is the determining of it as positedness. It is necessity itself,
therefore, that determines itself as contingency: in its being it repels itself
from itself, in this very repelling has only returned to itself, and in this
turning back which is its being has repelled itself from itself.
Thus has form pervaded in its realization all its distinctions; it has made
itself transparent and, as absolute necessity, is only this simple self-identity of
being in its negation, or in essence. – The distinction itself of content and
form has thus equally vanished; for that unity of possibility in actuality
and actuality in possibility is the form which in its determinateness or
in positedness is indifferent towards itself: it is the fact full of content on
which the form of necessity externally ran its course. But necessity is thus
this reflected identity of the two determinations as indifferent to them,
and hence the form determination of the in-itself as against the positedness,
and this possibility constitutes the limitation of the content which real
necessity had. The resolution of this difference is however the absolute
necessity whose content is this difference which in this necessity penetrates
itself.
Absolute necessity is therefore the truth in which actuality and possibility
in general as well as formal and real necessity return. – As we have just
seen, it is being which in its negation, in essence, refers itself to itself and
is being. It is equally simple immediacy or pure being and simple immanent
reflection or pure essence; it is this, that the two are one and the same. – The
absolutely necessary only is because it is; it otherwise has neither condition
nor ground. – But it equally is pure essence, its being the simple immanent
reflection; it is because it is. As reflection, it has a ground and a condition
but has only itself for this ground and condition. It is in-itself, but its
in-itself is its immediacy, its possibility is its actuality. – It is, therefore,
because it is; as the rejoining of being with itself, it is essence; but because
this simple is equally immediate simplicity, it is being.
Absolute necessity is thus the reflection or form of the absolute, the unity
of being and essence, simple immediacy which is absolute negativity. On
the one hand, therefore, its differences are not like the determinations
of reflection but an existing manifoldness, a differentiated actuality in the
shape of others independently subsisting over against each other. On the
other hand, since its connection is that of absolute identity, it is the abso-
lute conversion of its actuality into its possibility and its possibility into
its actuality. – Absolute necessity is therefore blind. On the one hand,
the two different terms determined as actuality and possibility have the
shape of immanent reflection as being; they are therefore free actualities,
neither of which reflectively shines in the other, nor will either allow in
it a trace of its reference to the other; grounded in itself, each is inher-
ently necessary. Necessity as essence is concealed in this being; the reciprocal
contact of these actualities appears, therefore, as an empty externality; the
actuality of the one in the other is the possibility which is only possibility, con-
tingency. For being is posited as absolutely necessary, as the self-mediation
which is the absolute negation of mediation-through-other, or being which
is identical only with being; consequently, an other that has actuality
in being, is therefore determined as something merely possible, as empty
positedness.
But this contingency is rather absolute necessity; it is the essence of those
free, inherently necessary actualities. This essence is averse to light, because
there is no reflective shining in these actualities, no reflex – because they
are grounded purely in themselves, are shaped for themselves, manifest
themselves only to themselves – because they are only being. – But their
essence will break forth in them and will reveal what it is and what they are.
The simplicity of their being, their resting just on themselves, is absolute
negativity; it is the freedom of their reflectionless 20 immediacy. This negative
breaks forth in them because being, through this same negativity which
is its essence, is self-contradiction; it will break forth against this being in
the form of being, hence as the negation of those actualities, a negation
absolutely different from their being; it will break forth as their nothing, as an
otherness which is just as free towards them as their being is free. – Yet this
negative was not to be missed in them. In their self-based shape they are
indifferent to form, are a content and consequently different actualities and a
determinate content. This content is the mark that necessity impressed upon
them by letting them go free as absolutely actual – for in its determination
it is an absolute turning back into itself. It is the mark to which necessity
appeals as witness to its right, and, overcome by it, the actualities now
perish. This manifestation of what determinateness is in its truth, that it is
negative self-reference, is a blind collapse into otherness; in the sphere of
immediate existence, the shining or the reflection that breaks out in it is a
becoming, a transition of being into nothing. But, conversely, being is equally
essence, and becoming is reflection or a shining. Thus the externality is its
inwardness; their connection is one of absolute identity; and the transition
of the actual into the possible, of being into nothing, is a self-rejoining;
contingency is absolute necessity; it is itself the presupposing of that first
absolute actuality.
This identity of being with itself in its negation is now substance. It is this
unity as in its negation or as in contingency; and so, as relation to itself, it
is substance. The blind transition of necessity is rather the absolute’s own
exposition, its movement in itself which, in its externalization, reveals itself
instead.

CHAPTER 3

The absolute relation

Absolute necessity is not so much the necessary, even less a necessary, but
necessity – being simply as reflection. It is relation because it is a distin-
guishing whose moments are themselves the whole totality of necessity, and
therefore subsist absolutely, but do so in such a way that their subsisting is
one subsistence, and the difference only the reflective shine of the movement
of exposition, and this reflective shine is the absolute itself. – Essence as
such is reflection or a shining; as absolute relation, however, essence is the
reflective shine posited as reflective shine, one which, as such self-referring,
is absolute actuality. – The absolute, first expounded by external reflection,
as absolute form or as necessity now expounds itself; this self-exposition is
its self-positing, and is only this self-positing. – Just as the light of nature
is not a something, nor is it a thing, but its being is rather only its shining,
so manifestation is self-identical absolute actuality.
The sides of the absolute relation are not, therefore, attributes. In the
attribute the absolute reflectively shines only in one of its moments, as
in a presupposition that external reflection has simply assumed. But the
expositor of the absolute is the absolute necessity which, as self-determining,
is identical with itself. Since this necessity is the reflective shining posited
as reflective shining, the sides of this relation, because they are as shine,
are totalities; for as shine, the differences are themselves and their opposite,
that is, they are the whole; and, conversely, they thus are only shine because
they are totalities. Thus this distinguishing, this reflecting shining of the
absolute, is only the identical positing of itself.
This relation in its immediate concept is the relation of substance and
accidents, the immediate internal disappearing and becoming of the abso-
lute reflective shine. If substance determines itself as a being-for-itself over
against an other or is absolute relation as something real, then we have
the relation of causality. Finally, when this last relation passes over into
reciprocal causality by referring itself to itself, we then have the absolute
relation also posited in accordance with the determination it contains;
this posited unity of itself in its determinations, which are posited as
the whole itself and consequently equally as determinations, is then the
concept.

A. THE RELATION OF SUBSTANTIALITY

Absolute necessity is absolute relation because it is not being as such but
being that is because it is, being as the absolute mediation of itself with
itself. This being is substance; as the final unity of essence and being,
it is the being in all being. It is neither the unreflected immediate, nor
something abstract standing behind concrete existence and appearance,
but the immediate actuality itself, and it is this actuality as being absolutely
reflected into itself, as a subsisting that exists in and for itself. – Substance,
as this unity of being and reflection, is essentially the shining and the
positedness of itself. The shining is a self-referring shining, thus it is; this
being is substance as such. Conversely, this being is only the self-identical
positedness, and as such it is shining totality, accidentality.
This shining is identity as identity of form – the unity of possibility and
actuality. It is becoming at first, contingency as the sphere of coming-to-be
and passing-away; for in the determination of immediacy the connection
of possibility and actuality is the immediate conversion of the two into each
other as existents, of each into its other as only an other to it. – But because
being is shine, their relation is also one of identical terms or of terms
shining in one another, that is, reflection. The movement of accidentality,
therefore, exhibits in each of its moments the mutual reflective shine of
the categories of being and of the reflective determinations of essence. –
The immediate something has a content; its immediacy is at the same time
reflected indifference towards the form. This content is determinate, and
because this determinateness is one of being, the something passes over into
an other. But quality is also a determinateness of reflection; as such, it
is indifferent diversity. But this diversity is animated into opposition, and
returns to the ground which is the nothing, but also immanent reflection.
This reflection sublates itself; but it is itself also reflected in-itselfness: so
it is possibility, and this in-itselfness, in its transition which is equally
immanent reflection, is necessary actuality.
This movement of accidentality is the actuosity of substance as the tran-
quil coming forth of itself. It is not active against something, but only against
itself as a simple unresisting element. The sublating of a presupposition is
the disappearing shine; only in the act of sublating the immediate does this
immediate itself come to be, or is that shining; the beginning that begins
from itself is first of all the positing of this itself from which the beginning
is made.
Substance, as this identity of the reflective shining, is the totality of the
whole and embraces accidentality in itself, and accidentality is the whole
substance itself. Its differentiation into the simple identity of being and the
flux of accidents within it is one form of its shining. That simple being
is the formless substance of the imagination for which the shine has not
determined itself as shine, but which holds on, as on an absolute, to this
indeterminate identity that has no truth but only is the determinateness
of immediate actuality, or equally so of in-itselfness or possibility – form
determinations that fall into accidentality. –
The other determination, the flux of accidents, is the absolute form-unity
of accidentality, substance as absolute power. – The ceasing-to-be of the
accident is its return as actuality into itself, as into its in-itself or into its
possibility; but this, its in-itself, is itself only a positedness and therefore also
actuality, and because these form determinations are equally determinations
of content, this possible is an actual differently determined also according to
content. Substance manifests itself through the actuality, with the content
of the latter into which it translates the possible, as creative power, and,
through the possibility to which it reduces the actual, as destructive power;
the creating is destructive and the destructing creative, for the negative and
the positive, possibility and negativity are in substantial necessity absolutely
united.
The accidents as such – and there are several of them, because plurality is
one of the determinations of being – have no power over each other. They
are the immediately existent something, or the something that immedi-
ately exists for itself; concretely existing things of manifold properties; or
wholes consisting of parts, self-subsisting parts; forces in need of reciprocal
solicitation and conditioning each other. In so far as such an accidental
being seems to exercise a power over an other, that power is that of sub-
stance that encompasses them both within itself and, as negativity, posits
an inequality of value: one it determines as ceasing-to-be and another as
having a different content and as coming-to-be, the one as passing over into
its possibility and the other into actuality accordingly – ever dividing itself
into this difference of form and content and ever purifying itself of this
one-sidedness, but in this purification ever falling back into determination
and division. – One accident thus drives out another only because its own
subsisting is this very totality of form and content into which it, as well as
its other, equally perishes.
Because of this immediate identity and presence of substance in the
accidents, there is still no real difference present. In this first determi-
nation, substance is not yet manifested according to its whole concept.
When substance, as self-identical being-in-and-for-itself, is differentiated
from itself as a totality of accidents, it is substance itself, as power, that
mediates the difference. This power is necessity, the positive persistence of
the accidents in their negativity and their mere positedness in their sub-
sistence; this middle is thus the unity of substantiality and accidentality
themselves, a middle whose extremes have no subsistence of their own.
Substantiality is, therefore, only the relation as immediately vanishing; it
refers to itself not as a negative and, as the immediate unity of power with
itself, is in the form only of its identity, not of its negative essence; only
one of its moments, that of negativity or of difference, vanishes altogether;
the other moment of identity does not. – Another way of considering the
matter is this. The shine or the accidentality is indeed in itself substance
by virtue of the power, but is not thus posited as this self-identical shine;
and therefore substance has only the accidentality, not itself, for its shape
or positedness; it is not substance as substance. The relation of substan-
tiality is at first, therefore, only this, that substance manifests itself as a
formal power whose differences are not substantial; in fact, substance only
is as the inner of the accidents, and these only are in the substance. Or
this relation is only the shining of totality as becoming; but it is equally
reflection and, for this reason, the accidentality which substance is in
itself is also posited as such; it is thus determined as self-referring nega-
tivity over against itself, determined as self-referring simple identity with
itself; and it is substance that exists in and for itself, substance endowed with
power. Thus the relation of substantiality passes over into the relation of
causality.

B. THE RELATION OF CAUSALITY

Substance is power – power reflected into itself, not transitive power but
power that posits determinations and distinguishes them from itself. As self-
referring in its determining, it is itself that which it posits as a negative or
makes into a positedness. This positedness is, as such, sublated substantiality,
the merely posited, the effect; the substance that exists for itself is, however,
cause.
This relation of causality is in the first place only this relation of cause
and effect; as such, it is the formal relation of causality.

a. Formal causality

1. Cause is originative as against the effect. – As power, substance is the
reflective shining, or it has accidentality. But in this shining, as power, it
equally is an immanent reflection; it thus expounds its transition, and this
reflective shine is determined as reflective shine, or the accident is posited
as being just this, something posited. – But in its determining substance
does not proceed from accidentality, as if the latter were an other before-
hand and were determined as determinateness only then, but the two are
one actuosity. Substance as power determines itself; but this determining is
immediately itself the sublation of the determining and a turning back.
It determines itself: substance, that which determines, is thus the immediate
and that which is itself already determined; in determining itself it therefore
posits the already determined as determined; and thus it has sublated the
positedness and has returned into itself. – Conversely, because this turning
back is the negative reference of substance to itself, it is itself a determining
or the repelling of itself from itself; it is through this turning back that the
determinate comes to be from which substance seems to begin and now to
posit as something which it has found already determined. – Absolute actu-
osity is thus cause – the power of substance in its truth as the manifestation
by which that which is in itself, the accident or the positedness, is imme-
diately expounded in its becoming, is posited as positedness, as effect. – This
effect is, therefore, first the same as what the accidentality of the relation
of substance is, namely substance as positedness; but, second, an accident
is substantially such only by vanishing, only as transient; but as effect it
is positedness as self-identical; in the effect the cause is manifested as the
whole substance, that is to say, as reflected into itself in the positedness
itself as such.
2. Over against this positedness reflected into itself, this determined as
determined, there stands substance as the non-posited original. Because
substance is as absolute power a turning back into itself, yet this turning
back is itself a determining, it is no longer the mere in-itself of its accident
but is also posited as this in-itself. Substance has actuality, therefore, only
as cause. But this actuality in which its in-itself, its determinateness in the
relation of substantiality, is now posited as determinateness, is effect; there-
fore substance has the actuality which it has as cause only in its effect. –
This is the necessity which is cause. – It is actual substance, because as
power substance determines itself; but it is at the same time cause, because
it expounds this determinateness or posits it as positedness and thus
posits its actuality as positedness or effect. This is the other of cause, the
positedness as against the original and as mediated through it. But cause,
as necessity, equally sublates this mediating and, in determining itself as
the originally self-referring term, as against the mediated, turns back to
itself; for positedness is determined as positedness, and consequently as
self-identical; therefore, cause is truly actual and self-identical only in its
effect. – The effect is therefore necessary, because it is the manifestation of
the cause or is this necessity which the cause is. – Only as this necessity is
cause self-moving, self-initiating without being solicited by another, self-
subsisting source of production out of itself; it must effect; its originariness is
this, that it is because its immanent reflection is a positing that determines
and conversely; the two are one unity.
Consequently, an effect contains nothing whatever that the cause does not
contain. Conversely, a cause contains nothing that is not in its effect. A cause
is cause only to the extent that it produces an effect; to be cause is nothing
but this determination of having an effect, and to be effect is nothing but this
determination of having a cause. Cause as such entails its effect, and the
effect entails the cause; in so far as a cause has not acted yet or has ceased
to act, it is not a cause; and the effect, in so far as its cause is no longer
present, is no longer an effect but an indifferent actuality.
3. Now in this identity of cause and effect the form distinguishing them
respectively, as that which exists in itself and that which is posited, is
sublated. The cause is extinguished in its effect and the effect too is thereby
extinguished, for it only is the determinateness of the cause. Hence this
causality which has been extinguished in the effect is an immediacy which
is indifferent to the relation of cause and effect and comes to it externally.

b. The determinate relation of causality

1. The self-identity of cause in its effect is the sublation of its power and
negativity, hence a unity which is indifferent to differences of form, that
is to say, content. – This content, therefore, refers to form (here causality)
only implicitly. The two are thus posited as diverse, and with respect to
content the form is itself a causality which is only immediately efficient, a
contingent causality.
Further, the content is as thus determined an internally diversified con-
tent; and the cause is determined in accordance with its content, and so
is therefore also the effect. – The content, since reflectedness here is also
immediate actuality, is to this extent actual, but finite, substance.
This is now the relation of causality in its reality and finitude. As formal,
it is the infinite relation of absolute power, the content of which is pure
manifestation or necessity. As finite causality, on the contrary, it has a given
content and, as an external difference, it runs its course here and there over
it, this identical content which in its determination is one and the same
substance.
Because of this identity of content, this causality is an analytic propo-
sition. It is the same fact that comes up once as cause and then again as
effect, in one case as something subsisting on its own and in the other
as positedness or determination. Since these determinations of form are
an external reflection, it is up to the essentially tautological consideration
of a subjective understanding to determine an appearance as effect and to
rise from it to its cause in order to comprehend and explain it. The same
content is being repeated; there is nothing else in the cause which is not in
the effect. – For instance, rain is the cause of wetness which is its effect; “the
rain makes wet,” this is an analytical proposition; the same water which is
rain is wetness; as rain, this water is only in the form of a subject by itself;
as wetness or moisture, it is on the contrary in adjectival form, something
posited no longer meant to have a subsistence on its own; and the one
determination, just like the other, is external to water. – Again, the cause of
this color is a coloring agent, a pigment which is one and the same actuality,
once in the form of an agent external to it, that is, is externally linked to
an agent different from it; but again in the determination, equally external
to it, of an effect. – The cause of an act is the inner intention of the subject
who is the agent, and this intention is the same in content and value as the
existence which it attains through the action. If the movement of a body
is considered as effect, the cause of this effect is then a propulsive force;
but it is the same quantum of movement which is present before and after
the propulsion, the same concrete existence which the propulsive body
contained and which it communicated to the one propelled; and what it
communicated, it lost in equal measure.
The cause, say the painter or the propulsive body, does have yet another
content than, in the case of the painter, the colors and the form combining
these into a painting; and, in the other case, the movement of specific
strength and direction. But this further content is a contingent side-being
which has nothing to do with the cause; whatever other qualities the painter
might possess besides being the painter of this painting, this does not enter
into the painting; only those of his properties which are displayed in the
effect are present in him as cause; as for the rest, he is not a cause. Likewise,
whether the propulsive body is of stone or wood, green, yellow, and so
on, all this does not enter into its propulsion and, to this extent, is not a
cause.
It is worth noting in regard to this tautology of the relation of causality
that the tautology does not seem to occur whenever it is not the proximate,
but the remote cause which is at issue. The alteration of form which the
basic fact undergoes as it passes through several middle terms hides the
identity which it preserves across them. In this proliferation of causes
introduced between it and the last effect, that fact is linked to other things
and circumstances, so that it is not that first term, which is declared the
cause, but all these several causes together, that contain the complete effect. –
For instance, if a man developed his talents in circumstances due to the
loss of his father who was hit by a bullet in battle, then this shot (or still
further back, the war or some cause of the war, and on to infinity) could
be adduced as the cause of the man’s skillfulness. But it is clear that the
shot, for one, is not the cause by itself but only in conjunction with the
other efficient determinations. Or more precisely, the shot is not the cause
at all, but only a single moment that pertained to the circumstances of the
possibility.
But it is the inadmissible application of the relation of causality to the
relations of physico-organic and spiritual life that must be noted above all.
Here that which is called the cause does indeed show itself to be of a different
content than the effect, but this is because anything that has an effect on a
living thing is independently determined, altered, and transmuted by the
latter, for the living thing will not let the cause come to its effect, that is, it
sublates it as cause. Thus it is inadmissible to say that nourishment is the
cause of blood, or that such and such a dish, or chill and humidity, are the
causes of fever or of what have you; it is equally inadmissible to give the Ionic
climate as the cause of Homer’s works, or Caesar’s ambition as the cause of
the fall of Rome’s republican constitution. In history in general there are
indeed spiritual masses and individuals at play and influencing each other;
but it is of the nature of spirit, in a much higher sense than it is of the
character of living things, that it will not admit another originative principle
within itself, or that it will not let a cause continue to work its causality
in it undisturbed but will rather interrupt and transmute it. – But these
relations belong to the idea, and will come up for discussion then. – This
much can still be noted here, namely that in so far as the relation of cause
and effect is admitted, albeit in an inappropriate sense, the effect cannot
be greater than the cause. It has become a common witticism in history
to let great effects arise from small causes and to cite as the first cause of an
event of far-reaching and profound consequence an anecdote. Any such so-
called first cause is to be regarded as no more than an occasion, an external
stimulus, of which the inner spirit of the event had no need, or could have
used a countless number of others, in order to make its first appearance,
to give itself a first breath and announce itself. The converse is rather the
case. It is by the spirit that any such triviality and contingency is determined
in the first place to be the occasion of spirit. Historical arabesques that draw
a full-blown figure out of a slender stalk are no doubt an ingenious, but
highly superficial, practice. It is true that in the rise of the great out of
the small we witness everywhere the conversion that spirit works on the
external; but precisely for this reason the external is not the cause within
spirit; rather, that conversion itself sublates the relation of causality.
2. But this determinateness of the relation of causality, that content
and form are different and indifferent to each other, extends further. The
determination of form is also content determination; cause and effect, the
two sides of the relation, are therefore also another content. Or the content,
because it is only as the content of a form, has the difference of this form
within it and is essentially different. But this form of the content is the
relation of causality, which is a content identical in cause and the effect,
and consequently the different content is externally connected, on the one
hand with the cause and on the other with the effect; hence the content itself
does not enter into the effective action and into the relation.
This external content is therefore relationless – an immediate concrete
existence, or because it is as content the implicit identity of cause and effect,
it is also immediate, existent identity. This content is, therefore, anything
at all which has manifold determinations of its existence, among them also
this, that it is in some respect or other cause or also effect. In it, the form
determinations of cause and effect have their substrate, that is to say, their
essential subsistence – and each has a particular subsistence (since their
identity is their subsistence); but it is a subsistence which is at the same
time immediate, not their subsistence as unity of form or as relation.
But this thing is not only substrate but also substance, for it is identical
subsistence only as subsistence of the relation. Moreover, the substance is
finite substance, for it is determined as immediate over against its causality.
But it has causality at the same time, for it is just as much an identity as this
relation. 21 – Now this substrate is, as cause, negative reference to itself. But
this “itself” to which it refers is, first, a positedness because it is determined
as immediately actual; this positedness, as content, is any determination
whatever. – Second, causality is external to the substrate and itself constitutes,
therefore, its positedness. Now since it is causal substance, its causality consists
in negatively referring itself to itself, hence to its positedness and external
causality. The effective action of this substance thus begins from something
external, frees itself from this external determination, and its turning back
into itself is the preservation of its immediate concrete existence and the
sublation of the one which is posited, and consequently of its causality as
such.
Take a stone that moves. It is a cause. Its movement is a determination
which it has. But, besides it, it contains yet many other determinations
(color, shape, and so on) that do not enter into its causality. Because
its immediate concrete existence is separated from its form-connection,
namely the form of causality, the latter is something external; the stone’s
movement and the causality attaching to it is in it only positedness. – But
the causality of the stone is also the stone’s own causality, as follows from
the fact that its substantial subsistence is the stone’s identical self-reference,
but that this is now determined as positedness and is therefore at the same
time negative self-reference. – Its causality, which is directed against itself as
a positedness or as an externality, consists therefore in sublating this and
through its removal in returning to itself – to this extent, therefore, in not
being self-identical in its positedness but only in restoring its originariness. –
Or again, rain is the cause of wetness, which is the same water as the rain.
This water has the determination of being rain and cause because this
determination has been posited in it by another; another force, or what
have you, has lifted it into the air and compressed it into a mass, the weight
of which makes it fall. Its being removed from the earth is a determination
alien to its original self-identity, to its gravity; its causality consists in
removing such determination and in restoring its original identity; but this
means also sublating its causality.
We now consider the second determinateness of causality which concerns
form; this relation is causality external to itself, as the originariness which is
within just as much positedness or effect. This union of opposite determi-
nation in an existent substrate constitutes the infinite regress from cause to
cause. – We start from an effect; the latter has as effect a cause; but this
cause has a cause in turn, and so on. Why does the cause have a cause in
turn? That is to say, why is the same side which was previously determined
as cause now determined as effect and therefore demands a new cause? –
Because the cause is something finite, a determinate in general; determined
as one moment of the form as against the effect; so it has its determinate-
ness or negation outside it; but for this very reason it is itself finite, has its
determinateness within it and is thereby positedness or effect. Its identity as
this positedness is also posited, but it is a third term, the immediate sub-
strate; causality is therefore external to itself, because its originariness is here
an immediacy. The difference of form is therefore a first determinateness,
not yet determinateness posited as determinateness; it is existent otherness.
Finite reflection, on the one hand, stops short at this immediate, removes
the unity of form from it and makes it be cause in one respect and effect in
another; on the other hand, it transfers the unity of form into the infinite,
and through the endless progression expresses its impotence in attaining
and holding fast to this unity.
Exactly the same is the case of the effect, or rather the endless progression
from effect to effect is one and the same as the regression from cause to cause.
Just as in the latter a cause becomes an effect which has another cause
in turn, so too, conversely, the effect becomes a cause which has another
effect in turn. – The determinate cause under consideration begins from
an externality and returns in its effect back to itself, but not as cause;
on the contrary, it loses its causality in that process. But, conversely, the
effect arrives at a substrate which is substance, an original self-referring
subsistence; in it, therefore, that positedness becomes a positedness, that is
to say, this substance, as the effect is posited in it, behaves as cause. But
that first effect, the positedness that accrues to the substance externally, is
other than the second which the substance produces; for this second effect is
determined as the immanent reflection of substance whereas the first is in it
as an externality. – But because causality is here causality external to itself,
it also equally fails to return in its effect back to itself but becomes therein
external to itself; its effect becomes again a positedness in a substrate – as
in another substance which however equally makes this positedness into a
positedness, in other words, manifests itself as cause, again repels its effect
from itself, and so on, into bad infinity.
3. We now have to see what has resulted from the movement of deter-
minate causality. – Formal causality expires in the effect and the element
of identity of these two moments emerges as a result, but it does so only
as an implicit unity of cause and effect to which the form connection is
external. – For this reason, the element of identity is immediate also with
respect to both of the two determinations of immediacy, first as in-itself,
as a content on which causality is deployed externally; second, as a concrete
existent substrate in which cause and effect inhere as different determina-
tions of form. In this substrate, the two determinations are implicitly one,
but, on account of this implicitness or of the externality of form, each is
external to itself and hence, in its unity with the other, is also determined
as other with respect to it. Consequently, the cause has indeed an effect and
is at the same time itself effect; and the effect not only has a cause but is itself
also cause. But the effect which the cause has, and the effect which it is, are
different – as are also the cause which the effect has and the cause which it
is.
The outcome of the movement of the determinate relation of causality
is then this, that the cause does not just expire in the effect, and thereby the
effect as well, as in formal causality, but that by expiring in the effect the
cause comes to be again; that the effect vanishes in the cause, but equally
comes to be again in it. Each of these determinations sublates itself in its
positing, and posits itself in its sublating; what we have is not an external
transition of causality from one substrate to another, but its becoming-
other is at the same time its own positing. Causality thus pre-supposes itself
or conditions itself. The previously only implicit identity, the substrate, is
therefore now determined as presupposition or posited as against the efficient
causality, and the reflection hitherto only external to the identity is now in
relation to it.

c. Action and reaction

Causality is a presupposing activity. The cause is conditioned; it is a negative
reference to itself as a presupposed, as an external other which in itself, but
only in itself, is causality itself. This other is, as we have seen, 22 the substantial
identity into which formal causality passes over, which now has determined
itself as against this causality as its negative. Or it is the same as the substance
of the causal relation, but a substance which is confronted by the power
of accidentality as itself substantial activity. – It is the passive substance. –
Passive is that which is immediate, or which exists-in-itself but is not also
for itself – pure being or essence in just this determinateness of abstract self-
identity. – Confronting the passive substance is the negatively self-referring
substance, the efficient substance. It is cause inasmuch as in determinate
causality it has restored itself out of the effect through the negation of itself –
a reflected being 23 which in its otherness or as an immediate behaves
essentially as a positing activity and through its negation mediates itself.
Here, therefore, causality no longer has a substrate in which it inheres; it is
not a determination of form as against this identity but is itself substance,
or in other words, causality alone is at the origin. – The substrate is the
passive substance which causality has presupposed for itself.
This cause now acts, for it is the negative power over itself; at the same
time it is its own presupposition; thus it acts upon itself as upon an other,
upon the passive substance. – Hence, it first sublates the otherness of this
substance and returns in it back to itself; second, it determines this same
substance, posits this sublation of its otherness or the substance’s turning
back into itself as a determinateness. This positedness, because it is at the
same time the substance’s turning back into itself, is at first its effect.
But conversely, because as presupposing it determines itself as its other, it
then posits the effect in this other, in the passive substance. – Or again,
because the passive substance is itself this double – namely a self-subsistent
other, and at the same time something presupposed and already implicitly
identical with the efficient cause – because of this, the action of the passive
substance is therefore itself double. It is at once both the sublation of
its determinateness, namely of its condition, or the sublation of the self-
subsistence of the passive substance; and also, in sublating its identity as
it sublates this substance, the pre-supposing of itself, that is, the positing
or supposing of itself as other. – Through this last moment, the passive
substance is preserved; that first sublation of it appears in this respect at
the same time also in this way, namely that only some determinations are
sublated in it, and its identity in the effect with the efficient cause occurs
in it externally.
To this extent it suffers violence. – Violence is the appearance of power,
or power as external. But power is something external only in so far as in its
action, that is, in the positing of itself, the causal substance is at the same
time a presupposing, that is, posits itself as sublated. Conversely, the act of
violence is therefore equally an act of power. The violent cause acts only on
an other which it presupposes; its effect on it is its negative self-reference,
or the manifestation of itself. The passive is the self-subsistent which is only
a posited, something internally fractured – an actuality which is condition,
though a condition that now is in its truth as an actuality that is only a
possible, or, conversely, an in-itself that is only the determinateness of the
in-itself, is only passive. To that which suffers violence, therefore, not only
is it possible to do violence, but violence must be done to it; that which
has dominion over an other, only has it because its power is that of the
other, a power which in that dominion manifests both itself and the other.
Through violence the passive substance is only posited as what it is in truth,
namely, that because it is the simple positive or the immediate substance,
for that very reason it is only something posited; the “pre-” 24 that it has as
condition is the reflective shine of immediacy that the efficient causality
strips off from it.
Passive substance, therefore, is only given its due by the action on it of
another power. What it loses is the immediacy it had, the substantiality alien
to it. What comes to it as an alien something, namely that it is determined
as a positedness, is its own determination. – But now in being determined
in its positedness, or in its own determination, the result is that it is not
sublated but rather that it only rejoins itself and in its being determined
is, therefore, an originariness. – On the one hand, therefore, the passive
substance is preserved or posited by the active, namely in so far as the latter
sublates itself; but, on the other hand, it is the act of the passive substance
itself to rejoin itself and thus to make itself into what is originary and a
cause. The being posited by an other and its own becoming are one and the
same.
Now, because the passive substance has been converted into a cause, it
follows, first, that the effect is sublated in it; therein consists its reaction in
general. As passive substance, it is in itself as positedness; also, positedness
has been posited in it by the other substance, namely in so far as it received
its effect within it. Its reaction contains, therefore, a twofold aspect. For
one, what it is in itself is posited. And two, what it is as posited displays itself
as its in-itself; it is positedness in itself, hence through the other substance it
receives an effect within; but, conversely, this positedness is its own in-itself,
it is thus its own effect, it itself displays itself as a cause.
Second, the reaction is directed at the first efficient cause. For the effect
which the hitherto passive substance sublates within itself is precisely the
effect of that other cause. But a cause has its substantial actuality only in
its effect; inasmuch as this effect is sublated, so is also the causal substan-
tiality of the other cause. This happens first in itself through itself, in that
the cause makes itself into an effect; its negative determination disappears
in this identity and the cause becomes passive; and, second, it happens
through the hitherto passive, but now reacting substance, which sublates its
effect. – Now in determinate causality the substance acted upon becomes
a cause, for it acts against the positing of an effect in it. But it did not
react against the cause of that effect but posited its effect rather in another
substance, and thus there arose the progression to infinity of effects –
for here the cause is only implicitly identical with itself in the effect, and
hence, on the one hand, it expires into an immediate identity as it comes
to rest, but, on the other hand, it revives in another substance. – In con-
ditioned causality, on the contrary, the cause refers back to itself in the
effect, for the latter is as a condition, as a presupposition, its other, and its
act is therefore just as much a becoming as a positing and sublating of the
other.
Further, causality behaves in all this as passive substance; but, as we
have seen, 25 the latter becomes causal through the effect it incurs. That first
cause, the one which acts first and receives its effect back into itself as a
reaction, thus comes up again as a cause, whereby the activity which in
finite causality runs into the bad infinite progression is bent around and
becomes an action that returns to itself, an infinite reciprocal action.

C. RECIPROCITY OF ACTION

In finite causality it is substances that actively relate to each other. Mech-
anism consists in this externality of causality, where the cause’s reflection
in its effect into itself is at the same time a repelling being, or where, in
the self-identity which the causal substance has in its effect, the substance
is equally immediately external to itself and the effect is transposed into
another substance. In reciprocity of action this mechanism is now sublated,
for it contains first the disappearing of that original persistence of immediate
substantiality; second, the coming to be of the cause, and hence originariness
mediating itself with itself through its negation.
At first, the reciprocity of action takes on the form of a reciprocal causality
of substances that are presupposed and that condition each other; each is with
respect to the other both active and passive substance. Since the two are thus
passive and active at once, their difference is thereby already sublated; it
is a totally transparent reflective shine; they are substances only in being
the identity of the active and the passive. The reciprocity of action is itself,
therefore, only a still empty way and manner, and all that is still needed
is merely the external bringing together of what is already there, both in
itself and as posited. First of all, it is no longer substrates that are referred to
each other but substances; in the movement of conditional causality, the
still left over presupposed immediacy has been sublated, and what conditions
the causing activity is only an influence, or its own passivity. But this
influence, moreover, does not come from another substance originating it
but from precisely a causality which is conditioned by influence, or one
which is mediated. This at first external factor that accrues to the cause
and constitutes the side of its passivity is therefore mediated through the
causality itself, is produced through its own activity and is, consequently,
a passivity posited by its own very activity. – Causality is conditioned and
conditioning. As conditioning, it is passive; but it is equally so as conditioned.
This conditioning or passivity is the negation of the cause through itself in
that it makes itself essentially into an effect and is cause precisely for that
reason. Reciprocity of action is, therefore, only causality itself; the cause does
not just have an effect but, in the effect, refers as cause back to itself.
Causality has thereby returned to its absolute concept and has at the
same time attained the concept itself. At first, it is real necessity, absolute
self-identity in which the difference between it and the determinations
referring to each other within it are substances, free actualities, over against
one another. Necessity is in this way inner identity; causality is the mani-
festation of it in which its reflective shine of substantial otherness has been
sublated, and necessity is elevated to freedom. – In the reciprocity of action,
originative causality displays itself as arising from its negation, from pas-
sivity, and as passing away into it, as a becoming, but in such a way that
this becoming is at the same time equally only shining; the transition into
otherness is reflection-into-itself; negation, which is the ground of the cause,
is its positive rejoining with itself.
In the reciprocity of action, therefore, necessity and causality have dis-
appeared; they contain both the immediate identity as combination and
reference and the absolute substantiality of the differences, consequently their
contingency, the original unity of substantial difference and therefore the
absolute contradiction. Necessity is being, because being is; it is the unity
of being with itself that has itself as ground, but, conversely, because this
being has a ground, it is not being; it is simply and solely reflective shin-
ing, reference or mediation. Causality is this posited transition of original
being, of cause, into reflective shine or mere positedness, and, conversely, of
positedness into originariness; but the identity itself of being and reflective
shine still is the inner necessity. This inwardness or this in-itself sublates
the movement of causality; the result is that the substantiality of the sides
that stand in relation is lost, and necessity unveils itself. Necessity does not
come to be freedom by vanishing but in that its still only inner identity is
manifested, and this manifestation is the identical movement immanent to
the different sides, the immanent reflection of shine as shine. – Conversely,
contingency thereby comes to be freedom at the same time, for the sides
of necessity, which have the shape of independent, free actualities that do
not reflectively shine into each other, are now posited as an identity, so that
now these totalities of immanent reflection, in their differences, also shine
as identical, in other words, they are also posited as only one and the same
reflection.

No longer, therefore, does absolute substance as self-differentiating abso-
lute form repel itself as necessity from itself, nor does it fall apart as
contingency into indifferent, external substances, but, on the contrary, it
differentiates itself: on the one hand, into the totality (the heretofore passive
substance) which is at the origin, as the reflection from internal deter-
minateness, as simple whole that contains its positedness within itself and
in this positedness is posited as self-identical – this is the universal –; on the
other hand, into the totality (the hitherto causal substance) which is the
reflection, equally from internal determinateness, into the negative deter-
minateness which, just as the self-identical determinateness, equally is the
whole, but posited as the self-identical negativity – the singular. But, because
the universal is self-identical only in that the determinateness that it holds
within is sublated, hence it is the negative as negative, it immediately is the
same negativity that singularity is. And the singularity, because it equally is
the determinedly determined, the negative as negative, immediately is the
same identity that universality is. This, their simple identity, is the particular-
ity that, from the singular, holds the moment of determinateness; from the
universal, that of immanent reflection – the two in immediate unity. These
three totalities are therefore one and the same reflection that, as negative
self-reference, differentiates itself into the other two totalities – but as into
a perfectly transparent difference, namely into the determinate simplicity, or
into the simple determinateness, which is their one same identity. – This is
the concept, the realm of subjectivity or of freedom.
