SECTION I

Essence as Reflection Within

Essence issues from being;
hence it is not immediately in and for itself
but is a result of that movement.
Or, since essence is taken at first as something immediate,
it is a determinate existence to which another stands opposed;
it is only essential existence, as against the unessential.
But essence is being which has been sublated in and for itself;
what stands over against it is only shine.
The shine, however, is essence’s own positing.
First, essence is reflection.
Reflection determines itself;
its determinations are a positedness
which is immanent reflection at the same time.
Second, these reflective determinations
or essentialities are to be considered.
Third, as the reflection of its immanent determining,
essence turns into foundation and passes over
into concrete existence and appearance.

CHAPTER 1

Shine

As it issues from being, essence seems to stand over against it;
this immediate being is, first, the unessential.
But, second, it is more than just the unessential;
it is being void of essence; it is shine.
Third, this shine is not something external,
something other than essence, but is essence’s own shining.
This shining of essence within it is reflection.

A. THE ESSENTIAL AND THE UNESSENTIAL

Essence is sublated being.

It is simple equality with itself
but is such as the negation of the sphere of being in general.
And so it has immediacy over against it,
as something from which it has come to be but which has
preserved and maintained itself in this sublating.
Essence itself is in this determination
an existent immediate essence,
and with reference to it
being is only something negative,
nothing in and for itself:
essence, therefore, is a determined negation.
Being and essence relate to each other in this fashion
as again others in general which are mutually indifferent,
for each has a being, an immediacy,
and according to this being they stand in equal value.
But as contrasted with essence,
being is at the same time the unessential;
as against essence, it has the determination of something sublated.
And in so far as it thus relates to essence
as an other only in general,
essence itself is not essence proper
but is just another existence, the essential.
The distinction of essential and unessential has made
essence relapse into the sphere of existence,
for as essence is at first,
it is determined with
respect to being as an existent and therefore as an other.
The sphere of existence is thus laid out as foundation,
and that in this sphere being is
being-in-and-for-itself is a further determination external to existence,
just as, contrariwise, essence is indeed being-in-and-for-itself,
but only over against an other, in a determinate respect.
Consequently, inasmuch as essential and unessential aspects are
distinguished in an existence from each other,
this distinguishing is an external positing,
a taking apart that leaves the existence itself untouched;
it is a separation which falls on the side of
a third and leaves undetermined what belongs to the essential and what
belongs to the unessential.
It is dependent on some external standpoint or consideration
and the same content can therefore sometimes be considered
as essential, sometimes as unessential.
On closer consideration, essence becomes something only essential as
contrasted with an unessential because essence is only taken, is
as sublated being or existence.
In this fashion, essence is only the first negation,
or the negation, which is determinateness,
through which being becomes only existence,
or existence only an other.
But essence is the absolute negativity of being;
it is being itself, but not being determined only as an other:
it is being rather that has sublated itself both as immediate being
and as immediate negation, as the negation
which is affected by an otherness.
Being or existence, therefore, does not persist except as what essence is,
and the immediate which still differs from essence is not
just an unessential existence but an immediate
which is null in and for itself;
it only is a non-essence, shine.

B. SHINE

1. Being is shine.

The being of shine consists solely in the sublatedness of being,
in being’s nothingness;
this nothingness it has in essence, apart from its nothingness,
apart from essence, it does not exist.
It is the negative posited as negative.
Shine is all that remains of the sphere of being.
But it still seems to have an immediate side
which is independent of essence and to be, in general,
an other of essence.
Other entails in general the two moments of existence and non-existence.
Since the unessential no longer has a being,
what is left to it of otherness is only the pure moment of non-existence;
shine is this immediate non-existence,
a non-existence in the determinateness of being,
so that it has existence only with reference to another,
in its non-existence;
it is the non-self-subsistent which exists only in its negation.
What is left over to it is thus only the pure determinateness of immediacy;
it is as reflected immediacy, that is, one which is
only by virtue of the mediation of its negation
and which, over against this mediation, is nothing
except the empty determination of the immediacy of non-existence.
Shine, the “phenomenon” of skepticism,
and also the “appearance” of idealism, is thus this immediacy
which is not a something nor a thing in general,
not an indifferent being that would exist apart from
its determinateness and connection with the subject.
Skepticism did not permit itself to say “It is,”
and the more recent idealism did not permit itself
to regard cognitions as a knowledge of the thing-in-itself.
The shine of the former was supposed absolutely
not to have the foundation of a being:
the thing-in-itself was not supposed to enter into these cognitions.
But at the same time skepticism allowed
a manifold of determinations for its shine,
or rather the latter turned out to have
the full richness of the world for its content.
Likewise for the appearance of idealism:
it encompassed the full range of these manifold determinacies.
So, the shine of skepticism and the appearance of idealism
do immediately have a manifold of determination.
This content, therefore, might well have no being as foundation,
no thing or thing-in-itself;
for itself, it remains as it is;
it is simply transposed from being into shine,
so that the latter has within itself those manifold
determinacies that exist immediately,
each an other to the other.
The shine is thus itself something immediately determined.
It can have this or that content;
but whatever content it has, it has not posited it
but possesses it immediately.
Idealism, whether Leibnizian, Kantian, Fichtean, or in any
other form, has not gone further than skepticism in this:
it has not advanced beyond being as determinateness.
Skepticism lets the content of its shine to be given to it;
the shine exists for it immediately,
whatever content it might have.
The Leibnizian monad develops its representations from itself
but is not their generating and controlling force;
they rise up in it as a froth, indifferent,
immediately present to each other and to the monad as well.
Likewise Kant’s appearance is a given content of perception that
presupposes affections, determinations of the subject which are immediate
to each other and to the subject.
As for the infinite obstacle of Fichte’s Idealism,
it might well be that it has no thing-in-itself for foundation,
so that it becomes a determinateness purely within the “I.”
But this determinateness that the “I” makes its own,
sublating its externality,
is to the “I” at the same time an immediate determinateness,
a limitation of the “I” which the latter may transcend
but which contains a side of indifference,
and on account of this indifference,
although internal to the “I,”
it entails an immediate non-being of it.
2. Shine thus contains an immediate presupposition,
an independent side vis-à-vis essence.
But the task, inasmuch as this shine is distinct from essence,
is not to demonstrate that it sublates itself
and returns into essence, for being has returned into essence in its totality;
shine is the null as such.
The task is to demonstrate that the determinations which
distinguish it from essence are the determinations of essence itself;
further, that this determinateness of essence,
which shine is, is sublated in essence itself.
What constitutes the shine is the immediacy of non-being;
this non-being, however, is nothing else than
the negativity of essence within essence itself.
In essence, being is non-being.
Its inherent nothingness is the negative nature of essence itself.
But the immediacy or indifference which this non-being contains is
essences’s own absolute in-itself.
The negativity of essence is its self-equality or
its simple immediacy and indifference. Being has
preserved itself in essence inasmuch as this latter, in its infinite negativity,
has this equality with itself;
it is through this that essence is itself being.
The immediacy that the determinateness has in shine against essence is
thus none other than essence’s own immediacy,
though not the immediacy of an existent
but rather the absolutely mediated
or reflective immediacy which is shine;
being, not as being, but only as the determinateness of
being as against mediation;
being as moment.
These two moments [nothingness but as subsisting],
and being but as moment;
or again, negativity existing in itself and reflected immediacy,
these two moments that are the moments of shine,
are thus the moments of essence itself;
it is not that there is a shine of being in essence,
or a shine of essence in being:
the shine in the essence is not the shine of an other
but is rather shine as such, the shine of essence itself.
Shine is essence itself in the determinateness of being.
Essence has a shine because it is determined within itself
and is therefore distinguished from its absolute unity.
But this determinateness is as determinateness
just as absolutely sublated in it.
For essence is what stands on its own:
it exists as self-mediating through a negation which it itself is.
It is, therefore, the identical unit of absolute negativity and immediacy.
The negativity is negativity in itself;
it is its reference to itself and thus immediacy in itself.
But it is negative reference to itself,
a self-repelling negating;
thus the immediacy existing in itself is
the negative or the determinate over against the negativity.
But this determinateness is itself absolute negativity
and this determining, which as determining immediately sublates itself,
is a turning back into itself.
Shine is the negative which has a being,
but in another, in its negation;
it is a non-self-subsisting-being
which is sublated within and null.
And so it is the negative which returns into itself,
the non-subsistent as such, internally non-subsistent.
This reference of the negative or
the non-subsistent to itself is
the immediacy of this non-subsistent;
it is an other than it;
it is its determinateness over against it,
or the negation over against the negative.
But this negation which stands over against the negative is negativity as
referring solely to itself, the absolute sublation of the determinateness itself.
The determinateness that shine is in essence is, therefore, infinite deter-
minateness;
it is only the negative which coincides with itself and hence
a determinateness that, as determinateness, is self-subsistence and not
determined. – Contrariwise, the self-subsistence, as self-referring immedi-
acy, equally is just determinateness and moment, negativity solely referring
to itself. – This negativity which is identical with immediacy, and thus the
immediacy which is identical with negativity, is essence. Shine is, therefore,
essence itself, but essence in a determinateness, in such a way, however,
that the determinateness is only a moment, and the essence is the shining
of itself within itself.

In the sphere of being, non-being arises over against being,
each equally an immediate, and the truth of both is becoming.
In the sphere of essence, we have the contrast first of
essence and the non-essential, then of essence and shine,
the non-essential and the shine being both the leftover of being.
But these two, and no less the distinction of essence from them,
consist solely in this:
that essence is taken at first as an immediate, not as it is in itself,
namely as an immediacy which is immediacy
as pure mediacy or absolute negativity.
This first immediacy is thus only the determinateness of immediacy.
The sublating of this determinateness of essence consists, therefore,
in nothing further than showing that the unessential is only shine,
and that essence rather contains this shine within itself.
For essence is an infinite self-contained movement
which determines its immediacy as negativity
and its negativity as immediacy,
and is thus the shining of itself within itself.
In this, in its self-movement,
essence is reflection.

C. REFLECTION

Shine is the same as what reflection is;
but it is reflection as immediate.
For this shine which is internalized
and therefore alienated from its immediacy,
the German has a word from an alien language, “Reflexion.”
Essence is reflection, the movement of becoming and transition
that remains within itself, wherein that which is distinguished is
determined simply and solely as the negative in itself, as shine.
In the becoming of being, it is being which lies
at  the foundation of determinateness,
and determinateness is reference to an other.
Reflective movement is by contrast the other as negation in itself,
a negation which has being only as self-referring.
Or, since this self-referring is precisely this negating of negation,
what we have is negation as negation,
negation that has its being in its being-negated, as shine.
Here, therefore, the other is not being with negation or limit,
but negation with negation.
But the first over against this other, the immediate or being, is
only this self-equality itself of negation,
the negated negation, the absolute negativity.
This self-equality or immediacy, therefore, is
not a first from which the beginning is made
and which would pass over into its negation;
nor is there an existent substrate which would
go through the moves of reflection;
immediacy is rather just this movement itself.
In essence, therefore, the becoming,
the reflective movement of essence,
is the movement from nothing to nothing
and thereby back to itself.
Transition or becoming sublates itself in its transition;
the other which comes to be in this transition is
not the non-being of a being but the nothingness of a nothingness,
and this, to be the negation of a nothingness, constitutes being.
Being is only as the movement of nothingness to nothingness,
and so it is essence;
and this essence does not have this movement in itself,
but the movement is rather the absolute shine itself,
the pure negativity which has nothing outside it
which it would negate but which rather negates only its negative,
the negative which is only in this negating.
This pure absolute reflection, which is the movement
from nothing to nothing, further determines itself.
It is, first, positing reflection.
Second, it takes as its starting point the presupposed immediate,
and then it is external reflection.
Third, it sublates however this presupposition,
and because in the sublating of the presupposition
it presupposes at the same time,
it is determining reflection.

1. Positing reflection

Shine is a nothingness or a lack of essence.
But a nothingness or that which
is void of essence does not have its being in an other in which it shines,
but its being is its own equality with itself;
this conversion of the negative with itself has been determined
as the absolute reflection of essence.

This self-referring negativity is therefore the negating of itself. It is thus
just as much sublated negativity as it is negativity. Or again, it is itself
the negative and the simple equality with itself or immediacy. It consists,
therefore, in being itself and not being itself, and the two in one unity. –

Reflection is at first the movement of the nothing to the nothing, and
thus negation coinciding with itself. This self-coinciding is in general
simple equality with itself, immediacy. But this falling together is not the
transition of negation into equality as into a being other than it; reflection
is transition rather as the sublating of transition, for it is the immediate
falling together of the negative with itself. And so this coinciding is, first,
self-equality or immediacy; but, second, this immediacy is the self-equality
of the negative, and hence self-negating equality, immediacy which is in
itself the negative, the negative of itself: its being is to be what it is not.

The self-reference of the negative is therefore its turning back into itself;
it is immediacy as the sublating of the negative,
but immediacy simply and solely as this reference
or as turning back from a one, and hence as self-sublating immediacy.
This is positedness, immediacy purely as
determinateness or as self-reflecting.
This immediacy, which is only as the
turning back of the negative into itself, is the immediacy which consti-
tutes the determinateness of shine, and from which the previous reflective
movement seemed to begin.
But, far from being able to begin with this
immediacy,
the latter first is rather as the turning back or as the reflection
itself.
Reflection is therefore the movement which, since it is the turning
back, only in this turning is that which starts out or returns.

It is a positing, inasmuch as it is immediacy as a turning back; that is to
say, there is not an other beforehand, one either from which or to which it
would turn back; it is, therefore, only as a turning back or as the negative of
itself.
But further, this immediacy is sublated negation and sublated return
into itself.
Reflection, as the sublating of the negative, is the sublating of its
other, of the immediacy.
Because it is thus immediacy as a turning back,
the coinciding of the negative with itself, it is equally the negation of the
negative as negative.
And so it is presupposing.
Or immediacy is as a
turning back only the negative of itself, just this, not to be immediacy;
but reflection is the sublating of the negative of itself, coincidence with
itself;
it therefore sublates its positing, and inasmuch as it is in its positing
the sublating of positing, it is presupposing.
In presupposing, reflection
determines the turning back into itself as the negative of itself, as that of
which essence is the sublating.
It is its relating to itself, but to itself as to
the negative of itself;
only so is it negativity which abides with itself, self-
referring negativity.
Immediacy comes on the scene simply and solely as a
turning back and is that negative which is the semblance of a beginning,
the beginning which the return negates.
The turning back of essence is therefore its self-repulsion.
Or inner directed reflection is essentially the
presupposing of that from which the reflection is the turning back.

It is only by virtue of the sublating of its equality with itself that essence
is equality with itself. Essence presupposes itself, and the sublating of this
presupposing is essence itself; contrariwise, this sublating of its presuppo-
sition is the presupposition itself. – Reflection thus finds an immediate
before it which it transcends and from which it is the turning back. But
this turning back is only the presupposing of what was antecedently found.
This antecedent comes to be only by being left behind; its immediacy is sub-
lated immediacy. – The sublated immediacy is, contrariwise, the turning
back into itself, essence that arrives at itself, simple being equal to itself.
This arriving at itself is thus the sublating of itself and self-repelling, pre-
supposing reflection, and its repelling of itself from itself is the arriving at
itself.

It follows from these considerations that the movement of reflection is to
be taken as an absolute internal counter-repelling. For the presupposition of
the turning back into itself – that from which essence arises, essence being
only as this coming back – is only in the turning back itself. Transcending
the immediate from which reflection begins occurs rather only through
this transcending; and the transcending of the immediate is the arriving at
the immediate. The movement, as forward movement, turns immediately
around into itself and so is only self-movement – a movement which comes
from itself in so far as positing reflection is presupposing reflection, yet, as
presupposing reflection, is simply positing reflection.

Thus is reflection itself and its non-being, and only is itself by being the
negative of itself, for only in this way is the sublating of the negative at the
same time a coinciding with itself.

The immediacy which reflection, as a process of sublating, presupposes
for itself is simply and solely a positedness, something in itself sublated
which is not diverse from reflection’s turning back into itself but is itself
only this turning back. But it is at the same time determined as a negative, as
immediately in opposition to something, and hence to an other. And so is
reflection determined. According to this determinateness, because reflection
has a presupposition and takes its start from the immediate as its other, it
is external reflection.

2. External reflection

Reflection, as absolute reflection, is essence shining within, essence that
posits only shine, only positedness, for its presupposition; and as presup-
posing reflection, it is immediately only positing reflection. But external
or real reflection presupposes itself as sublated, as the negative of itself. In
this determination, it is doubled. At one time it is as what is presupposed,
or the reflection into itself which is the immediate. At another time, it is as
the reflection negatively referring to itself; it refers itself to itself as to that
its non-being.
External reflection thus presupposes a being, at first not in the sense that
its immediacy is only positedness or moment, but in the sense rather that
this immediacy refers to itself and the determinateness is only as moment.
Reflection refers to its presupposition in such a way that the latter is its
negative, but this negative is thereby sublated as negative. – Reflection,
in positing, immediately sublates its positing, and so it has an immediate
presupposition. It therefore finds this presupposition before it as something
from which it starts, and from which it only makes its way back into
itself, negating it as its negative. But that this presupposition is a negative
or a positedness is not its concern; this determinateness belongs only to
positing reflection, whereas in the presupposing positedness it is only as
sublated. What external reflection determines and posits in the immediate
are determinations which to that extent are external to it.
In the sphere of being, external reflection was the infinite;
the finite stands as the first,
as the real from which the beginning is made as from a foundation that
abides, whereas the infinite is the reflection into itself standing over against
it.
This external reflection is the syllogism in which the two extremes are
the immediate and the reflection into itself; the middle term is the reference
connecting the two, the determinate immediate, so that one part of this
connecting reference, the immediate, falls to one extreme alone, and the
other, the determinateness or the negation, only to the other extreme.
But if one takes a closer look at what the external reflection does, it
turns out that it is, secondly, the positing of the immediate, an immediate
which thus becomes the negative or the determined; but it is immediately
also the sublating of this positing, for it pre-supposes the immediate; in
negating, it is the negating of its negating. But thereby it immediately is
equally a positing, the sublating of the immediate which is its negative; and
this negative, from which it seemed to begin as from something alien, only
is in this its beginning. In this way, the immediate is not only implicitly
in itself (that is, for us or in external reflection) the same as what reflection
is, but is posited as being the same. For the immediate is determined by
reflection as the negative of the latter or as the other of it, but it is reflection
itself which negates this determining. – The externality of reflection vis-
à-vis the immediate is consequently sublated; its self-negating positing is
its coinciding with its negative, with the immediate, and this coinciding is
the immediacy of essence itself. – It thus transpires that external reflection
is not external but is just as much the immanent reflection of immediacy
itself; or that the result of positing reflection is essence existing in and for
itself. External reflection is thus determining reflection.

3. Determining reflection

Determining reflection is in general
the unity of positing and external reflection.
This is now to be examined more closely.

1. External reflection begins from immediate being,
positing reflection from nothing.
In its determining, external reflection posits another in the
place of the sublated being, but this other is essence;
the positing does not posit its determination in the place of an other;
it has no presupposition.
But, precisely for this reason,
it is not complete as determining reflection;
the determination which it posits is consequently only a posited;
this is an immediate, not however as equal to itself
but as self-negating;
its connection with the turning back into itself is absolute;
it is only in the reflection-into-itself
but is not this reflection itself.

The posited is therefore an other,
but in such a manner that the self-equality
of reflection is retained;
for the posited is only as sublated,
as reference to the turning back into itself.
In the sphere of being, existence was the being
that had negation in it, and being was the immediate ground
and element of this negation which was,
therefore, itself immediate negation.
In the sphere of essence,
positedness is what corresponds to existence.
Positedness is equally an existence,
but its ground is being as essence
or as pure negativity;
it is a determinateness or a negation,
not as existent but immediately as sublated.
Existence is only positedness;
this is the principle of the essence of existence.
Positedness stands on the one side over against existence,
and over against essence on the other:
it is to be regarded as the means which conjoins
existence with essence and essence with existence.
If it is said, a determination is only a positedness,
the claim can thus have a twofold meaning,
according to whether the determination is such
in opposition to existence or in opposition to essence.
In either meaning, existence is taken for something superior to positedness,
which is attributed to external reflection, to the subjective.
In fact, however, positedness is the superior, because, as posited,
existence is what it is in itself something negative,
something that refers simply and solely to the turning back into itself.
For this reason positedness is only a positedness
with respect to essence:
it is the negation of this turning back
as achieved return into itself.

2. Positedness is not yet a determination of reflection;
it is only determinateness as negation in general.
But the positing is now united with external reflection;
in this unity, the latter is absolute presupposing, that is,
the repelling of reflection from itself
or the positing of determinateness as its own.
As posited, therefore, positedness is negation;
but as presupposed, it is reflected into itself.
And in this way positedness is a determination of reflection.

The determination of reflection is distinct from the determinateness
of being, of quality; the latter is immediate reference to other in general;
positedness also is reference to other, but to immanently reflected being.
Negation as quality is existent negation; being constitutes its ground and
element. The determination of reflection, on the contrary, has for this
ground immanent reflectedness. Positedness gets fixed in determination
precisely because reflection is self-equality in its negatedness; the latter
is therefore itself reflection into itself. Determination persists here, not
by virtue of being but because of its self-equality. Since the being which
sustains quality is unequal to the negation, quality is consequently unequal
within itself, and hence a transient moment which disappears in the other.
The determination of reflection is on the contrary positedness as negation
negation which has negatedness for its ground, is therefore not unequal to
itself within itself, and hence essential rather than transient determinateness.
What gives subsistence to it is the self-equality of reflection which has the
negative only as negative, as something sublated or posited.

Because of this reflection into themselves, the determinations of reflection
appear as free essentialities, sublated in the void without reciprocal attraction
or repulsion. In them the determinateness has become entranced and
infinitely fixed by virtue of the reference to itself. It is the determinate
which has subjugated its transitoriness and its mere positedness to itself,
that is to say, has deflected its reflection-into-other into reflection-into-
itself. These determinations hereby constitute the determinate shine as it
is in essence, the essential shine. Determining reflection is for this reason
reflection that has exited from itself; the equality of essence with itself is
lost in the negation, and negation predominates.

Thus there are two distinct sides to the determination of reflection.
First, reflection is positedness, negation as such;
second, it is immanent reflection.
According to the side of positedness,
it is negation as negation,
and this already is its unity with itself.
But it is this unity at first only implicitly or
in itself, an immediate which sublates itself within, is the other of itself.
To this extent, reflection is a determining that abides in itself. In it essence
does not exit from itself; the distinctions are solely posited, taken back
into essence. But, from the other side, they are not posited but are rather
reflected into themselves; negation as negation is equality with itself, not
in its other, not reflected into its non-being.

3. Now keeping in mind that the determination of reflection is both
immanently reflected reference and positedness as well, its nature imme-
diately becomes more transparent. For, as positedness, the determination
is negation as such, a non-being as against another, namely, as against the
absolute immanent reflection or as against essence. But as self-reference, it
is reflected within itself.
This, the reflection of the determination,
and that positedness are distinct;
its positedness is rather the sublatedness of the determination
whereas its immanent reflectedness is its subsisting.
In so far as now the positedness is at the same time immanent reflection, the
determinateness of the reflection is the reference in it to its otherness.
It is not a determinateness that exists quiescent, one which would be referred
to an other in such a way that the referred term and its reference would be
different, each something existing in itself, each a something that excludes
its other and its reference to this other from itself. Rather, the determi-
nation of reflection is within it the determinate side and the reference of
this determinate side as determinate, that is, the reference to its negation.
Quality, through its reference, passes over into another; its alteration begins
in its reference. The determination of reflection, on the contrary, has taken
its otherness back into itself.
It is positedness, negation which has however deflected
the reference to another into itself, and negation which,
equal to itself, is the unity of itself and its other,
and only through this is an essentiality.
It is, therefore, positedness, negation,
but as reflection into itself it is at the same time
the sublatedness of this positedness, infinite reference to itself.

CHAPTER 2

The essentialities or the determinations of reflection

Reflection is determined reflection;
accordingly, essence is determined essence, or it is essentiality.

Reflection is the shining of essence within itself.

Essence, as infinite immanent turning back is
not immediate simplicity, but negative simplicity;
it is a movement across moments that are distinct,
is absolute mediation with itself.
But in these moments it shines;
the moments are, therefore, themselves determinations reflected into themselves.

First, essence is simple self-reference, pure identity. This is its determi-
nation, one by which it is rather the absence of determination.

Second, the specifying determination is difference,
difference which is either external or indefinite,
diversity in general, or opposed diversity or opposition.

Third, as contradiction this opposition is reflected into itself
and returns to its foundation.

A. IDENTITY

1. Essence is simple immediacy as sublated immediacy.
Its negativity is its being;
it is equal to itself in its absolute negativity
by virtue of which otherness and reference to other
have as such simply disappeared into pure self-equality.
Essence is therefore simple self-identity.

This self-identity is the immediacy of reflection.
It is not that self-equality which being is,
or also nothing, but a self-equality which,
in producing itself as unity,
does not produce itself over again, as from another,
but is a pure production, from itself and in itself,
essential identity.

It is not, therefore, abstract identity or
an identity which is the result of a relative negation preceding it,
one that separates indeed what it distinguishes from it
but, for the rest, leaves it existing outside it,
the same after as before.

Being, and every determinateness of being,
has rather sublated itself not relatively,
but in itself, and this simple negativity,
the negativity of being in itself, is
the identity itself.

In general, therefore, it is still the same as essence.

Thus the result of this consideration is this:
(1) the principle of identity or contradiction,
when meant to express merely abstract identity in opposition
to difference as a truth, is not a law of thought but expresses rather the
opposite of it;
(2) these two principles contain more than is meant by them,
namely this opposite, absolute difference itself.

B. DIFFERENCE

1. Absolute difference

Difference is the negativity that reflection possesses in itself, the nothing
which is said in identity discourse, the essential moment of identity itself
which, as the negativity of itself, at the same time determines itself and is
differentiated from difference.

1. This difference is difference in and for itself, absolute difference, the
difference of essence. – It is difference in and for itself, not difference through
something external but self-referring, hence simple, difference. – It is essen-
tial that we grasp absolute difference as simple. In the absolute difference
of A and not-A from each other, it is the simple “not” which, as such,
constitutes the difference. Difference itself is a simple concept. “In this,”
so it is said, “two things differ, in that etc.” – “In this,” that is, in one and
the same respect, relative to the same basis of determination. It is the dif-
ference of reflection, not the otherness of existence. One existence and another
existence are posited as lying outside each other; each of the two existences
thus determined over against each other has an immediate being for itself.
The other of essence, by contrast, is the other in and for itself, not the other
of some other which is to be found outside it; it is simple determinateness
in itself. Also in the sphere of existence did otherness and determinateness
prove to be of this nature, simple determinateness, identical opposition;
but this identity showed itself only as the transition of a determinateness
into the other. Here, in the sphere of reflection, difference comes in as
reflected, so posited as it is in itself.
2. Difference in itself is the difference that refers itself to itself; thus
it is the negativity of itself, the difference not from another but of itself
from itself; it is not itself but its other. What is different from differ-
ence, however, is identity. Difference is, therefore, itself and identity.
The two together constitute difference; difference is the whole and its
moment. – One can also say that difference, as simple difference, is no
difference; it is such only with reference to identity; even better, that as dif-
ference it entails itself and this reference equally. – Difference is the whole
and its own moment, just as identity equally is its whole and its moment. –
This is to be regarded as the essential nature of reflection and as the deter-
mined primordial origin of all activity and self-movement. – Both difference
and identity make themselves into moment or positedness because, as reflec-
tion, they are negative self-reference.
Difference, thus as unity of itself and of identity, is internally determined
difference. It is not the transition into another, not reference to another
outside it; it has its other, identity, within, and in like manner identity, in
being included in the determination of difference, has not lost itself in it as
its other but retains itself therein – is the reflection-into-itself of difference,
its moment.
3. Difference has both these moments, identity and difference; thus
the two are both a positedness, determinateness. But in this positedness
each refers to itself. The one, identity, is itself immediately the moment of
immanent reflection; but no less is the other, difference, difference in itself,
reflected difference. Difference, inasmuch as it has two such moments
which are themselves reflections into themselves, is diversity.

2. Diversity

1. Identity internally breaks apart into diversity because, as absolute dif-
ference in itself, it posits itself as the negative of itself and these, its two
moments (itself and the negative of itself ), are reflections into themselves,
are identical with themselves; or precisely because it itself immediately sub-
lates its negating and is in its determination reflected into itself. The different
subsists as diverse, indifferent to any other, because it is identical with itself,
because identity constitutes its base and element; or, the diverse remains
what it is even in its opposite, identity.
Diversity constitutes the otherness as such of reflection. The other of
existence has immediate being, where negativity resides, for its foundation.
But in reflection it is self-identity, the reflected immediacy, that constitutes
the subsistence of the negative and its indifference.
The moments of difference are identity and difference itself. These
moments are diverse when reflected into themselves, referring themselves to
themselves; thus, in the determination of identity, they are only self-referring;
identity is not referred to difference, nor is difference referred to identity;
hence, inasmuch as each of these moments is referred only to itself, the two
are not determined with respect to each other. – Now because in this way
the two are not differentiated within, the difference is external to them. The
diverse moments, therefore, conduct themselves with respect to each other,
not as identity and difference, but only as moments different in general,
indifferent to each other and to their determinateness.
2. In diversity, as the indifference of difference, reflection has in general
become external; difference is only a positedness or as sublated, but is itself the
whole reflection. – On closer consideration, both, identity and difference
are reflections, as we have just established; each is the unity of it and
its other, each is the whole. But the determinateness, to be only identity
or only difference, is thus a sublated something. They are not, therefore,
qualities, since their determinateness, because of the immanent reflection,
is at the same time only as negation. What we have is therefore this
duplicity, immanent reflection as such and determinateness as negation or
positedness. Positedness is the reflection that is external to itself; it is negation
as negation and consequently, indeed in itself self-referring negation and
immanent reflection, but only in itself, implicitly; its reference is to a
something external.
Reflection in itself and external reflection are thus the two determi-
nations in which the moments of difference, identity and difference, are
posited. They are these moments themselves as they have determined them-
selves at this point. – Immanent reflection is identity, but determined to
be indifferent to difference, not to have difference at all but to conduct
itself towards difference as identical with itself; it is diversity. It is identity
that has so reflected itself into itself that it truly is the one reflection of
the two moments into themselves; both are immanent reflections. Identity
is this one reflection of the two, the identity which has difference within
it only as an indifferent difference and is diversity in general. – External
reflection, on the contrary, is their determinate difference, not as absolute
immanent reflection, but as a determination towards which the implicitly
present reflection is indifferent; its two moments, identity and difference
themselves, are thus externally posited, are not determinations that exist in
and for themselves.
Now this external identity is likeness, and external difference is
unlikeness. – Likeness is indeed identity, but only as a positedness, an
identity which is not in and for itself. – Unlikeness is equally difference,
but an external difference which is not, in and for itself, the difference of
the unlike itself. Whether something is like or unlike something else is
not the concern of either the like of the unlike; each refers only to itself,
each is in and for itself what it is; identity or non-identity, in the sense of
likeness or unlikeness, depend on the point of view of a third external to
them.
3. External reflection connects diversity by referring it to likeness and
unlikeness. This reference, which is a comparing, moves back and forth from
likeness to unlikeness and from unlikeness to likeness. But this back and
forth referring of likeness and unlikeness is external to these determinations
themselves; moreover, they are not referred to each other, but each, for itself,
is referred to a third. In this alternation, each immediately stands out on its
own. – External reflection is as such external to itself; determinate difference
is negated absolute difference; it is not simple difference, therefore, not an
immanent reflection, but has this reflection outside it; hence its moments
come apart and both refer, each also outside the other, to the immanent
reflection confronting them.

In reflection thus alienated from itself, likeness and unlikeness present
themselves, therefore, as themselves unconnected, and reflection keeps them
apart, for it refers them to one and the same something by means of “in so
far,” “from this side or that,” and “from this view or that.” Thus diverse
things that are one and the same, when likeness and unlikeness are said of
them, are from one side like each other, but from another side unlike, and
in so far as they are alike, to that extent they are not unlike. Likeness thus
refers only to itself, and unlikeness is equally only unlikeness.
Because of this separation from each other, they sublate themselves.
Precisely that which should save them from contradiction and dissolution,
namely that something is like another in one respect but unlike in another –
precisely this keeping of likeness and unlikeness apart, is their destruction.
For both are determinations of difference; they are references to each other,
each intended to be what the other is not; the like is not the unlike, and the
unlike is not the like; both have this connecting reference essentially, and
have no meaning outside it; as determinations of difference, each is what it
is as different from its other. But because of their indifference to each other,
the likeness is referred to itself, and similarly is unlikeness a point of view of
its own and a reflection unto itself; each, therefore, is like itself; difference
has vanished, since they have no determinateness to oppose them; in other
words, each is consequently only likeness.

Accordingly, this indifferent viewpoint or the external difference sublates
itself and it is in itself the negativity of itself. It is the negativity which in
comparing belongs to that which does the comparing. This latter oscillates
from likeness to unlikeness and back again; hence it lets the one disappear
into the other and is in fact the negative unity of both. This negative
unity transcends at first what is compared as well as the moments of the
comparing as a subjective operation that falls outside them. But the result
is that this unity is in fact the nature of likeness and unlikeness themselves.
Even the independent viewpoint that each of these is, is rather the self-
reference that sublates their distinctness and so, too, themselves.
From this side, as moments of external reflection and as external to
themselves, likeness and unlikeness disappear together into their likeness.
But this, their negative unity, is in addition also posited in them; for their
reflection implicitly exists outside them, that is, they are the likeness and
unlikeness of a third, of another than they themselves are. Thus the like is
not the like of itself, and the unlike, as the unlike not of itself but of an
unlike to it, is itself the like. The like and the unlike is each therefore the
unlike of itself. Each is thereby this reflection: likeness, that it is itself and
the unlikeness; unlikeness, that it is itself and the likeness.
Likeness and unlikeness constituted the side of positedness as against
what is being compared or the diverse which, as contrasted with them,
had determined itself as implicitly existent reflection. But this positedness
has consequently equally lost its determinateness as against this reflection.
Likeness and unlikeness, the determinations of external reflection, are pre-
cisely the merely implicitly existent reflection which the diverse as such was
supposed to be, its only indeterminate difference. Implicitly existent reflec-
tion is self-reference without negation, abstract self-identity and therefore
positedness itself. – The merely diverse thus passes over through the posit-
edness into negative reflection. The diverse is difference which is merely
posited, hence a difference which is no difference, hence a negation that
negates itself within. Likeness and unlikeness themselves, the positedness,
thus return through indifference or through implicitly existing reflection
back into negative unity with themselves, into the reflection which is the
implicit difference of likeness and unlikeness. Diversity, the indifferent sides
of which are just as much simply and solely moments of a negative unity, is
opposition.

3. Opposition

In opposition, the determinate reflection, difference, is brought to comple-
tion. Opposition is the unity of identity and diversity; its moments are
diverse in one identity, and so they are opposites.
Identity and difference are the moments of difference as held inside dif-
ference itself; they are reflected moments of its unity. Likeness and unlikeness
are instead the externalized reflection; their self-identity is not only the
indifference of each towards the other differentiated from it, but towards
being-in-and-for-itself as such; theirs is a self-identity that contrasts with
identity reflected into itself, hence an immediacy which is not reflected into
itself. The positedness of the sides of external reflection is therefore a being,
just as their non-positedness is a non-being.
On closer consideration, the moments of opposition are positedness
reflected into itself or determination in general. Positedness is likeness and
unlikeness; these two, reflected into themselves, constitute the determi-
nations of opposition. Their immanent reflection consists in that each is
within it the unity of likeness and unlikeness. Likeness is only in a reflec-
tion which compares according to the unlikeness and is therefore mediated
by its other indifferent moment; similarly, unlikeness is only in the same
reflective reference in which likeness is. – Each of these moments, in its
determinateness, is therefore the whole. It is the whole because it also con-
tains its other moment; but this, its other, is an indifferent existent; thus
each contains a reference to its non-being, and it is reflection-into-itself, or
the whole, only as essentially referring to its non-being.
This self-likeness, reflected into itself and containing the reference to
unlikeness within it, is the positive; and the unlikeness that contains within
itself the reference to its non-being, to likeness, is the negative. – Or again,
both are positedness; now in so far as the differentiated determinateness is
taken as a differentiated determinate reference of positedness to itself, oppo-
sition is, on the one hand, positedness reflected into its likeness with itself;
and, on the other hand, it is the same positedness reflected into its inequal-
ity with itself: the positive and the negative. – The positive is positedness as
reflected into self-likeness; but what is reflected is positedness, that is, the
negation as negation, and so this immanent reflection has the reference to
the other for its determination. The negative is positedness as reflected into
unlikeness; but positedness is the unlikeness itself, and so this reflection is
therefore the identity of unlikeness with itself and absolute self-reference. –
Each, therefore, equally has the other in it: positedness reflected into self-
likeness has the unlikeness; and positedness reflected into self-unlikeness,
the likeness.
The positive and the negative are thus the sides of opposition that
have become self-subsisting. They are self-subsisting because they are the
reflection of the whole into itself, and they belong to opposition in so far
as the latter is determinateness which, as the whole, is reflected into itself.
Because of their self-subsistence, the opposition which they constitute is
implicitly determinate. Each is itself and its other; for this reason, each
has its determinateness not in an other but within. – Each refers itself to
itself only as referring itself to its other. This has a twofold aspect. Each
is the reference to its non-being as the sublating of this otherness in itself;
its non-being is thus only a moment in it. But, on the other hand, here
positedness has become a being, an indifferent subsistence; the other of
itself which each contains is therefore also the non-being of that in which
it should be contained only as a moment. Each is, therefore, only to the
extent that its non-being is, the two in an identical reference.
The determinations which constitute the positive and the negative con-
sist, therefore, in that the positive and the negative are, first, absolute
moments of opposition; their subsistence is indivisibly one reflection; it is
one mediation in which each is by virtue of the non-being of its other,
hence by virtue of its other or its own non-being. – Thus they are simply
opposites; or each is only the opposite of the other; the one is not yet the
positive and the other not yet the negative, but both are negative with
respect to each other. Each, therefore, simply is, first, to the extent that the
other is; it is what it is by virtue of the other, by virtue of its own non-being;
it is only positedness. Second, it is to the extent that the other is not; it is what
it is by virtue of the non-being of the other; it is reflection into itself. – The
two, however, are both the one mediation of opposition as such in which
they simply are only posited moments.
Moreover, this mere positedness is reflected into itself in general and,
according to this moment of external reflection, the positive and the negative
are indifferent towards this first identity where they are only moments; or
again, because that first reflection is the positive’s and the negative’s own
reflection into itself, each is indifferent towards its reflection into its non-
being, towards its own positedness. The two sides are thus merely diverse,
and because their determinateness – that they are positive or negative –
constitutes their positedness as against each other, each is not specifically so
determined internally but is only determinateness in general; to each side,
therefore, there belongs indeed one of the two determinacies, the positive
or the negative; but the two can be interchanged, and each side is such as
can be taken equally as positive or negative.
But, in third place, the positive and the negative are not only a posited
being, nor are they something merely indifferent, but their positedness, or
the reference to the other in the one unity which they themselves are not, is
rather taken back into each. Each is itself positive and negative within; the
positive and the negative are the determination of reflection in and for
itself; only in this reflection of the opposite into itself is the opposite either
positive or negative. The positive has within it the reference to the other in
which the determinateness of the positive consists. And the same applies
to the negative: it is not negative as contrasted with another but has the
determinateness by which it is negative within.
Each is thus self-subsistent unity existing for itself. The positive is indeed
a positedness, but in such a way that the positedness is for it posited being
as sublated. It is the non-opposed, the sublated opposition, but as the side
of the opposition itself. – As positive, it is indeed a something which is
determined with reference to an otherness, but in such a way that its nature
is not to be something posited; it is the immanent reflection that negates
otherness. But its other, the negative, is itself no longer positedness or a
moment but itself a self-subsisting being; and so the negating reflection
of the positive is internally determined to exclude this being, which is its
non-being, from itself.
Thus the negative, as absolute reflection, is not the immediate negative
but is the negative as sublated positedness, the negative in and for itself
which positively rests upon itself. As immanent reflection, it negates its
reference to its other; its other is the positive, a self-subsisting being –
hence its negative reference to this positive is the excluding of it from
itself. The negative is the independently existing opposite, over against the
positive which is the determination of the sublated opposition – the whole
opposition resting upon itself, opposed to the self-identical positedness.
The positive and the negative are such, therefore, not just in themselves,
but in and for themselves. They are in themselves positive and negative when
they are abstracted from their excluding reference to the other and are taken
only in accordance with their determination. Something is in itself positive
or negative when it is not supposed to be determined as positive or negative
merely in contrast with the other. But the positive and the negative, taken
not as a positedness and hence not as opposed, are each an immediate, being
and non-being. They are, however, moments of opposition: their in-itself
constitutes only the form of their immanent reflectedness. Something is
said to be positive in itself, outside the reference to something negative, and
something negative in itself, outside the reference to something negative: in
this determination, merely the abstract moment of this reflectedness is held
on to. However, to say that the positive and the negative exist in themselves
essentially implies that to be opposed is not a mere moment, nor that it is
just a matter of comparison, but that it is the determination of the sides
themselves of the opposition. The sides, as positive or negative in themselves,
are not, therefore, outside the reference to the other; on the contrary, this
reference, precisely as exclusive, constitutes their determination or their in-
itselfness; in this, therefore, they are at the same time in and for themselves.

C. CONTRADICTION

1. Difference in general contains both its sides as moments; in diversity, these
sides fall apart as indifferent to each other; and in opposition as such, they
are the moments of difference, each determined by the other and hence
only moments. But in opposition these moments are equally determined
within, indifferent to each other and mutually exclusive, self-subsisting
determinations of reflection.
One is the positive and the other the negative, but the former as a positive
which is such within, and the latter as a negative which is such within. Each
has indifferent self-subsistence for itself by virtue of having the reference to
its other moment within it; each moment is thus the whole self-contained
opposition. – As this whole, each moment is self-mediated through its other
and contains this other. But it is also self-mediated through the non-being of
its other and is, therefore, a unity existing for itself and excluding the other
from itself.
Since the self-subsisting determination of reflection excludes the other
in the same respect as it contains it and is self-subsisting for precisely
this reason, in its self-subsistence the determination excludes its own self-
subsistence from itself. For this self-subsistence consists in that it contains
the determination which is other than it in itself and does not refer to
anything external for just this reason; but no less immediately in that it is
itself and excludes from itself the determination that negates it. And so it
is contradiction.
Difference as such is already implicitly contradiction; for it is the unity
of beings which are, only in so far as they are not one – and it is the
separation of beings which are, only in so far as they are separated in the
same reference connecting them. The positive and the negative, however, are
the posited contradiction, for, as negative unities, they are precisely their
self-positing and therein each the sublating of itself and the positing of
its opposite. – They constitute determining reflection as exclusive; for the
excluding is one act of distinguishing and each of the distinguished beings,
as exclusive, is itself the whole act of excluding, and so each excludes itself
internally.
If we look at the two self-subsisting determinations of reflection on
their own, the positive is positedness as reflected into likeness with itself –
positedness which is not reference to another, hence subsistence inasmuch
as the positedness is sublated and excluded. But with this the positive makes
itself into the reference of a non-being – into a positedness. – In this way the
positive is contradiction – in that, as the positing of self-identity by the
excluding of the negative, it makes itself into a negative, hence into the other
which it excludes from itself. This last, as excluded, is posited free of the
one that excludes; hence, as reflected into itself and itself as excluding. The
reflection that excludes is thus the positing of the positive as excluding
the other, so that this positing immediately is the positing of its other
which excludes it.
This is the absolute contradiction of the positive; but it is immediately
the absolute contradiction of the negative; the positing of both in one
reflection. – Considered in itself as against the positive, the negative is
positedness as reflected into unlikeness to itself, the negative as negative. But
the negative is itself the unlike, the non-being of another; consequently,
reflection is in its unlikeness its reference rather to itself. – Negation in
general is the negative as quality or immediate determinateness; but taken
as negative, it is referred to the negative of itself, to its other. If this second
negative is taken only as identical with the first, then it is also only imme-
diate, just like the first; they are not taken, therefore, as each the other of
the other, hence not as negatives: the negative is not at all an immediate.
But now, since each is moreover equally the same as what the other is,
this reference connecting them as unequal is just as much their identical
connection.
This is therefore the same contradiction which the positive is, namely
positedness or negation as self-reference. But the positive is only implicitly
this contradiction, is contradiction only in itself; the negative, on the
contrary, is the posited contradiction; for in its reflection into itself, as a
negative which is in and for itself or a negative which is identical with
itself, its determination is to be the not-identical, the exclusion of identity.
The negative is this, to be identical with itself over against identity, and
consequently, because of this excluding reflection, to exclude itself from
itself.
The negative is therefore the whole opposition – the opposition which,
as opposition, rests upon itself; distinction that absolutely does not refer
itself to another; distinction which, as opposition, excludes identity from
itself, but thereby also excludes itself, for as reference to itself it determines
itself as the very identity which it excludes.
2. Contradiction resolves itself.
In the self-excluding reflection we have just considered, the positive and
the negative, each in its self-subsistence, sublates itself; each is simply the
passing over, or rather the self-translating of itself into its opposite. This
internal ceaseless vanishing of the opposites is the first unity that arises by
virtue of contradiction; it is the null.
But contradiction does not contain merely the negative; it also contains
the positive; or the self-excluding reflection is at the same time positing
reflection; the result of contradiction is not only the null. – The positive
and the negative constitute the positedness of the self-subsistence; their own
self-negation sublates it. It is this positedness which in truth founders to the
ground in contradiction.
The immanent reflection by virtue of which the sides of opposition are
turned into self-subsistent self-references is, first of all, their self-subsistence
as distinct moments; thus they are this self-subsistence only in themselves,
for they are still opposites, and that they are in themselves self-subsistent
constitutes their positedness. But their excluding reflection sublates this
positedness, turns them into self-subsistent beings existing in and for them-
selves, such as are self-subsistent not only in themselves but by virtue of their
negative reference to their other; in this way, their self-subsistence is also
posited. But, further, by thus being posited as self-subsistent, they make
themselves into a positedness. They fate themselves to founder, since they
determine themselves as self-identical, yet in their self-identity they are
rather the negative, a self-identity which is reference-to-other.
However, on closer examination, this excluding reflection is not only
this formal determination. It is self-subsistence existing in itself, and the
sublating of this positedness – is only through this sublating a unity that
exists for itself and is in fact self-subsistent. Of course, through the sublating
of otherness or positedness, positedness or the negative of an other is indeed
present again. But in fact, this negation is not just a return to the first
immediate reference to the other, is not positedness as sublated immediacy,
but positedness as sublated positedness. The excluding reflection of self-
subsistence, since it is excluding, makes itself a positedness but is just as
much the sublation of its positedness. It is sublating reference to itself; in
that reference, it first sublates the negative and it secondly posits itself as a
negative, and it is only this posited negative that it sublates; in sublating
the negative, it both posits and sublates it at the same time. In this way the
exclusive determination is itself that other of itself of which it is the negation;
the sublation of this positedness is not, therefore, once more positedness
as the negative of an other, but is self-withdrawal, positive self-unity. Self-
subsistence is thus unity that turns back into itself by virtue of its own
negation, for it turns into itself through the negation of its positedness. It
is the unity of essence – to be identical with itself through the negation not
of an other, but of itself.
3. According to this positive side, since self-subsistence in opposition,
as excluding reflection, makes itself into a positedness and equally sublates
this positedness, not only has opposition foundered but in foundering it has
gone back to its foundation, to its ground. – The excluding reflection of the
self-subsisting opposition turns it into a negative, something only posited;
it thereby reduces its formerly self-subsisting determinations, the positive
and the negative, to determinations which are only determinations; and the
positedness, since it is now made into positedness, has simply gone back
to its unity with itself; it is simple essence, but essence as ground. Through
the sublating of the determinations of essence, which are in themselves
self-contradictory, essence is restored, but restored in the determination of
an exclusive, reflective unity – a simple unity which determines itself as
negation, but in this positedness is immediately like itself and withdrawn
into itself.
In the first place, therefore, because of its contradiction, the self-
subsisting opposition goes back into a ground; this opposition is what comes
first, the immediate from which the beginning is made, while the sublated
opposition or the sublated positedness is itself a positedness. Accordingly,
essence is as ground a positedness, something that has become. But conversely,
only this has been posited, namely that the opposition or the posited-
ness is something sublated, only is as positedness. As ground, therefore,
essence is excluding reflection because it makes itself into a positedness;
because the opposition from which the start was just now made and
was the immediate is the merely posited determinate self-subsistence of
essence; because opposition only sublates itself within, whereas essence is
in its determinateness reflected into itself. As ground, therefore, essence
excludes itself from itself, it posits itself; its positedness – which is what
is excluded – is only as positedness, as identity of the negative with
itself. This self-subsistent is the negative posited as the negative, something
self-contradictory which, consequently, remains in the essence as in its
ground.
The resolved contradiction is therefore ground, essence as unity of the
positive and the negative. In opposition, determinateness has progressed
to self-subsistence; but ground is this self-subsistence as completed; in it,
the negative is self-subsistent essence, but as negative; and, as self-identical
in this negativity, ground is thus equally the positive. In ground, therefore,
opposition and its contradiction are just as much removed as preserved.
Ground is essence as positive self-identity which, however, at the same time
refers itself to itself as negativity and therefore determines itself, making
itself into an excluded positedness; but this positedness is the whole self-
subsisting essence, and essence is ground, self-identical in its negation
and positive. The self-contradictory self-subsistent opposition was itself,
therefore, already ground; all that was added to it was the determination
of self-unity which emerges as each of the self-subsisting opposites sublates
itself and makes itself into its other, thereby founders and sinks to the
ground but therein also reunites itself with itself; thus in this foundering,
that is, in its positedness or in the negation, it rather is for the first time
the essence that is reflected into itself and self-identical.

The upshot of this examination of the nature of contradiction is that,
if a contradiction can be pointed out in something, by itself this is still
not, as it were, a blemish, not a defect or failure. On the contrary, every
determination, anything concrete, every concept, is essentially a unity of
distinguished and distinguishable elements which, by virtue of the deter-
minate, essential difference, pass over into elements which are contradictory.
This contradictoriness of course resolves itself into nothing: it goes back
into its negative unity. A thing, a subject, a concept, is then precisely this
negative unity; it is something inherently self-contradictory, but it is no
less the resolved contradiction; it is the ground which contains the determi-
nations it bears. The thing, the subject or the concept, each as reflected
into itself within its sphere, is their contradiction as resolved; but the whole
sphere of each is in turn determinate, diverse, and therefore finite, and this
means contradictory. This sphere is not itself the resolution of its higher
contradiction but has yet a higher sphere for its negative unity, for its
ground. Finite things, in their indifferent variety, are therefore just this: to
be contradictory, internally fractured and bound to return to their ground. –
As we shall see in due course, 9 the true inference from the finite and acci-
dental to an absolutely necessary being does not consist in inferring the
latter from that finite and accidental as a being which is and remains the
ground of the inference, but from it as a being which is only “falling” [as
accidentality, from the Latin cadere, immediately conveys], a being inher-
ently self-contradictory; or rather, the inference consists in demonstrating
that accidental being makes in itself its return to its ground and is there
sublated – in addition, that in this return to the ground it posits the latter
in such a manner that it makes itself rather into a positedness. In customary
inference, the being of the finite appears to be the ground of the absolute;
because the finite is, the absolute is. But the truth is that the absolute is
because the finite is the immanently self-contradictory opposition, because
it is not. In the former meaning, the conclusion is that “the being of the
finite is the being of the absolute”; but in the latter, that “the non-being of
the finite is the being of the absolute.”

CHAPTER 3

Ground

Essence determines itself as ground.

Just as nothing is at first in simple immediate unity with being, so here
too the simple identity of essence is at first in simple unity with its absolute
negativity. Essence is only this negativity which is pure reflection. It is this
pure reflection as the turning back of being into itself; hence it is determined,
in itself or for us, as the ground into which being resolves itself. But this
determinateness is not posited by the essence itself; in other words, essence is
not ground precisely because it has not itself posited this determinateness
that it possesses. Its reflection, however, consists in positing itself as what
it is in itself, as a negative, and in determining itself. The positive and
the negative constitute the essential determination in which essence is lost
in its negation. These self-subsisting determinations of reflection sublate
themselves, and the determination that has foundered to the ground is the
true determination of essence.

Consequently, ground is itself one of the reflected determinations of essence,
but it is the last, or rather, it is determination determined as sublated deter-
mination. In foundering to the ground, the determination of reflection
receives its true meaning – that it is the absolute repelling of itself within
itself; or again, that the positedness that accrues to essence is such only
as sublated, and conversely that only the self-sublating positedness is the
positedness of essence. In determining itself as ground, essence determines
itself as the not-determined, and only the sublating of its being determined
is its determining. – Essence, in thus being determined as self-sublating,
does not proceed from an other but is, in its negativity, identical with
itself.

Since the advance to the ground is made starting from determination
as an immediate first (is done by virtue of the nature of determination
itself that founders to the ground through itself ), the ground is at first
determined by that immediate first. But this determining is, on the one
hand, as the sublating of the determining, the merely restored, purified or
manifested identity of essence which the determination of reflection is in
itself; on the other hand, this negating movement is, as determining, the
first positing of that reflective determinateness that appeared as immediate
determinateness, but which is posited only by the self-excluding reflection
of ground and therein is posited as only something posited or sublated. –
Thus essence, in determining itself as ground, proceeds only from itself.
As ground, therefore, it posits itself as essence, and its determining consists
in just this positing of itself as essence. This positing is the reflection of
essence that sublates itself in its determining; on that side is a positing, on
this side is the positing of essence, hence both in one act.

Reflection is pure mediation in general; ground, the real mediation of
essence with itself. The former, the movement of nothing through nothing
back to itself, is the reflective shining of one in an other; but, because in
this reflection opposition does not yet have any self-subsistence, neither is
the one, that which shines, something positive, nor is the other in which it
reflectively shines something negative. Both are substrates, actually of the
imagination; they are still not self-referring. Pure mediation is only pure
reference, without anything being referred to. Determining reflection, for
its part, does posit such terms as are identical with themselves; but these
are at the same time only determined references. Ground, on the contrary,
is mediation that is real, since it contains reflection as sublated reflection;
it is essence that turns back into itself through its non-being and posits itself.
According to this moment of sublated reflection, what is posited receives
the determination of immediacy, of an immediate which is self-identical
outside its reference or its reflective shining. This immediacy is being as
restored by essence, the non-being of reflection through which essence
mediates itself. Essence returns into itself as it negates; therefore, in its
turning back into itself, it gives itself the determinateness that precisely
for this reason is the self-identical negative, is sublated positedness, and
consequently, as the self-identity of essence as ground, equally an existent.
The ground is, first, absolute ground – one in which the essence is first
of all the general substrate for the ground-connection. It then further
determines itself as form and matter and gives itself a content.
Second, it is determinate ground, the ground of a determinate content.
Because the ground-connection, in being realized, becomes as such exter-
nal, it passes over into conditioning mediation.
Third, ground presupposes a condition; but the condition equally pre-
supposes the ground; the unconditioned is the unity of the two, the fact
itself that, by virtue of the mediation of the conditioning reference, passes
over into concrete existence.

A. ABSOLUTE GROUND

a. Form and essence

The determination of reflection, inasmuch as this determination returns
into ground, is a first immediate existence in general from which the
beginning is made. But existence still has only the meaning of positedness
and essentially presupposes a ground, in the sense that it does not really
posit a ground; that the positing is a sublating of itself; that it is rather the
immediate that is posited, and the ground the non-posited. As we have
seen, this presupposing is the positing that rebounds on that which posits;
as sublated determinate being, the ground is not an indeterminate but is
rather essence determined through itself, but determined as indeterminate or
as sublated positedness. It is essence that in its negativity is identical with itself.
The determinateness of essence as ground is thus twofold: it is the deter-
minateness of the ground and of the grounded. It is, first, essence as ground,
essence determined to be essence as against positedness, as non-positedness.
Second, it is that which is grounded, the immediate that, however, is not
anything in and for itself: is positedness as positedness. Consequently, this
positedness is equally identical with itself, but in an identity which is that
of the negative with itself. The self-identical negative and the self-identical
positive are now one and the same identity. For the ground is the self-identity
of the positive or even also of positedness; the grounded is positedness as
positedness, but this its reflection-into-itself is the identity of the ground. –
This simple identity, therefore, is not itself ground, for the ground is essence
posited as the non-posited as against positedness. As the unity of this deter-
minate identity (the ground) and of the negative identity (the grounded),
it is essence in general distinct from its mediation.
For one thing, this mediation, compared with the preceding reflections
from which it derives, is not pure reflection, which is undistinguished from
essence and still does not have the negative in it, consequently also does
not as yet contain the self-subsistence of the determinations. These have
their subsistence, rather, in the ground understood as sublated reflection. –
And it is also not the determining reflection whose determinations have
essential self-subsistence, for that reflection has foundered, has sunk to the
ground, and in the unity of the latter the determinations are only posited
determinations. – This mediation of the ground is thus the unity of pure
reflection and determining reflection; their determinations or that which
is posited has self-subsistence, and conversely the self-subsistence of the
determinations is a posited subsistence. Since this subsistence of the deter-
minations is itself posited or has determinateness, the determinations are
consequently distinguished from their simple identity, and they constitute
the form as against essence.
Essence has a form and determinations of this form. Only as ground
does it have a fixed immediacy or is substrate. Essence as such is one with
its reflection, inseparable from its movement. It is not essence, therefore,
through which this movement runs its reflective course; nor is essence
that from which the movement begins, as from a starting point. It is this
circumstance that above all makes the exposition of reflection especially
difficult, for strictly speaking one cannot say that essence returns into itself,
that essence shines in itself, for essence is neither before its movement nor
in the movement: this movement has no substrate on which it runs its
course. A term of reference arises in the ground only following upon the
moment of sublated reflection. But essence as the referred-to term is deter-
minate essence, and by virtue of this positedness it has form as essence. –
The determinations of form, on the contrary, are now determinations in
the essence; the latter lies at their foundation as an indeterminate which
in its determination is indifferent to them; in it, they are reflected into
themselves. The determinations of reflection should have their subsistence
in them and be self-subsistent. But their self-subsistence is their dissolu-
tion, which they thus have in an other; but this dissolution is itself this
self-identity or the ground of the subsistence that they give to themselves.
Everything determinate belongs in general to form; it is a form deter-
mination inasmuch as it is something posited and hence distinguished
from that of which it is the form. As quality, determinateness is one with
its substrate, being; being is the immediate determinate, not yet distinct
from its determinateness or, in this determinateness, still unreflected into
itself, just as the determinateness is, therefore, an existent determinate-
ness, not yet one that is posited. – Moreover, the form determinations
of essence are, in their more specific determinateness, the previously con-
sidered moments of reflections: 12 identity and difference, the latter as both
diversity and opposition. But also the ground-connection belongs among
these form determinations of essence, because through it, though itself the
sublated determination of reflection, essence is at the same time as posited.
By contrast, the identity that has the ground immanent in it does not per-
tain to form, because positedness, as sublated and as such (as ground and
grounded), is one reflection, and this reflection constitutes essence as simple
substrate which is the subsistence of form. But in ground this subsistence is
posited, or this essence is itself essentially as determinate and, consequently,
is in turn also the moment of the ground-connection and form. – This is
the absolute reciprocal connecting reference of form and essence: essence
is the simple unity of ground and grounded but, in this unity, is itself
determined, or is a negative, and it distinguishes itself as substrate from
form, but at the same time it thereby becomes itself ground and moment
of form.
Form is therefore the completed whole of reflection; it also contains this
determination of reflection, that it is sublated; just like reflection, therefore,
it is one unity of its determining, and it is also referred to its sublatedness,
to another that is not itself form but in which the form is. As essential self-
referring negativity, in contrast with that simple negative, form is positing
and determining; simple essence, on the contrary, is indeterminate and
inert substrate in which the determinations of form have their subsistence
or their reflection into themselves. – External reflection normally halts at
this distinction of essence and form; the distinction is necessary, but the
distinguishing itself of the two is their unity, just as this unity of ground is
essence repelling itself from itself and making itself into positedness. Form
is absolute negativity itself or the negative absolute self-identity by virtue
of which essence is indeed not being but essence. This identity, taken
abstractly, is essence as against form, just as negativity, taken abstractly
as positedness, is the one determination of form. But this determination
has shown itself to be in truth the whole self-referring negativity which
within, as this identity, thus is simple essence. Consequently, form has
essence in its own identity, just as essence has absolute form in its negative
nature. One cannot therefore ask, how form comes to essence, for form is
only the internal reflective shining of essence, its own reflection inhabiting
it. Form equally is, within it, the reflection turning back into itself or the
identical essence; in its determining, form makes the determination into
positedness as positedness. – Form, therefore, does not determine essence,
as if it were truly presupposed, separate from essence, for it would then be
the unessential, constantly foundering determination of reflection; here it
rather is itself the ground of its sublating or the identical reference of its
determinations. That the form determines the essence means, therefore,
that in its distinguishing form sublates this very distinguishing and is the
self-identity that essence is as the subsistence of the determinations; form
is the contradiction of being sublated in its positedness and yet having
subsistence in this sublatedness; it is accordingly ground as essence which
is self-identical in being determined or negated.
These distinctions, of form and of essence, are therefore only moments
of the simple reference of form itself. But they must be examined and fixed
more closely. Determining form refers itself to itself as sublated positedness;
it thereby refers itself to its identity as to another. It posits itself as sublated;
it therefore pre-supposes its identity; according to this moment, essence is
the indeterminate to which form is an other. It is not the essence which is
absolute reflection within, but essence determined as formless identity: it
is matter.

b. Form and matter

Essence becomes matter in that its reflection is determined as relating itself
to essence as to the formless indeterminate. Matter, therefore, is the simple
identity, void of distinction, that essence is, with the determination that
it is the other of form. Hence it is the proper base or substrate of form,
since it constitutes the immanent reflection of the determinations of form,
or the self-subsistent term, to which such determinations refer as to their
positive subsistence.
If abstraction is made from every determination, from every form of
a something, matter is what is left over. Matter is the absolutely abstract.
(One cannot see, feel, etc. matter; what one sees or feels is a determinate
matter, that is, a unity of matter and form.) This abstraction from which
matter derives is not, however, an external removal and sublation of form;
it is rather the form itself which, as we have just seen, reduces itself by
virtue of itself to this simple identity.
Further, form presupposes a matter to which it refers. But for this reason
the two do not find themselves confronting each other externally and
accidentally; neither matter nor form derives from itself, is a se, or, in
other words, is eternal. Matter is indifferent with respect to form, but this
indifference is the determinateness of self-identity to which form returns
as to its substrate. Form pre-supposes matter for the very reason that it
posits itself as a sublated, hence refers to this, its identity, as to something
other. Contrariwise, form is presupposed by matter; for matter is not
simple essence, which immediately is itself absolute reflection, but is essence
determined as something positive, that is to say, which only is as sublated
negation. – But, on the other hand, since form posits itself as matter only in
sublating itself, hence in presupposing matter, matter is also determined as
groundless subsistence. Equally so, matter is not determined as the ground
of form; but rather, inasmuch as matter posits itself as the abstract identity
of the sublated determination of form, it is not that identity as ground,
and form is therefore groundless with respect to it. Form and matter are
consequently alike determined as not to be posited each by the other, each
not to be the ground of the other. Matter is rather the identity of the
ground and the grounded, as the substrate that stands over against this
reference of form. This determination of indifference that the two have in
common is the determination of matter as such and also constitutes their
reciprocal reference. The determination of form, that it is the connection
of the two as distinct, equally is also the other moment of the relating of
the two to each other. – Matter, determined as indifferent, is the passive as
contrasted to form, which is determined as the active. This latter, as self-
referring negative, is inherently contradiction, self-dissolving, self-repelling,
and self-determining. It refers to matter, and it is posited to refer to this
matter, which is its subsistence, as to another. Matter is posited, on the
contrary, as referring only to itself and as indifferent to the other; but,
implicitly, it does refer to the form, for it contains the sublated negativity
and is matter only by virtue of this determination. It refers to it as an other
only because form is not posited in it, because it is form only implicitly.
It contains form locked up inside it, and it is an absolute receptivity for
form only because it has the latter within it absolutely, because to be form
is its implicit vocation. 13 Hence matter must be informed, and form must
materialize itself; it must give itself self-identity or subsistence in matter.
2. Consequently, form determines matter, and matter is determined by
form. – Because form is itself absolute self-identity and hence implicitly
contains matter; and equally because matter in its pure abstraction or
absolute negativity possesses form within it, the activity of the form on the
matter and the reception by the latter of the form determination is only
the sublating of the semblance of their indifference and distinctness. Thus
the determination referring each to the other is the self-mediation of each
through its own non-being. But the two mediations are one movement,
and the restoration of their original identity is the inner recollection 14 of
their exteriorization.
First, form and matter pre-suppose each other. As we have seen, 15 this
only means that the one essential unity is negative self-reference, and that
it therefore splits, determined as an indifferent substrate in the essential
identity, and as determining form in essential distinction or negativity.
That unity of essence and form, the two opposed to each other as form
and matter, is the absolute self-determining ground. Inasmuch as this unity
differentiates itself, the reference connecting the two diverse terms, because
of the unity that underlies them, becomes a reference of reciprocal presup-
position.
Second, the form already is, as self-subsisting, self-sublating contradic-
tion; but it is also posited as in this way self-sublating, for it is self-subsisting
and at the same time essentially referred to another, and consequently it
sublates itself. Since it is itself two-sided, its sublating also has two sides. For
one, form sublates its self-subsistence and transforms itself into something
posited, something that exists in an other, and this other is in its case matter.
For the other, form sublates its determinateness vis-à-vis matter, sublates
its reference to it, consequently its positedness, and it thereby gives itself
subsistence. Its reflection in thus sublating its positedness is its own identity
into which it passes over. But since form at the same time externalizes this
identity and posits it over against itself as matter, that reflection of the
positedness into itself is a union with a matter in which it obtains subsis-
tence. In this union, therefore, it is equally both: is united with matter as
with something other (in accordance with the first side, viz. in that it makes
itself into a positedness), and, in this other, is united with its own identity.
The activity of form by which matter is determined consists, therefore,
in a negative relating of the form to itself. But, conversely, form thereby
negatively relates itself to matter also; the movement, however, by which
matter becomes determined is just as much the form’s own movement.
Form is free of matter, but it sublates its self-subsistence; but this, its self-
subsistence, is matter itself, for it is in this matter that it has its essential
identity. It makes itself into a positedness, but this is one and the same
as making matter into something determinate. – But, considered from
the other side, the form’s own identity is at the same time externalized,
and matter is its other; for this reason, because form sublates its own
self-subsistence, matter is also not determined. But matter only subsists
vis-à-vis form; as the negative sublates itself, so does the positive also. And
as the form sublates itself, the determinateness of matter that the latter has
vis-à-vis form also falls away – the determinateness, namely, of being the
indeterminate subsistence.
What appears here as the activity of form is, moreover, just as much the
movement that belongs to matter itself. The determination that implicitly
exists in matter, what matter is supposed to be, is its absolute negativity.
Through it matter does not just refer to form simply as to an other, but this
external other is the form rather that matter itself contains locked up within
itself. Matter is in itself the same contradiction that form contains, and this
contradiction, like its resolution, is only one. But matter is thus in itself
self-contradictory because, as indeterminate self-identity, it is at the same
time absolute negativity; it sublates itself within: its identity disintegrates
in its negativity while the latter obtains in it its subsistence. Since matter is
therefore determined by form as by something external, it thereby attains
its determination, and the externality of the relating, for both form and
matter, consists in that each, or rather in that the original unity of each,
in positing is at the same time presupposing: the result is that self-reference
is at the same time a reference to the self as sublated or is reference to its
other.
Third, through this movement of form and matter, the original unity of
the two is, on the one hand, restored; on the other hand, it is henceforth a
posited unity. Matter is just as much a self-determining as this determining
is for it an activity of form external to it; contrariwise, form determines
only itself, or has the matter that it determines within it, just much as in
its determining it relates itself to another; and both, the activity of form
and the movement of matter, are one and the same thing, only that the
former is an activity, that is, it is the negativity as posited, while the latter
is movement or becoming, the negativity as determination existing in itself.
The result, therefore, is the unity of the in-itself and positedness. Matter is
as such determined or necessarily has a form, and form is simply material,
subsistent form.
Inasmuch as form presupposes a matter as its other, it is finite. It is
not a ground but only the active factor. Equally so, matter, inasmuch as
it presupposes form as its non-being, is finite matter; it is not the ground
of its unity with form but is for the latter only the substrate. But neither
this finite matter nor the finite form have any truth; each refers to the
other, or only their unity is their truth. The two determinations return to
this unity and there they sublate their self-subsistence; the unity thereby
proves to be their ground. Consequently, matter is the ground of its form
determination not as matter but only inasmuch as it is the absolute unity
of essence and form; similarly, form is the ground of the subsistence of
its determinations only to the extent that it is that same one unity. But
this one unity, as absolute negativity, and more specifically as exclusive
unity, is, in its reflection, a presupposing; or again, that unity is one act,
of preserving itself as positedness in positing, and of repelling itself from
itself; of referring itself to itself as itself and to itself as to another. Or,
the act by which matter is determined by form is the self-mediation of
essence as ground, in one unity: through itself and through the negation of
itself.
Informed matter or form that possesses subsistence is now, not only this
absolute unity of ground with itself, but also unity as posited. The move-
ment just considered is the one in which the absolute ground has exhibited
its moments at once as self-sublating and consequently as posited. Or the
restored unity, in withdrawing into itself, has repelled itself from itself and
has determined itself; for its unity has been established through negation
and is, therefore, also negative unity. It is, therefore, the unity of form
and matter, as the substrate of both, but a substrate which is determi-
nate: it is formed matter, but matter at the same time indifferent to form
and matter, indifferent to them because sublated and unessential. This is
content.

c. Form and content

Form stands at first over against essence; it is then the ground-connection
in general, and its determinations are the ground and the grounded. It then
stands over against matter, and so it is determining reflection, and its deter-
minations are the determination of reflection itself and the subsistence of
the latter. Finally, it stands over against content, and then its determinations
are again itself and matter. What was previously the self-identical – at first
the ground, then subsistence in general, and finally matter – now passes
under the dominion of form and is once more one of its determinations.
Content has, first, a form and a matter that belong to it essentially; it
is their unity. But, because this unity is at the same time determinate or
posited unity, content stands over against form; the latter constitutes the
positedness and is the unessential over against content. The latter is therefore
indifferent towards form; form embraces both the form as such as well as
the matter, and content therefore has a form and a matter, of which it
constitutes the substrate and which are to it mere positedness.
Content is, second, what is identical in form and matter, so that these
would be only indifferent external determinations. They are positedness in
general, but a positedness that has returned in the content to its unity or its
ground. The identity of the content with itself is, therefore, in one respect
that identity which is indifferent to form, but in another the identity of
ground. The ground has at first disappeared into content; but content is
at the same time the negative reflection of the form determinations into
themselves; its unity, at first only the unity indifferent to form, is therefore
also the formal unity or the ground-connection as such. Content, therefore,
has this ground-connection as its essential form, and, contrariwise, the
ground has a content.
The content of the ground is therefore the ground that has returned
into its unity with itself; the ground is at first the essence that in its
positedness is identical with itself; as diverse from and indifferent to its
positedness, the ground is indeterminate matter; but as content it is at
the same time informed identity, and this form becomes for this reason a
ground-connection, since the determinations of its oppositions are posited
in the content also as negated. – Content is further determined within,
not like matter as an indifferent in general, but like informed matter, so
that the determinations of form have a material, indifferent subsistence.
On the one hand, content is the essential self-identity of the ground in its
positedness; on the other hand, it is posited identity as against the ground-
connection; this positedness, which is in this identity as determination of
form, stands over against the free positedness, that is to say, over against
the form as the whole connection of ground and grounded. This form is
the total positedness returning into itself; the other form, therefore, is only
the positedness as immediate, the determinateness as such.
The ground has thus made itself into a determinate ground in general,
and the determinateness is itself twofold: of form first, and of content
second. The former is its determinateness of being external to the content
as such, the content that remains indifferent to this external reference. The
latter is the determinateness of the content that the ground has.

B. DETERMINATE GROUND

a. Formal ground

The ground has a determinate content. For the form, as we have seen, 16 the
determinateness of content is the substrate, the simple immediate as against
the mediation of form. The ground is negatively self-referring identity
which, for this reason, makes itself into a positedness; it negatively refers to
itself because in its negativity it is identical with itself; this identity is the
substrate or the content which thus constitutes the indifferent or positive
unity of the ground-connection and, in this connection, is the mediating
factor.
In this content, the determinateness that the ground and the grounded
have over against one another has at first disappeared. The mediation,
however, is also negative unity. The negative implicit in that indifferent
substrate is this substrate’s immediate determinateness through which the
ground has a determinate content. But then, the negative is the negative
reference of form to itself. What has been posited sublates itself on its
side and returns to its ground; the ground, however, the essential self-
subsistence, refers negatively to itself and makes itself into a positedness.
This negative mediation of ground and grounded is the mediation that
belongs to form as such, formal mediation. Now both sides of form, because
each passes over into the other, thereby mutually posit themselves into one
identity as sublated; in this, they pre-suppose the identity. The latter is
the determinate content to which the formal mediation thus refers itself
through itself as to the positive mediating factor. That content is the
identical element of both, and because the two are distinct, yet in their
distinction each is the reference to the other, it is their subsistence, the
subsistence of each as the whole itself.
Accordingly, the result is that in the determinate ground we have the
following. First, a determinate content is considered from two sides, once
in so far as it is ground, then again in so far as it is grounded. The content
itself is indifferent to these forms; it is in each simply and solely one
determination. Second, the ground is itself just as much a moment of form
as what is posited by it; this is its identity according to form. It is a matter of
indifference which of the two determinations is made the first, whether the
transition is from the one as posited to the other as ground or from the one
as ground to the other as posited. The grounded, considered for itself, is
the sublating of itself; it thereby makes itself on the one side into a posited,
and is at the same time the positing of the ground. The same movement
is the ground as such; it makes itself into something posited, and thereby
becomes the ground of something, that is to say, is present therein both
as a posited and also first as ground. That there be a ground, of that the
posited is the ground, and, conversely, the ground is thereby the posited.
The mediation begins just as much from the one as from the other; each
side is just as much ground as posited, and each is the whole mediation
or the whole form. – Further, this whole form is itself, as self-identical,
the substrate of the two determinations that constitute the two sides of the
ground and the grounded; form and content are thus themselves one and
the same identity.
Because of this identity of the ground and the grounded, according both
to content and form, the ground is sufficient (the sufficiency being limited
to this relation); there is nothing in the grounded which is not in the ground.
Whenever one asks for a ground, one expects to see the same determination
which is the content doubled, once in the form of that which is posited, and
again in the form of existence reflected into itself, of essentiality.
Now inasmuch as in the determined ground, the ground and the
grounded are each the whole form, and their content, though determi-
nate, is nevertheless one and the same, the two sides of the ground do
not as yet have a real determination, do not have a different content; the
determinateness is only one simple determinateness that has yet to pass
over into the two sides; the determinate ground is present only in its pure
form, as formal ground. – Because the content is only this simple determi-
nateness, one that does not have in it the form of the ground-connection,
the determinateness is a self-identical content indifferent to form, and the
form is external to it; the content is other than the form.

b. Real ground

The determinateness of ground is, as we have seen, on the one hand deter-
minateness of the substrate or content determination; on the other hand,
it is the otherness in the ground-connection itself, namely the distinctness
of its content and the form; the connection of ground and grounded strays
in the content as an external form, and the content is indifferent to these
determinations. – But in fact the two are not external to each other; for
this is what the content is: to be the identity of the ground with itself in
the grounded, and of the grounded in the ground. The side of the ground
has shown itself to be itself a posited, and the side of the grounded to be
itself ground; each side is this identity of the whole within it. But since
they equally belong to form and constitute its determinate difference, each
is in its determinateness the identity of the whole with itself. Consequently,
each has a diverse content as against the other. – Or, considering the matter
from the side of the content, since the latter is the self-identity of the
ground-connection, it essentially possesses this difference of form within,
and is as ground something other than what it is as grounded.
Now the moment ground and grounded have a diverse content, the
ground-connection has ceased to be a formal one; the turning back to the
ground and the procession forward from ground to posited is no longer
a tautology; the ground is realized. Henceforth, whenever we ask for a
ground, we actually demand another content determination for it than the
determination of the content whose ground we are asking for.
This connection now determines itself further. For inasmuch as its two
sides are of different content, they are indifferent to each other; each is
an immediate, self-identical determination. Moreover, as referred to each
other as ground and grounded, the ground reflects itself in the other, as
in something posited by it, back to itself; the content on the side of the
ground, therefore, is equally in the grounded; the latter, as the posited, has
its self-identity and subsistence only in the ground. But besides this content
of the ground, the grounded also now possesses a content of its own and
is accordingly the unity of a twofold content. Now this unity, as the unity
of sides that are different, is indeed their negative unity; but since the two
determinations of content are indifferent to each other, that unity is only
their empty reference to each other, in itself void of content, and not their
mediation; it is a one or a something externally holding them together.
In the real grounding connection there is present, therefore, a twofold.
For one thing, the content determination which is ground extends
continuously 18 into the positedness, so that it constitutes the simple identity
of the ground and the grounded; the grounded thus contains the ground
fully within itself; their connection is one of undifferentiated essential
compactness. Anything else in the grounded added to this simple essence is,
therefore, only an unessential form, external determinations of the content
that, as such, are free from the ground and constitute an immediate man-
ifold. Of this unessential more, therefore, the essential is not the ground,
nor is it the ground of any connection between it and the unessential in
the grounded. The unessential is a positively identical element that resides
in the grounded but does not posit itself there in any distinctive form;
as self-referring content, it is rather an indifferent positive substrate. – For
another thing, that which in the something is linked with this substrate is an
indifferent content, but as the unessential side. The main thing is the con-
nection of the substrate and the unessential manifold. But this connection,
since the determinations that it connects are an indifferent content, is also
not a ground; true, one determination is determined as essential content
and the other as only unessential or as posited; but this form is to each,
as a self-referring content, an external one. The one of the something that
constitutes their connection is for this reason not a reference of form, but
only an external tie that does not hold the unessential manifold content as
posited; it too is therefore likewise only a substrate.
Ground, in determining itself as real, because of the diversity of the
content that constitutes its reality, thus breaks down into external deter-
minations. The two connections of the essential reality – content, as
the simple immediate identity of ground and grounded; and then the
something connecting distinct contents – are two different substrates. The
self-identical form of ground, according to which one and the same thing
is at one time the essential and at another the posited, has vanished. The
ground-connection has thus become external to itself.
Consequently, it is an external ground that now holds together a diversi-
fied content and determines what is ground and what is posited by it; this
determination is not to be found in the two-sided content itself. The real
ground is therefore the reference to another, on the one hand, of a content
to another content and, on the other, of the ground-connection itself (the
form) to another, namely to an immediate, to something not posited by it.

c. Complete ground

1. In real ground, ground as content and ground as connection are only
substrates. The former is only posited as essential and as ground; the con-
nection is what the grounded immediately is as the indeterminate substrate
of a diversified content, a linking of this content which is not the content’s
own reflection but is rather external and consequently a reflection which
is only posited. The real ground-connection is ground, therefore, rather as
sublated; consequently, it rather makes up the side of the grounded or of
the positedness. As positedness, however, the ground itself has now returned
to its ground; it is now something grounded: it has another ground. This
ground will therefore be so determined that, first, it is identical with the
ground by which it is grounded; both sides have in this determination one
and the same content; the two content determinations and their linkage in
a something are equally to be found in the new ground. But, second, the
new ground into which the previously merely posited and external link is
now sublated is the immanent reflection of this link: the absolute reference
of the two content determinations to each other.
Because real ground has itself thus returned to its ground, the identity
of ground and grounded or the formality of ground reasserts itself in it.
The newly arisen ground-connection is therefore the one which is complete,
which contains the formal and real ground in itself at the same time and
mediates the content determinations which in the real ground confronted
each other immediately.
2. Thus the ground-connection has more precisely determined itself as
follows. First, something has a ground; it contains the content determination
which is the ground and, in addition, a second determination as posited by
the ground. But, because of the indifference of content, the one determi-
nation is not ground in itself, nor is the other in itself one that is grounded
by the first; this connection of ground and grounded is rather sublated in
the immediacy of their content, is posited, and as such has its ground
in another such connection. Since this second connection is distinguished
only according to form, it has the same content as the first; it still has the
same two determinations of content but is now their immediate linking
together. This linking, however, is of a general nature, and the content,
therefore, is diversified into determinations that are indifferent to each
other. The linking is not, therefore, their true absolute connection that
would make one determination the element of self-identity in the posited-
ness, and the other determination the positedness of this same self-identity;
on the contrary, the two are supported by a something and this something
is what connects them, but in a connection which is not reflected, is rather
only immediate and, therefore, only a relative ground as against the linking
in the other something. The two somethings are therefore the two distinct
connections of content that have transpired. They stand in the identi-
cal ground-connection of form; they are one and the same whole content,
namely the two content determinations and their connection; they are dis-
tinct only by the kind of this connection, which in the one is an immediate
and in the other a posited connection; through this, they are distinguished
one from another as ground and grounded only according to form. – Second,
this ground-connection is not only formal, but also real. Formal ground
passes over into real ground, as has been shown; the moments of the form
reflect themselves into themselves; they are a self-subsistent content, and
the ground-connection contains also one content with the character of
ground and another with that of grounded. The content constitutes at first
the immediate identity of both sides of the formal ground; so the two sides
have one and the same content. But the content also has the form in it, and
so it is a twofold content that behaves as ground and grounded. One of the
two content determinations of the two somethings is therefore determined,
not merely as being common to them according to external comparison,
but as their identical substrate and the foundation of their connection.
As against the other determination of the content, this determination is
essential and is the ground of the other which is posited, that is, posited
in the something, the connection of which is the grounded. In the first
something, which is the ground-connection, this second determination of
the content is also immediately and in itself linked with the first. But the
other something only contains the one determination in itself as that in
which it is immediately identical with the first something, but the other
as the one which is posited in it. The former content determination is its
ground by virtue of its being originally linked in the first something with
the other content determination.
The ground-connection of the content determinations in the second
something is thus mediated through the connection present in the first
something. The inference is this: since determination B is implicitly linked
with determination A in a something, in a second something to which only
the one determination A immediately belongs, also B is linked with it. In
the second something, not only is this second determination mediated;
also mediated is that its immediate ground is mediated, namely by virtue
of its original connection with B in the first something. This connection
is thus the ground of the ground A, and the whole ground-connection is
present in the second something as posited or grounded.
3. Real ground shows itself to be the self-external reflection of ground;
its complete mediation is the restoration of its identity with itself. But
because this identity has in the process equally acquired the externality of
real ground, the formal ground-connection in this unity of itself and real
ground is just as much self-positing as self-sublating ground; the ground-
connection mediates itself with itself through its negation. The ground is at
first, as the original connection, the connection of immediate content deter-
minations. The ground-connection, being essential form, has for sides such
that are sublated or are as moments. Consequently, as the form of imme-
diate determinations, it connects itself with itself as self-identical while at
the same time connecting with their negation; accordingly, it is ground not
in and for itself but as connected with the sublated ground-connection. –
Second, the sublated connection or the immediate, which in the original
and in the posited connection is the identical substrate, is likewise real
ground not in and for itself; that it is ground is rather posited by virtue of
that original link. –
Thus the ground-connection is in its totality essentially presupposing
reflection; formal ground presupposes the immediate content determi-
nation, and this content presupposes form as real ground. Ground is
therefore form as an immediate linkage but in such a manner that it
repels itself from itself and rather presupposes immediacy, referring itself
therein as to another. This immediate is the content determination, the
simple ground; but as such, that is, as ground, it is equally repelled
from itself and refers itself to itself equally as to an other. – Thus the
total ground-connection has taken on the determination of conditioning
mediation.

C. CONDITION

a. The relatively unconditioned

1. Ground is the immediate, and the grounded the mediated. But ground
is positing reflection; as such, it makes itself into positedness and is presup-
posing reflection; as such it refers itself to itself as to something sublated,
to an immediate through which it is itself mediated. This mediation, as an
advance from the immediate to the ground, is not an external reflection
but, as we have seen, the ground’s own doing or, what is the same, the
ground-connection, as reflection into its self-identity, is just as essentially
self-externalizing reflection. The immediate to which ground refers as to its
essential presupposition is condition; real ground is accordingly essentially
conditioned. The determinateness that it contains is the otherness of itself.
Condition is therefore, first, an immediate, manifold existence. Second,
it is this existence referred to an other, to something which is ground, not
of this existence but in some other respect, for existence itself is immediate
and without ground. According to this reference, it is something posited;
as condition, the immediate existence is supposed to be not for itself
but for another. But this, that it thus is for another, is at the same time
itself only a positedness; that it is posited is sublated in its immediacy: an
existence is indifferent to being a condition. Third, condition is something
immediate in the sense that it constitutes the presupposition of ground. In
this determination, it is the form-connection of ground withdrawn into
self-identity, hence the content of ground. But content is as such only the
indifferent unity of ground, as in the form: without form, no content.
It nevertheless frees itself from this indifferent unity in that the ground-
connection, in the complete ground, becomes a connection external to
its identity, whereby content acquires immediacy. In so far, therefore, as
condition is that in which the ground-connection has its identity with itself,
it constitutes the content of ground; but since this content is indifferent
to form, it is only implicitly the content of form, is something which has
yet to become content and hence constitutes the material for the ground.
Posited as condition, and in accordance with the second moment, existence
is determined to lose its indifferent immediacy and to become the moment
of another. By virtue of its immediacy, it is indifferent to this connection;
inasmuch as it enters into it, however, it constitutes the in-itself of the
ground and is for it the unconditioned. In order to be condition, it has its
presupposition in the ground and is itself conditioned; but this condition
is external to it.
2. Something is not through its condition; its condition is not its ground.
Condition is for the ground the moment of unconditioned immediacy, but
is not itself the movement and the positing that refers itself to itself neg-
atively and that makes itself into a positedness. Over against condition
there stands, therefore, the ground-connection. Something has, besides its
condition, also a ground. – This ground is the empty movement of reflec-
tion, for the latter has the immediacy which is its presupposition outside
it. But it is the whole form and the self-subsistent process of mediation,
for the condition is not its ground. Since this mediating refers itself to
itself as positing, it equally is according to this side something immediate
and unconditioned; it does indeed presuppose itself, but as an externalized
or sublated positing; whatever it is in accordance with its determination,
that it is, on the contrary, in and for itself. – Inasmuch as the ground-
connection is thus a self-subsisting self-reference and has within it the
identity of reflection, it has a content which is peculiarly its own as against
the content of the condition. The one content is that of the ground and is
therefore essentially informed; the other content, that of the condition, is
on the contrary only an immediate material whose connecting reference to
the ground, while at the same time constituting the in-itself of the latter, is
also equally external to it; it is thus a mingling of a self-subsisting content
that has no reference to the content of the ground determination and of
the content that enters into the latter and, as its material, should become
a moment of it.
3. The two sides of the whole, condition and ground, are thus, on the one
hand, indifferent and unconditioned with respect to each other: the one as
the non-referred-to side, to which the connecting reference in which it is
the condition is external; the other as the connecting reference, or form,
for which the determinate existence of the condition is only a material,
something passive whose form, such as it possesses on its own account, is
unessential. On the other hand, the two sides are also mediated. Condition
is the in-itself of the ground; so much is it the essential moment of the
ground-connection, that it is the simple self-identity of the ground. But this
also is sublated; this in-itself is only something posited; immediate existence
is indifferent to being a condition. The fact, therefore, that condition is the
in-itself of the ground constitutes the side of it by which it is a mediated
condition. Likewise, the ground-connection has in its self-subsistence also
a presupposition; it has its in-itself outside itself. – Consequently, each
of the two sides is this contradiction, that they are indifferent immediacy
and essential mediation, both in one reference – or the contradiction of
independent subsistence and of being determined as only moments.

b. The absolutely unconditioned

At first, each of the two relatively unconditioned sides reflectively shines in
the other; condition, as an immediate, is reflected in the form connection
of the ground, and this form in the immediate existence as its positedness;
but each, apart from this reflective shine of its other in it, stands out on its
own and has a content of its own.
Condition is at first immediate existence; its form has these two moments:
that of positedness, according to which it is, as condition, material and
moment of the ground; and that of the in-itself, according to which it
constitutes the essentiality of ground or its simple reflection into itself.
Both sides of the form are external to immediate existence, for the latter
is the sublated ground-connection. – But, first, existence is in it only this:
to sublate itself in its immediacy and to founder, going to the ground.
Being is as such only the becoming of essence; it is its essential nature to
make itself into a positedness and into an identity which is an immediacy
through the negation of itself. The form determinations of positedness and
of self-identical in-itself, the form through which immediate existence is
condition, are not, therefore, external to that existence; the latter is, rather,
this very reflection. Second, as condition, being is now posited as that which
it essentially is, namely as a moment and consequently as the being of an
other, and at the same time as the in-itself of an other; it is in itself but only
through the negation of itself, namely through the ground and through
its self-sublating and consequent presupposing reflection; the in-itself of
being is thus only something posited. This in-itself of the condition has
two sides: one side is its essentiality as essentiality of the ground, while the
other is the immediacy of its existence. Or rather, both sides are the same
thing. Existence is an immediate, but immediacy is essentially something
mediated, namely through the self-sublating ground. Existence, as this
immediacy mediated by a self-sublating mediating, is at the same time
the in-itself of the ground and its unconditioned side; but again, this in-
itself is at the same time itself equally only moment or positedness, since
it is mediated. – Condition is, therefore, the whole form of the ground-
connection; it is the presupposed in-itself of the latter, but, consequently,
is itself a positedness and its immediacy is this, to make itself into a
positedness and thereby to repel itself from itself, in such as way that it
both founders to the ground and is ground, the ground that makes itself
into a positedness and thereby into a grounded, and both are one and the
same.
Likewise in the conditioned ground, the in-itself is not just as the
reflective shining of an other in it. This ground is the self-subsistent,
that is, self-referring reflection of the positing, and consequently the self-
identical; or it is in it its in-itself and its content. But it is at the same time
presupposing reflection; it negatively refers to itself and posits its in-itself
as an other opposite to it, and condition, according to both its moment
of in-itself and of immediate existence, is the ground-connection’s own
moment; the immediate existence essentially is only through its ground
and is a moment of itself as a presupposing. This ground, therefore, is
equally the whole itself.
What we have here, therefore, is only one whole of form, but equally so
only one whole of content. For the proper content of condition is essential
content only in so far as it is the self-identity of reflection in the form, or the
ground-connection is in it this immediate existence. Further, this existence
is condition only through the presupposing reflection of the ground; it is
the ground’s self-identity, or its content, to which the ground posits itself
as opposite. Therefore, the existence is not a merely formless material for
the ground-connection; on the contrary, because it has this form in it, it is
informed matter, and because in its identity with it it is at the same time
indifferent to it, it is content. Finally, it is the same content as that possessed
by the ground, for it is precisely content as that which is self-identical in
the form connection.
The two sides of the whole, condition and ground, are therefore one
essential unity, as content as well as form. They pass into one another, or,
since they are reflections, they posit themselves as sublated, refer themselves
to this their negation, and reciprocally presuppose each other. But this is at
the same time only one reflection of the two, and their presupposing
is, therefore, one presupposing only; the reciprocity of this presupposing
ultimately amounts to this, that they both presuppose one identity for their
subsistence and their substrate. This substrate, the one content and unity
of form of both, is the truly unconditioned; the fact in itself. – Condition
is, as it was shown above, 20 only the relatively unconditioned. It is usual,
therefore, to consider it as itself something conditioned and to ask for a new
condition, whereby the customary progression ad infinitum from condition
to condition is set in motion. But now, why is it that at one condition
a new condition is asked for, that is, why is that condition assumed to
be something conditioned? Because it is some finite determinate existence
or other. But this is a further determination of condition that does not
enter into its concept. Condition is as such conditioned solely because
it is the posited in-itselfness; it is, therefore, sublated in the absolutely
unconditioned.
Now this contains within itself the two sides, condition and ground,
as its moments; it is the unity to which they have returned. Together, the
two constitute its form or its positedness. The unconditioned fact is the
condition of both, but the condition which is absolute, that is to say, one
which is itself ground. – As ground, the fact is now the negative identity
that has repelled itself into those two moments: first, in the shape of the
sublated ground-connection, the shape of an immediate manifold void of
unity and external to itself, one that refers to the ground as an other to
it and at the same time constitutes its in-itself; second, in the shape of an
inner, simple form which is ground, but which refers to the self-identical
immediate as to an other, determining it as condition, that is, determining
the in-itself of it as its own moment. – These two sides pre-suppose the
totality, presuppose that it is that which posits them. Contrariwise, because
they presuppose the totality, the latter seems to be in turn also conditioned
by them, and the fact to spring forth from its condition and its ground. But
since these two sides have shown themselves to be an identity, the relation
of condition and ground has disappeared; the two are reduced to a mere
reflective shine; the absolutely unconditioned is in its movement of positing
and presupposing only the movement in which this shine sublates itself. It
is the fact’s own doing that it conditions itself and places itself as ground
over against its conditions; but in connecting conditions and ground, the
fact is a reflection shining in itself; its relation to them is a rejoining itself.

c. Procession of the fact into concrete existence

The absolutely unconditioned is the absolute ground that is identical with
its condition, the immediate fact as the truly essential. As ground, it refers
negatively to itself and makes itself into a positedness; but this positedness
is a reflection that is complete in both its sides and is in them the self-
identical form of connection, as has transpired from its concept. This
positedness is therefore first the sublated ground, the fact as an immediacy
void of reflection, the side of the conditions. This is the totality of the
determinations of the fact, the fact itself, but the fact as thrown into the
externality of being, the restored circle of being. In condition, essence lets
go of the unity of its immanent reflection; but it lets it go as an immediacy
that now carries the character of being a conditioning presupposition and of
essentially constituting only one of its sides. – For this reason the conditions
are the whole content of the fact, because they are the unconditioned in
the form of formless being. But because of this form, they also have yet
another shape besides the conditions of the content as this is in the fact
as such. They appear as a manifold without unity, mingled with extra-
essential elements and other circumstances that do not belong to the circle
of existence as constituting the conditions of this determinate fact. – For
the absolute, unrestricted fact, the sphere of being itself is the condition.
The ground, returning into itself, posits that sphere as the first immediacy
to which it refers as to its unconditioned. This immediacy, as sublated
reflection, is reflection in the element of being, which thus forms itself as
such into a whole; form proliferates as determinateness of being and thus
appears as a manifold distinct from the determination of reflection and as
a content indifferent to it. The unessential, which is in the sphere of being
but which the latter sheds in so far as it is condition, is the determinateness
of the immediacy into which the unity of form has sunk. This unity of
form, as the connection of being, is in the latter at first as becoming – the
passing over of a determinateness of being into another. But the becoming
of being is also the coming to be of essence and a return to the ground.
The existence that constitutes the conditions, therefore, is in truth not
determined as condition by an other and is not used by it as material; on
the contrary, it itself makes itself, through itself, into the moment of an
other. – Further, the becoming of this existence does not start off from itself
as if it were truly the first and immediate; on the contrary, its immediacy
is something only presupposed, and the movement of its becoming is the
doing of reflection itself. The truth of existence is thus that it is condition;
its immediacy is solely by virtue of the reflection of the ground-connection
that posits itself as sublated. Consequently, like immediacy, becoming is
only the reflective shine of the unconditioned inasmuch as this presupposes
itself and has its form in this presupposing, and hence the immediacy of
being is essentially only a moment of the form.
The other side of this reflective shining of the unconditioned is the
ground-connection as such, determined as form as against the immediacy
of the conditions and the content. But this side is the form of the absolute
fact that possesses the unity of its form with itself or its content within
it, and, in determining this content as condition, in this very positing
sublates the diversity of the content and reduces it to a moment; just
as, contrariwise, as a form void of essence, in this self-identity it gives
itself the immediacy of subsistence. The reflection of the ground sublates
the immediacy of the conditions, connecting them and making them
moments within the unity of the fact; but the conditions are that which the
unconditioned fact itself presupposes and the latter, therefore, sublates its
own positing; consequently, its positing converts itself just as immediately
into a becoming. – The two, therefore, are one unity; the internal movement
of the conditions is a becoming, the return into the ground and the
positing of the ground; but the ground as posited, and this means as
sublated, is the immediate. The ground refers negatively to itself, makes
itself into a positedness and grounds the conditions; in this, however, in that
the immediate existence is thus determined as a positedness, the ground
sublates it and only then makes itself into a ground. – This reflection is
therefore the self-mediation of the unconditioned fact through its negation.
Or rather, the reflection of the unconditioned is at first a presupposing,
but this sublating of itself is immediately a positing which determines;
secondly, in this positing the reflection is immediately the sublating of
the presupposed and a determining from within itself; this determining is
thus in turn the sublating of the positing: it is a becoming within itself.
In this, the mediation as a turning back to itself through negation has
disappeared; mediation is simple reflection reflectively shining within itself
and groundless, absolute becoming. The fact’s movement of being posited,
on the one hand through its conditions, and on the other hand through
its ground, now is the disappearing of the reflective shine of mediation. The
process by which the fact is posited is accordingly a coming forth, the simple
self-staging of the fact in concrete existence, the pure movement of the fact
to itself.
When all the conditions of a fact are at hand, the fact steps into concrete
existence. The fact is, before it exists concretely; it is, first, as essence or
as unconditioned; second, it has immediate existence or is determined,
and this in the twofold manner just considered, on the one hand in its
conditions and on the other in its ground. In the former case, it has given
itself the form of the external, groundless being, for as absolute reflection
the fact is negative self-reference and makes itself into its presupposition.
This presupposed unconditioned is, therefore, the groundless immediate
whose being is just to be there, without grounds. If, therefore, all the
conditions of the fact are at hand, that is, if the totality of the fact is
posited as a groundless immediate, then this scattered manifold internally
recollects itself. – The whole fact must be there, within its conditions, or
all the conditions belong to its concrete existence; for the all of them
constitutes the reflection of the fact. Or again, immediate existence, since
it is condition, is determined by form; its determinations are therefore
determinations of reflection and with the positing of one the rest also
are essentially posited. – The recollecting of the conditions is at first the
foundering to the ground of immediate existence and the coming to be
of the ground. But the ground is thereby a posited ground, that is, to
the extent that it is ground, to that extent it is sublated as ground and is
immediate being. If, therefore, all the conditions of the fact are at hand, they
sublate themselves as immediate existence and as presupposition, and the
ground is equally sublated. The latter proves to be only a reflective shine
that immediately disappears; this coming forth is thus the tautological
movement of the fact to itself: its mediation through the conditions and
through the ground is the disappearing of both of these. The coming forth
into concrete existence is therefore so immediate, that it is mediated only
by the disappearing of the mediation.

The fact proceeds from the ground.
It is not grounded or posited by it in
such a manner that the ground would still stay underneath, as a substrate;
on the contrary,
the positing is the outward movement of ground to itself
and the simple disappearing of it.
Through its union with the conditions, it
obtains the external immediacy and the moment of being.
But it does not
obtain them as a something external,
nor by referring to them externally;
rather,
as ground it makes itself into a positedness;
its simple essentiality
rejoins itself in the positedness
and, in this sublating of itself,
it is the
disappearing of its difference from its positedness,
and is thus simple essential immediacy.
It does not, therefore, linger on as something distinct
from the grounded;
on the contrary, the truth of the grounding is that in
grounding the ground unites with itself,
and its reflection into another is
consequently its reflection into itself.
The fact is thus the unconditioned
and, as such, equally so the groundless;
it arises from the ground only in so
far as the latter has foundered and is no longer ground:
it rises up from the groundless, that is,
from its own essential negativity or pure form.
This immediacy, mediated by ground and condition
and self-identical through the sublating of mediation,
is concrete existence.
