SECTION III

Measure

Abstractly expressed, quality and quantity are in measure united.
Being as such is the immediate equality of determinateness with itself.
This immediacy of determinateness has sublated itself.
Quantity is being that has returned to itself in such a way that it is
a simple self-equality indifferent to determinateness.
But this indifference is only the externality of having
the determinateness not in itself but in an other.
As third, we now have the externality that refers itself to itself;
as self-reference, it is at the same time sublated externality and
carries the difference from itself in it;
a difference which, as externality, is the quantitative moment, and,
as taken back into itself, the qualitative.

Since among the categories of transcendental idealism
modality comes after quantity and quality,
with relation inserted in between,
this is an appropriate place to say something about it.
In transcendental idealism, this category has the meaning that
it is the connection of the subject matter to thought.
As understood by that idealism, thought is as such
essentially external to the thing-in-itself.
Hence, inasmuch as the other categories have
the transcendental determination of belonging only to consciousness,
but as its objective moment,
so modality, which is the category of the connection to the subject,
possesses the determination of reflection in itself in a relative sense,
that is to say, the objectivity which is granted to the other categories
is lacking in those of modality;
these, according to Kant's words, do not add in the least to
the concept as a determination of the object but only express
its relation to the faculty of cognition
(Cr. of Pure R., 2nd edn, pp. 99, 266).
The categories which Kant groups under modality
(possibility, actuality, and necessity)
will come up later in their proper place.
Kant did not apply the form of triplicity
(an infinitely important form even though
with him it occurred only as a formal spark of light)
to the genera of his categories (to quantity, quality, etc.),
but only to their species to which he also gave the name of genera.
He was therefore unable to hit upon the third to quality and quantity.

With Spinoza, the mode is likewise the third after substance and attribute;
Spinoza defines it as the affections of substance, or as that which
is in another through which it is also comprehended.
In this way of conceiving it, this third is externality as such;
as has already been mentioned, with Spinoza generally,
the rigidity of substance lacks the turning back into itself.

The remark just made extends to any of the systems of pantheism which
thought has in one way or other produced.
Being, the one, substance, the infinite, essence, is the first;
opposite this abstraction is the second which
can be mustered in an equally abstract form, as is habitually done as the
next step in any purely formalistic thinking,
namely all determinateness generally taken as the mere finite,
the mere accidental, the transitory, the extraneous and unessential, etc.
But the bond connecting this second with the first is too invasive
for the second not to be not equally grasped with the first;
thus with Spinoza the attribute is the whole of substance,
though as comprehended by the understanding,
which is itself a restriction of substance or mode;
and so the mode, the insubstantial as such which can
be grasped only through an other,
constitutes the opposite extreme of substance, the third.
Also Indian pantheism, taken abstractly, has attained
in its monstrous fantasies this refinement which runs like a moderating
thread across its excesses as its one point of interest
namely that Brahma, the one of abstract thought,
progresses through the shape of Vishnu,
particularly in the form of Krishna, to the third, Shiva.
The determination of this third is that of mode, alteration,
coming-to-be and passing-away;
it is the field of externality in general.
This Indian trinity has tempted a comparison with the Christian, and
one must indeed acknowledge a common element in them.
But it is essential to be aware of the difference that separates them.
It is not just that this difference is infinite but that the
true infinity makes the difference.
The determination of the Indian third principle is that
it is the dispersal of the substantial unity into its opposite,
not its turning back to itself – a spiritual void rather, not spirit.
In the true trinity, there is not only unity but unification;
the syllogism is brought to a unity which is full of content and actual,
a unity which in its totally concrete determination is spirit.
The principle of the mode and of alteration
does not, of course, exclude unity altogether.
In Spinozism, for instance, precisely the mode is as such untrue
while substance alone is what truly is;
everything is supposed to be reduced to substance, and
this is then a sinking of all content into an only formal unity void of content.
As for Shiva, it too is again the great whole,
not distinct from Brahma, from Brahma itself, that is,
the distinction and the determinateness just disappear
without being preserved, without being sublated, and
the unity does not become concrete unity, nor is the diremption reconciled.
The supreme goal of the human being, relegated as he is to
the sphere of coming-to-be and passing-away, of modality in general,
is to sink into unconsciousness, into unity with Brahma, annihilation;
the Buddhist Nirvana, Nibbana, etc., is the same.

Now although the mode is as such abstract externality, indifference
to qualitative as well as quantitative determinations, and nothing in the
essence should depend on the external, the unessential, it is nevertheless
conceded that in the many all depends on the how;
but this is to concede that the mode itself essentially belongs to
the substance of a thing, a very indefinite connection but
one which at least implies that the externality of
the mode is not all that abstract an externality after all.
Here the mode has the definite meaning of being measure.
The Spinozistic mode, just like the Indian principle of alteration,
is the measureless.
The Greeks were aware that everything has a measure.
Parmenides himself introduces necessity after abstract being,
as the ancient limit which is imposed on all.
And this awareness, although still vague, is the beginning of a much
higher concept than is contained in substance and
in the distinction of the mode from it.

Measure in its more developed, more reflected form is necessity.
Fate, Nemesis, ultimately comes down to a determination of measure.
Whatever renders itself beyond the pale, becomes too great, too high,
is brought down to the other extreme of being reduced to nothing,
so that the mean of measure, the medium, is restored.
That the Absolute, God, is the measure of all things, is not
a stronger statement of pantheism than the definition,
“the Absolute, God, is being,” but is infinitely truer.
Measure is indeed an external way of things, a more or less,
but one which is as at the same time reflected into itself,
not merely an indifferent and external determinateness
but one which exists in itself; thus it is the concrete truth of being.
For this reason the nations have revered in it the presence of
something inviolable and sacred.

Already present in measure is the idea of essence,
namely of being self-identical in the immediacy of being determined,
so that this immediacy is reduced through the self-identity to
something mediated, just as the self-identity is equally mediated
only through this externality, but the mediation is one with itself:
this is reflection, the moments of which indeed are, but
in this being are absolutely nothing but moments of their negative unity.
In measure, the qualitative element is quantitative;
the determinateness or the difference is indifferent and
therefore a difference which is none;
it is sublated and this quantitativeness,
as an immanent turning back in which it is qualitative,
constitutes the being-in-and-for-itself which is essence.
But measure is essence only implicitly in itself or in its concept;
this concept of measure is not yet posited.
Measure is as such still the existent unity of the
qualitative and the quantitative element;
its moments are an existence, a quality and some quanta of this quality
which, in themselves, are indeed only indivisible,
but do not yet have the meaning of this reflected determination.
In the development of measure, these moments are differentiated
but at the same time referred to each other,
so that the identity which they are in themselves becomes
their connection explicitly, that is, is posited.
The meaning of this development is the realization of measure in which the
latter posits itself in relation to itself and consequently as moment as well;
through this mediation, measure is determined as sublated; its immediacy
as well as that of its moments disappears; these moments are reflected and
thus measure, having disclosed what it is according to its concept, has
passed over into essence.

Measure is at first the immediate unity of the qualitative and the quan-
titative element, so that it is,

first, a quantum that has qualitative meaning and is as measure. As
so implicitly determined in itself, its further determination is that the
difference of its moments, of its qualitatively and quantitatively determined
being, is disclosed in it. These moments further determine themselves into
wholes of measure which as such are self-subsistent, and, inasmuch as they
refer to each other essentially, measure becomes,

second, a ratio of specific quanta, each an independent measure. But
their self-subsistence also rests essentially on a quantitative relation and a
difference of magnitude, and so the self-subsistence becomes a transition
of one measure into another. The result is that measure collapses into the
measureless. – But this beyond of measure is the negativity of measure only
in itself; thus,

third, the indifference of the determinations of measure is thereby posited,
and measure (real measure because of the negativity contained within it)
is posited as an inverse ratio of measures which, as self-subsistent qualities,
essentially rest on only their quantity and their negative reference to each
other, and consequently prove to be only moments of their truly self-
subsistent unity. This unity is the reflection-into-itself of each and the
positing of them; it is essence.

The development of measure, which we have attempted in what follows,
is among the most difficult of subject matters.
Starting with immediate,
external measure, it would have to proceed, on the one hand, to the further
abstract determination of the quantitative aspect of natural things (of a
mathematics of nature); on the other side, it would have to indicate the
link between this determination of measure and the qualitative aspect of
those things – at least in general, for the detailed demonstration of the
link between the qualitative and the quantitative aspects as they originate
in the concept of a concrete object belongs to the particular science of
the concrete (examples of which, concerning the law of falling bodies and
the free movement of the heavens, will be found in the Encyclopedia of the
Philosophical Sciences).
We may remark quite in general in this connection
that the different forms in which measure is realized also belong to different
spheres of natural reality.
The complete, abstract indifference of developed
measure, that is, of its laws, can only be found in the sphere of mechanism
where concrete corporeity is only abstract matter itself; the qualitative
differences of this matter are of an essentially quantitative nature; space
and time are nothing but pure externalities, and the aggregates of matters,
the masses, the intensity of weight, are determinations which are just as
external and have their proper determinateness in the quantitative element.
On the other hand, in physical things but even more so in the organic, this
quantitative determinateness of abstract materiality is already disturbed by
the multiplicity and consequently the conflict of qualities. And the thus
ensuing conflict is not just one of qualities as such, but measure itself
is subordinated here to higher relations and its immanent development is
reduced rather to the simple form of immediate measure. The limbs of
the animal organism have a measure which, as a simple quantum, stands
in a ratio to the other quanta of the other limbs; the proportions of the
human body are the fixed ratios of such quanta, and the science of nature
still has far to go in discovering anything about the link that connects
these magnitudes with the organic functions on which they are entirely
dependent. But the closest example of the reduction of an immanent
measure to a merely externally determined magnitude is motion. In the
heavenly bodies, motion is free motion, one which is only determined
by the concept from which alone, consequently, its magnitudes equally
depend (see above); but in the organic body this free motion is reduced
to one which is arbitrary or mechanically regular, that is, to one which is
totally abstract and formal.

And in the realm of spirit there is even less of a characteristic, free
development of measure to be found.
For instance, it is obvious that a
republican constitution like the Athenian, or an aristocratic constitution
mixed with democracy, is possible only in a state of a certain size;
it is also obvious that in civil society the multitudes of individuals
who belong to the different occupations stand in a certain ratio to each other.
But none of this yields either laws of measures or typical forms of it.
In the spiritual realm as such there are indeed distinctions of intensity of
character, strength of imagination, sensations, representations, and so on;
but in determining them one cannot go past this indefinite duo of
“strength” and “weakness.”
To see how lame and totally empty ultimately turn out to be
the so-called laws which have been established for
the relation of strength and weakness in sensations, representations,
and so on, one need only look at the psychologies
that busy themselves with just such matters.

CHAPTER 1

Specific quantity

Qualitative quantity is,

first, an immediate, specific quantum; and this quantum,

second, in relating itself to another,
becomes a quantitative specifying,
a sublating of the indifferent quantum.
This measure is to this extent a rule
and contains the two moments of measure as different;
namely, the quantitative determinateness
and the external quantum as existing in themselves.
In this difference, however, the two sides become qualities,
and the rule becomes a relation of the two; measure presents itself thereby,

third, as a relation of qualities that have one single measure at first;
a measure, however, which further specifies itself in itself into
a difference of measures.

A. THE SPECIFIC QUANTUM

1. Measure is the simple self-reference of quantum, its own determinateness
in itself; quantum is thus qualitative. At first, as an immediate measure
it is an immediate quantum and hence some specific quantum; equally
immediate is the quality that belongs to it; it is some specific quality or
other. – Thus quantum, as this no longer indifferent limit but as self-
referring externality, is itself quality and, although distinguished from it,
it does not extend past it, just as quality does not extend past quantum.
Quantum is thus the determinateness that has returned into simple self-
equality – which is at one with determinate existence just as determinate
existence is at one with it.

If a proposition is to be made of the determination just obtained, it
could be expressed thus: “Whatever is, has a measure.” Every existence has
a magnitude, and this magnitude belongs to the very nature of a something;
it constitutes its determinate nature and its in-itself. The something is not
indifferent to this magnitude, as if, were the latter to alter, it would remain
the same; rather the alteration of the magnitude alters its quality. As measure
the quantum has ceased to be a limit which is none; it is from now on the
determination of a thing, so that, were the latter to exceed or fall short of
this quantum, it would perish. –

A measure, in the usual sense of a standard, is a quantum which is
arbitrarily assumed as the unit determinate in itself as against an external
amount. Of course, such a unit can in fact also be determinate in itself,
like a foot or some such other original measure; to the extent, however,
that it is used as the measuring standard for other things, it is with respect
to them only an external measure, not their original measure. – Thus the
diameter of the earth or the length of a pendulum may be taken as a
specific quantum on their own account. But the choice of a fraction of the
earth's diameter or of the pendulum's length, and this last under which
degree of latitude, for use as a standard of measure is arbitrary. And for
other things such a standard is something even more external. These have
further specified the universal specific quantum in some particular way, and
they have thereby been made into particular things. It is therefore foolish
to speak of a natural standard of things. Anyway, a universal standard
is meant for use only as an external comparison of things, and in this
superficial sense of universal standard it is quite a matter of indifference
what is used for the purpose. It is not meant to be a fundamental measure
in the sense that in it the natural measures of particular things would be
displayed and recognized, according to a rule, as the specifications of one
universal measure, the measure of the things's universal body. But without
this meaning the sole interest and significance of an absolute standard of
measure is that of something common, and any such standard is a universal
not in itself but only by convention.

Immediate measure is a simple determination of magnitude as, for exam-
ple, the size of organic beings, of their limbs, and so forth. But any concrete
existent has the size required for being what it is, and for having existence
in the first place. – As a quantum, the existent is an indifferent magnitude
open to external determination and capable of fluctuating increases and
decreases. As a measure, however, it is at the same time distinct from itself
as quantum, itself as such an indifferent determination, and is a restriction
on that indifferent fluctuation of a limit.
Since in the existence of anything the quantitative determinateness is
thus twofold, in the sense that quality is tied to it and yet the quantity
can fluctuate without prejudice to quality – so the demise of anything that
has a measure occurs through the alteration of its quantum. On the one
hand, the demise appears unexpected, inasmuch as there can be alteration
in the quantum without the measure and the quality being altered; but, on
the other hand, it is made into something quite simple to grasp by means
namely of the concept of gradualness. It is easy to turn to this category for
visualizing or “explaining” the disappearance of a quality or of a something,
for it gives the impression that one can witness this disappearance as if
before one's eyes: since the quantum is posited as the external limit which
is by nature alterable, the alteration (of quantum only) then follows by
itself. But in fact nothing is thereby explained, for the alteration is at the
same time essentially the transition of one quality into another, or the more
abstract transition of one existence into a non-existence, and therein lies
another determination than just gradualness, which is only a decrease or
increase, and the one-sided holding fast to magnitude.
2. The ancients had already taken notice of this coincidence, that an
alteration which appears to be only quantitative suddenly changes into a
qualitative one, and they used popular examples to illustrate the inconsis-
tencies that arise when such a coincidence is not understood. Two such
examples go under the familiar names of “the bald” and “the heap.” They
are elenchi, that is, according to Aristotle's explanation, two ways in which
one is compelled to say the opposite of what one has previously asserted. 12
The question was put: does the plucking of one hair from someone's head
or from a horse's tail produce baldness, or does a heap cease to be a heap if
one grain is removed? The expected answer can safely be conceded, for the
removal amounts to a merely quantitative difference, and an insignificant
one at that. And so one hair is removed, one grain, and this is repeated
with only one hair and one grain being removed each time the answer is
conceded. At last the qualitative alteration is revealed: the head or the tail
is bald; the heap has vanished. In conceding the answer, it was not only
the repetition that was each time forgotten, but also that the individually
insignificant quantities (like the individually insignificant disbursements
from a patrimony) add up, and the sum constitutes the qualitative whole,
so that at the end this whole has vanished: the head is bald, the purse is
empty.

The embarrassment, the contradiction, produced by the result, is not
anything sophistic in the usual sense of the word, as if the contradiction
were a pretense. The mistake is committed by the assumed interlocutor
(that is, our ordinary consciousness), and that is of assuming a quantity to
be only an indifferent limit, that is, of taking it in the narrowly defined
sense of a quantity. But this assumption is confounded by the truth to
which it is brought, namely that quantity is a moment of measure and
is linked to quality; refuted is the one-sided stubborn adherence to the
abstract determinateness of quantum. – Also those elenchi are, therefore,
not anything frivolous or pedantic but basically correct: they attest to a
mind which has an interest in the phenomena that come with thinking.
Quantum, when it is taken as indifferent limit, is the side from which
an existence is unsuspectedly attacked and laid low. It is the cunning of
the concept that it would seize on an existence from this side where its
quality does not seem to come into play – and it does it so well that the
aggrandizement of a State or of a patrimony, etc., which will bring about
the misfortune of the State or the owner, even appears at first to be their
good fortune.

3. Measure is in its immediacy an ordinary quality of a specific mag-
nitude appropriate to it. Now there is also the distinction between the
side according to which quantum is an indifferent limit that can fluctuate
without the quality altering and the other side according to which it is
qualitative and specific. Both sides are the magnitude determinations of
one and the same thing; but because of the original immediacy of measure,
this distinction too is to be taken as immediate, and accordingly the two
sides each also have a diverse concrete existence. The concrete existence of
measure, which is the side of magnitude determinate in itself, then behaves
towards the concrete existence of the alterable external side by sublating
the indifference of the latter; this is a specifying of measure.

B. SPECIFYING MEASURE

First, this measure is a rule, a measure external to the mere quantum.

Second, it is a specific quantity determining the external quantum.

Third, the two sides, both as qualities of specific quantitative determinacy,
relate to one another as one measure.

a. The rule

The rule, or the standard which we have just mentioned, is in the first
instance as a magnitude which is determinate in itself and is a unit with
respect to a quantum of a particular concrete existence: this is a quantum
with a concrete existence which is other than the something of the rule –
is measured by the latter, that is, is determined as the amount of the said
unit. This comparison is an external act, and the unit itself is an arbitrary
magnitude which can in turn be posited as an amount (the foot as an
amount of inches). But measure is not only an external rule; as a specific
measure its intrinsic nature is that it relates to its other which is a quantum.

b. Specifying measure

Measure is a specific determining of the external magnitude, that is, of the
indifferent magnitude which is now posited in the measuring something by
some other concrete existence in general.
The something of the measure is indeed itself a quantum, but
with the difference that it is the qualitative side
determining the merely indifferent, external quantum.
It has intrinsically this side of being-for-other
to which the fluctuation in size belongs.
The immanent measuring is a quality of the something, and this something
is confronted by the same quality in another something; in the latter,
however, the quality is at first relative, with a measureless quantum in
general as against the something determined as measuring. 14
Inasmuch as a something has an internal measure, an alteration of the
magnitude of its quality comes to it from outside, and the something
does not take on the arithmetical aggregate of the alteration. Its measure
reacts against it, behaves towards the aggregate as an intensive measure and
assimilates it in a way typically its own; it alters the externally imposed
alteration, makes something else out of this quantum and demonstrates
through this specifying function that in this externality it is for-itself. – This
specifically assimilated aggregate is itself a quantum which is also dependent
on the other, that is, on the other aggregate which is only external to it.
The specified aggregate is therefore also alterable, but is not for that reason
a quantum as such but the external quantum as specified in a constant
manner. Measure thus has its determinate existence as a ratio, and its
specificity is in general the exponent of this ratio.

In intensive and extensive quantum, as we saw when considering these
determinations, 15 it is the same quantum which is present, once in the
form of intensity and again in the form of extension. In this difference the
underlying quantum does not suffer any alteration; the difference is only
an external form. In the specifying measure, on the contrary, the quantum
is taken in one instance in its immediate magnitude, but through the
exponent of the ratio it is taken in a second instance in another amount.

The exponent which constitutes the element of specificity can appear
at first to be a fixed quantum, as a quotient of the ratio between the
external and the qualitatively determined quantum. But it would then be
nothing more than an external quantum whereas by the exponent we are to
understand here nothing but the qualitative moment itself that specifies the
quantum as such. The only strictly immanent qualitative determination of
the quantum is (as we saw earlier) 16 that of the exponent, and it must be such
an exponential determination which now constitutes the ratio and which,
as the internally existent determination, comes to confront the quantum as
externally constituted. The principle of this quantum is the numerical one
which constitutes its internal determinateness: the mode of connection of
this numerical one is external, and the alteration which is determined only
through the nature of the immediate quantum as such consists essentially
in the addition of one such numerical one and then another and so forth.
As the external quantum alters in arithmetical progression in this way, the
specifying reaction of the qualitative nature produces another series which
refers to the first, increases and decreases with it, not however in a ratio
determined by a numerical exponent but in a ratio which is numerically
incommensurable, in the manner of a power determination.

c. Relation of the two sides as qualities

1. The qualitative side of the quantum, in itself determined, exists only
as a reference to the external quantitative side; as specifying the latter, it
is a sublating of its externality through which quantum as such is. This
qualitative side thus has a quantum for the presupposition from which it
starts. But this quantum is also qualitatively distinguished from quality,
and this difference between the two must now be posited in the immediacy
of being in general which still characterizes measure. The two sides thus
stand to each other in a qualitative respect, each a qualitative existence for
itself, and the one side that was at first only an internally indeterminate
formal quantum is the quantum of a something and of its quality, and,
just as their reciprocal reference is now determined as measure in general,
so too is the specific magnitude of these qualities. These qualities stand
in relation to each other according to a determination of measure. This
determination is their exponent, but they are already implicitly connected
to each other in the being-for-itself of measure: the quantum is in its double
being external quantum and specific quantum, so that each of the distinct
quantities has this double determination in it and is at the same time
inextricably interwoven with the other; it is in this way alone that the
qualities are determined. They are not, therefore, a determinate being in
general existing for each other but are rather posited as indivisible, and the
specific magnitude tied to them is a qualitative unity – one determination
of measure in which they are implicitly bound together in accordance with
their concept. Measure is thus the immanent quantitative relating of two
qualities to each other.

2. In measure we have the essential determination of variable magnitude,
for measure is sublated quantum – quantum, therefore, no longer as that
which it ought to be in order to be quantum, but quantum as quantum
and as something other besides. This other is the qualitative element and,
as we have established, is nothing else than the relation of powers of the
quantum. In immediate measure, this variability is not yet posited; it
is just one single quantum or other to which a quality is attached. In the
specifying of measure (the preceding determination), which is an alteration
of the merely external quantum by the qualitative moment, what is posited
is the distinctness of two determinate magnitudes and hence in general the
plurality of measures in one common, external quantum. And it is in this
differentiation of the quantum from itself that the latter shows itself for the
first time to be a real measure, for it appears as a being which is one and the
same (e.g. the constant temperature of the medium) and at the same time of
diversified and indeed quantitative existence (in the various temperatures of
the bodies found in the medium). This differentiation of quantum into the
diverse qualities – the different bodies – yields a further form of measure,
one in which the two sides relate to each other as qualitatively determined
quanta, and this can be called the realized measure.

Magnitude, simply as such, is alterable, for its determinateness is a limit
which at the same time is none; the alteration only affects, therefore, a
particular quantum in place of which another is posited. But the genuine
alteration is that of the quantum as such. Here we have the determina-
tion, interesting when so understood, of the variable magnitude of higher
mathematics in which we neither need to stop short at the formal deter-
mination of alteration or variability in general nor introduce any other
determination except the simple determination of the concept by which
the other of the quantum is only the qualitative. The genuine determination,
therefore, of real variable magnitude is that it is qualitative, that is, as we
have sufficiently shown, that it is determined by a ratio of powers. Posited
in this variable magnitude is that quantum has no value as such but only
as determined in conformity with its other, that is, qualitatively.
The two sides in this relating have, in keeping with their abstract side as
qualitites in general, some particular meaning or other, for instance, space
and time. Taken at first simply as determinacies of magnitude in their
ratio of measure, one of them is the amount which increases and decreases
in external arithmetical progression; the other is an amount specifically
determined by the other amount, which for it is the unit. If each of these
two sides were only a particular quality in general, there would be no
way of distinguishing which of them is to be taken with respect to their
determination of magnitude as merely externally quantitative and which as
varying in quantitative specification. If they are related, for instance, as root
and square, it is indifferent which is regarded as increasing or decreasing
in merely external arithmetical progression, and which on the contrary has
its specific determination in this quantum.

But the two qualities are not indeterminate in their difference, for they
are the moments of measure and the qualification of the latter ought to
rest on them. The closest determinateness of the qualitites themselves
is, of the one, the extensive, that it is an externality within; and of the
other, the intensive, that it exists in itself or is the negative as against the
other. Accordingly, amount is the quantitative moment that pertains to the
former, and unit the one that pertains to the latter; in simple direct ratio,
the former is to be taken as the dividend and the latter as the divisor; in
specifying ratio, the former as the power or the becoming-other and the
latter as the root. Inasmuch as we still count here, that is, we reflect on the
external quantum (which is thus the totally accidental determinateness of
magnitude which we call empirical), and accordingly also equally take the
alteration as advancing in external arithmetical progression, then this falls
on the side of the unit, the intensive quality; the external extensive side, by
contrast, is to be represented as altering in the specified series. But the direct
ratio (like velocity as such, st ) is reduced here to a formal determination
which has no concrete existence but belongs rather only to the abstraction
of reflection; and even though in the ratio of root and square (as in s = at 2 ),
the root is to be taken as an empirical quantum varying in an arithmetical
progression, the other side as specified instead, the higher realization of
the qualification of the quantitative moment, one which would be more
in keeping with the concept, is this: that both sides are related in higher
determinations of powers (as in s 3 = at 2 ).

C. THE BEING-FOR-ITSELF IN MEASURE

1. In the form of specified measure just considered, the quantitative moment
of each side is qualitatively determined (both in the ratio of powers); they
are thus moments of one measure-determinateness of qualitative nature.
Here, however, the qualities are still posited immediately, only as diverse;
they are not related in the manner of their quantitative determinacies, that
is, that outside their relation they would have neither meaning nor existence,
as is the case for the quantitative determinacies as a ratio of powers. The
qualitative moment thus disguises itself as specifying, not itself, but the
determinateness of magnitude. Only within the latter is it posited; for
itself it is instead immediate quality as such which, besides the fact that it
posits the magnitude as non-indifferent, and besides its connection with
its other, still has existence subsisting on its own. Thus space and time,
outside that specification which their quantitative determinateness obtains
in the motion of falling bodies or in the absolutely free motion, both have
the value of space in general and time in general, space subsisting on its
own outside and without the duration of time, and time flowing on its
own independently of space.

This immediacy of the qualitative moment as against its specific measure-
relation is, however, equally bound up with a quantitative immediacy and
with the indifference to this same measure-relation of the quantitative ele-
ment in it; the immediate quality also has only an immediate quantum. For
that reason, the specific measure also has a side of an at first external alter-
ation which advances in merely arithmetical progression undisturbed by it
and in which falls the external and hence merely empirical determination
of magnitude. Quality and quantum, even as they extend outside measure,
are at the same time connected with it; the immediacy is a moment of
their belonging to measure. Thus the immediate qualities also belong to
measure, are likewise connected and stand in a ratio which, outside the
specified one of power determination, is itself only a direct ratio and an
immediate measure. This conclusion and its consequences must now be
indicated.

2. Although the immediately determined quantum is as a moment of
measure otherwise implicitly grounded in a conceptual nexus, in connec-
tion with specific measure it is, as such, externally given. But the immediacy
which is thereby posited is the negation of the determination of qualitative
measure; the same immediacy was shown just now on the sides of the deter-
mination of measure which appeared for that reason to be self-subsisting
qualities. Such a negation and the return to immediate quantitative deter-
minateness are present in the qualitatively determined relation because a
relation of distinct terms entails as such the connection of such terms in
one determinateness, and here, in the quantitative sphere, in distinction
from the determination of the relation, this determinateness is a quantum.
As the negation of the distinct, qualitatively determined sides, this expo-
nent is a being-for-itself, an absolutely determined being, but is so only
in itself; as existence, it is a simple immediate quantum, the quotient or
exponent of a ratio between the sides of measure which is taken as direct; as
such, however, it appears in the quantitative side of measure as an empirical
unit. – In the motion of falling bodies, the spaces traversed are proportional
to the square of the elapsed time, s = at 2 . This is a specifically determinate
ratio, one between space and time; the other, the direct ratio, would per-
tain to space and time as mutually indifferent qualities; it is supposed
to be the ratio of the space traversed to the first moment of time. The
same coefficient, a, remains in all the succeeding time-points – the unit
of the amount, determined for its part by the specifying measure, being
an ordinary quantum. This unit counts at the same time as the exponent
of that direct ratio which pertains to the merely imagined, the bad, that is,
the reflectively formal velocity which is not specifically determined by the
concept. Such a velocity does not have concrete existence here, no more
than does the one previously mentioned which is supposed to pertain to a
falling body at the end of a moment of time. That velocity is ascribed to
the first temporal moment in the fall, but this so-called temporal moment
is itself only an assumed unit which has no existence as such an atomic
point. The beginning of the motion (its alleged smallness would make
no difference) is straight away a magnitude, and one which is specified
by the law of falling bodies. The said empirical quantum is attributed to
the force of gravity, and this force thus has itself no connection with the
specification at hand (the determinateness of powers), that is, with the
determinateness characteristic of measure. The immediate moment, that
in the motion of falling bodies the amount of some fifteen spatial units
which are assumed as feet enter into one unit of time (call it a second, the
so-called first unit), is an immediate measure, just like the size of human
limbs, the distances and diameters of planets, etc. The determination of
such measures falls elsewhere than within the qualitative determination of
measure, here of the law itself of falling bodies. On what such numbers
depend, these merely immediate and therefore empirical appearances of a
measure – on this the concrete sciences have yet to give us any information.
Here we are only concerned with this conceptual determination, namely
that in any determination of measure the empirical coefficient constitutes
the being-for-itself, but only the moment of the being-for-itself, in so far
as it is in principle and therefore as immediate. The other moment is the
development of this being-for-itself, the specific measure-determinateness
of the sides. – According to this second moment, in the ratio expressing
the motion of falling bodies (this motion that is still half conditioned and
half free) gravity is to be regarded as a force of nature. Its ratio is thus
determined by the nature of time and space, and the said specification –
the ratio of powers – therefore falls in it; the other moment mentioned
above, the simple direct ratio, expresses only a mechanical relation of time
and space, a reflectively formal velocity externally produced and externally
determined.

3. Measure has now taken on the determination of being a specified
quantitative relation, one which, as qualitative, has within it the usual
external quantum; but this quantum is not a quantum in general but
essentially the determining moment of the relation as such; it is thus an
exponent and, because of the immediacy now of its determinateness, an
invariable exponent, consequently an exponent of the already mentioned
direct ratio of the same qualities whose reciprocal quantitative relation is
at the same time specifically determined by the ratio. In the example of
measure that we have used, that of the motion of falling bodies, this direct
ratio is, as it were, anticipated and assumed as given, but, as remarked, in
this motion it still does not have concrete existence. – However, that matter
is now realized – in that its sides are both measures, one distinguished as
immediate and external and the other as internally specified, while measure
itself is the unity of the two – this constitutes a further determination. As
the unity of its two sides, measure contains the relation in which the mag-
nitudes, by virtue of the nature of the qualities, are posited as determined
and non-indifferent; its determinateness, therefore, is entirely immanent
and self-subsisting, and has at the same time collapsed into the being-for-
itself of an immediate quantum, the exponent of a direct ratio. In this the
self-determination of measure is negated, for in this immediate quantum,
its other, measure has a final self-existent determinateness; conversely, the
immediate measure which ought to be internally qualitative assumes truly
qualitative determinateness only in that other. This negative unity is real
being-for-itself, the category of a something which is the unity of qualities in a
relation of measure – a complete self-subsistence. The two resulting diverse
relations also immediately yield a twofold existence, or, more precisely,
that self-subsisting whole, as a being that exists for itself, is as such in itself
a repulsion into distinct self-subsisting somethings whose qualitative nature
and subsistence (materiality) lies in their determinateness of measure.

CHAPTER 2

Real measure

Measure is now determined as a connection of measures that make up the
quality of distinct self-subsisting somethings, or, in more common lan-
guage, things. The relations of measures just considered belong to abstract
qualities like space and time; further examples of these now to be considered
are specific gravity and then chemical properties, that is, determinations
of concrete material existence. Space and time are also moments of these
measures, but are now subordinated to other determinations and no longer
behave relative to one another only according to their own conceptual
determination. In the case of sound, for instance, the time within which a
certain number of vibrations occur, the spatial width and thickness of the
sounding body, are moments of its determination. But the magnitudes of
such idealized moments are externally determined; they no longer assume
the form of a ratio of powers but relate in the usual direct way, and har-
mony is reduced to the strictly external simplicity of numbers in relations
which are most easy to grasp; they therefore afford a satisfaction which is
the exclusive reserve of the senses, for there is nothing there of representa-
tion, imagery, thought, or the like, that would satisfy spirit. Since the sides
which now constitute the relation of measure are themselves measures, but
at the same time real somethings, their measures are, in the first instance,
immediate measures, and the relations in them direct relations. We now
have to examine the further determination of the relation of such relations.
Measure, now real measure, is as follows.
First, it is the independent measure of a type of body, a measure which
relates to other measures and, in thus relating to them, specifies them as
well as their self-subsistent materiality. This specification, as an external
connecting reference to many others in general, produces other relations,
and consequently other measures; the specific self-subsistence, for its part,
does not remain fixed in one direct relation but passes over into specific
determinacies, and this is a series of measures.
Second, the direct relations that thus result are in themselves determinate
and exclusive measures (elective affinities). 20 But because they are at the
same time only quantitatively different from one another, what we have
is a progression of relations which is in part merely external, but is also
interrupted by qualitative relations, forming a nodal line of specifically self-
subsisting things.
Third, what emerges in this progression for measure, however, is the
measureless: the measureless in general and more specifically the infinitude
of measure in which the mutually exclusive forms of self-subsistence are one
with each other, and anything self-subsistent comes to stand in negative
reference to itself.

A. THE RELATION OF INDEPENDENT MEASURES

By measures we no longer mean now merely immediate measures,
but measures that are self-subsistent
because they become within themselves
relations which are specified,
and in this being-for-itself they are thus
a something, things that are physical and at first material.

(a) However, the whole, which is a relation of such measures
is at first itself immediate.
Thus the two sides, which are as such independent measures,
have their subsistence in things external to each other,
and are posited in combination externally.

(b) But the self-subsistent materialities are
what they qualitatively are only in virtue of
the quantitative determination that they have as measures,
and are posited in virtue of this same quantitative connection
with others as non-indifferent from them (the so-called affinity);
they are the members of a series of such quantitative relations.

(c) This indifferent multifarious relating concludes
at the same time by cutting itself off as
an exclusive being-for-itself:
the so-called elective affinity.

a. Combination of two measures

As a measure relation, something is in itself determined by quanta to
which further qualities accrue; the something is the connection of such
qualities. One of them is the something's in-itselfness, according to which
This is a term which was introduced around the middle of the eighteenth century to describe the
tendency of chemical species to combine with some substances in preference to others. Goethe
popularized the term by extending its meaning to include human relations, in his novella Elective
Affinities (1809).
it is something that exists for itself, something material (a weight, if taken
intensively; or, taken extensively, an aggregate of parts); the other is however
the externality of this itselfness (something abstract and idealized, space).
These qualities are quantitatively determined, and their mutual relation
is what constitutes the qualitative nature of the material something – the
ratio of weight to volume, the specifically determined gravity. The volume,
the idealized element, is the one to be taken as unit, whereas the intensive
aspect, which in quantitative determinateness and in comparison with
the other appears as extensive magnitude, that is, an aggregate of ones
existing for themselves, is the amount. – The purely qualitative relating of
the two determinacies of magnitude in a ratio of powers has disappeared,
because with the self-subsistence of the being-for-itself (the material being)
immediacy has come back, and in this immediacy the determinateness
of magnitude is a quantum whose relation to the other side is likewise
determined in the ordinary exponent of a direct ratio.

This exponent is the specific quantum of the something, but it is an
immediate quantum and this is determined – as is also the specific nature
of the something – only in being compared with other exponents of like
ratios. The exponent constitutes the specific way in which the something is
determined in itself, the inner measure characteristic of it; but since this,
the something's measure, rests on a quantum, it is also an external and
indifferent determinateness, and the something, despite its inner deter-
mination as measure, is for this reason subject to alteration. The other to
which it can relate as alterable is not an aggregate of matters, not a quantum
in general; on the contrary, against these it holds out its specific intrinsic
determination of being; rather, it is a quantum which is at the same time
it too the exponent of such a specific ratio. Take two things of different
inner measure that stand connected and enter into composition, say, two
metals of different specific gravity. (In what way the two must otherwise
be homogeneous in nature in order for the composition to be possible, e.g.
that we cannot be speaking here of a metal combining with water, this is not
at issue here). – Now, on the one hand each of the two measures preserves
itself in the alteration they will incur because of the externality of quantum,
for each is a measure; but, on the other hand, this self-preservation is itself
a negative relating to the quantum, a specification of it, and, since this
quantum is the exponent of the ratio of measure, the self-preservation is
an alteration of the measure itself and indeed a reciprocal specification.

According to its merely quantitative determination, the compound
would be the mere sum of the two magnitudes of the one quality and
the two magnitudes of the other quality, e.g. the sum of the two weights
and the two volumes in a compound of two matters of different specific
weights, so that not only would the weight of the mixture remain equal to
the said sum but also the space which the mixture occupies would remain
equal to the sum of the two spaces. But only the weight of the mixture is
found to be in fact the sum of the weights present prior to combination;
the side capable of addition is the one which, as the one existing for itself,
has attained stability of being and consequently a permanent immediate
quantum – the weight of matter, or, what amounts to the same thing from
a quantitative standpoint, the aggregate of material parts. Alteration falls
instead in the exponents, for they are the expression of the qualitative side
of the compound, of its being-for-itself as relations of measure, and, since
the quantum as such is subject to accidental alteration by an increase which
is summed, this being-for-itself proves at the same time to be a negating
factor with respect to this externality. Since this immanent determining of
the quantitative element cannot appear, as we have seen, in the weight, it
turns up in the other quality which is the idealized side of the relation.
It might be surprising to sense-perception that, upon the mixing of two
specifically diverse materials, what results is an alteration of the total vol-
ume, normally a diminution of it, for it is space which gives stability to
matters otherwise existing outside one another. But this stability, in face of
the negativity which is in the being-for-itself, is that which has no existence
in itself, is the alterable as such. In this manner, space is posited as what it
is in truth, an idealization.

But then, not only is one of the qualitative sides posited as alterable but
measure itself, and so also the qualitative determinateness of the something
which is based on it, has shown itself not to be something stable within
but, like quantum in general, to have its determinateness in other measure-
relations.

b. Measure as a series of measure-relations

1. If something united with an other were what it is only by virtue of
simple qualitative determination, and the same also applied to the other,
then the two would only sublate themselves in being compounded. But
when something is in itself a measure-relation, it is independent yet at
the same time capable of union with another which is also in itself a
measure-relation. For in being sublated in union, each preserves itself
through the permanence of its quantitative indifference and it behaves as
the specifying moment of a new measure-relation. Its quality is enveloped
in the quantitative element; it is therefore equally indifferent towards the
exponent of the new measure is itself only some quantum or other, an
external determinateness, and its indifference is displayed by the fact that
the specifically determined something runs, with other such measures, into
precisely the same kind of neutralizations of reciprocal measure-relations;
the specific property of the something is not expressed in the one measure-
relation alone which is formed by it and another something.

2. This combination with a plurality of others which are in them-
selves likewise measures yields different ratios that therefore have different
exponents. The measure which is independent has the exponent of its
being-determined-in-itself only in the comparison with other measures;
neutrality with the others constitutes, however, its real comparison with
them; it is its comparison with them through itself. The exponents of these
ratios are however diverse, and the independent measure consequently dis-
plays its qualitative exponent as the series of these different amounts of which
it is the unit – a series of specific ways of relating to others. As one immedi-
ate quantum, the qualitative exponent expresses one single relation. The
independent measure truly differentiates itself in the characteristic series of
exponents which it, taken as a unit, forms with other equally independent
measures; for each of these measures, also brought in connection with
any other and taken as a unit, forms another series. – The relation of
such series among themselves now constitutes the qualitative aspect of the
independent measure.

Now inasmuch as this independent measure produces a series of expo-
nents with a series of such independent measures, it appears at first, when
compared with another independent measure outside this series itself, that
it is differentiated from it by virtue of the fact that this other indepen-
dent measure produces another series of exponents with the same measures
opposite to it. But in this way these two independent measures would not
be comparable – not in so far as each is regarded as a unit with respect to
its exponents, and the two series thus resulting on the basis of this connec-
tion of unit and exponents do not differ in any determinate way. The two
measures which, as independent, are supposed to be compared, are at first
differentiated from each other only as quanta; in order that their ratio be
determined, the ratio requires a unit which exists for itself and is common
to both. This determinate unit is to be sought, as shown, only where the
specific existence of the two independent measures to be compared is to
be found, namely in the ratio which the exponents of the ratio of the
series have to each other. But this ratio of the exponents is itself a unit
existing for itself and in fact determinate only in so far as the members of
the series have it as a constant ratio between them; in that way it can be
their common unit. In it alone, therefore, lies the possibility of comparing
the two independent measures which were assumed not to neutralize each
other but to be rather reciprocally indifferent. Each, taken by itself outside
the comparison, is the unit of the ratios that it establishes with the opposite
members: these are the amounts relative to this unit and hence represent
the series of exponents. But conversely, this series is the unit for the two
independent measures which when compared with each other are as quanta
to each other; as such they are themselves different amounts of their just
indicated unit.
But further, the measures which together with the two or rather indefi-
nitely many reciprocally opposing and contrasting measures yield the series
of the exponents of the ratio of these same measures, are equally inde-
pendent in themselves; each is a specific something with its own relevant
measure-relation. To this extent, they are similarly to be taken each as a
unit so that they have a series of exponents in the two or rather the inde-
terminate plurality of measures, which are first named and compared only
among themselves, and these exponents are the comparative numbers of
precisely these named measures; conversely, the comparative numbers of
the independent measures now taken singly are similarly the series of the
exponents for the members of the first series. In this way, both sides are
a series in which, first, each number is simply a unit with respect to the
opposite series in which it has its specifically determined being as a series
of exponents; second, each number is itself one of the exponents for each
member of the opposite series; and, third, it is a comparative number for
the rest of the numbers of its series and, as such an amount which belongs
to it also as an exponent, it has its unit, determined for itself, in the opposite
series.
3. In this relation there is a return to the mode in which quantum was
posited as existing for itself, namely as degree, as being simple and yet having
the magnitude of its determinateness in a quantum that exists outside it,
one which is a circle of quanta. In measure, however, this externality is not
merely a quantum and a circle of quanta, but a series of numerical ratios, and
it is in the entirety of these that the being-determined-for-itself of measure
lies. Just as it is the case in the being-for-itself of quantum as degree, the
nature of the independent measure has turned into this self-externality.
Its self-reference is at first an immediate relation, and by that very fact its
indifference to an other lies only in the quantum. Hence its qualitative
side falls in this externality, and its relating to the other becomes that which
constitutes the specific determination of this independent measure. Such a
determination thus consists solely in the quantitative mode of this relating,
and this mode is determined just as much by the other as by the measure
itself, and this other is a series of quanta while the measure, for its part,
is one such quantum. But this connection in which two specific measures
specify themselves in a third thing (the exponent) further entails that, in
the exponent, the one measure has not gone over into the other; that there
is not, therefore, only one negation, but that in the connection the two
measures are rather each negatively posited, and since each preserves itself
in it indifferently, their negation is itself also negated. This, their qualitative
unity, is thus an exclusive unit existing for itself. It is only in this moment
of exclusion that the exponents, at first comparative numbers, first gain
in them the determinateness truly specifying them reciprocally and their
difference thus becomes of qualitative nature. But such a difference has a
quantitative basis: first, the independent measure relates to a plurality of its
qualitatively other side only because in this relating it is at the same time
indifferent; second, this neutral connection is in virtue of its quantitative
basis not only alteration but is now posited as a negation of the negation
and as exclusive unit. Consequently, the affinity of an independent measure
with the plurality of measures of the other side is no longer an indifferent
connection but an elective affinity.

c. Elective affinity

The expression elective affinity used here refers to a chemical relation, as
also do the preceding neutrality and affinity. For in the chemical sphere
a matter receives its specific determinateness essentially in the connection
with its other; it exists concretely only as this non-indifference. Moreover,
this specific connection is bound up with quantity and is at the same
time not only a connection with a single other, but with a series of such
non-indifferent matters standing over and against it; combinations with
this series rest on a so-called affinity with every member of the series, even
though in this indifference each member excludes every other. We now
have to consider this connection of opposite determinations. – But it is
not just in the chemical sphere that specificity is exhibited in a circle of
combinations; the meaning of a single note also depends on its relation
to, and combination with, another note and a series of other notes; the
harmony or disharmony in such a circle of combinations constitutes its
qualitative nature which is at the same time based on quantitative ratios
that form a series of exponents, and the ratios of the two specific ratios
is what each of the combined notes is within it. The single note is the
keynote of a system, but at the same time also one member in the system
of every other note. The harmonies are exclusive elective affinities whose
characteristic quality, however, equally dissolves again in the externality of
a mere quantitative progression. – Where to find, however, the principle
of a measure for such affinities which among others and against others
(whether chemical or musical or whatever) are elective affinities, on this
more will be said in the following Remark in connection with chemical
affinity; but this is a question of a higher order which is very closely bound
up with the specific nature of what is strictly qualitative and belongs to
particular parts of the concrete science of nature.
Inasmuch as the member of a series has its qualitative unity in the way it
relates to the whole of an opposite series, the members of which differ from
each other only by virtue of the quantum required for being neutralized
with that member, the more specific determinateness in this manifold affin-
ity is likewise only quantitative. In elective affinity which is an exclusive,
qualitative connection, the relating escapes this quantitative difference. The
next determination which offers itself is this: that in accordance with the
difference in the number, hence the extensive magnitudes, of the members
of one side required for the neutralization of a member in another side, the
elective affinity of this member would be directed to the members of the
first series with all of which it has affinity. Thus transformed, the exclusion
that would thereby be established in the form of a firmer bonding against
other possibilities of combination would appear in a proportionally greater
intensity, in keeping with the previously demonstrated identity of the forms
of extensive and intensive magnitude, 21 inasmuch as the determinateness
of magnitude is in both one and the same. However, this sudden conver-
sion of the one-sided form of extensive magnitude also into its other, the
intensive form, changes nothing in so far as the nature of the fundamental
determination is concerned, which is one and the same quantum. In fact,
therefore, no exclusiveness would thereby be posited, but instead of one
single bonding a combination of any number of members could just as
well take place, provided that the portions of such members entering into
the combination corresponded to the required quantum proportionately
to the ratios between them.

But the combination which we have also called neutralization is not only
the form of intensity; the exponent is essentially a determination of measure
and, as such, exclusive. On this side of exclusive relating, numbers have lost
their continuity and their tendency to combine; their relating is one of more
or less, and this acquires a negative character; and the advantage that one
exponent has over another does not remain confined to the determinateness
of magnitude. Also equally present is the other side which makes it again
a matter of indifference to a moment of measure whether it receives its
neutralizing quantum from several opposite moments, from each according
to the determinateness that specifies it as against an other; the exclusive,
negative relating is at the same time vulnerable to this incursion of the
quantitative side. – What is posited here is thus the sudden conversion
of an indifferent, merely quantitative relating into a qualitative one, and
conversely the transition of a specifically determinate being into merely
external relation – a series of relations, sometimes of a merely quantitative
nature, sometimes specific relations and measures.

B. NODAL LINES OF MEASURE-RELATIONS

The last determination of the measure-relation was that, as specific, it is
exclusive; exclusiveness accrues to neutrality as a negative unity of the dis-
tinct moments. For this unity existing for itself, for this elective affinity, no
further principle has become available specifying its connection with other
neutralities; the specification resides only in the quantitative determination
of affinity in general, according to which it is specific amounts that neu-
tralize themselves and consequently stand opposed to other relative elective
affinities of their moments. But further, because the basic determination is
quantitative, the exclusive elective affinity continues also into these opposite
neutralities, and this continuity is not just an external comparative connec-
tion of the different ratios of the neutralities but the neutrality, as such, has
an element of separability in it, for it is as self-subsisting somethings that
the moments from whose union it has come to be enter into connection
with one or the other of the opposite series: they enter into the connection
indifferently, even though they combine in different specific determinate
amounts. This measure, internally based on such a relation, is for this rea-
son affected by an indifference of its own; it is something which is external
within and alterable in its reference to itself.

This self-reference of the measure-relation differs from the externality and
alterability which belong to its quantitative side. In contrast to these, its
self-reference is an existent qualitative foundation – a permanent, material
substrate which, since it is also the continuity of the measure with itself
in its externality, would have to contain in its quality the said principle of
specification of this externality.

Now in more developed form, as external to itself in its being-for-itself,
the exclusive measure repels itself from itself and posits itself both as another
merely quantitative relation and as another relation which is as such at the
same time another measure; it is determined as in itself a specifying unity
which produces measure-relations within itself. These relations differ from
affinities of the kind mentioned above in which an independent measure
relates itself to independent measures of another quality and to a series of
these. They occur in one and the same substrate within the same moments
of the neutrality; the self-repelling measure takes on the determination of
other only qualitatively diverse relations, and these relations likewise form
affinities and measures, alternating with those that remain only quantitative
diversities. They form in this way a nodal line of measures on a scale of
more and less.

Here we have a measure-relation, a self-subsistent reality qualitatively
distinguished from others. Such a being-for-itself, since it essentially is at
the same time a relation of quanta, is open to externality and quantitative
alteration; it has a margin within which it remains indifferent to this
alteration and does not alter its quality. But there comes a point in the
quantitative alteration at which this quality alters and the quantum shows
itself to be specifying, so that the altered quantitative relation is suddenly
turned into a measure and thereby into a new quality, a new something.
The relation that has now replaced the first is determined by the latter, both
because of the qualitative sameness of the moments which are in affinity and
because of the quantitative continuity. But since the distinction falls on the
side of this quantitative moment, the new something stands indifferently
related to the preceding one; the difference between the two is the external
one of quantum. The new something has not therefore emerged out of the
preceding one; it has emerged rather immediately from itself, that is, from
the internal specifying unity which has not yet entered into existence. –
The new quality or the new something is subjected to the same progression
of alteration, and so on, into infinity.

Inasmuch as the advance from a quality proceeds in the steady con-
tinuity of quantity, the ratios approaching the one qualifying point are
distinguished, quantitatively considered, by a more or less. In this respect,
the alteration is a gradual one. But the gradualness concerns merely the
externality of the alteration, not its qualitative moment; the preceding
quantitative relation, though infinitely near to the succeeding one, is still
another qualitative existence. From the qualitative side, therefore, the grad-
ual, merely quantitative progression which has no limits in itself, is abso-
lutely interrupted; and since in its merely quantitative connection the
newly emerging quality is with respect to the vanishing one an indetermi-
nate other, one which is indifferent to it, the transition is a leap; the two
are posited as wholly external to each other. – It is a favorite practice to
try to make an alteration conceptually comprehensible by the gradualness of
the transition leading up to it; but gradualness is rather alteration precisely
as merely indifferent, the opposite of a qualitative alteration. Rather, in
gradualness the connecting link between two realities – be they states or
self-subsistent things – is sublated, what is posited is that neither reality is
the limit of the other, but that each is absolutely external to the other. Thus,
the very point necessary for the conceptual comprehension of the alteration
is missed, although little enough is required for that purpose.

C. THE MEASURELESS

The exclusive measure, even in its realized being-for-itself, remains affected
by the moment of quantitative existence and is therefore susceptible to the
movement up and down the quantum scale of varying ratios. Something,
or a quality, based on such a ratio is driven beyond itself into the measureless,
collapsing under the mere mutation of its magnitude. Magnitude is that
constitutional aspect where an existence can be caught up in an apparently
harmless entanglement and be destroyed by it.
The abstract measureless is the quantum in general inasmuch as it lacks
internal order and is only indifferent determinateness which does not alter
measure. In the nodal lines of measures this determinateness is at the
same time posited as specifying; the abstract measureless sublates itself
into qualitative determinateness; the new measure-relation into which the
original passes over is measureless with respect to the latter but, in it,
it equally is a quality that exists for itself; what is thus posited is the
alternation of specific existences with one another and equally of them
with relations that still remain merely quantitative – an alternation ad
infinitum. Therefore, present in this transition is both the negation of the
specific relations and the negation of the quantitative progression; this is the
infinite existing for itself. – In the sphere of existence, the qualitative infinite
was the irruption of the infinite into the finite, the immediate transition
and vanishing of the “this here” into its “beyond there.” In contrast to
it, the quantitative infinite is in its very determinateness the continuity
of quantum, its continuing beyond itself. The qualitatively finite becomes
infinite; the quantitatively finite is its beyond in it: it points beyond itself. But
this infinite of the specification of measure posits both the qualitative and the
quantitative as each sublating itself into the other, and it thereby posits their
first immediate unity, which is measure in general, as returned into itself
and consequently as itself posited. The transition of the qualitative, of one
specific concrete existence into another, is such that what happens is only
an alteration of magnitude determinateness; the alteration of the qualitative
as such into the qualitative is thus posited as an external and indifferent
alteration, as a coming together with itself; the quantitative, for its part,
sublates itself by suddenly turning into the qualitative, that is, a being
which is determined in-and-for-itself. This unity which thus continues
in itself in its alternating measures is the self-subsistent matter that truly
persists, the fact.

What we have here is (a) one and the same substantial matter which
is posited as the perennial substrate of its differentiations. This detaching
of being from its determinateness already begins in quantum in general;
something is great indifferently as against the determinateness it has. In
measure, the persisting matter is itself already in itself the unity of the
qualitative and the quantitative – the two moments into which the general
sphere of being is distinguished, each as the beyond of the other; in this
way the perennial substrate begins to possess in it the determination of an
existent infinity. (b) This self-sameness of the substrate is posited in that
the qualitative independent measures into which the determining unit is
dispersed consist of only quantitative differences, so that the substrate per-
sists while being internally distinguished. (g) In the infinite progression of
the nodal series there is posited the continuation of the qualitative into the
quantitative advance as into an indifferent alteration, but equally too, there
is posited the negation of the qualitative contained in the progression and
consequently, at the same time, of the merely quantitative externality. The
quantitative pointing beyond itself to an other which is itself quantitative
perishes with the emergence of a measure-relation, of a quality, and the
qualitative transition is sublated in the very fact that the new quality is
itself only a quantitative relation. This reciprocal transition into the other
of the qualitative and the quantitative moments occurs on the basis of their
unity, and the meaning of this process is only the existence which is the
demonstration or the positing that such a substrate does underlie the process
and is the unity of its moments.

In the series of indepedent measure relations the one-sided members
of the series are immediate qualitative somethings
(specific gravities or chemical materials, bases or alkalis, acids, for instance),
and their neutralizations (by which we must understand here also the combinations
of materials of different specific gravities) are in turn self-subsistent and
themselves exclusive measure-relations, mutually indifferent totalities exist-
ing for themselves. Such relations are now determined only as nodes of one
and the same substrate. The measures and the self-subsistent forms posited
with them are consequently demoted to states. Alteration is only the muta-
tion of a state, and that which passes over is posited as remaining the same
in the mutation.

In reviewing the progressive determination which measure has gone
through, the moments of the progression can be summed up as follows.
Measure is at first itself the immediate unity of quality and quantity in the
form of a common quantum which is, however, specific. As a determi-
nateness of quantity which does not refer to an other but refers to itself,
measure is thus essentially a ratio. Consequently, it also holds its moments
as sublated and unseparated in itself; as is always the case in a concept, the
difference in the ratio is such that each of its moments is itself the unity of
the qualitative and the quantitative moment. The difference is thus real,
and it yields a multitude of measure-relations which, as formal totalities,
are each self-subsistent in themselves. The series which form the two sides
of these ratios are for each single member the same constant ordering
by which the member, as belonging to the one side, relates itself to the
whole series standing opposite it. This ordering, although a mere order,
a still wholly external unity, demonstrates itself to be, as the immanently
specifying unity of a measure which has a determinate being-for-itself,
distinguished from its specifications; but this specifying principle is not
yet the free concept, which alone gives its differences an immanent deter-
mination; the principle is at first, rather, only a substrate, a matter, for
whose differences, in order that they be totalities, that is, that they should
have in themselves the nature of their stable, self-equal substrate, there is
only available the external quantitative determination which shows itself
at the same time to be a qualitative differentiation. In this self-unity of
the substrate the determination of measure is a sublated determination, its
quality an external state determined by quantum. – This course is equally
the progressive determination of measure and its demotion to a moment.

CHAPTER 3

The becoming of essence

A. ABSOLUTE INDIFFERENCE

Being is abstract indifference,
and when this trait is to be thought by itself as being,
the abstract expression “indifferentness” has been used;
in which there is not supposed to be as yet
any kind of determinateness.
Pure quantity is this indifference in the sense of
being open to any determinations,
provided that these are external to it
and that quantity itself does not have
any link with them originating in it.
The indifference which can be called absolute,
however, is one which, through the negation of
every determinateness of being, of quality and quantity
and of their at first immediate unity, that is, of measure,
mediates itself with itself to form a simple unity.
Determinateness is in it still only a state, that is,
something qualitative and external
which has the indifference as a substrate.

But that which has thus been determined
as qualitative and external
is only a vanishing something;
as thus external with respect to being,
the qualitative sphere is the opposite of itself
and, as such, only the sublating of itself.
In this way, determinateness is still
only posited in the substrate
as an empty differentiation.
But it is precisely this empty differentiation
which is the indifference itself as result.
And this indifference is indeed concrete,
in the sense that it is self-mediated
through the negation of all the determinations of being.
As such a mediation, it contains negation and relation,
and what was called “state” is a differentiation
which is immanent to it and self-referring.
It is precisely this externality and its vanishing
which make the unity of being into an indifference:
consequently, they are inside this indifference,
which thereby ceases to be only a substrate
and, within, only abstract.

B. INDIFFERENCE AS INVERSE RATIO OF ITS FACTORS

We now have to see how this determination of indifference is posited in
the indifference itself and the latter is posited, therefore, as existing for itself.

1. The reduction of at first independently accepted measure-relations
establishes their one substrate; this substrate is their continuing into one
another and is, therefore, the one indivisible independent measure which
is wholly present in its differentiation. Present for this differentiation are
the two determinations contained in the measure, quality and quantity,
and everything depends on how these two are posited in it. But this is in
turn determined by the fact that the substrate is at first posited as result
and, though in itself mediation, this mediation is not yet posited as such
in it; for this reason, it is in the first instance substrate and, with respect to
determinateness, indifference.

Consequently, the difference present in it is at first essentially one which
is only quantitative and external; there simply are two different quanta of
one and the same substrate which would thus be their sum, itself posited as a
quantum. But the indifference is this fixed measure, the implicitly existent
absolute limit which, as connected to those differences, would not itself
be in itself a quantum, and would not in any way enter into opposition
with others, whether as sum or also as exponent, be those others sums
or indifferences. It is only the abstract determinateness which falls into
the indifference; the two quanta, in order that they may be posited in it
as moments, are alterable, indifferent, greater or smaller relative to one
another. However, inasmuch as they are restricted by the fixed limit of
their sum, they are at the same time related to each other not externally,
but negatively, and this is now the qualitative determination in which they
stand to each other. Accordingly, they stand in inverse ratio to each other.
This relation differs from the earlier formal inverted ratio 35 inasmuch as
the limit is here a real substrate, and each of the two sides is posited as
having to be in itself the whole.

According to the qualitative determinacy just stated, the difference is
present, further, in the form of two qualities, each of which is sublated
by the other and yet, since the two are held together in the one unity
which they constitute, is inseparable from it. The substrate itself, as the
indifference, is in itself likewise the unity of the two qualities; consequently,
each of the sides of the relation equally contains both sides within itself
and is distinguished from the other by a more of one quality and a less
of the other, or conversely. The one quality, through its quantum, only
predominates on the one side, as does the other quality on the other side.
Thus each side is in it an inverted relation which, as formal, recurs in
the two distinguished sides. The sides themselves thus continue into each
other also according to their qualitative determinations; each of the qual-
ities relates itself in the other to itself and is present in each of the two
sides, only in a different quantum. Their quantitative difference is that
indifference in accordance with which they continue into each other, and
this continuation is the self-sameness of the qualities in each of the two
unities. – The sides, however, each containing the whole of the determi-
nations and consequently the indifference itself, are thus at the same time
posited as self-subsistent vis-a-vis each other.

2. As this indifference, being is now the determinateness of measure no
longer in its immediacy but in the developed manner just indicated; it is
indifference because it is in itself the whole of the determinations of being
now resolved into this unity; and it is existence as well, as a totality of the
posited realization, in which the moments themselves are the totality of
the indifference existing in itself, sustained by the latter as their unity. But
because the unity is held fast only as indifference and consequently only
implicitly in itself, and the moments are not yet determined as existing for
themselves, that is, are not yet determined as sublating themselves into unity
internally and through each other, what is here present is therefore simply the
indifference of the unity itself towards itself as a developed determinateness.
This thus indivisible independent measure is now to be more closely
examined. It is immanent in all its determinations and in them it remains in
unity with itself and undisturbed by them. But, (a) since the determinacies
sublated in it implicitly remain the totality, they emerge in it groundlessly.
The implicit being of indifference and its existence are thus unconnected;
the determinacies show up in the indifference in an immediate manner
and the indifference is in each of them entirely the same. The difference
between them is thus posited at first as sublated, hence as quantitative –
for this reason, therefore, not as a self-repelling; and the indifference not as
self-determining, but as having and becoming the determinate being that
it has only externally.
(b) The two moments are in inverse quantitative relation – a fluctuat-
ing on the scale of magnitude which is not however determined by the
indifference, which is precisely the indifference of the fluctuation, but only
externally. For the determining appeal is made to an other which lies out-
side the indifference. The absolute, as indifference, has in this respect the
second defect of quantitative form, namely that the determinateness of
the difference is not determined by the absolute itself, just as it has the
first defect in that the differences emerge in it only in general, that is, the
positing of them is something immediate, not a self-mediation.
(g) The quantitative determinateness of the moments which are now
sides of the relation constitutes the mode of their subsistence; their existence is
by virtue of this indifference withdrawn from the transitoriness of quality.
But they do have a subsistence of their own in themselves, one that differs
from this quantitative existence, for they are in themselves the indifference
itself, each the unity itself of the two qualities into which the qualitative
moment splits itself. The difference of the two moments is restricted by
the fact that the one quality is posited on the one side with a more and in
the other with a less, and the other is posited in inverse order accordingly.
Each side is thus in it the totality of the indifference. – Each of the two
qualities taken singly for itself likewise remains the same sum which the
indifference is; it continues from one side into the other without being
restricted by the quantitative limit which is thereby posited in it. At this,
the determinations come into immediate opposition, an opposition which
develops into contradiction, as we must now see.

3. Namely, each quality enters inside each side in connection with the
other, and it does so in such a manner that, as has been determined, this
connection also is supposed to be only a quantitative difference. If the
two qualities are both self-subsistent – something like sensible materials
independent of each other – then the whole determinateness of indifference
falls apart; their unity and totality would be empty names. But they are
at the same time determined as comprised into one unity, as inseparable,
each having meaning and reality only in this one qualitative connecting
reference to the other. But now, because their quantitativeness 36 is simply
and solely of this qualitative nature, each reaches only as far as the other. If
they are assumed to differ as quanta, then the one would reach beyond
the other and would have in this more an indifferent existence which the
other would not have. As qualitatively connected, however, each is only in
so far as the other is. – The result is that they are in equilibrium, so that to
the extent that one increases or decreases, the other likewise increases or
decreases and would do so in the same proportion.
On the basis, therefore, of their qualitative connection, there is no
question of a quantitative difference or of a more of the one quality. The
more by which the one of the two connected moments would exceed the
other would be only an unstable determination, or would only be the other
itself again; but, in this equality of the two, neither would then be there,
for their existence would have to rest on the inequality of their quantum. –
Each of these supposed factors vanishes, whether the one factor is assumed
to exceed the other or to be equal to it. From the standpoint of quantitative
representation, the vanishing appears as a disturbance of the equilibrium,
one factor becoming greater than the other; the sublation of the quality
of the other and its instability are thus posited; the first factor becomes
the predominant one as the other diminishes with accelerated velocity and
is overcome by it; this in turn constitutes itself as the one self-subsistent
factor; with this, however, there are no longer two specific moments as
factors but only the one whole.
This unity thus posited as the totality of the process of determining, itself
determined in this process as indifference, is a contradiction all around.
It must therefore be posited as this self-sublating contradiction, and be
determined as subsistence existing for itself, one which no longer has a
merely indifferent unity for result but a unity immanently negative and
absolute. This is essence.

C. TRANSITION INTO ESSENCE

Absolute indifference is the final determination of being
before the latter becomes essence;
but it does not attain essence.
It shows that it still belongs to the sphere of being
because it is still determined as indifferent,
and therefore difference is external to it, quantitative.
This externality is its existence, by which
it finds itself at the same time in the opposition of
being determined over against it as existing in itself,
not as being thought as the absolute that exists for itself.
Or again, it is external reflection
which insists that specific determinations,
whether in themselves or in the absolute,
are one and the same;
that their difference is only an indifferent one,
not a difference in itself.
What is still missing here is that
this reflection should sublate itself,
that it would cease to be the external
reflection of thought, of a subjective consciousness,
but that it would be rather the very determination of
the difference of that unity;
a unity which would then prove itself to be the absolute negativity,
the unity's indifference towards itself,
towards its own indifference no less than towards otherness.

But this self-sublation of the determination of
indifference has already manifested itself;
in the progressive positing of its being
it has shown itself on all sides to be contradiction.
Indifference is in itself the totality
in which all the determinations of being are sublated and contained;
thus it is the substrate,
but at first only in the one-sided determination of being-in-itself,
and consequently the differences,
the quantitative difference and the inverseratio of factors,
are present in it as external.
As thus the contradiction of itself and its determinateness,
of its implicitly existent determination
and of its posited determinateness,
it is the negative totality whose determinacies
have internally sublated themselves,
consequently, have also sublated the one-sidedness
of their substrate, their in-itselfness.
Indifference, now posited as what it in fact is,
is simple and infinitely negative self-reference,
the incompatibility of itself with itself,
the repelling of itself from itself.
Determining and being determined are
not a transition,
nor an external alteration,
nor again an emergence of determinations in it,
but its own referring to itself
which is the negativity of itself,
of its in-itselfness.

But as so repelled,
the determinations are not self-possessed;
do not emerge as self-subsistent or external
but are rather as moments:
first, as belonging to the unity
whose existence is still only implicit,
they are not let go by it but are rather borne by it
as their substrate and are filled by it alone;
and, second, as determinations immanent
to the unity as it exists for itself,
they are only through their repulsion from themselves.
Instead of some existent or other,
as they are in the whole sphere of being,
they now are simply and solely as posited,
with the sole determination and significance of
being referred to their unity
and hence each to the other and to negation,
marked by this their relativity.

Being in general and the being or immediacy
of the different determinacies have thereby vanished
just as much as the in-itselfness,
and the unity is being,
immediately presupposed totality,
so that it is only this simple self-reference,
mediated by the sublation of this presupposition,
and this pre-supposedness, the immediate being,
is itself only a moment of its repelling:
the original self-subsistence and self-identity are only
as the resulting infinite self-rejoining.
And so is being determined as essence:
being which, through the sublation of being,
is simple being with itself.
