C. MEASURE

Measure is the qualitative quantum,
at first in the immediate sense as a quantum
with which an existence or a quality is bound up.

Insofar as quality and quantity only exist in immediate unity in measure,
their difference emerges in them in an equally immediate manner.
Insofar as this is the case, the specific quantum is in part mere quantum,
and existence is capable of increase and decrease without the measure
being sublated thereby (the measure is in this respect a rule).
In part, however, the alteration of the quantum is
likewise an alteration of quality.

The measureless is initially this process of a measure,
by virtue of its quantitative nature,
of passing beyond its qualitative determinacy.
However, since the other quantitative relationship, [namely,]
the measurelessness of the first,
is equally qualitative,
the measureless is likewise a measure.
Both transitions, from quality to quantum
and from the latter to the former,
can be represented again as an infinite progression,
as the suspending and re-establishing of
the measure in the measureless.

What in fact happens here is that the immediacy,
which still belongs to the measure as such, is sublated.
Quality and quantity themselves are first
in the measure as immediate,
and it is merely their relative identity.
Yet measure turns out to sublate itself in the measureless.
The latter, while it is the negation of measure,
is nonetheless itself the unity of quantity and quality,
and hence displays itself just as much as
simply coming together with itself.

The infinite, the affirmation as negation of negation,
now has for its sides quality and quantity
instead of the more abstract sides of being and nothing,
something and an other, and so on.
Those [sides of the infinite] have made the transition
(a) first of quality into quantity
and of quantity into quality
and been demonstrated by this means to be negations.
(b) But in their unity (measure)
they are at first distinct
and one is only by being mediated by the other.
And (c), after the immediacy of this unity
has proved to be self-sublating,
this unity is now posited as what it is in itself,
as simple relation-to-itself that contains
being in general and its sublated forms within itself.
Being or the immediacy that is mediation with itself
and relation to itself through the negation of itself,
and thus equally mediation that sublates itself
towards relation-to-itself, towards immediacy, is essence.
