Book One

The Doctrine of Being

WITH WHAT MUST THE BEGINNING OF SCIENCE BE MADE?

It is only in recent times that there has been a new awareness of
the difficulty of finding a beginning in philosophy,
and the reason for this difficulty,
and so also the possibility of resolving it,
have been discussed in a variety of ways.
The beginning of philosophy must be
either something mediated or something immediate,
and it is easy to show that it can be neither
the one nor the other; so either way of beginning runs into contradiction.
The principle of a philosophy also expresses a beginning, of course,
but not so much a subjective as an objective one,
the beginning of all things.
The principle is a somehow determinate content:
“water,” “the one,” “nous,” “idea,” or “substance,” “monad,” etc. or,
if it designates the nature of cognition and is therefore meant simply
as a criterion rather than an objective determination,
as “thinking,” “intuition,” “sensation,” “I,” even “subjectivity,” then
here too the interest still lies in the content determination.
The beginning as such, on the other hand, as something
subjective in the sense that it is an accidental way of introducing the
exposition, is left unconsidered, a matter of indifference, and consequently
also the need to ask with what a beginning should be made remains of no
importance in face of the need for the principle in which alone the interest
of the fact seems to lie, the interest as to what is the truth, the absolute
ground of everything.
But the modern perplexity about a beginning proceeds from
a further need which escapes those who are either busy demonstrating
their principle dogmatically or skeptically looking for
a subjective criterion against dogmatic philosophizing,
and is outright denied by those who begin, like a shot from a pistol,
from their inner revelation, from faith, intellectual intuition, etc. and
who would be exempt from method and logic.
If earlier abstract thought is at first interested only
in the principle as content, but is driven as philosophical culture
advances to the other side to pay attention to the conduct of
the cognitive process, then the subjective activity has
also been grasped as an essential moment of objective truth, and
with this there comes the need to unite the method with the content,
the form with the principle.
Thus the principle ought to be also the beginning, and
that which has priority for thinking ought to be also
the first in the process of thinking.
Here we only have to consider how the logical beginning appears.
The two sides from which it can be taken have already been named, namely
either by way of mediation as result, or immediately as beginning proper.
This is not the place to discuss the question apparently so important
to present-day culture, whether the knowledge of truth is an immediate
awareness that begins absolutely, a faith, or rather a mediated knowledge.
In so far as the issue allows passing treatment, this has already been done
elsewhere (in my Encyclopedia of the Philosophical Sciences, 3rd edn, in the
Prefatory Concept, §§21ff.). Here we may quote from it only this, that
there is nothing in heaven or nature or spirit or anywhere else that does
not contain just as much immediacy as mediation, so that both these
determinations prove to be unseparated and inseparable and the opposition
between them nothing real. As for a scientific discussion, a case in point is
every logical proposition in which we find the determinations of immediacy
and mediacy and where there is also entailed, therefore, a discussion of their
opposition and their truth. This opposition, when connected to thinking,
to knowledge, to cognition, assumes the more concrete shape of immediate
or mediated knowledge, and it is then up to the science of logic to consider
the nature of cognition in general, while the more concrete forms of
the same cognition fall within the scope of the science of spirit and the
phenomenology of spirit. But to want to clarify the nature of cognition
prior to science is to demand that it should be discussed outside science,
and outside science this cannot be done, at least not in the scientific manner
which alone is the issue here.
A beginning is logical in that it is to be made in the element of a free,
self-contained thought, in pure knowledge; it is thereby mediated, for pure
knowledge is the ultimate and absolute truth of consciousness. We said in
the Introduction that the Phenomenology of Spirit is the science of con-
sciousness, its exposition; that consciousness has the concept of science,
that is, pure knowledge, for its result. 3 To this extent, logic has for its
presupposition the science of spirit in its appearance, a science which
contains the necessity, and therefore demonstrates the truth, of the stand-
point which is pure knowledge and of its mediation. In this science of
spirit in its appearance the beginning is made from empirical, sensuous
consciousness, and it is this consciousness which is immediate knowledge
in the strict sense; there, in this science, is where its nature is discussed.
Any other consciousness, such as faith in divine truths, inner experience,
knowledge through inner revelation, etc., proves upon cursory reflection
to be very ill-suited as an instance of immediate knowledge. In the said
treatise, immediate consciousness is also that which in the science comes
first and immediately and is therefore a presupposition; but in logic the
presupposition is what has proved itself to be the result of that preceding
consideration, namely the idea as pure knowledge. Logic is the pure science,
that is, pure knowledge in the full compass of its development. But in
that result the idea has the determination of a certainty that has become
truth; it is a certainty which, on the one hand, no longer stands over and
against a subject matter confronting it externally but has interiorized it, is
knowingly aware that the subject matter is itself; and, on the other hand,
has relinquished any knowledge of itself that would oppose it to objectivity
and would reduce the latter to a nothing; it has externalized this subjectivity
and is at one with its externalization.
Now starting with this determination of pure knowledge, all that we have
to do to ensure that the beginning will remain immanent to the science
of this knowledge is to consider, or rather, setting aside every reflection,
simply to take up, what is there before us.
Pure knowledge, thus withdrawn into this unity, has sublated every
reference to an other and to mediation; it is without distinctions and as
thus distinctionless it ceases to be knowledge; what we have before us is
only simple immediacy.
Simple immediacy is itself an expression of reflection; it refers to the
distinction from what is mediated. The true expression of this simple
immediacy is therefore pure being. Just as pure knowledge should mean
nothing but knowledge as such, so also pure being should mean nothing
but being in general; being, and nothing else, without further determination
and filling.
Being is what makes the beginning here; it is presented indeed as origi-
nating through mediation, but a mediation which at the same time sublates
itself, and the presupposition is of a pure knowledge which is the result
of finite knowledge, of consciousness. But if no presupposition is to be
made, if the beginning is itself to be taken immediately, then the only
determination of this beginning is that it is to be the beginning of logic, of
thought as such. There is only present the resolve, which can also be viewed
as arbitrary, of considering thinking as such. The beginning must then be
absolute or, what means the same here, must be an abstract beginning; and
so there is nothing that it may presuppose, must not be mediated by anything
or have a ground, ought to be rather itself the ground of the entire science.
It must therefore be simply an immediacy, or rather only immediacy itself.
Just as it cannot have any determination with respect to an other, so too it
cannot have any within; it cannot have any content, for any content would
entail distinction and the reference of distinct moments to each other, and
hence a mediation. The beginning is therefore pure being.
After this simple exposition of what alone first belongs to this simplest of
all simples, the logical beginning, we may add the following further reflec-
tions which should not serve, however, as elucidation and confirmation of
the exposition – this is complete by itself – but are rather occasioned by
notions and reflections which may come our way beforehand and yet, like
all other prejudices that antedate the science of logic, must be disposed
of within the science itself and are therefore to be patiently deferred until
then.
The insight that absolute truth must be a result, and conversely, that a
result presupposes a first truth which, because it is first, objectively con-
sidered is not necessary and from the subjective side is not known – this
insight has recently given rise to the thought that philosophy can begin only
with something which is hypothetically and problematically true, and that
at first, therefore, philosophizing can be only a quest. This is a view that
Reinhold has repeatedly urged in the later stages of his philosophizing, 4
and which must be given credit for being motivated by a genuine interest
in the speculative nature of philosophical beginning. A critical examination
of this view will also be an occasion for introducing a preliminary under-
standing of what progression in logic generally means, for the view has
direct implications for the nature of this advance. Indeed, as portrayed by
it, progression in philosophy would be rather a retrogression and a ground-
ing, only by virtue of which it then follows as result that that, with which
the beginning was made, was not just an arbitrary assumption but was in
fact the truth, and the first truth at that.
It must be admitted that it is an essential consideration – one which
will be found elaborated again within the logic itself – that progression is a
retreat to the ground, to the origin and the truth on which that with which
the beginning was made, and from which it is in fact produced, depends. –
Thus consciousness, on its forward path from the immediacy with which
it began, is led back to the absolute knowledge which is its innermost
truth. This truth, the ground, is then also that from which the original
first proceeds, the same first which at the beginning came on the scene as
something immediate. – It is most of all in this way that absolute spirit
(which is revealed as the concrete and supreme truth of all being) comes
to be known, as at the end of the development it freely externalizes itself,
letting itself go into the shape of an immediate being – resolving itself into
the creation of a world which contains all that fell within the development
preceding that result and which, through this reversal of position with
its beginning, is converted into something dependent on the result as
principle. Essential to science is not so much that a pure immediacy should
be the beginning, but that the whole of science is in itself a circle in which
the first becomes also the last, and the last also the first.
Conversely, it follows that it is just as necessary to consider as result that
into which the movement returns as to its ground. In this respect, the first
is just as much the ground, and the last a derivative; since the movement
makes its start from the first and by correct inferences arrives at the last as the
ground, this last is result. Further, the advance from that which constitutes
the beginning is to be considered only as one more determination of the
same advance, so that this beginning remains as the underlying ground of
all that follows without vanishing from it. The advance does not consist in
the derivation of an other, or in the transition to a truly other: inasmuch
as there is a transition, it is equally sublated again. Thus the beginning
of philosophy is the ever present and self-preserving foundation of all
subsequent developments, remaining everywhere immanent in its further
determinations.
In this advance the beginning thus loses the one-sidedness that it has
when determined simply as something immediate and abstract; it becomes
mediated, and the line of scientific forward movement consequently turns
into a circle. – It also follows that what constitutes the beginning, because it
is something still undeveloped and empty of content, is not yet truly known
at that beginning, and that only science, and science fully developed, is the
completed cognition of it, replete with content and finally truly grounded.
But for this reason, because it is as absolute ground that the result finally
emerges, the progression of this cognition is not anything provisory, still
problematic and hypothetical, but must be determined through the nature
of the matter at issue and of the content itself. Nor is the said beginning an
arbitrary and only temporary assumption, 5 or something which seems to
be an arbitrary and tentative presupposition but of which it is subsequently
shown that to make it the starting point was indeed the right thing to do;
this is not as when we are instructed to make certain constructions in order
to aid the proof of a geometrical theorem, and only in retrospect, in the
course of the proof, does it become apparent that we did well to draw
precisely these lines and then, in the proof itself, to begin by comparing
them or the enclosed angles – though the line-drawing or the comparing
themselves escape conceptual comprehension.
So we have just given, right within science itself, the reason why in pure
science the beginning is made with pure being. This pure being is the unity
into which pure knowledge returns, or if this knowledge, as form, is itself
still to be kept distinct from its unity, then pure being is also its content.
It is in this respect that this pure being, this absolute immediate, is just as
absolutely mediated. However, just because it is here as the beginning, it is
just as essential that it should be taken in the one-sidedness of being purely
immediate. If it were not this pure indeterminacy, if it were determined, it
would be taken as something mediated, would already be carried further
than itself: a determinate something has the character of an other with
respect to a first. It thus lies in the nature of a beginning itself that it should
be being and nothing else. There is no need, therefore, of other preparations
to enter philosophy, no need of further reflections or access points.
Nor can we derive a more specific determination or a more positive content
for the beginning of philosophy from the fact that it is such a beginning. 6
For here, at the beginning, where the fact itself is not yet at hand, philosophy
is an empty word, a received and yet unjustified notion. Pure knowledge
yields only this negative determination, namely that the beginning ought
to be abstract. If pure being is taken as the content of pure knowledge,
then the latter must step back from its content, allowing it free play and
without determining it further. – Or again, inasmuch as pure being is
to be considered as the unity into which knowledge has collapsed when
at the highest point of union with its objectification, knowledge has then
disappeared into this unity, leaving behind no distinction from it and hence
no determination for it. – Nor is there anything else present, any content
whatever, that could be used to make a more determinate beginning with
But, it may be said, the determination of being assumed so far as the
beginning can also be let go, so that the only requirement would be that a
pure beginning should be made. Nothing would then be at hand except the
beginning itself, and we must see what this would be. – This position could
be suggested also for the benefit of those who are either not comfortable,
for whatever reason, with beginning with being and even less with the
transition into nothing that follows from being, or who simply do not
know how else to make a beginning in a science except by presupposing
a representation which is subsequently analyzed, the result of the analysis
then yielding the first determinate concept in the science. If we also want
to test this strategy, we must relinquish every particular object that we may
intend, since the beginning, as the beginning of thought, is meant to be
entirely abstract, entirely general, all form with no content; we must have
nothing, therefore, except the representation of a mere beginning as such.
We have, therefore, only to see what there is in this representation.
As yet there is nothing, and something is supposed to become. The
beginning is not pure nothing but a nothing, rather, from which something
is to proceed; also being, therefore, is already contained in the beginning.
Therefore, the beginning contains both, being and nothing; it is the unity
of being and nothing, or is non-being which is at the same time being, and
being which is at the same time non-being.
Further, being and nothing are present in the beginning as distinguished;
for the beginning points to something other – it is a non-being which refers
to an other; that which begins, as yet is not; it only reaches out to being.
The being contained in the beginning is such, therefore, that it distances
itself from non-being or sublates it as something which is opposed to it.
But further, that which begins already is, but is also just as much not yet.
The opposites, being and non-being, are therefore in immediate union in
it; or the beginning is their undifferentiated unity.
An analysis of the beginning would thus yield the concept of the unity of
being and non-being – or, in a more reflected form, the concept of the unity
of differentiated and undifferentiated being – or of the identity of identity
and non-identity. 7 This concept could be regarded as the first, purest, that
is, most abstract, definition of the absolute – as it would indeed be if the
issue were just the form of definitions and the name of the absolute. In this
sense, just as such an abstract concept would be the first definition of the
absolute, so all further determinations and developments would be only
more determinate and richer definitions of it. But let those who are not
satisfied with being as the beginning, since being passes over into nothing
and what emerges is the unity of the two – let them consider what is more
likely to satisfy them: this beginning that begins with the representation
of the beginning and an analysis of it (an analysis that is indeed correct yet
equally leads to the unity of being and non-being) or a beginning which
makes being the beginning.
But, regarding this strategy, there is still a further observation to be made.
The said analysis presupposes that the representation of the beginning is
known; its strategy follows the example of other sciences. These presuppose
their object and presume that everyone has the same representation of it and
will find in it roughly the same determinations which they have collected
here or there, through analysis, comparison, and sundry argumentation,
and they then offer as its representations. But that which constitutes the
absolute beginning must likewise be something otherwise known; now,
if it is something concrete and hence in itself variously determined, then
this connectedness which it is in itself is presupposed as a known; the con-
nectedness is thereby adduced as something immediate, which however it is
not; for it is connectedness only as a connection of distinct elements and
therefore contains mediation within itself. Further, the accidentality and
the arbitrariness of the analysis and the specific mode of determination
affect the concrete internally. Which determinations are elicited depends
on what each individual happens to discover in his immediate accidental
representation. The connection contained within a concrete something,
within a synthetic unity, is necessary only in so far as it is not found already
given but is produced rather by the spontaneous return of the moments
back into this unity, a movement which is the opposite of the analytical
procedure that occurs rather within the subject and is external to the fact
itself.
Here we then have the precise reason why that with which the begin-
ning is to be made cannot be anything concrete, anything containing a
connection within its self. It is because, as such, it would presuppose within
itself a process of mediation and the transition from a first to an other, of
which process the concrete something, now become a simple, would be
the result. But the beginning ought not itself to be already a first and an
other, for anything which is in itself a first and an other implies that an
advance has already been made. Consequently, that which constitutes the
beginning, the beginning itself, is to be taken as something unanalyzable,
taken in its simple, unfilled immediacy; and therefore as being, as complete
emptiness.
If, impatient with this talk of an abstract beginning, one should say that
the beginning is to be made, not with the beginning, but directly with the
fact itself, well then, this subject matter is nothing else than that empty
being. For what this subject matter is, that is precisely what ought to result
only in the course of the science, what the latter cannot presuppose to
know in advance.
On any other form otherwise assumed in an effort to have a beginning
other than empty being, that beginning would still suffer from the same
defects. Let those who are still dissatisfied with this beginning take upon
themselves the challenge of beginning in some other way and yet avoiding
such defects.
But we cannot leave entirely unmentioned a more original beginning
to philosophy which has recently gained notoriety, the beginning with the
“I.” It derived from both the reflection that all that follows from the first
truth must be deduced from it, and the need that this first truth should be
something with which one is already acquainted, and even more than just
acquainted, something of which one is immediately certain. This proposed
beginning is not, as such, an accidental representation, or one which might
be one thing to one subject and something else to another. For the “I,” this
immediate consciousness of the self, appears from the start to be both itself
an immediate something and something with which we are acquainted in a
much deeper sense than with any other representation; true, anything else
known belongs to this “I,” but it belongs to it as a content which remains
distinct from it and is therefore accidental; the “I,” by contrast, is the simple
certainty of its self. But the “I” is, as such, at the same time also a concrete,
or rather, the “I” is the most concrete of all things – the consciousness of
itself as an infinitely manifold world. Before the “I” can be the beginning
and foundation of philosophy, this concreteness must be excised, and this
is the absolute act by virtue of which the “I” purifies itself and makes its
entrance into consciousness as abstract “I.” But this pure “I” is now not
immediate, is not the familiar, ordinary “I” of our consciousness to which
everyone immediately links science. Truly, that act of excision would be
none other than the elevation to the standpoint of pure knowledge in
which the distinction between subject and object has disappeared. But
as thus immediately demanded, this elevation is a subjective postulate;
before it proves itself as a valid demand, the progression of the concrete “I”
from immediate consciousness to pure knowledge must be demonstratively
exhibited within the “I” itself, through its own necessity. Without this
objective movement, pure knowledge, also when defined as intellectual
intuition, appears as an arbitrary standpoint, itself one of those empirical
states of consciousness for which everything depends on whether someone,
though not necessarily somebody else, discovers it within himself or is able
to produce it there. But inasmuch as this pure “I” must be essential, pure
knowledge – and pure knowledge is however one which is only posited in
individual consciousness through an absolute act of self-elevation, is not
present in it immediately – we lose the very advantage which was to derive
from this beginning of philosophy, namely that it is something with which
everyone is well acquainted, something which everyone finds within himself
and to which he can attach further reflection; that pure “I,” on the contrary,
in its abstract, essential nature, is to ordinary consciousness an unknown,
something that the latter does not find within itself. What comes with it is
rather the disadvantage of the illusion that we are speaking of something
supposedly very familiar, the “I” of empirical self-consciousness, whereas
at issue is in fact something far removed from the latter. Determining
pure knowledge as “I” acts as a continuing reminder of the subjective “I”
whose limitations should rather be forgotten; it leads to the belief that
the propositions and relations which result from the further development
of the “I” occur within ordinary consciousness and can be found pre-
given there, indeed that the whole issue is about this consciousness. This
mistake, far from bringing clarity, produces instead an even more glaring
and bewildering confusion; among the public at large, it has occasioned
the crudest of misunderstandings.
Further, as regards the subjective determinateness of the “I” in general,
pure knowledge does remove from it the restriction that it has when under-
stood as standing in unsurmountable opposition to an object. But for this
reason it would be at least superfluous still to hold on to this subjective
attitude by determining pure knowledge as “I.” For this determination not
only carries with it that troublesome duality of subject and object; on closer
examination, it also remains a subjective “I.” The actual development of
the science that proceeds from the “I” shows that in the course of it the
object has and retains the self-perpetuating determination of an other with
respect to the “I”; that therefore the “I” from which the start was made
does not have the pure knowledge that has truly overcome the opposition
of consciousness, but is rather still entangled in appearance.
In this connection, there is the further essential observation to be made
that, although the “I” might well be determined to be in itself pure knowl-
edge or intellectual intuition and declared to be the beginning, in science
we are not concerned with what is present in itself or as something inner, bu
with the external existence 9 rather of what in thought is inner and with the
determinateness which this inner assumes in that existence. But whatever
externalization there might be of 10 intellectual intuition at the beginning of
science, or – if the subject matter of science is called the eternal, the divine,
the absolute – of the eternal or absolute, this cannot be anything else than a
first, immediate, simple determination. Whatever richer name be given to
it than is expressed by mere being, the only legitimate consideration is how
such an absolute enters into discursive 11 knowledge and the enunciation of
this knowledge. Intellectual intuition might well be the violent rejection
of mediation and of demonstrative, external reflection. However, anything
which it says over and above simple immediacy would be something con-
crete, and this concrete would contain a diversity of determinations in it.
But, as already remarked, the enunciation and exposition of this concrete
something is a process of mediation which starts with one of the determina-
tions and proceeds to another, even though this other returns to the first –
and this is a movement which, moreover, is not allowed to be arbitrary
or assertoric. Consequently, that from which the beginning is made in any
such exposition is not something itself concrete but only the simple imme-
diacy from which the movement proceeds. Besides, what is lacking if we
make something concrete the beginning is the demonstration which the
combination of the determinations contained in it requires.
Therefore, if in the expression of the absolute, or the eternal, or God (and
God would have the perfectly undisputed right that the beginning be made
with him), if in the intuition or the thought of them, there is more than
there is in pure being, then this more should first emerge in a knowledge
which is discursive 12 and not figurative; 13 as rich as what is implicitly
contained in knowledge may be, the determination that first emerges in
it is something simple, for it is only in the immediate that no advance is
yet made from one thing to an other. Consequently, whatever in the richer
representations of the absolute or God might be said or implied over and
above being, all this is at the beginning only an empty word and only
being; this simple determination which has no further meaning besides,
this empty something, is as such, therefore, the beginning of philosophy.
This insight is itself so simple that this beginning is as beginning in
no need of any preparation or further introduction, and the only possible
purpose of this preliminary disquisition regarding it was not to lead up to
it but to dispense rather with all preliminaries.

GENERAL DIVISION OF BEING

Being is determined, first, as against another in general;
secondly, it is internally self-determining;
thirdly, as this preliminary division is cast off,
it is the abstract indeterminateness and immediacy
in which it must be the beginning.
According to the first determination,
being partitions itself off from essence,
for further on in its development it proves to be
in its totality only one sphere of the concept, and
to this sphere as moment it opposes another sphere.
According to the second, it is the sphere within which
fall the determinations and the entire movement of its reflection.

In this, being will posit itself in three determinations:

I. as determinateness; as such, quality;

II. as sublated determinateness; magnitude, quantity;

III. as qualitatively determined quantity; measure.

This division, as was generally remarked of such divisions in the
Introduction, is here a preliminary statement;
its determinations must first arise from the movement of being itself,
and receive their definitions and justification by virtue of it.
As regards the divergence of this division from the usual listing
of the categories,  namely quantity, quality, relation and modality,
(for Kant, incidentally, these are supposed to be only classifications
of his categories, but are in fact themselves categories,
only more abstract ones; about this, there is nothing to remark here,
since the entire listing will diverge from the usual ordering and meaning
of the categories at every point.)
This only can perhaps be remarked, that the determination of quantity
is ordinarily listed ahead of quality and as a rule
this is done for no given reason.
It has already been shown that the beginning is made with being as such,
and hence with qualitative being.
It is clear from a comparison of quality with quantity that
the former is by nature first.
For quantity is quality which has already become negative;
magnitude is the determinateness which, no longer one with being but
already distinguished from it, is the sublated quality
that has become indifferent.
It includes the alterability of being without altering the fact itself,
namely being, of which it is the determination;
qualitative determinateness is on the contrary one with its being,
it neither transcends it nor stays within it
but is its immediate restrictedness.
Hence quality, as the determinateness which is immediate,
is the first and it is with it that the beginning is to be made.
Measure is a relation, not relation in general but specifically
of quality and quantity to each other;
the categories dealt with by Kant under relation
will come up elsewhere in their proper place.
Measure, if one so wishes, can be considered also a modality;
but since with Kant modality is no longer supposed to make up
a determination of content, but only concerns the reference of
the content to thought, to the subjective, the result is a
totally heterogeneous reference that does not belong here.
The third determination of being falls within the section Quality
inasmuch as being, as abstract immediacy, reduces itself to
one single determinateness as against its other determinacies
inside its sphere.

Determinateness (Quality)

Being is the indeterminate immediate;
it is free of determinateness with respect to essence,
just as it is still free of any determinateness that it can
receive within itself.
This reflectionless being is being as it immediately is only within.
Since it is immediate, it is being without quality;
but the character of indeterminateness attaches to it in itself
only in opposition to what is determinate or qualitative.
Determinate being thus comes to stand over and against being in general;
with that, however, the very indeterminateness
of being constitutes its quality.
It will therefore be shown that the first being is
in itself determinate, and therefore secondly,
that it passes over into existence, is existence;
that this latter, however, as finite being, sublates itself and
passes over into the infinite reference of being to itself;
it passes over, thirdly, into being-for-itself.

CHAPTER 1

Being

A. BEING

Being, pure being – without further determination.
In its indeterminate immediacy it is equal only to itself and
also not unequal with respect to another;
it has no difference within it, nor any outwardly.
If any determination or content were posited in it as distinct,
or if it were posited by this determination or
content as distinct from an other,
it would thereby fail to hold fast to its purity.
It is pure indeterminateness and emptiness.
There is nothing to be intuited in it,
if one can speak here of intuiting; or, it is
only this pure empty intuiting itself.
 Just as little is anything to be thought
in it, or, it is equally only this empty thinking.
Being, the indeterminate immediate is in fact nothing,
and neither more nor less than nothing.

B. NOTHING

Nothing, pure nothingness; it is simple equality with itself, complete empti-
ness, complete absence of determination and content; lack of all distinction
within. – In so far as mention can be made here of intuiting and thinking,
it makes a difference whether something or nothing is being intuited or
thought. To intuit or to think nothing has therefore a meaning; the two
are distinguished and so nothing is (concretely exists) in our intuiting or
thinking; or rather it is the empty intuiting and thinking itself, like pure
being. – Nothing is therefore the same determination or rather absence of
determination, and thus altogether the same as what pure being is.

C. BECOMING

1. Unity of being and nothing

Pure being and pure nothing are therefore the same. The truth is neither
being nor nothing, but rather that being has passed over into nothing and
nothing into being – “has passed over,” not passes over. But the truth is
just as much that they are not without distinction; it is rather that they are
not the same, that they are absolutely distinct yet equally unseparated and
inseparable, and that each immediately vanishes in its opposite. Their truth
is therefore this movement of the immediate vanishing of the one into the
other: becoming, a movement in which the two are distinguished, but by a
distinction which has just as immediately dissolved itself.

2. The moments of becoming

Becoming is the unseparatedness of being and nothing, not the unity that
abstracts from being and nothing; as the unity of being and nothing it is
rather this determinate unity, or one in which being and nothing equally
are. However, inasmuch as being and nothing are each unseparated from
its other, each is not. In this unity, therefore, they are, but as vanishing, only
as sublated. They sink from their initially represented self-subsistence into
moments which are still distinguished but at the same time sublated.
Grasped as thus distinguished, each is in their distinguishedness a unity
with the other. Becoming thus contains being and nothing as two such
unities, each of which is itself unity of being and nothing; the one is being
as immediate and as reference to nothing; the other is nothing as immediate
and as reference to being; in these unities the determinations are of unequal
value.
Becoming is in this way doubly determined. In one determination,
nothing is the immediate, that is, the determination begins with nothing
and this refers to being; that is to say, it passes over into it. In the other
determination, being is the immediate, that is, the determination begins
with being and this passes over into nothing – coming-to-be and ceasing-
to-be.
Both are the same, becoming, and even as directions that are so different
they interpenetrate and paralyze each other. The one is ceasing-to-be; being
passes over into nothing, but nothing is just as much the opposite of
itself, the passing-over into being, coming-to-be. This coming-to-be is the
other direction; nothing goes over into being, but being equally sublates
itself and is rather the passing-over into nothing; it is ceasing-to-be. –
They do not sublate themselves reciprocally – the one sublating the other
externally – but each rather sublates itself in itself and is within it the
opposite of itself.

3. Sublation of becoming

The equilibrium in which coming-to-be and ceasing-to-be are poised is in
the first place becoming itself. But this becoming equally collects itself in
quiescent unity. Being and nothing are in it only as vanishing; becoming
itself, however, is only by virtue of their being distinguished. Their vanish-
ing is therefore the vanishing of becoming, or the vanishing of the vanishing
itself. Becoming is a ceaseless unrest that collapses into a quiescent result.
This can also be expressed thus: becoming is the vanishing of being
into nothing, and of nothing into being, and the vanishing of being and
nothing in general; but at the same time it rests on their being distinct. It
therefore contradicts itself in itself, because what it unites within itself is
self-opposed; but such a union destroys itself.
This result is a vanishedness, but it is not nothing; as such, it would be
only a relapse into one of the already sublated determinations and not the
result of nothing and of being. It is the unity of being and nothing that has
become quiescent simplicity. But this quiescent simplicity is being, yet no
longer for itself but as determination of the whole.
Becoming, as transition into the unity of being and nothing, a unity
which is as existent or has the shape of the one-sided immediate unity of
these moments, is existence.

CHAPTER 2

Existence

Existence is determinate being; its determinateness is existent determinate-
ness, quality. Through its quality, something is opposed to an other; it is
alterable and finite, negatively determined not only towards an other, but
absolutely within it. This negation in it, in contrast at first with the finite
something, is the infinite; the abstract opposition in which these determina-
tions appear resolves itself into oppositionless infinity, into being-for-itself.
The treatment of existence is therefore in three divisions:

A. existence as such
B. something and other, finitude
C. qualitative infinity.

a. existence as such

In existence (a) as such, its determinateness is first (b) to be distinguished as
quality. The latter, however, is to be taken in both the two determinations of
existence as reality and negation. In these determinacies, however, existence
is equally reflected into itself, and, as so reflected, it is posited as (c)
something, an existent.
a. Existence in general
Existence proceeds from becoming. It is the simple oneness of being and
nothing. On account of this simplicity, it has the form of an immediate. Its
mediation, the becoming, lies behind it; it has sublated itself, and existence
therefore appears as a first from which the forward move is made. It is
at first in the one-sided determination of being; the other determination
which it contains, nothing, will likewise come up in it, in contrast to the
first.
It is not mere being but existence, or Dasein [in German]; according to
its [German] etymology, it is being (Sein) in a certain place (da). But the
representation of space does not belong here. As it follows upon becoming,
existence is in general being with a non-being, so that this non-being is taken
up into simple unity with being. Non-being thus taken up into being with
the result that the concrete whole is in the form of being, of immediacy,
constitutes determinateness as such.
The whole is likewise in the form or determinateness of being, since in
becoming being has likewise shown itself to be only a moment – something
sublated, negatively determined. It is such, however, for us, in our reflection;
not yet as posited in it. What is posited, however, is the determinateness as
such of existence, as is also expressed by the da (or “there”) of the Dasein. –
The two are always to be clearly distinguished. Only that which is posited in
a concept belongs in the course of the elaboration of the latter to its content.
Any determinateness not yet posited in the concept itself belongs instead to
our reflection, whether this reflection is directed to the nature of the concept
itself or is a matter of external comparison. To remark on a determinateness
of this last kind can only be for the clarification or anticipation of the whole
that will transpire in the course of the development itself. That the whole,
the unity of being and nothing, is in the one-sided determinateness of being
is an external reflection; but in negation, in something and other, and so
forth, it will become posited. – It was necessary here to call attention to
the distinction just given; but to comment on all that reflection can allow
itself, to give an account of it, would lead to a long-winded anticipation
of what must transpire in the fact itself. Although such reflections may
serve to facilitate a general overview and thus facilitate understanding,
they also bring the disadvantage of being seen as unjustified assertions,
unjustified grounds and foundations, of what is to follow. They should
be taken for no more than what they are supposed to be and should be
distinguished from what constitutes a moment in the advance of the fact
itself.
Existence corresponds to being in the preceding sphere. But being is the
indeterminate; there are no determinations that therefore transpire in it.
But existence is determinate being, something concrete; consequently, sev-
eral determinations, several distinct relations of its moments, immediately
emerge in it.

b. Quality

On account of the immediacy with which being and nothing are one in
existence, neither oversteps the other; to the extent that existence is existent,
to that extent it is non-being; it is determined. Being is not the universal,
determinateness not the particular. 42 Determinateness has yet to detach itself
from being; nor will it ever detach itself from it, since the now underlying
truth is the unity of non-being with being; all further determinations will
transpire on this basis. But the connection which determinateness now has
with being is one of the immediate unity of the two, so that as yet no
differentiation between the two is posited.
Determinateness thus isolated by itself, as existent determinateness, is
quality – something totally simple, immediate. Determinateness in general
is the more universal which, further determined, can be something quan-
titative as well. On account of this simplicity, there is nothing further to
say about quality as such.
Existence, however, in which nothing and being are equally contained,
is itself the measure of the one-sidedness of quality as an only immedi-
ate or existent determinateness. Quality is equally to be posited in the
determination of nothing, and the result is that the immediate or existent
determinateness is posited as distinct, reflected, and the nothing, as thus
the determinate element of determinateness, will equally be something
reflected, a negation. Quality, in the distinct value of existent, is reality;
when affected by a negating, it is negation in general, still a quality but one
that counts as a lack and is further determined as limit, restriction.
Both are an existence, but in reality, as quality with the accent on being
an existent, that it is determinateness and hence also negation is concealed;
reality only has, therefore, the value of something positive from which
negating, restriction, lack, are excluded. Negation, for its part, taken as
mere lack, would be what nothing is; but it is an existence, a quality, only
determined with a non-being.

c. Something

In existence its determinateness has been distinguished as quality; in this
quality as something existing, the distinction exists – the distinction of
reality and negation. Now though these distinctions are present in existence,
they are just as much null and sublated. Reality itself contains negation; it
is existence, not indeterminate or abstract being. Negation is for its part
equally existence, not the supposed abstract nothing but posited here as it
is in itself, as existent, as belonging to existence. Thus quality is in general
unseparated from existence, and the latter is only determinate, qualitative
being.
This sublating of the distinction is more than the mere retraction and
external re-omission of it, or a simple return to the simple beginning, to
existence as such. The distinction cannot be left out, for it is. Therefore,
what de facto is at hand is this: existence in general, distinction in it, and the
sublation of this distinction; the existence, not void of distinctions as at the
beginning, but as again self-equal through the sublation of the distinction;
the simplicity of existence mediated through this sublation. This state of
sublation of the distinction is existence’s own determinateness; existence is
thus being-in-itself; it is existent, something.
Something is the first negation of negation, as simple existent self-
reference. Existence, life, thought, and so forth, essentially take on the
determination of an existent being, a living thing, a thinking mind (“I”),
and so forth. This determination is of the highest importance if we do not
wish to halt at existence, life, thought, and so forth, as generalities – also
not at Godhood (instead of God). In common representation, something
rightly carries the connotation of a real thing. Yet it still is a very superficial
determination, just as reality and negation, existence and its determinate-
ness, though no longer the empty being and nothing, still are quite abstract
determinations. For this reason they also are the most common expressions,
and a reflection that is still philosophically unschooled uses them the most;
it casts its distinctions in them, fancying that in them it has something
really well and firmly determined. – As something, the negative of the neg-
ative is only the beginning of the subject – its in-itselfness is still quite
indeterminate. It determines itself further on, at first as existent-for-itself
and so on, until it finally obtains in the concept the intensity of the sub-
ject. At the base of all these determinations there lies the negative unity
with itself. In all this, however, care must be taken to distinguish the first
negation, negation as negation in general, from the second negation, the
negation of negation which is concrete, absolute negativity, just as the first
is on the contrary only abstract negativity.
Something is an existent as the negation of negation, for such a negation
is the restoration of the simple reference to itself – but the something is
thereby equally the mediation of itself with itself. Present in the simplicity
of something, and then with greater determinateness in being-for-itself, in
the subject, and so forth, this mediation of itself with itself is also already
present in becoming, but only as totally abstract mediation; mediation
with itself is posited in the something in so far as the latter is determined as
a simple identity. – Attention can be drawn to the presence of mediation
in general, as against the principle of the alleged bare immediacy of a
knowledge from which mediation should be excluded. But there is no
further need to draw particular attention to the moment of mediation,
since it is to be found everywhere and on all sides, in every concept.
This mediation with itself which something is in itself, when taken only
as the negation of negation, has no concrete determinations for its sides;
thus it collapses into the simple unity which is being. Something is, and
is therefore also an existent. Further, it is in itself also becoming, but a
becoming that no longer has only being and nothing for its moments.
One of these moments, being, is now existence and further an existent.
The other moment is equally an existent, but determined as the negative of
something – an other. As becoming, something is a transition, the moments
of which are themselves something, and for that reason it is an alteration – a
becoming that has already become concrete. – At first, however, something
alters only in its concept; it is not yet posited in this way, as mediated and
mediating, but at first only as maintaining itself simply in its reference to
itself; and its negative is posited as equally qualitative, as only an other in
general.

b. finitude

(a) Something and other: at first they are indifferent to one another; an
other is also an immediate existent, a something; the negation thus falls
outside both. Something is in itself in contrast to its being-for-other.
But the determinateness belongs also to its in-itself, and
(b) the determination of this in-itself in turn passes over into constitution,
and this latter, as identical with determination, constitutes the imma-
nent and at the same time negated being-for-another, the limit of
something which
(c) is the immanent determination of the something itself, and the some-
thing thus is the finite.
In the first division where existence in general was considered, this exis-
tence had, as at first taken up, the determination of an existent. The
moments of its development, quality and something, are therefore of
equally affirmative determination. The present division, on the contrary,
develops the negative determination which is present in existence and was
there from the start only as negation in general. It was then the first nega-
tion but has now been determined to the point of the being-in-itself of the
something, the point of the negation of negation.
a. Something and an other
1. Something and other are, first, both existents or something.
Second, each is equally an other. It is indifferent which is named first,
and just for this reason it is named something (in Latin, when they occur
in a proposition, both are aliud, or “the one, the other,” alius alium; in the
case of an alternating relation, the analogous expression is alter alterum).
If of two beings we call the one A and the other B, the B is the one which
is first determined as other. But the A is just as much the other of the
B. Both are other in the same way. “This” serves to fix the distinction
and the something which is to be taken in the affirmative sense. But
“this” also expresses the fact that the distinction, and the privileging of one
something, is a subjective designation that falls outside the something itself.
The whole determinateness falls on the side of this external pointing; also
the expression “this” contains no distinctions; each and every something is
just as good a “this” as any other. By “this” we mean to express something
completely determinate, overlooking the fact that language, as a work of
the understanding, only expresses the universal, albeit naming it as a single
object. But an individual name is something meaningless in the sense that
it does not express a universal. It appears as something merely posited and
arbitrary for the same reason that proper names can also be arbitrarily
picked, arbitrarily given as well as arbitrarily altered. 52
Otherness thus appears as a determination alien to the existence thus
pointed at, or the other existence as outside this one existence, partly
because the one existence is determined as other only by being compared
by a Third, and partly because it is so determined only on account of the
other which is outside it, but is not an other for itself. At the same time,
as has been remarked, even for ordinary thinking every existence equally
determines itself as an other existence, so that there is no existence that
remains determined simply as an existence, none which is not outside an
existence and therefore is not itself an other.
Both are determined as something as well as other : thus they are the
same and there is as yet no distinction present in them. But this sameness of
determinations, too, falls only within external reflection, in the comparison
of the two; but the other, as posited at first, though an other with reference
to something, is other also for itself apart from the something.
Third, the other is therefore to be taken in isolation, with reference to
itself, has to be taken abstractly as the other, the t¼ teron of Plato who
opposes it to the one as a moment of totality, and in this way ascribes
to the other a nature of its own. Thus the other, taken solely as such, is
not the other of something, but is the other within, that is, the other of
itself. – Such an other, which is the other by its own determination, is
physical nature; nature is the other of spirit; this, its determination, is at first
a mere relativity expressing not a quality of nature itself but only a reference
external to it. But since spirit is the true something, and hence nature is
what it is within only in contrast to spirit, taken for itself the quality of
nature is just this, to be the other within, that which-exists-outside-itself (in
the determinations of space, time, matter).
The other which is such for itself is the other within it, hence the other
of itself and so the other of the other – therefore, the absolutely unequal in
itself, that which negates itself, alters itself. But it equally remains identical
with itself, for that into which it alters is the other, and this other has no
additional determination; but that which alters itself is not determined in
any other way than in this, to be an other; in going over to this other, it only
unites with itself. It is thus posited as reflected into itself with sublation of
the otherness, a self-identical something from which the otherness, which is
at the same time a moment of it, is therefore distinct, itself not appertaining
to it as something.

2. The something preserves itself in its non-being; it is essentially one
with it, and essentially not one with it. It therefore stands in reference to
an otherness without being just this otherness. The otherness is at once
contained in it and yet separated from it; it is being-for-other.
Existence as such is an immediate, bare of references; or, it is in the
determination of being. However, as including non-being within itself,
existence is determinate being, being negated within itself, and then in the
first instance an other – but, since in being negated it preserves itself at the
same time, it is only being-for-other.
It preserves itself in its non-being and is being; not, however, being
in general but being with reference to itself in contrast to its reference
to the other, as self-equality in contrast to its inequality. Such a being is
being-in-itself.

Being-for-other and being-in-itself constitute the two moments of some-
thing. There are here two pairs of determinations: (1) something and
other; (2) being-for-other and being-in-itself. The former contain the non-
connectedness of their determinateness; something and other fall apart.
But their truth is their connection; being-for-other and being-in-itself are
therefore the same determinations posited as moments of one and the same
unity, as determinations which are connections and which, in their unity,
remain in the unity of existence. Each thus itself contains within it, at the
same time, also the moment diverse from it.
Being and nothing in their unity, which is existence, are no longer
being and nothing (these they are only outside their unity); so in their
restless unity, in becoming, they are coming-to-be and ceasing-to-be. – In
the something, being is being-in-itself. Now, as self-reference, self-equality,
being is no longer immediately, but is self-reference only as the non-being
of otherness (as existence reflected into itself ). – The same goes for non-
being: as the moment of something in this unity of being and non-being:
it is not non-existence in general but is the other, and more determinedly –
according as being is at the same time distinguished from it – it is reference
to its non-existence, being-for-other.
Hence being-in-itself is, first, negative reference to non-existence; it has
otherness outside it and is opposed to it; in so far as something is in itself,
it is withdrawn from being-other and being-for-other. But, second, it has
non-being also right in it; for it is itself the non-being of being-for-other.
But being-for-other is, first, the negation of the simple reference of being
to itself which, in the first place, is supposed to be existence and something;
in so far as something is in an other or for an other, it lacks a being of
its own. But, second, it is not non-existence as pure nothing; it is non-
existence that points to being-in-itself as its being reflected into itself, just
as conversely the being-in-itself points to being-for-other.
3. Both moments are determinations of one and the same, namely of
something. Something is in-itself in so far as it has returned from the
being-for-other back to itself. But something has also a determination or
circumstance, whether in itself (here the accent is on the in) or in it; in so
far as this circumstance is in it externally, it is a being-for-other.
This leads to a further determination. Being-in-itself and being-for-other
are different at first. But that something also has in it what it is in itself and
conversely is in itself also what it is as being-for-other – this is the identity of
being-in-itself and being-for-other, in accordance with the determination
that the something is itself one and the same something of both moments,
and these are in it, therefore, undivided. – This identity already occurs
formally in the sphere of existence, but more explicitly in the treatment of
essence and later of the relations of interiority and externality, and in the
most determinate form in the treatment of the idea, as the unity of concept
and actuality. – Opinion has it that with the in-itself something lofty is
being said, as with the inner; but what something is only in itself, is also only
in it; in-itself is a merely abstract, and hence itself external determination.
The expressions: there is nothing in it, or there is something in it, imply,
though somewhat obscurely, that what is in a thing also pertains to its
in-itselfness, to its inner, true worth.
It may be observed that here we have the meaning of the thing-in-itself.
It is a very simple abstraction, though it was for a while a very important
determination, something sophisticated, as it were, just as the proposition
that we know nothing of what things are in themselves was a much valued
piece of wisdom. – Things are called “in-themselves” in so far as abstraction
is made from all being-for-other, which really means, in so far as they are
thought without all determination, as nothing. In this sense, of course, it
is impossible to know what the thing-in-itself is. For the question “what?”
calls for determinations to be produced; but since the things of which the
determinations are called for are at the same time presumed to be things-in-
themselves, which means precisely without determination, the impossibility
of an answer is thoughtlessly implanted in the question, or else a senseless
answer is given. – The thing-in-itself is the same as that absolute of which
nothing is known except that in it all is one. What there is in these things-
in-themselves is therefore very well known; they are as such nothing but
empty abstractions void of truth. What, however, the thing-in-itself in
truth is, what there basically is in it, of this the Logic is the exposition.
But in this Logic something better is understood by the in-itself than an
abstraction, namely, what something is in its concept; but this concept is
in itself concrete: as concept, in principle conceptually graspable; and, as
determined and as the connected whole of its determinations, inherently
cognizable.
Being-in-itself has at first the being-for-other as a moment standing
over against it. But positedness also comes to be positioned over against
it, and, although in this expression being-for-other is also included, the
expression still contains the determination of the bending back, which
has already occurred, of that which is not in itself into that wherein it is
positive, and this is its being-in-itself. Being-in-itself is normally to be taken
as an abstract way of expressing the concept; positing, strictly speaking, first
occurs in the sphere of essence, of objective reflection; the ground posits
that which is grounded through it; more strongly, the cause produces an
effect, an existence whose subsistence is immediately negated and which
carries the meaning that it has its substance, its being, in an other. In the
sphere of being, existence only emerges out of becoming. Or again, with
the something an other is posited; with the finite, an infinite; but the finite
does not bring forth the infinite, does not posit it. In the sphere of being,
the self-determining of the concept is at first only in itself or implicit, and
for that reason it is called a transition or passing over. And the reflecting
determinations of being, such as something and other, or finite and infinite,
although they essentially point to one another, or are as being-for-other,
also stand on their own qualitatively; the other exists; the finite, like the
infinite, is equally to be regarded as an immediate existent that stands firm
on its own; the meaning of each appears complete even without its other.
The positive and the negative, on the contrary, cause and effect, however
much they are taken in isolation, have at the same time no meaning eachE
without the other; their reflective shining in each other, the shine in each of
its other, is present right in them. – In the different cycles of determination
and especially in the progress of the exposition, or, more precisely, in the
progress of the concept in the exposition of itself, it is of capital concern
always to clearly distinguish what still is in itself or implicitly and what is
posited, how determinations are in the concept and how they are as posited
or as existing-for-other. This is a distinction that belongs only to the
dialectical development and one unknown to metaphysical philosophizing
(to which the critical also belongs); the definitions of metaphysics, like
its presuppositions, distinctions, and conclusions, are meant to assert and
produce only the existent and that, too, as existent-in-itself.
In the unity of the something with itself, being-for-other is identical with
its in-itself; the being-for-other is thus in the something. The determinate-
ness thus reflected into itself is therefore again a simple existent and hence
again a quality – determination.

b. Determination, constitution, and limit

The in-itself, in which the something is reflected into itself from its being-
for-other, no longer is an abstract in-itself but, as the negation of its
being-for-other, is mediated through this latter, which is thus its moment.
It is not only the immediate identity of the something with itself, but
the identity by virtue of which the something also has present in it what
it is in itself; the being-for-other is present in it because the in-itself is the
sublation of it, is in itself from it; but, because it is still abstract, and therefore
essentially affected with negation, it is equally affected with being-for-other.
We have here not only quality and reality, existent determinateness, but
determinateness existent-in-itself; and the development consists in positing
such determinateness as thus immanently reflected.
1. The quality which in the simple something is an in-itself essentially in
unity with the something’s other moment, its being-in-it, can be named its
determination, provided that this word is distinguished, in a more precise
signification, from determinateness in general. Determination is affirmative
determinateness; it is the in-itself by which a something abides in its
existence while involved with an other that would determine it, by which
it preserves itself in its self-equality, holding on to it in its being-for-
other. Something fulfills its determination to the extent that the further
determinateness, which variously accrues to it in the measure of its being-
in-itself as it relates to an other, becomes its filling. Determination implies
that what something is in itself is also present in it.
The determination of the human being, its vocation, is rational thought:
thinking in general is his simple determinateness; by it the human being
is distinguished from the brute; he is thinking in himself, in so far as
this thinking is distinguished also from his being-for-other, from his own
natural and sensuous being that brings him in immediate association with
the other. But thinking is also in him; the human being is himself thinking,
he exists as thinking, thought is his concrete existence and actuality; and,
further, since thinking is in his existence and his existence is in his thinking,
thinking is concrete, must be taken as having content and filling; it is rational
thought and as such the determination of the human being. But even this
determination is again only in itself, as an ought, that is to say, it is, together
with the filling embodied in its in-itself, in the form of an in-itself in general
as against the existence which is not embodied in it but still lies outside
confronting it, immediate sensibility and nature.
2. The filling of the being-in-itself with determinateness is also dis-
tinct from the determinateness which is only being-for-other and remains
outside the determination. For in the sphere of the qualitative, the dis-
tinguished terms are left, in their sublated being, also with an immediate,
qualitative being contrasting them. That which the something has in it
thus separates itself and is from this side the external existence of the some-
thing and also its existence, but not as belonging to its being-in-itself. –
Determinateness is thus constitution.
Constituted in this or that way, the something is caught up in external
influences and in external relationships. This external connection on which
the constitution depends, and the being determined through an other,
appear as something accidental. But it is the quality of the something to
be given over to this externality and to have a constitution.
In so far as something alters, the alteration falls on the side of its con-
stitution; the latter is that in the something which becomes an other. The
something itself preserves itself in the alteration; the latter affects only this
unstable surface of the something’s otherness, not its determination.
Determination and constitution are thus distinct from each other; some-
thing, according to its determination, is indifferent to its constitution.
But that which the something has in it is the middle term of this syllo-
gism connecting the two, determination and constitution. Or, rather, the
being-in-the-something showed itself to fall apart into these two extremes.
The simple middle term is determinateness as such; its identity belongs to
determination just as well as to constitution. But determination passes over
into constitution on its own, and constitution into determination. This
is implied in what has been said. The connection, upon closer consider-
ation, is this: in so far as that which something is in itself is also in it,
the something is affected with being-for-other; determination is therefore
open, as such, to the relation with other. Determinateness is at the same
time moment, but it contains at the same time the qualitative distinction
of being different from being-in-itself, of being the negative of the some-
thing, another existence. This determinateness which thus holds the other
in itself, united with the being-in-itself, introduces otherness in the latter or
in determination, and determination is thereby reduced to constitution. –
Conversely, the being-for-other, isolated as constitution and posited on its
own, is in it the same as what the other as such is, the other in it, that is,
the other of itself; but it consequently is self-referring existence, thus being-
in-itself with a determinateness, therefore determination. – Consequently,
inasmuch as the two are also to be held apart, constitution, which appears
to be grounded in something external, in an other in general, also depends
on determination, and the determining from outside is at the same time
determined by the something’s own immanent determination. And fur-
ther, constitution belongs to that which something is in itself: something
alters along with its constitution.
This altering of something is no longer the first alteration of something
merely in accordance with its being-for-other. The first was an alteration
only implicitly present, one that belonged to the inner concept; now the
alteration is also posited in the something. – The something itself is further
determined, and negation is posited as immanent to it, as its developed
being-in-itself.
The transition of determination and constitution into each other is at
first the sublation of their distinction, and existence or something in general
is thereby posited; moreover, since this something in general results from a
distinction that also includes qualitative otherness within it, there are two
somethings. But these are, with respect to each other, not just others in
general, so that this negation would still be abstract and would occur only
in the comparison of the two; rather the negation now is immanent to the
somethings. As existing, they are indifferent to each other, but this, their
affirmation, is no longer immediate: each refers itself to itself through the
intermediary of the sublation of the otherness which in determination is
reflected into the in-itselfness.
Something behaves in this way in relation to the other through itself;
since otherness is posited in it as its own moment, its in-itselfness holds
negation in itself, and it now has its affirmative existence through its
intermediary alone. But the other is also qualitatively distinguished from
this affirmative existence and is thus posited outside the something. The
negation of its other is only the quality of the something, for it is in
this sublation of its other that it is something. The other, for its part,
truly confronts an existence only with this sublation; it confronts the first
something only externally, or, since the two are in fact inherently joined
together, that is, according to their concept, their connectedness consists
in this, that existence has passed over into otherness, something into other;
that something is just as much an other as the other is. Now in so far as the
in-itselfness is the non-being of the otherness that is contained in it but
is at the same time also distinct as existent, something is itself negation,
the ceasing to be of an other in it; it is posited as behaving negatively in
relation to the other and in so doing preserving itself. This other, the in-
itselfness of the something as negation of the negation, is the something’s
being-in-itself, and this sublation is as simple negation at the same time in
it, namely, as its negation of the other something external to it. It is one
determinateness of the two somethings that, on the one hand, as nega-
tion of the negation, is identical with the in-itselfness of the somethings,
and also, on the other hand, since these negations are to each other as
other somethings, joins them together of their own accord and, since each
negation negates the other, equally separates them. This determinateness is
limit.
3. Being-for-other is indeterminate, affirmative association of something
with its other; in limit the non-being-for-other is emphasized, the quali-
tative negation of the other, which is thereby kept out of the something
that is reflected into itself. We must see the development of this concept –
a development that will rather look like confusion and contradiction.
Contradiction immediately raises its head because limit, as an internally
reflected negation of something, ideally holds in it the moments of some-
thing and other, and these, as distinct moments, are at the same time
posited in the sphere of existence as really, qualitatively, distinct.
a. Something is therefore immediate, self-referring existence and at first
it has a limit with respect to an other; limit is the non-being of the other,
not of the something itself; in limit, something marks the boundary of its
other. – But other is itself a something in general. The limit that something
has with respect to an other is, therefore, also the limit of the other as a
something; it is the limit of this something in virtue of which the something
holds the first something as its other away from itself, or is a non-being of
that something. The limit is thus not only the non-being of the other, but of
the one something just as of the other, and consequently of the something
in general.
But the limit is equally, essentially, the non-being of the other; thus,
through its limit, something at the same time is. In limiting, something
is of course thereby reduced to being limited itself; but, as the ceasing of
the other in it, its limit is at the same time itself only the being of the
something; this something is what it is by virtue of it, has its quality in it. –
This relation is the external appearance of the fact that limit is simple
negation or the first negation, whereas the other is, at the same time, the
negation of the negation, the in-itselfness of the something.
Something, as an immediate existence, is therefore the limit with respect
to another something; but it has this limit in it and is something through
the mediation of that limit, which is just as much its non-being. The limit
is the mediation in virtue of which something and other each both is and
is not.
b. Now in so far as something in its limit both is and is not, and these
moments are an immediate, qualitative distinction, the non-existence and
the existence of the something fall outside each other. Something has its
existence outside its limit (or, as representation would also have it, inside it);
in the same way the other, too, since it is something, has it outside it. The
limit is the middle point between the two at which they leave off. They have
existence beyond each other, beyond their limit; the limit, as the non-being
of each, is the other of both.
– It is in accordance with this difference of the something from its limit
that the line appears as line outside its limit, the point; the plane as plane
outside the line; the solid as solid only outside its limiting plane. – This
is the aspect of limit that first occurs to figurative representation (the self-
external-being of the concept) and is also most commonly assumed in the
context of spatial objects.
g. But further, something as it is outside the limit, as the unlimited
something, is only existence in general. As such, it is not distinguished
from its other; it is only existence and, therefore, it and its other have the
same determination; each is only something in general or each is other;
and so both are the same. But this, their at first immediate existence, is now
posited in them as limit: in it both are what they are, distinct from each
other. But it is also equally their common distinguishedness, the unity and
the distinguishedness of both, just like existence. This double identity of
the two, existence and limit, contains this: that something has existence
only in limit, and that, since limit and immediate existence are each at the
same time the negative of each other, the something, which is now only
in its limit, equally separates itself from itself, points beyond itself to its
non-being and declares it to be its being, and so it passes over into it. To
apply this to the preceding example, the one determination is this: that
something is what it is only in its limit. Therefore, the point is the limit of
line, not because the latter just ceases at the point and has existence outside
it; the line is the limit of plane, not because the plane just ceases at it; and
the same goes for the plane as the limit of solid. Rather, at the point the line
also begins; the point is its absolute beginning, and if the line is represented
as unlimited on both its two sides, or, as is said, as extended to infinity, the
point still constitutes its element, just as the line constitutes the element
of the plane, and the plane that of the solid. These limits are the principle
of that which they delimit; just as one, for instance, is as hundredth the
limit, but also the element, of the whole hundred.
The other determination is the unrest of the something in its limit in
which it is immanent, the contradiction that propels it beyond itself. Thus
the point is this dialectic of itself becoming line; the line, the dialectic of
becoming plane; the plane, of becoming total space. A second definition
is given of line, plane, and whole space which has the line come to be
through the movement of the point; the plane through the movement of
the line, and so forth. This movement of the point, the line, and so forth, is
however viewed as something accidental, or as movement only in figurative
representation. In fact, however, this view is taken back by supposing that
the determinations from which the line, and so forth, originate are their
elements and principles, and these are, at the same time, nothing else but
their limits; the coming to be is not considered as accidental or only as
represented. That the point, the line, the plane, are per se self-contradictory
beginnings which on their own repel themselves from themselves, and
consequently that the point passes over from itself into the line through
its concept, moves in itself and makes the line come to be, and so on – all
this lies in the concept of the limit which is immanent in the something.
The application itself, however, belongs to the treatment of space; as an
indication of it here, we can say that the point is the totally abstract limit,
but in a determinate existence; this existence is still taken in total abstraction,
it is the so-called absolute, that is, abstract space, the absolutely continuous
being-outside-one-another. Inasmuch as the limit is not abstract negation,
but is rather in this existence, inasmuch as it is spatial determinateness, the
point is spatial, is the contradiction of abstract negation and continuity
and is, for that reason, the transition as it occurs and has already occurred
into the line, and so forth. And so there is no point, just as there is no line
or plane.
The something, posited with its immanent limit as the contradiction of
itself by virtue of which it is directed and driven out and beyond itself, is
the finite.

c. Finitude

Existence is determinate. Something has a quality, and in this quality it is
not only determined but delimited; its quality is its limit and, affected by it,
something remains affirmative, quiescent existence. But, so developed that
the opposition of its existence and of the negation as the limit immanent to
this existence is the very in-itselfness of the something, and this is thus only
becoming in it, this negation constitutes the finitude of the something.
When we say of things that they are finite, we understand by this that
they not only have a determinateness, that their quality is not only reality
and existent determination, that they are not merely limited and as such
still have existence outside their limit, but rather that non-being consti-
tutes their nature, their being. Finite things are, but in their reference to
themselves they refer to themselves negatively – in this very self-reference
they propel themselves beyond themselves, beyond their being. They are,
but the truth of this being is (as in Latin) their finis, their end. 54 The finite
does not just alter, as the something in general does, but perishes, and its
perishing is not just a mere possibility, as if it might be without perish-
ing. Rather, the being as such of finite things is to have the germ of this
transgression 55 in their in-itselfness: the hour of their birth is the hour of
their death.

a. The immediacy of finitude

The thought of the finitude of things brings this mournful note with it
because finitude is qualitative negation driven to the extreme, and in the
simplicity of such a determination there is no longer left to things an
affirmative being distinct from their determination as things destined to
ruin. 56 Because of this qualitative simplicity of negation that has returned
to the abstract opposition of nothing and perishing to being, finitude
is the most obstinate of the categories of the understanding; negation in
general, constitution, limit, are compatible with their other, with existence;
even the abstract nothing, by itself, is given up as an abstraction; but
finitude is negation fixed in itself and, as such, stands in stark contrast to its
affirmative. The finite thus does indeed let itself be submitted to flux; this
is precisely what it is, that it should come to an end, and this end is its only
determination. Its refusal is rather to let itself be brought affirmatively to its
affirmative, the infinite, to be associated with it; it is therefore inseparably
posited with its nothing, and thereby cut off from any reconciliation with
its other, the affirmative. The determination of finite things does not go
past their end. The understanding persists in this sorrow of finitude, for
it makes non-being the determination of things and, at the same time,
this non-being imperishable and absolute. Their transitoriness would only
pass away in their other, in the affirmative; their finitude would then be
severed from them; but this finitude is their unalterable quality, that is,
their quality which does not pass over into their other, that is, not into the
affirmative; and so finitude is eternal.
This is a very important consideration. But that the finite is absolute is
certainly not a standpoint that any philosophy or outlook, or the under-
standing, would want to endorse. The opposite is rather expressly present
in the assertion of finitude: the finite is the restricted, the perishable, the
finite is only the finite, not the imperishable; all this is immediately part
and parcel of its determination and expression. But all depends on whether
in one’s view of finitude its being is insisted on, and the transitoriness thus
persists, or whether the transitoriness and the perishing perish. The fact is
that this perishing of the perishing does not happen on precisely the view
that would make the perishing the final end of the finite. The official claim
is that the finite is incompatible with the infinite and cannot be united
with it; that the finite is absolutely opposed to the infinite. Being, absolute
being, is ascribed to the infinite. The finite remains held fast over against it
as its negative; incapable of union with the infinite, it remains absolute on
its own side; from the affirmative, from the infinite, it would receive affir-
mation and thus it would perish; but a union with the infinite is precisely
what is declared impossible. If the finite were not to persist over against
the infinite but were to perish, its perishing, as just said, would then be
the last of it – not its affirmative, which would be only a perishing of the
perishing. However, if it is not to perish into the affirmative but its end
is rather to be grasped as a nothing, then we are back at that first, abstract
nothing that itself has long since passed away.
With this nothing, however, which is supposed to be only nothing
but to which a reflective existence is nevertheless granted in thought, in
representation or in speech, the same contradiction occurs as we have
just indicated in connection with the finite, except that in the nothing
it just occurs but in the finite it is instead expressed. In the one case, the
contradiction appears as subjective; in the other, the finite is said to stand
in perpetual opposition to the infinite, in itself to be null, and to be as null
in itself. This is now to be brought to consciousness. The development
of the finite will show that, expressly as this contradiction, it collapses
internally, but that, in this collapse, it actually resolves the contradiction; it
will show that the finite is not just perishable, and that it perishes, but that
the perishing, the nothing, is rather not the last of it; that the perishing
rather perishes.

b. Restriction and the ought

This contradiction is indeed abstractly present by the very fact that the
something is finite, or that the finite is. But something or being is no longer
posited abstractly but reflected into itself, and developed as being-in-itself
that has determination and constitution in it, or, more determinedly still,
in such a way that it has a limit within it; and this limit, as constituting
what is immanent to the something and the quality of its being-in-itself,
is finitude. It is to be seen what moments are contained in this concept of
the finite something.
Determination and constitution arose as sides for external reflection,
but determination already contained otherness as belonging to the in-itself
of something. On the one side, the externality of otherness is within the
something’s own inwardness; on the other side, it remains as otherness
distinguished from it; it is still externality as such, but in the something.
But further, since otherness is determined as limit, itself as negation of
the negation, the otherness immanent to the something is posited as the
connection of the two sides, and the unity of the something with itself (to
which both determination and constitution belong) is its reference turned
back upon itself, the reference to it of its implicitly existing determination
that in it negates its immanent limit. The self-identical in-itself thus refers
itself to itself as to its own non-being, but as negation of the negation, as
negating that which at the same time retains existence in it, for it is the
quality of its in-itselfness. Something’s own limit, thus posited by it as a
negative which is at the same time essential, is not only limit as such, but
restriction. But restriction is not alone in being posited as negative; the
negation cuts two ways, for that which it posits as negated is limit, and
limit is in general what is common to something and other, and is also
the determinateness of the in-itself of determination as such. This in-itself,
consequently, as negative reference to its limit (which is also distinguished
from it), as negative reference to itself as restriction, is the ought.
In order for the limit that is in every something to be a restriction, the
something must at the same time transcend it in itself – must refer to it
from within as to a non-existent. The existence of something lies quietly
indifferent, as it were, alongside its limit. But the something transcends
its limit only in so far as it is the sublatedness of the limit, the negative
in-itselfness over against it. And inasmuch as the limit is as restriction in
the determination itself, the something thereby transcends itself.
The ought therefore contains the double determination: once, as a deter-
mination which has an in-itselfness over against negation; and again, as a
non-being which, as restriction, is distinguished from the determination
but is at the same time itself a determination existing in itself. 57
The finite has thus determined itself as connecting determination and
limit; in this connection, the determination is the ought and the limit is the
restriction. Thus the two are both moments of the finite, and therefore both
themselves finite, the ought as well as the restriction. But only restriction is
posited as the finite; the ought is restricted only in itself, and therefore only
for us. It is restricted by virtue of its reference to the limit already immanent
within it, though this restriction in it is shrouded in in-itselfness, for
according to its determinate being, that is, according to its determinateness
in contrast to restriction, it is posited as being-in-itself.
What ought to be is, and at the same time is not. If it were, it would
not be what merely ought to be. The ought has therefore a restriction
essentially. This restriction is not anything alien; that which only ought to
be is determination now posited as it is in fact, namely as at the same time
only a determinateness.
The being-in-itself of the something is thus reduced in its determination
to the ought because the very thing that constitutes the something’s in-
itselfness is, in one and the same respect, as non-being; or again, because
in the in-itselfness, in the negation of the negation, the said being-in-itself
is as one negation (what negates) a unity with the other, and this other, as
qualitatively other, is the limit by virtue of which that unity is as reference
to it. 58 The restriction of the finite is not anything external, but the finite’s
own determination is rather also its restriction; and this restriction is both
itself and the ought; it is that which is common to both, or rather that in
which the two are identical.
But further, as “ought” the finite transcends its restriction; the same
determinateness which is its negation is also sublated, and is thus its in-
itself; its limit is also not its limit.
As ought something is thus elevated above its restriction, but conversely
it has its restriction only as ought. The two are indivisible. Something has
a restriction in so far as it has negation in its determination, and the
determination is also the being sublated of the restriction.

g. Transition of the finite into the infinite

The ought contains restriction explicitly, for itself, and restriction contains
the ought. Their mutual connection is the finite itself, which contains them
both in its in-itself. These moments of its determination are qualitatively
opposed; restriction is determined as the negative of the ought, and the
ought equally as the negative of restriction. The finite is thus in itself the
contradiction of itself; it sublates itself, it goes away and ceases to be. 61
But this, its result, the negative as such, is (a) its very determination; for
it is the negative of the negative. So, in going away and ceasing to be,
the finite has not ceased; it has only become momentarily an other finite
which equally is, however, a going-away as a going-over into another finite,
and so forth to infinity. But, (b) if we consider this result more closely, in
its going-away and ceasing-to-be, in this negation of itself, the finite has
attained its being-in-itself; in it, it has rejoined itself. Each of its moments
contains precisely this result; the ought transcends the restriction, that is,
it transcends itself; but its beyond, or its other, is only restriction itself.
Restriction, for its part, immediately points beyond itself to its other, and
this is the ought; but this ought is the same diremption of in-itselfness and
determinateness as is restriction; it is the same thing; in going beyond itself,
restriction thus equally rejoins itself. This identity with itself, the negation
of negation, is affirmative being, is thus the other of the finite which is
supposed to have the first negation for its determinateness; this other is the
infinite.

c. infinity

The infinite in its simple concept can be regarded, first of all, as a fresh
definition of the absolute; as self-reference devoid of determination, it is
posited as being and becoming. The forms of existence have no place in the
series of determinations that can be regarded as definitions of the absolute,
since the forms of that sphere are immediately posited for themselves
only as determinacies, as finite in general. But the infinite is accepted
unqualifiedly as absolute, since it is explicitly determined as the negation
of the finite; the restrictedness – to which being and becoming would
somehow be susceptible even if they do not have it or exhibit it – is thereby
both explicitly referred to and denied in it.
But, in fact, by just this negation the infinite is not already free from
restrictedness and finitude. It is essential to distinguish the true concept
of infinity from bad infinity, the infinite of reason from the infinite of the
understanding. The latter is in fact a finitized infinite, and, as we shall now
discover, in wanting to maintain the infinite pure and distant from the
finite, the infinite is by that very fact only made finite.
The infinite
(a) in simple determination, is the affirmative as negation of the finite;
(b) but is thereby in alternating determination with the infinite, and is
abstract, one-sided infinite;
(c) is the self-sublation of this infinite and of the finite in one process. This
is the true infinite.
a. The infinite in general
The infinite is the negation of negation, the affirmative, being that has
reinstated itself out of restrictedness. The infinite is, in a more intense
sense than the first immediate being; it is the true being; the elevation
above restriction. At the mention of the infinite, soul and spirit light up,
for in the infinite the spirit is at home, and not only abstractly; rather, it
rises to itself, to the light of its thinking, its universality, its freedom.
What is first given with the concept of the infinite is this, that in its
being-in-itself existence is determined as finite and transcends restriction.
It is the very nature of the finite that it transcend itself, that it negate its
negation and become infinite. Consequently, the infinite does not stand
above the finite as something ready-made by itself, as if the finite stood fixed
outside or below it. Nor is it we only, as a subjective reason, who transcend
the finite into the infinite – as if, in saying that the infinite is a concept of
reason and that through reason we elevate ourselves above things temporal,
we did this without prejudice to the finite, without this elevation (which
remains external to the finite) affecting it. In so far as the finite itself is
being elevated to infinity, it is not at all an alien force that does this for it;
it is rather its nature to refer itself to itself as restriction (both restriction
as such and as ought) and to transcend this restriction, or rather, in this
self-reference, to have negated the restriction and gone above and beyond
it. It is not in the sublation of the finite in general that infinity in general
comes to be, but the finite is rather just this, that through its nature it
comes to be itself the infinite. Infinity is its affirmative determination, its
vocation, what it truly is in itself.
The finite has thus vanished into the infinite and what is, is only the
infinite.

b. Alternating determination of finite and infinite

The infinite is; in this immediacy it is at the same time the negation of
an other, of the finite. And so, as existent and at the same time as the
non-being of an other, it has fallen back into the category of the something,
of something determinate in general. More precisely: the infinite is the
existence reflected into itself which results from the mediating sublation of
determinateness in general and is consequently posited as existence distinct
from its determinateness; therefore, it has fallen back into the category of
something with a limit. In accordance with this determinateness, the finite
stands over against the infinite as real existence; they thus remain outside
each other, standing in qualitative mutual reference; the immediate being of
the infinite resurrects the being of its negation, of the finite again, which
seemed at first to have vanished into the infinite.
But the infinite and the finite are not in these referential categories only;
the two sides are further determined in addition to being as mere others
to each other. Namely, the finite is restriction posited as restriction; it is
existence posited with the determination that it passes over into what is its
in-itself and becomes infinite. Infinity is the nothing of the finite, the in-itself
that the latter ought to be, but it is this at the same time as reflected within
itself, as realized ought, as only affirmative self-referring being. In infinity
we have the satisfaction that all determinateness, alteration, all restriction
and the ought itself together with it, have vanished, are sublated, and the
nothing of the finite is posited. As this negation of the finite is the being-
in-itself determined which, as negation of negation, is in itself affirmative.
Yet this affirmation is qualitatively immediate self-reference, being; and,
because of this, the affirmative is led back to the category of being that has
the finite confronting it as an other; its negative nature is posited as existent
negation, and hence as first and immediate negation. The infinite is in this
way burdened with the opposition to the finite, and this finite, as an other,
remains a real existence even though in its being-in-itself, in the infinite,
it is at the same time posited as sublated; this infinite is that which is not
finite – a being in the determinateness of negation. Contrasted with the
finite, with the series of existent determinacies, of realities, the infinite is
indeterminate emptiness, the beyond of the finite, whose being-in-itself is
not in its existence (which is something determinate).
As thus posited over against the finite, the two connected by the quali-
tative mutual reference of others, the infinite is to be called the bad infinite,
the infinite of the understanding, for which it counts as the highest, the
absolute truth. The understanding believes that it has attained satisfaction
in the reconciliation of truth while it is in fact entangled in unreconciled,
unresolved, absolute contradictions. And it is these contradictions, into
which it falls on every side whenever it embarks on the application and
explication of these categories that belong to it, that must make it conscious
of the fact.
This contradiction is present in the very fact that the infinite remains
over against the finite, with the result that there are two determinacies.
There are two worlds, one infinite and one finite, and in their connection
the infinite is only the limit of the finite and thus only a determinate, itself
finite infinite.
This contradiction develops its content into more explicit forms. – The
finite is the real existence which persists as such even when it has gone
over into its non-being, the infinite. As we have seen, this infinite has
for its determinateness, over against the finite, only the first, immediate
negation, just as the finite, as negated, has over against this negation
only the meaning of an other and is, therefore, still a something. When,
therefore, the understanding, elevating itself above this finite world, rises
to what is the highest for it, to the infinite, the finite world remains for it
as something on this side here, and, thus posited only above the finite, the
infinite is separated from the finite and, for the same reason, the finite from
the infinite: each is placed in a different location, the finite as existence here,
and the infinite, although the being-in-itself of the finite, there as a beyond,
at a nebulous, inaccessible distance outside which there stands, enduring,
the finite.
As thus separated, they are just as much essentially connected with each
other, through the very negation that divides them. This negation connect-
ing them – these somethings reflected into themselves – is the common
limit of each over against the other; and that, too, in such a way that each
does not merely have this limit in it over against the other, but the negation
is rather the in-itselfness of each; each thus has for itself, in its separation
from the other, the limit in it. But the limit is the first negation; both are
thus limited, finite, in themselves. Yet, as each affirmatively refers itself to
itself, each is also the negation of its limit; each thus immediately repels
the negation from itself as its non-being, and, qualitatively severed from it,
posits it as an other being outside it: the finite posits its non-being as this
infinite, and the infinite likewise the finite. It is readily conceded that the
finite passes over into the infinite necessarily (that is, through its determi-
nation) and is thereby elevated to what is its in-itself, for while the finite
is indeed determined as subsistent existence, it is at the same time also a
null in itself and therefore destined to self-dissolution; whereas the infinite,
although burdened with negation and limit, is equally also determined as
the existent in-itself, so that this abstraction of self-referring affirmation is
what constitutes its determination, and hence finite existence is not present
in it. But it has been shown that the infinite itself attains affirmative being
only by the mediation of negation, as negation of negation, and that when
its affirmation thus attained is taken as just simple, qualitative being, the
negation contained in it is demoted to simple immediate negation and,
therefore, to determinateness and limit; and these, then, are excluded
from the infinite as contradicting its in-itself; they are posited as not
belonging to it but rather as opposed to its in-itself, as the finite. Since each
is in it and through its determination the positing of its other, the two are
inseparable. But this unity rests hidden in their qualitative otherness; it is
their inner unity, one that lies only at their base.
The manner of the appearance of this unity has thereby been defined.
The unity is posited in existence as a turning over or transition of the
finite into the infinite, and vice-versa; so that the infinite only emerges in
the finite, and the finite in the infinite, the other in the other; that is
to say, each arises in the other independently and immediately, and their
connection is only an external one.
The process of their transition has the following, detailed shape. We
have the finite passing over into the infinite. This passing over appears as
an external doing. In this emptiness beyond the finite, what arises? What is
there of positive in it? On account of the inseparability of the infinite and
the finite (or because this infinite, which stands apart, is itself restricted),
the limit arises. The infinite has vanished and the other, the finite, has
stepped in. But this stepping in of the finite appears as an event external
to the infinite, and the new limit as something that does not arise out of
the infinite itself but is likewise found given. And with this we are back
at the previous determination, which has been sublated in vain. This new
limit, however, is itself only something to be sublated or transcended. And
so there arises again the emptiness, the nothing, in which we find again the
said determination – and so forth to infinity.
We have before us the alternating determination of the finite and the
infinite; the finite is finite only with reference to the ought or the infinite,
and the infinite is only infinite with reference to the finite. The two are
inseparable and at the same time absolutely other with respect to each
other; each has in it the other of itself; each is thus the unity of itself and
its other, and, in its determinateness – not to be what itself and what its
other is – it is existence.
This alternating determination of self-negating and of negating the
negating is what passes as the progress to infinity, which is accepted in
so many shapes and applications as an unsurpassable ultimate at which
thought, having reached this “and so on to infinity,” has usually achieved
its end. – This progress breaks out wherever relative determinations are
pressed to the point of opposition, so that, though in inseparable unity,
each is nevertheless attributed an independent existence over against the
other. This progress is therefore the contradiction which is not resolved but
is rather always pronounced simply as present.
What we have before us is an abstract transcending which remains
incomplete because the transcending itself has not been transcended. Before
us we have the infinite; of course, this infinite is transcended, for another
limit is posited, but just because of that only a return is instead made back
to the finite. This bad infinite is in itself the same as the perpetual ought;
it is indeed the negation of the finite, but in truth it is unable to free
itself from it; the finite constantly resurfaces in it as its other, since this
infinite only is with reference to the finite, which is its other. The progress
to infinity is therefore only repetitious monotony, the one and the same
tedious alternation of this finite and infinite.
The infinity of the infinite progress remains burdened by the finite as
such, is thereby restricted, and is itself finite. In fact, however, it is thereby
posited as the unity of the finite and the infinite. Only, this unity is not
reflected upon. Yet it alone rouses the finite in the infinite, and the infinite
in the finite; it is, so to speak, the impulse driving the infinite progress. This
progress is the outside of this unity at which representation remains fixated –
fixated at that perennial repetition of one and the same alternation; at the
empty unrest of a progression across the limit towards the infinite which, in
this infinity, finds a new limit but is just as unable to halt at it as it is at the
infinite. This infinite has the rigid determination of a beyond that cannot
be attained, for the very reason that it ought not be attained, since the
determinateness of the beyond, of an existent negation, has not been let go.
In this determination, the infinite has the finite as a this-side over against it –
a finite that is likewise unable to raise itself up to the infinite just because
it has this determination of an other, that is, of an existence that perennially
regenerates itself in that beyond precisely by being different from it.
In this reciprocal determination of the finite and the infinite alternating
back and forth as just indicated, the truth of these two is already implicitly
present in itself, and all that is needed is to take up what is there. This back
and forth movement constitutes the external realization of the concept in
which the content of the latter is posited, but externally, as a falling out of
the two; all that is needed is the comparing of these two different moments
in which the unity is given which the concept itself gives. “Unity of the
finite and the infinite” – as has often been already noted but must especially
be kept in mind at this juncture – is the uneven expression for the unity as
it is in truth; but also the removal of this uneven determination must be
found in the externalization of the concept that lies ahead of us.
Taken in their first, only immediate determination, the infinite is the
transcending of the finite; according to its determination, it is the negation
of the finite; the finite, for its part, is only that which must be transcended,
the negation in it of itself, and this is the infinite. In each, therefore, there is
the determinateness of the other, whereas, according to the viewpoint of the
infinite progression, the two should be mutually excluded and would have
to follow one another only alternately; neither can be posited and grasped
without the other, the infinite without the finite, the finite without the
infinite. In saying what the infinite is, namely the negation of the finite,
the finite itself is said also; it cannot be avoided in the determination
of the infinite. One need only know what is being said in order to find
the determination of the finite in the infinite. Regarding the finite, it
is readily conceded that it is the null; this very nothingness is however
the infinite from which it is inseparable. – Understood in this way, they
may seem to be taken according to the way each refers to its other. Taken
without this connecting reference, and thus joined only through an “and,”
they subsist independently, each only an existent over against the other.
We have to examine how they would be constituted in this way. The
infinite, thus positioned, is one of the two; but, as only one of them, it
is itself finite, it is not the whole but only one side; it has its limit in
that which stands over against it; and so it is the finite infinite. We have
before us only two finites. The finitude of the infinite, and therefore its
unity with the finite, lies in the very fact that it is separated from the finite
and placed, consequently, on one side. – The finite, for its part, removed
from the infinite and positioned for itself, is this self-reference in which the
relativity, its dependence and transitoriness, are removed; it is the same
self-subsistence and self-affirmation which the infinite is presumed to be.
The two pathways of consideration, even though they seem at first to
have each a different determinateness for their point of departure – the
former inasmuch as it assumes it to be only the reference of infinite and
finite to each other, of each to the other; and the latter their complete
separation from each other – yield one and the same result. The infinite
and the finite, taken together as referring to each other in a connection
which is presumed external but is in fact essential to them (for without it,
neither is what it is), each contains its other in its own determination, just
as, when each is taken for itself, when looked at on its terms, each has the
other present in it as its own moment.
This yields, then, the scandalous unity of the finite and the infinite –
the unity which is itself the infinite that embraces both itself and the
finite – the infinite, therefore, understood in a sense other than when the
finite is separated from it and placed on the other side from it. Since they
must now also be distinguished, each is within it, as just shown, itself
the unity of both; there are thus two such unities. The common element,
the unity of both determinacies, as such a unity, posits them at first as
negated, for each is to be what it is in being distinguished; in their unity,
therefore, they lose their qualitative nature – an important reflection for
countering the incorrigible habit of representing the infinite and the finite,
in their unity, as still holding on to the quality that they would have when
taken apart from each other; of seeing in that unity, therefore, nothing
except contradiction, and not also the resolution of the contradiction by
the negation of the qualitative determinateness of each. And so is the unity
of the infinite and the finite, at first simple and universal, falsified.
But further, since the two are now to be taken also as distinguished, the
unity of the infinite which is itself both of these moments is determined
differently in each. The infinite, determined as such, has in it the finitude
which is distinct from it; in this unity, the infinite is the in-itself while the
finite is only determinateness, the limit in the infinite. But such a limit is
the absolute other of the infinite, its opposite. The infinite’s determination,
which is the in-itself as such, is corrupted by being saddled with a quality
of this sort; the infinite is thus a finitized infinite. Likewise, since the finite
is as such only the non-in-itself but equally has its opposite in it by virtue
of the said unity, it is elevated above its worth and, so to speak, infinitely
elevated; it is posited as the infinitized finite.
Likewise, just as the simple unity of infinite and finite was falsified before
by the understanding, so too is the double unity. Here also this happens
because the infinite is taken in one of the two unities not as negated but,
rather, as the in-itself in which, therefore, determinateness and restriction
should not be posited, for they would debase and corrupt it. Conversely,
the finite is equally held fixed as not negated, although null in itself; so
that, in combination with the infinite, it is elevated to what it is not and is
thereby infinitized notwithstanding its determination that has not vanished
but is rather perpetuated.
The falsification that the understanding perpetrates with respect to the
finite and the infinite, of holding their reciprocal reference fixed as quali-
tative differentiation, of maintaining that their determination is separate,
indeed, absolutely separate, comes from forgetting what for the under-
standing itself is the concept of these moments. According to this concept,
the unity of the finite and the infinite is not an external bringing together of
them, nor an incongruous combination that goes against their nature, one
in which inherently separate and opposed terms that exist independently
and are consequently incompatible, would be knotted together. Rather,
each is itself this unity, and this only as a sublating of itself in which neither
would have an advantage over the other in in-itselfness and affirmative
existence. As has earlier been shown, finitude is only as a transcending
of itself; it is therefore within it that the infinite, the other of itself, is
contained. Similarly, the infinite is only as the transcending of the finite;
it therefore contains its other essentially, and it is thus within it that it is
the other of itself. The finite is not sublated by the infinite as by a power
present outside it; its infinity consists rather in sublating itself.
This sublating is not, consequently, alteration or otherness in general,
not the sublating of something. That into which the finite is sublated is
the infinite as the negating of finitude. But the latter has long since been
only existence, determined as a non-being. It is only the negation, therefore,
that in the negation sublates itself. Thus infinity is determined on its side
as the negative of the finite and thereby of determinateness in general, as
an empty beyond; its sublating of itself into the finite is a return from an
empty flight, the negation of the beyond which is inherently a negative.
Present in both, therefore, is the same negation of negation. But this
negation of negation is in itself self-reference, affirmation but as turning
back to itself, that is, through the mediation that the negation of negation is.
These are the determinations that it is essential to bring to view; the second
point, however, is that in the infinite progression they are also posited, and
how they are posited therein, namely, not in their ultimate truth.
First, both are negated in that progression, the infinite as well as the finite;
both are equally transcended. Second, they are also posited as distinct, one
after the other, each positive for itself. We sort out these two determinations
while comparing them, just as in the comparison (in an external comparing)
we have separated the two ways of considering them: the finite and the
infinite as referring to one another, and each taken for itself. The infinite
progression, however, says more than this. Also posited in it, though at
first still only as transition and alternation, is the connectedness of the terms
being distinguished. We now only need to see, in one simple reflection,
what is in fact present in it.
In the first place, the negation of the finite and the infinite which is
posited in the infinite progression can be taken as simple, and hence as
mutual externality, only a following of one upon the other. Starting from
the finite, the limit is thus transcended, the finite negated. We now have
its beyond, the infinite, but in this the limit rises up again; so we have the
transcending of the infinite. This twofold sublation is nonetheless partly
only an external event and an alternating of moments in general, and partly
still not posited as one unity; each of these moves beyond is an independent
starting point, a fresh act, so that the two fall apart. – But, in addition, their
connection is also present in the infinite progression. The finite comes first;
then there is the transcending of it, and this negative, or this beyond of the
finite, is the infinite; third, this negation is transcended in turn, a new limit
comes up, a finite again. – This is the complete, self-closing movement that
has arrived at that which made the beginning; what emerges is the same as
that from which the departure was made, that is, the finite is restored; the
latter has therefore rejoined itself, in its beyond has only found itself again.
The same is the case regarding the infinite. In the infinite, in the beyond
of the limit, only a new limit arises which has the same fate, namely, that
as finite it must be negated. Thus what is again at hand is the same infinite
that just now disappeared in the new limit; by being sublated, by traversing
the new limit, the infinite has not therefore advanced one jot further: it
has distanced itself neither from the finite (for the finite is just this, to pass
over into the infinite), nor from itself, for it has arrived at itself.
Thus the finite and the infinite are both this movement of each returning
to itself through its negation; they are only as implicit mediation, and the
affirmative of each contains the negative of each, and is the negation of the
negation. – They are thus a result and, as such, not in the determination
that they had at the beginning: neither is the finite an existence on its side
nor the infinite an existence or a being-in-itself beyond that existence, that
is, beyond existence in the determination of finitude. The understanding
strongly resists the unity of the finite and the infinite only because it pre-
supposes restriction and finitude to remain, like being-in-itself, constants.
It thereby overlooks the negation of both which is in fact present in the
infinite progression, just as it equally overlooks that the two occur in this
progression only as moments of a whole – that each emerges only through
the mediation of its opposite but, essentially, equally by means of the
sublation of its opposite.
If this immanent turning back has for the moment been reckoned to
be just as much the turning back of the finite to itself and of the infinite
to itself, noticeable in this very result is an error connected with the one-
sidedness just criticized: the finite and then the infinite is each taken as
the starting point, and only in this way two results ensue. But it is a matter
of total indifference which is taken as the starting point and, with this,
the distinction caused by the duality of results dissolves of itself. This is
likewise posited in the line of the infinite progression, open-ended on both
sides, wherein each of the moments recurs in equal alternation, and it is
totally extraneous at which position the progression is arrested and taken
as beginning. – The moments are distinguished in the progression but each
is equally only moment of the other. Since both, the finite and the infinite,
are themselves moments of the progress, they are jointly the finite, and,
since they are equally jointly negated in it and in the result, this result as
the negation of their joint finitude is called with truth the infinite. Their
distinction is thus the double meaning which they both have. The finite has
the double meaning, first, of being the finite over against the infinite which
stands over against it, and, second, of being at the same time the finite
and the infinite over against the infinite. Also the infinite has the double
meaning of being one of the two moments (it is then the bad infinite)
and of being the infinite in which the two moments, itself and its other,
are only moments. Therefore, as in fact we now have it, the nature of the
infinite is that it is the process in which it lowers itself to be only one of its
determinations over against the finite and therefore itself only one of the
finites, and elevates this distinction of itself and itself to be self-affirmation
and, through this mediation, the true infinite.
This determination of the true infinite cannot be captured in the already
criticized formula of a unity of the finite and the infinite; unity is abstract,
motionless self-sameness, and the moments are likewise unmoved beings.
But, like both its moments, the infinite is rather essentially only as becoming,
though a becoming now further determined in its moments. Becoming has
for its determinations, first, abstract being and nothing; as alteration, it has
existence, something and other; now as infinite, it has finite and infinite,
these two themselves as in becoming.
This infinite, as being-turned-back-unto-itself, as reference of itself
to itself, is being – but not indeterminate, abstract being, for it is
posited as negating the negation; consequently, it is also existence or
“thereness,” 62 for it contains negation in general and consequently deter-
minateness. It is, and is there, present, before us. Only the bad infinite is
the beyond, since it is only the negation of the finite posited as real and,
as such, it is abstract first negation; thus determined only as negative, it
does not have the affirmation of existence in it; held fast only as something
negative, it ought not to be there, it ought to be unattainable. However, to
be thus unattainable is not its grandeur but rather its defect, which is at
bottom the result of holding fast to the finite as such, as existent. It is the
untrue which is the unattainable, and what must be recognized is that such
an infinite is the untrue. – The image of the progression in infinity is the
straight line; the infinite is only at the two limits of this line, and always
only is where the latter (which is existence) is not but transcends itself, in
its non-existence, that is, in the indeterminate. As true infinite, bent back
upon itself, its image becomes the circle, the line that has reached itself,
closed and wholly present, without beginning and end.
True infinity, thus taken in general as existence posited as affirmative in
contrast to abstract negation, is reality in a higher sense than it was earlier
as simply determined; it has now obtained a concrete content. It is not
the finite which is the real, but rather the infinite. Thus reality is further
determined as essence, concept, idea, and so forth. In connection with the
more concrete, it is however superfluous to repeat such earlier and more
abstract categories as reality, and to use them for determinations more
concrete than they are by themselves. Such a repetition, as when it is said
that essence, or that the concept, is real, has its origin in the fact that to
uneducated thought the most abstract categories such as being, existence,
reality, finitude, are the most familiar.
The more immediate occasion, however, for recalling here the categories
of reality is that the negation, against which reality is the affirmative, is here
the negation of negation, and consequently itself posited over against that
reality which finite existence is. – Negation is thus determined as ideality;
the idealized h is the finite as it is in the true infinite – as a determination, a
content, a distinct but not a subsistent existent, a moment rather. Ideality has
this more concrete signification which is not fully expressed through the
negation of finite existence. – As regards reality and ideality, the opposition
The ideal [das Ideale] has a broader meaning (such as of the beautiful and its associations) than
the idealized [das Ideelle]. The former does not belong here yet, and for this reason the expression
idealized is being used. There is no such distinction made in language usage for “reality”; in German
reelle and reale are used as roughly synonymous and no interest is served in nuancing the two in
some sort of opposition.
of finite and infinite is, however, so grasped that the finite assumes the value
of “the real,” whereas the infinite that of “the idealized”; in the same way,
further on, also the concept is regarded as an idealization, that is, as a mere
idealization, in contrast to existence in general, which is regarded as “the
real.” When contrasted in this way, it is of course of no use to have reserved
for the said concrete determination of negation the distinctive expression
of “idealization”; in that opposition of finite and infinite, we are back to
the one-sidedness of the abstract negative characteristic of the bad infinite
and still fixed in the affirmative existence of the finite.

Transition

Ideality can be called the quality of the infinite;
but it is essentially the process of becoming,
and hence a transition, like the transition
of becoming into existence.
We must now explicate this transition.
This immanent turning back, as the sublating of finitude,
of finitude as such and equally of the negative finitude that
only stands opposite to it, is only negative finitude,
is self-reference, being.
Since there is negation in this being, the latter is existence;
but, further, since the negation is essentially
negation of the negation, self-referring negation,
it is the existence that carries the name of being-for-itself.

CHAPTER 3

Being-for-itself

In being-for-itself, qualitative being is brought to completion;
it is infinite being;
the being of the beginning is void of determination;
existence is sublated but only immediately sublated being;
it thus contains, to begin with, only the first negation, itself immediate;
being is of course retained as well,
and the two are united in existence in simple unity;
for this reason, however, each is in itself still unlike the other,
and their unity is still not posited.
Existence is therefore the sphere of differentiation,
of dualism, the domain of finitude.
Determinateness is determinateness as such;
being which is relatively, not absolutely, determined.
In being-for-itself, the distinction between being and determinateness,
or negation, is posited and equalized.
Quality, otherness, limit, as well as reality, in-itselfness,
ought, and so forth, are the incomplete configurations of negation in being
which are still based on the differentiation of the two.
But since in finitude
negation has passed over into infinity, in the posited negation of negation,
negation is simple self-reference and in it, therefore, the equalization with
being – absolutely determinate being.
First, being-for-itself is immediately an existent-for-itself, the one.
Second, the one passes over into a multiplicity of ones – repulsion or the
otherness of the one which sublates itself into its ideality, attraction.
Third, we have the alternating determination of repulsion and attraction
in which the two sink into a state of equilibrium; and quality, driven to a
head in being-for-itself, passes over into quantity.

A. BEING-FOR-ITSELF AS SUCH

The general concept of being-for-itself has come to light. The justification
for using the expression “being-for-itself” for that concept would depend on
showing that the representation associated with the expression corresponds
to the concept. So indeed it appears to do. We say that something is for itself
inasmuch as it sublates otherness, sublates its connection and community
with other, has rejected them by abstracting from them. The other is in
it only as something sublated, as its moment; being-for-itself consists in
having thus transcended limitation, its otherness; it consists in being, as
this negation, the infinite turning back into itself. – In representing to itself
an intended object which it feels, or intuits, and so forth, consciousness
already contains in itself as consciousness the determination of being-
for-itself; that is, it has in it the content of that object, which is thus
an idealization; even as it intuits, or in general becomes involved in the
negative of itself, in the other, it abides with itself. Being-for-itself is the
polemical, negative relating to the limiting other and, through this negation
of the other, is being-reflected-within-itself – even though, side by side with
this immanent turning back of consciousness and the ideality of its object,
the reality of this object is also retained, for the object is at the same time
known as an external existence. Consciousness is thus phenomenal, 64 or
it is this dualism: on the one side, it knows an external object which is
other than it; on the other side, it is for-itself, has this intended object
in it as idealized, abides not only by this other but therein abides also
with itself. Self-consciousness, on the contrary, is being-for-itself brought to
completion and posited; the side of reference to another, to an external
object, is removed. Self-consciousness is thus the nearest example of the
presence of infinity – granted, of a still abstract infinity, but one which is
of a totally different, concrete determination than the being-for-itself in
general, whose infinity still has only qualitative determinateness.

a. Existence and being-for-itself

As already mentioned, being-for-itself is infinity that has sunk into simple
being; it is existence in so far as in the now posited form of the immediacy
of being the negative nature of infinity, which is the negation of negation,
is only as negation in general, as infinite qualitative determinateness. But
in such a determinateness, wherein it is existence, being is at once also
distinguished from this very being-for-itself which is such only as infinite
qualitative determinateness; nevertheless, existence is at the same time a
moment of being-for-itself, for the latter certainly contains being affected
by negation. So the determinateness which in existence as such is an other,
and a being-for-other, is bent back into the infinite unity of being-for-itself,
and the moment of existence is present in the being-for-itself as being-for-
one.

b. Being-for-one

This moment gives expression to how the finite is in its unity with the
infinite or as an idealization.
Being-for-itself does not have negation in it as
a determinateness or limit, and consequently also not as
reference to an existence other than it.
Although this moment is now being designated
as being-for-one, there is yet nothing at hand for which it would be –
there is not the one of which it would be the moment.
There is in fact nothing of the sort yet fixed in being-for-itself;
that for which something (and there is no something here)
would be, what the other side in general should be,
is likewise a moment, itself only being-for-one, not yet a one.
What we have before us, therefore, is still an undistinguishedness of two
sides that may suggest themselves in the being-for-one; there is only one
being-for-another, and since this is only one being-for-another, it is also only
being-for-one; there is only the one ideality, of that for which or in which
there should be a determination as moment, and of that which should
be the moment in it.
Being-for-one and being-for-itself do not therefore
constitute two genuine determinacies, each as against the other. Inasmuch
as the distinction is momentarily assumed and we speak of a-being-for-
itself, it is this very being-for-itself, as the sublated being of otherness,
that refers itself to itself as to the sublated other, is therefore for-one; in
its other it refers itself only to itself. An idealization is necessarily for-one,
but it is not for an other; the one, for which it is, is only itself.
The “I,” therefore, spirit in general, or God, are idealizations,
because they are infinite; as existents which are for-themselves, however,
they are not ideationally different from that which is for-one.
For if they were different, they would be only immediate, or, more precisely, they would only be
existence and a being-for-another; for if the moment of being for-one did
not attach to them, it is not they themselves but an other that would be
that which is for them.
God is therefore for himself, in so far he is himself
that which is for him.
Being-for-itself and being-for-one are not, therefore, diverse significations of
ideality but essential, inseparable, moments of it.

c. The one

Being-for-itself is the simple unity of itself and its moments, of the
being for-one. There is only one determination present, the self-reference
itself of the sublating. The moments of being-for-itself have sunk into
an indifferentiation which is immediacy or being, but an immediacy that
is based on the negating posited as its determination. Being-for-itself is
thus an existent-for-itself, and, since in this immediacy its inner meaning
vanishes, it is the totally abstract limit of itself – the one.
Attention may be drawn in advance to the difficulties that lie ahead
in the exposition of the development of the one, and to the source of
these difficulties. The moments that constitute the concept of the one as
being-for-itself occur in it one outside the other; they are (1) negation in
general; (2) two negations that are, therefore, (3) the same, (4) absolutely
opposed; (5) self-reference, identity as such; (6) negative reference which
is nonetheless self-reference. These moments occur here apart because the
form of immediacy, of being, enters into the being-for-itself as existent-for-
itself; because of this immediacy, each moment is posited as a determination
existent on its own, and yet they are just as inseparable. Hence, of each
determination the opposite must equally be said; it is this contradiction
that causes the difficulty that goes with the abstract nature of the moments.

B. THE ONE AND THE MANY

The one is the simple reference of being-for-itself to itself in which its
moments have fallen together – in which, therefore, being-for-itself has the
form of immediacy and its moments, therefore, are now there as existents.
As the self-reference of the negative, the one is a determining – and, as
self-reference, it is infinite self-determining. However, because of the present
immediacy, these distinctions are no longer only moments of one and the
same self-determination but are at the same time posited as existents. The
ideality of the being-for-itself as a totality thus turns at first into reality – a
reality, moreover, of the most fixed and abstract kind, as a one. In the one,
the being-for-itself is the posited unity of being and existence, as the absolute
union of the reference to another and the reference to itself; but also the
determinateness of being then enters into opposition to the determination
of the infinite negation, to self-determination, so that what the one is in
itself, it is that now only in it, and the negative consequently is an other
distinct from it. What shows itself to be present as distinct from the one is
the one’s own self-determining; its unity with itself, as thus distinct from
itself, is demoted to reference, and, as negative unity, it is negation of itself
as other, the excluding of the one as an other from itself, from the one.

a. The one within

Within it, the one just is; this, its being, is not an existence, not a deter-
mination as reference to an other, not a constitution; it is rather its having
negated this circle of categories. The one is not capable, therefore, of
becoming any other; it is unalterable.
It is indeterminate, yet no longer like being; its indeterminateness is the
determinateness of self-reference, absolutely determined being; posited in-
itselfness. As negation which, in accordance with its concept, is self-referring,
it has distinction in it: it directs away from itself towards another, but this
direction is immediately reversed, because, according to this moment of
self-determining, there is no other to which it would be addressed, and the
directing reverts back to itself.
In this simple immediacy, even the mediation of existence and ideality,
and with it all diversity and manifoldness, have vanished. In the one there is
nothing; this nothing, the abstraction of self-reference, is here distinguished
from the in-itselfness of the one; it is a posited nothing, for this in-itselfness
no longer has the simplicity of the something, but, as mediation, has
rather the determination of being concrete; taken in abstraction, it is
indeed identical with the one, but different from its determination. So this
nothing, posited as in the one, is the nothing as the void. – The void is thus
the quality of the one in its immediacy.
b. The one and the void
The one is the void as the abstract self-reference of negation. But the void,
as nothing, is absolutely diverse from the simple immediacy of the one,
from the being of the latter which is also affirmative, and because the two
stand in one single reference, namely to the one, their diversity is posited;
however, as distinct from the affirmative being, the nothing stands as void
outside the one as existent.
Being-for-itself, determined in this way as the one and the void, has
again acquired an existence. – The one and the void have their negative
self-reference as their common and simple terrain. The moments of being-
for-itself come out of this unity, become external to themselves; for through
the simple unity of the moments the determination of being comes into
play, and the unity itself thus withdraws to one side, is therefore lowered
to existence, and there it is confronted by its other determination standing
over against it, negation as such and likewise as the existence of the nothing,
as the void.

c. Many ones

Repulsion

The one and the void constitute the first existence of being-for-itself. Each
of these moments has negation for its determination, and is posited at the
same time as an existence. In accordance with this determination, the one
and the void are each the reference of negation to negation as of an other
to its other: the one is negation in the determination of being; the void,
negation in the determination of non-being. Essentially, however, the one
is only self-reference as referring negation, that is, it is itself the same as the
void outside it is supposed to be. Both are, however, also posited as each
an affirmative existence – the one as being-for-itself as such, the other as
indeterminate existence in general – and each as referring to the other as
to an other existence. Essentially, however, the being-for-itself of the one
is the ideality of the existence and of the other; it does not refer to an
other but only to itself. But inasmuch as the being-for-itself is fixed as the
one, as existent for itself, as immediately present, its negative reference to
itself is at the same time reference to an existent; and since the reference
is just as much negative, that to which the being-for-itself refers remains
determined as an existence and as an other; as essentially self-reference, the
other is not indeterminate negation like the void, but is likewise a one. The
one is consequently a becoming of many ones.
Strictly speaking, however, this is not just a becoming; for becoming is
a transition of being into nothing; the one, by contrast, becomes only a
one. The one, as referred to, contains the negative as reference; it has this
reference, therefore, in it. Hence, instead of a becoming, the one’s own
immanent reference is, first, present; and, second, since this reference is
negative and the one is at the same time an existent, the one repels itself
from itself. This negative reference of the one to itself is repulsion.
This repulsion, as thus the positing of many ones but through the one
itself, is the one’s own coming-forth-from-itself, but to such outside it as
are themselves only ones. This is repulsion according to the concept, as it
exists implicitly in itself. The second repulsion is distinguished from it. It
is the one that first occurs to the representation of external reflection, not
as the generation of ones but only as the mutual holding off of ones which
are presupposed as already there. To be seen now is how the first repulsion
that exists in itself determines itself as the second, the external repulsion.
We must first establish the determinations that the many ones have as
such. The becoming of the many, or their being produced, immediately
vanishes as the product of a positing; what is produced are the ones, not
for another, but as infinitely referring to themselves. The one repels only
itself from itself; it does not come to be but it already is; that which is
represented as the repelled is equally a one, an existent; repelling and being
repelled applies in like manner to both, and makes no difference.
The ones are thus presupposed with respect to each other – posited through
the repulsion of the one from itself; pre-supposed, posited as non-posited;
their being-posited is sublated, they are existents with respect to each other,
such as refer only to themselves.
Thus plurality appears not as an otherness, but as a determination com-
pletely external to the one. The one, in repelling itself, remains reference to
itself, just like that which is taken as repelled at the start. That the ones are
other to one another, that they are brought together in the determinateness
of plurality, does not therefore concern the one. If the plurality were a
reference of the ones to one another, the ones would then limit each other
and would have the being-for-other affirmatively in them. Their connect-
ing reference (and this they have through their unity which is in itself ),
as posited here, is determined as none; it is again the previously posited
void. This void is their limit, but an external limit in which they are not
supposed to be for one another. The limit is that in which the limited are
just as much as are not; but the void is determined as pure non-being, and
this alone constitutes the limit of the ones.
The repulsion of the one from itself is the making explicit of what the
one is implicitly in itself; but, thus laid out as one-outside-the-other, infinity
is here an infinity that has externalized itself, and this it has done through
the immediacy of the infinite, of the one. Infinity is just as much the simple
reference of the one to the one as, on the contrary, the one’s absolute lack
of reference; it is the former according to the simple affirmative reference
of the one to itself; it is the latter according to the same reference as
negative. Or again, the plurality of the ones is the one’s own positing of
the one; the one is nothing but the negative reference of the one to itself,
and this reference – hence the one itself – is the plural one. But equally,
plurality is utterly external to the one, for the one is precisely the sublating
of otherness; repulsion is its self-reference and simple equality with itself.
The plurality of the ones is infinity as a contradiction that unconstrainedly
produces itself.

C. REPULSION AND ATTRACTION

a. Exclusion of the one

The many ones are each a being; their existence or their reference to
one another is a non-reference, it is external to them: the abstract void.
But they themselves are now this negative reference to themselves as to
existent others: the demonstrated contradiction, the infinity posited in the
immediacy of being. With this, repulsion now finds immediately before it
that which is repelled by it. In this determination, it is an excluding; the one
repels from itself only the many not generated by it, the ones not posited
by it. This repelling is mutual or from all sides – relative, limited by the
being of the ones.
Plurality is not at first posited otherness; limit is only the void, only that
in which the ones are not. But in the limit they also are; they are in the
void, or their repulsion is their common connecting reference.
This mutual repulsion is the posited existence of the many ones; it is not
their being-for-itself, in accordance with which they would be distinguished
as many only in a third, but is rather their own distinguishing which
preserves them. – They mutually negate themselves, posit one another as
being only for-one. But at the same time they negate this being only for-
one just as much; they repel the ideality that they have and are. – So the
moments which in ideality are absolutely united come apart. In its being-
for-itself, the one is also for-one; but this one, for which it is, is itself; its
distinguishing from itself is immediately sublated. But in the plurality the
distinguished one has a being; the being-for-one as has been determined
in exclusion is therefore a being-for-other. Each thus comes to be repelled
by an other, is sublated and made into a one which is not for itself but
for-one, and an other one at that.
The being-for-itself of the many ones thus shows itself to be their self-
preservation through the mediation of their mutual repulsion in which
they sublated themselves reciprocally and posit the others as mere being-
for-another. But the self-preservation consists at the same time in repelling
this ideality and positing the ones as not being for-an-other. This self-
preservation of the ones through their negative reference to one another is,
however, rather their dissolution.
The ones not only are but maintain themselves through their recipro-
cal exclusion. First, it is in their being, and indeed their being-in-itself as
contrasted with their reference to the other, that they should now have a
firm point of support for their diversity as against their being negated; this
in-itselfness rests on their being ones. But they all are this; in their being-in-
itself, instead of having there their firm point of support for their diversity,
they are all the same. Second, their existence and their way of relating
to one another, that is, their positing themselves as one, is their recipro-
cal negating; this, however, is likewise one and the same determination of
all through which they therefore posit themselves as identical; just as, by
being in themselves the same, the ideality that should be posited in them
through others is their own, and they thus repel just as little. – According
to their being and positing, they are, consequently, only one affirmative
unity.
This consideration regarding the ones – that from either side of their
determination, whether they just are or refer to one another, they show
themselves to be only one and the same, indistinguishable – is a comparison
that belongs to us. – Also to be seen, therefore, is what is posited in them
in their mutual reference itself. – They are – this much is presupposed
in this reference – and they are only inasmuch as they negate themselves
reciprocally and at the same time keep away this ideality, their being
negated, from themselves, that is, they negate the reciprocal negating.
But they are only inasmuch as they negate, and so, since their reciprocal
negating is negated, their being is negated. To be sure, since they are,
nothing would be negated through this negating which for them is only
something external; this negating of the other rebounds off them, coming
their way only by striking their surface. And yet, they turn back upon
themselves only by negating the others; they are only as this mediation,
this turning back of theirs is their self-preservation and their being-for-
itself. Since their negating is ineffectual because of the resistance offered by
the others, whether as existents or as negating, they do not return back to
themselves, do not preserve themselves, and so are not.
It was previously remarked that the ones themselves are each a one like
any other. 77 This is not just a matter of our connecting them by way of
reference, of bringing them together externally; repulsion is itself a referring;
the one that excludes the ones refers itself to them, to the ones, that is,
to itself. The negative relating of the ones to one another is consequently
only a coming-together-with-oneself. This identity in which their repelling
crosses over is the sublation of their diversity and externality which they
should have rather asserted with respect to each other by excluding each
other.
This self-positing-in-a-one of the many ones is attraction.

b. The one one of attraction

Repulsion is the fragmentation of the one, first into the many of which it
is the negative relating, since they presuppose each other as each existent;
it is only the ought of ideality; this ideality will, however, be realized in
attraction. Repulsion passes over into attraction, the many ones into one
one. Both, repulsion and attraction, are at first distinguished from each
other, repulsion as the reality of the ones, attraction as their posited ideality.
Attraction refers to repulsion by having it for a presupposition. Repulsion
delivers the material for attraction. If there were no ones, there would
be nothing to attract; the representation of continuing attraction, of the
consumption of the ones, presupposes an equally continuing generation
of the ones; the sense representation of spatial attraction gives continuity
to the flow of ones to be attracted; to replace the atoms that vanish at the
point of attraction, another multitude comes forth from the void, infinitely
if one so wishes. If attraction were represented as accomplished, that is,
the many as brought to the point of the one one, the result would be just
an inert one, no longer any attraction. The ideality immediately present in
attraction still also has in it the determination of the negation of itself, the
many ones to which it refers; attraction is inseparable from repulsion.
To attract pertains at first in equal measure to each of the many ones as
immediately present; none has advantage over an other; what would result
then is an equilibrium in the attraction, or more precisely, an equilibrium
in the attraction and the repulsion themselves, and an inert state of rest
without any ideality present there. But there can be no question here of
any such immediately present one taking precedence over another, for
this would presuppose a determinate distinction between them; attraction
is rather the positing of the given lack of distinction among the ones.
Attraction is itself the positing in the first place of a one distinct from other
ones; these are only the immediate ones that are to preserve themselves
through repulsion; through their posited negation, however, what proceeds
is the one of attraction which is therefore determined as the mediated one,
the one posited as one. The first ones, as immediate, do not in their ideality
return into themselves, but have this ideality each in another.
The one one is, however, ideality that has been realized, posited in the
one; it attracts through the mediation of repulsion; it contains in itself this
mediation as its determination. It thus does not swallow the attracted ones
within it as into one point, that is, does not sublate them abstractly. Since
it contains repulsion in its determination, the latter equally preserves the
ones as many within it; by its attracting, it musters, so to speak, something
before it, gains an area or a filling. Thus there is in it the unity of repulsion
and attraction in general.

c. The connection of repulsion and attraction

The difference of the one and the many has determined itself as a difference
of their mutual reference connecting them which breaks down into two,
repulsion and attraction, each of which stands at first outside the other
on its own, in such a way that the two are essentially joined together
nevertheless. Their still indeterminate unity must be brought out in greater
detail.
As the fundamental determination of the one, repulsion appears first, and
it appears as immediate, like its ones which are indeed generated by it and
yet are at the same time posited as immediate, and it is therefore indifferent
to the attraction which is added to it externally as thus presupposed. Rather,
attraction is not presupposed by repulsion: it is not supposed to have any
part in the positing and in the being of the latter, that is, as if repulsion
were not, already in it, the negation of itself, or the ones were not already
negated in it. In this way, we have repulsion in abstraction, by itself, and
attraction likewise holds out to the ones, as each an existent, the side of an
immediate existence which comes to them by itself as an other.
If we take mere repulsion in this way, for itself, it is then the dispersion
of the many ones in indeterminacy, outside the sphere of repulsion itself;
for repulsion is the negating of the connection of the many to one another;
lack of connection is their determination when abstractly taken. But repul-
sion is not just the void; the ones, although unconnected, do not repel
what constitutes their determination, do not exclude it. Although nega-
tive, repulsion is nonetheless essentially connection; the mutual repulsion
and flight is not a liberation from what is repelled and fled from; that which
is excluded still stands in connection with what is excluded from it. But this
moment of connection is attraction, which is thus within repulsion itself;
it is the negating of that abstract repulsion by which the ones would each
be an existent referring only to itself without mutual exclusion.
But in starting with the repulsion of the ones as immediately present
there, and with attraction consequently also posited as intruding on them
externally, the two, repulsion and attraction, are held apart as diverse
determinations despite their inseparability. But it has been established that
it is not just repulsion which is presupposed by attraction, but that there
equally is present also a reverse connection of repulsion to attraction, and
that repulsion no less has attraction for its presupposition.Being-for-itself
As thus determined, they are inseparable, and at the same time each
is determined as an ought and a limitation with respect to the other.
Their ought is their abstract determinateness as each an existent in itself –
a determinateness, however, which is thereby directed beyond itself and
refers to the other. And so, through the mediation of the other, each is as
other; their self-subsistence consists in their being mutually posited in this
mediation as an other determining. – Thus, repulsion is the positing of the
many; attraction the positing of the one; this latter is equally the negation
of the many and the former the negation of the ideality of such a many in
the one; so that attraction too is attraction only through the mediation of
repulsion, just as repulsion is repulsion through the mediation of attraction.
In all this, however, the mediation of each with itself through the other is
in fact negated; each of the two determinations is its own self-mediation.
This will result from a closer examination of the two determinations and
will bring us back to the unity of their concept.
In the first place, that each presupposes itself, that in its presupposition
each refers only to itself, this is already present in the way the still relative
repulsion and attraction behave at first.
Relative repulsion is the mutual repulsion of many ones which are already
at hand, supposedly immediately given. But that there be many ones, this is
repulsion itself; any presupposition that it would have is only its own posit-
ing. Moreover, the determination of the being that would accrue to the ones
apart from their being posited – whereby they would already be – belongs
likewise to repulsion. Repelling is that through which the ones manifest
themselves and maintain themselves as ones; through which they are as
such. Their being is their repulsion itself, which is thus not some relative
existence against another other but relates itself throughout only to itself.
Attraction is the positing of the one as such, of the real one, with respect
to which the existence of the many is determined as only a vanishing
idealization. Attraction thus directly presupposes itself; it presupposes itself
in the determination namely, of the many ones to be an idealization, the
same ones which are otherwise supposed to have existence for themselves
and to repel others, including therefore any other that attracts. Against this
determination of repulsion, the ones do not attain ideality only through
the relation to attraction; on the contrary, the ideality is presupposed: it
is the ideality of the ones as an existent in itself, inasmuch as they, as ones
(including the one conceived as attracting), are not distinguished from one
another but are one and the same.
This self-presupposing of the two determinations, each for itself,
implies further that each contains within itself the other as moment.
Self-presupposing in general is the positing of oneself in a one as the neg-
ative of oneself (repulsion), and what is presupposed in this positing is
the same as that which presupposes (attraction). That each is in itself only
a moment, this is the transition of each from itself into the other, the
negation of itself in the other and the positing of itself as the other of
itself. The one, as such, is thus a coming-out-of-itself; is itself only the
positing of itself as its other, as the many. And the many, for its part, is
only the falling back upon itself and the positing of itself as its other, as a
one, and is in this equally only the connecting of itself to itself, each con-
tinuing itself in its other. Therefore, the coming-out-of-itself (repulsion)
and the self-positing-as-one (attraction) are already inherently present as
undivided. But in the repulsion and attraction which are relative, that is,
which presuppose immediate, determinedly existent ones, it is posited that
the two are each, within it, this negation of itself, and consequently also the
continuity of itself in its other. The repulsion of the determinedly existent
ones is the self-preservation of the one through the mutual holding off of
the others, so that (1) the other ones are negated in it (this is the side of
its existence or of its being-for-another and is therefore attraction as the
ideality of the ones); and (2) the one is in itself, without reference to the
others (however, not only has the in-itself in general long since passed over
into being-for-itself; the one in itself, according to its determination, is
the coming to be of many). – The attraction of the existent ones is their
ideality and the positing of the one, and in this, as both the negating and
the producing of the one, attraction sublates itself, and as a positing within
it of the one, is the negative of itself: it is repulsion.
With this, the development of being-for-itself is completed and has
attained its result. In connecting itself to itself infinitely, that is, as the
posited negation of negation, the one is the mediation by which it repels
itself as its absolute (that is, abstract) otherness (the many) from itself, and
in thus negatively connecting itself to this, its non-being, it sublates it and is
in it precisely only the connection to itself. The one is only this becoming
in which the determination “it begins,” that is, its being posited as an
immediate existent, and equally that, as result, it has restored itself as the
one, that is, the equally immediate and exclusive one, have vanished; the
process which it is, posits and contains it from all sides only as something
sublated. The sublation, determined at first only as a relative sublating of
the connection to another existent, a connection which is therefore itself not
an indifferent repulsion and attraction, equally proves itself to pass over into
the infinite connection of mediation through the negation of the external
connection of immediate and determinate existents, and to have for result
precisely that becoming which, in the instability of its moments, is the
collapse, or rather the going-together-with-itself, into simple immediacy.
This being, according to the determination which it has now acquired, is
quantity.
If we briefly review the moments of this transition of quality into quantity,
we find that the qualitative has being and immediacy for its fundamental
determination, and the limit and the determinateness are in this immediacy
so identical with the being of something, that the something itself vanishes
along with its alteration; as thus posited, it is determined as finite. Because
of the immediacy of this unity in which the distinction has disappeared,
although it is implicitly present in the unity of being and nothing, the
distinction falls outside that unity as otherness in general. This reference to
the other contradicts the immediacy in which qualitative determinateness
is self-reference. This otherness is sublated in the infinity of the being-for-
itself, the being-for-itself that has realized the distinction implicitly present
in it in the negation of negation: has realized it as the one and the many
and as their connecting references, and has also elevated the qualitative to
true unity, that is, a unity which is no longer immediate but posited as
accordant with itself.
This unity is, therefore, (a) being, only as affirmative, that is, immediacy
self-mediated through the negation of negation: being is posited as a unity
permeating its determinacies, limits, etc., which are posited in it as sublated;
(b) existence: in this determination it is negation or determinateness as
moment of the affirmative being; yet this determinateness is no longer
immediate but reflected into itself, refers not to another but to itself;
absolutely-determined-being, absolute in-itselfness, the one; otherness as
such is itself being-for-itself; (g) being-for-itself : as that being which persists
across the determinateness and in which the one and even the being-
determined-in-itself are posited as sublated. The one is simultaneously
determined as having gone beyond itself and as unity; the one, the absolutely
determined limit, is consequently posited as a limit which is none, a limit
which is in being but is indifferent to it.
SECTION II

Magnitude (Quantity)

The difference between quantity and quality has been indicated.
Quality is the first, immediate determinateness.
Quantity is the determinateness that has become indifferent to being;
a limit which is just as much no limit;
being-for-itself which is absolutely identical with being-for-another:
the repulsion of the many ones which is immediate non-repulsion,
their continuity.
Because that which exists for itself is now so posited
that it does not exclude its other but rather affirmatively
continues in it, it is then otherness,
inasmuch as existence surfaces again on this continuity and
its determinateness is at the same time no longer simple self-reference, no longer
the immediate determinateness of the existent something, but is posited
as repelling itself from itself, as referring to itself in the determinateness
rather of an other existence (a being which exists for itself ); and since they
are at the same time indifferent limits, reflected into themselves and uncon-
nected, determinateness is as such outside itself, an absolute externality and
a something just as external; such a limit, the indifference of the limit as
limit and the indifference of the something to the limit, constitutes the
quantitative determinateness of the something.
In the first place, we have to distinguish pure quantity from quantity
as determinate, from quantum. First, pure quantity is real being-for-itself
turned back into itself, with as yet no determinateness in it: a compact,
infinite unity which continues itself into itself.
Second, this quantity proceeds to determinateness, and
this is posited in it as a determinateness that
at the same time is none, is only external.
Quantity becomes quantum. Quantum is indifferent determinateness,
that is, one that transcends itself, negates itself; as this otherness of other-
ness, it lapses into infinite progress.
Infinite quantum, however, is sublated indifferent determinateness:
it is the restoration of quality.
Third, quantum in qualitative form is quantitative ratio. Quantum tran-
scends itself only in general; in the ratio, however, it transcends itself
into its otherness, in such a way that this otherness in which it has its
determination is at the same time posited, is another quantum. With this
we have quantum as turned back into itself and referring to itself as into
its otherness.
At the foundation of this relation there still lies the externality of quantum;
it is indifferent quanta that relate themselves to each other, that is,
they have the reference that mutually connects them in this being-outside-itself.
The ratio is, therefore, only a formal unity of quality and quantity, and
dialectic is its transition into their absolute unity, in measure.

CHAPTER 1

Quantity

a. pure quantity

Quantity is sublated being-for-itself. The repelling one that behaved only
negatively towards the excluded one, now that it has gone over in connec-
tion with it, behaves towards the other as identical to itself and has there-
fore lost its determination; being-for-itself has passed over into attraction.
The absolute obduracy of the one has melted away into this unity which,
however, as containing the one, is at the same time determined by the
repulsion residing in it; as unity of the self-externality, it is unity with itself.
Attraction is in this way the moment of continuity in quantity.
Continuity is therefore simple, self-same reference to itself unbroken by
any limit or exclusion – not, however, immediate unity but the unity of
ones which have existence for themselves. Still contained in it is the outside-
one-another of plurality, though at the same time as something without
distinctions, unbroken. Plurality is posited in continuity as it implicitly is
in itself; the many are each what the others are, each is like the other, and the
plurality is, consequently, simple and undifferentiated equality. Continuity
is this moment of self-equality of the outsideness-of-one-another, the self-
continuation of the different ones into the ones from which they are
distinguished.
In continuity, therefore, magnitude immediately possesses the moment
of discreteness – repulsion as now a moment in quantity. – Steady continuity 1
is self-equality, but of many that do not become exclusive; it is repulsion
that first expands self-equality to continuity. Hence discreteness is, for its
part, a discreteness of confluents – of ones that do not have the void to
connect them, not the negative, but their own steady advance and, in the
many, do not interrupt this self-equality.
Quantity is the unity of these moments, of continuity and discreteness.
At first, however, it is this continuity in the form of one of them, of
continuity, as a result of the dialectic of the being-for-itself which has col-
lapsed into the form of self-equal immediacy. Quantity is as such this simple
result in so far as the being-for-itself has not yet developed its moments and
has not posited them within it. – Quantity contains these moments at first
as being-for-itself posited in its truth. It was the determination of being-in-
itself to be self-sublating self-reference, a perpetual coming-out-of-itself.
But what is repelled is itself; repulsion is thus a creative flowing away
from itself. On account of the sameness of what is repelled, this discerning
is unbroken continuity; and on account of the coming-out-of-itself, this
continuity is at the same time, without being broken off, a plurality – a
plurality which persists just as immediately in its equality with itself.

b. continuous and discrete magnitude

1. Quantity contains the two moments of continuity and discreteness. It
is to be posited in both, in each as its determination. – It is already from
the start the immediate unity of the two, that is, quantity is itself posited at
first only in one of the two determinations, that of continuity, and as such
is continuous magnitude.
Or continuity is indeed one of the moments of quantity which is brought
to completion only with the other, discreteness. But quantity is concrete
unity only in so far as it is the unity of distinct moments. These are to
be taken, therefore, also as distinct, without however resolving them again
into attraction and repulsion but, rather, as they truly are, each remaining
in its unity with the other, that is, remaining the whole. Continuity is only
the compact unity holding together as unity of the discrete; posited as such,
it is no longer only moment but the whole quantity: continuous magnitude.
2. Immediate quantity is continuous magnitude. Quantity, however, is
not as such an immediate; immediacy is a determinateness, the sublated
being of which is precisely quantity. Quantity is to be posited, therefore,
in the determinateness immanent to it, and this is the one. Quantity is
discrete magnitude.
Discreteness is, like continuity, a moment of quantity, but is itself also
the whole quantity just because it is a moment in it, the whole, and
therefore as distinct moment does not diverge from its unity with the
other moment. – Quantity is the outsideness-of-one-another as such, and
continuous magnitude is this outsideness-of-one-another onwardly posit-
ing itself without negation as an internally self-same connectedness. On
the other hand, discrete magnitude is this outsideness-of-one-another as
discontinuous, as broken off. With this aggregate of ones, however, the
aggregate of atom and void, repulsion in general, is not thereby reinstated.
Because discrete magnitude is quantity, its discreteness is itself continuous.
Such a continuity in the discrete consists in the ones being the same as
one another, or in that they have the same unity. Discrete magnitude is
therefore the one-outside-the-other of the many ones as of a same – not the
many ones in general, but posited rather as the many of a unity.

c. the limiting of quantity

Discrete magnitude has, first, the one for its principle and, second, is a
plurality of ones; third, it is essentially continuous, it is the one as at the
same time sublated, as unity, the self-continuing as such in the discreteness
of the ones. Consequently, it is posited as one magnitude, and the “one”
is its determinateness – a “one” which, in this posited and determinate
existence, excludes, is a limit to the unity. Discrete magnitude as such is not
supposed to be immediately limited, but, when distinguished from contin-
uous magnitude, it is an existence and a something, the determinateness
of which, and in it also the first negation and limit, is the “one.”
This limit, besides referring to the unity and being the moment of
negation in it, is also, as one, self-referred; thus it is enclosing, encompassing
limit. The limit here is not at first distinct from the something of its
existence, but, as one, is essentially this negative point itself. But the being
which is here limited is essentially as continuity, and in virtue of this
continuity it transcends the limit, transcends this one, and is indifferent to
it. Real, discrete quantity is thus one quantity, or quantum – quantity as
an existence and a something.
Since the one which is a limit encompasses within it the many ones
of discrete quantity, it posits them equally as sublated in it; it is a limit
to continuity simply as such and, consequently, the distinction between
continuous and discrete magnitude is here indifferent; or, more precisely,
it is a limit to the continuity of the one just as much as of the other; in this,
both pass over into being quanta.
Quantum, which in the first instance is quantity with a determinateness or
limit in general, in its complete determinateness is number. Second, quan-
tum divides first into extensive quantum, in which limit is the limitation of a
determinately existent plurality; and then, inasmuch as the existence of this
plurality passes over into being-for-itself, into intensive quantum or degree.
This last is for-itself but also, as indifferent limit, equally outside itself. It thus
has its determinateness in an other. Third, as this posited contradiction of
being determined simply in itself yet having its determinateness outside
itself and pointing outside itself for it, quantum, as thus posited outside
itself within itself, passes over into quantitative infinity.

A. NUMBER

Quantity is quantum, or has a limit, both as continuous and discrete
magnitude. The distinction between these two species has here, in the first
instance, no significance.
As the sublated being-for-itself, quantity is already in and for itself
indifferent to its limit. But, equally, the limit or to be a quantum is not
thereby indifferent to quantity; for quantity contains within itself as its
own moment the absolute determinateness of the one, and this moment,
posited in the continuity or unity of quantity, is its limit, but a limit which
remains as the one that quantity in general has become.
This one is therefore the principle of quantum, but as the one of quantity.
For this reason it is, first, continuous, it is a unity; second, it is discrete,
a plurality (implicit in continuous magnitude or posited in discrete mag-
nitude) of ones that have equality with one another, the said continuity,
the same unity. Third, this one is also the negation of the many ones as
a simple limit, an excluding of its otherness from itself, a determination
of itself in opposition to other quanta. The one is thus (a) self-referring,
(b) enclosing, and (g) other-excluding limit.
Thus completely posited in these determinations, quantum is number.
The complete positedness lies in the existence of the limit as a plurality
and so in its being distinguished from the unity. Number appears for this
reason as a discrete magnitude, but in unity it has continuity as well. It is,
therefore, also quantum in complete determinateness, for in it the limit is
the determinate plurality that has the one, the absolutely determined, for
its principle. Continuity, in which the one is only implicitly present as a
sublated moment – posited as unity – is the form of the indeterminateness.
Quantum, only as such, is limited in general; its limit is its abstract,
simple determinateness. But as number, this limit is posited as in itself
manifold. It contains the many ones that make up its existence, but does
not contain them in an indeterminate manner, for the determinateness of
the limit falls rather in it; the limit excludes the existence of other ones,
that is, of other pluralities, and those which it encloses are a determinate
aggregate: they are the amount or the how many times 24 with respect to
which, taken as discreteness in the way it is in number, the other is the
unit, the continuity of the amount. Amount and unit constitute the
moments of number.
Regarding amount, we must examine yet more closely how the many
ones in which it consists are in the limit. It is rightly said of amount that it
consists of the many, for the ones are not in it as sublated but are rather present
in it, only posited with the excluding limit to which they are indifferent.
But the limit is not indifferent to them. In the sphere of existence, the
limit was at first so placed in relation to existence that the latter was left on
this side of its limit, standing there as the affirmative, while the limit, the
negation, stood outside on the border of existence; similarly, with respect
to the many ones, their being truncated and the exclusion of the remaining
ones appears in them as a determination that falls outside the enclosed
ones. But it was found in that sphere of existence that the limit pervades
existence, that it extends so far as existence does, and that the something is
for this reason limited by its very determination, that is, is finite. – Now,
in the quantitative sphere, a number, say a hundred, is so represented that
only the hundredth unit brings to the many the limit that makes them a
hundred. In one respect this is correct; but, in another respect, none of the
ones in the hundred has precedence over any other, for they are only equal;
each is just as much the hundredth; they all thus belong to the limit that
makes the number a hundred; this number cannot dispense with any of
them for its determinateness; with respect to the hundredth, therefore, the
rest do not constitute a determinate existence which is in any way different
from it whether inside or outside the limit. Consequently, the number
is not a plurality over against the limiting one that encloses it, but itself
constitutes this delimitation which is a determinate quantum; the many
constitute a number, a one, a two, a ten, a hundred, etc.
Now, the limiting one is a discriminating determinateness, the distin-
guishing of a number from other numbers. But this distinguishing does
not become a qualitative determinateness; it remains rather quantitative,
falling only within the compass of comparing, of external reflection; as
one, number remains turned back onto itself and indifferent to others.
This indifference of number to others is its essential determination; it con-
stitutes its being-determined-in-itself, but at the same time also its own
exteriority. – Number is thus a numerical one that is absolutely determined
but which has at the same time the form of simple immediacy, and to
which, therefore, the connecting reference to an other remains completely
external. Further, as numerical, the one possesses the determinateness (such
as consists in the reference to other) as a moment in it, in its distinction
of unit and amount; and amount is itself the plurality of the ones, that is,
this absolute exteriority is in the one itself. – This intrinsic contradiction
of number or of quantum in general is the quality of quantum, and the
contradiction will develop in the further determinations of this quality.

b. extensive and intensive quantum

a. Their difference

1. We have seen that quantum has its determinateness as limit in amount. In
itself quantum is discrete, a plurality which does not have a being different
from its limit or its limit outside it. Quantum, thus with its limit which as
limit is a plurality, is extensive magnitude.
Extensive magnitude is to be distinguished from continuous magnitude;
its direct opposite is not the discrete, but the intensive magnitude. Exten-
sive and intensive magnitudes are determinacies of the quantitative limit
itself, whereas quantum is identical with its limit; continuous and discrete
magnitudes are, on the contrary, determinations of magnitude in itself, that
is, of quantity as such, in so far as in quantum abstraction is made from
the limit. – Extensive magnitude has the moment of continuity in it and
in its limit, for its many is everywhere continuous; the limit as negation
appears, therefore, in this equality of the many as a limiting of the unity.
Continuous magnitude is quantity that continues without regard to any
limit, and in so far as it is represented with one such limit, the latter is
a limitation in general, without discreteness being posited in it. Determined
as only continuous magnitude, quantum is not yet determined for itself
because the magnitude lacks the one (in which the determinateness-for-
itself lies) and number. Similarly, a discrete magnitude is immediately only
a differentiated many in general which, if it were to have a limit as a many,
would be only an aggregate, that is, would be only indeterminately limited;
in order for quantum to be determinate, the many must be concentrated
into one and thereby be posited as identical with the limit. Continuous
and discrete magnitude, taken as quantum in general, have each posited in
it only one of the two sides by virtue of which quantum is fully determined
and a number. Taken immediately, this latter is extensive quantum – the
simple determinateness which is amount essentially, but the amount of one
and the same unit; extensive quantum is distinguished from number only
because in the latter the determinateness is explicitly posited as plurality.
2. However, the determinateness through number – how much there is
of something – does not require being distinguished from how much thereQuantum
183
is of something else, as if to the determinateness of one thing belonged
how much there is of it and how much there is of an other, for the
determinateness of magnitude as such is a limit determined for itself,
indifferent and simply self-referring; and in number this limit is posited
as enclosed within the one existing for itself: the externality that it has,
the reference to other, is inside it. Further, like the many in general, this
many of the limit is not internally unequal but continuous; each many is
what any other many is; consequently, the many as a many of existents
outside one another, or as discrete, does not constitute the determinateness
as such. Thus this many collapses for itself into its continuity and becomes
simple unity. – Amount is but a moment of number; but, as an aggregate
of numerical ones, it does not constitute the determinateness of number;
on the contrary, these ones as indifferent and self-external are sublated
in number whose being has turned back into itself; the externality that
constituted the ones of plurality vanishes in the one as the self-reference of
number.
The limit of quantum which, as extensive, had its existent determinate-
ness as self-external number, thus passes over into simple determinateness.
In this simple determination of limit, quantum is intensive magnitude; and
the limit or the determinateness which is identical with quantum is now
also posited as simple: it is degree.
Degree is thus a determinate magnitude, a quantum, but at the same
time it is not an aggregate or several within itself; it is only a plurality; 43
plurality is a severality that has gathered together into simple determination,
it is existence that has returned into being-for-itself. It is true that its
determinateness must be expressed by a number, which is the being of the
quantum as completely determined, but the number is not an amount or a
how many times but is rather a onefold, only a degree. When we speak of
10 or 20 degrees, the quantum which has that many degrees – the tenth,
the twentieth degree – is not the amount and sum of the degrees; if that
were the case, it would be an extensive quantum; it is rather only that one
degree, the tenth, the twentieth. It does contain the determinateness found
in the number ten or twenty, but not as several ones: the number is there
as a sublated amount, as a simple determinateness.
3. In number, quantum is posited in its complete determinateness; but
as intensive quantum, in the being-for-itself of number, it is posited as
it is according to its concept or implicitly in itself. For the form of self-
reference which it has in degree is at the same time the externality of degree
to itself. As extensive quantum, number is a numerical plurality and thus
has externality inside it. This externality, as plurality in general, collapses
into a state of undifferentiatedness and is sublated into the numerical
one, the self-reference of number. But quantum has its determinateness as
number; it contains this determinateness, as we have said, whether the latter
is posited in it or not. 44 Degree, therefore, which as internally simple no
longer has this external otherness in it, has it outside it, and refers to it as to its
determinateness. A plurality external to it constitutes the determinateness
of the simple limit which the degree is for itself. In so far as in the extensive
quantum amount was supposed to be found within number, it was sublated
there; now, as thus sublated, amount is posited outside number. Number,
in being posited as a one, as self-reference reflected into itself, excludes from
itself the indifference and the externality of amount and is self-reference as
reference to an external through itself.
In this, quantum has the reality which is adequate to its concept. The
indifference of the determinateness constitutes its quality, that is, a determi-
nateness which is in it as a determinateness external to itself. – Accordingly,
degree is a unitary quantitative determinateness among a plurality of such
intensities which, though diverse, each being only a simple reference to
itself, are at the same time in essential connection with each other, so
that each has its determinateness in this continuity with the others. This
reference connecting a degree through itself to its other makes ascent and
descent on the scale of degrees a continuous progress, a flow, which is
an uninterrupted and indivisible alteration; none of the “more or less” 45
differentiated within it is separate from the others but each has its determi-
nateness only in these others. As a self-referring quantitative determination,
each degree is indifferent towards the others; but, in itself, it equally refers
to this externality; it is what it is only through the intermediary of this
externality; in short, its reference to itself is not an indifferent reference to
externality but in this externality it possesses its quality.
b. Identity of extensive and intensive magnitude
Degree is not inherently external to itself. It is not, however, the indeter-
minate one, is not the principle of number as such which is not amount
44
45
In the preceding Remark.
Mehrern. Here and in what follows Hegel is playing on the root meaning of the German word
for “several” (mehrere) which is “more” (mehr). This wordplay is lost in English. I am using
circumlocutions in an attempt to retain it as much as possible.Quantum
185
except negatively, that is, only in the sense of not being an amount. Inten-
sive magnitude is at first a simple one of many “more or less”; there are several
degrees; but they are not determined either as a unitary one or as a more or
less but only as referring to each other as outside each other, or in the identity
of the one and the “more or less than.” Thus, although the several “more
or less” are as such indeed outside the unitary degree, the determinateness
of the latter lies nonetheless in its connection with them; the degree thus
contains amount. Just as twenty contains as extensive magnitude twenty
ones as discrete, the specific degree contains them as continuity, a continu-
ity which simply is this determinate plurality; it is the twentieth degree, and
it is this twentieth degree only through the intermediate of this amount
which, as such, is outside it.
The determinateness of the intensive magnitude is to be considered,
therefore, from two sides. It is determined through other intensive quanta
and is continuous with its otherness, so that its determinateness consists
in this connection with it. Now, in so far as this determinateness is, first,
a simple determinateness, it is determined as against the other degrees; it
excludes them from itself and has its determinateness in this exclusion.
But, second, it is determined within; this it is in the amount as its amount,
not in the amount as excluded, or not in the amount of the other degrees.
The twentieth degree contains the twenty within itself; it is not only
determined as distinguished from the nineteenth, the twenty-first, etc.,
but its determinateness is rather its amount. But, inasmuch as the amount
is its own, and the determinateness is at the same time essentially as amount,
the degree is extensive quantum.
Extensive and intensive magnitude are, therefore, one and the same
determinateness of quantum; they are distinguished only inasmuch as the
one has the amount within and the other has the same without. Extensive
magnitude passes over into intensive magnitude because its many collapses
in and for itself into oneness and steps outside it. But, conversely, this simple
one has its determinateness only in the amount, its amount; indifferent to
the otherwise determined intensities, it has the externality of amount in it;
thus intensive magnitude is just as essentially extensive magnitude.
With this identity, the qualitative something comes on the scene; for
the identity is the unity that refers back to itself through the negation of
its distinct terms; these terms, however, make up the determinateness of
the existent magnitude. The something is a quantum, but its qualitative
existence is now posited as indifferent to it as it is in itself. One can speak
of quantum, number as such, etc., without any mention of a something as
their substrate. But the something, self-mediated by virtue of the negation
of its determinations, now confronts these as existing for itself, and, since
it has a quantum, it confronts them as something which has an extensive
and intensive quantum. Its one determinateness which it has as quantum
is posited in the distinct moments of unity and amount; it is in itself not
only one and the same determinateness, but the positing of it in these
differences as extensive and intensive quantum is the return into this unity
which, as negative, is the something posited as indifferent to them.

c. Alteration of quantum

The difference between extensive and intensive quantum is indifferent to
the determinateness of quantum as quantum. But quantum is in general
determinateness posited as sublated: the indifferent limit, the determinate-
ness which is just as much the negation of itself. In extensive quantum,
this difference is developed; but intensive magnitude is the existence of
this externality which quantum is in itself. The difference is posited as the
contradiction which it is in itself, of being a simple self-referring determi-
nateness which is the negation of itself, of having its determinateness not
in it but in another quantum.
A quantum, according to its quality, is therefore in absolute continuity
with its externality, with its otherness. Consequently, not only can every
determinateness of magnitude be transcended, not only can it be altered:
that it must alter is now posited. The determination of magnitude continues
into its otherness in such a way that it has its being only in this continuity
with an other; it is not just a limit that exists but one that becomes.
The one is infinite or self-referring negation, and hence the repulsion of
itself from itself. Quantum is equally infinite, posited as the self-referring
negation; it repels itself from itself. But it is a determinate “one,” the
one which has passed over into existence and limit, thus the repulsion
of determinateness from itself, not the generation of something that is
like itself (as the repulsion of the one) but of its otherness; quantum is
now posited in it as sending itself 50 beyond itself. It consists in this, that it
increases or decreases; it is within it the externality of determinateness.
Thus quantum sends itself beyond itself; this other which it becomes
is at first itself a quantum, but a quantum which is not a static limit but
one that impels itself beyond itself. The limit which arises in this beyond
is therefore only one that again sublates itself and sends itself to a further
limit, and so on to infinity.

c. quantitative infinity

a. Its concept

Quantum alters and becomes another quantum; the further determination
of this alteration, that it goes on to infinity, lies in that it is positioned as
inherently self-contradictory. – Quantum becomes an other; but it continues
in its otherness; the other is therefore also a quantum. This latter, however,
is the other, not of a quantum, but of the quantum as such, the negative
of itself as a limited something, and hence its own unlimitedness, infinity.
Quantum is an ought; it implies that it be determined-for-itself, and this
being-determined-for-itself is rather the being determined in an other; and,
conversely, it is the being-determined in an other as sublated, is indifferent
subsisting-for-itself.
In this way, finitude and infinity each at once acquires within it a double
though opposite meaning. Quantum is finite, first, as limited in general;
second, as sending itself beyond itself, as being-determined in an other. On
the other hand, its infinity, is, first, the unlimitedness of quantum; second,
its being-turned-back-into-itself, the indifferent being-for-itself. If we now
compare these moments with each other, we find that the determination
of quantum’s finitude, its sending itself beyond itself into an other that
constitutes its determination, is equally the determination of the infinite;
the negation of limit is this same transcendence of determinateness, so that
in this negation, in the infinite, quantum has its final determinateness.
The other moment of infinity is the for-itself which is indifferent to the
limit; but the quantum itself is so limited, as to be indifferent with respect
to its limit, and hence with respect to other quanta and its “beyond.”
In quantum, finitude and infinity (the latter supposedly separate from
finitude, as bad infinity) each already possesses within it the moment of
the other.
The qualitative and quantitative infinite are distinguished inasmuch as
in the former the opposition of the finite and infinite is qualitative, and the
transition of the finite into the infinite, or the reference of each to the other,
is present only in the in-itself, in their concept. Qualitative determinateness
is immediate; it refers to otherness essentially as to a being which is other
than it; it is not posited as having its negation, its other, in it. By contrast,
magnitude is as such sublated determinateness; it is posited as being unlike
and indifferent to itself, and hence as something alterable. The qualitative
finite and infinite, therefore, stand opposed to each other absolutely, that
is, abstractly; their unity is the inner connection underlying them; hence
the finite continues in its other only in itself, not in it. By contrast, in the
infinite in which the quantitative finite has its absolute determinateness,
this finite refers to itself in it. This, their mutual reference, is first displayed
in the quantitatively infinite process.

b. The quantitative infinite process

The process to infinity is in general the expression of contradiction, here,
of the contradiction contained in the quantitative finite or in quantum in
general. It is the reciprocal determination of the finite and the infinite that
came up for consideration in the sphere of the qualitative, 51 with the differ-
ence that, as just indicated, in the sphere of quantity the limit inherently
sends itself beyond itself and continues there, and hence, conversely, the
quantitative infinite is also posited as having the quantum in it, for in its
externality quantum is itself; its externality belongs to its determination.
The infinite progress is now the expression of this contradiction, not the
resolution of it; however, because of the continuity of one determinateness
in the other, the progress gives rise to the semblance of a resolution in a
union of the two. As at first posited, such a progress is the task of attaining
the infinite but not the attainment of it; it is a perpetual generation of the
infinite, without the progress of ever getting beyond the quantum itself, and
without the infinite ever becoming something which is positively present.
It belongs to the concept of quantum to have a beyond of itself. This
beyond is, first, the abstract moment of the non-being of quantum; this
resolves itself in it; it thus refers to its beyond as to its infinity in accordance
with the qualitative moment of the opposition. But, second, quantum is
continuous with this beyond; it consists precisely in being the other of
itself, external to itself; this externality equally is, therefore, no more an
other than the quantum; the beyond or the infinite is thus itself a quantum.
The beyond is thus recalled from its flight and the infinite is attained. But
because the infinite, now become a “this-side,” is again a quantum, what
is posited is again only a new limit; this limit, as quantum, has also fled
again from itself, is as such beyond itself, and has repelled itself from itself
into its non-being, into its beyond, and as the quantum repels itself into
the beyond, so does the beyond perpetually become a quantum.
The continuity of quantum with its other brings about the conjunction
of the two in the expression of an infinitely great or infinitely small. Since
they both still have in them the determination of quantum, they remain
alterable and the absolute determinateness which would be a being-for-itself
is thus not attained. This being-outside-itself of the determination is posited
in the double infinity (posited in the relative opposition of the “more”
and the “less”) of the infinitely great and the infinitely small. In each, the
quantum is maintained in perpetual opposition to its beyond. No matter
how much the “great” is enlarged, it shrinks to insignificance; since it refers
to the infinite as to its non-being, the opposition is qualitative; the enlarged
quantum has gained nothing, therefore, from the infinite; the latter is its
nothing now just as before. Or again, the increase in the quantum is not
an approximation to the infinite, for the distinction between the quantum
and its infinity essentially has also the moment of being non-quantitative.
This moment is only the sharpened expression of the contradiction that the
quantum ought to be something great, that is, a quantum, and non-finite,
that is, not a quantum. – Equally, the infinitely small is, as something small,
a quantum and therefore remains absolutely, that is, qualitatively, too great
for the infinite and opposed to it. In both, there remains the contradiction
of the infinite progress which in them should have reached its goal.
This infinity, which persists in the determination of the beyond of
the finite, is to be characterized as the bad quantitative infinity. Like the
qualitatively bad infinity, it is the perpetual movement back and forth from
one side of the persistent contradiction to the other, from the limit to its
non-being, and from the latter back again to the other, the limit. To be
sure, the term to which the advance is made in the quantitative progress
is not an abstract “other” in general but a quantum which is explicitly
posited as different; but this quantum remains opposed to its negation in
the same way. Also the progress, therefore, is neither an advance nor a gain
but rather a repetition of one and the same move, a positing, a sublating,
and then again a positing and a sublating: an impotence of the negative
to which what it sublates continuously comes back by its very sublation of
it. The two, the positing and the sublation, are so bonded to each other
that they absolutely flee from each other and yet, in thus fleeing, they are
unable to part but rather become bonded in their very flight from each
other.

c. The infinity of quantum

The infinite quantum as infinitely great or infinitely small is itself, in itself,
the infinite progress; as great or small it is a quantum and at the same time
the non-being of quantum. The infinitely great and the infinitely small are,
therefore, figurative representations which on closer inspection prove to be
but unsubstantial nebulous shadows. In the infinite progress, however, this
contradiction is explicitly present and with it that which constitutes the
nature of quantum which, as intensive magnitude, has attained its reality
and is now posited in its existence as it is in its concept. We must now consider
this identity.
Quantum is as degree simple, self-referred, and determined within it.
Because the otherness and the determinateness are sublated in it through
this simplicity, the determinateness is external to it; it has its determinate-
ness outside it. This, its being-outside-itself, is at first the abstract non-being
of quantum in general, the bad infinity. But further, this non-being is also
a magnitude; quantum continues in its non-being, for it is precisely in
its externality that it has its determinateness, and this, its externality, is
itself therefore equally a quantum; the non-being of quantum, the infinity,
is thus limited, that is, this beyond is sublated, is itself determined as a
quantum which, consequently, in its negation is with itself.
But this is what quantum as such is in itself. For through its externality it
is precisely itself; the externality constitutes that in virtue of which quantum
is quantum, where it is with itself. In the infinite progress, therefore, the
concept of quantum is posited.
If we now first look at this progress in its abstract determinations as
they are displayed before us, what we find in it is the sublating of quantum,
but no less also of its beyond; what we find, therefore, is the negation of
quantum as well as the negation of this negation. Its truth is the unity of
these two negations in which the negations are, but as moments. – This
unity is the resolution of the contradiction of which the infinite progress
is the expression; its most immediate meaning, therefore, is that of the
restoration of the concept of magnitude, of being an indifferent or external
limit. On the subject of the infinite progress as such, the only reflection
which is usually made is that each quantum, however great or small, can
disappear, that it must be possible to transcend it – not, however, that
this sublating of the quantum, the beyond, the bad infinite itself, also
disappears.
Even the first sublating, the negation of quality as such whereby the
quantum is posited, is in itself the sublation of negation – quantum is
sublated qualitative limit, consequently sublated negation – but it is at the
same time only in itself; the sublating is posited as an existence, and its
negation is then fixed as the infinite, as the side beyond quantum, while the
latter remains on its side as an immediate; thus the infinite is determinedQuantum
203
only as first negation and it is in this way that it appears in the infinite
progress. But there is more to it, as has just been shown: there is the negation
of negation or what the infinite is in truth. And this we have just seen with
the restoration of the concept of quantum. Such a restoration means, in
the first place, that to the existence of the quantum there has accrued a
more precise determination. What we now have is quantum determined
according to its concept, and this quantum is different from the immediate
quantum: externality is now the opposite of itself, is posited as a moment of
magnitude; quantum, for its part, is posited as having its determinateness in
another quantum, through the intermediary of its non-being, of infinity,
that is, that it is qualitatively what it is. Yet this comparison of the concept
of quantum with its existence belongs more to our reflection, to a relation
which is not yet present here. The next determination, rather, which is
present here is that quantum has returned to quality, is from now on
qualitatively determined. For its defining property, its quality, is externality,
the indifference of the determinateness; and quantum is now posited rather
as being itself in its externality, of referring to itself therein, of being in
simple unity with itself, that is, of being qualitatively determined. – This
qualitative being is still more closely determined, namely as being-for-itself;
for the very self-reference which quantum has attained has proceeded from
mediation, from the negation of the negation. Quantum no longer has
infinity, the being-determined-for-itself, outside it, but in it.
The infinite, which in the infinite progress only has the empty meaning
of a non-being, of an unattained but sought beyond, is in fact nothing
other than quality. Quantum, as indifferent limit, surpasses itself into the
infinite; it thereby seeks nothing else than its being-determined-for-itself,
the qualitative moment which, however, is only an ought. Its indifference
towards the limit, and hence its lack of a determinateness which is an
existent-for-itself, its surpassing itself, is that which makes the quantum
what it is. This, its surpassing, is to be negated and quantum is to find in
infinity its absolute determinateness.
Quite generally: quantum is sublated quality; but quantum is infinite, it
surpasses itself, is the negation of itself; this, its surpassing, is therefore in
itself the negation of the negated quality, the restoration of it; and what is
posited is that the externality, which seemed to be a beyond, is determined
as quantum’s own moment.
Quantum is thus posited as repelled from itself, and with that there are
two quanta which are however sublated, only moments of one unity, and
this unity is the determinateness of quantum. – Quantum, self-referred as
indifferent limit and hence qualitatively posited, is the quantitative relation
or ratio. – In ratio quantum is external to itself, different from itself; this,
its externality, is the reference connecting a quantum to another quantum,
each quantum acquiring value only in this connection with its other; and
this reference constitutes the determinateness of the quantum which is this
unity. In this unity quantum possesses not an indifferent but a qualitative
determination; in this its externality has turned back into itself; it is in it
what it is.

Further forms associated with the qualitative determinateness of magnitude
The affirmative meaning of the infinitesimal of differential calculus is
that it is the qualitative determinateness of magnitude, and, regarding this
determinateness, we have shown in particular that it is present in this
calculus not only as a power determinateness in general, but as specifically
determining the relation of a function of powers to the power of the
expansion. This qualitative determinateness is also present, however, in
another weaker form, so to speak, and this form, together with the use of
the infinitesimal associated with it and the meaning that the latter has in
this use, will be the subject of the present Remark.
Before we proceed, however, it should be recalled at this juncture that the
different power determinations first come in from the analytical side, that
they are there merely formal and totally homogeneous, that they signify
numerical magnitudes which, as such, are not qualitatively differentiated
from each other. It is in the application to spatial objects that the analytical
relation fully shows itself in its qualitative determinateness as the transition
from linear to plane determinations, from the determinations of straight
n
In the already cited critique (Jahrb. für wissensch. Krit. II, Vol. 1827, Nr. 155, 6ff.), 145 are to be
found interesting opinions of a learned specialist in the subject, Herr Spehr. 146 They are quoted
from his Neue Principien des Fluentencalculs (Braunschweig, 1826), and concern a matter which has
contributed substantially to the issue of obscurities and unscientific claims in differential calculus;
moreover, they accord with what we have said concerning the general orientation of the theory of this
calculus. To quote: “Purely arithmetical investigations – admittedly those which more than all other
have direct bearing on differential calculus – have not been separated from this calculus proper.
Indeed, they have even been taken, as by Lagrange, for what is of the essence of the calculus while
calculus itself has been regarded as only the application of them. These arithmetical investigations
include the rules of differentiation, the derivation of Taylor’s theorems, and so on, even the various
methods of integration. But the case is quite the reverse, for it is these applications that make up the
subject matter of differential calculus proper, which presupposes all those arithmetical developments
and operations from analysis.” – We have shown how, with Lagrange, the separation of the so-
called application from the procedure of the general part which takes its start from the series serves
precisely to bring to light the proper subject of differential calculus. However, it is strange that the
author, despite entertaining the interesting view that it is precisely the so-called applications that
constitute the subject matter of differential calculus proper, would then himself divagate (ibid.)
into the formalistic metaphysics of continuous magnitudes, becoming, flow, etc., and would want to
add new ballast to old. These determinations are formal, in the sense that they are only general
categories which fail to give precisely that which is specific to the subject matter. But this is what was
to be recognized in, and be abstracted from, the concrete propositions, the applications.
Cf. above, 21.267, Hegel’s footnote.
Friedrich Wilhelm Spehr (1799–1833), mathematician and cartographer in Braunschweig, best
known for his Vollständiger Lehrbegriff der reinen Kombinationslehre (1824).Quantum
261
line to those of curves, etc. This application also brings in train that
the spatial objects which are given in the form of continuous magnitudes
in accordance with their nature will be taken discretely, the plane as an
aggregate of lines, the line as an aggregate of points, etc. The sole interest
of this dissolution is to determine the points and the lines into which
the lines and the planes have respectively been resolved in order to be
able, starting from this determination, to proceed analytically, that is, in
truly arithmetical fashion; for the sought magnitude determinations these
starting points are the elements from which the function and the equation
for the concrete, that is, the continuous, magnitude are to be derived. For
the problems where the use of this procedure is of special interest, it is
requisite that the element which is the starting point be self-determined,
in contrast with the indirect procedure which can start only with limits
and seek between them the self-determined element which is its goal.
But if it is only the law of progressively determining magnitudes without
however ever possibly attaining the required perfect, that is, the so-called
finite, determination, then in both methods the result amounts to the
same thing. To Kepler is ascribed the honor of first having thought of
that reversal of procedure and of having made the discrete the starting
point. His explanation of how he understands the first proposition of
Archimedes’ cyclometry conveys this quite simply. 147 This proposition, as
we know, is that the circle is equal to a right-angle triangle, one side of
which enclosing the right angle equals the diameter of the circle and the
other its circumference. Since Kepler takes this proposition to mean that
the periphery of the circle has as many parts as it has points, that is, infinitely
many, each of which can be regarded as the base of an isosceles triangle,
he thereby gives expression to the resolution of the continuous into the
form of the discrete. The expression “infinite” which occurs here is still far
removed from the meaning which it is supposed to assume in differential
calculus. – Now when a determinateness, a function, has been found for
such discrete elements, these should be summed up further, as essentially
elements of the continuous. But since a sum of points will not yield a
line, nor a sum of lines a surface, the points are straight away taken as
already linear, just as the lines are taken as two-dimensional. But again,
since linear elements ought not to be lines yet, which they would be if they
were taken as quantum, they are represented as infinitely small. The discrete
is only capable of an external combination in which the moments retain the
meaning of discrete units; the only transition analytically made from these
147
See note 141, above.
21.301262
21.302
Georg Wilhelm Friedrich Hegel
units is to their sum; nor is there a geometrical transition from point to line,
or from line to surface, etc.; therefore the element which is determined as
a point or a line is also given, together also with this determination, the
quality of being as point linear or as line two-dimensional, in order that
their sum would indeed become, as the sum of tiny lines, a line, and, as
the sum of tiny planes, a plane.
It is the need to get hold of this moment of qualitative transition and
for this reason to have recourse to the infinitely small that must be regarded
as the source of all the representations which, though meant to smooth
out every difficulty, are themselves the greatest difficulty. But to dispense
with these aids, it would have had to be possible to show that contained
even in the analytical procedure, which otherwise appears as a mere sum-
mation, there is in fact already a multiplication. But a fresh assumption
is trotted out at this juncture which provides the basis for this applica-
tion of arithmetical relations to geometrical figures. The assumption is
that arithmetical multiplication is also for the geometrical determination
a transition to a higher dimension, that the arithmetical multiplication of
magnitudes spatially determined as lines is at the same time a promotion of
the linear to the status of two-dimensional determination; three times four
linear feet gives twelve linear feet, but three linear feet times four linear feet
gives twelve plane-feet, that is, square-feet, since the unit in both factors,
as discrete quanta, is the same. The multiplication of lines by lines appears
at first sight to make no sense, for multiplication has to do simply with
numbers, that is, it is an alteration of elements totally homogeneous with
that into which they are altered, with the product, only the quantity alters.
In contrast to this alteration, what has been called the multiplication of
a line as such by a line (it has been named ductus lineæ in lineam, like
the ductus plani in planum; there is also the ductus puncti in lineam) is
an alteration not merely of magnitude, but of magnitude as a qualitative
determination of spatiality, of dimensionality; the transition of the line into
plane is to be understood as the self-externalization of the line, just as the
self-externalization of the point is a line and the self-externalization of the
plane a whole space. This is the same as representing the motion of the
point as being the line, and so on; but since motion entails time determi-
nation, it is more likely to appear in this way of representing it as merely
accidental, an external alteration of state, whereas what is to be captured is
the conceptual determination as expressed by self-externalization, that is,
the qualitative alteration which, arithmetically, is the multiplication of unit
(the point, etc.) by amount (the line, etc.). – It may also be remarked here
that in the case of the self-externalization of the plane, which would appearQuantum
263
as a multiplying of surface by surface, there seems to emerge a difference
of arithmetical and geometrical products, for that self-externalization, as
ductus plani in planum, would arithmetically give a multiplication of a
two-dimensional factor by another factor equally two-dimensional, and
hence a product of four dimensions which, however, is reduced to three
in geometrical determination. For its part, since number has the one as
its principle, it yields the fixed determination for the external, quantita-
tive element; but the product of its operation is to that extent formal.
3 · 3, taken as a numerical determination, when it reproduces itself, yields
3 · 3 × 3 · 3. However, the same magnitude, taken as the determination of a
plane, in reproducing itself is held back at 3 · 3 · 3, because space, imagined
as a movement of transcendence which starts from the point, the limit
which is only abstract, attains its true limit as concrete determinateness in
the third dimension, starting from the line. The difference just referred to
could prove relevant in the context of free motion, where the one side, the
spatial, is subject to geometrical determination (according to Kepler’s law,
x 3 : t 2 ), and the other side, the temporal, to arithmetical determination.
How the qualitative moment being considered here differs from the
subject of the previous Remark, should now be self-evident without fur-
ther comment. In the earlier Remark, the qualitative moment lay in the
determinateness of the powers; here it is the same as the infinitesimal,
except only as a factor arithmetically related to product, or as point to line,
line to plane, etc. The qualitative transition to be made from the discrete,
into which the continuous magnitude is imagined to be resolved, to the
continuous is now effected as a process of summing.
But that in fact the alleged mere summation contains in itself a multi-
plication, therefore the transition from the linear to the two-dimensional
determination, appears in the simplest form in the way in which, for
instance, it is shown that the area of a trapezium is equal to the product
of the sum of the two opposite parallel lines and half the height. This
height is represented as being only the amount of an aggregate of discrete
magnitudes which must be summed up. These magnitudes are lines lying
parallel between the two parallels limiting them. There are infinitely many
of them, for they are supposed to make up the plane. And yet they are lines,
and therefore, in order to be two-dimensional, they must at the same time
be posited with negation. In order to escape the difficulty of a sum of lines
supposedly yielding a plane, the lines are assumed as planes from the start
but also as infinitely thin, for they derive their determination solely from
the linear character of the parallel limits of the trapezium. As parallel and
bounded by the other pair of rectilinear sides of the trapezium, they can
21.303264
21.304
Georg Wilhelm Friedrich Hegel
be represented as the terms of a uniformly differentiated but in principle
indeterminate arithmetical progression whose first and last terms are the
two limiting parallels. The sum of such a series, as we know, is the product
of these parallels and half the amount of the terms. This last quantum is
called “amount” simply and solely with reference to the representation of
infinitely many lines; it is the determinateness as the magnitude in general
of a continuous – the height. It is clear that what is called a “sum” is at
the same time a ductus lineæ in lineam, the multiplying of the linear by the
linear which is, according to the above determination, 148 the emergence of
two-dimensionality. In the simplest case now of any rectangle in general, a
b, each of the two factors is a simple magnitude; but already in the other
still elementary example of the trapezium, only one factor is simple as
one half of the height. The other factor is determined by a progression;
it is still something linear, but its determinateness as magnitude is more
complex; in so far as it can be expressed by a series, the task of summing it
is called analytical, that is, arithmetical; but the geometrical moment in it
is the multiplication, and this is the qualitative moment of transition from
the dimension of line to that of plane; the one factor has been taken as
discrete only for the arithmetical determination of the other; by itself, like
the other, it is the magnitude of something linear.
But the method of representing planes as sums of lines is also commonly
employed where no multiplication as such is being performed in view of
the result. This is the case when the task is to give the magnitude in the
equation not as quantum but as a proportion. There is, for instance, a
well-known way of showing that the area of a circle bears the same relation
to the area of an ellipse, the major axis of which is the diameter of the
circle, as the major axis does to the minor axis, by taking each of these
areas as the sum of the ordinates pertaining to it; each ordinate of the ellipse
stands to the corresponding ordinate of the circle as the minor axis stands
to the major; therefore, it is concluded, also the sums of the ordinates, that
is, the areas, stand in equal proportion. Those who would avoid in this
connection the representation of areas or planes as a sum of lines resort
to the usual, quite superfluous ad hoc aid of making the ordinates into
trapezia of infinitely small width; since the equation is only a proportion,
only one of the two linear elements of the area enters the comparison.
The other element, the abscissa axis, is assumed equal in the ellipse and
the circle – therefore, as a factor of arithmetical magnitude-determination,
equal to 1 – and the proportion is, consequently, totally dependent only
148
Cf. above, 21.301.Quantum
265
on the relation of the one determining moment. The two dimensions are
necessary to the representation of a plane; the determination of magnitude,
however, as is to be given in this proportion, extends only to the one
moment. Therefore, to yield to the representation, or to try to improve
it by adding the representation of sum to this one moment, is in truth a
failure to recognize what mathematical determinateness is all about.
The foregoing discussion contains also the criterion for Cavalieri’s
method of indivisibles referred to above 149 – a method which the crite-
rion also justifies and one that does not need the recourse to the infinitely
small. These indivisibles are lines when Cavelieri considers a plane, or
squares and circles when he considers a pyramid or a cone, etc. He calls the
base line or base plane assumed as determined the regula; it is the constant
and, with reference to a series, its first or last term; the indivisibles are
regarded as parallel to it, therefore in equal determination with respect to
the figure. Now Cavalieri’s general fundamental proposition is that (Exerc.
Geometr. VI; in the later work, Exerc. I, p. 6) “all figures, plane as well as
solid, are proportionate to all their indivisibles, these being compared with
each other collectively and, if there happens to be in them some common
proportion, distributively.” 150 For this purpose, he compares in figures of
equal base line and height the proportions between the lines drawn parallel
to the base and equidistant from it; all such lines of a figure have one and the
same determination and constitute its whole content. In this way Cavalieri
proves, for example, also the elementary proposition that parallelograms
of equal height are proportional to their base; any two lines drawn in the
two figures at equal distance from the base and parallel to it are in the
same proportion as the bases, and so too, therefore, are the whole figures.
In fact, the lines do not constitute the content of the figure as continuous
but constitute it only in so far as they are arithmetically determinable; the
linear figure is the element of the content and through it alone must the
determinateness of the latter be grasped.
With this we are led to reflect on a difference regarding where to place
the determinateness of a figure, whether it is constructed, like the height of
the figure here, or whether it is the external limit. In so far as it is an external
limit, then one concedes that the continuity of the figure follows upon, so
to speak, the equality or the proportion of the limit; e.g. the equality of
congruent figures rests on the fact that the boundaries, the limiting lines,
coincide. But in parallelograms of equal height and base, only this last
determinateness is an external limit; the height on which the second main
149
Cf. above, 21.253.
150
Cavalieri, Exercitationes geometricæ sex (1647), p. 6.
21.305266
21.306
Georg Wilhelm Friedrich Hegel
determination of the figures rests, the proportion of the two, and not the par-
allelism as such, brings a second principle of determination to the external
limits. Euclid’s proof of the equality of parallelograms that have the same
height and base reduces them to triangles, to externally limited continu-
ous figures. In Cavalieri’s proof, first of all concerning the proportionality
of parallelograms, the limit is simply a determinateness of magnitude as
such, which is explicated as taken at every pair of lines drawn at the same
distance from each other in both figures. These lines, equal or in equal
proportion to the base, taken collectively, yield the figures that stand in
equal proportion. The image of an aggregate of lines is incompatible with
the continuity of the figure; consideration of the lines alone fully exhausts
the determinateness which is at issue. Cavalieri has a standard reply to the
supposed difficulty, that the representation of indivisibles requires the
comparison of lines or planes which are infinite in number (Geom. Lib. II,
Prop. I, schol.). 151 He rightly points out that he is not comparing the num-
ber of such lines or planes, which we do not know (that is, it is rather an
empty representation, as just remarked, 152 introduced as prop), but only the
magnitude, that is, the quantitative determinateness as such which is equal
to the space occupied by these lines; because this space is enclosed within
limits, its magnitude is also enclosed within these limits; the continuous is
nothing other than the indivisibles themselves, he says; if it were anything
apart from them, it would not be comparable; but it would be absurd to
say that bounded continuous figures are not comparable with each other.
One can see that Cavalieri’s intention is to distinguish what belongs
to the external existence of the continuous from that in which its deter-
minateness lies and which alone must be stressed when comparing it or
devising theorems about it. The categories which he used for this purpose,
namely the continuous as composed of, or consisting in, indivisibles, and the
like, are admittedly not satisfactory, for implied in them is the claim to
an intuition of the continuous or, as already said, of its external existence;
instead of saying “that the continuous is nothing other than the indivisibles
themselves,” 153 it would have been more correct to say, and consequently
also self-explanatory, that the determinateness of the magnitude of the con-
tinuous is none other than that of the indivisibles themselves. – Cavalieri is
not comfortable with the dubious consequence drawn by the schools from
the representation of the continuous as made up of indivisibles, namely
that there is a greater and smaller infinite; he expressly states further in the
151
152
See Cavalieri, Geometria indivisibilibus continuorum nova (1635). Book 2, Proposition 1, scholium.
153 Cf. 21.305, note 151 above.
Cf. above, 21.304.Quantum
267
treatise (Geom. Lib. VII, Pref.) 154 that he is in no way forced by his method
of proof to represent the continuous as made up of indivisibles; continuous
figures only follow the proportion of the indivisibles. He says that he did not
take up the aggregate of indivisibles in the way they seem to end up in
the determination of the infinite, for the sake of an infinite heap of lines
or planes; he took them rather because of the way they are constituted, for
the specific nature of limitedness that they have in them. Nevertheless, in
order to remove this stumbling block, he does not spare himself the trouble
of demonstrating the main propositions of his geometry, in the seventh
book specially added for this purpose, in a way that remains free of any
admixture of infinity. – This way reduces the proofs to the just cited usual
form of congruence of figures, that is, as we remarked, to the representation
of determinateness as external spatial limit. 155
A further remark that can be made here concerning this form of congru-
ence is that it is no more than, so to speak, a childish aid for the intuition
of the senses. In the elementary propositions concerning the triangles, two
such triangles are represented side by side; on the assumption that, of the
six parts of each, three in the one triangle are equal to the corresponding
three in the other, it is shown that the two triangles are congruent, that
is, that in each the remaining three parts are equal to the corresponding
remaining three in the other – because by virtue of the equality of the first
three parts the two triangles cover each other. In more abstract terms, all that
this equality of corresponding pairs of parts in the two triangles amounts
to is that there is just one single triangle in which three parts are assumed
as already determined, whence follows the determinateness of the remaining
three parts. The determinateness of the one triangle is thus demonstrated
to be completed with the three parts assumed; for the determinateness as
such the remaining three parts are therefore a superfluity – the superfluity of
sensible existence, that is, of the intuition of continuity. When the matter is
expressed in these terms, the qualitative determinateness stands out in its
distinction from what is given to intuition, from the whole as continuous
in itself; congruity does not let this distinction show through.
As we have observed, 156 a new circumstance comes into play in the case of
parallel lines and parallelograms: on the one hand the equality of the angles
only, and on the other the height of the figures from which the external
limits of the latter (the sides of the parallelograms) are distinct. There is an
ambiguity here as to whether in these figures, besides the determinateness
of the side which is one external limit, the base, we should take as the other
154
Ibid.
155
Cf. above, 21.305.
156
Cf. above, 21.304.
21.307268
21.308
Georg Wilhelm Friedrich Hegel
external limit the other side of the parallelogram or the height instead. In
the case of two parallelograms that have the same base and height, but
one is rectangular whereas the other has two angles sharply acute and the
opposite two very obtuse accordingly, intuitively the latter figure can easily
appear greater than the former, for intuitively its given longer side is taken
to be the determinant one and, as in Cavalieri’s figurative method, the planes
are compared according to the aggregate of parallel lines intersecting them;
the longer side could be seen as offering the possibility of more lines than
the vertical side of the rectangle does. This intuitive ambiguity, however,
is no argument against Cavalieri’s method, for the aggregate of parallel
lines imagined in the two parallelograms for the purpose of comparison
equally presupposes the equidistance of the lines from each other or from
the base, from which it follows that the height, and not the other side of
the parallelogram, is the other determining moment. The situation changes
again, however, when the two parallelograms being compared are of equal
height and equal base but do not lie on one plane and make different angles
with a third plane. Here the parallel cross-sections that arise when the third
plane is imagined cutting across the parallelograms in parallel motions are
no longer equidistant from each other and the two planes are unequal.
Cavalieri is very careful to call attention to this difference which he defines
as one between transitus rectus and transitus obliquus of the indivisibles (in
Exercit. I, n. XIIff., and already in Geometr. 1.II), 157 thus staying at the
source any superficial misunderstanding which could arise at this juncture.
I remember Barrow citing in his above mentioned work (Lect. Geom. II,
p. 21) 158 an objection of Taquet 159 – an astute geometer who was also
working at this time on new methods – bearing on precisely this subject.
This Barrow is the one who also used the method of the indivisibles, though
already adulterated by the assumption (which he passed on to his pupil
Newton and to other contemporary mathematicians, Leibniz included)
that a curvilinear triangle, like the so-called characteristic triangle, and a
rectilinear triangle can be said to be equal if both are infinite, that is, very
small. The difficulty raised by Taquet likewise concerns the question of
which line, in the calculation of conic and spherical surfaces, should be
taken as the determining factor in a method based on the application of
the discrete. Taquet’s objection to the method of the indivisibles is that
in the calculation of conic and spheric surfaces, the triangle of the cone
is represented according to this atomistic method as made up of straight
157
159
158 Cf. above, 21.284.
See Cavalieri, Exercitationes geometricæ sex, §§15–17.
Andrew Taquet (1611–1660), Jesuit mathematician at Antwerp.Quantum
269
lines, parallel to the base and perpendicular to the axis, which are at the
same time the radii of the circles of which the surface of the cone consists.
If now this surface is defined as the sum of the circumferences, and this
sum is determined from the number of the radii of these circumferences,
that is, from the length of the axis and the height of the cone, then such
a result would contradict the truth which Archimedes has taught and
demonstrated. To this objection of Taquet, Barrow’s response is that, for
determining the surface, it is not the axis but the side of the triangle of the
cone that must be taken as the line the revolution of which generates the
surface, and this line – not the axis – must therefore be taken as determining
the magnitude for the aggregate of the circumferences.
Objections or uncertainties of this kind have their origin solely in the
vagueness of the representation used for the infinite aggregate of points of
which the line, or of lines of which the plane, etc., supposedly consists; it
is this representation which obscures the essence of the determinateness,
whether of the lines or the planes. – It was the aim of these Remarks to bring
attention to the affirmative determinations of the category of the infinitesi-
mal which, in the various uses made of this category in mathematics, remain
in the background so to speak, and to extract them from the nebulosity in
which they are shrouded when that category is kept at a merely negative
level of determination. In infinite series, as in Archimedes’ cyclometry, the
infinite means no more than that, if the law of the progression of the series
is known, but the so-called finite, that is, the arithmetical expression, is not
given, the reduction of the arc to the straight line cannot be carried out; this
incommensurability is their qualitative difference. Likewise the qualitative
difference between the discrete and the continuous contains in general a
negative determination which makes them appear incommensurable: it
thus introduces the infinite, in the sense that the continuous which is now
to be taken as discrete no longer is supposed to have, as continuous, any
quantum. The continuous, which arithmetically is to be taken as product,
is thereby posited as discrete in itself, that is to say, is broken down into
the elements which are its factors; its magnitude-determinateness lies in
these factors, and they, precisely by virtue of being factors or elements,
are of a lower dimension and, as the determinateness of power goes, of a
lower power than the magnitude of which they are the elements or factors.
This difference appears arithmetically as merely quantitative, a difference
of the root or the power, of whatever degree they might be; however, if
the expression does not go further than just the quantitative sphere as such
(for example, a : a 2 or d a 2 = 2a : a 2 = 2 : a ; or for the law of gravity,
t : at 2 ), it then yields the ratios 1 : a, 2 : a, 1 : at, which say nothing
21.309270
Georg Wilhelm Friedrich Hegel
whatsoever; as against their merely quantitative determination, the sides
would have to be held apart by the qualitative meaning differentiating
them, as s : at 2 , the magnitude being thereby spoken of as a quality, as
a function of the magnitude of another quality. Here then, we are only
conscious of the quantitative determinateness, and it is easy enough, in its
own way, to operate with it. Nor is there anything terrible in multiplying
the magnitude of a line by the magnitude of another line. However, the
multiplication of these magnitudes yields at the same time the qualitative
alteration of the transition of line into plane, and to this extent a negative
determination comes into play. It is this determination that occasions the
difficulty, a difficulty which can be resolved by an insight into its peculiarity
and the simple nature of the matter, but which, when the attempt is made
to eliminate it by the aid of the infinite, only degenerates unresolved into
confusion.

CHAPTER 3

Ratio or the quantitative relation

The infinity of quantum has been determined up to the point where it is
the negative beyond of quantum, a beyond which quantum, however, has
within it. This beyond is the qualitative moment in general. The infinite
quantum, as the unity of the two moments, of the quantitative and the
qualitative determinateness, is in the first instance ratio.
In ratio, quantum no longer has a merely indifferent determinateness but
is qualitatively determined as simply referring to its beyond. It continues
in its beyond, and this beyond is at first just an other quantum. Essentially,
however, the two do not refer to each other as external quanta but each
rather possesses its determinateness in this reference to the other. In this, in
their otherness, they have thus returned into themselves; what each is,
that it is in its other; the other constitutes the determinateness of each. –
The quantum’s self-transcendence does not now mean, therefore, that
quantum has simply changed either into some other or into its abstract
other, into its abstract beyond, but that there, in the other, it has attained
its determinateness; in its other, which is an other quantum, it finds itself.
The quality of quantum, its conceptual determinateness, is its externality
as such, and in ratio quantum is now posited as having its determinateness
in this externality, in another quantum – as being in its beyond what it is.
It is quanta that stand to each other in the connection that has now come
on the scene. This connection is itself also a magnitude; quantum is not only
in relation, but is itself posited as relation; it is a quantum as such that has
that qualitative determinateness in itself. So, as relation (as ratio), quantum
gives expression to itself as self-enclosed totality and to its indifference to
limit by containing the externality of its being-determined in itself: in this
externality it is only referred back to itself and is thus infinite within.
Ratio in general is:
1. direct ratio. In this, the qualitative moment does not yet emerge explicitly
as such; in no other way except still as quantum is quantum posited
as having its determinateness in its externality. – In itself, however, the
quantitative relation is the contradiction of externality and self-reference,
the persistence of quanta and their negation. Such a contradiction next
sublates itself:
2. first inasmuch as in indirect or inverse ratio the negation of each of
the quanta is as such co-posited in the alteration of the other, and the
variability of the direct ratio is itself posited;
3. but in the ratio of powers, the unity, which in its difference refers back
to itself, proves to be a simple self-production of the quantum; this
qualitative moment itself, finally posited in a simple determination and
as identical with the quantum, becomes measure.
– About the nature of the following ratios, much was anticipated in
the preceding remarks concerning the infinity of quantity, that is, the
qualitative moment in it; it only remains to analyze, therefore, the abstract
concept of this ratio.

A. THE DIRECT RATIO

1. In the ratio which, as immediate, is direct, the determinateness of each
quantum lies in the reciprocal determinateness of the other. There is only
one determinateness or limit of both – one which is itself a quantum, namely
the exponent of the ratio.
2. The exponent is some quantum or other; however, in referring itself
to itself in the otherness which it has within it, it is only a qualitatively
determined quantum, for its difference, its beyond and otherness, is in
it. This difference in the quantum is the difference of unit and amount –
the unit, which is the being-determined-for-itself; the amount, which is
the indifferent fluctuation of determinateness, the external indifference of
quantum. Unit and amount were at first the moments of quantum; now,
in the ratio, in quantum as realized so far, each of its moments appears as a
quantum on its own and as determinations of the existence of the quantum,
as delimitations against the otherwise external, indifferent determinateness
of magnitude.
The exponent is this difference as simple determinateness, that is, it
has the meaning of both determinations immediately in it. First, it is a
quantum and thus an amount. If the one side of the ratio which is taken as
unit is expressed in a numerical one, and has only the value of one, then the
other, the amount, is the quantum of the exponent itself. Second, it is simple
determinateness as the qualitative moment of the sides of the ratio. When
the quantum of the one side is determined, the other is also determined
by the exponent and it is a matter of total indifference how the first is
determined; it no longer has any meaning as a quantum determined for
itself but can just as well be any other quantum without thereby altering the
determinateness of the ratio, which rests solely on the exponent. The one
side which is taken as unit always remains unit however great it becomes,
and the other, however great it too thereby becomes, must remain the same
amount of that unit.
3. Accordingly, the two truly constitute only one quantum; the one side
has only the value of unit with respect to the other, not of an amount; and
the other only that of amount. According to their conceptual determinateness,
therefore, they are themselves not complete quanta. But this incompleteness
is in them a negation, and it is so not because of their general variability,
according to which one of them (any of the two) can assume all possible
magnitudes, but because they are so determined that, as one is altered, the
other is increased or decreased in corresponding measure. This means that,
as indicated, only one of them, the unit, is altered as quantum; the other
side, the amount, remains the same quantum of units, and the first side too
retains the value of a unit, however much it is altered as quantum. Each side
is thus only one of the two moments of quantum, and the self-subsistence
which is their proper characteristic is in principle negated; in this qualitative
combination they are to be posited as negative with respect to each other.
The exponent ought to be the complete quantum, since the determi-
nations of both sides come together in it; but in fact, even as quotient the
exponent only has the value of amount, or of unit. There is nothing avail-
able for determining which of the two sides of the relation would have to
be taken as the unit or as the amount; if one side, quantum B, is measured
against quantum A as unit, then the quotient C is the amount of such
units; but if A is itself taken as amount, the quotient C is the unit which is
required by the amount A for the quantum B. As exponent, therefore, this
quotient is not posited for what it ought to be, namely the determinant
of the ratio, or the ratio’s qualitative unity. It is posited as such only to
the extent that its value is that of the unity of the two moments, of unit
and amount. And since these two sides, as quanta, are indeed present as
they should be in the explicated quantum, in the ratio, but at the same
time have the value, which is specific to them as the sides of the ratio, of
being incomplete quanta and of counting only as one of those qualitative
moments, they are to be posited with this negation qualifying them. Thus
there arises a more real ratio, one more in accordance with its definition,
one in which the exponent has the meaning of the product of the sides. In
this determinateness, it is the inverse ratio.

B. THE INVERSE RATIO

1. The ratio as now before us is the sublated direct ratio. It was an immediate
relation and therefore not yet truly determinate; henceforth, the newly
introduced determinateness gives the exponent the value of a product, the
unity of unit and amount. In immediacy, as we have just seen, it was
possible for the exponent to be indifferently taken as unit or amount.
Moreover, it also was only a quantum in general and therefore an amount
by choice. One side was the unit, and this was to be taken as a numerical
one with respect to which the other side would be a fixed amount and at
the same time the exponent. The quality of the latter, therefore, was only
that this quantum is taken as fixed, or rather that the constant only has the
meaning of quantum.
Now in the inverse ratio, the exponent is as quantum likewise immediate,
a quantum or other which is assumed as fixed. But to the one of the
other quantum in the ratio, this quantum is not a fixed amount; the ratio,
previously taken as fixed, is now posited instead as alterable; if another
quantum is taken as the unit of the one side, the other side now no longer
remains the same amount of units of the first side. In direct ratio, this unit
is only the common element of both sides; as such, it continues into the
other side, the amount; the amount itself, or the exponent, is by itself
indifferent to the unit.
But as the determinateness of the ratio now is, the amount as such alters
relative to the unit with respect to which it makes up the other side of the
ratio; whenever another quantum is taken as the unit, that amount alters.
Consequently, although the exponent is still only an immediate quantum
and only arbitrarily assumed as fixed, it does not remain fixed in the side of
the ratio: rather this side, and with it the direct ratio of the sides, is alterable.
In the ratio now before us, the exponent as the determining quantum is
thus posited as negative towards itself as a quantum of the ratio, and hence
as qualitative, as limit; the result is that the qualitative moment distinctly
comes to the fore for itself as against the quantitative moment. – In the
direct ratio, the alteration of the two sides is only the one alteration of the
quantum from which the unity which is the common element is taken; by as
much, therefore, as the one side is increased or decreased, so also is the other;
the ratio itself is indifferent to this alteration and the alteration external
to the ratio. In the indirect ratio, on the contrary, although still arbitrary
according to the moment of quantitative indifference, the alteration is
contained within the ratio, and its arbitrary quantitative extension is limited
by the negative determinateness of the exponent as by a limit.Ratio or the quantitative relation
275
2. We must now consider this qualitative nature of the indirect ratio more
closely, as it is realized, and sort out the entanglement of the affirmative
and the negative moments that are contained in it. – Quantum is posited
as being quantum qualitatively, that is, as self-determining, as displaying
its limit within it. Accordingly, first, it is an immediate magnitude as simple
determinateness; it is the whole as existent, an affirmative quantum. But,
second, this immediate determinateness is at the same time limit; for that
purpose it is distinguished into two quanta which are at first the other of
each other; but as the qualitative determinateness of these quanta, and as
a determinateness which is moreover complete, quantum is the unity of
the unit and the amount, a product of which the two are the factors. Thus
on the one hand the exponent of their ratio is in them identical with itself
and is their affirmative moment by which they are quanta; on the other
hand, as the negation posited in them, it is in them the unity according
to which each, at first an immediate and limited quantum in general, is at
the same time so limited as to be only implicitly in itself identical with its
other. Third, the exponent as the simple determinateness is the negative
unity of this differentiation of it into the two quanta, and the limit of their
reciprocal limiting.
In accordance with these determinations, the two moments limit them-
selves inside the exponent and each is the negative of the other, for the
exponent is their determinate unity; the one moment becomes as many
times smaller as the other becomes greater; each possesses a magnitude of
its own to the extent that this magnitude is in it that of their other, that
is, is the magnitude that the other lacks. The magnitude of each in this
way continues into the other negatively; how much it is in amount, that
much it sublates in the other as amount and is what it is only through
this negation or limit which is posited in it by the other. In this way,
each contains the other as well, and is proportioned to it, for each is sup-
posed to be only that quantum which the other is not; the magnitude of
the other is indispensable to the value of each, and therefore inseparable
from it.
This continuity of each in the other constitutes the moment of unity
through which the two magnitudes stand in relation – the moment of
the one determinateness, of the simple limit which is the exponent. This
unity, the whole, constitutes the in-itself of each from which their given
magnitude is distinct: this is the magnitude according to which each is
only to the extent that it takes from the other a part of their common
in-itself which is the whole. But each can take from the other only as much
as will make it equal to this in-itself; it has its maximum in the exponent276
which, in accordance with the stated second determination, is the limit of
the reciprocal delimitation. And since each is a moment of the ratio only
to the extent that it limits the other and is thereby limited by it, it loses
this, its determination, by making itself equal to its in-itself; in this loss,
the other magnitude will not only become a zero, but itself vanishes, for
what it ought to be is not just a quantum but what it is as such a quantum,
namely only the moment of a ratio. Thus each side is the contradiction
between the determination as its in-itself, that is, as the unity of the whole
which is the exponent, and the determination as the moment of a ratio;
this contradiction is infinity again, in a new form peculiar to it.
The exponent is the limit of the sides of its ratio, within which limit
the sides increase and decrease proportionately to each other; but they
cannot become equal to this exponent because of the latter’s affirmative
determinateness as quantum. Thus, as the limit of their reciprocal limiting,
the exponent is (a) their beyond which they infinitely approximate but can
never attain. This infinity in which the sides approximate their beyond is
the bad infinity of the infinite progression; it is an infinity which is itself
finite, that finds its restriction in its opposite, in the finitude of each side
and of the exponent itself, and for this reason is only approximation. But
(b) the bad infinite is equally posited here as what it is in truth, namely
the negative moment in general, in accordance with which the exponent
is the simple limit as against the distinct quanta of the ratio: it is the in-
itself to which, as the absolutely alterable, the finitude of the quanta is
referred, but which, as the quanta’s negation, remains absolutely different
from them. This infinite, which the quanta can only approximate, is then
equally found affirmatively present on their side: the simple quantum of
the exponent. In it is attained the beyond with which the sides of the
ratio are burdened; it is in itself the unity of the two or, consequently,
in itself the other side of each side; for each side has only as much value
as the other does not have; its whole determinateness thus rests in the
other, and this, their being-in-itself, is as affirmative infinity simply the
exponent.
3. With this, however, we have the transition of the inverse ratio into
a determination other than the one it had at first. This consisted in the
fact that the quantum is immediate but at the same time so connected to
an other that the greater it is, the smaller is the other, that it is what it is
by virtue of negatively relating to the other; also, a third magnitude is the
common restriction on this alteration in magnitude that the two quanta
undergo. This reciprocal alteration, as contrasted to the fixed qualitative

limit, is here their distinctive property; they have the determination of

sum up are not only that this infinite beyond is at the same time some

present finite quantum or other, but that its fixity – which makes it with
respect to the quantitative moment the infinite beyond that it is, and
which is the qualitative moment of being only as abstract self-reference –
has developed itself as a mediation of itself with itself in its other, the
finite moments of the relation. The general point is that the whole is as
such the limit of the reciprocal limiting of the two terms, and that the
negation of the negation (and consequently infinity, the affirmative self-
relation) is therefore posited. The more particular point is that, as product,
the exponent already implicitly is already the unity of unit and amount,
whereas each of these two terms is only one of two moments, and for
this reason the exponent encloses them in itself and in them it implicitly
refers itself to itself. But in the inverse ratio the difference has developed
into the externality of quantitative being, and the qualitative being is not
only something fixed, nor does it simply enclose the two moments of the
ratio immediately in it, but in the externally existent otherness it rejoins itself.
It is this determination that stands out as a result in the moments we
have seen. The exponent, namely, is found to be the implicit being whose
moments are realized in quanta and in their generalized alterability. The
indifference of the magnitudes of these moments in the course of their
alteration displays itself as an infinite progression, the basis of which is
that in their indifference their determinateness is to have their value each
in the value of the other. Thus, (a) according to the affirmative side of
their quantum, the determinateness of the moments is that each is in itself
the whole of the exponent; equally, (b) they have the magnitude of the
exponent for their negative moment, for their reciprocal limiting; their
limit is that of the exponent. The fact that such moments do not have
any other immanent limit, any fixed immediacy, is posited in the infinite
progression of their existence and in their limitation, in the negation of
every particular value. This is, accordingly, the negation of the externality
of the exponent which is displayed in them, and the exponent – itself
equally a quantum as such and also expanded into quanta – is thereby
posited as preserving itself in the negation of the indifferent subsistence of
the moments, as rejoining itself, and thus as the determining factor in this
movement of self-surpassing.
The ratio is hereby determined as the ratio of powers.

C. THE RATIO OF POWERS

1. Quantum, in positing itself as self-identical in its otherness and in
determining its own movement of self-surpassing, has come to be a being-
for-itself. As such a qualitative totality, in positing itself as developed, it
has for its moments the conceptual determinations of number: the unit
and the amount. This last, amount, is in the inverse ratio still an aggregate
which is not determined as amount by the unit itself but from elsewhere,
by a third determinate aggregate; but now it is posited as determined only
by the unit. This is the case in the ratio of powers where the unit, which in
it is amount, is at the same time the amount as against itself as unit. The
otherness, the amount of units, is the unit itself. The power is an aggregate
of units, each of which is this aggregate itself. The quantum, as indifferent
determinateness, changes; but inasmuch as the alteration is the raising to a
power, the otherness of the quantum is determined purely by itself. – The
quantum is thus posited in the power as having returned into itself; it is
immediately itself and also its otherness.
The exponent of this ratio is no longer an immediate quantum, as in the
direct ratio and also in the inverse ratio. In the ratio of powers, the exponent
is of an entirely qualitative nature; it is this simple determinateness: that the
amount is the unit itself, that the quantum is self-identical in its otherness.
And the side of its quantitative nature is to be found in this: that the limit or
negation is not an immediate existent, but that existence is posited rather
as continuing in its otherness. For the truth of quality is precisely to be
quantity, or immediate determinateness as sublated.
2. The ratio of powers appears at first as an external alteration to which a
given quantum is subjected; but it has a closer connection with the concept
of quantum, namely, that in the existence into which the quantum has
developed in the ratio of powers, quantum has attained that concept, has
realized its concept to the fullest. The ratio of powers is the display of
what the quantum is implicitly in itself; it expresses its determinateness
of quantum or the quality by which it is distinguished from another.
Quantum is indifferent determinateness posited as sublated, that is to say,
determinateness as limit, one which is just as much no determinateness,
which continues in its otherness and in it, therefore, remains identical
with itself. Thus is quantum posited in the ratio of powers: its otherness,
the surpassing of itself in another quantum, as determined through the
quantum itself.
If we compare the progressive realization of quantum in the preceding
ratios, we find that quantum’s quality of being the difference of itself from
itself is simply this: that it is a ratio. As the direct ratio, quantum is this
posited difference only in the first instance or immediately, so that the self-
reference which it has as exponent, in contrast to its differences, counts
only as the fixity of an amount of the unit. In the inverse ratio, as negatively
determined, quantum is a relating of itself to itself (to itself as to its negation
in which, however, it has its value); as an affirmative self-reference, it is an
exponent which, as quantum, is only implicitly in itself the determinant of
its moments. But in the ratio of powers quantum is present in the difference
as a difference of itself from itself. This externality of determinateness is the
quality of quantum and is thus posited, in conformity to the concept
of quantum, as quantum’s own determining, as its reference to itself, its
quality.
3. By being thus posited as it is in conformity to its concept, quantum
has passed over into another determination; or, as we can also say, its
determination is now also as the determinateness, the in-itself also as existence.
It is quantum in so far as the externality or the indifference of its determining
(as we say, it is that which can be increased or decreased) is simply accepted
and immediately posited; it has become the other of itself, namely quality,
in so far as that same externality is now posited as mediated by quantum
itself and hence as a moment of quantum, so that in that very externality
quantum refers itself to itself – is being as quality.
At first quantity as such thus appears in opposition to quality; but
quantity is itself a quality, self-referring determinateness as such, distinct
from the determinateness which is its other, from quality as such. Except
that quantity is not only a quality, but the truth of quality itself is quantity,
and quality has demonstrated itself as passing over into it. Quantity, in
its truth, is instead the externality which has returned into itself, which is
no longer indifferent. Thus is quantity quality itself, in such a way that
outside this determination quality as such would yet not be anything at
all. – For the totality to be posited, a double transition is required, not
only the transition of one determinateness into the other, but equally the
transition of this other into the first, its going back into it. Through the
first transition, the identity of the two is present at first only in itself:
quality is contained in quantity, but the latter still is only a one-sided
determinateness. Conversely, that quantity is equally contained in quality,
that it is equally also only as sublated, this results in the second transition,
the going back into the first determinateness. This remark regarding the
necessity of the double transition is everywhere of great importance for
scientific method.

Quantum is henceforth no longer an indifferent or external determina-
tion but is sublated as such, and it is a quality and that by virtue of which
anything is what it is; the truth of quantum is to be measure.
SECTION III

Measure

Abstractly expressed, quality and quantity are in measure united.
Being as such is the immediate equality of determinateness with itself.
This immediacy of determinateness has sublated itself.
Quantity is being that has returned to itself in such a way that it is
a simple self-equality indifferent to determinateness.
But this indifference is only the externality of having
the determinateness not in itself but in an other.
As third, we now have the externality that refers itself to itself;
as self-reference, it is at the same time sublated externality and
carries the difference from itself in it;
a difference which, as externality, is the quantitative moment, and,
as taken back into itself, the qualitative.

Since among the categories of transcendental idealism modality comes
after quantity and quality, with relation inserted in between,
this is an appropriate place to say something about it.
In transcendental idealism, this category has the meaning that
it is the connection of the subject matter to thought.
As understood by that idealism, thought is as such essentially
external to the thing-in-itself.
Hence, inasmuch as the other categories have the
transcendental determination of belonging only to consciousness,
but as its objective moment,
so modality, which is the category of the connection to the subject,
possesses the determination of reflection in itself in a relative sense,
that is to say, the objectivity which is granted to the other categories
is lacking in those of modality;
these, according to Kant’s words, do not add in the least to
the concept as a determination of the object but only express
its relation to the faculty of cognition
(Cr. of Pure R., 2nd edn, pp. 99, 266).
The categories which Kant groups under modality
(possibility, actuality, and necessity)
will come up later in their proper place.
Kant did not apply the form of triplicity
(an infinitely important form even though
with him it occurred only as a formal spark of light)
to the genera of his categories (to quantity, quality, etc.),
but only to their species to which he also gave the name of genera.
He was therefore unable to hit upon the third to quality and quantity.

With Spinoza, the mode is likewise the third after substance and attribute;
Spinoza defines it as the affections of substance, or as that which
is in another through which it is also comprehended.
In this way of conceiving it, this third is externality as such;
as has already been mentioned, with Spinoza generally,
the rigidity of substance lacks the turning back into itself.

The remark just made extends to any of the systems of pantheism which
thought has in one way or other produced.
Being, the one, substance, the infinite, essence, is the first;
opposite this abstraction is the second which
can be mustered in an equally abstract form, as is habitually done as the
next step in any purely formalistic thinking,
namely all determinateness generally taken as the mere finite,
the mere accidental, the transitory, the extraneous and unessential, etc.
But the bond connecting this second with the first is too invasive
for the second not to be not equally grasped with the first;
thus with Spinoza the attribute is the whole of substance,
though as comprehended by the understanding,
which is itself a restriction of substance or mode;
and so the mode, the insubstantial as such which can
be grasped only through an other,
constitutes the opposite extreme of substance, the third.
Also Indian pantheism, taken abstractly, has attained
in its monstrous fantasies this refinement which runs like a moderating
thread across its excesses as its one point of interest
namely that Brahma, the one of abstract thought,
progresses through the shape of Vishnu,
particularly in the form of Krishna, to the third, Shiva.
The determination of this third is that of mode, alteration,
coming-to-be and passing-away;
it is the field of externality in general.
This Indian trinity has tempted a comparison with the Christian, and
one must indeed acknowledge a common element in them.
But it is essential to be aware of the difference that separates them.
It is not just that this difference is infinite but that the
true infinity makes the difference.
The determination of the Indian third principle is that
it is the dispersal of the substantial unity into its opposite,
not its turning back to itself – a spiritual void rather, not spirit.
In the true trinity, there is not only unity but unification;
the syllogism is brought to a unity which is full of content and actual,
a unity which in its totally concrete determination is spirit.
The principle of the mode and of alteration
does not, of course, exclude unity altogether.
In Spinozism, for instance, precisely the mode is as such untrue
while substance alone is what truly is;
everything is supposed to be reduced to substance, and
this is then a sinking of all content into an only formal unity void of content.
As for Shiva, it too is again the great whole,
not distinct from Brahma, from Brahma itself, that is,
the distinction and the determinateness just disappear
without being preserved, without being sublated, and
the unity does not become concrete unity, nor is the diremption reconciled.
The supreme goal of the human being, relegated as he is to
the sphere of coming-to-be and passing-away, of modality in general,
is to sink into unconsciousness, into unity with Brahma, annihilation;
the Buddhist Nirvana, Nibbana, etc., is the same.

Now although the mode is as such abstract externality, indifference
to qualitative as well as quantitative determinations, and nothing in the
essence should depend on the external, the unessential, it is nevertheless
conceded that in the many all depends on the how;
but this is to concede that the mode itself essentially belongs to
the substance of a thing, a very indefinite connection but
one which at least implies that the externality of
the mode is not all that abstract an externality after all.
Here the mode has the definite meaning of being measure.
The Spinozistic mode, just like the Indian principle of alteration,
is the measureless.
The Greeks were aware that everything has a measure.
Parmenides himself introduces necessity after abstract being,
as the ancient limit which is imposed on all.
And this awareness, although still vague, is the beginning of a much
higher concept than is contained in substance and
in the distinction of the mode from it.

Measure in its more developed, more reflected form is necessity.
Fate, Nemesis, ultimately comes down to a determination of measure.
Whatever renders itself beyond the pale, becomes too great, too high,
is brought down to the other extreme of being reduced to nothing,
so that the mean of measure, the medium, is restored.
That the Absolute, God, is the measure of all things, is not
a stronger statement of pantheism than the definition,
“the Absolute, God, is being,” but is infinitely truer.
Measure is indeed an external way of things, a more or less,
but one which is as at the same time reflected into itself,
not merely an indifferent and external determinateness
but one which exists in itself; thus it is the concrete truth of being.
For this reason the nations have revered in it the presence of
something inviolable and sacred.

Already present in measure is the idea of essence,
namely of being self-identical in the immediacy of being determined,
so that this immediacy is reduced through the self-identity to
something mediated, just as the self-identity is equally mediated
only through this externality, but the mediation is one with itself:
this is reflection, the moments of which indeed are, but
in this being are absolutely nothing but moments of their negative unity.
In measure, the qualitative element is quantitative;
the determinateness or the difference is indifferent and
therefore a difference which is none;
it is sublated and this quantitativeness,
as an immanent turning back in which it is qualitative,
constitutes the being-in-and-for-itself which is essence.
But measure is essence only implicitly in itself or in its concept;
this concept of measure is not yet posited.
Measure is as such still the existent unity of the
qualitative and the quantitative element;
its moments are an existence, a quality and some quanta of this quality
which, in themselves, are indeed only indivisible,
but do not yet have the meaning of this reflected determination.
In the development of measure, these moments are differentiated
but at the same time referred to each other,
so that the identity which they are in themselves becomes
their connection explicitly, that is, is posited.
The meaning of this development is the realization of measure in which the
latter posits itself in relation to itself and consequently as moment as well;
through this mediation, measure is determined as sublated; its immediacy
as well as that of its moments disappears; these moments are reflected and
thus measure, having disclosed what it is according to its concept, has
passed over into essence.

Measure is at first the immediate unity of the qualitative and the quan-
titative element, so that it is,

first, a quantum that has qualitative meaning and is as measure. As
so implicitly determined in itself, its further determination is that the
difference of its moments, of its qualitatively and quantitatively determined
being, is disclosed in it. These moments further determine themselves into
wholes of measure which as such are self-subsistent, and, inasmuch as they
refer to each other essentially, measure becomes,

second, a ratio of specific quanta, each an independent measure. But
their self-subsistence also rests essentially on a quantitative relation and a
difference of magnitude, and so the self-subsistence becomes a transition
of one measure into another. The result is that measure collapses into the
measureless. – But this beyond of measure is the negativity of measure only
in itself; thus,

third, the indifference of the determinations of measure is thereby posited,
and measure (real measure because of the negativity contained within it)
is posited as an inverse ratio of measures which, as self-subsistent qualities,
essentially rest on only their quantity and their negative reference to each
other, and consequently prove to be only moments of their truly self-
subsistent unity. This unity is the reflection-into-itself of each and the
positing of them; it is essence.

The development of measure, which we have attempted in what follows,
is among the most difficult of subject matters.
Starting with immediate,
external measure, it would have to proceed, on the one hand, to the further
abstract determination of the quantitative aspect of natural things (of a
mathematics of nature); on the other side, it would have to indicate the
link between this determination of measure and the qualitative aspect of
those things – at least in general, for the detailed demonstration of the
link between the qualitative and the quantitative aspects as they originate
in the concept of a concrete object belongs to the particular science of
the concrete (examples of which, concerning the law of falling bodies and
the free movement of the heavens, will be found in the Encyclopedia of the
Philosophical Sciences).
We may remark quite in general in this connection
that the different forms in which measure is realized also belong to different
spheres of natural reality.
The complete, abstract indifference of developed
measure, that is, of its laws, can only be found in the sphere of mechanism
where concrete corporeity is only abstract matter itself; the qualitative
differences of this matter are of an essentially quantitative nature; space
and time are nothing but pure externalities, and the aggregates of matters,
the masses, the intensity of weight, are determinations which are just as
external and have their proper determinateness in the quantitative element.
On the other hand, in physical things but even more so in the organic, this
quantitative determinateness of abstract materiality is already disturbed by
the multiplicity and consequently the conflict of qualities. And the thus
ensuing conflict is not just one of qualities as such, but measure itself
is subordinated here to higher relations and its immanent development is
reduced rather to the simple form of immediate measure. The limbs of
the animal organism have a measure which, as a simple quantum, stands
in a ratio to the other quanta of the other limbs; the proportions of the
human body are the fixed ratios of such quanta, and the science of nature
still has far to go in discovering anything about the link that connects
these magnitudes with the organic functions on which they are entirely
dependent. But the closest example of the reduction of an immanent
measure to a merely externally determined magnitude is motion. In the
heavenly bodies, motion is free motion, one which is only determined
by the concept from which alone, consequently, its magnitudes equally
depend (see above); but in the organic body this free motion is reduced
to one which is arbitrary or mechanically regular, that is, to one which is
totally abstract and formal.

And in the realm of spirit there is even less of a characteristic, free
development of measure to be found.
For instance, it is obvious that a
republican constitution like the Athenian, or an aristocratic constitution
mixed with democracy, is possible only in a state of a certain size;
it is also obvious that in civil society the multitudes of individuals
who belong to the different occupations stand in a certain ratio to each other.
But none of this yields either laws of measures or typical forms of it.
In the spiritual realm as such there are indeed distinctions of intensity of
character, strength of imagination, sensations, representations, and so on;
but in determining them one cannot go past this indefinite duo of
“strength” and “weakness.”
To see how lame and totally empty ultimately turn out to be
the so-called laws which have been established for
the relation of strength and weakness in sensations, representations,
and so on, one need only look at the psychologies
that busy themselves with just such matters.

CHAPTER 1

Specific quantity

Qualitative quantity is, first, an immediate, specific quantum;
and this quantum, second, in relating itself to another,
becomes a quantitative specifying, a sublating of the indifferent quantum.
This measure is to this extent a rule
and contains the two moments of measure as different;
namely, the quantitative determinateness and
the external quantum as existing in themselves.
In this difference, however, the two sides become qualities, and
the rule becomes a relation of the two; measure presents itself thereby,
third, as a relation of qualities that have one single measure at first;
a measure, however, which further specifies itself in itself into
a difference of measures.

A. THE SPECIFIC QUANTUM

1. Measure is the simple self-reference of quantum, its own determinateness
in itself; quantum is thus qualitative. At first, as an immediate measure
it is an immediate quantum and hence some specific quantum; equally
immediate is the quality that belongs to it; it is some specific quality or
other. – Thus quantum, as this no longer indifferent limit but as self-
referring externality, is itself quality and, although distinguished from it,
it does not extend past it, just as quality does not extend past quantum.
Quantum is thus the determinateness that has returned into simple self-
equality – which is at one with determinate existence just as determinate
existence is at one with it.

If a proposition is to be made of the determination just obtained, it
could be expressed thus: “Whatever is, has a measure.” Every existence has
a magnitude, and this magnitude belongs to the very nature of a something;
it constitutes its determinate nature and its in-itself. The something is not
indifferent to this magnitude, as if, were the latter to alter, it would remain
the same; rather the alteration of the magnitude alters its quality. As measure
the quantum has ceased to be a limit which is none; it is from now on the
determination of a thing, so that, were the latter to exceed or fall short of
this quantum, it would perish. –

A measure, in the usual sense of a standard, is a quantum which is
arbitrarily assumed as the unit determinate in itself as against an external
amount. Of course, such a unit can in fact also be determinate in itself,
like a foot or some such other original measure; to the extent, however,
that it is used as the measuring standard for other things, it is with respect
to them only an external measure, not their original measure. – Thus the
diameter of the earth or the length of a pendulum may be taken as a
specific quantum on their own account. But the choice of a fraction of the
earth’s diameter or of the pendulum’s length, and this last under which
degree of latitude, for use as a standard of measure is arbitrary. And for
other things such a standard is something even more external. These have
further specified the universal specific quantum in some particular way, and
they have thereby been made into particular things. It is therefore foolish
to speak of a natural standard of things. Anyway, a universal standard
is meant for use only as an external comparison of things, and in this
superficial sense of universal standard it is quite a matter of indifference
what is used for the purpose. It is not meant to be a fundamental measure
in the sense that in it the natural measures of particular things would be
displayed and recognized, according to a rule, as the specifications of one
universal measure, the measure of the things’s universal body. But without
this meaning the sole interest and significance of an absolute standard of
measure is that of something common, and any such standard is a universal
not in itself but only by convention.

Immediate measure is a simple determination of magnitude as, for exam-
ple, the size of organic beings, of their limbs, and so forth. But any concrete
existent has the size required for being what it is, and for having existence
in the first place. – As a quantum, the existent is an indifferent magnitude
open to external determination and capable of fluctuating increases and
decreases. As a measure, however, it is at the same time distinct from itself
as quantum, itself as such an indifferent determination, and is a restriction
on that indifferent fluctuation of a limit.
Since in the existence of anything the quantitative determinateness is
thus twofold, in the sense that quality is tied to it and yet the quantity
can fluctuate without prejudice to quality – so the demise of anything that
has a measure occurs through the alteration of its quantum. On the one
hand, the demise appears unexpected, inasmuch as there can be alteration
in the quantum without the measure and the quality being altered; but, on
the other hand, it is made into something quite simple to grasp by means
namely of the concept of gradualness. It is easy to turn to this category for
visualizing or “explaining” the disappearance of a quality or of a something,
for it gives the impression that one can witness this disappearance as if
before one’s eyes: since the quantum is posited as the external limit which
is by nature alterable, the alteration (of quantum only) then follows by
itself. But in fact nothing is thereby explained, for the alteration is at the
same time essentially the transition of one quality into another, or the more
abstract transition of one existence into a non-existence, and therein lies
another determination than just gradualness, which is only a decrease or
increase, and the one-sided holding fast to magnitude.
2. The ancients had already taken notice of this coincidence, that an
alteration which appears to be only quantitative suddenly changes into a
qualitative one, and they used popular examples to illustrate the inconsis-
tencies that arise when such a coincidence is not understood. Two such
examples go under the familiar names of “the bald” and “the heap.” They
are elenchi, that is, according to Aristotle’s explanation, two ways in which
one is compelled to say the opposite of what one has previously asserted. 12
The question was put: does the plucking of one hair from someone’s head
or from a horse’s tail produce baldness, or does a heap cease to be a heap if
one grain is removed? The expected answer can safely be conceded, for the
removal amounts to a merely quantitative difference, and an insignificant
one at that. And so one hair is removed, one grain, and this is repeated
with only one hair and one grain being removed each time the answer is
conceded. At last the qualitative alteration is revealed: the head or the tail
is bald; the heap has vanished. In conceding the answer, it was not only
the repetition that was each time forgotten, but also that the individually
insignificant quantities (like the individually insignificant disbursements
from a patrimony) add up, and the sum constitutes the qualitative whole,
so that at the end this whole has vanished: the head is bald, the purse is
empty.
The embarrassment, the contradiction, produced by the result, is not
anything sophistic in the usual sense of the word, as if the contradiction
were a pretense. The mistake is committed by the assumed interlocutor
(that is, our ordinary consciousness), and that is of assuming a quantity to
be only an indifferent limit, that is, of taking it in the narrowly defined
sense of a quantity. But this assumption is confounded by the truth to
which it is brought, namely that quantity is a moment of measure and
is linked to quality; refuted is the one-sided stubborn adherence to the
abstract determinateness of quantum. – Also those elenchi are, therefore,
not anything frivolous or pedantic but basically correct: they attest to a
mind which has an interest in the phenomena that come with thinking.
Quantum, when it is taken as indifferent limit, is the side from which
an existence is unsuspectedly attacked and laid low. It is the cunning of
the concept that it would seize on an existence from this side where its
quality does not seem to come into play – and it does it so well that the
aggrandizement of a State or of a patrimony, etc., which will bring about
the misfortune of the State or the owner, even appears at first to be their
good fortune.
3. Measure is in its immediacy an ordinary quality of a specific mag-
nitude appropriate to it. Now there is also the distinction between the
side according to which quantum is an indifferent limit that can fluctuate
without the quality altering and the other side according to which it is
qualitative and specific. Both sides are the magnitude determinations of
one and the same thing; but because of the original immediacy of measure,
this distinction too is to be taken as immediate, and accordingly the two
sides each also have a diverse concrete existence. The concrete existence of
measure, which is the side of magnitude determinate in itself, then behaves
towards the concrete existence of the alterable external side by sublating
the indifference of the latter; this is a specifying of measure.

B. SPECIFYING MEASURE

First, this measure is a rule, a measure external to the mere quantum.
Second, it is a specific quantity determining the external quantum.
Third, the two sides, both as qualities of specific quantitative determinacy,
relate to one another as one measure.

a. The rule

The rule, or the standard which we have just mentioned, is in the first
instance as a magnitude which is determinate in itself and is a unit with
respect to a quantum of a particular concrete existence: this is a quantum
with a concrete existence which is other than the something of the rule –
is measured by the latter, that is, is determined as the amount of the said
unit. This comparison is an external act, and the unit itself is an arbitrary
magnitude which can in turn be posited as an amount (the foot as an
amount of inches). But measure is not only an external rule; as a specific
measure its intrinsic nature is that it relates to its other which is a quantum.

b. Specifying measure

Measure is a specific determining of the external magnitude, that is, of the
indifferent magnitude which is now posited in the measuring something by
some other concrete existence in general.
The something of the measure is indeed itself a quantum, but
with the difference that it is the qualitative side
determining the merely indifferent, external quantum.
It has intrinsically this side of being-for-other
to which the fluctuation in size belongs.
The immanent measuring is a quality of the something, and this something
is confronted by the same quality in another something; in the latter,
however, the quality is at first relative, with a measureless quantum in
general as against the something determined as measuring. 14
Inasmuch as a something has an internal measure, an alteration of the
magnitude of its quality comes to it from outside, and the something
does not take on the arithmetical aggregate of the alteration. Its measure
reacts against it, behaves towards the aggregate as an intensive measure and
assimilates it in a way typically its own; it alters the externally imposed
alteration, makes something else out of this quantum and demonstrates
through this specifying function that in this externality it is for-itself. – This
specifically assimilated aggregate is itself a quantum which is also dependent
on the other, that is, on the other aggregate which is only external to it.
The specified aggregate is therefore also alterable, but is not for that reason
a quantum as such but the external quantum as specified in a constant
manner. Measure thus has its determinate existence as a ratio, and its
specificity is in general the exponent of this ratio.
In intensive and extensive quantum, as we saw when considering these
determinations, 15 it is the same quantum which is present, once in the
form of intensity and again in the form of extension. In this difference the
underlying quantum does not suffer any alteration; the difference is only
an external form. In the specifying measure, on the contrary, the quantum
is taken in one instance in its immediate magnitude, but through the
exponent of the ratio it is taken in a second instance in another amount.
The exponent which constitutes the element of specificity can appear
at first to be a fixed quantum, as a quotient of the ratio between the
external and the qualitatively determined quantum. But it would then be
nothing more than an external quantum whereas by the exponent we are to
understand here nothing but the qualitative moment itself that specifies the
quantum as such. The only strictly immanent qualitative determination of
the quantum is (as we saw earlier) 16 that of the exponent, and it must be such
an exponential determination which now constitutes the ratio and which,
as the internally existent determination, comes to confront the quantum as
externally constituted. The principle of this quantum is the numerical one
which constitutes its internal determinateness: the mode of connection of
this numerical one is external, and the alteration which is determined only
through the nature of the immediate quantum as such consists essentially
in the addition of one such numerical one and then another and so forth.
As the external quantum alters in arithmetical progression in this way, the
specifying reaction of the qualitative nature produces another series which
refers to the first, increases and decreases with it, not however in a ratio
determined by a numerical exponent but in a ratio which is numerically
incommensurable, in the manner of a power determination.

c. Relation of the two sides as qualities

1. The qualitative side of the quantum, in itself determined, exists only
as a reference to the external quantitative side; as specifying the latter, it
is a sublating of its externality through which quantum as such is. This
qualitative side thus has a quantum for the presupposition from which it
starts. But this quantum is also qualitatively distinguished from quality,
and this difference between the two must now be posited in the immediacy
of being in general which still characterizes measure. The two sides thus
stand to each other in a qualitative respect, each a qualitative existence for
itself, and the one side that was at first only an internally indeterminate
formal quantum is the quantum of a something and of its quality, and,
just as their reciprocal reference is now determined as measure in general,
so too is the specific magnitude of these qualities. These qualities stand
in relation to each other according to a determination of measure. This
determination is their exponent, but they are already implicitly connected
to each other in the being-for-itself of measure: the quantum is in its double
being external quantum and specific quantum, so that each of the distinct
quantities has this double determination in it and is at the same time
inextricably interwoven with the other; it is in this way alone that the
qualities are determined. They are not, therefore, a determinate being in
general existing for each other but are rather posited as indivisible, and the
specific magnitude tied to them is a qualitative unity – one determination
of measure in which they are implicitly bound together in accordance with
their concept. Measure is thus the immanent quantitative relating of two
qualities to each other.
2. In measure we have the essential determination of variable magnitude,
for measure is sublated quantum – quantum, therefore, no longer as that
which it ought to be in order to be quantum, but quantum as quantum
and as something other besides. This other is the qualitative element and,
as we have established, is nothing else than the relation of powers of the
quantum. In immediate measure, this variability is not yet posited; it
is just one single quantum or other to which a quality is attached. In the
specifying of measure (the preceding determination), which is an alteration
of the merely external quantum by the qualitative moment, what is posited
is the distinctness of two determinate magnitudes and hence in general the
plurality of measures in one common, external quantum. And it is in this
differentiation of the quantum from itself that the latter shows itself for the
first time to be a real measure, for it appears as a being which is one and the
same (e.g. the constant temperature of the medium) and at the same time of
diversified and indeed quantitative existence (in the various temperatures of
the bodies found in the medium). This differentiation of quantum into the
diverse qualities – the different bodies – yields a further form of measure,
one in which the two sides relate to each other as qualitatively determined
quanta, and this can be called the realized measure.
Magnitude, simply as such, is alterable, for its determinateness is a limit
which at the same time is none; the alteration only affects, therefore, a
particular quantum in place of which another is posited. But the genuine
alteration is that of the quantum as such. Here we have the determina-
tion, interesting when so understood, of the variable magnitude of higher
mathematics in which we neither need to stop short at the formal deter-
mination of alteration or variability in general nor introduce any other
determination except the simple determination of the concept by which
the other of the quantum is only the qualitative. The genuine determination,
therefore, of real variable magnitude is that it is qualitative, that is, as we
have sufficiently shown, that it is determined by a ratio of powers. Posited
in this variable magnitude is that quantum has no value as such but only
as determined in conformity with its other, that is, qualitatively.
The two sides in this relating have, in keeping with their abstract side as
qualitites in general, some particular meaning or other, for instance, space
and time. Taken at first simply as determinacies of magnitude in their
ratio of measure, one of them is the amount which increases and decreases
in external arithmetical progression; the other is an amount specifically
determined by the other amount, which for it is the unit. If each of these
two sides were only a particular quality in general, there would be no
way of distinguishing which of them is to be taken with respect to their
determination of magnitude as merely externally quantitative and which as
varying in quantitative specification. If they are related, for instance, as root
and square, it is indifferent which is regarded as increasing or decreasing
in merely external arithmetical progression, and which on the contrary has
its specific determination in this quantum.
But the two qualities are not indeterminate in their difference, for they
are the moments of measure and the qualification of the latter ought to
rest on them. The closest determinateness of the qualitites themselves
is, of the one, the extensive, that it is an externality within; and of the
other, the intensive, that it exists in itself or is the negative as against the
other. Accordingly, amount is the quantitative moment that pertains to the
former, and unit the one that pertains to the latter; in simple direct ratio,
the former is to be taken as the dividend and the latter as the divisor; in
specifying ratio, the former as the power or the becoming-other and the
latter as the root. Inasmuch as we still count here, that is, we reflect on the
external quantum (which is thus the totally accidental determinateness of
magnitude which we call empirical), and accordingly also equally take the
alteration as advancing in external arithmetical progression, then this falls
on the side of the unit, the intensive quality; the external extensive side, by
contrast, is to be represented as altering in the specified series. But the direct
ratio (like velocity as such, st ) is reduced here to a formal determination
which has no concrete existence but belongs rather only to the abstraction
of reflection; and even though in the ratio of root and square (as in s = at 2 ),
the root is to be taken as an empirical quantum varying in an arithmetical
progression, the other side as specified instead, the higher realization of
the qualification of the quantitative moment, one which would be more
in keeping with the concept, is this: that both sides are related in higher
determinations of powers (as in s 3 = at 2 ).

C. THE BEING-FOR-ITSELF IN MEASURE

1. In the form of specified measure just considered, the quantitative moment
of each side is qualitatively determined (both in the ratio of powers); they
are thus moments of one measure-determinateness of qualitative nature.
Here, however, the qualities are still posited immediately, only as diverse;
they are not related in the manner of their quantitative determinacies, that
is, that outside their relation they would have neither meaning nor existence,
as is the case for the quantitative determinacies as a ratio of powers. The
qualitative moment thus disguises itself as specifying, not itself, but the
determinateness of magnitude. Only within the latter is it posited; for
itself it is instead immediate quality as such which, besides the fact that it
posits the magnitude as non-indifferent, and besides its connection with
its other, still has existence subsisting on its own. Thus space and time,
outside that specification which their quantitative determinateness obtains
in the motion of falling bodies or in the absolutely free motion, both have
the value of space in general and time in general, space subsisting on its
own outside and without the duration of time, and time flowing on its
own independently of space.
This immediacy of the qualitative moment as against its specific measure-
relation is, however, equally bound up with a quantitative immediacy and
with the indifference to this same measure-relation of the quantitative ele-
ment in it; the immediate quality also has only an immediate quantum. For
that reason, the specific measure also has a side of an at first external alter-
ation which advances in merely arithmetical progression undisturbed by it
and in which falls the external and hence merely empirical determination
of magnitude. Quality and quantum, even as they extend outside measure,
are at the same time connected with it; the immediacy is a moment of
their belonging to measure. Thus the immediate qualities also belong to
measure, are likewise connected and stand in a ratio which, outside the
specified one of power determination, is itself only a direct ratio and an
immediate measure. This conclusion and its consequences must now be
indicated.
2. Although the immediately determined quantum is as a moment of
measure otherwise implicitly grounded in a conceptual nexus, in connec-
tion with specific measure it is, as such, externally given. But the immediacy
which is thereby posited is the negation of the determination of qualitative
measure; the same immediacy was shown just now on the sides of the deter-
mination of measure which appeared for that reason to be self-subsisting
qualities. Such a negation and the return to immediate quantitative deter-
minateness are present in the qualitatively determined relation because a
relation of distinct terms entails as such the connection of such terms in
one determinateness, and here, in the quantitative sphere, in distinction
from the determination of the relation, this determinateness is a quantum.
As the negation of the distinct, qualitatively determined sides, this expo-
nent is a being-for-itself, an absolutely determined being, but is so only
in itself; as existence, it is a simple immediate quantum, the quotient or
exponent of a ratio between the sides of measure which is taken as direct; as
such, however, it appears in the quantitative side of measure as an empirical
unit. – In the motion of falling bodies, the spaces traversed are proportional
to the square of the elapsed time, s = at 2 . This is a specifically determinate
ratio, one between space and time; the other, the direct ratio, would per-
tain to space and time as mutually indifferent qualities; it is supposed
to be the ratio of the space traversed to the first moment of time. The
same coefficient, a, remains in all the succeeding time-points – the unit
of the amount, determined for its part by the specifying measure, being
an ordinary quantum. This unit counts at the same time as the exponent
of that direct ratio which pertains to the merely imagined, the bad, that is,
the reflectively formal velocity which is not specifically determined by the
concept. Such a velocity does not have concrete existence here, no more
than does the one previously mentioned which is supposed to pertain to a
falling body at the end of a moment of time. That velocity is ascribed to
the first temporal moment in the fall, but this so-called temporal moment
is itself only an assumed unit which has no existence as such an atomic
point. The beginning of the motion (its alleged smallness would make
no difference) is straight away a magnitude, and one which is specified
by the law of falling bodies. The said empirical quantum is attributed to
the force of gravity, and this force thus has itself no connection with the
specification at hand (the determinateness of powers), that is, with the
determinateness characteristic of measure. The immediate moment, that
in the motion of falling bodies the amount of some fifteen spatial units
which are assumed as feet enter into one unit of time (call it a second, the
so-called first unit), is an immediate measure, just like the size of human
limbs, the distances and diameters of planets, etc. The determination of
such measures falls elsewhere than within the qualitative determination of
measure, here of the law itself of falling bodies. On what such numbers
depend, these merely immediate and therefore empirical appearances of a
measure – on this the concrete sciences have yet to give us any information.
Here we are only concerned with this conceptual determination, namely
that in any determination of measure the empirical coefficient constitutes
the being-for-itself, but only the moment of the being-for-itself, in so far
as it is in principle and therefore as immediate. The other moment is the
development of this being-for-itself, the specific measure-determinateness
of the sides. – According to this second moment, in the ratio expressing
the motion of falling bodies (this motion that is still half conditioned and
half free) gravity is to be regarded as a force of nature. Its ratio is thus
determined by the nature of time and space, and the said specification –
the ratio of powers – therefore falls in it; the other moment mentioned
above, the simple direct ratio, expresses only a mechanical relation of time
and space, a reflectively formal velocity externally produced and externally
determined.
3. Measure has now taken on the determination of being a specified
quantitative relation, one which, as qualitative, has within it the usual
external quantum; but this quantum is not a quantum in general but
essentially the determining moment of the relation as such; it is thus an
exponent and, because of the immediacy now of its determinateness, an
invariable exponent, consequently an exponent of the already mentioned
direct ratio of the same qualities whose reciprocal quantitative relation is
at the same time specifically determined by the ratio. In the example of
measure that we have used, that of the motion of falling bodies, this direct
ratio is, as it were, anticipated and assumed as given, but, as remarked, in
this motion it still does not have concrete existence. – However, that matter
is now realized – in that its sides are both measures, one distinguished as
immediate and external and the other as internally specified, while measure
itself is the unity of the two – this constitutes a further determination. As
the unity of its two sides, measure contains the relation in which the mag-
nitudes, by virtue of the nature of the qualities, are posited as determined
and non-indifferent; its determinateness, therefore, is entirely immanent
and self-subsisting, and has at the same time collapsed into the being-for-
itself of an immediate quantum, the exponent of a direct ratio. In this the
self-determination of measure is negated, for in this immediate quantum,
its other, measure has a final self-existent determinateness; conversely, the
immediate measure which ought to be internally qualitative assumes truly
qualitative determinateness only in that other. This negative unity is real
being-for-itself, the category of a something which is the unity of qualities in a
relation of measure – a complete self-subsistence. The two resulting diverse
relations also immediately yield a twofold existence, or, more precisely,
that self-subsisting whole, as a being that exists for itself, is as such in itself
a repulsion into distinct self-subsisting somethings whose qualitative nature
and subsistence (materiality) lies in their determinateness of measure.

CHAPTER 2

Real measure

Measure is now determined as a connection of measures that make up the
quality of distinct self-subsisting somethings, or, in more common lan-
guage, things. The relations of measures just considered belong to abstract
qualities like space and time; further examples of these now to be considered
are specific gravity and then chemical properties, that is, determinations
of concrete material existence. Space and time are also moments of these
measures, but are now subordinated to other determinations and no longer
behave relative to one another only according to their own conceptual
determination. In the case of sound, for instance, the time within which a
certain number of vibrations occur, the spatial width and thickness of the
sounding body, are moments of its determination. But the magnitudes of
such idealized moments are externally determined; they no longer assume
the form of a ratio of powers but relate in the usual direct way, and har-
mony is reduced to the strictly external simplicity of numbers in relations
which are most easy to grasp; they therefore afford a satisfaction which is
the exclusive reserve of the senses, for there is nothing there of representa-
tion, imagery, thought, or the like, that would satisfy spirit. Since the sides
which now constitute the relation of measure are themselves measures, but
at the same time real somethings, their measures are, in the first instance,
immediate measures, and the relations in them direct relations. We now
have to examine the further determination of the relation of such relations.
Measure, now real measure, is as follows.
First, it is the independent measure of a type of body, a measure which
relates to other measures and, in thus relating to them, specifies them as
well as their self-subsistent materiality. This specification, as an external
connecting reference to many others in general, produces other relations,
and consequently other measures; the specific self-subsistence, for its part,
does not remain fixed in one direct relation but passes over into specific
determinacies, and this is a series of measures.
Second, the direct relations that thus result are in themselves determinate
and exclusive measures (elective affinities). 20 But because they are at the
same time only quantitatively different from one another, what we have
is a progression of relations which is in part merely external, but is also
interrupted by qualitative relations, forming a nodal line of specifically self-
subsisting things.
Third, what emerges in this progression for measure, however, is the
measureless: the measureless in general and more specifically the infinitude
of measure in which the mutually exclusive forms of self-subsistence are one
with each other, and anything self-subsistent comes to stand in negative
reference to itself.
a. the relation of independent measures
By measures we no longer mean now merely immediate measures, but
measures that are self-subsistent because they become within themselves
relations which are specified, and in this being-for-itself they are thus a
something, things that are physical and at first material.
(a) However, the whole, which is a relation of such measures is at first
itself immediate. Thus the two sides, which are as such independent
measures, have their subsistence in things external to each other, and
are posited in combination externally.
(b) But the self-subsistent materialities are what they qualitatively are only
in virtue of the quantitative determination that they have as measures,
and are posited in virtue of this same quantitative connection with
others as non-indifferent from them (the so-called affinity); they are the
members of a series of such quantitative relations.
(c) This indifferent multifarious relating concludes at the same time by
cutting itself off as an exclusive being-for-itself – the so-called elective
affinity.
a. Combination of two measures
As a measure relation, something is in itself determined by quanta to
which further qualities accrue; the something is the connection of such
qualities. One of them is the something’s in-itselfness, according to which
This is a term which was introduced around the middle of the eighteenth century to describe the
tendency of chemical species to combine with some substances in preference to others. Goethe
popularized the term by extending its meaning to include human relations, in his novella Elective
Affinities (1809).
it is something that exists for itself, something material (a weight, if taken
intensively; or, taken extensively, an aggregate of parts); the other is however
the externality of this itselfness (something abstract and idealized, space).
These qualities are quantitatively determined, and their mutual relation
is what constitutes the qualitative nature of the material something – the
ratio of weight to volume, the specifically determined gravity. The volume,
the idealized element, is the one to be taken as unit, whereas the intensive
aspect, which in quantitative determinateness and in comparison with
the other appears as extensive magnitude, that is, an aggregate of ones
existing for themselves, is the amount. – The purely qualitative relating of
the two determinacies of magnitude in a ratio of powers has disappeared,
because with the self-subsistence of the being-for-itself (the material being)
immediacy has come back, and in this immediacy the determinateness
of magnitude is a quantum whose relation to the other side is likewise
determined in the ordinary exponent of a direct ratio.
This exponent is the specific quantum of the something, but it is an
immediate quantum and this is determined – as is also the specific nature
of the something – only in being compared with other exponents of like
ratios. The exponent constitutes the specific way in which the something is
determined in itself, the inner measure characteristic of it; but since this,
the something’s measure, rests on a quantum, it is also an external and
indifferent determinateness, and the something, despite its inner deter-
mination as measure, is for this reason subject to alteration. The other to
which it can relate as alterable is not an aggregate of matters, not a quantum
in general; on the contrary, against these it holds out its specific intrinsic
determination of being; rather, it is a quantum which is at the same time
it too the exponent of such a specific ratio. Take two things of different
inner measure that stand connected and enter into composition, say, two
metals of different specific gravity. (In what way the two must otherwise
be homogeneous in nature in order for the composition to be possible, e.g.
that we cannot be speaking here of a metal combining with water, this is not
at issue here). – Now, on the one hand each of the two measures preserves
itself in the alteration they will incur because of the externality of quantum,
for each is a measure; but, on the other hand, this self-preservation is itself
a negative relating to the quantum, a specification of it, and, since this
quantum is the exponent of the ratio of measure, the self-preservation is
an alteration of the measure itself and indeed a reciprocal specification.
According to its merely quantitative determination, the compound
would be the mere sum of the two magnitudes of the one quality and
the two magnitudes of the other quality, e.g. the sum of the two weights
and the two volumes in a compound of two matters of different specific
weights, so that not only would the weight of the mixture remain equal to
the said sum but also the space which the mixture occupies would remain
equal to the sum of the two spaces. But only the weight of the mixture is
found to be in fact the sum of the weights present prior to combination;
the side capable of addition is the one which, as the one existing for itself,
has attained stability of being and consequently a permanent immediate
quantum – the weight of matter, or, what amounts to the same thing from
a quantitative standpoint, the aggregate of material parts. Alteration falls
instead in the exponents, for they are the expression of the qualitative side
of the compound, of its being-for-itself as relations of measure, and, since
the quantum as such is subject to accidental alteration by an increase which
is summed, this being-for-itself proves at the same time to be a negating
factor with respect to this externality. Since this immanent determining of
the quantitative element cannot appear, as we have seen, in the weight, it
turns up in the other quality which is the idealized side of the relation.
It might be surprising to sense-perception that, upon the mixing of two
specifically diverse materials, what results is an alteration of the total vol-
ume, normally a diminution of it, for it is space which gives stability to
matters otherwise existing outside one another. But this stability, in face of
the negativity which is in the being-for-itself, is that which has no existence
in itself, is the alterable as such. In this manner, space is posited as what it
is in truth, an idealization.
But then, not only is one of the qualitative sides posited as alterable but
measure itself, and so also the qualitative determinateness of the something
which is based on it, has shown itself not to be something stable within
but, like quantum in general, to have its determinateness in other measure-
relations.

B. MEASURE AS A SERIES OF MEASURE-RELATIONS

1. If something united with an other were what it is only by virtue of
simple qualitative determination, and the same also applied to the other,
then the two would only sublate themselves in being compounded. But
when something is in itself a measure-relation, it is independent yet at
the same time capable of union with another which is also in itself a
measure-relation. For in being sublated in union, each preserves itself
through the permanence of its quantitative indifference and it behaves as
the specifying moment of a new measure-relation. Its quality is enveloped
in the quantitative element; it is therefore equally indifferent towards the
exponent of the new measure is itself only some quantum or other, an
external determinateness, and its indifference is displayed by the fact that
the specifically determined something runs, with other such measures, into
precisely the same kind of neutralizations of reciprocal measure-relations;
the specific property of the something is not expressed in the one measure-
relation alone which is formed by it and another something.
2. This combination with a plurality of others which are in them-
selves likewise measures yields different ratios that therefore have different
exponents. The measure which is independent has the exponent of its
being-determined-in-itself only in the comparison with other measures;
neutrality with the others constitutes, however, its real comparison with
them; it is its comparison with them through itself. The exponents of these
ratios are however diverse, and the independent measure consequently dis-
plays its qualitative exponent as the series of these different amounts of which
it is the unit – a series of specific ways of relating to others. As one immedi-
ate quantum, the qualitative exponent expresses one single relation. The
independent measure truly differentiates itself in the characteristic series of
exponents which it, taken as a unit, forms with other equally independent
measures; for each of these measures, also brought in connection with
any other and taken as a unit, forms another series. – The relation of
such series among themselves now constitutes the qualitative aspect of the
independent measure.
Now inasmuch as this independent measure produces a series of expo-
nents with a series of such independent measures, it appears at first, when
compared with another independent measure outside this series itself, that
it is differentiated from it by virtue of the fact that this other indepen-
dent measure produces another series of exponents with the same measures
opposite to it. But in this way these two independent measures would not
be comparable – not in so far as each is regarded as a unit with respect to
its exponents, and the two series thus resulting on the basis of this connec-
tion of unit and exponents do not differ in any determinate way. The two
measures which, as independent, are supposed to be compared, are at first
differentiated from each other only as quanta; in order that their ratio be
determined, the ratio requires a unit which exists for itself and is common
to both. This determinate unit is to be sought, as shown, only where the
specific existence of the two independent measures to be compared is to
be found, namely in the ratio which the exponents of the ratio of the
series have to each other. But this ratio of the exponents is itself a unit
existing for itself and in fact determinate only in so far as the members of
the series have it as a constant ratio between them; in that way it can be
their common unit. In it alone, therefore, lies the possibility of comparing
the two independent measures which were assumed not to neutralize each
other but to be rather reciprocally indifferent. Each, taken by itself outside
the comparison, is the unit of the ratios that it establishes with the opposite
members: these are the amounts relative to this unit and hence represent
the series of exponents. But conversely, this series is the unit for the two
independent measures which when compared with each other are as quanta
to each other; as such they are themselves different amounts of their just
indicated unit.
But further, the measures which together with the two or rather indefi-
nitely many reciprocally opposing and contrasting measures yield the series
of the exponents of the ratio of these same measures, are equally inde-
pendent in themselves; each is a specific something with its own relevant
measure-relation. To this extent, they are similarly to be taken each as a
unit so that they have a series of exponents in the two or rather the inde-
terminate plurality of measures, which are first named and compared only
among themselves, and these exponents are the comparative numbers of
precisely these named measures; conversely, the comparative numbers of
the independent measures now taken singly are similarly the series of the
exponents for the members of the first series. In this way, both sides are
a series in which, first, each number is simply a unit with respect to the
opposite series in which it has its specifically determined being as a series
of exponents; second, each number is itself one of the exponents for each
member of the opposite series; and, third, it is a comparative number for
the rest of the numbers of its series and, as such an amount which belongs
to it also as an exponent, it has its unit, determined for itself, in the opposite
series.
3. In this relation there is a return to the mode in which quantum was
posited as existing for itself, namely as degree, as being simple and yet having
the magnitude of its determinateness in a quantum that exists outside it,
one which is a circle of quanta. In measure, however, this externality is not
merely a quantum and a circle of quanta, but a series of numerical ratios, and
it is in the entirety of these that the being-determined-for-itself of measure
lies. Just as it is the case in the being-for-itself of quantum as degree, the
nature of the independent measure has turned into this self-externality.
Its self-reference is at first an immediate relation, and by that very fact its
indifference to an other lies only in the quantum. Hence its qualitative
side falls in this externality, and its relating to the other becomes that which
constitutes the specific determination of this independent measure. Such a
determination thus consists solely in the quantitative mode of this relating,
and this mode is determined just as much by the other as by the measure
itself, and this other is a series of quanta while the measure, for its part,
is one such quantum. But this connection in which two specific measures
specify themselves in a third thing (the exponent) further entails that, in
the exponent, the one measure has not gone over into the other; that there
is not, therefore, only one negation, but that in the connection the two
measures are rather each negatively posited, and since each preserves itself
in it indifferently, their negation is itself also negated. This, their qualitative
unity, is thus an exclusive unit existing for itself. It is only in this moment
of exclusion that the exponents, at first comparative numbers, first gain
in them the determinateness truly specifying them reciprocally and their
difference thus becomes of qualitative nature. But such a difference has a
quantitative basis: first, the independent measure relates to a plurality of its
qualitatively other side only because in this relating it is at the same time
indifferent; second, this neutral connection is in virtue of its quantitative
basis not only alteration but is now posited as a negation of the negation
and as exclusive unit. Consequently, the affinity of an independent measure
with the plurality of measures of the other side is no longer an indifferent
connection but an elective affinity.

c. Elective affinity

The expression elective affinity used here refers to a chemical relation, as
also do the preceding neutrality and affinity. For in the chemical sphere
a matter receives its specific determinateness essentially in the connection
with its other; it exists concretely only as this non-indifference. Moreover,
this specific connection is bound up with quantity and is at the same
time not only a connection with a single other, but with a series of such
non-indifferent matters standing over and against it; combinations with
this series rest on a so-called affinity with every member of the series, even
though in this indifference each member excludes every other. We now
have to consider this connection of opposite determinations. – But it is
not just in the chemical sphere that specificity is exhibited in a circle of
combinations; the meaning of a single note also depends on its relation
to, and combination with, another note and a series of other notes; the
harmony or disharmony in such a circle of combinations constitutes its
qualitative nature which is at the same time based on quantitative ratios
that form a series of exponents, and the ratios of the two specific ratios
is what each of the combined notes is within it. The single note is the
keynote of a system, but at the same time also one member in the system
of every other note. The harmonies are exclusive elective affinities whose
characteristic quality, however, equally dissolves again in the externality of
a mere quantitative progression. – Where to find, however, the principle
of a measure for such affinities which among others and against others
(whether chemical or musical or whatever) are elective affinities, on this
more will be said in the following Remark in connection with chemical
affinity; but this is a question of a higher order which is very closely bound
up with the specific nature of what is strictly qualitative and belongs to
particular parts of the concrete science of nature.
Inasmuch as the member of a series has its qualitative unity in the way it
relates to the whole of an opposite series, the members of which differ from
each other only by virtue of the quantum required for being neutralized
with that member, the more specific determinateness in this manifold affin-
ity is likewise only quantitative. In elective affinity which is an exclusive,
qualitative connection, the relating escapes this quantitative difference. The
next determination which offers itself is this: that in accordance with the
difference in the number, hence the extensive magnitudes, of the members
of one side required for the neutralization of a member in another side, the
elective affinity of this member would be directed to the members of the
first series with all of which it has affinity. Thus transformed, the exclusion
that would thereby be established in the form of a firmer bonding against
other possibilities of combination would appear in a proportionally greater
intensity, in keeping with the previously demonstrated identity of the forms
of extensive and intensive magnitude, 21 inasmuch as the determinateness
of magnitude is in both one and the same. However, this sudden conver-
sion of the one-sided form of extensive magnitude also into its other, the
intensive form, changes nothing in so far as the nature of the fundamental
determination is concerned, which is one and the same quantum. In fact,
therefore, no exclusiveness would thereby be posited, but instead of one
single bonding a combination of any number of members could just as
well take place, provided that the portions of such members entering into
the combination corresponded to the required quantum proportionately
to the ratios between them.
But the combination which we have also called neutralization is not only
the form of intensity; the exponent is essentially a determination of measure
and, as such, exclusive. On this side of exclusive relating, numbers have lost
their continuity and their tendency to combine; their relating is one of more
or less, and this acquires a negative character; and the advantage that one
exponent has over another does not remain confined to the determinateness
of magnitude. Also equally present is the other side which makes it again
a matter of indifference to a moment of measure whether it receives its
neutralizing quantum from several opposite moments, from each according
to the determinateness that specifies it as against an other; the exclusive,
negative relating is at the same time vulnerable to this incursion of the
quantitative side. – What is posited here is thus the sudden conversion
of an indifferent, merely quantitative relating into a qualitative one, and
conversely the transition of a specifically determinate being into merely
external relation – a series of relations, sometimes of a merely quantitative
nature, sometimes specific relations and measures.

B. NODAL LINES OF MEASURE-RELATIONS

The last determination of the measure-relation was that, as specific, it is
exclusive; exclusiveness accrues to neutrality as a negative unity of the dis-
tinct moments. For this unity existing for itself, for this elective affinity, no
further principle has become available specifying its connection with other
neutralities; the specification resides only in the quantitative determination
of affinity in general, according to which it is specific amounts that neu-
tralize themselves and consequently stand opposed to other relative elective
affinities of their moments. But further, because the basic determination is
quantitative, the exclusive elective affinity continues also into these opposite
neutralities, and this continuity is not just an external comparative connec-
tion of the different ratios of the neutralities but the neutrality, as such, has
an element of separability in it, for it is as self-subsisting somethings that
the moments from whose union it has come to be enter into connection
with one or the other of the opposite series: they enter into the connection
indifferently, even though they combine in different specific determinate
amounts. This measure, internally based on such a relation, is for this rea-
son affected by an indifference of its own; it is something which is external
within and alterable in its reference to itself.
This self-reference of the measure-relation differs from the externality and
alterability which belong to its quantitative side. In contrast to these, its
self-reference is an existent qualitative foundation – a permanent, material
substrate which, since it is also the continuity of the measure with itself
in its externality, would have to contain in its quality the said principle of
specification of this externality.
Now in more developed form, as external to itself in its being-for-itself,
the exclusive measure repels itself from itself and posits itself both as another
merely quantitative relation and as another relation which is as such at the
same time another measure; it is determined as in itself a specifying unity
which produces measure-relations within itself. These relations differ from
affinities of the kind mentioned above in which an independent measure
relates itself to independent measures of another quality and to a series of
these. They occur in one and the same substrate within the same moments
of the neutrality; the self-repelling measure takes on the determination of
other only qualitatively diverse relations, and these relations likewise form
affinities and measures, alternating with those that remain only quantitative
diversities. They form in this way a nodal line of measures on a scale of
more and less.
Here we have a measure-relation, a self-subsistent reality qualitatively
distinguished from others. Such a being-for-itself, since it essentially is at
the same time a relation of quanta, is open to externality and quantitative
alteration; it has a margin within which it remains indifferent to this
alteration and does not alter its quality. But there comes a point in the
quantitative alteration at which this quality alters and the quantum shows
itself to be specifying, so that the altered quantitative relation is suddenly
turned into a measure and thereby into a new quality, a new something.
The relation that has now replaced the first is determined by the latter, both
because of the qualitative sameness of the moments which are in affinity and
because of the quantitative continuity. But since the distinction falls on the
side of this quantitative moment, the new something stands indifferently
related to the preceding one; the difference between the two is the external
one of quantum. The new something has not therefore emerged out of the
preceding one; it has emerged rather immediately from itself, that is, from
the internal specifying unity which has not yet entered into existence. –
The new quality or the new something is subjected to the same progression
of alteration, and so on, into infinity.
Inasmuch as the advance from a quality proceeds in the steady con-
tinuity of quantity, the ratios approaching the one qualifying point are
distinguished, quantitatively considered, by a more or less. In this respect,
the alteration is a gradual one. But the gradualness concerns merely the
externality of the alteration, not its qualitative moment; the preceding
quantitative relation, though infinitely near to the succeeding one, is still
another qualitative existence. From the qualitative side, therefore, the grad-
ual, merely quantitative progression which has no limits in itself, is abso-
lutely interrupted; and since in its merely quantitative connection the
newly emerging quality is with respect to the vanishing one an indetermi-
nate other, one which is indifferent to it, the transition is a leap; the two
are posited as wholly external to each other. – It is a favorite practice to
try to make an alteration conceptually comprehensible by the gradualness of
the transition leading up to it; but gradualness is rather alteration precisely
as merely indifferent, the opposite of a qualitative alteration. Rather, in
gradualness the connecting link between two realities – be they states or
self-subsistent things – is sublated, what is posited is that neither reality is
the limit of the other, but that each is absolutely external to the other. Thus,
the very point necessary for the conceptual comprehension of the alteration
is missed, although little enough is required for that purpose.

C. THE MEASURELESS

The exclusive measure, even in its realized being-for-itself, remains affected
by the moment of quantitative existence and is therefore susceptible to the
movement up and down the quantum scale of varying ratios. Something,
or a quality, based on such a ratio is driven beyond itself into the measureless,
collapsing under the mere mutation of its magnitude. Magnitude is that
constitutional aspect where an existence can be caught up in an apparently
harmless entanglement and be destroyed by it.
The abstract measureless is the quantum in general inasmuch as it lacks
internal order and is only indifferent determinateness which does not alter
measure. In the nodal lines of measures this determinateness is at the
same time posited as specifying; the abstract measureless sublates itself
into qualitative determinateness; the new measure-relation into which the
original passes over is measureless with respect to the latter but, in it,
it equally is a quality that exists for itself; what is thus posited is the
alternation of specific existences with one another and equally of them
with relations that still remain merely quantitative – an alternation ad
infinitum. Therefore, present in this transition is both the negation of the
specific relations and the negation of the quantitative progression; this is the
infinite existing for itself. – In the sphere of existence, the qualitative infinite
was the irruption of the infinite into the finite, the immediate transition
and vanishing of the “this here” into its “beyond there.” In contrast to
it, the quantitative infinite is in its very determinateness the continuity
of quantum, its continuing beyond itself. The qualitatively finite becomes
infinite; the quantitatively finite is its beyond in it: it points beyond itself. But
this infinite of the specification of measure posits both the qualitative and the
quantitative as each sublating itself into the other, and it thereby posits their
first immediate unity, which is measure in general, as returned into itself
and consequently as itself posited. The transition of the qualitative, of one
specific concrete existence into another, is such that what happens is only
an alteration of magnitude determinateness; the alteration of the qualitative
as such into the qualitative is thus posited as an external and indifferent
alteration, as a coming together with itself; the quantitative, for its part,
sublates itself by suddenly turning into the qualitative, that is, a being
which is determined in-and-for-itself. This unity which thus continues
in itself in its alternating measures is the self-subsistent matter that truly
persists, the fact.
What we have here is (a) one and the same substantial matter which
is posited as the perennial substrate of its differentiations. This detaching
of being from its determinateness already begins in quantum in general;
something is great indifferently as against the determinateness it has. In
measure, the persisting matter is itself already in itself the unity of the
qualitative and the quantitative – the two moments into which the general
sphere of being is distinguished, each as the beyond of the other; in this
way the perennial substrate begins to possess in it the determination of an
existent infinity. (b) This self-sameness of the substrate is posited in that
the qualitative independent measures into which the determining unit is
dispersed consist of only quantitative differences, so that the substrate per-
sists while being internally distinguished. (g) In the infinite progression of
the nodal series there is posited the continuation of the qualitative into the
quantitative advance as into an indifferent alteration, but equally too, there
is posited the negation of the qualitative contained in the progression and
consequently, at the same time, of the merely quantitative externality. The
quantitative pointing beyond itself to an other which is itself quantitative
perishes with the emergence of a measure-relation, of a quality, and the
qualitative transition is sublated in the very fact that the new quality is
itself only a quantitative relation. This reciprocal transition into the other
of the qualitative and the quantitative moments occurs on the basis of their
unity, and the meaning of this process is only the existence which is the
demonstration or the positing that such a substrate does underlie the process
and is the unity of its moments.
This is a reminder, at the end of the dialectic of being, that the task we have been performing all
along, and are still engaged in at the moment, is to bring stability to the fluidity of becoming.
We first did this with the category of “existence” or the “being this or that” (Dasein). That first
determination has now developed into an objective structure with an internal distinction between
a permanent substrate and all the other determinations introduced so far under the rubrics of
“quality” and “quantity.” This distinction will next develop into the distinction between essence
and being.
chemical materials, bases or alkalis, acids, for instance), and their neu-
tralizations (by which we must understand here also the combinations
of materials of different specific gravities) are in turn self-subsistent and
themselves exclusive measure-relations, mutually indifferent totalities exist-
ing for themselves. Such relations are now determined only as nodes of one
and the same substrate. The measures and the self-subsistent forms posited
with them are consequently demoted to states. Alteration is only the muta-
tion of a state, and that which passes over is posited as remaining the same
in the mutation.
In reviewing the progressive determination which measure has gone
through, the moments of the progression can be summed up as follows.
Measure is at first itself the immediate unity of quality and quantity in the
form of a common quantum which is, however, specific. As a determi-
nateness of quantity which does not refer to an other but refers to itself,
measure is thus essentially a ratio. Consequently, it also holds its moments
as sublated and unseparated in itself; as is always the case in a concept, the
difference in the ratio is such that each of its moments is itself the unity of
the qualitative and the quantitative moment. The difference is thus real,
and it yields a multitude of measure-relations which, as formal totalities,
are each self-subsistent in themselves. The series which form the two sides
of these ratios are for each single member the same constant ordering
by which the member, as belonging to the one side, relates itself to the
whole series standing opposite it. This ordering, although a mere order,
a still wholly external unity, demonstrates itself to be, as the immanently
specifying unity of a measure which has a determinate being-for-itself,
distinguished from its specifications; but this specifying principle is not
yet the free concept, which alone gives its differences an immanent deter-
mination; the principle is at first, rather, only a substrate, a matter, for
whose differences, in order that they be totalities, that is, that they should
have in themselves the nature of their stable, self-equal substrate, there is
only available the external quantitative determination which shows itself
at the same time to be a qualitative differentiation. In this self-unity of
the substrate the determination of measure is a sublated determination, its
quality an external state determined by quantum. – This course is equally
the progressive determination of measure and its demotion to a moment.

CHAPTER 3

The becoming of essence

a. absolute indifference

Being is abstract indifference, and when this trait is to be thought by itself as
being, the abstract expression “indifferentness” has been used – in which
there is not supposed to be as yet any kind of determinateness. Pure quantity
is this indifference in the sense of being open to any determinations,
provided that these are external to it and that quantity itself does not
have any link with them originating in it. The indifference which can
be called absolute, however, is one which, through the negation of every
determinateness of being, of quality and quantity and of their at first
immediate unity, that is, of measure, mediates itself with itself to form a
simple unity. Determinateness is in it still only a state, that is, something
qualitative and external which has the indifference as a substrate.
But that which has thus been determined as qualitative and external
is only a vanishing something; as thus external with respect to being, the
qualitative sphere is the opposite of itself and, as such, only the sublating
of itself. In this way, determinateness is still only posited in the substrate
as an empty differentiation. But it is precisely this empty differentiation
which is the indifference itself as result. And this indifference is indeed
concrete, in the sense that it is self-mediated through the negation of all
the determinations of being. As such a mediation, it contains negation and
relation, and what was called “state” is a differentiation which is immanent
to it and self-referring. It is precisely this externality and its vanishing which
make the unity of being into an indifference: consequently, they are inside
this indifference, which thereby ceases to be only a substrate and, within,
only abstract.

The becoming of essence

b. indifference as inverse ratio of its factors
We now have to see how this determination of indifference is posited in
the indifference itself and the latter is posited, therefore, as existing for itself.
1. The reduction of at first independently accepted measure-relations
establishes their one substrate; this substrate is their continuing into one
another and is, therefore, the one indivisible independent measure which
is wholly present in its differentiation. Present for this differentiation are
the two determinations contained in the measure, quality and quantity,
and everything depends on how these two are posited in it. But this is in
turn determined by the fact that the substrate is at first posited as result
and, though in itself mediation, this mediation is not yet posited as such
in it; for this reason, it is in the first instance substrate and, with respect to
determinateness, indifference.
Consequently, the difference present in it is at first essentially one which
is only quantitative and external; there simply are two different quanta of
one and the same substrate which would thus be their sum, itself posited as a
quantum. But the indifference is this fixed measure, the implicitly existent
absolute limit which, as connected to those differences, would not itself
be in itself a quantum, and would not in any way enter into opposition
with others, whether as sum or also as exponent, be those others sums
or indifferences. It is only the abstract determinateness which falls into
the indifference; the two quanta, in order that they may be posited in it
as moments, are alterable, indifferent, greater or smaller relative to one
another. However, inasmuch as they are restricted by the fixed limit of
their sum, they are at the same time related to each other not externally,
but negatively, and this is now the qualitative determination in which they
stand to each other. Accordingly, they stand in inverse ratio to each other.
This relation differs from the earlier formal inverted ratio 35 inasmuch as
the limit is here a real substrate, and each of the two sides is posited as
having to be in itself the whole.
According to the qualitative determinacy just stated, the difference is
present, further, in the form of two qualities, each of which is sublated
by the other and yet, since the two are held together in the one unity
which they constitute, is inseparable from it. The substrate itself, as the
indifference, is in itself likewise the unity of the two qualities; consequently,
each of the sides of the relation equally contains both sides within itself
and is distinguished from the other by a more of one quality and a less
of the other, or conversely. The one quality, through its quantum, only
predominates on the one side, as does the other quality on the other side.
Thus each side is in it an inverted relation which, as formal, recurs in
the two distinguished sides. The sides themselves thus continue into each
other also according to their qualitative determinations; each of the qual-
ities relates itself in the other to itself and is present in each of the two
sides, only in a different quantum. Their quantitative difference is that
indifference in accordance with which they continue into each other, and
this continuation is the self-sameness of the qualities in each of the two
unities. – The sides, however, each containing the whole of the determi-
nations and consequently the indifference itself, are thus at the same time
posited as self-subsistent vis-a-vis each other.
2. As this indifference, being is now the determinateness of measure no
longer in its immediacy but in the developed manner just indicated; it is
indifference because it is in itself the whole of the determinations of being
now resolved into this unity; and it is existence as well, as a totality of the
posited realization, in which the moments themselves are the totality of
the indifference existing in itself, sustained by the latter as their unity. But
because the unity is held fast only as indifference and consequently only
implicitly in itself, and the moments are not yet determined as existing for
themselves, that is, are not yet determined as sublating themselves into unity
internally and through each other, what is here present is therefore simply the
indifference of the unity itself towards itself as a developed determinateness.
This thus indivisible independent measure is now to be more closely
examined. It is immanent in all its determinations and in them it remains in
unity with itself and undisturbed by them. But, (a) since the determinacies
sublated in it implicitly remain the totality, they emerge in it groundlessly.
The implicit being of indifference and its existence are thus unconnected;
the determinacies show up in the indifference in an immediate manner
and the indifference is in each of them entirely the same. The difference
between them is thus posited at first as sublated, hence as quantitative –
for this reason, therefore, not as a self-repelling; and the indifference not as
self-determining, but as having and becoming the determinate being that
it has only externally.
(b) The two moments are in inverse quantitative relation – a fluctuat-
ing on the scale of magnitude which is not however determined by the
indifference, which is precisely the indifference of the fluctuation, but only
externally. For the determining appeal is made to an other which lies out-
side the indifference. The absolute, as indifference, has in this respect the
second defect of quantitative form, namely that the determinateness of
the difference is not determined by the absolute itself, just as it has the
first defect in that the differences emerge in it only in general, that is, the
positing of them is something immediate, not a self-mediation.
(g) The quantitative determinateness of the moments which are now
sides of the relation constitutes the mode of their subsistence; their existence is
by virtue of this indifference withdrawn from the transitoriness of quality.
But they do have a subsistence of their own in themselves, one that differs
from this quantitative existence, for they are in themselves the indifference
itself, each the unity itself of the two qualities into which the qualitative
moment splits itself. The difference of the two moments is restricted by
the fact that the one quality is posited on the one side with a more and in
the other with a less, and the other is posited in inverse order accordingly.
Each side is thus in it the totality of the indifference. – Each of the two
qualities taken singly for itself likewise remains the same sum which the
indifference is; it continues from one side into the other without being
restricted by the quantitative limit which is thereby posited in it. At this,
the determinations come into immediate opposition, an opposition which
develops into contradiction, as we must now see.
3. Namely, each quality enters inside each side in connection with the
other, and it does so in such a manner that, as has been determined, this
connection also is supposed to be only a quantitative difference. If the
two qualities are both self-subsistent – something like sensible materials
independent of each other – then the whole determinateness of indifference
falls apart; their unity and totality would be empty names. But they are
at the same time determined as comprised into one unity, as inseparable,
each having meaning and reality only in this one qualitative connecting
reference to the other. But now, because their quantitativeness 36 is simply
and solely of this qualitative nature, each reaches only as far as the other. If
they are assumed to differ as quanta, then the one would reach beyond
the other and would have in this more an indifferent existence which the
other would not have. As qualitatively connected, however, each is only in
so far as the other is. – The result is that they are in equilibrium, so that to
the extent that one increases or decreases, the other likewise increases or
decreases and would do so in the same proportion.
On the basis, therefore, of their qualitative connection, there is no
question of a quantitative difference or of a more of the one quality. The
more by which the one of the two connected moments would exceed the
other would be only an unstable determination, or would only be the other
itself again; but, in this equality of the two, neither would then be there,
for their existence would have to rest on the inequality of their quantum. –
Each of these supposed factors vanishes, whether the one factor is assumed
to exceed the other or to be equal to it. From the standpoint of quantitative
representation, the vanishing appears as a disturbance of the equilibrium,
one factor becoming greater than the other; the sublation of the quality
of the other and its instability are thus posited; the first factor becomes
the predominant one as the other diminishes with accelerated velocity and
is overcome by it; this in turn constitutes itself as the one self-subsistent
factor; with this, however, there are no longer two specific moments as
factors but only the one whole.
This unity thus posited as the totality of the process of determining, itself
determined in this process as indifference, is a contradiction all around.
It must therefore be posited as this self-sublating contradiction, and be
determined as subsistence existing for itself, one which no longer has a
merely indifferent unity for result but a unity immanently negative and
absolute. This is essence.

c. transition into essence

Absolute indifference is the final determination of being before the latter
becomes essence; but it does not attain essence. It shows that it still belongs to
the sphere of being because it is still determined as indifferent, and therefore
difference is external to it, quantitative. This externality is its existence, by
which it finds itself at the same time in the opposition of being determined
over against it as existing in itself, 39 not as being thought as the absolute that
exists for itself. Or again, it is external reflection which insists that specific
determinations, whether in themselves or in the absolute, are one and the
same – that their difference is only an indifferent one, not a difference in
itself. What is still missing here is that this reflection should sublate itself,
that it would cease to be the external reflection of thought, of a subjective
consciousness, but that it would be rather the very determination of the
difference of that unity – a unity which would then prove itself to be the
absolute negativity, the unity’s indifference towards itself, towards its own
indifference no less than towards otherness.
But this self-sublation of the determination of indifference has already
manifested itself; in the progressive positing of its being it has shown itself
on all sides to be contradiction. Indifference is in itself the totality in which
all the determinations of being are sublated and contained; thus it is the
substrate, but at first only in the one-sided determination of being-in-itself,
and consequently the differences, the quantitative difference and the inverse
ratio of factors, are present in it as external. As thus the contradiction of
itself and its determinateness, of its implicitly existent determination and of
its posited determinateness, it is the negative totality whose determinacies
have internally sublated themselves – consequently, have also sublated
the one-sidedness of their substrate, their in-itselfness. Indifference, now
posited as what it in fact is, is simple and infinitely negative self-reference,
the incompatibility of itself with itself, the repelling of itself from itself.
Determining and being determined are not a transition, nor an external
alteration, nor again an emergence of determinations in it, but its own
referring to itself which is the negativity of itself, of its in-itselfness.
But as so repelled, the determinations are not self-possessed – do not
emerge as self-subsistent or external but are rather as moments: first, as
belonging to the unity whose existence is still only implicit, they are not
let go by it but are rather borne by it as their substrate and are filled by
it alone; and, second, as determinations immanent to the unity as it exists
for itself, they are only through their repulsion from themselves. Instead of
some existent or other, as they are in the whole sphere of being, they now are
simply and solely as posited, with the sole determination and significance of
being referred to their unity and hence each to the other and to negation –
marked by this their relativity.
Being in general and the being or immediacy of the different deter-
minacies have thereby vanished just as much as the in-itselfness, and the
unity is being, immediately presupposed totality, so that it is only this simple
self-reference, mediated by the sublation of this presupposition, and this pre-
supposedness, the immediate being, is itself only a moment of its repelling:
the original self-subsistence and self-identity are only as the resulting infinite
self-rejoining. And so is being determined as essence – being which, through
the sublation of being, is simple being with itself.
