SECTION II

Magnitude (Quantity)

The difference between quantity and quality has been indicated. Quality
is the first, immediate determinateness. Quantity is the determinateness
that has become indifferent to being; a limit which is just as much no
limit; being-for-itself which is absolutely identical with being-for-another:
the repulsion of the many ones which is immediate non-repulsion, their
continuity.
Because that which exists for itself is now so posited that it does not
exclude its other but rather affirmatively continues in it, it is then other-
ness, inasmuch as existence surfaces again on this continuity and its deter-
minateness is at the same time no longer simple self-reference, no longer
the immediate determinateness of the existent something, but is posited
as repelling itself from itself, as referring to itself in the determinateness
rather of an other existence (a being which exists for itself ); and since they
are at the same time indifferent limits, reflected into themselves and uncon-
nected, determinateness is as such outside itself, an absolute externality and
a something just as external; such a limit, the indifference of the limit as
limit and the indifference of the something to the limit, constitutes the
quantitative determinateness of the something.
In the first place, we have to distinguish pure quantity from quantity
as determinate, from quantum. First, pure quantity is real being-for-itself
turned back into itself, with as yet no determinateness in it: a compact,
infinite unity which continues itself into itself.
Second, this quantity proceeds to determinateness, and this is posited
in it as a determinateness that at the same time is none, is only exter-
nal. Quantity becomes quantum. Quantum is indifferent determinateness,
that is, one that transcends itself, negates itself; as this otherness of other-
ness, it lapses into infinite progress. Infinite quantum, however, is sublated
indifferent determinateness: it is the restoration of quality.
Third, quantum in qualitative form is quantitative ratio. Quantum tran-
scends itself only in general; in the ratio, however, it transcends itself
into its otherness, in such a way that this otherness in which it has its
determination is at the same time posited, is another quantum. With this
we have quantum as turned back into itself and referring to itself as into
its otherness.
At the foundation of this relation there still lies the externality of quan-
tum; it is indifferent quanta that relate themselves to each other, that is, they
have the reference that mutually connects them in this being-outside-itself.
The ratio is, therefore, only a formal unity of quality and quantity, and its
dialectic is its transition into their absolute unity, in measure.

CHAPTER 1

Quantity

a. pure quantity

Quantity is sublated being-for-itself. The repelling one that behaved only
negatively towards the excluded one, now that it has gone over in connec-
tion with it, behaves towards the other as identical to itself and has there-
fore lost its determination; being-for-itself has passed over into attraction.
The absolute obduracy of the one has melted away into this unity which,
however, as containing the one, is at the same time determined by the
repulsion residing in it; as unity of the self-externality, it is unity with itself.
Attraction is in this way the moment of continuity in quantity.
Continuity is therefore simple, self-same reference to itself unbroken by
any limit or exclusion – not, however, immediate unity but the unity of
ones which have existence for themselves. Still contained in it is the outside-
one-another of plurality, though at the same time as something without
distinctions, unbroken. Plurality is posited in continuity as it implicitly is
in itself; the many are each what the others are, each is like the other, and the
plurality is, consequently, simple and undifferentiated equality. Continuity
is this moment of self-equality of the outsideness-of-one-another, the self-
continuation of the different ones into the ones from which they are
distinguished.
In continuity, therefore, magnitude immediately possesses the moment
of discreteness – repulsion as now a moment in quantity. – Steady continuity 1
is self-equality, but of many that do not become exclusive; it is repulsion
that first expands self-equality to continuity. Hence discreteness is, for its
part, a discreteness of confluents – of ones that do not have the void to
connect them, not the negative, but their own steady advance and, in the
many, do not interrupt this self-equality.
Quantity is the unity of these moments, of continuity and discreteness.
At first, however, it is this continuity in the form of one of them, of
continuity, as a result of the dialectic of the being-for-itself which has col-
lapsed into the form of self-equal immediacy. Quantity is as such this simple
result in so far as the being-for-itself has not yet developed its moments and
has not posited them within it. – Quantity contains these moments at first
as being-for-itself posited in its truth. It was the determination of being-in-
itself to be self-sublating self-reference, a perpetual coming-out-of-itself.
But what is repelled is itself; repulsion is thus a creative flowing away
from itself. On account of the sameness of what is repelled, this discerning
is unbroken continuity; and on account of the coming-out-of-itself, this
continuity is at the same time, without being broken off, a plurality – a
plurality which persists just as immediately in its equality with itself.

b. continuous and discrete magnitude

1. Quantity contains the two moments of continuity and discreteness. It
is to be posited in both, in each as its determination. – It is already from
the start the immediate unity of the two, that is, quantity is itself posited at
first only in one of the two determinations, that of continuity, and as such
is continuous magnitude.
Or continuity is indeed one of the moments of quantity which is brought
to completion only with the other, discreteness. But quantity is concrete
unity only in so far as it is the unity of distinct moments. These are to
be taken, therefore, also as distinct, without however resolving them again
into attraction and repulsion but, rather, as they truly are, each remaining
in its unity with the other, that is, remaining the whole. Continuity is only
the compact unity holding together as unity of the discrete; posited as such,
it is no longer only moment but the whole quantity: continuous magnitude.
2. Immediate quantity is continuous magnitude. Quantity, however, is
not as such an immediate; immediacy is a determinateness, the sublated
being of which is precisely quantity. Quantity is to be posited, therefore,
in the determinateness immanent to it, and this is the one. Quantity is
discrete magnitude.
Discreteness is, like continuity, a moment of quantity, but is itself also
the whole quantity just because it is a moment in it, the whole, and
therefore as distinct moment does not diverge from its unity with the
other moment. – Quantity is the outsideness-of-one-another as such, and
continuous magnitude is this outsideness-of-one-another onwardly posit-
ing itself without negation as an internally self-same connectedness. On
the other hand, discrete magnitude is this outsideness-of-one-another as
discontinuous, as broken off. With this aggregate of ones, however, the
aggregate of atom and void, repulsion in general, is not thereby reinstated.
Because discrete magnitude is quantity, its discreteness is itself continuous.
Such a continuity in the discrete consists in the ones being the same as
one another, or in that they have the same unity. Discrete magnitude is
therefore the one-outside-the-other of the many ones as of a same – not the
many ones in general, but posited rather as the many of a unity.

c. the limiting of quantity

Discrete magnitude has, first, the one for its principle and, second, is a
plurality of ones; third, it is essentially continuous, it is the one as at the
same time sublated, as unity, the self-continuing as such in the discreteness
of the ones. Consequently, it is posited as one magnitude, and the “one”
is its determinateness – a “one” which, in this posited and determinate
existence, excludes, is a limit to the unity. Discrete magnitude as such is not
supposed to be immediately limited, but, when distinguished from contin-
uous magnitude, it is an existence and a something, the determinateness
of which, and in it also the first negation and limit, is the “one.”
This limit, besides referring to the unity and being the moment of
negation in it, is also, as one, self-referred; thus it is enclosing, encompassing
limit. The limit here is not at first distinct from the something of its
existence, but, as one, is essentially this negative point itself. But the being
which is here limited is essentially as continuity, and in virtue of this
continuity it transcends the limit, transcends this one, and is indifferent to
it. Real, discrete quantity is thus one quantity, or quantum – quantity as
an existence and a something.
Since the one which is a limit encompasses within it the many ones
of discrete quantity, it posits them equally as sublated in it; it is a limit
to continuity simply as such and, consequently, the distinction between
continuous and discrete magnitude is here indifferent; or, more precisely,
it is a limit to the continuity of the one just as much as of the other; in this,
both pass over into being quanta.
Quantum, which in the first instance is quantity with a determinateness or
limit in general, in its complete determinateness is number. Second, quan-
tum divides first into extensive quantum, in which limit is the limitation of a
determinately existent plurality; and then, inasmuch as the existence of this
plurality passes over into being-for-itself, into intensive quantum or degree.
This last is for-itself but also, as indifferent limit, equally outside itself. It thus
has its determinateness in an other. Third, as this posited contradiction of
being determined simply in itself yet having its determinateness outside
itself and pointing outside itself for it, quantum, as thus posited outside
itself within itself, passes over into quantitative infinity.

A. NUMBER

Quantity is quantum, or has a limit, both as continuous and discrete
magnitude. The distinction between these two species has here, in the first
instance, no significance.
As the sublated being-for-itself, quantity is already in and for itself
indifferent to its limit. But, equally, the limit or to be a quantum is not
thereby indifferent to quantity; for quantity contains within itself as its
own moment the absolute determinateness of the one, and this moment,
posited in the continuity or unity of quantity, is its limit, but a limit which
remains as the one that quantity in general has become.
This one is therefore the principle of quantum, but as the one of quantity.
For this reason it is, first, continuous, it is a unity; second, it is discrete,
a plurality (implicit in continuous magnitude or posited in discrete mag-
nitude) of ones that have equality with one another, the said continuity,
the same unity. Third, this one is also the negation of the many ones as
a simple limit, an excluding of its otherness from itself, a determination
of itself in opposition to other quanta. The one is thus (a) self-referring,
(b) enclosing, and (g) other-excluding limit.
Thus completely posited in these determinations, quantum is number.
The complete positedness lies in the existence of the limit as a plurality
and so in its being distinguished from the unity. Number appears for this
reason as a discrete magnitude, but in unity it has continuity as well. It is,
therefore, also quantum in complete determinateness, for in it the limit is
the determinate plurality that has the one, the absolutely determined, for
its principle. Continuity, in which the one is only implicitly present as a
sublated moment – posited as unity – is the form of the indeterminateness.
Quantum, only as such, is limited in general; its limit is its abstract,
simple determinateness. But as number, this limit is posited as in itself
manifold. It contains the many ones that make up its existence, but does
not contain them in an indeterminate manner, for the determinateness of
the limit falls rather in it; the limit excludes the existence of other ones,
that is, of other pluralities, and those which it encloses are a determinate
aggregate: they are the amount or the how many times 24 with respect to
which, taken as discreteness in the way it is in number, the other is the
unit, the continuity of the amount. Amount and unit constitute the
moments of number.
Regarding amount, we must examine yet more closely how the many
ones in which it consists are in the limit. It is rightly said of amount that it
consists of the many, for the ones are not in it as sublated but are rather present
in it, only posited with the excluding limit to which they are indifferent.
But the limit is not indifferent to them. In the sphere of existence, the
limit was at first so placed in relation to existence that the latter was left on
this side of its limit, standing there as the affirmative, while the limit, the
negation, stood outside on the border of existence; similarly, with respect
to the many ones, their being truncated and the exclusion of the remaining
ones appears in them as a determination that falls outside the enclosed
ones. But it was found in that sphere of existence that the limit pervades
existence, that it extends so far as existence does, and that the something is
for this reason limited by its very determination, that is, is finite. – Now,
in the quantitative sphere, a number, say a hundred, is so represented that
only the hundredth unit brings to the many the limit that makes them a
hundred. In one respect this is correct; but, in another respect, none of the
ones in the hundred has precedence over any other, for they are only equal;
each is just as much the hundredth; they all thus belong to the limit that
makes the number a hundred; this number cannot dispense with any of
them for its determinateness; with respect to the hundredth, therefore, the
rest do not constitute a determinate existence which is in any way different
from it whether inside or outside the limit. Consequently, the number
is not a plurality over against the limiting one that encloses it, but itself
constitutes this delimitation which is a determinate quantum; the many
constitute a number, a one, a two, a ten, a hundred, etc.
Now, the limiting one is a discriminating determinateness, the distin-
guishing of a number from other numbers. But this distinguishing does
not become a qualitative determinateness; it remains rather quantitative,
falling only within the compass of comparing, of external reflection; as
one, number remains turned back onto itself and indifferent to others.
This indifference of number to others is its essential determination; it con-
stitutes its being-determined-in-itself, but at the same time also its own
exteriority. – Number is thus a numerical one that is absolutely determined
but which has at the same time the form of simple immediacy, and to
which, therefore, the connecting reference to an other remains completely
external. Further, as numerical, the one possesses the determinateness (such
as consists in the reference to other) as a moment in it, in its distinction
of unit and amount; and amount is itself the plurality of the ones, that is,
this absolute exteriority is in the one itself. – This intrinsic contradiction
of number or of quantum in general is the quality of quantum, and the
contradiction will develop in the further determinations of this quality.

b. extensive and intensive quantum

a. Their difference

1. We have seen that quantum has its determinateness as limit in amount. In
itself quantum is discrete, a plurality which does not have a being different
from its limit or its limit outside it. Quantum, thus with its limit which as
limit is a plurality, is extensive magnitude.
Extensive magnitude is to be distinguished from continuous magnitude;
its direct opposite is not the discrete, but the intensive magnitude. Exten-
sive and intensive magnitudes are determinacies of the quantitative limit
itself, whereas quantum is identical with its limit; continuous and discrete
magnitudes are, on the contrary, determinations of magnitude in itself, that
is, of quantity as such, in so far as in quantum abstraction is made from
the limit. – Extensive magnitude has the moment of continuity in it and
in its limit, for its many is everywhere continuous; the limit as negation
appears, therefore, in this equality of the many as a limiting of the unity.
Continuous magnitude is quantity that continues without regard to any
limit, and in so far as it is represented with one such limit, the latter is
a limitation in general, without discreteness being posited in it. Determined
as only continuous magnitude, quantum is not yet determined for itself
because the magnitude lacks the one (in which the determinateness-for-
itself lies) and number. Similarly, a discrete magnitude is immediately only
a differentiated many in general which, if it were to have a limit as a many,
would be only an aggregate, that is, would be only indeterminately limited;
in order for quantum to be determinate, the many must be concentrated
into one and thereby be posited as identical with the limit. Continuous
and discrete magnitude, taken as quantum in general, have each posited in
it only one of the two sides by virtue of which quantum is fully determined
and a number. Taken immediately, this latter is extensive quantum – the
simple determinateness which is amount essentially, but the amount of one
and the same unit; extensive quantum is distinguished from number only
because in the latter the determinateness is explicitly posited as plurality.
2. However, the determinateness through number – how much there is
of something – does not require being distinguished from how much thereQuantum
183
is of something else, as if to the determinateness of one thing belonged
how much there is of it and how much there is of an other, for the
determinateness of magnitude as such is a limit determined for itself,
indifferent and simply self-referring; and in number this limit is posited
as enclosed within the one existing for itself: the externality that it has,
the reference to other, is inside it. Further, like the many in general, this
many of the limit is not internally unequal but continuous; each many is
what any other many is; consequently, the many as a many of existents
outside one another, or as discrete, does not constitute the determinateness
as such. Thus this many collapses for itself into its continuity and becomes
simple unity. – Amount is but a moment of number; but, as an aggregate
of numerical ones, it does not constitute the determinateness of number;
on the contrary, these ones as indifferent and self-external are sublated
in number whose being has turned back into itself; the externality that
constituted the ones of plurality vanishes in the one as the self-reference of
number.
The limit of quantum which, as extensive, had its existent determinate-
ness as self-external number, thus passes over into simple determinateness.
In this simple determination of limit, quantum is intensive magnitude; and
the limit or the determinateness which is identical with quantum is now
also posited as simple: it is degree.
Degree is thus a determinate magnitude, a quantum, but at the same
time it is not an aggregate or several within itself; it is only a plurality; 43
plurality is a severality that has gathered together into simple determination,
it is existence that has returned into being-for-itself. It is true that its
determinateness must be expressed by a number, which is the being of the
quantum as completely determined, but the number is not an amount or a
how many times but is rather a onefold, only a degree. When we speak of
10 or 20 degrees, the quantum which has that many degrees – the tenth,
the twentieth degree – is not the amount and sum of the degrees; if that
were the case, it would be an extensive quantum; it is rather only that one
degree, the tenth, the twentieth. It does contain the determinateness found
in the number ten or twenty, but not as several ones: the number is there
as a sublated amount, as a simple determinateness.
3. In number, quantum is posited in its complete determinateness; but
as intensive quantum, in the being-for-itself of number, it is posited as
it is according to its concept or implicitly in itself. For the form of self-
reference which it has in degree is at the same time the externality of degree
43
Mehrheit; literally, “majority.”
21.210
21.211184
21.212
Georg Wilhelm Friedrich Hegel
to itself. As extensive quantum, number is a numerical plurality and thus
has externality inside it. This externality, as plurality in general, collapses
into a state of undifferentiatedness and is sublated into the numerical
one, the self-reference of number. But quantum has its determinateness as
number; it contains this determinateness, as we have said, whether the latter
is posited in it or not. 44 Degree, therefore, which as internally simple no
longer has this external otherness in it, has it outside it, and refers to it as to its
determinateness. A plurality external to it constitutes the determinateness
of the simple limit which the degree is for itself. In so far as in the extensive
quantum amount was supposed to be found within number, it was sublated
there; now, as thus sublated, amount is posited outside number. Number,
in being posited as a one, as self-reference reflected into itself, excludes from
itself the indifference and the externality of amount and is self-reference as
reference to an external through itself.
In this, quantum has the reality which is adequate to its concept. The
indifference of the determinateness constitutes its quality, that is, a determi-
nateness which is in it as a determinateness external to itself. – Accordingly,
degree is a unitary quantitative determinateness among a plurality of such
intensities which, though diverse, each being only a simple reference to
itself, are at the same time in essential connection with each other, so
that each has its determinateness in this continuity with the others. This
reference connecting a degree through itself to its other makes ascent and
descent on the scale of degrees a continuous progress, a flow, which is
an uninterrupted and indivisible alteration; none of the “more or less” 45
differentiated within it is separate from the others but each has its determi-
nateness only in these others. As a self-referring quantitative determination,
each degree is indifferent towards the others; but, in itself, it equally refers
to this externality; it is what it is only through the intermediary of this
externality; in short, its reference to itself is not an indifferent reference to
externality but in this externality it possesses its quality.
b. Identity of extensive and intensive magnitude
Degree is not inherently external to itself. It is not, however, the indeter-
minate one, is not the principle of number as such which is not amount
44
45
In the preceding Remark.
Mehrern. Here and in what follows Hegel is playing on the root meaning of the German word
for “several” (mehrere) which is “more” (mehr). This wordplay is lost in English. I am using
circumlocutions in an attempt to retain it as much as possible.Quantum
185
except negatively, that is, only in the sense of not being an amount. Inten-
sive magnitude is at first a simple one of many “more or less”; there are several
degrees; but they are not determined either as a unitary one or as a more or
less but only as referring to each other as outside each other, or in the identity
of the one and the “more or less than.” Thus, although the several “more
or less” are as such indeed outside the unitary degree, the determinateness
of the latter lies nonetheless in its connection with them; the degree thus
contains amount. Just as twenty contains as extensive magnitude twenty
ones as discrete, the specific degree contains them as continuity, a continu-
ity which simply is this determinate plurality; it is the twentieth degree, and
it is this twentieth degree only through the intermediate of this amount
which, as such, is outside it.
The determinateness of the intensive magnitude is to be considered,
therefore, from two sides. It is determined through other intensive quanta
and is continuous with its otherness, so that its determinateness consists
in this connection with it. Now, in so far as this determinateness is, first,
a simple determinateness, it is determined as against the other degrees; it
excludes them from itself and has its determinateness in this exclusion.
But, second, it is determined within; this it is in the amount as its amount,
not in the amount as excluded, or not in the amount of the other degrees.
The twentieth degree contains the twenty within itself; it is not only
determined as distinguished from the nineteenth, the twenty-first, etc.,
but its determinateness is rather its amount. But, inasmuch as the amount
is its own, and the determinateness is at the same time essentially as amount,
the degree is extensive quantum.
Extensive and intensive magnitude are, therefore, one and the same
determinateness of quantum; they are distinguished only inasmuch as the
one has the amount within and the other has the same without. Extensive
magnitude passes over into intensive magnitude because its many collapses
in and for itself into oneness and steps outside it. But, conversely, this simple
one has its determinateness only in the amount, its amount; indifferent to
the otherwise determined intensities, it has the externality of amount in it;
thus intensive magnitude is just as essentially extensive magnitude.
With this identity, the qualitative something comes on the scene; for
the identity is the unity that refers back to itself through the negation of
its distinct terms; these terms, however, make up the determinateness of
the existent magnitude. The something is a quantum, but its qualitative
existence is now posited as indifferent to it as it is in itself. One can speak
of quantum, number as such, etc., without any mention of a something as
their substrate. But the something, self-mediated by virtue of the negation
of its determinations, now confronts these as existing for itself, and, since
it has a quantum, it confronts them as something which has an extensive
and intensive quantum. Its one determinateness which it has as quantum
is posited in the distinct moments of unity and amount; it is in itself not
only one and the same determinateness, but the positing of it in these
differences as extensive and intensive quantum is the return into this unity
which, as negative, is the something posited as indifferent to them.

c. Alteration of quantum

The difference between extensive and intensive quantum is indifferent to
the determinateness of quantum as quantum. But quantum is in general
determinateness posited as sublated: the indifferent limit, the determinate-
ness which is just as much the negation of itself. In extensive quantum,
this difference is developed; but intensive magnitude is the existence of
this externality which quantum is in itself. The difference is posited as the
contradiction which it is in itself, of being a simple self-referring determi-
nateness which is the negation of itself, of having its determinateness not
in it but in another quantum.
A quantum, according to its quality, is therefore in absolute continuity
with its externality, with its otherness. Consequently, not only can every
determinateness of magnitude be transcended, not only can it be altered:
that it must alter is now posited. The determination of magnitude continues
into its otherness in such a way that it has its being only in this continuity
with an other; it is not just a limit that exists but one that becomes.
The one is infinite or self-referring negation, and hence the repulsion of
itself from itself. Quantum is equally infinite, posited as the self-referring
negation; it repels itself from itself. But it is a determinate “one,” the
one which has passed over into existence and limit, thus the repulsion
of determinateness from itself, not the generation of something that is
like itself (as the repulsion of the one) but of its otherness; quantum is
now posited in it as sending itself 50 beyond itself. It consists in this, that it
increases or decreases; it is within it the externality of determinateness.
Thus quantum sends itself beyond itself; this other which it becomes
is at first itself a quantum, but a quantum which is not a static limit but
one that impels itself beyond itself. The limit which arises in this beyond
is therefore only one that again sublates itself and sends itself to a further
limit, and so on to infinity.

c. quantitative infinity

a. Its concept

Quantum alters and becomes another quantum; the further determination
of this alteration, that it goes on to infinity, lies in that it is positioned as
inherently self-contradictory. – Quantum becomes an other; but it continues
in its otherness; the other is therefore also a quantum. This latter, however,
is the other, not of a quantum, but of the quantum as such, the negative
of itself as a limited something, and hence its own unlimitedness, infinity.
Quantum is an ought; it implies that it be determined-for-itself, and this
being-determined-for-itself is rather the being determined in an other; and,
conversely, it is the being-determined in an other as sublated, is indifferent
subsisting-for-itself.
In this way, finitude and infinity each at once acquires within it a double
though opposite meaning. Quantum is finite, first, as limited in general;
second, as sending itself beyond itself, as being-determined in an other. On
the other hand, its infinity, is, first, the unlimitedness of quantum; second,
its being-turned-back-into-itself, the indifferent being-for-itself. If we now
compare these moments with each other, we find that the determination
of quantum’s finitude, its sending itself beyond itself into an other that
constitutes its determination, is equally the determination of the infinite;
the negation of limit is this same transcendence of determinateness, so that
in this negation, in the infinite, quantum has its final determinateness.
The other moment of infinity is the for-itself which is indifferent to the
limit; but the quantum itself is so limited, as to be indifferent with respect
to its limit, and hence with respect to other quanta and its “beyond.”
In quantum, finitude and infinity (the latter supposedly separate from
finitude, as bad infinity) each already possesses within it the moment of
the other.
The qualitative and quantitative infinite are distinguished inasmuch as
in the former the opposition of the finite and infinite is qualitative, and the
transition of the finite into the infinite, or the reference of each to the other,
is present only in the in-itself, in their concept. Qualitative determinateness
is immediate; it refers to otherness essentially as to a being which is other
than it; it is not posited as having its negation, its other, in it. By contrast,
magnitude is as such sublated determinateness; it is posited as being unlike
and indifferent to itself, and hence as something alterable. The qualitative
finite and infinite, therefore, stand opposed to each other absolutely, that
is, abstractly; their unity is the inner connection underlying them; hence
the finite continues in its other only in itself, not in it. By contrast, in the
infinite in which the quantitative finite has its absolute determinateness,
this finite refers to itself in it. This, their mutual reference, is first displayed
in the quantitatively infinite process.

b. The quantitative infinite process

The process to infinity is in general the expression of contradiction, here,
of the contradiction contained in the quantitative finite or in quantum in
general. It is the reciprocal determination of the finite and the infinite that
came up for consideration in the sphere of the qualitative, 51 with the differ-
ence that, as just indicated, in the sphere of quantity the limit inherently
sends itself beyond itself and continues there, and hence, conversely, the
quantitative infinite is also posited as having the quantum in it, for in its
externality quantum is itself; its externality belongs to its determination.
The infinite progress is now the expression of this contradiction, not the
resolution of it; however, because of the continuity of one determinateness
in the other, the progress gives rise to the semblance of a resolution in a
union of the two. As at first posited, such a progress is the task of attaining
the infinite but not the attainment of it; it is a perpetual generation of the
infinite, without the progress of ever getting beyond the quantum itself, and
without the infinite ever becoming something which is positively present.
It belongs to the concept of quantum to have a beyond of itself. This
beyond is, first, the abstract moment of the non-being of quantum; this
resolves itself in it; it thus refers to its beyond as to its infinity in accordance
with the qualitative moment of the opposition. But, second, quantum is
continuous with this beyond; it consists precisely in being the other of
itself, external to itself; this externality equally is, therefore, no more an
other than the quantum; the beyond or the infinite is thus itself a quantum.
The beyond is thus recalled from its flight and the infinite is attained. But
because the infinite, now become a “this-side,” is again a quantum, what
is posited is again only a new limit; this limit, as quantum, has also fled
again from itself, is as such beyond itself, and has repelled itself from itself
into its non-being, into its beyond, and as the quantum repels itself into
the beyond, so does the beyond perpetually become a quantum.
The continuity of quantum with its other brings about the conjunction
of the two in the expression of an infinitely great or infinitely small. Since
they both still have in them the determination of quantum, they remain
alterable and the absolute determinateness which would be a being-for-itself
is thus not attained. This being-outside-itself of the determination is posited
in the double infinity (posited in the relative opposition of the “more”
and the “less”) of the infinitely great and the infinitely small. In each, the
quantum is maintained in perpetual opposition to its beyond. No matter
how much the “great” is enlarged, it shrinks to insignificance; since it refers
to the infinite as to its non-being, the opposition is qualitative; the enlarged
quantum has gained nothing, therefore, from the infinite; the latter is its
nothing now just as before. Or again, the increase in the quantum is not
an approximation to the infinite, for the distinction between the quantum
and its infinity essentially has also the moment of being non-quantitative.
This moment is only the sharpened expression of the contradiction that the
quantum ought to be something great, that is, a quantum, and non-finite,
that is, not a quantum. – Equally, the infinitely small is, as something small,
a quantum and therefore remains absolutely, that is, qualitatively, too great
for the infinite and opposed to it. In both, there remains the contradiction
of the infinite progress which in them should have reached its goal.
This infinity, which persists in the determination of the beyond of
the finite, is to be characterized as the bad quantitative infinity. Like the
qualitatively bad infinity, it is the perpetual movement back and forth from
one side of the persistent contradiction to the other, from the limit to its
non-being, and from the latter back again to the other, the limit. To be
sure, the term to which the advance is made in the quantitative progress
is not an abstract “other” in general but a quantum which is explicitly
posited as different; but this quantum remains opposed to its negation in
the same way. Also the progress, therefore, is neither an advance nor a gain
but rather a repetition of one and the same move, a positing, a sublating,
and then again a positing and a sublating: an impotence of the negative
to which what it sublates continuously comes back by its very sublation of
it. The two, the positing and the sublation, are so bonded to each other
that they absolutely flee from each other and yet, in thus fleeing, they are
unable to part but rather become bonded in their very flight from each
other.

c. The infinity of quantum

The infinite quantum as infinitely great or infinitely small is itself, in itself,
the infinite progress; as great or small it is a quantum and at the same time
the non-being of quantum. The infinitely great and the infinitely small are,
therefore, figurative representations which on closer inspection prove to be
but unsubstantial nebulous shadows. In the infinite progress, however, this
contradiction is explicitly present and with it that which constitutes the
nature of quantum which, as intensive magnitude, has attained its reality
and is now posited in its existence as it is in its concept. We must now consider
this identity.
Quantum is as degree simple, self-referred, and determined within it.
Because the otherness and the determinateness are sublated in it through
this simplicity, the determinateness is external to it; it has its determinate-
ness outside it. This, its being-outside-itself, is at first the abstract non-being
of quantum in general, the bad infinity. But further, this non-being is also
a magnitude; quantum continues in its non-being, for it is precisely in
its externality that it has its determinateness, and this, its externality, is
itself therefore equally a quantum; the non-being of quantum, the infinity,
is thus limited, that is, this beyond is sublated, is itself determined as a
quantum which, consequently, in its negation is with itself.
But this is what quantum as such is in itself. For through its externality it
is precisely itself; the externality constitutes that in virtue of which quantum
is quantum, where it is with itself. In the infinite progress, therefore, the
concept of quantum is posited.
If we now first look at this progress in its abstract determinations as
they are displayed before us, what we find in it is the sublating of quantum,
but no less also of its beyond; what we find, therefore, is the negation of
quantum as well as the negation of this negation. Its truth is the unity of
these two negations in which the negations are, but as moments. – This
unity is the resolution of the contradiction of which the infinite progress
is the expression; its most immediate meaning, therefore, is that of the
restoration of the concept of magnitude, of being an indifferent or external
limit. On the subject of the infinite progress as such, the only reflection
which is usually made is that each quantum, however great or small, can
disappear, that it must be possible to transcend it – not, however, that
this sublating of the quantum, the beyond, the bad infinite itself, also
disappears.
Even the first sublating, the negation of quality as such whereby the
quantum is posited, is in itself the sublation of negation – quantum is
sublated qualitative limit, consequently sublated negation – but it is at the
same time only in itself; the sublating is posited as an existence, and its
negation is then fixed as the infinite, as the side beyond quantum, while the
latter remains on its side as an immediate; thus the infinite is determinedQuantum
203
only as first negation and it is in this way that it appears in the infinite
progress. But there is more to it, as has just been shown: there is the negation
of negation or what the infinite is in truth. And this we have just seen with
the restoration of the concept of quantum. Such a restoration means, in
the first place, that to the existence of the quantum there has accrued a
more precise determination. What we now have is quantum determined
according to its concept, and this quantum is different from the immediate
quantum: externality is now the opposite of itself, is posited as a moment of
magnitude; quantum, for its part, is posited as having its determinateness in
another quantum, through the intermediary of its non-being, of infinity,
that is, that it is qualitatively what it is. Yet this comparison of the concept
of quantum with its existence belongs more to our reflection, to a relation
which is not yet present here. The next determination, rather, which is
present here is that quantum has returned to quality, is from now on
qualitatively determined. For its defining property, its quality, is externality,
the indifference of the determinateness; and quantum is now posited rather
as being itself in its externality, of referring to itself therein, of being in
simple unity with itself, that is, of being qualitatively determined. – This
qualitative being is still more closely determined, namely as being-for-itself;
for the very self-reference which quantum has attained has proceeded from
mediation, from the negation of the negation. Quantum no longer has
infinity, the being-determined-for-itself, outside it, but in it.
The infinite, which in the infinite progress only has the empty meaning
of a non-being, of an unattained but sought beyond, is in fact nothing
other than quality. Quantum, as indifferent limit, surpasses itself into the
infinite; it thereby seeks nothing else than its being-determined-for-itself,
the qualitative moment which, however, is only an ought. Its indifference
towards the limit, and hence its lack of a determinateness which is an
existent-for-itself, its surpassing itself, is that which makes the quantum
what it is. This, its surpassing, is to be negated and quantum is to find in
infinity its absolute determinateness.
Quite generally: quantum is sublated quality; but quantum is infinite, it
surpasses itself, is the negation of itself; this, its surpassing, is therefore in
itself the negation of the negated quality, the restoration of it; and what is
posited is that the externality, which seemed to be a beyond, is determined
as quantum’s own moment.
Quantum is thus posited as repelled from itself, and with that there are
two quanta which are however sublated, only moments of one unity, and
this unity is the determinateness of quantum. – Quantum, self-referred as
indifferent limit and hence qualitatively posited, is the quantitative relation
or ratio. – In ratio quantum is external to itself, different from itself; this,
its externality, is the reference connecting a quantum to another quantum,
each quantum acquiring value only in this connection with its other; and
this reference constitutes the determinateness of the quantum which is this
unity. In this unity quantum possesses not an indifferent but a qualitative
determination; in this its externality has turned back into itself; it is in it
what it is.

Further forms associated with the qualitative determinateness of magnitude
The affirmative meaning of the infinitesimal of differential calculus is
that it is the qualitative determinateness of magnitude, and, regarding this
determinateness, we have shown in particular that it is present in this
calculus not only as a power determinateness in general, but as specifically
determining the relation of a function of powers to the power of the
expansion. This qualitative determinateness is also present, however, in
another weaker form, so to speak, and this form, together with the use of
the infinitesimal associated with it and the meaning that the latter has in
this use, will be the subject of the present Remark.
Before we proceed, however, it should be recalled at this juncture that the
different power determinations first come in from the analytical side, that
they are there merely formal and totally homogeneous, that they signify
numerical magnitudes which, as such, are not qualitatively differentiated
from each other. It is in the application to spatial objects that the analytical
relation fully shows itself in its qualitative determinateness as the transition
from linear to plane determinations, from the determinations of straight
n
In the already cited critique (Jahrb. für wissensch. Krit. II, Vol. 1827, Nr. 155, 6ff.), 145 are to be
found interesting opinions of a learned specialist in the subject, Herr Spehr. 146 They are quoted
from his Neue Principien des Fluentencalculs (Braunschweig, 1826), and concern a matter which has
contributed substantially to the issue of obscurities and unscientific claims in differential calculus;
moreover, they accord with what we have said concerning the general orientation of the theory of this
calculus. To quote: “Purely arithmetical investigations – admittedly those which more than all other
have direct bearing on differential calculus – have not been separated from this calculus proper.
Indeed, they have even been taken, as by Lagrange, for what is of the essence of the calculus while
calculus itself has been regarded as only the application of them. These arithmetical investigations
include the rules of differentiation, the derivation of Taylor’s theorems, and so on, even the various
methods of integration. But the case is quite the reverse, for it is these applications that make up the
subject matter of differential calculus proper, which presupposes all those arithmetical developments
and operations from analysis.” – We have shown how, with Lagrange, the separation of the so-
called application from the procedure of the general part which takes its start from the series serves
precisely to bring to light the proper subject of differential calculus. However, it is strange that the
author, despite entertaining the interesting view that it is precisely the so-called applications that
constitute the subject matter of differential calculus proper, would then himself divagate (ibid.)
into the formalistic metaphysics of continuous magnitudes, becoming, flow, etc., and would want to
add new ballast to old. These determinations are formal, in the sense that they are only general
categories which fail to give precisely that which is specific to the subject matter. But this is what was
to be recognized in, and be abstracted from, the concrete propositions, the applications.
Cf. above, 21.267, Hegel’s footnote.
Friedrich Wilhelm Spehr (1799–1833), mathematician and cartographer in Braunschweig, best
known for his Vollständiger Lehrbegriff der reinen Kombinationslehre (1824).Quantum
261
line to those of curves, etc. This application also brings in train that
the spatial objects which are given in the form of continuous magnitudes
in accordance with their nature will be taken discretely, the plane as an
aggregate of lines, the line as an aggregate of points, etc. The sole interest
of this dissolution is to determine the points and the lines into which
the lines and the planes have respectively been resolved in order to be
able, starting from this determination, to proceed analytically, that is, in
truly arithmetical fashion; for the sought magnitude determinations these
starting points are the elements from which the function and the equation
for the concrete, that is, the continuous, magnitude are to be derived. For
the problems where the use of this procedure is of special interest, it is
requisite that the element which is the starting point be self-determined,
in contrast with the indirect procedure which can start only with limits
and seek between them the self-determined element which is its goal.
But if it is only the law of progressively determining magnitudes without
however ever possibly attaining the required perfect, that is, the so-called
finite, determination, then in both methods the result amounts to the
same thing. To Kepler is ascribed the honor of first having thought of
that reversal of procedure and of having made the discrete the starting
point. His explanation of how he understands the first proposition of
Archimedes’ cyclometry conveys this quite simply. 147 This proposition, as
we know, is that the circle is equal to a right-angle triangle, one side of
which enclosing the right angle equals the diameter of the circle and the
other its circumference. Since Kepler takes this proposition to mean that
the periphery of the circle has as many parts as it has points, that is, infinitely
many, each of which can be regarded as the base of an isosceles triangle,
he thereby gives expression to the resolution of the continuous into the
form of the discrete. The expression “infinite” which occurs here is still far
removed from the meaning which it is supposed to assume in differential
calculus. – Now when a determinateness, a function, has been found for
such discrete elements, these should be summed up further, as essentially
elements of the continuous. But since a sum of points will not yield a
line, nor a sum of lines a surface, the points are straight away taken as
already linear, just as the lines are taken as two-dimensional. But again,
since linear elements ought not to be lines yet, which they would be if they
were taken as quantum, they are represented as infinitely small. The discrete
is only capable of an external combination in which the moments retain the
meaning of discrete units; the only transition analytically made from these
147
See note 141, above.
21.301262
21.302
Georg Wilhelm Friedrich Hegel
units is to their sum; nor is there a geometrical transition from point to line,
or from line to surface, etc.; therefore the element which is determined as
a point or a line is also given, together also with this determination, the
quality of being as point linear or as line two-dimensional, in order that
their sum would indeed become, as the sum of tiny lines, a line, and, as
the sum of tiny planes, a plane.
It is the need to get hold of this moment of qualitative transition and
for this reason to have recourse to the infinitely small that must be regarded
as the source of all the representations which, though meant to smooth
out every difficulty, are themselves the greatest difficulty. But to dispense
with these aids, it would have had to be possible to show that contained
even in the analytical procedure, which otherwise appears as a mere sum-
mation, there is in fact already a multiplication. But a fresh assumption
is trotted out at this juncture which provides the basis for this applica-
tion of arithmetical relations to geometrical figures. The assumption is
that arithmetical multiplication is also for the geometrical determination
a transition to a higher dimension, that the arithmetical multiplication of
magnitudes spatially determined as lines is at the same time a promotion of
the linear to the status of two-dimensional determination; three times four
linear feet gives twelve linear feet, but three linear feet times four linear feet
gives twelve plane-feet, that is, square-feet, since the unit in both factors,
as discrete quanta, is the same. The multiplication of lines by lines appears
at first sight to make no sense, for multiplication has to do simply with
numbers, that is, it is an alteration of elements totally homogeneous with
that into which they are altered, with the product, only the quantity alters.
In contrast to this alteration, what has been called the multiplication of
a line as such by a line (it has been named ductus lineæ in lineam, like
the ductus plani in planum; there is also the ductus puncti in lineam) is
an alteration not merely of magnitude, but of magnitude as a qualitative
determination of spatiality, of dimensionality; the transition of the line into
plane is to be understood as the self-externalization of the line, just as the
self-externalization of the point is a line and the self-externalization of the
plane a whole space. This is the same as representing the motion of the
point as being the line, and so on; but since motion entails time determi-
nation, it is more likely to appear in this way of representing it as merely
accidental, an external alteration of state, whereas what is to be captured is
the conceptual determination as expressed by self-externalization, that is,
the qualitative alteration which, arithmetically, is the multiplication of unit
(the point, etc.) by amount (the line, etc.). – It may also be remarked here
that in the case of the self-externalization of the plane, which would appearQuantum
263
as a multiplying of surface by surface, there seems to emerge a difference
of arithmetical and geometrical products, for that self-externalization, as
ductus plani in planum, would arithmetically give a multiplication of a
two-dimensional factor by another factor equally two-dimensional, and
hence a product of four dimensions which, however, is reduced to three
in geometrical determination. For its part, since number has the one as
its principle, it yields the fixed determination for the external, quantita-
tive element; but the product of its operation is to that extent formal.
3 · 3, taken as a numerical determination, when it reproduces itself, yields
3 · 3 × 3 · 3. However, the same magnitude, taken as the determination of a
plane, in reproducing itself is held back at 3 · 3 · 3, because space, imagined
as a movement of transcendence which starts from the point, the limit
which is only abstract, attains its true limit as concrete determinateness in
the third dimension, starting from the line. The difference just referred to
could prove relevant in the context of free motion, where the one side, the
spatial, is subject to geometrical determination (according to Kepler’s law,
x 3 : t 2 ), and the other side, the temporal, to arithmetical determination.
How the qualitative moment being considered here differs from the
subject of the previous Remark, should now be self-evident without fur-
ther comment. In the earlier Remark, the qualitative moment lay in the
determinateness of the powers; here it is the same as the infinitesimal,
except only as a factor arithmetically related to product, or as point to line,
line to plane, etc. The qualitative transition to be made from the discrete,
into which the continuous magnitude is imagined to be resolved, to the
continuous is now effected as a process of summing.
But that in fact the alleged mere summation contains in itself a multi-
plication, therefore the transition from the linear to the two-dimensional
determination, appears in the simplest form in the way in which, for
instance, it is shown that the area of a trapezium is equal to the product
of the sum of the two opposite parallel lines and half the height. This
height is represented as being only the amount of an aggregate of discrete
magnitudes which must be summed up. These magnitudes are lines lying
parallel between the two parallels limiting them. There are infinitely many
of them, for they are supposed to make up the plane. And yet they are lines,
and therefore, in order to be two-dimensional, they must at the same time
be posited with negation. In order to escape the difficulty of a sum of lines
supposedly yielding a plane, the lines are assumed as planes from the start
but also as infinitely thin, for they derive their determination solely from
the linear character of the parallel limits of the trapezium. As parallel and
bounded by the other pair of rectilinear sides of the trapezium, they can
21.303264
21.304
Georg Wilhelm Friedrich Hegel
be represented as the terms of a uniformly differentiated but in principle
indeterminate arithmetical progression whose first and last terms are the
two limiting parallels. The sum of such a series, as we know, is the product
of these parallels and half the amount of the terms. This last quantum is
called “amount” simply and solely with reference to the representation of
infinitely many lines; it is the determinateness as the magnitude in general
of a continuous – the height. It is clear that what is called a “sum” is at
the same time a ductus lineæ in lineam, the multiplying of the linear by the
linear which is, according to the above determination, 148 the emergence of
two-dimensionality. In the simplest case now of any rectangle in general, a
b, each of the two factors is a simple magnitude; but already in the other
still elementary example of the trapezium, only one factor is simple as
one half of the height. The other factor is determined by a progression;
it is still something linear, but its determinateness as magnitude is more
complex; in so far as it can be expressed by a series, the task of summing it
is called analytical, that is, arithmetical; but the geometrical moment in it
is the multiplication, and this is the qualitative moment of transition from
the dimension of line to that of plane; the one factor has been taken as
discrete only for the arithmetical determination of the other; by itself, like
the other, it is the magnitude of something linear.
But the method of representing planes as sums of lines is also commonly
employed where no multiplication as such is being performed in view of
the result. This is the case when the task is to give the magnitude in the
equation not as quantum but as a proportion. There is, for instance, a
well-known way of showing that the area of a circle bears the same relation
to the area of an ellipse, the major axis of which is the diameter of the
circle, as the major axis does to the minor axis, by taking each of these
areas as the sum of the ordinates pertaining to it; each ordinate of the ellipse
stands to the corresponding ordinate of the circle as the minor axis stands
to the major; therefore, it is concluded, also the sums of the ordinates, that
is, the areas, stand in equal proportion. Those who would avoid in this
connection the representation of areas or planes as a sum of lines resort
to the usual, quite superfluous ad hoc aid of making the ordinates into
trapezia of infinitely small width; since the equation is only a proportion,
only one of the two linear elements of the area enters the comparison.
The other element, the abscissa axis, is assumed equal in the ellipse and
the circle – therefore, as a factor of arithmetical magnitude-determination,
equal to 1 – and the proportion is, consequently, totally dependent only
148
Cf. above, 21.301.Quantum
265
on the relation of the one determining moment. The two dimensions are
necessary to the representation of a plane; the determination of magnitude,
however, as is to be given in this proportion, extends only to the one
moment. Therefore, to yield to the representation, or to try to improve
it by adding the representation of sum to this one moment, is in truth a
failure to recognize what mathematical determinateness is all about.
The foregoing discussion contains also the criterion for Cavalieri’s
method of indivisibles referred to above 149 – a method which the crite-
rion also justifies and one that does not need the recourse to the infinitely
small. These indivisibles are lines when Cavelieri considers a plane, or
squares and circles when he considers a pyramid or a cone, etc. He calls the
base line or base plane assumed as determined the regula; it is the constant
and, with reference to a series, its first or last term; the indivisibles are
regarded as parallel to it, therefore in equal determination with respect to
the figure. Now Cavalieri’s general fundamental proposition is that (Exerc.
Geometr. VI; in the later work, Exerc. I, p. 6) “all figures, plane as well as
solid, are proportionate to all their indivisibles, these being compared with
each other collectively and, if there happens to be in them some common
proportion, distributively.” 150 For this purpose, he compares in figures of
equal base line and height the proportions between the lines drawn parallel
to the base and equidistant from it; all such lines of a figure have one and the
same determination and constitute its whole content. In this way Cavalieri
proves, for example, also the elementary proposition that parallelograms
of equal height are proportional to their base; any two lines drawn in the
two figures at equal distance from the base and parallel to it are in the
same proportion as the bases, and so too, therefore, are the whole figures.
In fact, the lines do not constitute the content of the figure as continuous
but constitute it only in so far as they are arithmetically determinable; the
linear figure is the element of the content and through it alone must the
determinateness of the latter be grasped.
With this we are led to reflect on a difference regarding where to place
the determinateness of a figure, whether it is constructed, like the height of
the figure here, or whether it is the external limit. In so far as it is an external
limit, then one concedes that the continuity of the figure follows upon, so
to speak, the equality or the proportion of the limit; e.g. the equality of
congruent figures rests on the fact that the boundaries, the limiting lines,
coincide. But in parallelograms of equal height and base, only this last
determinateness is an external limit; the height on which the second main
149
Cf. above, 21.253.
150
Cavalieri, Exercitationes geometricæ sex (1647), p. 6.
21.305266
21.306
Georg Wilhelm Friedrich Hegel
determination of the figures rests, the proportion of the two, and not the par-
allelism as such, brings a second principle of determination to the external
limits. Euclid’s proof of the equality of parallelograms that have the same
height and base reduces them to triangles, to externally limited continu-
ous figures. In Cavalieri’s proof, first of all concerning the proportionality
of parallelograms, the limit is simply a determinateness of magnitude as
such, which is explicated as taken at every pair of lines drawn at the same
distance from each other in both figures. These lines, equal or in equal
proportion to the base, taken collectively, yield the figures that stand in
equal proportion. The image of an aggregate of lines is incompatible with
the continuity of the figure; consideration of the lines alone fully exhausts
the determinateness which is at issue. Cavalieri has a standard reply to the
supposed difficulty, that the representation of indivisibles requires the
comparison of lines or planes which are infinite in number (Geom. Lib. II,
Prop. I, schol.). 151 He rightly points out that he is not comparing the num-
ber of such lines or planes, which we do not know (that is, it is rather an
empty representation, as just remarked, 152 introduced as prop), but only the
magnitude, that is, the quantitative determinateness as such which is equal
to the space occupied by these lines; because this space is enclosed within
limits, its magnitude is also enclosed within these limits; the continuous is
nothing other than the indivisibles themselves, he says; if it were anything
apart from them, it would not be comparable; but it would be absurd to
say that bounded continuous figures are not comparable with each other.
One can see that Cavalieri’s intention is to distinguish what belongs
to the external existence of the continuous from that in which its deter-
minateness lies and which alone must be stressed when comparing it or
devising theorems about it. The categories which he used for this purpose,
namely the continuous as composed of, or consisting in, indivisibles, and the
like, are admittedly not satisfactory, for implied in them is the claim to
an intuition of the continuous or, as already said, of its external existence;
instead of saying “that the continuous is nothing other than the indivisibles
themselves,” 153 it would have been more correct to say, and consequently
also self-explanatory, that the determinateness of the magnitude of the con-
tinuous is none other than that of the indivisibles themselves. – Cavalieri is
not comfortable with the dubious consequence drawn by the schools from
the representation of the continuous as made up of indivisibles, namely
that there is a greater and smaller infinite; he expressly states further in the
151
152
See Cavalieri, Geometria indivisibilibus continuorum nova (1635). Book 2, Proposition 1, scholium.
153 Cf. 21.305, note 151 above.
Cf. above, 21.304.Quantum
267
treatise (Geom. Lib. VII, Pref.) 154 that he is in no way forced by his method
of proof to represent the continuous as made up of indivisibles; continuous
figures only follow the proportion of the indivisibles. He says that he did not
take up the aggregate of indivisibles in the way they seem to end up in
the determination of the infinite, for the sake of an infinite heap of lines
or planes; he took them rather because of the way they are constituted, for
the specific nature of limitedness that they have in them. Nevertheless, in
order to remove this stumbling block, he does not spare himself the trouble
of demonstrating the main propositions of his geometry, in the seventh
book specially added for this purpose, in a way that remains free of any
admixture of infinity. – This way reduces the proofs to the just cited usual
form of congruence of figures, that is, as we remarked, to the representation
of determinateness as external spatial limit. 155
A further remark that can be made here concerning this form of congru-
ence is that it is no more than, so to speak, a childish aid for the intuition
of the senses. In the elementary propositions concerning the triangles, two
such triangles are represented side by side; on the assumption that, of the
six parts of each, three in the one triangle are equal to the corresponding
three in the other, it is shown that the two triangles are congruent, that
is, that in each the remaining three parts are equal to the corresponding
remaining three in the other – because by virtue of the equality of the first
three parts the two triangles cover each other. In more abstract terms, all that
this equality of corresponding pairs of parts in the two triangles amounts
to is that there is just one single triangle in which three parts are assumed
as already determined, whence follows the determinateness of the remaining
three parts. The determinateness of the one triangle is thus demonstrated
to be completed with the three parts assumed; for the determinateness as
such the remaining three parts are therefore a superfluity – the superfluity of
sensible existence, that is, of the intuition of continuity. When the matter is
expressed in these terms, the qualitative determinateness stands out in its
distinction from what is given to intuition, from the whole as continuous
in itself; congruity does not let this distinction show through.
As we have observed, 156 a new circumstance comes into play in the case of
parallel lines and parallelograms: on the one hand the equality of the angles
only, and on the other the height of the figures from which the external
limits of the latter (the sides of the parallelograms) are distinct. There is an
ambiguity here as to whether in these figures, besides the determinateness
of the side which is one external limit, the base, we should take as the other
154
Ibid.
155
Cf. above, 21.305.
156
Cf. above, 21.304.
21.307268
21.308
Georg Wilhelm Friedrich Hegel
external limit the other side of the parallelogram or the height instead. In
the case of two parallelograms that have the same base and height, but
one is rectangular whereas the other has two angles sharply acute and the
opposite two very obtuse accordingly, intuitively the latter figure can easily
appear greater than the former, for intuitively its given longer side is taken
to be the determinant one and, as in Cavalieri’s figurative method, the planes
are compared according to the aggregate of parallel lines intersecting them;
the longer side could be seen as offering the possibility of more lines than
the vertical side of the rectangle does. This intuitive ambiguity, however,
is no argument against Cavalieri’s method, for the aggregate of parallel
lines imagined in the two parallelograms for the purpose of comparison
equally presupposes the equidistance of the lines from each other or from
the base, from which it follows that the height, and not the other side of
the parallelogram, is the other determining moment. The situation changes
again, however, when the two parallelograms being compared are of equal
height and equal base but do not lie on one plane and make different angles
with a third plane. Here the parallel cross-sections that arise when the third
plane is imagined cutting across the parallelograms in parallel motions are
no longer equidistant from each other and the two planes are unequal.
Cavalieri is very careful to call attention to this difference which he defines
as one between transitus rectus and transitus obliquus of the indivisibles (in
Exercit. I, n. XIIff., and already in Geometr. 1.II), 157 thus staying at the
source any superficial misunderstanding which could arise at this juncture.
I remember Barrow citing in his above mentioned work (Lect. Geom. II,
p. 21) 158 an objection of Taquet 159 – an astute geometer who was also
working at this time on new methods – bearing on precisely this subject.
This Barrow is the one who also used the method of the indivisibles, though
already adulterated by the assumption (which he passed on to his pupil
Newton and to other contemporary mathematicians, Leibniz included)
that a curvilinear triangle, like the so-called characteristic triangle, and a
rectilinear triangle can be said to be equal if both are infinite, that is, very
small. The difficulty raised by Taquet likewise concerns the question of
which line, in the calculation of conic and spherical surfaces, should be
taken as the determining factor in a method based on the application of
the discrete. Taquet’s objection to the method of the indivisibles is that
in the calculation of conic and spheric surfaces, the triangle of the cone
is represented according to this atomistic method as made up of straight
157
159
158 Cf. above, 21.284.
See Cavalieri, Exercitationes geometricæ sex, §§15–17.
Andrew Taquet (1611–1660), Jesuit mathematician at Antwerp.Quantum
269
lines, parallel to the base and perpendicular to the axis, which are at the
same time the radii of the circles of which the surface of the cone consists.
If now this surface is defined as the sum of the circumferences, and this
sum is determined from the number of the radii of these circumferences,
that is, from the length of the axis and the height of the cone, then such
a result would contradict the truth which Archimedes has taught and
demonstrated. To this objection of Taquet, Barrow’s response is that, for
determining the surface, it is not the axis but the side of the triangle of the
cone that must be taken as the line the revolution of which generates the
surface, and this line – not the axis – must therefore be taken as determining
the magnitude for the aggregate of the circumferences.
Objections or uncertainties of this kind have their origin solely in the
vagueness of the representation used for the infinite aggregate of points of
which the line, or of lines of which the plane, etc., supposedly consists; it
is this representation which obscures the essence of the determinateness,
whether of the lines or the planes. – It was the aim of these Remarks to bring
attention to the affirmative determinations of the category of the infinitesi-
mal which, in the various uses made of this category in mathematics, remain
in the background so to speak, and to extract them from the nebulosity in
which they are shrouded when that category is kept at a merely negative
level of determination. In infinite series, as in Archimedes’ cyclometry, the
infinite means no more than that, if the law of the progression of the series
is known, but the so-called finite, that is, the arithmetical expression, is not
given, the reduction of the arc to the straight line cannot be carried out; this
incommensurability is their qualitative difference. Likewise the qualitative
difference between the discrete and the continuous contains in general a
negative determination which makes them appear incommensurable: it
thus introduces the infinite, in the sense that the continuous which is now
to be taken as discrete no longer is supposed to have, as continuous, any
quantum. The continuous, which arithmetically is to be taken as product,
is thereby posited as discrete in itself, that is to say, is broken down into
the elements which are its factors; its magnitude-determinateness lies in
these factors, and they, precisely by virtue of being factors or elements,
are of a lower dimension and, as the determinateness of power goes, of a
lower power than the magnitude of which they are the elements or factors.
This difference appears arithmetically as merely quantitative, a difference
of the root or the power, of whatever degree they might be; however, if
the expression does not go further than just the quantitative sphere as such
(for example, a : a 2 or d a 2 = 2a : a 2 = 2 : a ; or for the law of gravity,
t : at 2 ), it then yields the ratios 1 : a, 2 : a, 1 : at, which say nothing
21.309270
Georg Wilhelm Friedrich Hegel
whatsoever; as against their merely quantitative determination, the sides
would have to be held apart by the qualitative meaning differentiating
them, as s : at 2 , the magnitude being thereby spoken of as a quality, as
a function of the magnitude of another quality. Here then, we are only
conscious of the quantitative determinateness, and it is easy enough, in its
own way, to operate with it. Nor is there anything terrible in multiplying
the magnitude of a line by the magnitude of another line. However, the
multiplication of these magnitudes yields at the same time the qualitative
alteration of the transition of line into plane, and to this extent a negative
determination comes into play. It is this determination that occasions the
difficulty, a difficulty which can be resolved by an insight into its peculiarity
and the simple nature of the matter, but which, when the attempt is made
to eliminate it by the aid of the infinite, only degenerates unresolved into
confusion.

CHAPTER 3

Ratio or the quantitative relation

The infinity of quantum has been determined up to the point where it is
the negative beyond of quantum, a beyond which quantum, however, has
within it. This beyond is the qualitative moment in general. The infinite
quantum, as the unity of the two moments, of the quantitative and the
qualitative determinateness, is in the first instance ratio.
In ratio, quantum no longer has a merely indifferent determinateness but
is qualitatively determined as simply referring to its beyond. It continues
in its beyond, and this beyond is at first just an other quantum. Essentially,
however, the two do not refer to each other as external quanta but each
rather possesses its determinateness in this reference to the other. In this, in
their otherness, they have thus returned into themselves; what each is,
that it is in its other; the other constitutes the determinateness of each. –
The quantum’s self-transcendence does not now mean, therefore, that
quantum has simply changed either into some other or into its abstract
other, into its abstract beyond, but that there, in the other, it has attained
its determinateness; in its other, which is an other quantum, it finds itself.
The quality of quantum, its conceptual determinateness, is its externality
as such, and in ratio quantum is now posited as having its determinateness
in this externality, in another quantum – as being in its beyond what it is.
It is quanta that stand to each other in the connection that has now come
on the scene. This connection is itself also a magnitude; quantum is not only
in relation, but is itself posited as relation; it is a quantum as such that has
that qualitative determinateness in itself. So, as relation (as ratio), quantum
gives expression to itself as self-enclosed totality and to its indifference to
limit by containing the externality of its being-determined in itself: in this
externality it is only referred back to itself and is thus infinite within.
Ratio in general is:
1. direct ratio. In this, the qualitative moment does not yet emerge explicitly
as such; in no other way except still as quantum is quantum posited
as having its determinateness in its externality. – In itself, however, the
quantitative relation is the contradiction of externality and self-reference,
the persistence of quanta and their negation. Such a contradiction next
sublates itself:
2. first inasmuch as in indirect or inverse ratio the negation of each of
the quanta is as such co-posited in the alteration of the other, and the
variability of the direct ratio is itself posited;
3. but in the ratio of powers, the unity, which in its difference refers back
to itself, proves to be a simple self-production of the quantum; this
qualitative moment itself, finally posited in a simple determination and
as identical with the quantum, becomes measure.
– About the nature of the following ratios, much was anticipated in
the preceding remarks concerning the infinity of quantity, that is, the
qualitative moment in it; it only remains to analyze, therefore, the abstract
concept of this ratio.

A. THE DIRECT RATIO

1. In the ratio which, as immediate, is direct, the determinateness of each
quantum lies in the reciprocal determinateness of the other. There is only
one determinateness or limit of both – one which is itself a quantum, namely
the exponent of the ratio.
2. The exponent is some quantum or other; however, in referring itself
to itself in the otherness which it has within it, it is only a qualitatively
determined quantum, for its difference, its beyond and otherness, is in
it. This difference in the quantum is the difference of unit and amount –
the unit, which is the being-determined-for-itself; the amount, which is
the indifferent fluctuation of determinateness, the external indifference of
quantum. Unit and amount were at first the moments of quantum; now,
in the ratio, in quantum as realized so far, each of its moments appears as a
quantum on its own and as determinations of the existence of the quantum,
as delimitations against the otherwise external, indifferent determinateness
of magnitude.
The exponent is this difference as simple determinateness, that is, it
has the meaning of both determinations immediately in it. First, it is a
quantum and thus an amount. If the one side of the ratio which is taken as
unit is expressed in a numerical one, and has only the value of one, then the
other, the amount, is the quantum of the exponent itself. Second, it is simple
determinateness as the qualitative moment of the sides of the ratio. When
the quantum of the one side is determined, the other is also determined
by the exponent and it is a matter of total indifference how the first is
determined; it no longer has any meaning as a quantum determined for
itself but can just as well be any other quantum without thereby altering the
determinateness of the ratio, which rests solely on the exponent. The one
side which is taken as unit always remains unit however great it becomes,
and the other, however great it too thereby becomes, must remain the same
amount of that unit.
3. Accordingly, the two truly constitute only one quantum; the one side
has only the value of unit with respect to the other, not of an amount; and
the other only that of amount. According to their conceptual determinateness,
therefore, they are themselves not complete quanta. But this incompleteness
is in them a negation, and it is so not because of their general variability,
according to which one of them (any of the two) can assume all possible
magnitudes, but because they are so determined that, as one is altered, the
other is increased or decreased in corresponding measure. This means that,
as indicated, only one of them, the unit, is altered as quantum; the other
side, the amount, remains the same quantum of units, and the first side too
retains the value of a unit, however much it is altered as quantum. Each side
is thus only one of the two moments of quantum, and the self-subsistence
which is their proper characteristic is in principle negated; in this qualitative
combination they are to be posited as negative with respect to each other.
The exponent ought to be the complete quantum, since the determi-
nations of both sides come together in it; but in fact, even as quotient the
exponent only has the value of amount, or of unit. There is nothing avail-
able for determining which of the two sides of the relation would have to
be taken as the unit or as the amount; if one side, quantum B, is measured
against quantum A as unit, then the quotient C is the amount of such
units; but if A is itself taken as amount, the quotient C is the unit which is
required by the amount A for the quantum B. As exponent, therefore, this
quotient is not posited for what it ought to be, namely the determinant
of the ratio, or the ratio’s qualitative unity. It is posited as such only to
the extent that its value is that of the unity of the two moments, of unit
and amount. And since these two sides, as quanta, are indeed present as
they should be in the explicated quantum, in the ratio, but at the same
time have the value, which is specific to them as the sides of the ratio, of
being incomplete quanta and of counting only as one of those qualitative
moments, they are to be posited with this negation qualifying them. Thus
there arises a more real ratio, one more in accordance with its definition,
one in which the exponent has the meaning of the product of the sides. In
this determinateness, it is the inverse ratio.

B. THE INVERSE RATIO

1. The ratio as now before us is the sublated direct ratio. It was an immediate
relation and therefore not yet truly determinate; henceforth, the newly
introduced determinateness gives the exponent the value of a product, the
unity of unit and amount. In immediacy, as we have just seen, it was
possible for the exponent to be indifferently taken as unit or amount.
Moreover, it also was only a quantum in general and therefore an amount
by choice. One side was the unit, and this was to be taken as a numerical
one with respect to which the other side would be a fixed amount and at
the same time the exponent. The quality of the latter, therefore, was only
that this quantum is taken as fixed, or rather that the constant only has the
meaning of quantum.
Now in the inverse ratio, the exponent is as quantum likewise immediate,
a quantum or other which is assumed as fixed. But to the one of the
other quantum in the ratio, this quantum is not a fixed amount; the ratio,
previously taken as fixed, is now posited instead as alterable; if another
quantum is taken as the unit of the one side, the other side now no longer
remains the same amount of units of the first side. In direct ratio, this unit
is only the common element of both sides; as such, it continues into the
other side, the amount; the amount itself, or the exponent, is by itself
indifferent to the unit.
But as the determinateness of the ratio now is, the amount as such alters
relative to the unit with respect to which it makes up the other side of the
ratio; whenever another quantum is taken as the unit, that amount alters.
Consequently, although the exponent is still only an immediate quantum
and only arbitrarily assumed as fixed, it does not remain fixed in the side of
the ratio: rather this side, and with it the direct ratio of the sides, is alterable.
In the ratio now before us, the exponent as the determining quantum is
thus posited as negative towards itself as a quantum of the ratio, and hence
as qualitative, as limit; the result is that the qualitative moment distinctly
comes to the fore for itself as against the quantitative moment. – In the
direct ratio, the alteration of the two sides is only the one alteration of the
quantum from which the unity which is the common element is taken; by as
much, therefore, as the one side is increased or decreased, so also is the other;
the ratio itself is indifferent to this alteration and the alteration external
to the ratio. In the indirect ratio, on the contrary, although still arbitrary
according to the moment of quantitative indifference, the alteration is
contained within the ratio, and its arbitrary quantitative extension is limited
by the negative determinateness of the exponent as by a limit.Ratio or the quantitative relation
275
2. We must now consider this qualitative nature of the indirect ratio more
closely, as it is realized, and sort out the entanglement of the affirmative
and the negative moments that are contained in it. – Quantum is posited
as being quantum qualitatively, that is, as self-determining, as displaying
its limit within it. Accordingly, first, it is an immediate magnitude as simple
determinateness; it is the whole as existent, an affirmative quantum. But,
second, this immediate determinateness is at the same time limit; for that
purpose it is distinguished into two quanta which are at first the other of
each other; but as the qualitative determinateness of these quanta, and as
a determinateness which is moreover complete, quantum is the unity of
the unit and the amount, a product of which the two are the factors. Thus
on the one hand the exponent of their ratio is in them identical with itself
and is their affirmative moment by which they are quanta; on the other
hand, as the negation posited in them, it is in them the unity according
to which each, at first an immediate and limited quantum in general, is at
the same time so limited as to be only implicitly in itself identical with its
other. Third, the exponent as the simple determinateness is the negative
unity of this differentiation of it into the two quanta, and the limit of their
reciprocal limiting.
In accordance with these determinations, the two moments limit them-
selves inside the exponent and each is the negative of the other, for the
exponent is their determinate unity; the one moment becomes as many
times smaller as the other becomes greater; each possesses a magnitude of
its own to the extent that this magnitude is in it that of their other, that
is, is the magnitude that the other lacks. The magnitude of each in this
way continues into the other negatively; how much it is in amount, that
much it sublates in the other as amount and is what it is only through
this negation or limit which is posited in it by the other. In this way,
each contains the other as well, and is proportioned to it, for each is sup-
posed to be only that quantum which the other is not; the magnitude of
the other is indispensable to the value of each, and therefore inseparable
from it.
This continuity of each in the other constitutes the moment of unity
through which the two magnitudes stand in relation – the moment of
the one determinateness, of the simple limit which is the exponent. This
unity, the whole, constitutes the in-itself of each from which their given
magnitude is distinct: this is the magnitude according to which each is
only to the extent that it takes from the other a part of their common
in-itself which is the whole. But each can take from the other only as much
as will make it equal to this in-itself; it has its maximum in the exponent276
which, in accordance with the stated second determination, is the limit of
the reciprocal delimitation. And since each is a moment of the ratio only
to the extent that it limits the other and is thereby limited by it, it loses
this, its determination, by making itself equal to its in-itself; in this loss,
the other magnitude will not only become a zero, but itself vanishes, for
what it ought to be is not just a quantum but what it is as such a quantum,
namely only the moment of a ratio. Thus each side is the contradiction
between the determination as its in-itself, that is, as the unity of the whole
which is the exponent, and the determination as the moment of a ratio;
this contradiction is infinity again, in a new form peculiar to it.
The exponent is the limit of the sides of its ratio, within which limit
the sides increase and decrease proportionately to each other; but they
cannot become equal to this exponent because of the latter’s affirmative
determinateness as quantum. Thus, as the limit of their reciprocal limiting,
the exponent is (a) their beyond which they infinitely approximate but can
never attain. This infinity in which the sides approximate their beyond is
the bad infinity of the infinite progression; it is an infinity which is itself
finite, that finds its restriction in its opposite, in the finitude of each side
and of the exponent itself, and for this reason is only approximation. But
(b) the bad infinite is equally posited here as what it is in truth, namely
the negative moment in general, in accordance with which the exponent
is the simple limit as against the distinct quanta of the ratio: it is the in-
itself to which, as the absolutely alterable, the finitude of the quanta is
referred, but which, as the quanta’s negation, remains absolutely different
from them. This infinite, which the quanta can only approximate, is then
equally found affirmatively present on their side: the simple quantum of
the exponent. In it is attained the beyond with which the sides of the
ratio are burdened; it is in itself the unity of the two or, consequently,
in itself the other side of each side; for each side has only as much value
as the other does not have; its whole determinateness thus rests in the
other, and this, their being-in-itself, is as affirmative infinity simply the
exponent.
3. With this, however, we have the transition of the inverse ratio into
a determination other than the one it had at first. This consisted in the
fact that the quantum is immediate but at the same time so connected to
an other that the greater it is, the smaller is the other, that it is what it is
by virtue of negatively relating to the other; also, a third magnitude is the
common restriction on this alteration in magnitude that the two quanta
undergo. This reciprocal alteration, as contrasted to the fixed qualitative

limit, is here their distinctive property; they have the determination of

sum up are not only that this infinite beyond is at the same time some

present finite quantum or other, but that its fixity – which makes it with
respect to the quantitative moment the infinite beyond that it is, and
which is the qualitative moment of being only as abstract self-reference –
has developed itself as a mediation of itself with itself in its other, the
finite moments of the relation. The general point is that the whole is as
such the limit of the reciprocal limiting of the two terms, and that the
negation of the negation (and consequently infinity, the affirmative self-
relation) is therefore posited. The more particular point is that, as product,
the exponent already implicitly is already the unity of unit and amount,
whereas each of these two terms is only one of two moments, and for
this reason the exponent encloses them in itself and in them it implicitly
refers itself to itself. But in the inverse ratio the difference has developed
into the externality of quantitative being, and the qualitative being is not
only something fixed, nor does it simply enclose the two moments of the
ratio immediately in it, but in the externally existent otherness it rejoins itself.
It is this determination that stands out as a result in the moments we
have seen. The exponent, namely, is found to be the implicit being whose
moments are realized in quanta and in their generalized alterability. The
indifference of the magnitudes of these moments in the course of their
alteration displays itself as an infinite progression, the basis of which is
that in their indifference their determinateness is to have their value each
in the value of the other. Thus, (a) according to the affirmative side of
their quantum, the determinateness of the moments is that each is in itself
the whole of the exponent; equally, (b) they have the magnitude of the
exponent for their negative moment, for their reciprocal limiting; their
limit is that of the exponent. The fact that such moments do not have
any other immanent limit, any fixed immediacy, is posited in the infinite
progression of their existence and in their limitation, in the negation of
every particular value. This is, accordingly, the negation of the externality
of the exponent which is displayed in them, and the exponent – itself
equally a quantum as such and also expanded into quanta – is thereby
posited as preserving itself in the negation of the indifferent subsistence of
the moments, as rejoining itself, and thus as the determining factor in this
movement of self-surpassing.
The ratio is hereby determined as the ratio of powers.

C. THE RATIO OF POWERS

1. Quantum, in positing itself as self-identical in its otherness and in
determining its own movement of self-surpassing, has come to be a being-
for-itself. As such a qualitative totality, in positing itself as developed, it
has for its moments the conceptual determinations of number: the unit
and the amount. This last, amount, is in the inverse ratio still an aggregate
which is not determined as amount by the unit itself but from elsewhere,
by a third determinate aggregate; but now it is posited as determined only
by the unit. This is the case in the ratio of powers where the unit, which in
it is amount, is at the same time the amount as against itself as unit. The
otherness, the amount of units, is the unit itself. The power is an aggregate
of units, each of which is this aggregate itself. The quantum, as indifferent
determinateness, changes; but inasmuch as the alteration is the raising to a
power, the otherness of the quantum is determined purely by itself. – The
quantum is thus posited in the power as having returned into itself; it is
immediately itself and also its otherness.
The exponent of this ratio is no longer an immediate quantum, as in the
direct ratio and also in the inverse ratio. In the ratio of powers, the exponent
is of an entirely qualitative nature; it is this simple determinateness: that the
amount is the unit itself, that the quantum is self-identical in its otherness.
And the side of its quantitative nature is to be found in this: that the limit or
negation is not an immediate existent, but that existence is posited rather
as continuing in its otherness. For the truth of quality is precisely to be
quantity, or immediate determinateness as sublated.
2. The ratio of powers appears at first as an external alteration to which a
given quantum is subjected; but it has a closer connection with the concept
of quantum, namely, that in the existence into which the quantum has
developed in the ratio of powers, quantum has attained that concept, has
realized its concept to the fullest. The ratio of powers is the display of
what the quantum is implicitly in itself; it expresses its determinateness
of quantum or the quality by which it is distinguished from another.
Quantum is indifferent determinateness posited as sublated, that is to say,
determinateness as limit, one which is just as much no determinateness,
which continues in its otherness and in it, therefore, remains identical
with itself. Thus is quantum posited in the ratio of powers: its otherness,
the surpassing of itself in another quantum, as determined through the
quantum itself.
If we compare the progressive realization of quantum in the preceding
ratios, we find that quantum’s quality of being the difference of itself from
itself is simply this: that it is a ratio. As the direct ratio, quantum is this
posited difference only in the first instance or immediately, so that the self-
reference which it has as exponent, in contrast to its differences, counts
only as the fixity of an amount of the unit. In the inverse ratio, as negatively
determined, quantum is a relating of itself to itself (to itself as to its negation
in which, however, it has its value); as an affirmative self-reference, it is an
exponent which, as quantum, is only implicitly in itself the determinant of
its moments. But in the ratio of powers quantum is present in the difference
as a difference of itself from itself. This externality of determinateness is the
quality of quantum and is thus posited, in conformity to the concept
of quantum, as quantum’s own determining, as its reference to itself, its
quality.
3. By being thus posited as it is in conformity to its concept, quantum
has passed over into another determination; or, as we can also say, its
determination is now also as the determinateness, the in-itself also as existence.
It is quantum in so far as the externality or the indifference of its determining
(as we say, it is that which can be increased or decreased) is simply accepted
and immediately posited; it has become the other of itself, namely quality,
in so far as that same externality is now posited as mediated by quantum
itself and hence as a moment of quantum, so that in that very externality
quantum refers itself to itself – is being as quality.
At first quantity as such thus appears in opposition to quality; but
quantity is itself a quality, self-referring determinateness as such, distinct
from the determinateness which is its other, from quality as such. Except
that quantity is not only a quality, but the truth of quality itself is quantity,
and quality has demonstrated itself as passing over into it. Quantity, in
its truth, is instead the externality which has returned into itself, which is
no longer indifferent. Thus is quantity quality itself, in such a way that
outside this determination quality as such would yet not be anything at
all. – For the totality to be posited, a double transition is required, not
only the transition of one determinateness into the other, but equally the
transition of this other into the first, its going back into it. Through the
first transition, the identity of the two is present at first only in itself:
quality is contained in quantity, but the latter still is only a one-sided
determinateness. Conversely, that quantity is equally contained in quality,
that it is equally also only as sublated, this results in the second transition,
the going back into the first determinateness. This remark regarding the
necessity of the double transition is everywhere of great importance for
scientific method.

Quantum is henceforth no longer an indifferent or external determina-
tion but is sublated as such, and it is a quality and that by virtue of which
anything is what it is; the truth of quantum is to be measure.
