SECTION I

Subjectivity

The concept is, to start with, formal, the concept in its beginning or as
the immediate concept. – In this immediate unity, its difference or its
positedness is, first, itself initially simple and only a reflective shine, so that
the moments of the difference are immediately the totality of the concept
and only the concept as such.
But, second, because it is absolute negativity, the concept divides and
posits itself as the negative or the other of itself; yet, because it is still
immediate concept, this positing or this differentiation is characterized by
the reciprocal indifference of its moments, each of which comes to be on
its own; in this division the unity of the concept is still only an external
connection. Thus, as the connection of its moments posited as self-subsisting
and indifferent, the concept is judgment. 29
Third, although the judgment contains the unity of the concept that
has been lost in its self-subsisting moments, this unity is not posited. It
will become posited by virtue of the dialectical movement of the judgment
which, through this movement, becomes syllogistic inference, 30 and this is
the fully posited concept, for in the inference the moments of the concept
as self-subsisting extremes and their mediating unity are both equally posited.
But since this unity itself, as unifying middle, and the moments, as self-
subsisting extremes, stand at first immediately opposite one another, this
contradictory relation that occurs in the formal inference sublates itself, and
the completeness of the concept passes over into the unity of totality; the
subjectivity of the concept into its objectivity.

CHAPTER 1

The concept

The faculty of concepts is normally associated with the understanding, and
the latter is accordingly distinguished from the faculty of judgment and
from the faculty of syllogistic inferences which is formal reason. But it is
particularly with reason that the understanding is contrasted, and it signifies
then, not the faculty of concepts in general, but the faculty of determinate
concepts, as if, as the prevailing opinion has it, the concept were only a
determinate. When distinguished in this meaning from the formal faculty
of judgment and from formal reason, the understanding is accordingly to
be taken as the faculty of the single determinate concept. For the judgment
and the syllogism or reason, as formal, are themselves only a thing of the
understanding, 31 since they are subsumed under the form of the abstract
determinateness of the concept. Here, however, we are definitely not taking
the concept as just abstractly determined; the understanding is therefore
to be distinguished from reason only in that it is the faculty of the concept
as such.
This universal concept that we now have to consider contains the three
moments of universality, particularity, and singularity. The difference and
the determinations which the concept gives itself in its process of distin-
guishing constitute the sides formerly called positedness. Since this posit-
edness is in the concept identical with being-in-and-for-itself, each of the
moments is just as much the whole concept as it is determinate concept and
a determination of the concept.
It is at first pure concept, or the determination of universality. But the
pure or universal concept is also only a determinate or particular concept
that takes its place alongside the other concepts. Because the concept is
a totality, and therefore in its universality or pure identical self-reference
is essentially a determining and a distinguishing, it possesses in itself the
norm by which this form of its self-identity, in pervading all the moments
Georg Wilhelm Friedrich Hegel
and comprehending them within, equally determines itself immediately as
being only the universal as against the distinctness of the moments.
Second, the concept is thereby posited as this particular or determinate
concept, distinct from others.
Third, singularity is the concept reflecting itself out of difference into
absolute negativity. This is at the same time the moment at which it has
stepped out of its identity into its otherness and becomes judgment.

a. the universal concept

The pure concept is the absolutely infinite, unconditioned and free. It is
here, as the content of our treatise begins to be the concept itself, that we
must look back once more at its genesis. Essence came to be out of being, and
the concept out of essence, therefore also from being. But this becoming
has the meaning of a self-repulsion, so that what becomes is rather the uncon-
ditional and the originative. In passing over into essence, being became a
reflective shine or a positedness, and becoming or the passing over into an
other became a positing; conversely, the positing or the reflection of essence
sublated itself and restored itself to a non-posited, an original being. The
concept is the mutual penetration of these moments, namely, the qualita-
tive and the originative existent is only as positing and as immanent turn-
ing back, and this pure immanent reflection simply is the becoming-other
or determinateness which is, consequently, no less infinite, self-referring
determinateness.
Thus the concept is absolute self-identity by being first just this, the
negation of negation or the infinite unity of negativity with itself. This
pure self-reference of the concept, which is such by positing itself through
the negativity, is the universality of the concept.
Universality seems incapable of explanation, because it is the simplest
of determinations; explanation must rely on determinations and differen-
tiations and must apply predicates to its subject matter, and this would
alter rather than explain the simple. But it is precisely of the nature of the
universal to be a simple that, by virtue of absolute negativity, contains dif-
ference and determinateness in itself in the highest degree. Being is simple
as an immediate; for this reason we can only intend it without being able
to say what it is; therefore, it is immediately one with its other, non-being.
The concept of being is just this, that it is so simple as to vanish into its
opposite immediately; it is becoming. The universal is, on the contrary, a
simple that is at the same time all the richer in itself, for it is the concept.
First, therefore, it is simple self-reference; it is only in itself. But, second,
this identity is in itself absolute mediation but not anything mediated. Of
the universal which is mediated, that is to say, the abstract universal, the one
opposed to the particular and the singular, of that we shall have to speak
only in connection with the determinate concept. – Yet, even the abstract
universal entails this much, that in order to obtain it there is required the
leaving aside of other determinations of the concrete. As determinations in
general, these determination are negations, and leaving them aside is a further
negating. Even in the abstract universal, therefore, the negation of negation
is already present. But this double negation comes to be represented as if
it were external to it, both as if the properties of the concrete that are left
out were different from the ones that are retained as the content of the
abstraction, and as if this operation of leaving some aside while retaining
the rest went on outside them. With respect to this movement, the universal
has not yet acquired the determination of externality; it is still in itself that
absolute negation which is, precisely, the negation of negation or absolute
negativity.
Accordingly, because of this original unity, the first negative, or the
determination, is not, to begin with, a restriction for the universal; rather,
the latter maintains itself in it and its self-identity is positive. The categories
of being were, as concepts, essentially these identities of the determinations
with themselves in their restriction or their otherness; but this identity was
only implicitly the concept, was not yet made manifest. Consequently, the
qualitative determination perished as such in its other and had as its truth
a determination diverse from it. The universal, on the contrary, even when
it posits itself in a determination, remains in it what it is. It is the soul
of the concrete which it inhabits, unhindered and equal to itself in its
manifoldness and diversity. It is not swept away in the becoming but persists
undisturbed through it, endowed with the power of unalterable, undying
self-preservation.
It also does not simply shine reflectively in its other, as does the deter-
mination of reflection. This determination, as something relative, does not
refer only to itself but is a relating. It lets itself be known in its other, but
at first it only shines reflectively in it, and this reflective shining of each in
the other, or their reciprocal determination, has the form of an external
activity alongside their self-subsistence. – The universal is posited, on the
contrary, as the essence of its determination, as this determination’s own
positive nature. For the determination that constitutes the negative of the
universal is in the concept simply and solely a positedness; essentially, in
other words, it is at the same time the negative of the negative, and only
is as this self-identity of the negative which is the universal. To this extent,
the universal is also the substance of its determinations, but in such a way
that what for the substance as such was an accident, is the concept’s own
self mediation, its own immanent reflection. But this mediation, which first
raises the accidental to necessity, is the manifested reference; the concept is
not the abyss of formless substance, or the necessity which is the inner iden-
tity of things or circumstances different from each other and reciprocally
constricting; rather, as absolute negativity, it is the informing and creative
principle, and since the determination is not as limitation but is just as
much simply sublated as determination, is positedness, so is the reflective
shine the appearance as appearance of the identical.
The universal is therefore free power; it is itself while reaching out to its
other and embracing it, but without doing violence to it; on the contrary,
it is at rest in its other as in its own. Just as it has been called free power, it
could also be called free love and boundless blessedness, for it relates to that
which is distinct from it as to itself; in it, it has returned to itself.
Mention has just been made of determinateness, even though the concept
has not yet progressed to it, being at first only as the universal and only self-
identical. But one cannot speak of the universal apart from determinateness
which, to be more precise, is particularity and singularity. For in its absolute
negativity the universal contains determinateness in and for itself, so that,
when speaking of determinateness in connection with the universal, the
determinateness is not being imported into the latter from outside. As
negativity in general, that is, according to the first immediate negation, the
universal has determinateness in it above all as particularity; as a second
universal, as the negation of negation, it is absolute determinateness, that
is, singularity and concreteness. – The universal is thus the totality of the
concept; it is what is concrete, is not empty but, on the contrary, has content
by virtue of its concept – a content in which the universal does not just
preserve itself but is rather the universal’s own, immanent to it. It is of
course possible to abstract from this content, but what we have then is not
the universal element of the concept but the abstract universal, which is an
isolated and imperfect moment of the concept, void of truth.
More precisely, the universal shows itself to be this totality as follows.
In so far as the universal possesses determinateness, this determinateness
is not only the first negation but also the reflection of this negation into
itself. According to that first negation, taken by itself, the universal is a
particular, and in this guise we shall consider it in a moment. In the other
determinateness, however, the universal is still essentially universal, and
this side we have here still to consider. – For this determinateness, as it
is in the concept, is the total reflection – a doubly reflective shine, both
outwards, as reflection into the other, and inwards, as reflection into itself.
The outward shining establishes a distinction with respect to an other; the
universal accordingly takes on a particularity which is resolved in a higher
universality. Inasmuch as it now is also only a relative universal, it does not
lose its character of universality; it preserves itself in its determinateness,
not just because it remains indifferent to it – for then it would be only
posited together with it – but because of what has just been called the inward
shining. The determinateness, as determinate concept, is bent back into itself;
it is the concept’s own immanent character, a character made essential by
being taken up into the universality and by being pervaded by it, just as
it pervades it in turn, equal in extension and identical with it. This is
the character that belongs to the genus as the determinateness which is
not separated from the universal. To this extent, it is not an outwardly
directed limitation, but is positive, for by virtue of the universality it stands
in free self-reference. Thus even the determinate concept remains in itself
infinitely free concept.
But in regard to the other side in which the genus is limited because of
its determinate character, we have just said that, as a lower genus, it has
its resolution in a higher universal. This universal can also be grasped as
a genus but as a more abstract one; it always pertains, however, only to
the side of the determinate concept which is outwardly directed. The truly
higher universal is the one in which this outwardly directed side is redirected
inwardly; this is the second negation in which the determinateness is present
simply and solely as something posited, or as reflective shine. Life, the “I,”
spirit, absolute concept, are not universals only as higher genera, but are
rather concretes whose determinacies are also not mere species or lower
genera but determinacies which, in their reality, are self-contained and self-
complete. Of course, life, the “I,” finite spirit, are also only determinate
concepts. To this extent, however, their resolution is in a universal which,
as the truly absolute concept, is to be grasped as the idea of infinite spirit –
the spirit whose posited being is the infinite, transparent reality in which it
contemplates its creation and, in this creation, itself.
The true, infinite universal, the one which, immediately in itself, is just
as much particularity as singularity, is now to be more closely examined
as particularity. It determines itself freely; the process by which it becomes
finite is not a transition, the kind that occurs only in the sphere of being;
it is creative power as self-referring absolute negativity. As such, it differen-
tiates itself internally, and this is a determining, because the differentiating
is one with the universality. Accordingly, it is a positing of differences
that are themselves universals, self-referring. They become thereby fixed,
isolated differences. The isolated subsistence of the finite that was earlier
determined 32 as its being-for-itself, also as thinghood, as substance, is in
its truth universality, the form with which the infinite concept clothes its
differences – a form which is equally itself one of its differences. Herein
consists the creativity of the concept, a creativity which is to be compre-
hended only in the concept’s innermost core.

b. the particular concept

Determinateness as such belongs to being and the qualitative; as the deter-
minateness of the concept, it is particularity. It is not a limit, as if it were
related to an other beyond it, but is rather, as just shown, the universal’s
own immanent moment; in particularity, therefore, the universal is not in
an other but simply and solely with itself.
The particular contains the universality that constitutes its substance;
the genus is unaltered in its species; these do not differ from the uni-
versal but only from each other. The particular has one and the same
universality as the other particulars to which it is related. The diversity
of these particulars, because of their identity with the universal, is as
such at the same time universal; it is totality. – The particular, there-
fore, does not only contain the universal but exhibits it also through its
determinateness; accordingly the universal constitutes a sphere that the par-
ticular must exhaust. This totality, inasmuch as the determinateness of
the particular is taken as mere diversity, appears as completeness. In this
respect, the species are complete simply in so far as there are no more
of them. There is no inner standard or principle available for them, for
their diversity is just the dispersed 33 difference for which the universality,
which is for itself absolute unity, is a merely external reflex and an uncon-
strained, contingent completeness. But diversity passes over into opposition,
into an immanent connection of diverse moments. Particularity, however,
because it is universality, is this immanent connection, not by virtue of
a transition, but in and for itself. It is totality intrinsically, and simple
determinateness, essential principle. It has no other determinateness than
that posited by the universal itself and resulting from it in the following
manner.
The particular is the universal itself, but it is its difference or reference
to an other, its outwardly reflecting shine; but there is no other at hand from
which the particular would be differentiated than the universal itself. – The
universal determines itself, and so is itself the particular; the determinateness
is its difference; it is only differentiated from itself. Its species are therefore
only (a) the universal itself and (b) the particular. The universal is as concept
itself and its opposite, and this opposite is in turn the universal itself as
its posited determinateness; the universal overreaches it and, in it, it is
with itself. Thus it is the totality and the principle of its diversity, which is
determined wholly and solely through itself.
There is, therefore, no other true logical division than this, that the
concept sets itself on one side as the immediate, indeterminate universality;
it is this very indeterminateness that makes its determinateness, or that it
is a particular. The two are both a particular and are therefore coordinated.
Both, as particular, are also determinate as against the universal, and in
this sense they are subordinated to it. But even this universal, as against
which the particular is determined, is for that reason itself also just one
of the opposing sides. When we speak of two opposing sides, we must
repeat that the two constitute the particular, not just together, as if they
were alike in being particular only for external reflection, but because
their determinateness over against each other is at the same time essentially
only one determinateness; it is the negativity which in the universal is
simple.
Difference, as it presents itself here, is in its concept and therefore in
its truth. All previous difference has this unity in the concept. As it is
present immediately in being, difference is the limit of an other; as present
in reflection, it is relative, posited as referring essentially to its other; here is
where the unity of the concept thus begins to be posited; at first, however,
the unity is only a reflective shine in an other. – The true significance of the
transitoriness and the dissolution of these determinations is just this, that
they attain to their concept, to their truth; being, existence, something, or
whole and part, and so on, substance and accidents, cause and effect, are
thought determinations on their own; as determinate concepts, however,
they are grasped in so far as each is cognized in unity with its others or
in opposition to them. – Whole and parts, for example, or cause and
effect, and so on, are not yet diverse terms that are determined as particular
relatively to each other, for although they implicitly constitute one concept,
their unity has not yet attained the form of universality; thus the difference
as well which is in these relations, does not yet have the form of being
one determinateness. Cause and effect, for example, are not two diverse
concepts but only one determinate concept, and causality is, like every
concept, a simple concept.
With respect to completeness, we have just seen that the determinate
moment 34 of particularity is complete in the difference of the universal and
the particular, and that only these two make up the particular species. To
be sure, there are more than two species to be found in any genus in nature,
and these many species cannot stand in the same relation to each other as
we have shown. This is the impotence of nature, that it cannot abide by
and exhibit the rigor of the concept and loses itself in a blind manifoldness
void of concept. We can wonder at nature, at the manifoldness of its genera
and species, in the infinite diversity of its shapes, for wonder is without
concept and its object is the irrational. It is allowed to nature, since nature
is the self-externality of the concept, to indulge in this diversity, just as
spirit, even though it possesses the concept in the shape of concept, lets
itself go into pictorial representation and runs wild in the infinite mani-
foldness of the latter. The manifold genera and species of nature must not
be esteemed to be anything more than arbitrary notions of spirit engaged
in pictorial representations. Both indeed show traces and intimations of
the concept, but they do not exhibit it in trustworthy copy, for they are the
sides of its free self-externality; the concept is the absolute power precisely
because it can let its difference go free in the shape of self-subsistent diver-
sity, external necessity, accidentality, arbitrariness, opinion – all of which,
however, must not be taken as anything more than the abstract side of
nothingness.
As we have just seen, the determinateness of the particular is simple as
principle, but it is also simple as a moment of the totality, determinateness
as against the other determinateness. The concept, in determining or dif-
ferentiating itself, behaves negatively towards its unity and gives itself the
form of one of its ideal moments of being; as a determinate concept, it has
a determinate existence in general. But this being no longer has the signifi-
cance of mere immediacy, but has the significance rather of an immediacy
which is equal to itself by virtue of absolute mediation, an immediacy that
equally contains in itself the other moment of essence or of reflection. This
universality, with which the determinate clothes itself, is abstract univer-
sality. The particular has this universality in it as its essence; but in so far
as the determinateness of the difference is posited and thereby has being,
the universality is form in it, and the determinateness as such is its con-
tent. Universality becomes form inasmuch as the difference is something
essential, just as in the pure universal it is, on the contrary, only absolute
negativity and not a difference posited as such.
Now the determinateness is indeed an abstraction, as against the other
determinateness; but the other determinateness is only universality itself,
and this too is therefore abstract universality; and the determinateness
of the concept, or particularity, is again nothing more than determinate
universality. In this universality, the concept is outside itself, and because it is
it, the concept, which is there outside itself, the abstract-universal contains all
the moments of the concept. It is (a) universality, (b) determinateness, (g)
the simple unity of the two; but this unity is immediate, and the particularity
is not therefore as totality. Implicitly it is this totality also, and mediation;
it is essentially a reference to the other excluding it, or the sublation of
negation, namely of the other determinateness – an other that lingers on
only as an intention, for it vanishes immediately revealing itself to be the
same as its other is supposed to be. Therefore, what makes this universality
an abstraction is that the mediation is only a condition, or is not posited
in it. Because it is not posited, the unity of the abstraction has the form of
immediacy, and the content has the form of indifference to its universality,
for the content is nothing but this totality which is the universality of
absolute negativity. Hence the abstract universal is indeed the concept, but
the unconceptualized concept, the concept not posited as such.
When we speak of the determinate concept, what we ordinarily mean is
precisely just this abstract universal. Even by concept as such, what is gener-
ally understood is only this unconceptualized concept, and the understand-
ing is designated as its faculty. Demonstration belongs to this understanding
inasmuch as it proceeds by way of concepts, that is to say, only in deter-
minations. This progression by way of concepts does not therefore reach
past finitude and necessity; the highest it reaches is the negative infinite,
the abstraction of the highest essence which is itself the determinateness
of the indeterminateness. Absolute substance, too, although not this empty
abstraction but on the contrary a totality according to content, is still
abstract, for since it is without absolute form, its innermost truth is not
constituted by the concept; although it is the identity of universality and
particularity, or of thought and externality, this identity is not the deter-
minateness of the concept; there is rather an understanding outside it – an
understanding which is contingent precisely because it is outside it – in
which and for which substance exists in diverse attributes and modes.
Moreover, abstraction is not as empty as it is usually said to be; it is the
determinate concept; it has some determinateness or other for its content;
the highest essence also, the pure abstraction, has the determinateness of
indeterminateness, as just mentioned; but indeterminateness is a determi-
nateness because it is supposed to stand opposite the determinate. But the
moment one says what it is, its intended meaning sublates itself by itself;
for it is spoken of on a par with determinateness, and from this abstraction
the concept and its truth are brought out. – To be sure, any determinate
concept is empty in so far as it does not contain the totality, but only a
one-sided determinateness. Even when it has otherwise concrete content
such as, for instance, humankind, the state, animal, etc., it remains an
empty concept inasmuch as its determinateness is not the principle of its
differentiation; the principle contains the beginning and the essence of its
development and realization; any other determinateness of the concept is
however otiose. To reproach the concept as such for being empty is to
ignore its absolute determinateness which is the difference of the concept
and the only true content in the element of the concept.
Here we have the circumstance that explains why the understanding
is nowadays held in such a low repute and is so much discredited when
measured against reason; 35 it is the fixity which it imparts to determinacies
and consequently to anything finite. This fixity consists in the form of
the abstract universality just considered that makes them unalterable. For
qualitative determinateness, and also the determination of reflection, are
essentially limited, and because of their limitation they entail a reference
to their other; hence the necessity of their transition and passing away. But
the universality which they possess in the understanding gives them the
form of immanent reflection and, because this form removes from them
the reference to the other, they have become unalterable. Now although
this eternity belongs to the pure concept by nature, the determinations
of the concept are eternal essentialities only according to form; but their
content is not proportionate to this form and, therefore, they are not truth,
or imperishable. Their content is not proportionate to the form because it
is not the determinateness itself as universal, that is, not as totality of the
difference of the concept, or is not itself the whole form; the form of the
limited understanding is for this reason itself imperfect universality, that is
to say, abstract universality. – But further, we must pay due respect to the
infinite force of the understanding in splitting the concrete into abstract
determinacies and plumbing the depth of the difference – this force which
alone is at the same time the mighty power causing the transition of
the determinacies. The concrete of intuition is a totality, but a sensuous
totality, a real material that subsists in space and time, part outside part,
each indifferent to the other; surely this lack of unity in a manifold that
makes it the content of intuition ought not to be credited as privileging it
over the universal of the understanding. The mutability that the manifold
exhibits in intuition already points to the universal; but all of the manifold
that comes to intuition is just more of the same, an equally alterable other –
not the universal that one would expect to appear and take its place. But
least of all should we reckon to the credit of such sciences as for example
Geometry and Arithmetic that their material carries an intuitive element
with it, or imagine that their propositions are grounded by it. On the
contrary, the presence of that element renders the material of these sciences
of an inferior nature; the intuition of figures or numbers is of no help to
the science of figures and numbers; only the thought of them produces this
science. – But if by intuition we understand not merely a sensuous material
but the objective totality, then the intuition is an intellectual one, that is,
its subject matter is not existence in its externalization but that element
in existence which is unalterable reality and truth – the reality only in so
far as it is essentially in the concept and is determined by it; the idea, of
whose more precise nature more will be said later. What intuition as such
is supposed to have over the concept is external reality, the reality that lacks
the concept and receives value only through the concept.
Consequently, since the understanding exhibits the infinite force that
determines the universal, or conversely, since it is the understanding that
through the form of universality imparts stable subsistence to the otherwise
inherent instability of determinateness, then it is not the fault of the
understanding if there is no further advance. It is a subjective impotence of
reason that allows these determinacies to remain so dispersed, and is unable
to bring them back to their unity through the dialectical force opposed
to that abstract universality, that is to say, through the determinacies’ own
nature which is their concept. To be sure, the understanding does give
them through the form of abstract universality a rigidity of being, so to
speak, which they do not otherwise possess in the qualitative sphere and in
the sphere of reflection; but by thus simplifying them, the understanding
at the same time quickens them with spirit, and it so sharpens them that
only at that point, only there, do they also obtain the capacity to dissolve
themselves and to pass over into their opposite. The ripest maturity, the
highest stage, that anything can attain is the one at which its fall begins.
The fixity of the determinacies which the understanding appears to run up
against, the form of the imperishable, is that of self-referring universality.
But this universality belongs to the concept as its own, and for this reason
what is found expressed in it, infinitely close at hand, is the dissolution of
the finite. This universality directly contradicts the determinateness of the
finite and makes explicit its disproportion with respect to it. – Or rather,
that disproportion is already at hand; the abstract determinate is posited as
one with universality and, for this reason, not for itself (for it would then
be only a determinate) but, on the contrary, only as the unity of itself and
the universal, that is, as concept.
Therefore the common practice of separating understanding and reason
is to be rejected on all counts. On the contrary, to consider the concept
as void of reason should itself be considered as an incapacity of reason to
recognize itself in the concept. The determinate and abstract concept is
the condition, or rather an essential moment, of reason; it is form quickened
by spirit in which the finite, through the universality in which it refers
to itself, is internally kindled, is posited as dialectical and thereby is the
beginning of the appearance of reason.
Since in the foregoing the determinate concept has been presented in its
truth, it is only left to indicate what, as so presented, it has already been
posited as. – Difference, which is an essential moment of the concept but
in the pure universal is not yet posited as such, receives its due in the deter-
minate concept. Determinateness in the form of universality is united with
the latter to form a simple; determinate universality is self-referring deter-
minateness, determinate determinateness or absolute negativity posited for
itself. But self-referring determinateness is singularity. Just as universality
immediately is particularity in and for itself, no less immediately is par-
ticularity also singularity in and for itself; this singularity is at first to be
regarded as the third moment of the concept, inasmuch as it is held fast
in opposition to the other two, but also as the absolute turning back of the
concept into itself, and at the same time as the posited loss of itself.
Remark
Universality, particularity, and singularity are, according to the forego-
ing, the three determinate concepts, that is, if one wants to count them.
We have already shown that number is a form unsuited to conceptual
determinations, 36 but for the determination of the concept itself it is
unsuited the most; number, since the unit is its principle, turns the counted
36
Cf. GW 11, 129–130; above 21.203ff.The concept
541
into totally separated units indifferent to each other. We have seen from the
foregoing that the diverse determinate concepts, rather than falling apart
as they do when counted, are only one and the same concept.
In the customary treatment of logic, a variety of classifications and species
of concepts are adduced. It immediately strikes one as inconsequential
that the species are introduced in this way: “There are, as regards quality,
quantity, etc., the following concepts.” The “there are” conveys no other
justification than that we find the named species and that they show up
in experience. What we have in this manner is an empirical logic – an odd
science indeed, an irrational cognition of the rational. In this the logic sets a
very bad precedent for compliance to its own teaching; it allows itself to do
the opposite of what it prescribes as a rule, namely, that concepts should be
derived, and scientific propositions (therefore also the proposition: “There
are such and such species of concepts”) demonstrated. – In this context,
the Kantian philosophy incurs a further inconsequence by borrowing the
categories for the transcendental logic, as so-called root concepts, from the
subjective logic where they were assumed empirically. 37 Since the Kantian
philosophy admits the latter fact, it is hard to see why transcendental logic
resorts to borrowing from such a science rather than directly helping itself
from experience.
By way of example, concepts are normally classified according to their
clarity, namely, as clear and obscure, distinct and indistinct, adequate and
inadequate. We can also add to the list perfect and redundant and other
suchlike superfluities. – Now as regards this classification according to
clarity, it immediately transpires that this standpoint and its connected dis-
tinctions are taken from psychological and not logical determinations. The
so-called clear concept is supposed to be one that suffices to differentiate
one intended object from another. But this cannot be called a concept yet;
it is nothing more than a subjective representation. What an obscure concept
might be must be left to itself, for otherwise it would not be obscure but
a distinct concept. – The distinct concept is supposed to be one whose
mark can be given. But then it is, strictly speaking, the determinate concept.
The mark, when taken in its strict signification, is nothing else than the
determinateness or the simple content of the concept in so far as the latter is
distinguished from the form of universality. But the mark does not quite
have at first this more precise meaning; it is generally taken as only a deter-
mination by which a third party takes note of a subject matter or of the
concept; it can therefore be a very contingent circumstance. It expresses in
37
Cf. A80–81/B106–107.
12.44542
12.45
Georg Wilhelm Friedrich Hegel
general not so much the immanence and essentiality of a determination as
its reference to an external understanding. If the latter really is an under-
standing, then it has the concept before it and makes a note of it by nothing
else than by what is in the concept itself. In case the mark is different from the
concept, then the mark is a sign or some other determination that belongs
to the representation of the matter at hand, not to its concept. – What an
indistinct concept might be, this can be passed over as superfluous.
But the adequate concept is something higher; it properly denotes the
agreement of the concept with reality, and this is not the concept as such
but the idea.
If the mark of a distinct concept were really to be the determination
itself of the concept, then logic would have trouble with the simple concept
which, according to another classification, is opposed to the composite.
For if for a simple concept a true, that is, immanent mark is given, then
the concept is no longer regarded as simple; and if no mark is given for
it, then the concept is not distinct. But the clear concept now comes to
the rescue. Unity, reality, and suchlike determinations, are supposed to be
simple concepts, perhaps because logicians were unable to come up with a
determination for them and had to be content, therefore, with just a clear
concept of them, that is to say, with no concept at all. A definition, that
is, the statement of a concept, requires as a general rule the statement of
genus and specific difference. It thus presents the concept, not as something
simple, but in two enumerable components. Yet surely nobody will suppose
that the concept is for that reason a composite. – There is an allusion in the
mention of the simple concept to abstract simplicity, to a unity that does
not entail difference and determinateness – a unity, therefore, that does not
pertain to the concept. Inasmuch as an object is present in representation,
especially in memory, or is also an abstract thought determination, it can
be quite simple. Even the object that is richest in content, as for example
spirit, nature, world, even God, when non-conceptually apprehended in
a simple representation of the equally simple expression: spirit, nature,
world, God, is of course something simple at which consciousness can stop
short without proceeding to extract the proper determination or a defining
mark. But the objects of consciousness ought not to remain so simple,
ought not to remain representations or abstract thought determinations,
but should rather become conceptualized, that is, their simplicity should
be determined together with their inner difference. – A composite concept,
however, is but the equivalent of a wooden iron. We can of course have the
concept of a composition; but a composite concept would be something
worse than materialism, which assumes only the substance of the soul to beThe concept
543
a composite, yet takes thought to be simple. The uneducated reflection first
stumbles upon the notion of composition because it is the most completely
external connection, the worst form in which things can be considered; even
the lowest of natures must be an inner unity. That, to top it off, this form
of untruest existence would be extended to the “I,” to the concept, is more
than one should have expected and it must be regarded as an inept form
of barbarism.
Concepts are further divided into contrary and contradictory, a principal
division. – If the task of a treatise on the concept were to give all the
determinate concepts that there are, then we would have to cite all possible
determinations – for all determinations are concepts, hence determinate
concepts – and we would have to list as species of concepts all the categories
of being as well as all the determinations of essence. And this is what is done
in the textbooks on logic, where we are told – in greater or lesser detail,
according to the whim of the author – that there are affirmative, negative,
identical, conditional, necessary concepts. But these determinations have
already been left behind by the nature of the concept itself and are therefore
misplaced when applied to the concept itself, thus admitting only the kind
of superficial nominal definitions that are of no interest here. – The under-
lying basis of the distinction between contrary and contradictory concepts
with which we are specifically dealing here is the reflective determination
of diversity and opposition. They are viewed as two particular species, each
fixed for itself and indifferent towards the other, without any thought being
given to the dialectic and the inner nothingness of these differences, as if
that which is contrary would not equally have to be determined as contra-
dictory. The nature and the transition essential to the forms of reflection
which they express have been considered in their proper place. 38 In the
concept, identity has developed into universality, difference into particu-
larity, opposition (which returns to the ground) into singularity. In these
forms, those determinations of reflection are present as they are in their
concept. The universal has proved itself to be not only the identical, but
at the same time the diverse or contrary as against the particular and the
singular, and then also to be opposed to them, or contradictory; but in this
opposition it is identical with them, and it is their true ground in which
they are sublated. The same applies to particularity and singularity, which
are likewise the totality of the determinations of reflection.
Concepts are further divided into subordinate and coordinate – a distinc-
tion that comes closer to a determination of the concept, namely that of
38
Cf. above, 11.267–278.
12.46544
12.47
Georg Wilhelm Friedrich Hegel
the relation of universality and particularity, in the context of which we did
mention them albeit incidentally. 39 But it is customary to consider them
like the rest as rigidly fixed relations, and then to produce several otiose
propositions regarding them. The most long-winded disquisition in this
regard has to do again with the connection of contrariety and contradiction
with subordination and coordination. Since judgment is the connection of
determinate concepts, the true relation will have to come first into view only
with reference to it. That fashion of comparing these determinations with-
out a thought to their dialectic and to the continuing alteration of their
determination, or rather to the conjunction in them of opposite determina-
tions, makes the whole disquisition of what may or may not be consonant in
them – as if this consonance or dissonance were something separate by itself
and permanent – into an otiose exercise void of content. – The great Euler,
infinitely fertile and sharp of mind in detecting and arranging the deep
relations of algebraic quantities, 40 the dry, prosaic Lambert in particular, 41
and others, have attempted to construct a notation for this class of relations
between determinations of the concept based on lines, figures, and the like,
the general intention being to elevate – or in fact rather to debase – the
logical modes of relation to the status of a calculus. One need only compare
the nature of a sign with what the sign ought to indicate immediately to
see that even the project of a logical notation is unworkable. The determi-
nations of the concept, universality, particularity, and singularity, certainly
are, like lines or the letters of algebra, diverse; and they are also opposed
and allow, therefore, the signs of plus and minus. But they themselves and
especially their connections, even if we stop short at just subsumption and
inherence, are in their essential nature entirely different from algebraic let-
ters, from lines and their connections, from the equality and diversity of
magnitudes, the plus and minus, or the superimposition of lines, or the
joining of them in angles and the resulting disposition of space that they
enclose. It is characteristic of objects of this kind, as contrasted with the
determinations of the concept, that they are mutually external, that they
have a fixed determination. Now when concepts are made to conform to
such signs, they cease to be concepts. Their determinations are not inert
things, like numbers and lines whose connections lie outside them; they
39
40
41
Cf above, 12.38.
See Leonhard Euler, Lettres à une princesse allemande sur divers sujets de physique et de philosophie
(Leipzig: Dyck, 1792–4), Letters 52–55.
See Johann Heinrich Lambert, Neues Organon oder Gedanken über die Erforschung und Bezeichnung
des Wahren und dessen Unterscheidung vom Irrtum und Schein, Vol. 1 (Berlin: Akademie, 1990),
§229f.The concept
545
are living movements; the distinguished determinateness of the one side is
immediately also internal to the other side; what would be a complete
contradiction for numbers and lines is essential to the nature of the
concept. – The higher mathematics, which also proceeds to the infinite
and allows itself contradictions, can no longer employ its customary signs
for representing such determinations. In order to indicate the still concep-
tually uncomprehended representation of the infinite approximation of two
ordinates, or when it equates a curve to an infinite number of infinitely
small straight lines, all it does is to design two straight lines outside each
other or to draw straight lines inside but still distinct from a curve; for the
infinite, which is the point at issue here, higher mathematics falls back on
pictorial representation.
What first led to this wayward attempt is above all the quantitative
relation in which universality, particularity, and singularity are supposed
to stand to one another: universal means, more extensive than particular
and singular; and particular, more extensive than singular. The concept
is concrete and the richest in determination, because it is the ground and
the totality of the previous determinations, of the categories of being and
the determinations of reflection; these, therefore, are certain also to come
up in it. But its nature is totally misunderstood if such determinations are
retained in it in their former abstraction – if the wider extent of the universal
is understood to mean that the universal is a more, or a greater quantum,
than the particular and the singular. As absolute ground, it is the possibility
of quantity, but no less so of quality, that is, its determinations are no less
qualitatively distinct; therefore they are already viewed in contravention to
their truth when they are posited in the form of quantity alone. So, too,
a reflective determination is a relative, something in which the opposite
shines reflectively; unlike a quantum, its relation is not external. But the
concept is more than all this; its determinations are determinate concepts,
themselves essentially the totality of all determinations. It is, therefore,
entirely inappropriate, in order to grasp such an inner totality, to want to
apply numerical and spatial relations in which the terms fall apart; such
relations are rather the last and the worst medium that could be used.
Natural relations, as for instance “magnetism” or “color tonality” would
make infinitely higher and truer symbols for the purpose. Since the human
being has in language a means of designation that is appropriate to reason,
it is otiose to look for a less perfect means of representation to bother oneself
with. It is essentially only spirit that can grasp the concept as concept, for
the latter is not just the property of spirit but its pure self. It is futile to
want to fix it by means of spatial figures and algebraic signs for the sake
12.48546
12.49
Georg Wilhelm Friedrich Hegel
of the outer eye and a non-conceptual, mechanical manipulation, such as a
calculus. Also anything else that might be supposed to serve as symbol, like
the symbols for the nature of God, can at best elicit only intimations and
echoes of the concept; if, however, one insists on employing such symbols
for expressing and cognizing the concept, then it is not their external nature
which is fit for the task; the reverse relation applies, namely that what in the
symbols is the echo of a higher determination is recognized to be such only
by virtue of the concept, and it is only by shedding the sensuous standbys
that were supposed to express it that one comes closer to the concept.
c. the singular
Singularity, as we have seen, 42 is already posited through particularity; this
is determinate universality and hence self-referring determinateness, the
determinate determinate.
1. At first, therefore, singularity appears as the reflection of the concept out
of its determinateness into itself. It is the concept’s self-mediation by virtue of
which, since its otherness has once more been made into an other, it restores
itself as self-equal, but in the determination of absolute negativity. – The
negative in the universal, by virtue of which this universal is a particular,
was earlier determined as a doubly reflective shine. 43 In so far as the reflective
shining is inward, the particular remains a universal; through the outward
shining, it is a determinate particular; the turning back of this side into the
universal is twofold, either by virtue of an abstraction that lets the particular
fall away and climbs to a higher and the highest genus, or by virtue of the
singularity to which the universality in the determinateness itself descends. –
Here is where the false start is made that makes abstraction stray away
from the way of the concept, abandoning the truth. Its higher and highest
universal to which it rises is only a surface that becomes progressively more
void of content; the singularity which it scorns is the depth in which the
concept grasps itself and where it is posited as concept.
Universality and particularity appeared, on the one hand, as moments of
the becoming of singularity. But it has already been shown that the two are
in themselves the total concept; consequently, that in singularity they do
not pass over into an other but that, on the contrary, what is posited in it is
what they are in and for themselves. 44 The universal is for itself because it
is absolute mediation in itself, self-reference only as absolute negativity. It
is an abstract universal inasmuch as this sublating is an external act and so
42
Cf. above, 12.43.
43
Cf. above, 12.35.
44
Cf. above, 12.35, 37.The concept
547
a dropping off of the determinateness. This negativity, therefore, attaches
indeed to the abstract universal, but it remains outside it, as a mere condition
of it; it is the abstraction itself that holds its universal opposite it, and so the
universal does not have singularity in itself and remains void of concept. –
Life, spirit, God, as well as the pure concept, are for this reason beyond
the grasp of abstraction, for abstraction keeps singularity away from its
products, and singularity is the principle of individuality and personality.
And so it comes to nothing but lifeless universalities, void of spirit, color,
and content.
But the unity of the concept is so indissoluble that these products of
abstraction also, though they are supposed to drop singularity, are rather
themselves singulars. For in elevating the concrete to universality, abstrac-
tion grasps the universal as only a determinate universality, and this is
precisely the singularity that presented itself as self-referring determinate-
ness. Thus abstraction is a partitioning of the concrete and an isolating of its
determinations; only singular properties or moments are picked out by it,
for its product must contain what it itself is. But the difference between this
singularity of its products and the singularity of the concept is that in the
former the singular and the universal differ from each other as content and
form respectively, precisely because the content is not the absolute form, is
not the concept itself, or this form is not the totality of form. – However,
this closer consideration shows that the product of abstraction is itself the
unity of the singular content and of abstract universality, therefore that it
is something concrete, the opposite of what it is supposed to be.
The particular, for the same reason that makes it only a determinate uni-
versal, is also a singular, and conversely, because the singular is a determinate
universal, it is equally a particular. If we stay at this abstract determinate-
ness, then the concept has the three particular determinations of universal,
particular, and singular, whereas earlier we gave only the universal and the
particular as species of the particular. Because singularity is the turning of
the concept as a negative back to itself, this turning back from abstraction,
which in the turning is truly sublated, can itself be placed as an indifferent
moment alongside the others and be counted with them.
If singularity is listed as one of the particular determinations of the
concept, then particularity is the totality which embraces them all and,
precisely as this totality, it is the concretion of the determinations or singu-
larity itself. But it is a concrete also according to the previously mentioned
side, as determinate universality; and then it is the immediate unity in
which none of these moments is posited as distinct or as the determinant,
and in this form it will constitute the middle term of the formal syllogism.
12.50548
12.51
Georg Wilhelm Friedrich Hegel
It follows that each of the determinations established in the preceding
exposition of the concept has immediately dissolved itself and has lost
itself in its other. Each distinction is confounded in the course of the very
reflection that should isolate it and hold it fixed. Only a way of thinking that
is merely representational, for which abstraction has isolated them, is capable
of holding the universal, the particular, and the singular rigidly apart. Then
they can be counted; and for a further distinction this representation relies
on one which is entirely external to being, on their quantity, and nowhere
is such a distinction as inappropriate as here. – In singularity, the earlier
true relation, the inseparability of the determinations of the concept, is
posited; for as the negation of negation, singularity contains the opposition
of those determinations and this opposition itself at its ground or the unity
where the determinations have come together, each in the other. Because
in this reflection universality is in and for itself, singularity is essentially
the negativity of the determinations of the concept, but not merely as if it
stood as a third something distinct from them, but because what is now
posited is that positedness is being-in-and-for-itself; that is, what is posited is
that each of the distinct determinations is the totality. The turning back of
the determinate concept into itself means that its determination is to be in
its determinateness the whole concept.
2. Singularity is not, however, only the turning back of the concept into
itself, but the immediate loss of it. Through singularity, where it is inter-
nal to itself, the concept becomes external to itself and steps into actuality.
Abstraction, which is the soul of singularity and so the self-reference of the
negative, is, as we have seen, 45 nothing external to the universal and the par-
ticular but is immanent in them, and these are concreted through it, they
become a content, a singular. But, as this negativity, singularity is the deter-
minate determinateness, differentiation as such, and through this reflection
of the difference into itself, the difference becomes fixed; the determining
of the particular occurs only by virtue of singularity, for singularity is that
abstraction which, precisely as singularity, is now posited abstraction.
The singular, therefore, is as self-referring negativity the immediate iden-
tity of the negative with itself; it exists for itself. Or it is the abstraction
determining the concept as an immediate, according to its ideal moment of
being. – Thus the singular is a one which is qualitative, or a this. In accor-
dance with this qualitative character, it is, first, the repulsion of itself from
itself by virtue of which many other ones are presupposed; second, it is now
a negative reference with respect to these presupposed others, and to this
45
i.e., just above, 12.49–50.The concept
549
extent the singular is exclusive. Universality, when referred to these singulars
as indifferent ones – and it must be referred to them, for they are a moment
of the concept of singularity – is only their commonality. If by the universal
one understands that which is common to several singulars, the indifferent
subsistence of these singulars is then taken as the starting point, thus mixing
in the immediacy of being into the determination of the concept. The low-
est conception one can have of the universal as connected with the singular
is this external relation that it has to the latter as a mere commonality.
The singular, which in the reflective sphere of concrete existence is as a
this, does not have the excluding reference to the other that is characteristic
of qualitative being-for-itself. A this is a one reflected into itself, without
repulsion; or the repulsion is in this reflection one with abstraction, a
reflective mediation present in the this that makes it a posited immediacy
pointed at by someone external to it. The this is; it is immediate, it is a
this, however, only in so far as it is pointed at. This “pointing at” is the
reflective movement that takes hold of itself and posits the immediacy, but
as something external to itself. – Now the singular surely is also a this,
as an immediate which is the result of mediation, but does not have this
mediation outside it; it is itself repelling separation, posited abstraction, yet
is, precisely in its separation, a positive connection.
This act of abstraction by the singular is, as the immanent reflection
of difference, the first positing of the differences as self-subsisting, reflected
into themselves. They exist immediately; but, further, this separating is
reflection in general, the reflective shining of one in the other; the differences
thus stand in essential relation. They are, moreover, not singulars that
just exist next to each other; a plurality of this kind belongs to being; the
singularity that posits itself as determinate does not posit itself in an external
difference but in a difference of the concept; singularity thus excludes the
universal from itself, but since this universal is a moment of it, it refers to
it just as essentially.
The concept, as this connection of its self-subsistent determinations, has
lost itself, for the concept itself is no longer the posited unity of these deter-
minations, and these no longer are moments, the reflective shining of the
concept, but subsist rather in and for themselves. – As singularity, the con-
cept returns in determinateness into itself, and therewith the determinate
has itself become totality. The concept’s turning back into itself is thus the
absolute, originative partition of itself, that is, as singularity it is posited as
judgment. 46
46
There is a play on words here: “partition” = Teilung; “judgment” = Urteil.
12.52ch a p ter 2
Judgment
12.53
Judgment is the determinateness of the concept posited in the concept itself.
The determinations of the concept, or, what amounts to the same thing
as shown, 47 the determinate concepts, have already been considered on
their own; but this consideration was rather a subjective reflection or a
subjective abstraction. But the concept is itself this act of abstracting;
the positioning of its determinations over against each other is its own
determining. Judgment is this positing of the determinate concepts through
the concept itself.
Judging is therefore another function than conceiving; or rather, it is the
other function of the concept, for it is the determining of the concept through
itself. The further progress of judgment into a diversity of judgments is
this progressive determination of the concept. What kind of determinate
concepts there are, and how they prove to be necessary determinations of
it – this has to be exhibited in judgment.
Judgment can therefore be called the first realization of the concept,
for reality denotes in general the entry into existence as determinate being.
More precisely, the nature of this realization has presented itself in such
a way that the moments of the concept are totalities which, on the one
hand, subsist on their own through the concept’s immanent reflection or
through its singularity; on the other hand, however, the unity of the concept
is their connection. 48 The immanently reflected determinations are determi-
nate totalities that exist just as essentially disconnected, indifferent to each
other, as mediated through each other. The determining itself is a totality
only as containing these totalities and their connections. This totality is the
judgment. – The latter contains, therefore, the two self-subsistents which
go under the name of subject and predicate. What each is cannot yet be said;
they are still indeterminate, for they are to be determined only through the
judgment. Inasmuch as judgment is the concept as determinate, the only
47
Cf. above, 12.46.
48
Cf. above, 12.51.
550Judgment
551
determination at hand is the difference that it contains between determinate
and still indeterminate concept. As contrasted to the predicate, the subject
can at first be taken, therefore, as the singular over against the universal, or
also as the particular over against the universal, or the singular over against
the particular; so far, they stand to each other only as the more determinate
and the more universal in general.
It is therefore fitting and unavoidable to have these names, “subject”
and “predicate,” for the determinations of the judgment; as names, they
are something indeterminate, still in need of determination, and therefore
nothing but names. It is partly for this reason that the determinations
themselves of the concept could not be used for the two sides of judgment;
but a still stronger reason is because of the nature of a concept determination
which is nothing abstract, nothing fixed, but contains its opposite in it,
explicitly posited there; since the sides of the judgment are themselves
concepts and therefore the totality of the determinations of the concept,
each side must run through all these determinations, exhibiting them
within whether in abstract or concrete form. But now, if in this altering of
determination we want to fix the two sides in some general way, names will
be the most useful means, for they can be kept the same throughout the
process. – But a name remains distinct from the fact or the concept, and this
is a distinction that transpires within the judgment as such; since the subject
is in general the determinate term and more, therefore, of an immediate
existent, whereas the predicate expresses the universal, the essence or the
concept, the subject as such is at first only a kind of name; what it is, is first
enunciated only by the predicate which contains being in the sense of the
concept. When we ask, “What is this?,” or “What kind of plant is this?,”
the being we are enquiring about is often just a name, and once we learn
this name, we are satisfied that we now know what the fact is. This is being
in the sense of the subject. The concept, however, or at least the essence and
the universal in general, is only given by the predicate, and when we ask for
it, we do it in the sense of the judgment. – God, therefore, or spirit, nature,
or what have you, is as the subject of a judgment only a name at first; what
any such subject is in accordance with the concept, is first found only in
the predicate. When we ask for the predicate that belongs to such subjects,
the required judgment must be based on a concept that is presupposed; yet
it is the predicate that first gives this concept. It is, therefore, the mere
representation that in fact makes up the presupposed meaning, and this
yields only a nominal definition whereby it is a mere accident, a historical
fact, what is understood by a name. So many disputes about whether a
predicate does or does not belong to a subject are, therefore, nothing more
12.54552
12.55
Georg Wilhelm Friedrich Hegel
than verbal disputes, for they proceed from this form; what lies at the base
(subjectum, Ëpokeimenon) is still nothing more than a name.
Secondly, we now have to examine more closely how the connection of
subject and predicate in judgment is determined, and how the two are
themselves thereby determined. Judgment has in general totalities for its
sides, totalities that are at first essentially self-subsistent. The unity of the
concept is at first, therefore, only a connection of self-subsistent terms; it is
not yet the concrete, the fulfilled unity that has returned into itself from this
reality but is a unity rather outside which the two terms persist as extremes
yet unsublated in it. – Now any consideration of the judgment can start
either from the originative unity of the concept or from the self-subsistence
of the extremes. Judgment is the self-diremption of the concept; therefore,
it is by starting from the unity of the concept as ground that the judgment is
considered in accordance with its true objectivity. In this respect, judgment
is the originative division (or Teilung, in German) of an originative unity; the
German word for judgment, Urteil (or “primordial division”), thus refers
to what judgment is in and for itself. But the concept is present in the
judgment as appearance, since its moments have attained self-subsistence
there, and it is to this side of externality that ordinary representation is more
likely to fasten.
From this subjective standpoint, the subject and the predicate are there-
fore treated as ready-made, each for itself outside the other – the subject as
a subject matter that would exist even if it did not have that predicate, and
the predicate as a universal determination that would exist even without
accruing to this subject. The act of judgment accordingly brings with it
the further reflection whether this or that predicate which is in someone’s
head can and should be attached to the subject matter that exists outside it
on its own; the judgment itself is simply the act that combines the predicate
with the subject, so that, if this combination did not occur, the subject
and predicate would still each remain what it is, the one concretely existing
as thing in itself, the other as a representation in someone’s head. – But
the predicate which is combined with the subject should also pertain to it,
which is to say, should be in and for itself identical with it. The significance
of their being combined is that the subjective sense of judgment, and the
indifferent external persistence of the subject and predicate, are again sub-
lated. Thus in “this action is good,” the copula indicates that the predicate
belongs to the being of the subject and is not merely externally combined
with it. Of course, grammatically speaking this kind of subjective relation
that proceeds from the indifferent externality of subject and predicate is
perfectly valid, for it is words that are here externally combined. – It can alsoJudgment
553
be mentioned in this context that a proposition can indeed have a subject
and predicate in a grammatical sense without however being a judgment
for that. The latter requires that the predicate behave with respect to the
subject in a relation of conceptual determination, hence as a universal with
respect to a particular or singular. And if what is said of a singular sub-
ject is itself only something singular, as for instance, “Aristotle died at the
age of 73 in the fourth year of the 115th Olympiad,” then this is a mere
proposition, not a judgment. There would be in it an element of judgment
only if one of the circumstances, say, the date of death or the age of the
philosopher, came into doubt even though the stated figures were asserted
on the strength of some ground or other. In that case, the figures would be
taken as something universal, as a time that, even without the determinate
content of Aristotle’s death, would still stand on its own filled with some
other content or simply empty. Likewise would the news that my friend
N. has died be a proposition, and a judgment only if there were a question
as to whether he is actually dead and not just apparently dead.
In the usual definition of judgment, that it is the combination of two
concepts, we may indeed accept the vague expression of “combination” for
the external copula, and also accept that the terms combined are at least
meant to be concepts. But the definition is otherwise a highly superficial
one. It is not just that in the disjunctive judgment, for instance, there are
more than two so-called concepts that are combined; more to the point is
rather that the definition is much better than the matter defined, for it is not
determinations of concepts, but determinations of representation that are in
fact meant; it was remarked in connection with the concept in general, and
with the concept as determinate, that what usually goes under this name
of concept does not deserve the name at all; 49 where should concepts then
come from in the case of judgment? – Above all this definition of judgment
ignores what is essential to it, namely the difference of its determinations;
still less does it take into account its relation to the concept.
As regards the further determination of the subject and predicate, we
have remarked above 50 that it is in judgment that they must first receive
their determination. But since judgment is the posited determinateness
of the concept, this determinateness possesses the given differences imme-
diately and abstractly as singularity and universality. – But inasmuch as
judgment is in general the immediate existence or the otherness of the con-
cept that has not yet restored itself to the unity through which it exists as
concept, there also emerges the determinateness that is void of concept,
49
Cf. above, 12.40.
50
Cf. above, 12.53, 54.
12.56554
12.57
Georg Wilhelm Friedrich Hegel
the opposition of being and reflection or the in-itself. But since the concept
constitutes the essential ground of judgment, these determinacies are at least
indifferent in the sense that, when one accrues to the subject and the other
to the predicate, the converse relation equally holds. The subject, being
the singular, appears at first as the existent or as the one that exists for itself
with the determinate determinateness of a singular on which judgment is
passed – as an actual object even when it is such in representation only –
as for instance in the case of bravery, right, agreement, etc. The predicate,
which is the universal, appears on the contrary as the reflection of this
judgment on that object, or rather as the object’s immanent reflection that
transcends the immediacy of the judgment and sublates its determinacies
as mere existents – appears, that is, as the object’s in-itselfness. – In this way,
the start is made from the singular as the first, the immediate, and through
the judgment this singular is raised to universality, just as, conversely, the
universal that exists only in itself descends in the singular into existence or
becomes a being that exists for itself.
This significance of the judgment is to be taken as its objective meaning
and at the same time as the true significance of the previous forms of
transition. The existent comes to be and becomes another, the finite passes
over into the infinite and in it passes away; the existent comes forth into
appearance out of its ground and to this ground it founders; the accidents
manifest the wealth of substance as well as its might; in being, there is
transition into an other; in essence, there is the reflective shining in an
other that manifests the necessity of a connection. This transition and this
reflective shining have now passed over into the originative division of the
concept in judgment, and this division, in bringing the singular back to the
in-itselfness of its universality, equally determines the universal as something
actual. These two are one and the same – the positing of singularity in its
immanent reflection and of the universal as determinate.
But equally pertaining to this objective meaning is that the said differ-
ences, as they re-occur in the determinateness of the concept, are at the
same time posited as only appearing, that is to say, that they are nothing
fixed but accrue rather just as much to one determination of the concept as
to the other. The subject is therefore equally to be taken as the in-itself, and
the predicate as determinate existence in contrast to it. The subject without
the predicate is what the thing without properties, the thing-in-itself, is in
the sphere of appearance, an empty indeterminate ground; it is then the
implicit concept that receives a difference and a determinateness only in
the predicate; the predicate thus constitutes the side of the determinate
existence of the subject. Through this determinate universality the subjectJudgment
555
refers to the outside, is open to the influence of other things and thereby
confronts them actively. What is there comes forth from its in-itselfness into
the universal element of combination and relations, into negative references
and into the interplay of actuality which is a continuation of the singular
into other singulars and is, therefore, universality.
Yet the identity just indicated, the fact that the determination of the
subject accrues equally to the predicate and vice versa, is not just a matter
for our consideration; it is not only in itself but is also posited in the
judgment; for the judgment is the reference connecting the two; the copula
expresses that the subject is the predicate. The subject is the determinate
determinateness, and the predicate is this determinateness of the subject as
posited; the subject is determined only in its predicate, or is subject only in
it; in the predicate, it is turned back into itself and is therein the universal. –
Now in so far as the subject is the self-subsistent term, this identity has the
relation that the predicate does not possess a self-subsistence of its own but
has its subsistence only in the subject; it inheres in the subject. Accordingly,
since the predicate is distinguished from the subject, it is only a singularized
determinateness of the subject, only one of its properties; the subject itself
is however the concrete, the totality of manifold determinacies, just as the
predicate contains one of them; the subject is the universal. – But, on the
other hand, the predicate also is self-subsistent universality, and the subject
conversely only one determination of it. The predicate thus subsumes the
subject; the singularity and the particularity are not for themselves but
have their essence and their substance in the universal. The predicate
expresses the subject in its concept; the singular and the particular are to the
subject accidental determinations; the subject is their absolute possibility.
When by “subsumption” an external connection of subject and predicate
is thought, and the subject is represented as something self-subsistent,
then subsumption refers to the subjective act of judging mentioned above,
namely the judging that starts off from the self-subsistence of both subject
and predicate. 51 Subsumption is then only the application of the universal to
a particular or singular posited under it in accordance with an indeterminate
representation, one of lesser quantity.
When we treat the identity of subject and predicate as meaning that at
one time one determination of the concept belongs to the subject and the
other to the predicate, and at another time the converse equally applies,
then the identity is as yet still implicit; on account of the self-subsistent
diversity of the two sides of judgment, their posited connection also has
51
Cf. above, 12.55.
12.58556
12.59
Georg Wilhelm Friedrich Hegel
the two at first as diverse. But it is the identity void of difference that in fact
constitutes the true connection of the subject and predicate. The determi-
nation of the concept is itself essentially a connection, for it is a universal; the
same determinations, therefore, which the subject and the predicate each
have, are also had by their connection. The connection is universal, for it
is the positive identity of both, of the subject and predicate; but it is also
determinate, for the determinateness of the predicate is the determinateness
of the subject; it is singular as well, for in it the self-subsisting extremes are
sublated as in their negative unity. – In judgment, however, this identity
is not posited yet; the copula is as the still indeterminate connection of
being in general, “A is B,” for the self-subsistence of the concept’s deter-
minacies, or the extremes, is in judgment the reality that the concept has
within. If the “is” of the copula were already posited as the determinate
and fulfilled unity of subject and predicate earlier mentioned, 52 were
posited as their concept, it would then already be the conclusion of syllogistic
inference. 53
To restore again this identity of the concept, or rather to posit it – this
is the goal of the movement of the judgment. What is already present in
the judgment is, on the one hand, the self-subsistence but also reciprocal
determinateness of the subject and predicate, and, on the other hand,
their still abstract connection. “The subject is the predicate” – this is what
the judgment says at first. But since the predicate is not supposed to be
what the subject is, a contradiction is at hand that must resolve itself, must
pass over into a result. Or rather, since the subject and predicate are in
and for themselves the totality of the concept, and judgment is the reality
of the concept, the judgment’s forward movement is only development;
what comes forth from it is already present in it, and to this extent the
demonstration is a display, 54 a reflection as the positing of that which is already
at hand in the extreme terms of the judgment; but even this positing is
already present; it is the connection of the extremes.
First, as immediate, judgment is the judgment of existence; its subject is
immediately an abstract, existent singular, and the predicate is an immediate
determinateness or property of it, an abstract universal.
Second, as this qualitative character of the subject and predicate is sub-
lated, the determination of the one begins to shine reflectively in the other;
the judgment is now the judgement of reflection.
52
54
53 “syllogistic inference” = Schluß.
Cf. above, 12.55.
“display” = Monstration. Hegel is playing on the Latinate words Demonstration and Monstration,
the Latin root of which means “displaying” or “exhibiting.”Judgment
557
But this external combination passes over into the essential identity of a
substantial, necessary combination; and so we have, third, the judgment of
necessity.
Fourth, since in this essential identity the difference of subject and
predicate has become a form, the judgment becomes subjective; it entails
the opposition of the concept and its reality and the comparison of the two;
it is the judgment of the concept.
This emergence of the concept grounds the transition of judgment into
syllogistic inference.
a. the judgment of existence
In the subjective judgment we expect to see one and the same object double,
once in its singular actuality, and again in its essential identity or in its
concept: the singular raised into its universality or, what is the same thing,
the universal made singular into its actuality. The judgment is thus truth,
for it is the agreement of the concept and reality. But it is not at first
constituted in this way, for at first the judgment is immediate, since as yet
no reflection and no movement of the determinations has been found
in it. This immediacy renders the first judgment a judgment of immediate
existence; we can also call it a qualitative judgment, but only in so far as
quality does not apply to the determinateness of being alone but also extends
to the universality which, because of its simplicity, likewise has the form of
immediacy.
The judgment of existence is also the judgment of inherence because,
though immediacy is its determination, it is the subject that in the distinc-
tion between subject and predicate is the immediate and hence the first
and the essential term in the judgment, and the predicate consequently
takes on the form of something that does not subsist on its own but has its
foundation in the subject.
a. The positive judgment
1. The subject and predicate, as we have just said, 55 are names at first that
receive their actual determination only as the judgment runs its course.
However, as sides of the judgment – the judgment being the posited deter-
minate concept – they have the determination of moments of the concept,
but, on account of their immediacy, this determination is as yet quite
55
Cf. above, 12.54.
12.60558
12.61
Georg Wilhelm Friedrich Hegel
simple, still not enriched by mediation and also still caught up in the
abstract opposition of abstract singularity and abstract universality. – The
predicate, to speak of it first, is the abstract universal; this abstract is con-
ditioned by mediation, by the sublation of singularity and particularity,
but so far such a mediation is here only a presupposition. In the sphere of
the concept there can be no other immediacy than the one that contains
mediation in and for itself and has arisen only through its sublation; this is
the immediacy of the universal. Thus qualitative being also is in its concept
a universal; as being, however, the immediacy is not yet posited as such;
it is only as universality that immediacy is the concept determination in
which it is posited that negativity essentially belongs to it. This connection
is given in the judgment in which universality is the predicate of a subject. –
Similarly the subject is an abstract singular, or the immediate which is sup-
posed to be such and therefore the singular as a something in general. The
subject constitutes, therefore, the abstract side of the judgment, the side
in it according to which the concept has passed over into externality. – As
these two concept determinations are determined, so is also their connec-
tion, the “is” or the copula; it too can have no other meaning than that of
an immediate, abstract being. It is because of this connection, which still
does not contain any mediation or negation, that this judgment is called
“positive.”
2. The first pure expression of the positive judgment is, therefore, the
proposition: the singular is universal. This expression must not be put in
the form of “A is B,” for A and B are totally formless and hence meaningless
names, whereas judgment in general, and therefore already the judgment
of existence, has determinations of the concept for its extremes. “A is B”
can stand just as well for any mere proposition as for a judgment. But what
is asserted in every judgment, even one more richly determined in form,
is the proposition that has this determined content, namely, “the singular
is universal,” for every judgment is in principle also an abstract judgment.
(Regarding the negative judgment, how far it likewise comes under this
expression, of this we shall speak presently.) 56 – However, if no thought
is given to the fact that with every judgment, the positive at least, the
assertion is made that the singular is universal, this happens either because
no attention is given to the determinate form differentiating subject and
object – for it is taken for granted that the judgment is nothing but the
connecting of two concepts – or also likely because the further content of
the judgment, “Gaius is learned,” or “the rose is red,” comes drifting in
56
Cf. below, 12.65.Judgment
559
before the mind, and the latter, busy with the picture of Gaius etc., fails to
reflect on the form – even though, at least, such a content as “Gaius,” which
is the one that usually has to be dragged in as an example, is much less
interesting than the form, and is indeed chosen because it is uninteresting,
not to divert attention from the form to itself.
The objective meaning of the proposition stating that the singular is
universal conveys, as already incidentally noted, 57 both the perishableness of
singular things and their positive subsistence in the concept in general. The
concept itself is imperishable, but that which emerges from it in its division
is subjected to alteration and to falling back into its universal nature. But
the universal, conversely, gives itself a determinate existence. Just as essence
goes out into reflective shine in its determinations; or ground into concrete
existence in appearance; and substance into manifestation in its accidents, so
does the universal resolve itself into the singular; judgment is this resolution
of the universal, the development of the negativity which, implicitly, it
already is. – This last circumstance is expressed by the converse proposition,
“the universal is singular,” which is also equally spoken in the positive
judgment. The subject, the immediate singular at first, is in the judgment
itself referred to its other, namely the universal; it is thereby posited as
the concrete – according to the category of being, as a something of many
qualities; or as the concrete of reflection, a thing of manifold properties, an
actual of manifold possibilities, a substance of precisely such accidents. Because
these manifolds here belong to the subject of the judgment, the something,
the thing, etc., is in its qualities, properties, or accidents, reflected into
itself, or continues across them, maintaining itself in them and them in
itself. Positedness or determinateness belongs to being which is in and
for itself. The subject is therefore inherently the universal. – The predicate,
on the contrary, being this universality not as real or concrete, but as
abstract, is in contrast to the subject the determinateness; it contains only
one moment of the subject’s totality to the exclusion of the others. On
account of this negativity, which as an extreme of the judgment is at the
same time self-referring, the predicate is an abstract singular. – For instance,
in the proposition, “the rose is fragrant,” the predicate expresses only one
of the many properties of the rose; it isolates it, whereas in the subject
the property is joined with the others; likewise in the dissolution of the
thing, the manifold properties that inhere in it become isolated in acquiring
self-subsistence as materials. From this side, then, the proposition of the
judgment says: the universal is singular.
57
Cf. above, 12.57.
12.62560
12.63
Georg Wilhelm Friedrich Hegel
By juxtaposing this reciprocal determination of subject and predicate in
the judgment, we thus obtain this twofold result. (1) Immediately, the
subject is indeed an existent or the singular, while the predicate is the
universal. But because the judgment connects the two, and the subject is
determined as universal by the predicate, the subject is then the universal.
(2) The predicate is determined in the subject, for it is not a determination
in general but the determination rather of the subject. “The rose is fragrant.”
This fragrance is not some indeterminate fragrance or other, but the fra-
grance of the rose. The predicate is therefore a singular. – Now since the
subject and predicate stand related in the judgment, they should retain
the opposition of concept determinations; likewise, in the reciprocity of
causality, before the latter attains its truth, the two sides are still supposed
to remain self-subsistent and mutually opposed as against the equality of
their determination. Therefore, when the subject is determined as universal,
the predicate should not also be taken in its determination of universality,
for then we would have no judgment; it must rather be taken only in its
determination of singularity. And if the subject is determined as singular,
then the predicate is to be taken as universal. – If we reflect on the mere
identity above, then we have these two identical propositions, “the singular
is singular,” “the universal is universal,” in which the sides of the judg-
ment would have completely fallen apart; only the self-reference of each is
expressed while the reference connecting them to each other is dissolved;
and thus the judgment would be sublated. – Of the two propositions we
drew, the first, “the universal is singular,” expresses the judgment according
to its content, as an isolated determination in the predicate and as the total-
ity of determinations in the subject. The other, “the singular is universal,”
expresses it according to form as immediately given through the judgment
itself. – In the immediate positive judgment, the extremes are still simple:
form and content are therefore still united. Or, in other words, it does not
consist of two propositions; the twofold connection that it yielded imme-
diately constitutes the one positive judgment. For its extremes are (a) the
self-subsisting abstract determinations of judgment, and (b) each side of
the determination is determined through the other by virtue of the copula
connecting them. Implicitly, however, the difference of form and content
is for this reason present in it, as we have seen; and indeed, what the first
proposition contains, that the singular is universal, belongs to form, for the
proposition expresses the immediate determinateness of the judgment. The
relation, on the contrary, which the other proposition expresses, that the
universal is singular or that the subject is determined as universal whereas
the predicate is determined as particular or singular, concerns the content;Judgment
561
for its determinations are only the result of an immanent reflection by
virtue of which the immediate determinacies of the judgment are sublated
and the form is thereby converted into an identity that has withdrawn into
itself and persists over against the distinction of form: it converts itself into
content.
3. If now the two propositions, the one of form and the other of content,
(Subject) (Predicate)
The singular is universal
The universal is singular,
were to be united because they are contained in the one positive judgment,
so that both, the subject as well as the predicate, were determined as the
unity of singularity and universality, then both the subject and predicate
would be the particular, and this must be recognized as implicitly their
inner determination. However, this combination would be arrived at only
through an external reflection; moreover, the proposition that results from
it, “the particular is the particular,” would no longer be a judgment but
an empty identical proposition as were the two propositions already found
in the positive judgment, “the singular is singular,” and “the universal
is universal.” – Singularity and universality cannot yet be united into
particularity, because in the positive judgment they are still posited as
immediate. – Or again, the judgment must still be distinguished according
to its form and its content, because the subject and predicate are themselves
still distinguished as immediacy and mediated, or because the judgment,
according to its connection, is both the self-subsistence of the connected
terms and their reciprocal determination or mediation.
In first place, then, the meaning of the judgment when considered accord-
ing to its form is that the singular is universal. But in fact such an immediate
singular is definitely not universal; its predicate is of wider extension, does
not correspond to it. The subject is a being existing immediately for itself, and
hence the opposite of that abstraction, of that universality posited through
mediation that was supposed to be predicated of it.
In second place, if the judgment is considered according to its content,
or as the proposition, “the universal is singular,” then the subject is a
universe of qualities, an infinitely determined concrete universe, and since
its determinacies are as yet qualitites, properties, or accidents, its totality
is the bad infinite plurality of them. Such a subject, therefore, is not at
all the one single property that its predicate declares. Consequently, both
propositions must be united, and the positive judgment must be posited as
negative instead.
12.64562
Georg Wilhelm Friedrich Hegel
b. The negative judgment
12.65
1. We spoke earlier of the common notion that whether the content of
a judgment is true or false depends solely on the content itself, since
logical truth concerns only the form and its only requirement is that such
content shall not contradict itself. 58 Nothing else is reckoned as the form
of judgment except that the latter is a connection of two concepts. But we
have seen that these two concepts are not just the relationless determination
of a sum, but that they relate to each other as singular and universal. 59 These
are the determinations that constitute the truly logical content and also,
abstracted in that way, the content of the positive judgment; whatever
other content is in a judgment (“the sun is round,” “Cicero was a great
Roman orator,” “it is daytime now,” etc.) does not concern the judgment
as such; the judgment only says that the subject is predicate, or, since these
are only names, that the singular is universal and vice versa. – It is because
of this purely logical content that the positive judgment is not true but
has its truth in the negative judgment. – In judgment, so it is required,
the content simply ought not to contradict itself; but it does contradict
itself in the positive judgment, as we have just seen. – At any rate, it
makes absolutely no difference if that logical content is called form, and
by content is understood only the remaining empirical filling, for even
then the form would not contain a mere empty identity outside which the
content determination would then lie. The positive judgment has in fact
no truth through its form as positive judgment; whoever calls truth the
correctness of an intuition or a perception, the agreement of representation
with the subject matter, has for a minimum no expression left for that
which is the subject matter and the aim of philosophy. We should at least
say of these that they are the truth of reason, and it will surely be granted
that such judgments as “Cicero was a great orator,” that “it is daytime
now,” are definitely not truths of reason. But they are not such truths, not
because they have an empirical content as it were contingently, but because
they are only positive judgments that can have, and ought to have, no other
content than an immediate singular and an abstract determinateness.
The positive judgment first attains its truth in the negative judgment:
the singular is not abstractly universal – but rather, the predicate of the
singular, because it is such a predicate, or because, if considered by itself
without reference to the subject, it is an abstract universal, is for that very
reason itself something determinate; from the start, therefore, the singular
58
Cf. above, 12.27.
59
Cf. above, 12.61.Judgment
563
is a particular. Furthermore, with respect to the other proposition that the
positive judgment contains, the meaning of the negative judgment is that
the universal is not abstractly singular but that this predicate, “singular,” by
the very fact that it is a predicate, or because it refers to a universal subject,
is more than just mere singularity, and the universal, accordingly, is from
the start equally a particular. – Since this universal, as subject, is itself in the
judgment determination of singularity, the two propositions both reduce
to one: “the singular is a particular.”
We may remark that (a) the particularity that here comes to the predi-
cate has already come up for consideration before; 60 here, however, it is not
posited by external reflection but has arisen rather as mediated by the neg-
ative connection indicated in the judgment. (b) This determination results
here only for the predicate. In the immediate judgment, the judgment of
existence, the subject is the underlying basis; the determination seems at
first, therefore, to occur in the predicate. But in fact this first negation cannot
as yet be a determination, or cannot truly be the positing of the singular, for
such a positing is only a second moment, the negative of the negative.
The singular is a particular: this is the positive expression of the negative
judgment. This expression, therefore, is not the positive judgment itself, for
the latter, because of its immediacy, has an abstraction for its extremes, while
the particular, precisely through the positing of the judgment connection,
results as the first mediated determination. – But this determination is not
to be taken only as a moment of the extremes, but also as the determination
of the connection, as it truly is from the start; in other words, the judgment
is also to be considered as negative.
This transition is founded on the relation of the extremes and on their
connection in the judgment as such. The positive judgment is the connec-
tion of the singular and the universal which are such immediately and each,
therefore, is not at the same time what the other is. The connection is there-
fore just as essentially separation, or negative; for this reason the positive
judgment was to be posited as negative. There was no need, therefore, for
the logicians to make such a fuss about the not of the negative judgment
being attached to the copula. In the judgment, the determination of the
extremes is equally a determinate connection. The judgment determination,
or the extreme, is not the purely qualitative one of immediate being that
only stands over against an other outside it. Nor is it the determination of
reflection, which, in accordance with its general form, behaves positively
and negatively, posited in either case as exclusive, only implicitly identical
60
Cf. above, 12.63.
12.66564
12.67
Georg Wilhelm Friedrich Hegel
with the other. The judgment determination, as the determination of the
concept, is a universal within, posited as extending continuously in its other.
Conversely, the judgment connection is the same determination as the
extremes have; for it is precisely this universality and continuous extension
of each into the other; in so far as these are distinguished, the connection
also has negativity in it.
The just stated transition from the form of the connection to the form of
the determination has the immediate consequence that the not of the copula
must just as equally be attached to the predicate and that the latter must be
determined as the not-universal. But, through a no less immediate conse-
quence, the not-universal is the particular. – If the focus is on the negative
according to the totally abstract determination of immediate non-being,
then the predicate is the totally indeterminate not-universal. This is the
determination which is normally treated in logic in connection with the
contradictory concepts, and the further point is made – a point considered
important – that in the negative of a concept one should only focus on the
negative, taking it as the mere indeterminate extent of the other of the pos-
itive concept. Thus the mere not-white would be just as much red, yellow,
blue, etc. as black. White, however, is an unconceptualized determination
of intuition; the not of white is equally, then, unconceptualized not-being,
the abstraction that came in for consideration at the very beginning of the
Logic where becoming was recognized to be its closest truth. To use as an
example, in the consideration of judgment determinations, an unconcep-
tualized content of this sort, drawn from intuition and the imagination,
and to take the determinations of being, and of reflection, as such judgment
determinations, is the same uncritical practice as when Kant applies the
concepts of the understanding to the infinite idea of reason, the so-called
thing-in-itself; 61 the concept, to which the judgment proceeding from it also
belongs, is the true thing-in-itself or the rational; those other determina-
tions belong to being and essence; they are not yet forms developed into the
shape where they are in their truth, in the concept. – If we stop at white,
red, as representations of the senses, then we call concept what is only a
determination of pictorial representation. This is common practice. But
then, surely, the not-white, the not-red, will be nothing positive, just as the
not-triangular will be something totally indeterminate, for a determination
based as such on number and quantum is essentially something indifferent,
void of concept. Yet, like non-being itself, such a sensuous content ought to
be conceptualized; ought to shed that indifference and abstract immediacy
61
Cf. B166, note.Judgment
565
with which it is affected in the blind immobility of pictorial representa-
tion. Already in the sphere of immediate existence, the non-being which is
otherwise void of thought becomes limit, and by virtue of this limit the
something refers to an other despite itself. In the sphere of reflection, on the
other hand, it is the negative that refers essentially to a positive, and is thereby
determined; a negative is no longer that indeterminate non-being, for it is
posited to be only to the extent that the positive stands over against it, and
as third comes their ground; the negative is thus held circumscribed in a
sphere within which the non-being of one is something determinate. – But
it is all the more in the absolutely fluid continuity of the concept that the
not is immediately a positive, and the negation is not just determinateness
but is taken up into universality and is posited as identical with it. The
non-universal is therefore directly the particular.
2. Since negation has to do with the connection of judgment, and we are
considering the negative judgment still as such, the latter is in the first instance
still a judgment; we thus have the relation of subject and predicate, or of
singularity and universality, and their connection, the form of the judgment.
The subject, as the immediate underlying basis, remains untouched by the
negation; it retains, therefore, its determination of having a predicate, or its
reference to the universality. Consequently, what is negated in the predicate
is not the universality as such, but the abstraction or the determinateness
of the predicate that appeared as content in contrast to that universality. –
The negative judgment is not, therefore, total negation; the universal
sphere which contains the predicate remains standing; the connection
of subject and the predicate is therefore still essentially positive; the yet
remaining determination of the predicate is no less connection. – When
it is said that, for instance, the rose is not red, only the determinateness
of the predicate is thereby denied and thus separated from the universal-
ity which equally attaches to it; the universal sphere, color, is retained; if
the rose is not red, it is nonetheless assumed that it has a color, though
another color. From the side of this universal sphere, the judgment is still
positive.
“The singular is a particular.” This positive form of the negative judgment
immediately expresses that the particular contains universality. In addition,
it also expresses that the predicate is not just a universal but also one which
is still determinate. The negative form contains the same, for although
the rose, for instance, is not red, it is supposed, nevertheless, not only
still to retain the universal sphere of color as predicate, but to have some
other determinate color as well; the singularity of determinateness of the
rose is therefore only sublated; and not only is the universal sphere left
12.68566
12.69
Georg Wilhelm Friedrich Hegel
standing but determinateness too is retained, although transformed into
an indeterminate determinateness, a universal determinateness, that is to
say, into particularity.
3. The particularity that has resulted as the positive determination of the
negative judgment is the term mediating singularity and universality; so
the negative judgment is now that which provides in general the mediation
for the third step, that of the reflection of the judgment of existence into itself.
This judgment is according to its objective meaning only the moment of
the alteration of accidents, or, in the sphere of existence, of the singularized
properties of the concrete. Through this alteration, the full determinateness
of the predicate, or the concrete, emerges as posited.
“The singular is particular” is what the positive expression of the negative
judgment says. But the singular is also not particular, for particularity is of
wider extension than singularity; it is a predicate, therefore, that does not
correspond to the subject, one in which the latter, therefore, does not as yet
have its truth. “The singular is only a singular”: this is a negativity that refers
to nothing else, be it positive or negative, except itself. – The rose is not a
thing of some color or other, but one that only has the one determinate color
which is the rose-color. The singular is not an indeterminate determinate
but the determinate determinate.
This negation of the negative judgment appears, when one starts from
its positive form, to be again a first negation. But this is not what it is. The
negative judgment is again, in and for itself, already the second negation
or the negation of negation, and this, what it is in and for itself, is to be
posited. To wit: the judgment negates the determinateness of the predicate
of the positive judgment, its abstract universality, or, considered as content,
the singular quality that it possesses of the subject. But the negation of the
determinateness is already the second negation, hence the infinite turning
back of the singularity into itself. With this, therefore, the restoration of
the concrete totality of the subject has taken place, or rather, the subject
is now for the first time posited as singular, for through the negation and
the sublation of that negation it is mediated with itself. The predicate,
for its part, has thereby passed over from the first universality to absolute
determinateness and made itself equal to the subject. Thus the judgment
says: “the singular is singular.” – From the other side, since the subject
was equally to be taken as a universal, and since in the negative judgment
the predicate, which as against that subject is the singular, expanded into
particularity; moreover, since now the negation of this determinateness is
equally the purification of the universality contained in the predicate, this
judgment also says: “the universal is the universal.”Judgment
567
In these two judgments, which were earlier obtained through external
reflection, 62 the predicate is already expressed in its positivity. But the
negation of the negative judgment must itself first appear in the form of a
negative judgment. It has just been shown that there still remained in this
judgment a positive connection of subject and predicate as well the universal
sphere of the latter. 63 From this side, the negative judgment thus contains
a universality which is more purified of limitation than was contained by
the positive judgment and is for this reason all the more to be negated of
the subject as a singular. In this manner, the whole extent of the predicate
is negated, and there is no longer any positive connection between it and
the subject. This is the infinite judgment.
c. The infinite judgment
The negative judgment is as little of a true judgment as the positive. But
the infinite judgment which is supposed to be its truth is, according to
its negative expression, the negative infinite, a judgment in which even
the form of judgment is sublated. – But this is a nonsensical judgment.
It ought to be a judgment, and hence contains a connection of subject
and predicate; but any such connection ought not at the same time to be
there. – The name of the infinite judgment does indeed occur in the com-
mon textbooks of logic, but without any clarification as to its meaning. –
Examples of negatively infinite judgments are easy to come by. It is a
matter of picking determinations, one of which does not contain not just
the determinateness of the other but its universal sphere as well, and of
combining them negatively as subject and predicate, as when we say, for
example, that spirit is not red, yellow, etc., is not acid, not alkali, etc., or that
the rose is not an elephant, the understanding is not a table, and the like. –
These judgments are correct or true, as it is said, and yet, any such truth
notwithstanding, nonsensical and fatuous. – Or, more to the point, they are
not judgments at all. – A more realistic example of the infinite judgment is
the evil action. In civil litigation, when a thing is negated as the property of
another party, it is still conceded that the same thing would indeed belong
to that party if the latter had a right to it. It is only under the title of right
that the possession of it is challenged; in the negative judgment, therefore,
the universal sphere, “right,” is still acknowledged and maintained. But
crime is the infinite judgment that negates, not only the particular right, but
the universal sphere, the right as right. It has correctness, in the sense that it
62
Cf. above, 12.63.
63
In the preceding page.
12.70568
12.71
Georg Wilhelm Friedrich Hegel
is an effective action, but since it stands in a thoroughly negative fashion
with respect to the morality that constitutes its sphere, it is nonsensical.
The positive element of the infinite judgment, the negation of the nega-
tion, is the reflection of singularity into itself by virtue of which the sin-
gularity is first posited as the determinate determinate. “The singular is
singular” is what the infinite judgment said according to that reflection. In
the judgment of existence, the subject is as the immediate singular, hence
more of just a something in general. Through the mediation of the negative
and infinite judgment, it is posited as singular for the first time.
The singular is thus posited as expanding into its predicate, which is
identical with it; to the same extent, therefore, universality is also no longer
anything immediate but a summing of distincts. The positively infinite
judgment equally says, “the universal is universal,” and in this the universal
is posited also as a turning back into itself.
Now through the reflection of the judgment determinations into them-
selves, the judgment has sublated itself; in the negatively infinite judgment,
the difference is, so to speak, too great for it still to remain a judgment; sub-
ject and predicate have no positive connection whatsoever to each other; in
the positively infinite judgment, on the contrary, only identity is present,
and because of this total lack of difference there is no longer a judgment.
More precisely, it is the judgment of existence that has sublated itself and,
consequently, there is posited what the copula of the judgment contains,
namely that in its identity the qualitative extremes are sublated. But since
this unity is the concept, it is immediately torn apart and is a judgment, but
one whose terms are no longer immediately determined but are reflected
into themselves. The judgment of existence has passed over into the judgment
of reflection.
b. the judgment of reflection
In the judgment that has now arisen, the subject is a singular as such;
and similarly, the universal is no longer an abstract universality, or a singu-
lar property, but is posited as a universal that has collected itself together
into a unity through the connection of different terms, or, regarded from
the standpoint of the content of diverse determinations in general, as the
coalescing of manifold properties and concrete existences. – If examples
of predicates of judgments of reflection are to be given, they must be of
another kind than for the judgments of existence. It is only in the judgment
of reflection that we first have a determinate content strictly speaking, that is,
a content as such; for the content is the form determination reflected intoJudgment
569
identity as distinct from the form in so far as this is a distinct determinate-
ness – as it still is as judgment. In the judgment of existence, the content
is merely an immediate, or abstract, indeterminate content. – These may
therefore serve as examples of judgments of reflection: the human being
is mortal, things are perishable, this thing is useful, harmful; hardness, elas-
ticity of bodies, happiness, etc., are predicates of this particular kind. They
express an essentiality which is however a relational determination, or a
comprehensive universality. This universality, which will further determine
itself in the movement of the judgment of reflection, is still distinct from
the universality of the concept as such; although it is no longer the abstract
universality of the qualitative judgment, it still has a connection to the
immediate from which it proceeds and has the latter at the basis of its neg-
ativity. – The concept determines immediate existence, in the first instance,
to relational determinations that extend across the diverse multiplicity of
concrete existence, so that the true universal is indeed the inner essence of
that multiplicity, but is such in the sphere of appearance, and this relative
nature or even its mark is not as yet the element of the multiplicity that
exists in and for itself.
It may seem fitting to define the judgment of reflection as a judgment of
quantity, just as the judgment of existence was defined also as qualitative
judgment. But just as the immediacy in the latter was not just there, but was
an immediacy which is also essentially mediated and abstract, so, here also,
that same immediacy which is now sublated is not just sublated quality,
and therefore not merely quantity; on the contrary, just as quality is the
most external immediacy, so is quantity, in the same way, the most external
determination belonging to mediation.
Also to be noted concerning the determination as it appears in the move-
ment of the judgment of reflection is that, in the judgment of immediate
existence, the movement of the determination showed itself in the predi-
cate, for this kind of judgment was in the determination of immediacy and
its subject, therefore, appeared as the underlying basis. For a similar reason,
in the judgment of reflection the onward movement of determination runs
its course in the subject, for this judgment has the reflected in-itselfness for
its determination. Hence the essential is here the universal or the predicate,
and it is the latter, therefore, that constitutes the basis against which the
subject is to be measured and determined accordingly. – Yet the predicate
also receives a further determination through the further development of
the form of the subject, but it receives it indirectly, whereas the progression
of the subject manifests itself, for the reason just given, as a direct advance
in determination.
12.72570
Georg Wilhelm Friedrich Hegel
As regards the objective signification of the judgment, the singular enters
into existence by virtue of its universality, but it does so in an essential
determination which is relational, in an essentiality that maintains itself
across the manifold of appearance; the subject is supposed to be that which
is determined in and for itself; this is the determinateness which it has in its
predicate. The singular, for its part, is reflected into this predicate which is
its universal essence; to this extent, the subject is a concrete existence and a
phenomenal something. 64 In this judgment, the predicate no longer inheres
in the subject, for it is rather the implicit being under which the singular
subject is subsumed as an accidental. If the judgments of existence can also
be defined as judgments of inherence, then the judgments of reflection are
by contrast judgments of subsumption.
a. The singular judgment
12.73
Now the immediate judgment of reflection is again, “the singular is univer-
sal,” but with the subject and predicate in the signification just explained.
More accurately, therefore, it can also be expressed thus, “this is an essential
universal.”
But a “this” is not an essential universal. That positive judgment – positive
according to form – must as judgment be taken negatively. But inasmuch
as the judgment of reflection is not merely something positive, the negation
does not directly affect the predicate – a predicate which does not inhere in
the subject but is rather its implicit being. On the contrary, it is the subject
that is alterable and needs determination. The negative judgment is there-
fore to be understood as saying: “‘not a this’ is a universal of reflection”; 65
such an in-itself has a more universal concrete existence than it would have
in a “this.” Accordingly, the singular judgment has its proximate truth in
the particular judgment.
b. The particular judgment
The non-singularity of the subject that must be posited in the first judg-
ment of reflection instead of the subject’s singularity is particularity. But
particularity is determined in the judgment of reflection as essential singu-
larity; particularity cannot be, therefore, a simple, abstract determination in
which the singular would be sublated and the concrete existent dissolved,
but is rather only an extension of this singular in external reflection. Thus
the subject is: “these ones,” or “a particular number of singulars.”
64
“phenomenal something” = Erscheinende.
65
“Nicht ein Dieses” ist ein Allgemeines der Reflexion.Judgment
571
The judgment, “some singulars are a universal of reflection,” appears at
first to be a positive judgment, but it is just as well also negative; for “some”
contains universality and may, accordingly, be regarded as comprehensive;
but since it is particularity, it is equally disproportionate with respect to
universality. The negative determination which the subject has obtained
through the transition of the singular judgment also is, as we have shown
above, the determination of the connection, the copula. 66 – Implicated
in the judgment, “some humans are happy,” is the immediate consequence:
“some humans are not happy.” When some things are useful, then, pre-
cisely for that reason, there also are some that are not useful. The positive
and the negative judgment no longer fall outside one another, but the
particular immediately contains both at the same time, precisely because
it is a judgment of reflection. – But the particular judgment is therefore
indeterminate.
If, in the example of such a judgment, we consider further the subject,
“some humans,” “some animals,” etc., we find that it contains, besides the
particular form determination of “some,” also the content determination
of “humans,” etc. By the subject of the singular judgment one could mean,
“this human,” a singularity that properly pertains to external pointing; it
would best be expressed, therefore, by something like “Gaius.” But the
subject of the particular judgment can no longer be “some Gaiuses,” for
Gaius is supposed to be a singular as singular. To the “some,” therefore,
there is added a more universal content, say “humans,” “animals,” etc. This
is not a mere empirical content, but one which is determined by the form of
the judgment; it is universal, that is, because “some” contains universality,
and the latter must at the same time be separated from the singulars which
the reflected singularity has as a basis. More precisely, this universality is
also the universal nature or species “human,” “animal” – the universality
which is the result of the judgment of reflection, but anticipated; just as
the positive judgment, since it has the singular for subject, also anticipates
the determination which is the result of the judgment of existence.
Thus the subject that contains the singulars, their connection to par-
ticularity, and the universal nature, is already posited as the totality of the
determinations of the concept. But, to be precise, this consideration is an
external one. What is at first already posited in the subject by virtue of its
form, in reciprocal connection, is the extension of the “this” to particularity;
but this generalization is not commensurate to the “this”; the latter is per-
fectly determinate, but “some” is indeterminate. The extension ought to
66
i.e. in section a. immediately preceding: “But a ‘this’ is not an essential universal.”
12.74572
Georg Wilhelm Friedrich Hegel
be appropriate to the “this” and therefore, in conformity with it, it ought
to be completely determined; such an extension is totality, or, in the first
instance, universality in general.
This universality has the “this” for its basis, for the singular is here the
singular reflected into itself; its further determinations run their course,
therefore, outside it, and just as for this reason particularity determined itself
as a “some,” so the universality which the subject has attained is an “allness,”
and thus the particular judgment has passed over into the universal.
c. The universal judgment
12.75
The universality of the subject of the universal judgment is the external
universality of reflection, “allness”; the “all” is the all of all the singulars in
which the singular remains unchanged. This universality is therefore only
a commonality of self-subsisting singulars, an association of such singulars
as comes about only by way of comparison. – This is the association that
first comes to mind at a subjective level of representation when there is talk
of universality. The most obvious reason given for viewing a determination
as universal is because it fits many. – Also in analysis is this conception of
universality the one most prevalent, as when, for instance, the development
of a function in a polynomial is taken to have greater universal value than
its development in a binomial, because the polynomial displays more single
terms than the binomial. The demand that the function should be resolved
in its full universality would require, strictly speaking, a pantonomial, the
exhausted infinity. But here is where the limitation of that demand becomes
apparent, and where the display of the infinite number of terms must rest
satisfied with the ought it commands, and therefore also with a polynomial.
But in fact the binomial is already the pantonomial in those cases where
the method or the rule concerns only the dependence of one member on
another, and the dependence of several terms on those that precede them
does not particularize itself but remains one and the same underlying
function. It is the method or the rule which is to be regarded as the true
universal; in the progress of the development or in the development of a
polynomial, the rule is only repeated, so that it gains nothing in universality
through the increased number of terms. We have already spoken earlier of
the bad infinity and its deception; 67 the universality of the concept is the
achieved beyond, whereas that bad infinity remains afflicted with a beyond
which is unattainable but remains a mere progression to infinity. If it is allness
67
Cf. GW 11, 79–81; above 21.127ff.Judgment
573
that universality brings to mind, a universality that ought to be exhausted
in singulars as singulars, then there has been a relapse into that bad infinity;
or else it is mere plurality which is taken for allness. But plurality, however
great it might be, remains inescapably only particularity: it is not allness. –
Yet there is in all this an obscure intimation of the universality of the concept
as it exists in and for itself; it is the concept that violently strives to reach
beyond the stubborn singularity to which pictorial representation clings
and beyond the externality of its reflection, passing off allness as totality or
rather as the category of the in-and-for-itself.
This is apparent in other ways as well in the allness which is above all
empirical universality. Inasmuch as the singular is presupposed as something
immediate and is therefore pre-given and externally picked, the reflection
which collects it into an allness is equally external to it. But because the
singular, as a “this,” is absolutely indifferent to such a reflection, the uni-
versality and the collected singularity cannot combine to form a unity.
The empirical allness thus remains a task; it is an ought which, as such,
cannot be represented in the form of being. Now an empirically universal
proposition – for nevertheless such are advanced – rests on the tacit agree-
ment that, if no instance of the contrary can be adduced, a plurality of cases
ought to count for an allness; or that a subjective allness, namely the known
cases, may be taken for an objective allness.
Now a closer examination of the universal judgment before us shows
that the subject, as we have just noted, contains the achieved universality as
presupposed; it even contains it as posited in it. “All humans” expresses, first,
the species “human”; second, this species in its singularization, but in such
a way that the singulars are at the same time expanded to the universality
of the species; conversely, through this conjunction with singularity, the
universality is just as perfectly determined as singularity, and the posited
universality has thereby become equal to what was presupposed.
But, strictly speaking, we should not anticipate the presupposed but
should rather consider the result for itself in the form determination. –
The singularity, inasmuch as it is expanded to allness, is posited as negativ-
ity, and this is identical self-reference. It has not remained, therefore, that
first singularity (of Gaius, for instance) but is a determination identical
with universality, or the absolute determinateness of the universal. – That
first singularity of the singular judgment was not the immediate singularity
of the positive judgment, but came about through the dialectical move-
ment of the judgment of existence in general; 68 it was already determined
68
Cf. above, 12.71.
12.76574
12.77
Georg Wilhelm Friedrich Hegel
to be the negative identity of the determinations of that judgment. This is
the true presupposition in the judgment of reflection; as contrasted to the
positing that runs its course in that judgment, that first determinateness
of singularity was the latter’s in-itself; consequently, what singularity is in
itself, through the movement of the judgment of reflection is now posited –
posited, that is, as the identical self-reference of the determinate. Therefore
the reflection that expanded the singularity to allness is not external to it;
on the contrary, it only makes explicit what was before implicit. – Hence
the result is in truth the objective universality. The subject has thus shed the
form determination of the judgment of reflection that made its way from
the “this” to the “allness” through the “some.” Instead of “all humans,” we
now have to say “the human being.” 69
The universality that has thereby arisen is the genus, or the universality
which is concrete in its universality. The genus does not inhere in the subject;
it is not one property of it or a property at all; it contains all singular
determinacies dissolved into its substantial purity. – Because it is thus
posited as this negative self-identity, it is for that reason essentially subject,
but one that is no longer subsumed under its predicate. Consequently the
nature of the judgment of reflection is now altogether altered.
This judgment was essentially a judgment of subsumption. The predi-
cate was determined, in contrast to its subject, as the implicit universal;
according to its content, it could be taken as an essentially relational deter-
mination or also as a mark – a determination which makes the subject
essentially only an appearance. But when determined to objective universal-
ity, the subject ceases to be subsumed under such a relational determination
or the collecting grasp of reflection; with respect to this objective univer-
sality, a predicate of this sort is rather a particular. The relation of subject
and predicate has thus reversed itself, and to this extent the judgment has
at this point sublated itself.
This sublation of the judgment coincides with what the determination
of the copula becomes, as we still have to consider; 70 the sublation of the
determinations of judgment and their transition into the copula are one
and same. – For inasmuch as the subject has raised itself to universality,
it has become in this determination equal to the predicate which, as the
reflected universality, also contains particularity within itself; subject and
predicate are therefore identical, that is, in the copula they have come to
coincide. This identity is the genus or the nature of a thing in and for itself.
Inasmuch as this identity, therefore, again divides, it is the inner nature by
69
der Mensch.
70
Cf. below, 12.89.Judgment
575
virtue of which a subject and predicate are connected to each other. This
is a connection of necessity wherein the two terms of the judgment are only
unessential distinctions. – That what belongs to all the singulars of a genus
belongs to the genus by nature, is an immediate consequence. It expresses
what we have just seen – that the subject, e.g. “all humans,” sheds its form
determination and “the human being” is what it should say instead. –
This combination, implicit and explicit, constitutes the basis of a new
judgment – the judgment of necessity.
c. the judgment of necessity
The determination to which universality has advanced is, as we have seen,
the universality that exists in and for itself or the objective universality that
in the sphere of essence corresponds to substantiality. It is distinguished
from the latter because it belongs to the concept and for this reason is not
only the inner but also the posited necessity of its determinations, or in
other words, the distinction is immanent to it, whereas substance has its
distinction only in its accidents, does not have it as a principle within it.
In the judgment now, this objective universality is posited – first, posited
with this determinateness as essential to it, immanent to it; second, posited
with it as diverse from it, a particularity for which the said universality
constitutes the substantial basis. In this way the universality is determined
as genus and species.
a. The categorical judgment
The genus essentially divides or repels itself into species; it is genus only
in so far as it comprehends the species under it; the species is a species
only in so far as, on the one side, it exists in singulars, and, on the other
side, it possesses in the genus a higher universality. – Now the categorical
judgment has for predicate such a universality as in it the subject possesses
its immanent nature. But the categorical judgment is itself the first or
the immediate judgment of necessity; consequently, the determinateness
of the subject, by virtue of which the latter is a singular as contrasted to
the genus or the species, belongs to the immediacy of external concrete
existence. – But objective universality also has here only its first immediate
particularization; on the one hand, therefore, it is itself a determinate genus
with respect to which there are higher genera; on the other hand, it is not
the most proximate genus, that is, its determinateness is not directly the
principle of the specific particularity of the subject. But what is necessary in
12.78576
Georg Wilhelm Friedrich Hegel
it is the substantial identity of subject and predicate, in view of which the
distinguishing mark of each is only an unessential positedness or even only
a name; in its predicate, the subject is reflected into its being-in-and-for-
itself. Such a predicate ought not to be classed with the predicates of the
preceding judgments. For example, to throw together into one class these
judgments:
The rose is red,
The rose is a plant,
or This ring is yellow,
It is gold,
12.79
and thus to take such an external property as the color of a flower as a
predicate equal to its vegetable nature, is to overlook a difference which
the dullest mind would not miss. – The categorical judgment, therefore, is
definitely to be distinguished from the positive and the negative judgment;
in these, what is said of the subject is a singular accidental content; in the
former, the content is the totality of the form reflected into itself. In this
content, therefore, the copula has the meaning of necessity, whereas in that
of the other two it has only the meaning of abstract, immediate being.
The determinateness of the subject, which makes it a particular with
respect to the predicate, is at first still something contingent; subject and
predicate are not connected with necessity by the form or the determinateness;
the necessity is therefore still an inner one. – The subject is subject, however,
only as a particular, and to the extent that it possesses objective universality,
it has to possess it essentially in accordance with that at first immediate
determinateness. The objective universal, in determining itself, that is, in
positing itself in a judgment, is in a connection of identity with this repelled
determinateness as such – essentially, that is, this determinateness is not to
be posited as merely accidental. Only through this necessity of its immediate
being does the categorical judgment conform to its objective universality
and, in this way, has passed over into the hypothetical judgment.
b. The hypothetical judgment
“If A is, then B is”; or “The being of A is not its own being but the being
of an other, of B.” – What is posited in this judgment is the necessary
connectedness of immediate determinacies, a connectedness which in the
categorical judgment is not yet posited. – There are here two immediate,
or externally contingent concrete existences, of which in the categorical
judgment there is at first only one, the subject; but since one is externalJudgment
577
to the other, this other is immediately also external with respect to the
first. – On account of this immediacy, the contents of both sides are still
indifferent to each other; at first, therefore, this judgment is a proposition
of empty form. Now, first, the immediacy is as such indeed self-subsistent,
a concrete being; but, second, what is essential is its connection; this being
is therefore just as much mere possibility; the hypothetical judgment does
not say either that A is, or that B is, but only that if the one is, then the
other is; only the connectedness of the extremes is posited as existing, not
the extremes themselves. Indeed, each extreme is posited in this necessity
as equally the being of an other. – The principle of identity asserts that A is
only A, not B; and B is only B, not A. In the hypothetical judgment, on
the contrary, the being of finite things is posited through the concept in
accordance with their formal truth, namely that the finite is its own being,
but equally is not its own being but is the being of an other. In the sphere
of being, the finite alters and comes to be an other. In the sphere of essence,
it is appearance; its being is posited to consist in the reflective shining of an
other in it, and the necessity is the inner connection not yet posited as such.
But the concept is this: that this identity is posited; that the existent is not
abstract self-identity but concrete self-identity and is, immediately within it,
the being of an other.
The hypothetical judgment can be more closely determined in terms of
the relations of reflection as a relation of ground and consequence, condi-
tion and conditioned, causality etc. Just as substantiality is present in the
categorical judgment in the form of its concept, so is the connectedness
of causality in the hypothetical judgment. This and the other relations all
recur in it, but they are there essentially only as moments of one and the
same identity. – However, in it they are as yet not opposed as singular or
particular and universal according to the determinations of the concept,
but are only as moments in general at first. The hypothetical judgment,
therefore, has a shape which is more that of a proposition; just as the
particular judgment is of indeterminate content, so is the hypothetical of
indeterminate form, for the determination of its content does not conform
to the relation of subject and predicate. – Yet the being, since it is the
being of the other, is for that very reason in itself the unity of itself and
the other, and therefore universality; by the same token it is in fact only a
particular, for it is a determinate being and does not refer in its determi-
nateness merely to itself. But it is not the simple, abstract particularity that
is posited; on the contrary, through the immediacy which the determinacies
possess, the moments of particularity are differentiated; at the same time,
through the unity of these moments as constituted by their connection,
12.80578
Georg Wilhelm Friedrich Hegel
the particularity is also their totality. – In truth, therefore, what is posited
in this judgment is universality as the concrete identity of the concept
whose determinations do not have any subsistence of their own but are
only particularities posited in that identity. So it is the disjunctive judgment.
c. The disjunctive judgment
12.81
In the categorical judgment, the concept is objective universality and an
external singularity. In the hypothetical, the concept manifests its presence
in this externality, in its negative identity. Through this identity, the objec-
tive universality and the external singularity obtain the determinateness,
now posited in the disjunctive judgment, which in the hypothetical they
possess immediately. Hence the disjunctive judgment is objective univer-
sality at the same time posited in union with the form. It thus contains,
first, the concrete universality or the genus in simple form, as the subject;
second, the same universality but as the totality of its differentiated determi-
nations. “A is either B or C.” This is the necessity of the concept in which,
first, the self-identity of the two extremes is of the same extent, content,
and universality. Second, they are differentiated according to the form of
conceptual determination, but, because of that identity, this determination
is a mere form. Third, the identical objective universality appears for that
reason reflected into itself as against the non-essential form, as a content
which however has the determinateness of form in it – once as the sim-
ple determination of genus; then again, as this determinateness developed
in its difference, and in this way it is the particularity of the species and
their totality, the universality of the genus. – The particularity constitutes in
its development the predicate, because, in containing the whole universal
sphere of the subject, and in containing it, however, also in the articulation
of particularity, it is to that extent the greater universal.
Upon closer consideration of this particularization, it is the genus that
constitutes first of all the substantial universality of the species; the subject
is thus B as well as C; this “as well as” indicates the positive identity of the
particular with the universal; this objective universal maintains itself fully
in its particularity. Secondly, the species mutually exclude one another; “A is
either B or C”; for they are the specific difference of the universal sphere.
This “either or” is their negative connection. In this negative connection
they are just as identical as in the positive; the genus is their unity as a unity
of determinate particulars. – If the genus were an abstract universality,
as in the judgments of existence, then the species would also have to
be taken as diverse and mutually indifferent; this universality, however, isJudgment
579
not the external one that arises only through comparison and abstraction
but is, on the contrary, the universality which is immanent to the genus
and concrete. – An empirical disjunctive judgment is without necessity;
A is either B or C or D, etc., because the species B, C, D, etc., are
found beforehand; strictly speaking, therefore, there is no question here of
an “either or,” for the completeness of these species is only a subjective
one; of course, one species excludes the other, but the “either or” excludes
every other species and excludes within itself an entire sphere. This totality
has its necessity in the negative unity of the objective universal which
has dissolved singularity within itself and possesses, immanent in it, the
simple principle of differentiation by which the species are determined and
connected. The empirical species, on the contrary, have their differences
in some accidentality or other which is a principle external to them and
is not, therefore, their principle, and consequently also not the immanent
determinateness of the genus; for this reason, they are also not reciprocally
connected according to their determinateness. – Yet it is by virtue of
their determinateness that the species constitute the universality of the
predicate. – Here is where the so-called contrary and contradictory concepts
should find their proper place, for the disjunctive judgment is where the
essential difference of the concept is posited; but here they also equally find
their truth, namely that contrariness and contradictoriness are themselves
differentiated both as contraries and as contradictory. Species are contrary
inasmuch as they are merely diverse, that is to say, inasmuch as they possess
an immediate existence as subsisting in and for themselves by virtue of
the genus which is their nature. They are contradictory, inasmuch as they
exclude one another. But each of these determinations is by itself one-sided
and void of truth. In the “either or” of the disjunctive judgment, their
unity is posited as their truth, which is that the independent subsistence of
the species as concrete universality is itself also the principle of the negative
unity by which they mutually exclude one another.
Through the identity just demonstrated of subject and predicate in
accordance with the negative unity, the genus is determined in the dis-
junctive judgment as the proximate genus. This expression indicates at first
the mere quantitative difference of the more or less determinations which
a universal contains as contrasted to a particularity coming under it. On
this account, which is the truly proximate genus remains contingent. But
then, if the genus is taken as a universal arrived at by the mere abstraction
of determinations, it cannot strictly speaking form a disjunctive judgment;
for it is contingent whether, as it were, there is still left in it the determi-
nateness that constitutes the principle of the “either or”; the genus would
12.82580
12.83
Georg Wilhelm Friedrich Hegel
not be displayed in the species according to its determinateness, and these
would only be capable of contingent completeness. In the categorical judg-
ment, the genus stands at first over against the subject only in this abstract
form – is not, therefore, necessarily its proximate genus and, to this extent,
is external to it. But when the genus is a concrete, essentially determined
universality, then, as simple determinateness, it is the unity of the moments
of the concept – moments that, only sublated in that simplicity, have their
real difference in the species. Hence the genus is the proximate genus of a
species, for the latter possesses its specific difference in the essential deter-
minateness of the genus, and the species have as such the determination
differentiating them in the nature of the genus.
What we have just considered constitutes the identity of subject and
predicate from the aspect of determinateness in general. This is an aspect
that was posited by the hypothetical judgment, the necessity of which is
an identity of immediate and diverse things and is, therefore, essentially a
negative unity. It is this negative unity that in principle separates subject
and predicate but is now posited as itself differentiated – in the subject,
as simple determinateness; in the predicate, as totality. That parting of
subject and predicate is the difference of the concept; the totality of the
species in the predicate can then be none other than this difference. – The
reciprocal determination of the disjunctive terms is therefore hereby given.
It reduces to the difference of the concept, for it is the concept alone
that disjoins itself and manifests in its determination its negative unity.
Of course, the species comes up for consideration here only under the
aspect of its simple conceptual determinateness, not according to the shape
in which, proceeding from the idea, it steps into a further self-subsistent
reality. This reality is of course dropped in the simple principle of the genus;
but the essential differentiation must be a moment of the concept. In the
judgment here considered, it is really now the concept’s own progressive
determination that itself posits its disjunction, just as was the case for the
concept itself, as we saw when it was determined in and for itself and was
differentiated into determinate concepts. 71 – Now because the concept is
the universal, the positive as well as the negative totality of the particulars,
for that reason it is immediately itself also one of its disjunctive members; the
other member, however, is this universality resolved into its particularity, or
the determinateness of the concept as determinateness, in which the very
universality displays itself as totality. – If the disjunction of a genus into
species has not yet attained this form, this is proof that the disjunction has
71
Cf. above, 12.38–39.Judgment
581
not risen to the determinateness of the concept and has not proceeded from
it. – Color is either violet, indigo, blue, green, yellow, orange, or red; even
empirically, the confusion and impurity of such a disjunction are at once
apparent; it is a barbarism even from this standpoint. If color is conceived
as the concrete unity of light and darkness, then this genus has within it the
determinateness that constitutes the principle of its particularization into
species. Of these, however, one must be the utterly simple color that holds
the opposition in balance, contained and negated in the color’s intensity;
the relation of the opposition of light and darkness must then take its
place over against it, and, since this relation is a natural phenomenon,
the indifferent neutrality of the opposition must be further added to it. –
Taking for genus such mixtures as violet, and orange, or shades of difference
like indigo blue and light blue, betrays a totally inconsiderate procedure
that shows too little reflection even for empiricism. 72 – But this is not
the place to discuss the different and more finely determined forms that
disjunction may indeed assume in the element of nature or spirit.
In the first instance, the disjunctive judgment has the members of the
disjunction in the predicate. But the judgment is itself equally disjoined;
its subject and predicate are the members of the disjunction; they are the
moments of the concept posited in their determinateness but at the same
time as identical – identical, (a) in the objective universality which is in
the subject as the simple genus, and in the predicate as the universal sphere
and totality of the moments of the concept; and (b) in the negative unity,
the developed connectedness of necessity, in accordance with which the
simple determinateness in the subject has fallen apart into the difference of
the species and these, in this very difference, have their essential connection
and self-identity.
This unity, the copula of this judgment in which the extremes have
come together through their identity, is thus the concept itself, indeed the
concept as posited; the mere judgment of necessity has thereby risen to the
judgment of the concept.
d. the judgment of the concept
To know how to form judgments of existence, such as “the rose is red,” “the
snow is white,” etc., hardly counts as a sign of great power of judgment.
The judgments of reflection are more in the nature of propositions; to be sure,
72
The criticism is directed at Goethe’s theory of colors. See Goethe, Zur Farbenlehre: Didaktischer
Teil in Goethes Werke, Vol. 13 (Hamburg: Christian Wegner, 1966), §175.
12.84582
12.85
Georg Wilhelm Friedrich Hegel
in the judgment of necessity the subject matter is present in its objective
universality, but it is only in the judgment now to be considered that its
connection with the concept is to be found. The concept is at the basis of this
judgment, and it is there with reference to the subject matter, as an ought to
which reality may or may not conform. – This is the judgment, therefore,
that first contains true adjudication; the predicates, “good,” “bad,” “true,”
“right,” etc., express that the fact is measured against the concept as an
ought which is simply presupposed, and is, or is not, in agreement with it.
The judgment of the concept has been called the judgment of modality,
and has been regarded as containing the form of the connection of subject
and predicate as this obtains in an external understanding, and as concerned
with the value of the copula only in connection with thought. 73 Accordingly,
judgment is said to be problematic when the affirmation or negation is
taken as optional or possible; assertoric, when it is taken as true, that is,
actual, and apodictic when it is taken as necessary. – It is easy to see why it
would be an easy step in this judgment to go outside the judgment itself
and to regard its determination as something merely subjective. For it is the
concept here, the subjective, that comes into play again in judgment and
relates to an immediate actuality. But this subjectivity is not to be confused
with external reflection, which is of course also something subjective but in
a different sense than the concept itself; on the contrary, the concept that
has again emerged out of the disjunctive judgment is the very opposite of
a mere mode or manner. The earlier judgments are subjective in this sense,
for they rest on an abstract one-sidedness in which the concept is lost.
But the judgment of the concept is instead objective and, as contrasted
with the others, it is the truth, for it rests on the concept precisely in its
determinateness as concept, not in some external reflection or with reference
to some subjective, that is, accidental, thought.
In the disjunctive judgment, the concept was posited as the identity of
universal nature and its particularization, and with that the relation of the
judgment was sublated. This concretion of universality and particularization
is at first a simple result; it must now further develop itself into totality,
for its moments have at first collapsed into it and do not as yet stand over
against each other in determinate self-subsistence. – The shortcoming of
that result may also be stated more incisively by saying that although in
the disjunctive judgment the objective universality has attained completion
in its particularization, the negative unity of the latter has only retreated
into it and has not as yet determined itself as the third moment, that of
73
Cf. A74/B99ff.Judgment
583
singularity. – But to the extent that the result is itself negative unity, it is
already this singularity; it is then this one determinateness alone that must
now posit its negativity, that must part itself into extremes and in this way
concludes its development in the syllogistic conclusion. 74
The proximate diremption of this unity is the judgment in which the
unity is posited first as subject, as an immediate singular, and then as
predicate, as the determinate connection of its moments.
a. The assertoric judgment
The judgment of the concept is at first immediate; as such, it is the assertoric
judgment. The subject is a concrete singular in general, and the predicate
expresses this same singular as the connection of its actuality, its determi-
nateness or constitution, to its concept. (“This house is bad,” “this action
is good.”) More closely considered, it contains, therefore, (a) that the sub-
ject ought to be something; its universal nature has posited itself as the
self-subsistent concept; (b) that particularity is something constituted or an
external concrete existence, not only because of its immediacy, but because it
expressly differs from its self-subsisting universal nature; its external con-
crete existence, for its part, because of this self-subsistence of the concept, is
also indifferent with respect to the universal and may or may not conform
to it. – This constitution is the singularity which in the disjunctive judg-
ment escapes the necessary determination of the universal, a determination
that exists only as the particularization of the species and as the negative
principle of the genus. Thus the concrete universality that has come out
of the disjunctive judgment divides in the assertoric judgment into the
form of extremes to which the concept itself, as the posited unity connecting
them, is still lacking.
For this reason the judgment is so far only assertoric; its credential is only
a subjective assurance. That something is good or bad, right, suitable or
not, hangs on an external third. But to say that the connectedness is thus
externally posited is the same as saying that it is still only in itself or internal. –
When we say that something is good or bad, etc., we certainly do not mean
to say that it is good only in a subjective consciousness but may perhaps
be bad in itself, or that “good and bad,” “right,” “suitable,” etc. may not
be predicates of the object itself. The merely subjective character of the
assertion of this judgment consists, therefore, in the fact that the implicitly
present connectedness of subject and predicate has not been posited yet, or,
74
There is in German a play on words here. Schluß means both “conclusion” and “syllogism.”
12.86584
Georg Wilhelm Friedrich Hegel
what amounts to the same thing, that it is only external; the copula still is
an immediate abstract being.
Thus the assurance of the assertoric judgment can with right be con-
fronted by an opposing one. When the assurance is given that “this action
is good,” the opposite, “this action is bad,” has equal justification. – Or,
considering the judgment in itself, since its subject is an immediate singular,
in this abstraction it still does not have, posited in it, the determinateness
that would contain its connection with the universal concept; it still is a
contingent matter, therefore, whether there is or there is not conformity to
the concept. Essentially, therefore, the judgment is problematic.
b. The problematic judgment
12.87
The problematic judgment is the assertoric judgment in so far as the latter
must be taken positively as well as negatively. – According to this qualitative
side, the particular judgment is likewise a problematic one, for it has
positive just as much as negative value (equally problematic is also the
being of the subject and predicate in the hypothetical judgment), and also
posited through this side is that the singular judgment and the categorical
are still something merely subjective. In the problematic judgment as such,
however, this positing is more immanent than it is in these others, for in
it the content of the predicate is the connection of the subject to the concept –
here, therefore, the determination of the immediate as something contingent
is itself present.
Whether the predicate ought to be or not to be coupled with a cer-
tain subject appears at first only as problematic, and to this extent the
indeterminateness falls on the side of the copula. The predicate has no
determination to gain from this coupling, since it is already the objec-
tive, concrete universality. The problematic element falls, therefore, on the
immediacy of the subject, which is thereby determined as a contingency. –
But further, we must not for that reason abstract from the singularity of
the subject; purified of such a singularity, the subject would be only a
universal, whereas the predicate entails precisely this, that the concept of
the subject ought to be posited with reference to its singularity. – We may
not say, “the house or a house is good,” but, “so indeed it is in the way
it is made.” – The problematic element in the subject itself constitutes its
moment of contingency, the subjectivity of the fact it expresses as contrasted
with its objective nature or its concept, its mere mode and manner or its
constitution.Judgment
585
Consequently the subject is itself differentiated into its universality or
objective nature, that is, its ought, and the particularized constitution
of immediate existence. It thereby contains the ground for being or not
being what it ought to be. In this way, it is equated with the predicate. –
Accordingly, the negativity of the problematic character of the judgment,
inasmuch as it implicates the immediacy of the subject, only amounts to
this original partition of the latter into its moments of universal and partic-
ular of which it is already the unity – a partition which is the judgment
itself.
One more comment that can be made is that both sides of the subject, its
concept and the way it is constituted, could each be called its subjectivity.
The concept is the universal essence of a fact, withdrawn into itself, the
fact’s negative self-unity; this unity constitutes the fact’s subjectivity. But a
fact is also essentially contingent and has an external constitution; this last
may also be called its mere subjectivity, as contrasted with the objectivity of
the concept. The fact consists just in this, that its concept, as self-negating
unity, negates its universality and projects itself into the externality of
singularity. – As this duplicity, the subject of the judgment is here posited;
the truth of those two opposite meanings of subjectivity is that they are in
one. – The meaning of subjective has itself become problematic by having
lost the immediate determinateness that it had in the immediate judgment
and its determinate opposition to the predicate. – These opposite meanings
of subjectivity that surface even in the ratiocination of ordinary reflection
should by themselves at least call attention to the fact that subjectivity has
no truth in one of them alone. The duplicity of meaning is the manifestation
of the one-sidedness of each when taken by itself.
When this problematic character of the judgment is thus posited as the
character of the fact, the fact with its constitution, the judgment itself is no
longer problematic but apodictic.
c. The apodictic judgment
The subject of the apodictic judgment (“the house, as so and so constituted,
is good,” “the action, as so and so constituted, is right”) includes, first, the
universal, or what it ought to be; second, its constitution; the latter contains
the ground why a predicate of the judgment of the concept does or does
not pertain to it, that is, whether the subject corresponds to its concept or
not. This judgment is now truly objective; or it is the truth of the judgment
in general. Subject and predicate correspond to each other, and have the
same concept, and this content is itself posited concrete universality; that is
12.88586
12.89
Georg Wilhelm Friedrich Hegel
to say, it contains the two moments, the objective universal or the genus
and the singularized universal. Here we have, therefore, the universal that
is itself and continues through its opposite, and is a universal only in unity
with the latter. – Such a universal, like “good,” “fitting,” “right,” etc., has
an ought for its ground, and contains at the same time the correspondence
of existence; it is not the ought or the genus by itself, but this correspondence
which is the universality that constitutes the predicate of the apodictic
judgment.
The subject likewise contains these two moments in immediate unity as
fact. The truth of the latter, however, is that it is internally fractured into
its ought and its being; this is the absolute judgment on all actuality. 75 – That
this original partition, which is the omnipotence of the concept, is equally
a turning back into the concept’s unity and the absolute connection of
“ought” and “being” to each other, is what makes the actual into a fact; the
fact’s inner connection, this concrete identity, constitutes its soul.
The transition from the immediate simplicity of the fact to the correspon-
dence which is the determinate connection of its ought and its being – the
copula – now shows itself upon closer examination to lie in the particular
determinateness of the fact. The genus is the universal existing in and for
itself which, to that extent, appears as unconnected; the determinateness,
however, is that which in that universality is reflected into itself but at the
same time into an other. The judgment, therefore, has its ground in the
constitution of the subject and is thereby apodictic. Consequently, we now
have the determinate and accomplished copula which hitherto consisted in
the abstract “is” but has now further developed into ground in general. It
first attaches to the subject as immediate determinateness, but it is equally
the connection to the predicate – a predicate that has no other content than
this correspondence itself, or the connection of the subject to the universality.
Thus the form of judgment has passed away, first, because subject and
predicate are in themselves the same content; but, second, because through
its determinateness the subject points beyond itself and connects itself to
the predicate; but again, third, this connecting has equally passed over into
the predicate, only constitutes the content of it, and so it is the connecting
as posited or the judgment itself. – The concrete identity of the concept
that was the result of the disjunctive judgment and constitutes the inner
foundation of the judgment of the concept – the identity that was posited
at first only in the predicate – is thus recovered in the whole.
75
Like the Greek kr©siv (cf. crisis), the German Urteil (judgment) connotes “partition.”Judgment
587
On closer examination, the positive factor in this result which is respon-
sible for the transition of the judgment into another form is that, as we
have just seen, 76 the subject and predicate are in the apodictic judgment
each the whole concept. – The unity of the concept, as the determinateness
constituting the copula that connects them, is at the same time distinct
from them. At first, it stands only on the other side of the subject as the
latter’s immediate constitution. But since its essence is to connect, it is not
only that immediate constitution but the universal that runs through the
subject and predicate. – While subject and predicate have the same con-
tent, it is the form of their connection that is instead posited through the
determinateness of the copula, the determinateness as a universal or the par-
ticularity. – Thus it contains in itself both the form determinations of the
extremes and is the determinate connection of the subject and predicate:
the accomplished copula of the judgment, the copula replete of content, the
unity of the concept that re-emerges from the judgment wherein it was lost
in the extremes. – By virtue of this repletion of the copula, the judgment has
become syllogism.
76
In the preceding section 12.88.c h a p ter 3
12.90
The syllogism
The syllogism is the result of the restoration of the concept in the judgment,
and consequently the unity and the truth of the two. The concept as such
holds its moments sublated in this unity; in judgment, the unity is an
internal or, what amounts to the same, an external one, and although the
moments are connected, they are posited as self-subsisting extremes. In the
syllogism, the determinations of the concept are like the extremes of
the judgment, and at the same time their determinate unity is posited.
Thus the syllogism is the completely posited concept; it is, therefore, the
rational. – The understanding is taken to be the faculty of the determinate
concept which is held fixed for itself by virtue of abstraction and the form
of universality. But in reason the determinate concepts are posited in their
totality and unity. Therefore, it is not just that the syllogism is rational but
that everything rational is a syllogism. Syllogistic inference has long since
been ascribed to reason; but, on the other hand, reason in and for itself,
and rational principles and laws, are so spoken of that no light is thrown
on why the one reason that syllogizes, and the other which is the source
of laws and otherwise eternal truths and absolute thoughts, hang together.
If the former is supposed to be only a formal reason while the latter is
supposed to be the one that generates content, then one would expect on
this distinction that precisely the form of reason, the inference, would not
be missing in the latter. And yet, the two are commonly held so far apart,
the one without mention of the other, that it seems as though the reason
of absolute thoughts were ashamed, so to speak, of inferential reason, and
the syllogism were listed as also an activity of reason merely as matter of
tradition. But surely, as we have just remarked, logical reason must be
essentially recognizable, when regarded as formal, also in the reason that
is concerned with a content; indeed, no content can be rational except
by virtue of the rational form. In this matter we cannot rely on what is
commonly said about reason, for common views fail to tell us what we are
to understand by reason; this would-be rational wisdom is so busy with its
588The syllogism
589
objects that it forgets to pay attention to reason itself but only identifies it
by characterizing it through the objects that it is said to have. If reason is
supposed to be a cognition that would know about God, freedom, right and
duty, the infinite, the unconditional, the suprasensible, or even gives only
representations and feelings of such objects, then for one thing these objects
are only negative, and for another the original question still stands, what
is there in all these objects that makes them rational? – The answer is that
the infinitude in them is not the empty abstraction from the finite, is not a
universality which is void of content and determination, but is the fulfilled
universality, the concept which is determined and is truly in possession of its
determinateness, namely, in that it differentiates itself internally and is the
unity of its thus intelligible and determined differences. Only in this way
does reason rise above the finite, the conditioned, the sensuous, or however
one might define it, and is in this negativity essentially replete with content,
for as unity it is the unity of determinate extremes. And so the rational is
nothing but the syllogism.
Now the syllogism, like judgment, is at first immediate; as such, its
determinations (termini) are simple, abstract determinacies; it is then the
syllogism of the understanding. If one stays at this configuration of the
syllogism, then its rationality, though present there and posited, is not
apparent. The essential element of the syllogism is the unity of the extremes,
the middle term that unites them and the ground that supports them.
Abstraction, by holding fast to the self-subsistence of the extremes, posits this
unity opposite them, as a determinateness with just as fixed an existence of its
own, thus grasping it more as a non-unity than as a unity. The expression,
“middle term” (medius terminus), is derived from spatial representation,
and has its share of responsibility for why we stop short at the externality
of the terms. Now if the syllogism consists in the positing in it of the unity
of the extremes, but if this unity is simply taken on the one hand as a
particular by itself, and on the other hand as only an external connection,
and non-unity is made the essential relation of syllogism, then the reason
of the syllogism is of no help to rationality.
First, the syllogism of existence, in which the terms are thus immediately
and abstractly determined, demonstrates internally that, since like judg-
ment it is the connection of those terms, these are not in fact abstract but
each contains in it the reference connecting it to the others, and the deter-
mination of the middle term is not just a determinateness opposed to the
determinations of the extremes but contains these extremes posited in it.
Through this dialectic, the syllogism of existence becomes the syllogism
of reflection, the second syllogism. Its terms are such that in each the other
12.91
12.92590
Georg Wilhelm Friedrich Hegel
shines essentially reflected in it, or are posited as mediated, as they are indeed
supposed to be in accordance with the nature of syllogistic inference in
general.
Third, inasmuch as this reflective shining or this mediatedness is reflected
into itself, syllogism is determined as the syllogism of necessity, one in which
the mediating factor is the objective nature of the fact. As this syllogism
determines the extremes of the concept also as totalities, it has attained the
correspondence of its concept (or the middle term) and its existence (or
the difference of the extremes). It has attained its truth – and with that it
has stepped forth out of subjectivity into objectivity.
a. the syllogism of existence
12.93
1. The syllogism in its immediate form has for its moments the determi-
nations of the concept as immediate. Accordingly, these are the abstract
determinacies of form, such as have not yet been developed by mediation
into concretion but are only singular determinacies. The first syllogism is thus
the one which is strictly formal. The formalism of syllogistic inference con-
sists in stopping short at the form of this first syllogism. The concept, when
partitioned into its abstract moments, has singularity and universality for its
extremes, and itself appears as the particularity that stands between them.
Because of their immediacy, these determinacies only refer to themselves,
one and all a single content. Particularity constitutes at first the middle term
by uniting within itself, immediately, the two moments of singularity and
universality. Because of its determinateness, on the one hand it is subsumed
under the universal; on the other hand, the singular with respect to which it
possesses universality is subsumed under it. This concretion is at first, how-
ever, only a double-sidedness; the middle term, because of the immediacy
that affects it in the immediate syllogism, is a simple determinateness, and
the mediation which it constitutes is not as yet posited. Now the dialectical
movement of the syllogism of existence consists in positing the moments
of the mediation that alone constitutes the syllogism.
a. First figure of the syllogism
S-P-U is the general schema of the determinate syllogism. Singularity con-
nects with 77 universality through particularity; the singular is not universal
immediately but by means of particularity; and conversely, universality is
77
schliesst . . . zusammen: note the schliessen which is connected with Schluss, “syllogism,” “conclusion,”
“inference.” There is a constant play on words in this section which is difficult to render in English
without departing from the text too far.The syllogism
591
likewise not singular immediately but lowers itself to it through particular-
ity. – These determinations stand over against each other as extremes and are
one in a third term which is diverse from them. The two are both determi-
nateness; in this they are identical; this, their universal determinateness, is
particularity. But they are no less extremes with respect to this particularity
than they are to each other, for each is in its immediate determinateness.
The general meaning of this syllogism is that the singular, which as such
is infinite self-reference and consequently would be only an inwardness,
emerges through the medium of particularity into existence, into a uni-
versality wherein it no longer belongs just to itself but stands in external
conjunction; conversely, since in its determinateness the singular sets itself
apart as particularity, in this separation it is a concreted term and, because
of the self-reference of the determinateness, it is a self-referring universal,
and consequently also a true singular; in the extreme of universality the
singular has gone from externality into itself. – The objective significance
of the syllogism is in this first figure only superficially present at first, for the
determinations are not as yet posited in it as the unity which constitutes
the essence of the syllogistic inference. The syllogism is still something
subjective inasmuch as the abstract meaning which its terms have has no
being in and for itself but is rather only in a subjective consciousness, and
is thus isolated. – Moreover, as we have seen, 78 the relation of singularity,
particularity, and universality is the necessary and essential form-relation of
the determinations of the syllogism; the deficiency does not rest in the
determinateness of the form but in that each single determination is not
at the same time richer under it. – Aristotle confined himself rather to
the mere relation of inherence by defining the nature of the syllogism as
follows: When three terms are so related to each other that the one extreme
is in the entire middle term, and this middle term is in the entire other
extreme, then these two extremes are necessarily united in the conclusion. 79
What is here expressed is the repetition of the equal relation of inherence
of the one extreme to the middle term, and then again of this last to the
other extreme, rather than the determinateness of the three terms to each
other. – Now since the syllogistic inference rests on this determinateness of
the terms to each other, it is immediately apparent that the other relations
of terms as are given by the other figures can have validity as inferences of
the understanding only to the extent that they let themselves be reduced to
that original relation; these other are not diverse species of figures that stand
alongside the first but, on the one hand, to the extent that they are assumed
78
In section 12.92.
79
Prior Analytics, 25b32–35.
12.94592
12.95
Georg Wilhelm Friedrich Hegel
to be correct inferences, they rest on the form of syllogistic inference in
general; and, on the other hand, to the extent that they deviate from it,
they are variant forms into which the first abstract form necessarily passes
over and thereby further determines itself and becomes totality. How this
occurs, we must now see in greater detail.
S-P-U is thus the general schema of the syllogism in its determinateness.
The singular is subsumed under the particular and the particular under the
universal; therefore, the singular is also subsumed under the universal. Or
the particular inheres in the singular and the universal in the particular;
therefore, the universal also inheres in the singular. With respect to the side
of the universal, the particular is the subject; with respect to that of the
singular, it is predicate; or as against the one it is singular, as against the
other it is universal. Since both these determinations are united in it, by
virtue of this unity of determinations the extremes are joined together. The
“therefore” appears as an inference that has taken place in the subject and
derives from the subjective insight into the relation of the two immediate
premises. Since subjective reflection expresses the two connections of the
middle to the extremes as particular and indeed immediate judgments or
propositions, the conclusion as the mediated connection is of course also a
particular proposition, and the “hence” or the “therefore,” is the expression
that it is the one which is mediated. But this “therefore” is not to be
regarded as a determination which is external to this proposition, one that
would have its ground and seat in subjective reflection, but as grounded
rather in the nature of the extremes themselves whose connection is again
enunciated as a mere judgment or proposition only for the sake of, and by
virtue of, abstractive reflection, but whose true connection is posited as the
middle term. – “Therefore S is A”: that this is a judgment is a merely
subjective circumstance; that it is not a merely subjective judgment, that is,
not a connection drawn through the mere copula or the empty “is” but one
drawn rather through a determinate middle which is replete with content,
that is precisely the meaning of the syllogistic inference.
For this reason, to regard the syllogism as merely consisting of three judg-
ments is a formalistic view that ignores the relation of the determinations
which alone is at issue in the inference. It is altogether a merely subjective
reflection that splits the connection of the terms into isolated premises and
a conclusion distinct from them:
All humans are mortal,
Gaius is a human
Therefore Gaius is mortal.The syllogism
593
One is immediately seized by boredom the moment one hears this infer-
ence being trotted out, a boredom brought on by the futility of a form that
by means of separate propositions gives the illusion of a diversity which
is immediately dissolved in the fact itself. It is mostly because of this sub-
jective attire that the inference appears as a subjective expedient in which
reason or understanding take refuge when they are incapable of immediate
cognition. – The nature of things, the rational, certainly does not operate
in this way, first by drawing up a major premise for itself, the connection of
some particularity to a subsistent universal; then by discovering in a second
moment the separate connection of a singularity to the particularity, out
of which in a third and final moment a new proportion comes to light. –
This syllogistic inference from one separate proposition to another is noth-
ing but a subjective form; the nature of the fact is that its various determi-
nations are united in a unity of essence. This rationality is not an expedient;
on the contrary, in contrast to the immediacy of the connection that still
obtains in judgment, it is the objective element; it is the prior immediacy
of cognition that rather is mere subjectivity, in contrast to the syllogistic
inference which is the truth of the judgment. – All things are a syllogism, a
universal united through particularity with singularity; surely not a whole
made up of three propositions.
2. In the immediate syllogism of the understanding, the terms have the
form of immediate determinations; we must now consider that syllogism
from this side, according to which the terms are content. We may then
regard it as qualitative, just as we did the judgment of existence 80 which
has the same side of qualitative determination. The terms of this syllogism,
just like the terms of that judgment, are accordingly singular determinacies,
for the determinateness is posited because of its self-reference as indifferent
to form and hence as content. The singular is some immediate concrete
subject matter or other; particularity, one of its determinacies, properties
or relations; universality, a yet more abstract, more singularized determi-
nateness in the particular. – Since the subject, as something immediately
determined, is as yet not posited in its concept, its concretion is not reduced
to its essential determinacies; its self-referring determinateness is therefore
indeterminate, an infinite manifoldness. In this immediacy, the singular
has an infinite multitude of determinacies that belong to its particularity
and any may serve in a syllogism as the middle term for it. Through each
middle term, however, the singular attaches to another universal; through
each of its properties it enters into a different arrangement and context of
80
Cf. above, 12.60.
12.96594
12.97
Georg Wilhelm Friedrich Hegel
existence. – Moreover, in comparison with the universal, the middle term
is also a concreted term; it itself contains several predicates, and through
the same middle term the singular can again attach to several universals.
In general, therefore, it is entirely accidental and arbitrary which of the
many properties of a thing is taken for the purpose of connecting it with a
predicate; other middle terms are transitions to other predicates, and even
the same middle term may by itself be the transition to different predicates,
for as a particular against the universal it contains several determinations.
But not only is an indeterminate number of syllogisms equally possible
for a subject and not only is any single syllogism contingent as regards
content, but these syllogisms that concern the same subject must also run
into contradiction. For difference as such, which is at first an indifferent
diversity, is in essence equally opposition. The concrete is no longer merely
phenomenal but is concrete through the unity in the concept of opposites
that have determined themselves as moments of the latter. Now inasmuch
as in the formal syllogism, in keeping with the qualitative nature of the
terms, the concrete is taken according to one of the single determinations
that pertain to it, the syllogistic inference assigns to it the predicate cor-
responding to this middle term; but inasmuch as from another side the
opposite determinateness is inferred, the previous conclusion turns out to
be false, even though its premises and equally so its consequences are by
themselves quite correct. – If from the middle term that a wall was painted
blue it is concluded that it is blue, this is a correct inference. But, this con-
clusion notwithstanding, the wall can be green if it was also covered over
with a yellow color, a circumstance from which alone it would follow that
the wall is yellow. – If from the senses as middle term it is concluded that
the human being is neither good nor bad, for neither the one nor the other
can be predicated of the senses, the inference is correct; yet the conclusion
is wrong, because to the human being, taken concretely, spirituality also
applies as the middle term. – From the middle term of the gravitation of
the planets, the satellites and comets towards the sun, it follows correctly
that these bodies fall into the sun; but they do not fall into it, because they
are equally their own center of gravity or, as it is said, are driven by
the centrifugal force. Likewise, from sociability as the middle term, the
community of goods among citizens can be inferred; however, from indi-
viduality as the middle term, if the term is pressed with equal abstractness,
there follows the dissolution of the state, as for example it did follow for
the German Empire from adhering to that middle term.– It is only fair
to hold that nothing is as unsatisfactory as such a formal syllogism, since
which middle term is employed is a matter of chance or arbitrariness. NoThe syllogism
595
matter how elegantly a deduction is run through inferences of this kind,
however fully its correctness is to be conceded, all this still amounts to
nothing, for the possibility is still there that other middle terms may be
found from which the opposite can be deduced with equal correctness. –
Kant’s antinomies of reason amount to nothing more than that from a
concept one of its determinations is laid down as ground at one time, and
another determination at another time, both with equal necessity. 81 – The
insufficiency and contingency of an inference must not be blamed in these
cases on the content, as if they were independent of the form and the latter
alone were the concern of logic. On the contrary, it lies in the form of the
formal syllogism that the content is such a one-sided quality; the content
is destined to this one-sidedness because of the form’s abstractness. It is,
namely, one single quality of the many qualities or determinations of a
concrete subject matter, or of a concept, because according to the form it
is not supposed to be anything more than just such an immediate, single
determinateness. The extreme of singularity is, as abstract singularity, the
immediate concrete, consequently an infinite or indeterminate manifold;
the middle term is the equally abstract particularity, consequently a sin-
gle one of these manifold qualities, and likewise the other extreme is the
abstract universal. It is therefore because of its form that the formal syllo-
gism is totally contingent as regards its content, not indeed because to the
syllogism it is accidental whether this or that subject matter is subject to it
(logic abstracts from content), but because, in so far as a subject is laid at
its basis, it is contingent which content determinations it will infer from it.
3. The determinations of the syllogism are determinations of content
inasmuch as they are immediate and abstract determinations reflected
into themselves. But their essence is to be, not immanently reflected and
mutually indifferent determinations, but determinations of form, and to this
extent they are essentially connections. These connections are, first, those
of the extremes to the middle term. These are immediate connections, the
propositiones præmissæ – namely, the connection of the particular to the
universal, the propositio major, and that of the singular to the particular,
the propositio minor. Second, there is the connection of the extremes to one
another, and this is the mediated connection, the conclusio. The immediate
connections, the premises, are propositions or judgments in general, and
they contradict the nature of the syllogism, for according to the latter the
different concept determinations should not be immediately connected
but also their unity should be posited; the truth of the judgment is the
81
For the antinomies, cf. A426/B454ff.
12.98596
12.99
Georg Wilhelm Friedrich Hegel
syllogistic conclusion. And there is all the more reason why the premises
cannot remain immediate connections as their content is made up of
immediately differentiated determinations which, as such, are not in and
for themselves identical – unless the premises are identical propositions,
that is, empty tautologies that lead to nothing.
Accordingly, the normal expectation is that the premises will be proved,
that is, that they ought likewise to be exhibited as conclusions. The two
premises, therefore, yield two further syllogisms. But these two new syllo-
gisms together yield four premises that require four new syllogisms; these
have eight premises whose eight conclusions yield in turn sixteen conclusions
for their sixteen premises, and so on in a geometrical progression to infinity.
Thus we have again the progress to infinity that occurred in the lower
sphere of being but we would not expect now in the domain of the concept,
the domain of the absolute reflection from the finite to the self, the region of
free infinity and truth. It was shown in the sphere of being 82 that whenever
the bad infinity that runs away into a progression raises its head, what we
have is the contradiction of a qualitative being and of an impotent ought
that would transcend it; the progression itself is the repeated demand that
there be unity that intervenes to confront the qualitative, and the constant
fall back into the limitation which is inadequate to the demand. Now in
the formal syllogism the immediate connection or the qualitative judgment
is the basis, and the mediation of the syllogism is the higher truth posited
over against it. The infinite progression of the proof of the premises does
not resolve this contradiction but only perpetually renews it and is the
repetition of one and the same original deficiency. – The truth of the
infinite progression is rather the sublation of it and of the form which the
progression itself has already determined as deficient. – This form is that
of the mediation S-P-U. The two connections, S-P and P-U are supposed
to be mediated; if this is done in the same manner, only the deficient form
S-P-U is replicated, and so on to infinity. With respect to S, P also has the
form determination of a universal; and with respect to U that of a singular,
for these connections are as such judgments. As such, they are in need
of mediation; but in that form of mediation, only the relation that was
supposed to be sublated comes up again.
The mediation must therefore occur in some other way. For the medi-
ation of P-U, there is S available; hence the mediation must be given the
shape of P-S-U. To mediate S-P, there is U available; accordingly, this
mediation becomes S-U-P.
82
Cf. GW 11.79; above 21.127ff.The syllogism
597
If this transition is examined more closely in light of its concept, then,
as was shown earlier, 83 the mediation of the formal syllogism is in the first
instance contingent according to content. The immediate singular has in
its determinacies an indeterminate number of middle terms, and these
have in general equally as many determinacies in turn; it is, therefore,
entirely a matter of external arbitrary choice, or in general of an external
circumstance and accidental determination, as to which universal the subject
of the syllogism should be annexed. As regards its content, therefore, the
mediation is not anything necessary, nor universal; it is not grounded in
the concept of the fact but the basis of the inference is something external
to it, that is, the immediate; but of the determinations of the concept, it is
the singular which is the immediate.
With respect to form, the mediation likewise presupposes the immediacy
of connection; the mediation itself is thus mediated – mediated indeed
by means of the immediate, that is, the singular. – More precisely, the
singular has become a mediating term through the conclusion of the first
syllogism. That conclusion is S-U; the singular is thereby posited as a
universal. In one premise, that is, the minor S-P, it is already as a particular;
consequently, the singular is that in which these two determinations are
united. – Or the conclusion expresses in and for itself the singular as a
universal, and it does it, not in any immediate manner, but mediatedly,
hence as a necessary connection. The simple particularity was the middle
term; in the conclusion, this particularity is posited as developed as the
connection of singular and universality. But the universal is still a qualitative
determinateness, the predicate of the singular; in being determined as
universal, the singular is posited as the universality of the extremes or as
the middle; it is for itself the extreme of singularity, but since it is now
determined as a universal, it is at the same time the unity of the two
extremes.
b. The second figure: P-S-U
1. The truth of the first qualitative syllogism is that something is not in
and for itself united to a qualitative determinateness which is a universal,
but is united to it by means of a contingency or in a singularity. The subject
of the syllogism has not returned in such a quality to its concept but is
conceived only in its externality; the immediacy constitutes the basis of the
connection and hence the mediation; to this extent, the singular is in truth
the middle.
83
Cf. above, 12.97.
12.100598
12.101
Georg Wilhelm Friedrich Hegel
But further, the syllogistic connection is the sublation of the immediacy;
the conclusion is a connection drawn not immediately but through a third
term; therefore, it contains a negative unity; therefore, the mediation is
now determined as containing a negative moment within it.
In this second syllogism, the premises are: P-S and S-U; only the first
of these premises is still an immediate one; the second, S-U, is already
mediated, namely through the first syllogism; the second syllogism thus
presupposes the first just as, conversely, the first presupposes the second. –
The two extremes are here determined, the one as against the other, as
particular and universal. The latter thus retains its place; it is predicate.
But the particular has exchanged places; it is subject or is posited in the
determination of the extreme of singularity, just as the singular is posited
with the determination of the middle term or of particularity. The two no
longer are, therefore, the abstract immediacies which they were in the first
syllogism. However, they are not yet posited as concrete somethings; in
standing in the place of the other, each is thereby posited – in its own
determination and at the same time, although only externally – into that
of the other.
The determinate and objective meaning of this syllogism is that the
universal is not in and for itself a determinate particular (it is rather the
totality of its particulars) but that it is one of its species through the mediation
of singularity; the rest of its species are excluded from it by the immediacy
of externality. Likewise the particular is not for its part immediately, and
in and for itself, the universal; the negative unity is rather what removes
the determinateness from it and thereby raises it to universality. – The
singularity thus relates to the particular negatively in so far as it is supposed
to be its predicate; it is not the predicate of the particular.
2. But the terms are at first still immediate determinacies; they have not
advanced of their own to any objective signification; the positions which
two of them have exchanged and now occupy is the form, and this is as yet
only external to them. Therefore they are still, as in the first syllogism, each
a content indifferent as such to the other – two qualities linked together,
not in and for themselves, but through the mediation of an accidental
singularity.
The syllogism of the first figure was the immediate syllogism, or again,
the syllogism in so far as its concept is an abstract form that has not yet
realized itself in all its determinations. The transition of this pure form into
another figure is on the one hand the beginning of the realization of the
concept, in that the negative moment of the mediation, and thereby one
further determinateness of the form, is posited in the originally immediate,The syllogism
599
qualitative determinateness of the terms. – But, on the other hand, this
is at the same time an alteration of the pure form of the syllogism; the
latter no longer conforms to it fully, and the determinateness posited in its
terms is at variance with that original form determination. – In so far as
it is regarded as only a subjective syllogism that runs its course in external
reflection, we can then take it as a species of syllogistic inference that should
conform to the genus, namely the general schema S-P-U. But it does not
at the moment conform to it; its two premises are P-S or S-P and S-U; the
middle term is in both cases the one which is subsumed or is the subject
in which the two other terms thus inhere – is not therefore a middle
term that in one case would subsume or be predicate, and in the other
would be subsumed or be subject, or a middle in which one of the terms
would inhere but would itself inhere in the other. – The true meaning
of this syllogism’s lack of conformity to the general form of the syllogism
is that the latter has passed over into it, for its truth consists in being a
subjective, contingent conjoining of terms. If the conclusion in this second
figure is correct (that is, without recurring to the restriction, to which we
shall presently turn, that makes of it something indeterminate), then it is
correct because it is so on its own, not because it is the conclusion of this
syllogism. But the same is the case for the conclusion of the first figure;
it is this, the truth of that first figure, which is posited by the second. –
On the view that the second figure is only one species, we overlook the
necessary transition of the first figure into this second and stop short at the
first as the true form. Hence, if in the second figure (which from ancient
custom is referred to, without further ground, as the third) we are equally
supposed to find a correct syllogism in this subjective sense, this syllogism
would have to be commensurate with the first; consequently, since the one
premise S-U has the relation of the subsumption of the middle term under
one extreme, then it would have to be possible for the other premise S-P to
receive the opposite relation to that which it has, and for P to be subsumed
under S. But such a relation would be the sublation of the determinate
judgment S is P, and could only occur in an indeterminate judgment, a
particular judgment; consequently, the conclusion in this figure can only
be particular. But the particular judgment, as we remarked above, 84 is
positive as well as negative – a conclusion, therefore, to which no great
value can be ascribed. – Since the particular and universal are also the
extremes, and are immediate determinacies indifferent to each other, their
relation itself is indifferent; each can be the major or the minor term,
84
Cf. above, 12.73.
12.102600
Georg Wilhelm Friedrich Hegel
indifferently the one or the other, and consequently either premise can also
be taken as major or minor.
3. Since the conclusion is positive as well as negative, it is a connec-
tion which for that reason is indifferent to these determinacies, hence a
universal connection. More precisely, the mediation of the first syllogism
was implicitly a contingent one; in the second syllogism, this contingency is
posited. Consequently, the mediation is self-sublating; it has the determina-
tion of singularity and immediacy; what this syllogism joins together must,
on the contrary, be in itself and immediately identical, for that mediating
middle, the immediate singularity, is an infinitely manifold and external
determining. Posited in it, therefore, is rather the self-external mediation.
The externality of singularity, however, is universality; that mediation by
means of the immediate singular points beyond itself to the mediation
which is the other than it, one which therefore occurs by means of the
universal. – In other words, what is supposed to be united by means of
the second syllogism, must be immediately conjoined; the immediacy on
which it is based does not allow any definite conclusion. The immediacy
to which this syllogism points is the opposite of its own: it is the sublated
first immediacy of being, therefore the immediacy reflected into itself or
the abstract universal existing in itself.
From the standpoint of the present consideration, the transition of
this syllogism was like the transition of being an alteration, for its base
is qualitative; it is the immediacy of singularity. But according to the
concept, singularity conjoins the particular and the universal by sublating
the determinateness of the particular – and this is what presents itself as the
contingency of this syllogistic inference. The extremes are not conjoined
by the specific connective which they have in the middle term; this term
is not, therefore, their determinate unity, and the positive unity that yet
pertains to it is abstract universality. But inasmuch as the middle term is
posited in this determination which is its truth, we have another form of
the syllogism.
c. The third figure: S-U-P
12.103
1. This third syllogism no longer has any single immediate premise; the
connection S-U has been mediated by the first syllogism; the connection
P-U by the second. It thus presupposes both these syllogisms; but conversely
it is presupposed by them, just as in general each presupposes the other
two. In this third figure, therefore, it is the determination of the syllogism
as such that is brought to completion. – This reciprocal mediation meansThe syllogism
601
just this, that each syllogism, although for itself a mediation, does not
possess the totality of mediation but is affected by an immediacy whose
mediation lies outside it.
Considered in itself, the syllogism S-U-P is the truth of the formal
syllogism; it expresses the fact that its mediating middle is the abstract
universal and that the extremes are not contained in it according to their
essential determinateness but only according to their universality – that
precisely that is not conjoined in it, which was supposed to be mediated.
Posited here, therefore, is that wherein the formalism of the syllogism
consists – that its terms have an immediate content which is indifferent
towards the form, or, what amounts to the same, that they are such form
determinations as have not yet reflected themselves into determinations of
content.
2. The middle of this syllogism is indeed the unity of the extremes,
but a unity in which abstraction is made from their determinateness, the
indeterminate universal. But in so far as this universal is at the same time
distinguished from the extremes as the abstract from the determinate, it is
itself also a determinate as against them, and the whole is a syllogism whose
relation to its concept needs examining. As the universal, the middle term is
with respect to both its extremes the term that subsumes or the predicate –
not a term for once also subsumed or the subject. Now as a species of
syllogism, it ought to conform to the latter, and this can only happen
on condition that, inasmuch as the one connection S-U already possesses
the appropriate relation, the other connection P-U contains it too. This
occurs in a judgment in which the relation of subject and predicate is an
indifferent one, in a negative judgment. Thus does the syllogism become
legitimate, but the conclusion is necessarily negative.
Consequently, also indifferent is now which of the two determinations
of this proposition is taken as predicate or subject, and whether the deter-
mination is taken in the syllogism as the extreme of singularity or the
extreme of particularity, hence as the minor or major term. Since on the
usual assumption which of the premises is supposed to be the major or
the minor depends on this distinction, this too has now become a matter
of indifference. – This is the ground of the customary fourth figure of the
syllogism which was unknown to Aristotle and has to do with an entirely
void and uninteresting distinction. In it the immediate position of the
terms is the reverse of their position in the first figure; since from the point
of view of the formal treatment of judgment the subject and predicate of
the negative conclusion do not have the determinate relation of subject
and predicate, but each can take the place of the other, it is a matter of
12.104602
Georg Wilhelm Friedrich Hegel
indifference which term is taken as subject and which as predicate; and just
as indifferent is therefore which premise is taken as the major and which as
the minor. – This indifference, to which the determination of particularity
also contributes (especially if it is noted that this particularity can be taken
in a comprehensive sense), makes of this fourth figure something totally
idle.
3. The objective significance of the syllogism in which the universal is
the middle is that the mediating term, as the unity of the extremes, is
essentially a universal. But since the universality is at first only qualitative or
abstract, the determinateness of the extremes is not contained in it; their
being conjoined in the conclusion, if the conjunction is to take place, must
likewise have its ground in a mediation that lies outside this syllogism and
is, with respect to the latter, just as contingent as it is in the preceding forms
of the syllogism. But now, since the universal is determined as the middle
term, and since the determinateness of the extremes is not contained in
this middle, the latter is posited as one which is wholly indifferent and
external. – It is here, by virtue indeed of a bare abstraction, that a fourth
figure of the syllogism arose in the first place, namely the figure of the
relationless syllogism, U-U-U, which abstracts from the qualitative differ-
entiation of the terms and therefore has their merely external unity, their
equality, for its determination.
d. The fourth figure: U-U-U, or the mathematical syllogism
12.105
1. The mathematical syllogism goes like this: if two things or two determi-
nations are equal to a third, then they are equal to each other. – The relation
of inherence or subsumption of terms is done away with.
A “third” is in general the mediating term; but this third has absolutely
no determination as against the extremes. Each of the three terms can
therefore be the mediating term just as well as any other. Which is needed
for the job, which of the three connections are therefore to be taken as
immediate, and which as mediated, depends on external circumstances
and other conditions, namely which two of the three are immediately
given. But this determination does not concern the syllogism and is wholly
external.
2. The mathematical syllogism ranks in mathematics as an axiom, as a first
self-explanatory proposition which is neither capable nor in need of proof,
i.e of any mediation – which neither presupposes anything else nor can be
derived from anything else. – If we take a closer look at this prerogative that
the proposition claims, of being immediately self-evident, we find that it liesThe syllogism
603
in its formalism, in the fact that it abstracts from every qualitative diversity
of determinations and only admits their quantitative equality or inequal-
ity. But for this very reason it is not without presupposition or mediation;
the quantitative determination, which alone comes into consideration in
it, is only by virtue of the abstraction from qualitative differentiation and
from the concept determinations. – Lines, figures, posited as equal to each
other, are understood only according to their magnitude. A triangle is
posited as equal to a square, not however as triangle to square but only
according to magnitude, etc. Nor does the concept and its determinations
enter into this syllogism; there is in it, therefore, no conceptual comprehen-
sion at all; the understanding is also not faced here by even the formal,
abstract determinations of the concept. The self-evidence of this syllogism
rests, therefore, solely on the indigence and abstractness of its mode of
thought.
3. But the result of the syllogism of existence is not just this abstraction
from all determinateness of the concept; the negativity of the immediate
and abstract determinations that emerged from it has yet another positive
side, namely that in the abstract determinateness its other has been posited
and the determinateness has thereby become concrete.
In the first place, the syllogisms of existence all have one another for
presupposition, and the extremes conjoined in the conclusion are truly
conjoined, in and for themselves, only inasmuch as they are otherwise
united by an identity grounded elsewhere; the middle term, as constituted
in the syllogisms we have examined, ought to be the conceptual unity of
these syllogisms but is in fact only a formal determinateness that is not
posited as their concrete unity. But what is thus presupposed by each and
every of these mediations is not merely a given immediacy in general, as is
the case for the mathematical syllogism, but is itself a mediation, namely
of each of the other two syllogisms. Therefore, what is truly present here
is not a mediation based on a given immediacy, but a mediation based on
mediation. And this mediation is not quantitative, not one that abstracts
from the form of mediation, but is rather a self-referring mediation, or the
mediation of reflection. The circle of reciprocal presupposing which these
syllogisms bring to closure is the turning back of this presupposing into
itself – a presupposing that in this turning back forms a totality, and has
the other to which every single syllogism refers, not outside by virtue of
abstraction, but included within the circle.
Further, from the side of the single determinations of form it has been
shown that in this whole of formal syllogisms each single determination
has in turn occupied the place of the middle term. As immediate, this term
12.106604
Georg Wilhelm Friedrich Hegel
was determined as particularity; thereupon, through dialectical movement
it determined itself as singularity and universality. Likewise did each of
these determinations occupy the places of both of the two extremes. The
merely negative result is the dissolution of the qualitative determinations
of form into the merely quantitative, mathematical syllogism. But what
we truly have here is the positive result, namely that mediation occurs, not
through any single qualitative determinateness of form, but through the
concrete identity of the determinacies. The deficiency and formalism of the
three figures of the syllogism just considered consists precisely in this, that
one such single determinateness was supposed to constitute the middle
term in it. – Mediation has thus determined itself as the indifference of the
immediate or abstract determinations of form and the positive reflection
of one into the other. The immediate syllogism of existence has thereby
passed over into the syllogism of reflection.
Remark
In the account here given of the nature of the syllogism and its various
forms, passing reference was also made to what constitutes in the ordinary
examination and treatment of the syllogisms the main object of interest,
namely how the right conclusion may be drawn in each figure; but only
the main point came in for consideration, and no mention was made of
the cases and the intricacies that arise when the distinction of positive
and negative judgment is also brought in alongside the determination
of quantity, of particularity especially. – A few comments on how the
syllogism is ordinarily viewed and treated in logic have their place here. –
This doctrine was famously elaborated to such precision of detail that its
hair-splittings, as they came to be called, have been the object of universal
aversion and disgust. The natural understanding 85 in asserting itself over
the unsubstantial forms of reflection in all areas of humanistic culture also
turned against this artificial knowledge of the forms of reason, believing
that it could dispense with the science of such forms on the ground that
it performed the individual operations of thought treated there naturally
and spontaneously, without specialized training. Indeed, with respect to
rational thinking, if its pre-condition were the laborious study of syllogistic
formulas, humankind would be in just as sorry a state as they would (as we
already remarked in the Preface) 86 if they could not walk or digest without
the previous study of anatomy and physiology. But if we admit that there
might be some use to the study of these sciences for dietary purposes,
85
The reference here is to common-sense philosophy.
86
Cf. GW 11, 6; above, 21.6.The syllogism
605
surely we must credit the study of the forms of reason with an even more
important influence on the correctness of thinking. But without going
here into this aspect of the study that concerns the education of subjective
thinking and hence, strictly speaking, pedagogics, it must be granted that
a study that has for its subject matter the modes of operation and the laws
of reason must be of the greatest interest in and of itself – an interest at
least not inferior to the knowledge of the laws and the particular shapes
of nature. If it is not reputed a small matter to have discovered some sixty
species of parrots, one hundred and thirty-seven species of veronica, and
so on, much less ought it to be reputed a small matter to have discovered
the forms of reason. Is not the figure of a syllogism something infinitely
higher than a species of parrot or veronica?
Yet, although contempt for the knowledge of the forms of reason must
be viewed as nothing short of barbarism, it must equally be granted that the
customary exposition of the syllogism and of its particular configurations
is not a rational cognition, not an exposition of them as forms of reason,
and that syllogistic wisdom has by its own unworthiness brought upon
itself the disparagement that it has experienced. Its deficiency consists in
the fact that it simply stops short at the form which the understanding
gives to the syllogism and in which the determinations of the concept are
taken as abstract formal determinations. It is all the more inconsequent
to cling to them as abstract qualities, since in the syllogism it is their
connections that constitute the essential element, and the inherence and the
subsumption already imply that the singular, since the universal inheres in
it, is itself universal, and the universal, since it subsumes the singular, is
itself singular; more to the point, the syllogism expressly posits this very
unity as the middle term, and its determination is mediation itself, that
is, the concept determinations no longer have, as they did in judgment,
their reciprocal externality for their basis, but rather their unity. – It is
thus the concept of the syllogism that gives away the imperfection of the
formal syllogism in which the middle is supposed to be held fixed, not
as the unity of the extremes, but as a formal and abstract determination,
qualitatively at variance with them. – The treatment is made even more
vacuous by the fact that such connections or judgments in which the
formal determinations become indifferent, as in negative and particular
judgments, and which therefore approximate propositions, are also still
regarded as perfect relations. – Now since the qualitative form S-P-U is
generally accepted as ultimate and absolute, the dialectical treatment of the
syllogism falls entirely by the wayside, and the rest of the syllogisms are
thereby treated, not as necessary alterations of that form, but as species. – It is
12.107606
12.108
Georg Wilhelm Friedrich Hegel
then indifferent whether the first formal syllogism is regarded as itself only
one species alongside the others or as rather genus and species at the same
time. This last is the case when the rest of the syllogisms are reduced to the
first. But even when this reduction is not explicitly carried out, the basis is
yet always the same formal relation of external subsumption expressed by
the first figure.
This formal syllogism is the contradiction that the middle term ought
to be the determinate unity of the extremes – not, however, as this unity
but as a determination qualitatively distinct from the terms whose unity
it ought to be. Because the syllogism is this contradiction, it is inherently
dialectical. Its dialectical movement displays it in the full range of the
moments of the concept, so that not only the said relation of subsumption
or particularity, but just as essentially the negative unity and the universality,
are moments in the process of closing this unity. In so far as each of these
equally is by itself only a one-sided moment of particularity, they are
likewise imperfect middle terms, but at the same time they constitute the
developed determinations of the middle term; the entire course across the
three figures displays this middle in each of these determinations, one after
the other, and the true result that emerges from it is that the middle is not
any single one of them but the totality of them all.
The deficiency of the formal syllogism does not rest, therefore, in the
form of the syllogism (which is, on the contrary, the form of rationality) but
in that the form is only an abstract one, and hence void of the concept. It
was shown that the abstract determination, on account of its abstract self-
reference, can equally be regarded as content. 87 But then, all that the formal
syllogism achieves is that a connection of subject and predicate follows or
does not follow from only this middle term. It does not help to prove a
proposition by means of it; on account of the abstract determinateness of
the middle term which is a quality without concept, there can just as well be
other middle terms from which the opposite would follow; even from the
same middle term opposite predicates can also be deduced in turn through
further middle terms. – Besides being of little use, the formal syllogism is
also something very simple; the many rules that have been fabricated are
already tiresome because they contrast so strongly with the simplicity of
the fact at issue, but then also because they apply to cases where the formal
worth of the syllogism is especially diminished by the externality of the
form determination, notably that of particularity (especially because for
87
Hegel is very likely referring to the classical “atoms” which, as empty self-references, acquire an
immediate, external, existence. Cf. the Remark above, 21.153ff.; GW 11, 93.The syllogism
607
this purpose the latter must be taken in a comprehensive sense), and where
even with respect to the form only totally empty results are produced. –
But the most justified and important aspect of the disfavor into which
syllogistic theory has fallen is that, as such a long-drawn-out occupation
with a subject matter whose single content is none other than the concept,
it is itself without concept. – The many syllogistic rules remind one of the
procedure of the arithmeticians who also give a great many rules about
arithmetical operations, all of which presuppose that one has not the
concept of operation. – But numbers are a material with no concept and
the operations of arithmetic are an external combining or separating, a
mechanical procedure – indeed, calculating machines have been invented
that execute these operations perfectly well. The form determinations of
the syllogism are on the contrary concepts, and it is the most glaring of
scandals when they are treated as a material with no concept.
The extreme example of this mechanical treatment of the concept deter-
minations of the syllogism is surely that of Leibniz. q He subjected the
syllogism to a combinatory calculus, thereby reckoning the number of
possible positions of the syllogism – that is, with respect to the distinc-
tion of positive and negative, then of universal, particular, indeterminate
and singular judgments. He found that there are 2,048 such possible com-
binations, of which, after the exclusion of the useless figures, 24 useful
ones remain. – Leibniz makes much of the usefulness of this combina-
tory analysis, not only in order to discover the forms of the syllogism but
also the combinations of other concepts. The operation by which this is
accomplished is the same as calculating how many combinations of let-
ters an alphabet allows, how many throws are possible in a game of dice,
how many plays with an ombre card, etc. We see here the determina-
tions of the syllogism thus placed in the same class as the points of the
dice and the ombre card; we see the rational treated as something dead
and empty of concept, the neglect of that which characterizes the concept
and its determinations most, namely that these, as spiritual essences, enter
into connections and thereby sublate their immediate determination. – This
Leibnizian application of combinatory calculus to the syllogism and to the
combination of other concepts differs from the disreputable Art of Lully 89
solely because it is more methodical on the numerical side, but for the rest
it equals it in meaninglessness. – Connected with this was an idea dear to
q
88
89
Opp. Tom. II, P. I. 88
Leibniz, Dissertatio de arte combinatoria, Opera (ed. Dutens), Tomus II, Pars I, 352–361.
Raymond Lully, c. 1232–1315. His calculation tables were reproduced in the Ars magna sciendi (1669)
by the Jesuit Athanasius Kircher.
12.109608
12.110
Georg Wilhelm Friedrich Hegel
Leibniz, one which he conceived in his youth and, despite its immaturity
and shallowness, never abandoned even in later life. 90 This was the idea of
a characteristica universalis of concepts – a standard language in which each
concept is presented as a connection of other concepts or as connecting
with others – as if in a rational combination, which is essentially dialectical,
a content would still retain the same determinations that it has when fixed
in isolation.
Ploucquet’s calculus has undoubtedly hit upon the most consequent
method for preparing the syllogistic relation to be subjected to calculus. 91
It is based on abstracting in a judgment from the difference of relation,
the difference of singularity, particularity, and universality, and restricting
oneself to the abstract identity of subject and predicate whereby the two
are in mathematical equality – a connection that makes of the syllogistic
inference a totally empty and tautological construal of propositions. – In
the proposition, “the rose is red,” the predicate is taken to mean not the
universal red but only the determinate red of the rose; in the proposition, “all
Christians are human beings,” the predicate is taken to mean only those
humans who are Christian, from which proposition and the proposition,
“the Jews are not Christian,” there follows the conclusion (which did not
recommend this syllogistic calculus to Mendelssohn) that “therefore the
Jews are not humans” (namely, not those humans that the Christians
are). – Ploucquet states as a consequence of his invention: posse etiam rudes
mechanice totam logicam doceri, uti pueri arithmeticam docentur, ita quidem,
ut nulla formidine in ratiociniis suis errandi torqueri, vel fallaciis circumveniri
possint, si in calculo non errant. 92 – This recommendation, that by means
of calculus the whole of logic can mechanically be made available to the
uneducated, is surely the worst that can be said of an invention that bears
on the presentation of the science of logic.
90
91
92
See Leibniz, Nouveaux essais sur l’entendement par l’auteur du système de l’harmonie préestablie, in
Die philosophischen Schriften von Gottfried Wilhelm Leibniz, Vol. 5, ed. C. J. Gerhardt (Hildesheim:
Georg Olms, 1960), pp. 365f.; New Essays on Human Understanding, trans., ed. Peter Remnant and
Jonathan Bennett (Cambridge: Cambridge University Press, 1981), pp. 384f.
Gottfried Plouquet (1716–1790), Expositiones Philosophiæ Theoreticæ, Logik (1782), ed., trans. into
German (original Latin text included), Michael Franz, Studien und Materialen zur Geschichte der
Philosophie (Hildesheim: Olms, 2006).
“I can teach the whole logic mechanically even to the uneducated, just as children are taught
arithmetic, so that, if there is no error of calculation, it would be possible not to be tormented by
the fear of erring in reasoning or of being deceived by falsities.” According to the editors of the
critical edition, this quote is not taken directly from Plouquet. GW 12, 347 (to page 110.14–18). In
§70 of the 1782 edition (see previous note 91), Plouquet only says that “by means of this [following]
theory, the whole business of syllogistic discovery and judgment . . . can be taken care of with the
greatest facility and be reduced without fear of error to a logical calculus.”The syllogism
609
b. the syllogism of reflection
The course of the qualitative syllogism has sublated the abstractness of
its terms; the syllogistic term has thus posited itself as a determinate-
ness in which also the other determinateness shines reflectively. Besides
the abstract terms, there is also present in the syllogism the connection of
the terms, and in the conclusion this connection is posited as one which
is mediated and necessary; in truth, therefore, each determinateness is
posited, not singly by itself, but with reference to the others, as concrete
determinateness.
The middle term was the abstract particularity, an isolated simple deter-
minateness, and was a middle only externally and relative to the self-
subsisting extremes. This term is now posited as the totality of the deter-
minations; thus it is the posited unity of the extremes; but this unity
is at first that of a reflection embracing the extremes within itself –
an embracing which, as a first sublating of immediacy and a first con-
necting of the determinations, is not yet the absolute identity of the
concept.
The extremes are the determinations of the judgment of reflection,
singularity proper, and universality as a determination of relation, or a
reflection that embraces a manifold within itself. But, as was shown in
connection with the judgment of reflection, 93 the singular subject also
contains, besides the mere singularity that belongs to form, determinateness
as universality absolutely reflected into itself, as presupposed, that is, here
still immediately assumed, genus.
From this determinateness of the extremes, which belongs to the
course of the determination of the judgment, there results the more pre-
cise content of the middle, which is what counts most in the syllogism,
for it is the middle that distinguishes the syllogism from judgment. The
middle contains (1) singularity; (2) but singularity expanded into uni-
versality, as an “all”; (3) the universality that lies at the basis, uniting
singularity and abstract universality in itself, the genus. – The syllogism
of reflection is thus the first to possess genuine determinateness of form,
for the middle is posited as the totality of determinations; the immedi-
ate syllogism is by contrast indeterminate because the middle is still only
abstract particularity in which the moments of its concept are not yet
posited. – This first syllogism of reflection may be called the syllogism of
allness.
93
Cf. above, 12.73–74.
12.111610
Georg Wilhelm Friedrich Hegel
a. The syllogism of allness
12.112
1. The syllogism of allness is the syllogism of the understanding in its
perfection, but more than that it is not yet. That the middle in it is not
abstract particularity but is developed into its moments and is therefore
concrete, is indeed an essential requirement of the concept. But at first the
form of the allness gathers the singular into universality only externally, and
conversely the singular behaves in the universality still as an immediate that
subsists on its own. The negation of the immediacy of the determinations
which was the result of the syllogism of existence is only the first negation,
not yet the negation of the negation, or absolute immanent reflection. The
singular determinations that the universality of reflection holds within still
lie, therefore, at the basis of that universality – in other words, allness is
not yet the universality of the concept, but the external universality of
reflection.
The syllogism of existence was contingent because its middle term was
one single determinateness of the concrete subject and as such admitted
of a multitude of other such middle terms, and consequently the subject
could be conjoined in conclusion with an indeterminate number of other
predicates, with opposite predicates as well. But since the middle term
now contains singularity and is thereby itself concrete, only a predicate that
concretely belongs to the subject can be attached to the latter by means of
it. – For instance, if from the middle term “green” the conclusion is made to
follow that a painting is pleasing, because green is pleasing to the eye, or if a
poem, a building, etc., is said to be beautiful because it possesses regularity,
the painting, the poem, the building, etc., may nonetheless still be ugly on
account of other determinations from which this predicate “ugly” might
be deduced. By contrast, when the middle term has the determination of
allness, it contains the green, the regularity, as a concreted term which for
that very reason is not the abstraction of a mere green, a mere regular, etc.;
only predicates commensurate with concrete totality may now be attached to
this concreted term. – In the judgment, “what is green or regular is pleasing,”
the subject is only the abstraction of green, regularity; in the proposition,
“all things green or regular are pleasing,” the subject is on the contrary all
actual concrete things that are green or regular – things, therefore, that are
intended as concreted with all the properties that they may also have besides
the green or the regularity.
2. However, this very reflective perfection of the syllogism makes of
it a mere illusion. The middle term has the determinateness of “all,” to
which there is immediately attached in the major the predicate which in theThe syllogism
611
conclusion is then conjoined with the subject. But the “all” is “all singulars”;
in it, therefore, the subject already possesses that predicate immediately –
it does not first obtain it by means of the syllogistic inference. – Or again,
the subject obtains a predicate as a consequence through the conclusion;
but the major premise already contains this conclusion in it; therefore the
major premise is not correct on its own account, or is not an immediately
presupposed judgment, but itself already presupposes the conclusion of which
it should be the ground. – In the much cited syllogism:
All humans are mortal,
Now Gaius is a human,
Therefore Gaius is mortal,
the major premise is correct only because and to the extent that the conclu-
sion is correct; were Gaius by chance not mortal, the major premise would
not be correct. The proposition which was supposed to be the conclusion
must be correct on its own, immediately, for otherwise the major premise
would not include all singulars; before the major premise can be accepted
as correct, the antecedent question is whether the conclusion may not be a
counter-instance of it.
3. It followed from the concept of the syllogism, with regard to the
syllogism of existence, that the premises, as immediate, contradicted the
conclusion, that is to say, contradicted the mediation that the concept
of the syllogism requires; that the first syllogism thus presupposed other
syllogisms, and conversely these presupposed the first. 94 In the syllogism
of reflection this result is posited in the syllogism itself: the major premise
presupposes its conclusion, for it contains the union of the singular with a
predicate that would have to be a conclusion first.
What we have here in fact can therefore be expressed by saying that
the syllogism of reflection is only an external, empty reflective semblance
of syllogistic inference; that therefore the essence of the inference rests on
subjective singularity; this singularity thus constitutes the middle term and
is to be posited as such: singularity which is singularity as such and pos-
sesses universality only externally. – Or what has been shown on closer
inspection of the content of the syllogism of reflection is that the sin-
gular stands connected to its predicate immediately, not by way of an
inference, and that the major premise, the union of a particular with a
universal, or more precisely of a formal universal with a universal in itself,
is mediated through the connection of the singularity that is present in
94
Cf. above, 12.97–98.
12.113612
Georg Wilhelm Friedrich Hegel
the formal universal, of singularity as allness. But this is the syllogism of
induction.
b. The syllogism of induction
1. The syllogism of allness comes under the schema of the first figure, S-P-U;
the syllogism of induction under that of the second, U-S-P, because it again
has singularity for its middle term, not abstract singularity but singularity
as completed, that is to say, posited with its opposite determination, that of
universality. – The one extreme is some predicate or other which is common
to all these singulars; its connection with them makes up the kind of
immediate premises, of which one was supposed to be the conclusion in
the preceding syllogism. – The other extreme may be the immediate genus,
as it is in the middle term of the preceding syllogism, or in the subject
of the universal judgment, and which is exhausted in the collection of
singulars or also species of the middle term. Accordingly, the syllogism has
this configuration:
s
s
U– –P
s
s
ad
infinitum.
12.114
2. The second figure of the formal syllogism, U-S-P, does not correspond
to this schema, because the S that constitutes the middle term did not
subsume or was not a predicate. In induction this deficiency is eliminated;
here the middle term is “all singulars”; the proposition, U-S, which contains
as the subject the objective universal or the genus set apart as an extreme,
has a predicate which is of at least equal extension as the subject and is
consequently identical with it for external reflection. Lion, elephant, etc.,
constitute the genus of quadruped; the difference, that the same content
is posited once in singularity and again in universality, is thus just an
indifferent determination of form – an indifference which in the syllogism
of reflection is the posited result of the formal syllogism and is posited here
through the equality of extension.
Induction, therefore, is not the syllogism of mere perception or of contin-
gent existence, like the second figure corresponding to it, but the syllogism
of experience – of the subjective gathering together of singulars in the genus,The syllogism
613
and of the conjoining of the genus with a universal determinateness on
the ground that the latter is found in all singulars. It also has the objective
significance that the immediate genus has determined itself through the
totality of singularity as a universal property and possesses its existence
in a universal relation or mark. – But the objective significance of this
syllogism, as it was of the others, is at first only its inner concept, and is
not as yet posited in it.
3. On the contrary, induction is essentially still a subjective syllogism.
The middle terms are the singulars in their immediacy, the collecting of
them into a genus through the allness is an external reflection. Because of
the persisting immediacy of the singulars and because of the externality that
derives from it, the universality is only completeness, or rather, it remains
a task. – In induction, therefore, there recurs the progression into the bad
infinity; singularity ought to be posited as identical with universality, but
since the singulars are equally posited as immediate, the intended unity
remains only a perpetual ought; it is a unity of likeness; the terms which are
supposed to be identical are at the same time supposed not to be identical.
The a, b, c, d, e, constitute the genus only further on, in the infinite;
they do not yield a complete experience. The conclusion of induction thus
remains problematic.
But induction, by expressing that perception, in order to become expe-
rience, ought to be carried on to infinity, presupposes that the genus is in
and for itself conjoined with its determinateness. In this, it in fact rather
presupposes its conclusion as something immediate, just as the syllogism
of allness presupposes the conclusion for one of its premises. – An experi-
ence that rests on induction is assumed as valid even though the perception
is admittedly not complete; it may be assumed, however, that there is no
counter-instance to the experience only if the latter is true in and for itself.
Inference by induction, therefore, is based indeed on an immediacy, but
not on the immediacy on which it is supposed to be based, not on a
singularity that exists immediately, but on one that exists in and for itself,
on the universal. – The fundamental character of induction is that it is a
syllogistic inference; if singularity is taken as the essential determination of
the middle term, but universality as only the external determination, then
the middle term would fall apart into two disjoined parts, and there would
be no inference; this externality belongs rather to the extremes. Singularity
can only be a middle term if immediately identical with the universality;
such a universality is in truth objective universality, the genus. – The matter
can also be viewed in this way: universality is external but essential to the
determination of the singularity which is at the basis of the middle term of
12.115614
Georg Wilhelm Friedrich Hegel
induction; such an external is just as much immediately its opposite, the
internal. – The truth of the syllogism of induction is therefore a syllogism
that has for its middle term a singularity which is immediately in itself
universality. This is the syllogism of analogy.
c. The syllogism of analogy
1. This syllogism has the third figure of the immediate syllogism, S-U-P, for
its abstract schema. But its middle term is no longer some single quality or
other but a universality which is the immanent reflection of a concreted term
and is therefore its nature; and conversely, since it is thus the universality of
a concreted term, it is at the same time in itself this concreted term. – Here,
therefore, a singular is the middle term, but a singular taken in its universal
nature; there is moreover another singular, an extreme term, which has the
same universal nature as the other which is the middle term. For example:
The earth has inhabitants,
The moon is an earth,
Therefore the moon has inhabitants. 95
12.116
2. Analogy is all the more superficial, the more the universal in which the
two extremes are united, and in accordance with which the one extreme
becomes the predicate of the other, is a mere quality or, since quality is a
matter of subjectivity, is some distinctive mark or other and the identity of
the extremes is therein taken as just a similarity. But this kind of superfi-
ciality to which a form of understanding or of reason is reduced by being
debased to the sphere of mere representation should have no place in logic. –
Also unacceptable is to present the major premise of this syllogism as though
it should run: “That which is similar to an object in one distinctive mark
is similar to it in other such marks as well.” On this formulation, the form
of the syllogism is expressed in the shape of a content while the empirical
content, the content properly so called, is together relegated to the minor
premise. So, for example, could also the whole form of the first syllogism
be expressed as its major premise: “That which is subsumed under another
thing in which a third thing inheres has that third thing inhering in it
too; but now . . . etc.” But what matters in the syllogism as such is not the
95
This example is not original with Hegel and very likely was a common trope at the time. One can
find it, for instance, in Antonio Genovesi, Elementa artis logico-criticæ, Liber V (Venice, 1749), p. 179,
and also, by the same author, Gli elementi dell’arte logico-critica, 2nd edn (Venice, 1783), p. 148.
In Genovesi’s treatise, syllogism of analogy follows immediately upon the treatment of probability,
exactly as it does in Hegel.The syllogism
615
empirical content, and to make its own form the content of a major premise
makes just as little difference as to take any other empirical content for that
purpose. Nothing of consequence follows for the syllogism of analogy from
a content that contains nothing but the form peculiar to that syllogism,
just as nothing of consequence would have followed for the first syllogism
from having as its content the form that makes the syllogism a syllogism. –
What counts is always the form of the syllogism, whether the latter has
itself or something else for its empirical content. So the syllogism of anal-
ogy is a form peculiarly its own, and it is vacuous not to want to regard it
as such on the ground that that form could be made into the content or
matter of a major premise whereas matter is no concern in logic. – What
might tempt one to this view in regard to the syllogism of analogy, and
perhaps in regard to the syllogism of induction too, is that the middle term
in them, and also the extremes, are more determined than they are in the
merely formal syllogism, and therefore the determinations of form, since
they are no longer simple and abstract, must also take on the appearance
of a content determination. But that the form determines itself to content is
first of all a necessary advance on the part of the formal side, and therefore
an advance that touches the nature of the syllogism essentially; secondly,
such a content determination cannot, therefore, be regarded as any other
empirical content, and abstraction cannot be made from it.
When we consider the syllogism of analogy with its major premise
expressed as above, namely, “if two subject matters agree in one or more
properties, then a further property of one also belongs to the other,” it may
seem that this syllogism contains four terms, the quaternio terminorum –
a circumstance that brings with it the difficulty of how to bring analogy
into the form of a formal syllogism. – There are two singulars; for a third,
a property immediately assumed as common, and, for a fourth, the other
properties that one singular possesses immediately but the other first comes
to possess only by means of the syllogism. – This is so because, as we have
seen, in the syllogism of analogy the middle term is posited as singularity but
immediately also as the true universality of the singularity. – In induction,
the middle term is, apart from the extremes, an indeterminate number of
singulars; this syllogism, therefore, required the enumeration of an infinite
number of terms. – In the syllogism of allness the universality in the
middle term is still only the external form determination of the allness;
in the syllogism of analogy, on the contrary, it is as essential universality.
In the above example, the middle term, “the earth,” is taken as something
concrete which, in truth, is just as much a universal nature or genus as it is
a singular.
12.117616
12.118
Georg Wilhelm Friedrich Hegel
From this aspect, the quaternio terminorum would not make analogy
an imperfect syllogism. But it would make it so from another aspect; for
although the one subject has the same universal nature as the other, it is
undetermined whether the determinateness, which is inferred to pertain
also to the second subject, pertains to the first because of its nature in
general or because of its particularity; for example, whether the earth has
inhabitants as a heavenly body in general or only as this particular heavenly
body. – Analogy is still a syllogism of reflection inasmuch as singularity
and universality are united in its middle term immediately. Because of this
immediacy, the externality of the unity of reflection is still there; the singular
is the genus only in itself, implicitly; it is not posited in this negativity by
which its determinateness would be the genus’s own determinateness. For
this reason the predicate that belongs to the singular of the middle term
is not already the predicate of the other singular, even though the two
singulars both belong to the one genus.
3. S-P (“the moon is inhabited”) is the conclusion; but the one premise
(“the earth is inhabited”) is likewise S-P; in so far as S-P is supposed to
be a conclusion, it entails the requirement that that premise also be S-P.
This syllogism is thus in itself the demand to counter the immediacy that it
contains; or again, it presupposes its conclusion. One syllogism of existence
has its presupposition in the other syllogism of existence. In the syllogisms
just considered, the presupposition has been moved into them, because
they are syllogisms of reflection. Since the syllogism of analogy is therefore
the demand that it be mediated as against the immediacy with which its
mediation is burdened, what it demands is the sublation of the moment of
singularity. Thus there remains for the middle term the objective universal,
the genus purified of immediacy. – In the syllogism of analogy the genus
was a moment of the middle term only as immediate presupposition; since
the syllogism itself demands the sublation of the presupposed immediacy,
the negation of singularity and hence the universal is no longer immediate
but posited. – The syllogism of reflection contained the first negation of
immediacy; the second has now come on the scene, and with it the exter-
nal universality of reflection is determined as existing in and for itself. –
Regarded from the positive side, the conclusion shows itself to be identical
with the premises, the mediation to have rejoined its presupposition, and
what we have is thus an identity of the universality of reflection by virtue
of which it becomes a higher universality.
Reviewing the course of the syllogism of reflection, we find that media-
tion is in general the posited or concrete unity of the form determinations of
the extremes; reflection consists in this positing of the one determinationThe syllogism
617
in the other; the mediating middle is thus allness. But it is singularity that
proves to be the essential ground of mediation while universality is only as
an external determination in it, as completeness. But universality is essential
to the singular if the latter is to be the conjoining middle term; the singular
is therefore to be taken as an implicitly existing universal. But the singular
is not united with it in just this positive manner but is sublated in it and
is a negative moment; thus the universal is the genus posited as existing in
and for itself, and the singular as immediate is rather the externality of the
genus, or it is an extreme. – The syllogism of reflection, taken in general,
comes under the schema P-S-U in which the singular is still as such the
essential determination of the middle term; but since its immediacy has
been sublated, the syllogism has entered under the formal schema S-U-P,
and the syllogism of reflection has thus passed over into the syllogism of
necessity.
c. the syllogism of necessity
The mediating middle has now determined itself (1) as simple determinate
universality, like the particularity in the syllogism of existence, but (2) as
objective universality, that is to say, one that, like the allness of the syllogism
of reflection, contains the whole determinateness of the different extremes;
this is a completed but simple universality, the universal nature of the fact,
the genus.
This syllogism is full of content, because the abstract middle term of
the syllogism of existence has posited itself to be determinate difference, in
the way it is as the middle term of the syllogism of reflection, but this
difference has again reflected itself into simple identity. – This syllogism
is for this reason the syllogism of necessity, because its middle term is not
any adventitious immediate content but is the immanent reflection of the
determinateness of the extremes. These have their inner identity in the
middle term, whose content determinations are the form determinations
of the extremes. – Consequently, what differentiates the terms is a form
which is external and unessential and the terms themselves are as moments
of a necessary existence.
This syllogism is at first immediate and formal in the sense that what
holds the terms together is the essential nature, as content, and this content
is in the distinguished terms only in different form, and the extremes are
by themselves only an unessential subsistence. – The realization of this
syllogism is a matter of determining it in such a way that the extremes are
equally posited as this totality which initially the middle term is, and the
12.119618
Georg Wilhelm Friedrich Hegel
necessity of the connection, which is at first only the substantial content,
shall be a connection of the posited form.
a. The categorical syllogism
12.120
1. The categorical syllogism has the categorical judgment for one or for
both of its premises. – Associated with this syllogism, just as with that
judgment, is the more specific signification that its middle term is the
objective universality. Superficially, the categorical syllogism is also taken
for nothing more than a mere syllogism of inherence.
Taken in its full import, the categorical syllogism is the first syllogism
of necessity, one in which a subject is conjoined with a predicate through
its substance. But when elevated to the sphere of the concept, substance
is the universal, so posited to be in and for itself that it has for its form
or mode of being, not accidentality, as it has in the relation specific to
it, but the determination of the concept. Its differences are therefore the
extremes of the syllogism, specifically universality and singularity. This
universality, as contrasted with the genus that more closely defines the
middle term, is abstract or is a universal determinateness: it is the acci-
dentality of substance summed up in a simple determinateness which
is, however, the substance’s essential difference, its specific difference. –
Singularity, for its part, is the actual, in itself the concrete unity of genus
and determinateness – though here, in the immediate syllogism, it is
immediate singularity at first, accidentality summed up in the form of
a subsistence existing for itself. – The connection of this extreme term
to the middle term constitutes a categorical judgment; but since the
other extreme term also, as just determined, expresses the specific dif-
ference of the genus or its determinate principle, this other premise is also
categorical.
2. This syllogism, as the first and therefore immediate syllogism of neces-
sity, comes in the first instance under the schema of the formal syllogism,
S-P-U. – But since the middle term is the essential nature of the singular
and not just one or other of its determinacies or properties, and likewise the
extreme of universality is not any abstract universal, nor just any singular
quality either, but is rather the universal determinateness of the genus, its
specific difference, we no longer have the contingency of a subject being con-
joined with just any quality through just any middle term. – Consequently,
since the connections of the extremes with the middle term also do not
have the external immediacy that they have in the syllogism of existence,
we do not have coming into play the demand for proof in the sense inThe syllogism
619
which it occurred in the case of that other syllogism and led to an infinite
progression.
Further, this syllogism does not presuppose its conclusion for its
premises, as in the syllogism of reflection. The terms, in keeping with
the substantial content, stand to one another in a connection of identity
that exists in and for itself; we have here one essence running through
the three terms – an essence in which the determinations of singularity,
particularity, and universality are only formal moments.
To this extent, therefore, the categorical syllogism is no longer subjective;
in that connection of identity, objectivity begins; the middle term is the
identity, full of content, of its extremes, and these are contained in it in their
self-subsistence, for their self-subsistence is the said substantial universality
which is the genus. The subjective element of the syllogism consists in the
indifferent subsistence of the extremes with respect to the concept or the
middle term.
3. But there is still a subjective element in this syllogism, for that identity
is still the substantial identity or content but is not yet identity of form at the
same time. The identity of the concept still is an inner bond and therefore,
as connection, still necessity; the universality of the middle term is solid,
positive identity, but is not equally the negativity of its extremes.
The immediacy of this syllogism, which is not yet posited as what it is in
itself, is more precisely present in this way. The truly immediate element
of the syllogism is the singular. This singular is subsumed under its genus
as middle term; but subsumed under the same genus are also an indeter-
minate number of many other singulars; it is therefore contingent that only
this singular is posited as subsumed under it. – But further, this contin-
gency does not belong only to an external reflection that finds the singular
posited in the syllogism to be contingent by comparison with others; on
the contrary, it is because the singular is itself connected to the middle
term as its objectivity universality that it is posited as contingent, as a sub-
jective actuality. From the other side, because the subject is an immediate
singular, it contains determinations that are not contained in the middle
term as the universal nature; it also has, therefore, a concrete existence
which is indifferent to the middle term, determined for itself and with a
content of its own. Therefore, conversely, this other term also has an indif-
ferent immediacy and a concrete existence distinct from the former. –
The same relation also obtains between the middle term and the
other extreme; for this too likewise has the determination of immedi-
acy, hence of a being which is contingent with respect to the middle
term.
12.121620
Georg Wilhelm Friedrich Hegel
Accordingly, what is posited in the categorical syllogism are, on the
one hand, extremes that are so related to the middle term that they have
objective universality or self-subsistent nature in themselves, and are at the
same time immediate actualities, hence indifferent to one another. On the
other hand, they are equally contingent, or their immediacy is as sublated in
their identity. But this identity, because of the self-subsistence and totality
of the actuality, is only formal, inner identity, and the syllogism of necessity
has thereby determined itself to the hypothetical syllogism.
b. The hypothetical syllogism
1. The hypothetical judgment contains only the necessary connection with-
out the immediacy of the connected terms. “If A is, so is B”; or, the being
of A is also just as much the being of an other, of the B; with this, it is not
as yet said either that A is, or that B is. The hypothetical syllogism adds
this immediacy of being:
If A is, so is B,
But A is,
Therefore B is.
12.122
The minor premise expresses by itself the immediate being of the A.
But it is not only this that is added to the judgment. The conclusion
contains the connection of subject and predicate, not as the abstract copula,
but as the accomplished mediating unity. The being of the A is to be
taken, therefore, not as mere immediacy but essentially as middle term of the
syllogism. This needs closer examination.
2. In the first place, the connection of the hypothetical judgment is
the necessity or the inner substantial identity associated with the external
diversity of concrete existence – an identical content lying internally as its
basis. The two sides of the judgment are both, therefore, not an immediate
being, but a being held in necessity, hence one which is at the same time
sublated or only being as appearance. The two behave, moreover, as sides
of the judgment, as universality and singularity; the one, therefore, is the
above content as totality of determinations, the other as actuality. Yet it is
a matter of indifference which side is taken as universality and which as
singularity. That is to say, inasmuch as the conditions are still the inner,
abstract element of an actuality, they are the universal, and it is by being
held together in one singularity that they step into actuality. Conversely, the
conditions are a dismembered and dispersed appearance that gains unity and
meaning, and a universally valid existence, only in actuality.The syllogism
621
The relation that is here being assumed between the two sides of con-
dition and conditioned may however also be taken to be one of cause and
effect, ground and consequence. This is a matter of indifference here. The
relation of condition, however, corresponds more closely to the one that
obtains in the hypothetical judgment and syllogism inasmuch as condi-
tion is essentially an indifferent concrete existence, whereas ground and
cause are inherently a transition; moreover, condition is a more universal
condition in that it comprehends both sides of the relation, since effect,
consequence, etc., are just as much the condition of cause and ground as
these are the condition of them. –
Now A is the mediating being in so far as it is, first, an immediate being,
an indifferent actuality, but, second, in so far as it is equally inherently contin-
gent, self-sublating being. What translates the conditions into the actuality
of the new shape of which they are the conditions is the fact that they
are not being as an abstract immediacy, but being according to its concept –
becoming in the first instance, but more determinedly (since the concept
is no longer transition) singularity as self-referring negative unity. – The
conditions are a dispersed material awaiting and requiring application; this
negativity is the mediating means, the free unity of the concept. It deter-
mines itself as activity, for this middle term is the contradiction of objective
universality, or of the totality of the identical content and the indifferent
immediacy. – This middle term is no longer, therefore, merely inner but
existent necessity; the objective universality contains its self-reference as sim-
ple immediacy, as being. In the categorical syllogism this moment is at first
a determination of the extremes; but as against the objective universality
of the middle term, it determines itself as contingency, hence as something
which is only posited and also sublated, something that has returned into
the concept or into the middle terms as unity, a unity which is now in its
objectivity also being.
The conclusion, “therefore B is,” expresses the same contradiction – that
B exists immediately but at the same time through an other or as mediated.
According to its form, it is therefore the same concept that the middle term
is, distinguished from necessity only as the necessary, in the totally superficial
form of singularity as contrasted with universality. The absolute content of
A and B is the same; for ordinary representation, they are two different
names for the same basic thing, since representation fixes the appearances
of the diversified shape of existence and distinguishes the necessary from
its necessity; but to the extent that necessity were to be separated from B,
the latter would not be the necessary. What we have here, therefore, is the
identity of the mediating term and the mediated.
12.123622
Georg Wilhelm Friedrich Hegel
3. The hypothetical syllogism is the first to display the necessary connection
as a connectedness through form or negative unity, just as the categorical
syllogism displays it through positive unity, the solid content, the objective
universality. But necessity merges with the necessary; the form-activity of trans-
lating the conditioning actuality into the conditioned is in itself the unity
into which the determinacies of the oppositions previously let free into
indifferent existence are sublated, and where the difference of A and B is an
empty name. The unity is therefore a unity reflected into itself, and hence an
identical content, and is this content not only implicitly in itself but, through
this syllogism, it is also posited, for the being of A is also not its own being
but that of B and vice versa, and in general the being of the one is the being
of the other and, as determined in the conclusion, their immediate being or
indifferent determinateness is a mediated one – therefore, their externality
has been sublated, and what is posited is their unity withdrawn into itself.
The mediation of the syllogism has thereby determined itself as singular-
ity, immediacy, and self-referring negativity, or as a differentiating identity
that retrieves itself into itself out of this differentiation – as absolute form,
and for that very reason as objective universality, self-identical existent
content. In this determination, the syllogism is the disjunctive syllogism.
c. The disjunctive syllogism
12.124
As the hypothetical syllogism comes in general under the schema of the
second figure of the formal syllogism, U-S-P, so the disjunctive comes
under the schema of the third, S-U-P. The middle term, however, is a
universality replete with form; it has determined itself as totality, as developed
objective universality. The middle term, therefore, is universality as well as
particularity and singularity. As that universality, it is in the first place the
substantial identity of the genus, but this identity is secondly one in which
particularity is included, but again, included as equal to it – therefore as a
universal sphere that contains its total particularity, the genus sorted out
in its species, an A which is B as well as C and D. But particularization is
differentiation and as such equally the either-or of B, C, D – negative unity,
the reciprocal exclusion of the determinations. – This excluding, moreover,
is now not just reciprocal, the determination not merely relative, but is also
just as much self-referring determination, the particular as singularity to the
exclusion of the others.
A is either B or C or D,
But A is B,
Therefore A is neither C nor D.The syllogism
623
Or also:
A is either B or C or D,
But A is neither C nor D,
Therefore A is B.
A is subject not only in the two premises but also in the conclusion. It
is a universal in the first premise and in its predicate the universal sphere
particularized in the totality of its species; in the second premise, it is as a
determinate, or as a species; in the conclusion it is posited as the excluding,
singular determinateness. – Or again, in the minor it is already exclusive
singularity, and in the conclusion it is positively posited as the determinate
that it is.
Consequently, what as such appears to be meditated is the universality of
A with the singularity. But the mediating means is this A which is the uni-
versal sphere of its particularizations and is determined as a singular. What
is posited in the disjunctive syllogism is thus the truth of the hypothetical
syllogism, the unity of the mediator and the mediated, and for that reason
the disjunctive syllogism is equally no longer a syllogism at all. For the
middle term which is posited in it as the totality of the concept itself con-
tains the two extremes in their complete determinateness. The extremes,
as distinct from this middle term, are only a positedness to which there no
longer accrues any proper determinateness of its own as against the middle
term.
If we consider the matter with narrower reference to the hypothetical
syllogism, we find that there was in the latter a substantial identity as the
inner bond of necessity, and a negative unity distinct from it, namely the
activity or the form that translated one existence into another. The disjunc-
tive syllogism is in general in the determination of universality, its middle
term is the A as genus and as perfectly determined; also posited through this
unity is the earlier inner content and, conversely, the positedness or the
form is not the external negative unity over against an indifferent existence
but is identical with that solid content. The whole form determination of
the concept is posited in its determinate difference and at the same time in
the simple identity of the concept.
In this way the formalism of the syllogistic inference, and consequently
the subjectivity of the syllogism and of the concept in general, has sub-
lated itself. This formal or subjective factor consisted in that the middle
mediating the extremes is the concept as an abstract determination and is
therefore distinct from the terms whose unity it is. In the completion of the
syllogism, where the objective universality is equally posited as the totality
12.125624
12.126
Georg Wilhelm Friedrich Hegel
of the form determinations, the distinction of mediating and mediated has
on the contrary fallen away. That which is mediated is itself an essential
moment of what mediates it, and each moment is the totality of what is
mediated.
The figures of the syllogism exhibit each determinateness of the con-
cept singly as the middle term, a middle term which is at the same time
the concept as an ought, the requirement that the mediating factor be the
concept’s totality. The different genera of the syllogism exhibit instead the
stages in the repletion or concretion of the middle term. In the formal syl-
logism the middle is posited as totality only through all the determinacies,
but each singly, discharging the function of mediation. In the syllogism of
reflection, the middle term is the unity gathering together externally the
determinations of the extremes. In the syllogism of necessity the middle
has determined itself as a unity which is just as developed and total as it is
simple, and the form of the syllogism, which consisted in the difference of
the middle term over against its extremes, has thereby sublated itself.
With this the concept in general has been realized; more precisely, it
has gained the kind of reality which is objectivity. The first reality was that
the concept, in itself negative unity, partitions itself and as judgment posits
its determinations in determinate and indifferent difference, and in the
syllogism it then sets itself over against them. Since it is still in this way the
inwardness of this now acquired externality, in the course of the syllogisms
this externality is equated with the inner unity; the different determinations
return into the latter through the mediation that unites them at first in
a third term, and as a result the externality exhibits, in itself, the concept
which, for its part, is no longer distinct from it as inner unity.
Conversely, however, that determinateness of the concept which was
considered as reality is equally a positedness. For the identity of the con-
cept’s inwardness and externality has been exhibited as the truth of the
concept not only in this result; on the contrary, already in the judgment
the moments of the concept remain, even in their reciprocal indifference,
determinations that have significance only in their connection. The syllo-
gism is mediation, the complete concept in its positedness. Its movement is
the sublation of this mediation in which nothing is in and for itself, but
each thing is only through the meditation of an other. The result is there-
fore an immediacy that has emerged through the sublation of the mediation,
a being which is equally identical with mediation and is the concept that
has restored itself out of, and in, its otherness. This being is therefore a fact
which is in and for itself – objectivity.