SECTION II Objectivity

In Book One of the Objective Logic, abstract being was presented as passing
over into existence, but at the same time as retreating into essence. In Book
Two, essence shows itself as determining itself as ground, thereby stepping
into concrete existence and realizing itself as substance, but at the same
time retreating into the concept. Of the concept, we have now first shown
that it determines itself as objectivity. It should be obvious that this latter
transition is essentially the same as the proof from the concept, that is to
say, from the concept of God to his existence, that was formerly found in
Metaphysics, or the so-called ontological proof. – Equally well known is that
Descartes’s sublimest thought, that God is that whose concept includes his
being within itself, 1 after having degenerated into the bad form of the formal
syllogism, namely into the form of the said proof, finally succumbed to
the Critique of Reason and to the thought that existence cannot be extracted
from the concept. 2 Some elucidations concerning this proof have already
been made earlier. In Volume I, pp. 47 ff., 3 where being has vanished into
its closest opposite, non-being, and becoming has shown itself to be the
truth of both, attention was called to the confusion that arises in the case
of a determinate existence when we concentrate, not on its being, but on
its determinate content, and then imagine – if we compare this determinate
content (e.g. one hundred dollars) with another determinate content (e.g.
the context of my perception, of my financial situation) and discover that
it makes indeed a difference whether the one content is added to the other
or not – that we are dealing with the distinction of being and non-being, or
even the distinction of being and the concept. Further, in the same Volume
Philosophical Writings of Descartes, Vol. 2, trans. John Cottingham,
Robert Stoothoff, and Dugald Murdoch (Cambridge: Cambridge University Press, 1984)
(this English edition also uses the pagination of Adam and Tannery).
on pp. 64ff. and on p. 289 of Volume II, the definition of a sum-total of
all reality which occurs in the ontological proof was elucidated. 4 – But
the essential subject matter of that proof, the connectedness of concept and
existence, is the concern of the treatment of the concept just concluded and of
the entire course that the latter traverses in determining itself to objectivity.
The concept, as absolutely self-identical negativity, is self-determining; it
was noted that the concept, in resolving itself into judgment in singularity,
already posits itself as something real, an existent; this still abstract reality
completes itself in objectivity.
Now it might appear that the transition from the concept into objectivity
is quite another thing than the transition from the concept of God to
God’s existence. But, on the one hand, it must be borne in mind that the
determinate content, God, makes no difference in a logical progression, and
that the ontological proof is only one application of this logical progression
to that particular content. On the other hand, it is essential to be reminded
of the remark made above 5 that the subject obtains determinateness and
content only in its predicate; that prior to the predicate, whatever that
content might otherwise be for feeling, intuition, and representation, so
far as conceptual cognition is concerned it is only a name; but in the
predicate, with determinateness, there begins at the same time the process
of realization in general. – The predicates, however, must be grasped as
themselves still confined within the concept, hence as something subjective
with which no move to existence has yet been made; even for this reason,
in judgment the realization of the concept is certainly not completed yet.
But there is the further reason that the mere determination of a subject
matter through predicates, without this determination being at the same
time the realization and objectification of the concept, remains something
so subjective that it is not even a true cognition and determination of the
concept of the subject matter – “subjective” in the sense of abstract reflection
and non-conceptual representation. – God as living God, and better still as
absolute spirit, is only recognized in what he does. Humankind were directed
early to recognize God in his works; only from these can the determinations
proceed that can be called his properties, and in which his being is also
contained. It is thus the conceptual comprehension of God’s activity, that
is to say, of God himself, that recognizes the concept of God in his being and
his being in his concept. Being by itself, or even existence, are such a poor and
restricted determination, that the difficulty of finding them in the concept
may well be due to not having considered what being or existence themselves
are. – Being as entirely abstract, immediate self-reference, is nothing but the
abstract moment of the concept; it is its moment of abstract universality
that also provides what is required of being, namely that it be outside the
concept, for inasmuch as universality is a moment of the concept, it is
also its difference or the abstract judgment wherein the concept opposes
itself to itself. The concept, even as formal, already immediately contains
being in a truer and richer form, in that, as self-referring negativity, it is
singularity.
But of course the difficulty of finding being in the concept in general, and
equally so in the concept of God, becomes insuperable if we expect being to
be something that we find in the context of external experience or in the form
of sense-perception, like the one hundred dollars in the context of my finances,
as something graspable only by hand, not by spirit, essentially visible to
the external and not the internal eye; in other words, if the name of being,
reality, truth, is given to that which things possess as sensuous, temporal,
and perishable. – The consequence of a philosophizing that in regard to
being fails to rise above the senses is that, in regard to the concept, it also
fails to let go of merely abstract thought; such thought stands opposed to
being.
The customary practice of regarding the concept as something just as
one-sided as abstract thought will already stand in the way of accepting what
has just been suggested, namely, that we regard the transition of the concept
of God to his being as an application of the logical course of objectification
of the concept presented above. Yet if it is granted, as it commonly is,
that the logical element, as the formal element, constitutes the form for
the cognition of every determinate content, then that application at least
would have to be conceded, unless even at the opposition of concept and
objectivity in general one stops short at the untrue concept and an equally
untrue reality as an ultimate. – But in the exposition of the pure concept it
was further indicated that the latter is the absolute divine concept itself. 6
In truth, therefore, what takes place is not a relation of application but
the immediate display in the logical course of God’s self-determination as
being. But on this point it is to be remarked that inasmuch as the concept
is to be presented as the concept of God, it ought be apprehended as it is
when already taken up in the idea. The said pure concept passes through
the finite forms of the judgment and the syllogism precisely because it is
not yet posited in and for itself as one with objectivity, but is conceived
rather only in the process of becoming that objectivity. The latter, too, is
not yet the divine concrete existence, not yet the reality reflectively shining
in the idea. And yet objectivity is just that much richer and higher than the
being or existence of the ontological proof, as the pure concept is richer and
higher than that metaphysical vacuum of the sum-total of all reality. – But
I reserve for another occasion the task of elucidating in greater detail the
manifold misunderstanding brought upon the ontological proof of God’s
existence, and also on the rest of the other so-called proofs, by logical
formalism. We shall also elucidate Kant’s critique of such proofs in order
to establish their true meaning and thus restore the thoughts on which
they are based to their worth and dignity. 7
We have previously called attention to the several forms of immediacy
that have already come on the scene, but in different determinations. 8 In
the sphere of being, immediacy is being itself and existence; in the sphere of
essence, it is concrete existence and then actuality and substantiality; in the
sphere of the concept, besides being immediacy as abstract universality, it is
now objectivity. – These expressions, when the exactitude of philosophical
conceptual distinctions is not at stake, may be used as synonymous; but
the determinations are derived from the necessity of the concept. Being
is as such the first immediacy, and existence is the same immediacy with
a first determinateness. Concrete existence, along with the thing, is the
immediacy that proceeds from ground, from the self-sublating mediation
of the simple reflection of essence. But actuality and substantiality are the
immediacy that proceeds from the sublated difference of the still unessential
concrete existence as appearance and its essentiality. Finally, objectivity is the
immediacy as which the concept has determined itself by the sublation of its
abstraction and mediation. – It is the privilege of philosophy to choose such
expressions from the language of ordinary life, which is made for the world
of imaginary representations, as seem to approximate the determinations
of the concept. There is no question of demonstrating for a word chosen
from ordinary life that in ordinary life too the same concept is associated
with that for which philosophy uses it, for ordinary life has no concepts,
only representations of the imagination, and to recognize the concept in
what is otherwise mere representation is philosophy itself. It must therefore
suffice if representation, for those of its expressions that philosophy uses
for its definitions, has only some rough approximation of their distinctive
difference; it may also be the case that in these expressions one recognizes
pictorial adumbrations which, as approximations, are close indeed to the
corresponding concepts. – One will be hard pressed, perhaps, to concede
that something can be without actually existing; but at least nobody will
mistake, for instance, being as the copula of the judgment for the expression
“to exist actually,” and nobody will say that “this article exists dear, suitable,
etc.,” “gold exists a metal or metallic,” instead of “this article is dear, suitable,
etc.,” “gold is a metal.” r And surely it is common to distinguish being from
appearing, appearance from actuality, as also being as contrasted to actuality,
and still more all these expressions from objectivity. – But even if such
expressions were used synonymously, philosophy would in any case have
the freedom to take advantage of such empty superfluity of language for
the purpose of its distinctions.
Mention was made in connection with the apodictic judgment – where
judgment attains completion and the subject thus loses its determinateness
as against the predicate – of the double meaning of subjectivity originating
from it, namely the subjectivity of the concept and equally so of the exter-
nality and contingency confronting the concept. 9 A similar objectivity also
appears for the double meaning, of standing opposed to the self-subsistent
concept yet of also existing in and for itself. 10 In the former sense, the object
stands opposed to the “I = I” which in subjective idealism is declared to be
the absolute truth. It is then the manifold world in its immediate existence
with which the “I” or the concept is engaged in endless struggle, in order,
by the negation of the inherently nullity of this other, to give to its first
certainty of being a self the actual truth of its equality with itself. – In a
broader sense, it means a subject matter in general for whatever interest or
activity of the subject.
In the opposite sense, however, the objective signifies that which exists
in and for itself, without restriction and opposition. Rational principles,
perfect works of art, etc., are said to be objective to the extent that they are
free and above every accidentality. Although rational principles, whether
theoretical or ethical, only belong to the sphere of the subjective, to con-
sciousness, this aspect of the latter of existing in and for itself is nonetheless
called objective; the cognition of truth is made to rest on the cognition of
the object as free of any addition by subjective reflection, and right conduct
on the adherence to objective laws, such as are not of subjective origin and
In a French report in which the officer in command states that before making for land he would
wait for the wind, which in the region of the island usually arose towards morning, we find the
expression: “le vent ayant été longtemps sans exister”; here the distinction simply arises from the
other common idiom as, for example, in “il a été longtemps sans m’écrire.”
are immune to arbitrariness and to treatment that would compromise their
necessity.
At the present standpoint of our treatise, objectivity has the meaning
first of all of the being in and for itself of the concept that has sublated
the mediation posited in its self-determination, raising it to immediate
self-reference. This immediacy is therefore itself immediately and entirely
pervaded by the concept, just as its totality is immediately identical with
its being. But further, since the concept equally has to restore the free
being-for-itself of its subjectivity, it enters with respect to objectivity into
a relation of purpose in which the immediacy of the objectivity becomes
a negative for it, something to be determined through its activity. This
immediacy thus acquires the other significance, namely that in and for
itself, in so far as it stands opposed to the concept, it is a nullity.
First, then, objectivity is in its immediacy. Its moments, on account of
the totality of all moments, stand in self-subsistent indifference as objects
each outside the other, and as so related they possess the subjective unity of
the concept only as inner or as outer. This is mechanism.
But, second, inasmuch as in mechanism that unity reveals itself to be
the immanent law of the objects, their relation becomes one of non-
indifference, each specifically different according to law; a connection in
which the objects’ determinate self-subsistence is sublated. This is chemism.
Third, this essential unity of the objects is thereby posited as distinct
from their self-subsistence. It is the subjective concept, but posited as
referring in and for itself to the objectivity, as purpose. This is teleology.
Since purpose is the concept posited as within it referring to objectivity,
and through itself sublating its defect of being subjective, the at first external
purposiveness becomes, through the realization of the purpose, internal. It
becomes idea.

CHAPTER 1 Mechanism

Since objectivity is the totality of the concept that has returned into its
unity, an immediate is thereby posited which is in and for itself that
totality, and is also posited as such, but in it the negativity of the concept
has as yet not detached itself from the immediacy of the totality; in other
words, the objectivity is not yet posited as judgment. In so far as it has
the concept immanent in it, the difference of the concept is present in it;
but on account of the objective totality, the differentiated moments are
complete and self-subsistent objects that, consequently, even in connection
relate to one another as each standing on its own, each maintaining itself
in every combination as external. – This is what constitutes the character
of mechanism, namely, that whatever the connection that obtains between
the things combined, the connection remains one that is alien to them,
that does not affect their nature, and even when a reflective semblance
of unity is associated with it, the connection remains nothing more than
composition, mixture, aggregate, etc. Spiritual mechanism, like its material
counterpart, also consists in the things connected in the spirit remaining
external to one another and to spirit. A mechanical mode of representation,
a mechanical memory, a habit, a mechanical mode of acting, mean that the
pervasive presence that is proper to spirit is lacking in what spirit grasps
or does. Although its theoretical or practical mechanism cannot take place
without its spontaneous activity, without an impulse and consciousness,
the freedom of individuality is still lacking in it, and since this freedom
does not appear in it, the mechanical act appears as a merely external
one.

A. THE MECHANICAL OBJECT

The object is, as we have seen, the syllogism whose mediation has attained
equilibrium and has therefore come to be immediate identity. It is therefore
in and for itself a universal – universality, not in the sense of a commonality
of properties, but a universality that pervades particularity and in it is
immediate singularity.
1. To begin with, therefore, the object does not differentiate itself into
matter and form, matter being its presumed self-subsistent universal aspect
and form the particular and singular instead; according to its concept,
any such abstract differentiation of singularity and universality has no place
in the object; if regarded as matter, the object must then be taken to be in
itself informed matter. One can just as well take it as a thing with properties,
as a whole consisting of parts, as substance with accidents, or as determined
by the other relations of reflection. But these are all past relations that in the
concept have come to an end. The object, therefore, has neither properties
nor accidents, for these are separable from the thing or the substance,
whereas in the object particularity is absolutely reflected into the totality.
In the parts of a whole, there is indeed present that self-subsistence that
pertains to the differences of the object, but these differences are at once
themselves essentially objects, totalities which, unlike parts, are not such
as against the whole.
At first, therefore, the object is indeterminate, for it has no determi-
nate opposition within, because it is the mediation that has collapsed into
immediate identity. Inasmuch as the concept is essentially determined, the
object has in it the determinateness of a manifold which, although com-
plete, is otherwise indeterminate, that is, relationless, one that constitutes
a totality also not further determined at first; sides or parts that may be
distinguished within it belong to an external reflection. This totally inde-
terminate difference thus amounts just to this, that there are several objects,
each of which only contains its determinateness reflected into its universal-
ity and does not reflectively shine outwardly. – Because this indeterminate
determinateness is essential to the object, the object is in itself a plurality,
and must therefore be regarded as a composite, an aggregate. – Yet it does not
consist of atoms, for atoms are not objects because they are not totalities.
Leibniz’s monad would be more of an object. It is a total representation
of the world which, shut up within its intensive subjectivity, in essence at
least is supposed to be a one. Yet the monad, determined as an exclusive
one, is a principle only assumed by reflection. It is an object, however, both
because the ground of its manifold representations – of the developed, that
is, the posited determinations of its merely implicit totality – lies outside
it, and because it is equally a matter of indifference for the monad that it
constitutes an object together with other objects; in fact, therefore, it is not
exclusive, not self-determined for itself.
2. Since the object is now a totality of determinateness, yet, because of
its indeterminateness and immediacy, it is not the totality’s negative unity,
it is indifferent towards the determinations as singulars, determined in and
for themselves, just as these are themselves indifferent to each other. These,
therefore, are not comprehensible from it nor from one another; the object’s
totality is the form of the overall reflectedness of its manifoldness into a
singularity in general which is not in itself determinate. The determinacies,
therefore, which are in it do indeed pertain to it; but the form that consti-
tutes their difference and combines them into a unity is an external one,
indifferent to them; whether it be a mixture, or again an order, a certain
arrangement of parts and sides, these are combinations that are indifferent
to what they connect.
Consequently, like an existence in general, the object has the determi-
nateness of its totality outside it, in other objects, and these again outside
them, and so forth to infinity. The immanent turning back of this pro-
gression in infinitum must indeed be likewise assumed, and it must be
represented as a totality, as a world, but one which is nothing but a univer-
sality brought to closure through a singularity that remains indeterminate,
a universe.
Since the object is thus determinate yet indifferent to its determinateness,
through itself it points for its determinateness outside and beyond itself,
constantly to objects for which it is however likewise a matter of indifference
that they do the determining. Consequently, nowhere is a principle of self-
determination to be found. Determinism, which is the standpoint that
cognition adopts when it assumes as truth the object as we first have it
here, assigns for each determination of the object that of another object;
but this other object is likewise indifferent both to its determinateness and
its determining. – For this reason determinism is itself so indeterminate as
to be bound to an infinite progression; it can halt at will anywhere, and be
satisfied there, because the object to which it has progressed, being a formal
totality, is shut up within itself and indifferent to its being determined by
another. For this reason to explain the determination of an object, and
to this end to extend the representation of it beyond it, is only an empty
word, for there is no self-determination in the other object to which the
explanation has been extended.
3. Now since the determinateness of an object lies in an other, there is
no determinate diversity separating the two; the determinateness is merely
doubled, once in the one object and then again in the other; it is something
utterly identical and the explanation or comprehension is, therefore, a
tautology. This tautology is an external back and forth movement; since
the determinateness fails to obtain from objects that are indifferent to it
any proper differentiation and is therefore only identical, there is only
one determinateness at hand, and that it should be doubled only expresses
precisely the externality and vacuity of a difference. But the objects are
at the same time self-subsistent in regard to one another; in that identity,
therefore, they remain utterly external. – Thus there arises the contradiction
of a perfect indifference of objects to one another and of an identity of
determinateness of such objects, or of the objects’ perfect externality in the
identity of their determinateness. This contradiction is consequently the
negative unity of a plurality of objects reciprocally repelling each other in
the unity. This is the mechanical process.

B. THE MECHANICAL PROCESS

If objects are regarded only as self-enclosed totalities, they cannot act on
one another. Regarded in this way, they are the same as the monads which,
precisely for that reason, were thought of as having no influence on each
other. But the concept of a monad is for just this reason a deficient reflec-
tion. For, in the first place, the monad is a determinate representation of its
only implicit totality; as a certain degree of development and positedness of its
representation of the world, it is determinate; but since it is a self-enclosed
totality, it is also indifferent to this determinateness and is, therefore, not its
own determinateness but a determinateness posited through another object.
In second place, it is an immediate in general, for it is supposed to be just a
mirroring; 11 its self-reference is therefore abstract universality and hence an
existence open to others. – It does not suffice, in order to gain the freedom of
substance, to represent the latter as a totality that, complete in itself, would
have nothing to receive from the outside. On the contrary, a self-reference
that grasps nothing conceptually but is only a mirroring 11 is precisely a
passivity towards the other. – Likewise the determinateness, whether we now
take it as the determinateness of a being that exists or that mirrors, 11 as a
degree of the monad’s own internally generated development, is something
external; the degree that the development achieves has its limit in an other.
To project the reciprocal influence of substances into a predetermined har-
mony means nothing more than to make it a presupposition, in effect to
remove it from the scope of the concept. – The need to avoid the interac-
tion of substances was founded on the moment of absolute self-subsistence
and originariness which was made a fundamental assumption. But since
the positedness, the degree of development, does not correspond to this
assumed in-itselfness, it has for this reason its ground in an other.
In connection with the relation of substantiality, we showed that it passes
over into the relation of causality. 12 But now the existent no longer has the
determination of a substance but that of an object; the causal relation has
come to an end in the concept; the originariness of one substance vis-à-vis
another has shown itself to be a reflective shine, the substance’s action a
transition into the opposite substance. This relation has therefore no objec-
tivity. Hence in so far as one object is posited in the form of subjective
unity, as efficient cause, this no longer counts as an originary determination
but as something mediated; the active object has this determination only
by means of another object. – Mechanism, since it belongs to the sphere of
the concept, has that posited within it which proved to be the truth of the
relation of causality, namely, that the cause which is supposed to be some-
thing existing in and for itself is in fact effect just as well, positedness. In
mechanism, therefore, the originary causality of the object is immediately a
non-originariness; the object is indifferent to this determination attributed
to it; that it is a cause is therefore something accidental to it. – To this
extent, it can be said that the causality of substances is only the product
of representation. But precisely this causality as product of representation
is what mechanism is; for mechanism is this, that causality, as identical
determinateness of a diversity of substances and hence as the foundering
into this identity of their self-subsistence, is mere positedness; the objects
are indifferent to this unity and maintain themselves in the face of it. But
this also, their indifferent self-subsistence, is a mere positedness, and for this
reason they are capable of mixing and aggregating, and as an aggregate of
becoming one object. Through this indifference both to their transition and
to their self-subsistence, the substances are objects.

A. THE FORMAL MECHANICAL PROCESS

The mechanical process is the positing of that which is contained in
the concept of mechanism, hence the positing in the first place of a
contradiction.
1. It follows from the just indicated concept that the interaction of objects
is the positing of their identical connection. This positing consists simply in
giving to the determinateness which is generated the form of universality –
and this is communication, which occurs without transition into the
opposite. – Spiritual communication, which however takes place in an
element of universality in the form of universality, is an idealized connec-
tion for itself, one in which a determinateness continues undisturbed from
one person to another, generalizing itself unaltered – like a scent freely
spreading in the unresisting atmosphere. But also in the communication
between material objects does their determinateness widen, so to speak,
in an equally idealizing manner; personality is an infinitely more intensive
hardness than objects possess. The formal totality of an object in general – a
totality indifferent to determinateness and hence not a self-determination –
renders the object indistinct from another object and thus makes interac-
tion at first an unimpeded continuing of the determinateness of the one
into the other.
Now in the region of the spirit there is an infinitely manifold content
capable of communication, for by being taken up into intelligence the
content receives this form of universality in which it becomes communica-
ble. But that which is a universal not only by virtue of form, but in and
for itself, is the objective as such, both in the region of the spirit and of
the body, and as against it the singularity of external objects, as of persons
also, is an unessential factor unable to offer any resistance to it. Laws,
morals, rational conceptions in general, are in the region of the spirit com-
municable entities of this kind; they pervade individuals unconsciously
imposing themselves on them. In the region of the body, such entities
are motion, heat, magnetism, electricity, and the like, all of which, even
when one wants to imagine them as stuffs or materials, must be termed as
imponderable agents, for they lack that aspect of materiality that grounds
its singularization.
2. Now although in the interaction of objects their identical universality
is posited first, it is equally necessary to posit the other moment of the
concept, that of particularity; the objects thus also demonstrate their self-
subsistence; they hold themselves outside each other, and in that universality
they produce singularity. This production is reaction in general. To begin
with, this reaction is not to be conceived of as a mere sublation of action
and of the communicated determinateness; what is communicated is as
universal positively present in the particular objects and particularizes itself
only in their diversity. To this extent, therefore, what is communicated
remains what it is, only distributed among the objects or determined by
their particularity. – The cause gets lost in its other, in the effect; the activity
of the causal substance in its action; but the active object only becomes a
universal; its action is from the start not a loss of its determinateness but
a particularization by virtue of which the object, which was at first that
whole determinateness present in it as single, now becomes a species of it, and
the determinateness is thereby posited for the first time as a universal. The
two – the raising in communication of the singular determinateness into
universality; and the particularization of it in distribution, the reduction
of what was only a one to a species – are one and the same.
Now reaction is equal to action. – First, this is manifested by the other
object taking over the entire universal; and so it is now active against the
first. Thus its reaction is the same as the action, a reciprocal repulsion of
the impulse. Second, what is communicated is the objective; it therefore
remains the substantial determination of the object on the presupposi-
tion of their diversity; the universal thus at the same time specifies itself
in them, and consequently each object does not simply give back the
whole action but possesses its specific share. But, third, reaction is a wholly
negative action in so far as each object, because of the elasticity of its self-
subsistence, repels within it the positedness of an other and retains its self-
reference. The specific particularity of the determinateness communicated
in the objects, what was before called species, returns to singularity, and
the object asserts its externality as against the communicated universality.
The action thereby passes over into rest. It proves to be only a superfi-
cial, transient alteration within the self-enclosed indifferent totality of the
object.
3. This return constitutes the product of the mechanical process. Imme-
diately, the object is presupposed as a singular; then as a particular as against
another particular; but finally as indifferent towards its particularity, as
universal. The product is the totality of the concept previously presupposed
but now posited. It is the conclusion in which the communicated universal
is united with singularity through the particularity of the object. In rest,
however, the mediation is posited at the same time as sublated; or again,
what is posited is that the product is indifferent to this determining of it
and that the received determinateness is external in it.
Accordingly the product is the same as the object that first enters the
process. But at the same time that object is first determined through this
movement; the mechanical object is, as such, an object only as product,
for what it is, is only by virtue of the mediation of an other in it. It is as
product that it thus is what it was supposed to be in and for itself, a
composite, a mixture, a certain arrangement of parts, in general such that its
determinateness is not self-determination but something posited.
Yet the result of the mechanical process is not already there ahead of that
process itself; its end is not in its beginning, as in the case of purpose. The
product is in the object a determinateness which is externally posited in
it. Hence this product is indeed according to its concept the same as what
the object already is at the beginning. But at the beginning the external
determinateness is not yet there as posited. The result is therefore something
quite other than the first existence of the object, and is something utterly
accidental for it.

B. THE REAL MECHANICAL PROCESS

The mechanical process passes over into rest. That is to say, the determi-
nateness that the object obtains through that process is only an external
one. Just as external to it is this rest, for although the latter is a determi-
nateness opposed to the activity of the object, the two are each indifferent
to the object. Rest can also be viewed, therefore, as brought about by an
external cause, just as much as it was indifferent to the object to be active.
Now further, since the determinateness is a posited one, and the concept
of the object has gone back to itself through the process of mediation, the
object contains the determinateness as one that is reflected into itself.
Hence in the mechanical process the objects and the process itself now
have a more closely determined relation. They are not merely diverse, but
are determinedly differentiated as against one another. Consequently the
result of the formal process, on the one hand a determinationless rest, is,
on the other hand, through the immanently reflected determinateness, the
distribution among several objects mechanically relating to one another of
the opposition which is in the object as such. The object that on the one
hand lacks all determination, showing no elasticity and no self-subsistence
in its relations, has, on the other hand, a self-subsistence impenetrable to
other objects. Objects now have also as against one another this more
determined opposition of the self-subsistent singularity and the non-self-
subsistent universality. – The precise difference between any two may be
had merely quantitatively as a difference in a body of diverse magnitudes of
mass, or of intensity, or in various other ways. But in general the difference
cannot be fixed at just this abstract level; also as objects, both are positively
self-subsistent.
Now the first moment of this real process is, as before, communication.
The weaker can be seized and invaded by the stronger only in so far as
it accepts the stronger and constitutes one sphere with it. Just as in the
material realm the weaker is secured against the disproportionately strong
(as a sheet hanging freely in the air is not penetrated by a musket ball; a
weak organic receptivity is not as vulnerable to strong stimuli as it is to
weak), so is the wholly feeble spirit safer facing the strong than one who
stands closer to the strong. Imagine, if you will, someone dull-witted and
ignoble; lofty intelligence will make no impression on such a one, nor
will nobility. The one single effective defense against reason is not to get
involved with it at all. – To the extent that an object that has no standing
of its own is unable to make contact with one which is self-subsistent, and
no communication can take place between them, the latter is also unable
to offer resistance, that is, cannot specify the communicated universal for
itself. – If they were not in the same sphere, their mutual connection would
be an infinite judgment and no process would be possible between them.
Resistance is the precise moment of the overpowering of the one object
by the other, for it is the initial moment in the distribution of the com-
municated universal and in the positing of the self-referring negativity, of
the singularity to be established. Resistance is overpowered when its deter-
minateness is not commensurate to the communicated universal which the
object has accepted and which is supposed to be singularized in the latter.
The object’s relative lack of self-subsistence is manifested in the fact that its
singularity lacks the capacity for what is communicated to it and is therefore
shattered by it, for it is unable to constitute itself as subject in this universal,
cannot make the latter its predicate. – Violence against an object is for the
latter something alien only according to this second aspect. Power becomes
violence when power, an objective universality, is identical with the nature
of the object, yet its determinateness or negativity is not the object’s own
immanent negative reflection according to which the object is a singular.
In so far as the negativity of the object is not reflected back into itself in
the power, and the latter is not the object’s own self-reference, the negativ-
ity, as against the power, is only abstract negativity whose manifestation is
extinction.
Power, as objective universality and as violence against the object is what is
called fate – a concept that falls within mechanism in so far as fate is called
blind, that is, its objective universality is not recognized by the subject in
its own specific sphere. – To add a few more remarks on the subject, the
fate of a living thing is in general the genus, for the genus manifests itself
through the fleetingness of the living individuals that do not possess it as
genus in their actual singularity. Merely animate natures, as mere objects,
like other things at lower levels on the scale of being, do not have fate.
What befalls them is a contingency; however, in their concept as objects
they are self-external; hence the alien power of fate is simply and solely
their own immediate nature, externality and contingency itself. Only self-
consciousness has fate in a strict sense, because it is free, and therefore in the
singularity of its “I” it absolutely exists in and for itself and can oppose itself
to its objective universality and alienate itself from it. By this separation,
however, it excites against itself the mechanical relation of a fate. Hence,
for the latter to have violent power over it, it must have given itself some
determinateness or other over against the essential universality; it must
have committed a deed. Self-consciousness has thereby made itself into a
particular, and this existence, like abstract universality, is at the same time
the side open to the communication of its alienated essence; it is from this
side that it is drawn into the process. A people without deeds is without
blame; it is wrapped up in objective, ethical universality, is dissolved into
it, is without the individuality that moves the unmoved, that gives itself
a determinateness on the outside and an abstract universality separated
from the objective universality; yet in this individuality the subject is also
divested of its essence, becomes an object and enters into the relation of
externality towards its nature, into that of mechanism.

C. THE PRODUCT OF THE MECHANICAL PROCESS

The product of formal mechanism is the object in general, an indifferent
totality in which determinateness is as posited. The object has hereby entered
the process as a determinate thing, and, in the extinction of this process,
the result is, on the one hand, rest, the original formalism of the object, the
negativity of its determinateness-for-itself. But, on the other hand, it is the
sublation of the determinateness, the positive reflection of it into itself, the
determinateness that has withdrawn into itself, or the posited totality of the
concept, the true singularity of the object. The object, determined at first in
its indeterminate universality, then as particular, is now determined as an
objective singular, so that in it that reflective semblance of singularity, which
is only a self-subsistence opposing itself to the substantial universality, is
sublated.
This resulting immanent reflection, the objective oneness of the objects,
is now a oneness which is an individual self-subsistence – the center.
Secondly, the reflection of negativity is the universality which is not a
fate standing over against determinateness, but a rational fate, immanently
determined – a universality that particularizes itself from within, the differ-
ence that remains at rest and fixed in the unstable particularity of the objects
and their process; it is the law. This result is the truth, and consequently
also the foundation, of the mechanical process.

C. ABSOLUTE MECHANISM

A. THE CENTER

The empty manifoldness of the object is now gathered first into objec-
tive singularity, into the simple self-determining middle point. Secondly,
in so far as the object retains as an immediate totality its indifference
to determinateness, the latter too is present in it as unessential or as an
outside-one-another of many objects. As against this immediate totality,
the prior or the essential determinateness constitutes the real middle term
between the many interacting objects; it unites them in and for them-
selves and is their objective universality. Universality exhibited itself first in
the relation of communication, as present only through positing; as objec-
tive universality, however, it is the pervading immanent essence of the
objects.
In the material world it is the central body which is the genus or rather
the individualized universality of the single objects and their mechanical
process. The unessential single bodies relate to one another by impact
and pressure; this kind of relation does not hold between the central body
and the objects of which it is the essence; for their externality no longer
constitutes their fundamental determination. Hence their identity with the
central body is rather rest, namely the being at their center; this unity is their
concept existing in and for itself. It nevertheless remains only an ought, since
the objects’ externality, still posited at the same time, does not conform
to that unity. The striving which the objects consequently have towards
the center is their absolute universality, one which is not posited through
communication; it constitutes the true rest, itself concrete and not posited from
the outside, into which the process of instability must find its way back. – It
is for this reason an empty abstraction to assume in mechanics that a body
set in motion would go on moving in a straight line to infinity if it did not
lose movement because of external resistance. Friction, or whatever other
form resistance takes, is only a phenomenon of centrality; it is the latter
that in principle brings the body back to itself, since that against which
the body rubs and incurs friction has its power of resistance only because
it is united with the center. – In things spiritual the center, and the union
with it, assume higher forms; but the unity of the concept and the reality
of that unity, which is here in a first instance mechanical centrality, must
there too constitute the fundamental determination.
The central body has therefore ceased to be a mere object, for in the latter
the determinateness is something unessential, whereas now the central body
no longer has only being-in-itself, in-itselfness, but also has the being-for-itself
of the objective totality. For this reason it can be regarded as an individual.
Its determinateness is essentially different from a mere order or arrangement
and external combination of diverse parts; as a determinateness that exists
in and for itself it is an immanent form, a self-determining principle to
which the objects inhere and in virtue of which they are bound together
in a true One.
But this central individual is at first only a middle term that as yet has no
true extremes; as the negative unity of the total concept it dirempts itself
rather into such extremes. Or again: the previously non-self-subsistent,
self-external objects become likewise determined as individuals by the
retreat of the concept; the self-identity of the central body, still a striving,
is burdened by an externality to which, in being taken up into the body’s
objective singularity, the latter is communicated. The objects, through this
centrality of their own, are positioned outside the original center and are
themselves centers for the non-self-subsistent objects. These second centers
and the non-self-subsistent objects are brought into unity by the absolute
middle term.
But the relative individual centers themselves also constitute the middle
term of a second syllogism. This middle term is on the one hand subsumed
under a higher extreme, the objective universality and power of the absolute
center; on the other hand, it subsumes under it the non-self-subsistent
objects whose superficiality and formal singularization it supports. – These
non-self-subsistent objects are in turn the middle term of a third syllogism,
the formal syllogism, for since the central individuality obtains through
them the externality by virtue of which, in referring to itself, is also strives
towards an absolute middle point, those non-self-subsistent objects are the
link between absolute and relative central individuality. The formal objects
have for their essence the identical gravity of their immediate central body
in which they inhere as in their subject and the extreme of singularity;
through the externality which they constitute, this immediate central body
is subsumed under the absolute central body; they therefore are the formal
middle term of particularity. – But the absolute individual is the objectively
universal middle term that brings into unity and holds firm the inwardness
of the relative individual and its externality. – Similarly, the government, the
individual citizens, and the needs or the external life of these, are also three
terms, of which each is the middle term of the other two. The government is
the absolute center in which the extreme of the singulars is united with their
external existence; the singulars are likewise the middle term that incites
that universal individual into external concrete existence and transposes
their ethical essence into the extreme of actuality. The third syllogism is
the formal syllogism, the syllogism of reflective shine in which the singular
citizens are tied by their needs and external existence to this universal
absolute individuality; this is a syllogism that, as merely subjective, passes
over into the others and has its truth in them.
This totality, whose moments are themselves the completed relations of
the concept, the syllogisms in which each of the three different objects runs
through the determination of the middle term and the extreme, consti-
tutes free mechanism. In it the different objects have objective universality
for their fundamental determination, the pervasive gravity that persists
self-identical in the particularization. The connections of pressure, impact,
attraction, and the like, as also of aggregations or mixtures, belong to the
relation of externality which is at the basis of the third of the three syllo-
gisms. Order, which is the merely external determinateness of the objects,
has passed over into immanent and objective determination. This is the
law.

B. THE LAW

In law, the more specific difference of the idealized reality of objectivity
versus the external reality comes into view. The object, as the immediate
totality of the concept, does not yet possess an externality differentiated
from the concept, and the latter is not posited for itself. Now that through
the mediation of the process the object has withdrawn into itself, there
has arisen the opposition of simple centrality as against an externality now
determined as externality, that is, one posited as not existing in and for itself.
That moment of identity or idealization of individuality is, on account
of the reference to externality, an ought; it is the unity of the concept,
determined in-and-for-itself and self-determining, to which that external
reality does not correspond, and therefore does not go past the mere striving
towards it. But individuality is, in and for itself, the concrete principle of
negative unity, and as such is itself totality; it is a unity that dirempts itself
into the specific differences of the concept while abiding within its self-equal
universality; it is thus the central point expanded inside its pure ideality by
difference. – This reality that corresponds to the concept is the idealized
reality, distinct from the reality that is only a striving; it is difference, earlier
a plurality of objects but now in its essential nature, and taken up into pure
universality. This real ideality is the soul of the hitherto developed objective
totality, the identity of the system which is now determined in and for itself.
The objective being-in-and-for-itself thus manifests itself more precisely
in its totality as the negative unity of the center, a unity that divides into
subjective individuality and external objectivity, maintains the former in the
latter and determines it in an idealized difference. This self-determining
unity that absolutely reduces external objectivity to ideality is a principle
of self-movement; the determinateness of this animating principle, which
is the difference of the concept itself, is the law. – Dead mechanism
Georg Wilhelm Friedrich Hegel
appeared as self-subsisting, but precisely for that reason are in truth non-
self-subsistent and have their center outside them; this process that passes
over into rest exhibits either contingency and indeterminate difference or
formal uniformity. This uniformity is indeed a rule, but not law. Only
free mechanism has a law, the determination proper to pure individuality
or to the concept existing for itself. As difference, the law is in itself the
inexhaustible source of a self-igniting fire and, since in the ideality of its
difference it refers only to itself, it is free necessity.

C. TRANSITION OF MECHANISM

This soul is however still immersed in its body. The now determined but
inner concept of objective totality is free necessity in the sense that the
law has not yet stepped in opposite its object; it is concrete centrality as
a universality immediately diffused in its objectivity. Such an ideality does
not have, therefore, the objects themselves for its determinate difference;
these are self-subsistent individuals of the totality, or also, if we look back
at the formal stage, non-individual, external objects. The law is indeed
immanent in them and it does constitute their nature and power; but
its difference is shut up in its ideality and the objects are not themselves
differentiated in the idealized non-indifference of the law. But the object
possesses its essential self-subsistence solely in the idealized centrality and its
laws; it has no power, therefore, to put up resistance to the judgment of the
concept and to maintain itself in abstract, indeterminate self-subsistence
and remoteness. Because of the idealized difference which is immanent in
it, its existence is a determinateness posited by the concept. Its lack of self-
subsistence is thus no longer just a striving towards a middle point, with
respect to which, precisely because its connection with it is only that of a
striving, it still has the appearance of a self-subsistent external object; it is
rather a striving towards the object determinedly opposed to it; and likewise
the center has itself for that reason fallen apart and its negativity has passed
over into objectified opposition. Centrality, therefore, is now the reciprocally
negative and tense connection of these objectivities. Thus free mechanism
determines itself to chemism.

CHAPTER 2 Chemism

In objectivity as a whole chemism constitutes the moment of judgment, of
the difference that has become objective, and of process. Since it already
begins with determinateness and positedness, and the chemical object is
at the same time objective totality, the course it follows next is simple and
perfectly determined by its presupposition.

A. THE CHEMICAL OBJECT

The chemical object is distinguished from the mechanical in that the latter
is a totality indifferent to determinateness, whereas in the chemical object
the determinateness, and hence the reference to other, and the mode and
manner of this reference, belong to its nature. – This determinateness is
at the same time essentially a particularization, that is, it is taken up into
universality; thus it is a principle – a determinateness which is universal, not
only the determinateness of the one singular object but also of the other.
In the chemical object there is now, therefore, a distinction in its concept,
between the inner totality of the two determinacies and the determinateness
that constitutes the nature of the singular object in its externality and
concrete existence. Since in this way the object is implicitly the whole concept,
it has within it the necessity and the impulse to sublate its opposed, one-
sided subsistence, and to bring itself in existence to the real whole which it
is according to its concept.
Regarding the expression “chemism” for the said relation of the non-
indifference of objectivity, it may be further remarked that the expression
is not to be understood here as though the relation were only to be found
in that form of elemental nature that strictly goes by that name. Already
the meteorological relation must be regarded as a process whose parts have
more the nature of physical than chemical elements. 13 In animate things,
the sex relation falls under this schema, and the schema also constitutes the
formal basis for the spiritual relations of love, friendship, and the like.
On closer examination, the chemical object is at first a self-subsistent
totality in general, one reflected into itself and therefore distinct from
its reflectedness outwards – an indifferent basis, the individual not yet
determined as non-indifferent; the person, too, is in the first instance a basis
of this kind, one that refers only to itself. But the immanent determinateness
that constitutes the object’s non-indifference is, first, reflected into itself in
such a manner that this retraction of the reference outwards is only a formal
abstract universality; the outwards reference is thus a determination of the
object’s immediacy and concrete existence. From this side the object does
not return, within it, to individual totality: the negative unity has its two
moments of opposition in two particular objects. Accordingly, a chemical
object is not comprehensible from itself, and the being of one object is the
being of another. – But, second, the determinateness is absolutely reflected
into itself and is the concrete moment of the individual concept of the
whole which is the universal essence, the real genus of the particular objects.
The chemical object, which is thus the contradiction of its immediate
positedness and its immanent individual concept, is a striving to sublate the
immediate determinateness of its existence and to give concrete existence
to the objective totality of the concept. Hence it does still remain a non-
self-subsistent object, but in such a way that it is by nature in tension with
this lack of self-subsistence and initiates the process as a self-determining.

B. THE CHEMICAL PROCESS

1. It begins with the presupposition that the objects in tension, as much as
they are tensed against themselves, just as much are they by that very fact
at first tensed against each other – a relation which is called their affinity.
Each stands through its concept in contradiction to its concrete existence’s
own one-sidedness and each consequently strives to sublate it, and in this
there is immediately posited the striving to sublate the one-sidedness of
the other and, through this reciprocal balancing and combining, to posit a
reality conformable to the concept that contains both moments.
Since each is posited within it as self-contradictory and self-sublating,
they are held apart from each other and from their reciprocal complementa-
tion only by external violent force. The middle term whereby these extremes
are now concluded into a unity is, first, the implicitly existent nature of both,
the whole concept containing both within. But, second, since in concrete
existence the two stand over against each other, their absolute unity is also a
still formal element that concretely exists distinct from them – the element
of communication wherein they enter into external community with each
other. Since the real difference belongs to the extremes, this middle term
is only the abstract neutrality, the real possibility of those extremes – the
theoretical element, as it were, of the concrete existence of the chemical
objects, of their process and its result. In the realm of bodies, water fulfills
the function of this medium; in that of spirit, inasmuch as there is in it an
analog of such a relation, the sign in general, and language more specifically,
can be regarded as fulfilling it.
The relation of the objects, as mere communication in this element, is
on the one hand a tranquil coming-together, but on the other it is equally
a negative relating, for in communication the concrete concept which is
their nature is posited in reality, and the real differences of the object are
thereby reduced to its unity. Their prior self-subsistent determinateness
is thus sublated in the union that conforms to the concept, which is one
and the same in both; their opposition and tension are thereby blunted,
with the result that in this reciprocal complementation the striving attains
its tranquil neutrality.
The process is in this way dissolved; since the contradiction between
concept and reality has been resolved, the extremes of the syllogism have
consequently lost their opposition and have ceased to be extremes as against
each other and the middle term. The product is something neutral, that
is, something in which the ingredients, which can no longer be called
objects, are no longer in tension and therefore no longer have the properties
that accrued to them in tension, though in the product the capacity for
their prior self-subsistence and tension is retained. For the negative unity
of the neutral product proceeds from a presupposed non-indifference; the
determinateness of the chemical object is identical with its objectivity; it
is original. Through the process just considered, this non-indifference is
only immediately sublated; the determinateness, therefore, is not as yet
absolutely reflected into itself, and consequently the product of the process
is only a formal unity.
2. In this product the tension of opposition, and the negative unity
which is the activity of the process, are now indeed dissolved. But since
this unity is essential to the concept and has also itself come into concrete
existence, it is still present but has stepped outside the neutral object. The
process does not spontaneously re-start itself, for it had non-indifference
only as its presupposition – it did not posit it. – This self-subsistent negativity
outside the object, the concrete existence of the abstract singularity whose
being-for-itself has its reality in the non-indifferent object, is in itself now in
tension with its abstraction, an inherently restless activity outwardly bent
on consuming. It connects immediately with the object whose tranquil
neutrality is the real possibility of an opposition to this neutrality; the same
object is now the middle term of the prior formal neutrality, now concrete
in itself and determined.
The more precise immediate connection of the extreme of negative unity
with the object is in that the latter is determined by it and is thereby
disrupted. This disruption may at first be regarded as the restoration of
the opposition of the objects in tension with which chemism began. But
this determination does not constitute the other extreme of the syllogism
but belongs to the immediate connection of the differentiating principle
with the middle in which this principle gives itself its immediate reality;
it is the determinateness which the middle term, besides at the same time
being the universal nature of the subject matter, possesses in the disjunctive
syllogism, whereby that object is both objective universality and determi-
nate particularity. The other extreme of the syllogism stands opposed to the
external self-subsistent extreme of singularity; it is, therefore, the equally self-
subsisting extreme of universality; hence the disruption that the real neu-
trality of the middle term undergoes in it is that it breaks up into moments
that are not non-indifferent but, on the contrary, neutral. 14 Accordingly
these moments are, on the one side, the abstract and indifferent 15 base,
and, on the other, this base’s activating principle which, separated from it,
equally attains the form of indifferent objectivity.
This disjunctive syllogism is the totality of chemism in which the same
objective whole is exhibited as self-standing negative unity; then, in the
middle term, as real unity; and finally as the chemical reality resolved
into its abstract moments. In these moments the determinateness has
not reached its immanent reflection in an other as in the neutral prod-
uct, but has in itself returned into its abstraction, an originally determined
element.
3. These elemental objects are therefore liberated from chemical tension;
in them, the original basis of that presupposition with which chemism began
has been posited through the real process. Now further, their inner deter-
minateness is as such essentially the contradiction of their simple indifferent
subsistence and themselves as determinateness, and is the outward impulse
that disrupts itself and posits tension in its determined object and in an
other, in order that the object may have something to which it can relate as
non-indifferent, with which it can neutralize itself and give to its simple
determinateness an existent reality. Consequently, on the one hand
chemism has gone back to its beginning in which objects in a state of
reciprocal tension seek one another and then combine in a neutral product
by means of a formal and external middle term; and, on the other hand,
by thus going back to its concept, chemism sublates itself and has gone over
into a higher sphere.

C. TRANSITION OF CHEMISM

Even ordinary chemistry shows examples of chemical alterations in which
a body, for example, imparts a higher oxidation to one part of its mass and
thereby reduces another part to a lower degree of the same, at which degree
alone it can enter into a neutral combination with another differing body
brought into contact with it, a combination to which it would not have
been receptive at that other first immediate degree. What happens here
is that the object does not connect with another in accordance with an
immediate, one-sided determinateness, but, in accordance with the inner
totality of an original relation, posits the presupposition which it needs for a
real connection and thereby gives itself a middle term by virtue of which it
unites its concept with its reality in conclusion; it is a singularity determined
in and for itself, the concrete concept as the principle of the disjunction
into extremes whose re-union is the activity of that same negative principle
that thereby returns to its first determination, but returns to it objectified.
Chemism is itself the first negation of the indifferent objectivity and
of the externality of determinateness; it is still burdened, therefore, by the
immediate self-subsistence of the object and with externality. Consequently
it is not yet for itself that totality of self-determination that proceeds from it
and in which it is rather sublated. – The three syllogisms that have resulted
constitute its totality. The first has formal neutrality for its middle term and
for extremes the objects in tension. The second has for its middle term the
product of the first, real neutrality; and for extremes the disrupting activity
and its product, the indifferent element. But the third is the self-realizing
concept that posits for itself the presupposition by virtue of which the
process of its realization is conditioned – a syllogism that has the universal
for its essence. Yet, on account of the immediacy and externality by which
the chemical objectivity is still determined, these three syllogisms fall apart.
The first process whose product is the neutrality of the tensed objects is
extinguished in this product and is re-activated only by a differentiation
that comes to it from outside; conditioned by an immediate presupposition,
the process is exhausted in it. – The excretion out of the neutral product
of the non-indifferent extremes, as also their decomposition into their
abstract elements, must likewise proceed from conditions and stimulations
of activity brought in from the outside. But the two essential moments of
the process, neutralization on the one hand and dissolution and reduction
on the other, since they too are bound together in one and the same process
and the union blunting the tension of the extremes is also a separation into
these, constitute on account of the still underlying externality two diverse
sides; the extremes that are separated in that same process are other than the
objects or matters uniting in it; in so far as the former proceed from it again
as non-indifferent, they must turn outwards; their renewed neutralization
is a process other than the one that took place in the first.
But these various processes, which have demonstrated themselves to be
necessary, are equally so many stages by which externality and conditionality
are sublated, and from which the concept emerges as determined in and
for itself, a totality unconditioned by externality. In the first process, what
is sublated is the externality of the mutually non-indifferent extremes that
constitute the whole reality, or the distinction between the implicitly deter-
minate concept and its existing determinateness. Sublated in the second
process is the externality of the real unity, union as merely neutral. Or
more precisely, the formal activity sublates itself in bases that are equally
formal, neutral 16 determinacies whose inner concept is now the absolute
activity that has withdrawn into itself and now realizes itself internally, that
is, posits the determinate difference within itself and through this media-
tion constitutes itself as real unity; this is a mediation which is thus the
concept’s own mediation, its self-determination and, considering that in it
the concept reflects itself back into itself, an immanent presupposing. The
third syllogism, which on the one hand is the restoration of the preceding
processes, sublates on the other hand the last remaining moment of indif-
ferent bases: it sublates the whole abstract external immediacy that becomes
in this way the concept’s own moment of self-mediation. The concept that
has thus sublated as external all the moments of its objective existence, and
has posited them in its simple unity, is thereby completely liberated from
the objective externality to which it refers only as an unessential reality.
This objective free concept is purpose.

CHAPTER 3 Teleology

Where there is the perception of a purposiveness, an intelligence is assumed
as its author; required for purpose is thus the concept’s own free concrete
existence. Teleology is above all contrasted with mechanism, in which the
determinateness posited in the object, being external, is one that gives
no sign of self-determination. The opposition between causæ efficientes and
causæ finales, between merely efficient and final causes, refers to this distinc-
tion, just as, at a more concrete level, the enquiry whether the absolute
essence of the world is to be conceived as blind mechanism or as an intelli-
gence that determines itself in accordance with purposes also comes down
to it. The antinomy of fatalism, along with determinism, and freedom is
equally concerned with the opposition of mechanism and teleology; for
the free is the concept in its concrete existence.
Earlier metaphysics has dealt with these concepts as it dealt with others.
It presupposed a certain picture of the world and strived to show that
one or the other concept of causality was adequate to it, and the opposite
defective because not explainable from the presupposed picture, all the while
not examining the concept of mechanical cause and that of purpose to see
which possesses truth in and for itself. If this is established independently, it
may turn out that the objective world exhibits mechanical and final causes;
its actual existence is not the norm of what is true, but what is true is rather
the criterion for deciding which of these concrete existences is its true one.
Just as the subjective understanding exhibits also errors in it, so the objective
world exhibits also aspects and stages of truth that by themselves are still
one-sided, incomplete, and only relations of appearances. If mechanism
and purposiveness stand opposed to each other, then by that very fact they
cannot be taken as indifferent concepts, as if each were by itself a correct
concept and had as much validity as the other, the only question being
where the one or the other may apply. This equal validity of the two rests
only on the fact that they are, that is to say, that we have them both. But
since they do stand opposed, the necessary first question is, which of the
two concepts is the true one; and the higher and truly telling question is,
whether there is a third which is their truth, or whether one of them is the
truth of the other. – But purposive connection has proved to be the truth of
mechanism. – Regarding chemism, what came under it can be taken together
with mechanism, for purpose is the concept in free concrete existence, and
the concept’s state of unfreedom, its being sunk into externality, stands
opposed to it in any form. Both, mechanism as well as chemism, are
therefore included under natural necessity: mechanism, because in it the
concept does not exist in the object concretely, for as mechanical the latter
lacks self-determination; chemism, either because the concept has in it a
one-sided concrete existence in a state of tension, or because, emerging
as the unity that creates in the neutral object a tension of extremes, it is
external to itself in so far as it sublates this divide.
The closer the teleological principle is associated with the concept of an
extra-mundane intelligence, and the more it has therefore enjoyed the favor
of piety, all the more it has seemed to depart from the true investigation
of nature, which aims at a cognition of the properties of nature not as
extraneous, but as immanent determinacies, and accepts only such cognition
as a valid conceptual comprehension. Since purpose is the concept itself
in its concrete existence, it may seem strange that a cognition of objects
based on their concept rather appears as an unjustified trespass into a
heterogeneous element, whereas mechanism, for which the determinateness
of an object is posited in it externally and by an other, is accepted as a more
immanent view of things than teleology. Of course mechanism, at least the
ordinary unfree mechanism, and chemism as well, must be regarded as an
immanent principle in so far as the externally determining object is itself
again just another such object, externally determined and indifferent to its
being determined, or, in the case of chemism, in so far as the other object
must likewise be one that is chemically determined; in general, in so far
as an essential moment of the totality always lies in something external.
These principles remain confined, therefore, within the same natural form
of finitude; but although they do not wish to transcend the finite and,
as regards appearances, lead only to finite causes that themselves demand
further causes, they nonetheless equally expand themselves, partly into a
formal totality in the concept of force, cause, or of such determinations of
reflection that are supposed to signify originariness, and partly, through the
medium of abstract universality, also into a sum total of forces, a whole of
reciprocal causes. Mechanism thus reveals itself to be a striving for totality
by the very fact that it seeks to comprehend nature by itself as a whole that
has no need of an other for its concept – a totality that is not found in
purpose and the extra-mundane intelligence associated with it.
Now purposiveness presents itself from the first as something of a gen-
erally higher nature, as an intelligence that externally determines the man-
ifoldness of objects through a unity that exists in and for itself, so that the
indifferent determinacies of the objects become essential by virtue of this
connection. In mechanism they become so through the mere form of neces-
sity that leaves their content indifferent, for they are supposed to remain
external and only the understanding as such is expected to find satisfaction
by recognizing its principle of union, the abstract identity. In teleology,
on the contrary, the content becomes important, for teleology presupposes
a concept, something determined in and for itself and consequently self-
determining, and has therefore extracted from the connection of differences
and their reciprocal determinateness, from the form, a unity that is reflected
into itself, something that is determined in and for itself and is consequently
a content. But if this content is otherwise finite and insignificant, then it
contradicts what it is supposed to be, for according to its form purpose is
a totality infinite within itself – especially when the activity operating in
accordance with it is assumed to be an absolute will and intelligence. For
this reason has teleology drawn the reproach of triviality so much upon
itself, for the purposes that it has espoused are, as the case may be, more
important or more trivial [than the content], and it was inevitable that
the connection of purposiveness in objects would so often appear just a
frivolity, since it appears external and therefore contingent. Mechanism,
on the contrary, leaves to the determinacies of the objects, as regards their
content, their status as accidents indifferent to the object, and these deter-
minacies are not supposed to have, whether for the objects or the subjective
understanding, any value higher than that. This principle, combined with
external necessity, yields therefore a consciousness of infinite freedom that
contrasts with teleology, which sets up as something absolute bits of its
content that are trivial and even contemptible, where the more universal
thought can only find itself infinitely constricted, even to the point of
feeling disgust.
The formal disadvantage from which this teleology immediately suffers
is that it only goes as far as external purposiveness. The content of concept,
since the latter is thereby posited as something formal, is for teleology also
externally given to it in the manifoldness of the objective world – in those
very determinacies that are also the content of mechanism, but are there as
something external and accidental. Because of this commonality of content,
only the form of purposiveness constitutes by itself the essential element of
the teleological. In this respect, without as yet considering the distinction
between external and internal purposiveness, the connection of purpose in
general has proven itself to be the truth of mechanism. – Teleology possesses
in general the higher principle, the concept in its concrete existence, which
is in and for itself the infinite and absolute – a principle of freedom which,
utterly certain of its self-determination, is absolutely withdrawn from the
external determining of mechanism.
One of Kant’s greatest services to philosophy was in drawing the distinc-
tion between relative or external purposiveness and internal purposiveness;
in the latter he opened up the concept of life, the idea, and with that he
positively raised philosophy above the determinations of reflection and the
relative world of metaphysics, something that the Critique of Reason does
only imperfectly, ambiguously, and only negatively. – We have remarked
that the opposition of teleology and mechanism is first of all the general
opposition of freedom and necessity. 17 Kant treated the opposition in this
form, among the antinomies of reason, namely, as the third conflict of the
transcendental ideas. 18 – I cite his exposition, to which reference was made
earlier, 19 very briefly because its essential point is so simple that it does
not need extensive explanation – and moreover, the peculiarities of Kant’s
antinomies have been elucidated in greater detail elsewhere.
The thesis of the antinomy now in question runs thus: Causality accord-
ing to the laws of nature is not the only one from which the appearance of
the world can exhaustively be derived. For their explanation, it is necessary
to assume yet another causality through freedom.
The antithesis: There is no freedom, but everything in the world happens
solely according to laws of nature.
As in the other antinomies, the proof starts off apagogically by assuming
the opposite of each thesis; secondly, in order to show the contradiction of
this assumption, its opposite – which is then the proposition to be proved –
is assumed in turn and presupposed as valid. This whole roundabout proof
could therefore be spared, for the proof consists in nothing but the assertoric
assertion of the two opposite propositions.
Thus to prove the thesis we should first assume that there is no other
causality than that according to the laws of nature, that is, according to
the necessity of mechanism in general, chemism being included. This
proposition contradicts itself, because the law of nature consists just in
this, that nothing happens without a cause sufficiently determined a priori, a
cause that would have to contain an absolute spontaneity within it, that is,
the assumption opposed to the thesis is contradictory, for the reason that
it contradicts the thesis.
In support of the proof of the antithesis we should assume that there is a
freedom as a particular kind of causality for absolutely initiating a situation,
and together with it also a series of consequences following upon it. But
now, since such a beginning presupposes a situation that has no causal link
with the one preceding it, it contradicts the law of causality that alone
makes the unity of experience and experience in general possible – that is,
the assumption of the freedom that is opposed to the antithesis cannot be
made, for the reason that it contradicts the antithesis.
We find in essence the same antinomy in the Critique of the Teleological
Judgement as the opposition between the proposition that every generation
of material things happens according to merely mechanical laws, and the
proposition that some cases of generation of material things are not possible
according to such laws. 20 – Kant’s resolution of this antinomy is the same
as the general resolution of the rest, namely that reason cannot prove
either the one or the other proposition because we cannot have a priori
any determining principle of the possibility of things according to merely
empirical laws of nature; further, that therefore the two propositions must
be regarded not as objective propositions but as subjective maxims; that
I ought to reflect on the events of nature every time according to the
principle of the mechanism of nature alone, but that this does not prevent,
when occasion permits, following up certain natural forms in accordance
with another maxim, namely in accordance with the principal of final
causes – as if now these two maxims, which moreover are supposed to be
necessary only for human reason, did not stand in the same opposition as
the two propositions in antinomy. – Missing in all this, as we remarked
above, 21 is the one thing that alone is of philosophical interest, namely
the investigation of which of the two principles has truth in and for
itself. On this standpoint, it makes no difference whether the principles
should be regarded as objective, which means here, as externally existing
determinations of nature, or as mere maxims of a subjective cognition; what
is subjective here is rather the contingent cognition that applies one or the
other maxim as occasion demands, indeed, according to whether it deems
them fitting for given objects, but for the rest does not ask about the truth
of these determinations themselves, whether they both are determinations
of the objects or of cognition.
However unsatisfactory is for this reason Kant’s discussion of the teleo-
logical principle with respect to its essential viewpoint, still worthy of note
is the place that Kant assigns to it. Since he ascribes it to a reflective faculty of
judgment, he makes it into a mediating link between the universal of reason
and the singular of intuition; 22 further, he distinguishes this reflective judg-
ment from the determining judgment, the latter one that merely subsumes
the particular under the universal. 23 Such a universal that only subsumes
is an abstraction that becomes concrete only in an other, in the particular.
Purpose, on the contrary, is the concrete universal containing within itself
the moment of particularity and of externality; it is therefore active and
the impulse to repel itself from itself. The concept, as purpose, is of course
an objective judgment in which one determination, the subject, namely the
concrete concept, is self-determined, while the other is not only a predicate
but external objectivity. But for that reason the connection of purpose is
not a reflective judgment that considers external objects only according to
a unity, as though an intelligence had given them to us for the convenience
of our faculty of cognition; on the contrary, it is the truth that exists in
and for itself and judges objectively, determining the external objectivity
absolutely. The connection of purpose is therefore more than judgment; it
is the syllogism of the self-subsistent free concept that through objectivity
unites itself with itself in conclusion.
Purpose has resulted as the third to mechanism and chemism; it is their
truth. Inasmuch as it still stands inside the sphere of objectivity or of the
immediacy of the total concept, it is still affected by externality as such and
has an objective world over against it to which it refers. From this side,
mechanical causality, to which chemism is also in general to be added,
still makes its appearance in this purposive connection which is the external
one, but as subordinated to it and as sublated in and for itself. As regards
the more precise relation, the mechanical object is, as immediate totality,
indifferent to its being determined and consequently, conversely, to its
being a determinant. This external determinateness has now progressed to
self-determination and accordingly the concept that in the object was only
inner or, which amounts to the same, only outer, is now posited; purpose
is, in the first instance, precisely this concept which is external to the
mechanical object. And so for chemism also, purpose is the self-determining
which brings the external determinateness conditioning it back to the unity
of the concept. – We have here the nature of the subordination of the two
preceding forms of the objective process. The other, which in those forms
lies in the infinite progress, is the concept posited at first as external to them,
and this is purpose; not only is the concept their substance but externality
is for them also an essential moment constituting their determinateness.
Thus mechanical or chemical technique, because of its character of being
externally determined, naturally offers itself to the connection of purpose,
which we must now examine more closely.

A. THE SUBJECTIVE PURPOSE

In the centrality of the objective sphere, which is an indifference to deter-
minateness, the subjective concept has first rediscovered and posited the
negative point of unity, and in chemism it has first rediscovered and posited
the objectivity of the determinations of the concept by which it is first posited
as concrete objective concept. Its determinateness or its simple difference now
has the determinateness of externality within it, and its simple unity is there-
fore the unity that repels itself from itself and in this repelling maintains
itself. Purpose, therefore, is the subjective concept as an essential striving
and impulse to posit itself externally. In this, it is exempt from transition.
It is neither a force expressing itself, nor a substance or a cause manifesting
itself in its accidents or effects. To the extent that force has not expressed
itself, it is only an abstract inner; or again, it first has existence in an exter-
nalization to which it has to be solicited. The same applies to cause and to
substance. Since they have actuality only in the accidents and in the effects,
their activity is a transition against which they do not maintain themselves
in freedom. Purpose can of course also be defined as a force or a cause, but
these expressions cover only an incomplete side of its signification; if they
are to be said of purpose according to its truth, this can be done only in a
way that sublates their concept – as a cause that solicits itself to expression,
or a cause that is a cause of itself or whose effect is immediately the cause.
When purposiveness is attributed to an intelligence, as was said above, 24
this is done with specific reference to a certain content. But, as such,
purpose is to be taken as the rational in its concrete existence. It manifests
rationality by being the concrete concept that holds the objective difference
in its absolute unity. Within, therefore, it is essentially syllogism. It is the
self-equal universal; more precisely, inasmuch as it contains self-repelling
negativity, it is universal though at first still indeterminate activity. But since
this activity is negative self-reference, it determines itself immediately and
gives itself the moment of particularity, and this particularity, as likewise
the totality of the form reflected into itself, is content as against the posited
differences of the form. The same negativity, through its self-reference, is
just as immediately the reflection of the form into itself and singularity.
From the one side, this reflection is the inner universality of the subject;
from the other side, however, it is outwards reflection; and to this extent
purpose is still something subjective, its activity still directed to an external
objectivity.
For purpose is the concept that has come to itself in objectivity; the
determinateness that it has given itself there is that of objective indifference
and externality of determinateness; its self-repelling negativity is therefore
one whose moments, being only determinations of the concept itself, also
have the form of objective indifference to one another. – Already in the
formal judgment are subject and predicate determined as self-subsistent over
against each other; but their self-subsistence is still only abstract universality.
It has now attained the determination of objectivity, but, as a moment of
the concept, this complete difference is enclosed within the simple unity
of the concept. Now in so far as purpose is this total reflection of objectivity
into itself and is such immediately, in the first place, the self-determination
or the particularity as simple reflection into itself is distinguished from the
concrete form, and is a determinate content. Accordingly, purpose is finite,
even though according to form it is equally infinite subjectivity. Secondly,
since its determinateness has the form of objective indifference, it has the
shape of a presupposition, and from this side its finitude consists in its
having before it an objective, mechanical and chemical world to which
its activity is directed as to something already there; its self-determining
activity is in its identity thus immediately external to itself, reflection into
itself just as much as reflection outwards. To this extent purpose still has
a truly extra-mundane concrete existence – to the extent, namely, that this
objectivity stands opposed to it, just as the latter, as a mechanical and
chemical whole still not determined and not pervaded by purpose, stands
on its side opposed to it.
Consequently, the movement of purpose can now be expressed as being
directed at sublating its presupposition, that is, the immediacy of the object,
and at positing it as determined by the concept. This negative relating to
the object is equally a negative attitude towards itself, a sublating of the
subjectivity of purpose. Positively, this is the realization of purpose, namely
the unification of the objective being with it, so that this being, which as a
moment of purpose is immediately the determinateness identical with it,
shall be as external determinateness, and conversely the objective, as presup-
position, shall be posited rather as determined by the concept. – Purpose659
Teleology
is in it the impulse to its realization; the determinateness of the moments of
the concept is externality; the simplicity of these moments within the unity
of the concept is however incommensurable with what this unity is, and the
concept therefore repels itself from itself. This repulsion is in general the
resolution 25 of the self-reference of the negative unity by virtue of which
the latter is exclusive 26 singularity; but by this excluding 27 the unity resolves
itself, that is to say, it discloses itself, 28 for it is self-determination, the posit-
ing of itself. On the one hand, in determining itself, subjectivity makes
itself into particularity, gives itself a content which, enclosed within the
unity of the concept, is still an inner content; but this positing, the simple
reflection into itself, is, as we have seen, 29 at the same time immediately
a presupposing; and at the same moment in which the subject of purpose
determines itself, it is referred to an indifferent, external objectivity which
is to be made equal by it with the determinateness of that inner content,
that is to say, posited as something determined by the concept – first of all as
meansu

B. THE MEANS

The first immediate positing in purpose is equally the positing of some-
thing internally determined, that is, determined as posited, and, at the
same time, the presupposing of an objective world, one indifferent to the
determination of purpose. But the subjectivity of purpose is the absolutely
negative unity; its second determining is, therefore, the sublation of this pre-
supposition as such; this sublation is an immanent turning back inasmuch
as that moment of the first negation which is the positing of the negative
over against the subject, the external object, is sublated by it. But as against
the presupposition or the immediacy of the determining, as against the
objective world, it is as yet only the first, itself immediate and hence exter-
nal negation. This positing is therefore not yet the realized purpose itself
but only the beginning of this realization. The object so determined is now
the means.
Through a means the purpose unites with objectivity and in objectivity
unites with itself. This means is the middle term of the syllogism. Purpose is
in need of a means for its realization, because it is finite – in need of a means,
that is to say, of a middle term that has at the same time the shape of an
external existence indifferent towards the purpose itself and its realization.
The absolute concept has mediation within itself in such a manner that the
first positing of it is not a presupposition in whose object the fundamental
determination would be an indifferent externality; on the contrary, the
world as creation has only the form of such an externality; it is its negativity
and the positedness that rather constitute its fundamental determination. –
Accordingly, the finitude of purpose consists in this, that its determining
is as such external to itself; accordingly, its first determining, as we have
seen, 30 falls apart into a supposing and a presupposing; the negation of
this determining is therefore only according to one side already immanent
reflection; according to the other side, it is rather only first negation. Or
again, the immanent reflection is itself also self-external and a reflection
outwards.
The means is therefore the formal middle term of a formal syllogism;
it is something external to the extreme of the subjective purpose as also,
therefore, to the extreme of the objective purpose; just as particularity in
the formal syllogism is an indifferent medius terminus that can be replaced
by others. Moreover, just as this particularity is a middle term by virtue
of being determinateness with reference to one extreme but universality
with reference to the other extreme, and therefore obtains its mediating
determination by being related to an other, so too the means is a mediating
middle term only because it is, first, an immediate object, and, second,
because it is a means by virtue of a reference connecting it with the extreme
of purpose external to it – a reference which is for it a form to which it is
indifferent.
Concept and objectivity, therefore, are in the means only externally
linked; hence the means is only a merely mechanical object. The reference
of the object to purpose is a premise or the immediate reference which,
as we have seen, 31 is with respect to purpose an immanent reflection; the
means is an inhering predicate; its objectivity is subsumed under the deter-
mination of purpose which, on account of its concreteness, is universality.
Through this purposive determination present in it, the means is now
also subsumptive with respect to the other extreme, the at the moment
still indeterminate objectivity. – Conversely, as contrasted with the sub-
jective purpose, the means has as immediate objectivity a universality of
existence which the subjective singularity of purpose still misses. – Thus,
since purpose is in the means as only an external determinateness at first, it
is itself, as the negative unity, outside the means; the means, for its part, is
a mechanical object that possesses purpose only as a determinateness, not
as the simple concretion of totality. But as the unifying means, the middle
term must itself be the totality of the purpose. It has been shown that the
determination of purpose is in the middle term at the same time immanent
reflection; 32 as this reflection, it is a formal self-reference, since the determi-
nateness is posited as real indifference, as the objectivity of the middle term.
But precisely for this reason this subjectivity, which is in one respect pure
subjectivity, is at the same time also activity. – In the subjective purpose the
negative self-reference is still identical with determinateness as such, with
the content and the externality. However, in the initial objectification of
purpose which is a becoming-other of the simple concept, those moments
come apart, each outside the other, or, conversely, the becoming-other or
the externality itself consists in this coming apart.
This whole middle term is thus the totality of the syllogism in which
the abstract activity and the external means constitute the extremes, while
the determinateness of the object through the purpose, by virtue of which
it is a means, constitutes the middle term. – But further, universality is the
connection of purposiveness and the means. This means is object, in itself
the totality of the concept; it does not have with respect to purpose any
of the power of resistance that it initially has against another immediate
object. To the purpose, therefore, which is the posited concept, it is utterly
penetrable, and it is receptive to this communication because it is in itself
identical with it. But it is now also posited that it is penetrable by the
concept, for in centrality it is an object striving towards negative unity;
in chemism, too, whether as neutral or non-indifferent, it is no longer
self-subsistent. – Its non-self-subsistence consists precisely in its being the
totality of the concept only implicitly; but the concept is being-for-itself.
Consequently, with respect to purpose the object has the character of being
powerless and of serving it; purpose is the subjectivity or soul of the object
that has in the latter its external side.
The object, immediately subjected to purpose in this way, is not an
extreme of the syllogism; on the contrary, this connection between the two
constitutes a premise of it. But the means has also one side from which
it still has self-subsistence as against the purpose. The objectivity which
in the means is bound with the purpose is still external to it, because
it is only immediately so connected; and therefore the presupposition still
persists. The activity of the purpose through the means is for that reason
still directed against this presupposition, and the purpose is activity, no
longer mere impulse and striving, because in the means the moment of
objectivity is posited in its determinateness as something external, and the
simple unity of the concept now has this objectivity as such within it.

C. THE REALIZED PURPOSE

1. Purpose is in its connection to the means already reflected into itself,
but its objective immanent turning back is not yet posited. The activity of
purpose through its means is still directed against objectivity as an initial
presupposition; this is precisely what that activity is, to be indifferent to
determinateness. If it were again to consist in determining the immediate
objectivity, the product would again be only a means, and so forth into
infinity; only a purposeful means would result, but not the objectivity of
the purpose itself. In being active in its means, therefore, purpose must not
determine the immediate object as something external to it, and the object,
accordingly, must merge with it in the unity of the concept through itself;
or again, the otherwise external activity of purpose through its means must
determine itself as mediation and thus sublate itself as external.
The connection of the activity of purpose with the external object
through the means is first of all the second premise of the syllogism –
an immediate connection of the middle term with the other extreme. It is
immediate because the middle term has within it an external object and
the other extreme is likewise an external object. The means is effective and
potent against this latter object because its own is linked with the self-
determining activity, whereas for the other the immediate determinateness
that it possesses is an indifferent one. Their process in this connection is
none other than the mechanical or chemical one; the previous relations
come up again in this objective externality, but under the dominance of
purpose. – But these processes, as they themselves showed, return into
purpose on their own. 34 If, therefore, the connection of the means to the
external object which it has to work upon is at first an immediate one, that
connection has earlier exhibited itself already as a syllogism, for purpose
proved to be their true middle term and unity. Since the means is therefore
the object that stands on the side of purpose and has the latter’s activity
within it, the mechanism that occurs here is at the same time the turning
back of objectivity into itself, into the concept which, however, is already
presupposed as purpose; the negative attitude of the purposeful activity
towards the object is therefore not an external attitude but, on the contrary,
the objectivity’s own alteration and internal transition into it.
That the purpose immediately refers to an object and makes it into
a means, as also that through this means it determines another object,
may be regarded as violence inasmuch as purpose appears of an entirely
different nature than the object, and the two objects are in like matter
mutually independent totalities. But that the purpose posits itself in a
mediate connection with the object, and between itself and this object inserts
another object, may be regarded as the cunning of reason. As remarked,
the finitude of rationality has this side, that purpose relates to the object
as a presupposition, that is, as external. In an immediate connection with
that object, purpose would itself enter into the sphere of mechanism and
chemism and would therefore be subject to accidentality and to the loss
of its determining vocation 35 to be the concept that exists in and for itself.
But in this way, by sending an object as a means ahead of it, it lets it do
the slavish work of externality in its stead, abandons it to the wear and tear
while preserving itself behind it against mechanical violence.
Since it is finite, the purpose further has a finite content; accordingly, it
is not rational absolutely, or simply in and for itself. But the means is the
external middle term of the syllogism which is the realization of purpose;
in the means, therefore, the rationality in the purpose manifests itself as
such by maintaining itself in this external other, and precisely through this
externality. To this extent the means is higher than the finite purposes of
external purposiveness: the plough is more honorable than are immediately
the enjoyments which it procures and which are the purposes. The tool
lasts while the immediate enjoyments pass away and are forgotten. It is
in their tools that human beings possess power over external nature, even
though with respect to their purposes they are subjected to it.
But the purpose does not just keep outside the mechanical process; on
the contrary, it keeps itself in it and is its determination. For purpose –
as the concept that concretely exists freely over against the object and its
process, and is self-determining activity – since it is equally the truth of
mechanism existing in and for itself, in the latter only rejoins itself. The
power of purpose over the object is this identity existing for itself, and
its activity is the manifestation of this identity. The purpose as content is
the determinateness as it exists in and for itself, present in the object as
indifferent and external; but the activity of the purpose is the truth of the
process on the one side, and, as negative unity, the sublation of the reflective
shine of externality. The indifferent determinateness of the object is one
that can abstractly be replaced by another just as externally; but the truth of
the simple abstraction of the determinateness is the totality of the negative,
the concrete concept that posits the externality within itself.
The content of the purpose is its negativity as simple determinateness
reflected into itself, distinguished from its totality as form. On account of
this simplicity, the determinateness of which is in and for itself the totality
of the concept, the content appears as that which remains identical in the
realization of the purpose. The teleological process is the translation of
the concept that concretely exists distinctly as concept into objectivity;
as we see, this translation into a presupposed other is the rejoining of
the concept through itself with itself. The content of the purpose is now
this identity concretely existing in the form of the identical. In every
transition the concept maintains itself; for instance, when the cause comes
to effect, it is the cause that in the effect only comes to itself. But in
the teleological transition, what maintains itself is the concept that as
such already concretely exists as cause, as the free concrete unity as against
objectivity and its external determinateness. The externality into which
the purpose translates itself is, as we have seen, 36 itself already posited as
a moment of the concept, as the form of its inner differentiation. In the
externality, therefore, the purpose has its own moment; and the content, as
the content of the concrete unity, is its simple form that does not remain
in its different moments only implicitly equal with itself (as subjective
purpose, as means and mediating activity, and as objective purpose) but
also exists concretely as abidingly self-equal.
Of the teleological activity one can say, therefore, that in it the end
is the beginning, the consequence the ground, the effect the cause; that
it is a becoming of what has become; that in it only that which already
concretely exists comes into existence, and so on; that is to say, that quite
in general all the relation determinations that belong to the sphere of
reflection or of immediate being have lost their distinction, and what, like
end, consequence, effect, and so on, is spoken of as an other, no longer has
in purpose this determination of other, but is rather posited as identical
with the simple concept.
2. If we now examine the product of teleological activity more closely,
we see that purpose comes to it only externally if we take it as an absolute
presupposition over against a purpose which is subjective, that is to say,
in so far as we stop short at a purposive activity that relates to the object
through its means only mechanically, positing in place of one indifferent
determinateness of the object an other which is just as external to it. A
determinateness such as an object possesses through purpose differs in
general from one which is merely mechanical in that it is a moment of
a unity and consequently, although external to the object, is yet not in
itself something merely external. The object that exhibits such a unity is a
whole with respect to which its parts, its own externality, are indifferent;
it is a determinate, concrete unity that unites different connections and
determinacies within itself. This unity, which cannot be comprehended
from the specific nature of the object and, as regards determinate content,
is of another content than the object’s own, is for itself not a mechanical
determinateness, yet still is in the object mechanically. Just as in this product
of purposive activity the content of the purpose and the content of the
object are external to each other, so too do the determinations in the other
moments of the syllogism relate to each other externally – in the connecting
middle, the purposive activity and the object which is the means; and in the
subjective purpose, which is the other extreme, the infinite form as totality
of the concept and the content of the concept. According to the connection
by which the subjective purpose is syllogistically united with objectivity,
both premises are an immediate connection – namely the connection of
the object determined as middle term with the still external object, and the
connection of the subjective purpose with the object made into means. The
syllogism is therefore affected by the deficiency of the formal syllogism in
general, namely that the connections in which it consists are not themselves
conclusions or mediations but already presuppose the conclusion for the
production of which they are supposed to serve as means.
If we consider the one premise, that of the immediate connection of the
subjective purpose and the object that thereby becomes a means, then the
purpose cannot connect with the object immediately, for the latter is just
as immediate as the object of the other extreme in which the purpose is
to be realized through mediation. Since the two are thus posited as diverse,
a means for their connection must be interjected between this objectivity
and the subjective purpose; but such a means is equally an object already
determined by purpose, and between this objectivity and the teleological
determination a new means is to be interjected, and so on to infinity. The
infinite progress of mediation is thereby set in motion. – The same happens
as regards the other premise, the connection of the means with the yet
indeterminate object. Since the two terms are utterly self-subsistent, they
can be united only in a third, and so on to infinity. – Or conversely,
since the premises already presuppose the conclusion, the latter can only be
imperfect, for it is based on those only immediate premises. The conclusion
or the product of the purposive activity is nothing but an object determined
by a purpose that is external to it; thus it is the same as what the means is.
In such a product itself, therefore, only a means has been derived, not a
realized purpose; or again: purpose has not truly attained any objectivity in
it. – It is therefore entirely a matter of indifference whether we consider
an object determined by external purpose as realized purpose or only as
means; what we have is not an objective determination but a relative one,
external to the object itself. All objects in which an external purpose is
realized equally are, therefore, only a means of purpose. Anything which
is intended for the realization of a purpose and is taken essentially as a
means, is such a means by virtue of its vocation 37 that it be used up. But
also the object that is supposed to contain the realized purpose and show
itself to be its objectivity is perishable; it likewise fulfills its purpose not by
a tranquil, self-preserving existence, but only to the extent that it is worn
out, for only to this extent does it conform to the unity of the concept,
namely in so far as its externality, that is, its objectivity, sublates itself in
that unity. – A house, a clock, may appear as purposes with respect to
the instruments employed in their production; but the stones, the cross-
beams, or the wheels, the axles, and the rest that make up the actuality
of the purpose, fulfill this purpose only through the pressure which they
suffer, through the chemical processes to which they are exposed with air,
light, and water, and from which they shield the human being; through
their friction, and so on. They fulfill their vocation, therefore, only through
their being used up and worn out, and only by virtue of their negation do
they correspond to what they are supposed to be. They are not united with
purpose positively, because they possess self-determination only externally
and are only relative purposes, or essentially only means.
These purposes thus in general have a restricted content; their form is
the infinite self-determination of the concept, which through that content
has restricted itself to external singularity. The restricted content renders
these purposes inadequate to the infinity of the concept, relegating them to
untruth; such a determinateness is through the sphere of necessity, through
being, already at the mercy of becoming and alteration and passes away.
3. The result now is that external purposiveness, which only has so far
the form of teleology, only goes so far as to be a means, not to be an objec-
tive purpose, because subjective purpose remains an external, subjective
determination. Or in so far as purpose is active and attains completion,
albeit only in a means, it is still bound up with objectivity immediately; it
is sunk into it. Purpose is itself an object and, as one may say, it does not
attain a means because its realization is needed before such a realization
can be brought about through a means.
But the result is in fact not only an external purposive connection, but
the truth of such a connection, inner purposive connection and an objective
purpose. The self-subsistence of the object over against the concept that
purpose presupposes is posited in this presupposition as an unessential
reflective shine and as already sublated in and for itself; the activity of the
purpose truly is, therefore, only the exposure of this reflective shine and the
sublation of it. – As the concept has demonstrated, 38 the first object becomes
by virtue of communication a means, for it implicitly is the totality of the
concept, and its determinateness, which is none other than the externality
itself, is posited as something only external and unessential – is posited
in purpose itself, therefore, as the latter’s own moment, not as anything
that stands on its own over against it. As a result, the determination of the
object as a means is altogether immediate. There is no need, therefore, for
the subjective purpose to exercise any violence to make the object into a
means, no need of extra reinforcement; the resolution, 39 the resolve, 40 this
determination of itself, is the only posited externality of the object, which is
therein immediately subjected to purpose, and has no other determination
as against it than that of the nothingness of the being-in-and-for-itself.
The second sublating of objectivity through objectivity differs from
this first sublation in that the latter, being the first, is the purpose in
objective immediacy; the second, therefore, is not only the sublating of a
first immediacy but of both, of the objective as something merely posited
and of the immediate. The negativity thus returns to itself in such a way
that it is equally the restoration of objectivity, but of an objectivity which
is identical with it, and in this it is at the same time also the positing of it
as an external objectivity which is only determined by purpose. Because of
this positing, the product remains as before also a means; because of the
identity with negativity, the product is an objectivity which is identical with
the concept, is the realized purpose in which the side of being a means is
the reality itself of purpose. In the completed purpose the means disappears
because it would be simply and solely the objectivity immediately subsumed
under that purpose, an objectivity which in the realized purpose is the
turning back of the purpose into itself; further, there also disappears with
it mediation itself, as the relating of an external; it disappears into both
the concrete identity of objective purpose, and into the same identity as
abstract identity and immediacy of existence.
Herein is also contained the mediation that was required for the first
premise, the immediate connection of the purpose with the object. The
realized purpose is also a means; conversely, the truth of the means is just
this, to be the real purpose itself, and the first sublation of objectivity
is already also the second, just as the second proved to contain the first
also. 41 For the concept determines itself, its determinateness is the external
indifference which is immediately determined in the resolution as sublated,
that is to say, as inner, subjective indifference and at the same time as pre-
supposed object. Its further procession out of itself that appeared – namely
as the immediate communication and subsumption of the presupposed
object under it – is at one and the same time the sublating of that deter-
minateness of externality which was internal, shut up in the concept, that is,
posited as sublated, and the sublating of the presupposition of an object;
consequently, this apparently first sublating of the indifferent objectivity is
already the second as well, an immanent reflection that has gone through
mediation, and the realized purpose.
Since the concept is here, in the sphere of objectivity where its deter-
minateness has the form of indifferent externality, in reciprocal action with
itself, the exposition of its movement becomes doubly difficult and intri-
cate, for such a movement is itself immediately doubled and a first is always
also a second. In the concept taken for itself, that is, in its subjectivity, the
difference of itself from itself is as an immediate identical totality on its
own; but since its determinateness here is indifferent externality, its self-
identity is in this externality immediately also self-repulsion again, so that
what is determined as external and indifferent to the identity is rather this
identity itself, and the identity as identity, as self-reflected, is rather its
other. Only by firmly attending to this shall we comprehend the objective
turning back of the concept into itself, that is, its true objectification; only
then shall we see that every one of the single moments through which this
mediation runs its course is itself the whole syllogism of the mediation.
Thus the original inner externality of the concept, by virtue of which the
concept is self-repelling unity, purpose and the striving of purpose towards
objectivity, is the immediate positing or the presupposition of an exter-
nal object; the self-determination is also the determination of an external
object not determined by the concept; and conversely this determination
is self-determination, that is, the sublated externality posited as inner, or
the certainty of the unessentiality of the external object. – Of the second
connection, that of the determination of the object as a means, we have
just shown 42 how it is within itself the self-mediation of purpose in the
object. – Likewise the third mode of connection, mechanism, which pro-
ceeds under the dominance of purpose and sublates the object by virtue
of the object, is on the one hand the sublating of the means, of the
object already posited as sublated, and consequently a second sublation
and immanent reflection, and on the other hand a first determining of
the external object. This last, as we remarked, 43 is in the realized purpose
again the production of only a means; the subjectivity of the finite concept,
by contemptuously rejecting the means, has attained nothing better in its
goal. But this reflection, namely that purpose is attained in the means and
that the means and the mediation are preserved in the fulfilled purpose, is
the final result of the external connection of purpose – a result in which this
connection has sublated itself and which it has exhibited as its truth. – The
last considered third syllogism differs from the rest in that it is in the first
instance the subjective purposive activity of the preceding syllogism, but
also the sublation of external objectivity and consequently of externality in
general; it is this through itself, and is, therefore, the totality in its positedness.
We have now seen subjectivity, the being-for-itself of the concept, pass over
into the concept’s being-in-itself, into objectivity, and then the negativity of
that being-for-itself reassert itself in objectivity; the concept has so deter-
mined itself in that negativity that its particularity is an external objectivity,
or has determined itself as the simple concrete unity whose externality is its
self-determination. The movement of purpose has now attained this much,
namely that the moment of externality is not just posited in the concept,
the purpose is not just an ought and a striving, but as a concrete totality is
identical with immediate objectivity. This identity is on the one hand the
simple concept, and the equally immediate objectivity, but, on the other
hand, it is just as essentially mediation, and it is that simple immediacy only
through this mediation sublating itself as mediation. Thus the concept is
essentially this: to be distinguished, as an identity existing for itself, from
its implicitly existent objectivity, 44 and thereby to obtain externality, but in
this external totality to be the totality’s self-determining identity. So the
concept is now the idea.
