SECTION III 

The idea

The idea is the adequate concept, the objectively true, or the true as such.
If anything has truth, it has it by virtue of its idea, or something has truth
only in so far as it is idea. – The expression “idea” has otherwise also
often been used in philosophy as well as in ordinary life for “concept,”
or even for just a “representation.” To say that I still have no idea of this
lawsuit, this building, this region, means nothing more than I still have
no representation of it. It is Kant who reclaimed the expression “idea”
for the “concept of reason.” 2 – Now according to Kant the concept of
reason should be the concept of the unconditional, but a concept which is
transcendent with respect to appearances, that is, one for which no adequate
empirical use can be made. The concepts of reason are supposed to serve
for the comprehension of perceptions, those of the understanding for the
understanding of them. 3 – In fact, however, if these last concepts of the
understanding are truly concepts, then they are comprehensions, which means
concepts; 4 they will make comprehending 5 possible, and an understanding of
perceptions through concepts of the understanding will be a comprehending.
But if understanding is only the determining of perceptions by categories
such as whole and parts, force, cause, and the like, then it signifies only
a determining by means of reflection, just as by understanding one may
mean only the determinate representation of a fully determined sensuous
content; as when someone is being shown the way, that at the end of the
wood he must turn left, and he replies “I understand,” understanding means
nothing more than a grasp in pictorial representation and in memory. –
“Concept of reason,” too, is a somewhat clumsy expression; for the concept
is in general something rational, and in so far as reason is distinguished
from the understanding and the concept as such, it is the totality of the
concept and objectivity. – The idea is the rational in this sense; it is the
unconditioned, because only that has conditions which essentially refers to
an objectivity that it does not determine itself but which still stands over
against it in the form of indifference and externality, just as the external
purpose had conditions.

If we now reserve the expression “idea” for the objective or real concept
and we distinguish it from the concept itself and still more from mere
representation, then we must also even more definitely reject that estimate
of it according to which the idea is something with no actuality, and true
thoughts are accordingly said to be only ideas. If thoughts are something
merely subjective and contingent, then they certainly have no further value;
but in this they do not stand lower than the temporal and contingent
actualities which likewise have no further value than that of accidentalities
and appearances. But if, on the contrary, the idea is supposed not to have the
value of truth because in regard to appearances it is transcendent, because no
congruent object can be given for it in the world of the senses, then this is
indeed an odd misunderstanding, for objective validity is being denied to it
on the ground that it lacks precisely what makes of appearances the untrue
being of the objective world. In regard to the practical ideas, Kant recognizes
that “nothing can be more harmful and unworthy of a philosopher than
the vulgar appeal to experience, which supposedly contradicts the idea.
Any such alleged contradiction would not be there at all if, for example,
political institutions were set up at the right time in accordance with ideas,
and if crude concepts, crude just because they are drawn from experience, had
not usurped the place of ideas thus thwarting all good intentions.” 6 Kant
regards the idea as something necessary, the goal which, as the archetype, we
must strive to set up as a maximum and to which we must bring actuality
as it presently stands ever closer.

But since the result now is that the idea is the unity of the concept and
objectivity, the true, we must not regard it as just a goal which is to be
approximated but itself remains always a kind of beyond; we must rather
regard everything as being actual only to the extent that it has the idea in
it and expresses it. It is not just that the subject matter, the objective and
the subjective world, ought to be in principle congruent with the idea; the
two are themselves rather the congruence of concept and reality; a reality
that does not correspond to the concept is mere appearance, something
subjective, accidental, arbitrary, something which is not the truth. When
it is said that there is no subject matter to be found in experience which
is perfectly congruent with the idea, the latter is opposed to the actual as
a subjective standard; but there is no saying what anything actual might
possibly be in truth, if its concept is not in it and its objectivity does
not measure up to this concept; it would then be a nothing. Indeed, the
mechanical and the chemical object, like a subject devoid of spirit and a
spirit conscious only of finitude and not of its essence, do not, according
to their various natures, have their concept concretely existing in them in
its own free form. But they can be something at all true only in so far as
they are the union of their concept and reality, of their soul and their body.
Wholes like the state and the church cease to exist in concreto when the
unity of their concept and their reality is dissolved; the human being, the
living thing, is dead when soul and body are parted in it; dead nature, the
mechanical and the chemical world – that is, when “the dead” is taken
to mean the inorganic world, for the expression would otherwise have no
positive meaning at all – this dead nature, then, if it is separated into its
concept and its reality, is nothing but the subjective abstraction of a thought
form and a formless matter. Spirit that were not idea, not the unity of the
concept with itself, not the concept that has the concept itself as its reality,
would be dead spirit, spiritless spirit, a material object.

Since the idea is the unity of the concept and reality, being has attained the
significance of truth; it now is, therefore, only what the idea is. Finite things
are finite because, and to the extent that, they do not possess the reality of
their concept completely within them but are in need of other things for it –
or, conversely, because they are presupposed as objects and consequently
the concept is in them as an external determination. The highest to which
they attain on the side of this finitude is external purposiveness. That actual
things are not congruent with the idea constitutes the side of their finitude,
of their untruth, and it is according to this side that they are objects, each
in accordance with its specific sphere, and, in the relations of objectivity,
determined as mechanical, chemical, or by an external purpose. That the
idea has not perfectly fashioned their reality, that it has not completely
subjugated it to the concept, the possibility of that rests on the fact that
the idea itself has a restricted content; that, as essentially as it is the unity
of the concept and reality, just as essentially it is also their difference; for
only the object is the immediate unity, that is, the unity that only exists in
itself. But if a subject matter, say the state, did not at all conform to its idea,
that is to say, if it were not rather the idea of the state; if its reality, which is
the self-conscious individuals, did not correspond at all to the concept, its
soul and body would have come apart; the soul would have taken refuge
in the secluded regions of thought, the body been dispersed into singular
individualities. But because the concept of the state is essential to the nature
of these individualities, it is present in them as so mighty an impulse that
they are driven to translate it into reality, be it only in the form of external
purposiveness, or to to put up with it as it is, or else they must needs perish.
The worst state, one whose reality least corresponds to the concept, in so
far as it still has concrete existence, is yet idea; the individuals still obey the
power of a concept.

But the idea has not only the general meaning of true being, of the unity of
concept and reality, but also the more particular one of the unity of subjective
concept and objectivity. For the concept is as such itself already the identity of
itself and reality; for the indeterminate expression “reality” means nothing
but determinate being, and this the concept possesses in its particularity
and singularity. Objectivity, moreover, is likewise the total concept that
has withdrawn into identity with itself out of its determinateness. In the
subjectivity of the concept, the determinateness or the difference of the
latter is a reflective shine which is immediately sublated, withdrawn into
being-for-itself or into negative unity, an inhering predicate. But in this
objectivity the determinateness is posited as immediate totality, as external
whole. Now the idea has shown itself to be the concept liberated again
into its subjectivity from the immediacy into which it has sunk in the
object; it is the concept that distinguishes itself from its objectivity –
but an objectivity which is no less determined by it and possesses its
substantiality only in that concept. This identity has therefore rightly been
designated as a subject-object, for it is just as well the formal or subjective
concept as it is the object as such. 7 But this is a point that needs further
precision. The concept, inasmuch as it has truly attained its reality, is
this absolute judgment whose subject distinguishes itself as self-referring
negative unity from its objectivity and is the latter’s being-in-and-for-itself;
but it refers to it essentially through itself and is, therefore, self-directed
purpose and impulse. For this very reason, however, the subject does not
possess objectivity immediately in it (it would then be only the totality of
the object as such, a totality lost in the objectivity) but is the realization
of the purpose – an objectivity posited by virtue of the activity of the
purpose, one which, as positedness, has its subsistence and its form only as
permeated by its subject. As objectivity, it has the moment of the externality
of the concept in it and is in general, therefore, the side of finitude, of
alteration and appearance; but this side retreats into the negative unity of
the concept and there it perishes; the negativity whereby its indifferent
externality of being 8 manifests itself as unessential and as a positedness is
the concept itself. Despite this objectivity, the idea is therefore absolutely
simple and immaterial, for the externality has being only as determined
by the concept and as taken up into its negativity; in so far as it exists as
indifferent externality, it is not only abandoned to mechanism in general
but exists only as the transitory and untrue. – Thus although the idea
has its reality in a materiality, 9 the latter is not an abstract being standing
over against the concept but, on the contrary, it exists only as becoming,
as simple determinateness of the concept by virtue of the negativity of the
indifferent being.

This yields the following closer determinations of the idea. – First,
the idea is the simple truth, the identity of concept and objectivity as a
universal in which the opposition, the presence of the particular, is dissolved
in its self-identical negativity and is equality with itself. Second, it is the
connection of the subjectivity of the simple concept, existing for itself, and
of the concept’s objectivity which is distinguished from it; the former is
essentially the impulse to sublate this separation, and the latter is indifferent
positedness, subsistence which in and for itself is null. As this connection,
the idea is the process of disrupting itself into individuality and into the
latter’s inorganic nature, and of then bringing this inorganic nature again
under the controlling power of the subject and back to the first simple
universality. The identity of the idea with itself is one with the process; the
thought that liberates actuality from the seeming of purposeless mutability
and transfigures it into idea must not represent this truth of actuality as
dead repose, as a mere picture, numb, without impulse and movement, as
a genius or number, or as an abstract thought; the idea, because of the
freedom which the concept has attained in it, also has the most stubborn
opposition within it; its repose consists in the assurance and the certainty
with which it eternally generates that opposition and eternally overcomes
it, and in it rejoins itself.

But the idea is at first again only immediate or only in its concept; the
objective reality is indeed conformable to the concept but has not yet been
liberated into the concept, and it does not concretely exist explicitly as
the concept. Thus the concept is indeed the soul, but the soul is in the
guise of an immediate, that is, it is not determined as soul itself, has not
comprehended itself as soul, does not have its objective reality within itself;
the concept is as a soul that is not yet fully animated.

Thus the idea is, first of all, life. 
It is the concept which, distinct from its objectivity,
simple in itself, permeates that objectivity
and, as self-directed purpose, has its means within it
and posits it as its means, yet is immanent in this means
and is therein the realized purpose identical with itself.
The idea, on account of its immediacy, has singularity
for the form of its concrete existence.
But the reflection within it of its absolute process is
the sublating of this immediate singularity;
thereby the concept, which as universality is
in this singularity the inner,
transforms externality into universality,
or posits its objectivity as a self-equality.

Thus is the idea, in second place, 
the idea of the true and the good,
as cognition and will. 
It is at first finite cognition and finite will,
where the true and the good are still distinguished 
and the two are at first only as a goal.
The concept has first liberated itself into itself,
giving itself only a still abstract objectivity for its reality.
But the process of this finite cognition and this finite action transforms 
the initially abstract universality into totality,
whereby it becomes complete objectivity.
Or considered from the other side, finite, that is, subjective spirit,
makes for itself the presupposition of an objective world, 
such a presupposition as life only has;
but its activity is the sublating of this presupposition
and the turning of it into something posited.
Thus its reality is for it the objective world,
or conversely the objective world is
the ideality in which it knows itself.

Third, spirit recognizes the idea as its absolute truth, 
as the truth that is in and for itself:
the infinite idea in which cognizing and doing are equalized, 
and which is the absolute knowledge of itself.

A. LIFE

The idea of life has to do with a subject matter so concrete, and if you will so
real, that in dealing with it one may seem according to the common notion
of logic to have overstepped its boundaries. Needless to say, if the logic were
to contain nothing but empty, dead forms of thought, then there could be
no talk in it at all of such a content as the idea, or life, are. But if the subject
matter of logic is the absolute truth, and truth as such lies essentially in
cognition, then cognition at least would have to come in for consideration. –
It is common practice to have the so-called pure logic be followed by an
applied logic – a logic that has to do with concrete cognition, quite apart from
all the psychology and anthropology that is commonly deemed necessary to
interpolate into logic. But the anthropological and psychological side of
cognition is concerned with the form in which cognition appears when the
concept does not as yet have an objectivity equal to it, that is, when it
does not have itself as object. The part of the logic that deals with this
concrete cognition does not belong to applied logic as such; if it did, then
every science would have to be dragged into logic, for each is an applied
logic in so far as it consists in apprehending its subject matter in forms of
thought and of concepts. – The subjective concept has presuppositions that
are exhibited in psychological, anthropological, and other forms. But the
presuppositions of the pure concept belong in logic only to the extent that
they have the form of pure thoughts, of abstract essentialities, such as are the
determinations of being and essence. The same goes for cognition, which is
the concept’s comprehension of itself: no other shape of its presupposition
but the one which is itself idea is to be dealt with in the logic; this, however,
is a presupposition which is necessarily treated in logic. This presupposition
is now the immediate idea; for while cognition is the concept, in so far as
the latter exists for itself but as a subjectivity referring to an objectivity,
then the concept refers to the idea as presupposed or as immediate. But the
immediate idea is life.

Life

To this extent the necessity of considering the idea of life in logic would
be based on the necessity, itself recognized in other ways, of treating the
concrete concept. But this idea has arisen through the concept’s own
necessity; the idea, that which is true in and for itself, is essentially the
subject matter of the logic; since it is first to be considered in its immediacy,
so that this treatment be not an empty affair devoid of determination, it is
to be apprehended and cognized in this determinateness in which it is life.
A comment may be in order here to differentiate the logical view of life
from any other scientific view of it, though this is not the place to concern
ourselves with how life is treated in non-philosophical sciences but only
with how to differentiate logical life as idea from natural life as treated in
the philosophy of nature, and from life in so far as it is bound to spirit. – As
treated in the philosophy of nature, as the life of nature and to that extent
exposed to the externality of existence, life is conditioned by inorganic nature
and its moments as idea are a manifold of actual shapes. Life in the idea is
without such presuppositions, which are in shapes of actuality; its presuppo-
sition is the concept as we have considered it, on the one hand as subjective,
and on the other hand as objective. In nature life appears as the highest stage
that nature’s externality can attain by withdrawing into itself and sublating
itself in subjectivity. It is in logic the simple in-itselfness which in the idea
of life has attained the externality truly corresponding to it; the concept
that came on the scene earlier as a subjective concept is the soul of life itself;
it is the impulse that gives itself reality through a process of objectification.
Nature, as it reaches this idea starting from its externality, transcends itself;
its end is not its beginning but is for it as a limit in which it sublates itself. –
Similarly, in the idea of life the moments of life’s reality do not receive the
shape of external actuality but remain enveloped in conceptual form.
In spirit, however, life appears both as opposed to it and as posited as at
one with it, in a unity reborn as the pure product of spirit. For life is here
to be taken generally in its proper sense as natural life, for what is called
the life of spirit as spirit, is spirit’s own peculiar nature that stands opposed
to mere life; just as we speak of the nature of spirit, even though spirit is
nothing natural but stands rather in opposition to nature. Thus life as such
is for spirit in one respect a means, and then spirit holds it over against
itself; in another respect, spirit is an individual, and then life is its body; in
yet another respect, this unity of spirit and its living corporeality is born
of spirit into ideality. None of these connections of life to spirit concerns
logical life, and life is to be considered here neither as the instrument of
a spirit, nor as a living body, nor again as a moment of the ideal and of
beauty. – In both cases, as natural life and as referring to spirit, life obtains a
determinateness from its externality, in one case through its presuppositions,
such as are other formations of nature, and in the other case through the
purposes and the activity of spirit. The idea of life by itself is free from both
the conditioning objectivity presupposed in the first case and the reference
to subjectivity of the second case.
Life, considered now more closely in its idea, is in and for itself absolute
universality; the objectivity which it possesses is throughout permeated by
the concept, and this concept alone it has as substance. Whatever is dis-
tinguished as part, or by some otherwise external reflection, has the whole
concept within it; the concept is the soul omnipresent in it, a soul which is
simple self-reference and remains one in the manifoldness that accrues to
the objective being. This manifoldness, as self-external objectivity, has an
indifferent subsistence which in space and time, if these could already be
mentioned here, is a mutual externality of entirely diverse and atomistic
matters. But externality is in life at the same time as the simple determi-
nateness of its concept; thus the soul flows omnipresently in this manifold
but remains at the same time the simple oneness of the concrete concept
with itself. – That way of thinking that clings to the determinations of
reflective relations and of the formal concept, when it comes to consider
life, the unity of its concept in the externality of objectivity, the absolute
multiplicity of atomistic matter, finds that all its thoughts are absolutely of
no avail; the omnipresence of the simple in the manifold externality is for
reflection an absolute contradiction and also, since it cannot at the same
time avoid witnessing this omnipresence in the perception of life and must
therefore grant the actuality of this idea, an incomprehensible mystery – for
reflection does not grasp the concept, nor does it grasp it as the substance
of life. – But this simple life is not only omnipresent; it is the one and
only subsistence and immanent substance of its objectivity; but as subjective
substance it is impulse, more precisely the specific impulse of particular dif-
ference, and no less essentially the one and universal impulse of the specific
that leads its particularization back to unity and holds it there. Only as this
negative unity of its objectivity and particularization is life self-referring, life
that exists for itself, a soul. As such, it is essentially a singular that refers to
objectivity as to an other, an inanimate nature. The originative judgment of
life consists therefore in this, that it separates itself off as individual subject
from the objective and, since it constitutes itself as the negative unity of
the concept, makes the presupposition of an immediate objectivity.
First, life is therefore to be considered as a living individual that is for itself
the subjective totality and is presupposed as indifferent to an objectivity
that stands indifferent over against it.
Second, it is the life-process of sublating its presupposition, of positing
as negative the objectivity indifferent to it, and of actualizing itself as the
power and negative unity of this objectivity. By so doing, it makes itself
into the universal which is the unity of itself and its other.
Third, consequently life is the genus-process, the process of sublating
its singularization and relating itself to its objective existence as to itself.
Accordingly, this process is on the one hand the turning back to its concept
and the repetition of the first forcible separation, the coming to be of a
new individuality and the death of the immediate first; but, on the other
hand, the withdrawing into itself of the concept of life is the becoming of
the concept that relates itself to itself, of the concept that exists for itself,
universal and free, the transition into cognition.

A. THE LIVING INDIVIDUAL

1. The concept of life or universal life is the immediate idea, the concept that
has an objectivity corresponding to it; but the objectivity corresponds to it
only to the extent that the concept is the negative unity of this externality,
that is to say, posits it as corresponding to it. The infinite reference of
the concept to itself is as negativity a self-determining, the diremption of
itself within itself as subjective singularity and itself as indifferent universality.
The idea of life in its immediacy is as yet only the creative universal soul.
Because of this immediacy, the first internal negative reference of the idea
is the self-determination of itself as concept – an implicit positing which is
explicit only as a turning back into itself; this is the creative presupposing.
By virtue of this self-determining, universal life is particularized; it has
thus split itself into the two extremes of the judgment which immediately
becomes syllogism.
The determinations of the opposition are the universal determinations
of the concept, for the splitting into two is the affair of the concept; the
filling of them, however, is the idea. One determination is the unity of
the concept and of reality, which is the idea as the immediate unity that
earlier assumed the form of objectivity. Here, however, it is in another
determination. There, it was the unity of the concept and of reality in
so far as the concept has gone over into reality and is lost in it; it did
not stand opposite to it, or, since it is for the reality only an inner, it is
itself for it only an external reflection. That earlier objectivity is therefore
the immediate itself in immediate mode. Here, on the contrary, it has
proceeded only from the concept, so that its essence is positedness, or that
it exists as negative. – It is to be regarded as the side of the universality of
the concept, hence as an abstract universality, essentially only inhering in
the subject and in the form of immediate being which, posited for itself, is
indifferent to the subject. Hence the totality of the concept that attaches
to the objectivity is, as it were, only lent to it; the last self-subsistence that
objectivity has over against the subject is this being whose truth is only
that moment of the concept in which the latter, as a presupposing, is in the
first determinateness of a positing that exists implicitly and is not yet as
positing, as immanently reflected unity. Having proceeded from the idea,
self-subsisting objectivity is therefore immediate being only as the predicate
of the judgment of the concept’s self-determination – a being that is indeed
distinct from the subject but is at the same time essentially posited as a
moment of the concept.
According to content, this objectivity is the totality of the concept – a
totality, however, that has the subjectivity of the concept, or its negative
unity, standing over against it, and this subjectivity or negativity is what
constitutes the true centrality, that is to say, the concept’s free unity with
itself. This subject is the idea in the form of singularity, as simple but
negative self-identity – the living individual.
This individual is in the first place life as soul, as the concept of itself,
fully determined within itself, the initiating self-moving principle. In its
simplicity the concept contains determinate externality, as a simple moment
enclosed within itself. – But, further, this soul is in its immediacy imme-
diately external, and has an objective being within it – a reality which is
subjugated to purpose, the immediate means, at first the objectivity which
is predicated of the subject, but then also the middle term of the syllogism,
for the corporeity of the soul is that whereby the soul links itself to exter-
nal objectivity. – The living being has this corporeity at first as a reality
immediately identical with the concept; to this extent, the corporeity has
this reality in general by nature.
Now because this objectivity is the predicate of the individual and is
taken up in the subjective unity, the earlier determinations of the objects
do not attach to it, not the relation of mechanism or of chemism, and
even less so the reflective relations of whole and part, and the like. As
externality, it is indeed capable of such relations, but to that extent it is no
longer a living being; when a living thing is taken to be a whole consisting
of parts, something exposed to the action of mechanical or chemical causes,
itself a mechanical or chemical product (whether merely as such or as also
determined by some external purpose), then the concept is taken as external
to it, the individual itself as something dead. Since the concept is immanent
in it, the purposiveness of the living being is to be grasped as inner; it is
present in it as a determinate concept, distinguished from its externality
but, in thus distinguishing itself from it, pervading it thoroughly and self-
identical. This objectivity of the living being is the organism; it is the means
and instrument of purpose, fully purposive, for the concept constitutes
its substance; but precisely for this reason this means and instrument
is itself the accomplished purpose in which the subjective purpose thus
immediately closes in upon itself. 14 As for its externality, the organism is
a manifold, not of parts but of members. (a) These members exist as such
only in the individuality; they are separable inasmuch as they are external
and can be grasped in this externality, but as thus separated they revert to
the mechanical and chemical relations of common objectivity. (b) Their
externality is opposed to the negative unity of the living individuality. This
individuality is therefore the impulse to posit as a concretely real 15 difference
the otherwise abstract moment of the determinateness of the concept; and
since this concretely real difference exists as immediate, it is the impulse
of each singular, specific moment to produce itself and equally to raise its
particularity to universality, to sublate the other moments external to it
and promote itself at their cost, but no less to sublate itself and make itself
a means for the other.
2. This process of the living individuality is restricted to itself and still
falls entirely within the individuality. – The first premise of the syllogism
of external purposiveness, where the purpose immediately refers to objec-
tivity and makes it a means, was earlier 16 taken in the sense that although
in it the purpose remains self-equal and has gone back into itself, the
objectivity has not yet sublated itself within, and consequently the purpose
is not in it in and for itself but becomes such only in the conclusion.
The process of the living being with itself is this same premise, but in
so far as the premise is also the conclusion, in so far as the immediate
reference of the subject to the objectivity, by virtue of which the latter
becomes means and instrument, is at the same time the negative unity of
the concept within itself, the purpose realizes itself in this externality by
being the subjective power over it and the process in which the external-
ity displays its self-dissolution and its return into this negative unity of
the purpose. The unrest and the mutability of the external side of the
living being is the manifestation in it of the concept, and the concept,
as in itself negativity, has objectivity in so far as this objectivity’s indif-
ferent subsistence shows itself to be self-sublating. Thus the concept pro-
duces itself through its impulse in such a way that the product, being its
essence, is itself the producing factor: is product, in other words, only
as an externality that equally posits itself negatively, or as the process of
production.
3. Now the idea as just considered is the concept of the living subject and
of its process; the determinations that stand in relation to one another are
the self-referring negative unity of the concept and the objectivity which is
the concept’s means but also where the concept has returned into itself.
But since these moments of the idea of life do not go past the concept
of life, they are not the determinate conceptual moments of the living
individual in its reality. This individual’s objectivity or its corporeity is
a concrete totality; those moments are the sides out of which the living
reality 17 constitutes itself; they are not, therefore, the moments of this
reality as already constituted by the idea. But the living objectivity of the
individual, since as objectivity it is ensouled by the concept and has the
latter for its substance, has also in it, for its essential difference, such
determinations as pertain to the concept, universality, particularity, and
singularity; hence the shape in which the determinations in it are exter-
nally differentiated is divided or incised (insectum) in accordance with
these.
Thus that shape is in this instance universality, the purely internal pul-
sating of living reality, sensibility. The concept of universality, as we saw it
earlier, is the simple immediacy which is such, however, only as inherently
absolute negativity. This concept of absolute difference with its negativity
dissolved into simplicity and self-equal, is brought to intuition in sensibility.
It is the in-itselfness not as abstract simplicity, but as an infinite deter-
minable receptivity that does not become in its determinateness anything
manifold and external but is absolutely reflected into itself. Determinate-
ness is present in this universality as simple principle; the singular external
determinateness, the so-called impression, goes back from its external and
manifold determination into this simplicity of self-feeling. Sensibility may
therefore be regarded as the external existence of the inward soul, for it
takes in all externality but reduces it to the complete simplicity of self-equal
universality.
The second determination of the concept is particularity. This is the
moment of posited difference, the opening up of the negativity otherwise
locked up in simple self-feeling or present in it as abstractly ideal, not yet
concretely real determinateness. It is irritability. Because of the abstraction
of its negativity, feeling is impulse; it determines itself; the self-determination
of the living being is its judgment or the self-limiting whereby it refers to
the outside as to a presupposed objectivity with which it is in reciprocal
activity. – The living being, as a particular, now stands on one side as one
species next to other species; the formal reflection of this indifferent diversity
into itself is the formal genus and its systematization; but the individual
reflection is this, that as outwardly directed the particularity, the negativity
of the living being’s determinateness, is the self-referring negativity of the
concept.
According to this third determination, the living being is a singular. This
immanent reflection further determines itself in such a way that in irritabil-
ity the living being is the externality of itself as against itself, as against the
objectivity that it possesses immediately as its means and instrument and
which is externally determinable. The immanent reflection sublates this
immediacy: on the one side as theoretical reflection, that is, in so far as the
negativity is the simple moment of sensibility as was considered in the lat-
ter, and which constitutes feeling; on the other side as real reflection, in that
the unity of the concept posits itself in its externality as negative unity, and
this is reproduction. – The two first moments, sensibility and irritability, are
abstract determinations; in reproduction life is something concrete and vital;
in it alone does it also have feeling and power of resistance. Reproduction
is the negativity as simple moment of sensibility, and irritability is only a
vital power of resistance, so that the relation to the external is reproduction
and identity of the individual with itself. Each singular moment is essen-
tially the totality of all; their difference constitutes the ideal determination
of form which is posited in reproduction as the concrete totality of the
whole. On the one hand, therefore, this whole is opposed to the previous
determinate totalities as a third, namely as a concretely real totality; on the
other hand, however, it is their implicit essentiality and also that in which
they are comprehended as moments and where they have their subject and
subsistence.
With reproduction as a moment of singularity, the living being posits
itself as actual individuality, a self-referring being-for-itself; but it is at the
same time a real outward reference, the reflection of particularity or irri-
tability as against an other, as against the objective world. The life-process
enclosed within the individual passes over into a reference to the presup-
posed objectivity as such, by virtue of the fact that, as the individual posits
itself as subjective totality, the moment of its determinateness, its reference to
externality, also becomes a totality.

B. THE LIFE-PROCESS

In shaping itself inwardly, the living individual comes into tension with
its original presupposing and, as a subject existing in and for itself, sets
itself in opposition to the presupposed objective world. The subject is a
purpose unto itself, the concept that has its means and subjective reality
in the objectivity subjugated to it. As such, it is constituted as the idea
existing in and for itself and as an essentially self-subsistent being, as against
which the presupposed external world has the value only of something
negative and without self-subsistence. In its self-feeling the living being
has the certainty of the intrinsic nullity of the otherness confronting it. Its
impulse is the need to sublate this otherness and to give itself the truth
of this certainty. At first the individual is, as subject, only the concept
of the idea of life; its inner subjective process in which it feeds upon
itself, and the immediate objectivity which it posits as a natural means
in conformity with its concept, are mediated by the process that refers to
the fully posited externality, to the objective totality standing indifferently
alongside it.
This process begins with need, that is, the twofold moment of self-
determination of the living being by which the latter posits itself as negated
and thereby refers itself to an other than it, to the indifferent objectivity, but
in this self-loss it is equally not lost, preserves itself in it and remains the
identity of the self-equal concept. The living being is thereby the impulse
to posit as its own this world which is other than it, to posit itself as equal
to it, to sublate the world and objectify itself. Its self-determination has
therefore the form of objective externality, and since it is at the same time
self-identical, it is the absolute contradiction. The immediate shape of the
living being is the idea in its simple concept, the objectivity conforming
to the concept; as such the shape is good by nature. But since its negative
moment realizes itself as an objective particularity, that is, since the essential
moments of its unity are each realized as a totality for itself, the concept
splits into two, becoming an absolute inequality with itself; and since even
in this rupture the concept remains absolute identity, the living being is for
itself this rupture, has the feeling of this contradiction which is pain. Pain
is therefore the prerogative of living natures; since they are the concretely
existing concept, they are an actuality of infinite power, so that they are
in themselves the negativity of themselves, that this their negativity exists
for them, that in their otherness they preserve themselves. – It is said that
contradiction cannot be thought; but in the pain of the living being it is
even an actual, concrete existence.

This internal rupture of the living being, when taken up into the simple
universality of the concept, in sensibility, is feeling. From pain begin the
need and the impulse that constitute the transition by which the individual,
in being for itself the negation of itself, also becomes for itself identity – an
identity which only is as the negation of that negation. – The identity which
is in the impulse as such is the individual’s subjective certainty of itself,
in accordance with which it relates to the indifferent, concrete existence
of its external world as to an appearance, to an actuality intrinsically void
of concept and unessential. This actuality is to obtain its concept only
through the subject, which is the immanent purpose. The indifference
of the objective world to determinateness and hence to purpose is what
constitutes its external aptitude to conform to the subject; whatever other
specifications there might be in it, its mechanical determinability, the lack
of the freedom of the immanent concept, constitute its impotence in
preserving itself against the living being. – In so far as the object confronts
the living being at first as something external and indifferent, it can affect
it mechanically, but without in this way affecting it as a living thing; and
in so far as it does relate to it as a living thing, it does not affect it as a cause
but it rather excites it. Because the living being is an impulse, externality
impinges upon it and penetrates it only to the extent that in principle it
is already in it; hence the effect on the subject consists only in that the
latter finds that the externality at its disposal accords with it. And should
this externality not accord with it as a totality, then it must at least accord
with a particular side of it – a possibility lodged in the very fact that, in its
relation to the outside, the subject is a particular.
Now the subject, in so far as in being determined in its need it connects
with the outside and consequently is itself something external or an instru-
ment, exercises violence over the object. Its particular character, its finitude
in general, falls into the more determinate appearance of this relation. – The
external factor in this is the process of objectivity in general, mechanism
and chemism. But this same process is immediately interrupted and the
externality transformed into interiority. The external purposiveness which
is at first elicited in the indifferent object by the activity of the subject is
thereby sublated, for as against the concept the object is not a substance:
the concept, therefore, cannot become for it just an external form but must
rather posit itself as its essence and as the determinateness immanently per-
vading it through and through in conformity with the concept’s original
identity.
By seizing hold of the object, the mechanical process passes over into
an internal process by which the individual appropriates the object in such
a manner that it takes away from it its distinctive make-up, makes it into
a means, and confers upon it its own subjectivity as its substance. This
assimilation thus coincides with the individual’s process of reproduction
considered above; 18 in this process the individual feeds on itself, in the sense
that it makes its own objectivity its object; the mechanical and chemical
conflict of its members with external things is an objective moment of
itself. The mechanical and chemical factor in the process is a beginning of
the dissolution of the living thing. Since life is the truth of these processes,
and as a living being it is therefore the concrete existence of this truth and
the power over the processes, it infringes upon the latter, permeates them
as their universality, and their product is entirely determined by it. This
transformation of them into the living individual constitutes the turning
back of this individual into itself, with the result that the production that
as such would be the transition into an other becomes reproduction, a
reproduction in which the living being posits itself as self-identical for
itself.
The immediate idea is also the immediate identity of concept and reality
but one that does not exist for itself; through the objective process, the
living being gives itself its feeling of self; for in that process it posits itself
as it is in and for itself, namely, as self-identical in an otherness posited as
indifferent to it, as the negative unity of the negative. The individual, in thus
rejoining the objectivity at first presupposed as indifferent to it, has equally
constituted itself as actual singularity and has sublated its particularity,
raising it to universality. Its particularity consisted in the disruption whereby
life posited the individual life and the objectivity external to it as its species.
Through the external life-process, it has consequently posited itself as real
universal life, as genus.

C. THE GENUS

The living individual, at first cut off from the universal concept of life,
is a presupposition yet unproven through itself. Through its process with
the simultaneously presupposed world, it has posited itself for itself as the
negative unity of its otherness, as the foundation of itself; thus it is the
actuality of the idea, so that the individual now brings itself forth out
of actuality, whereas before it proceeded only from the concept, and its
coming to be, which was a presupposing, now becomes its production.
But the further determination that it has attained by the sublation
of the opposition is that it is genus, identity of itself with its hitherto
indifferent otherness. This idea of the individual, since it is this essential
identity, is essentially the particularization of itself. This particularization,
its disruption, in keeping with the totality from which it proceeds, is the
duplication of the individual – the presupposing of an objectivity which
is identical with it, and a relating of the living being to itself as to another
living being.
This universal is the third stage, the truth of life in so far as life is still
shut up within itself. This stage is the process of the individual as it refers
to itself, where the externality is the individual’s immanent moment and
is, besides, itself a living totality, an objectivity which for the individual is
the individual itself – an externality in which the individual has certainty
of itself not as being sublated, but as subsisting.
Now because the relation of genus is the identity of individual self-feeling
in such a one who is at the same time another self-subsistent individual,
it is a contradiction; accordingly, the living being is once more impulse. –
The genus is indeed now the completion of the idea of life, but at first it is
still within the sphere of immediacy; this universality is therefore actual in
a singular shape; it is the concept whose reality has the form of immediate
objectivity. And consequently the individual, although it is the genus, it is
the genus in itself rather than for itself; what is for it is as yet only another
living individual; the concept distinguished from itself has for object, with
which it is identical, not itself as concept, but a concept rather that as a
living being has at the same time external objectivity for it, a form which
is therefore immediately reciprocal.
The identity with the other individual, the universality of the individual,
is thus still only inner or subjective; it therefore has the longing to posit
this identity and to realize itself as universal. But this impulse of the genus
can realize itself only through the sublation of the singular individualities
which are still particular to each other. At first, in so far as it is these
individualities which, in themselves universal, themselves satisfy the tension
of their longing and dissolve themselves into the universality of their genus,
their realized identity is the negative unity of the genus reflecting itself into
itself out of its rupture. To this extent, it is the individuality of life itself,
no longer generated out of its concept but out of the actual idea. At first, it
is itself only the concept that still has to objectify itself, but a concept which
is actual – the germ of a living individual. To ordinary perception what the
concept is, and that the subjective concept has external actuality, are visibly
present in it. For the germ of the living being is the complete concretion
of individuality: it is where all the living being’s diverse sides, its properties
and articulated differences, are contained in their entire determinateness;
where the at first immaterial, subjective totality is present undeveloped,
simple and non-sensuous. Thus the germ is the whole living being in the
inner form of the concept.
From this side the genus obtains actuality through its reflection into
itself, for the moment of negative unity and individuality is thereby posited
in it – the propagation of the living species. The idea, which as life is still
in the form of immediacy, thus falls back into actuality, and its reflection
is now only the repetition and the infinite process in which it does not
step outside the finitude of its immediacy. But this going back to its first
concept also has the higher side that the idea has not only run through the
mediation of its processes inside immediacy, but, just because it has run
through them, has sublated this immediacy and has thereby elevated itself
to a higher form of its existence.
That is to say, the process of the genus in which the single individuals
sublate in one another their indifferent, immediate, concrete existence,
and in this negative unity die away, has further the realized genus that has
posited itself as identical with the concept for the other side of its product. –
In the process of the genus, the isolated singularities of individual life perish;
the negative identity in which the genus turns back into itself is on the
one side the generation of singularity just as it is also, on the other side,
the sublation of it – is thus the genus rejoining itself, the universality of the
idea as it comes to be explicitly for itself. In copulation, 19 the immediacy of
living individuality perishes; the death of this life is the coming to be of
spirit. The idea, implicit as genus, becomes explicit in that it has sublated its
particularity that constituted the living species, and has thereby given itself
a reality which is itself simple universality; thus it is the idea that relates itself
to itself as idea, the universal that has universality for its determinateness
and existence. This is the idea of cognition.

CHAPTER 2

THE IDEA OF COGNITION

Life is the immediate idea, or the idea as its still internally unrealized
concept. In its judgment, the idea is cognition in general.
The concept is for itself as concept inasmuch as it freely and concretely
exists as abstract universality or a genus. As such, it is its pure self-identity
that internally differentiates itself in such a way that the differentiated is
not an objectivity but is rather equally liberated into subjectivity or into the
form of simple self-equality; consequently, the object facing the concept is
the concept itself. Its reality in general is the form of its existence; all depends
on the determination of this form; on it rests the difference between what
the concept is in itself, or as subjective, and what it is when immersed in
objectivity, and then in the idea of life. In this last, the concept is indeed
distinguished from its external reality and posited for itself; however, this
being-for-itself which it now has, it has only as an identity that refers to itself
as immersed in the objectivity subjugated to it, or to itself as indwelling,
substantial form. The elevation of the concept above life consists in this,
that its reality is the concept-form liberated into universality. Through this
judgment the idea is doubled, into the subjective concept whose reality is
the concept itself, and the objective concept which is as life. – Thought,
spirit, self-consciousness, are determinations of the idea inasmuch as the latter
has itself as the subject matter, and its existence, that is, the determinateness
of its being, is its own difference from itself.
The metaphysics of the spirit or, as was more commonly said in the past,
of the soul, revolved around the determinations of substance, simplicity,
immateriality. These were determinations for which spirit was supposed to
be the ground, but as a subject drawn from empirical consciousness, and
the question then was which predicates agreed with the perceived facts.
But this was a procedure that could go no further than the procedure
of physics, which reduces the world of appearance to general laws and
determinations of reflection, for it is spirit still as phenomenal that is taken
as the foundation. In fact, in so far as scientific stringency goes, it also
had to fall short of physics. For not only is spirit infinitely richer than
nature; since its essence is constituted by the absolute unity in the concept
of opposites, and in its appearance, therefore, and in its connection with
externality, it exhibits contradiction at its most extreme form, it must be
possible to adduce an experience in support of each of the opposite deter-
minations of reflection, or, starting from experiences, to proceed by way
of formal inference to the opposite determinations. Since the predicates
immediately drawn from the appearances still belong to empirical psychol-
ogy, so far as metaphysical consideration goes, all that is in truth left are
the entirely inadequate determinations of reflection. – In his critique of
rational psychology, 20 Kant insists that, since this metaphysics is supposed
to be a rational science, the least addition of anything drawn from per-
ception to the universal representation of self-consciousness would alter it
into an empirical science, thus compromising its rational purity and its
independence from all experience. – Accordingly, all that is left on this
view is the simple representation “I,” a representation entirely devoid of
content, of which one cannot even say that it is a concept, but must say that
it is a mere consciousness, one that accompanies every concept. Now, as Kant
argues further, this “I,” or, if you prefer, this “it” (the thing) that thinks,
takes us no further than the representation of a transcendental subject of
thoughts = x, a subject which is known only through the thoughts that
are its predicates, and of which, taken in isolation, we cannot ever have
the least concept. This “I” has the associated inconvenience that, as Kant
expresses it, in order to judge anything about it, we must every time already
make use of it, for it is not so much one representation by which a particular
object is distinguished, as it is rather a form of representation in general, in
so far as representation can be said to be cognition. – Now the paralogism
that rational psychology incurs, as Kant expresses it, consists in this: that
modes of self-consciousness in thinking are converted into concepts of the
understanding, as if they were the concepts of an object; that that “I think”
is taken to be a thinking being, a thing-in-itself; that in this way, because I
am present in consciousness always as a subject, am indeed as a singular sub-
ject, identical in all the manifoldness of representation, and distinguishing
myself from this manifoldness as external to it, the illegitimate inference
is thereby drawn that I am a substance, and a qualitatively simple being on
top of that, and a one, and a being that concretely exists independently of the
things of space and time.
I have cited this position in some detail because one can clearly recognize
in it both the nature of the former metaphysics of the soul and also, more to
the point, of the Critique that put an end to it. – The former was intent on
determining the abstract essence of the soul; it went about this starting from
observation, and then converting the latter’s empirical generalizations, and
the determination of purely external reflection attaching to the singularity
of the actual, into the form of the determinations of essence just cited. –
What Kant generally has in mind here is the state of the metaphysics of
his time which, as a rule, stayed at these one-sided determinations with no
hint of dialectic; he neither paid attention to, nor examined, the genuinely
speculative ideas of older philosophers on the concept of spirit. In his
critique of those determinations he then simply abided by the Humean
style of skepticism; that is to say, he fixes on how the “I” appears in
self-consciousness, but from this “I,” since it is its essence (the thing in
itself ) that we want to cognize, he removes everything empirical; nothing
then remains but this appearance of the “I think” that accompanies all
representations and of which we do not have the slightest concept. – It must
of course be conceded that, as long as we are not engaged in comprehending
but confine ourselves to a simple, fixed representation or to a name, we
do not have the slightest concept of the “I,” or of anything whatever, not
even of the concept itself. – Peculiar indeed is the thought (if one can call
it a thought at all) that I must make use of the “I” in order to judge the
“I.” The “I” that makes use of self-consciousness as a means in order to
judge: this is indeed an x of which, and also of the relation involved in this
“making use,” we cannot possibly have the slightest concept. But surely it is
laughable to label the nature of this self-consciousness, namely that the “I”
thinks itself, that the “I” cannot be thought without the “I” thinking it, an
awkwardness and, as if it were a fallacy, a circle. The awkwardness, the circle,
is in fact the relation by which the eternal nature of self-consciousness and
of the concept is revealed in immediate, empirical self-consciousness – is
revealed because self-consciousness is precisely the existent and therefore
empirically perceivable pure concept; because it is the absolute self-reference
that, as parting judgment, makes itself into an intended object and consists
in simply making itself thereby into a circle. – This is an awkwardness that
a stone does not have. When it is a matter of thinking or judging, the stone
does not stand in its own way; it is dispensed from the burden of making
use of itself for the task; something else outside it must shoulder that effort.
The defect, which these surely barbarous notions place in the fact that
in thinking the “I” the latter cannot be left out as a subject, then also
appears the other way around, in that the “I” occurs only as the subject of
consciousness, or in that I can use myself only as a subject, and no intuition
is available by which the “I” would be given as an object; but the concept
of a thing capable of existence only as a subject does not as yet carry any
objective reality with it. – Now if external intuition as determined in time
and space is required for objectivity, and it is this objectivity that is missed,
it is then clear that by objectivity is meant only sensuous reality. But to
have risen above such a reality is precisely the condition of thinking and of
truth. Of course, if the “I” is not grasped conceptually but is taken as a mere
representation, in the way we talk about it in everyday consciousness, then
it is an abstract self-determination, and not the self-reference that has itself
for its subject matter. Then it is only one of the extremes, a one-sided subject
without its objectivity; or else just an object without subjectivity, which it
would be were it not for the awkwardness just touched upon, namely that
the thinking subject will not be left out of the “I” as object. But as a matter
of fact this awkwardness is already found in the other determination, that of
the “I” as subject; the “I” does think something, whether itself or something
else. This inseparability of the two forms in which the “I” opposes itself to
itself belongs to the most intimate nature of its concept and of the concept
as such; it is precisely what Kant wants to keep away in order to retain
what is only a representation that does not internally differentiate itself and
consequently, of course, is void of concept. Now this kind of conceptual
void may well oppose itself to the abstract determinations of reflection or
to the categories of the previous metaphysics, for in one-sidedness it stands
at the same level with them, though these are in fact on a higher level of
thought; but it appears all the more lame and empty when compared with
the profounder ideas of ancient philosophy concerning the concept of the
soul or of thinking, as for instance the truly speculative ideas of Aristotle. 22
If the Kantian philosophy subjected the categories of reflection to critical
investigation, all the more should it have investigated the abstraction of the
empty “I” that he retained, the supposed idea of the thing-in-itself. The
experience of the awkwardness complained of is itself the empirical fact in
which the untruth of that abstraction finds expression.
The Kantian critique of rational psychology only refers to Mendelssohn’s
proof of the persistence of the soul, 23 and I now also cite its refutation of that
proof because of the oddness of what it adduces against it. Mendelssohn’s
proof is based on the simplicity of the soul, by virtue of which it is supposed
to be incapable of alteration in time, of transition into an other. Qualitative
simplicity is in general the form of abstraction earlier considered; as quali-
tative determinateness, it was investigated in the sphere of being and it was
then proved that the qualitative, which is as such abstractly self-referring
determinateness, is precisely for that reason dialectical, mere transition into
an other. 24 In the case of the concept, however, it was shown that, when
considered in connection with persistence, indestructibility, imperishable-
ness, it is that which exists for itself, which is eternal, just because it is not
abstract but concrete simplicity – because it is not a determinateness that
refers to itself abstractly but is the unity of itself and its other, and it cannot
therefore pass over into this other as if it thereby altered in it; it cannot
precisely because it is itself the other, the determinateness, and hence in this
passing over it only comes to itself. 25 – Now the Kantian critique opposes
to this qualitative determination of the unity of the concept a quantitative
one. As it says, although the soul is not a manifold of reciprocally external
parts and contains no extensive magnitude, yet consciousness has a degree,
and the soul, like every concretely existing being, is an intensive magnitude;
with this magnitude, however, there is posited the possibility of a transition
into nothing through gradual vanishing. 26 – Now what is this refutation
but the application to spirit of a category of being, of intensive magnitude –
a determination that has no truth in itself but on the contrary is sublated
in the concept?
Metaphysics – even one that restricted itself to the fixed concepts of the
understanding without rising to speculation, to the nature of the concept
and of the idea – did have for its aim the cognition of truth; it did probe its
subject matter to ascertain whether they were something true or not, whether
substances or phenomena. The triumph of the Kantian critique over this
metaphysics consists, on the contrary, in side-lining any investigation that
would have truth for its aim and this aim itself; it simply does not pose the
one question which is of interest, namely whether a determinate subject,
in this case the abstract “I” of representation, has truth in and for itself.
But to stay at appearances and at the mere representations of ordinary
consciousness is to give up on the concept and on philosophy. Anything
beyond that is branded by the Kantian critique as high-flown, something
to which reason has no claim. As a matter of fact, the concept does fly
high, rising above what has no concept, and the immediate justification
for going beyond it is, for one thing, the concept itself, and, for another,
on the negative side, the untruth of appearance and of representation, and
also of such abstractions as the thing-in-itself and the said “I” which is not
supposed to be an object to itself.
In the context of this logical exposition, it is from the idea of life that the
idea of spirit has emerged, or what is the same thing, that has demonstrated
itself to be the truth of the idea of life. As this result, the idea possesses
its truth in and for itself, with which one may then also compare the
empirical reality or the appearance of spirit to see how far it accords with
it. We have seen regarding life that it is the idea, 27 but at the same time
it has shown itself not to be as yet the true presentation or the true mode
of its existence. For in life, the reality of the idea is singularity; universality
or the genus is the inwardness. The truth of life as absolute negative unity
consists, therefore, in this: to sublate the abstract or, what is the same, the
immediate singularity, and as identical to be self-identical, as genus, to be
self-equal. Now this idea is spirit. – In this connection, we may further
remark that spirit is here considered in the form that pertains to this idea
as logical. For the idea also has other shapes which we may now mention
in passing; in these it falls to the concrete sciences of spirit to consider it,
namely as soul, consciousness, and spirit as such.
The name “soul” was used formerly to mean singular finite spirit in
general, and rational or empirical psychology was supposed to be synony-
mous with doctrine of spirit. The expression, “soul,” evokes an image of it
as if it were a thing like other things. One enquires regarding its seat, the
spatial location from which its forces operate; still more, how this thing
can be imperishable, subjected to the conditions of temporality yet exempt
from alteration in it. The system of monads elevates matter by making all
of it in principle a soul; 28 on this way of representing it, the soul is an
atom like the atoms of matter; the atom that rises from a cup of coffee as
vapor is capable in favorable circumstances of developing into a soul; only
the greater obscurity of its ideation distinguishes it from the kind of thing
that is manifestly soul. – The concept that is for itself is necessarily also
in immediate existence; in this substantial identity with life, immersed in
its externality, the concept is the subject matter of anthropology. But even
anthropology would find alien a metaphysics in which this form of imme-
diacy is made into a soul-thing, into an atom like the atoms of matter. –
To anthropology must be left only that obscure region where spirit,
under influences which were once called sidereal and terrestrial, lives as a
natural spirit in sympathy with nature and has presentiments of the latter’s
alterations in dreams and presentiments, and indwells the brain, the heart,
the liver, and so forth. To the liver, according to Plato, God gave the gift
of prophesy above which the self-conscious human is exalted, so that even
the irrational part of the soul would be provided for by his bounty and
made to share in higher things. 29 To this irrational side belongs further the
behavior of figurative representation, and of higher spiritual activity in so
far as the latter is subject to the play of an entirely corporeal constitution,
of external influences and particular circumstances.
This lowest of the concrete shapes in which spirit is sunk into
materiality 30 has the one immediately superior to it in consciousness. In
this form the free concept, as the “I” existing for itself, is withdrawn from
objectivity, but it refers to the latter as its other, a subject matter that con-
fronts it. Since spirit is here no longer as soul, but, in the certainty that it
has of itself, the immediacy of being has for it the significance rather of a
negative, its identity with itself in the objectivity confronting it is at the
same time still only a reflective shining, for that objectivity still also has the
form of a being that exists in itself. This stage is the subject matter of the
phenomenology of spirit – a science that stands midway between the science
of the natural spirit and of the spirit as such. It considers spirit as it exists
for itself, but at the same time as referring to its other – an other which, as we
have just said, is thereby determined both as an object existing in itself and
as a negative. The science thus considers spirit as appearing, as exhibiting
itself in its contrary.
But the higher truth of this form is spirit for itself. For this spirit, the
subject matter 31 which for consciousness exists in itself has the form of its
own determination, the form of representation in general; this spirit, which
acts on the subject matter’s determinations as on its own, on feelings, on
representations and thought, is thus infinite in itself and in its form. The
consideration of this stage belongs to the doctrine of spirit proper, which
would embrace the subject matter of ordinary empirical psychology but
which, in order to be the science of spirit, must not go about its work
empirically but must be conceived scientifically. – At this stage spirit is
finite spirit in so far as the content of its determinateness is an immediately
given content; the science of this finite spirit has to display the course along
which it liberates itself from this determinateness and goes on to grasp its
truth, the infinite spirit.
The idea of spirit which is the subject matter of logic already stands,
on the contrary, inside pure science; it has no need, therefore, to observe
spirit’s tracing that course, to see how it gets entangled with nature, with
immediate determinateness, with matter, or in other words with pictorial
representation; this is what the other three sciences investigate. The idea of
spirit has this course already behind it, or what is the same, it has it rather
ahead of it – behind in so far as logic is taken as the final science; ahead
in so far as it is taken as the first science from which the idea first passes
over into nature. In the logical idea of spirit, therefore, the “I” is from the
start in the way it has emerged from the concept of nature as the truth of
nature, the free concept which in its judgment is itself the subject matter
confronting it, the concept as its idea. Also in this shape, however, the idea
is still not consummated.
Although the idea is indeed the free concept that has itself as its subject
matter, it is nonetheless immediate, and just because it is immediate, it is
still the idea in its subjectivity, and hence in its finitude in general. It is
the purpose that ought to realize itself, or the absolute idea itself still in its
appearance. What the idea seeks is the truth, this identity of the concept
itself and reality; but at first it only seeks it; for it is here as it is at first, still
something subjective. Consequently, although the subject matter that is for
the concept is here also a given subject matter, it does not enter into the
subject as affecting it, or as confronting it with a constitution of its own as
subject matter, or as a pictorial representation; on the contrary, the subject
transforms it into a conceptual determination; it is the concept which is the
active principle in it – which therein refers itself to itself, and, by thus
giving itself its reality in the object, finds truth.
Initially, therefore, the idea is one extreme of a syllogism, the concept
that as purpose has itself at first for its subjective reality; the other extreme is
the restriction of the subjective, the objective world. The two extremes are
identical in that they are the idea. Their unity is, first, that of the concept,
a unity which in the one extreme is only for itself and in the other only in
itself. Second, it is reality, abstract in the one extreme and in the other in
its concrete externality. – This unity is now posited through cognition, and,
because the latter is the subjective idea which as purpose proceeds from
itself, it is at first only a middle term. – The knowing subject, through the
determinateness of its concept which is the abstract being-for-itself, refers
to an external world; nevertheless, it does this in the absolute certainty
of itself, in order to elevate its implicit reality, this formal truth, to real
truth. It has the entire essentiality of the objective world in its concept; its
process consists in positing for itself the concrete reality of that world as
identical with the concept, and conversely in positing the latter as identical
with objectivity.
Immediately, the idea of appearance is the theoretical idea, cognition as
such. For to the concept that exists for itself, the objective world immedi-
ately has the form of immediacy or of being, just as that concept is to itself
at first only the abstract concept of itself, is still shut up within itself. The
concept is therefore only as form, of which only its simple determinations
of universality and particularity are the reality that it possesses within, while
the singularity or the determinate determinateness, the content, is received
by it from the outside.

A. THE IDEA OF THE TRUE

At first the subjective idea is impulse. For it is the contradiction of the
concept that it has itself for the subject matter and is to itself the reality
without, however, the subject matter being an other that subsists on its
own over against it, or without the differentiation of itself from itself
having at the same time the essential determination of diversity and of
indifferent existence. The specific nature of this impulse is therefore to
sublate its own subjectivity, to make that first abstract reality a concrete
one, filling it with the content of the world presupposed by its subjectivity. –
From the other side, the impulse is determined in this way: the concept
is indeed the absolute certainty of itself; however, opposite its being-for-
itself there stands its presupposition of a world that exists in itself, but one
whose indifferent otherness has for the concept’s certainty of itself the status
of something merely unessential; the concept is therefore the impulse to
sublate this otherness and, in the object, to intuit its identity with itself.
This immanent reflection is the sublated opposition and the singularity
that originally makes its appearance as the presupposed being-in-itself but
is now posited and made actual for the subject; accordingly, this being-in-
itself is the self-identity of the form as it has issued from the opposition –
an identity which is therefore determined as indifferent towards the form
in its differentiation. It is content.
Consequently this impulse is the impulse of truth in so far as the truth
is in cognition, and therefore of truth in its strict sense as theoretical idea. –
Although objective truth is the idea itself as the reality that corresponds to
the concept, and to this extent a subject matter may or may not possess
truth, nevertheless the more precise meaning of truth is that it is such for
or in the subjective concept, in knowledge. 32 Truth is the relation of the
judgment of the concept, the concept that proved to be the formal judgment
of truth; 33 for the predicate in this judgment is not only the objectivity
of the concept, but the comparison connecting the concept of the fact with
the actuality of it. – This realization of the concept is theoretical in so far as
the concept still has, as form, the determination of subjectivity, or in so far
as it has for the subject the determination of being its own determination.
Because cognition is the idea as purpose or as subjective idea, the negation
of the world, presupposed as existing in itself, is the first negation; the
conclusion in which the objective is posited in the subjective has at first,
therefore, only the meaning that what exists in itself is posited only as
something subjective, only in conceptual determination, and consequently
does not exist as so posited in and for itself. Thus the conclusion only
attains to a neutral unity, or a synthesis, that is, to a unity of terms that
are originally separate, only externally conjoined. – Hence, since in this
cognition the concept posits the object as its own, the idea gives itself at first
only a content of which the foundation is given, in which only the form
of externality has been sublated. To this extent, this cognition still retains
its finitude in its realized purpose; in the realized purpose, it has at the
same time not attained its purpose, and in its truth it has not arrived at the
truth. For in so far as in the result the content still has the determination
of a given, the presupposed being-in-itself confronting the concept is not
sublated; the unity of concept and reality, the truth, is thereby equally not
contained in it. – Remarkable is that this side of finitude is the one that of
late has been clung to and accepted as the absolute relation of cognition –
as if the finite as such were to be the absolute! On this view, an unknown
thinghood-in-itself is attributed to the object, behind cognition, and this
thinghood, and the truth also along with it, are regarded for cognition as
an absolute beyond. Thought determinations in general, the categories, the
determinations of reflection, as well as the formal concept and its moments,
acquire on this view the status of determinations that are finite, not in and
for themselves, but in the sense of being something subjective as against
this empty thinghood-in-itself; the fallacy of taking this untrue relation of
cognition as the true relation has become the universal opinion of modern
times.
It is immediately clear from this definition of finite cognition that it is
a contradiction that sublates itself; it is the contradiction of a truth that
is supposed at the same time not to be truth, of a cognition of what is
that at the same time does not know the thing-in-itself. In the collapse of
this contradiction, its content, subjective cognition and the thing-in-itself,
collapses, that is, proves itself to be an untruth. But it is incumbent upon
cognition itself to resolve its finitude by its own forward movement and
along with it its contradiction. What we have said is a consideration which
we bring to it and remains a reflection external to it. But cognition is
itself the concept which is a purpose unto itself and, therefore, through its
realization fulfills itself, and precisely in this fulfillment sublates its subjec-
tivity and the presupposed being-in-itself. – We must examine cognition,
therefore, in its positive activity within it. Because this idea, as we have
shown, 34 is the concept’s impulse to realize itself for itself, its activity con-
sists in determining the object, and by virtue of this determining to refer
itself to itself in it as identical. The object is simply the determinable as
such, and in the idea it has this essential side of not being in and for itself
opposed to the concept. Because this cognition is still finite, not speculative
cognition, the presupposed objectivity does not as yet have for it the shape
of something which is inherently the concept simply and solely and does
not hold anything particular for itself as against the cognition. But by thus
having the status of a beyond that exists in itself, the determination of being
determinable through the concept is essential to it; for the idea is the concept
that exists for itself, is that which is absolutely infinite in itself, in which the
object is implicitly sublated, and the aim is still to sublate it explicitly. The
object, therefore, is indeed presupposed by the idea of cognition as existing
in itself, but as so essentially related to the idea that the latter, certain of
itself and of the nothingness of this opposition, arrives in the object at the
realization of its concept.
In the syllogism whereby the subjective idea now rejoins objectivity, the
first premise is the same form of immediate seizure and connection of the
concept with respect to the object as we see in the purposive connection. 35
The determining activity of the concept upon the object is an immediate
communication of itself to the object, an unresisted invasion of it. In all this
the concept remains in pure self-identity; but this immediate immanent
reflection equally has the determination of objective immediacy; that which
for the concept is its own determination, is equally a being, for it is the
first negation of the presupposition. The posited determination equally
has the status, therefore, of a presupposition which is merely found, the
apprehension of a given wherein the activity of the concept consists rather
in being negative towards itself, in holding itself back away from what is
found and passive towards it, in order that the latter be allowed to show
itself, not as determined by the subject, but as it is in itself.
Georg Wilhelm Friedrich Hegel
In this premise, therefore, this cognition does not in any way appear as an
application of logical determinations, but as a reception and apprehension
of such determinations as already found, and its activity appears restricted
simply to the removing from the subject matter of a subjective obstacle, an
external veil. This cognition is analytic cognition.

a. Analytic cognition

The difference between analytic and synthetic cognition is sometimes said
to be that the one proceeds from the known to the unknown and the other
from the unknown to the known. On closer examination, however, it is
difficult to find any definite thought behind this difference, even less a
concept. It may be said that in general cognition begins with ignorance,
for one does not learn to know something with which one is already
acquainted. Conversely, it also begins with the known, for it is a tautological
proposition that that with which cognition begins, what it therefore actually
knows, is for that reason a known; what is as yet not known, and is expected
to be known only later, is still an unknown. In this respect it must be said
that cognition, once it has begun, always proceeds from the known to the
unknown.
The specific difference of analytic cognition is already established by the
fact that, since it is the first premise of the whole syllogism, mediation
does not as yet belong to it; analytic cognition is rather the immediate
communication of the concept, a communication that does not as yet
contain otherness and in which activity divests itself of its negativity. Yet
this immediacy of the connection is for that reason itself mediation, for it
is a negative reference of the concept to the object that annuls itself and
thereby makes itself simple and identical. This immanent reflection is only
subjective, because in its mediation the difference is present still in the
form of a presupposition existing in itself, as the object’s difference 36 within
itself. The determination that results through this connection, therefore,
is the form of simple identity, of abstract universality. Accordingly, analytic
cognition has in general this identity for its principle, and the transition
into an other, the linking of different terms is excluded from it and from
its activity.
If we look now more closely at analytic cognition, we see that it starts
from a presupposition, hence from some singular, concrete subject matter,
whether for representation this subject matter is already completed or in
the form of a task, that is, given to it only under certain circumstances
and conditions rather than disengaged from these on its own and pre-
sented in simple independence. Now the analysis of this subject matter
cannot consist just in resolving it into the particular representations possibly
contained within it; such a resolution and the apprehension of the partic-
ular representations is an affair that would not belong to cognition, but
would rather be a matter of closer acquaintance, a determination within
the sphere of representing. Analysis, since it is based on the concept, has for
its products determinations that are essentially conceptual, though such
as are contained in the subject matter immediately. We have seen from the
nature of the idea of cognition that the activity of the subjective concept
must be regarded from one side only as the explication of what is already in
the object, 37 for the object itself is nothing but the totality of the concept.
It is just as one-sided to portray analysis as though there were nothing
in the subject matter that is not imported into it, as it is to suppose that
the resulting determinations are only extracted from it. The former way
of stating the case corresponds, as is well known, to subjective idealism,
which takes the activity of cognition in analysis to be only a one-sided
positing, beyond which the thing-in-itself remains hidden. The other way
belongs to the so-called realism, for which the subjective concept is an
empty identity that imports the thought determinations from outside. –
Since analytical cognition, the transformation of the given material into
logical determinations, has shown itself to be a positing that immediately
determines itself to be equally a presupposing, to be both in one, the logical
element can appear on account of this presupposing to be in the subject
matter as something already completed, just as because of the positing it can
appear as the product of a merely subjective activity. But the two moments
are not to be separated. In the abstract form to which analysis raises it,
the logical element is of course only to be found in cognition, just as con-
versely it is not only something posited but something that rather exists in
itself.
Now in so far as analytical cognition is the indicated transformation, it
does not go through further middle terms; on the contrary, the determina-
tion is to that extent immediate and has precisely the meaning of being the
subject matter’s own determination, of belonging to it in itself, and there-
fore of being apprehended directly from it without subjective mediation.
But further, cognition is also supposed to be a progress, an explication of
differences. But because, according to the determination that it has here,
it is void of concept and undialectical, it only possesses a given difference,
and its progression happens solely in the determinations of the material. It
seems to have an immanent progress only in so far as the derived thought
determinations can in turn be analyzed, in so far as they are still some-
thing concrete; the highest and final term of this analyzing is the abstract
highest essence – or the abstract subjective identity and, over against it,
the difference. This progress, however, is nothing but just the repetition
of the one original activity of analysis, namely, the re-determination as a
concretion of what has already been taken up in abstract conceptual form,
and following upon that the analysis of this concretion, and then again
the renewed determination of the resulting abstraction as concrete, and so
forth. – But the thought determinations also seem to contain a transition
in themselves. If the subject matter is defined as a whole, then of course
one advances from it to the other determination of part; from cause to the
other determination of effect, and so forth. But this is no progress, for part
and whole, cause and effect, are relations – indeed, in the context of this
formal cognition they are such consummate relations that the one determi-
nation is already found essentially linked to the other. The subject matter
that has been determined whether as cause or as part is thus determined
by the whole relation, already by both sides of it. Although this relation
is in itself something synthetic, this connection is for analytical cognition
just as much of a mere given as is any other connection in its material,
and therefore outside its sphere of competence. Whether this connection
is otherwise determined as a priori or a posteriori is here indifferent, for it
is apprehended as already given, or, as it has also been called, as a fact of
consciousness – namely the fact that with the determination of whole there
is linked the determination of part, and so on. Kant made the profound
observation that there are synthetic principles a priori, and he recognized
as their root the unity of self-consciousness, hence the self-identity of the
concept. 38 However, he takes the specific connection, the relational con-
cepts, and the synthetic principles, from formal logic as given; the deduction
of these should have been the exposition of the transition of that sim-
ple unity of self-consciousness into these determinations and distinctions;
but Kant spared himself the effort of demonstrating this truly synthetic
progression, that of the self-producing concept.
It is well known that “analytical science” and “analysis” are the names
of preference of arithmetic and the sciences of discrete magnitude in gen-
eral. And in fact their typical method of cognition is most immanently
analytical and we must now briefly consider why this is so. – Any other
analytical cognition begins from a concrete material that has an accidental
manifoldness within; every distinction of content and every advance to
further content depend on this material. The material of arithmetic and
of algebra is, on the contrary, an already totally abstract and indeterminate
product from which every peculiarity of relation has been eliminated, and
to which, therefore, every determination and every joining is something
external. This product is the principle of discrete magnitude, the one. This
relationless atom can be increased to a plurality and externally determined
and unified into a sum; the increasing and the limiting are an empty pro-
gression and an empty determining that never gets past the same principle
of the abstract one. How the numbers are further combined and separated
depends solely on the positing activity of the knowing subject. Magnitude
is in general the category within which these determinations are conducted;
it is the determinateness that has become indifferent, so that the subject
matter has no determinateness which is immanent to it and is therefore
given to cognition. Since cognition has from the start provided itself with
an accidental assortment of numbers, these now constitute the material
for further elaboration and manifold relations. Such relations, their dis-
covery and elaboration, do not seem, it is true, to be anything immanent
in analytical cognition, but seem rather something accidental and given;
moreover, these same relations and the operations connected with them
are also routinely conducted one after the other, as diverse, with no notice
of any internal connectedness. But it is easy to recognize the presence of a
guiding principle; it is none other than the immanent principle of analyt-
ical identity, an identity that in diversity appears as equality; progression is
the reduction of the unequal to ever greater equality. To give an example
from the first elementary operations, addition is the combining of quite
accidentally unequal numbers; multiplication, on the contrary, is the com-
bination of equal numbers, upon which there then follows the relation of
equality of number of times and unit, and the relation of powers then comes
in.
Now because the determinateness of the subject matter and of the
relations is a posited one, any further operation with them is also wholly
analytic; accordingly, analytical science has not so much theorems as it has
problems. The analytical theorem contains the problem as already resolved
on its own terms; the wholly external distinction that attaches to the two
sides which it equates is so unessential that, as theorem, it would appear to
be a trivial identity. To be sure, Kant has declared the proposition 5 + 7 =
12 to be synthetic, because the same is exhibited on the one side in the form
of a plurality, 5 + 7, and on the other in the form of a unity, 12. 39 But, if the
analytic proposition is to mean more than just the totally abstract identity
and tautology of 12 = 12 and is to contain any progression within it at
all, then there must be present some sort of distinction, though not one
based on a quality, on a reflective and still less conceptual determinateness.
5 + 7 and 12 are absolutely the very same content; but the first side also
expresses the demand that 5 and 7 be combined into one expression, that
is to say, that just as 5 is the product of a process of counting that was
arbitrarily interrupted but might just as well have been carried farther, so
now the counting is to be resumed as before with the stipulation that the
ones to be added should be seven. The 12 is therefore the result of 5 and
7 and of a pre-set operation which is by nature also a completely external
and thoughtless act, one that a machine can also therefore perform. Here
there is not the slightest transition to an other; what there is, is the mere
continuation, that is, the repetition, of the same operation that produced 5
and 7.
The proof of a theorem of this kind – and it would require a proof if
it were a synthetic proposition – would consist simply in the operation of
continuing counting, starting from 5, up to 7 as the pre-determined limit,
and in the recognition of the agreement of the product of this counting
with what is otherwise called 12, a figure which is again nothing more
than that same counting up to a determined limit. For this reason we state
the proof in the form of a problem rather than a theorem as a matter of
course. We demand to perform an operation, that is to say, we state only
one side of the equation that would constitute the theorem and whose
other side is now to be found. The problem contains the content and
assigns the specific operation to be performed with it. The operation is not
constrained by any recalcitrant material endowed with specifying relations,
but is rather an external subjective act, the determinations of which are
received with indifference by the material where they are posited. The
whole difference between the conditions stipulated in the problem and the
result in the solution is only this, that the union or separation as stipulated
in the problem is actual in the solution
It is, therefore, a supremely superfluous piece of scaffolding to apply here
the form of geometrical method that goes with synthetic propositions, and
to add to the problem, over and above the solution, a proof as well. Such a
proof can express no more than the tautology that the solution is correct
because the prescribed operation has been performed. If the problem is to
add several numbers, then the solution is to add them; the proof shows
that the solution is correct because addition was prescribed and addition
was performed. If the problem involves more complex expressions and
operations, as for instance the multiplication of decimal numbers, and the
solution only states the mechanical procedure, a proof will then indeed
be necessary; but it can consist in nothing more than the analysis of the
expressions and of the operation from which the solution proceeds of
itself. By this separation of the solution as a mechanical procedure, and of
the proof as a reminder of the nature of the subject matter to be treated
and of the operation itself, we lose precisely the advantage of the analytic
problem, namely that the construction can be derived directly from the
problem and presented, therefore, as intelligible in and for itself; in the
other way, the construction is expressly given a defect which is typical of
the synthetic method. – In higher analysis, especially in connection with
the relations of powers, where qualitative relations of discrete magnitudes
dependent on conceptual determinacies come into play, the problems and
the theorems do of course contain synthetic determinations; in these cases,
other expressions and relations than are given by the problem or theorem
must be taken as intermediary links. But here also, the determinations
enlisted as an aid must be such as to be based on recalling or developing
one side or other of the problem or theorem; the look of synthesis comes
solely from the fact that the problem or theorem has not as yet already
identified that side. – For instance, the problem of finding the sum of the
powers of the roots of an equation is solved through the examination and
then the joining of the functions that are the coefficients of the equation
of the roots. The determination of the functions of these coefficients and
their link here enlisted as an aid is not already expressed in the problem, but
for the rest the development is totally analytical. The same applies to the
solution of the equation x m – 1 = 0 with the aid of the sine, and also to its
immanent algebraic solution, famously discovered by Gauß, which takes
into consideration as an aid the residuum of x m –1 – 1 divided by 1, and the
so-called primitive roots – one of the most important extensions of analysis
in modern times. 40 These solutions are synthetic, for the determinations
enlisted in their aid, the sine or the examination of the residua, are not a
determination of the problem itself.
We gave in the first part of this Logic a detailed account of the nature
of the analysis which is dedicated to the so-called infinite differentiations
of variable magnitudes, the analysis of differential and integral calculus.
It was shown there that underlying this analysis there is a fundamental
qualitative determination of magnitude that can be comprehended only
by the concept. 42 The transition to this determination from magnitude
as such is no longer analytic; to this day, therefore, mathematics has been
incapable of justifying internally, that is, mathematically, the operations
based on it, for the transition is not of a mathematical nature. We said in
the same place that Leibniz, famed for having rendered the calculation of
infinitesimals into a calculus, executed that transition in a way which is
utterly deficient, just as totally void of concept as unmathematical. 43 But
of course, once the transition is presupposed – and in the present state of
the science it is no more than a presupposition – the further course is only
a series of ordinary analytical operations.
We have said that analysis becomes synthetic when it comes to deter-
minations that are no longer posited by the problems themselves. But the
general transition from analytic to synthetic cognition lies in the necessary
transition from the form of immediacy to mediation, from abstract identity
to difference. Analysis in general restricts its activity to determinations in
so far as these are self-referential; yet by virtue of their determinateness they
also essentially refer to an other by nature. We have already remarked that
analytic cognition remains such even when it advances to relations that
are not an externally given material but are rather thought determinations,
since for it these relations are also given. But because the abstract identity
which this cognition knows to be solely its own is essentially an identity
in difference, 44 even as such it must be cognition’s own identity, and the
connection as well must become for the concept one which is posited by it
and is identical with it.

B. SYNTHETIC COGNITION

Analytic cognition is the first premise of the whole syllogism – the imme-
diate reference of the concept to the object. Identity is, therefore, the
determination which analytic cognition recognizes as its own, and analytic
cognition is the apprehension of what is. Synthetic cognition aims at the
comprehension of what is, that is, at grasping the manifoldness of determi-
nations in their unity. It is, therefore, the second premise of the syllogism,
the one in which the diverse as such is connected. Its aim, therefore, is
necessity in general. – The diverse terms that are combined stand, on the
one hand, in a relation in which they are both connected yet mutually indif-
ferent and self-subsistent; but, on the other hand, they are linked together
in the concept which is their simple yet determinate unity. Now inasmuch
as in a first moment synthetic cognition passes over from abstract identity
to relation, or from being to reflection, it is not the absolute reflection
of the concept that the latter recognizes in its subject matter; the reality
that the concept gives itself is the next stage, 45 namely the said identity
in diversity as such, an identity that equally is, therefore, still inner, and
only necessity; it is not the subjective identity existing for itself, hence not
as yet the concept as such. Synthetic cognition, therefore, does also have
for its content the determinations of the concept, the object is posited in
them; but they stand only in relation to one another or in immediate unity,
and for that very reason not in the unity by which the concept exists as
subject.
This is what constitutes the finitude of this cognition. Because the iden-
tity which this real side of the idea has in it is still an inner one, the
determinations of that identity are still external to themselves; and because
the identity is not as subjectivity, the concept’s own specific presence in
the subject matter still lacks singularity; although what in the object corre-
sponds to the concept is no longer the abstract but the determinate form of
the concept and hence the concept’s particularity; the singularizing element
in the object is nevertheless still a given content. Consequently, although
this cognition transforms the objective world into concepts, what it gives
to it in accordance with conceptual determinations is only the form; as for
the object in its singularity, in its determinate determinateness, this it must
find; the cognition is not yet self-determining. It likewise finds propositions
and laws, and proves their necessity; but it proves the latter not as a necessity
inherent in a fact in and for itself, that is to say, it does not demonstrate it
from the concept; it proves it rather as the necessity inherent to a cognition
that delves into given determinations, into phenomenal differences, and
cognizes for itself the proposition as a unity and relation, or cognizes the
ground of appearance from the appearance itself.
We must now examine the detailed moments of synthetic cognition.

1. Definition

To start with, the still given objectivity is transformed into simple form, as
the first form and therefore as the form of the concept; the moments of this
apprehension are none other, therefore, than the moments of the concept;
universality, particularity, and singularity. – The singular is the object itself as
an immediate representation; it is that which is to be defined. The universal
of this singular object took the form of genus in the determination of
the objective judgment, or the judgment of necessity; 46 more precisely,
it took the form of the proximate genus, that is to say, of the universal
with the determinateness which is at the same time the principle for the
differentiation of the particular. This is a difference that the subject matter
receives in its specific non-indifference, 47 the one that makes it a determinate
species and is the basis of its disjunction from the remaining species.
Definition, in thus reducing the subject matter to its concept, gets rid of
the externalities that are requisite for its concrete existence; it abstracts from
what is added to the concept in its realization, whereby the concept issues
first into idea and secondly into external concrete existence. Description is
for representation; it collects this extra content that belongs to reality. But
definition reduces this wealth of manifold determinations of the intuited
existence to its simplest moments; what is contained in the concepts are the
form of these simple elements and how they are determined with respect
to one another. The subject matter is thus apprehended, as we just said,
as a universal which is determined at the same time. The subject matter is
the third, the singular in which genus and particularization are posited in
one – an immediate which is posited outside the concept, for it is not yet
self-determining.
In these determinations, in the difference of form of the definition, the
concept finds itself; there it finds the reality that corresponds to it. But
since the reflection of the moments of the concept into themselves which
is singularity is not as yet contained in this reality, and since the object, in so
far as it is in cognition, is consequently not as yet determined as subjective,
it is cognition which, on the contrary, is subjective and has an external
beginning; that is to say, because of its external beginning in a singular it
is subjective. The content of the concept is therefore something given and
contingent. The concrete concept itself is thus contingent in two respects:
once because its content is contingent; and again because it is a matter
of accident which content determinations, from the many qualities which
the intended object has in external existence, are chosen for the concept as
constituting its moments.
This last respect requires closer consideration. Since singularity is a
determinate way of existing in and for itself, it escapes the conceptual
determination proper of synthetic cognition. There is in fact no principle,
therefore, for determining which aspects of the subject matter are to be
regarded as belonging to its conceptual determination and which only to
its external reality. In the case of definitions, this constitutes a difficulty
which for synthetic cognition cannot be eliminated. A distinction must
nonetheless be made here. – In the first place, so far as the products of self-
conscious purposiveness are concerned, it is easy enough to discover their
definition, for the purpose which they should serve is a determination that
is generated by a subjective resolution and constitutes the essential partic-
ularization, the form of the concrete existent, on which alone everything
depends here. The further nature of the material of the existent thing or
its other external properties, in so far as they correspond to the purpose,
are contained in the thing’s determination; the rest are unessential for it.
Secondly, geometrical objects are abstract determinations of space; the
underlying abstraction, the so-called absolute space, has lost all other
concrete determinations and now possesses no further shapes and con-
figurations than are posited in it; essentially, therefore, such shapes and
configurations are only what they are intended to be; their conceptual deter-
mination in general, and more proximately their specific difference, have
unfettered reality in them; in this respect, therefore, they are the same as
the products of external purposiveness, and in this they also agree with
the objects of arithmetic in which the underlying determination is also
only one that has been posited in them. – Of course, space has yet further
determinations: its tri-dimensionality, its continuity and divisibility which
are not first posited in it by external determination. But these belong to
the material under consideration and are immediate presuppositions; syn-
thetic relations and laws are produced only through the combination and
the interweaving of these subjective determinations with this distinctive
nature of their field into which they have been imported. – In the case of
number determinations, since they are based on the simple principle of
the one, their combination and further determination is only an entirely
posited product; on the other hand, determinations in space, which for its
part is a continuous externality, 48 run a further course of their own and
have a reality that exceeds their concept, but it no longer belongs to the
immediate definition.
But, thirdly, in the case of the definitions of concrete objects, of nature as
well as of spirit, the situation is quite different. For representation, such sub-
ject matters are in general things of many properties. In their case all depends
on apprehending what is their proximate genus, and then what is their
specific difference. We have to determine, therefore, which of the many
properties pertains to the subject matter as genus, which as species, and
which among these properties is the essential one; this further involves rec-
ognizing how the properties hang together, whether one is already posited
with the other. For this, however, no other criterion is yet available than
existence itself. – For the definition, in which the property is to be posited
as simple undeveloped determinateness, the essentiality of the property is
its universality. But in existence this universality is empirical; it is a univer-
sality in time (whether the property persists while the rest ostensibly come
and go within the permanence of the whole), or a universality resulting
from comparison with other concrete wholes, in which case it does not get
beyond commonality. Now if comparison gives as a common foundation
the total habitus, such as is empirically given, then reflection must gather
it together into one simple thought determination and grasp the simple
character of the resulting totality. But the only possible attestation that
a thought determination, or any single one of the immediate properties,
constitutes the simple and determinate essence of the subject matter is
its derivation from the concrete constitution of the latter. But this would
require an analysis that transforms the immediate elements of this consti-
tution into thoughts and reduces their concreteness to a simple thought
determination; and this is an analysis of a higher order than the one just
considered, for it would not be abstractive; on the contrary, in the universal
it should still retain the singular character of the concrete, should unify it
and show that it is dependent on the simple thought determination.
The connections of the manifold determinations of immediate exis-
tence to the simple concept would however require theorems, and these
need proof. But definition is the first, still undeveloped concept, and in
that it has to apprehend the simple determinateness of the subject matter,
and this apprehension is to be something simple, it can employ for the
purpose only one of the subject’s immediate so-called properties, a deter-
mination of sensuous existence or of representation; the singling out, then,
of this property through abstraction is what constitutes the simplicity, and
for universality and essentiality the concept must resort to empirical uni-
versality, to persistence under altered circumstances and to the reflection
that seeks the determination of the concept in external existence and in
pictorial representation – seeks it, that is, where it is not to be found. –
Defining, therefore, by its own doing also forfeits the true concept deter-
minations that would by essence be the principles of the subject matter,
and contents itself with marks, that is, determinations in which that they
are essential to the subject matter is a matter of indifference and whose
only purpose is rather to be markers for external reflection. – Any such
single, external determinateness is too disproportionate with respect to the
concrete totality and to the nature of its concept to justify its being singled
out or to assume that a concrete whole would find in it its true expression
and determination. – For example, as Blumenbach observes, the lobe of
the ear is something lacking to all other animals and is therefore perfectly
entitled, in accordance with ordinary ways of speaking about common and
distinguishing markers, to be used as the distinctive characteristic in the
definition of the physical human being. 49 But how disproportionate such a
totally external determination at once appears when measured against the
representation of the total habitus of the physical human being, and against
the demand that the concept determination shall be something essential!
It is entirely accidental whether the markers taken up into the definition
are pure makeshifts like this one or approximate the nature of a principle
instead. From their externality one can also see that the cognition based
on concepts did not begin with them; it was rather an obscure feeling,
an indeterminate but profound sense, an intimation of the essential that
preceded the discovery of genera in nature and spirit, and only afterwards
was a specific externality sought for the understanding. – Since in existence
the concept has entered into externality, it has unfolded into its differences
and cannot be absolutely attached to any single one of such properties. The
properties, as the externality of the thing, are external to themselves; for this
reason, as we demonstrated in the sphere of appearance in connection with
the thing of many properties, 50 do the properties essentially become even
self-subsistent matters; spirit, regarded from the standpoint of appearance,
turns into an aggregate of many independent forces. Regarded in this way,
the single property or force, even where it is posited as indifferent to the
other properties, ceases to be a characterizing principle, with the result
that the determinateness, as the determinateness of the concept, vanishes
completely.
In the concrete things, together with the diversity of the properties
among themselves, there also enters the difference between the concept and
its realization. The concept has an external presentation in nature and
spirit wherein its determinateness manifests itself as dependence on the
external, as transitoriness and inadequacy. Therefore, although an actual
thing will indeed manifest in itself what it ought to be, yet, in accordance
with the negative judgment of the concept, it may equally also show that
its actuality only imperfectly corresponds with this concept, that it is bad.
Now the definition is supposed to indicate the determinateness of the
concept in an immediate property; yet there is no property against which
an instance could not be adduced where the whole habitus indeed allows
the recognition of the concrete thing to be defined, yet the property taken
for its character shows itself to be immature and stunted. In a bad plant, a
bad animal type, a contemptible human individual, a bad state, there are
aspects of their concrete existence that are defective or entirely missing but
that might otherwise be picked out for the definition as the distinctive mark
and essential determinateness in the existence of any such concrete entity.
A bad plant, a bad animal, etc., remains a plant, an animal just the same.
If, therefore, the bad specimens are also to be covered by the definition,
then the empirical search for essential properties is ultimately frustrated,
because of the instances of malformation in which they are missing; for
instance, in the case of the physical human being, the essentiality of the
brain is missing in the instance of acephalous individuals; or, in the case
of the state, the essentiality of the protection of life and of property is
missing in the instance of despotic states and tyrannical governments. –
If the concept is maintained despite the contradicting instance and the
latter is declared, as measured by the concept, to be a bad specimen, then
the attestation of the concept is no longer based on appearance. But that
the concept stands on its own goes against the meaning of definition;
for definition is supposed to be the immediate concept, and can therefore
derive its determinations of the subject matter only from the immediacy of
existence and justify itself only in what it already finds there. – Whether its
content is in and for itself truth or contingency, this lies outside the sphere
of definition; but for this reason, because the singular subject matter under
consideration may well be a bad specimen, formal truth, or the agreement
of the concept subjectively posited in the definition and the actual subject
matter outside it, cannot be established.
The content of a definition is taken in general from immediate existence,
and because it is immediate, it has no justification; the question regarding
its necessity is precluded by its origination; by the very fact that the
definition voices the concept as something merely immediate, it renounces
comprehending it conceptually. What it exhibits, therefore, is nothing but
the form determination of the concept in a given content, without the
reflection of the concept within itself, that is, without its being-for-itself.
But immediacy proceeds as such only from mediation, and must there-
fore pass over into it. Or the determinateness of the content contained in
the definition is, for the very reason that it is determinateness, not only
immediate but something mediated by its other. Consequently definition
can apprehend its subject matter only by virtue of the opposite determina-
tion and must therefore pass over into division.

2. Division

The universal must particularize itself; to this extent, the necessity of divi-
sion lies in the universal. But because definition itself already begins with
the particular, its necessity for passing over into division lies in the particu-
lar that points, as particular, to an other. Conversely, the particular separates
itself off from the universal precisely by holding on to its determinateness
for the sake of keeping it distinct from an other than it; the universal is
therefore presupposed for division. The way to proceed is therefore this: the
singular content of definition is raised through particularity to the extreme
of universality; but universality must from now on be assumed as the objec-
tive foundation and, with it as the starting point, division presents itself as
the disjunction of the universal, the latter being the first.
A transition is now introduced which, since it takes place from the
universal to the particular, is determined by the form of the concept.
Definition is as such something singular; a greater number of definitions
pertains to a greater number of subject matters. The advance from the
universal to the particular characteristic of the concept constitutes the
basis and the possibility of a synthetic science, of a system, and of systematic
cognition.
The first requirement for this is that, as indicated, the beginning be made
with the subject matter in the form of a universal. In the realm of actuality,
whether of nature or spirit, it is the concrete singularity that is given to
subjective, natural cognition as first. But in a cognition which is a concep-
tual comprehension, at least inasmuch as it has the form of the concept
for its basis, it is the simple, abstracted from the concrete, that on the
contrary comes first, for only in this form does the subject matter have
the form of a self-referring universality and of an immediacy that accords
with the concept. It may perhaps be objected to this way of proceeding
in matters scientific that, since intuition is easier than cognition, what can
be intuited, that is, concrete actuality, should be made the starting point
of science; that this way of proceeding would be more natural than one
that starts from an abstract subject matter and then proceeds from it to its
particularization and concrete singularization. – But inasmuch as cognition
is the issue, any comparison with intuition has already been decided and
dismissed; the only question allowed here is what should be the first inside
cognition, and how one should then go from there; what is required is not
a method appropriate to nature but one appropriate to cognition. – If the
issue is merely one of easiness, then it goes without saying that it is easier
for cognition to grasp the abstract simple thought determination than to
grasp a concrete subject matter which is a complex web of such thought
determinations and relations; and it is in this manner, no longer as it is
given in intuition, that the matter should be grasped. The universal is in
and for itself the first moment of the concept, because it is the simple,
and the particular only comes after it, because it is the mediated; and
conversely the simple is the more universal, and the concrete, since it is
internally differentiated and hence mediated, is what already presupposes
the transition from a first. – This remark applies not only to the ordering
of the whole into the specific forms of definitions, divisions, and propo-
sitions, but also to the ordering of cognition as a whole and simply with
respect to the difference of abstract and concrete in general. – Thus in
learning to read for example, the more rational way to begin is also not
with reading whole words or even syllables, but with the elements of words
and syllables and with the signs of abstract sound inflections; in alphabetic
script the analysis of concrete words into their abstract sound inflections
and their signs is already accomplished, and for this reason learning to read
is a primary occupation with abstractions. In geometry, the beginning has
to be made not with a concrete space configuration but with point and line
and then with plane figures, and among the latter not with polygons but
with the triangle, and among curves with the circle. In physics, the singular
natural properties or matters must be freed from the manifold entangle-
ments in which they are found in concrete actuality, and presented with
their simple, necessary conditions; they too, like space configurations, are
accessible to intuition, but this intuition has to be prepared for, in such a
way that they finally appear and are held free of all the modifications due
to circumstances which are extraneous to their specific nature. Magnetism,
electricity, various kinds of gases, etc., are such subject matters as come
to be known in their specificity only by being apprehended as removed
from the concrete conditions under which they appear in actuality. Of
course, an experiment will exhibit them to intuition in some concrete case;
but, in order for the experiment to be scientific, it must admit only such
conditions as are necessary to it, and it must be repeated in various forms
in order to demonstrate that the concrete cluster of conditions that cannot
be separated from the matters under consideration are inessential, for these
can appear in one concrete configuration and then again in another, so that
only their abstract form is left for cognition. – To mention yet one more
example, it might appear natural and reasonable to regard colors, first, in
the concrete appearance of the animal subjective sense; next, as a spectral
phenomenon hovering outside the subject, and finally fixed in objects in
external actuality. But for cognition, the universal and consequently truly
primary form is the middle one of the three mentioned, color as it hovers
between subjectivity and objectivity in the well-known form of the spec-
trum, still unentangled with subjective and objective circumstances. For
the pure consideration of the nature of this object such circumstances are
at first only a source of interference, because they behave as efficient causes
and therefore make it uncertain whether specific alterations of colors, spe-
cific transitions and relations, are based on their own specific nature as color
or are rather to be attributed to the specific pathology of the circumstances
themselves, to the healthy and sick particular affections and effects of the
organs of the subject, or to the chemical, vegetable, and animal forces of the
objects. – Numerous other examples could be adduced from the cognition
of organic nature and of the world of spirit; everywhere the abstract must
constitute the starting point and the element in which and from which the
particularities and rich shapes of the concrete spread out.
Now, although with division or with the particular the distinction of
universal and particular is duly introduced, this universal is nevertheless
itself already something determinate and therefore itself only a member
of a division. Hence there is a higher universal for it, and a higher yet
for this other universal, and the same for the next all the way to infinity.
There is no immanent limit to the cognition under consideration here,
because it proceeds from a given and it is the form of abstract universality
that defines its “first.” Any object, therefore, that seems to possess an
elementary universality is made the subject matter of a specific science:
it makes an absolute beginning because ordinary acquaintance with it is
presupposed, and the assumption made is that it stands on its own with no
need of derivation. Definition takes it as an immediate.
Division is the immediately next step after this starting point. For this
advance, only an immanent principle would be required, that is, a begin-
ning from the universal and the concept; but the cognition under consid-
eration here lacks any such principle, for it follows only upon the form
determination of the concept without that form’s immanent reflection,
and therefore takes the determinateness of the content from what is given.
There is no specific reason for the particular that enters into the divi-
sion, whether with respect to what constitutes the basis of the division, or
with respect to the specific relation that the members of the disjunction
are supposed to have to one another. Consequently, in this respect the
business of cognition can only consist partly in orderly arranging the par-
ticularities found in the empirical material, and partly also in discovering
universal determinations of this material by means of comparison. Such
determinations then count as grounds of division, of which there can be
a variety, just as there can be an equal variety of divisions based on them.
The relation of the members of a division to one another, the relation
of the species, has only this one universal determination, namely that the
members, the species, are determined relative to one another in accordance
with the assumed ground of division; if their differentiation were to rest on
some other consideration, their order would be arranged along different
lines accordingly.
Because of the lack of a principle of self-determination, the only possible
laws for this business of division consist in formal, empty rules that lead
nowhere. – Thus we see laid down as a rule that division should exhaust the
concept; but in fact each single member of the division must exhaust the
concept. Actually, what is meant is that the determinateness of the concept
should be exhausted; but there is nothing in an empirical manifold of
species, internally void of determination, that contributes to this exhaustion
whether few or many of them have been discovered; it is indifferent, for
example, to the exhaustion of the concept whether in addition to the
sixty-seven species of parrots another dozen are discovered. The demand
for exhaustion can only mean this tautological proposition, that all the
species should be listed in their completeness. – Now as empirical cognition
expands, it can well happen that species are discovered that do not fit the
assumed determination of the genus, for the genus is usually adopted more
in accordance with some obscure representation of the whole habitus than
in accordance with the more or less singular mark that should expressly
serve to determine it. – In such a case, the genus would have to be altered
and a justification would have to be given for regarding another group of
species as the species of the one new genus; that is to say, the genus would
receive its determination from what we group together on the basis of
some standpoint or other that we choose to assume as a principle of unity;
this standpoint thus becomes itself the ground of division. Conversely,
should we hold on to the determinateness originally assumed to define the
genus, then the material that we wanted to bring as species in unity with the
earlier ones, would have to be excluded. This way of carrying on without the
concept – at one time by assuming a certain determinateness as the essential
moment of the genus, subordinating the particulars to it or excluding them
from it accordingly; at another time by starting with the particulars and
letting oneself be guided in grouping them by some other determinateness –
this way gives the appearance of a game in which it is left up to chance to
decide which part or which side of the concrete to fix on in order then to
bring order in accordance with it. – It is physical nature itself that presents
such a contingency in the principles of division; because of the external
dependency of its actuality, it stands in a manifold of connectedness which
for it is likewise given; there is, therefore, an assortment of principles to
which it has to adapt itself, following one principle in one series of its
forms but another in another series, while also producing hybrids that go
in different directions at once. Thus it happens that in one series of natural
things certain marks come to the fore as especially significant and essential
that in another series become inconspicuous and purposeless, the result
being that it is impossible to abide by any such principle of division.
The general determinateness of empirical species can only consist in this:
that they are simply diverse from one another without being opposed. The
disjunction of the concept was presented earlier in a determinate form; 51
if particularity is taken without the negative unity of the concept, as a
particularity which is immediate and given, then difference stays at only
the reflective form of diversity considered earlier. The externality in which
the concept is pre-eminently to be found in nature brings with it the total
indifference of difference; for this reason it is common to take from number
the determination for division.
Such is the contingency here of the particular with respect to the uni-
versal, and therefore of division in general, that it may be attributed to
an instinct of reason when we discover in this cognition bases of division
and divisions which, to the extent that sensuous properties allow it, show
themselves to be more adequate to the concept. For example, in the case
of animals, the instruments for eating, the teeth and the claws, are used
in the systems as a far-reaching criterion of division. They may be taken
at first only as features in which it is easier, for the subjective purpose of
cognition, to detect distinguishing marks. But in fact the differentiation
embodied in those organs is not one that pertains just to external reflection;
such organs are rather the vital point of animal individuality, where the
latter posits itself as self-referring singularity by cutting itself loose from
the otherness of its external nature and from continuity with the other. –
In the case of the plant, it is the reproductive parts that constitute the
highest point of vegetable life, the point at which the plant points to the
transition into sexual difference and thereby into individual singularity. For
this reason the system has rightly turned to this point for a base of division
which, though not sufficient, is nonetheless far-reaching, and has thereby
laid down for foundation a determinateness which is such not merely for
external reflection, for the purpose of comparison, but is in and for itself
the highest of which the plant is capable.

3. The theorem

1. The third stage in this advance of cognition based on concept determi-
nations is the transition of particularity to singularity; this stage constitutes
the content of the theorem. To be considered here, therefore, is the self-
referring determinateness, the internal differentiation of the subject matter
and the connection of the differentiated determinacies to one another. Defi-
nition contains only one determinateness, division contains determinateness
as against the other; in singularization the subject matter has parted inter-
nally. Whereas definition stops at the universal concept, in theorems the
subject matter is known in its reality, in the conditions and the forms of its
real existence. Together with the definition, therefore, the subject matter
exhibits the idea, which is the unity of the concept and reality. But the
cognition being considered here, a cognition that is still a seeking, does not
attain this presentation, for in it reality does not proceed from the concept,
and therefore the dependency of reality on the concept and consequently
the unity itself is not cognized.
Now according to the definition just given, the theorem is the properly
synthetic element of a subject matter, because the relations of its determina-
cies are necessary, that is, are grounded in the inner identity of the concept.
In definition and division the synthetic element is a connectedness held
together externally; what is found given is brought into the form of the
concept, but, as given, the entire content is only displayed; in the theorem,
on the contrary, it ought to be demonstratively displayed. Since this cogni-
tion does not deduce the content of its definitions and of the principles of
division, it seems that it might also spare itself the proof of the relations
expressed by the theorem and be satisfied here too just with perception. But
what distinguishes cognition from mere perception and representation is
the form of the concept in general that it imparts to the content; this is done
in definition and division; but since the content of the theorem proceeds
from the concept’s moment of singularity, it consists in determinations of
reality that no longer have as their relation just the simple and immediate
determinations of the concept; in singularity the concept has gone over to
otherness, to reality, thereby becoming idea. The synthesis contained in the
theorem no longer has, therefore, the form of the concept to justify it; it
is a joining together of such as are diverse; consequently, the unity not yet
thereby posited still remains to be demonstrated; here proof thus becomes
necessary to this cognition itself.
Now the first difficulty that we encounter here is of distinctly distin-
guishing between which of the determinations of the subject matter can be
admitted into the definitions, and which must rather be relegated to the
theorems. In this matter, there is no principle readily available. There might
seem to be one in assuming, perhaps, that what pertains to a subject matter
immediately also belongs to the definition, while for the rest, since it is
something mediated, the mediation must first be demonstrated. But the
content of the definition is as such determined and therefore itself essen-
tially mediated; its immediacy is only a subjective one, that is to say, the
subject makes an arbitrary beginning letting a subject matter count as a
presupposition. Now since this subject matter is as such in itself concrete,
and must also be divided, the result is a number of determinations that
are mediated by nature and are taken as immediate and unproved, not on
the basis of any principle, but only subjectively. – Even in Euclid, who has
always been justly recognized to be the master of this synthetic kind of
cognition, we find under the name of axiom a presupposition about parallel
lines that some have held to be in need of proof, and various attempts have
been made to fill this lack. And there are several other theorems in which
people have thought to have discovered presuppositions that should not
have been immediately assumed but ought to have been proved. So far
as the axiom of parallel lines is concerned, it may be noted that precisely
there is where Euclid gives evidence of his good sense; he had duly appre-
ciated the nature of his science. The proof of that axiom would have to be
derived from the concept of parallel lines; any such proof, however, has no
place in his science, no more than does the deduction of his definitions,
of his axioms and his subject matter in general, of space itself and its first
determinations, the dimensions. For any such deduction would have to be
carried out on the basis of the concept, and this lies outside the proper
domain of Euclidean science; these must remain for it, therefore, necessary
presuppositions, relative firsts.
Axioms, to take this opportunity of mentioning them, belong to the
same class. They are commonly but wrongly taken to be absolute firsts, as
if they were not in need of proof in and for themselves. If such were in
fact the case, they would then be mere tautologies, for it is only in abstract
identity that there is no diversity and that therefore also no mediation is
needed. But if the axioms are more than just tautologies, then they are
propositions drawn from another science, since within the science for which
they serve as axioms they are meant as presuppositions. Strictly speaking,
therefore, they are theorems, and are indeed mostly drawn from logic. The
axioms of geometry are lemmas of this kind, logical propositions, and they
come close, moreover, to being tautologies because they are concerned with
quantity alone and every qualitative difference has therefore been purged
from it. Of the principal axiom, the purely quantitative syllogism, we spoke
earlier. 52 – Axioms, therefore, when considered in and for themselves, are
just as much in need of proof as are definitions and divisions, and they
are not made into theorems only for the reason that, since they are relative
firsts, they are assumed for a certain standpoint as presuppositions.
As regards the content of theorems, there is one further precision to be
made. Because this content consists in a connection of determinacies of
the concept’s reality, such connections may be more or less incomplete
and single relations of the subject matter, or, on the contrary, they may
be one such relation that encompasses the whole content of reality and
expresses the content’s determinate connection. But the unity of all the
content determinacies is equivalent to the concept; a proposition that contains
them is therefore itself a definition again – not one, however, that expresses
the concept only as immediately assumed, but one that expresses it rather
as developed into its determinate, real differences, or one that expresses
the concept’s complete existence. The two together, therefore, present the
idea.
If we closely compare the theorems of a synthetic science, of geometry
in particular, we find this distinction, namely that some of the science’s
theorems contain only singular relations of the subject matter, whereas
others contain relations in which its full determinateness is expressed. It
is a very superficial view that attributes equal value to all propositions, on
the ground that each generally contains a truth, and in the formal progres-
sion from step to step of a proof each is equally essential. 53 The difference
with respect to the content of theorems is most intimately connected with
this progression itself, and some further remarks concerning the latter will
serve to clarify that difference further, as well as the nature of synthetic
cognition. To start with, Euclidean geometry (which, as the representa-
tive of the synthetic method of which it delivers the most accomplished
exemplar, shall serve as an example) has from ancient times been renowned
for the order in the progression of its theorems, whereby for each theo-
rem the propositions required for its construction and proof are always
found already proved. This circumstance concerns the formal sequence
of inference; yet, important as it is, it still has primarily to do with the
external ordering of purposiveness, and bears on its own no connection
to the essential difference of the concept and the idea in which there lies
a higher principle of the necessity of the progression. – That is to say,
the definitions with which the beginning is made apprehend the sensuous
subject matter as immediately given, and they determine it according to its
proximate genus and specific difference; 54 these are equally the simple and
immediate determinacies of the concept, the universality and particularity
whose relation is developed no further. Now the initial theorems themselves
have nothing at their disposal except such immediate determinations as are
found in the definitions; similarly their reciprocal dependence can only be
a matter at first of each being in general determined through the other.
Thus Euclid’s first propositions regarding the triangle have to do only with
congruence, that is, how many parts must be determined in a triangle in order
that the remaining parts of the one and same triangle, or the whole of it, be
determined in full. 55 That two triangles are compared with one another and
their congruence posited in the fact that they coincide is a detour needed
by a method that must rely on sensuous coincidence instead of the thought
of determinateness. Otherwise considered for themselves, these theorems
themselves contain two parts, one of which can be regarded as the concept,
and the other as the reality that completes the concept by realizing it. For the
understanding, whatever suffices for a full determination, that is to say, the
two sides and the enclosed angle in this case, is already the entire triangle;
nothing further is needed for its complete determinateness; the remaining
two angles and the third side are the superfluity of reality over and above
the determinateness of the concept. Thus all that those theorems in fact
do is to reduce the sensuous triangle, which of course requires three sides
and three angles, to its simplest conditions. The definition had mentioned,
quite in general, only the three lines that enclose the plane figure and
make it a triangle; it is a theorem that first expresses the determinateness of
the angles through the determinateness of the sides, just as the remaining
theorems express the dependence of three other parts on three others. –
But the complete determinateness of the magnitude of a triangle in terms
of its sides is contained in the Pythagorean theorem; it is in this theorem
that we first have the equation of the sides of the triangle, for the preceding
sides 56 bring the triangle to a reciprocal determinateness of part to part only
in general, not to an equation. This proposition is therefore the perfect,
real definition of the triangle – of the right-angled triangle in the first place,
the simplest in its differences and hence the most regular. – Euclid brings
the first book to a close with this proposition, for it does in fact attain a
perfect determinateness. And after he has reduced to a uniform type those
triangles which are not right-angles and are affected by greater inequality,
he concludes the second book with the reduction of the rectangle to the
square, with an equation between the self-equal, or the square, and the
internally unequal, or the rectangle; similarly, in the Pythagorean theorem,
the hypotenuse that corresponds to the right-angle, the self-equal, con-
stitutes one side of the equation, while the other side is made up by the
self-unequal, the two perpendicular sides. The equation between the square
and the rectangle is at the basis of the second definition of the circle, and
this is again the Pythagorean theorem, except that the two perpendicular
sides of the right-angle are assumed to be alterable; the first equation of the
circle is in precisely the relationship of sensuous determinateness to equation
as holds between the two different definitions of conic sections in general.
This truly synthetic progression is a transition from universal to singu-
larity, namely to that which is determined in and for itself, or to the unity of
the subject matter in itself inasmuch as this has come apart, differentiated
into its essential real determinacies. In other science, however, the common
and quite imperfect way of advancing from universality to singularity is
indeed to start from a universal, but then to singularize and concretize it
by applying it to a material brought in from elsewhere; in this way, the
singularity of the idea is strictly speaking an empirical addition.
Now whatever the content of the theorem, whether imperfect or perfect,
it must be proved. It is a relation of real determinations that do not have
the relation of concept determinations; when they do have this relation,
as it can be shown to be the case for the propositions we have called
second or real definitions, such definitions are for that very reason in one
respect definitions; but since their content consists at the same time of
relations of real determinations, not just of the relation of universal and
simple determinateness, in comparison with such a first definition they
are also in need and capable of proof. As real determinacies, they have the
form of indifferent subsistence and indifferent diversity; hence they are not
immediately one and therefore their mediation is to be demonstrated. The
immediate unity in the first definition is the one in accordance with which
the particular is in the universal.
2. Now the mediation which we must now consider more closely may be
simple or may go through several mediations. The mediating members are
joined together with those to be mediated; but since it is not on the basis of
the concept, to which the transition into an opposite is altogether alien, that
the mediation and the theorem are retraced in this cognition, in the absence
of any concept of connectedness, the mediating determinations must be
imported from somewhere as a provisory material for the scaffolding of the
proof. This preparation is the construction.
Now among the connections of the content of the theorem, of which
there can be a great number, only those must be adduced and made
to work that are of service to the proof. The supply of material only
acquires meaning in this context; in itself it appears blind and meaningless.
In retrospect it will of course become apparent in the proof that there
was a purpose to drawing, for example, such or such additional lines to
a geometrical figure as the construction specifies; in the course of the
construction itself, however, this must be done blindly; by itself, therefore,
this operation is without understanding, since the purpose motivating it is
yet to be declared. – It is a matter of indifference whether the operation is
undertaken for the purpose of a theorem in the strict sense or a problem;
as it first appears before the proof, the operation is not anything derived
from the given specification of the theorem or the problem – a meaningless
act, therefore, for anyone as yet not acquainted with its purpose, and then
always only directed by an external purpose.
This hidden purpose becomes apparent in the proof. This contains, as
stated, 57 the mediation of what the theorem declares as bound together, and
it is only by virtue of this mediation that this connectedness first appears
as necessary. Just as the construction lacks on its own the subjectivity of the
concept, so is the proof a subjective act lacking in objectivity. For since the
content determinations of the theorem are not posited at the same time as
determinations of the concept, but are posited instead as indifferent parts
standing in a multitude of external relations to one another, it is only in
the formal, external concept that the necessity manifests itself. The proof
is not a genesis of the relation that constitutes the content of the theorem;
the necessity is present only to insight, and the whole proof is only for
the subjective interest of cognition. It is for this reason a thoroughly external
reflection that proceeds from the outside to the inside, that is, arrives in
conclusion at the inner constitution of the relation on the basis of external
circumstances. These circumstances, which the construction has presented,
are a consequence of the nature of the subject matter; here they are converted
instead into the ground and the mediating relations. The middle term, the
third term in which the terms linked in the theorem present themselves
in their unity and that provides the nerve of the proof, is therefore only
something in which the connectedness appears and is external. Because the
sequence which the proof goes through is rather the reverse of the nature of
the fact, what is considered in the proof as ground is a subjective ground,
one that brings out the nature of the fact only for cognition.
The foregoing considerations make clear the necessary limit of this cog-
nition, a limit that usually goes unrecognized. The science of geometry is the
most illustrious example of the synthetic method – but it has been inappro-
priately applied to other sciences as well, even to philosophy. Geometry is
a science of magnitude; hence formal inference is the one most appropriate
to it; since it treats the quantitative determination alone, abstracting from
anything qualitative, it can confine itself to formal identity, to the unity void
of concept which is equality and belongs to external, abstractive reflection.
The determinations of space that are its subject matter are already abstract
objects, suitably prepared for the purpose of obtaining a perfectly finite,
external determinateness. This science, because of its abstract subject mat-
ter, on the one hand has an aura of sublimity about it, for in these empty
silent spaces color is extinguished and the other sensuous properties have
equally vanished, and further, every other interest that would appeal to a
living individuality is silenced. On the other hand, this abstract subject
matter is still space, a non-sensuous sensuous. To be sure, intuition is raised
to a higher level in this abstraction; space is now a form of intuition,
but it is still intuition – sensuous intuition, the externality 58 of the senses
themselves, 59 their pure absence of concept. – Enough has been heard lately
of the pre-eminence of geometry in this respect. There are those who say
that geometry’s foremost advantage is that it is based on the intuition of
the senses, even believe that its scientific pre-eminence depends on this
circumstance and that its proofs rest on intuition. This shallow view must
be countered with the plain reminder that no science can be brought about
by intuition, but only by thought. The intuitive character that geometry
possesses because of its still sensuous material only gives to it that level of
evidence that the senses generally provide to thoughtless spirit. It is therefore
regrettable that the same sensuousness of material which is a sign of the
inferiority of its standpoint has been reckoned instead to its advantage. It is
solely to the abstraction of its sensuous subject matter that geometry owes
its aptitude for a higher scientific reach and the advantage that it has over
the collections of information that people are also want to call sciences but
have for content only the concrete perceptible material of the senses, and
only because of the order that they seek to bring to it do they give any sign
of a remote inkling and hint of the requirements of the concept.
It is only because the space of geometry is the abstract emptiness of
externality that it is possible for figures to be drawn in its indeterminateness
in such a way that their determinations remain perfectly at rest outside one
another with no immanent transition to the opposite. The science of these
figures is therefore plainly and simply the science of the finite which is
compared according to magnitude and has for its unity the external one of
equality. But now, since with these figures the start is made from a variety of
sides and points at once, and the various figures fall into place of themselves,
in comparing them their qualitative unlikeness and incommensurability also
come into view. Geometry is thus driven, beyond the finitude within which
it advanced step by step orderly and securely, to infinity – to the positing
as equal of such as are qualitatively diverse. Here it loses the evidence that
it derived from being otherwise based on fixed finitude without having
to deal with the concept and the transition to the opposite which is its
manifestation. As a finite science, geometry reaches its limit at this point,
for the necessity and the mediation of the synthetic realm is no longer
grounded in merely positive identity, but in negative identity.
If geometry, like algebra, quickly runs up against its limit with its abstract
subject matter, suited as this is only to the understanding, it is evident from
the start that the synthetic method is all the more insufficient for other sci-
ences, and most insufficient of all for philosophy. Regarding definition and
division, we have already made the relevant points, and we should be left
here to speak only of theorems and proofs. But, besides the presupposi-
tion of definition and division that itself requires proof and presupposes
it, also problematic is the very position of definition and division with
respect to the theorems. This position is especially noteworthy in the
empirical sciences, as for example physics, whenever they want to give
themselves the form of synthetic sciences. This is how they go about it.
The reflective determinations of particular forces, or of otherwise inner and
essential forms, which are the results of an analysis of experience and can be
justified only as such results, must be placed at the top, in order to obtain
from them a general foundation that can then be applied to the singular
and be instantiated there. Since these general foundations have no hold of
their own, we must simply grant them in the meantime; it is only in the
derived consequences that we notice that the latter are in fact the ground
of those presuppositions. The so-called explanation, and the proof of the
concrete brought into theorems, turn out to be partly a tautology, partly
an obfuscation of the true relation, and partly also an obfuscation that
serves to hide the deception of cognition. For cognition has collected expe-
riences tendentiously, only so that it could attain its simple definitions and
principles; and it has pre-empted the possibility of empirical refutation
by taking experiences and accepting them as valid, not in their concrete
totality but selectively, as examples that can then be used on behalf of its
hypotheses and theories. In this subordination of concrete experience to
presupposed determinations, the foundation of the theory is obscured and
is only indicated according to the side that suits the theory; and, quite in
general, the unprejudiced examination of concrete perceptions for their
own sake is thereby much impeded. Only by turning the whole procedure
upside down does the whole thing acquire the right relation in which the
link of ground and consequence can come into view. One of the principal
obstacles in the study of these sciences is thus the way we enter into them,
which we can only do by blindly taking the presuppositions for granted and,
without being able to form any further concept of them, often not even
an exact representation, at best by conjuring up in phantasy a confused
picture of them, we right there impress in our memory the determinations
of the forces and matters that we have assumed, their hypothetical shapes,
their directions and rotations. If we are asked to produce the necessity and
the concept of these assumptions in order to justify assuming their validity,
we discover that we are incapable of making a step beyond the starting
point.
We had occasion above to speak of the inappropriateness of applying the
synthetic method to strictly analytic science. 60 Wolff extended this applica-
tion to every kind of bits of knowledge that he dragged into philosophy and
mathematics – cognitions which were partly of a wholly analytical nature,
and partly also devoted to practical matters of an incidental kind. The
incongruity between this material, easy to grasp and by nature incapable of
rigorous and scientific treatment, and the pompous scientific roundabouts
icing it, has alone demonstrated the clumsiness of such an application,
finally discrediting it. s Yet, this misuse has not sufficed to shake the belief
that this method is both suited and essential to attaining scientific rigor
in philosophy. Spinoza’s example, the way he presented his philosophy, has
long served as model in this regard. But the fact is that Kant and Jacobi
did do away with this whole style of the previous metaphysics and its
method along with it. As for the content of that metaphysics, Kant has
in his own fashion shown that it leads by strict demonstration to anti-
nomies, the same whose nature we have in other respects elucidated at the
appropriate places. 63 But Kant did not reflect on the nature of the demon-
stration associated with them, on the fact that such a demonstration is
inextricably bound to a finite content. In his Principles of Natural Science,
he gave himself an example of how to deal with a science of reflection on
its own methodological terms, in a way that he thought would vindicate
it for philosophy. 64 – While Kant attacked previous metaphysics for the
most part from the side of its content, Jacobi did it especially from the side
of its method of demonstration and, with great clarity and profundity, he
put his finger on precisely the point at issue, namely that such a method
of demonstration is strictly bound to the cycle of rigid necessity of finite
reality, and that freedom, that is, the concept and with it everything that truly
exists, lies beyond it and is unattainable by it. 65 – According to Kant’s result,
For example, Wolff’s First Principles of Architecture, the Eighth Theorem, runs as follows:
Theorem: A window must be wide enough that two persons can comfortably stand at it side by side.
Proof: For it is common for one to stand at a window with somebody else to view the outside.
Now since the fabric architect should in every respect satisfy the main intentions of the owner of the
fabric (§1), he must also construct the window in such a manner that two persons can comfortably
stand at it side by side. Q.E.D. 61
From the same author’s Principles of Fortification, the Second Theorem:
Theorem: If the enemy encamps in the vicinity, and the attempt at a rush to relieve the siege
is expected, a line of circumvallation must be drawn around the whole siege. Proof: Lines of
circumvallation prevent anyone from penetrating into an area from outside (§311). But those who
wish to relieve the area of siege will attempt to penetrate the area from outside. If, therefore, one
wants to keep them out, a line of circumvallation must be drawn around the area. Therefore, if the
enemy encamps in the vicinity and the attempt at a rush to relieve the siege is expected, the area
must be enclosed in lines of circumvallation. Q.E.D.
it is the peculiar content of metaphysics that leads it into contradictions;
the inadequacy of cognition is due to its subjectivity. Jacobi’s result is that
the inadequacy is due instead to the method and the whole nature of cog-
nition itself that only grasps a concatenation of conditions and dependency
and therefore proves itself inadequate to what exists in and for itself, to
what is absolutely true. And in fact, since the principle of philosophy is
the infinite free concept and all its content rests on that alone, the method
suited to a finitude empty of concept is inadequate to it. The synthesis and
the mediation of this method, the process of proving, goes no further than a
necessity which is opposed to freedom, that is, an identity of the dependent
which is only implicit, whether it is apprehended as internal or as external,
and in which that which in it constitutes reality, the differentiation that
has emerged in concrete existence, remains simply self-subsistent diversity
and therefore something finite. 66 In this reality, therefore, this identity does
not itself attain concrete existence but remains only internal, or again, is only
external, because its determinate content is given to it. Either way, whether
internal or external, the identity is something abstract that does not possess
within it the side of reality, is not posited as determinate identity in and for
itself; therefore the concept, which alone is the issue here and which is the
infinite in and for itself, is precluded from this cognition.
In synthetic cognition, therefore, the idea achieves its purpose only to the
extent that the concept becomes for the concept according to its moments of
identity and real determinations, or of universal and particular differences –
further also as an identity which is connectedness and dependence in diversity.
But this, its subject matter, is not adequate to the concept; for in it, in this
subject matter or in its reality, the concept does not come to be the unity
of itself with itself; in necessity its identity is for it, but in this identity the
necessity is not itself the determinateness but is on the contrary a material
external to it, that is to say, is not determined by the concept and the
concept, therefore, does not recognize itself in it. Thus in general the
concept is not for itself, is not at the same time determined in and for itself
according to its unity. For this reason the idea does not as yet attain the
truth in this cognition: it does not because of the disproportion between
subject matter and subjective concept. – But the sphere of necessity is the
highest point of being and reflection; of itself, in and for itself, it passes
over into the freedom of the concept, inner identity passes over into its
manifestation which is the concept as concept. How this transition from
the sphere of necessity to the concept occurs in itself has been shown
when considering necessity, 67 and we saw it also at the beginning of this
Book as the genesis of the concept. 68 In the present context, necessity has the
position of being the reality or the subject matter of the concept, just as the
concept into which it passes is now the concept’s subject matter. But the
transition itself is the same. Here, too, it is at first only in itself, still lying
in our reflection outside cognition, that is, itself still the inner necessity
of cognition. Only the result is for cognition. The idea, in so far as the
concept is now for itself determined in and for itself, is the practical idea,
action.

B. THE IDEA OF THE GOOD

Inasmuch as the concept, which is its own subject matter, is determined in
and for itself, the subject is determined as singular. As subjective it again has
an implicit otherness for its presupposition; it is the impulse to realize itself,
the purpose that on its own wants to give itself objectivity in the objective
world and realize itself. In the theoretical idea the subjective concept, as a
universal that in and for itself lacks determination, stands opposed to the
objective world from which it derives determinate content and filling. But
in the practical idea it is as actual that it stands over against the actual;
but the certainty of itself that the subject possesses in being determined
in and for itself is a certainty of its actuality and of the non-actuality of
the world; it is the singularity of this world, and the determinateness of its
singularity, not just its otherness as abstract universality, which is a nullity
for the subject. The subject has here vindicated objectivity for itself; its
inner determinateness is the objective, for it is the universality which is just
as much absolutely determined; the previously objective world is on the
contrary only something still posited, an immediate which is determined in
a multitude of ways but which, because it is only immediately determined,
in itself eludes the unity of the concept and is of itself a nullity.
This determinateness which is in the concept, is equal to the concept,
and entails a demand for singular external actuality, is the good. It comes
on the scene with the dignity of being absolute, because it is intrinsically
the totality of the concept, the objective which is at the same time in the
form of free unity and subjectivity. This idea is superior to the idea of
cognition just considered, for it has not only the value of the universal but
also of the absolutely actual. – It is impulse, in so far as this actual is still
subjective, self-positing, without at the same time the form of immediate
presupposition; its impulse to realize itself is not, strictly speaking, to give
itself objectivity, for this it possesses within itself, but to give itself only
this empty form of immediacy. – The activity of purpose, therefore, is
not directed at itself, is not a matter of letting in a given determination
and making it its own, but of positing rather its own determination and,
by means of sublating the determinations of the external world, giving
itself reality in the form of external actuality. – The idea of the will as a
self-determining explicitly possesses content within itself. Now this content
is indeed a determinate content, and to this extent finite and restricted;
self-determination is essentially particularization, since the reflection of
the will is in itself, as negative unity as such, also singularity in the sense
that it excludes an other while presupposing it. Yet the particularity of the
content is at first infinite by virtue of the form of the concept, of which
it is the proper determinateness, and which in that content possesses its
negative self-identity, and consequently not only a particularity but its
infinite singularity. The mentioned finitude of the content in the practical
idea only means, therefore, that the idea is at first not yet realized; the
concept is for the content that which exists in and for itself; it is here the
idea in the form of objectivity existing for itself; on the one hand, the
subjective is for this reason no longer just something posited, arbitrary or
accidental, but is an absolute; but, on the other hand, this form of concrete
existence, this being-for-itself, does not as yet have the form of the being-in-
itself. Thus what from the side of the form as such appears as opposition,
appears in the form of the concept reflected into simple identity, that is,
appears in the content as its simple determinateness; the good, although
valid in and for itself, is thereby a certain particular purpose, but not one
that first receives its truth by being realized; on the contrary, it is for itself
already the true.
The syllogism of immediate realization does not itself require closer
exposition here; it is none other than the previously considered syllogism
of external purposiveness; 69 only the content constitutes the difference. In
external as in formal purposiveness it was an indeterminate finite content
in general; here, though also finite, it is as such at the same time absolutely
valid. But in regard to the conclusion, the realized purpose, a further dif-
ference enters in. In being realized the finite purpose still attains only the
status of a means; since it is not a purpose determined in and for itself
already from the beginning, as realized it also remains something that does
not exist in and for itself. If the good is again also fixed as something finite,
and is essentially such, then, notwithstanding its inner infinity, it too can-
not escape the fate of finitude – a fate that manifests itself in several forms.
The realized good is good by virtue of what it already is in the subjective
purpose, in its idea; the realization gives it an external existence, but since
this existence has only the status of an externality which is in and for itself
null, what is good in it has attained only an accidental, fragile existence,
not a realization corresponding to the idea. – Further, since this good is
restricted in content, there are several kinds of it; in concrete existence a
good is subject to destruction not only due to external contingency and
to evil, but also because of collision and conflict in the good itself. From
the side of the objective world presupposed for it (in the presupposition
of which consists the subjectivity and the finitude of the good, and which
as a distinct world runs its own course), the realization itself of the good
is exposed to obstacles, indeed, might even be made impossible. The good
thus remains an ought; it exists in and for itself, but being, as the ultimate
abstract immediacy, remains over against it also determined as a non-being.
The idea of the fulfilled good is indeed an absolute postulate, but no more
than a postulate, that is, the absolute encumbered with the determinate-
ness of subjectivity. There still are two worlds in opposition, one a realm of
subjectivity in the pure spaces of transparent thought, the other a realm of
objectivity in the element of an externally manifold actuality, an impervious
realm of darkness. The complete development of this unresolved contra-
diction, between that absolute purpose and the restriction of this reality that
stands opposed to it, has been examined in detail in the Phenomenology
of Spirit (pp. 323ff.). – Inasmuch as the idea has within it the moment
of complete determinateness, the other concept to which the concept in
it relates possesses in its subjectivity at the same time the moment of an
object; consequently the idea enters here into the shape of self-consciousness,
and in this one respect coincides with its exposition.
But what the practical idea still lacks is the moment of real consciousness
itself, namely that the moment of actuality in the concept would have
attained for itself the determination of external being. – This lack can also
be regarded in this way, namely that the practical idea still lacks the moment
of the theoretical idea. That is to say, in the latter there stands on the side
of the subjective concept – the concept that is in process of being intuited
in itself by the concept – only the determination of universality; cognition
only knows itself as apprehension, as the identity of the concept with
itself which, for itself, is indeterminate; the filling, that is, the objectivity
determined in and for itself, is for this identity a given; what truly exists is for
it the actuality present there independently of any subjective positing. For
the practical idea, on the contrary, this actuality constantly confronting it
as an insuperable restriction is in and for itself a nullity that ought to receive
its true determination and intrinsic value only through the purposes of the
good. It is the will, therefore, that alone stands in the way of attaining its
goal, because it separates itself from cognition and because for it external
actuality does not receive the form of a true existence. The idea of the good
can therefore find its completion only in the idea of the true.
But it makes this transition through itself. In the syllogism of action, one
premise is the immediate reference of the good purpose to the actuality which
it appropriates and which, in the second premise, it directs as external
means against the external actuality. The good is for the subjective concept
the objective; actuality confronts it in existence as an insuperable restriction
only in so far as it still has the determination of immediate existence, not
of something objective in the sense that it is being in and for itself; it is
rather either the evil or the indifferent, the merely determinable, whose
worth does not lie within it. But this abstract being that confronts the
good in the second premise has already been sublated by the practical idea
itself; the first premise of this idea’s action is the immediate objectivity of
the concept, according to which purpose is communicated to actuality
without any resistance and is in the simple connection of identity with it.
To this extent, therefore, what remains is to bring together the thoughts of
the two premises of the practical idea. All that is added to what is already
accomplished in the first premise by the objective concept is that in the
second it is posited by way of mediation, hence for it. Just as in purposive
connection in general, where the realized purpose is again only a means
but the means is conversely also the realized purpose, so too now in the
syllogism of the good the second premise is already immediately present in
the first in itself, except that this immediacy is not sufficient and the second
premise is for the first already postulated – the realization of the good in the
face of another actuality confronting it is the mediation which is essentially
necessary for the immediate connection and consummation of the good.
For the first premise is only the first negation or the otherness of the concept,
an objectivity that would be a state of immersion of the concept into
externality; the second premise is the sublation of this otherness, whereby
the immediate realization of the purpose first becomes the actuality of
the good as concept existing for itself, for in that actuality the concept is
posited as identical with itself, not with an other, and in this way alone as
free concept. If it is now claimed that the purpose of the good is thereby
still not realized, what we have is a relapse of the concept to the standpoint
that it assumes prior to its activity, when the actual is determined as
worthless and yet presupposed as real. This is a relapse that gives rise to the
progression to bad infinity. Its sole ground is that in the sublating of that
abstract reality the sublating itself is just as immediately forgotten, or what
is forgotten is that this reality is rather already presupposed as an actuality
which is in and for itself worthless, nothing objective. This repetition of
the presupposition of the unrealized purpose after the actual realization of
the purpose also means that the subjective attitude of the objective concept
is reproduced and perpetuated, with the result that the finitude of the good,
with respect to both content and form, appears as the abiding truth, and
its actualization always as only a singular, never universal, act. – As a matter
of fact this state has already sublated itself in the realization of the good;
what still limits the objective concept is its own view of itself, and this view
vanishes in the reflection on what its realization is in itself. By this view the
concept only stands in its own way, and all that it has to do about it is to
turn, not against an external actuality, but against itself.
That is to say, the activity in the second premise produces only a one-
sided being-for-itself, and its product therefore appears as something subjec-
tive and singular, and the first presupposition is consequently repeated in it.
But this activity is in truth just as much the positing of the implicit identity
of the objective concept and the immediate actuality. This actuality is by
presupposition determined to have only the reality of an appearance, to be
in and for itself a nullity, entirely open to determination by the objective
concept. As the external actuality is altered by the activity of the objective
concept and its determination is consequently sublated, the merely appar-
ent reality, the external determinability and worthlessness, are by that very
fact removed from it and it is thereby posited as having existence in and for
itself. In this the presupposition itself is sublated, namely the determina-
tion of the good as a merely subjective purpose restricted in content, the
necessity of first realizing it by subjective activity, and this activity itself.
In the result the mediation itself sublates itself; the result is an immedi-
acy which is not the restoration of the presupposition, but is rather the
presupposition as sublated. The idea of the concept that is determined in
and for itself is thereby posited, no longer just in the active subject but
equally as an immediate actuality; and conversely, this actuality is posited
as it is in cognition, as an objectivity that truly exists. The singularity of
the subject with which the subject was burdened by its presupposition has
vanished together with the presupposition. Thus the subject now exists
as free, universal self-identity for which the objectivity of the concept is a
given, just as immediately present to the subject as the subject immediately
knows itself to be the concept determined in and for itself. Accordingly,
in this result cognition is restored and united with the practical idea; the
previously discovered reality is at the same time determined as the realized
absolute purpose, no longer an object of investigation, a merely objective
world without the subjectivity of the concept, but as an objective world
whose inner ground and actual subsistence is rather the concept. This is
the absolute idea.

CHAPTER 3 THE ABSOLUTE IDEA

The absolute idea has shown itself to be the identity of the theoretical and
the practical idea, each of which, of itself still one-sided, possesses the idea
only as a sought-for beyond and unattained goal; each is therefore a synthesis
of striving, each possessing as well as not possessing the idea within it, passing
over from one thought to the other without bringing the two together but
remaining fixed in the contradiction of the two. The absolute idea, as the
rational concept that in its reality only rejoins itself, is by virtue of this
immediacy of its objective identity, on the one hand, a turning back to life;
on the other hand, it has equally sublated this form of its immediacy and
harbors the most extreme opposition within. The concept is not only soul,
but free subjective concept that exists for itself and therefore has personality –
the practical objective concept that is determined in and for itself and
is as person impenetrable, atomic subjectivity – but which is not, just the
same, exclusive singularity; it is rather explicitly universality and cognition,
and in its other has its own objectivity for its subject matter. All the rest
is error, confusion, opinion, striving, arbitrariness, and transitoriness; the
absolute idea alone is being, imperishable life, self-knowing truth, and is all
truth.
It is the sole subject matter and content of philosophy. Since it contains
all determinateness within it, and its essence consists in returning through
its self-determination and particularization back to itself, it has various
shapes, and the business of philosophy is to recognize it in these. Nature
and spirit are in general different modes of exhibiting its existence, art
and religion its different modes of apprehending itself and giving itself
appropriate existence. Philosophy has the same content and the same
purpose as art and religion, but it is the highest mode of apprehending the
absolute idea, because its mode, that of the concept, is the highest. Hence
it seizes those shapes of real and ideal finitude, as well of infinity and
holiness, and comprehends them and itself. The derivation and cognition
of these particular modes are now the further business of the particular
philosophical sciences. Also the logicality of the absolute idea can be called
a mode of it; but mode signifies a particular kind, a determinateness of
form, whereas the logicality of the idea is the universal mode in which
all particular modes are sublated and enveloped. The logical idea is the
idea itself in its pure essence, the idea which is enclosed in simple identity
within its concept and in reflective shining has as yet to step into a form-
determinateness. The Logic thus exhibits the self-movement of the absolute
idea only as the original word, a word which is an utterance, but one that
in being externally uttered has immediately vanished again. The idea is,
therefore, only in this self-determination of apprehending itself; it is in pure
thought, where difference is not yet otherness, but is and remains perfectly
transparent to itself. – The logical idea thus has itself, as the infinite form,
for its content – form that constitutes the opposite of content inasmuch as
the latter is the form determination that has withdrawn into itself and has
been so sublated in identity that this concrete identity stands over against
the identity developed as form; the content has the shape of an other and of
something given as against the form that as such stands simply in reference,
and whose determinateness is posited at the same time as reflective shine. –
More exactly, the absolute idea itself has only this for its content, namely
that the form determination is its own completed totality, the pure content.
Now the determinateness of the idea and the entire course traversed by this
determinateness has constituted the subject matter of the science of logic,
and out of this course the absolute idea has come forth for itself; thus to
be for itself, however, has shown itself to amount to this, namely that
determinateness does not have the shape of a content, but that it is simply
as form, and that accordingly the idea is the absolutely universal idea. What
is left to be considered here, therefore, is thus not a content as such, but
the universal character of its form – that is, method.
Method may appear at first to be just the manner in which cognition
proceeds, and this is in fact its nature. But as method this manner of
proceeding is not only a modality of being determined in and for itself; it is a
modality of cognition, and as such is posited as determined by the concept
and as form, since form is the soul of all objectivity and all otherwise
determined content has its truth in form alone. If the content is again
assumed as given to the method and of a nature of its own, then method,
so understood, is just like the logical realm in general a merely external
form. But against this assumption appeal can be made, not only to the
fundamental concept of what constitutes logic, but to the entire logical
course in which all the shapes of a given content and of objects came up for
consideration. This course has shown the transitoriness and the untruth of
all such shapes; also that no given object is capable of being the foundation
to which the absolute form would relate as only an external and accidental
determination; that, on the contrary, it is the absolute form that has proved
itself to be the absolute foundation and the ultimate truth. For this course
the method has resulted as the absolutely self-knowing concept, as the concept
that has the absolute, both as subjective and objective, as its subject matter,
and consequently as the pure correspondence of the concept and its reality,
a concrete existence that is the concept itself.
Accordingly, what is to be considered as method here is only the move-
ment of the concept itself. We already know the nature of this movement,
but it now has, first, the added significance that the concept is all, and
that its movement is the universal absolute activity, the self-determining
and self-realizing movement. The method is therefore to be acknowledged
as the universal, internal and external mode, free of restrictions, and as
the absolutely infinite force to which no object that may present itself as
something external, removed from reason and independent of it, could
offer resistance, or be of a particular nature opposite to it, and could not be
penetrated by it. It is therefore soul and substance, and nothing is conceived
and known in its truth unless completely subjugated to the method; it is the
method proper to each and every fact because its activity is the concept.
This is also the truer meaning of its universality; according to the universal-
ity of reflection, it is taken only as the method for all things; but according
to the universality of the idea, it is both the manner of cognition, of the
concept subjectively aware of itself, and the objective manner, or rather the
substantiality of things – that is, of concepts as they first appear as others
to representation and reflection. It is therefore not only the highest force of
reason, or rather its sole and absolute force, but also reason’s highest and
sole impulse to find and recognize itself through itself in all things. – Second,
here we also have the distinction of the method from the concept as such, the
particularization of the method. As the concept was considered for itself,
it appeared in its immediacy; the reflection, or the concept considering it,
fell on the side of our knowledge. The method is this knowledge itself, for
which the concept is not only as subject matter but is as its own subjec-
tive act, the instrument and the means of cognitive activity, distinct from
this activity and yet the activity’s own essentiality. In cognition as enquiry,
the method likewise occupies the position of an instrument, as a means
that stands on the side of the subject, connecting it with the object. The
subject in this syllogism is one extreme, the object is the other, and in
conclusion the subject unites through its method with the object without
however uniting with itself there. The extremes remain diverse, because
subject, method, and object are not posited as the one identical concept; the
syllogism is therefore always the formal syllogism; the premise in which
the subject posits the form on its side as its method is an immediate deter-
mination and contains therefore the determinations of the form – as we
have seen, of definition, division, and so forth 71 – as matters of fact found
ready-made in the subject. In true cognition, on the contrary, method is
not only an aggregate of certain determinations, but the determinateness
in-and-for-itself of the concept, and the concept is the middle term only
because it equally has the significance of the objective; in the conclusion,
therefore, the objective does not attain only an external determinateness by
virtue of the method, but is posited rather in its identity with the subjective
concept.
1. Accordingly, what constitutes the method are the determinations of
the concept itself and their connections, and these we must now examine
in the significance that they have as determinations of the method. – In
this, we must begin from the beginning. We spoke of this beginning at
the very beginning of the Logic, 72 and also in connection with subjective
cognition, 73 and we showed that, when not performed arbitrarily and in
the absence of categorial sensitivity, though it may seem to present many
difficulties, it is nevertheless of an extremely simple nature. Because it is the
beginning, its content is an immediate, but one that has the meaning and
the form of abstract universality. Or be it a content of being, or of essence or
of the concept, inasmuch as it is something immediate, it is assumed, found in
advance, assertoric. But first of all it is not an immediate of sense-intuition or
of representation, but of thought, which because of its immediacy can also be
called a supersensuous, inner intuiting. The immediate of sense-intuition
is a manifold and a singular. Cognition, on the contrary, is a thinking that
conceptualizes; its beginning, therefore, is also only in the element of thought,
a simple and a universal. – We spoke of this form earlier, in connection with
definition. 74 At the beginning of finite cognition universality is likewise
recognized as an essential determination, but only as thought – and concept
determination in opposition to being. In fact this first universality is an
immediate universality, and for that reason it has equally the significance of
being, for being is precisely this abstract self-reference. Being has no need
of further derivation, as if it came to the abstract element of definition
only because taken from the intuition of the senses or elsewhere, and
in so far as it can be pointed at. This pointing and deriving involve a
mediation that is more than a mere beginning, and is a mediation of a kind
that does not belong to the comprehension of thought, but is rather the
elevation of representation, of empirical and ratiocinative consciousness, to
the standpoint of thinking. According to the currently accepted opposition
of thought, or concept, and being, it passes as a very important truth that
no being belongs as yet to thought as thought, and that being has a ground
of its own independent of thought. But the simple determination of being
is in itself so poor that, if for that reason alone, not much fuss ought to be
made about it; the universal is immediately itself this immediate because,
as abstract, it is also the abstract self-reference which is being. In fact, the
demand that being should be exhibited has a further, inner meaning in
which more is at issue than just this abstract determination; implied in it
is the demand for the realization of the concept, a realization that is missing
at the beginning itself but is rather the goal and the business of the entire
subsequent development of cognition. Further, inasmuch as the content
of the beginning is to be justified and authenticated as something true or
correct by being exhibited in inner or outer perception, it is no longer
the form of universality as such that is meant, but its determinateness,
about which more in a moment. The authentication of the determinate
content with which the beginning is made seems to lie behind it, but is in
fact to be regarded as an advance, in so far as it is a matter of conceptual
cognition.
The beginning, therefore, has for the method no other determinateness
than that of being the simple and universal; this is precisely the determinate-
ness that makes it deficient. Universality is the pure, simple concept, and
the method, as the consciousness of this concept, is aware that universality
is only a moment and that in it the concept is still not determined in and for
itself. But with this consciousness that would want to carry the beginning
further only for the sake of method, the method is only a formal procedure
posited in external reflection. Where the method, however, is the objective
and immanent form, the immediate character of the beginning must be
a lack inherent in the beginning itself, which must be endowed with the
impulse to carry itself further. But in the absolute method the universal
has the value not of a mere abstraction but of the objective universal, that
is, the universal that is in itself the concrete totality, but a totality as yet not
posited, not yet for itself. Even the abstract universal is as such, when con-
sidered conceptually, that is, in its truth, not just anything simple, but is,
as abstract, already posited afflicted by a negation. For this reason also there
is nothing so simple and so abstract, be it in actuality or in thought, as is
commonly imagined. Anything as simple as that is a mere presumption that
has its ground solely in the lack of awareness of what is actually there. – We
said earlier that the beginning is made with the immediate; 75 the immediacy
of the universal is the same as what is here expressed as the in-itself that is
without being-for-itself. – One may well say, therefore, that every beginning
must be made with the absolute, just as every advance is only the exposition
of it, in so far as implicit in existence is the concept. 76 But because the
absolute exists at first only implicitly, in itself, it equally is not the absolute
nor the posited concept, and also not the idea, for the in-itself is only an
abstract, one-sided moment, and this is what they are. The advance is not,
therefore, a kind of superfluity; this is what it would be if that which is
at the beginning were already the absolute; the advance consists rather in
this, that the universal determines itself and is the universal for itself, that
is, equally a singular and a subject. Only in its consummation is it the
absolute.
It may also be mentioned that a beginning which is in itself a concrete
totality may as such also be free and its immediacy have the determination
of an external existence; the germ of anything living, and subjective purpose
in general, have shown themselves to be such beginnings; hence both are
themselves impulses. The non-spiritual and inanimate, on the contrary, are
the concrete concept only as real possibility; cause is the highest stage in
which the concrete concept has, as the beginning in the sphere of necessity,
an immediate existence; but it is not yet a subject that maintains itself
as such in the course of its effective realization. The sun, for instance,
and in general all things inanimate, are determinate concrete existences
in which real possibility remains an inner totality; the moments of the
latter are not posited in them in subjective form and therefore, in so far
as they are realized, they attain concrete existence through other corporeal
individuals.
2. The concrete totality which makes the beginning possesses as such,
within it, the beginning of the advance and development. As concrete,
it is differentiated in itself, but because of its initial immediacy, this first
differentiation is to start with a diversity. However, as self-referring univer-
sality, as subject, the immediate is also the unity of this diversity. – This
reflection is the first stage of the forward movement – the emergence of
non-indifference, judgment, and determining in general. Essential is that the
method find, and recognize, the determination of the universal within it.
Whatever in this abstractive generation of the universal is left out of the
concrete is then picked up, still externally, by the finite cognition of the
understanding. This is how the latter operates. The absolute method, on
the contrary, does not behave in this manner of external reflection but
takes the determinate from its subject matter, for it is itself its immanent
principle and its soul. – This is what Plato demanded of cognition, that it
should consider things in and for themselves; on the one hand, that it should
consider them in their universality; on the other hand, that it should not
stray away from them while it grasps at circumstances, examples, and com-
parisons, but, on the contrary, should keep only them in view before it
and bring to consciousness what is immanent in them. – To this extent the
method of absolute cognition is analytic. That the method finds the further
determinations of its initial universal simply and solely in this universal,
constitutes the concept’s absolute objectivity, of which the method is the
certainty. – Equally so, however, is the method synthetic, for its subject
matter, while immediately determined as the simple universal, through the
determinateness which it has in its very immediacy and universality, proves
to be an other. Yet this connection in diversity that the subject matter is
thus in itself, is no longer a synthesis as understood in finite cognition; the
no less thoroughly analytic determination of the subject matter, the fact
that the connection is within the concept, already distinguishes it fully from
the latter synthesis.
This no less synthetic than analytic moment of the judgment through
which the initial universal determines itself from within itself as the other
of itself is to be called the dialectical moment. Dialectic is one of those
ancient sciences that have been the most misjudged in the metaphysics of
the moderns, and in general also by popular philosophy, both ancient and
recent. Diogenes Laërtius said of Plato that, just as Thales was the founder
of natural philosophy and Socrates of moral philosophy, so Plato was the
founder of the third of the sciences that belong to philosophy, of dialectic –
a contribution for which he was highly esteemed by the ancients but that
often goes quite unnoticed by those who have the most to say about him. 77
Dialectic has often been regarded as an art, as if it rested on a subjective
talent and did not belong to the objectivity of the concept. What shape
it received in Kantian philosophy, and with what result, has already been
indicated in representative examples of that philosophy’s view. 78 It must be
regarded as an infinitely important step that dialectic is once more being
recognized as necessary to reason, although the result that must be drawn
from it is the opposite than Kant drew.
When dialectic is not presented, as it generally is, as something inci-
dental, it usually assumes the following more precise form. It is shown of
a subject matter or other (for instance: world, movement, point, and so
on) that a certain determination accrues to it (for instance, in the order of
the just mentioned examples: finitude in space or time, being at this place,
absolute negation of space), also the opposite determinations can then just
as necessarily be shown to accrue to it (for example: infinity in space and
time, not-being at this place, reference to space and hence spatiality). The
older Eleatic school directed its dialectic especially against motion; Plato
commonly did it against accepted notions and concepts of his time, in par-
ticular those of the Sophists, but also against the pure categories and the
determinations of reflection; the later and more sophisticated form of skep-
ticism extended it not only to the immediate so-called facts of consciousness
and the maxims of ordinary life, but also to all scientific concepts. Now the
conclusion drawn from this kind of dialectic is in general the contradiction
and nullity of the asserted claims. But this can happen in two ways – either
in the objective sense, that the subject matter that thus contradicts itself
internally cancels itself and is a non-thing (this was, for instance, the con-
clusion of the Eleatics, who denied the truth of the world, of movement,
of the point); or in the subjective sense, that cognition is deficient. Now
understood in this last subjective sense, the conclusion may be taken in
two further ways. It may mean that it is this dialectic itself that generates
the artifice of an illusion. This is the common view of the so-called healthy
common sense that takes its stand on the evidence of the senses and on cus-
tomary notions and claims, at times quietly, like Diogenes the cynic did, who
demonstrated the vacuity of the dialectic of motion by silently walking up
and down; but often by getting itself all worked up, declaring that dialectic
is mere foolery or, when important ethical matters are at issue, the criminal
attempt at unsettling essentially solid norms and providing excuses for the
wicked – a view we see directed in the Socratic dialectic against that of
the Sophists, with an ire that, turned into the opposite direction, even cost
Socrates his life. As for the vulgar refutation that opposes to thinking, as
Diogenes did, sensuous consciousness and in this latter believes that it finds
the truth, this we must leave to itself; but in so far as dialectic sublates
ethical determinations, we must have confidence in reason that it will
know how to reinstate them, but reinstate them in their truth and in the
consciousness of their right, though also of their limitations. – Yet another
view is that the result of subjective nullity has nothing to do with dialectic
itself, but that it affects the cognition against which it is directed and, in
the view of skepticism and likewise of the Kantian philosophy, cognition in
general.
The fundamental prejudice here is that dialectic has only a negative
result – a point about which more in a moment. First, regarding the said
form in which dialectic usually makes its appearance, it is to be observed
that according to that form the dialectic and its result affect a subject matter
which is previously assumed or also the subjective cognition of it, and
declare either the latter or the subject matter to be null and void, while,
on the contrary, no attention is given to the determinations which are
exhibited in the subject matter as in a third thing and presupposed as valid
for themselves. To have called attention to this uncritical procedure has
been the infinite merit of the Kantian philosophy, and in so doing to have
given the impetus to the restoration of logic and dialectic understood as the
examination of thought determinations in and for themselves. The subject
matter, as it is apart from thought and conceptualization, is a picture
representation or also a name; it is in the determinations of thought and of
the concept that it is what it is. In fact, therefore, everything rests on these
determinations; they are the true subject matter and content of reason,
and anything else that might be understood by subject matter and content
in distinction from them has meaning only through them and in them.
It must not therefore be taken as the fault of a subject matter or of the
cognition that these determinations, because of what they are and the way
they are externally joined, prove to be dialectical. On this assumption,
the subject matter and the cognition are imagined to be a subject on
which the determinations are brought to bear, in the form of predicates,
properties, or self-subsistent universals, as fixed and independently correct,
so that these determinations are brought into dialectical relations and incur
contradiction only by extraneous and contingent conjunction in and by a
third thing. But this kind of external and fixed subject of imagination and
understanding, and also these abstract determinations, cannot be regarded
as ultimates, as secure and permanent substrates. On the contrary, they are to
be regarded as themselves immediate, precisely the kind of presuppositions
and starting points which, as we have shown above, 79 must succumb to
dialectic in and for themselves, because they are to be taken as in themselves
the concept. The same applies to all oppositions that are assumed as fixed,
as for example the finite and the infinite, the singular and the universal.
These are not in contradiction through some external conjoining; on the
contrary, as an examination of their nature shows, they are a transition in
and for themselves; the synthesis and the subject in which they appear is
the product of their concept’s own reflection. If a consideration that avoids
the concept stops short at their external relation, isolates them and leaves
them as fixed presuppositions, it is the concept that, on the contrary, will
fix its sight on them, move them as their soul and bring out their dialectic.
Now this is the very standpoint indicated above 80 from which a universal
prius, considered in and for itself, proves to be the other of itself. Taken quite
generally, this determination can be taken to mean that what is at first
immediate is therewith posited as mediated, as referred to an other, or
that the universal is posited as a particular. The second universal that has
thereby arisen is thus the negative of that first and, in view of subsequent
developments, the first negative. From this negative side, the immediate has
perished in the other; but the other is essentially not an empty negative, the
nothing which is normally taken to be the result of dialectic, but is rather
the other of the first, the negative of the immediate; it is therefore determined
as the mediated – contains as such the determination of the first in it. The
first is thus essentially preserved and contained also in the other. – To hold
fast to the positive in its negative, to the content of the presupposition in
the result, this is the most important factor in rational cognition; what
is more, it takes only the simplest of reflections to be convinced of the
absolute truth and necessity of this requirement, and as for examples of
proofs that testify to this, the whole Logic consists of such proofs.
So what we now have, taken first or also immediately, is the mediated,
also a simple determination, for the first has perished in it, and only the
second is therefore at hand. Now since the first is contained in the second,
and this second is the truth of the first, this unity of the two can be expressed
in the form of a proposition in which the immediate is placed as the subject
but the mediated as its predicate; for example, “the finite is infinite,” “one
is many,” “the singular is the universal.” The inadequacy of the form
of such propositions and judgments is however obvious. In connection
with judgment it was shown that its form in general, and most of all the
immediate form of the positive judgment, is incapable of holding within its
grasp the speculative content and the truth. 81 Its closest complement, the
negative judgment, would have to be brought in at least in equal measure.
In judgment the first, as subject, conveys the reflective semblance of an
independent subsistence, whereas it is in fact sublated in the predicate as
in its other; this negation is indeed contained in the content of the above
propositions, but their positive form contradicts the content; consequently,
what is contained in them is not posited – whereas this was precisely the
intent behind the use of a proposition.
The second determination, the negative or mediated determination, is
moreover at the same time the one that mediates. At first it may be taken as
a simple determination, but its truth is that it is a reference or relation; for
it is the negative, but the negative of the positive, and it includes this positive
within itself. It is the other, therefore, not of a one to which it is indifferent;
in that case it would not be an other, nor a reference or relation. It is rather
the other in itself, the other of an other; hence it includes its own other within
itself and is consequently the contradiction, the posited dialectic, of itself. –
Because the first or the immediate is the concept in itself or implicitly,
and therefore is the negative also only implicitly, the dialectical moment
in it consists in the positing of the difference that is implicitly contained
in it. The second is on the contrary itself the determinate, the difference or
relation; hence the dialectical moment consists in its case in the positing
of the unity contained within it. – For this reason, if the negative, the
determinate, relation, judgment, and all the determinations falling under
this second moment, do not appear by themselves already as contradiction,
as dialectical, this is solely a defect on the part of thinking that fails to bring
its thoughts together. For the material, the opposed determinations in one
connection, are already posited, already present for thought. But formal
thinking makes identity its law, lets the contradictory content that it has
before it fall into the sphere of representation, in space and time, where
the contradictory is held in external moments, next to and following each
other, parading before consciousness without reciprocal contact. The firm
principle that formal thinking lays down for itself here is that contradiction
cannot be thought. But in fact the thought of contradiction is the essential
moment of the concept. Formal thought does in fact think it, only it at
once looks away from it and stating its principle it only passes over from it
into abstract negation.
Now the negativity just considered constitutes the turning point of the
movement of the concept. It is the simple point of the negative self-reference,
the innermost source of all activity, of living and spiritual self-movement;
it is the dialectical soul which everything true possesses and through which
alone it is true; for on this subjectivity alone rests the sublation of the
opposition between concept and reality, and the unity which is truth. –
The second negative at which we have arrived, the negative of the negative,
is this sublating of contradiction, and it too, just like contradiction, is
not an act of external reflection; for it is on the contrary the innermost,
objective moment of the life of spirit by virtue of which a subject is a person,
is free. – The self-reference of the negative is to be regarded as the second
premise of the entire syllogism. If the terms analytic and synthetic are used
as opposites, the first premise may be regarded as the analytic moment, for
in it the immediate relates to its other immediately and therefore passes
over, or rather has passed over, into it – though this connection, as already
remarked, is for this very reason also synthetic, for it is its other that it passes
over into. The second premise considered here may be defined as synthetic,
because it is the connection of the differentiated, as differentiated, to that
from which it is differentiated. – Just as the first premise is the moment of
universality and communication, so is the second determined by singularity –
a singularity which in referring to the other is at first exclusive, for itself,
and different. The negative appears as the mediating factor, because it holds
itself and the immediate of which it is the negation within itself. In so far
as these two determinations are taken as referring to each other externally
in some relation or other, the negative is only the formal mediating factor;
but, as absolute negativity, the negative moment of absolute mediation is
the unity which is subjectivity and soul.
In this turning point of the method, the course of cognition returns at the
same time back into itself. This negativity is as self-sublating contradiction
the restoration of the first immediacy, of simple universality; for the other
of the other, the negative of the negative, is immediately the positive, the
identical, the universal. In the whole course, if one at all cares to count, this
second immediate is third to the first immediate and the mediated. But it
is also third to the first or formal negative and to the absolute negativity or
second negative; now in so far as that first negative is already the second
term, the term counted as third can also be counted as fourth, and instead
of a triplicity, the abstract form may also be taken to be a quadruplicity; in
this way the negative or the difference is counted as a duality. – The third or
the fourth is in general the unity of the first and the second moment, of the
immediate and the mediated. – That it is this unity, or that the entire form
of the method is a triplicity, is indeed nothing but the merely superficial,
external side of cognition; but to have also demonstrated this superficiality,
and to have done it in the context of a specific application (for the abstract
form of number has been around for a long time, as is well known, but
without conceptual comprehension and therefore without any result) is
again to be regarded as an infinite merit of the Kantian philosophy. 82 The
syllogism, or the threefold, has always been recognized to be the universal
form of reason; but it has had in general the value of a wholly external form
that does not determine the nature of the content; moreover, since in its
formalism it gets caught up in the understanding’s determination of mere
identity, it lacks the essential dialectical moment of negativity; and yet this
moment enters into the triplicity of the determinations, because the third
term is the unity of the two first determinations and these, since they are
diverse, can be in unity only as sublated. – Formalism, it is true, has also
seized hold of triplicity, attending to its empty schema; the shallow nonsense
and the barrenness of the so-called construction of modern philosophy, that
consists in nothing but fastening that formal schema everywhere for the
sake of external order, with no concept or immanent determination, has
rendered that form tedious and has given it a bad name. Yet the insipidity
of this use cannot rob it of its inner worth, and the fact that the shape of
reason was discovered, albeit without conceptual comprehension at first, is
always to be highly valued.
Now, on closer examination, the third is the immediate, but the imme-
diate through sublation of mediation, the simple through the sublating of
difference, the positive through the sublating of the negative; it is the
concept that has realized itself through its otherness, and through the sub-
lating of this reality has rejoined itself and has restored its absolute reality,
its simple self-reference. This result is therefore the truth. It is just as much
immediacy as mediation – though these forms of judgments, that the third
is immediacy and mediation, or that it is the unity of the two, are not
capable of grasping it, for it is not a dormant third but, exactly like this
unity, self-mediating movement and activity. – Just as that with which we
began was the universal, so the result is the singular, the concrete, the subject;
what the former is in itself, the latter is now equally for itself: the universal
is posited in the subject. The two first moments of triplicity are abstract,
untrue moments that are dialectical for that very reason, and through this
their negativity make themselves into the subject. For us at first, the con-
cept itself is both the universal that exists in itself and the negative that
exists for itself, and also the third term that exists in and for itself, the
universal that runs through all the moments of the syllogism; but this third
is the conclusion in which the concept mediates itself with itself through
its negativity and is thereby posited for itself as the universal and the identity
of its moments.
Now this result, as the whole that has withdrawn into itself and is identi-
cal with itself, has given itself again the form of immediacy. Consequently,
it is now itself all that the starting point had determined itself to be. As sim-
ple self-reference it is a universal, and in this universal the negativity that
constituted its dialectic and mediation has likewise withdrawn into simple
determinateness, which can again be a beginning. It may seem at first that
this cognition of the result is an analysis of it and would therefore have to
dissect these determinations again, and the course that it went through in
order to come to be – the course that we have examined. But if the subject
matter were in fact treated analytically in this manner, it would belong to
that stage of the idea considered above, 83 a mode of cognition that searches
for its subject matter and only states of it what it is, without the necessity
of its concrete identity and of its concept. But the method of truth that
comprehends the subject matter, though analytic as we have seen, since
it remains strictly within the concept, is however equally synthetic, for
through the concept the subject matter is determined as dialectical and as
other. 84 On the new foundation that the result has now constituted as the
subject matter, the method remains the same as in the preceding subject
matter. The difference concerns solely the status of the foundation as such;
although it is certainly still a foundation, its immediacy is only form, since
it was a result as well; hence its determinateness as content is no longer
something merely taken up but is deduced and proved.
It is here that the content of cognition first enters as such into the circle
of consideration, because as deduced it now belongs to the method. The
method itself expands with this moment into a system. – With respect to
content, the beginning has to be for the method at first wholly indeter-
minate; to this extent the method appears as the merely formal soul, for
which and by which the beginning was determined simply and solely only
according to form, that is to say, as the immediate and universal. In the
course of the movement we have indicated, the subject matter has received
a determinateness for itself – and this determinateness is a content, for the
negativity that has withdrawn into simplicity is the sublated form, and
stands as simple determinateness over against its development, and in the
first instance against its very opposition to universality.
Now since this determinateness is the proximate truth of the indetermi-
nate beginning, it denounces the incompleteness of the latter, and it also
denounces the method itself which, starting from that beginning, was only
formal. This can now be expressed as the henceforth determinate demand
that the beginning, since as against the determinateness of the result it
is itself something determinate, ought to be taken not as immediate, but
as mediated and deduced. This may appear as the demand for an infi-
nite retrogression in proof and deduction; just as from the newly obtained
beginning a result likewise emerges as the method runs its course, so that
the movement would roll on forwards to infinity as well.
It has been repeatedly shown that the infinite progression as such belongs
to a reflection void of concept; the absolute method, which has the con-
cept for its soul and content, cannot lead into it. Even such beginnings as
being, essence, universality, may seem at first to be of the kind that possess
the full universality and complete absence of content that is required for
an entirely formal beginning, such as the beginning is supposed to be,
and therefore not to require or allow, as absolutely first beginnings, fur-
ther regress. Since they refer purely to themselves, they are immediate and
indeterminate, and so they do not of course have in them the difference
which is straightaway posited in some other beginning between the uni-
versality of its form and its content. But the very indeterminacy which
these logical beginnings have as their sole content is what constitutes their
determinateness; this determinateness consists in their negativity, as sub-
lated mediation; the particularity of this negativity gives a particularity
also to their indeterminacy, and it is by virtue of it that being, essence, and
universality, are differentiated. Now the determinateness that accrues to
them when taken for themselves is their immediate determinateness, and
this is just as immediate as that of any content and in need, therefore,
of derivation; for the method it is a matter of indifference whether the
determinateness is taken as determinateness of form or of content. That it
gives itself a determination by the first of its results does not mean that,
in fact, it is thereby set on a new footing; it remains neither more nor less
formal than before. For since the method is the absolute form, the con-
cept that knows itself and everything as concept, there is no content that
would stand out over against it and determine it as a one-sided external
form. Hence, just as the lack of content of the said beginnings does not
make them absolute beginnings, so too it is not the content that would
as such lead the method into the infinite progress forwards or backwards.
In one respect, the determinateness that the method generates for itself in
its result is the moment through which it is self-mediation and converts
the immediate into a mediated beginning. But conversely, it is through that
determinateness that this mediation of the method runs its course; it goes
through a content, as through a seeming other of itself, back to its begin-
ning, in such a way that it does not merely restore that beginning, albeit as
determinate, but that the result is equally the sublated determinateness, and
hence also the restoration of the first immediacy in which it began. This
it accomplishes as a system of totality. We now have to consider it in this
determination.
The determinateness which was the result is, as we have shown, 85 itself
a new beginning because of the form of simplicity into which it has
withdrawn; since this beginning is distinguished from the one preceding
it by this very determinateness, cognition rolls onwards from content to
content. First of all, this forward movement determines itself in that it
begins from simple determinacies, and the following become ever richer
and more concrete. For the result contains its beginning and its course
has enriched it with a new determinateness. The universal constitutes the
foundation; the advance is not to be taken, therefore, as a flowing from
other to other. In the absolute method, the concept maintains itself in its
otherness, the universal in its particularization, in judgment and reality; at
each stage of further determination, the universal elevates the whole mass of
its preceding content, not only not losing anything through its dialectical
advance, or leaving it behind, but, on the contrary, carrying with itself all
that it has gained, inwardly enriched and compressed.
This expansion may be regarded as the moment of content, and in the
whole as the first premise; the universal is communicated to the wealth of
content, is immediately received in it. But the relation has also a second,
negative or dialectical side. The enrichment proceeds in the necessity of the
concept, it is contained by it, and every determination is a reflection into
itself. Each new stage of exteriorization, that is, of further determination, is
also a withdrawing into itself, and the greater the extension, just as dense
is the intensity. The richest is therefore the most concrete and the most
subjective, and that which retreats to the simplest depth is the mightiest
and the most all-encompassing. The highest and most intense point is the
pure personality that, solely by virtue of the absolute dialectic which is its
nature, equally embraces and holds everything within itself, for it makes
itself into the supremely free – the simplicity which is the first immediacy
and universality.
It is in this manner that each step of the advance in the process of further
determination, while getting away from the indeterminate beginning, is
also a getting back closer to it; consequently, that what may at first appear
to be different, the retrogressive grounding of the beginning and the pro-
gressive further determination of it, run into one another and are the same.
The method, which thus coils in a circle, cannot however anticipate in
a temporal development that the beginning is as such already something
derived; sufficient for an immediate beginning is that it be simple uni-
versality. Inasmuch as this is what it is, it has its complete condition; and
there is no need to deprecate the fact that it may be accepted only provi-
sionally and hypothetically. Whatever might be adduced against it – about
the limitations of human cognition; about the need to reflect critically
on the instrument of cognition before getting to the fact itself – all these
are themselves presuppositions, concrete determinations that as such carry
with them the demand for mediation and grounding. Therefore, since
they formally have no advantage over beginning with the fact itself as they
protest against, and, because of their more concrete content, are on the
contrary all the more in need of derivation, singling them out for special
attention is to be considered as empty presumption. They have an untrue
content, for they make into something incontestable and absolute what is
known to be finite and untrue, namely a restricted cognition determined as
form and instrument in opposition to its content; this untrue cognition is itself
also the form, the retroactive search for grounds. – The method of truth also
knows that the beginning is incomplete, because it is a beginning; but at
the same time it knows that this incompleteness is necessary, because truth
is but the coming-to-oneself through the negativity of immediacy. The
impatience that would merely transcend the determinate – be it called begin-
ning, object, the finite, or in whatever other form it is otherwise taken –
in order that one would find oneself immediately in the absolute, has
nothing before it as cognition but the empty negative, the abstract infinite.
Or what it has before it is a presumed absolute, presumed because not
posited, not comprehended; comprehended it will be only through the medi-
ation of cognition, of which the universal and immediate are a moment,
and as for the truth itself, it resides only in the extended course of media-
tion and at the end. – To meet the subjective need and the impatience that
come with not knowing, one may well provide an overview of the whole in
advance – by means of a division for reflection that, in the manner of finite
cognition, gives the particular of the universal as already there, to be waited
for as the science progresses. Yet this affords nothing more than a picture
for representation; for the true transition from the universal to the particular
and to the whole which is determined in and for itself and in which that
first universal is in truth itself again a moment – this transition is alien to
the division of reflection and is the exclusive mediation of science itself.
By virtue of the nature of the method just indicated, the science presents
itself as a circle that winds around itself, where the mediation winds the
end back to the beginning which is the simple ground; the circle is thus a
circle of circles, for each single member ensouled by the method is reflected
into itself so that, in returning to the beginning it is at the same time the
beginning of a new member. Fragments of this chain are the single sciences,
each of which has a before and an after – or, more accurately said, has in
possession only the before and in its conclusion points to its after.
So the logic also has returned in the absolute idea to this simple unity
which is its beginning; the pure immediacy of being, in which all deter-
mination appears at first as extinguished or removed by abstraction, is the
idea that through mediation, that is, the sublation of mediation, has come
to the likeness corresponding to it. The method is the pure concept that
only relates to itself; it is, therefore, the simple self-reference which is being.
But it now is also the fulfilled concept, the concept that comprehends itself
conceptually, being as the concrete and just as absolutely intensive totality. –
In conclusion, there remains only this to be said of this idea, that in it, in
the first place, the science of logic has apprehended its own concept. In the
sphere of being, at the beginning of its content, its concept appears as a
knowledge external to that content in subjective reflection. But in the idea
of absolute cognition, the concept has become the idea’s own content. The
idea is itself the pure concept that has itself as its subject matter and which,
as it runs itself as subject matter through the totality of its determinations,
builds itself up to the entirety of its reality, to the system of science, and
concludes by apprehending this conceptual comprehension of itself, hence
by sublating its position as content and subject matter and cognizing the
concept of science. – In second place, this idea is still logical; it is shut
up in pure thought, the science only of the divine concept. Its systematic
exposition is of course itself a realization, but one confined within the same
sphere. Because the pure idea of cognition is to this extent shut up within
subjectivity, it is the impulse to sublate it, and pure truth becomes as final
result also the beginning of another sphere and science. It only remains here
to indicate this transition.
The idea, namely, in positing itself as the absolute unity of the pure
concept and its reality and thus collecting itself in the immediacy of being,
is in this form as totality – nature. – This determination, however, is
nothing that has become, is not a transition, as was the case above when the
subjective concept in its totality becomes objectivity, or the subjective purpose
becomes life. The pure idea into which the determinateness or reality of the
concept is itself raised into concept is rather an absolute liberation for which
there is no longer an immediate determination which is not equally posited
and is not concept; in this freedom, therefore, there is no transition that
takes place; the simple being to which the idea determines itself remains
perfectly transparent to it: it is the idea that in its determination remains
with itself. The transition is to be grasped, therefore, in the sense that the
idea freely discharges 87 itself, absolutely certain of itself and internally at
rest. On account of this freedom, the form of its determinateness is just as
absolutely free: the externality of space and time absolutely existing for itself
without subjectivity. – Inasmuch as this externality is only in the abstract
determinateness of being and is apprehended by consciousness, it is as mere
objectivity and external life; within the idea, however, it remains in and
for itself the totality of the concept, and science in the relation of divine
cognition to nature. But what is posited by this first resolve of the pure idea
to determine itself as external idea is only the mediation out of which the
concept, as free concrete existence that from externality has come to itself,
raises itself up, completes this self-liberation in the science of spirit, and in
the science of logic finds the highest concept of itself, the pure concept
conceptually comprehending itself.