
The Doctrine of the Concept

OF THE CONCEPT IN GENERAL

What the nature of the concept is cannot be given right away,
not any more than can the concept of any other subject matter.
It might perhaps seem that, in order to state the concept of a subject matter,
the logical element can be presupposed, and that this element would not
therefore be preceded by anything else, or be something deduced,
just as in geometry logical propositions, when they occur applied to
magnitudes and employed in that science, are premised in the form of axioms,
underived and underivable determinations of cognition.
Now the concept is to be regarded indeed, not just as
a subjective presupposition but as absolute foundation;
but it cannot be the latter except to the extent that
it has made itself into one.
Anything abstractly immediate is indeed a first;
but, as an abstraction, it is rather something mediated, the foundation of which,
if it is to be grasped in its truth, must therefore first be sought.
And this foundation will indeed be something immediate,
but an immediate which has made itself such by the sublation of mediation.

From this aspect the concept is at first to be regarded simply
as the third to being and essence, to the immediate and to reflection.
Being and essence are therefore the moments of its becoming;
but the concept is their foundation and truth as the identity
into which they have sunk and in which they are contained.
They are contained in it because the concept is their result,
but no longer as being and essence;
these are determinations which they have only in so far as
they have not yet returned into the identity which is their unity.
Hence the objective logic, which treats of being and essence,
constitutes in truth the genetic exposition of the concept.

More precisely, substance already is real essence,
or essence in so far as it is united with being
and has stepped into actuality.
Consequently, the concept has substance
for its immediate presupposition;
substance is implicitly what the concept is explicitly.
The dialectical movement of substance
through causality and reciprocal affection
is thus the immediate genesis of the concept
by virtue of which its becoming is displayed.
But the meaning of its becoming,
like that of all becoming,
is that it is the reflection of something
which passes over into its ground,
and that the at first apparent other
into which this something has passed over
constitutes the truth of the latter.

Thus the concept is the truth of substance,
and since necessity is the determining relational mode of substance,
freedom reveals itself to be the truth of necessity
and the relational mode of the concept.
The necessary forward course of determination
characteristic of substance is the positing
of that which is in and for itself.
The concept is now this absolute
unity of being and reflection
whereby being-in-and-for-itself
only is by being equally reflection or positedness,
and positedness only is by being equally in-and-for-itself.
This abstract result is elucidated by the
exposition of its concrete genesis
which contains the nature of the concept
but had to precede its treatment.
We must briefly sum up here, therefore,
the main moments of this exposition
(which has been treated in detail in
Book Two of the Objective Logic).

Substance is the absolute,
the actual in-and-for-itself:

in itself, because it is the simple
identity of possibility and actuality;
absolute, because it is the essence containing
all actuality and possibility within itself;
for itself, because it is this identity
as absolute power or absolutely self-referring negativity.

The movement of substantiality posited by these moments
consists in the following stages:

1. Substance, as absolute power or self-referring negativity, differentiates
itself into a relation in which what are at first only simple moments are
substances and original presuppositions. – Their specific relation is that
of a passive substance, of the originariness of the simple in-itself which,
powerless to posit itself, is only originary positedness, and of an active
substance, the self-referring negativity which has as such posited itself as
an other and refers to it. This other is precisely the passive substance
which the active substance, as originative power, has presupposed for itself
as its condition. – This presupposing is to be understood in the sense
that the movement of substance is at first in the form of one moment of
its concept, that of the in-itself – that the determinateness of one of the
substances standing in relation is itself also the determinateness of this
relation.

2. The other moment is the being-for-itself or the power positing itself as
self-referring negativity and thereby again sublating what it presupposes. –
The active substance is cause; it acts; this means that it is now a positing,
just as before it was a presupposing, that (a) power is also given the reflective
shine of power, positedness also the reflective shine of positedness. What
in the presupposition was the originary becomes in causality, by virtue
of the reference to an other, what it is in itself. The cause brings about
an effect. But it does so in another substance and it is now power with
reference to an other; it thus appears as cause but is cause only by virtue
of this appearing. – (b) 2 The effect enters the passive substance and
by virtue of it the latter now also appears as positedness, but is passive
substance only in this positedness.

3. But there is more still present here than just this appearance, namely:
(a) the cause acts upon the passive substance, alters its determination;
but this determination is its positedness, for otherwise there is noth-
ing else to alter; the other determination which it obtains is however
that of causality; the passive substance thus comes to be cause, power,
and activity;
(b) the effect is posited in it by the cause; but that which is posited by
the cause is the cause itself which, in acting, is identical with itself;
it is this cause that posits itself in the place of the passive substance.

Similarly, with respect to the active substance:

(a) the action is the translation of the cause into the effect,
into its other, the positedness;
(b) the cause reveals itself in the effect as what it is;
the effect is identical with the cause, is not an other;
in acting the cause thus reveals the positedness to be
that which it (the cause) essentially is.
Each side, therefore, in accordance with how it refers to the other both
as identical with it and as the negative of it, becomes the opposite of
itself, but, in becoming this opposite, the other, and therefore also each,
remains identical with itself.
But both, the identical and the negative
reference, are one and the same; substance is self-identical only in its
opposite and this constitutes the absolute identity of the two substances
posited as two. It is by its act that active substance is manifested as cause
or originary substantiality, that is, by positing itself as the opposite of
itself, a positing which is at the same time the sublating of its presupposed
otherness, of passive substance. Contrariwise, it is by being acted upon that
the positedness is manifested as positedness, the negative as negative, and
consequently the passive substance as self-referring negativity, and in this
other the cause simply rejoins itself.

Through this positing, therefore, what is presupposed,
that is, the implicit originariness, becomes explicit;
but this being, which is now in and for itself,
is only by virtue of a positing
which is equally the sublation of what is presupposed,
or because the absolute substance has returned to itself
only out of, and in, its positedness
and for that reason is absolute.
Hence this reciprocal action is
appearance that again sublates itself;
the revelation that the reflective shine of causality,
in which the cause is as cause,
is just that, that it is reflective shine.
This infinite immanent reflection
(that the being-in-and-for-itself is
only such by being a positedness)
is the consummation of substance.
But this consummation is no longer the substance itself
but is something higher, the concept, the subject.
The transition of the relation of substantiality
occurs through its own immanent necessity
and is nothing more than the manifestation of itself,
that the concept is its truth,
and that freedom is the truth of necessity.

Earlier, in Book Two of the Objective Logic,
I have already called attention to the fact that
the philosophy that assumes its position at the standpoint of
substance and stops there is the system of Spinoza.
I have also indicated there the defect of this system,
both with respect to form and matter.
Something else, however, is the refutation of it.
Elsewhere, in connection with the refutation of a philosophical system,
I have also remarked quite in general that we must get over the distorted
idea that that system has to be represented as if thoroughly false, and as
if the true system stood to the false as only opposed to it. It is on the basis
of the context within which the system of Spinoza is presented here that
we can see its true standpoint and ask whether the system is true or false.
The relation of substantiality was generated by the nature of essence; this
relation and also its exposition as an expanded totality in the form of system
is, therefore, a necessary standpoint at which the absolute positions itself.
Such a standpoint, therefore, is not to be regarded as just an opinion, an
individual's subjective, arbitrary way of representing and thinking, as an
aberration of speculation; on the contrary, speculation necessarily runs into
it and, to this extent, the system is perfectly true. – But it is not the highest
standpoint. By itself alone, therefore, the system cannot be regarded as false,
as either requiring or being capable of refutation. This alone is rather to
be considered false in it: that it would be the highest standpoint. It also
follows that the true system cannot be related to it as just its opposite, for
as so opposed it would itself be one-sided. Rather, as the higher system, it
must contain it within as its subordinate.

Further, any refutation would have to come not from outside, that is, not
proceed from assumptions lying outside the system and irrelevant to it. The
system need only refuse to recognize those assumptions; the defect is such
only for one who starts from such needs and requirements as are based on
them. For this reason it has been said that there cannot be any refutation of
Spinozism for anyone who does not presuppose a commitment to freedom
and the independence of a self-conscious subject. Besides, a standpoint
so lofty and inherently so rich as that of the relation of substance does not
ignore those assumptions but even contains them: one of the attributes of
the Spinozistic substance is thought. The system knows how to resolve and
assimilate the determinations in which these assumptions conflict with it,
so that they re-emerge in it, but duly modified accordingly. The nerve,
therefore, of any external refutation consists solely in obstinately clinging
to the opposite categories of these assumptions, for example, to the absolute
self-subsistence of the thinking individual as against the form of thought
which in the absolute substance is posited as identical with extension.
Effective refutation must infiltrate the opponent's stronghold and meet
him on his own ground; there is no point in attacking him outside his
territory and claiming jurisdiction where he is not. The only possible
refutation of Spinozism can only consist, therefore, in first acknowledging
its standpoint as essential and necessary and then raising it to a higher
standpoint on the strength of its own resources. The relation of substantiality,
considered simply on its own, leads to its opposite: it passes over into the
concept. The exposition in the preceding Book of substance as leading to
the concept is, therefore, the one and only true refutation of Spinozism. It is
the unveiling of substance, and this is the genesis of the concept the principal
moments of which we have documented above. – The unity of substance is
its relation of necessity. But this unity is thus only inner necessity. By positing
itself through the moment of absolute negativity, it becomes manifested
or posited identity, and also, therefore, the freedom which is the identity
of the concept. This concept, the totality resulting from the relation of
reciprocity, is the unity of the two substances that stand in that relation,
but in such a way now that the two belong to freedom: they no longer
possess their identity blindly, that is to say, internally; on the contrary, the
substances now explicitly have the determination that they are essentially
reflective shine or moments of reflection, and for that reason that each has
immediately rejoined its other or its positedness, that each contains this
positedness in itself and in its other, therefore, is posited simply and solely
as identical with itself.

In the concept, therefore, the kingdom of freedom is disclosed.
The concept is free because the identity that exists in and for itself
and constitutes the necessity of substance exists
at the same time as sublated or as positedness,
and this positedness, as self-referring, is that very identity.
Vanished is the obscurity which the causally related substances have for each other, for the
originariness of their self-subsistence that makes them causes has passed
over into positedness and has thereby become self-transparently clear;
the “originary fact” is “originary” because it is a “self-causing fact,”
and this is the substance that has been let go freely into the concept.

The direct result for the concept is the following more detailed determination.
Because being which is in and for itself is immediately a positedness,
the concept is in its simple self-reference an absolute determinateness which,
by referring only to itself, is however no less immediately simple identity.
But this self-reference of the determinateness in which the latter rejoins itself
is just as much the negation of determinateness, 8 and thus the concept, as
this equality with itself, is the universal. But this identity equally has the
determination of negativity; it is a negation or determinateness that refers
to itself and as such the concept is the singular. 9 Each, the universal and the
singular, is a totality; each contains the determination of the other within
it and therefore the two are just as absolutely one totality as their oneness
is the diremption of its self into the free reflective shine of this duality.
And this is a duality which in the differentiation of singular and universal
appears to be perfect opposition, but an opposition which is so much of a
reflective shine that, in that the one is conceptualized and said, immediately
the other is therein conceptualized and said.
The determinateness does not go past itself,
even excludes the possibility of going past itself.
In this sense, because it precludes reference to anything else besides itself,
it ceases to be a determinateness
and becomes a universal. It negates its own determinateness, i.e. itself as negation.
In precluding reference to anything besides itself,
the universal regains negativity.
It is just itself and nothing else.
In this sense, it is a universe by itself, a singular.

The foregoing is to be regarded as
the concept of the concept.
To some it may seem to depart
from the common understanding of “concept,”
and they might require that we indicate
how our result fits with other ways
of representing or defining it.
But, for one thing, this cannot be an issue of proof
based on the authority of ordinary understanding.
In the science of the concept,
the content and determination of the latter
can be proven solely on the basis
of an immanent deduction which contains its genesis,
and such a deduction lies behind us.
And also, whereas the concept of the concept
as deduced here should in principle
be recognized in whatever else
is otherwise adduced as such a concept,
it is not as easy to ascertain what
others have said about its nature.
For in general they do not bother at all
enquiring about it but presuppose
that everyone already understands what
the concept means when speaking of it.
Of late especially one may indeed
believe that it is not worth pursuing
any such enquiry because, just as it was
for a while the fashion to say all things bad
about the imagination, then about memory,
it became in philosophy the habit some time ago,
and is still the habit now,
to heap every kind of defamation on the concept,
to hold it in contempt,
the concept which is the highest form of thought,
while the incomprehensible and the non-comprehended are
regarded as the pinnacle of both science and morality.

I confine myself to one remark which may contribute to the compre-
hension of the concept here developed and facilitate one's way into it. The
concept, when it has progressed to a concrete existence which is itself free,
is none other than the “I” or pure self-consciousness. True, I have con-
cepts, that is, determinate concepts; but the “I” is the pure concept itself,
the concept that has come into determinate existence. It is fair to suppose,
therefore, when we think of the fundamental determinations which con-
stitute the nature of the “I,” that we are referring to something familiar,
that is, a commonplace of ordinary thinking. But the “I” is in the first place
purely self-referring unity, and is this not immediately but by abstracting
from all determinateness and content and withdrawing into the freedom
of unrestricted equality with itself. As such it is universality, a unity that
is unity with itself only by virtue of its negative relating, which appears as
abstraction, and because of it contains all determinateness within itself as
dissolved. In second place, the “I” is just as immediately self-referring nega-
tivity, singularity, absolute determinateness that stands opposed to anything
other and excludes it – individual personality. This absolute universality
which is just as immediately absolute singularization – a being-in-and-
for-itself which is absolute positedness and being-in-and-for-itself only by
virtue of its unity with the positedness – this universality constitutes the
nature of the “I” and of the concept; neither the one nor the other can
be comprehended unless these two just given moments are grasped at the
same time, both in their abstraction and in their perfect unity.

When I say of the understanding that I have it, according to ordinary
ways of speaking, what is being understood by it is a faculty or a property
that stands in relation to my I in the same way as the property of a thing
stands related to that thing – as to an indeterminate substrate which is not
the true ground or the determining factor of the property. In this view, I
have concepts, and I have the concept, just as I also have a coat, complexion,
and other external properties. – Kant went beyond this external relation
of the understanding, as the faculty of concepts and of the concept, to
the “I.” It is one of the profoundest and truest insights to be found in
the Critique of Reason that the unity which constitutes the essence of the
concept is recognized as the original synthetic unity of apperception, the
unity of the “I think,” or of self-consciousness. 10 – This proposition is
all that there is to the so-called transcendental deduction of the categories
which, from the beginning, has however been regarded as the most difficult
piece of Kantian philosophy – no doubt only because it demands that we
should transcend the mere representation of the relation of the “I” and the
understanding, or of the concepts, to a thing and its properties or accidents,
and advance to the thought of it. – The object, says Kant in the Critique of
Pure Reason (2nd edn, p. 137), is that, in the concept of which the manifold
of a given intuition is unified. But every unification of representations
requires a unity of consciousness in the synthesis of them. Consequently, this
unity of consciousness is alone that which constitutes the reference of the
representations to an object, hence their objective validity, and that on
which even the possibility of the understanding rests. Kant distinguishes this
objective unity from the subjective unity of consciousness by which the
“I” becomes conscious of a manifold, whether simultaneously or successively
depending on empirical conditions. 11 In contrast to this subjective unity,
the principles of the objective determination of representations are only to
be derived from the principle of the transcendental unity of apperception. It
is by virtue of the categories, which are these objective determinations, that
the manifold of given representations is so determined as to be brought
to the unity of consciousness. 12 – On this explanation, the unity of the
concept is that by virtue of which something is not the determination of
mere feeling, is not intuition or even mere representation, but an object, and
this objective unity is the unity of the “I” with itself. – In point of fact,
the conceptual comprehension of a subject matter consists in nothing else
than in the “I” making it its own, in pervading it and bringing it into its
own form, that is, into a universality which is immediately determinateness,
or into a determinateness which is immediately universality. As intuited
or also as represented, the subject matter is still something external, alien.
When it is conceptualized, the being-in-and-for-itself that it has in intuition
and representation is transformed into a positedness; in thinking it, the “I”
pervades it. But it is only in thought that it is in and for itself; as it
is in intuition or representation, it is appearance. Thought sublates the
immediacy with which it first comes before us and in this way transforms
it into a positedness; but this, its positedness, is its being-in-and-for-itself or
its objectivity. This is an objectivity which the subject matter consequently
attains in the concept, and this concept is the unity of self-consciousness into
which that subject matter has been assumed; consequently its objectivity
or the concept is itself none other than the nature of self-consciousness,
has no other moments or determinations than the “I” itself.

Accordingly, we find in a fundamental principle of Kantian philosophy
the justification for turning to the nature of the “I” in order to learn
what the concept is. But conversely, it is necessary to this end that we have
grasped the concept of the “I” as stated. If we cling to the mere representation
of the “I” as we commonly entertain it, then the “I” is only the simple thing
also known as the soul, a thing in which the concept inheres as a possession
or a property. This representation, which does not bother to comprehend
either the “I” or the concept, is of little use in facilitating or advancing the
conceptual comprehension of the concept.

The position of Kant just cited contains two other points which concern
the concept and necessitate some further comments.
First of all, preceding
the stage of understanding are the stages of feeling and of intuition. It is
an essential proposition of Kant's Transcendental Philosophy that concepts
without intuition are empty, and that they have validity only as references
connecting the manifold given by intuition.
Second, the concept is given
as the objective element of cognition, consequently as the truth. Yet it is
taken to be something merely subjective, and we are not allowed to extract
reality from it, 15 for by reality objectivity is to be understood, since reality
is contrasted with subjectivity.
Moreover, the concept and anything logical are declared to be something
merely formal which, since it abstracts from content, does not contain truth.

Now, in the first place, as regards the relation of the understanding or concept
to the stages presupposed by it, the determination of the form of these stages
depends on which science is being considered. In our science, which is pure
Logic, they are being and essence. In Psychology, the stages preceding the
understanding are feeling and intuition, and then representation generally.
In the Phenomenology of Spirit, which is the doctrine of consciousness,
the ascent to the understanding is made through the stages of sensuous
consciousness and then of perception. Kant places ahead of it only feeling
and intuition. But, for a start, he himself betrays the incompleteness of
this progression of stages by appending to the Transcendental Logic or the
Doctrine of the Understanding a treatise on the concepts of reflection – a
sphere lying between intuition and understanding, or being and concept. 16
And if we consider the substance itself of these stages, it must first be
said that such shapes as intuition, representation, and the like, belong to
the self-conscious spirit which, as such, does not fall within the scope of
logical science. Of course, the pure determinations of being, essence, and
the concept, also constitute the substrate and the inner sustaining structure
of the forms of spirit; spirit, as intuiting as well as sensuous consciousness, is
in the form of immediate being, just as spirit as representational and also
perceptual consciousness has risen from being to the stage of essence or
reflection. But these concrete shapes are of as little interest to the science
of logic as are the concrete forms that logical determinations assume in
nature. These last would be space and time, then space and time as assuming
a content, as inorganic and then organic nature. Similarly, the concept is also
not to be considered here as the act of the self-conscious understanding,
not as subjective understanding, but as the concept in and for itself which
constitutes a stage of nature as well as of spirit. Life, or organic nature, is
the stage of nature where the concept comes on the scene, but as a blind
concept that does not comprehend itself, that is, is not thought; only as
self-aware and as thought does it belongs to spirit. Its logical form, however,
is independent of such shapes, whether unspiritual or spiritual. This is a
point which was already duly adumbrated in the Introduction, 17 and one
that one must be clear about before undertaking Logic, not when one is
already in it.

But, in second place, how the forms that precede the concept might
ever be shaped depends on how the concept is thought in relation to them.
This relation, as assumed in ordinary psychology as well as in Kant's
Transcendental Philosophy, 18 is that the empirical material, the manifold
of intuition and representation, is at first just there by itself, and that the
understanding then comes into it, brings unity to it, and raises it through
abstraction to the form of universality. The understanding is in this way
an inherently empty form which, on the one hand, obtains reality only by
virtue of that given content, and, on the other, abstracts from it, that is
to say, discards it as something useless, but useless only for the concept.
In both operations, the concept is not the one which is independent, is
not what is essential and true about that presupposed material; rather, this
material is the reality in and for itself, a reality that cannot be extracted
from the concept.

Now it must certainly be conceded that the concept is as such not yet
complete, that it must rather be raised to the idea which alone is the unity
of the concept and reality; and this is a result which will have to emerge
in what follows from the nature of the concept itself.
For the reality that
the concept gives itself cannot be picked up as it were from the outside
but must be derived from the concept itself in accordance with scientific
requirements. But the truth is that it is not the material given by intuition
and representation which must be validated as the real in contrast to the
concept. “It is only a concept,” people are wont to say, contrasting the
concept, as superior to it, not only with the idea, but with sensuous, spatial
and temporal, palpable existence. For this reason the abstract is then held to
be of less significance than the concrete, because so much of this palpable
material has been removed from it. In this view, to abstract means to
select from a concrete material this or that mark, but only for our subjective
purposes, without in any way detracting from the value and the status of
the many other properties and features that are left out; on the contrary,
by retaining them as reality, but yonder on the other side, still as fully
valid as ever. It is only because of its incapacity that the understanding
thus does not draw from this wealth and is forced rather to make do with
the impoverished abstraction. But now, to regard the given material of
intuition and the manifold of representation as the real, in contrast to what
is thought and the concept, is precisely the view that must be given up as
condition of philosophizing, and that religion, moreover, presupposes as
having already been given up. How could there be any need of religion, how
could religion have any meaning, if the fleeting and superficial appearance
of the sensuous and the singular were still regarded as the truth? But it
is philosophy that yields the conceptually comprehended 19 insight into the
status of the reality of sensuous being. Philosophy assumes indeed that the
stages of feeling, intuition, sense consciousness, and so forth, are prior to
the understanding, for they are the conditions of the genesis of the latter,
but they are conditions only in the sense that the concept results from
their dialectic and their nothingness and not because it is conditioned by
their reality. Abstractive thought, therefore, is not to be regarded as the
mere discarding of a sensuous material which does not suffer in the process
any impairment of reality; it is rather the sublation and reduction of that
material as mere appearance to the essential, which is manifested only in
the concept. Of course, if what is to be taken up into the concept from the
concrete appearance is intended to serve only as a mark or sign, then it may
well be anything at all, any mere sensuous singular determination of the
subject matter will do, selected from the others because of some external
interest but of like kind and nature as the rest.

In this conjunction, the prevailing fundamental
misunderstanding is that the natural principle,
or the starting point in the natural development
or the history of an individual in the process of self-formation,
is regarded as the truth and conceptually the first.
Intuition or being are no doubt first in the order of nature,
or are the condition for the concept,
but they are not for all that the unconditioned in and for itself;
on the contrary, in the concept their reality is sublated
and, consequently, so is also the reflective shine
that they had of being the conditioning reality.
If it is not the truth which is at issue but only narration,
as it is the case in pictorial and phenomenal thinking,
then we might as well stay with the story
that we begin with feelings and intuitions,
and that the understanding then extracts a universal
or an abstraction from their manifold,
for which purpose it quite understandably needs
a substrate for these feelings and intuitions
which, in the process of abstraction, retains
for representation the same complete reality
with which it first presented itself.
But philosophy ought not to be a narrative of what happens,
but a cognition of what is true in what happens,
in order further to comprehend on the basis of this truth
what in the narrative appears as a mere happening.

If on the superficial view of what the concept is all manifoldness falls
outside it, and only the form of abstract universality or of empty reflective
identity stays with it, we can at once call attention to the fact that any
statement or definition of a concept expressly requires, besides the genus
which in fact is already itself more than just abstract universality, also a
specific determinateness. And it does not take much thoughtful reflection on
the implication of this requirement to see that differentiation is an equally
essential moment of the concept. Kant introduced this line of reflection
with the very important thought that there are synthetic judgments
a priori. His original synthesis of apperception is one of the most pro-
found principles for speculative development; it contains the beginning of
a true apprehension of the nature of the concept and is fully opposed to any
empty identity or abstract universality which is not internally a synthesis. –
The further development, however, did not live up to this beginning. The
term itself, “synthesis,” easily conjures up again the picture of an external
unity, of a mere combination of terms that are intrinsically separate. Then,
again, the Kantian philosophy has never got over the psychological reflex
of the concept and has once more reverted to the claim that the concept is
permanently conditioned by the manifold of intuition. It has declared the
content of the cognitions of the understanding, and of experience, to be
phenomenal, not because of the finitude of the categories as such but, on the
ground of a psychological idealism, because they are only determinations
derived from self-consciousness. Here accordingly we have again the suppo-
sition that apart from the manifoldness of intuition the concept is without
content, empty, despite the fact that the concept is said to be a synthesis a
priori; as such, it surely contains determinateness and differentiation within
itself. And because this determinateness is the determinateness of the con-
cept, and hence the absolute determinateness, singularity, the concept is the
ground and the source of all finite determinateness and manifoldness.

The formal position that the concept never abandons as understanding
is completed in Kant's exposition of what reason is. One should expect that
in reason, which is the highest stage of thought, the concept would lose
the conditionality with which it still appears at the stage of understanding
and would attain perfect truth. But this expectation is disappointed. For
Kant defines the relation of reason to the categories as merely dialectical.
Indeed, he even takes the result of this dialectic to be simply and solely
an infinite nothingness, the result being that the synthesis is again lost, lost
also to the infinite unity of reason, and lost with it is whatever beginning
there was of a speculative, truly infinite, concept; reason becomes the well-
known, totally formal, merely regulative unity of the systematic employment
of the understanding. It is declared an abuse when Logic, which is supposed
to be a mere canon of judgment, is considered instead as an organon for
the production of objective insights. The concepts of reason, in which
we would have expected a higher power and a deeper content, no longer
possess anything constitutive as still do the categories; they are mere ideas
which we are of course are quite at liberty to use, provided that by these
intelligible entities in which all truth was to be revealed we mean nothing
more than hypotheses to which it would be the height of arbitrariness and
recklessness to ascribe absolute truth, for they – cannot be found in any
experience.
– Would anyone have ever thought that philosophy would
deny truth to intelligible entities on the ground that they lack the spatial
and temporal material of the senses?

Directly connected with this is the issue of how to view the concept
and the character of logic generally, the issue namely of the relation of
the concept and its science to truth itself. This is an issue on which the
Kantian philosophy holds the same position as is commonly taken. We
cited earlier from Kant's deduction of the categories to the effect that,
according to it, the object in which the manifold of intuition is unified is
this unity only by virtue of the unity of self-consciousness. 22 The objectivity of
thought is here, therefore, specifically defined: it is an identity of concept
and thing which is the truth. In the same way it is also commonly accepted
that, as thought appropriates a given subject matter, this subject matter
thereby undergoes an alteration and is made from something sensuous
into something thought. But nothing is changed in this alteration in so
far as the essentiality of the object goes; on the contrary, it is accepted that
the object is in its truth only in its concept, whereas in the immediacy in
which it is given it is only appearance and accidentality; that the cognition
conceptualizing the subject matter is a cognition of it as it is in and for itself,
and the concept is its very objectivity. But, on the other hand, it is also
equally claimed that we cannot know things as they are in and for themselves
and that truth is inaccessible to rational cognition; that the aforesaid truth
that would consist in the unity of the object and the concept is in fact
only appearance, again on the ground now that the content is only the
manifold of intuition. But we have just remarked, regarding this point,
that it is precisely in the concept that the manifold is sublated inasmuch
as it pertains to intuition as opposed to the concept, and that through the
concept the subject matter is reduced to its non-contingent essentiality;
the latter does enter into appearance, and this is why appearance is not
something merely essenceless but is the manifestation of essence. When
this manifestation of essence is set free, then we have the concept. – These
propositions that we are now recalling are not dogmatic assertions, for
they are results that obtained on their own out of the whole development
of essence. The present position to which this development has led is that
the form of the absolute which is higher than being and essence is the
concept. Viewed from this side, the concept has subjugated the spheres of
being and essence to which, from other starting points, feeling, intuition,
and representation, which appeared to be its antecedent conditions, also
belong; it has demonstrated itself to be their unconditional foundation. But
this is one side alone. There is a second side left to which this third book of
the Logic is devoted, namely the demonstration of how the concept forms
within and from itself the reality that has vanished in it. It is conceded, in
other words, that the cognition that does not go past the concept, purely
as concept, is still incomplete, that it has only arrived at abstract truth. But
its incompleteness does not lie in its lack of that alleged reality as would be
given in feeling and intuition, but in the fact that the concept has yet to give
to itself its own reality, one that it generates out of itself. The demonstrated
absoluteness of the concept as against the material of experience and, more
exactly, the categorial and the reflective determinations of it, consists in
this, that as this material appears outside and before the concept, it has no
truth but that it has it only in its ideality or in its identity with the concept.
The derivation of the real from the concept, if “derivation” is what we
want to call it, consists at first essentially in this, that the concept in its
formal abstraction reveals itself to be incomplete and through a dialectic
immanently grounded in it passes over into reality: it passes over into it,
however, as into something which it generates out of itself, not as if it
were falling back again onto a ready-made reality which it finds opposite
it, or as if it were taking refuge, because it sought for something better
but found none, into something that has already been proven to be the
unessential element of appearance. – It will always be a source of wonder
how the Kantian philosophy did acknowledge that the relation of thought
to sensuous existence (the relation at which it stopped) is only a relation of
mere appearance, and also well recognized in the idea in general a higher
unity of those two terms, even gave expression to it, as for example in the
idea of an intuitive understanding, and yet stopped short at that relative
relation and at the claim that the concept remains utterly separate from
reality – thus asserting as truth what it declared to be finite cognition, and
explaining away as extravagant and illegitimate figments of thought what
it recognized as truth and had specifically defined as such.

Since it is logic above all and not science generally whose relation to
truth is the issue here, it must be further conceded that logic as the formal
science cannot also contain, nor should contain, the kind of reality which
is the content of the other parts of philosophy, of the sciences of nature
and of spirit. These concrete sciences do attain to a more real form of the
idea than logic does, but not because they have turned back to the reality
which consciousness abandoned as it rose above the appearance of it to
science, or because they have again resorted to the use of such forms as
are the categories and the determinations of reflection, the finitude and
untruth of which were demonstrated in the logic. The logic rather exhibits
the rise of the idea up to the level from which it becomes the creator
of nature and passes over into the form of a concrete immediacy whose
concept, however, again shatters this shape also in order to realize itself as
concrete spirit. These sciences, just as they had the logic as their prototype,
hold on to its logical principle or the concept as in them their formative
factor. As contrasted with them, the logic is of course the formal science,
yet the science of the absolute form which is implicit totality and contains
the pure idea of truth itself. This absolute form has in it a content or reality
of its own; the concept, since it is not a trivial, empty identity, obtains its
differentiated determinations in the moment of negativity or of absolute
determining; and the content is only these determinations of the absolute
form and nothing else – a content posited by the form itself and therefore
adequate to it. – This form is for this reason of quite another nature than
logical form is ordinarily taken to be. It is truth already on its own account,
because this content is adequate to its form or this reality to its concept,
and it is pure truth, because the determinations of the content do not yet
have the form of an absolute otherness or of absolute immediacy. – When
Kant in the Critique of Pure Reason (p. 83), 23 in connection with logic
comes to discuss the old and famous question: What is truth?, he starts by
passing off as a triviality the nominal definition that it is the agreement of
cognition with its subject matter – a definition which is of great, indeed of
supreme value. If we recall this definition together with the fundamental
thesis of transcendental idealism, namely that rational cognition is incapable
of comprehending things in themselves, 24 that reality lies absolutely outside
the concept, it is then at once evident that such a reason, one which is
incapable of setting itself in agreement with its subject matter, and the things
in themselves, such as are not in agreement with the rational concept – a
concept that does not agree with reality and a reality that does not agree
with the concept – that these are untrue conceptions. If Kant had measured
the idea of an intuitive understanding against that first definition of truth,
he would have treated that idea which expresses the required agreement,
not as a figment of thought but rather as truth.
“What we would want to know,” Kant proceeds to say, “is a universal
and certain criterion of truth of any cognition whatever, one that would be
valid for all cognitions without distinction of their subject matters; but since
any such criterion would abstract from all content of cognition (the reference
to its object), and truth has to do precisely with this content, it would be quite
impossible, even absurd, to ask for a mark of the truth of this content of
cognitions.” 25 – Here we have, clearly expressed, the ordinary conception
of the formal function of logic which gives to the adduced argument the air
of convincing. But first of all it is to be noted what usually happens to this
kind of formal argumentation: it forgets as it speaks that on which it is based
and of which it speaks. It would be absurd, it says, to ask for a criterion
of the truth of the content of cognition. But according to the definition on
which it is based it is not the content that constitutes the truth, but the
agreement of it with the concept. Such a content as is here spoken of, one
without the concept, is something void of concept and therefore void of
essence; of course, we cannot ask of such a content for a criterion of truth,
but for the opposite reason, namely, not because it cannot be the required
agreement on account of its being void of concept, but because it cannot
be anything more than just another truthless opinion. – Let us leave aside
any talk about content, which is the cause of the confusion here – the
confusion in which formalism invariably falls, and which is responsible for
making it say, every time it tries to explain itself, the opposite of what it
wants to say – and let us just stay with the abstract view that the logic
is only formal, that it abstracts from all content. What we then have is a
one-sided cognition which is not supposed to contain any subject matter,
an empty form void of determination which is therefore just as little an
agreement (for it necessarily takes two for an agreement) as it is truth. –
In the a priori synthesis of the concept, Kant did have a higher principle in
which it was possible to recognize a duality and therefore what is required
for truth; but the material of the senses, the manifoldness of intuition, was
too strong for him to be able to wrest himself away from it and turn to a
consideration of the concept and the categories in and for themselves, and
to a speculative form of philosophizing.

Since logic is the science of the absolute form, this formal discipline,
in order to be true, must have a content in it which is adequate to its form;
all the more so, because logical form is pure form and
logical truth, accordingly, the pure truth itself.

This formal discipline must therefore be thought of
as inherently much richer in determinations and content,
and also of infinitely greater efficacy over the concrete,
than it is normally taken to be.

The laws of logic by themselves
(extraneous elements aside, such as applied logic and
the rest of the psychological and anthropologicalmaterial)
are commonly restricted, apart from the law of contradiction,
to a few meager propositions concerning the conversion of judgments
and the forms of inference.

And the forms, too, that come up in this context,
 as well
as their further specifications,
 are only taken up historically as it were,
  not
subjected to criticism to see whether they are in and for themselves true.
For example, the form of the positive judgment is accepted as something
perfectly correct in itself,
and whether the judgment is true is made to
depend solely on the content.
 No thought is given to investigating whether
this form of judgment is a form of truth in and for itself;
whether the
proposition it enunciates, “
the individual is a universal,” is not inherently
dialectical.
It is at once assumed that the judgment is capable of possessing
truth on its own account,
 and that every proposition expressed in a positive
judgment is true, even though it is patently evident that the judgment lacks
what is required by the definition of truth, namely the agreement of the
concept with its subject matter; for if the predicate, which here is the
universal, is taken as the concept,
and the subject, which is the singular,
as the subject matter,
then the concept does not agree with it.
 But if the
abstract universal which is the predicate does not yet amount to a concept
(for surely there is more that belongs to it);
or if the subject, for its part,
still is not much more than a grammatical one, how should the judgment
possibly contain truth seeing that its concept and the intended object do
not agree, as also that the concept is missing and indeed the object as well?
This rather is then where the impossible and the absurd lie, in the attempt
to grasp the truth in such forms as are the positive judgment or a judgment
in general.
Just as the Kantian philosophy did not consider
the categories in and for themselves,
but declared them to be finite determinations
unfit to hold the truth,
on the only inappropriate ground that
they are subjective forms of self-consciousness,
still less did it subject to criticism the forms
of the concepts that make up the content of ordinary logic.
What it did, rather, is to pick a portion of them,
namely the functions of judgments,
for the determination of categories,
and simply accepted them as valid presuppositions.
Even if there were nothing more to the forms of logic
than these formal functions of judgment,
for that reason alone they would already be
worthwhile investigating to see how far, by themselves,
they correspond to the truth.
A logic that does not perform this task can at most claim
the value of a natural description of the phenomena of thought
as they simply occur.
It is an infinite merit of Aristotle,
one that must fill us with the highest admiration
for the power of his genius,
that he was the first to undertake this description.
But it is necessary to go further
and determine both the systematic connection
of these forms and their value.

DIVISION

The concept, as considered so far,
has demonstrated itself to be
the unity of being and essence.
Essence is the first negation of being,
which has thereby become reflective shine;
the concept is the second negation,
or the negation of this negation,
and is therefore being
which has been restored once more, but as in itself
the infinite mediation and negation of being.
In the concept, therefore, being and essence
no longer have determination as being and essence,
nor are they only in such a unity in which
each would reflectively shine in the other.
Consequently, the concept does not differentiate
itself into these determinations.
The concept is the truth of the substantial relation
in which being and essence attain their perfect
self-subsistence and determination each through the other.
The truth of substantiality proved to be the substantial identity,
an identity that equally is, and only is, positedness.
Positedness is determinate existence and differentiation;
in the concept, therefore, being-in-and-for-itself
has attained a true existence adequate to it,
for that positedness is itself being-in-and-for-itself.
This positedness constitutes the difference
of the concept in the concept itself;
and because the concept is immediately being-in-and-for-itself,
its differences are themselves the whole concept,
universal in their determinateness
and identical in their negation.

This is now the concept itself of the concept,
but at first only the concept
of the concept or also itself only concept.
Since the concept is being-in-and-
for-itself
 by being a positedness, or is absolute substance,
  and substance
manifests the necessity of distinct substances as an identity,
this identity must
itself posit what it is.
The moments of the movement of the substantial
relation through which the concept came to be and the reality thereby
exhibited are only in the transition to the concept;
that reality is not yet the
concept’s own determination,
one that has emerged out of it;
it fell in the
sphere of necessity whereas the reality of the concept can only be its free
determination, a determinate existence in which the concept is identical
with itself
and whose moments are themselves concepts posited through the
concept itself.

At first, therefore, the concept is only implicitly the truth;
because it is only something inner, it is equally only something outer.
It is at first simply
an immediate and in this shape its moments have the form of immediate,
fixed determinations. It appears as the determinate concept, as the sphere of
mere understanding. – Because this form of immediacy is an existence still
inadequate to the nature of the concept, for the concept is free and only
refers to itself, it is an external form in which the concept does not exist
in-and-for-itself, but can only count as something posited or subjective. –
The shape of the immediate concept constitutes the standpoint that makes
of the concept a subjective thinking, a reflection external to the subject
matter. This stage constitutes, therefore, subjectivity, or the formal concept.
Its externality is manifested in the fixed being of its determinations that
makes them come up each by itself, isolated and qualitative, and each only
externally referred to the other. But the identity of the concept, which is
precisely their inner or subjective essence, sets them in dialectical movement,
and through this movement their singleness is sublated and with it also
the separation of the concept from the subject matter, and what emerges
as their truth is the totality which is the objective concept.

Second, in its objectivity the concept is the fact itself as it exists in-and-
for-itself. The formal concept makes itself into the fact by virtue of the
necessary determination of its form, and it thereby sheds the relation
of subjectivity and externality that it had to that matter. Or, conversely,
objectivity is the real concept that has emerged from its inwardness and has
passed over into existence. – In this identity with the fact, the concept
thus has an existence which is its own and free. But this existence is still a
freedom which is immediate and not yet negative. Being at one with the
subject matter, the concept is submerged into it; its differences are objective
determinations of existence in which it is itself again the inner. As the soul
of objective existence, the concept must give itself the form of subjectivity
that it immediately had as formal concept; and so, in the form of the free
concept which in objectivity it still lacked, it steps forth over against that
objectivity and, over against it, it makes therein the identity with it, which
as objective concept it has in and for itself, into an identity that is also posited.

In this consummation in which the concept has the form of freedom
even in its objectivity, the adequate 27 concept is the idea. Reason, which is
the sphere of the idea, is the self-unveiled 28 truth in which the concept
attains the realization absolutely adequate to it, and is free inasmuch as in
this real world, in its objectivity, it recognizes its subjectivity, and in this
subjectivity recognizes that objective world.

SECTION I

Subjectivity

The concept is, to start with, formal,
the concept in its beginning
or as the immediate concept.
In this immediate unity, its difference
or its positedness is, first,
itself initially simple and only a reflective shine,
so that the moments of the difference are immediately
the totality of the concept and only the concept as such.

But, second, because it is absolute negativity, the concept divides and
posits itself as the negative or the other of itself; yet, because it is still
immediate concept, this positing or this differentiation is characterized by
the reciprocal indifference of its moments, each of which comes to be on
its own; in this division the unity of the concept is still only an external
connection. Thus, as the connection of its moments posited as self-subsisting
and indifferent, the concept is judgment.

Third, although the judgment contains the unity of the concept that
has been lost in its self-subsisting moments, this unity is not posited. It
will become posited by virtue of the dialectical movement of the judgment
which, through this movement, becomes syllogistic inference, 30 and this is
the fully posited concept, for in the inference the moments of the concept
as self-subsisting extremes and their mediating unity are both equally posited.

But since this unity itself, as unifying middle, and the moments, as self-
subsisting extremes, stand at first immediately opposite one another, this
contradictory relation that occurs in the formal inference sublates itself, and
the completeness of the concept passes over into the unity of totality; the
subjectivity of the concept into its objectivity.

CHAPTER 1

The concept

The faculty of concepts is normally associated with the understanding, and
the latter is accordingly distinguished from the faculty of judgment and
from the faculty of syllogistic inferences which is formal reason. But it is
particularly with reason that the understanding is contrasted, and it signifies
then, not the faculty of concepts in general, but the faculty of determinate
concepts, as if, as the prevailing opinion has it, the concept were only a
determinate. When distinguished in this meaning from the formal faculty
of judgment and from formal reason, the understanding is accordingly to
be taken as the faculty of the single determinate concept. For the judgment
and the syllogism or reason, as formal, are themselves only a thing of the
understanding, 31 since they are subsumed under the form of the abstract
determinateness of the concept. Here, however, we are definitely not taking
the concept as just abstractly determined; the understanding is therefore
to be distinguished from reason only in that it is the faculty of the concept
as such.

This universal concept that we now have to consider contains the three
moments of universality, particularity, and singularity. The difference and
the determinations which the concept gives itself in its process of distin-
guishing constitute the sides formerly called positedness. Since this posit-
edness is in the concept identical with being-in-and-for-itself, each of the
moments is just as much the whole concept as it is determinate concept and
a determination of the concept.

It is at first pure concept, or the determination of universality. But the
pure or universal concept is also only a determinate or particular concept
that takes its place alongside the other concepts. Because the concept is
a totality, and therefore in its universality or pure identical self-reference
is essentially a determining and a distinguishing, it possesses in itself the
norm by which this form of its self-identity, in pervading all the moments
Georg Wilhelm Friedrich Hegel
and comprehending them within, equally determines itself immediately as
being only the universal as against the distinctness of the moments.

Second, the concept is thereby posited as this particular or determinate
concept, distinct from others.

Third, singularity is the concept reflecting itself out of difference into
absolute negativity. This is at the same time the moment at which it has
stepped out of its identity into its otherness and becomes judgment.

CHAPTER 2

Judgment

Judgment is the determinateness of the concept posited in the concept itself.
The determinations of the concept, or, what amounts to the same thing
as shown, 47 the determinate concepts, have already been considered on
their own; but this consideration was rather a subjective reflection or a
subjective abstraction. But the concept is itself this act of abstracting;
the positioning of its determinations over against each other is its own
determining. Judgment is this positing of the determinate concepts through
the concept itself.

Judging is therefore another function than conceiving; or rather, it is the
other function of the concept, for it is the determining of the concept through
itself. The further progress of judgment into a diversity of judgments is
this progressive determination of the concept. What kind of determinate
concepts there are, and how they prove to be necessary determinations of
it – this has to be exhibited in judgment.

Judgment can therefore be called the first realization of the concept,
for reality denotes in general the entry into existence as determinate being.
More precisely, the nature of this realization has presented itself in such
a way that the moments of the concept are totalities which, on the one
hand, subsist on their own through the concept's immanent reflection or
through its singularity; on the other hand, however, the unity of the concept
is their connection. 48 The immanently reflected determinations are determi-
nate totalities that exist just as essentially disconnected, indifferent to each
other, as mediated through each other. The determining itself is a totality
only as containing these totalities and their connections. This totality is the
judgment. – The latter contains, therefore, the two self-subsistents which
go under the name of subject and predicate. What each is cannot yet be said;
they are still indeterminate, for they are to be determined only through the
judgment. Inasmuch as judgment is the concept as determinate, the only
determination at hand is the difference that it contains between determinate
and still indeterminate concept. As contrasted to the predicate, the subject
can at first be taken, therefore, as the singular over against the universal, or
also as the particular over against the universal, or the singular over against
the particular; so far, they stand to each other only as the more determinate
and the more universal in general.

It is therefore fitting and unavoidable to have these names, “subject”
and “predicate,” for the determinations of the judgment; as names, they
are something indeterminate, still in need of determination, and therefore
nothing but names. It is partly for this reason that the determinations
themselves of the concept could not be used for the two sides of judgment;
but a still stronger reason is because of the nature of a concept determination
which is nothing abstract, nothing fixed, but contains its opposite in it,
explicitly posited there; since the sides of the judgment are themselves
concepts and therefore the totality of the determinations of the concept,
each side must run through all these determinations, exhibiting them
within whether in abstract or concrete form. But now, if in this altering of
determination we want to fix the two sides in some general way, names will
be the most useful means, for they can be kept the same throughout the
process. – But a name remains distinct from the fact or the concept, and this
is a distinction that transpires within the judgment as such; since the subject
is in general the determinate term and more, therefore, of an immediate
existent, whereas the predicate expresses the universal, the essence or the
concept, the subject as such is at first only a kind of name; what it is, is first
enunciated only by the predicate which contains being in the sense of the
concept. When we ask, “What is this?,” or “What kind of plant is this?,”
the being we are enquiring about is often just a name, and once we learn
this name, we are satisfied that we now know what the fact is. This is being
in the sense of the subject. The concept, however, or at least the essence and
the universal in general, is only given by the predicate, and when we ask for
it, we do it in the sense of the judgment. – God, therefore, or spirit, nature,
or what have you, is as the subject of a judgment only a name at first; what
any such subject is in accordance with the concept, is first found only in
the predicate. When we ask for the predicate that belongs to such subjects,
the required judgment must be based on a concept that is presupposed; yet
it is the predicate that first gives this concept. It is, therefore, the mere
representation that in fact makes up the presupposed meaning, and this
yields only a nominal definition whereby it is a mere accident, a historical
fact, what is understood by a name. So many disputes about whether a
predicate does or does not belong to a subject are, therefore, nothing more
than verbal disputes, for they proceed from this form; what lies at the base
(subjectum, Ëpokeimenon) is still nothing more than a name.
Secondly, we now have to examine more closely how the connection of
subject and predicate in judgment is determined, and how the two are
themselves thereby determined. Judgment has in general totalities for its
sides, totalities that are at first essentially self-subsistent. The unity of the
concept is at first, therefore, only a connection of self-subsistent terms; it is
not yet the concrete, the fulfilled unity that has returned into itself from this
reality but is a unity rather outside which the two terms persist as extremes
yet unsublated in it. – Now any consideration of the judgment can start
either from the originative unity of the concept or from the self-subsistence
of the extremes. Judgment is the self-diremption of the concept; therefore,
it is by starting from the unity of the concept as ground that the judgment is
considered in accordance with its true objectivity. In this respect, judgment
is the originative division (or Teilung, in German) of an originative unity; the
German word for judgment, Urteil (or “primordial division”), thus refers
to what judgment is in and for itself. But the concept is present in the
judgment as appearance, since its moments have attained self-subsistence
there, and it is to this side of externality that ordinary representation is more
likely to fasten.

From this subjective standpoint, the subject and the predicate are there-
fore treated as ready-made, each for itself outside the other – the subject as
a subject matter that would exist even if it did not have that predicate, and
the predicate as a universal determination that would exist even without
accruing to this subject. The act of judgment accordingly brings with it
the further reflection whether this or that predicate which is in someone's
head can and should be attached to the subject matter that exists outside it
on its own; the judgment itself is simply the act that combines the predicate
with the subject, so that, if this combination did not occur, the subject
and predicate would still each remain what it is, the one concretely existing
as thing in itself, the other as a representation in someone's head. – But
the predicate which is combined with the subject should also pertain to it,
which is to say, should be in and for itself identical with it. The significance
of their being combined is that the subjective sense of judgment, and the
indifferent external persistence of the subject and predicate, are again sub-
lated. Thus in “this action is good,” the copula indicates that the predicate
belongs to the being of the subject and is not merely externally combined
with it. Of course, grammatically speaking this kind of subjective relation
that proceeds from the indifferent externality of subject and predicate is
perfectly valid, for it is words that are here externally combined. – It can also
be mentioned in this context that a proposition can indeed have a subject
and predicate in a grammatical sense without however being a judgment
for that. The latter requires that the predicate behave with respect to the
subject in a relation of conceptual determination, hence as a universal with
respect to a particular or singular. And if what is said of a singular sub-
ject is itself only something singular, as for instance, “Aristotle died at the
age of 73 in the fourth year of the 115th Olympiad,” then this is a mere
proposition, not a judgment. There would be in it an element of judgment
only if one of the circumstances, say, the date of death or the age of the
philosopher, came into doubt even though the stated figures were asserted
on the strength of some ground or other. In that case, the figures would be
taken as something universal, as a time that, even without the determinate
content of Aristotle's death, would still stand on its own filled with some
other content or simply empty. Likewise would the news that my friend
N. has died be a proposition, and a judgment only if there were a question
as to whether he is actually dead and not just apparently dead.
In the usual definition of judgment, that it is the combination of two
concepts, we may indeed accept the vague expression of “combination” for
the external copula, and also accept that the terms combined are at least
meant to be concepts. But the definition is otherwise a highly superficial
one. It is not just that in the disjunctive judgment, for instance, there are
more than two so-called concepts that are combined; more to the point is
rather that the definition is much better than the matter defined, for it is not
determinations of concepts, but determinations of representation that are in
fact meant; it was remarked in connection with the concept in general, and
with the concept as determinate, that what usually goes under this name
of concept does not deserve the name at all; 49 where should concepts then
come from in the case of judgment? – Above all this definition of judgment
ignores what is essential to it, namely the difference of its determinations;
still less does it take into account its relation to the concept.
As regards the further determination of the subject and predicate, we
have remarked above 50 that it is in judgment that they must first receive
their determination. But since judgment is the posited determinateness
of the concept, this determinateness possesses the given differences imme-
diately and abstractly as singularity and universality. – But inasmuch as
judgment is in general the immediate existence or the otherness of the con-
cept that has not yet restored itself to the unity through which it exists as
concept, there also emerges the determinateness that is void of concept,
the opposition of being and reflection or the in-itself. But since the concept
constitutes the essential ground of judgment, these determinacies are at least
indifferent in the sense that, when one accrues to the subject and the other
to the predicate, the converse relation equally holds. The subject, being
the singular, appears at first as the existent or as the one that exists for itself
with the determinate determinateness of a singular on which judgment is
passed – as an actual object even when it is such in representation only –
as for instance in the case of bravery, right, agreement, etc. The predicate,
which is the universal, appears on the contrary as the reflection of this
judgment on that object, or rather as the object's immanent reflection that
transcends the immediacy of the judgment and sublates its determinacies
as mere existents – appears, that is, as the object's in-itselfness. – In this way,
the start is made from the singular as the first, the immediate, and through
the judgment this singular is raised to universality, just as, conversely, the
universal that exists only in itself descends in the singular into existence or
becomes a being that exists for itself.

This significance of the judgment is to be taken as its objective meaning
and at the same time as the true significance of the previous forms of
transition. The existent comes to be and becomes another, the finite passes
over into the infinite and in it passes away; the existent comes forth into
appearance out of its ground and to this ground it founders; the accidents
manifest the wealth of substance as well as its might; in being, there is
transition into an other; in essence, there is the reflective shining in an
other that manifests the necessity of a connection. This transition and this
reflective shining have now passed over into the originative division of the
concept in judgment, and this division, in bringing the singular back to the
in-itselfness of its universality, equally determines the universal as something
actual. These two are one and the same – the positing of singularity in its
immanent reflection and of the universal as determinate.

But equally pertaining to this objective meaning is that the said differ-
ences, as they re-occur in the determinateness of the concept, are at the
same time posited as only appearing, that is to say, that they are nothing
fixed but accrue rather just as much to one determination of the concept as
to the other. The subject is therefore equally to be taken as the in-itself, and
the predicate as determinate existence in contrast to it. The subject without
the predicate is what the thing without properties, the thing-in-itself, is in
the sphere of appearance, an empty indeterminate ground; it is then the
implicit concept that receives a difference and a determinateness only in
the predicate; the predicate thus constitutes the side of the determinate
existence of the subject. Through this determinate universality the subject
refers to the outside, is open to the influence of other things and thereby
confronts them actively. What is there comes forth from its in-itselfness into
the universal element of combination and relations, into negative references
and into the interplay of actuality which is a continuation of the singular
into other singulars and is, therefore, universality.
Yet the identity just indicated, the fact that the determination of the
subject accrues equally to the predicate and vice versa, is not just a matter
for our consideration; it is not only in itself but is also posited in the
judgment; for the judgment is the reference connecting the two; the copula
expresses that the subject is the predicate. The subject is the determinate
determinateness, and the predicate is this determinateness of the subject as
posited; the subject is determined only in its predicate, or is subject only in
it; in the predicate, it is turned back into itself and is therein the universal. –
Now in so far as the subject is the self-subsistent term, this identity has the
relation that the predicate does not possess a self-subsistence of its own but
has its subsistence only in the subject; it inheres in the subject. Accordingly,
since the predicate is distinguished from the subject, it is only a singularized
determinateness of the subject, only one of its properties; the subject itself
is however the concrete, the totality of manifold determinacies, just as the
predicate contains one of them; the subject is the universal. – But, on the
other hand, the predicate also is self-subsistent universality, and the subject
conversely only one determination of it. The predicate thus subsumes the
subject; the singularity and the particularity are not for themselves but
have their essence and their substance in the universal. The predicate
expresses the subject in its concept; the singular and the particular are to the
subject accidental determinations; the subject is their absolute possibility.
When by “subsumption” an external connection of subject and predicate
is thought, and the subject is represented as something self-subsistent,
then subsumption refers to the subjective act of judging mentioned above,
namely the judging that starts off from the self-subsistence of both subject
and predicate. 51 Subsumption is then only the application of the universal to
a particular or singular posited under it in accordance with an indeterminate
representation, one of lesser quantity.

When we treat the identity of subject and predicate as meaning that at
one time one determination of the concept belongs to the subject and the
other to the predicate, and at another time the converse equally applies,
then the identity is as yet still implicit; on account of the self-subsistent
diversity of the two sides of judgment, their posited connection also has
the two at first as diverse. But it is the identity void of difference that in fact
constitutes the true connection of the subject and predicate. The determi-
nation of the concept is itself essentially a connection, for it is a universal; the
same determinations, therefore, which the subject and the predicate each
have, are also had by their connection. The connection is universal, for it
is the positive identity of both, of the subject and predicate; but it is also
determinate, for the determinateness of the predicate is the determinateness
of the subject; it is singular as well, for in it the self-subsisting extremes are
sublated as in their negative unity. – In judgment, however, this identity
is not posited yet; the copula is as the still indeterminate connection of
being in general, “A is B,” for the self-subsistence of the concept's deter-
minacies, or the extremes, is in judgment the reality that the concept has
within. If the “is” of the copula were already posited as the determinate
and fulfilled unity of subject and predicate earlier mentioned, 52 were
posited as their concept, it would then already be the conclusion of syllogistic
inference.

To restore again this identity of the concept, or rather to posit it – this
is the goal of the movement of the judgment. What is already present in
the judgment is, on the one hand, the self-subsistence but also reciprocal
determinateness of the subject and predicate, and, on the other hand,
their still abstract connection. “The subject is the predicate” – this is what
the judgment says at first. But since the predicate is not supposed to be
what the subject is, a contradiction is at hand that must resolve itself, must
pass over into a result. Or rather, since the subject and predicate are in
and for themselves the totality of the concept, and judgment is the reality
of the concept, the judgment's forward movement is only development;
what comes forth from it is already present in it, and to this extent the
demonstration is a display, 54 a reflection as the positing of that which is already
at hand in the extreme terms of the judgment; but even this positing is
already present; it is the connection of the extremes.

First, as immediate, judgment is the judgment of existence; its subject is
immediately an abstract, existent singular, and the predicate is an immediate
determinateness or property of it, an abstract universal.

Second, as this qualitative character of the subject and predicate is sub-
lated, the determination of the one begins to shine reflectively in the other;
the judgment is now the judgement of reflection.

But this external combination passes over into the essential identity of a
substantial, necessary combination; and so we have, third, the judgment of
necessity.

Fourth, since in this essential identity the difference of subject and
predicate has become a form, the judgment becomes subjective; it entails
the opposition of the concept and its reality and the comparison of the two;
it is the judgment of the concept.

This emergence of the concept grounds the transition of judgment into
syllogistic inference.

CHAPTER 3

The syllogism

The syllogism is the result of the restoration of the concept in the judgment,
and consequently the unity and the truth of the two. The concept as such
holds its moments sublated in this unity; in judgment, the unity is an
internal or, what amounts to the same, an external one, and although the
moments are connected, they are posited as self-subsisting extremes. In the
syllogism, the determinations of the concept are like the extremes of
the judgment, and at the same time their determinate unity is posited.

Thus the syllogism is the completely posited concept; it is, therefore, the
rational. – The understanding is taken to be the faculty of the determinate
concept which is held fixed for itself by virtue of abstraction and the form
of universality. But in reason the determinate concepts are posited in their
totality and unity. Therefore, it is not just that the syllogism is rational but
that everything rational is a syllogism. Syllogistic inference has long since
been ascribed to reason; but, on the other hand, reason in and for itself,
and rational principles and laws, are so spoken of that no light is thrown
on why the one reason that syllogizes, and the other which is the source
of laws and otherwise eternal truths and absolute thoughts, hang together.
If the former is supposed to be only a formal reason while the latter is
supposed to be the one that generates content, then one would expect on
this distinction that precisely the form of reason, the inference, would not
be missing in the latter. And yet, the two are commonly held so far apart,
the one without mention of the other, that it seems as though the reason
of absolute thoughts were ashamed, so to speak, of inferential reason, and
the syllogism were listed as also an activity of reason merely as matter of
tradition. But surely, as we have just remarked, logical reason must be
essentially recognizable, when regarded as formal, also in the reason that
is concerned with a content; indeed, no content can be rational except
by virtue of the rational form. In this matter we cannot rely on what is
commonly said about reason, for common views fail to tell us what we are
to understand by reason; this would-be rational wisdom is so busy with its
objects that it forgets to pay attention to reason itself but only identifies it
by characterizing it through the objects that it is said to have. If reason is
supposed to be a cognition that would know about God, freedom, right and
duty, the infinite, the unconditional, the suprasensible, or even gives only
representations and feelings of such objects, then for one thing these objects
are only negative, and for another the original question still stands, what
is there in all these objects that makes them rational? – The answer is that
the infinitude in them is not the empty abstraction from the finite, is not a
universality which is void of content and determination, but is the fulfilled
universality, the concept which is determined and is truly in possession of its
determinateness, namely, in that it differentiates itself internally and is the
unity of its thus intelligible and determined differences. Only in this way
does reason rise above the finite, the conditioned, the sensuous, or however
one might define it, and is in this negativity essentially replete with content,
for as unity it is the unity of determinate extremes. And so the rational is
nothing but the syllogism.

Now the syllogism, like judgment, is at first immediate; as such, its
determinations (termini) are simple, abstract determinacies; it is then the
syllogism of the understanding. If one stays at this configuration of the
syllogism, then its rationality, though present there and posited, is not
apparent. The essential element of the syllogism is the unity of the extremes,
the middle term that unites them and the ground that supports them.
Abstraction, by holding fast to the self-subsistence of the extremes, posits this
unity opposite them, as a determinateness with just as fixed an existence of its
own, thus grasping it more as a non-unity than as a unity. The expression,
“middle term” (medius terminus), is derived from spatial representation,
and has its share of responsibility for why we stop short at the externality
of the terms. Now if the syllogism consists in the positing in it of the unity
of the extremes, but if this unity is simply taken on the one hand as a
particular by itself, and on the other hand as only an external connection,
and non-unity is made the essential relation of syllogism, then the reason
of the syllogism is of no help to rationality.

First, the syllogism of existence, in which the terms are thus immediately
and abstractly determined, demonstrates internally that, since like judg-
ment it is the connection of those terms, these are not in fact abstract but
each contains in it the reference connecting it to the others, and the deter-
mination of the middle term is not just a determinateness opposed to the
determinations of the extremes but contains these extremes posited in it.

Through this dialectic, the syllogism of existence becomes the syllogism
of reflection, the second syllogism. Its terms are such that in each the other
shines essentially reflected in it, or are posited as mediated, as they are indeed
supposed to be in accordance with the nature of syllogistic inference in
general.

Third, inasmuch as this reflective shining or this mediatedness is reflected
into itself, syllogism is determined as the syllogism of necessity, one in which
the mediating factor is the objective nature of the fact. As this syllogism
determines the extremes of the concept also as totalities, it has attained the
correspondence of its concept (or the middle term) and its existence (or
the difference of the extremes). It has attained its truth – and with that it
has stepped forth out of subjectivity into objectivity.
SECTION II

Objectivity

In Book One of the Objective Logic,
abstract being was presented as
passing over into existence,
but at the same time as
retreating into essence.

In Book Two, essence shows itself as
determining itself as ground,
thereby stepping into concrete existence
and realizing itself as substance,
but at the same time
retreating into the concept.

Of the concept, we have now first shown
that it determines itself as objectivity.
It should be obvious that this latter transition is
essentially the same as the proof from the concept,
that is to say, from the concept of God to his existence,
that was formerly found in
Metaphysics, or the so-called ontological proof.
Equally well known is that
Descartes's sublimest thought,
that God is that whose concept includes his
being within itself,
after having degenerated into the bad form of the formal
syllogism, namely into the form of the said proof, finally succumbed to
the Critique of Reason and to the thought that existence cannot be extracted
from the concept.
Some elucidations concerning this proof have already
been made earlier. In Volume I, pp. 47 ff., 3 where being has vanished into
its closest opposite, non-being,
and becoming has shown itself to be the
truth of both,
attention was called to the confusion that arises in the case
of a determinate existence when we concentrate, not on its being,
but on its determinate content,
and then imagine if we compare this determinate content
(e.g. one hundred dollars) with another determinate content
(e.g. the context of my perception, of my financial situation)
and discover that it makes indeed a difference whether
the one content is added to the other or not;
that we are dealing with the distinction of being and non-being,
or even the distinction of being and the concept.
Further, in the same Volume
on pp. 64ff. and on p. 289 of Volume II,
the definition of a sum-total of all reality
which occurs in the ontological proof was elucidated.
But the essential subject matter of that proof,
the connectedness of concept and existence,
is the concern of the treatment of the concept just concluded
and of the entire course that the latter traverses
in determining itself to objectivity.
The concept, as absolutely self-identical negativity,
is self-determining;
it was noted that the concept,
in resolving itself into judgment in singularity,
already posits itself as something real, an existent;
this still abstract reality completes itself in objectivity.

Now it might appear that the transition from the concept into objectivity
is quite another thing than the transition from the concept of God to
God's existence.
But, on the one hand, it must be borne in mind that the
determinate content, God, makes no difference in a logical progression
, and
that the ontological proof is only one application of this logical progression
to that particular content.
On the other hand, it is essential to be reminded
of the remark made above that the subject obtains determinateness and
content only in its predicate;
that prior to the predicate, whatever that
content might otherwise be for feeling, intuition, and representation,
 so
far as conceptual cognition is concerned it is only a name;
 but in the
predicate, with determinateness, there begins at the same time the process
of realization in general.
The predicates, however, must be grasped as
themselves still confined within the concept,
 hence as something subjective
with which no move to existence has yet been made;
even for this reason,
in judgment the realization of the concept is certainly not completed yet.
But there is the further reason that the mere determination of a subject
matter through predicates,
 without this determination being at the same
time the realization and objectification of the concept,
 remains something
so subjective that it is not even a true cognition and determination of the
concept of the subject matter –
“subjective” in the sense of abstract reflection
and non-conceptual representation. –
God as living God, and better still as
absolute spirit, is only recognized in what he does.
 Humankind were directed
early to recognize God in his works; only from these can the determinations
proceed that can be called his properties, and in which his being is also
contained.
It is thus the conceptual comprehension of God's activity, that
is to say, of God himself,
that recognizes the concept of God in his being and
his being in his concept.
Being by itself, or even existence, are such a poor and
restricted determination, that the difficulty of finding them in the concept
may well be due to not having considered what being or existence themselves
are.
 Being as entirely abstract, immediate self-reference, is nothing but the
abstract moment of the concept;
 it is its moment of abstract universality
that also provides what is required of being,
 namely that it be outside the
concept, for inasmuch as universality is a moment of the concept,
 it is
also its difference or the abstract judgment
wherein the concept opposes itself to itself.
The concept, even as formal, already immediately contains
being in a truer and richer form,
in that, as self-referring negativity, it is singularity.

But of course the difficulty of finding
being in the concept in general,
and equally so in the concept of God,
becomes insuperable if we expect being to
be something
that we find in the context of external experience
or in the form
of sense-perception,
like the one hundred dollars in the context of my finances,
as something graspable only by hand,
not by spirit, essentially visible to
the external and not the internal eye;
in other words, if the name of being,
reality, truth, is given to that which things possess as sensuous, temporal,
and perishable.
The consequence of a philosophizing that in regard to
being fails to rise above the senses is
that, in regard to the concept, it also
fails to let go of merely abstract thought;
such thought stands opposed to being.

The customary practice of regarding the concept as something just as
one-sided as abstract thought will already stand in the way of accepting what
has just been suggested,
namely, that we regard the transition of the concept
of God to his being as an application of the logical course of objectification
of the concept presented above.
 Yet if it is granted, as it commonly is,
that the logical element, as the formal element, constitutes the form for
the cognition of every determinate content, then that application at least
would have to be conceded, unless even at the opposition of concept and
objectivity in general one stops short at the untrue concept and an equally
untrue reality as an ultimate.
– But in the exposition of the pure concept it
was further indicated that the latter is the absolute divine concept itself.
In truth, therefore, what takes place is not a relation of application but
the immediate display in the logical course of God's self-determination as
being.
 But on this point it is to be remarked that inasmuch as the concept
is to be presented as the concept of God,
 it ought be apprehended as it is
when already taken up in the idea.
The said pure concept passes through
the finite forms of the judgment and the syllogism precisely because it is
not yet posited in and for itself as one with objectivity,
but is conceived
rather only in the process of becoming that objectivity.
 The latter, too, is
not yet the divine concrete existence,
not yet the reality reflectively shining
in the idea.
 And yet objectivity is just that much richer and higher than the
being or existence of the ontological proof,
as the pure concept is richer and
higher than that metaphysical vacuum of the sum-total of all reality.
But
I reserve for another occasion the task of elucidating in greater detail the
manifold misunderstanding brought upon the ontological proof of God's
existence,
and also on the rest of the other so-called proofs,
by logical formalism.
We shall also elucidate Kant's critique
of such proofs in order to establish their true meaning
and thus restore the thoughts on which they are based
to their worth and dignity.

We have previously called attention to the several forms of immediacy
that have already come on the scene, but in different determinations. 8 In
the sphere of being, immediacy is being itself and existence; in the sphere of
essence, it is concrete existence and then actuality and substantiality; in the
sphere of the concept, besides being immediacy as abstract universality, it is
now objectivity. – These expressions, when the exactitude of philosophical
conceptual distinctions is not at stake, may be used as synonymous; but
the determinations are derived from the necessity of the concept. Being
is as such the first immediacy, and existence is the same immediacy with
a first determinateness. Concrete existence, along with the thing, is the
immediacy that proceeds from ground, from the self-sublating mediation
of the simple reflection of essence. But actuality and substantiality are the
immediacy that proceeds from the sublated difference of the still unessential
concrete existence as appearance and its essentiality. Finally, objectivity is the
immediacy as which the concept has determined itself by the sublation of its
abstraction and mediation. – It is the privilege of philosophy to choose such
expressions from the language of ordinary life, which is made for the world
of imaginary representations, as seem to approximate the determinations
of the concept. There is no question of demonstrating for a word chosen
from ordinary life that in ordinary life too the same concept is associated
with that for which philosophy uses it, for ordinary life has no concepts,
only representations of the imagination, and to recognize the concept in
what is otherwise mere representation is philosophy itself. It must therefore
suffice if representation, for those of its expressions that philosophy uses
for its definitions, has only some rough approximation of their distinctive
difference; it may also be the case that in these expressions one recognizes
pictorial adumbrations which, as approximations, are close indeed to the
corresponding concepts. – One will be hard pressed, perhaps, to concede
that something can be without actually existing; but at least nobody will
mistake, for instance, being as the copula of the judgment for the expression
“to exist actually,” and nobody will say that “this article exists dear, suitable,
etc.,” “gold exists a metal or metallic,” instead of “this article is dear, suitable,
etc.,” “gold is a metal.” r And surely it is common to distinguish being from
appearing, appearance from actuality, as also being as contrasted to actuality,
and still more all these expressions from objectivity. – But even if such
expressions were used synonymously, philosophy would in any case have
the freedom to take advantage of such empty superfluity of language for
the purpose of its distinctions.

Mention was made in connection with the apodictic judgment – where
judgment attains completion and the subject thus loses its determinateness
as against the predicate – of the double meaning of subjectivity originating
from it, namely the subjectivity of the concept and equally so of the exter-
nality and contingency confronting the concept. 9 A similar objectivity also
appears for the double meaning, of standing opposed to the self-subsistent
concept yet of also existing in and for itself. 10 In the former sense, the object
stands opposed to the “I = I” which in subjective idealism is declared to be
the absolute truth. It is then the manifold world in its immediate existence
with which the “I” or the concept is engaged in endless struggle, in order,
by the negation of the inherently nullity of this other, to give to its first
certainty of being a self the actual truth of its equality with itself. – In a
broader sense, it means a subject matter in general for whatever interest or
activity of the subject.

In the opposite sense, however, the objective signifies
that which exists in and for itself,
without restriction and opposition.
Rational principles, perfect works of art, etc.,
are said to be objective to the extent
that they are free and above every accidentality.
Although rational principles, whether theoretical or ethical,
only belong to the sphere of the subjective, to consciousness,
this aspect of the latter of existing in and for itself
is nonetheless called objective;
the cognition of truth is made to rest on the cognition of
the object as free of any addition by subjective reflection, and right conduct
on the adherence to objective laws, such as are not of subjective origin and
are immune to arbitrariness and to treatment that would compromise their
necessity.

At the present standpoint of our treatise,
objectivity has the meaning first of all of
the being in and for itself of the concept
that has sublated the mediation posited
in its self-determination,
raising it to immediate self-reference.
This immediacy is therefore itself immediately
and entirely pervaded by the concept,
just as its totality is immediately identical with its being.
But further, since the concept equally has to restore the free
being-for-itself of its subjectivity,
it enters with respect to objectivity
into a relation of purpose
in which the immediacy of the objectivity
becomes a negative for it,
something to be determined through its activity.
This immediacy thus acquires the other significance,
namely that in and for itself, in so far as it
stands opposed to the concept, it is a nullity.

First, then, objectivity is in its immediacy.
Its moments, on account of the totality of all moments,
stand in self-subsistent indifference
as objects each outside the other,
and as so related they possess
the subjective unity of the concept
only as inner or as outer.
This is mechanism.

But, second, inasmuch as in mechanism that unity
reveals itself to be the immanent law of the objects,
their relation becomes one of non-indifference,
each specifically different according to law;
a connection in which the objects'
determinate self-subsistence is sublated.
This is chemism.

Third, this essential unity of the objects is
thereby posited as distinct from their self-subsistence.
It is the subjective concept,
but posited as referring in and for itself
to the objectivity, as purpose.
This is teleology.

Since purpose is the concept posited
as within it referring to objectivity,
and through itself sublating its defect
of being subjective,
the at first external purposiveness becomes,
through the realization of the purpose, internal.
It becomes idea.

CHAPTER 1 Mechanism

Since objectivity is the totality of the concept
that has returned into its unity,
an immediate is thereby posited which is in and for itself
that totality, and is also posited as such,
but in it the negativity of the concept has as yet
not detached itself from the immediacy of the totality;
in other words, the objectivity is not yet posited as judgment.
In so far as it has the concept immanent in it,
the difference of the concept is present in it;
but on account of the objective totality,
the differentiated moments are
complete and self-subsistent objects
that, consequently, even in connection
relate to one another as each standing on its own,
each maintaining itself in every combination as external.
This is what constitutes the character of mechanism,
namely,  that whatever the connection that
obtains between the things combined,
the connection remains one that is alien to them,
that does not affect their nature,
and even when a reflective semblance
of unity is associated with it,
the connection remains nothing more than
composition, mixture, aggregate, etc.
Spiritual mechanism, like its material counterpart,
also consists in the things connected in the spirit
remaining external to one another and to spirit.
A mechanical mode of representation,
a mechanical memory, a habit, a mechanical mode of acting,
mean that the pervasive presence that is proper to spirit
is lacking in what spirit grasps or does.
Although its theoretical or practical mechanism
cannot take place without its spontaneous activity,
without an impulse and consciousness,
the freedom of individuality is still lacking in it,
and since this freedom does not appear in it,
the mechanical act appears as a merely external one.

CHAPTER 2 Chemism

In objectivity as a whole
chemism constitutes the moment of judgment,
of the difference that has become objective,
and of process.

Since it already begins with
determinateness and positedness,
and the chemical object is
at the same time objective totality,
the course it follows next is
simple and perfectly determined
by its presupposition.

CHAPTER 3

Teleology

Where there is the perception of a purposiveness,
an intelligence is assumed as its author;
required for purpose is thus the concept's
own free concrete existence.
Teleology is above all contrasted with mechanism,
in which the determinateness posited in the object,
being external, is one that gives no sign of self-determination.
The opposition between causæ efficientes and causæ finales,
between merely efficient and final causes,
refers to this distinction, just as,
 at a more concrete level, the enquiry whether the absolute
essence of the world is to be conceived as blind mechanism or as an intelli-
gence that determines itself in accordance with purposes also comes down
to it.
The antinomy of fatalism, along with determinism,
and freedom is equally concerned with
the opposition of mechanism and teleology;
for the free is the concept in its concrete existence.

Earlier metaphysics has dealt with these concepts as it dealt with others.
It presupposed a certain picture of the world and strived to show that
one or the other concept of causality was adequate to it, and the opposite
defective because not explainable from the presupposed picture, all the while
not examining the concept of mechanical cause and that of purpose to see
which possesses truth in and for itself. If this is established independently, it
may turn out that the objective world exhibits mechanical and final causes;
its actual existence is not the norm of what is true, but what is true is rather
the criterion for deciding which of these concrete existences is its true one.
Just as the subjective understanding exhibits also errors in it, so the objective
world exhibits also aspects and stages of truth that by themselves are still
one-sided, incomplete, and only relations of appearances. If mechanism
and purposiveness stand opposed to each other, then by that very fact they
cannot be taken as indifferent concepts, as if each were by itself a correct
concept and had as much validity as the other, the only question being
where the one or the other may apply. This equal validity of the two rests
only on the fact that they are, that is to say, that we have them both. But
since they do stand opposed, the necessary first question is, which of the
two concepts is the true one; and the higher and truly telling question is,
whether there is a third which is their truth, or whether one of them is the
truth of the other. – But purposive connection has proved to be the truth of
mechanism. – Regarding chemism, what came under it can be taken together
with mechanism, for purpose is the concept in free concrete existence, and
the concept's state of unfreedom, its being sunk into externality, stands
opposed to it in any form. Both, mechanism as well as chemism, are
therefore included under natural necessity: mechanism, because in it the
concept does not exist in the object concretely, for as mechanical the latter
lacks self-determination; chemism, either because the concept has in it a
one-sided concrete existence in a state of tension, or because, emerging
as the unity that creates in the neutral object a tension of extremes, it is
external to itself in so far as it sublates this divide.

The closer the teleological principle is associated with the concept of an
extra-mundane intelligence, and the more it has therefore enjoyed the favor
of piety, all the more it has seemed to depart from the true investigation
of nature, which aims at a cognition of the properties of nature not as
extraneous, but as immanent determinacies, and accepts only such cognition
as a valid conceptual comprehension. Since purpose is the concept itself
in its concrete existence, it may seem strange that a cognition of objects
based on their concept rather appears as an unjustified trespass into a
heterogeneous element, whereas mechanism, for which the determinateness
of an object is posited in it externally and by an other, is accepted as a more
immanent view of things than teleology. Of course mechanism, at least the
ordinary unfree mechanism, and chemism as well, must be regarded as an
immanent principle in so far as the externally determining object is itself
again just another such object, externally determined and indifferent to its
being determined, or, in the case of chemism, in so far as the other object
must likewise be one that is chemically determined; in general, in so far
as an essential moment of the totality always lies in something external.
These principles remain confined, therefore, within the same natural form
of finitude; but although they do not wish to transcend the finite and,
as regards appearances, lead only to finite causes that themselves demand
further causes, they nonetheless equally expand themselves, partly into a
formal totality in the concept of force, cause, or of such determinations of
reflection that are supposed to signify originariness, and partly, through the
medium of abstract universality, also into a sum total of forces, a whole of
reciprocal causes. Mechanism thus reveals itself to be a striving for totality
by the very fact that it seeks to comprehend nature by itself as a whole that
has no need of an other for its concept – a totality that is not found in
purpose and the extra-mundane intelligence associated with it.

Now purposiveness presents itself from the first as something of a gen-
erally higher nature, as an intelligence that externally determines the man-
ifoldness of objects through a unity that exists in and for itself, so that the
indifferent determinacies of the objects become essential by virtue of this
connection. In mechanism they become so through the mere form of neces-
sity that leaves their content indifferent, for they are supposed to remain
external and only the understanding as such is expected to find satisfaction
by recognizing its principle of union, the abstract identity. In teleology,
on the contrary, the content becomes important, for teleology presupposes
a concept, something determined in and for itself and consequently self-
determining, and has therefore extracted from the connection of differences
and their reciprocal determinateness, from the form, a unity that is reflected
into itself, something that is determined in and for itself and is consequently
a content. But if this content is otherwise finite and insignificant, then it
contradicts what it is supposed to be, for according to its form purpose is
a totality infinite within itself – especially when the activity operating in
accordance with it is assumed to be an absolute will and intelligence. For
this reason has teleology drawn the reproach of triviality so much upon
itself, for the purposes that it has espoused are, as the case may be, more
important or more trivial [than the content], and it was inevitable that
the connection of purposiveness in objects would so often appear just a
frivolity, since it appears external and therefore contingent. Mechanism,
on the contrary, leaves to the determinacies of the objects, as regards their
content, their status as accidents indifferent to the object, and these deter-
minacies are not supposed to have, whether for the objects or the subjective
understanding, any value higher than that. This principle, combined with
external necessity, yields therefore a consciousness of infinite freedom that
contrasts with teleology, which sets up as something absolute bits of its
content that are trivial and even contemptible, where the more universal
thought can only find itself infinitely constricted, even to the point of
feeling disgust.

The formal disadvantage from which this teleology immediately suffers
is that it only goes as far as external purposiveness. The content of concept,
since the latter is thereby posited as something formal, is for teleology also
externally given to it in the manifoldness of the objective world – in those
very determinacies that are also the content of mechanism, but are there as
something external and accidental. Because of this commonality of content,
only the form of purposiveness constitutes by itself the essential element of
the teleological. In this respect, without as yet considering the distinction
between external and internal purposiveness, the connection of purpose in
general has proven itself to be the truth of mechanism. – Teleology possesses
in general the higher principle, the concept in its concrete existence, which
is in and for itself the infinite and absolute – a principle of freedom which,
utterly certain of its self-determination, is absolutely withdrawn from the
external determining of mechanism.

One of Kant's greatest services to philosophy was
in drawing the distinction between relative or
external purposiveness and internal purposiveness;
in the latter he opened up the concept of life, the idea,
and with that he positively raised philosophy
above the determinations of reflection
and the relative world of metaphysics,
something that the Critique of Reason does
only imperfectly, ambiguously, and only negatively.
We have remarked that the opposition of teleology and mechanism is
first of all the general opposition of freedom and necessity.
Kant treated the opposition in this form, among the antinomies of reason,
namely, as the third conflict of the transcendental ideas.
I cite his exposition, to which reference was made earlier,
very briefly because its essential point is so simple
that it does not need extensive explanation,
and moreover, the peculiarities of Kant's antinomies
have been elucidated in greater detail elsewhere.

The thesis of the antinomy now in question runs thus:
Causality according to the laws of nature is
not the only one from which the appearance of
the world can exhaustively be derived.
For their explanation, it is necessary to
assume yet another causality through freedom.

The antithesis: There is no freedom,
but everything in the world happens
solely according to laws of nature.

As in the other antinomies, the proof starts off apagogically
by assuming the opposite of each thesis;
secondly, in order to show the contradiction of this assumption,
its opposite (which is then the proposition to be proved)
is assumed in turn and presupposed as valid.
This whole roundabout proof could therefore be spared,
for the proof consists in nothing but the assertoric
assertion of the two opposite propositions.

Thus to prove the thesis we should first assume that there is no other
causality than that according to the laws of nature, that is, according to
the necessity of mechanism in general, chemism being included. This
proposition contradicts itself, because the law of nature consists just in
this, that nothing happens without a cause sufficiently determined a priori, a
cause that would have to contain an absolute spontaneity within it, that is,
the assumption opposed to the thesis is contradictory, for the reason that
it contradicts the thesis.

In support of the proof of the antithesis we should assume that there is a
freedom as a particular kind of causality for absolutely initiating a situation,
and together with it also a series of consequences following upon it. But
now, since such a beginning presupposes a situation that has no causal link
with the one preceding it, it contradicts the law of causality that alone
makes the unity of experience and experience in general possible – that is,
the assumption of the freedom that is opposed to the antithesis cannot be
made, for the reason that it contradicts the antithesis.

We find in essence the same antinomy in the Critique of the Teleological
Judgement as the opposition between the proposition that every generation
of material things happens according to merely mechanical laws, and the
proposition that some cases of generation of material things are not possible
according to such laws.
Kant's resolution of this antinomy is the same
as the general resolution of the rest, namely that reason cannot prove
either the one or the other proposition because we cannot have a priori
any determining principle of the possibility of things according to merely
empirical laws of nature; further, that therefore the two propositions must
be regarded not as objective propositions but as subjective maxims; that
I ought to reflect on the events of nature every time according to the
principle of the mechanism of nature alone, but that this does not prevent,
when occasion permits, following up certain natural forms in accordance
with another maxim, namely in accordance with the principal of final
causes – as if now these two maxims, which moreover are supposed to be
necessary only for human reason, did not stand in the same opposition as
the two propositions in antinomy. – Missing in all this, as we remarked
above, 21 is the one thing that alone is of philosophical interest, namely
the investigation of which of the two principles has truth in and for
itself. On this standpoint, it makes no difference whether the principles
should be regarded as objective, which means here, as externally existing
determinations of nature, or as mere maxims of a subjective cognition; what
is subjective here is rather the contingent cognition that applies one or the
other maxim as occasion demands, indeed, according to whether it deems
them fitting for given objects, but for the rest does not ask about the truth
of these determinations themselves, whether they both are determinations
of the objects or of cognition.

However unsatisfactory is for this reason Kant's discussion
of the teleological principle with respect to its essential viewpoint,
still worthy of note is the place that Kant assigns to it.
Since he ascribes it to a reflective faculty of judgment,
he makes it into a mediating link between
the universal of reason and the singular of intuition;
further, he distinguishes this reflective judgment
from the determining judgment, the latter one that
merely subsumes the particular under the universal.
Such a universal that only subsumes is an abstraction
that becomes concrete only in an other, in the particular.
Purpose, on the contrary, is the concrete universal containing within itself
the moment of particularity and of externality;
it is therefore active and
the impulse to repel itself from itself.
The concept, as purpose, is of course
an objective judgment in which one determination,
the subject, namely the concrete concept, is self-determined,
while the other is not only a predicate but external objectivity.
But for that reason the connection of purpose is not
a reflective judgment that considers external objects
only according to a unity,
as though an intelligence had given them to us
for the convenience of our faculty of cognition;
on the contrary, it is the truth that exists in
and for itself and judges objectively,
determining the external objectivity absolutely.
The connection of purpose is therefore more than judgment;
it is the syllogism of the self-subsistent free concept
that through objectivity unites itself with itself in conclusion.

Purpose has resulted as the third to mechanism and chemism; it is their
truth.
Inasmuch as it still stands inside the sphere of objectivity or of the
immediacy of the total concept,
 it is still affected by externality as such and
has an objective world over against it to which it refers.
From this side,
mechanical causality,
to which chemism is also in general to be added,
still makes its appearance in this purposive connection which is the external
one,
but as subordinated to it and as sublated in and for itself. As regards
the more precise relation,
 the mechanical object is, as immediate totality,
indifferent to its being determined and consequently, conversely,
to its
being a determinant.
This external determinateness has now progressed to
self-determination and accordingly the concept that in the object was only
inner or, which amounts to the same, only outer, is now posited;
 purpose
is, in the first instance, precisely this concept which is external to the
mechanical object.
And so for chemism also, purpose is the self-determining
which brings the external determinateness conditioning it back to the unity
of the concept. –
We have here the nature of the subordination of the two
preceding forms of the objective process.
 The other, which in those forms
lies in the infinite progress, is the concept posited at first as external to them,
and this is purpose; not only is the concept their substance but externality
is for them also an essential moment constituting their determinateness.
Thus mechanical or chemical technique,
because of its character of being
externally determined,
naturally offers itself to the connection of purpose,
which we must now examine more closely.
SECTION III

The idea

The idea is the adequate concept,
the objectively true,
or the true as such.
If anything has truth,
it has it by virtue of its idea,
or something has truth only
in so far as it is idea.
The expression “idea” has otherwise also often been used
in philosophy as well as in ordinary life for “concept,”
or even for just a “representation.”
To say that I still have no idea of
this lawsuit, this building, this region,
means nothing more than I still have no representation of it.
It is Kant who reclaimed the expression “idea”
for the “concept of reason.”
Now according to Kant the concept of reason
should be the concept of the unconditional,
but a concept which is transcendent with respect to appearances,
that is, one for which no adequate empirical use can be made.
The concepts of reason are supposed to serve
for the comprehension of perceptions,
those of the understanding for the understanding of them.
In fact, however, if these last
concepts of the understanding are truly concepts,
then they are comprehensions, which means concepts;
they will make comprehending possible,
and an understanding of perceptions
through concepts of the understanding
will be a comprehending.
But if understanding is only
the determining of perceptions by categories
such as whole and parts, force, cause, and the like,
then it signifies only a determining by means of reflection,
just as by understanding one may mean only the determinate
representation of a fully determined sensuous content;
as when someone is being shown the way, that at the end of the
wood he must turn left, and he replies “I understand,”
understanding means nothing more than a grasp
in pictorial representation and in memory.
“Concept of reason,” too, is a somewhat clumsy expression;
for the concept is in general something rational,
and in so far as reason is distinguished
from the understanding and the concept as such,
it is the totality of the concept and objectivity.
The idea is the rational in this sense;
it is the unconditioned,
because only that has conditions
which essentially refers to an objectivity
that it does not determine itself
but which still stands over against it
in the form of indifference and externality,
just as the external purpose had conditions.

If we now reserve the expression “idea”
for the objective or real concept
and we distinguish it from the concept itself
and still more from mere representation,
then we must also even more definitely reject
that estimate of it according to which
the idea is something with no actuality,
and true thoughts are accordingly said to be only ideas.
If thoughts are something merely subjective and contingent,
then they certainly have no further value;
but in this they do not stand lower
than the temporal and contingent actualities
which likewise have no further value than
that of accidentalities and appearances.
But if, on the contrary, the idea is
supposed not to have the value of truth
because in regard to appearances it is transcendent,
because no congruent object can be given for it
in the world of the senses,
then this is indeed an odd misunderstanding,
for objective validity is being denied to it
on the ground that it lacks precisely what makes
of appearances the untrue being of the objective world.
In regard to the practical ideas, Kant recognizes that
“nothing can be more harmful and unworthy of
a philosopher than the vulgar appeal to experience,
which supposedly contradicts the idea.
Any such alleged contradiction would not be there at all
if, for example, political institutions were set up
at the right time in accordance with ideas,
and if crude concepts, crude just because they are drawn from experience,
had not usurped the place of ideas thus thwarting all good intentions.”
Kant regards the idea as something necessary,
the goal which, as the archetype,
we must strive to set up as a maximum
and to which we must bring actuality
as it presently stands ever closer.

But since the result now is that the idea is
the unity of the concept and objectivity, the true,
we must not regard it as just a goal which is to be approximated
but itself remains always a kind of beyond;
we must rather regard everything as being actual
only to the extent that it has the idea
in it and expresses it.
It is not just that the subject matter,
the objective and the subjective world,
ought to be in principle congruent with the idea;
the two are themselves rather the congruence of concept and reality;
a reality that does not correspond to the concept is mere appearance,
something subjective, accidental, arbitrary, something which is not the truth.
When it is said that there is no subject matter
to be found in experience which is perfectly congruent with the idea,
the latter is opposed to the actual as a subjective standard;
but there is no saying what anything actual
might possibly be in truth, if its concept is not in it
and its objectivity does not measure up to this concept;
it would then be a nothing.
Indeed, the mechanical and the chemical object,
like a subject devoid of spirit
and a spirit conscious only of finitude and not of its essence,
do not, according to their various natures,
have their concept concretely existing in them in
its own free form.
But they can be something at all true only in so far as
they are the union of their concept and reality, of their soul and their body.
Wholes like the state and the church cease to exist in concreto when the
unity of their concept and their reality is dissolved; the human being, the
living thing, is dead when soul and body are parted in it; dead nature, the
mechanical and the chemical world
that is, when “the dead” is taken
to mean the inorganic world,
for the expression would otherwise have no positive meaning at all
this dead nature, then, if it is separated into its
concept and its reality,
is nothing but the subjective abstraction of
a thought form and a formless matter.
Spirit that were not idea,
not the unity of the concept with itself,
not the concept that has the concept itself as its reality,
would be dead spirit, spiritless spirit, a material object.

Since the idea is the unity of the concept and reality,
being has attained the significance of truth;
it now is, therefore, only what the idea is.
Finite things are finite because,
and to the extent that,
they do not possess the reality of
their concept completely within them
but are in need of other things for it
or, conversely, because they are presupposed as objects
and consequently the concept is in them as an external determination.
The highest to which they attain on
the side of this finitude is external purposiveness.
That actual things are not congruent with the idea
constitutes the side of their finitude, of their untruth,
and it is according to this side that they are objects,
each in accordance with its specific sphere,
and, in the relations of objectivity, determined
as mechanical, chemical, or by an external purpose.
That the idea has not perfectly fashioned their reality,
that it has not completely subjugated it to the concept,
the possibility of that rests on the fact that
the idea itself has a restricted content;
that, as essentially as it is
the unity of the concept and reality,
just as essentially it is also their difference;
for only the object is the immediate unity, that is,
the unity that only exists in itself.
But if a subject matter, say the state,
did not at all conform to its idea,
that is to say, if it were not rather the idea of the state;
if its reality, which is the self-conscious individuals,
did not correspond at all to the concept,
its soul and body would have come apart;
the soul would have taken refuge in the secluded regions of thought,
the body been dispersed into singular individualities.
But because the concept of the state is essential
to the nature of these individualities,
it is present in them as so mighty an impulse
that they are driven to translate it into reality,
be it only in the form of external purposiveness,
or to put up with it as it is,
or else they must needs perish.
The worst state, one whose reality least corresponds to the concept,
in so far as it still has concrete existence, is yet idea;
the individuals still obey the power of a concept.

But the idea has not only
the general meaning of true being,
of the unity of concept and reality,
but also the more particular one
of the unity of subjective concept and objectivity.
For the concept is as such itself
already the identity of itself and reality;
for the indeterminate expression “reality”
means nothing but determinate being,
and this the concept possesses
in its particularity and singularity.
Objectivity, moreover, is likewise the total concept
that has withdrawn into identity with itself out of its determinateness.
In the subjectivity of the concept,
the determinateness
or the difference of the latter is
a reflective shine which is immediately sublated,
withdrawn into being-for-itself
or into negative unity,
an inhering predicate.
But in this objectivity the determinateness is
posited as immediate totality, as external whole.
Now the idea has shown itself to be the concept
liberated again into its subjectivity
from the immediacy into which it has sunk in the object;
it is the concept that distinguishes itself from its objectivity
but an objectivity which is no less determined by it
and possesses its substantiality only in that concept.
This identity has therefore rightly been
designated as a subject-object,
for it is just as well the formal
or subjective concept
as it is the object as such.
But this is a point that needs further precision.
The concept, inasmuch as it has truly attained its reality,
is this absolute judgment whose subject distinguishes itself
as self-referring negative unity from its objectivity
and is the latter's being-in-and-for-itself;
but it refers to it essentially through itself
and is, therefore, self-directed purpose and impulse.
For this very reason, however,
the subject does not possess objectivity immediately in it
(it would then be only the totality of the object as such,
a totality lost in the objectivity)
but is the realization of the purpose,
an objectivity posited by virtue of
the activity of the purpose,
one which, as positedness, has its subsistence and its form only as
permeated by its subject.
As objectivity, it has the moment of the externality
of the concept in it and is in general,
therefore, the side of finitude,
of alteration and appearance;
but this side retreats into the negative unity of
the concept and there it perishes;
the negativity whereby its indifferent
externality of being manifests itself as unessential
and as a positedness is the concept itself.
Despite this objectivity, the idea is therefore
absolutely simple and immaterial,
for the externality has being only
as determined by the concept
and as taken up into its negativity;
in so far as it exists as indifferent externality,
it is not only abandoned to mechanism in general
but exists only as the transitory and untrue.
Thus although the idea
has its reality in a materiality,
the latter is not an abstract being standing
over against the concept
but, on the contrary, it exists only as becoming,
as simple determinateness of the concept
by virtue of the negativity of the indifferent being.

This yields the following closer determinations of the idea.

First, the idea is the simple truth,
the identity of concept and objectivity as a
universal in which the opposition,
the presence of the particular,
is dissolved in its self-identical negativity
and is equality with itself.

Second, it is the connection of
the subjectivity of the simple concept, existing for itself,
and of the concept's objectivity which is distinguished from it;
the former is essentially the impulse to sublate this separation,
and the latter is indifferent positedness,
subsistence which in and for itself is null.
As this connection, the idea is
the process of disrupting itself into individuality
and into the latter's inorganic nature,
and of then bringing this inorganic nature again
under the controlling power of the subject
and back to the first simple universality.
The identity of the idea with itself is one with the process;
the thought that liberates actuality from
the seeming of purposeless mutability
and transfigures it into idea
must not represent this truth of actuality
as dead repose, as a mere picture, numb, without impulse and movement,
as a genus or number, or as an abstract thought;
the idea, because of the freedom which the concept has attained in it,
also has the most stubborn opposition within it;
its repose consists in the assurance and the certainty
with which it eternally generates that opposition
and eternally overcomes it, and in it rejoins itself.

But the idea is at first again only immediate or only in its concept;
the objective reality is indeed conformable to the concept
but has not yet been liberated into the concept,
and it does not concretely exist explicitly as the concept.
Thus the concept is indeed the soul,
but the soul is in the guise of an immediate,
that is, it is not determined as soul itself,
has not comprehended itself as soul,
does not have its objective reality within itself;
the concept is as a soul that is not yet fully animated.

Thus the idea is, first of all, life.
It is the concept which, distinct from its objectivity,
simple in itself, permeates that objectivity
and, as self-directed purpose, has its means within it
and posits it as its means, yet is immanent in this means
and is therein the realized purpose identical with itself.
The idea, on account of its immediacy, has singularity
for the form of its concrete existence.
But the reflection within it of its absolute process is
the sublating of this immediate singularity;
thereby the concept, which as universality is
in this singularity the inner,
transforms externality into universality,
or posits its objectivity as a self-equality.

Thus is the idea, in second place,
the idea of the true and the good,
as cognition and will.
It is at first finite cognition and finite will,
where the true and the good are still distinguished
and the two are at first only as a goal.
The concept has first liberated itself into itself,
giving itself only a still abstract objectivity for its reality.
But the process of this finite cognition and this finite action
transforms the initially abstract universality into totality,
whereby it becomes complete objectivity.
Or considered from the other side,
finite, that is, subjective spirit,
makes for itself the
presupposition of an objective world,
such a presupposition as life only has;
but its activity is the sublating of this presupposition
and the turning of it into something posited.
Thus its reality is for it the objective world,
or conversely the objective world is
the ideality in which it knows itself.

Third, spirit recognizes the idea as its absolute truth,
as the truth that is in and for itself:
the infinite idea in which cognizing and doing are equalized,
and which is the absolute knowledge of itself.

CHAPTER 1

Life

The idea of life has to do with a subject matter so concrete, and if you will so
real, that in dealing with it one may seem according to the common notion
of logic to have overstepped its boundaries. Needless to say, if the logic were
to contain nothing but empty, dead forms of thought, then there could be
no talk in it at all of such a content as the idea, or life, are. But if the subject
matter of logic is the absolute truth, and truth as such lies essentially in
cognition, then cognition at least would have to come in for consideration.
It is common practice to have the so-called pure logic be followed by an
applied logic
 a logic that has to do with concrete cognition, quite apart from
all the psychology and anthropology that is commonly deemed necessary to
interpolate into logic. But the anthropological and psychological side of
cognition is concerned with the form in which cognition appears when the
concept does not as yet have an objectivity equal to it, that is, when it
does not have itself as object. The part of the logic that deals with this
concrete cognition does not belong to applied logic as such; if it did, then
every science would have to be dragged into logic, for each is an applied
logic in so far as it consists in apprehending its subject matter in forms of
thought and of concepts.
 The subjective concept has presuppositions that
are exhibited in psychological, anthropological, and other forms. But the
presuppositions of the pure concept belong in logic only to the extent that
they have the form of pure thoughts, of abstract essentialities, such as are the
determinations of being and essence. The same goes for cognition, which is
the concept's comprehension of itself: no other shape of its presupposition
but the one which is itself idea is to be dealt with in the logic; this, however,
is a presupposition which is necessarily treated in logic. This presupposition
is now the immediate idea; for while cognition is the concept, in so far as
the latter exists for itself but as a subjectivity referring to an objectivity,
then the concept refers to the idea as presupposed or as immediate. But the
immediate idea is life.

To this extent the necessity of considering the idea of life in logic would
be based on the necessity, itself recognized in other ways, of treating the
concrete concept. But this idea has arisen through the concept's own
necessity; the idea, that which is true in and for itself, is essentially the
subject matter of the logic; since it is first to be considered in its immediacy,
so that this treatment be not an empty affair devoid of determination, it is
to be apprehended and cognized in this determinateness in which it is life.
A comment may be in order here to differentiate the logical view of life
from any other scientific view of it, though this is not the place to concern
ourselves with how life is treated in non-philosophical sciences but only
with how to differentiate logical life as idea from natural life as treated in
the philosophy of nature, and from life in so far as it is bound to spirit.
 As
treated in the philosophy of nature, as the life of nature and to that extent
exposed to the externality of existence, life is conditioned by inorganic nature
and its moments as idea are a manifold of actual shapes. Life in the idea is
without such presuppositions, which are in shapes of actuality; its presuppo-
sition is the concept as we have considered it, on the one hand as subjective,
and on the other hand as objective. In nature life appears as the highest stage
that nature's externality can attain by withdrawing into itself and sublating
itself in subjectivity. It is in logic the simple in-itselfness which in the idea
of life has attained the externality truly corresponding to it; the concept
that came on the scene earlier as a subjective concept is the soul of life itself;
it is the impulse that gives itself reality through a process of objectification.
Nature, as it reaches this idea starting from its externality, transcends itself;
its end is not its beginning but is for it as a limit in which it sublates itself.
Similarly, in the idea of life the moments of life's reality do not receive the
shape of external actuality but remain enveloped in conceptual form.

In spirit, however, life appears both as opposed to it and as posited as at
one with it, in a unity reborn as the pure product of spirit. For life is here
to be taken generally in its proper sense as natural life, for what is called
the life of spirit as spirit, is spirit's own peculiar nature that stands opposed
to mere life; just as we speak of the nature of spirit, even though spirit is
nothing natural but stands rather in opposition to nature. Thus life as such
is for spirit in one respect a means, and then spirit holds it over against
itself; in another respect, spirit is an individual, and then life is its body; in
yet another respect, this unity of spirit and its living corporeality is born
of spirit into ideality. None of these connections of life to spirit concerns
logical life, and life is to be considered here neither as the instrument of
a spirit, nor as a living body, nor again as a moment of the ideal and of
beauty.
 In both cases, as natural life and as referring to spirit, life obtains a
determinateness from its externality, in one case through its presuppositions,
such as are other formations of nature, and in the other case through the
purposes and the activity of spirit. The idea of life by itself is free from both
the conditioning objectivity presupposed in the first case and the reference
to subjectivity of the second case.

Life, considered now more closely in its idea, is in and for itself absolute
universality; the objectivity which it possesses is throughout permeated by
the concept, and this concept alone it has as substance. Whatever is dis-
tinguished as part, or by some otherwise external reflection, has the whole
concept within it; the concept is the soul omnipresent in it, a soul which is
simple self-reference and remains one in the manifoldness that accrues to
the objective being. This manifoldness, as self-external objectivity, has an
indifferent subsistence which in space and time, if these could already be
mentioned here, is a mutual externality of entirely diverse and atomistic
matters. But externality is in life at the same time as the simple determi-
nateness of its concept; thus the soul flows omnipresently in this manifold
but remains at the same time the simple oneness of the concrete concept
with itself.
 That way of thinking that clings to the determinations of
reflective relations and of the formal concept, when it comes to consider
life, the unity of its concept in the externality of objectivity, the absolute
multiplicity of atomistic matter, finds that all its thoughts are absolutely of
no avail; the omnipresence of the simple in the manifold externality is for
reflection an absolute contradiction and also, since it cannot at the same
time avoid witnessing this omnipresence in the perception of life and must
therefore grant the actuality of this idea, an incomprehensible mystery
 for
reflection does not grasp the concept, nor does it grasp it as the substance
of life.
 But this simple life is not only omnipresent; it is the one and
only subsistence and immanent substance of its objectivity; but as subjective
substance it is impulse, more precisely the specific impulse of particular dif-
ference, and no less essentially the one and universal impulse of the specific
that leads its particularization back to unity and holds it there. Only as this
negative unity of its objectivity and particularization is life self-referring, life
that exists for itself, a soul. As such, it is essentially a singular that refers to
objectivity as to an other, an inanimate nature. The originative judgment of
life consists therefore in this, that it separates itself off as individual subject
from the objective and, since it constitutes itself as the negative unity of
the concept, makes the presupposition of an immediate objectivity.

First, life is therefore to be considered as a living individual that is for itself
the subjective totality and is presupposed as indifferent to an objectivity
that stands indifferent over against it.

Second, it is the life-process of sublating its presupposition, of positing
as negative the objectivity indifferent to it, and of actualizing itself as the
power and negative unity of this objectivity. By so doing, it makes itself
into the universal which is the unity of itself and its other.

Third, consequently life is the genus-process, the process of sublating
its singularization and relating itself to its objective existence as to itself.
Accordingly, this process is on the one hand the turning back to its concept
and the repetition of the first forcible separation, the coming to be of a
new individuality and the death of the immediate first; but, on the other
hand, the withdrawing into itself of the concept of life is the becoming of
the concept that relates itself to itself, of the concept that exists for itself,
universal and free, the transition into cognition.

CHAPTER 2

The idea of cognition

Life is the immediate idea, or the idea as its still internally unrealized
concept. In its judgment, the idea is cognition in general.

The concept is for itself as concept inasmuch as it freely and concretely
exists as abstract universality or a genus. As such, it is its pure self-identity
that internally differentiates itself in such a way that the differentiated is
not an objectivity but is rather equally liberated into subjectivity or into the
form of simple self-equality; consequently, the object facing the concept is
the concept itself. Its reality in general is the form of its existence; all depends
on the determination of this form; on it rests the difference between what
the concept is in itself, or as subjective, and what it is when immersed in
objectivity, and then in the idea of life. In this last, the concept is indeed
distinguished from its external reality and posited for itself; however, this
being-for-itself which it now has, it has only as an identity that refers to itself
as immersed in the objectivity subjugated to it, or to itself as indwelling,
substantial form. The elevation of the concept above life consists in this,
that its reality is the concept-form liberated into universality. Through this
judgment the idea is doubled, into the subjective concept whose reality is
the concept itself, and the objective concept which is as life.
 Thought,
spirit, self-consciousness, are determinations of the idea inasmuch as the latter
has itself as the subject matter, and its existence, that is, the determinateness
of its being, is its own difference from itself.

The metaphysics of the spirit or, as was more commonly said in the past,
of the soul, revolved around the determinations of substance, simplicity,
immateriality. These were determinations for which spirit was supposed to
be the ground, but as a subject drawn from empirical consciousness, and
the question then was which predicates agreed with the perceived facts.
But this was a procedure that could go no further than the procedure
of physics, which reduces the world of appearance to general laws and
determinations of reflection, for it is spirit still as phenomenal that is taken
as the foundation. In fact, in so far as scientific stringency goes, it also
had to fall short of physics. For not only is spirit infinitely richer than
nature; since its essence is constituted by the absolute unity in the concept
of opposites, and in its appearance, therefore, and in its connection with
externality, it exhibits contradiction at its most extreme form, it must be
possible to adduce an experience in support of each of the opposite deter-
minations of reflection, or, starting from experiences, to proceed by way
of formal inference to the opposite determinations. Since the predicates
immediately drawn from the appearances still belong to empirical psychol-
ogy, so far as metaphysical consideration goes, all that is in truth left are
the entirely inadequate determinations of reflection.
 In his critique of
rational psychology, Kant insists that, since this metaphysics is supposed
to be a rational science, the least addition of anything drawn from per-
ception to the universal representation of self-consciousness would alter it
into an empirical science, thus compromising its rational purity and its
independence from all experience.
 Accordingly, all that is left on this
view is the simple representation “I,” a representation entirely devoid of
content, of which one cannot even say that it is a concept, but must say that
it is a mere consciousness, one that accompanies every concept. Now, as Kant
argues further, this “I,” or, if you prefer, this “it” (the thing) that thinks,
takes us no further than the representation of a transcendental subject of
thoughts = x, a subject which is known only through the thoughts that
are its predicates, and of which, taken in isolation, we cannot ever have
the least concept. This “I” has the associated inconvenience that, as Kant
expresses it, in order to judge anything about it, we must every time already
make use of it, for it is not so much one representation by which a particular
object is distinguished, as it is rather a form of representation in general, in
so far as representation can be said to be cognition.
 Now the paralogism
that rational psychology incurs, as Kant expresses it, consists in this: that
modes of self-consciousness in thinking are converted into concepts of the
understanding, as if they were the concepts of an object; that that “I think”
is taken to be a thinking being, a thing-in-itself; that in this way, because I
am present in consciousness always as a subject, am indeed as a singular sub-
ject, identical in all the manifoldness of representation, and distinguishing
myself from this manifoldness as external to it, the illegitimate inference
is thereby drawn that I am a substance, and a qualitatively simple being on
top of that, and a one, and a being that concretely exists independently of the
things of space and time.

I have cited this position in some detail because one can clearly recognize
in it both the nature of the former metaphysics of the soul and also, more to
the point, of the Critique that put an end to it.
 The former was intent on
determining the abstract essence of the soul; it went about this starting from
observation, and then converting the latter's empirical generalizations, and
the determination of purely external reflection attaching to the singularity
of the actual, into the form of the determinations of essence just cited.
What Kant generally has in mind here is the state of the metaphysics of
his time which, as a rule, stayed at these one-sided determinations with no
hint of dialectic; he neither paid attention to, nor examined, the genuinely
speculative ideas of older philosophers on the concept of spirit. In his
critique of those determinations he then simply abided by the Humean
style of skepticism; that is to say, he fixes on how the “I” appears in
self-consciousness, but from this “I,” since it is its essence (the thing in
itself ) that we want to cognize, he removes everything empirical; nothing
then remains but this appearance of the “I think” that accompanies all
representations and of which we do not have the slightest concept.
 It must
of course be conceded that, as long as we are not engaged in comprehending
but confine ourselves to a simple, fixed representation or to a name, we
do not have the slightest concept of the “I,” or of anything whatever, not
even of the concept itself.
 Peculiar indeed is the thought (if one can call
it a thought at all) that I must make use of the “I” in order to judge the
“I.” The “I” that makes use of self-consciousness as a means in order to
judge: this is indeed an x of which, and also of the relation involved in this
“making use,” we cannot possibly have the slightest concept. But surely it is
laughable to label the nature of this self-consciousness, namely that the “I”
thinks itself, that the “I” cannot be thought without the “I” thinking it, an
awkwardness and, as if it were a fallacy, a circle. The awkwardness, the circle,
is in fact the relation by which the eternal nature of self-consciousness and
of the concept is revealed in immediate, empirical self-consciousness
 is
revealed because self-consciousness is precisely the existent and therefore
empirically perceivable pure concept; because it is the absolute self-reference
that, as parting judgment, makes itself into an intended object and consists
in simply making itself thereby into a circle.
 This is an awkwardness that
a stone does not have. When it is a matter of thinking or judging, the stone
does not stand in its own way; it is dispensed from the burden of making
use of itself for the task; something else outside it must shoulder that effort.

The defect, which these surely barbarous notions place in the fact that
in thinking the “I” the latter cannot be left out as a subject, then also
appears the other way around, in that the “I” occurs only as the subject of
consciousness, or in that I can use myself only as a subject, and no intuition
is available by which the “I” would be given as an object; but the concept
of a thing capable of existence only as a subject does not as yet carry any
objective reality with it.
 Now if external intuition as determined in time
and space is required for objectivity, and it is this objectivity that is missed,
it is then clear that by objectivity is meant only sensuous reality. But to
have risen above such a reality is precisely the condition of thinking and of
truth. Of course, if the “I” is not grasped conceptually but is taken as a mere
representation, in the way we talk about it in everyday consciousness, then
it is an abstract self-determination, and not the self-reference that has itself
for its subject matter. Then it is only one of the extremes, a one-sided subject
without its objectivity; or else just an object without subjectivity, which it
would be were it not for the awkwardness just touched upon, namely that
the thinking subject will not be left out of the “I” as object. But as a matter
of fact this awkwardness is already found in the other determination, that of
the “I” as subject; the “I” does think something, whether itself or something
else. This inseparability of the two forms in which the “I” opposes itself to
itself belongs to the most intimate nature of its concept and of the concept
as such; it is precisely what Kant wants to keep away in order to retain
what is only a representation that does not internally differentiate itself and
consequently, of course, is void of concept. Now this kind of conceptual
void may well oppose itself to the abstract determinations of reflection or
to the categories of the previous metaphysics, for in one-sidedness it stands
at the same level with them, though these are in fact on a higher level of
thought; but it appears all the more lame and empty when compared with
the profounder ideas of ancient philosophy concerning the concept of the
soul or of thinking, as for instance the truly speculative ideas of Aristotle. 22
If the Kantian philosophy subjected the categories of reflection to critical
investigation, all the more should it have investigated the abstraction of the
empty “I” that he retained, the supposed idea of the thing-in-itself. The
experience of the awkwardness complained of is itself the empirical fact in
which the untruth of that abstraction finds expression.

The Kantian critique of rational psychology only refers to Mendelssohn's
proof of the persistence of the soul, and I now also cite its refutation of that
proof because of the oddness of what it adduces against it. Mendelssohn's
proof is based on the simplicity of the soul, by virtue of which it is supposed
to be incapable of alteration in time, of transition into an other. Qualitative
simplicity is in general the form of abstraction earlier considered; as quali-
tative determinateness, it was investigated in the sphere of being and it was
then proved that the qualitative, which is as such abstractly self-referring
determinateness, is precisely for that reason dialectical, mere transition into
an other. 24 In the case of the concept, however, it was shown that, when
considered in connection with persistence, indestructibility, imperishable-
ness, it is that which exists for itself, which is eternal, just because it is not
abstract but concrete simplicity
 because it is not a determinateness that
refers to itself abstractly but is the unity of itself and its other, and it cannot
therefore pass over into this other as if it thereby altered in it; it cannot
precisely because it is itself the other, the determinateness, and hence in this
passing over it only comes to itself.
 Now the Kantian critique opposes
to this qualitative determination of the unity of the concept a quantitative
one. As it says, although the soul is not a manifold of reciprocally external
parts and contains no extensive magnitude, yet consciousness has a degree,
and the soul, like every concretely existing being, is an intensive magnitude;
with this magnitude, however, there is posited the possibility of a transition
into nothing through gradual vanishing.
 Now what is this refutation
but the application to spirit of a category of being, of intensive magnitude,
a determination that has no truth in itself but on the contrary is sublated
in the concept?

Metaphysics even one that restricted itself
to the fixed concepts of the understanding
without rising to speculation,
to the nature of the concept and of the idea,
did have for its aim the cognition of truth;
it did probe its subject matter to ascertain
whether they were something true or not,
whether substances or phenomena.
The triumph of the Kantian critique
over this metaphysics consists, on the contrary,
in side-lining any investigation
that would have truth for its aim and this aim itself;
it simply does not pose the one question
which is of interest,
namely whether a determinate subject,
in this case the abstract “I” of representation,
has truth in and for itself.
But to stay at appearances and at the mere representations
of ordinary consciousness is to give up
on the concept and on philosophy.
Anything beyond that is branded
by the Kantian critique as high-flown,
something to which reason has no claim.
As a matter of fact, the concept does fly high,
rising above what has no concept,
and the immediate justification
for going beyond it is, for one thing, the concept itself,
and, for another, on the negative side,
the untruth of appearance and of representation,
and also of such abstractions as the thing-in-itself
and the said “I” which is not supposed to be an object to itself.

In the context of this logical exposition, it is from the idea of life that the
idea of spirit has emerged, or what is the same thing, that has demonstrated
itself to be the truth of the idea of life. As this result, the idea possesses
its truth in and for itself, with which one may then also compare the
empirical reality or the appearance of spirit to see how far it accords with
it. We have seen regarding life that it is the idea, 27 but at the same time
it has shown itself not to be as yet the true presentation or the true mode
of its existence. For in life, the reality of the idea is singularity; universality
or the genus is the inwardness. The truth of life as absolute negative unity
consists, therefore, in this: to sublate the abstract or, what is the same, the
immediate singularity, and as identical to be self-identical, as genus, to be
self-equal. Now this idea is spirit.
 In this connection, we may further
remark that spirit is here considered in the form that pertains to this idea
as logical. For the idea also has other shapes which we may now mention
in passing; in these it falls to the concrete sciences of spirit to consider it,
namely as soul, consciousness, and spirit as such.

The name “soul” was used formerly to mean singular finite spirit in
general, and rational or empirical psychology was supposed to be synony-
mous with doctrine of spirit. The expression, “soul,” evokes an image of it
as if it were a thing like other things. One enquires regarding its seat, the
spatial location from which its forces operate; still more, how this thing
can be imperishable, subjected to the conditions of temporality yet exempt
from alteration in it. The system of monads elevates matter by making all
of it in principle a soul; 28 on this way of representing it, the soul is an
atom like the atoms of matter; the atom that rises from a cup of coffee as
vapor is capable in favorable circumstances of developing into a soul; only
the greater obscurity of its ideation distinguishes it from the kind of thing
that is manifestly soul.
 The concept that is for itself is necessarily also
in immediate existence; in this substantial identity with life, immersed in
its externality, the concept is the subject matter of anthropology. But even
anthropology would find alien a metaphysics in which this form of imme-
diacy is made into a soul-thing, into an atom like the atoms of matter.
To anthropology must be left only that obscure region where spirit,
under influences which were once called sidereal and terrestrial, lives as a
natural spirit in sympathy with nature and has presentiments of the latter's
alterations in dreams and presentiments, and indwells the brain, the heart,
the liver, and so forth. To the liver, according to Plato, God gave the gift
of prophesy above which the self-conscious human is exalted, so that even
the irrational part of the soul would be provided for by his bounty and
made to share in higher things. 29 To this irrational side belongs further the
behavior of figurative representation, and of higher spiritual activity in so
far as the latter is subject to the play of an entirely corporeal constitution,
of external influences and particular circumstances.

This lowest of the concrete shapes in which spirit is sunk into
materiality 30 has the one immediately superior to it in consciousness. In
this form the free concept, as the “I” existing for itself, is withdrawn from
objectivity, but it refers to the latter as its other, a subject matter that con-
fronts it. Since spirit is here no longer as soul, but, in the certainty that it
has of itself, the immediacy of being has for it the significance rather of a
negative, its identity with itself in the objectivity confronting it is at the
same time still only a reflective shining, for that objectivity still also has the
form of a being that exists in itself. This stage is the subject matter of the
phenomenology of spirit
 a science that stands midway between the science
of the natural spirit and of the spirit as such. It considers spirit as it exists
for itself, but at the same time as referring to its other
 an other which, as we
have just said, is thereby determined both as an object existing in itself and
as a negative. The science thus considers spirit as appearing, as exhibiting
itself in its contrary.

But the higher truth of this form is spirit for itself. For this spirit, the
subject matter 31 which for consciousness exists in itself has the form of its
own determination, the form of representation in general; this spirit, which
acts on the subject matter's determinations as on its own, on feelings, on
representations and thought, is thus infinite in itself and in its form. The
consideration of this stage belongs to the doctrine of spirit proper, which
would embrace the subject matter of ordinary empirical psychology but
which, in order to be the science of spirit, must not go about its work
empirically but must be conceived scientifically.
 At this stage spirit is
finite spirit in so far as the content of its determinateness is an immediately
given content; the science of this finite spirit has to display the course along
which it liberates itself from this determinateness and goes on to grasp its
truth, the infinite spirit.

The idea of spirit which is the subject matter of logic already stands,
on the contrary, inside pure science; it has no need, therefore, to observe
spirit's tracing that course, to see how it gets entangled with nature, with
immediate determinateness, with matter, or in other words with pictorial
representation; this is what the other three sciences investigate. The idea of
spirit has this course already behind it, or what is the same, it has it rather
ahead of it
 behind in so far as logic is taken as the final science; ahead
in so far as it is taken as the first science from which the idea first passes
over into nature. In the logical idea of spirit, therefore, the “I” is from the
start in the way it has emerged from the concept of nature as the truth of
nature, the free concept which in its judgment is itself the subject matter
confronting it, the concept as its idea. Also in this shape, however, the idea
is still not consummated.

Although the idea is indeed the free concept that has itself as its subject
matter, it is nonetheless immediate, and just because it is immediate, it is
still the idea in its subjectivity, and hence in its finitude in general. It is
the purpose that ought to realize itself, or the absolute idea itself still in its
appearance. What the idea seeks is the truth, this identity of the concept
itself and reality; but at first it only seeks it; for it is here as it is at first, still
something subjective. Consequently, although the subject matter that is for
the concept is here also a given subject matter, it does not enter into the
subject as affecting it, or as confronting it with a constitution of its own as
subject matter, or as a pictorial representation; on the contrary, the subject
transforms it into a conceptual determination; it is the concept which is the
active principle in it
 which therein refers itself to itself, and, by thus
giving itself its reality in the object, finds truth.

Initially, therefore, the idea is one extreme of a syllogism, the concept
that as purpose has itself at first for its subjective reality; the other extreme is
the restriction of the subjective, the objective world. The two extremes are
identical in that they are the idea. Their unity is, first, that of the concept,
a unity which in the one extreme is only for itself and in the other only in
itself. Second, it is reality, abstract in the one extreme and in the other in
its concrete externality.
 This unity is now posited through cognition, and,
because the latter is the subjective idea which as purpose proceeds from
itself, it is at first only a middle term.
 The knowing subject, through the
determinateness of its concept which is the abstract being-for-itself, refers
to an external world; nevertheless, it does this in the absolute certainty
of itself, in order to elevate its implicit reality, this formal truth, to real
truth. It has the entire essentiality of the objective world in its concept; its
process consists in positing for itself the concrete reality of that world as
identical with the concept, and conversely in positing the latter as identical
with objectivity.

Immediately, the idea of appearance is
the theoretical idea, cognition as such.
For to the concept that exists for itself,
the objective world immediately has
the form of immediacy or of being,
just as that concept is to itself
at first only the abstract concept of itself,
is still shut up within itself.
The concept is therefore only as form,
of which only its simple determinations
of universality and particularity are
the reality that it possesses within,
while the singularity or the determinate determinateness,
the content, is received by it from the outside.

CHAPTER 3

The absolute idea

The absolute idea has shown itself to be
the identity of the theoretical and the practical idea,
each of which, of itself still one-sided, possesses the idea
only as a sought-for beyond and unattained goal;
each is therefore a synthesis of striving,
each possessing as well as not possessing the idea within it,
passing over from one thought to the other
without bringing the two together
but remaining fixed in the contradiction of the two.
The absolute idea, as the rational concept
that in its reality only rejoins itself,
is by virtue of this immediacy of its objective identity,
on the one hand, a turning back to life;
on the other hand, it has equally
sublated this form of its immediacy
and harbors the most extreme opposition within.
The concept is not only soul,
but free subjective concept
that exists for itself
and therefore has personality,
the practical objective concept
that is determined in and for itself
and is as person impenetrable, atomic subjectivity
but which is not, just the same, exclusive singularity;
it is rather explicitly universality and cognition,
and in its other has its own objectivity for its subject matter.
All the rest is error, confusion, opinion,
striving, arbitrariness, and transitoriness;
the absolute idea alone is being, imperishable life,
self-knowing truth, and is all truth.

It is the sole subject matter and content of philosophy.
Since it contains all determinateness within it,
and its essence consists in returning
through its self-determination
and particularization back to itself,
it has various shapes,
and the business of philosophy
is to recognize it in these.
Nature and spirit are in general
different modes of exhibiting its existence,
art and religion its different modes of apprehending itself
and giving itself appropriate existence.
Philosophy has the same content and the same
purpose as art and religion,
but it is the highest mode of apprehending the
absolute idea, because its mode, that of the concept, is the highest.
 Hence it seizes those shapes of real and ideal finitude, as well of infinity and
holiness, and comprehends them and itself.
The derivation and cognition
of these particular modes are now the further business of the particular
philosophical sciences.
Also the logicality of the absolute idea can be called a mode of it;
but mode signifies a particular kind,
a determinateness of form,
whereas the logicality of the idea is
the universal mode in which all particular modes are sublated and enveloped.
The logical idea is the idea itself in its pure essence,
the idea which is enclosed in simple identity within its concept
and in reflective shining has as yet to step into a form-determinateness.
The Logic thus exhibits the self-movement of the absolute idea
only as the original word, a word which is an utterance,
but one that in being externally uttered has immediately vanished again.
The idea is, therefore, only in this self-determination of apprehending itself;
it is in pure thought, where difference is not yet otherness,
but is and remains perfectly transparent to itself.
The logical idea thus has itself, as the infinite form,
for its content, form that constitutes the opposite of content inasmuch as
the latter is the form determination that has withdrawn into itself and has
been so sublated in identity that this concrete identity stands over against
the identity developed as form; the content has the shape of an other and of
something given as against the form that as such stands simply in reference,
and whose determinateness is posited at the same time as reflective shine.
More exactly, the absolute idea itself has only this for its content,
namely that the form determination is
its own completed totality, the pure content.
Now the determinateness of the idea
and the entire course traversed by this determinateness
has constituted the subject matter of the science of logic,
and out of this course the absolute idea has come forth for itself;
thus to be for itself, however, has shown itself to amount to this,
namely that determinateness does not have the shape of a content,
but that it is simply as form,
and that accordingly the idea is
the absolutely universal idea.
What is left to be considered here, therefore,
is thus not a content as such,
but the universal character of its form
that is, method.

Method may appear at first to be just the manner
in which cognition proceeds, and this is in fact its nature.
But as method this manner of proceeding is not only a modality
of being determined in and for itself;
it is a modality of cognition,
and as such is posited as determined
by the concept and as form,
since form is the soul of all objectivity
and all otherwise determined content has its truth in form alone.
If the content is again assumed as given to the method
and of a nature of its own, then method,
so understood, is just like the logical realm in general
a merely external form.
But against this assumption appeal can be made, not only to the
fundamental concept of what constitutes logic, but to the entire logical
course in which all the shapes of a given content and of objects came up for
consideration.
This course has shown the transitoriness and the untruth of
all such shapes;
also that no given object is capable of being the foundation
to which the absolute form would relate as only an external and accidental
determination;
that, on the contrary, it is the absolute form
that has proved itself to be the absolute foundation
and the ultimate truth.
For this course the method has resulted as
the absolutely self-knowing concept,
as the concept that has the absolute,
both as subjective and objective,
as its subject matter,
and consequently as the pure correspondence
of the concept and its reality,
a concrete existence that is the concept itself.

Accordingly, what is to be considered as method here is only the move-
ment of the concept itself. We already know the nature of this movement,
but it now has, first, the added significance that the concept is all, and
that its movement is the universal absolute activity, the self-determining
and self-realizing movement. The method is therefore to be acknowledged
as the universal, internal and external mode, free of restrictions, and as
the absolutely infinite force to which no object that may present itself as
something external, removed from reason and independent of it, could
offer resistance, or be of a particular nature opposite to it, and could not be
penetrated by it. It is therefore soul and substance, and nothing is conceived
and known in its truth unless completely subjugated to the method; it is the
method proper to each and every fact because its activity is the concept.
This is also the truer meaning of its universality; according to the universal-
ity of reflection, it is taken only as the method for all things; but according
to the universality of the idea, it is both the manner of cognition, of the
concept subjectively aware of itself, and the objective manner, or rather the
substantiality of things
 that is, of concepts as they first appear as others
to representation and reflection. It is therefore not only the highest force of
reason, or rather its sole and absolute force, but also reason's highest and
sole impulse to find and recognize itself through itself in all things.
 Second,
here we also have the distinction of the method from the concept as such, the
particularization of the method. As the concept was considered for itself,
it appeared in its immediacy; the reflection, or the concept considering it,
fell on the side of our knowledge. The method is this knowledge itself, for
which the concept is not only as subject matter but is as its own subjec-
tive act, the instrument and the means of cognitive activity, distinct from
this activity and yet the activity's own essentiality. In cognition as enquiry,
the method likewise occupies the position of an instrument, as a means
that stands on the side of the subject, connecting it with the object. The
subject in this syllogism is one extreme, the object is the other, and in
conclusion the subject unites through its method with the object without
however uniting with itself there. The extremes remain diverse, because
subject, method, and object are not posited as the one identical concept; the
syllogism is therefore always the formal syllogism; the premise in which
the subject posits the form on its side as its method is an immediate deter-
mination and contains therefore the determinations of the form
 as we
have seen, of definition, division, and so forth
 as matters of fact found
ready-made in the subject. In true cognition, on the contrary, method is
not only an aggregate of certain determinations, but the determinateness
in-and-for-itself of the concept, and the concept is the middle term only
because it equally has the significance of the objective; in the conclusion,
therefore, the objective does not attain only an external determinateness by
virtue of the method, but is posited rather in its identity with the subjective
concept.

1. Accordingly, what constitutes the method are the determinations of
the concept itself and their connections, and these we must now examine
in the significance that they have as determinations of the method.
In this, we must begin from the beginning.
We spoke of this beginning at
the very beginning of the Logic,
and also in connection with subjective
cognition,
 and we showed that, when not performed arbitrarily and in
the absence of categorial sensitivity,
though it may seem to present many difficulties,
it is nevertheless of an extremely simple nature.
Because it is the beginning, its content is an immediate,
but one that has the meaning and the form of abstract universality.
Or be it a content of being, or of essence or
of the concept, inasmuch as it is something immediate, it is assumed, found in
advance, assertoric.
 But first of all it is not an immediate of sense-intuition or
of representation, but of thought, which because of its immediacy can also be
called a supersensuous, inner intuiting. The immediate of sense-intuition
is a manifold and a singular. Cognition, on the contrary, is a thinking that
conceptualizes; its beginning, therefore, is also only in the element of thought,
a simple and a universal.
 We spoke of this form earlier, in connection with
definition. At the beginning of finite cognition universality is likewise
recognized as an essential determination, but only as thought
 and concept
determination in opposition to being. In fact this first universality is an
immediate universality, and for that reason it has equally the significance of
being, for being is precisely this abstract self-reference. Being has no need
of further derivation, as if it came to the abstract element of definition
only because taken from the intuition of the senses or elsewhere, and
in so far as it can be pointed at. This pointing and deriving involve a
mediation that is more than a mere beginning, and is a mediation of a kind
that does not belong to the comprehension of thought, but is rather the
elevation of representation, of empirical and ratiocinative consciousness, to
the standpoint of thinking. According to the currently accepted opposition
of thought, or concept, and being, it passes as a very important truth that
no being belongs as yet to thought as thought, and that being has a ground
of its own independent of thought. But the simple determination of being
is in itself so poor that, if for that reason alone, not much fuss ought to be
made about it; the universal is immediately itself this immediate because,
as abstract, it is also the abstract self-reference which is being. In fact, the
demand that being should be exhibited has a further, inner meaning in
which more is at issue than just this abstract determination; implied in it
is the demand for the realization of the concept, a realization that is missing
at the beginning itself but is rather the goal and the business of the entire
subsequent development of cognition. Further, inasmuch as the content
of the beginning is to be justified and authenticated as something true or
correct by being exhibited in inner or outer perception, it is no longer
the form of universality as such that is meant, but its determinateness,
about which more in a moment. The authentication of the determinate
content with which the beginning is made seems to lie behind it, but is in
fact to be regarded as an advance, in so far as it is a matter of conceptual
cognition.

The beginning, therefore, has for the method no other determinateness
than that of being the simple and universal; this is precisely the determinate-
ness that makes it deficient. Universality is the pure, simple concept, and
the method, as the consciousness of this concept, is aware that universality
is only a moment and that in it the concept is still not determined in and for
itself. But with this consciousness that would want to carry the beginning
further only for the sake of method, the method is only a formal procedure
posited in external reflection. Where the method, however, is the objective
and immanent form, the immediate character of the beginning must be
a lack inherent in the beginning itself, which must be endowed with the
impulse to carry itself further. But in the absolute method the universal
has the value not of a mere abstraction but of the objective universal, that
is, the universal that is in itself the concrete totality, but a totality as yet not
posited, not yet for itself. Even the abstract universal is as such, when con-
sidered conceptually, that is, in its truth, not just anything simple, but is,
as abstract, already posited afflicted by a negation. For this reason also there
is nothing so simple and so abstract, be it in actuality or in thought, as is
commonly imagined. Anything as simple as that is a mere presumption that
has its ground solely in the lack of awareness of what is actually there.
 We
said earlier that the beginning is made with the immediate; the immediacy
of the universal is the same as what is here expressed as the in-itself that is
without being-for-itself.
 One may well say, therefore, that every beginning
must be made with the absolute, just as every advance is only the exposition
of it, in so far as implicit in existence is the concept. But because the
absolute exists at first only implicitly, in itself, it equally is not the absolute
nor the posited concept, and also not the idea, for the in-itself is only an
abstract, one-sided moment, and this is what they are. The advance is not,
therefore, a kind of superfluity; this is what it would be if that which is
at the beginning were already the absolute; the advance consists rather in
this, that the universal determines itself and is the universal for itself, that
is, equally a singular and a subject. Only in its consummation is it the
absolute.

It may also be mentioned that a beginning which is in itself a concrete
totality may as such also be free and its immediacy have the determination
of an external existence; the germ of anything living, and subjective purpose
in general, have shown themselves to be such beginnings; hence both are
themselves impulses. The non-spiritual and inanimate, on the contrary, are
the concrete concept only as real possibility; cause is the highest stage in
which the concrete concept has, as the beginning in the sphere of necessity,
an immediate existence; but it is not yet a subject that maintains itself
as such in the course of its effective realization. The sun, for instance,
and in general all things inanimate, are determinate concrete existences
in which real possibility remains an inner totality; the moments of the
latter are not posited in them in subjective form and therefore, in so far
as they are realized, they attain concrete existence through other corporeal
individuals.

2. The concrete totality which makes the beginning possesses as such,
within it, the beginning of the advance and development.
As concrete, it is differentiated in itself,
but because of its initial immediacy, this first
differentiation is to start with a diversity.
However, as self-referring universality, as subject,
the immediate is also the unity of this diversity.
This reflection is the first stage of the forward movement
the emergence of non-indifference, judgment, and determining in general.
Essential is that the method find, and recognize,
the determination of the universal within it.
Whatever in this abstractive generation of the universal
is left out of the concrete is then picked up, still externally,
by the finite cognition of the understanding.
This is how the latter operates.
The absolute method, on the contrary, does not behave
in this manner of external reflection
but takes the determinate from its subject matter,
for it is itself its immanent principle and its soul.
This is what Plato demanded of cognition,
that it should consider things in and for themselves;
on the one hand, that it should consider them in their universality;
on the other hand, that it should not stray away from them
while it grasps at circumstances, examples, and comparisons,
but, on the contrary, should keep only them in view before it
and bring to consciousness what is immanent in them.
To this extent the method of absolute cognition is analytic.
That the method finds the further determinations
of its initial universal simply and solely in this universal,
constitutes the concept's absolute objectivity,
of which the method is the certainty.
Equally so, however, is the method synthetic,
for its subject matter,
while immediately determined as the simple universal,
through the determinateness which it has
in its very immediacy and universality,
proves to be an other.
Yet this connection in diversity
that the subject matter is thus in itself,
is no longer a synthesis as understood in finite cognition;
the no less thoroughly analytic determination of the subject matter,
the fact that the connection is within the concept,
already distinguishes it fully from the latter synthesis.

This no less synthetic than analytic moment of the judgment through
which the initial universal determines itself from within itself as the other
of itself is to be called the dialectical moment. Dialectic is one of those
ancient sciences that have been the most misjudged in the metaphysics of
the moderns, and in general also by popular philosophy, both ancient and
recent. Diogenes Laërtius said of Plato that, just as Thales was the founder
of natural philosophy and Socrates of moral philosophy, so Plato was the
founder of the third of the sciences that belong to philosophy, of dialectic
a contribution for which he was highly esteemed by the ancients but that
often goes quite unnoticed by those who have the most to say about him.
Dialectic has often been regarded as an art, as if it rested on a subjective
talent and did not belong to the objectivity of the concept. What shape
it received in Kantian philosophy, and with what result, has already been
indicated in representative examples of that philosophy's view. It must be
regarded as an infinitely important step that dialectic is once more being
recognized as necessary to reason, although the result that must be drawn
from it is the opposite than Kant drew.

When dialectic is not presented, as it generally is, as something inci-
dental, it usually assumes the following more precise form. It is shown of
a subject matter or other (for instance: world, movement, point, and so
on) that a certain determination accrues to it (for instance, in the order of
the just mentioned examples: finitude in space or time, being at this place,
absolute negation of space), also the opposite determinations can then just
as necessarily be shown to accrue to it (for example: infinity in space and
time, not-being at this place, reference to space and hence spatiality). The
older Eleatic school directed its dialectic especially against motion; Plato
commonly did it against accepted notions and concepts of his time, in par-
ticular those of the Sophists, but also against the pure categories and the
determinations of reflection; the later and more sophisticated form of skep-
ticism extended it not only to the immediate so-called facts of consciousness
and the maxims of ordinary life, but also to all scientific concepts. Now the
conclusion drawn from this kind of dialectic is in general the contradiction
and nullity of the asserted claims. But this can happen in two ways
 either
in the objective sense, that the subject matter that thus contradicts itself
internally cancels itself and is a non-thing (this was, for instance, the con-
clusion of the Eleatics, who denied the truth of the world, of movement,
of the point); or in the subjective sense, that cognition is deficient. Now
understood in this last subjective sense, the conclusion may be taken in
two further ways. It may mean that it is this dialectic itself that generates
the artifice of an illusion. This is the common view of the so-called healthy
common sense that takes its stand on the evidence of the senses and on cus-
tomary notions and claims, at times quietly, like Diogenes the cynic did, who
demonstrated the vacuity of the dialectic of motion by silently walking up
and down; but often by getting itself all worked up, declaring that dialectic
is mere foolery or, when important ethical matters are at issue, the criminal
attempt at unsettling essentially solid norms and providing excuses for the
wicked
 a view we see directed in the Socratic dialectic against that of
the Sophists, with an ire that, turned into the opposite direction, even cost
Socrates his life. As for the vulgar refutation that opposes to thinking, as
Diogenes did, sensuous consciousness and in this latter believes that it finds
the truth, this we must leave to itself; but in so far as dialectic sublates
ethical determinations, we must have confidence in reason that it will
know how to reinstate them, but reinstate them in their truth and in the
consciousness of their right, though also of their limitations.
 Yet another
view is that the result of subjective nullity has nothing to do with dialectic
itself, but that it affects the cognition against which it is directed and, in
the view of skepticism and likewise of the Kantian philosophy, cognition in
general.

The fundamental prejudice here is that dialectic has only a negative
result
 a point about which more in a moment. First, regarding the said
form in which dialectic usually makes its appearance, it is to be observed
that according to that form the dialectic and its result affect a subject matter
which is previously assumed or also the subjective cognition of it, and
declare either the latter or the subject matter to be null and void, while,
on the contrary, no attention is given to the determinations which are
exhibited in the subject matter as in a third thing and presupposed as valid
for themselves. To have called attention to this uncritical procedure has
been the infinite merit of the Kantian philosophy, and in so doing to have
given the impetus to the restoration of logic and dialectic understood as the
examination of thought determinations in and for themselves. The subject
matter, as it is apart from thought and conceptualization, is a picture
representation or also a name; it is in the determinations of thought and of
the concept that it is what it is. In fact, therefore, everything rests on these
determinations; they are the true subject matter and content of reason,
and anything else that might be understood by subject matter and content
in distinction from them has meaning only through them and in them.
It must not therefore be taken as the fault of a subject matter or of the
cognition that these determinations, because of what they are and the way
they are externally joined, prove to be dialectical. On this assumption,
the subject matter and the cognition are imagined to be a subject on
which the determinations are brought to bear, in the form of predicates,
properties, or self-subsistent universals, as fixed and independently correct,
so that these determinations are brought into dialectical relations and incur
contradiction only by extraneous and contingent conjunction in and by a
third thing. But this kind of external and fixed subject of imagination and
understanding, and also these abstract determinations, cannot be regarded
as ultimates, as secure and permanent substrates. On the contrary, they are to
be regarded as themselves immediate, precisely the kind of presuppositions
and starting points which, as we have shown above, 79 must succumb to
dialectic in and for themselves, because they are to be taken as in themselves
the concept. The same applies to all oppositions that are assumed as fixed,
as for example the finite and the infinite, the singular and the universal.
These are not in contradiction through some external conjoining; on the
contrary, as an examination of their nature shows, they are a transition in
and for themselves; the synthesis and the subject in which they appear is
the product of their concept's own reflection. If a consideration that avoids
the concept stops short at their external relation, isolates them and leaves
them as fixed presuppositions, it is the concept that, on the contrary, will
fix its sight on them, move them as their soul and bring out their dialectic.

Now this is the very standpoint indicated above from which a universal
prius, considered in and for itself, proves to be the other of itself. Taken quite
generally, this determination can be taken to mean that what is at first
immediate is therewith posited as mediated, as referred to an other, or
that the universal is posited as a particular. The second universal that has
thereby arisen is thus the negative of that first and, in view of subsequent
developments, the first negative. From this negative side, the immediate has
perished in the other; but the other is essentially not an empty negative, the
nothing which is normally taken to be the result of dialectic, but is rather
the other of the first, the negative of the immediate; it is therefore determined
as the mediated
 contains as such the determination of the first in it. The
first is thus essentially preserved and contained also in the other.
 To hold
fast to the positive in its negative, to the content of the presupposition in
the result, this is the most important factor in rational cognition; what
is more, it takes only the simplest of reflections to be convinced of the
absolute truth and necessity of this requirement, and as for examples of
proofs that testify to this, the whole Logic consists of such proofs.

So what we now have, taken first or also immediately, is the mediated,
also a simple determination, for the first has perished in it, and only the
second is therefore at hand. Now since the first is contained in the second,
and this second is the truth of the first, this unity of the two can be expressed
in the form of a proposition in which the immediate is placed as the subject
but the mediated as its predicate; for example, “the finite is infinite,” “one
is many,” “the singular is the universal.” The inadequacy of the form
of such propositions and judgments is however obvious. In connection
with judgment it was shown that its form in general, and most of all the
immediate form of the positive judgment, is incapable of holding within its
grasp the speculative content and the truth. 81 Its closest complement, the
negative judgment, would have to be brought in at least in equal measure.
In judgment the first, as subject, conveys the reflective semblance of an
independent subsistence, whereas it is in fact sublated in the predicate as
in its other; this negation is indeed contained in the content of the above
propositions, but their positive form contradicts the content; consequently,
what is contained in them is not posited
 whereas this was precisely the
intent behind the use of a proposition.

The second determination, the negative or mediated determination, is
moreover at the same time the one that mediates. At first it may be taken as
a simple determination, but its truth is that it is a reference or relation; for
it is the negative, but the negative of the positive, and it includes this positive
within itself. It is the other, therefore, not of a one to which it is indifferent;
in that case it would not be an other, nor a reference or relation. It is rather
the other in itself, the other of an other; hence it includes its own other within
itself and is consequently the contradiction, the posited dialectic, of itself.
Because the first or the immediate is the concept in itself or implicitly,
and therefore is the negative also only implicitly, the dialectical moment
in it consists in the positing of the difference that is implicitly contained
in it. The second is on the contrary itself the determinate, the difference or
relation; hence the dialectical moment consists in its case in the positing
of the unity contained within it.
 For this reason, if the negative, the
determinate, relation, judgment, and all the determinations falling under
this second moment, do not appear by themselves already as contradiction,
as dialectical, this is solely a defect on the part of thinking that fails to bring
its thoughts together. For the material, the opposed determinations in one
connection, are already posited, already present for thought. But formal
thinking makes identity its law, lets the contradictory content that it has
before it fall into the sphere of representation, in space and time, where
the contradictory is held in external moments, next to and following each
other, parading before consciousness without reciprocal contact. The firm
principle that formal thinking lays down for itself here is that contradiction
cannot be thought. But in fact the thought of contradiction is the essential
moment of the concept. Formal thought does in fact think it, only it at
once looks away from it and stating its principle it only passes over from it
into abstract negation.

Now the negativity just considered constitutes
the turning point of the movement of the concept.
It is the simple point of the negative self-reference,
the innermost source of all activity,
of living and spiritual self-movement;
it is the dialectical soul which everything true possesses
and through which alone it is true;
for on this subjectivity alone rests
the sublation of the opposition
between concept and reality,
and the unity which is truth.
The second negative at which we have arrived,
the negative of the negative,
is this sublating of contradiction,
and it too, just like contradiction,
is not an act of external reflection;
for it is on the contrary the innermost,
objective moment of the life of spirit
by virtue of which a subject is a person, is free.
The self-reference of the negative is to be regarded
as the second premise of the entire syllogism.
If the terms analytic and synthetic are used as opposites,
the first premise may be regarded as the analytic moment,
for in it the immediate relates to its other immediately
and therefore passes over, or rather has passed over,
into it though this connection, as already remarked,
is for this very reason also synthetic,
for it is its other that it passes over into.
The second premise considered here may be defined as synthetic,
because it is the connection of the differentiated, as differentiated,
to that from which it is differentiated.
Just as the first premise is
the moment of universality and communication,
so is the second determined by singularity;
a singularity which in referring to the other is at first exclusive, for itself,
and different. The negative appears as the mediating factor, because it holds
itself and the immediate of which it is the negation within itself. In so far
as these two determinations are taken as referring to each other externally
in some relation or other, the negative is only the formal mediating factor;
but, as absolute negativity, the negative moment of absolute mediation is
the unity which is subjectivity and soul.

In this turning point of the method,
the course of cognition returns at the
same time back into itself.
This negativity is as self-sublating contradiction
the restoration of the first immediacy, of simple universality;
for the other of the other, the negative of the negative,
is immediately the positive, the identical, the universal.
In the whole course, if one at all cares to count,
this second immediate is third to the first immediate and the mediated.
But it is also third to the first or formal negative
and to the absolute negativity or second negative;
now in so far as that first negative is already the second
term, the term counted as third can also be counted as fourth,
and instead of a triplicity, the abstract form may also be taken to be a quadruplicity;
in this way the negative or the difference is counted as a duality.
The third or the fourth is in general
the unity of the first and the second moment,
of the immediate and the mediated.
That it is this unity,
or that the entire form
of the method is a triplicity,
is indeed nothing but the merely superficial,
external side of cognition;
 but to have also demonstrated this superficiality,
and to have done it in the context of a specific application (for the abstract
form of number has been around for a long time, as is well known, but
without conceptual comprehension and therefore without any result) is
again to be regarded as an infinite merit of the Kantian philosophy.
The syllogism, or the threefold, has always been recognized to be the universal
form of reason;
but it has had in general the value of a wholly external form
that does not determine the nature of the content;
moreover, since in its formalism it gets caught up
in the understanding's determination of mere identity,
it lacks the essential dialectical moment of negativity;
and yet this moment enters into the triplicity of the determinations,
because the third term is the unity of the two first determinations
and these, since they are diverse, can be in unity only as sublated.
Formalism, it is true, has also
seized hold of triplicity, attending to its empty schema;
the shallow nonsense and the barrenness of the so-called construction of modern philosophy, that
consists in nothing but fastening that formal schema everywhere for the
sake of external order,
with no concept or immanent determination,
has rendered that form tedious and has given it a bad name.
Yet the insipidity of this use cannot rob it of its inner worth,
and the fact that the shape of reason was discovered,
albeit without conceptual comprehension at first,
is always to be highly valued.

Now, on closer examination, the third is the immediate, but the imme-
diate through sublation of mediation, the simple through the sublating of
difference, the positive through the sublating of the negative; it is the
concept that has realized itself through its otherness, and through the sub-
lating of this reality has rejoined itself and has restored its absolute reality,
its simple self-reference. This result is therefore the truth. It is just as much
immediacy as mediation
 though these forms of judgments, that the third
is immediacy and mediation, or that it is the unity of the two, are not
capable of grasping it, for it is not a dormant third but, exactly like this
unity, self-mediating movement and activity.
 Just as that with which we
began was the universal, so the result is the singular, the concrete, the subject;
what the former is in itself, the latter is now equally for itself: the universal
is posited in the subject. The two first moments of triplicity are abstract,
untrue moments that are dialectical for that very reason, and through this
their negativity make themselves into the subject. For us at first, the con-
cept itself is both the universal that exists in itself and the negative that
exists for itself, and also the third term that exists in and for itself, the
universal that runs through all the moments of the syllogism; but this third
is the conclusion in which the concept mediates itself with itself through
its negativity and is thereby posited for itself as the universal and the identity
of its moments.
Now this result, as the whole that has withdrawn into itself and is identi-
cal with itself, has given itself again the form of immediacy. Consequently,
it is now itself all that the starting point had determined itself to be. As sim-
ple self-reference it is a universal, and in this universal the negativity that
constituted its dialectic and mediation has likewise withdrawn into simple
determinateness, which can again be a beginning. It may seem at first that
this cognition of the result is an analysis of it and would therefore have to
dissect these determinations again, and the course that it went through in
order to come to be
 the course that we have examined. But if the subject
matter were in fact treated analytically in this manner, it would belong to
that stage of the idea considered above, 83 a mode of cognition that searches
for its subject matter and only states of it what it is, without the necessity
of its concrete identity and of its concept. But the method of truth that
comprehends the subject matter, though analytic as we have seen, since
it remains strictly within the concept, is however equally synthetic, for
through the concept the subject matter is determined as dialectical and as
other. 84 On the new foundation that the result has now constituted as the
subject matter, the method remains the same as in the preceding subject
matter. The difference concerns solely the status of the foundation as such;
although it is certainly still a foundation, its immediacy is only form, since
it was a result as well; hence its determinateness as content is no longer
something merely taken up but is deduced and proved.
It is here that the content of cognition first enters as such into the circle
of consideration, because as deduced it now belongs to the method. The
method itself expands with this moment into a system.
 With respect to
content, the beginning has to be for the method at first wholly indeter-
minate; to this extent the method appears as the merely formal soul, for
which and by which the beginning was determined simply and solely only
according to form, that is to say, as the immediate and universal. In the
course of the movement we have indicated, the subject matter has received
a determinateness for itself
 and this determinateness is a content, for the
negativity that has withdrawn into simplicity is the sublated form, and
stands as simple determinateness over against its development, and in the
first instance against its very opposition to universality.
Now since this determinateness is the proximate truth of the indetermi-
nate beginning, it denounces the incompleteness of the latter, and it also
denounces the method itself which, starting from that beginning, was only
formal. This can now be expressed as the henceforth determinate demand
that the beginning, since as against the determinateness of the result it
is itself something determinate, ought to be taken not as immediate, but
as mediated and deduced. This may appear as the demand for an infi-
nite retrogression in proof and deduction; just as from the newly obtained
beginning a result likewise emerges as the method runs its course, so that
the movement would roll on forwards to infinity as well.

It has been repeatedly shown that the infinite progression as such belongs
to a reflection void of concept; the absolute method, which has the con-
cept for its soul and content, cannot lead into it. Even such beginnings as
being, essence, universality, may seem at first to be of the kind that possess
the full universality and complete absence of content that is required for
an entirely formal beginning, such as the beginning is supposed to be,
and therefore not to require or allow, as absolutely first beginnings, fur-
ther regress. Since they refer purely to themselves, they are immediate and
indeterminate, and so they do not of course have in them the difference
which is straightaway posited in some other beginning between the uni-
versality of its form and its content. But the very indeterminacy which
these logical beginnings have as their sole content is what constitutes their
determinateness; this determinateness consists in their negativity, as sub-
lated mediation; the particularity of this negativity gives a particularity
also to their indeterminacy, and it is by virtue of it that being, essence, and
universality, are differentiated. Now the determinateness that accrues to
them when taken for themselves is their immediate determinateness, and
this is just as immediate as that of any content and in need, therefore,
of derivation; for the method it is a matter of indifference whether the
determinateness is taken as determinateness of form or of content. That it
gives itself a determination by the first of its results does not mean that,
in fact, it is thereby set on a new footing; it remains neither more nor less
formal than before. For since the method is the absolute form, the con-
cept that knows itself and everything as concept, there is no content that
would stand out over against it and determine it as a one-sided external
form. Hence, just as the lack of content of the said beginnings does not
make them absolute beginnings, so too it is not the content that would
as such lead the method into the infinite progress forwards or backwards.
In one respect, the determinateness that the method generates for itself in
its result is the moment through which it is self-mediation and converts
the immediate into a mediated beginning. But conversely, it is through that
determinateness that this mediation of the method runs its course; it goes
through a content, as through a seeming other of itself, back to its begin-
ning, in such a way that it does not merely restore that beginning, albeit as
determinate, but that the result is equally the sublated determinateness, and
hence also the restoration of the first immediacy in which it began. This
it accomplishes as a system of totality. We now have to consider it in this
determination.

The determinateness which was the result is, as we have shown, 85 itself
a new beginning because of the form of simplicity into which it has
withdrawn; since this beginning is distinguished from the one preceding
it by this very determinateness, cognition rolls onwards from content to
content. First of all, this forward movement determines itself in that it
begins from simple determinacies, and the following become ever richer
and more concrete. For the result contains its beginning and its course
has enriched it with a new determinateness. The universal constitutes the
foundation; the advance is not to be taken, therefore, as a flowing from
other to other. In the absolute method, the concept maintains itself in its
otherness, the universal in its particularization, in judgment and reality; at
each stage of further determination, the universal elevates the whole mass of
its preceding content, not only not losing anything through its dialectical
advance, or leaving it behind, but, on the contrary, carrying with itself all
that it has gained, inwardly enriched and compressed.

This expansion may be regarded as the moment of content, and in the
whole as the first premise; the universal is communicated to the wealth of
content, is immediately received in it. But the relation has also a second,
negative or dialectical side. The enrichment proceeds in the necessity of the
concept, it is contained by it, and every determination is a reflection into
itself. Each new stage of exteriorization, that is, of further determination, is
also a withdrawing into itself, and the greater the extension, just as dense
is the intensity. The richest is therefore the most concrete and the most
subjective, and that which retreats to the simplest depth is the mightiest
and the most all-encompassing. The highest and most intense point is the
pure personality that, solely by virtue of the absolute dialectic which is its
nature, equally embraces and holds everything within itself, for it makes
itself into the supremely free
 the simplicity which is the first immediacy
and universality.

It is in this manner that each step of the advance in the process of further
determination, while getting away from the indeterminate beginning,
is also a getting back closer to it;
consequently, that what may at first appear to be different,
the retrogressive grounding of the beginning
and the progressive further determination of it,
run into one another and are the same.
The method, which thus coils in a circle,
cannot however anticipate in a temporal development that the beginning is as such already something
derived; sufficient for an immediate beginning is that it be simple universality.
Inasmuch as this is what it is, it has its complete condition;
and there is no need to deprecate the fact that it may be accepted only provi-
sionally and hypothetically.
Whatever might be adduced against it about
the limitations of human cognition;
about the need to reflect critically
on the instrument of cognition
before getting to the fact itself
 all these are themselves presuppositions,
concrete determinations that as such carry
with them the demand for mediation and grounding.
Therefore, since they formally have no advantage over beginning with the fact itself as they
protest against, and, because of their more concrete content, are on the
contrary all the more in need of derivation, singling them out for special
attention is to be considered as empty presumption. They have an untrue
content, for they make into something incontestable and absolute what is
known to be finite and untrue, namely a restricted cognition determined as
form and instrument in opposition to its content; this untrue cognition is itself
also the form, the retroactive search for grounds.
 The method of truth also
knows that the beginning is incomplete, because it is a beginning; but at
the same time it knows that this incompleteness is necessary, because truth
is but the coming-to-oneself through the negativity of immediacy. The
impatience that would merely transcend the determinate
 be it called begin-
ning, object, the finite, or in whatever other form it is otherwise taken
in order that one would find oneself immediately in the absolute, has
nothing before it as cognition but the empty negative, the abstract infinite.
Or what it has before it is a presumed absolute, presumed because not
posited, not comprehended; comprehended it will be only through the medi-
ation of cognition, of which the universal and immediate are a moment,
and as for the truth itself, it resides only in the extended course of media-
tion and at the end.
 To meet the subjective need and the impatience that
come with not knowing, one may well provide an overview of the whole in
advance
 by means of a division for reflection that, in the manner of finite
cognition, gives the particular of the universal as already there, to be waited
for as the science progresses. Yet this affords nothing more than a picture
for representation; for the true transition from the universal to the particular
and to the whole which is determined in and for itself and in which that
first universal is in truth itself again a moment
 this transition is alien to
the division of reflection and is the exclusive mediation of science itself.

By virtue of the nature of the method just indicated,
the science presents itself as a circle that winds around itself,
where the mediation winds the end back to the beginning
which is the simple ground;
the circle is thus a circle of circles,
for each single member ensouled by the method
is reflected into itself so that, in returning to the beginning
it is at the same time the beginning of a new member.
Fragments of this chain are the single sciences,
each of which has a before and an after
or, more accurately said, has
in possession only the before
and in its conclusion points to its after.

So the logic also has returned in the absolute idea to this simple unity
which is its beginning; the pure immediacy of being, in which all deter-
mination appears at first as extinguished or removed by abstraction, is the
idea that through mediation, that is, the sublation of mediation, has come
to the likeness corresponding to it. The method is the pure concept that
only relates to itself; it is, therefore, the simple self-reference which is being.
But it now is also the fulfilled concept, the concept that comprehends itself
conceptually, being as the concrete and just as absolutely intensive totality.

In conclusion, there remains only this to be said of this idea, that in it, in
the first place, the science of logic has apprehended its own concept. In the
sphere of being, at the beginning of its content, its concept appears as a
knowledge external to that content in subjective reflection. But in the idea
of absolute cognition, the concept has become the idea's own content. The
idea is itself the pure concept that has itself as its subject matter and which,
as it runs itself as subject matter through the totality of its determinations,
builds itself up to the entirety of its reality, to the system of science, and
concludes by apprehending this conceptual comprehension of itself, hence
by sublating its position as content and subject matter and cognizing the
concept of science.
In second place, this idea is still logical;
it is shut up in pure thought,
the science only of the divine concept.
Its systematic exposition is of course itself a realization,
but one confined within the same sphere.
Because the pure idea of cognition is
to this extent shut up within subjectivity,
it is the impulse to sublate it,
and pure truth becomes as final result
also the beginning of another sphere and science.
It only remains here to indicate this transition.

The idea, namely, in positing itself
as the absolute unity of the pure concept and its reality
and thus collecting itself in the immediacy of being,
is in this form as totality:  nature.
This determination, however, is
nothing that has become, is not a transition, as was the case above when the
subjective concept in its totality becomes objectivity, or the subjective purpose
becomes life. The pure idea into which the determinateness or reality of the
concept is itself raised into concept is rather an absolute liberation for which
there is no longer an immediate determination which is not equally posited
and is not concept; in this freedom, therefore, there is no transition that
takes place; the simple being to which the idea determines itself remains
perfectly transparent to it: it is the idea that in its determination remains
with itself. The transition is to be grasped, therefore, in the sense that the
idea freely discharges 87 itself, absolutely certain of itself and internally at
rest. On account of this freedom, the form of its determinateness is just as
absolutely free: the externality of space and time absolutely existing for itself
without subjectivity.
 Inasmuch as this externality is only in the abstract
determinateness of being and is apprehended by consciousness, it is as mere
objectivity and external life; within the idea, however, it remains in and
for itself the totality of the concept, and science in the relation of divine
cognition to nature. But what is posited by this first resolve of the pure idea
to determine itself as external idea is only the mediation out of which the
concept, as free concrete existence that from externality has come to itself,
raises itself up, completes this self-liberation in the science of spirit, and in
the science of logic finds the highest concept of itself,
the pure concept conceptually comprehending itself.
